\bookStart{Greenlandish Speeches of Attle}[Atla mǫ́l in grǿn-lęndsku]
\setBookCode{Atlamal}

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: late C11th (0.472)

\textbf{Meter:} \Malahattr
\end{flushright}%

\section{Introduction}

NOTE: The following translation is currently incomplete and the edition has not been thoroughly checked.

The \textbf{Greenlandish Speeches of Attle} (abbrev.~\Atlamal) is a long Eddic heroic poem dealing with the deaths of Guther and Hawn at the hands of Attle.

\subsection{Textual preservation and transmission}

\Atlamal\ is only preserved in \Regius\ where it is found on foll~41r/9--44r/13.  It follows \textlink{Atlakvida} and precedes \textlink{Gudrunarhvot}.  The title \emph{Atla mǫ́l in grǿn-lęndsku} ‘The Greenlandish Speeches of Attle’ is preserved in \textlink{Atlakvida}[P2].

\subsection{Story and parallels}

The basic narrative of \Atlamal\ is largely the same as the preceding \textlink{Atlakvida}, and together with that poem it clearly serves as a source for \VolsungaSaga\ 36--38.  For much of that section \Atlamal\ is the chief source, probably due to its greater length.

Although both poems build on the same basic narrative frame, \Atlamal\ is far longer and contains a number of otherwise unparallelled details.
One of the most interesting differences is the \Atlamal-poet’s clear lack of familiarity with continental European circumstances.  Whereas the poet of \textlink{Atlakvida} has Guther and Hawn ride through the dark Mirkwood and over high mountains to get to Attle in the south (\textlink{Atlakvida}[14]), \Atlamal\ instead lets Guther and Hawn sail out of their home firth (\emph{fjǫrðr}) to the land of Budle (\textlink{Atlamal}[28]--35) in what is clearly an insular Norse adaptation of the original setting.

\subsection{Meter and formal concerns}

\Atlamal\ consists of 382 long-lines distributed across 102 stanzas.  The typical length of a stanza is 4 lines, but other lengths are also represented.

The poem is composed in \Malahattr\ throughout.

\subsection{Localisation}

In \Regius\ the adjective \emph{grǿn-lęndskr} ‘Greenlandish’ is applied to both \textlink{Atlakvida} and \Atlamal, but only the latter appears actually to have been composed in one of the Norse settlements on Greenland.  This is shown most clearly by st.~18, where the poet makes reference to a “white-bear” (\emph{hvíta-bjǫrn}) arriving in a storm from the east.
The polar bear (\emph{Ursus maritimus}) is indeed found on Greenland, where members of the species not infrequently drift onto the shores on ice sheets carried by the winds, whereas they never appear on Iceland or the Scandinavian peninsula.

\subsection{Dating}

A Greenlandish origin for the poem lets us to determine a \emph{terminus post quem} for its composition.  \Atlamal\ cannot have been composed before ca. 985 \ce, the time of Greenland’s settlement by Icelanders agreed upon both by the mediæval saws and modern archeology.

We may push this date some decades forward based on the treatment of the dead in the poem.  Being burned on a pyre is a punishment associated with stoning (st.~84), while a proper burial is in a painted coffin, though apparently still in a boat (st.~100).  The equation of stoning and cremation clearly points towards composition after conversion, whereas pre-Christian poems like \textlink{Havamal} show no such assocation (cf.~\textlink{Havamal}[71]/3, 81/1).  Iceland officially became Christian in 1000 \ce\ and Greenland probably did not follow long thereafter, but the old custom of cremation can hardly have become thoroughly taboo immediately in the way it clearly is here.

On the linguistic side of things \Atlamal\ shows noticeably younger characteristics than its predecessor \textlink{Atlakvida}.

Most notably the sound change \emph{vr-} > \emph{r-} is consistently applied in roots originally containing the glide, namely \emph{vręiðr, vręiði} > \emph{ręiðr, ręiði} ‘wroth, wrath’ (\textlink{Atlamal}[51]/1: \emph{rǫskr, rǿða} : \emph{ręiðr}, 75/4: \emph{ró, ręiði} : \emph{ręynir} vs. \textlink{Atlakvida}[2]/2: \emph{vïn, val-hǫllu} : \emph{vręiði}),
\emph{vrangr, vręngja} > \emph{rangr, ręngja} ‘twisted, twist’ (\textlink{Atlamal}[4]/1: \emph{ru̇nar, rísta} : \emph{ręngði} vs. Bragi \emph{Þórr} 6 (in \Skp\ 3): \emph{vildi-t, vrǫngum : vágs}),
and \emph{*vrǫskr} > \emph{rǫskr} ‘vigorous’ (\textlink{Atlamal}[51]/1, 57/2: \emph{rǫskr, ręyna⸗sk : ręynt}, 87/1: \emph{rǫskr, rǿða : rakði⸗sk} vs. \Thorsdrapa\ 22/1: \emph{vręiðr, Vrǫsku : vá}).  For the root \emph{vreka} > \emph{reka} ‘drive, chase’, which is not found in \Atlamal, cf.~\textlink{Atlakvida}[14]/4.

Another young feature is the comparative rarity of the pre-verbal particle \emph{of}, which appears three times.  Of all 37 Eddic poems or sections of poems treated separately by \textcite[70]{Sapp2022}, \Atlamal\ has the lowest frequency of this particle at 0.039 occurrences per 10 unique lines.  This contrasts with the 6 occurrences in the much shorter \textlink{Atlakvida}, which has a middling frequency of 0.1709 occurrences per 10 unique lines.

\newpage

\section{Text}

\subsection{Atla mǫ́l in grǿn-lendsku (‘The Greenlandish Speeches of Attle’)}

\bvg\bva%
\Ballnote{Sts.~1--2 describe how Attle and his advisors decide to trick Guther and Hawn and betray them.}%
\edtext{Frétt}{\Afootnote{The \emph{F} is an ornamented initial 3 lines high, with the left vertical stroke tail-shaped and descending an additional 6 lines in the margin in \Regius.}} hęfir \alst{ǫ}ld \alst{ó}fu \hld\ þȧ’s \alst{ę}ndr of gørðu &
\alst{s}ęggir \alst{s}am-kundu, \hld\ \alst{s}ú vas nýt fę́stum; &
\alst{ǿ}xtu \alst{ęi}n-mę́li, \hld\ \alst{y}ggt vas þęim síðan &
ok it \alst{s}ama \alst{s}onum Gjúka \hld\ es vǫ́ru \alst{s}ann-ráðnir.\eva

\bvb {\huge F}\textsc{olk have learned} of the enmity long ago \\
when men held an assembly; that was a boon to the fewest. \\
Their private speeches intensified; it was terrible for them afterwards, \\
and the same for the sons of Yivick who were truly deceived.\evb\evg


\bvg\bva%
\edtrans{\alst{Sk}ǫp ǿxtu \alst{sk}jǫldunga}{They intensified the Shapes of the noblemen}{\Bfootnote{Attle and his men caused cruel fates (\emph{skǫp} ‘Shapes’) to afflict Guther and Hawn.}} \hld\ ---\alst{sk}yldu⸗at fęigir--- &
\alst{i}lla réð⸗sk \alst{A}tla, \hld\ \alst{á}tti hann þó hyggju; &
\edtrans{fęlldi \alst{st}oð \alst{st}óra}{He toppled a great pillar}{\Bfootnote{Symbolic.  His own greed was to be his undoing, just like a man who topples a pillar only for it to fall upon himself.}}, \hld\ \alst{st}ríddi sér harð⸗la, &
af \alst{b}ragði \alst{b}oð sęndi \hld\ at kvę̇mi \alst{b}rátt mágar.\eva

\bvb They intensified the Shapes of the noblemen---they should not have been fey--- \\
evil was coming for Attle although he had wit. \\
He toppled a great pillar, injuried himself grievously, \\
hurriedly sent a summons that his inlaws should come soon.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\Ballnote{Guthrun overhears the plans of her husband but since she herself cannot come to warn her brothers she comes up with a plan.}%
\alst{H}orsk vas \alst{h}ús-fręyja, \hld\ \alst{h}ugði at man-viti, &
\alst{l}ag hęyrði hǫ̇n orða, \hld\ hvat þęir ȧ \alst{l}aun mę́ltu; &
þȧ vas \alst{v}ant \alst{v}itri, \hld\ \alst{v}ildi hǫ̇n þęim hjalpa, &
skyldu of \alst{s}ę́ \alst{s}igla \hld\ ęn \alst{s}jǫlf né kom⸗sk⸗at.\eva

\bvb Wise was the lady of the house; she thought cleverly; \\
she heard the gist of the words which the men spoke in secret. \\
Then the clever woman was pressed; she wished to help them; \\
the men were to sail o’er the sea but she herself would not come.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{R}u̇nar nam at \alst{r}ísta, \hld\ \alst{r}ęngði þę́r Vingi--- &
\alst{f}árs vas hann \alst{f}lýtandi--- \hld\ áðr hann \alst{f}ramm sęldi, &
fóru þȧ \alst{s}íðan \hld\ \alst{s}ęndi-męnn Atla &
umb \edtrans{\alst{f}jǫrð Lima}{Limefirth}{\Bfootnote{Limfjorden in the north of Jutland; the localisation of the poem’s events in Denmark rather than in Burgundy (as the more archaic \textlink{Atlakvida}[21]/2) is a secondary development of the story likely based on the confusion between the continental and originally Asiatic Huns and the similarly named tribe of Denmark (TODO).}} \hld\ þar’s \alst{f}rǿknir bjoggu.\eva

\bvb She took to carve runes; Winge distorted them--- \\
he was a hastener of dangers---before he handed them over. \\
Then thereafter journeyed Attle’s messengers \\
around the Limfirth whither the brave men dwelled.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{Ǫ}l-vę́rir \alst{u}rðu \hld\ ok \alst{ę}lda kyndu, &
hugðu \alst{v}ę́tr \alst{v}éla \hld\ es þęir \alst{v}ǫ́ru komnir; &
tóku þęir \edtrans{\alst{f}órnir}{gifts}{\Bfootnote{Apparently whereon the runes were inscribed as suggested by the implication in the last line of the stanza.  The fact that they can be hung on a pillar seems to suggest that Guthrun embroidered the runes into a tapestry.}} \hld\ es þęim \alst{f}ríðr sęndi, &
\alst{h}ęngðu ȧ súlu, \hld\ \alst{h}ugðu⸗t þat varða.\eva

\bvb They were generous with ale and kindled the fires, \\
nowise suspected wiles from those who had come. \\
They took the gifts which the beautiful woman \ken*{= Guthrun} had sent them, \\
hung them on a pillar, did not think it had meaning.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{K}om þȧ \alst{K}ost·bera, \hld\ \alst{k}vę̇n vas hǫ̇n Hǫgna, &
\alst{k}ona \alst{k}apps-gȧ⸗lig, \hld\ ok \alst{k}vaddi þȧ báða; &
\alst{g}lǫð vas ok \alst{G}laum·vǫr \hld\ es \alst{G}unn·arr átti, &
fell⸗sk⸗at \edtrans{\alst{s}aðr}{virtue}{\Bfootnote{Literally ‘truth’.}} \alst{s}viðri, \hld\ \alst{s}ýsti of þǫrf gęsta.\eva

\bvb Then came Costbearer---she was Hawn’s wife, \\
a very discerning woman---and greeted them both. \\
Also glad was Gleamware, she whom Guther had for wife; \\
virtue did not fail that wise woman; she looked after the needs of the guests.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
Buðu þęir \alst{h}ęim \alst{H}ǫgna \hld\ ef hann þȧ \alst{h}ęldr fǿri; &
\edtext{\alst{s}ýn vas \alst{s}vip-vísi \hld\ ef þęir \alst{s}ïn gę̇ði}{\Bfootnote{Repeated in 71/2 below.}}; &
\alst{h}ét þá Gunn·arr \hld\ ef \alst{H}ǫgni vildi, &
\alst{H}ǫgni því nítti \hld\ es \alst{h}inn of réði.\eva

\bvb They invited Hawn to their home if he would be more willing to go; \\
the deceit was apparent if they would have watched out for themselves. \\
Guther asked if Hawn wished to go; \\
Hawn turned down what the other one counselled.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
Bǫ́ru \alst{m}jǫð \alst{m}ę́rar, \hld\ \alst{m}args vas alls bęini, &
\alst{f}ór þar \alst{f}jǫlð horna \hld\ und’s þȯtti \alst{f}ull-drukkit.\eva

\bvb The famous women bore mead; the servings were all plentiful; \\
there went forth a multitude of horns before it seemed fully drunk.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\Ballnote{Sts.~9--19 are paraphrased in \VolsungaSaga\ 34.}%
\alst{H}jú gørðu \alst{h}vílu \hld\ sęm þęim \alst{h}ǿgst þȯtti; &
\alst{k}ęnnd vas \alst{K}ost·bera, \hld\ \alst{k}unni hǫ̇n skil ru̇na, &
\alst{i}nnti \alst{o}rð-stafi \hld\ at \alst{ę}ldi ljósum; &
\alst{g}ę̇ta varð hǫ̇n tungu \hld\ \edtrans{ï \alst{g}óma báða}{between her gums}{\Bfootnote{Lit.~‘in her gums’, that is, due to the difficulty of reading the runes.}}: &
\alst{v}ǫ́ru svá \alst{v}illtar \hld\ at vas \alst{v}ant at ráða.\eva

\bvb The couple made their place of rest as they thought most pleasing; \\
Costbearer was renowned; she knew the distinctions of runes; \\
she interpreted the word-staves by the bright fire. \\
She had to hold her tongue between her gums; \\
they were so confused that there was little to read.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{S}ę́ing fóru \alst{s}íðan \hld\ \alst{s}ïna þau Hǫgni; &
\alst{d}ręymði \alst{d}rótt-láta, \hld\ \alst{d}ulði þęss vę́ttki, &
sagði \alst{h}orsk \alst{h}ilmi \hld\ þegar’s \alst{h}ȯn réð vakna:\eva

\bvb Then they went to their bed, she and Hawn; \\
the courtly-faced woman dreamt, hid no part of it; \\
the wise woman told the prince when she awoke:\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\llap{„}\alst{H}ęiman gøri⸗sk þú, \alst{H}ǫgni, \hld\ \alst{h}ygg þú at rǫ́ðum! &
\alst{F}ár ’s \alst{f}ull-rýninn, \hld\ \alst{f}ar þú ï sinn annat! &
\alst{R}éð’k þę́r \alst{r}u̇nar \hld\ es \alst{r}ęist þïn systir: &
\alst{b}jǫrt hęfir þér ęigi \hld\ \alst{b}oðit ï sinn þetta.\eva

\bvb “Thou makest ready to leave home, Hawn; think about this counsel. \\
Few know runes fully; journey another time! \\
I read the runes which thy sister carved; \\
the bright woman has not summoned thee this time.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{Ęi}tt ek męst \alst{u}ndr⸗umk \hld\ ---má’k⸗at \alst{ę}nn hyggja--- &
hvat þȧ \alst{v}arð \alst{v}itri \hld\ es skyldi \alst{v}illt rísta; &
því’t svá vas \alst{ȧ} \emph{ví}sat \hld\ sęm \alst{u}ndir vę́ri &
\alst{b}ani ykkarr \alst{b}ęggja \hld\ ef it \alst{b}rál⸗la kvę̇mið. &
\alst{V}ant ’s stafs \alst{v}ífi \hld\ eða \alst{v}alda aðrir.“\eva

\bvb I wonder most about a single thing; I cannot get to the bottom of it: \\
what happened to the clever woman that she should carve so confused, \\
for it was so indicated as if it were underneath \\
that it would the bane of you both if ye came hurriedly. \\
The wife left out a stave---or others are at fault.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\llap{„}\alst{A}llar ’ro \alst{i}ll-úðgar,“ \hld[kvað Hǫgni,] „\alst{á}’kk⸗a’k þęss kynni, &
vil’k⸗a’k þęss \alst{l}ęita \hld\ nema \alst{l}auna ęigim; &
okkr mun \alst{g}ramr \alst{g}olli \hld\ ręifa \alst{g}lóð-rauðu, &
\alst{ó}⸗umk ek \alst{a}ldri⸗gi \hld\ þó’tt vér \alst{ó}gn fregnim.“\eva

\bvb “All women are suspicious;” quoth Hawn, “I have not such a mind. \\
I will not look out for such unless we have to repay it. \\
The chief will gladden us two with glowing-red gold; \\
I shall never fear, although we should learn of something fearful.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\Ballnote{Sts.~14--19 deal with Costbearer’s prophetic dreams as told to her husband Hawn.}%
\speakernote{[Kost·bera kvað:]}%
\llap{„}\alst{St}opalt munuð ganga \hld\ ef it \alst{st}undið þangat, &
\alst{y}kkr mun \alst{ǫ̇}st-kynni \hld\ \alst{ęi}gi ï sinn þętta. &
\alst{D}ręymði mik, Hǫgni, \hld\ \alst{d}ylj⸗umk þat ęigi: &
Ganga mun \alst{y}kkr \alst{a}nd·ǿris \hld\ eða \alst{ę}lla hrę́ð⸗umk.\eva

\bvb\speakernoteb{[Quoth Costbearer:]}%
“Stumbling will ye two go if ye proceed thither; \\
there will not be any loving welcome shown you this time. \\
I had a dream, Hawn; I will not hide it: \\
It will go ill for you two---or else I am just afraid.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{B}lę́ju hugða’k þïna \hld\ \alst{b}renna ï ęldi, &
\alst{h}ryti \alst{h}ǫ́r logi \hld\ \alst{h}ús mïn ï gǫgnum.“ &
\speakernote{[Hǫgni kvað:]}%
\llap{„}\alst{L}iggja hér \alst{l}ïn-klę́ði \hld\ \edtext{þau\emph{’s} ér}{\Afootnote{\emph{‘þau er’} \Regius}} \alst{l}ítt rǿkið, &
þau munu \alst{b}rátt \alst{b}renna \hld\ þar’s þú \alst{b}lę́ju sátt.“\eva

\bvb I thought thy cover were burning in fire; \\
the high flame roared throughout my halls.” \\
\speakernoteb{[Quoth Hawn:]}%
“Here lie linen clothes which ye little do mind; \\
they will soon burn where thou didst see the cover.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Kost·bera kvað:]}%
\llap{„}\alst{B}jǫrn hugða’k hér inn kominn, \hld\ \alst{b}ryti upp stokka, &
\alst{h}risti svá \alst{h}ramma \hld\ at vér \alst{h}rę́dd yrðim; &
\alst{m}unn oss \alst{m}ǫrg hęfði \hld\ svá’t vér \alst{m}ę́ttim ękki; &
\alst{þ}ar vas ok \alst{þ}rǫmmun \hld\ \alst{þ}ęy⸗gi svá lítil.“\eva

\bvb\speakernoteb{[Quoth Costbearer:]}%
“I thought a bear came in hither, broke up the rafters, \\
struck so fiercely that we were frightened, \\
held many of us in its mouth so that we could nowise resist; \\
there was also crashing not at all small.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Hǫgni kvað:]}%
\llap{„}\alst{V}eðr mun þar \alst{v}axa, \hld\ \alst{v}erða ótt snemma; &
\edtrans{\alst{h}víta-bjǫrn \alst{h}ugðir: \hld\ þar mun \alst{h}regg austan}{thou thoughtest of a white-bear---there comes a tempest from the east}{\Bfootnote{The image is undoubtedly Greenlandish; polar bears float from the east to the island on large sheets of ice.  There is a real power to these lines which gives a sense of the harshness of the Greenland climate, even by Icelandic standards.}}.“\eva

\bvb\speakernoteb{[Quoth Hawn:]}%
“The storm will grow yonder, turn violent shortly; \\
thou thoughtest of a white-bear---there comes a tempest from the east.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Kost·bera kvað:]}%
\llap{„}\alst{Ǫ}rn hugða’k hér \alst{i}nn fljúga \hld\ \edtext{at \alst{ę}nd-lǫngu húsi}{\Bfootnote{Also occurring at 24/1b below.  The verse appears to be a \Malahattr\ variant of the formulaic \Fornyrdislag\ b-verse \emph{ęnd-langan sal} (cf.~\textlink{Thrymskvida}[27]/2 n.)}}, &
þat mun oss \alst{d}rjúgt \alst{d}ęila⸗sk, \hld\ \alst{d}ręifði hann oss ǫll blóði; &
\alst{h}ugða’k af \alst{h}ęitum \hld\ at vę́ri \edtrans{\alst{h}amr Atla}{Attle’s hame}{\Bfootnote{I.e., Attle in the shape of an eagle.}}.“\eva

\bvb\speakernoteb{[Quoth Costbearer:]}%
“I thought an eagle flew in hither through the end-long hall; \\
that must be of weight for us; he spattered us all in blood; \\
I thought from the names that it were Attle’s \inx[P]{hame}!”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Hǫgni kvað:]}%
\llap{„}\alst{S}lǫ́trum \alst{s}ýs⸗liga, \hld\ \alst{s}\emph{éu}m þȧ roðru; &
\alst{o}pt ’s þat fyr \alst{ǫ}xnum \hld\ es \alst{ǫ}rnu dręymir; &
\alst{h}ęill ’s \alst{h}ugr Atla \hld\ \alst{h}vat-ki’s þik dręymir.“ &
\alst{L}okit því \alst{l}étu, \hld\ \alst{l}íðr hvęr rǿða.\eva

\bvb\speakernoteb{[Quoth Hawn:]}%
“We will go to slaughter shortly---\emph{then} we will see blood; \\
it is oft a sign of oxen when one dreams of eagles. \\
Attle’s heart is wholesome no matter what thou dreamest.” \\
They closed the matter; each speech comes to its end.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\Ballnote{Gleamware, Guther’s wife, also has prophetic dreams; she tells them to Guther in sts.~20--26.   Sts.~20--38 are paraphrased in \VolsungaSaga\ 35.}%
\alst{V}ǫknuðu \alst{v}el-borin, \hld\ \alst{v}as þar sams dø̇mi, &
\alst{g}ę̇tti⸗sk þęss \alst{G}laum·vǫr \hld\ at vę́ri \alst{g}rand svefna &
{[...]} við Gunn·arr \hld\ at fȧ tvę́r lęiðir.\eva

\bvb The well-born couple awoke; the same thing happened there; \\
Gleamware asked for counsel about the blemishes of her sleep \ken{nightmares}; \\
{[...]} with Guther that it may have two outcomes:\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\llap{„}\alst{G}ǫrvan hugða’k þér \alst{g}alga, \hld\ \alst{g}engir þú at hanga &
\alst{ę́}ti þik \alst{o}rmar, \hld\ \alst{y}rða’k þik kvikvan, &
gørði⸗sk \alst{r}ǫk \alst{r}agna; \hld\ \alst{r}áð hvat þat vę́ri!\eva

\bvb “I thought a gallows were readied for thee; thou wentest to hang; \\
snakes did eat thee; I lost thee in life; \\
it was [like] the Rakes of the Reins.  Interpret what that was!\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{B}lóðgan hugða’k mę́ki \hld\ \alst{b}orinn ór serk þïnum, &
illt ’s \alst{s}vefn \alst{s}·líkan \hld\ at \alst{s}ęgja \edtrans{nauð-manni}{husband}{\Bfootnote{Hapax.  TODO.}}; &
\alst{g}ęir hugða’k standa \hld\ ï \alst{g}ǫgnum þik miðjan, &
\alst{ę}mjuðu \alst{u}lfar \hld\ ȧ \alst{ę}ndum bǫ́ðum.“\eva

\bvb I thought a bloodied sword were pulled out from thy shirt; \\
it is ill to tell of such a sleep to one’s husband. \\
I thought a spear were piercing through thy stomach; \\
wolves bellowed on both of its sides.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Gunn·arr kvað:]}%
\llap{„}\alst{R}akkar þar \alst{r}enna, \hld\ \alst{r}áða⸗sk mjǫk gęyja, &
opt verðr \alst{g}laumr hunda \hld\ fyr \alst{g}ęira flaugun.“\eva

\bvb\speakernoteb{[Quoth Guther:]}%
“Yonder hounds will run; they are loudly barking; \\
oft the revel of dogs comes before the hurling of spears.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Glaum·vǫr kvað:]}%
\llap{„}\alst{Ǫ́} hugða’k hér \alst{i}nn renna \hld\ at \alst{ę}ndi-lǫngu húsi, &
\alst{þ}yti af \alst{þ}jósti \hld\ \alst{þ}ęysti⸗sk of bękki; &
\alst{b}ryti fǿtr ykkra \hld\ \alst{b}rǿðra hér tvęggja, &
gørði⸗t \alst{v}atn \alst{v}ę́gja; \hld\ \alst{v}esa mun þat fyr nøkkvi.\eva

\bvb\speakernoteb{[Quoth Gleamware:]}%
“I thought a river ran in hither through the end-long hall; \\
it roared furiously, gushed over the benches, \\
here broke the feet of you brothers twain; \\
the water did not cede.  That must mean something!\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{K}onur hugða’k dauðar \hld\ \alst{k}oma ï nǫ́tt hingat, &
\alst{v}ę́ri⸗t \alst{v}art búnar, \hld\ \alst{v}ildi þik kjósa, &
\alst{b}yði þér \alst{b}ráð⸗liga \hld\ til \alst{b}ękkja sinna; &
\edtrans{\alst{e}k kveð \alst{a}f·lima \hld\ \alst{o}rðnar þér dísir}{For thee, I say, the dises have become feeble}{\Bfootnote{The goddesses of fate have become powerless to protect him due to his imminent death; for the expression cf.~\textlink{Grimnismal}[55].}}.“\eva

\bvb I thought dead women came hither tonight; \\
they were not poorly dressed; they would choose thee, \\
invited thee hurriedly onto their benches. \\
For thee, I say, the dises have become feeble!”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Gunn·arr kvað:]}%
\Ballnote{Guther at last admits that Gleamware likely has seen something.  He will not try to explain away the dreams further, but even if what they imply is true, fate has already decided that he and Hawn must journey to Attle.}%
\llap{„}\alst{S}ęinat ’s at \alst{s}ęgja, \hld\ \alst{s}vá ’s nú ráðit; &
\alst{f}orð⸗umk⸗a \alst{f}ǫr þó \hld\ alls þó ’s \alst{f}ara ę́tlat; &
\alst{m}argt ’s \alst{m}jǫk glík⸗ligt \hld\ at \alst{m}unim skamm-ę́ir.“\eva

\bvb\speakernoteb{[Quoth Guther:]}%
“It is too late to tell; it has now been thusly interpreted, \\
yet I cannot refrain from the journey when it is intended I should journey. \\
Much makes it likely that we shall be short-lived.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{L}itu es \alst{l}ýsti \hld\ \alst{l}étu⸗sk þęir fu̇sir &
\alst{a}llir \alst{u}pp rísa, \hld\ \alst{ǫ}nnur þau lǫttu; &
\alst{f}óru \alst{f}imm saman, \hld\ \alst{f}lęiri til vǫ́ru &
\alst{h}ǫlfu \alst{h}ús-karlar, \hld\ \alst{h}ugat vas því illa. &
\alst{S}nę́v·arr ok \alst{S}ól·arr, \hld\ \alst{s}ynir vǫ́ru þęir Hǫgna, &
\alst{O}rkning þann hétu \hld\ es þęim \alst{ę}nn fylgdi; &
\alst{b}líðr vas \alst{b}ǫrr skjaldar, \hld\ \edtrans{\alst{b}róðir hans kvȧnar}{brother of his wife}{\Bfootnote{The brother of Hawn’s wife, as Orkning was apparently Costbearer’s brother.}}.\eva

\bvb They beheld that it grew light; they said they were all \\
eager to arise; others would have restrained them. \\
They journeyed five men together; there were more men at hand, \\
twice as many housecarls; it was ill thought out. \\
Snower and Soler; they were the sons of Hawn. \\
Orkning they called yet one who followed them; \\
blithe was that shield-tree \ken{warrior}, the brother of his wife.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\Ballnote{The scene of sts.~28--32 is that the men, Guther, Hawn, Snower, Soler, and Orkning are in the parting ship alongside Attle’s messenger Winge, while the women, ashore, run alongside it.}%
\alst{F}óru \alst{f}agr-búnar \hld\ und’s þau \alst{f}jǫrðr skilði, &
\alst{l}ǫttu ȧ·vallt \alst{l}jósar, \hld\ \alst{l}étu⸗at hęldr sęgja⸗sk.\eva

\bvb The women went fair-clothed until the firth parted them; \\
the light ones would restrain them; the men would not be convinced.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{G}laum·vǫr kvað at orði, \hld\ es \alst{G}unn·arr átti, &
mę́lti hǫ̇n við \alst{V}inga \hld\ sęm hęnni \alst{v}ert þȯtti: &
\llap{„}\alst{V}ęit’k⸗at hvárt \alst{v}erð launið \hld\ at \alst{v}ilja ossum, &
\alst{g}lǿpr ’s \alst{g}ęsts kvȧma \hld\ ef ï \alst{g}øri⸗sk nǫkkvat.“\eva

\bvb Gleamware took to speak, she whom Guther had for wife; \\
she spoke with Winge as she thought worthiest: \\
“I know not whether ye will repay us according to our will; \\
a guest’s coming is a disgrace if something [ill] comes out of it!”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{S}ór þȧ Vingi, \hld\ \alst{s}ér réð hann lítt ęira: &
\llap{„}\alst{Ęi}gi \edtrans{hann}{him}{\Bfootnote{Winge himself.}} \alst{jǫ}tnar \hld\ ef hann at \alst{y}ðr lygi, &
\alst{g}algi \alst{g}ǫrv-allan \hld\ ef hann ȧ \alst{g}rið hygði.“\eva

\bvb Then Winge swore; he chose not to protect himself little: \\
“Let the ettins have him if he lies to you, \\
on a full gallows, if he thought to break the grith!”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{B}era kvað at orði, \hld\ \alst{b}líð ï hug sïnum: &
\llap{„}\alst{S}iglið ér \alst{s}ę́lir \hld\ ok \alst{s}igr árnið, &
\alst{f}ari sęm \alst{f}yr·mę́li’k! \hld\ \alst{F}ę́st ęigi því níta.“\eva

\bvb Bearer took to speak, blithe in her heart: \\
“May ye sail blessed and win victory; \\
may it go as I proclaim! May the fewest deny that.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\Ballnote{Hawn thanks the women for accompanying them to the point of their departure.}%
\alst{H}ǫgni svaraði \hld\ ---\alst{h}ugði gótt nǫ́num: &
\llap{„}\alst{H}uggið⸗sk it, \alst{h}orskar, \hld\ \alst{h}vé⸗gi’s þat gørvi⸗sk! &
\alst{M}ę́la þat \alst{m}argir, \hld\ \alst{m}issir þó stórum, &
mǫrgum rę́ðr \alst{l}itlu \hld\ hvé verð \alst{l}ęiddr hęiman.“\eva

\bvb Hawn answered; he thought well of his kin: \\
“Comfort yourselves, wise women, however it may turn out! \\
Many men say---yet it is utterly wrong, \\
that for many a man it matters little how he leaves his home.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{S}ǫ́u⸗sk til \alst{s}íðan \hld\ áðr ï \alst{s}undr hyrfi; &
þȧ hygg \alst{sk}ǫp \alst{sk}iptu, \hld\ \alst{sk}ilðu⸗sk vegir þęira.\eva

\bvb They looked at each other then before they were driven apart; \\
then I think the Shapes split them up; their roads did part.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\Ballnote{The men row furiously and leave the boat untied; they know that they will not return.}%
\alst{R}óa nǫ́mu \alst{r}íki, \hld\ \alst{r}ifu kjǫl halfan, &
\alst{b}ęystu \alst{b}ak-fǫllum, \hld\ \alst{b}rugðu⸗sk hęldr ręiðir; &
\alst{h}ǫmlur slitnuðu, \hld\ \alst{h}áir brotnuðu, &
gørðu⸗t \alst{f}ar \alst{f}ęsta \hld\ áðr þęir \alst{f}rȧ hyrfi.\eva

\bvb They took to row with strength, tore up half the keel, \\
struck the water leaning backwards, became rather wrathful; \\
the oar-straps snapped; the tholes broke; \\
they did not fasten their boat before they turned from it.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{L}itlu ok \alst{l}ęngra \hld\ ---\alst{l}ok mun’k þęss sęgja--- &
\alst{b}ǿ sǫ́u þęir standa \hld\ es \alst{B}uðli átti; &
\alst{h}ǫ́tt \alst{h}rikðu grindr \hld\ es \alst{H}ǫgni kníði.\eva

\bvb And a little later---I will tell it to its end--- \\
they saw standing the town which Budle owned; \\
loudly creaked the gates when Hawn moved them.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{O}rð kvað þȧ Vingi \hld\ þat’s \alst{ȧ}n vę́ri: &
\llap{„}\alst{F}arið \alst{f}irr húsi! \hld\ \alst{F}látt ’s til sǿkja, &
\alst{b}rátt hęfi’k ykkr \alst{b}ręnnda, \hld\ \alst{b}ragðs skuluð hǫggnir; &
\alst{f}agrt bað’k ykkr kvámu, \hld\ \alst{f}látt vas þó undir, &
\alst{ę}lla heðan bíðið \hld\ meðan \alst{e}k hǫgg yðr galga!“\eva

\bvb Then quoth Winge a word which he were better without: \\
“Fare ye far from the house! It is false to visit; \\
I will soon have you burned; ye shall shortly be cut down; \\
fairly I asked for your coming though there was falsehood underneath--- \\
or else stay here while I cut you a gallows!”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
Orð kvað \alst{h}itt \alst{H}ǫgni \hld\ ---\alst{h}ugði lítt vę́gja--- &
\alst{v}arr at \alst{v}ę́ttu⸗gi \hld\ es \alst{v}arð at ręyna: &
\llap{„}\alst{H}irð⸗a þú oss \alst{h}rę́ða, \hld\ \alst{h}af þú þat framm sjaldan! &
Ef þú \alst{ęy}kr \alst{o}rði \hld\ \alst{i}llt munt þér lęngja.“\eva

\bvb A different word quoth Hawn---he thought little of withdrawing--- \\
wary of nothing which he might come to find out: \\
“Do not try to frighten us; thou wilt seldom be successful! \\
If thou addest to thy word thy suffering will be prolonged!”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{H}rundu þęir Vinga \hld\ ok ï \alst{h}ęl drǫ́pu, &
\alst{ø}xar at lǫgðu \hld\ meðan ï \alst{ǫ}nd hiksti.\eva

\bvb They threw Wing and beat him into Hell, \\
laid their axes to him while he gasped for his breath.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{F}lykkðu⸗sk þęir Atli \hld\ ok \alst{f}óru ï brynjur, &
\alst{g}engu svá \alst{g}ǫrvir \hld\ at vas \alst{g}arðr \edtrans{milli}{between them}{\Bfootnote{I.e., between the two groups, Guther and Hawn on the one side and Attle’s men on the other.}}; &
\alst{u}rpu⸗sk ȧ \alst{o}rðum \hld\ \alst{a}llir sęnn ręiðir: &
\llap{„}\alst{F}yrr vǫ́rum \alst{f}ull-ráða \hld\ at \alst{f}irra yðr lífi.“\eva

\bvb Attle and his men arrayed themselves and got in their byrnies, \\
then went forth ready so that a courtyard stood between them. \\
They threw words at each other, all wrathful at once: \\
“We have long since fully decided to deprive you of life!”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\llap{„}\alst{Ȧ} sér þat \alst{i}lla \hld\ ef hǫfðuð \alst{á}ðr ráðit; &
\alst{ę}nn eruð \alst{ȯ}·búnir \hld\ ok hǫfum \alst{ęi}nn fęlldan, &
\alst{l}amðan til hęljar--- \hld\ \alst{l}iðs vas sá yðars!“\eva

\bvb “One thinks it looks bad if ye had decided beforehand; \\
ye are still unprepared and we have felled one man, \\
beat him lame to Hell---he was one of yours!”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{Ó}ðir þȧ \alst{u}rðu \hld\ es þat \alst{o}rð hęyrðu, &
\alst{f}orðuðu \alst{f}ingrum \hld\ ok \alst{f}engu ï snǿri, &
\alst{sk}utu \alst{sk}arp⸗liga \hld\ ok \alst{sk}jǫldum hlífðu⸗sk.\eva

\bvb Then they became mad when they heard that word, \\
twitched their fingers and seized their spears, \\
hurled them furiously and hid behind their shields.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{I}nn kom þȧ \alst{a}nn·spilli \hld\ hvat \alst{ú}ti drýgðu &
\alst{h}ǫ́tt fyr \alst{h}ǫllu; \hld\ \alst{h}ęyrðu þrę́l sęgja.\eva

\bvb The news came inside of what they were waging outside, \\
loudly through the hall; they heard a thrall speak.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{Ǫ}tul vas þȧ Guð·ru̇n \hld\ es hǫ̇n \alst{ę}kka hęyrði, &
\alst{h}laðin \alst{h}als-męnjum \hld\ \alst{h}ręytti hǫ̇n þęim gǫrv-ǫllum, &
\alst{s}løngði svá \alst{s}ilfri \hld\ at ï \alst{s}undr hrutu baugar.\eva

\bvb Then Guthrun became terrible when she heard his crying; \\
laden with neckrings she shook them off one and all; \\
she threw the silver so that the bighs crashed asunder.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{Ú}t gekk hǫ̇n síðan, \hld\ \alst{y}pði⸗t lítt hurðum, &
\alst{f}ór⸗a \alst{f}ę́lt þęy⸗gi \hld\ ok \alst{f}agnaði komnum; &
\alst{h}varf til Niflunga, \hld\ sú vas \alst{h}indst kvęðja, &
fylgði \alst{s}aðr \alst{s}·líku, \hld\ \alst{s}agði hǫ̇n mun flęira:\eva

\bvb Then she went outside, did not open the doors slowly, \\
went forth, nowise timid, and heartily greeted the comers. \\
She turned to the Nivlings; that was their last meeting; \\
virtue followed that doing; she spoke more from her mind:\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\llap{„}\alst{L}ęitaða’k ï \alst{l}íkna \hld\ at \alst{l}ętja ykkr hęiman, &
\edtrans{\alst{sk}ǫpum viðr mann⸗gi}{no man withstands the Shapes}{\Bfootnote{I.e., ‘no man can beat fate.’  The same construction is found in \textlink{Gripisspa}[53]/1b.}} \hld\ ok \alst{sk}uluð þó hér komnir.“ &
\alst{M}ę́lti \edtrans{af \alst{m}an-viti}{with little man-wit}{\Bfootnote{Lit.~‘against man-wit’, i.e., ‘with a girl’s naive sense’.
The men can hardly reconcile with Attle when they have already slain one of his men.}} \hld\ ef \alst{m}undu sę́tta⸗sk, &
\alst{ę}kki at réðu⸗sk, \hld\ \alst{a}llir ní kvǫ́ðu.\eva

\bvb “I tried to care for you by keeping you from leaving your home; \\
no man withstands the Shapes---and yet ye shall come here.” \\
She asked with little man-wit if they might be reconciled; \\
they did not plan to do that; they all said no.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\Ballnote{\Atlamal\ has Guthrun throwing off her clothes, taking a sword and fighting alongside her brothers.  Nothing of the sort is found in the older \textlink{Atlakvida} and it is clearly a later addition to the story, although it does show the high regard in which the Greenlanders held their women.  A similar anecdote is found in the Saw of the Greenlanders (the famous episode of Freydís, who bares her breasts in the fight against the natives. TODO: reference).}%
\alst{S}á þȧ \alst{s}ę́l-borin \hld\ at þęir \alst{s}árt léku, &
\alst{h}ugði ȧ \alst{h}arð-rę́ði \hld\ ok \alst{h}rauð⸗sk ór skikkju, &
\alst{n}økkðan tók hǫ̇n mę́ki \hld\ ok \alst{n}iðja fjǫr varði, &
\alst{h}ǿg vas at \alst{h}jaldri \hld\ hvar’s hǫ̇n \alst{h}ęndr fęsti.\eva

\bvb Then the woman of blessed birth saw that they sorely fought; \\
she thought of a harsh counsel and cast off her robe. \\
She took a naked sword and defended her kinsmen’s lives; \\
she was skilled in the fight where’er she turned her hands.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{D}óttir lét Gjúka \hld\ \alst{d}ręngi tvȧ hníga, &
\edtext{\alst{b}róður hjó hǫ̇n Atla}{\lemma{bróður \dots\ Atla ‘Attle’s brother’}\Bfootnote{An unknown figure.}}, \hld\ \alst{b}era varð þann síðan, &
\alst{sk}apði hǫ̇n svá \alst{sk}ę́ru, \hld\ \alst{sk}ęlldi fót undan.\eva

\bvb The daughter of Yivick let two young warriors sink down. \\
She struck Attle’s brother; he had to be borne away thereafter; \\
thus did she handle that engagement: she severed his foot from under him.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{A}nnan réð hǫ̇n hǫggva \hld\ svá’t sá \alst{u}pp ręis⸗at, &
ï \alst{h}ęlju hǫ̇n þann \alst{h}afði; \hld\ þęy⸗gi hęnni \alst{h}ęndr skulfu.\eva

\bvb She struck the other man so that he did not rise up; \\
she let that man go to Hell; yet nowise her hands trembled.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{Þ}jǫrku þar gjǫrðu, \hld\ \alst{þ}ęiri vas við brugðit; &
þat brá umb \alst{a}llt \alst{a}nnat \hld\ es \alst{u}nnu bǫrn Gjúka; &
\alst{s}vá kvǫ́ðu Niflunga \hld\ meðan \alst{s}jalfir lifðu &
skapa \alst{s}ókn \alst{s}verðum, \hld\ \alst{s}líta⸗sk af brynjur, &
\alst{h}ǫggva svá \alst{h}jalma \hld\ sęm þęim \alst{h}ugr dygði.\eva

\bvb They made war there; it was up and off; \\
it beat every other thing which the children of Yivick had accomplished. \\
So they said that the Nivlings while they themselves were alive \\
waged war with swords, tore off their byrnies, \\
struck so the helmets as did their hearts availed them.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{M}orgin \alst{m}ęst vǫ́gu \hld\ und’s \alst{m}iðjan dag líddi, &
\alst{ó}ttu \alst{a}lla \hld\ ok \alst{ǫ}nd·urðan dag; &
\alst{f}yrr vas \alst{f}ull-vegit, \hld\ \alst{f}lóði vǫllr blóði; &
\alst{á}t-tján \alst{á}ðr fellu, \hld\ \alst{ø}fri þęir urðu, &
\edtrans{\alst{B}eru \edtext{tvęim svęin\emph{um}}{\Afootnote{gramm.~and sens.~emend.; \emph{‘ii. sveinar’} (nom.~pl.) \Regius}} \hld\ ok \alst{b}róð\emph{u}r hęnnar}{over Bearer’s two boys and her brother}{\Bfootnote{The sense of the poem requires Bearer’s “two boys” (i.e., Guther and Hawn) to lose which justifies replacing the nom.~forms with dat.}}.\eva

\bvb For most of the morning they fought, until mid-day had passed, \\
the whole dawn and early part of the day. \\
Before it was fully fought the plain flowed with blood; \\
eighteen had then fallen; they got the upper hand \\
over Bearer’s two boys and her brother.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{R}ǫskr tók at \alst{r}ǿða \hld\ þó’tt hann \alst{r}ęiðr vę́ri: &
\llap{„}\alst{I}llt ’s \edtrans{umb líta⸗sk}{to look around}{\Bfootnote{And see the bloodied corpses lying on the ground.}}, \hld\ \alst{y}ðr ’s þat kęnna; &
vǫ́rum \alst{þ}rír tigir, \hld\ \alst{þ}egnar víg⸗ligir, &
\alst{ę}ptir lifum \alst{ę}llifu; \hld\ \edtrans{\alst{ó}r ’s þar brunnit}{the rest have burned away}{\Bfootnote{Lit.~‘it has burned away there’, symbolic; their numbers have waned greatly.}}.\eva

\bvb The bold man \ken*{= Attle} took to speak although he was wroth: \\
“It is evil to look around; this is your doing! \\
We were thirty thanes made ready for battle; \\
afterwards but eleven live; the rest have burned away.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{B}rǿðr vǫ́rum fimm \hld\ es \alst{B}uðla misstum, &%NOTE. NK has fjórir
hęfir nú \alst{H}ęl \alst{h}alfa \hld\ ęn \alst{h}ǫggnir tvęir liggja.\eva

\bvb We were five brothers when we lost Budle; \\
now Hell has half of them when two lie cut down.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\Ballnote{Attle addresses Guthrun.}%
\alst{M}ę́gð \edtrans{gat’k}{I received}{\Bfootnote{From his marriage with Guthrun.}} \alst{m}ikla \hld\ ---\alst{m}á’k⸗a’k því lęyna--- &
\alst{k}onu vá⸗liga, \hld\ \alst{k}ná’k⸗a’k þęss njóta; &
\alst{h}ljótt ǫ́ttum sjaldan, \hld\ síðst komt ï \alst{h}ęndr ossar, &
\alst{f}irrðan mik \alst{f}rę̇ndum, \hld\ \alst{f}é opt svikinn; &
\edtrans{\alst{s}ęnduð \alst{s}ystr Hęlju}{ye sent my sister to Hell}{\Bfootnote{Byrnhild, who committed suicide after the Yivickings Guther and Hawn slew Siward as told in \textlink{Sigurdskamma}.}}, \hld\ \alst{s}·líks ek męst kęnn⸗umk.“\eva

\bvb I received strong in-laws---I can not deny that--- \\
{[but]} a woeful wife; I do not enjoy that. \\
We seldom had peace since thou camest to our hands; \\
thou didst distance me from my kinsmen, oft betrayed me for wealth; \\
ye sent my sister to Hell---I suffer the most from that.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Guð·ru̇n kvað:]}%
\llap{„}\alst{G}etr þú þęss, Atli, \hld\ \alst{g}ørðir svá fyrri: &
\edtext{\alst{M}óður tókt \alst{m}ïna \hld\ ok \alst{m}yrðir til hnossa, &
\alst{s}vinna \alst{s}ystrungu \hld\ \alst{s}vęltir þú ï hęlli;}{\lemma{Móður \dots\ hęlli ‘Thou didst take \dots\ to death.’}\Bfootnote{The content is otherwise unknown.}} &
\alst{h}lǿg⸗ligt mér þat þikkir \hld\ es þú þïnn \alst{h}arm tïnir, &
\alst{g}oðum ek þat þakka \hld\ es þér \alst{g}ęng⸗sk illa.“\eva

\bvb\speakernoteb{[Quoth Guthrun:]}%
“Thou tellest, Attle, of what thou thyself didst yet earlier! \\
Thou didst take my mother and murder her for gold; \\
my wise cousins didst thou starve to death. \\
I find it laughable that thou speakest of thy griefs; \\
I thank the Gods when it goes ill for thee!”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Atli kvað:]}%
\llap{„}\alst{Ę}ggja’k yðr, \edtrans{\alst{ja}rlar}{earls}{\Bfootnote{Attle’s surviving companions.}}, \hld\ \alst{au}ka harm stóran &
\alst{v}ífs ins \alst{v}eg⸗liga, \hld\ \alst{v}ilja’k þat líta; &
\alst{k}ostið svá \alst{k}ęppa \hld\ at \alst{k}løkkvi Guð·ru̇n, &
\alst{s}já ek þat mę́tta’k \hld\ at hǫ̇n \alst{s}ér né ynði⸗t.\eva

\bvb\speakernoteb{[Quoth Attle:]}%
“I incite you, earls!  Increase the great harm \\
for the glorious wife; I want to behold that. \\
Strive to achieve that Guthrun sobs, \\
so that I might see that she is unhappy.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\Ballnote{Sts.~56--64 deal with the deaths of Guther and Hawn, especially focusing on the death of Hawn.  The same episode is found in \textlink{Atlakvida}[22]--35.}%
Takið ér \alst{H}ǫgna \hld\ ok \edtrans{\alst{h}yldið}{pry him open}{\Bfootnote{Otherwise only used of cutting up animals, esp.~whales.  \textcite{FinnurEdda} is most likely right in seeing its use here as a Greenlandicism.}} með knífi, &
\alst{sk}erið ór hjarta, \hld\ \alst{sk}uluð þęss gǫrvir; &
\edtext{\alst{G}unn·ar \alst{g}rimm-úðgan \hld\ ȧ \alst{g}alga fęstið, &
\alst{b}ęllið því \alst{b}ragði, \hld\ \alst{b}jóðið til ormum!}{\lemma{Gunn·ar \dots\ ormum! ‘Guther \dots\ snakes!’}\Bfootnote{Fulfilling the prophecy of st.~21.  In \textlink{Atlakvida} Guther is thrown in a snake-pit, not hung on the gallows.}}“\eva

\bvb Take ye Hawn and pry him open with a knife; \\
cut out his heart---ye shall be quick about it! \\
Guther the grim-hearted fasten on the gallows--- \\
hasten to that deed!---offer him to the snakes!”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{Hǫgni kvað:}%
\llap{„}\alst{G}ør sem til lystir! \hld\ \alst{G}laðr mun’k þęss bíða, &
\alst{r}ǫskr mun þér \alst{r}ęyna⸗sk, \hld\ \alst{r}ęynt hefi’k fyrr brattara; &
\alst{h}ǫfðuð \edtrans{\alst{h}nękking}{a setback}{\Bfootnote{Namely the loss of the eighteen retainers (st.~50).}} \hld\ meðan \alst{h}ęilir vǫ́rum, &
nú erum \alst{s}vá \alst{s}árir \hld\ at þú mátt \alst{s}jalfr valda.“\eva

\bvb\speakernoteb{Quoth Hawn:}%
“Do as thou desirest! I will suffer it gladly; \\
I’ll be proven brave to thee; I have tempted harder things before. \\
Ye suffered a setback while we were still healthy; \\
now we are so wounded that thou thyself mayst decide.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{B}ęiti þat mę́lti \hld\ ---\alst{b}ryti vas hann Atla: &
\llap{„}Tǫkum vér \alst{H}jalla \hld\ ęn \alst{H}ǫgna forðum! &
\alst{H}ǫgum vér \alst{h}alft-yrkjum, \hld\ \alst{h}ann es skap-dauði, &
\alst{l}ifir⸗a svá \alst{l}ęngi, \hld\ \alst{l}ǫskr mun ę́ hęitinn.“\eva

\bvb Bote spoke this---he was Attle’s overseer: \\
“Let us take Helle, but Hawn let us spare! \\
Let us handle the half-worker; he is shaped for death; \\
he will not live for so long; he will always be known as lazy.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{H}rę́ddr vas \alst{h}vęr-gę́tir, \hld\ \alst{h}elt⸗a in lęngr ru̇mi, &
\alst{k}unni \alst{k}løkkr verða, \hld\ \alst{k}lęif ï rǫ́ hvęrja; &
\alst{v}e-sall lét⸗sk \alst{v}ígs þęira \hld\ es skyldi \alst{v}áss gjalda &
ok sïnn \alst{d}ag \alst{d}apran \hld\ at \alst{d}ęyja frȧ svïnum, &
\alst{a}llri \alst{ø}r·kostu \hld\ es hann \alst{á}ðr hafði.\eva

\bvb The kettle-keeper was afraid, could not run further into the room, \\
knew how to start sobbing, hid in every nook; \\
said he was wretched o’er the feud of those who should repay him for his toil, \\
and that his day was dismal if he should die away from his swine, \\
from all the great pleasure which he had previously had.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
Tóku þęir \edtrans{\alst{b}rás}{cookboy}{\Bfootnote{A hapax, but apparently in ablaut with Swedish \emph{brasa} ‘fire’, i.e., one who tends to the fire.}} \alst{B}uðla \hld\ ok \alst{b}rugðu til knífi, &
\alst{ǿ}pði \alst{i}ll-þrę́li \hld\ áðr \alst{o}dds kęnndi; &
\alst{t}óm lét⸗sk⸗at ęiga, \hld\ \alst{t}ęðja vel garða, &
\alst{v}inna it \alst{v}ergasta \hld\ ef hann \alst{v}ið rétti; &
\alst{f}ęginn lét⸗sk þó Hjalli \hld\ at hann \alst{f}jǫr þę́gi.\eva

\bvb They took Budle’s cookboy and brandished the knife, \\
the bad thrall screamed before he saw the point. \\
He said he would take no leisure, dung the fields well, \\
do the meanest work if he might but be let go; \\
Helle said he would rejoice if he might but keep his life.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{G}ę̇tti⸗sk þess Hǫgni \hld\ ---\alst{g}ørva svá fę́ri--- &
at \alst{á}rna \alst{ȧ}·nauðgum \hld\ at \alst{u}ndan gengi: &
\llap{„}\alst{F}yrir kveð’k mér minna \hld\ at \alst{f}ręmja lęik þęnna; &
\alst{h}ví mynim \alst{h}ér vilja \hld\ \alst{h}ęyra ȧ þȧ skrę́ktun?“\eva

\bvb Hawn gave counsel---few men do so!--- \\
to grant the enthralled man that he might get away. \\
“It would be easier, I say, for me to endure this game; \\
why should we here wish to listen to this screeching?”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{Þ}rifu þęir \alst{þ}jóð-góðan, \hld\ \alst{þ}ȧ vas kostr ęngi &
\alst{r}ekkum \alst{r}akk-látum \hld\ \alst{r}áð ęnn lęngr dvęlja; &
\alst{h}ló þȧ \alst{H}ǫgni, \hld\ \alst{h}ęyrðu dag-męgir, &
\alst{k}ęppa hann svá \alst{k}unni, \hld\ \alst{k}vǫl hann vel þolði.\eva

\bvb They seized the very good man; then there was no choice \\
for the stern-faced retainers to tarry with their orders. \\
Then Hawn laughed; the sons of daylight \ken{workmen} heard it; \\
so did he know how to struggle; he suffered well the torment.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\Ballnote{The poet turns to Guther, who is on the gallows and at the same time being tortured by the wyrms.  For the scene of Guther playing the harp with his toes see \textlink{Atlakvida}[35].}%
\alst{H}ǫrpu tók Gunn·arr, \hld\ \alst{h}rǿrði il-kvistum; &
\alst{s}lá hann \alst{s}vá kunni \hld\ at \alst{s}nótir grétu; &
\alst{k}lukku þęir \alst{k}arlar \hld\ es \alst{k}unnu gørst hęyra; &
\edtrans{\alst{r}íkri}{powerful woman}{\Bfootnote{Presumably Guthrun; cf.~65/2 below.}} \alst{r}ǫ́ð sagði, \hld\ \alst{r}aptar sundr brustu.\eva

\bvb Guther took the harp, plucked it with his footsole-twigs \ken{toes}, \\
struck as he knew so that the ladies wept; \\
the menfolk cried who heard it clearest; \\
he gave the powerful woman counsel; the rafters burst asunder.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{D}óu þȧ \alst{d}ýrir, \hld\ \alst{d}ags vas hęldr snemma, &
\alst{l}étu þęir ȧ \alst{l}ęsti \hld\ \alst{l}ifa ï·þrótta.\eva

\bvb Then the precious men died; the day was rather early; \\
to the very last they let their skills linger on.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{St}órr þȯtti⸗sk Atli, \hld\ \alst{st}é hann of þȧ báða, &
\alst{h}orskri \alst{h}arm sagði \hld\ ok réð \alst{h}ęldr at bręgða: &
\llap{„}\alst{M}orginn ’s nú, Guð·ru̇n, \hld\ \alst{m}isst hęfir þér hollra, &
\alst{s}ums est \alst{s}jalf-skapa \hld\ at hafi \alst{s}vá gęngit.“\eva

\bvb Attle thought himself great; he stood over them both, \\
he told the wise woman a thing of harm and did harshly upbraid her: \\
“It is morning now, Guthrun; thou hast lost those true to thee; \\
it is partly thy own doing that things have gone so.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Guð·ru̇n kvað:]}%
\llap{„}\alst{F}ęginn est, Atli, \hld\ \alst{f}ęrr þú víg lýsa, &
\alst{ȧ} munu þér \alst{i}ðrar \hld\ ef \alst{a}llt ręynir; &
sú mun \alst{ę}rfð \alst{ę}ptir \hld\ ---\alst{e}k kann þér sęgja: &
\alst{i}lls gęng⸗sk þér \alst{a}ldri \hld\ nema \alst{e}k ok dęyja.“\eva

\bvb\speakernoteb{[Quoth Guthrun:]}%
“Thou art rejoicing, Attle; thou goest to announce the slayings; \\
thou wilt come to regret them if thou dost come to prove it all. \\
There will be an inheritance afterwards---\emph{I} can tell thee: \\
evil will never part from thee unless I too die.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Atli kvað:]}%
\llap{„}Kann’k⸗a’k \alst{s}·líks \alst{s}ynja, \hld\ \alst{s}é’k til ráð annat, &
\alst{h}ǫlfu \alst{h}óg⸗ligra \hld\ ---\alst{h}ǫfnum opt góðu: &
\edtext{\alst{M}an*i}{\Afootnote{metr.~and sens.~emend.; \emph{manni} \Regius.}} \alst{m}un’k þik hugga, \hld\ \alst{m}ę́tum ȧ·gę́tum, &
\alst{s}ilfri \alst{s}nę́-hvítu \hld\ sem \alst{s}jǫlf vilir.“\eva

\bvb\speakernoteb{[Quoth Attle:]}%
“I cannot deny such a thing; I see another counsel, \\
twice as fitting; we often abandon what is good. \\
I will comfort thee with a handmaid and with excellent treasures, \\
with snow-white silver, whate’er thou thyself desirest!”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Guð·ru̇n kvað:]}%
\llap{„}\alst{Ȯ}n es þęss \alst{ę}ngi, \hld\ \alst{e}k vil því níta, &
\alst{s}lęit’k þȧ \alst{s}áttir \hld\ es vǫ́ru \alst{s}akar minni; &
\alst{a}f·kǫ́r ek \alst{á}ðr þȯtta’k, \hld\ \alst{ȧ} mun nú gǿða, &
\alst{h}rę́fða’k umb \alst{h}ot-vetna \hld\ meðan \alst{H}ǫgni lifði.\eva

\bvb\speakernoteb{[Quoth Guthrun:]}%
“That will hardly be expected; I will refuse it. \\
I tore up reconciliations when the crimes were lesser; \\
I seemed violent before---I will now exceed it; \\
I put up with anything while Hawn still lived.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{A}lin \edtrans{vit}{we two}{\Bfootnote{Guthrun and Hawn.}} \alst{u}pp vǫ́rum \hld\ ï \alst{ęi}nu húsi, &
\alst{l}ékum \alst{l}ęik margan \hld\ ok ï \alst{l}undi óxum, &
\alst{g}ǿddi okkr \alst{G}rím·hildr \hld\ \alst{g}olli ok hals-męnjum. &
\alst{B}ana munt mér \alst{b}rǿðra \hld\ \alst{b}ǿta aldri⸗gi &
né \alst{v}inna þęss ękki \hld\ at mér \alst{v}el þikki.\eva

\bvb We two were reared up in one house; \\
we played many a game and waxed in the glade; \\
Grimhild endowed us with gold and neckrings. \\
For the slaying of my brothers wilt thou never restore me, \\
nor accomplish anything that might seem good to me.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{K}ostum drepr \alst{k}venna \hld\ \alst{k}arla of·ríki, &
\edtrans{ï \alst{k}né gęngr hnefi \hld\ ef \alst{k}vistir þverra}{The fist gets on its knees if the branches thin out}{\Bfootnote{\textcite{PettitEdda} sees this as a reference to the game Neve-Table (\emph{Hnefa-tafl}), where the \emph{hnefi} ‘fist’ is the name of the King, surrounded by his soldiers (Pawns or Rooks) on all sides, which seem to resemble branches.  For board games see \textlink{Voluspa}[8]/1 n.}}, &
\alst{t}ré \alst{t}ękr at hníga \hld\ ef hǫggr \alst{t}ág undan; &
nú mátt \alst{ęi}nn, \alst{A}tli, \hld\ \alst{ǫ}llu hér ráða.“\eva

\bvb The choices of women are beat down by the tyranny of men. \\
The fist gets on its knees if the branches thin out; \\
the tree begins to lean if one cuts away its roots from underneath it; \\
now mayst thou alone, Attle, rule everything here.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{G}nótt vas \alst{g}runn-ýðgi \hld\ es \alst{g}ramr því trúði, &
\alst{s}ýn vas \alst{s}vęip-vísi \hld\ ef hann \alst{s}ïn gę̇ði. &
\alst{K}rǫpp vas þá Guð·ru̇n, \hld\ \alst{k}unni umb hug mę́la, &
\alst{l}ėtt hǫ̇n sér gørði, \hld\ \alst{l}ék hǫ̇n tvęim skjǫldum.\eva

\bvb There was gullibility enough when the chief believed that; \\
the deceit was apparent if he would have watched out for himself. \\
Then Guthrun was hard; she could speak with care; \\
she made herself out to be light; she played with two shields.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{Ǿ}xti hǫ̇n \alst{ǫ}l-drykkjur \hld\ at \alst{ę}rfa brǿðr sína, &
\alst{s}amr lét⸗sk ok Atli \hld\ at \alst{s}ïna gørva.\eva

\bvb She made the ale-vessels grow to drink the wake of her brothers; \\
Attle, too, did the same for his own.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{L}okit því \alst{l}étu, \hld\ \alst{l}agat vas drykkju; &
\alst{s}ú vas \alst{s}am-kunda \hld\ við \alst{s}vǫrfun of·mikla; &
\alst{st}rǫng vas \alst{st}ór-huguð, \hld\ \alst{st}ríddi hǫ̇n ę́tt Buðla, &
\edtext{\alst{v}ildi hǫ̇n \alst{v}er sïnum \hld\ \alst{o}fr·hęfndir}{\Bfootnote{Alliteration between \emph{v-} and vowels is by no means unheard of in Norse poetry but varies greatly between styles.  In the Scaldic genre, for instance, it was always entirely forbidden.}}.\eva

\bvb They closed the matter; the drink was brewed; \\
that was an assembly with too great destruction. \\
Cruel was the great-minded woman; she injured the lineage of Budle; \\
against her husband she wished to wreak great revenge.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\Ballnote{Guthrun kills her own children, Earp and Oatle.  The killing is not described directly in \textlink{Atlakvida} but is alluded to in \textlink{Atlakvida}[37] and mentioned in 40--41.}%
\alst{L}okkaði hǫ̇n \alst{l}itla \hld\ ok \alst{l}agði við stokki; &
\alst{g}lúpnuðu \alst{g}rimmir \hld\ ok \alst{g}rétu þęy⸗gi, &
\alst{f}óru ï \alst{f}aðm móður, \hld\ \alst{f}réttu hvat þȧ skyldi.\eva

\bvb She called on her little ones and laid them against a stump; \\
the grim ones were downcast and yet nowise cried; \\
they jumped to their mother’s bosom, asked what would happen.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Guð·ru̇n kvað:]}%
\llap{„}\alst{Sp}yrið lítt ęptir! \hld\ \alst{Sp}illa ę́tla’k bǫ́ðum, &
\alst{l}yst vǫr⸗umk þęss \alst{l}ęngi \hld\ at \alst{l}yfja ykkr ęlli.“ &
\speakernote{[Bǫrn’in:]}%
\llap{„}\edtrans{\alst{B}lótt}{Bloot}{\Bfootnote{I.e., ‘sacrifice’.}} sęm vilt \alst{b}ǫrnum, \hld\ \alst{b}annar þat mann⸗gi, &
skǫmm mun \alst{r}ó \alst{r}ęiði \hld\ ef þú \alst{r}ęynir gǫrva.“\eva

\bvb\speakernoteb{[Quoth Guthrun:]}%
“Ask no more what is coming! I intend to slay you both, \\
I have long had lust to deprive you of old age.” \\
\speakernoteb{[The children:]}%
“Bloot thy children if thou wilt; no man forbids thee; \\
the respite of thy wrath will be short if thou dost truly try it.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{B}rá þá \alst{b}arn-ø̇sku \hld\ \alst{b}rǿðra in kapp-svinna, &
\alst{sk}ipti⸗sk \alst{sk}ap⸗liga, \hld\ \alst{sk}ar hǫ̇n ȧ háls báða. &
Ęnn \alst{f}rétti Atli \hld\ hvęrt \alst{f}arnir vę́ri &
\alst{s}vęinar hans lęika \hld\ es hann \alst{s}á þȧ hvęr⸗gi.\eva

\bvb Then childhood was cut short for the battle-wise brothers, \\
it was overturned according to Shapes---she slit the throats of them both. \\
Soon Attle asked whither they had gone, \\
his swains in their play, when he nowhere saw them.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\llap{„}\alst{Y}fir rǫ́ð⸗umk ganga, \hld\ \alst{A}tla til sęgja, &
\alst{d}ylja mun’k þik ęigi, \hld\ \alst{d}óttir Grím·hildar; &
\alst{g}laða mun þik minnst, Atli, \hld\ ef þú \alst{g}ǫrva ręynir; &
\alst{v}akðir \alst{v}ǫ́ mikla \hld\ es þú \alst{v}átt brǿðr mïna.\eva

\bvb “I decide to come over to tell thee, Attle; \\
I, the daughter of Grimhild, will not hide it from thee. \\
It will gladden thee the least if thou dost truly find it out; \\
thou didst awaken great woe when thou slewest my brothers.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{S}vaf’k mjǫk \alst{s}jaldan, \hld\ \alst{s}íðan’s þęir fellu, &
\alst{h}ét’k þér \alst{h}ǫrðu, \hld\ \alst{h}ęfi’k þik nú minntan; &
\edtrans{\alst{m}orgin}{This morning}{\Bfootnote{Cf.~st.~65.}} \alst{m}ér sagðir, \hld\ \alst{m}an’k ęnn þann gǫrva, &
nú ’s \alst{o}k \alst{a}ptann, \hld\ \alst{á}tt s·líkt at frétta.\eva

\bvb I slept very seldom since they fell; \\
I vowed harsh things towards thee; I have now reminded thee. \\
This morning thou spokest to me; I still remember it clearly. \\
Now it is the evening; thou hast to learn such things:\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\Ballnote{In sts.~79--81 Guthrun recounts how she fed Attle the corpses of their children.  Cf.~the description of the same events in \textlink{Atlakvida}[40]--41; the strong correspondences suggest a direct line of influence.}%
\alst{M}aga hęfir þïnna \hld\ \alst{m}isst sęm síðst skyldir; &
\alst{h}ausa vęitst þęira \hld\ \alst{h}afða at ǫl-skǫ́lum, &
\alst{d}rýgða’k þér svá \alst{d}rykkju--- \hld\ \alst{d}ręyra blett’k þęira.\eva

\bvb Thou hast lost thy lads as thou least shouldst have. \\
I know their skulls have been made into ale-bowls; \\
thus I enriched thy drink: I mixed it with their blood.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{T}ók’k þęira hjǫrtu \hld\ ok ȧ \alst{t}ęini stęikta’k, &
\alst{s}ęlda’k þér \alst{s}íðan, \hld\ \alst{s}agða’k at kalfs vę́ri; &
\alst{ęi}nn þú því \alst{o}llir \hld\ \alst{ę}kki rétt lęifa, &
\alst{t}ǫggt \alst{t}íð⸗liga, \hld\ \alst{t}rúðir vel jǫxlum.\eva

\bvb I took their hearts and roasted them on a stick, \\
then I handed them to thee; I said they were a calf’s. \\
Thou alone didst do that; thou decidedst to leave nothing, \\
didst chew greedily; thou trustedst thy molars well!\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{B}arna vęitst þïnna \hld\ ---\alst{b}iðr sér fár verra--- &
\alst{h}lut veld’k mïnum, \hld\ \alst{h}ę́l⸗umk þó ękki.“\eva

\bvb Thou knowest of thy children; few will ask for worse; \\
I bear fault for my part although I do not boast of it.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\llap{„}\alst{G}rimm vast, \alst{G}uð·ru̇n, \hld\ es þú \alst{g}øra svá máttir, &
\alst{b}arna þïnna \alst{b}lóði \hld\ at \alst{b}landa mér drykkju; &
\alst{s}nýtt hęfir \alst{s}ifjungum \hld\ sęm \alst{s}íðst skyldir, &
\alst{m}ér lę́tr þú ok sjǫlfum \hld\ \edtrans{\alst{m}illum ills lítit}{little [time] between evil}{\Bfootnote{Referring to Guthrun’s recent killing of Attle’s brother (sts.~47--48).}}.“\eva

\bvb “Thou wast cruel, Guthrun, when thou couldst do so, \\
to blend thy own children’s blood into my drink. \\
Thou hast destroyed thy own relations as thou least shouldst have; \\
for me leavest thou little [time] between evil.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Guð·ru̇n kvað:]}%
\llap{„}\alst{V}ili mér ęnn \alst{v}ę́ri \hld\ at \alst{v}ega þik sjalfan, &
\alst{f}átt ’s \alst{f}ull-illa \hld\ \alst{f}arit við gram s·líkan.“ &
\speakernote{[Atli kvað:]}%
\llap{„}\alst{D}rýgt þú fyrr hafðir \hld\ þat’s męnn \alst{d}ø̇mi vissu⸗t, &
til \alst{h}ęimsku, \alst{h}arð-rę́ðis, \hld\ \edtrans{ï \alst{h}ęimi þessum}{in this Home}{\Bfootnote{I.e., in the world of mankind.}}; &
nú hęfir þú \alst{ę}nn aukit \hld\ þat’s nú \alst{á}ðan frǫ́gum, &
\alst{g}ręipt \alst{g}lǿp stóran, \hld\ \alst{g}ørt hęfir þú þitt ęrfi.\eva

\bvb\speakernoteb{[Quoth Guthrun:]}%
“If but I still had willpower to slay thy very self; \\
little is evil enough when done against such a prince!” \\
\speakernoteb{[Quoth Attle:]}%
“Thou hast already done that of which men knew no equivalent \\
in foolishness and harsh counsel in this Home. \\
Now hast thou yet added to what we earlier had learned; \\
thou didst accomplish a great disgrace; thou hast had thy wake!\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\edtrans{\alst{B}ręnnd munt ȧ \alst{b}áli \hld\ ok \alst{b}arið grjóti áðr}{Thou wilt be burned on a pyre and struck with stones beforehand}{\Bfootnote{A Christian, mediæval punishment.  That being burned on a pyre is seen as a punishment and associated with stoning reveals a mediæval Catholic sensibility horrified at the thought of burning a body rather than burying it.  In pagan times, however, burning was a neutral way of disposing of a body and had no negative associations (so e.g. \textlink{Havamal}[71]/3, 81/1); cf.~\textcite[93--96]{Males2024Hav}.}}, &
þá hęfir þú \alst{á}rnat \hld\ þat’s \alst{ę́} bęiði⸗sk.“ &
\speakernote{[Guð·ru̇n kvað:]}%
\llap{„}\alst{S}ęg þér \alst{s}·líkar \hld\ \alst{s}orgir ár morgin, &
\alst{f}ríðra vil’k dauða, \hld\ \alst{f}ara ï \edtrans{ljós annat}{the other light}{\Bfootnote{The next world; the world as ‘light’ is more commonly found in West Germanic heroic poetry.}}.“\eva

\bvb Thou wilt be burned on a pyre and struck with stones beforehand; \\
then hast thou earned what thou always begged for!” \\
\speakernoteb{[Quoth Guthrun:]}%
“Tell thyself such sorrows early in the morning; \\
by a fairer death I wish to go into the other light.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{S}ǫ́tu \alst{s}am-týnis, \hld\ \alst{s}ęndu⸗sk fár-hugi, &
\alst{h}ęndu⸗sk \alst{h}ęipt-yrði, \hld\ \alst{h}várt⸗ki sér unði. &
\alst{H}ęipt óx \edtrans{\alst{H}niflungi}{Nivling}{\Bfootnote{\emph{Hniflungr}, the son of Hawn.}}, \hld\ \alst{h}ugði ȧ stór-rę́ði, &
\alst{g}at fyr \alst{G}uð·ru̇nu \hld\ at hann vę́ri \alst{g}rimmr Atla.\eva

\bvb They sat in the same court, gave each other evil thoughts, \\
handed each other hateful words; neither was content. \\
Hatred grew in Nivling; he thought of great counsel; \\
mentioned to Guthrun that he was furious towards Attle.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
Kvǫ́mu ï \alst{h}ug \alst{h}ęnni \hld\ \alst{H}ǫgna við·farar, &
talði \alst{h}app \alst{h}ǫ́num \hld\ ef \alst{h}ęfnt ynni; &
\alst{v}eginn vas þȧ Atli, \hld\ \alst{v}as þęss skammt bíða, &
\alst{s}onr vá Hǫgna \hld\ ok \alst{s}jǫlf Guð·ru̇n.\eva

\bvb Into her mind came Hawn’s demise; \\
she counted it lucky for him if he might accomplish the revenge. \\
Then Attle was slain; it was short to abide; \\
Hawn’s son struck him and Guthrun herself.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{R}ǫskr tók at \alst{r}ǿða, \hld\ \alst{r}akði⸗sk ór svefni, &
kęnndi \alst{b}rátt \alst{b}ęnja, \hld\ \alst{b}ands kvað þǫrf ønga: &
\llap{„}\alst{S}ęgið it \alst{s}annasta: \hld\ Hvęrr vá \alst{s}on Buðla? &
Em’k⸗a’k \alst{l}ítt \alst{l}ęikinn, \hld\ \alst{l}ífs tęl’k vǫ́n ønga.“\eva

\bvb The bold man took to speak, roused from his sleep, \\
shortly felt his wounds; said dressings were of no use. \\
“Tell the whole truth!  Who struck Budle’s son? \\
I am not a little injured; I count no hopes for my life.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Guð·ru̇n kvað:]}%
\llap{„}\alst{D}ylja mun þik ęigi \hld\ \alst{d}óttir Grím·hildar: &
\alst{L}ǫt⸗umk því valda \hld\ es \alst{l}íðr þïna ę́vi &
ęn \alst{s}umu \alst{s}onr Hǫgna \hld\ es þik \alst{s}ǫ́r mǿða.“\eva

\bvb\speakernoteb{[Quoth Guthrun:]}%
“The daughter of Grimhild will not hide it from thee: \\
I made myself the cause for why thy age comes to its end, \\
but some part the son of Hawn [plays] when the wounds do pain thee.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Atli kvað:]}%
\llap{„}\alst{V}aðit hęfir þú at \alst{v}ígi \hld\ þó’tt \alst{v}ę́ri⸗t skap⸗ligt; &
illt ’s \alst{v}in \alst{v}éla, \hld\ þann’s þér \alst{v}el trúir; &
\edtrans{\alst{b}ęiddr}{urged-on}{\Bfootnote{Attle reminds Guthrun that it was not his choice to marry her.}} fór’k hęiman \hld\ at \alst{b}iðja þïn, Guð·ru̇n.\eva

\bvb\speakernoteb{[Quoth Attle:]}%
“Thou hast waded forth in killings although it was not shapely; \\
it is ill to betray a friend who does trust thee well; \\
urged-on, I journeyed from home to ask for thee, Guthrun.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{L}ęyfð vast-u, ękkja, \hld\ \alst{l}étu stór-ráða, &
\alst{v}arð⸗a \alst{v}ǫ́n lygi \hld\ es \alst{v}ér of ręyndum; &
fórt-u \alst{h}ęim \alst{h}ingat, \hld\ fylgði oss \alst{h}ęrr manna, &
\alst{a}llt vas \alst{í}tar⸗ligt \hld\ umb \alst{ȯ}rar fęrðir.\eva

\bvb Thou wast praised, woman; they said thou hadst great plans; \\
that expectation was no lie as we have found out. \\
Thou journeyedst hither from home; we were followed by a host of men; \\
everything was splendid about our journey.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{M}args vas alls sȯmi, \hld\ \alst{m}anna tíginna, &
\alst{n}aut vǫ́ru ǿrin, \hld\ \alst{n}utum af stórum; &
þar vas \alst{f}jǫlð \alst{f}éar, \hld\ \alst{f}engu til margir.\eva

\bvb There was much honour in everything for the distinguished men; \\
the kine were ample, we enjoyed them greatly; \\
there was a multitude of wealth; many received it.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{M}und galt ek \alst{m}ę́rri, \hld\ \alst{m}ęiðma fjǫlð þiggja, &
\alst{þ}rę́la \alst{þ}rjá-tøgu, \hld\ \alst{þ}ýjar sjau góðar; &
\alst{s}ø̇mð vas at \alst{s}·líku, \hld\ \alst{s}ilfr vas þó męira.\eva

\bvb I paid the bride-price for the famed woman who got a multitude of treasures: \\
thirty thralls, seven good handmaids. \\
There was honour in all such things; there was even more silver.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
Létst þér \alst{a}llt þikkja \hld\ sęm \alst{ę}kki vę́ri &
\edtext{meðan \alst{l}ǫnd þau \alst{l}ǫ́gu \hld\ es mér \alst{l}ęifði Buðli; &
\alst{g}róft svá undir, \hld\ \alst{g}ørði⸗t hlut þiggja}{\lemma{meðan \dots\ þiggja ‘while \dots\ a part.’}\Bfootnote{Difficult.  L.~2 seems to say that Guthrun was a nagging wife and chastised Attle for not working harder (namely for not cultivating all his fields---the very suggestion reveals the poverty of the Greenlandish poet, for whom Attle can be little more than a wealthy farmer).  L.~3 is more difficult.}}. &
\alst{S}vę́ru létst þïna \hld\ \alst{s}itja opt grátna, &
fann’k⸗a’k ï \alst{h}ug \alst{h}ęilum \hld\ \alst{h}jȯna vę́tr síðan.“\eva

\bvb Thou saidst it all seemed to thee as if it were nothing \\
while the lands which Budle left me lay [unused]; \\
thou didst undermine [them]; one did not receive a part. \\
Thou didst let thy mother-in-law sit often weeping; \\
I never found the household in good heart thereafter.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Guð ̇ru̇n kvað:]}%
\llap{„}\alst{L}ýgr þú nú, Atli, \hld\ þó’tt ek þat \alst{l}ítt rękja; &
\alst{h}ęldr vas’k \alst{h}ǿg sjaldan \hld\ \alst{h}óft⸗sk þó stórum; &
\alst{b}ǫrðuð⸗sk ér \alst{b}rǿðr ungir, \hld\ \alst{b}ǫ́ruð⸗sk róg milli. &
\alst{H}alft gekk til \alst{h}ęljar \hld\ ór \alst{h}úsi þïnu; &
\alst{h}roldi \alst{h}ot-vetna \hld\ þat’s til \alst{h}ags skyldi.\eva

\bvb\speakernoteb{[Quoth Guthrun:]}%
“Thou liest now, Attle, though I care little about it; \\
rather, I was seldom docile, yet thou didst elevate thyself haughtily. \\
Ye young brothers fought each other, carried strife between yourselves. \\
Half of thy household went to Hell; \\
everything faltered which should have been a boon.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{Þ}rjú vǫ́rum systkin, \hld\ \alst{þ}ȯttum ȯ·vę́gin, &
\alst{f}órum af landi, \hld\ \alst{f}ylgðum \edtext{Sig·\emph{vę}rði}{\Afootnote{metr. emend.; \emph{Sigurði} \Regius}}; &
\alst{sk}ę́va vér létum, \hld\ \alst{sk}ipi hvęrt vȧrt stýrði, &
\edtrans{\alst{ǫ}rkuðum at \alst{au}ðnu}{we went forth according to luck}{\Bfootnote{Proverbial; the verb \emph{arka} is only found in this expression which appears twice in prose (see \ONP) and nowhere else in poetry.}} \hld\ und’s vér \alst{au}str kvǫ́mum.\eva

\bvb We were three siblings; we seemed undefeatable; \\
we went away from the land; we followed Siward. \\
We let ourselves speed forth; we each steered a ship; \\
we went forth according to luck until we came to the east.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{K}onung drǫ́pum fyrstan, \hld\ \alst{k}urum land þaðra, &
\alst{h}ęrsar oss ȧ \alst{h}ǫnd gengu, \hld\ \alst{h}rę́ðslu þat vissi; &
\edtext{\alst{v}ǫ́gum ór skógi \hld\ þann’s \alst{v}ildum syknan, &
\alst{s}ęttum þann \alst{s}ę́lan \hld\ es \alst{s}ér né átti⸗t.}{\lemma{vǫ́gum \dots\ átti⸗t. ‘We fought \dots\ for himself.’}\Bfootnote{Siward, who was an orphaned outlaw before he met the Yivickings.}}\eva

\bvb We first slew a king, chose the lands there; \\
the chieftains yielded to us; that showed fear. \\
We fought the way out of the forest for him who we would see redeemed; \\
we set down blessedly him who owned naught for himself.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{D}auðr varð \edtrans{inn hu̇nski}{The Hunnish man \ken*{= Siward}}{\Bfootnote{Cf.~\textlink{Sigurdskamma}[4]/4 for this appellation.}}, \hld\ \alst{d}rap þȧ brátt kosti, &
strangt vas \alst{a}ngr \alst{u}ngri \hld\ \alst{ę}kkju nafn hljóta; &
\alst{k}vǫl þȯtti \alst{k}vikri \hld\ at \alst{k}oma ï hús Atla; &
\alst{á}tti \alst{á}ðr kappi, \hld\ \alst{i}llr vas sá missir.\eva

\bvb The Hunnish man \ken*{= Siward} died; then I was soon without choice; \\
it was a cruel sorrow for a young woman to get a widow’s name. \\
It seemed a torture for her in life to come to the house of Attle; \\
she had once had a champion---ill was that loss.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
Komt⸗a þú \alst{þ}ví \alst{þ}ingi \hld\ es vér \alst{þ}at frę́gim &
at \alst{s}ǫk \alst{s}óttir \hld\ né \alst{s}lękkðir aðra; &
\alst{v}ildir ȧ·vallt \alst{v}ę́gja \hld\ ęn \alst{v}ę́tt⸗ki halda, &
kyrrt of því láta.“ \hld\ \edtext{...}{\Afootnote{A blank space is left for this half-line in \Regius.}}\eva

\bvb Thou didst not come from that Thing, as we have learned it, \\
after thou hadst pursued one charge or quashed another. \\
Thou wantedst always to withdraw---but nowise to hold fast--- \\
{[and]} to stay quiet about it.” ...\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Atli kvað:]}%
\llap{„}\alst{L}ýgr þú nú, Guð·ru̇n, \hld\ \alst{l}ítt mun við bǿta⸗sk &
\alst{h}luti \alst{h}vár·igra, \hld\ \alst{h}ǫfum ǫll skarðan; &
\alst{g}ør þú nú, \alst{G}uð·ru̇n, \hld\ af \alst{g}ǿtsku þïnni &
\alst{o}kkr til \alst{ȧ}·gę́tis \hld\ es mik \alst{ú}t hefja.“\eva

\bvb\speakernoteb{[Quoth Attle:]}%
“Thou liest now, Guthrun; little remedy will be found \\
for the lot of either of us; we have all been bereaved! \\
Act now, Guthrun, out of thy goodness, \\
to the honour of us both when they carry me out.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\speakernote{[Guð·ru̇n kvað:]}%
\llap{„}\alst{K}nǫrr mun’k \alst{k}aupa \hld\ ok \alst{k}istu stęinda, &
\alst{v}ęxa \alst{v}el blę́ju \hld\ at \alst{v}ęrja þitt líki, &
\alst{h}yggja ȧ þǫrf \alst{h}vęrja \hld\ sęm vit \alst{h}oll vę́rim.“\eva

\bvb\speakernoteb{[Quoth Guthrun:]}%
“I will buy a cargo-ship and a painted coffin, \\
weave the cover great to protect thy corpse, \\
think of every necessity, like if we had been true to each other.”\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{N}ár varð þȧ Atli, \hld\ \alst{n}iðjum stríð ǿxti, &
\alst{ę}fndi \alst{í}tr-borin \hld\ \alst{a}llt þat’s réð hęita; &
\alst{f}róð vildi Guð·ru̇n \hld\ \alst{f}ara sér at spilla, &
\edtrans{urðu \alst{d}vǫl \alst{d}ǿgra}{there was a delay in her days}{\Bfootnote{Namely when she was carried over the waves to King Enacker as told in \textlink{Gudrunarhvot}[P1].}}, \hld\ \alst{d}ó hǫ̇n ï sinn annat.\eva

\bvb Then Attle became a corpse; grief swelled for his kinsmen; \\
the noble-born woman carried out all what she had vowed. \\
Wise, Guthrun wished to go to destroy her life; \\
there was a delay in her days; she died another time.\evb\evg


\bvg\bva\mssnote{\Regius~TODO}%
\alst{S}ę́ll ’s hvęrr \alst{s}íðan \hld\ es \alst{s}·líkt getr fǿða &
\alst{jó}ð at \alst{a}fr·eki \hld\ sęm’s \alst{ó}l Gjúki; &
\alst{l}ifa mun þat ęptir \hld\ ȧ \alst{l}andi hvęrju, &
\alst{þ}ęira \alst{þ}rǫ́-mę́li \hld\ hvar⸗gi’s \alst{þ}jóð hęyrir.\eva

\bvb Blessed is whoever thenceforth gets to raise \\
such children to heroism like those Yivick begot. \\
It will afterwards live upon every land: \\
their troubled tale, wherever men hear it.\evb\evg

\sectionline
