\bookStart{First Lay of Guthrun}[I Guð·rúnar kviða]
\setBookCode{GudrunOne}

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: C10th (0.988)

\textbf{Meter:} \Fornyrdislag%
\end{flushright}

\section{Introduction}

The \textbf{First Lay of Guthrun} (abbrev. \GudrunOne) directly follows the narrative of \textlink{Brot}, as is seen by the fact that the concluding prose of \textlink{Brot} directly leads into the introduction to \GudrunOne\ in \Regius.  As such the two are clearly intended to form a continuous narrative.

The poem shares several lines with \textlink{GudrunTwo} in sts. 1 and 18 and with \textlink{Sigurdskamma} in sts. 16 and 25.

After Siward’s death Guthrun is so upset that she cannot make herself weep until she sees his body.

\sectionline

\newpage

\section{Text}

\bpg\bpa%
\Aallnote{No break between this passage and the concluding prose of \textlink{Brot} is made in \Regius.}%
Guð·rún sat yfir Sig·urði dauðum. Hón grét eigi sem aðrar konur en hón var búin til at springa af harmi.  Til gengu bę́ði konur ok karlar at hugga hana en þat var eigi auð-velt.  Þat er sǫgn manna at Guð·rún hefði etit af Fáfnis hjarta ok hón skilði því fugls rǫdd. Þetta er enn kveðit um Guð·rúnu:\epa

\bpb Guthrun sat over Siward, dead.  She did not weep like other women, but she was ready to burst apart from grief.  Unto her came both women and men to console her, but that was not easily done.  It is the saying of men that Guthrun had eaten of Fathomer’s heart and she therefore understood the speech of birds.  This is further sung about Guthrun:\epb\epg

\subsection{Guð·rúnar kviða (‘A Lay of Guthrun’)}

\bvg\bva%
\edtext{Ár}{\Afootnote{The \emph{A} is an ornamented initial 3 lines in height in \Regius.}} vas þat’s \alst{G}uð·ru̇n \hld\ \alst{g}ørði⸗sk at dęyja, &
\edtext{es hǫ̇n \alst{s}at \alst{s}org-full \hld\ yfir \alst{S}ig·urði, &
gørði⸗t hǫ̇n \alst{h}júfra \hld\ né \alst{h}ǫndum slá\emph{a} &
né \alst{k}vęina of \hld\ sem \alst{k}onur aðrar}{\lemma{hǫ̇n sat \dots\ aðrar ‘she sat \dots\ women.’}\Bfootnote{Closely related to \textlink{GudrunTwo}[10]/3–5.}}.\eva

\bvb {\huge I}\textsc{t was of yore} that Guthrun made ready to die \\
as she sat sorrowful over Siward. \\
She did not pant nor beat her hands \\
nor wail over him like other women.\evb\evg


\bvg\bva%
Gingu \alst{ja}rlar \hld\ \alst{a}l-snotrir framm, &
þęir’s \alst{h}arðs \alst{h}ugar \hld\ \alst{h}ana lǫttu; &
þęy⸗gi \alst{G}uð·ru̇n \hld\ \alst{g}ráta mátti, &
svá vas hǫ̇n \alst{m}óðug; \hld\ \alst{m}undi hǫ̇n springa.\eva

\bvb There went forth all-clever earls, \\
they who would loosen her hard heart, \\
yet nowise could Guthrun weep; \\
so moody was she—she would burst apart.\evb\evg


\bvg\bva%
Sǫ́tu \alst{í}trar \hld\ \alst{ja}rla brúðir &
\alst{g}olli búnar \hld\ fyr \alst{G}uð·ru̇nu; &
hvęr \alst{s}agði þęira \hld\ \alst{s}ínn of-trega &
þann’s \alst{b}itrastan \hld\ of \alst{b}eðit hafði.\eva

\bvb There sat the splendid brides of the earls \\
adorned with gold before Guthrun. \\
Each one of them told her own great sorrow, \\
the bitterest one she had endured.\evb\evg


\bvg\bva%
Þȧ kvað \alst{G}jaf·laug, \hld\ \alst{G}júka systir: &
„\alst{M}ik vęit’k ȧ \alst{m}oldu \hld\ \alst{m}unar-lausasta; &
hęfi’k \alst{f}imm vera \hld\ \alst{f}or·spell beðit, &
\edtext{tvęggja dǿtra, \hld\ þriggja systra}{\Bfootnote{No alliteration is found in this line and no obvious emendation can be found; it seems that the strong parallelism and assonance between \emph{tvęggja} and \emph{þriggja} compensate for it.}}, &
\alst{á}tta brǿðra, \hld\ þó ek \alst{ęi}n lifi.“\eva

\bvb Then quoth Yeflie, Yivick’s sister: \\
“I know myself on the earth to be most joyless. \\
Of five husbands have I suffered the loss, \\
of two daughters, three sisters, \\
eight brothers—yet I alone live.”\evb\evg


\bvg\bva%
Þęy⸗gi \alst{G}uð·ru̇n \hld\ \alst{g}ráta mátti; &
svá vas hǫ̇n \alst{m}óðug \hld\ at \alst{m}ǫg dauðan &
ok \alst{h}arð-\alst{h}uguð \hld\ of \alst{h}rør fylkis.\eva

\bvb Yet nowise could Guthrun weep; \\
so moody was she after the lad’s death, \\
and hard-hearted over the marshal’s corpse.\evb\evg


\bvg\bva%
Þȧ kvað þat \alst{H}ęr·borg, \hld\ \alst{H}úna-lands dróttning: &
„\alst{H}ęfi’k \alst{h}arðara \hld\ \alst{h}arm at sęgja: &
mínir \alst{s}jau \alst{s}ynir \hld\ \alst{s}unnan lands, &
\alst{v}err inn átti, \hld\ ï \alst{v}al fellu.\eva

\bvb Then quoth this Harburg, Hunland’s queen: \\
“I have a harder harm to tell. \\
My seven sons to the south of their land, \\
—my husband eighth—in battle fell.”\evb\evg


\bvg\bva%
\alst{F}aðir ok móðir, \hld\ \alst{f}jórir brǿðr, &
þau ȧ \alst{v}ági \hld\ \alst{v}indr of lék, &
\alst{b}arði \alst{b}ára \hld\ við \alst{b}orð-þili.\eva

\bvb My father and mother, four brothers— \\
them on the wave the wind outplayed; \\
the breaker beat against the ship-side.\evb\evg


\bvg\bva%
Sjǫlf skylda’k \alst{g}ǫfga, \hld\ sjǫlf skylda’k \alst{g}ǫtva, &
sjǫlf skylda’k \alst{h}ǫndla, \hld\ \edtext{\alst{h}ęl-fǫr}{\Afootnote{emend.; \emph{hęr-fǫr} \Regius}} þęira; &
þat ek \alst{a}llt of bęið \hld\ \alst{ęi}n misseri &
svá’t \alst{m}ér \alst{m}aðr ęngi \hld\ \alst{m}unar lęitaði.\eva

\bvb I alone had to honour them; I alone had to bury them; \\
I alone had to handle their hell-journey \ken{death}. \\
This all I suffered in one half-year, \\
while noone found me any joy.\evb\evg


\bvg\bva%
Þȧ varð’k \alst{h}apta \hld\ ok \alst{h}ęr-numa &
\alst{s}ams misseris \hld\ \alst{s}íðan-verða; &
\alst{sk}ylda’k \alst{sk}ręyta \hld\ ok \alst{sk}úa binda &
\alst{h}ęrsis kván \hld\ \alst{h}vęrjan morgin.\eva

\bvb Then I became a captive and taken in war, \\
in the latter part of that same half-year. \\
I had to dress and bind the shoes \\
of the ruler’s wife every morning.\evb\evg


\bvg\bva%
Hǫ̇n \alst{ǿ}gði mér \hld\ af \alst{a}f·brýði &
ok \alst{h}ǫrðum mik \hld\ \alst{h}ǫggum kęyrði; &
fann’k \alst{h}ús-guma \hld\ \alst{h}vęr-gi inn bętra &
en \alst{h}ús-fręyju \hld\ \alst{h}vęr-gi verri.“\eva

\bvb She tortured me out of jealousy \\
and with hard blows drove me on. \\
A husband nowhere I’ve met better, \\
but a housewife nowhere worse.”\evb\evg


\bvg\bva%
Þęy⸗gi \alst{G}uð·ru̇n \hld\ \alst{g}ráta mátti; &
svá vas hǫ̇n \alst{m}óðug \hld\ at \alst{m}ǫg dauðan &
ok \alst{h}arð-\alst{h}uguð \hld\ of \alst{h}rør fylkis.\eva

\bvb Yet nowise could Guthrun weep; \\
so moody was she after the lad’s death, \\
and hard-hearted over the marshal’s corpse.\evb\evg


\bvg\bva%
Þȧ kvað þat \alst{G}oll·rǫnd, \hld\ \alst{G}júka dóttir: &
„\alst{F}ǫ́ kannt, \alst{f}óstra, \hld\ þó’tt \alst{f}róð séir, &
\alst{u}ngu vífi \hld\ \alst{a}nd-spjǫll bera.“ &
Varaði hǫ̇n at \alst{h}ylja \hld\ of \alst{h}rør fylkis.\eva

\bvb Then quoth this Goldrand, Yivick’s daughter: \\
“Little canst thou, foster-mother—though thou be wise— \\
to a young wife give answers.”— \\
She bade them uncover the marshal’s corpse.\evb\evg


\bvg\bva%
\alst{S}vipti hǫ̇n blę́ju \hld\ af \alst{S}ig·urði &
ok \alst{v}att \alst{v}ęngi \hld\ fyr \alst{v}ífs \edtext{kn\emph{éu}m}{\Afootnote{metr. emend. by restoration of old hiatus form; \emph{knjám} \Regius}}: &
„\alst{L}ít-tu ȧ \alst{l}júfan, \hld\ \alst{l}ęgg þú munn við grǫn &
sem þú \alst{h}alsaðir \hld\ \alst{h}ęilan stilli.“\eva

\bvb She drew the shroud off of Siward \\
and turned his cheeks before the wife’s knees: \\
“Look on thy beloved! Lay thy mouth to his lips \\
like thou didst embrace the hale prince.”\evb\evg


\bvg\bva%
\alst{Ȧ} lęit Guð·ru̇n \hld\ \alst{ęi}nu sinni; &
sá hǫ̇n \alst{d}ǫglings skǫr \hld\ \alst{d}ręyra runna, &
\alst{f}rȧnar sjȯnir \hld\ \alst{f}ylkis liðnar, &
\alst{h}ug-borg jǫfurs \hld\ \alst{h}jǫrvi skorna.\eva

\bvb On him looked Guthrun a single time; \\
she saw the noble’s locks run with blood, \\
the gleaming gaze of the marshal gone, \\
the heart-fort \ken{chest} of the ruler cut by sword.\evb\evg


\bvg\bva%
Þȧ \alst{h}né Guð·ru̇n \hld\ \alst{h}ǫll við bólstri; &
\alst{h}addr losnaði, \hld\ \alst{h}lýr roðnaði &
en \alst{r}egns dropi \hld\ \alst{r}ann niðr umb kné.\eva

\bvb Then Guthrun sank down, slooped against the bolster; \\
her hair loosened, her cheek reddened, \\
and a rain-drop ran down to her knee.\evb\evg


\bvg\bva%
Þȧ \alst{g}rét \alst{G}uð·ru̇n, \hld\ \alst{G}júka dóttir, &
svá’t \alst{t}ǫ́r flugu \hld\ \edtrans{\alst{t}resk}{veil(?)}{\Bfootnote{A guess translation; this word is an unexplained \emph{hapax}.}} ï gǫgnum &
\edtrans{ok \alst{g}ullu við \hld\ \alst{g}ę̇ss ï tu̇ni}{and in response shrieked the geese in the yard}{\Bfootnote{Shared identically with \textlink{Sigurdskamma}[29]/4.}}, &
\alst{m}ę́rir fuglar \hld\ es \alst{m}ę́r átti.\eva

\bvb Then wept Guthrun, Yivick’s daughter, \\
so that the tears flew through her veil(?) \\
and in response shrieked the geese in the yard, \\
the famèd fowls which the maiden owned.\evb\evg


\bvg\bva%
Þȧ kvað þat \alst{G}oll·rǫnd, \hld\ \alst{G}júka dóttir: &
„\alst{y}kkar vissa’k \hld\ \alst{ȧ}stir męstar &
\alst{m}anna allra \hld\ fyr \alst{m}old ofan; &
\alst{u}nðir þú hvár-ki \hld\ \alst{ú}ti né inni, &
\alst{s}ystir mïn, \hld\ nema hjá \alst{S}ig·urði.“\eva

\bvb Then quoth this Goldrand, Yivick’s daughter: \\
“I knew the love between you two was the greatest \\
of all men above the earth. \\
Thou wast never content outside or inside, \\
O sister of mine, save by Siward’s side.”\evb\evg


\bvg\bva%
„\edtext{\alst{S}vá vas mïnn \alst{S}ig·urðr \hld\ hjá \alst{s}onum Gjúka &
sęm vę́ri \edtrans{\alst{g}ęir-laukr}{garlic}{\Bfootnote{Or ‘spear-leek’. I have opted for this translation based on etymology (cf. OE \emph{gâr-léac} ‘spear-leek’), but the botanical identity is unclear. \textlink{GudrunTwo} 2 has \emph{grǿnn laukr} ‘green leek’ instead. For the cultural importance of leeks and onions see note to \textlink{Voluspa} 4.}} \hld\ ór \alst{g}rasi vaxinn,}{\lemma{Svá vas \dots\ vaxinn ‘So was \dots\ grown’}\Bfootnote{These two lines are almost identical to \textlink{GudrunTwo} 2/1–2. Since the present poem is probably older \parencite{Sapp2022}, it is likely the source.}} &
\edtext{eða vę́ri \alst{b}jartr stęinn \hld\ ȧ \alst{b}and dręginn: &
\alst{ja}rkna-stęinn \hld\ yfir \alst{ǫ}ðlingum.}{\lemma{eða vę́ri \dots\ ǫðlingum. ‘or were \dots\ athlings.’}\Bfootnote{Beaded necklaces were commonly worn by Scandinavian women of the time, and the beads were mostly of opaque coloured glass.  Siward is likened to a bright crystal, the sons of Yivick to dull glass.}}\eva

\bvb\speakernoteb{[Guthrun quoth:]}%
“So was my Siward beside the sons of Yivick \\
like were a garlic out of grass grown, \\
or were a bright stone on a string drawn: \\
an \inx[C]{arkenstone} over the athlings.\evb\evg


\bvg\bva%
Ek \alst{þ}ȯtta auk \hld\ \alst{þ}jóðans rekkum &
\alst{h}vęrri \alst{h}ę́rri \hld\ \alst{H}ęrjans dísi; &
nú em’k svá \alst{l}ítil \hld\ sem \alst{l}auf séa &
\alst{o}pt ï \alst{jǫ}lstrum \hld\ at \alst{jǫ}fur dauðan.\eva

\bvb I also seemed to the ruler’s champions \\
higher than any of the Lord of Hosts’ dises \ken{walkirries}. \\
Now am I so small as if a leaf I were, \\
high in the willows, after the ruler’s death.\evb\evg


\bvg\bva%
\alst{S}akna’k ï \alst{s}essi \hld\ ok ï \alst{s}ę́ingu &
\alst{m}ïns \alst{m}ál-vinar— \hld\ valda \alst{m}ęgir Gjúka; &
valda \alst{m}ęgir Gjúka \hld\ \alst{m}ïnu bǫlvi &
ok \alst{s}ystr \alst{s}innar \hld\ \alst{s}ǫ́rum gráti.\eva

\bvb I miss in the seat and in the bed \\
my confidant—at fault are the lads of Yivick; \\
the lads of Yivick are at fault for my bale \\
and for their sister’s [my] bitter weeping.\evb\evg


\bvg\bva%
Svá ér umb \alst{l}ýða \hld\ \alst{l}andi ęyðið &
sęm ér of \alst{u}nnuð \hld\ \alst{ęi}ða svarða; &
man⸗a þú, \alst{G}unn·arr, \hld\ \alst{g}olls of njóta; &
\edtrans{þęir munu þér \alst{b}augar \hld\ at \alst{b}ana verða}{those bighs will for thee become the bane}{\Bfootnote{I.e. “the wealh will be the end of you”.  Formulaic; cf. \textlink{Fafnismal} 9, 20.}} &
es þú \alst{S}ig·urði \hld\ \alst{s}varðir ęiða.\eva

\bvb Even so may ye ruin the land for folk, \\
as ye fulfilled the sworn oaths! \\
Thou wilt not, Guther, enjoy the gold; \\
those bighs will for thee become the bane \\
on which thou to Siward didst swear the oaths.\evb\evg


\bvg\bva%
Opt vas ï \alst{t}u̇ni \hld\ \alst{t}ęiti męiri &
þȧ’s mïnn \alst{S}ig·urðr \hld\ \alst{s}ǫðladi Grana, &
ok þęir \alst{B}ryn·hildar \hld\ \alst{b}iðja fóru, &
\alst{a}rmrar vę́ttar \hld\ \alst{i}llu hęilli.“\eva

\bvb Oft in the courtyard there was greater cheer \\
when my Siward saddled Grane \\
and they journeyed to ask for Byrnhild’s hand, \\
that wretched wight of ill omen.”\evb\evg


\bvg\bva%
Þȧ kvað þat \alst{B}ryn·hildr \hld\ \alst{B}uðla dóttir: &
„\alst{V}ǫn sé sú \alst{v}ę́ttr \hld\ \alst{v}ers ok barna &
es þik \alst{G}uð·ru̇n \hld\ \alst{g}ráts of bęiddi &
ok þér ï \alst{m}orgun \hld\ \alst{m}ál-ru̇nar gaf.“\eva

\bvb Then quoth this Byrnhild, Budle’s daughter: \\
“TODO.”\evb\evg


\bvg\bva%
Þȧ kvað þat \alst{G}oll·rǫnd \hld\ \alst{G}júka dóttir: &
„\edtrans{\alst{Þ}ęgi þú, \alst{þ}jóð-lęið, \hld\ \alst{þ}ęira orða}{Shut up, thou most loathsome woman, with those words}{\Bfootnote{Same line is found with \emph{Þȯrr} ‘Thunder’ for \emph{þjóð-lęið} ‘most loathsome woman’ in \textlink{Thrymskvida}[18]/2.}}! &
\alst{U}rðr \alst{ǫ}ðlinga \hld\ hęfir þú \alst{ę́} vesit; &
rekr þik \alst{a}lda hvęr \hld\ \alst{i}llrar skępnu &
\alst{s}org \alst{s}ára \hld\ \alst{s}jau konunga &
ok \alst{v}in-spell \hld\ \alst{v}ífa męst.“\eva

\bvb Then quoth this Goldrand, Yivick’s daughter:
“Shut up, thou most loathsome woman, with those words. \\
The Weird of athlings hast thou always been. \\
Thou art driven by every wave of evil fate, \\
{[art]} the sore sorrow of seven kings, \\
and the greatest spoiler of women’s friendship.”\evb\evg


\bvg\bva%
Þȧ kvað þat \alst{B}ryn·hildr \hld\ \alst{B}uðla dóttir: &
„\edtrans{Vęldr \alst{ęi}nn \alst{A}tli \hld\ \alst{ǫ}llu bǫlvi}{Alone is Attle at fault for all evil}{\Bfootnote{Related to \textlink{HelgakvidaTwo}[34]/3 and \textlink{Sigurdskamma}[27]/4.}} &
\edtrans{of \alst{b}orinn \alst{B}uðla \hld\ \alst{b}róðir mïnn}{born of Budle, my own brother}{\Bfootnote{Identically shared with \textlink{Sigurdskamma}[56]/5.}} &
þȧ’s vit ï \alst{h}ǫll \hld\ \alst{h}u̇nskrar þjóðar &
\alst{ę}ld ȧ \alst{jǫ}fri \hld\ \alst{o}rm-bęðs litum; &
þęss hęfi’k \alst{g}angs \hld\ \alst{g}oldit síðan, &
þęirar \alst{s}ýnar; \hld\ \alst{s}ǫ́umk ęy.“\eva

\bvb Then quoth this Byrnhild, Budle’s daughter: \\
“Alone is Attle at fault for all evil, \\
born of Budle, my own brother, \\
since we two in the hall of the Hunnish folk \\
beheld the fire of the snake-bed \ken{gold} on the prince. \\
For that journey I have paid ever since, \\
for that sight—I always worried over it.”\evb\evg


\bvg\bva%
\alst{St}óð hǫ̇n und \alst{st}oð \hld\ \alst{st}ręngði hǫ̇n ęlfi, &
\alst{b}rann \alst{B}ryn·hildi \hld\ \alst{B}uðla dóttur &
\alst{ę}ldr ór \alst{au}gum; \hld\ \alst{ęi}tri fnę́sti &
es hǫ̇n \alst{s}ǫ́r of lęit \hld\ ȧ \alst{S}ig·urði.\eva

\bvb She stood beneath a pillar, fueled her anger; \\
in Byrnhild Budle’s daughter burned \\
fire from her eyes; she spit venom \\
when she beheld the wounds on Siward.\evb\evg


\bpg\bpa%
Guð·rún gekk þaðan á braut til skógar á eyði-merkr ok fór allt til Danmarkar ok var þar með Þóru, Hákonar dóttur, sjau misseri.  Bryn·hildr vildi eigi lifa eptir Sig·urð.  Hón lét drepa þrę́la sína átta ok fimm ambóttir, þá lagði hón sik sverði til bana svá sem segir í Sig·urðar kviðu inni skǫmmu.\epa

\bpb Guthrun went away thence through the woods of the wastelands and journeyed all the way to Denmark and stayed there with Thure, Hathkin’s daughter, for seven half-years.  Byrnhild did not want to live after Siward.  She had her eight thralls and five handmaids slain; then she ran herself through with a sword unto her death, as it says in the Short Lay of Siward.\epb\epg

\sectionline
