\bookStart{Speeches of Hamthew}[Ham·ðis mǫ́l]
\setBookCode{Hamdismal}

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: C10th (0.885)

\textbf{Meter:} \Fornyrdislag, \Malahattr% TODO
\end{flushright}%

\section{Introduction}

The \textbf{Speeches of Hamthew} (signum \textlink{Hamdismal}) is the final poem in \Regius.  It is obviously closely related to the immediately preceding \textlink{Gudrunarhvot}, with which it shares many lines (3/2–4, 5/1–2, 6/1–2) in the first part.  Although the two poems deal with the same events, they differ in perspective.  In both, Guthrun incites her two young sons to avenge their sister Swanhild by killing king Ermenric, and they set out to do so.  It is here that they diverge, for \textlink{Gudrunarhvot} stays behind with Guthrun after the departure of her sons, whereas \textlink{Hamdismal} instead follows the boys as they leave.

In Norse sources the slaying of Ermenric is also found in \Ragnarsdrapa\ and \VolsungaSaga\ 42 (the latter being based on \textlink{Hamdismal}, as seen by its citation of stanza TODO).

TODO: The narrative is founded in historical fact; in Jordanes’ \emph{Getica} 129 we read the following: TODO.

The trampling of Swanhild by horses and the attack launched by her brothers Sarle and Hamthew upon Ermenric is thus historical, as is the detail that he was maimed but not fatally injured.  On this point the archaic \Ragnarsdrapa\ is clearly further advanced from the fact of history than \textlink{Hamdismal}, since it (\Ragnarsdrapa\ 4/3–4) seems to describe Ermenric as drowning in a well.

TODO: Finnur argues for two poems (one in \Malahattr, one in \Fornyrdislag).

\sectionline

\newpage

\section{Text}

\subsection{Ham·ðis mǫ́l (‘The Speeches of Hamthew’)}

\bvg\bva%
\edtext{Spruttu}{\Afootnote{The \emph{S} is an ornamented initial 2 lines high in \Regius.}} ȧ \alst{t}ȧi \hld\ \alst{t}regnar iðir &
\alst{g}rǿti alfa \hld\ in \alst{g}lý-stǫmu; &
ár of \alst{m}orgin \hld\ \alst{m}anna bǫlva &
\alst{s}útir hvęrjar \hld\ \alst{s}org of kvęykva.\eva

\bvb TODO. \\
In early morning each kind of pain \\
of the bales of men awakens sorrow.\evb\evg


\bvg\bva%
Vas⸗a þat \alst{n}ú \hld\ \alst{n}é ï gę́r, &
þat hęfir \alst{l}angt \hld\ \alst{l}iðit síðan; &
es \alst{f}átt \alst{f}ornara \hld\ \alst{f}ręmr vas þat hǫlfu &
\edtrans{es hvatti \alst{G}uð·ru̇n}{when Guthrun \dots\ goaded}{\Bfootnote{As also told in \textlink{Gudrunarhvot}.}} \hld\ \alst{G}júka borin &
\alst{s}onu \alst{s}ïna unga \hld\ at hęfna \alst{S}van·hildar.\eva

\bvb It was not now nor yesterday; \\
a long time has passed since. \\
Little is older, [yet] it was twice as long ago \\
when Guthrun, born of Yivick, goaded \\
her young sons to avenge Swanhild.\evb\evg


\bvg\bva%
„\alst{S}ystir vas ykkur \hld\ \alst{S}van·hildr of hęitin &
\edtext{sú’s \alst{Jǫ}rmun·rekr \hld\ \alst{jó}m of traddi &
\alst{h}vítum ok svǫrtum \hld\ ȧ \alst{h}ęr-vegi &
\alst{g}rǫ́m \alst{g}ang-tǫmum \hld\ \alst{g}otna hrossum.}{\lemma{sú’s \dots\ hrossum. ‘she whom \dots\ horses!’}\Bfootnote{Repeated almost identically in \textlink{Gudrunarhvot} 2/3–6.}}\eva

\bvb „The sister of you two was named Swanhild, \\
she whom Ermenric had trampled by steeds, \\
by white ones and black on the path of war, \\
by grey, pacing, Gotnish horses!\evb\evg


\bvg\bva%
Ęptir ’s ykkr \alst{þ}rungit \hld\ \alst{þ}jóð-konunga, &
lifið \alst{ęi}nir \alst{é}r þátta \hld\ \alst{ę́}ttar minnar.\eva

\bvb Ye two have been forced back by great kings; \\
ye are the lone living strands of my lineage.\evb\evg


\bvg\bva%
\alst{Ęi}n-stǿð em’k \alst{o}rðin \hld\ sęm \alst{ǫ}sp ï holti, &
\alst{f}allin at \alst{f}rę́ndum \hld\ sęm \alst{f}ura at kvisti, &
\alst{v}aðin at \alst{v}ilja \hld\ sęm \alst{v}iðr at laufi, &
þȧ’s in \alst{k}vist-skǿða \hld\ \alst{k}ømr umb dag varman.”\eva

\bvb I am become supportless as the asp in the grove, \\
as bereaved of kinsmen as the pine from its branch, \\
as deprived of willpower as the tree from its leaf \\
when the branch-scather comes on a warm day.”\evb\evg

\bvg\bva%
\alst{H}itt kvað þȧ \alst{H}am·ðir \hld\ inn \alst{h}ugum stóri: &
„\alst{L}ítt myndir þȧ, Guð·ru̇n, \hld\ \alst{l}ęyfa dǫ́ð Hǫgna &
es þęir \alst{S}ig·urð \hld\ \alst{s}vefni ór vǫkðu; &
sats-tu ȧ \alst{b}ęð \hld\ en \alst{b}anar hlógu.\eva

\bvb This then quoth Hamthew the great of heart: \\
“Thou hadst little cause, then, Guthrun, to praise Hain’s deeds \\
when they Siward from his sleep awoke; \\
thou wast in the bed, but his killers laughed.\evb\evg


\bvg\bva%
\alst{B}ǿkr vǫ́ru þïnar \hld\ inar \alst{b}lá-hvítu &
ofnar \alst{v}ǫlundum, \hld\ flutu ï \alst{v}ers dręyra; &
\alst{s}valt þá \alst{S}ig·urðr, \hld\ \alst{s}atst yfir dauðum, &
\alst{g}lýja né \alst{g}ȧðir; \hld\ \alst{G}unn·arr þér svá vildi.\eva

\bvb Thy blue-white bed-clothes artfully woven \\
floated in thy husband’s blood. \\
Then Siward died; thou satst above him, dead, \\
sawst no sake for joy.  Guther would see thee so.\evb\evg


\bvg\bva%
\alst{A}tla þȯtti⸗sk þú stríða \hld\ at \alst{E}rps morði &
ok at \alst{Ęi}tils \alst{a}ldr-lagi, \hld\ þat vas þér \alst{ę}nn verra! &
Svá skyldi hvęrr \alst{ǫ}ðrum \hld\ vęrja til \alst{a}ldr-laga &
\alst{s}verði \alst{s}ár-bęitu \hld\ at \alst{s}ér né stríddi⸗t.“\eva

\bvb TODO\evb\evg


\bvg\bva%
Hitt kvað þȧ \alst{S}ǫrli, \hld\ \alst{s}vinna hafði hann hyggju: &
„Vil’k⸗at við \alst{m}óður \hld\ \alst{m}ǫ́lum skipta; &
\alst{o}rðs þikkir \alst{ę}nn vant \hld\ \alst{y}kkru hvǫ́ru; &
hvęrs biðr nú, \alst{G}uð·ru̇n, \hld\ es at \alst{g}ráti né fǿr⸗at?\eva

\bvb This then quoth Sarle; he had a wise mind: \\
“I do not with my mother wish to bandy words. \\
TODO.\evb\evg


\bvg\bva%
\alst{B}rǿðr grát þú þïna \hld\ ok \alst{b}uri svása, &
\alst{n}iðja \alst{n}á-borna \hld\ lęidda \alst{n}ę́r rógi; &
okkr skalt ok, \alst{G}uð·ru̇n, \hld\ \alst{g}ráta báða &
es hér sitjum \alst{f}ęigir á mǫrum, \hld\ \alst{f}jarri munum dęyja.“\eva

\bvb TODO\evb\evg


\bvg\bva%
\alst{G}engu ór \alst{g}arði \hld\ \alst{g}ǫrvir at ęiskra; &
liðu þá \alst{y}fir \alst{u}ngir \hld\ \alst{ú}rig fjǫll &
\alst{m}ǫrum hún-lęndskum, \hld\ \alst{m}orðs at hefna.\eva

\bvb They went out of the settlement ready to wage fury. \\
Then passed the young ones over drizzling mountains \\
on Hunlandish horses to avenge the murder.\evb\evg


\bvg\bva%
Þȧ kvað þat \alst{E}rpr \hld\ \alst{ęi}nu sinni, &
\alst{m}ę́rr of lék \hld\ ȧ \alst{m}ars baki: &
„Illt ’s \alst{b}lauðum hal \hld\ \alst{b}rautir kenna;“ &
kóðu \alst{h}arðan mjǫk \hld\ \alst{h}ornung vesa.\eva

\bvb TODO\evb\evg


\bvg\bva%
Fundu ȧ \edtrans{\alst{st}rę́ti}{street}{\Bfootnote{A Roman loanword used specifically for a paved road or street in a town.  It probably emphasises the southern geography in which the events transpire.}} \hld\ \alst{st}ór-brǫgð-óttan: &
„Hvé mun \alst{ja}rp-skammr \hld\ \alst{o}kkr full-tingja?“\eva

\bvb They found on the street the very cunning man: \\
“How should this brown shorty assist us?”\evb\evg


\bvg\bva%
\alst{S}varaði inn \alst{s}undr-mǿðri, \hld\ \alst{s}vá kvaðsk vęita myndu &
\alst{f}ull-ting \alst{f}rę́ndum \hld\ sem \alst{f}ótr ǫðrum. &
„Hvat męgi \alst{f}ótr \hld\ \alst{f}ǿti vęita &
né \alst{h}old-gróin \hld\ \alst{h}ǫnd annarri?“\eva

\bvb He of sundry mother answered, said that he would give \\
assistance to his kinsmen like a foot to another.— \\
“What can a foot give to a foot, \\
or a flesh-grown hand to another?”\evb\evg


\bvg\bva%
Drógu ór \alst{sk}íði \hld\ \alst{sk}íði-éarn, &
\alst{m}ę́kis ęggjar \hld\ \edtrans{at \alst{m}un flagði}{at the behest of the ogress}{\Bfootnote{They were acting according to the will of a cruel norn (\emph{flagð}); cf. the next-to-last st./4b (TODO), \textlink{Reginsmal}[2], \textlink{Sigurdskamma}[7].}}; &
\alst{þ}vęrrðu \alst{þ}rótt sïnn \hld\ at \alst{þ}riðjungi, &
létu \alst{m}ǫg ungan \hld\ til \alst{m}oldar hníga.\eva

\bvb They drew from their sheaths their sheath-irons \ken{swords}, \\
the edges of the blade at the behest of the ogress. \\
They used their strength on the third one; \\
they let the young lad sink down to the earth.\evb\evg


\bvg\bva%
\alst{Sk}óku loða, \hld\ \alst{sk}almir fęstu &
ok \alst{g}oð-bornir \hld\ smugu ï \alst{g}uð-vęfi.\eva

\bvb TODO.\evb\evg


\bvg\bva%
\alst{F}ramm lǫ́gu brautir, \hld\ \alst{f}undu vǫ́-stigu &
ok \alst{s}ystur \alst{s}on \hld\ \alst{s}áran ȧ męiði, &
\alst{v}arg-tré \alst{v}ind-kǫld \hld\ \alst{v}estan bǿjar; &
\alst{t}rýtti ę́ \alst{t}rǫnu hvǫt, \hld\ \alst{t}ítt vas⸗at biðja.\eva

\bvb Forwards lay their roads; they found paths of woe \\
and their sister’s son wounded on a branch, \\
a wind-cold hanging tree to the west of the town; \\
...; they had no wish to tarry.\evb\evg


\bvg\bva%
Glaumr vas ï \alst{h}ǫllu, \hld\ \alst{h}alir ǫl-ręifir, &
ok til \alst{g}ota ękki \hld\ \alst{g}ørðu-t hęyra &
áðr \alst{h}alr \alst{h}ug-fullr \hld\ ï \alst{h}orn of þaut.\eva

\bvb There was revel in the hall, the men ale-rowdy, \\
and the sound of the horses they hardly heard, \\
before a man full of heart blew on a horn.\evb\evg


\bvg\bva%
Sęgja fóru \emph{\alst{ę́}rir} \hld\ \alst{Jǫ}rmun·rekki &
at \alst{s}énir vǫ́ru \hld\ \alst{s}ęggir und hjǫlmum: &
„\alst{R}ǿðið ér umb \alst{r}ǫ́ð! \hld\ \alst{R}íkir eru komnir, &
fyr \alst{m}ǫ́ttkum hafið ér \alst{m}ǫnnum \hld\ \alst{m}ęy of tradda.“\eva

\bvb Messengers went to tell Ermenric \\
that the youths were seen in their helmets: \\
“Tell your decision!  Strong men have come. \\
To the wrath of mighty men have ye trampled that maiden!”\evb\evg


\bvg\bva%
\alst{H}ló þá Jǫrmun·rekkr, \hld\ \alst{h}ęndi drap ȧ kampa, &
\alst{b}ęiddi⸗sk at \alst{b}rǫngu, \hld\ \alst{b}ǫðvaði⸗sk at víni; &
\alst{sk}ók hann \alst{sk}ǫr jarpa, \hld\ sá ȧ \alst{sk}jǫld hvítan, &
lét hann sér ï \alst{h}ęndi \hld\ \alst{h}varfa kęr gollit.\eva

\bvb Then Ermenric laughed, put his hand to his beard, \\
..., became ferocious on his wine. \\
He shook his brown locks, looked on his white shield; \\
he let spin in his hand a golden cup.\evb\evg


\bvg\bva%
„\alst{S}ę́ll ek þȧ þø̇tt⸗umk \hld\ ef \alst{s}éa knę́tta &
\alst{H}am·ði ok Sǫrla \hld\ ï \alst{h}ǫllu mïnni; &
\alst{b}uri mynda’k þȧ \alst{b}inda \hld\ með \alst{b}oga stręngjum, &
\alst{g}óð bǫrn \alst{G}júka \hld\ fęsta ȧ \alst{g}alga.“\eva

\bvb “I would have thought me blessed if I got to see \\
Hamthew and Sarle in my hall; \\
those boys I would then bind with bow-strings, \\
the good children of Yivick, and tie them to the gallows.”\evb\evg


\bvg\bva%
\alst{H}itt kvað þȧ \alst{H}róðr·glǫð, \hld\ stóð of \alst{h}léðum, &
\alst{m}ę́-fingr \alst{m}ę́lti \hld\ við \alst{m}ǫg þęnna: &
„Því’t þat \alst{h}ęita \hld\ at \alst{h}lýði-gi myni; &
mega \alst{t}vęir menn ęinir \hld\ \alst{t}íu hundruð Gotna &
\alst{b}inda eða \alst{b}ęrja \hld\ ï \alst{b}org inni hǫ́u?\eva

\bvb TODO. \\
“Since they swear that no army will, \\
how can two men alone ten hundred Gots \\
bind or beat in the high stronghold?”\evb\evg


\bvg\bva%
\edtrans{\alst{St}yrr varð ï ranni}{There was uproar in the house}{\Bfootnote{Strongly reminiscent of \Ragnarsdrapa\ 3/5: \emph{Rósta varð ï ranni} ‘There was tumult in the hall’, \Beowulf\ 1302a: \emph{Hréam wearð ǫn Heorote} ‘There was an outcry in the Hart’, both in the context of hall-fights, the former describing the same events as \Hamdismal.}}, \hld\ \alst{st}ukku ǫl-skálir, &
ï \alst{b}lóði \alst{b}ragnar lǫ́gu, \hld\ komit ór \alst{b}rjósti Gotna.\eva

\bvb There was uproar in the hall, the ale-bowls shattered; \\
champions lay in blood flown from the breasts of the Gots.\evb\evg


\bvg\bva%
\alst{H}itt kvað þȧ \alst{H}am·ðir \hld\ inn \alst{h}ugum stóri: &
„\alst{Ǿ}stir \alst{Jǫ}rmunrekkr \hld\ \alst{o}kkarrar kvǫ́mu &
\alst{b}rǿðra sam-mǿðra \hld\ innan \alst{b}orgar þïnnar; &
\edtext{fǿtr \alst{s}ér þú þïna \hld\ hǫndum \alst{s}ér þú þïnum  &
\alst{Jǫ}rmunrekkr \alst{o}rpit \hld\ ï \alst{ę}ld hęitan!}{\lemma{fǿtr \dots\ hęitan! ‘Thou seest \dots\ fire!’}\Bfootnote{The detail of Ermenric’s severed hands and feet is also found in \Ragnarsdrapa\ 4.}}“\eva

\bvb This then quoth Hamthew the great of heart: \\
“Thou, Ermenric, didst incite our coming, \\
we brothers of the same mother, into thy stronghold! \\
Thou seest thy feet; thou seest them with thy hands, \\
O Ermenric, thrown in the hot fire!”\evb\evg


\bvg\bva%
Þȧ \edtext{\alst{r}aut}{\Afootnote{metr. emend.; \emph{hraut} \Regius}} við \hld\ inn \edtrans{\alst{r}ęgin-kunngi}{rein-descended}{\Bfootnote{Descended from the Reins, i.e. the Gods.  The godly ancestry of the Gotnish Amals, the house to which Ermenric belonged, is described in Jordanes \emph{Getica} 78, 79 (excerpt): \emph{Magnaque potiti per loca victoria iam proceres suos, quorum quasi fortuna vincebant, non puros homines, sed semideos id est Ansis vocaverunt.  Quorum genealogia ut paucis percurram vel quis quo parente genitus est aut unde origo coepta, ubi finem effecit, [...]  79 Horum ergo heroum, ut ipsi suis in fabulis referunt, primus fuit Gapt, qui genuit Hulmul. Hulmul vero genuit Augis, et Augis genuit} etc. etc. ‘They thereafter called their leaders, by whose good fortune they seemed to have conquered, not mere men, but demigods, that is Ansis.  Their genealogy I shall run through briefly, telling the lineage of each and the beginning and the end of this line. [...] Now the first of these heroes, as they themselves relate in their legends, was Gapt, who begat Hulmul.  And Hulmul begat Augis; and Augis begat’ et c. et c.

Jordanes’ \emph{Ansis} is a cognate to ON \emph{ę̇sir} ‘the Eese’, i.e. the Gods, which are called demigods in a clear instance of Christian euhemerism.  The first of the \emph{Ansis} is \emph{Gapt}, i.e. Geat, whose name directly corresponds to OE \emph{Géat}, the progenitor of several Anglo-Saxon royal lineages, and ON \emph{Gautr}, a poetic synonym for the god \inx[P]{Weden} (\textlink{Grimnismal}[56]/4, \textlink{Baldrsdraumar}[2]/1).}} &
\alst{B}aldr ï \alst{b}rynju \hld\ sem \alst{B}jǫrn hryti: &
„\alst{G}rýtið ér ȧ \alst{g}umna \hld\ alls \alst{g}ęirar né bíta &
\alst{ę}ggjar né \alst{éǫ}rn \hld\ \alst{Jó}nakrs sonu!“\eva

\bvb Then roared back the rein-descended \\
Balder in his byrnie \ken*{\textsc{warrior} > = Ermenric} like a roaring bear: \\
“Stone ye the men! For spears will not bite, \\
nor edges nor irons, Enacker’s sons!”\evb\evg


\bvg\bva%
\alst{H}itt kvað þȧ \alst{H}am·ðir \hld\ inn \alst{h}ugum stóri: &
„\alst{B}ǫl vannt, \alst{b}róðir, \hld\ es þann \alst{b}ęlg lęystir; &
opt ór þęim \alst{b}ęlg \hld\ \alst{b}ǫll rǫ́ð koma.“\eva

\bvb TODO\evb\evg


\bvg\bva%
„\alst{H}ug hęfðir, \alst{H}am·ðir, \hld\ ef hęfðir \alst{h}yggjandi, &
\alst{m}ikils es ȧ \alst{m}ann hvęrn vant \hld\ es \alst{m}an-vits es.“\eva

\bvb TODO\evb\evg


\bvg\bva%
\edtext{„\alst{A}f vę́ri nú \edtext{hǫfuð}{\Afootnote{\emph{hǫfuð’it} \VolsungaMS}} \hld\ ef \alst{E}rpr lifði, &
\alst{b}róðir okkarr inn \alst{b}ǫð-frǿkni, \hld\ es vit ȧ \alst{b}raut vǫ́gum}{\lemma{Af \dots\ vǫ́gum}\Afootnote{Also in \VolsungaMS.}\Bfootnote{These lines are quoted in \VolsungaSaga\ 42, preceded by the text: \emph{Þá mę́lti Ham·ðir: „Af mundi nú hǫfuð’it, ef Erpr lifði, bróðir okkarr, er vit vágum á leið’inni, ok sám vit þat of síð,“ sem kveðit er.} ‘Then Hamthew spoke: “The head would now have been off if Earp had lived, our brother whom we slew on the road, and we worried about that for some time”, as it is sung.’}}, &
\alst{v}err hinn \alst{v}íð-frǿkni, \hld\ hvǫtt⸗umk at dísir, &
\alst{g}umi hinn \alst{g}unn-hęlgi, \hld\ \alst{g}ǫrðum⸗sk at vígi.\eva

\bvb TODO\evb\evg


\bvg\bva%
\alst{Ę}kki hygg \alst{o}kkr vesa \hld\ \alst{u}lfa dǿmi, &
\ind at vit mynim \alst{s}jalfir of \alst{s}aka⸗sk &
sęm \alst{g}ręy norna \hld\ þau’s \alst{g}rǫ́ðug eru &
\ind ï \alst{au}ðn of \alst{a}lin.\eva

\bvb TODO\evb\evg


\bvg\bva%
\alst{V}el hǫfum vit \alst{v}egit, \hld\ stǫndum ȧ \alst{v}al Gotna &
\alst{o}fan \alst{ę}gg-móðum \hld\ sem \alst{ę}rnir á kvisti; &
\alst{g}óðs hǫfum tírar fęngit \hld\ þó’tt skylim nú eða ï \alst{g}ę́r dęyja, &
\alst{k}vęld lifir maðr ękki \hld\ ęptir \alst{k}við norna.“\eva

\bvb “Well have we two fought, we stand on the slain of the Gots: \\
over the edge-wearied like eagles on a branch. \\
We’ve won great glory, even if we should die now or tomorrow— \\
man lives not one evening after the verdict of the norns!”\evb\evg


\bvg\bva%
Þar fell \alst{S}ǫrli \hld\ at \alst{s}alar gafli, &
en \alst{H}am·ðir \alst{h}né \hld\ at \alst{h}ús-baki.\eva

\bvb There fell Sarle by the hall’s gables, \\
and Hamthew sank down by the back of the house.\evb\evg


\bpg\bpa%
Þetta eru kǫlluð Ham·ðis mǫ́l in \edtext{fornu}{\Afootnote{add. \emph{ꝺxplꝺcꝺꞇ lꝺbꝺr ꝺﬅꝺ} (cipher for \emph{*explicit liber isti}) ‘Here this book ends’ \Regius}}.\epa

\bpb These are called the Ancient Speeches of Hamthew.\epb\epg

\sectionline
