\bookStart{Fragment of a Lay of Siward}[Brot af Sig·urðar kviða]
\setBookCode{Brot}

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: C10th (0.974)

\textbf{Meter:} \Fornyrdislag%
\end{flushright}

\section{Introduction}

Following the Great Lacuna the text of \Regius\ picks up in the middle of a lay about Siward and Byrnhild, namely the present \textbf{Fragment} (\Brot).  According to the following prose (see \textlink{GudrunOne}) the poem began with Siward’s death.  It is perhaps the original “long” Lay of Siward, contrasting with the Short Lay of Siward (\textlink{Sigurdskamma}).

Although most of \Brot\ is lost, parts of \VolsungaSaga\ up to ch. 32 are clearly based on it, A version of st. 4 is cited in \VolsungaSaga\ 30.

TODO: Translation is in progress.

\sectionline

\section{Fragment of a Lay of Siward}

\bvg\bva%
„\edtrans{\emph{hvat hęfir \alst{S}ig·urð}}{[What has Siward]}{\Afootnote{om. \Regius}\Bfootnote{A full reconstruction, which however can be held as reasonably secure owing to the parallelism in the following stanza.}} \hld\ \alst{s}aka unnit &
es þú \alst{f}rǿknan vill \hld\ \alst{f}jǫrvi nę́ma?“\eva

\bvb “[What has Siward] done for a crime, \\
that thou wilt deprive the brave of life?”\evb\evg


\bvg\bva%
„Mér hęfir \alst{S}ig·urðr \hld\ \alst{s}ęlda ęiða &
\alst{ęi}ða sęlda \hld\ \alst{a}lla logna; &
þȧ \alst{v}élti hann mik \hld\ es hann \alst{v}esa skyldi &
\alst{a}llra \alst{ęi}ða \hld\ \alst{ęi}nn full-trúi.“\eva

\bvb “To me has Siward given oaths, \\
oaths given—all broken; \\
he tricked me when he should have been \\
of all the oaths the one true keeper.”\evb\evg


\bvg\bva%
„Þik hęfir \alst{B}ryn·hildr \hld\ \alst{b}ǫl at gørva &
\alst{h}ęiptar \alst{h}vattan \hld\ \alst{h}arm at vinna; &
fyrr man hȯn \alst{G}uð·ru̇nu \hld\ \alst{g}óðra ráða &
en \alst{s}íðan þér \hld\ \alst{s}ïn at njóta.“\eva

\bvb TODO: Translation.\evb\evg


\bvg\bva%
\Ballnote{A related st. is cited in \VolsungaMS\ (\VolsungaSaga\ 30).  It is sufficiently different to be edited on its own; see \textlink{Lacuna}[6] above.}%
Sumir \alst{u}lf sviðu, \hld\ sumir \alst{o}rm sniðu, &
sumir \alst{G}oð·þormi \hld\ af \alst{g}ęra dęildu, &
áðr þęir \alst{m}ę́tti \hld\ \alst{m}ęins of lystir &
ȧ \alst{h}orskum \alst{h}al \hld\ \alst{h}ęndr of lęggja.\eva

\bvb Some roasted a wolf; some cut up a serpent; \\
some shared hound-flesh with Godthorm, \\
TODO..\evb\evg


\bvg\bva%
Úti stóð \alst{G}uð·ru̇n \hld\ \alst{G}júka dóttir &
ok hȯn þat \alst{o}rða \hld\ \alst{a}lls fyrst of kvað: &
„Hvar ’s nú \alst{S}ig·urðr \hld\ \alst{s}ęggja dróttinn &
es \alst{f}rę̇ndr mïnir \hld\ \alst{f}yrri ríða?“\eva

\bvb Outside stood Guthrun, Yivick’s daughter, \\
and she this word first of all did say: \\
“Where is now Siward, the lord of men, \\
when my kinsmen ride ahead?”\evb\evg


\bvg\bva%
\alst{Ęi}nn því Hǫgni \hld\ \alst{a}nd·svǫr vęitti: &
„\alst{S}undr hǫfum \alst{S}ig·urð \hld\ \alst{s}verði hǫgg\emph{v}inn; &
\alst{g}napir ę́ \alst{g}rár jór \hld\ yfir \alst{g}ram dauðum.“\eva

\bvb Alone did Hain this answer grant: \\
“We have cut Siward asunder by sword; \\
his grey steed ever neighs over the dead prince.”\evb\evg


\bvg\bva%
Þȧ kvað þat \alst{B}ryn·hildr \hld\ \alst{B}uðla dóttir: &
„\alst{v}ęl skuluð njóta \hld\ \alst{v}ápna ok níu landa; &
\alst{ęi}nn \edtext{myndi}{\Afootnote{\emph{myndiv} \Regius}} Sig·urðr \hld\ \alst{ǫ}llu ráða &
ef hann \alst{l}ęngr \alst{l}ítlu \hld\ \alst{l}ifi hęldi.“\eva

\bvb Then quoth this Byrnhild, Budle’s daughter: \\
“Well shall ye enjoy weapons and nine lands! \\
Alone would Siward rule them all \\
if a little longer he had held his life.\evb\evg


\bvg\bva%
Vę́ri-a þat \alst{s}ǿmt \hld\ at hann \alst{s}vá réði &
\alst{G}júka arfi \hld\ ok \alst{g}ota męngi, &
es hann \alst{f}imm sonu \hld\ at \alst{f}olk-rǿði &
\alst{g}unnar fu̇sa \hld\ \alst{g}etna hafði.“\eva

\bvb It would not have been fitting that he should so have ruled \\
Yivick’s inheritance and the multitude of the Gots, \\
when to rule the folk he had begot \\
five sons eager for war.\evb\evg


\bvg\bva%
\edtext{Hló þȧ \alst{B}ryn·hildr \hld\ —\alst{b}ǿr allr dunði— &
\alst{ęi}nu sinni \hld\ af \alst{ǫ}llum hug:}{\lemma{Hló \dots\ hug ‘Then \dots\ heart’}\Bfootnote{Found near-identically in \textlink{Sigurdskamma}[30]/1–2.}} &
„vęl skuluð njóta \hld\ landa ok þegna &
es þér \alst{f}rǿknan gram \hld\ \alst{f}alla létuð.“\eva

\bvb Then Byrnhild laughed—the town all resounded— \\
a single time out of all her heart: \\
“Well shall ye enjoy the lands and thanes, \\
since ye made the brave prince to fall.”\evb\evg


\bvg\bva%
Þȧ kvað þat \alst{G}uð·ru̇n \hld\ \alst{G}júka dóttir: &
„\alst{M}jǫk \alst{m}ę́lir þú \hld\ \alst{m}iklar firnar; &
\edtrans{\alst{g}ramir hafi}{let the fiends have}{\Bfootnote{Apparently a common curse; cf. \textlink{Harbardsljod}[60].}} \alst{G}unnar \hld\ \alst{g}ǫtvað Sig·urðar, &
\alst{h}ęipt-gjarns \alst{h}ugar \hld\ \alst{h}ęfnt skal verða.“\eva

\bvb Then quoth this Guthrun, Yivick’s daughter: \\
“Many great evils hast thou spoken; \\
let the fiends have Guther, Siward’s gravedigger! \\
For the evil heart there must be revenge.”\evb\evg


\bvg\bva%
\alst{S}oltinn varð \alst{S}ig·urðr \hld\ \alst{s}unnan Rínar, &
\alst{h}rafn at męiði \hld\ \alst{h}ǫ́tt kallaði: &
„\alst{Y}kkr mun \alst{A}tli \hld\ \alst{ę}ggjar rjóða, &
munu \alst{v}íg-ská\emph{a} \hld\ of \alst{v}iða ęiðar.“\eva

\bvb Siward expired to the south of the Rhine; \\
a raven on a branch loudly called out: \\
“On you two will Attle redden his blades; \\
the oaths will destroy the war-harried men!”\evb\evg


\bvg\bva%
\alst{F}ramm vas kvelda \hld\ \alst{f}jǫlð vas drukkit &
þȧ vas hví-\alst{v}etna \hld\ \alst{v}il-mál talit; &
\alst{s}ofnuðu allir \hld\ es ï \alst{s}ę́ing kvǫ́mu, &
\alst{ęi}nn vakði Gunnarr \hld\ \alst{ǫ}llum lęngr.\eva

\bvb The evening was late, much was drunk; \\
then were all kinds of pleasant speeches told. \\
All fell asleep when they had come to bed; \\
alone did Guther wake longer than all.\evb\evg


\bvg\bva%
\alst{F}ót nam at hrǿra \hld\ \alst{f}jǫlð nam at spjalla &
\alst{h}itt \alst{h}ęr-glǫtuðr \hld\ \alst{h}yggja téði, &
hvat þęir ï \alst{b}ǫðvi \hld\ \alst{b}áðir sǫgðu &
\alst{h}rafn ęy ok ǫrn \hld\ es þęir \alst{h}ęim riðu.\eva

\bvb His foot he took to move, much he took to consider. \\
Of this the troop-destroyer thought, \\
what in the fray those two had said, \\
the raven always and the eagle, when they rode home.\evb\evg


\bvg\bva%
Vaknaði \alst{B}ryn·hildr \hld\ \alst{B}uðla dóttir &
\alst{d}ís skjǫldunga \hld\ fyr \alst{d}ag lítlu: &
„\alst{h}vętið mik eða lętið mik \hld\ \alst{h}armr es unninn &
\alst{s}org at \alst{s}ęgja \hld\ eða \alst{s}vá láta.“\eva

\bvb Byrnhild awoke, Budle’s daughter, \\
the dise of shieldings, a little before day: \\
“TODO.”\evb\evg


\bvg\bva%
\Ballnote{Sts. 15–20 are transparently paraphrased in \VolsungaSaga\ 31: \emph{Nú þóttist engi kunna at svara, at Brynhildr beiddi þess hlę́jandi, er hón harmaði með gráti.  Þá mę́lti hón: „Þat dreymdi mik, Gunnarr, at ek átta kalda sę́ng, en þú ríðr í hendr ó·vinum þínum, ok ǫll ę́tt yður mun illa fara, er þér eruð eið-rofa, ok mundir þú þat ó·gløggt, er þit blǫnduðuð blóði saman, Sigurðr ok þú, er þú rétt hann, ok hefir þú honum allt illu launat þat, er hann gerði vel til þín ok lét þik fremstan vera, ok þá reyndi þat, er hann kom til vár, hvé hann helt sína eiða, at hann lagði okkar í milli it snarp-eggjaða sverð, þat er eitri var hert. [...]} ‘(15) Now nobody thought himself able to answer how Byrnhild had laughed when she asked for what she now grieved by weeping.  (16) Then she spoke: “I dreamt, Guther, that I had a cold bed, but thou didst ride into the hands of thy enemies, (17) and all your lineage will fare ill, for ye are oath-breakers, (18) and thou didst not kindly recall how ye two blended your blood together, thou and Siward, when thou betrayed him; and thou hast all cruelly repaid him for how he was good towards thee and let thee be the foremost; (19) and it was proven when he came to us how well he kept his oaths (20) by that he laid between us the sharp-edged sword which was hardened with venom. [...]’}%
\alst{Þ}ǫgðu allir \hld\ við \alst{þ}ví orði &
\alst{f}ár kunni þęim \hld\ \alst{f}ljóða lǫ́tum &
es hǫ̇n \alst{g}rátandi \hld\ \alst{g}ørðisk at sęgja &
þat’s \alst{h}lę́jandi \hld\ \alst{h}ǫlða bęiddi:\eva

\bvb All were silent at that word; \\
TODO. \\
when she, weeping, made ready to tell \\
that which, laughing, she had asked of men.\evb\evg


\bvg\bva%
„Hugða’k mér, \alst{G}unnarr, \hld\ \alst{g}rimmt ï svefni, &
\alst{s}valt allt ï \alst{s}al \hld\ ę́tta’k \alst{s}ę́ing kalda, &
en þú \alst{g}ramr riðir \hld\ \alst{g}laums and·vani &
\alst{f}jǫtri \alst{f}atlaðr \hld\ ï \alst{f}jánda lið.\eva

\bvb “I had a cruel thought, Guther, in my sleep: \\
everything died in the hall, I had a cold bed, \\
and thou, prince, didst ride without cheerful fellows, \\
bound by fetters, into a troop of foes.\evb\evg


\bvg\bva%
Svá mun \alst{ǫ}ll \alst{y}ður \hld\ \alst{ę́}tt niflunga &
\alst{a}fli gęngin— \hld\ eruð \alst{ęi}ð-rofa!\eva

\bvb So will all your line of Nivlings \\
part from power—ye are oath-breakers!\evb\evg


\bvg\bva%
Mant-at-tu \alst{G}unnarr \hld\ til \alst{g}ǫrva þat &
es \edtrans{þit \alst{b}lóði ï spor \hld\ \alst{b}áðir ręnnduð}{your blood in your tracks ye both did drive}{\Bfootnote{Referring to a ritual of blood-brotherhood, wherein the brothers-to-be would spill and mix their blood into their footprints on the ground.  Blood-brotherhood is also mentioned in \textlink{Lokasenna}[9], and the footprint-ritual is described in \textcite{Saxo} 1.6.7: \emph{Siquidem icturi foedus veteres vestigia sua mutui sanguinis aspersione perfundere consueverant, amicitiarum pignus alterni cruoris commercio firmaturi} ‘Now our ancestors, when they meant to strike a pact, would sprinkle their combined blood in their footprints and mingle it, so as to strengthen the pledge of their fellowship.’  For discussion see PCRN History and Structures III:845 ff.}}, &
nú hęfir þú hǫ̇num þat \alst{a}llt \hld\ \alst{i}llu launat &
es hann \alst{f}ręmstan \edtrans{\emph{þ}ik}{thee}{\Afootnote{so in paraphrase \VolsungaMS; \emph{sik} ‘himself’ \Regius}} \hld\ \alst{f}inna vildi.\eva

\bvb Thou didst not recall, Guther, clearly enough \\
that your blood in your tracks ye both did drive. \\
Now hast thou for all that cruelly repaid him, \\
when he would have \emph{thee} as the foremost man.\evb\evg


\bvg\bva%
Þȧ \alst{r}ęyndi þat \hld\ es \alst{r}iðit hafði &
\alst{m}óðigr ȧ vit \hld\ \alst{m}ïn at biðja, &
\alst{h}vé \alst{h}ęr-glǫtuðr \hld\ \alst{h}afði fyrri &
\alst{ęi}ðum haldit \hld\ við inn \alst{u}nga gram.\eva

\bvb Then it was proven when he had ridden, \\
brave, to visit and ask for my hand, \\
how the troop-destroyer \ken{warrior} had uhpeld \\
his side of the oaths against the young lord:\evb\evg


\bvg\bva%
\alst{B}ęn-vǫnd of lét \hld\ \alst{b}rugðinn \edtext{gulli}{\Afootnote{\emph{gylli} \Regius}} &
\alst{m}arg-dýrr konungr \hld\ ȧ \alst{m}eðal okkar; &
\alst{ę}ldi vǫ́ru \alst{ę}ggjar \hld\ \alst{ú}tan gǫrvar &
ęn \alst{ęi}tr-dropum \hld\ \alst{i}nnan fáðar.“\eva

\bvb He placed a wound-wand \ken{sword} wrapped in gold, \\
the much-wealthy king, between us two. \\
With fire was its blade made on the outside, \\
but with poison-droplets stained on the inside.”\evb\evg


\section{From the Death of Siward (\emph{Frá dauða Sig·urðar})}

\bpg\bpa Hér er sagt í þessi kviðu frá dauða Sig·urðar ok víkr hér svá til sem þeir drę́pi hann úti.  En sumir segja svá at þeir drę́pi hann inni í rekkju sinni sofanda. En þýðverskir menn segja svá at þeir drę́pi hann úti í skógi ok svá segir í Guð·ru̇nar kviðu inni fornu at Sig·urðr ok Gjúka synir hefði til þings riðit þá er hann var drepinn.  En þat segja allir einnig at þeir sviku hann í tryggð ok vógu at hánum liggjanda ok ó·búnum.\epa

\bpb Here it has been said in this lay from the death of Siward, and it picks up at the point when they slew him outside.  But some say that they slew him inside in his chamber, sleeping.  But the Germans say that they slew him outside in the forest, and so it says in the Ancient Lay of Guthrun that Siward and the sons of Yivick had ridden to the Thing when he was slain.  But this they all say in unison, that they betrayed him while he trusted them and struck at him lying down and unarmed.\epb\epg

\sectionline
