\bookStart{The Icelandic Rune Poem}
\setBookCode{IcelandicRunePoem}

\begin{flushright}%
\textbf{Dating:} Mediæval.%TODO

\textbf{Meter:} Unclear.
\end{flushright}%

\section{Introduction}

The poem is highly formulaic.  All lines begin with the respective rune’s name, followed by three kennings for it.  It is only attested in late manuscripts which often have major disagreements with each other.

\sectionline

\section{Text}

\bvg\bva%
\alst{F}é es \alst{f}rę́nda róg \hld\ ok \alst{f}lǿðar viti &
\ind ok \alst{g}raf-sęiðs \alst{g}ata.\eva

\bvb Wealth is strife of kinsmen and beacon of the sea \\
\ind and grave-saithe’s \ken{serpent’s} street.\evb\evg


\bvg\bva%
Úr es \alst{sk}ýja grátr \hld\ ok \alst{sk}ára þvęrrir &
\ind ok \alst{h}irðis \alst{h}atr.\eva

\bvb Drizzle is weeping of clouds and ... \\
\ind and shepherd’s hatred.\evb\evg


\bvg\bva%
Þurs es \alst{k}venna \alst{k}vǫl \hld\ ok \alst{k}letta í·búi &
\ind ok \alst{v}arð-rúnar \alst{v}err.\eva

\bvb Thurse is women’s torment and indweller of hills \\
\ind and husband of the cairn-whisperess \ken{trollwoman}.\evb\evg


\bvg\bva%
\alst{Ǫ̇}ss es \alst{a}ldinn gautr \hld\ ok \alst{Ǫ̇}s-garðs jǫfurr, &
\ind ok \alst{V}al-hallar \alst{v}ísi.\eva

\bvb Os is ancient Geat, and Osyard’s chief, \\
\ind and Walhall’s overseer.\evb\evg


\bvg\bva%
Ręið es \alst{s}itjandi \alst{s}ę́la \hld\ ok \alst{s}núðig fęrð &
\ind ok \alst{jó}s \alst{ę}rfiði.\eva

\bvb Chariot is sitting bliss and twirling journey \\
\ind and horse’s heavy work.\evb\evg


\bvg\bva%
Kaun es \alst{b}arna \alst{b}ǫl \hld\ ok \alst{b}ar-dagi &
\ind ok \alst{h}old-fúa \alst{h}ús.\eva

\bvb Boil is children’s curse and TODO \\
\ind and house of flesh-rot.\evb\evg


\bvg\bva%
Hagall es \alst{k}alda \alst{k}orn \hld\ ok \alst{k}nappa drífa &
\ind ok \alst{s}náka \alst{s}ótt.\eva

\bvb Hail is cold kernel and storm of beads \\
\ind and sickness of snakes.\evb\evg


\bvg\bva%
Nauð es \alst{þ}ýjar \alst{þ}rǫ́ \hld\ ok \alst{þ}ungr kostr &
\ind ok \alst{v}ás-samlig \alst{v}erk.\eva

\bvb Need is maidservant’s yearning and scant choice \\
\ind and working in wet-cold weather.\evb\evg


\bvg\bva%
\alst{Í}ss es \alst{á}ar bǫrkr \hld\ ok \alst{u}nnar þękja &
\ind ok \alst{f}ęigra manna \alst{f}ár.\eva

\bvb Ice is river’s bark and wave’s roof \\
\ind and fey men’s danger.\evb\evg


\bvg\bva%
Ár es \alst{g}umna \alst{g}óði \hld\ ok \alst{g}ótt sumar &
\ind \emph{ok} \alst{a}l-gróinn \alst{a}kr.\eva

\bvb Year is men’s boon and good summer \\
\ind (and) all-grown acre.\evb\evg


\bvg\bva%
Sól es \alst{sk}ýja \alst{sk}jǫldr \hld\ ok \alst{sk}ínandi rǫðull &
\ind ok \alst{í}sa \alst{a}ldr-tregi.\eva

\bvb Sun is the shield of clouds and shining wheel \\
\ind and ice-sheets’ life-sorrow.\evb\evg


\bvg\bva%
Týr es \alst{ęi}n-hęndr \alst{ǫ̇}ss \hld\ ok \alst{u}lfs lęifar &
\ind ok \alst{h}ofa \alst{h}ilmir.\eva

\bvb Tew is the one-handed Os and the wolf’s leftovers \\
\ind and lord of hoves.\evb\evg


\bvg\bva%
Bjarkan es \alst{l}aufgat \alst{l}im \hld\ ok \alst{l}ítit tré &
\ind ok \alst{u}ng-samligr \alst{v}iðr.\eva

\bvb Birch is leafy branch and little tree \\
\ind and youthful wood.\evb\evg


\bvg\bva%
\alst{M}aðr es \alst{m}anns gaman \hld\ ok \alst{m}oldar auki &
\ind ok \alst{sk}ipa \alst{sk}ręytir.\eva

\bvb Man is man’s joy and the product of dust \\
\ind and adorner of ships.\evb\evg


\bvg\bva%
Lǫgr es \alst{v}ellanda \alst{v}atn \hld\ ok \alst{v}íðr kętill &
\ind ok \alst{g}lǫmmungr \alst{g}rund.\eva

\bvb Liquid is boiling water and wide kettle \\
\ind and TODO.\evb\evg


\bvg\bva%
Ýr es \alst{b}ęndr bogi \hld\ ok \alst{b}rot-gjarnt járn &
\ind ok \alst{f}ęnju \alst{f}lęygir.\eva

\bvb Yew is a bent bow and easily broken iron \\
\ind and arrow’s hurler.\evb\evg

\sectionline
