%Runic poetry from Sweden and Gotland.

\section{Introduction}

The following inscriptions are sorted according to their signum in the standard corpus edition \emph{Sveriges runinskrifter} (‘Runic inscriptions of Sweden’).  They are numbered according to traditional province, so that Sm 39 is the 39th inscription from Småland.

The province of Scania, although part of Sweden since 1658, was in the Wiking Age and the mediæval period an integral part of Denmark and as such its metrical inscriptions are found in the Danish section.

\subsection{Abbreviations of Swedish provinces}

\begin{itemize}
  \item G = Gotland
  \item Sm = Småland
  \item Sö = Södermanland
  \item U = Uppland
  \item Vg = Västergötland
  \item Ög = Östergötland
\end{itemize}

\sectionline

\section{G 203}

\begin{flushright}%
\textbf{Dating:} C12th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

TODO.

\bpg\bpa[0]%
\startsub Sig·mundr lét raisa stain eptiʀ brýðr sïna, auk bró gierva eptiʀ Sig·biern—Sankta Mikál hielpi \emph{siál h}ans—auk at Bót·raif auk at Sig·raif auk at Ai·biern, faður þaiʀa aldra,\endsub\epa

\bpb Syemund had this stone raised after his brothers and the bridge made after Syebern—may Saint Michael help his soul—and after Bootraf and after Syeraf and after Eanbern, the father of them all,\epb\epg

\bvg\bva[][2]%
auk \alst{b}yggvi hann \hld\ ï \alst{b}ý sunnarst.\eva

\bvb and he lived on the southernmost farm.\evb\evg

\bpg\bpa[0]%
Gaiʀviðr lęgði orm-áluʀ; némʀ innti ýʀ.\epa

\bpb Garwith laid the serpent-tracks; TODO.\epb\epg

\bvg\bva[][4]%
\alst{S}ig·mundr \emph{hefiʀ} \hld\ \alst{s}líku unnit &
\alst{k}uml \alst{k}arl-mannum. \hld\ Þet aʀ †\alst{k}e...† kunn. &
Hier mun \alst{st}anda \hld\ \alst{st}ainn at merki, &
\alst{b}iertr á \alst{b}iergi, \hld\ en \alst{b}ró fyriʀ; &
Róð·biern \alst{r}ísti \hld\ \alst{r}úniʀ [þ]essaʀ, &
\alst{G}aiʀ·laifʀ sumaʀ, \hld\ aʀ \alst{g}arla kann.\eva

\bvb[] Syemund has accomplished such \\
a monument for men; that is known to ... \\
Here the stone will stand as a mark, \\
bright on the hill and the bridge below. \\
Rothbern carved these runes, \\
{[and]} Garlaf, who knows clearly, some.\evb\evg

\sectionline

\section{Sm 5}

\begin{flushright}%
\textbf{Dating:} C11th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

TODO.
The carver seems to struggle with consonant clusters.  Thus he spells 2b \emph{mestr} as \textbf{mesr} and 3b \emph{aldri} as \textbf{ali}(!).  Line 2 is found near-identically in Ög 77 and line 3 in Vg 187.

\bvg\bva[]%
Gø̂tr satti stên þęnna eptiʀ Ketil sun sinn. &
Hann vaʀ \alst{m}anna \hld\ \edtrans{\alst{m}estr ó·níðingʀ}{greatest un-nithing}{\Bfootnote{An instance of \emph{litotes}; Kettle was a very virtuous man.}} &
eʀ ȧ \alst{E}ng-landi \hld\ \edtext{\alst{a}l\emph{dr}i}{\Afootnote{\textbf{ali} Sm 5}} tẏnði.\eva

\bvb Geat raised this stone after Kettle, his son. \\
He was of men the greatest un-nithing, \\
who in England lost his life.\evb\evg

\sectionline

\section{Sm 16}

\begin{flushright}%
\textbf{Dating:} C11th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

A not very tall stone, inscribed on two sides.  Side A (1 \emph{Hró·stęinn} \dots\ 3 \emph{kęnni⸗ligt}) is marked with a large cross while side B (3 \emph{ęptir} \dots\ 5 \emph{ru̇na}) is entirely unadorned.

The inscription uses the stung runes \emph{e} and \emph{y} and writes the vowel \emph{o} with the old \emph{ą}-rune.

\bvg\bva[]%
Hró·stęinn auk \alst{Ęi}·lífʀ, \hld\ \alst{Á}ki auk Hǫ́·kun, &
ręistu þęiʀ \alst{s}vęinaʀ \hld\ ęptiʀ \alst{s}inn faður &
\alst{k}umbl \alst{k}ęnni⸗ligt \hld\ ęptiʀ \alst{K}ala dauðan. &
Þý mun \alst{g}ó\emph{ðs manns} \hld\ \emph{u}m \alst{g}etit verða, &
með \alst{s\emph{t}}\emph{ę}inn lifiʀ \hld\ ok \alst{st}afiʀ ru̇na.\eva

\bvb Rothstan and Anlif, Onick and Hathkin, \\
these lads raised after their father \\
a remarkable monument, after the dead Cale. \\
Therefore will the good man be spoken of, \\
while the stone lives and the staves of the runes.\evb\evg

\sectionline

\section{Sm 39}

\begin{flushright}%
\textbf{Dating:} C11th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

A standing stone inscribed on two sides, one of which has a large cross.  The expression is formulaic; cf. Sm 44, Sö 130, U 703, U 739, and U 805.  For “\inx[C]{good of meat}” cf. \textlink{Havamal}[39].  The first line is not poetic.

\bvg\bva[]%
Gunni satti stên þęnna eptiʀ Súna, fǫður sinn, &
\alst{m}ildan orða \hld\ ok \alst{m}ataʀ góðan.\eva

\bvb Guthe set this stone after Sown, his father, \\
generous of words and good of meat.\evb\evg

\sectionline

\section{Sm 44}

\begin{flushright}%
\textbf{Dating:} C11th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

A heavily damaged stone, inscription more visible in Bautil.
The expression is formulaic; cf. Sm 39, Sö 130, U 703, U 739, and U 805.

\bvg\bva[]%
TODO
\alst{m}ildan við sinna \hld\ ok \alst{m}ataʀ góðan, &
TODO.\eva

\bvb TODO \\
Generous with his men and good of meat. \\
TODO\evb\evg

\sectionline

\section{Sö 34–35 (Tjuvstigen)}

\begin{flushright}%
\textbf{Dating:} C11th–C12th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

Two paired stones standing next to each other.  The last two lines of Sö 35 is not poetic.

\bvg\bva[Sö 34]%
\alst{St}yr·laugʀ ok Holmbʀ \hld\ \alst{st}ęina ręistu &
at \alst{b}rǿðr sína, \hld\ \edtrans{\alst{b}rautu nę́sta}{nearest to the road}{\Bfootnote{Cf. \textlink{Havamal}[72].}}. &
Þęir \alst{ę}ndaðus \hld\ í \alst{au}str-vegi, &
\alst{Þ}ór·kęll ok Styr·bjǫrn, \hld\ \alst{þ}iagnar góðir.\eva

\bvb Sturley and Holm raised the stones, \\
after their brothers, nearest to the road. \\
They were ended in the Eastern way, \\
Thurkettle and Sturbern, good thanes.\evb\evg


\bvg\bva[Sö 35]%
Lét \alst{I}ngi·gęiʀʀ \hld\ \alst{a}nnan ręisa stęin &
at \alst{s}onu \alst{s}ïna, \hld\ \alst{s}ýna giǫrði. &
Guð hialpi ǫnd þęira. \\
Þóriʀ hjó.\eva

\bvb Inggar let another stone be raised \\
after his sons; he made it seen. \\
God help their spirit. \\
Thurer hewed.\evb\evg

\sectionline

\section{Sö 56 (Fyrby)}

\begin{flushright}%
\textbf{Dating:} C11th–C12th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

TODO: INTRODUCTION.

\bvg\bva[]%
Iak vęit \alst{H}á·stęin \hld\ þá \alst{H}olm·stęin brǿðr &
\alst{m}ęnnr rýnasta \hld\ ȧ \alst{M}ið-garði &
sęttu \alst{st}ęin \hld\ auk \alst{st}afa marga &
eptir \alst{F}ręy·stęin \hld\ \alst{f}ǫður sinn.\eva

\bvb I know Highstan and Holmstan, the brothers, \\
the most rune-cunning men in Middenyard. \\
They set the stone and many staves \\
after Freestan, their father.\evb\evg

\sectionline

\section{Sö 65 (Djulefors)}

\begin{flushright}%
\textbf{Dating:} C11th–C12th

\textbf{Meter:} \Fornyrdislag\ with hendings in the b-verses
\end{flushright}%

A standing stone inscribed on one side with a large cross.  Already on the earliest depictions the stone was damaged, but an even larger part has now gone missing.  Other stones that mention \inx[P]{Longbeardland} (Lombardy) include TODO...  The meter is highly unusual for runic Swedish poetry, relying on hendings (in line 2 an ethel-hending \emph{arð- : barð-}, in line 3 a shot-hending \emph{land- : ęnd-}).  Line 2b is formulaic; see note.

\bvg\bva[]%
Inga ręisti stęin þannsi at Ǫ́·lęif sinn \textbf{a}... &
Hann \alst{au}star⸗la \hld\ \edtrans{\alst{a}rði barði}{ploughed with the prow}{\Bfootnote{i.e. “sailed”.  A formulaic poetic expression shared with an anonymous line from the Third Grammatical Treatise, which reads: \emph{sá’s af Íslandi \hld\ arði barði} ‘he who [awawy] from Iceland ploughed with the prow’.}} &
auk ȧ \alst{L}ang-barði- \hld\ \alst{l}andi ęndaðis.\eva

\bvb Inge raised this stone after Anlaf, her ... \\
Easterly he ploughed with the prow, \\
and on Longbeardland was ended.\evb\evg

\sectionline

\section{Sö 130}

\begin{flushright}%
\textbf{Dating:} C11th–C12th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

A standing stone. TODO.  The expression is formulaic; cf. Sm 39, Sm 44, U 703, U 739, and U 805.

\bvg\bva[]%
\alst{F}iuriʀ gęrðu \hld\ at \alst{f}ǫður góðan &
\alst{d}ýrð \alst{d}ręngi⸗la \hld\ at \alst{D}ómara &
\alst{m}ildan orða \hld\ ok \alst{m}ataʀ góðan. &
Þat \dots\eva

\bvb Four men made after their good father, \\
honourably a mark of praise after Doomer \\
mild of words and good of meat. \\
This \dots\evb\evg

\sectionline

\section{Sö 154 (Skarpåker)}

\begin{flushright}%
\textbf{Dating:} C11th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

The couplet at the end, expressing a father’s grief for his son, also serves as a good example of the Wiking Age preoccupation with the End Times.  The stone is decorated with a cross, but the text has no signs of Christian influence, and the language is traditional.

Cf. especially Arn \emph{Hryn} 3/7–8 (in \Skp\ 2, see also note there): \emph{meiri verði þinn an þeira \hld\ þrifnuðr allr, und’s himinn rifnar.} ‘greater than theirs may all thy wealth be until heaven rends.’

\bpg\bpa[0]%
Gunn·arr ręisti stęin þannsi at Lýð·bjorn, son sinn.\epa

\bpb Guther raised this stone after Leodbern, his son.\epb\epg


\bvg\bva[][2]%
\alst{Jǫ}rð \edtrans{sal}{shall}{\Bfootnote{A Swedish dialectal form of \emph{skal} ‘id.,’ cf. dialectal Swedish \emph{sa}.}} rifna \hld\ ok \alst{u}pp-himinn.\eva

\bvb Earth shall rend, and Up-heaven.\evb\evg

\sectionline

\section{Sö 179 (Gripsholm)}

\begin{flushright}%
\textbf{Dating:} C11th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

TODO: INTRODUCTION.  The three-line stanza is a biographical addition following a typical prose memorial formula.

\bpg\bpa[0]%
Tóla lét ręisa stęin þennsa at son sinn Harald, bróður Ingvars.\epa

\bpb Toole had this stone raised after his son Harold, brother of Ingwar.\epb\epg

\bvg\bva[]%
Þęiʀ \alst{f}óru dręngi⸗la \hld\ \alst{f}iarri at gulli &
ok \alst{au}star⸗la \hld\ \edtrans{\alst{ę}rni gǫ́fu}{gave to the eagle}{\Bfootnote{They “provided a feast for the eagle”, namely with the carnage of slain foes; for eagles and ravens as eaters of corpses and drinkers of blood cf. \textcite[118,203,207--208]{Meissner1921}.  Similar things are said of kings in numerous Scaldic poems from Iceland and Norway, and the lack of an object to \emph{gǫ́fu} reveals that this expression must have been well known also in Sweden.}}, &
dóu \alst{s}unnar⸗la \hld\ ȧ \alst{S}ęrk-landi.\eva

\bvb They journeyed honourably far for gold, \\
and easterly gave to the eagle; \\
died southerly in Serkland.\evb\evg

\sectionline

\section{U 323}

\begin{flushright}%
\textbf{Dating:} C11th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

A standing stone from Sälna, Uppland.  No decorations, but explicitly Christian.  Note the loss of initial \emph{h}.

\bpg\bpa[0]%
\startsub Ęy·stên auk Iurundr auk Biorn, þêʀ brǿðr rêsþu \dots\ \emph{ęptir} \dots stên \textbf{trums} fǫður sinn. \\
Guð ’ialpi ’ans ǫnd auk sêlu; for·gifi ’ǫ̇num sakaʀ auk syndiʀ.\endsub\epa

\bpb Eanstan and Earend and Bern—those brothers raised \dots\ [after] -stan \dots\ their father. \\
God help his breath and soul, forgive him his crimes and sins.\epb\epg


\bvg\bva[][3]%
\emph{\alst{Ę}}i mun liggja \hld\ með \alst{a}ldr lifiʀ &
\alst{b}ró harð-slagin \hld\ \alst{b}rêð e\emph{ft g}óð\emph{an}; &
\alst{s}vênaʀ k\emph{ęrðu} \hld\ at \alst{s}inn fǫður &
má êgi \alst{b}rø̂taʀ-kuml \hld\ \alst{b}ętra verða.\eva

\bvb Ever will lie—while the age lives— \\
the hard-hammered bridge, broad, after a good man. \\
The swains made it after their father; \\
no roadside monument can be greater.\evb\evg

\sectionline

\section{U 703}

\begin{flushright}%
\textbf{Dating:} C11th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

A standing stone inscribed on one side.  There is no cross present, but a large four-legged beast with a long tail.  The stone is heavily damaged, but mostly readable, except for what is here taken to be the half of line 2, which is entirely lost.  The expression is formulaic; cf. Sm 39, Sm 44, Sö 130, U 739, and U 805.  For “\inx[C]{good of meat}”, which also occurs in \textlink{Havamal}; see Index.  The first line is not poetic.

\bvg\bva[]%
Ás·vi lét ręisa stęin þennsa at Ǫrn·ulf, son sinn góðan. &
Hann byggi hér \hld\ ..., &
\alst{m}andr \alst{m}atar góðr \hld\ ok \alst{m}áls risinn.\eva

\bvb Oswye let raise this stone after Arnolf, her good son. \\
He dwelled here ..., \\
a man good of meat and proud of speech.\evb\evg

\sectionline

\section{U 739}

\begin{flushright}%
\textbf{Dating:} C11th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

A standing stone inscribed on one side, with a large cross present.  There are no major difficulties with the reading.  The expression is formulaic; cf. Sm 39, Sm 44, Sö 130, U 703, and U 805.  “mild of meat” appears to be a variant of “\inx[C]{good of meat}”, which also occurs in \textlink{Havamal}; see Index.  The first line is not poetic.  For other stones raised by someone in memory of themselves, see TODO.

\bvg\bva[]%
Hol·bjǫrn lét ręisa stęin at sik sialfan. &
Hann vaʀ \alst{m}ildr \alst{m}ataʀ \hld\ ok \alst{m}áls risinn.\eva

\bvb Holbern had this stone raised after himself. \\
He was mild of meat and proud of speech.\evb\evg

\sectionline

\section{U 805}

\begin{flushright}%
\textbf{Dating:} C11th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

The stone has been lost, and only survives in old depictions, which makes the reading, especially two of the personal names, uncertain.  My transliteration follows Rundata.

The expression is formulaic; cf. Sm 39, Sm 44, Sö 130, U 703, and U 739.  For “\inx[C]{good of meat}”, which also occurs in \textlink{Havamal}; see Index.  The first line is not poetic.

\bpg\bpa[]%
\startsub Fylkir lét ręisa st\emph{ęin epti}r \textbf{iel}, bróður sinn, ok Gunn·marr eptir \textbf{menk}, fǫður sinn,\endsub \\
\alst{b}ónda góðan matar; \hld\ \alst{b}yggi ï Vík-bý.\epa

\bpb Filch had this stone raised after ..., his brother, and Guthmar after ..., his father, \\
a farmer good of meat; he lived in Wickby.\epb\epg

\sectionline

\section{Vg 59}

\begin{flushright}%
\textbf{Dating:} C11th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

A large standing stone.

\bpg\bpa[0]%
\startsub \emph{H}ręfningʀ auk Gialli auk Bryn·ulfʀ auk Gef·ulfʀ sǫttu stên þannsi ęptiʀ Fót, fǫður sinn, harða góðan þegn.\endsub\epa

\bpb Rafning and Yeller and Byrnulf and Yevulf set this stone after Foot, their father, a very good thane.\epb\epg

\bvg\bva[][2]%
Svá hafiʀ \alst{Ȧ}sa, \hld\ es \alst{ê}gi mun &
\alst{s}um kvę̇n ępt vęr \hld\ \alst{s}íðan gǫrva. &
\alst{H}ialmʀ ok \alst{H}ialli \hld\ \alst{h}iǫggu ru̇naʀ.\eva

\bvb So has Eesie [made a monument], like never a wife \\
henceforth after her husband will make. \\
Helm and Helle hewed the runes.\evb\evg

\sectionline

\section{Vg 187}

\begin{flushright}%
\textbf{Dating:} C11th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

A large standing stone decorated with a cross.  The runes are well preserved.

\bpg\bpa[0]%
\startsub Gêʀi satti stên þęnna ęptiʀ Goða, bróður sinn,\endsub
eʀ ȧ \alst{Ę}ng-landi \hld\ \alst{a}ldri tẏnði.\epa

\bpb Gare set this stone after Gid, his brother, \\
who in England lost his life.\epb\epg

\sectionline

\section{Ög 77}

\begin{flushright}%
\textbf{Dating:} C11th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

TODO.

\bpg\bpa[0]%
Tonna satti stên þ\emph{ęnna ęp}t\emph{i}ʀ sinn vęr Þȯr·fast. \\
\edtrans{Vas hann \edtext{\alst{m}an\emph{na}}{\Afootnote{\textbf{man} Ög 77}} \hld\ \edtext{\alst{m}ęst\emph{r}}{\Afootnote{\textbf{mist} Ög 77}} \edtext{ȯ·níðingʀ}{\Afootnote{\textbf{uniþik} Ög 77}}}{He was of men the greatest un-nithing}{\Bfootnote{Found near-identically in Sm 5 above.}} \epa

\bpb Tonne set this stone after her husband, Thurfast. \\
He was of men the greatest un-nithing.\epb\epg

\sectionline

\section{Ög 81}

\begin{flushright}%
\textbf{Dating:} C11th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

A stone standing at the former site of the mediæval Högby church, into which it was once bricked in.  The stone is inscribed on 2 sides, A and B.

The orthography is notable in that older \emph{ęi} is consistently written as a monophthong, while \emph{au} is still written as a diphthong.  This unusual variation is reflected in the standardized orthography.


\bvg\bva[A]%
Þor·gęr\emph{ðr} rêsþi stên þannsi ęptiʀ Ǫss·ur sên, móður-bróður sinn, &
iaʀ \alst{ę}ndaðis \hld\ \alst{au}str ï Grikkjum.\eva

\bvb Thurgerd raised this stone after Answer the slow, her uncle \\
who lost his life to the east in Greece.\evb\evg


\bvg\bva[B][3]%
\alst{G}óðr karl \alst{G}ulli \hld\ \alst{g}at fimm syni. &
\alst{F}iall ȧ \alst{F}ýri \hld\ \alst{f}rǿkn dręngʀ Ǫ́s·mundr, &
\alst{ę}ndaðis \alst{Ǫ}ss·urr \hld\ \alst{au}str ï Grikkjum, &
varð ȧ \alst{H}olmi \hld\ \alst{H}alf·dan drepinn, &
Kári varð at \alst{U}ddi \hld\ \alst{au}k dauðr Bói. &
Þȯr·kell rêst ru̇naʀ.\eva

\bvb The good churl Gull begat five sons. \\
On Fyre fell the brave young man Osmund, \\
Answer lost his life to the east in Greece, \\
upon Holm was Halfdene slain, \\
Cheer was [slain] by Orde, also dead is Bo. \\
Thurkettle carved the runes.\evb\evg

\sectionline

\section{Ög 165}

\begin{flushright}%
\textbf{Dating:} C11th

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

A tall standing stone from Skänninge.  The carver Thurkettle is the same as that of Ög 81.

\bvg\bva[]%
\alst{Þ}ȯr·unn \edtrans{rêsti}{raised}{\Bfootnote{I.e., ‘raised the stone’.  The wording is so formulaic that the object can be left out.}} \hld\ auk \alst{þ}êʀ brǿðr, &
\alst{s}yniʀ Tosta, \hld\ ęptiʀ \alst{s}inn fǫður. &
Þȯr·kell ’k rêst stên þannsi \edtext{\emph{ę}pti}{\Afootnote{\textbf{aufti} Ög 165}} Tosta.\eva

\bvb Thurun raised along with her brothers \\
the sons of Toste, after their father. \\
I, Thurkettle, carved this stone after Toste.\evb\evg

\sectionline
