\section{Against wyrms (\emph{Contra vermes})}\chapterStart{}
\setBookCode{ContraVermes}

\begin{flushright}%
\textbf{Dating:} 9th c.

\textbf{Meter:} \Fornyrdislag%para
\end{flushright}%

\subsection{Introduction}

\textbf{Against wyrms} (abbrev. \ContraVermes) is an Old Saxon manuscript charm against wyrms located in the bone-marrow, presumably thought to cause aches.

\subsubsection{Preservation}
\ContraVermes\ is found as marginalia in the 9th c. manuscript Cod. Vind. (ÖNB) 751 on fol. 188v.

\subsubsection{Contents}

The galder is a sort of exorcism.  It calls upon Nesse, the chief injurious wyrm-dæmon, to depart from the patient along with its nine offspring.  The galder lays out a path for the wyrms to leave the sufferer’s body and instead go into an arrow or sharp point (\emph{strála}), perhaps a ritual implement used to draw blood from the affected area.

A closely related galder is found in Old High German form under the title \emph{Pro Nessia} ‘Against Nesse’ (\textlink{ProNessia}), edited immediately below.

The structure “Go from X to Y, from Y to Z” may be very old, for it is also found in some of the Romani charms collected by \textcite[27,28,95]{Leland1891}.  One of Leland’s charms is particularly similar to \ContraVermes\ and \ProNessia.  In that charm, the wyrms (Romani \emph{kirmora} < Sanskrit \emph{kŕ̥mi}, which is probably related to PGmc. \emph{*wurmiz}, although the irregular variation in the initial consonant is unusual and perhaps due to a taboo deformation) are to leave the body and instead go into the ritual implement, in the Gypsy charm an ointment.  I take the freedom to reproduce that charm in full, with Leland’s introduction and translation \parencite[95]{Leland1891}:

\begin{small}“Before sunrise wolf’s milk (Wolfsmilch, rukeskro tçud) is collected, mixed with salt, garlic, and water, put into a pot, and boiled down to a brew. With a part of this the afflicted spot is rubbed, the rest is thrown into a brook, with the words:---

\begin{verse}
\emph{Kirmora jánen ándre tçud} \\
\emph{Andrál tçud, andré sir} \\
\emph{Andrál sir, andré páñi,} \\
\emph{Panensá kiyá dádeske,} \\
\emph{Kiyá Niváseske} \\
\emph{Pçándel tumen shelehá} \\
\emph{Eñávárdesh teñá!}
\end{verse}

\begin{verse}
‘Worms go in the milk, \\
From the milk into the garlic, \\
From the garlic into the water, \\
With the water to (your) father, \\
To the Nivasi, \\
He shall bind you with a rope, \\
Ninety-nine (yards long).’”
\end{verse}\end{small}

\sectionline

\subsection{Text}

\bvg\bva[]Gang út, \edtrans{\alst{N}esso}{Nesse}{\Bfootnote{An otherwise unknown dæmon, entitled \emph{Nessia} in the Latin title to the version of this spell (\textlink{ProNessia}).  The \emph{naming} of the dæmon or spirit which is to be excised is commonplace in ancient magic, including several other galders edited here.  The idea is that knowledge of the name of the entity gives the magician power over it.}}, \hld\ mid nigun \alst{n}essi-klínon, &
ut fana þemo marge an þat \alst{b}ên, \hld\ fan þemo \alst{b}êne an þat flęsg, &
ut fan þemo flęsgke an þia \alst{h}úd, \hld\ ut fan þera \alst{h}úd an þesa strála. &
Drohtin, werþe só.\eva

\bvb Go out, Nesse, with nine small Nesses! \\
Out from the marrow into the bone, from the bone into the flesh, \\
out from the flesh into the skin, out from the skin into this arrow. \\
Lord, may it be so.\evb\evg

\sectionline
\newpage

\section{For Nesse (\emph{Pro Nessia})}\chapterStart{}
\setBookCode{ProNessia}

\begin{flushright}%
\textbf{Dating:} 9th c.

\textbf{Meter:} \Fornyrdislag%para
\end{flushright}%

\subsection{Introduction}

\textbf{Against Nesse} (abbrev. \ProNessia) is an Old High German variant of the same spell as the Old Saxon \textlink{ContraVermes}.

\subsubsection{Preservation}
The galder is marginalia, found in the 9th c. ms. BSB Clm 18524b, on fol. 203v.

\subsubsection{Contents}
The language of \ProNessia\ is Bavarian, having undergone the second sound shift.  The wording is strikingly similar to the Old Saxon \textlink{ContraVermes}, which is likewise directed against the dæmon Nesse and his/her nine offspring.  The main difference between the two galders is in the use of the words describing the path Nesse is to take, but the core conception is identical: the dæmon is to be drawn out from the patient and into some external object.

\sectionline

\subsection{Text}

\bvg\bva Gang úz, \alst{N}esso, \hld\ mit niun \alst{n}essi-khịlínon, &
úz fonna marge in deo \alst{á}dra, \hld\ vonna den \alst{á}drun in daz flęisk, &
fonna demu \alst{f}lęiske in daz \alst{f}el, \hld\ fona demo \alst{v}elle in diz \edtrans{tulli}{quiver}{\Bfootnote{The translation is uncertain. TODO: look up.}}. &
Ter paternoster.\eva

\bvb Go out, Nesse, with nine small Nesses! \\
Out from the marrow into the veins, from the veins into the flesh, \\
from the flesh into the skin, from the skin into this quiver. \\
Thrice ‘pater noster’.\evb\evg
