\section{Against a Sudden Stitch (\emph{Wið fǽr-stice})}\chapterStart{}

\begin{flushright}%
\textbf{Dating:} ?

\textbf{Meter:} \Fornyrdislag%para
\end{flushright}%

Attested in \Lacnunga.

\sectionline

\bvg\bva \alst{H}lúde wǽran hý, lá, \alst{h}lúde, \hld\ þá hý ofer þone \alst{h}lǽw ridan, &
wǽran \alst{â}n-móde, \hld\ þá hý \alst{o}fer land ridan. &
Scyld þú þé nú, þú þysne \alst{n}íð \hld\ ge·\alst{n}esan móte. &
\alst{Ú}t, lýtel spere, \hld\ gif hér \alst{i}nne síe!\eva

\bvb Loud were they, lo, loud, when they rode over that mound; \\
they were steadfast, when they rode over land. \\
Shield thyself now; thou mayst escape this evil! \\
Out little spear, if here within it be!\evb\evg


\bvg\bva Stód under \alst{l}inde, \hld\ under \alst{l}eohtum scylde, &
þær þá \alst{m}ihtigan wíf \hld\ hýra \alst{m}ægen be·rǽddon &
and hý \alst{g}yllende \hld\ \alst{g}âras sændan; &
ic him \alst{ó}ðerne \hld\ \alst{e}ft wille sændan, &
\alst{f}léogende \alst{f}lâne \hld\ \alst{f}orane tó·géanes. &
\alst{Ú}t, lytel spere, \hld\ gif hit her \alst{i}nne sý!\eva

\bvb Stood under the linden \ken{shield}—under the light shield— \\
where those mighty wives their might arrayed, \\
and they yelling spears did send. \\
To them another [projectile] will I send back: \\
a flying arrow, aimed against [them]. \\
Out little spear, if here within it be!\evb\evg


\bvg\bva \alst{S}æt \alst{s}mið, \hld\ \alst{s}loh seax, &
lytel \alst{í}serna, \hld\ \alst{w}und swíðe. &
\alst{Ú}t, lytel spere, \hld\ gif her \alst{i}nne sý!\eva

\bvb Sat the smith, struck the sax: \\
a little iron-thing—a great wound. \\
Out little spear, if here within it be!\evb\evg


\bvg\bva \alst{S}yx \alst{s}miðas \alst{s}ætan, &
\alst{w}æl-spera \alst{w}orhtan. &
\alst{Ú}t, spere, \hld\ næs \alst{i}n, spere! &
Gif her \alst{i}nne sý \hld\ \alst{í}senes dǽl, &
\alst{h}æg-tessan ge·weorc, \hld\ \alst{h}it sceal ge·myltan.\eva

\bvb Six smiths sat, \\
wrought slaughter-spears. \\
Out, spear! Be not in, spear! \\
If here within be a part of iron, \\
the work of a \inx[C]{hag-tess}—\emph{it} shall melt!\evb\evg


\bvg\bva Gif þú wǽre on \alst{f}ell scoten \hld\ oððe wǽre on \alst{f}læsc scoten &
oððe wǽre on blód scoten \hld\ [...] &
oððe wǽre on \alst{l}ið scoten, \hld\ næfre ne sý þín \alst{l}íf atæsed;\eva

\bvb If thou wert shot in the skin, or wert shot in the flesh, \\
or wert shot in the blood, [...], \\
or wert shot in the limb—never be thy life injured.\evb\evg


\bvg\bva gif hit wǽre \alst{ė}sa ge·scot \hld\ oððe hit wǽre \alst{y}lfa ge·scot &
oððe hit wǽre \alst{h}æg-tessan ge·scot, \hld\ nú ic wille þín \alst{h}elpan: &
þis þé tó bóte \alst{ė}sa ge·scotes, \hld\ þis þé tó bóte \alst{y}lfa ge·scotes, &
þis þé tó bóte \alst{h}æg-tessan ge·scotes; \hld\ ic þín wille \alst{h}elpan.\eva

\bvb If it were Eese-shot, or it were Elf-shot,\footnoteB{Formulaic; see \inx[F]{Eese and Elves}. That they are held in the same category as the hag-tess—a witch—indicates Christian influence. Among the Germanic peoples the elves and Eese were originally beneficial, as seen by numerous names like Alfred (OE \emph{Ęlf-réd} ‘Elf-counsel’), Oswald (OE \emph{Ós-weald} ‘Os-power’), Elfwin (Lomb. \emph{Alb-oin} ‘Elf-friend’), Oshelm (Lomb. \emph{Anselm} ‘Os-helmet’).}  \\
or it were Hag-tess-shot—now I will help thee! \\
This for thee as cure against Eese-shot; this for thee as cure against Elf-shot;  \\
this for thee as cure against Hag-tess-shot—I will help thee!\evb\evg


\bvg\bva \alst{F}leo þær on \hld\ \alst{f}yrgen-hæfde! &
\alst{H}âl wes-tu, \hld\ \alst{h}elpe þín drihten! &
Nim þonne þæt seax, \hld\ ado on wætan.\eva

\bvb TODO. \\
Be thou hale, may the Lord help thee.\evb\evg

\sectionline
