\section{Against a Dwarf (\emph{Wið dweorh})}\chapterStart{}
\setBookCode{WidDweorh}

\begin{flushright}%
\textbf{Dating:} TODO

\textbf{Meter:} \Fornyrdislag%
\end{flushright}

\subsection{Introduction}

TODO: Introduction.

\sectionline

\subsection{Text}

\bpg\bpa Mann sceal niman \emph{seofon} lytle of-lætan swylce mann mid ofrað, ond wrítan þás naman on ælcre oflætan: Maximianus, Malchus, Johannes, Martinianus, Dionisius, Constantinus, Serafion.  Þænne eft þæt galdor þæt hér æfter cweð[eð] mann sceal singan, ærest on þæt wynstre éare, þænne on þæt swíðre éare, þænne búfan þæs mannes moldan; ond gá þænne ân mæden-mann tó, ond hó hit ǫn his sweoran, ond dó mann swá þrý dagas.  Him bið sóna sél.\epa

\bpb One shall take seven small wafers, such as one offers [during the Mass], and write these names on each wafer: Maximianus, Malchus, Johannes, Martinianus, Dionysius, Constantinus, Seraphion.  After that shall one sing this galder which is henceforth said; first into the left ear, then into the right ear, then over the man’s head; and thereafter a maiden go forth, and hang it on his neck; and one do so for three days.  He will soon be well.\epb\epg


\bvg\bva%
Hér cóm \alst{i}n·gangan \hld\ \alst{i}n·spiden wiht, &
hæfde him his \alst{h}aman ǫn \alst{h}anda; \hld\ cwæð þæt þú his \alst{h}æncgest wǽre, &
\alst{l}ęgeþe þé his téage \emph{ǫ}n sweoran; \hld\ ǫn·gunnan him ǫf þæm \alst{l}ande líðan. &
Sóna swá hý ǫf þæm \alst{l}ande cóman \hld\ þá ǫn·gunnan him þá \emph{\alst{l}eomu} cólian.— &
Þá cóm in·gangan \hld\ déores sweostar; &
þá ge·\alst{æ}ndode héo \hld\ ond \alst{â}ðas swór, &
þæt næfre þis þæm \alst{a}dlegan \hld\ \emph{\alst{e}gl}ian ne móste &
né þæm þe þis \alst{g}aldor \hld\ be·\alst{g}ýtan mihte &
oððe þe þis \alst{g}aldor \hld\ on·\alst{g}alan cu̇ðe. &
Amen fiað.\eva

\bvb Here an inspiden wight came walking in, \\
had his harness in his hands, said that thou wert his horse, \\
laid his reins on thy neck; they began to ride away from the land. \\
As soon as they came away from the land then they began to cool limbs. \\
Then the beast’s sister came walking in; \\
then she made an end to it and swore oaths \\
that this never should torment the ailing man, \\
nor him who this galder might get, \\
nor whomever this galder could gale. \\
Amen, let it be.\evb\evg

\sectionline
