\part{Galders (INCOMPLETE!)}

\bookStart{Introduction to Galders}

The Germanic word “galder” (PGmc. \emph{*galdrą}) is a verbal noun derived from the old verb “gale” (PGmc. \emph{*galaną}), meaning ‘to sing, to chant’.  A “galder” is therefore what is chanted or sung, more specifically a magical chant, whether for the purpose of healing or cursing.

In the present section the reader will find gathered a number of such \inx[C]{galder}[galders] attested across the Old Germanic languages.  Not all of these texts may be true galders, but they share a magical purpose and many formulae and stylistic features reoccur throughout.

Note that this section will only include written charms containing likely or certain pre-Christian tradition elements, e.g.~the invocation of the Heathen Gods or the use of old poetic formulae.  Syncretic texts which mix traditional and Christian elements (e.g.~DR EM85;493, most of the OE galders) are also included.  Thoroughly Christian prayers, however, even if metrical, are excluded from this section and may instead be found under “Poetry on Christian Subjects”.


\chapter{Continental Germanic galders}

\input{books/galders/Merseburg.tex}
\newpage
\input{books/galders/Against Worms.tex}


\bookStart{Introduction to Old English Galders}

There survive twelve OE metrical charms, although it is questionable whether they all qualify as galders.  Notably for these texts is that they are preserved as the main text in manuscripts rather than marginalia, and that they have substantial ritual instructions alongside the words of the galders themselves.

The first four, \textlink{NineHerbs}, \textlink{WidDweorh}, \textlink{WidFerstice}, and “Infants”, are from the text known as the \emph{Lâcnunga} (abbrev. \Lacnunga) in the manuscript Harley MS 585 (ca.~1000 \ce, siglum \LacnungaMS).  One galder from \Lacnunga, “For Loss of Cattle” (A), on fol.~180v--181r, is excluded due to being entirely Christian.

The following two, “For Theft of Cattle” and \textlink{WidYmbe}, both belong to Corpus Christi College, Cambridge MS 41 (abbrev.~CCCC 41).  Two other galders in CCCC 41, “For Loss of Cattle” (B), a variant of (A) same-named galder found in \Lacnunga, and “A Journey Charm” are thoroughly Christian and therefore excluded.

The last three galders, “Against Wen”, “Against Water-Elf-Disease”, and \textlink{Acerbot}, are from various sources which will be discussed for each individual galder.

\input{books/galders/Nine Herbs.tex}
\newpage
\input{books/galders/Against Dwarf.tex}
\newpage
\input{books/galders/Against Fearstitch.tex}
\newpage
\input{books/galders/Infants.tex}
\newpage
\input{books/galders/Cattle.tex}
\newpage
\input{books/galders/Against Swarm.tex}
\newpage
\input{books/galders/Against Wen.tex}
\newpage
\input{books/galders/Against Water-Elf.tex}
\newpage
\input{books/galders/Acreboot.tex}


\chapter{Old Norse galders}

\input{books/galders/Ribe Galder.tex}
\newpage
\input{books/galders/Canterbury galder.tex}
\newpage
\input{books/galders/Sigtuna rib.tex}
\newpage
\input{books/galders/Sigtuna Plate I.tex}
\newpage
\input{books/galders/Bryggen.tex}


%\section{Runic plates}
