\bookStart{The Merseburg Galders}

\begin{flushright}%
\textbf{Dating:} C9th--10th

\textbf{Meter:} \Fornyrdislag, \Galdralag%
\end{flushright}

\section{Introduction}

The two \textbf{Merseburg galders} are some of the only surviving examples of genuine Heathen galders from continental West Germanic Europe, and the only ones to actually invoke the pre-Christian Germanic gods.

\subsection{Preservation}

The galders are preserved on fol. 85r of the theological manuscript Merseburger Domstiftsbibliothek Codex 136 (in the present critical edition given the signum \textbf{A}) where they are written in a neat and well-trained hand.  Immediately below them follows a thoroughly Christian prayer written in the same hand \parencite[360]{Murdoch1988}, edited below as an Appendix to the galders.  The theological contents of the manuscript, the trained hand and the Latin prayer together suggest that the scribe was most likely a monk.

\subsection{Summary}

Both galders share a common two-part structure, each beginning with an \emph{historiola}, a mythological account describing the successful effects of the galder in the distant mythic past, followed by an imperative \emph{incantatio} or the galder-proper, commanding that the willed magic effect take place in the present.

The first galder (\MerseburgOne) begins with the historiola describing a group of supernatural women in the midst of a battle, affecting its outcome by fastening or loosening fetters.  The incantatio then commands that the present fetters be destroyed so that captives may escape the enemy.

The second galder (\MerseburgTwo) begins with the historiola describing a group of Gods riding through the woods.  Among them is \inx[P]{Balder}, whose young foal sprains its foot.  Three Gods---the otherwise unknown goddess \inx[P]{Sithguth}, the goddess \inx[P]{Sun}, the god \inx[P]{Weden}---in turn chant a healing galder over it.  The incantatio---apparently the galder sung by Weden---then commands that a present sprain be healed.

\subsection{Cultural context}

I generally agree with \textcite{Murdoch1988} when he convincingly argues that the galders ought to be seen as incantations written down to be sung alongside the application of salves or other instruments in the context of Christian(ized) medieval medicine.  This is important for the context of \MerseburgOne, since the preservation of a galder for actually removing shackles is quite unlikely.  For one its effectivity must have been quite limited, and it is very unlikely that it would have been of any use in the monastic medicinal context.  Considering that likely context a more reasonable use case is for a cramp, which would then be symbolically likened within the sphere of the galder to the fetters on the battlefield.

As mentioned above, the two galders are followed by a Latin liturgical formula written in the same hand.  As \textcite[362--363]{Murdoch1988} shows, the same formula is also found in a cure against possession in the Anglo-Saxon Bald’s Leechbook, and this firmly anchors the galders within the Christian medicinal framework.
Indeed, when one compares the Merseburg galders with OE galders like \NineHerbs\ and Norse galders like the Ribe galder stick (DR EM85;493) it becomes clear that the mixing of pagan and Christian elements in this type of text was not unusual.  Monk or not, the traditional invocation of Weden must have been felt to have a particularly strong healing effect, and it was only later that he was entirely excised and demonised.

Of course this does not mean that the galders themselves are fully Christian, like \textcite{Murdoch1988} suggests.  The gods they invoke are Germanic and pre-Christian, as is their alliterative poetic form and their contents (especially \MerseburgTwo).  As \textcite[360--361]{Murdoch1988} himself brings up, contemporary mediæval Christian authors condemn pagan charms; and how could a galder invoking Weden himself not be considered pagan when the god is even mentioned by name in the contemporary Old Saxon Baptismal Vow (edited under Christian Poetry)?  In any case, \citeauthor{Murdoch} rightly makes an important distinction: regardless of whether the galders themselves are pagan or not, they are preserved in a Christian manuscript and in that context their healing magic appears to function alongside a Latin prayer.

\subsection{Parallels}

Moving away from the immediate cultural context we ought to look at parallels for both galders.  These will help us anchor them in a broader European tradition of magical song or galder.

Close verbal parallels to \MerseburgOne\ are found in the Norse \textlink{Havamal}[150] and \textlink{Grougaldr}[10], but those are literary monuments.  As discussed above it is questionable whether it was ever actually meant to be used by shackled captives, and to my knowledge no actual galders to that effect survive.

\MerseburgTwo\ has an impressive number of later parallels across Celtic and Germanic-speaking Northern Europe (TODO: cite Hopkins: Merseburg echoes).  These charms generally mirror both the historiola and the incantatio of \MerseburgTwo, but almost always replace the healer with Jesus Christ.  They are generally used to heal sprains or broken bones.  Many are found as incantations to be recited together with the sign of the cross.  Further afield the incantatio has a notable parallel in the Vedic material, for which see ll. 6–9 n.


\sectionline

\subsection{Text}

\setBookCode{MerseburgOne}
\bvg\bva%
\mssnote{\textbf{A} 85r/1--4}%
Ęiris \alst{s}ázun idisi \hld\ \alst{s}ázun hera duo der; &
suma \alst{h}apt \alst{h}ęptidun \hld\ suma \alst{h}ęri lęzidun &
suma \alst{k}lubodun \hld\ umbi \edtrans{\alst{k}uonjo-widi}{chains}{\Bfootnote{A rare word (hapax? TODO) most likely cognate with Gothic \emph{kuna-wida} ‘fetter; \textgreek{ἅλῠσῐς}’ \parencite[76]{Streitberg}.  Note that \emph{kuonjo-} and \emph{kuna-} are not formally identical but probably reflect different ablaut grades.}} &
\alst{i}n-sprink hapt-bandun \hld\ \alst{i}n-var vígandun &
\edtext{.H.}{\Bfootnote{The meaning of this letter, which is very clearly visible and written in the same hand as the galders, is uncertain.  To me the most convincing suggestion is that it is to be read as \emph{.N.}, a common abbreviation for Latin \emph{nomen} ‘name’ (cf. the Appendix below), and here presumably the name of the patient for whom the healer is chanting \parencite[363]{Murdoch1988}.}}\eva

\bvb {\huge O}\textsc{f yore sat dises}, sat here, then there; \\
some fastened fetters, some hindered armies, \\
some cut chains of captives asunder.--- \\
Destroy the fetter-bonds, lead the way from the foes! \\
.H.\evb\evg

\setBookCode{MerseburgTwo}
\bvg\bva%
\mssnote{\textbf{A} 85r/5--12}%
\alst{Ph}ol ęnde Wuodan \hld\ \alst{v}uorun zi holza &
dú wart demo Balderes \alst{v}olon \hld\ sín \alst{v}uoz bi·ręnkit &
þú \edtrans{bi·guol en}{galed over it}{\Bfootnote{Sang a \inx[C]{galder} over the horse; \emph{bi·guol} the 3rd sg. past ind. of \emph{bi·galan} ‘sing (a galder) over, begale’, the transitive of \emph{galan} ‘gale, sing (a galder)’.  Cf. \textlink{Havamal}[153]/4, \textlink{Oddrunargratr}[7]/3--4.}} \alst{S}inhtgunt \hld\ \alst{S}unna era swister &
þú bi·guol en \alst{F}rija \hld\ \alst{V}olla era swister &
þú bi·guol en \alst{W}uodan \hld\ só hé \alst{w}ola konda: &
\edtext{„Só-se \alst{b}ên-ręnkí \hld\ só-se \alst{b}luot-ręnkí \hld\ só-se lidi-ręnkí &
\ind \alst{b}ên zi \alst{b}êna &
\ind \alst{b}luot zi \alst{b}luoda &
\alst{l}id zi ge·\alst{l}iden \hld\ só-se ge·\alst{l}ímida sín!}{\lemma{Só-se \dots\ sín! ‘Like \dots\ together!’}\Bfootnote{As discussed above the incantatio has numerous Celtic and Germanic parallels.  The oldest by far is however \Atharvaveda\ 4.12.3--5 (ca. 1200--900 \bce), used alongside an herb to heal a broken bone: “Let your marrow be united with marrow and [let] your joint [be] united with joint. Let your torn [piece] of flesh and [your] bone grow together. // Let the marrow be united with marrow; let the skin grow with skin; let your blood grow with blood [and] let flesh grow with flesh. // You, O herb, make hair join with hair, make skin join with skin. Unite what is severed. Let grow forth with bone.” \parencite[72--74]{Zysk1985}}}“\eva

\bvb {\huge F}\textsc{ol and Weden} journeyed through the woods; \\
then was the foot of Balder’s foal sprained. \\
Then \inx[P]{Sithguth} galed over it---\inx[P]{Sun} her sister; \\
then \inx[P]{Frie} galed over it---\inx[P]{Full} her sister; \\
then Weden galed over it as he could well do: \\
“Like bone-sprain, like blood-sprain, like joint-sprain! \\
\ind Bone to bone, \\
\ind blood to blood, \\
joint to joints, like were they glued together!”\evb\evg

\subsection{Appendix}

Immediately below the galders the following Latin prayer is written.  Abbreviations have been resolved according to \textcite[369]{Murdoch1988}.  The prayer is closely related to the Anglican \emph{Prayer for the Clergy and People} as found in the Book of Common Prayer \parencite[58]{ChurchOfEngland1892}, which the present translation aims to follow where possible.

\bpg\bpa%
Omnipotens sancte pater noster deus, qui facis mirabilia magna solus, pretende super \edtrans{famulum tuum N.}{thy servant N.}{\Bfootnote{N.~abbreviating \emph{nomen} ‘name’, viz. the subject of the blessing.  The singular “thy servant” appears to have taken place of an earlier plural subject, for several following words (\emph{illis} ‘(to) their (charge)’ etc.) are still written in the plural.  This hints that the German scribe, in spite of his pleasing hand, may not have been fully competent in Latin.}}~et super cunctas congregationes illis commissas spiritum gratie salutaris, et ut in veritate tibi conplaceant perpetuum eis rorem tue benedictionis infundo.  Per dominum nostrum Jesum Christum.\epa

\bpb Almighty, holy Father, our God, who alone workest great marvels: send down upon thy servant N., and upon all the congregations committed to their charge, the spirit of saving grace; and, that they may truly please thee, pour upon them the continual dew of thy blessing.  Through our Lord Jesus Christ.\epb\epg

\sectionline
