\bookStart{The Galder of Nine Worts}[Nigon wyrta galdor]
\setBookCode{NineHerbs}

\begin{flushright}%
\textbf{Dating:} ?

\textbf{Meter:} \Fornyrdislag%para
\end{flushright}%

\section{Introduction}

TODO: introduction

Numerous lines begin with small crosses which appear to serve as section markers.  The present division into stanzas is largely based on these.  TODO: note which.

Preserved in \Lacnunga.

\sectionline

\section{Text}

\bvg\bva%
Ge·\alst{m}yne ðú \alst{m}ug-wyrt \hld\ hwæt þú á·\alst{m}eldodest &
hwæt þu \alst{r}enadest \hld\ æt \alst{R}egen-melde?\eva

\bvb Rememberest thou, Mugwort, what thou didst declare, \\
what thou didst arrange at Reinmeld?\evb\evg


\bvg\bva \alst{U}na þú hâttest \hld\ \alst{y}ldost wyrta &
þú miht \edtrans{wið \emph{\alst{þ}rim} \hld\ and wið \emph{\alst{þ}rí·tigum}}{against three and against thirty}{\Bfootnote{I.e. ‘against a great number of foes’.  ‘Three and thirty’ is formulaic and appears in many later English and Scottish folk-songs, viz. Child 4EFG, 18B, 20C, 30, 53BCDEIKM, 63EFH, 73I, 97AC, 100AG, 110BGH, 156G, 185A, 187A, 187C, 190A, 192A, 193B, 203C, 211A, 217GHLN, 244A, 268A, 269C, 281ABC.  Things described include horses, heads of cattle, warriors, days, years, winters.}} &
þú miht wiþ \alst{a}ttre \hld\ and wið \alst{ǫ}n·flyge &
þú miht wiþ þâm \alst{l}âþan \hld\ ðe geond \alst{l}ǫnd færð.\eva

\bvb Un thou art called, the oldest of worts; \\
thou availest against three and against thirty; \\
thou availest against the venom and against the onflier; \\
thou availest against the loathsome one that journeys through the lands.\evb\evg


\bvg\bva Ond þú \alst{w}eg·bráde \hld\ \alst{w}yrta módor &%NOTE: introduced with a + sign
\alst{éa}stan \alst{o}pene \hld\ \alst{i}nnan mihtigu &
ofer ðy \alst{c}ræte \alst{c}urron \hld\ ofer ðy \alst{c}wéne reodon &
\ind ofer ðy \alst{b}rýde \alst{b}rýodedon &
\ind ofer ðy \alst{f}earras \alst{f}nærdon.\eva

\bvb And thou, Waybroad, mother of worts, \\
open from the east, mighty within. \\
Over thee TODO. \\
\ind TODO \\
\ind TODO\evb\evg


\bvg\bva Eallum þu þon wið·\alst{st}óde \hld\ and wið·\alst{st}unedest &
swá ðú wið·stonde \alst{a}ttre \hld\ and \alst{ǫ}n·flyge &
and þǽm \alst{l}âðan \hld\ þe geond \alst{l}ǫnd fereð.\eva

\bvb Them all didst thou then withstand, and didst stop; \\
so mayst thou withstand the venom and the onflier, \\
and the loathsome one that journeys through the lands.\evb\evg


\bvg\bva \alst{St}une hætte þéos wyrt, \hld\ héo on \alst{st}âne ge·weox &
\alst{st}ǫnd héo wið attre, \hld\ \alst{st}unað héo wærce &
\alst{St}iðe héo hatte, \hld\ wið·\alst{st}unað héo attre &
\alst{w}receð héo \alst{w}râðan, \hld\ \alst{w}eorpeð út attor.\eva

\bvb Stun is this wort called, she grew on stone; \\
she withstands venom, she stops aches. \\
Stithe is she called, she stops the venom; \\
she drives away the wroth one, casts out the venom.\evb\evg


\bvg\bva Þis is séo \alst{w}yrt \hld\ séo wiþ \alst{w}yrm ge·feaht &
þéos mæg wið \alst{a}ttre, \hld\ héo mæg wið \alst{ǫ}n·flyge; &
héo mæg wið ðâm \alst{l}âþan \hld\ ðe geond \alst{l}ǫnd fereþ.\eva

\bvb This is the wort that fought against the Wyrm; \\
this one avails against the venom, she avails against the onflier; \\
she avails against the loathsome one that journeys through the lands.\evb\evg


\bvg\bva Fleoh þú nú attor-\alst{l}âðe, \hld\ séo \alst{l}ǽsse ðá mâran &
séo mâre þá lǽssan, \hld\ oððæt him beigra bót sý!\eva

\bvb TODO \\
TODO.\evb\evg


\bvg\bva Ge·\alst{m}yne þú, \alst{m}ægðe, \hld\ hwæt þú á·\alst{m}eldodest &
hwæt ðú ge·\alst{æ}ndadest \hld\ æt \alst{A}lor-forda &
þæt nǽfre for ge·\alst{f}loge \hld\ \alst{f}eorh ne ge·sealde &
syþðan him \alst{m}ǫn \alst{m}ægðan \hld\ tú \alst{m}ęte ge·gyrede.\eva

\bvb TODO \\
TODO \\
TODO \\
TODO\evb\evg


\bvg\bva Þis is séo \alst{w}yrt \hld\ ðe \alst{w}er-gulu hatte &
ðás on·\alst{s}ænde \alst{s}eolh \hld\ ofer \alst{s}ǽs hrygc &
\alst{o}ndan \alst{a}ttres \hld\ \alst{ȯ}þres tó bóte.\eva

\bvb TODO \\
TODO \\
TODO\evb\evg


\bvg\bva Ðás \emph{\alst{n}ygon} magon \hld\ wið \alst{n}ygon attrum.\eva

\bvb These nine avail against nine venoms.\evb\evg


\bvg\bva Wyrm cóm \alst{s}nícan, \hld\ to·\alst{s}lât hé \emph{m}an &
ðá ge·nam \alst{W}óden \hld\ \emph{\alst{n}ygon} \alst{w}uldor-tânas &
slóh ðá þá \alst{n}ǽddran \hld\ þæt héo on \emph{\alst{n}ygon} tó·fléah &
Þǽr ge·\alst{æ}ndade \alst{æ}ppel \hld\ and \alst{a}ttor &
þæt héo nǽfre ne wolde \hld\ ǫn hús búgan.\eva

\bvb A \inx[C]{Wyrm} came crawling; he tore apart a man. \\
Then took Weden nine glory-twigs, \\
slew then that adder, that it sprung into nine [parts]. \\
There ended apple and venom, \\
that she would never wish to enter a house.\evb\evg


\bvg\bva \alst{F}ille and \alst{f}inule, \hld\ \alst{f}ela-mihtigu twá &
þá \alst{w}yrte ge·sceop \hld\ \alst{w}ítig drihten &
\alst{h}âlig on \alst{h}eofonum, \hld\ þá hé \alst{h}ǫngode; &
\alst{s}ętte and \alst{s}ęnde \hld\ on \emph{\alst{s}eofon} worulde &
\alst{ea}rmum and \alst{éa}digum \hld\ \alst{ea}llum tó bóte.\eva

\bvb Fill and Fennel, the many-mighty two; \\
those worts the wise lord shaped, \\
holy in heaven when he hung. \\
He set and sent them into seven worlds, \\
for wretched men and for wealthy, for all men as a cure.\evb\evg


\bvg\bva \alst{St}ǫnd héo wið wærce, \hld\ \alst{st}unað héo wið attre &
séo mæg wið \emph{\alst{þ}rim} \hld\ \emph{and} wið \emph{\alst{þ}rí·tigum} &
wið \emph{\alst{f}éondes} hǫnd \hld\ and wið \alst{f}ǽr-bregde &
wið \alst{m}alscrunge \hld\ \alst{m}ânra wihta.\eva

\bvb She stands against ache, she stands against venom;
she avails against three and against thirty;
against the TODO of evil wights.\evb\evg


\bvg\bva Nu magon þás \emph{nygon} \alst{w}yrta \hld\ wið nygon \alst{w}uldor-ge·flogenum &
wið \emph{nygon} \alst{a}ttrum \hld\ and wið nygon \alst{o}n·flygnum &
wið ðý \alst{r}éadan attre, \hld\ wið ðý \alst{r}unlan attre &
wið ðý \alst{h}witan attre, \hld\ wið ðý \emph{\alst{h}æwe}nan attre &
wið ðý \alst{g}eolwan attre, \hld\ wið ðý \alst{g}rénan attre &
wið ðý \alst{w}onnan attre, \hld\ wið ðý \alst{w}edenan attre &
wið ðý \alst{b}rúnan attre, \hld\ wið ðý \alst{b}asewan attre &
wið \alst{w}yrm-ge·blæd, \hld\ wið \alst{w}æter-ge·blæd &
wið \alst{þ}orn-ge·blæd, \hld\ wið \alst{þ}ystel-ge·blæd &
wið \alst{ý}s-ge·blæd, \hld\ wið \alst{a}ttor-ge·blæd\eva

\bvb Now these nine worts avail against glory-onfliers: \\
against nine venoms and against nine onfliers; \\
against the red venom; against the TODO venom; \\
against the white venom; against the TODO venom; \\
against the yellow venom; against the green venom; \\
against the TODO venom; against the TODO venom; \\
against the brown venom; against the TODO venom; \\
against worm-TODO; against water-TODO; \\
against thorn-TODO; against thistle-TODO; \\
against ice-TODO; against venom-TODO.\evb\evg


\bvg\bva Gif ænig \alst{a}ttor cume \hld\ \alst{éa}stan fleógan &
oððe ǽnig norðan cume &
oððe ǽnig \alst{w}estan \hld\ ofer \alst{w}er-þeóde\eva

\bvb If any venom should come flying from the east; \\
or any come from the north; \\
or any from the west, over mankind.\evb\evg


\bvg\bva Críst stód ofer \alst{á}dle \hld\ \alst{ǽ}ngan cundes. &
Ic \alst{â}na wât \hld\ \alst{éa} rinnende &
þǽr þá \alst{n}ygon \alst{n}ǽdran \hld\ \alst{n}éan be·healdað.\eva

\bvb Christ stood over TODO; \\
I know one river running, \\
where the nine adders TODO.\evb\evg


\bvg\bva Motan ealle \alst{w}éoda \hld\ nu \alst{w}yrtum á·springan &
\alst{s}ǽs tó·\alst{s}lúpan, \hld\ eal \alst{s}ealt wæter &
ðonne ic \alst{þ}is attor \hld\ of \alst{þ}é ge·bláwe.\eva

\bvb TODO\evb\evg


\bpg\bpa Mucgwyrt, weg-brade þe eastan open sy, lombes-cyrse, attor-laðan, mageðan, netelan, wudu-sur-æppel, fille and finul, ealde sapan. Ge·wyrc ða wyrta to duste, mængc wiþ þa sapan and wiþ þæs æpples gor. Wyrc slypan of wætere and of axsan, ge·nim finol, wyl on þære slyppan and beþe mid æggemongc, þonne he þa sealfe on do, ge ær ge æfter. Sing þæt galdor on æcre þara wyrta, :III: ær he hy wyrce and on þone æppel eal-swa; ond singe þon męn in þǫne mu̇ð and in þá éaran búta and on ðá wunde þæt ilce gealdor, ær he þá sealfe on dó.\epa

\bpb TODO.\epb\epg

\sectionline
