\book{The Speech of Web-strengthener. (Vafþrúðnismǫ́l)}

(Óðinn kvað:) &
\bv1 Ráð mér nú \alst{F}rigg \hld alls mik \alst{f}ara tíðir &
at \alst{v}itja \alst{V}afþrúðnis; &
\alst{f}orvitni mikla \hld kveð'k mér á \alst{f}ornum stǫfum &
við þann hinn \alst{a}lsvinna \alst{jǫ}tun.\\ \&

\bv2 \textbf{Woden} quoth: \\ “Counsel me now, \textbf{Frie}, as I desire to travel to visit \textbf{Web-strengthener}; I have great curiosity to say ancient staves\footnotemark[1] against that all-wise \textbf{ettin}." \\
\footnotetext[1]{Ancient (pieces of) lore; cf. v. 55.} 

(Frigg kvað:) &
\bv1 \alst{H}ęima lętja \hld mynda'k \alst{H}ęrjafǫðr &
í \alst{g}ǫrðum \alst{g}oða; &
\alst{ę}ngi \alst{jǫ}tun \hld hugða'k \alst{ja}fnramman &
sęm \alst{V}afþrúðni \alst{v}esa.\\ \&

\bv2 Frie quoth: \\ “I would encourage the \textbf{Leader of Armies} to [stay at] home in the yards of the gods, for I've judged no ettin be as strong as\footnotemark[3] Web-strengthener." \\
\footnotetext[3]{Lit. 'equal-strong'.}

(Óðinn kvað:) &
\bv1 Fjǫlð ek fór, \hld fjǫlð fręistaða'k, &
fjǫlð ek ręynda ręgin; &
hitt vil'k vita, \hld hvé Vafþrúðnis &
salakynni séi.\\ \&

\bv2 Woden quoth: \\ “Much I travelled, much I tried, much I tested the \textbf{Powers}\footnotemark[4]. \emph{This} I want to know, how the condition of the halls of Web-strengthener might be?" \\
\footnotetext[4]{The gods.}

(Frigg kvað:) &
\bv1 Hęill þú farir, \hld hęill þú aptr komir, &
hęill á sinnum séir; &
ǿði þér dugi \hld hvar's skalt, Aldafǫðr, &
orðum mæla jǫtun.\\ \&

\bv2 Frie quoth: \\ “Whole may thou travel, whole may thou return, whole may thou be on thy paths! May, thy wisdom suffice, \textbf{Leader of Men}, when thou go to exchange words with the ettin." \\

\bv1 Fór þá Óðinn \hld at fręista orðspęki &
þess hins alsvinna jǫtuns; &
at hǫllu hann kom, \hld es\footnotemark[1] átti Íms faðir; &
inn gekk Yggr þegar.\\ \&
\footnotetext[1]{Ms. \emph{ok} corrected to \emph{es}. Alliteration is lacking in this line, for which reason FJ emends \emph{Íms} to \emph{Hymis}.}

\bv2 Then went Woden, to try the word-wisdom of that all-wise ettin. To the hall he came, which the father of \textbf{Ime}\footnotemark[5] owned; shortly the \textbf{Frightener}\footnotemark[6] walked in. \\
\footnotetext[5]{Web-strengthener.}
\footnotetext[6]{Woden.}

(Óðinn kvað:) &
\bv1 Hęill þú nú, Vafþrúðnir, \hld nú em'k í hǫll kominn &
á þik sjalfan séa; &
hitt vilk fyrst vita, \hld ef fróðr séir &
eða alsviðr, jǫtunn.\\ \&

\bv2 Woden quoth: \\ “Hail thee now, Web-strengthener; now I have come into the hall, to see thee thyself. \emph{This} I want to know first, if knowing thou might be, or all-wise, ettin!" \\

(Vafþrúðnir kvað:) &
\bv1 Hvat's þat manna, \hld es í mínum sal &
verpumk orði á? &
út þú né kømr \hld órum hǫllum frá. &
nema þú inn snotrari séir.\\ \&

\bv2 Web-strengthener quoth: \\ “What is that of men\footnotemark[10], that in \emph{my} hall throws words at me? Thou will not come \emph{out}, from \emph{our}\footnotemark[11] halls, unless thou be the wiser [of us two]." \\
\footnotetext[10]{Ie., ‘what man is that'. The use of the neuter pronoun \emph{hvat} by Web-str. may be seen as an insult or a way of belittling the guest.}
\footnotetext[11]{Prob. again meaning 'my', unless Web-str. has allies present in the hall, but no such indication is given.}

(Óðinn kvað:) &
\bv1 Gagnráðr\footnotemark[5] hęiti'k, \hld nú em'k af gǫngu kominn, &
þyrstr til þinna sala; &
laðar þurfi \hld hęf'k lęngi farit &
ok þinna andfanga, jǫtunn.\\ \&
\footnotetext[5]{R's \emph{Gagnráðr} ‘Gain-adviser', is attested as Gangráðr ‘Journey-adviser' in \emph{Gylf}.}

\bv2 Woden quoth: \\ “\textbf{Gain-adviser} I am called, I am come from the journey, thirsty to thy halls. I have travelled for a long time in need of hospitality, and of thy reception, ettin!" \\

(Vafþrúðnir kvað:) &
\bv1 Hví þú þá, Gagnráðr, \hld mælisk af golfi fyrir? &
far þú í sess í sal; &
þá skal fręista, \hld hvárr flęira viti, &
gęstr eða hinn gamli þulr.\\ \&

\bv2 Web-strengthener quoth: \\ “Why then, Gain-adviser, art thou speaking from the floor before [me]? Take a seat in the hall! Then it shall be proven, which of the two might know more; the guest, or the old \textbf{thyle}." \\

(Gagnráðr kvað:) &
\bv1 Óauðigr maðr, \hld es til auðigs kømr, &
mæli þarft eða þęgi; &
ofrmælgi mikil \hld hygg at illa geti &
hvęim's við kaldrifjaðan kømr.\\ \&

\bv2 Gain-adviser quoth: \\ “An unwealthy man, who comes to a wealthy [one], ought to speak what is needed, or be silent.\footnotemark[14] Much over-speaking\footnotemark[15], I judge, will be bad for the one who comes to a cold-ribbed\footnotemark[16] [man]." \\
\footnotetext[14]{Line identical to \emph{High} 18/2. The whole verse strongly reminds of verses from the \emph{Guest-thread} portion of said poem.}
\footnotetext[15]{“Speaking too much".}
\footnotetext[16]{That is, ‘cold-hearted', ‘cunning'.}

(Vafþrúðnir kvað:) &
\bv1 Sęg mér, Gagnráðr, \hld alls á golfi vill &
þíns of fręista frama, &
hvé hęstr hęitir, \hld sá's hvęrjan dręgr &
dag of dróttmǫgu.\\ \&

\bv2 Web-strengthener quoth: \\ “Say to me, Gain-adviser, since on the floor I will to try thy fame: What is the horse called, which drags each \emph{day} above the sons of the retinue\footnotemark[20]?" \\
\footnotetext[20]{Kenning for ‘men', ‘mankind'.}

(Gagnráðr kvað:) &
\bv1 Skinfaxi hęitir, \hld es hinn skíra dręgr &
dag of dróttmǫgu; &
hęsta baztr \hld þykkir með Hręiðgotum; &
ęy lýsir mǫn af mari.\\ \&

\bv2 Gain-adviser quoth: \\ “\textbf{Shining-fax} [that one] is called, who drags the bright day above the sons of the retinue. The best of horses he seems among the \textbf{Rode-goths}; the mane of that stallion ever shines." \\

(Vafþrúðnir kvað:) &
\bv1 Sęg þat, Gagnráðr, \hld alls á golfi vill &
þíns of fręista frama, &
hvé jór hęitir, \hld sá's austan dręgr &
nótt of nýt ręgin.\\ \&

\bv2 Web-strengthener quoth: \\ “Say this, Gain-adviser, since on the floor I will to try thy fame: What is the horse called, which from the east drags night above the useful \textbf{Powers}?" \\

(Gagnráðr kvað:) &
\bv1 Hrímfaxi hęitir, \hld es hvęrja dręgr &
nótt of nýt ręgin; &
méldropa fęllir \hld morgin hvęrjan; &
þaðan kømr dǫgg of dala.\\ \&

\bv2 Gain-adviser quoth: \\ “\textbf{Frost-fax} [that one] is called, who drags each night above the useful Powers. Every morning he lets foam fall from his bit\footnotemark[26]; thence comes dew in the valleys." \\
\footnotetext[26]{Lit. “he fells bit-drops".}

(Vafþrúðnir kvað:) &
\bv1 Sęg þat, Gagnráðr, \hld alls á golfi vill &
þíns of fręista frama, &
hvé ǫ́ hęitir, \hld sú's dęilir með jǫtna sonum &
grund ok með goðum.\\ \&

\bv2 Web-strengthener quoth: \\ “Say this, Gain-adviser, since on the floor I will to try thy fame; How the river is called, which divides the ground between the sons of ettins and the gods?" \\

(Gagnráðr kvað:) &
\bv1 Ífing hęitir ǫ́, \hld es dęilir með jǫtna sonum &
grund ok með goðum; &
opin rinna \hld hón skal um aldrdaga; &
verðr-at íss á ǫ́.\\ \&

\bv2 Gain-adviser quoth: \\ “\textbf{Iving} the river is called, which divides the ground between the sons of ettins and the gods. Throughout [her] life-days she shall flow open; ice forms not on the river." \\

(Vafþrúðnir kvað:) &
\bv1 Sęg þat, Gagnráðr, \hld alls á golfi vill &
þíns of fręista frama, &
hvé vǫllr hęitir, \hld es finnask vigi at &
Surtr ok hin svǫ́su goð.\\ \&

\bv2 Web-strengthener quoth: \\ “Say this, Gain-adviser, since on the floor I will to try thy fame: How that valley is called, where \textbf{Surt} and the excellent gods find each other at war?" \\

(Gagnráðr kvað:) &
\bv1 Vígríðr hęitir vǫllr, \hld es finnask vígi at &
Surtr ok hin svǫ́su goð; &
hundrað rasta \hld hann's á hvęrjan veg; &
sá's þęim vǫllr vitaðr.\\ \&

\bv2 Gain-adviser quoth: \\ “\textbf{Battle-rider} is the valley called, where Surt and the cheerful gods find each other at war. A hundred rests\footnotemark[30], he stretches in each direction; that valley is known for them.\footnotemark[31]" \\
\footnotetext[30]{An old unit of length, from its name prob. the length a horse could travel before resting.}
\footnotetext[31]{That is, known for its great size.}

(Vafþrúðnir kvað:) &
\bv1 Fróðr est nú gęstr, \hld far á bękk jǫtuns, &
mælumk í sessi saman; &
hǫfði vęðja \hld vit skulum hǫllu í &
gęstr, of gęðspęki.\\ \&

\bv2 Web-strengthener quoth: \\ “Knowing art thou now, guest, sit down on the ettin's bench; let us speak while sitting together. In the hall we shall wager a head, guest, over [our] mind-wisdom." \\

(Gagnráðr kvað:) &
\bv1 Sęg þat hit ęina, \hld ef þitt ǿði\footnotemark[10] dugir &
ok þú Vafþrúðnir vitir, &
hvaðan jǫrð of kom \hld eða upphiminn &
fyrst, hinn fróði jǫtunn.\\ \&
\footnotetext[10]{Starting with \emph{ǿði}, the poem is also preserved in 748.}

\bv2 Gain-adviser quoth: \\ “Say the first\footnotemark[32], if thy wisdom suffices, and thou, Web-strengthener, might know: Whence, O knowing ettin, the earth first came, or \textbf{up-heaven}?" \\
\footnotemark[32]{Lit. ‘one'.}

(Vafþrúðnir kvað:) &
\bv1 Ór Ymis holdi \hld vas jǫrð of skǫpuð, &
ęn ór bęinum bjǫrg, &
himinn ór hausi \hld hins hrimkalda jǫtuns, &
ęn ór svęita sær.\\ \&

\bv2 Web-strengthener quoth: \\ “Out of \textbf{Yime's} hull\footnotemark[35], the earth was created, but the mountains out of his bones. Heaven out of the skull of the frost-cold ettin, but the sea out of his sweat.\footnotemark[36]" \\
\footnotetext[35]{His body.}
\footnotetext[36]{\emph{svęiti} ‘sweat' is a common kenning for blood. — This v. closely resembles \emph{Grím} 40.}

(Gagnráðr kvað:) &
\bv1 Sęg þat annat, \hld ef þitt ǿði dugir &
ok þú Vafþrúðnir vitir, &
hvaðan máni of kom, \hld svát fęrr menn yfir, &
eða sól hit sama.\\ \&

\bv2 Gain-adviser quoth: \\ “Say the second, if thy wisdom suffices, and thou, Web-strengthener, might know: Whence the moon came, so that it travels over men, or likewise the sun?" \\

(Vafþrúðnir kvað:) &
\bv1 Mundilfari hęitir, \hld hann's Mána faðir &
ok svá Solar hit sama; &
himin hverfa \hld þau skulu hvęrjan dag &
ǫldum at ártali.\\ \&

\bv2 Web-strengthener quoth: \\ “\textbf{Moundelfare} [that one] is called, he is the father of the Moon, and likewise of the Sun. They shall circle in the heavens every day, for men to reckon time\footnotemark[40]." \\
\footnotetext[40]{Lit. “for men to year-reckoning".}

(Gagnráðr kvað:) &
\bv1 Sęg þat þriðja, \hld alls þik svinnan kveða &
ok þú Vafþrúðnir vitir, &
hvaðan dagr of kom, \hld sá's fęrr drótt yfir, &
eða nótt með niðum.\\ \&

\bv2 Gain-adviser quoth: \\ “Say the third, since [they] call thee wise, and thou, Web-strengthener, might know: Whence the day came, the one that travels over the rettinue, or night with the moon-phases?" \\

(Vafþrúðnir kvað:) &
\bv1 Dęllingr hęitir, \hld hann's Dags faðir, &
ęn Nótt vas Nǫrvi borin; &
ný ok nið \hld skópu nýt ręgin &
ǫldum at ártali.\\ \&

\bv2 Web-strengthener quoth: \\ “\textbf{Delling} [that one] is called, he is the father of \textbf{Day}, but \textbf{Night} was born to \textbf{Nare}. The waxing and waning [of the moon], the useful Powers created, for men to reckon time." \\

(Gagnráðr kvað:) &
\bv1 Sęg þat fjórða, \hld alls þik fróðan kveða, &
ok þú Vafþrúðnir vitir, &
hvaðan vetr of kom \hld eða varmt sumar &
fyrst með fróð ręgin.\\ \&

\bv2 Gain-adviser quoth: \\ “Say the fourth, since [they] call thee knowing, and thou, Web-strengthener, might know: Whence winter first came, or the warm summer, among the knowing Powers?" \\

(Vafþrúðnir kvað:) &
\bv1 Vindsvalr hęitir, \hld hann's Vetrar faðir, &
ęn Svǫ́suðr Sumars.\footnotemark[15]\\ \&
\footnotetext[15]{Second half of the v. seems missing.}

\bv2 Web-strengthener quoth: \\ “\textbf{Wind-cool} [that one] is called, he is the father of \textbf{Winter}, but \textbf{Delightful} of \textbf{Summer}." \\

(Gagnráðr kvað:) &
\bv1 Sęg þat fimta, \hld alls þik fróðan kveða, &
ok þú Vafþrúðnir vitir, &
hvęrr ása ęlztr \hld eða Ymis niðja &
yrði í árdaga.\\ \&

\bv2 Gain-adviser quoth: \\ “Say the fifth, since [they] call thee knowing, and thou, Web-strengthener, might know: Who, in days of yore became the eldest of the \textbf{Anses}, or of the descendants of Yime?" \\

(Vafþrúðnir kvað:) &
\bv1 Ørófi vetra \hld áðr væri jǫrð skǫpuð, &
þá vas Bergęlmir borinn, &
Þrúðgęlmir \hld vas þess faðir, &
ęn Aurgęlmir afi.\\ \&

\bv2 Web-strengthener quoth: \\ “Uncountable winters before the earth would be created, then \textbf{Bear-yeller} was born. \textbf{Strength-yeller} was \emph{that one's} father, and \textbf{Mud-yeller} the grandfather." \\

(Gagnráðr kvað:) &
\bv1 Sęg þat sétta, \hld alls þik svinnan kveða, &
ok þú Vafþrúðnir vitir, &
hvaðan Aurgęlmir kom \hld með jǫtna sonum &
fyrst, hinn fróði jǫtunn.\\ \&

\bv2 Gain-adviser quoth: \\ “Say the sixth, since [they] call thee wise, and thou, Web-strengthener, might know: Whence, O knowing ettin, Mud-yeller first came among the sons of ettins?" \\

(Vafþrúðnir kvað:) &
\bv1 Ór Élivǫ́gum \hld stukku ęitrdropar, &
svá óx unz ór varð jǫtunn; &
órar ættir \hld kómu þar allar saman; &
því's þat æ alt til atalt.\footnotemark[20]\\ \&
\footnotetext[20]{Lines 3–4 missing in R and 748, but quoted in \emph{Gylf}.}

\bv2 Web-strengthener quoth: \\ “From the \textbf{Ell-waves}, poison-drops splashed; thus [it] grew until an ettin emerged. \emph{Our} family lines all together originated there, therefore our race\footnotemark[45] is forever fierce against all.\footnotemark[46]" \\
\footnotetext[45]{Lit. ‘it' or ‘that'.}
\footnotetext[46]{Somewhat strange phrasing, but the line does not appear damaged. It is clearly an explanation of the fierce and maleficent nature of the ettins, as their first ancestors were created from poison.}

(Gagnráðr kvað:) &
\bv1 Sęg þat sjaunda, \hld alls þik svinnan kveða, &
ok þú Vafþrúðnir vitir, &
hvé sá bǫrn gat \hld hinn baldni\footnotemark[25] jǫtunn, &
es hann hafði-t gýgjar gaman.\\ \&
\footnotetext[25]{R has \emph{aldni}, ‘aged, old'. This breaks alliteration; \emph{baldni} ‘bold, defiant' has been substituted from 748.}

\bv2 Gain-adviser quoth: \\ “Say the seventh, since [they] call thee wise, and thou, Web-strengthener, might know: How did that one, the defiant ettin, beget children, when he did not enjoy the [carnal] pleasure of a troll-woman?" \\

(Vafþrúðnir kvað:) &
\bv1 Und hęndi vaxa \hld kvǫ́ðu hrímþursi &
męy ok mǫg saman; &
fótr við fǿti \hld gat hins fróða jǫtuns &
sexhǫfðaðan son.\\ \&

\bv2 Web-strengthener quoth: \\ “Neath the hand\footnotemark[50] on the \textbf{frost-thurse}, [they] said that a girl and boy together grew. A foot against a foot begot, for the knowing ettin, a six-headed son." \\
\footnotetext[50]{\hęndi (dative of \emph{hǫnd}) means ‘hand', but might here be a poetic circumlocution for ‘arm'.}

(Gagnráðr kvað:) &
\bv1 Sęg þat áttunda, \hld alls þik fróðan kveða, &
ok þú Vafþrúðnir vitir, &
hvat fyrst of mant \hld eða fręmst of vęizt, &
þú est alsviðr jǫtunn.\\ \&

\bv2 Gain-adviser quoth: \\ “Say the eigth, since [they] call thee knowing, and thou, Web-strengthener, might know: What dost thou first remember, or earliest know?\footnotemark[55] Thou art all-wise, ettin." \\
\footnotetext[55]{Cf. Vsp 1.}

(Vafþrúðnir kvað:) &
\bv1 Ørófi vetra \hld áðr væri jǫrð of skǫpuð, &
þá vas Bergęlmir borinn; &
þat fyrst um man'k, \hld es hinn fróði jǫtunn &
á vas lúðr of lagiðr.\footnotemark[30]\\ \&
\footnotetext[30]{This verse is quoted in \emph{Gylf}.}

\bv2 Web-strengthener quoth: \\ “Uncountable winters before the earth would be created, then Bear-yeller was born. \emph{That} I first remember, when the knowing ettin\footnotemark[60] was laid down on the funeral-bed\footnotemark[61]." \\
\footnotetext[60]{That is, Bear-yeller. Cf. v. 29.}
\footnotetext[61]{\emph{lúðr}, a tricky word.}

(Gagnráðr kvað:) &
\bv1 Sęg þat níunda, \hld alls þik svinnan kveða, &
ok þú Vafþrúðnir vitir, &
hvaðan vindr of kømr \hld svát fęrr vág yfir, &
æ męnn hann sjalfan of séa.\\ \&

\bv2 Gain-adviser quoth: \\ “Say the ninth, since [they] call thee wise, and thou, Web-strengthener, might know: Whence the wind comes, so that he travels over the wave; forever men see him himself.\footnotemark[65]" \\
\footnotetext[65]{Perhaps a negation has been lost here; the wind is never seen by men.}

(Vafþrúðnir kvað:) &
\bv1 Hræsvęlgr hęitir, \hld es sitr á himins ęnda, &
jǫtunn í arnar ham; &
af hans vængjum \hld kveða vind koma &
alla męnn yfir.\\ \&

\bv2 Web-strengthener quoth: \\ “\textbf{Corpse-swallower} [that one] is called, which sits at the end of the heavens, an ettin in the shape of an eagle. From his wings, they say [that] the wind comes over all men." \\

(Gagnráðr kvað:) &
\bv1 Sęg þat tíunda, \hld alls þú tíva rǫk &
ǫll Vafþrúðnir vitir, &
hvaðan Njǫrðr of kom \hld með niðjum ása. &
Hófum ok hǫrgum \hld hann ræðr hundmǫrgum &
ok varð-at hann ǫ́sum alinn.\\ \&

\bv2 Gain-adviser quoth: \\ “Say the \emph{tenth}, since thou, Web-strengthener, of all the fates of the \textbf{Tues} might know: Whence \textbf{Nearth} came into the company of the kinsmen of the \textbf{Anses}? He rules an immense number\footnotemark[68] of \textbf{hoves} and \textbf{heargs}, and he was not begotten among the Anses." \\
\footnotetext[68]{Lit. “he rules hound-many".}

(Vafþrúðnir kvað:) &
\bv1 Í Vanahęimi \hld skópu hann vís ręgin &
ok sęldu at gíslingu goðum, &
í aldar rǫk \hld hann mun aptr koma &
hęim með vísum vǫnum.\\ \&

\bv2 Web-strengthener quoth: \\ “In \textbf{Wane-Home}, the wise \textbf{Powers}\footnotemark[69] created him, and gave\footnotemark[70] him as a hostage to the gods. In the fate of the age, he will come back, home among the wise \textbf{Wanes}." \\
\footnotetext[69]{The gods, or specifically the Wanes.}
\footnotetext[70]{Lit. “sold".}

(Gagnráðr kvað:) &
\bv1 Sęg þat ęllipta, \hld hvar ýtar túnum í &
hǫggvask hvęrjan dag; &
val þęir kjósa \hld ok ríða vígi frá, &
sitja męir of sáttir saman.\footnotemark[35]\\ \&
\footnotetext[35]{This and the next v. are damaged in both R and 748; R has only this verse, but splits it in two (the 2nd starting with \emph{val}), while 748 has 40:1 (Ms.: \emph{S. þ. e. XI}) and then jumps to the answer v. 41. They have here been reconstructed, but it is possible some lines are still missing.}

\bv2 Gain-adviser quoth: \\ “Say the eleventh, Where men in yards, hew away at each other every day? They choose those destined to die in war, and ride [away] from battle; [then] they sit more content together." \\

(Vafþrúðnir kvað:) &
\bv1 Allir ęinhęrjar \hld Óðins túnum í &
hǫggvask hvęrjan dag, &
val þeir kjósa \hld ok ríða vígi frá, &
sitja męir of sáttir saman.\\ \&

\bv2 Web-strengthener quoth: \\ “In Woden's yards, all the \textbf{Lone Warriors} hew away at each other every day. They choose those destined to die in war, and ride [away] from battle; [then] they sit more content together." \\

(Gagnráðr kvað:) &
\bv1 Sęg þat tolpta, \hld hví þú tíva rǫk &
ǫll Vafþrúðnir vitir, &
frá jǫtna rúnum \hld ok allra goða &
þú hit sannasta sęgir, &
hinn alsvinni jǫtunn.\\ \&

\bv2 Gain-adviser quoth: \\ “Say the twelfth, Why thou, Web-strengthener, should know all the fates of the \textbf{Tues}\footnotemark[73]? Regarding the \textbf{runes} of the ettins and of all the gods, thou, the all-wise ettin, speaks most truly." \\
\footnotetext[73]{The gods. Formation identical to \emph{ragna rǫk} ‘the fates of the Powers'.}

(Vafþrúðnir kvað:) &
\bv1 Frá jǫtna rúnum \hld ok allra goða &
ek kann sęgja satt, &
þvíat hvęrn hęf'k \hld heim of komit, &
níu kom'k hęima \hld fyr niflhęl neðan; &
hinig dęyja ór hęlju halir.\\ \&

\bv2 Web-strengthener quoth: \\ “Regarding the runes of the ettins and of all the gods, I can speak truly, for I have come into each \textbf{Home}. I came [into] \textbf{nine Homes} beneath \textbf{Fog-hell}; thereto men die out of \textbf{Hell}\footnotemark[75]." \\
\footnotetext[75]{A difficult verse. Does it imply that if a man dies when already dead, he goes to Fog-hell?}

(Gagnráðr kvað:) &
\bv1 Fjǫlð ek fór, \hld fjǫlð fręistaða'k, &
fjǫlð ek ręynda ręgin; &
hvat lifir manna, \hld þá's hinn mæra líðr &
fimbulvetr með firum?\\ \&

\bv2 Gain-adviser quoth: \\ “Much I travelled, much I tried, much I tested the \textbf{Powers}.\footnotemark[80] What remains\footnotemark[79] of men, when the famous \textbf{fimbol-winter} passes among firs\footnotemark[81]?" \\
\footnotetext[79]{Lit. “lives".}
\footnotetext[80]{Here begins the repetition of the same “mantra" used in v. 3, which continues until the final question (v. 54).}
\footnotetext[81]{Among men.}

(Vafþrúðnir kvað:) &
\bv1 Líf ok Lífþrasir, \hld ęn þau lęynask munu &
í holti Hoddmímis; &
morgindǫggvar \hld þau sér at mat hafa; &
þaðan af aldir alask.\\ \&

\bv2 Web-strengthener quoth: \\ “\textbf{Life} and \textbf{Life-striver}, and they will hide themselves in the woods of \textbf{Hoard-Mime}\footnotemark[85]. Morning-dew they [will] have as food; thence generations will be bred." \\
\footnotetext[85]{Prob. the same as \emph{Yggdrasill}, the \textbf{Terrifier-Steed.}}

(Gagnráðr kvað:) &
\bv1 Fjǫlð ek fór, \hld fjǫlð fręistaða'k, &
fjǫlð ek ręynda ręgin; &
hvaðan kømr sól \hld á hinn slétta himin, &
es þessa hęfr Fęnrir farit?\\ \&

\bv2 Gain-adviser quoth: \\ “Much I travelled, much I tried, much I tested the Powers. Whence comes sun onto the smooth heaven, when \textbf{Fenner} has killed this one\footnotemark[90]?" \\
\footnotetext[90]{That is, the current incarnation of the sun, as explained in the next v.}

(Vafþrúðnir kvað:) &
\bv1 Ęina dóttur \hld berr alfrǫðull, &
áðr hana Fęnrir fari; &
sú skal ríða, \hld þá's ręgin dęyja, &
móður brautir mær.\\ \&

\bv2 Web-strengthener quoth: \\ “One daughter the elf-wheel\footnotemark[95] bears, before \textbf{Fenner} might kill her. When the Powers die, that one, the girl, shall ride the paths of the mother." \\
\footnotetext[95]{The sun.}

(Gagnráðr kvað:) &
\bv1 Fjǫlð ek fór, \hld fjǫlð fręistaða'k, &
fjǫlð ek ręynda ręgin; &
hvęrjar 'ro męyjar, \hld es líða mar yfir, &
fróðgęðjaðar fara.\\ \&

\bv2 Woden quoth: \\ “Much I travelled, much I tried, much I tested the Powers. Which are the girls that pass over the ocean; wise-minded they go?" \\

(Vafþrúðnir kvað:) &
\bv1 Þríar þjóðár \hld falla þorp yfir &
męyja Mǫgþrasis; &
hamingjur ęinar \hld þær's í hęimi eru, &
þó þær með jǫtnum alask.\\ \&

\bv2 Web-strengthener quoth: \\ “Three great rivers fall over the settlement of the girls of \textbf{Boy-striver}; they are the only \textbf{Hamings} in the Home,\footnotemark[99] though they are raised among the ettins\footnotemark[100]." \\
\footnotetext[99]{In Ettin-Home, or in the entire world?}
\footnotetext[100]{An opaque verse. These three girls may be the same as the three \emph{thurse-maidens} in \emph{Vsp} 9, see note there. Their father, Boy-striver, is not mentioned anywhere else.}

(Gagnráðr kvað:) &
\bv1 Fjǫlð ek fór, \hld fjǫlð fręistaða'k, &
fjǫlð ek ręynda ręgin; &
hvęrir ráða æsir \hld ęignum goða, &
þá's sloknar Surtalogi?\\ \&

\bv2 Gain-adviser quoth: \\ “Much I travelled, much I tried, much I tested the Powers. Which properties of the gods will the \textbf{Anses} [still] rule\footnotemark[105], when the flame of \textbf{Surt} burns out?" \\
\footnotetext[105]{Or ‘control'.}

(Vafþrúðnir kvað:) &
\bv1 Víðarr ok Váli \hld byggva vé goða, &
þá's sloknar Surtalogi; &
Móði ok Magni \hld skulu Mjǫlni hafa &
Vingnis at vígþroti.\\ \&

\bv2 Web-strengthener quoth: \\ “\textbf{Wider} and \textbf{Weel} [will] inhabit the sanctuaries of the gods, when the \textbf{flame of Surt} burns out. \textbf{Mood} and \textbf{Main} will own \textbf{Mealen}, when \textbf{Wingen} can no longer fight\footnotemark[110]." \\
\footnotetext[110]{Lit. “at Wingen's fight-exhaustion", referring to his death.}

(Gagnráðr kvað:) &
\bv1 Fjǫlð ek fór, \hld fjǫlð fręistaða'k, &
fjǫlð ek ręynda ręgin; &
hvat verðr Óðni \hld at aldrlagi, &
þá's rjúfask ręgin?\\ \&

\bv2 Gain-adviser quoth: \\ “Much I travelled, much I tried, much I tested the Powers. What happens to Woden at the end of the age, when the Powers are broken?" \\

(Vafþrúðnir kvað:) &
\bv1 Ulfr glęypa \hld mun Aldafǫðr, &
þess mun Víðarr vreka; &
kalda kjapta \hld hann klyfja mun &
vitnis vígi at.\\ \&

\bv2 Web-strengthener quoth: \\ “The wolf will swallow the \textbf{Leader of Men}; Wider will avenge that. He will cleave the cold jaws of the wolf at the battle." \\

(Gagnráðr kvað:) &
\bv1 Fjǫlð ek fór, \hld fjǫlð fręistaða'k, &
fjǫlð ek ręynda ręgin; &
hvat mælti Óðinn, \hld áðr á bál stigi, &
sjalfr í ęyra syni?\\ \&

\bv2 Gain-adviser quoth: \\ “Much I travelled, much I tried, much I tested the Powers. What spoke Woden himself, before [he]\footnotemark[115] would step onto the funeral pyre, into the ear of the son?" \\
\footnotetext[115]{Prob. Woden's son, that is \textbf{Balder}.}

(Vafþrúðnir kvað:) &
\bv1 Ęy mann-gi\footnotemark[40] vęit, \hld hvat þú í árdaga &
sagðir í ęyra syni; &
fęigum munni \hld mælta'k mína forna stafi &
ok of ragna rǫk. &
Nú við Óðin \hld dęilda'k mína orðspęki; &
þú est æ vísastr vera.\\ \&
\footnotetext[40]{Emended from R, 748 \emph{manni}; the word must be in the nominative, but \emph{manni} is dative.}

\bv2 Web-strengthener quoth: \\ “No man may ever know\footnotemark[119], what thou in days of yore said into the ear of the son. With a death-doomed\footnotemark[120] mouth, I spoke my ancient staves, and about the \textbf{fates of the Powers}\footnotemark[121]. Now against Woden, I have shared my word-wisdom\footnotemark[122]; thou art forever the wisest of men\footnotemark[123]." \\
\footnotetext[119]{Lit. “eternally no man knows".}
\footnotetext[120]{Or '\emph{fey} mouth'. Web-strengthener here realizes that he was bound to die from the moment (v. 19) he proposed the wager; no man, nor god, nor ettin can outwit Woden.}
\footnotetext[121]{“fates of the Powers", that is, \emph{ragna rǫk}.}
\footnotetext[122]{The same word-wisdom Woden in v. 5 set out to try.}
\footnotetext[123]{Word used is \emph{ver} ‘husband', ‘man'. Perhaps in the broader sense of ‘male being'.}
