\part{Index}

\bookStart{Index (INCOMPLETE!)}

NOTE: This index is both incomplete and inconsistently formatted and should not be relied on.  New entries will be added and old ones corrected and expanded in the future.  For an example of a complete entry see \inx[C]{Yimer}.

%NOTE: NOM ACC GEN DAT

\section{Cultural and religious terms and expressions (C)}
\begin{itemize}

\inxitem[C]{ape} (ON \emph{api}, OE \emph{apa}, OS \emph{apo}, OHG \emph{affo}, PNWGmc. \emph{*apó})
  In the Old Norse the word seems to mean ‘fool, buffoon’, in the other old languages apparently ‘monkey’, though this sense should be a later development of the former; why would the early Germanic tribes have a word for an animal that they had never encountered?

\inxitem[C]{aught} (ON \emph{ę́tt}, OE \emph{ǽht} ‘possession, property’)
  The Nordic (paternal) clan or family line.

\inxitem[C]{begale} (OHG \emph{bi·galan})
  To enchant, bewitch something or someone by singing a \inx[C]{galder}. Transitive of \inx[C]{gale}.

\inxitem[C]{bigh} (ON \emph{baugr}, OE \emph{béag}, OHG \emph{boug})
  Armlets used as currency during the Migration Period. --- The giving of rings and armlets in exchange for loyalty (\inx[C]{holdness} being the word used for a warrior’s loyalty towards his lord, and of a lord’s grace towards his servants) was common across all of Germanic Europe, as seen in the many poetic ruler-kennings of the type “breaker of rings” (e.g. \emph{béaga brytta} ‘the breaker of bighs’ in \Beowulf\ ll. 35, 352, 1487). An illustrative example of this is \textlink{Hildebrandslied} 33--35.
  This is also connected with the oath-ring, and the famous ring-swords. TODO? reference some literature on this.

\inxitem[C]{bloot} (ON \emph{blót}, OE \emph{blót}, OHG \emph{bluoz})
  A sacrifice or a sacrificial feast, one of the best attested Germanic pagan practices. The animals would be sacrificed by the host, cooked in large kettles and eaten communally. See also \inx[C]{bloot-house}.

\inxitem[C]{bloot-house} (ON \emph{blót-hús}, OHG \emph{bluoz-hús})
  A heathen temple. Glosses Latin \emph{fānum} in OHG. See also \inx[C]{harrow}, \inx[C]{hove}, \inx[C]{wigh}.

\inxitem[C]{Doom} (ON \emph{dómr}, OE \emph{dóm})
  Base meaning ‘judgment, verdict’ (whence Doomsday, ‘judgment Day’), but in the Norse and Anglo-Saxon poetry often specifically referring to one’s fame or good reputation (that is, how others will judge one’s character and deeds), especially after death. It is clear that this verdict was of utmost importance to the ancient Germanic people. The clearest examples are \textlink{Havamal} 77 (see there): \emph{I know one that never dies: the \textbf{Doom} o’er each man dead.} and \Beowulf\ 1384--1389, where Beewolf consols king Rothgar after Grendle’s mother has slain his trusted advisor Asher (\emph{Æschere}):
  \emph{Ne sorga, snotor guma! \hld\ Sélre bið ǽg-hwǽm, /
  þæt hé his fréond wrece, \hld\ þonne hé fela murne. /
  Úre ǽghwylc sceal \hld\ ende ge·bídan /
  worolde lífes; \hld\ wyrce sé þe móte /
  \textbf{dómes} ǽr déaþe; \hld\ þæt bið driht-guman /
  un·lifgendum \hld\ æfter sélest.}

  ‘Grieve not, wise man! ’Tis better for each one / that he avenge his friend than that he mourn much. / Each one of us shall suffer the end / of worldly life---win he who might / \textbf{Doom} before death: that is for the warrior, / unliving, afterwards the best.’
  Other illustrative examples in \Beowulf\ include 884b--887a:
  \emph{[...] Sige-munde ge·sprǫng /
  æfter déað-dæge \hld\ \textbf{dóm} un-lýtel /
  syþðan wíges heard \hld\ wyrm á·cwealde /
  hordes hyrde [...]}
  ‘For \inx[P]{Syemund} sprang up / after his death-day an unlittle \ken*{great} \textbf{Doom}, / since hard in conflict he defeated the \inx[C]{wyrm}, / the hoard’s herder.’
  and 953b--955a: \emph{[...] þú þé self hafast /
  dę́dum ge·fręmed \hld\ þæt þín \textbf{dóm} lyfað /
  áwa tó aldre [...]}
  ‘Thou hast for thyself / by deeds accomplished that thy \textbf{Doom} lives / for ever and ever.’

\inxitem[C]{feather-hame} (ON \emph{fjaðr-hamr}, OE \emph{feðer-hama}, OS \emph{feðar-}, \emph{feðer-hamo})
  A plumage which when donned by the wearer lets him fly like, or become a bird.  One is owned by Frow and used by Lock to fly between the homes in \textlink{Thrymskvida}.  In the Christian \textlink{Heliand} feather-hames are donned by angels who fly from heaven to earth.  See also \inx[C]{hame}.

\inxitem[C]{fee} (ON \emph{fé}, OE \emph{féoh})
  Originally ‘cattle, kine’, however also used in a broader sense to refer to one’s mobile wealth; for that cf.~particularly \textlink{Havamal}.

\inxitem[C]{fey} (ON \emph{fęigr}, OE \emph{fǽge}, OHG \emph{feigi} ‘cowardly’)
  Being doomed or fated to die, with a sense of predestination and inevitability.  Its earliest documented Scandinavian use is on the Rök stone: \textbf{aft uamuþ stąnta runaʀ þaʀ ᛭ n uarin faþi faþiʀ aft \emph{faikiąn} sunu} \emph{Apt Vámóð standa rúnaʀ þáʀ, en Varinn fáði, faðir aft \emph{fęigjan} sonu} ‘After Woemood (\emph{Vámóðr}) stand these \inx[C]{rune}[runes], but Warren (\emph{Varinn}) painted, the father after the \textbf{fey} son.’  See \textciteshorthand{PCRN-HS} II:35, p. 928 ff. (TODO)

\inxitem[C]{feyness} (ON \emph{fęigð})
  The state of being \inx[C]{fey}.

\inxitem[C]{fimble-} (ON \emph{fimbul-})
  The ultimate, final, greatest.  See \inx[P]{Fimblethyle}, \inx[P]{Fimble-winter}.

\inxitem[C]{five days} (ON \emph{fimm dagar})
  The Old Scandinavian (and perhaps Germanic) week was originally five days long, the seven-day week being a later import, as seen by the names of the days, which are obviously calqued from the Latin (\emph{Dies Mercurii} = Weden’s day, et.c.).  According to the \Gulatingslog\ there were six weeks in a month, and “five days” is used as a generic period of time in \textlink{Havamal} 51 and 74; in st.~74 it is contrasted with month.  Related to this is the legal term \emph{fifth} (ON \emph{fimmt}, OSw. \emph{fæmt}), a meeting or gathering set to be held at a five-day notice.  See \emph{fimt} in \CV, \textciteshorthand{LMNL} for further discussion.

\inxitem[C]{galder} (ON \emph{galdr}, OE \emph{gealdor}, OHG \emph{galdar})
  A magical song or incantation, probably synonymous with \inx[C]{leed}.  Verbal noun formed to \inx[C]{gale} ‘to sing, chant’.

\inxitem[C]{gale} (ON \emph{gala}, OE \emph{galan}, OHG \emph{galan})
  To sing, chant, especially of magical songs; verbal root of \inx[C]{galder} ‘something sung, chanted’.

\inxitem[C]{gand} (ON \emph{gandr}, Latin \emph{gandus})
  A witch’s familiar or foul spirit sent out to do her bidding. See \textciteshorthand{PCRN-HS} I:17, p. 361 and II:26, p. 656. TODO

\inxitem[C]{gid} (ON \emph{goði}, OE \emph{Gydda} m.~given name)
  A heathen priest or master of ceremonies.

\inxitem[C]{gidden} (ON \emph{gyðja}, OE \emph{gyden} ‘goddess’)
  The womanly equivalent or wife of a \inx[C]{gid}.

\inxitem[C]{good of meat} (ON \emph{matar góðr}, \emph{góðr matar})
   An old formula appearing in \textlink{Havamal} 39 and numerous Swedish Wiking Age Runic inscriptions Sm 39, Sm 44, Sö 130, U 703, and U 805.  Cf.~U 739 which has the related \emph{mildr mataʀ} ‘mild of meat’.  Antonyms are \emph{matar illr} ‘evil of meat’ and \inx[C]{meat-nithing}.

\inxitem[C]{guest} (ON \emph{gęstr}, OE \emph{giest}, OS \emph{gast}, OHG \emph{gast}, Got. \emph{gasts}, PGmc. \emph{gastiz})
  Guests were often strangers, wanderers, who would come to beg for food and lodgings.  The Old Germanic peoples placed great value on hospitality. TODO.

\inxitem[C]{hame} (ON \emph{hamr})
  A skin, shape. People could “shift hames” (ON \emph{skipta hǫmum}), leaving their human hames behind and instead entering into the shapes of wolves, bears, birds.  During this process the original hame, that is, the human body, would be sleeping in a vulnerable state. A concise description of this is found in \YnglingaSaga\ 7: \emph{Óðinn \emph{skipti hǫmum}, lá þá búkr’inn sem sofinn eða dauðr, en hann var þá fugl eða dýr, fiskr eða ormr, ok fór á einni svipstund á fjarlæg lǫnd at sínum erendum eða annarra manna.} ‘Weden \emph{shifted hames}; then lay the trunk of his body as if sleeping or dead, but he was then a fowl or beast, a fish or serpent, and journeyed in a short while to foreign lands with his errands or those of other men.’.

  See also \inx[C]{feather-hame}, \inx[C]{fence-rideresses}, \inx[C]{evening-rideresses}.

\inxitem[C]{harrow} (ON \emph{hǫrgr}, OE \emph{hearg}, PNWGmc. \emph{*harugaʀ})
  A hallowed cairn or stone-heap.  \textlink{Hyndluljod} 10 describes the construction of one.  The Norwegian laws prescribe the “breaking of harrows and burning of hoves”.

  See also \inx[C]{hove}, \inx[C]{wigh}.

\inxitem[C]{hold} (ON \emph{hollr}, OE \emph{hold}, OS \emph{hold}, OHG \emph{hold})
  %TODO Mention: unhold wights, Old Saxon baptismal formula.
  ‘Favourable, loyal, gracious’, often of a ruler towards his subject (in the sense of ‘gracious, benevolent’) or vice-versa (in the sense of ‘loyal, devoted’).  Mirroring these earthly relationships, it is often used to refer to divine grace, both of the Christian God---thus in the \emph{Ecclesiastical Laws of King Cnut} \textciteshorthand[372]{ALIE1}: \emph{Þam byþ witod-líce God hold, þe bið his hlâforde riht-líce hold} ‘Indeed God is \textbf{hold} to him who is rightly \textbf{hold} to his lord’---but in the oldest Scandinavian material likewise of the Heathen gods.
  So \textlink{Lokasenna} 4 (e.): \emph{holl ręgin} ‘\textbf{hold} \inx[P]{Reins}’, and \textlink{Oddrunargratr} 9/1: \emph{Svá hjalpi þér \hld\ hollar véttir} ‘So help thee \textbf{hold} \inx[C]{wights}’. %TODO: Numbering of Oddrunargratr is very uncertain.

  This word is common in old Scandinavian oath formulæ, e.g. in the elder redaction of the West-Geatish Law: \emph{Svá sé mér goð holl} ‘So may the Gods(!) be \textbf{hold} to me,’ in mediæval Norwegian laws (\textciteshorthand{NGL2}[197,397]) and Grey-Goose (TODO: cite): \emph{Guð sé mér hollr ef ek satt segi, gramr ef ek lýg} ‘God be \textbf{hold} to me if I speak truly, wroth if I lie,’ in Grey-Goose (TODO) also: \emph{Sé guð hollr þeim er heldr griðum, en gramr þeim er grið rýfr} ‘God be \textbf{hold} to him who keeps the truce, but wroth against him who breaks the truce’. I refer to \textcite{Läffler1895} for further discussion on these formulæ.

\inxitem[C]{holdness} (ON \emph{hylli}, OE \emph{hyldu}, OHG \emph{huldí})
  Abstract noun formed to \inx[C]{hold}, meaning ‘favour, loyalty, grace,’ with the same semantics as the adjective.

  Notably, this word appears three times in connection with the grace of gods in the poetry, namely in \textlink{Grimnismal} 43, where (according to my interpretation) the preparer of food at the bloot is said to earn the “\textbf{holdness} of \inx[P]{Wolder} and of all the gods;” and \textlink{Grimnismal} 53 where the disgraced king Garfrith is said to have been bereft of the support \emph{gęngi} of Weden and all the Oneharriers, and of “Weden’s \textbf{holdness}” (\emph{Óðins hylli}).  “Weden’s holdness” is also mentioned in a stanza by Hallfrith (edited as Hfr Lv 7 by Diana Whaley in \Skp\ V), who laments that: “The whole race of man has wrought songs to win the \textbf{holdness} of Weden; I recall the fully rewarded works of our kinsmen/ancestors.”

  From the semantics of this word the Germanic view on heavenly grace is clear: the Gods are \textbf{hold} towards those who do good works, which include swearing true oaths, faithfully observing truces, partaking in the bloot, following rules of hospitality, and composing poetry---and \inx[C]{gram} ‘wroth’ towards those who do the opposite.

\inxitem[C]{Home} (ON \emph{hęimr}, OE \emph{hám}, PNWGmc. \emph{*haimaʀ})
  In the Norse often referring to a realm in the cosmology (\textlink{Voluspa} 2: “I remember nine \textbf{Homes}”, \textlink{Vafthrudnismal} TODO: “From the runes of the \inx[P]{Ettins} and of all the gods I can speak truly, for I have come into each \textbf{Home}”). Thus \inx[P]{Ettinham} is the ‘\textbf{Home}/realm of the ettins’. When used on its own it means ‘the world (that we inhabit)’. See also \inx[P]{Nine Homes}, \inx[P]{Thrithham}.

\inxitem[C]{leat} (ON \emph{hlaut})
  In some saws explained as the blood drained from the offered animal; the verbal root is \emph{hljóta} ‘to get by lot’ and this word certainly refers to the use of the blood for auguries.

\inxitem[C]{leat-twig} (ON \emph{hlaut-tęinn})
  A twig used to sprinkle the \inx[C]{leat}.  The pattern of the blood would presumably be inspected for the augury; cf.~\textlink{Hymiskvida} 1.

\inxitem[C]{leek} (ON \emph{laukr}, OE \emph{léac}, PNWGmc \emph{laukaʀ})
  The leek was a plant of great cultural importance.  It was seen as the noblest plant, so \textlink{GudrunTwo} 2, where \inx[P]{Siward}’s superiority to the \inx[P]{Yivickings} is compared to a stag among wild beasts, gold among silver, and a green leek in grass; and \textlink{Voluspa} 4, where the earth of the \inx[P]{Golden Age} was grown with green leek.

  The leek was highly valued in folk magic, as seen already on gold bracteates from the C5th and C6th, where it often appears as a charm word in the old form {ᛚᚨᚢᚲᚨᛉ} \emph{laukaʀ}; in one inscription also paired with {ᛚᛁᚾᚨ} \emph{lína} ‘linen’.  Classical Norse attestations of magic use include \textlink{Sigrdrifumal} 8, where the leek is thrown into mead against poison; and the \Volsathattr, where a horse penis is said to be \emph{\alst{l}íni gǿddr \hld\ en \alst{l}aukum studdr} ‘endowed with linen and supported by leeks’ in a poetic line.  The leek was particularly associated with women and domestic life, as seen by its pairing with “linen” and its frequent use as the determinant in women-kennings \parencite[418]{Meissner1921}).  Anon \emph{Sveinfl} 1 (\Skp\ 1) sarcastically states that a battle was not \emph{sem manni \hld\ mę́r lauk eða ǫl bę́ri} ‘as if a maiden brought a man leek or ale’.%TODO: bibliography Meissner, 96. Frau, h, \textgreek{ω}

\inxitem[C]{leed} (ON \emph{ljóð}, OE \emph{léod})
  A magical chant or incantation, as seen by \textlink{Havamal} 153 near-synonymous with \inx[C]{galder}.  See also \inx[C]{gale}, \inx[C]{begale}.

\inxitem[C]{manwit} (ON \emph{man-vit})
  Common sense and wits.

\inxitem[C]{manicunning} (ON \emph{fjǫl-kunnigr})
  ‘Skilled with sorcery’.  \textlink{Havamal}[113], \textlink{Grimnismal}[P2], \textlink{Reginsmal}[P1]

\inxitem[C]{meat-nithing} (ON \emph{mat-níðingr})
  One who is a \inx[C]{nithing} with food, i.e. one who does not properly furnish his \inx[C]{guest}.

  See also \inx[C]{good of meat}.

\inxitem[C]{moon} (ON \emph{máni})
  the moon. --- For the personified \inx[P]{Moon} see under Proper Nouns.

\inxitem[C]{nithe} (ON \emph{níð}, OE \emph{níþ}, OHG \emph{níd})
  Originally ‘hatred, emnity’.  In the Norse the sense has developed in the direction of ‘shame’, not just as a social abstract, but almost a tangible thing.  So the curse ritual of Eyel, where the curser will “turn nithe” (\emph{snýja níð} against his enemy to cause him misfortune.  \inx[C]{scold}[Scolds] would “compose nithe” (\emph{yrkja níð}) through singing slanderous verses, which likewise had an adverse supernatural effect on their subject.  See also \inx[C]{nithing}.

\inxitem[C]{nithing} (ON \emph{níðingr}, OE \emph{níþing})
  One afflicted with \inx[C]{nithe}; a villain, criminal.  Among the Scandinavians a legal term; a nithing could not swear oaths or bear witness and was forbidden to marry.

\inxitem[C]{orlay} (ON \emph{ørlǫg}, OE \emph{orlæg})
  One’s predetermined fate, destiny, purpose as decreed by the \inx[P]{Norns}.

\inxitem[C]{queer} (ON \emph{argr, ragr} (with metathesis), OE \emph{earg}, OHG \emph{arg}.
  TODO.

\inxitem[C]{queerness} (ON \emph{ęrgi, ręgi})
  See \inx[C]{queer} above.

\inxitem[C]{rest} (ON \emph{rǫst})
  The distance between two rest-stops, a geographical mile (about 1850 metres).  See \CV: \emph{rǫst}.

\inxitem[C]{rune} (ON \emph{rún}, OE \emph{rún}, OS \emph{rúna}, OHG \emph{rúna}, Got. \emph{rúna}, PNWGmc. \emph{rūnu})
  An (esoteric) secret message or formula. That this---rather than ‘letter (of a Runic alphabet)’---is the original and proper sense is apparent from among others the Finnish borrowing \emph{runo} ‘poem; poetry; a division of a poem (specifically of the \emph{Kalevala})’, and its use in the singular in the earliest Runic inscriptions (e.g. Noleby Vg 63, which contains the linguistically indecipherable string of letters {ᚢᚾᚨᚦᛟᚢᛊᚢᚺᚢᚱᚨᚺᛊᚢᛊᛁᚺ[--]ᚨᛁ\rotatebox[origin=c]{180}{ᛏ}ᛁᚾ}, a \emph{rune} in the proper sense or the recently discovered Svingerud fragment.) Thus, Weden’s taking of the \emph{runes} should not be interpreted as merely a myth for the invention of profane writing, but rather the origin of esoteric incantations, not at all unlike Indian \emph{mántrās}.
  The word for letter was instead \inx[C]{stave}, see also there.

\inxitem[C]{scold} (ON \emph{skald})
  A Scandinavian court poet.  The name probably comes from their ability to slander with words.

\inxitem[C]{simble} (ON \emph{sumbl}, OE \emph{symbol})
  A banquet, symposium.

\inxitem[C]{spae} (ON \emph{spǫ́})
  Prophecy, foresight.

\inxitem[C]{Tables} (ON \emph{tafl}, OE \emph{tæfl})
  Generic term for board games (e.g. chess).  In the \inx[C]{golden age} the \inx[P]{Eese} played such games (\textlink{Voluspa} 8).  Pre-Christian Germanic burials commonly feature boards and bricks (TODO: reference, maybe to the Salme ship burials).

\inxitem[C]{thill} (ON \emph{þylja})
  To recite poetry learned by heart.  Cf.~the so called \inx[C]{thule}[thules] (poetic lists) and the title \inx[C]{thyle}.

\inxitem[C]{Thing} (ON, OE \emph{þing}, OS \emph{thing}, OHG \emph{ding})
  The Old Germanic assembly, where cases were settled and the law determined.  In connection with the Thing certain rituals were in order, viz. the enclosing of the space wherein the judges sat by means of \inx[C]{wigh-bonds} or sacred ropes.  Cf.~\textlink{Havamal} 61 for an excerpt from \emph{Germania} ch. 22.  See also the \inx[P]{Thing of the Gods}.

\inxitem[C]{thule} (ON \emph{þula})
  A poetic list, typically of various items of a category (e.g. gods, legendary horses) or poetic synonyms (e.g. for swords, men, Weden). Degoratively also a ditty, poorly composed poem. See \inx[C]{thyle}.

\inxitem[C]{thyle} (ON \emph{þulr}, OE \emph{þyle}, PNWGmc. \emph{*þuliʀ})
  A sage who through rote learning has acquired a large amount of mythological lore (cf.~\inx[C]{thule} ‘a list in poetic form; a ditty, bad poem’ and \inx[C]{thill} ‘to recite, to chant’).  Thus \inx[P]{Weden} is the \inx[P]{Fimblethyle}, being the unbeaten master of lore, as can be seen in his wisdom contests (like \textlink{Vafthrudnismal}).  Runic inscription DR 248 (Snoldelev) suggests that the thyle may have tied to a specific place, and in \Beowulf\ it seems to have been a court position, with the poet Unferth being described (l. 1456) as the “Rothgar’s thyle”.

\inxitem[C]{wale} (ON \emph{vǫlr})
  The staff or sceptre of a \inx[C]{wallow}.  TODO: archeological finds, mention Sutton Hoo.

\inxitem[C]{wallow} (ON \emph{vǫlva}, OE \emph{*wealwe} (cf.~ON \emph{svǫlva}, OE \emph{swealwe} ‘swallow’))
  A sibyl, seeress, oracle.  The word derives from the \inx[C]{wale}, a staff or sceptre probably used for ritual purposes.

\inxitem[C]{waste} (ON \emph{sóa})
  To ritually waste, to slay in a sacrificial context.

  \textlink{Havamal}[109]/4, 144/4 (n.), 145/3 (in \emph{of·sóa} ‘waste, kill too much’).

\inxitem[C]{wigh} (ON \emph{vé}, OE \emph{wéoh}, \emph{wíh}, PNWGmc. \emph{*wīhą})
  A holy place or sanctuary. It seems that where the \inx[C]{harrow} was a pile of stones or cairn used for carrying out rituals, the \textbf{wigh} was an enclosed space. The earliest Norse attestation is the runic inscription Ög N288 (Oklunda), which reads: “Guther <= Gunn·arr> painted these runes, and he fled, charged (with a crime, sought out this wigh, and he fled into this clearing. [...]” The implication seems to be that the wigh was considered so sacred that Guther could not be apprehended or punished for his crime while in it.

  In OE the word means ‘pagan idol’. It is not immediately clear which meaning is the original one, but in the present edition the Norse sense has been adopted, since the Anglo-Saxon sources are all of a Christian nature.  The name \emph{Wighstone} (\emph{Wíh-} or \emph{Wéohstān}) as found in \Beowulf\ in any case suggests it is the Norse meaning, since ‘idol-stone’ makes little sense.

\inxitem[C]{wode} (ON \emph{óðr}, OE \emph{wód}, PNWGmc. \emph{*wóduʀ})
  \inx[P]{Heener}’s gift to men, though the name may suggest it be from \inx[P]{Weden}. The word has several related meanings: ‘mind, (poetic) inspiration, rage’. See also \inx[P]{Woderearer}.

\inxitem[C]{wyrm} (ON \emph{ormr}, OE \emph{wyrm}, PNWGmc. \emph{*wurmiʀ})
  A dragon, serpent.  The distinction between “wyrm” and “worm; snake” is purely editorial and not made in the original languages.

\inxitem[C]{yin-} (ON \emph{ginn-})
  A rare augmentative prefix. TODO.

\inxitem[C]{yin-holy} (ON \emph{ginn-hęilagr})
  High holy, sacrosanct.  Used of the Gods in the formula \emph{ginn-hęilǫg goð} ‘yin-holy Gods’.

\end{itemize}

\section{Proper nouns (P)}

Indirect mentions (e.g.~\textlink{Vafthrudnismal}[54]/4 for Balder) follow the main entry; mentions based on textual emendation are italicized.

For names of people in Runic inscriptions see the end of each section.

\begin{itemize}

\inxitem[P]{Attle} m.an.,
  \emph{king of the \inx[P]{Huns}, historically from 434--453 \ce.}
  ON Atli, TODO.
  --- OE Ætla, \textlink{Widsith} 18; gen.~sg.~Ætlan: \char`~\ léodum, \textlink{Widsith} 122.
  [PGmc. \emph{*Attilô} whence MHG. \emph{Etzel}.  Latin \emph{Attila} borrowed from Gothic.]

\inxitem[P]{Attle 2}[2] m.an., ON Atli:
  \emph{son of the earl Idmund;} TODO.

\inxitem[P]{Balder} m., \emph{god, son of Weden, husband of Nan}.
  ON Baldr, \textlink{Voluspa}[59]/2, 59/3, \textlink{Grimnismal}[12]/1, \textlink{Hyndluljod}[28]/2, \Gylfaginning~49:77 (\textlink{EddicFragments}[F5][V]), \textlink{Hamdismal}[25]/2 (kenning; cf.~\inx[P]{Ermenric}); %
  acc.~sg.~Baldr, \textlink{Lokasenna}[28]/4, \Gylfaginning~49:81 (\textlink{EddicFragments}[F5][V]); %
  gen.~sg.~Baldrs: \textlink{EddicFragments}[F5][V] 1/2; \char`~\ bróðir, \textlink{Voluspa}[32]/3; \char`~\ faðir, \textlink{Hyndluljod}[29]/1; \char`~\ and·skota, \textlink{Voluspa}[33]/2, \textlink{Baldrsdraumar}[11]/4; \char`~\ bana, \textlink{Baldrsdraumar}[10]/4; %
  dat.~sg.~Baldri, \textlink{Voluspa}[31]/1 (n.), \textlink{Baldrsdraumar}[1]/4, 7/1, 8/3 \& 9/2, \textlink{Lokasenna}[27]/2.
  --- Note: áðr ȧ bál of bar/berr Baldrs and·skota: \textlink{Voluspa}[33]/2, \textlink{Baldrsdraumar}[11]/4; hvęrr/hann man Baldri at bana verða: \textlink{Baldrsdraumar}[8]/3, 9/2.
  --- Cpds.: \textbf{host-Balder} \ken{warrior}, ON hęr-Baldr: \textlink{Sigurdskamma}[18]/3.
  --- Indirect: ON Óðins barni: \textlink{Voluspa}[31]/2; syni \textlink{Vafthrudnismal}[54]/4 \& 55/2; ungum Óðins syni: \textlink{Skirnismal}[21]/2 \& 22/2 (\emph{cf.} \inx[P]{Dreepner}).
  --- OHG Balter; gen.~sg.~Balderes, \textlink{MerseburgTwo} 2.
  [PNWGmc. *baldraʀ ‘lord’ (> OE bealdor).  OE Bældæg, son of Wóden in the AS genealogies, cannot be a direct cognate.]

\inxitem[P]{Balethorn} m., ON Bǫl·þorn:
  \emph{obscure ettin, father of Bestle;} gen.~sg.~Bǫl·þorns: syni \char`~, \textlink{Havamal}[140]/1--2 (n.).

\inxitem[P]{Barrey} m., ON Barri:
  \emph{holy grove;} \textlink{Skirnismal}[39]/1 (n.) \& 41/1 (Barri hęitir \emph{etc.}).

\inxitem[P]{Beadhild} f.
  \emph{king \inx[P]{Nithad}’s daughter.}
  ON Bǫðv·ildr, \textlink{Volundarkvida}[P1], 19/1, 26/1, 29/4, 36/3;
  acc.~sg.~Bǫðv·ild, \textlink{Volundarkvida}[39]/2;
  gen.~sg.~Bǫðv·ildar, \textlink{Volundarkvida}[17]/2;
  dat.~sg.~Bǫðv·ildi, \textlink{Volundarkvida}[P2], 25/4 \& 36/2 (sló(’k) brjóst-kringlur, sęndi/sęnda’k Bǫðv·ildi);
  voc.~sg.~Bǫðv·ildr, \textlink{Volundarkvida}[40]/1. ---
  OE Beado·hild, dat.~sg.~Beado·hilde, \textlink{Deor}[2][8].

\inxitem[P]{Beal} f., ON Bęyla:
  \emph{minor agricultural goddess, wife of Bewer (1);} \textlink{Lokasenna}[P1], \emph{55/pr;} voc.~sg.~\char`~, \textlink{Lokasenna}[56]/1.

\inxitem[P]{Beck} m.an., ON Bikki:
  \emph{a cruel general associated with Attle or Ermenric;} \textlink{Gudrunarhvot}[P1] (twice); gen.~sg.~Bikka, \textlink{Sigurdskamma}[64]/1, \textlink{Atlakvida}[15]/2. ---
  OE Becca: \emph{obscure king, but likely the same as the Norse figure;} \textlink{Widsith}[4][19]; acc.~sg.~Beccan, \textlink{Widsith}[23][115].

\inxitem[P]{Bellower} m.jan., ON Bęli:
  \emph{obscure monster slain by \inx[P]{Free};} gen.~sg.~Bęlja: bani \char`~, \textlink{Voluspa}[51]/3 (n.)

\inxitem[P]{Bestle} f., ON Bęstla:
  \emph{wife of \inx[P]{Byre}, daughter of \inx[P]{Balethorn}, mother of \inx[P]{Weden}, \inx[P]{Will} and \inx[P]{Wigh};} gen.~sg.~Bęstlu: \char`~\ fǫður, \textlink{Havamal}[140]/2 (n.).

\inxitem[P]{Bewer}[Bewer 1] m., ON Byggvir:
  \emph{minor god of barley, Free’s servant, husband of Beal;} \textlink{Lokasenna}[P1], 43/pr, 45/1; gen.~sg.~Byggvis: \char`~\ kvę̇n, \textlink{Lokasenna}[56]/1; voc.~sg.~Byggvir, \textlink{Lokasenna}[46]/1.

\inxitem[P]{Bewer 2}[2] m., OE Béow:
  \emph{legendary Danish king, son of Shield;} TODO.

\inxitem[P]{Byre} m.i., ON Burr:
  \emph{early god, father of \inx[P]{Weden}, \inx[P]{Will}, and \inx[P]{Wigh};} gen.~sg.~Burs: \char`~\ synir, \textlink{Voluspa}[4]/1 (n.), \textlink{Hyndluljod}[29]/1.
  [burr ‘son’ < PGmc. *buriz ‘id.’, whence OE byre ‘id.’]

\inxitem[P]{Danes} m.i.pl.,
  \emph{tribe in eastern Denmark and southern Sweden.}
  ON Danir: gen.~pl.~Dana, \textlink{Helreid}[11]/4; dat.~pl.~Dǫnum, \textlink{GudrunTwo}[18]/1. --- Cpds.: \textbf{mountain-Danes} \ken{ettins}, ON berg-Danir: gen.~pl.~berg-Dana, \textlink{Hymiskvida}[17]/4. ---
  OE Dęní: TODO.

  --- See \inx[C]{Danish}.

\inxitem[P]{Dene} m.i., ON Danr:
  \emph{legendary ancestor of the \inx[P]{Danes};} \textlink{Rigsthula}[46]/1 (n.).

\inxitem[P]{Dreepner} m.ja., OE Draupnir:
  \emph{mythological duplicating ring.}  Indirect: baug: \textlink{Skirnismal}[21]/1 (n.), 22/1.

\inxitem[P]{Dreepner 2}[2] m.ja., OE Draupnir:
  \emph{dwarf;} \textlink{Voluspa}[15]/1.

\inxitem[P]{Dwarfs} (ON \emph{dvergar}, OE \emph{dweorgas}, OHG \emph{twerca}, PNWGmc. \emph{*dwergóʀ})
  Earthly (chthonic) supernatural beings, living in rocks and the earth.
  Noted members: TODO
  Occurrences: TODO

\inxitem[P]{Earp} m., ON Erpr:
  \emph{son of \inx[P]{Enacker} and half-brother of \inx[P]{Hamthew} and \inx[P]{Sarle};} \textlink{Gudrunarhvot}[P1] (n.), \textlink{Hamdismal}[12]/1, 28/1.

\inxitem[P]{Earp and Oatle} m., a.~and a., ON Erpr ok Ęitill:
  \emph{the murdered sons of \inx[P]{Attle} and \inx[P]{Guthrun};} \textlink{GudrunTwo}[P1]; acc.~Erp né Ęitil, \textlink{Atlakvida}[41]/2; gen.~Erps \dots\ ok \dots\ Ęitils, \textlink{Hamdismal}[8]/1--2.

\inxitem[P]{Earth} (ON \emph{jǫrð}, OE \emph{eorþe}, OHG \emph{erda}, PNWGmc. \emph{*erþu}, PGmc. \emph{*erþó})
  The personified Earth.  By \inx[P]{Weden} the mother of \inx[P]{Thunder}.

\inxitem[P]{Eastern Way} (ON \emph{Austr-vegr})
  In the mythology the eastern lands of the \inx[P]{Ettins}, to which \inx[P]{Thunder} goes to fight the Ettins and protect the realms of Gods and Men; see also \inx[P]{Ettinham}.  In geography referring to Eastern Europe and Asia, but with a connotation of barbaric wilderness.

\inxitem[P]{Eese} (rhyming with \emph{geese}; ON \emph{ę́sir}, OE \emph{ése}, PNWGmc. \emph{*ansiwiʀ}; sg. \emph{os}, ON \emph{ǫ́ss}, OE \emph{ós}, PNWGmc. \emph{*ansuʀ})
  The (male) gods. \Gylfaginning\ and \YnglingaSaga\ present them as a separate tribe from the \inx[P]{Wanes}. See also \inx[P]{Bonds}, \inx[P]{Gods}, \inx[P]{Reins}, \inx[P]{Farrows} \inx[P]{Tews}; for the goddesses see \inx[P]{Ossens}.
  Noted members: \inx[P]{Balder}, \inx[P]{Free}, \inx[P]{Hath}, \inx[P]{Homedall}, \inx[P]{Tew}, \inx[P]{Thunder}, \inx[P]{Weden}
  Occurrences: TODO

\inxitem[P]{Elves} (ON \emph{alfar}, OE \emph{ielfe}, PNWGmc. \emph{*alβíʀ})
  Earthly (chthonic) minor deities.  Possibly ancestral spirits?
  Noted members: TODO
  Occurrences: TODO

\inxitem[P]{Enacker} m., ON Jȯn·akr:
  \emph{third husband of \inx[P]{Guthrun}, father of \inx[P]{Earp}, \inx[P]{Hamthew} and \inx[P]{Sarle};} gen.~sg.~Jȯn·akrs, \textlink{Sigurdskamma}[62]/4, \textlink{Gudrunarhvot}[P1]; \char`~\ sonum, \textlink{Sigurdskamma}[63]/1; \char`~\ sonu, \textlink{Gudrunarhvot}[14]/4, \textlink{Hamdismal}[25]/4.
  [? + *wakraz ‘quick, awake’.]

\inxitem[P]{Ermenric} m.
  \emph{king of the Eastern \inx[P]{Gots}, historically ca.~300--376 \ce.}
  ON Jǫrmun·rek(k)r, \textlink{Sigurdskamma}[64]/2, \textlink{Gudrunarhvot}[2]/3, \textlink{Hamdismal}[3]/2, 20/1;
  dat.~sg.~Jǫrmun·rekki, \textlink{Gudrunarhvot}[5]/3, \textlink{Hamdismal}[19]/1; Jǫrmun·rekk: \char`~ inum ríkja, \textlink{Gudrunarhvot}[P1];
  voc.~sg.~Jǫrmun·rek(k)r, \textlink{Hamdismal}[24]/2, 24/5.
  --- OE Eorman·ríc, \textlink{Widsith}[4][18]; gen.~sg.~Eorman·ríces, \textlink{Widsith}[1][8], 111, \textlink{Deor}[5][21]; dat.~sg.~Eorman·ríce, \textlink{Widsith}[19][88].
  [Latin (H)ermanaricus from Gothic *airman ‘great, immense’ + reiks ‘king’.  Cf.~MHG Ermen·rích in legendary ballads.]

\inxitem[P]{Ettinham} (ON \emph{Jǫtun-hęimr}, \emph{Jǫtna-hęimar})
  The ‘\inx[P]{Ettins}[Ettin]-\inx[C]{Home}’ or ‘home of the Ettins’; the eastern realm of chaotic and inhospitable beings.  See also \inx[P]{Eastern Way}, \inx[P]{Outyards}.

\inxitem[P]{Ettins} (ON \emph{jǫtnar}, OE \emph{eotenas}, PNWGmc. \emph{*etunóʀ})
  The fundamental enemies of the Gods, the agents of chaos and disorder. See \inx[P]{Rises}, \inx[P]{Thurses}.
  Noted members: \inx[P]{Hymer}, \inx[P]{Thrim}, \inx[P]{Webthrithner}, \inx[P]{Yimer}
  Occurrences: TODO

\inxitem[P]{Fallowhooven} m.ja., ON Fal·hófnir:
  \emph{one of the gods’ horses;} \textlink{Grimnismal}[30]/1.
  [fǫlr + hófr + -nir]

\inxitem[P]{Fathomer} m.ja., ON Fáfnir:
  \emph{the wyrm slain by \inx[P]{Siward}, son of \inx[P]{Rethmar}, brother of \inx[P]{Otter} and \inx[P]{Rein};} acc.~sg.~Fáfni, TODO; gen.~sg.~Fáfnis, \textlink{Gripisspa}[13]/1, 15/4, TODO, \textlink{GudrunTwo}[P1]; bani \char`~, \textlink{Oddrunargratr}[17]/4; dat.~sg.~Fáfni, \textlink{Gripisspa}[11]/4, TODO, \textlink{Helreid}[10]/4.

\inxitem[P]{Fenhalls} m.i.pl., ON Fęn-salir:
  \emph{\inx[P]{Frie}’s dwellings;} dat.~pl.~Fęn-sǫlum, \textlink{Voluspa}[33]/3 (n.).

\inxitem[P]{Fenrer} m.ja., ON Fęnrir:
  \emph{evil wolf, one of \inx[P]{Lock}’s three children;} \textlink{Vafthrudnismal}[46]/4 (n.), 47/2; \textlink{Lokasenna}[38]/4.

\inxitem[P]{Fenrerswolf} m., ON Fęnris-ulfr:
  TODO.

\inxitem[P]{Fimblethyle} m.i., ON Fimbul·þulr:
  \emph{name for \inx[P]{Weden}}; \textlink{Havamal}[80]/4 \& 142/4.
  [‘Great \inx[C]{thyle}[Thyle]’.]

\inxitem[P]{Fimble-winter} m., ON fimbul-vetr:
  \emph{the Great Winter before the destruction of the earth;} \textlink{Vafthrudnismal}[44]/4. [fimbul- ‘fimble-, great’ + vetr ‘winter’.]

\inxitem[P]{Fold} (ON \emph{Fold}, OE \emph{Folde})
  A poetic or ritual name of \inx[P]{Earth}, especially in her role as Mother Earth.  In Germanic poetry the word \emph{fold} is typically used to simply refer to ‘land’, however.  It is cognate with Sanskrit TODO.

\inxitem[P]{Folkwong} m., ON Folk-vangr:
  \emph{\inx[P]{Frow}’s heavenly dwelling;} \textlink{Grimnismal}[14]/1 (n.).
  [folk ‘troop, army’ + vangr ‘meadow’.]

\inxitem[P]{Foresitter} m.an., ON For·seti:
  \emph{god associated with legal proceedings (\inx[P]{Thing}[Things]), son of Balder and Nan;} \textlink{Grimnismal}[15]/3 (n.)
  [TODO. ‘Presider.’]

\inxitem[P]{Free} m.ja., ON Fręyr:
  \emph{god, son of \inx[P]{Nearth}, brother of \inx[P]{Frow};} \textlink{Skirnismal}[P1], 4/pr, 6/pr, 9/pr, 20/3 (vit \char`~), 33/2, P4, \emph{42/pr}, \textlink{Lokasenna}[P1], 37/1, 41/pr, \textlink{Hyndluljod}[29]/2;
  acc.~sg.~Fręy, \textlink{Skirnismal}[P1], 19/3;
  gen.~sg.~Fręys, \textlink{Lokasenna}[44]/3; skó-sveinn \char`~, \textlink{Skirnismal}[P1]; þjónustu-menn \char`~, \textlink{Lokasenna}[P1]; \char`~\ vinar, \textlink{Sigurdskamma}[24]/4;
  dat.~sg.~Fręy, \textlink{Grimnismal}[5]/3, 44/3 (skírum \char`~);
  voc.~sg.~Fręyr, \textlink{Skirnismal}[3]/1.
  [< PNWGmc. *frawjaʀ ‘lord’, = OE fréa ‘id.’.  Likely from earlier Yngvi-Fręyr ‘Lord Ing’, see \textlink{Lokasenna}[43]/1 n.]
  --- Cpds.: \textbf{Ingwin-Free}, ON Ingunar-Fręyr: \textlink{Lokasenna}[43]/1 (n.).
  --- Indirect: bani Bęlja, bjartr: \textlink{Voluspa}[51]/3 (n.); folk-valdi goða: \textlink{Skirnismal}[3]/1; nýtum Njarðar bur: \textlink{Grimnismal}[44]/4; inum þroska Njarðar syni: \textlink{Skirnismal}[38]/4; Njarðar syni: \textlink{Skirnismal}[39]/3 \& 41/3; okkarn mǫg: \textlink{Skirnismal}[1]/2; mǫg: \textlink{Skirnismal}[2]/2; ykkrum syni: \textlink{Skirnismal}[2]/1; inn fróði afi: \textlink{Skirnismal}[1]/3--4 \& 2/3--4; sęggr: \textlink{Skirnismal}[5]/2; mïnn dróttinn: \textlink{Skirnismal}[3]/4.
  --- See \inx[P]{Ing}, \inx[P]{Ingwin}, \inx[P]{Waning}.

\inxitem[P]{Frie} f.ō.
  \emph{goddess, wife of \inx[P]{Weden}, (in the Norse:) mother of \inx[P]{Balder}.}
  ON Frigg: \textlink{Voluspa}[33]/3, \textlink{Vafthrudnismal}[2][\emph{2/pr,}] \emph{4/pr,} \textlink{Grimnismal}[P2] (3 times), \textlink{Lokasenna}[P1], \emph{25/pr, 27/pr,} 29/3, G1/3, \Gylfaginning\ 35:1 (in \textlink{EddicFragments}[F4][IV]), \textlink{Oddrunargratr}[9]/2; gen.~sg.~Friggjar, \textlink{Voluspa}[51]/4; voc.~sg.~Frigg, \textlink{Vafthrudnismal}[1]/1, \textlink{Lokasenna}[26]/1, 28/1.
  --- OHG Frija: \textlink{MerseburgTwo} 4.
  --- Latin Frea: \emph{Origo Gentis Langbardorum} 1:9 (n.), 1:10, 1:12; acc.~sg.~Fream, \emph{Origo Gentis Langbardorum} 1:8.
  [PGmc. *Frijjō, orig.~f.~sg.~of lost adj.~*frijjaz ‘dear, beloved’.  Latin Frea is from Longbeardish.  Cf.~OE *Fríġ (gen.~sg.~fríġe, e.g. in fríġe-dæġ ‘Friday’, lit.~‘Frie’s day’).]
  --- See \inx[P]{Line}.

\inxitem[P]{Frow} f.ōn., ON Fręyja:
  \emph{goddess, daughter of \inx[P]{Nearth}, sister of \inx[P]{Free}, wife of \inx[P]{Wode};} \textlink{Grimnismal}[14]/1 (n.), \textlink{Lokasenna}[P1], \emph{29/pr, 31/pr,} \textlink{Thrymskvida}[13]/1, 26/3 \& 28/3, \textlink{Oddrunargratr}[9]/2, \textlink{Hyndluljod}[1][\emph{1/1,}] \emph{7/1};
  acc.~sg.~Fręyju, \textlink{Thrymskvida}[8]/4 \& 11/4, 12/1, 22/3;
  gen.~sg.~\char`~, \textlink{Thrymskvida}[3]/1, 23/4 27/3;
  voc.~sg.~Fręyja, \textlink{Lokasenna}[30]/1 \& 32/1, 32/4, \textlink{Thrymskvida}[3]/3, 12/3, \textlink{Hyndluljod}[6]/1.
  [PNWGmc. *frawjǭ ‘lady’, = OHG frouwa ‘id.’]
  --- Note: fǿri(ð \dots) Fręyju at kvę̇n/kvǫ̇n: \textlink{Thrymskvida}[8]/4, 11/4, 22/3; át/svaf vę́tr Fręyja átta nǫ́ttum: \textlink{Thrymskvida}[26]/3, 28/3; Þęgi þú, Fręyja: \textlink{Lokasenna}[30]/1, 32/1.
  --- Indirect: Óðs męy: \textlink{Voluspa}[24]/4 (n.); Njarðar dóttur ór Nóa-tu̇num: \textlink{Thrymskvida}[22]/4.

\inxitem[P]{Full} (ON \emph{Fulla}, OHG \emph{Folla})
  In the Norse sources the maid-servant of \inx[P]{Frie}.  \textlink{MerseburgTwo} has her as Frie’s sister, though this need not be literal (cf.~\textlink{Hyndluljod}[1]).

\inxitem[P]{Gangred} m.pl., ON Gang·ráðr:
  \emph{name for \inx[P]{Weden};} \textlink{Vafthrudnismal}[8]/1 (\char`~ hęiti’k); voc.~sg.~\char`~, \textlink{Vafthrudnismal}[9]/1, \textlink{Vafthrudnismal}[11]/1, \textlink{Vafthrudnismal}[13]/1, \textlink{Vafthrudnismal}[15]/1, \textlink{Vafthrudnismal}[17]/1.
  [gangr ‘a walk’ + -ráðr ‘decider, counsellor’, i.e., ‘one who chooses to wander’.]

\inxitem[P]{Gap of Hawks} (ON \emph{Ginnunga-gap})
  The air or lower atmospheric sky, the midspace between \inx[P]{Earth} and \inx[P]{Up-heaven}; not synonymous with the latter.
  Occurrences: \textlink{Voluspa}[3]/4 (n.)  It is more prevalent in \Gylfaginning.

\inxitem[P]{Geatland} (ON \emph{Gaut-land, Gauta-land})
  The land of the \inx[P]{Geats} in southwestern Sweden.

\inxitem[P]{Geats} (ON \emph{gautar}, OE \emph{géatas}, PNWGmc. \emph{*gautóʀ} from \emph{*geut-} ‘to pour’, perhaps ‘the libators’)
  A tribe in what is today southern-central Sweden. See also \inx[P]{Geatland}, \inx[P]{Swedes}.
  Noted members: TODO
  Occurrences: TODO

\inxitem[P]{god} n., ON goð:
  \emph{a god, regardless of gender;} dat.~sg.~goði, \textlink{Grimnismal}[40]/1--2 (inu skír-lęita goði); skïnanda \char`~, \textlink{Grimnismal}[39]/2, \textlink{Sigrdrifumal}[15]/1.
  [OE \emph{god}, OHG \emph{got}, PGmc. \emph{*gudą}.]

\inxitem[P]{Gods} n.pl., ON goð:
  \emph{the Gods, often used in a pantheistic sense in the pl.;}
  \textlink{Hymiskvida}[22]/3, \textlink{Lokasenna}[49]/4 \& 50/2, \textlink{Fafnismal}[15]/2, \textlink{Oddrunargratr}[9]/2;
  goð’in ǫll, \textlink{Voluspa}[22]/4;
  goð ǫll, \textlink{Lokasenna}[45]/2, 55/4;
  ǫll \char`~, \textlink{Hyndluljod}[49]/4;
  in svǫ́su \char`~, \textlink{Vafthrudnismal}[17]/4 \& 18/2;
  ginn-hęilǫg \char`~, \textlink{Voluspa}[6]/2, 9/2, 22/2, 24/2;
  acc.~pl.~goð, \textlink{Hymiskvida}[3]/2, \textlink{Lokasenna}[12]/4, \textlink{Lokasenna}[55]/4;
  gen.~pl.~goða, \textlink{Vafthrudnismal}[50]/3, \textlink{Skirnismal}[33]/4, \textlink{Reginsmal}[19]/2; gǫrðum \char`~, \textlink{Vafthrudnismal}[2]/2; vé \char`~, \textlink{Vafthrudnismal}[51]/1; hodd \char`~, \textlink{Grimnismal}[27]/5; þing \char`~, \textlink{Hymiskvida}[39]/1;
  skír brúðr \char`~\ (\emph{kenning for \inx[P]{Shede}}), \textlink{Grimnismal}[11]/3; vǫrðr \char`~\ (\emph{kenning for \inx[P]{Homedall}}), \textlink{Grimnismal}[13]/3, \textlink{Lokasenna}[48]/4; folk-valdi \char`~\ (\emph{kenning for \inx[P]{Free}}), \textlink{Skirnismal}[3]/1;
  þrúð-valdr \char`~\ (\emph{kenning for \inx[P]{Thunder}}), \textlink{Harbardsljod}[9]/4; allra \char`~, \textlink{Vafthrudnismal}[42]/3 \& 43/1, \textlink{Grimnismal}[43]/1;
  dat.~pl.~goðum, \textlink{Vafthrudnismal}[15]/4 \& 16/2, 39/2, \textlink{Grimnismal}[51]/2, 56/4, \textlink{Lokasenna} T1, 34/2 \& 35/2, \textlink{Alvissmal}[4]/4, 12/1, 14/1, 16/1, 18/1, 20/1, 22/1, 24/1, 28/1, 30/1, 32/1, \textlink{Hyndluljod}[8]/3, 27/5, \textlink{Atlamal}[54]/5; vǫrðr með \char`~, \textlink{Skirnismal}[28]/4;
  voc.~pl.~goð, \textlink{Lokasenna}[7]/1; ǫll ginn-hęilǫg \char`~, \textlink{Lokasenna}[11]/2.
  [OE \emph{godu}, OHG \emph{gota}, PNWGmc. \emph{*gudō}, pl. of \emph{*gudą}.]

\inxitem[P]{Guther} (ON \emph{Gunn·arr}, MHG \emph{Gunther})
  The lord of the \inx[P]{Gots}.  In the Norse sources the brother of \inx[P]{Hawn}.  Historically he is based on king \emph{Gundaharius} (\emph{*Gunþiharjaz}) of the Burgundians.

\inxitem[P]{Guthlathe} (ON \emph{Gunnlǫð})
  Daughter of the ettin \inx[P]{Sutting}; she guarded the \inx[C]{Mead of Poetry} in the mountain, but gave it to \inx[P]{Weden} after he seduced her.

  See \textlink{Havamal} 103--110.

\inxitem[P]{Guthrun} (ON \emph{Guð·ru̇n})
  Daughter of king \inx[P]{Yivick}, sister of \inx[P]{Guther} and \inx[P]{Hawn}. The wife of \inx[P]{Attle}.

\inxitem[P]{Hamthew} m.ja., ON Ham·ðir:
  \emph{son of \inx[P]{Enacker} and \inx[P]{Guthrun}, brother of \inx[P]{Sarle} and \inx[P]{Earp};} \textlink{Gudrunarhvot}[P1], 4/1, 8/1, \textlink{Hamdismal}[6]/1, 24/1, 26/1, 31/2; acc.~sg.~Ham·ði, \textlink{Hamdismal}[21]/2;
  gen.~sg.~Ham·ðis, \textlink{Hamdismal} T1, P1;
  voc.~sg.~Ham·ðir, \textlink{Hamdismal}[27]/1.
  --- Note: Þȧ kvað þat Ham·ðir inn hugum stóri, \textlink{Gudrunarhvot}[4]/1, 8/1; Hitt kvað þȧ Ham·ðir inn hugum stóri, \textlink{Hamdismal}[6]/1, 24/1, 26/1.
  [= Ammius in Jordanes, \emph{Getica}; see Introduction to \textlink{Gudrunarhvot}.]

\inxitem[P]{Hawn}[Hawn 1] (ON \emph{Hǫgni}, OE \emph{Haguna}, \emph{Hagena}, OHG \emph{Hagano}, Ger. \emph{Hagen}, PNWGmc. \emph{*Hagunó})
  A \inx[P]{Nivlings}[Nivling] and \inx[P]{Yivickings}[Yivicking], son of king \inx[P]{Yivick}, brother of \inx[P]{Guther} and \inx[P]{Guthrun}. In \textlink{Atlakvida} he defeats seven warriors before being captured by \inx[P]{Attle}, who has his heart cut out at the request of Guther.

\inxitem[P]{Hawn 2}[2]
  A petty king of \inx[P]{East Geatland}, contemporary with \inx[P]{Granmer}, the king of \inx[P]{Southmanland} and Ingeld Illred, the \inx[P]{Inglings}[Ingling] king of \inx[P]{Upland}.

\inxitem[P]{Hath} m.u., ON Hǫðr:
  \emph{god, killer of his brother \inx[P]{Balder};} \textlink{Voluspa}[32]/2, 59/3, \textlink{Baldrsdraumar}[9]/1; dat.~sg.~Hęði, \textlink{Baldrsdraumar}[10]/3.
  [*haþuz ‘war, battle’]

\inxitem[P]{Heener} (ON Hø̇nir, PNWGmc. *Hónijaʀ ‘the little swan(?)’)
  \emph{obscure (bird?-)god;} \textlink{Voluspa}[17]/3 (n.), 60/1, \textlink{Reginsmal}[P1].

\inxitem[P]{Hell}[Hell 1] (ON \emph{hęl}, OE \emph{hęll}, PNWGmc. \emph{*halju}, Got. \emph{halja})
  The Underworld, personified as and formally identical to \inx[P]{Hell}, the goddess.  After the arrival of Christianity this word came to refer to the Christian hell-fire (= \emph{Gehenna}), which is the case in all attested languages apart from the Old Norse.  Cf.~\inx[P]{Nivelhell}.

\inxitem[P]{Hell}[2] (ON \emph{Hęl})
  Goddess of death and Hell (the Underworld).

\inxitem[P]{Hindle} (ON \emph{Hyndla})
  A witch awoken by \inx[P]{Frow} in \textlink{Hyndluljod}.

\inxitem[P]{Homedall} ON Hęim·dallr:
  \emph{god born of nine mothers, Watchman of the Gods;} \textlink{Voluspa}[45]/3, \textlink{Lokasenna}[47][\emph{47/pr}], \textlink{Thrymskvida}[15]/1 (\char`~, hvítastr Ȧsa), \textlink{Rigsthula}[P1], \Gylfaginning~27:1 (under \textlink{EddicFragments}[F3][III]);
  acc.~sg.~Hęim·dall, \textlink{Grimnismal}[13]/1;
  gen.~sg.~Hęim·dalar, \textlink{Voluspa}[26]/1; mǫgu \char`~, \textlink{Voluspa}[1]/2; Hęim·dallar, \Gylfaginning~27:15 (under \textlink{EddicFragments}[F3][III]);
  voc.~sg.~Hęim·dallr, \textlink{Lokasenna}[48]/1.
  [hęimr ‘home, (the settled) world’ + *dallr ‘brilliant, shining’.  See \textlink{Vafthrudnismal}[25]/1 n., \textlink{Thrymskvida}[13]/3 n.] ---
  Indirect: TODO.

\inxitem[P]{Huns} (ON \emph{húnir}, OE \emph{Húne}, OHG \emph{Húni}, \emph{Hunni}, PNWGmc. \emph{*húníʀ})
  Invading Asiatic tribe in the Migration Period.  In the Scandinavian legends they have been assimilated into the Germanic framework, and are not presented as racially or culturally distinct; part of this is undoubtedly due to their confusion with a Danish tribe (TODO).
  Noted members: \inx[P]{Attle}, TODO
  Occurrences: TODO

\inxitem[P]{Hymer} m., ON Hymir:
  \emph{ettin, \inx[P]{Tew}’s father according to \textlink{Hymiskvida}}; 5/2, 7/4, 10/2, \emph{16/3pr, 17/3pr, 19/3pr}, 21/1, 25/3, \emph{26/pr, 32/pr}, 39/2; gen.~sg.~Hymis, \textlink{Hymiskvida} T1 (in \AM), 15/4, 30/3; \char`~\ męyjar, \textlink{Lokasenna}[34]/3; dat.~sg.~Hymi, \textlink{Hymiskvida} 29/4, 35/3; voc.~sg.~\textlink{Hymiskvida}[11]/1 (ves hęill, Hymir).
  --- Indirect: mïnn faðir: \textlink{Hymiskvida}[5]/3 (n.); mïnn fríi: \textlink{Hymiskvida}[9]/3; karl: \textlink{Hymiskvida}[32]/3; gen.~sg.~karls: \textlink{Hymiskvida}[10]/4 (n.); dat.~sg.~karli: \textlink{Hymiskvida}[31]/3; %
  jǫtunn, \textlink{Hymiskvida}[14]/4, 20/3, 28/1; gen.~sg.~jǫtuns: \textlink{Hymiskvida}[27]/4; forn jǫtunn: \textlink{Hymiskvida}[13]/3; ballr jǫtunn: \textlink{Hymiskvida}[17]/2; ȯ·tęitr jǫtunn: \textlink{Hymiskvida}[25]/1; kost-móðs jǫtuns: \textlink{Hymiskvida}[30]/4; %
  hǫ́rum Hrungnis spjalla: \textlink{Hymiskvida}[16]/1; kjóla valdi: \textlink{Hymiskvida}[19]/4; átt-runn apa: \textlink{Hymiskvida}[20]/2.

\inxitem[P]{Idewold} m.u., ON Iða-vǫllr:
  \emph{the ‘Plain of Industry’, where the \inx[P]{Gods} settle and build \inx[P]{Osyard};} dat.~sg.~-vęlli: \textlink{Voluspa}[7]/1, 57/1.

\inxitem[P]{Idin} f.i., ON Ið·unn:
  \emph{goddess, keeper of the golden apples of immortality (cf.~\textlink{Skirnismal}[19]/1 n.), wife of \inx[P]{Bray};} \textlink{Lokasenna}[P1], \emph{16/pr, 18/pr}; voc.~sg.~\char`~, \textlink{Lokasenna}[17]/1. [ið + -unnr.]

\inxitem[P]{Ilewaves} m.pl., ON Éli-vágar:
  \emph{chaotic (primæval) salty waters;} acc.~pl.~Éli-vága, \textlink{Hymiskvida}[5]/1 (n.); dat.~pl.~Éli-vǫ́gum, \textlink{Vafthrudnismal}[31]/1 (n.).

\inxitem[P]{Ing} (ON \emph{Yngvi}, OE \emph{Ing})
  Probably an older name of \inx[P]{Free}. The legendary ancestor of the \inx[P]{Inglings}. Cf.~Ingwin-Free, Ing-Free.
  Occurrences: TODO, the Old English Rune Poem.

\inxitem[P]{Inglings} (ON \emph{Ynglingar}, PNWGmc. \emph{*Ingwalingóʀ} ‘the descendants of \inx[P]{Ing}’)
  The oldest known Swedish kingly lineage.  The difference between Inglings and \inx[P]{Shelvings} is a bit unclear; they seem to be used synonymously in the Norse sources but \Beowulf\ knows them only as Shelvings.

\inxitem[P]{Leered} m.a., ON Lę́·raðr:
  \emph{name for \inx[P]{Ugdrassle’s Ash};} gen.~sg.~Lę́·raðs: \textlink{Grimnismal}[25]/2 (n.) \& 26/2 (ok bítr af \char`~\ limum).

\inxitem[P]{Life and Lafthrasher} f. \& m.ja., ON Líf ok Lęif·þrasir:
  \emph{the only surviving humans who repopulate the earth after \inx[P]{Rakes of the Reins};} \textlink{Vafthrudnismal}[45]/1 (n.).

\inxitem[P]{Lithshelf} f.ō., ON Hlið-skjǫlf:
  \emph{high lookout post of the Gods from which they can see the whole world;} acc.~sg.~Hlið-skjǫlf, \textlink{Skirnismal}[P1]; dat.~sg.~Hlið-skjǫlfu, \textlink{Grimnismal}[P2] (n.).
  [hlið ‘cliff, slope’ + \emph{skjǫlf} ‘shelf’, i.e. the ‘Cliffside Shelf’, probably situating it on a high mountain.]

\inxitem[P]{Lock} (ON \emph{Loki})
  The bound Os. TODO.

\inxitem[P]{Loride} m., ON Hlór·riði:
  \emph{poetic name for \inx[P]{Thunder};} \textlink{Hymiskvida}[27]/1, 29/1; acc.~sg.~Hlór·riða, \textlink{Lokasenna}[55]/2; gen.~sg.~Hlór·riða, \textlink{Hymiskvida}[16]/2, 37/2; \char`~\ hamar, \textlink{Thrymskvida}[7]/4 \& 8/1 (\char`~\ hamar of folginn), 14/4; dat.~sg.~Hlór·riða, \textlink{Hymiskvida}[4]/3, \textlink{Lokasenna}[54]/4, \textlink{Thrymskvida}[31]/1. ---
  [‘Loud/Roaring Rider’.]

\inxitem[P]{Lother} m., ON Lóðurr, OS Logaþore, PNWGmc. *Loga-þorjaʀ ‘Flame-darer(?)’,
\emph{obscure god;} \textlink{Voluspa}[17]/4 (n.).

\inxitem[P]{Middenyard} m.a.,
  \emph{(the ‘Middle Enclosure’), the inhabited world.}
  ON Mið·garðr: acc.~sg.~-garð, \textlink{Voluspa}[4]/2 (n.), \textlink{Grimnismal}[42]/2 (n.); gen.~sg.~-garðs: \char`~\ Vé·urr, \textlink{Voluspa}[53]/5; dat.~sg.~-garði, \textlink{Harbardsljod}[23]/4, \textlink{Hyndluljod}[11]/6, 16/4, Sö 56 2.
  --- OE Middan-ġeard: TODO.
  --- OS Middil-gard (\emph{also} f.cons.): TODO.
  --- OHG Mittila-gart: \textlink{Muspilli} 53.
  [Cf.~Gothic midjun-gards ‘the world’.]
  See \inx[P]{Osyard}, \inx[P]{Wighward}, \inx[P]{Middenyardswyrm}.

\inxitem[P]{Millner} m.ja., ON Mjǫllnir:
  \emph{\inx[P]{Thunder}’s hammer;} \textlink{Lokasenna}[57]/2, 59/2, 61/2, 63/2; acc.~sg.~Mjǫllni, \textlink{Vafthrudnismal}[51]/3, \textlink{Thrymskvida}[30]/3; dat.~sg.~Mjǫllni, \textlink{Hymiskvida}[36]/2 (Mjǫllni morð-gjǫrnum).
  --- Note: þér skal mïnn þrúð-hamarr, Mjǫllnir, mál fyr·nema: \textlink{Lokasenna}[57]/2, 59/2, 61/2, 63/2.
  --- [PGmc. *Meldunjaʀ.]

\inxitem[P]{Moon} m.an., ON Mȧni:
  \emph{the personified moon in the Eddas, son of \inx[P]{Mundlefarer} and brother of the \inx[P]{Sun};}
  \textlink{Voluspa}[5]/4, \textlink{Vafthrudnismal}[22]/3, \textlink{Alvissmal}[13]/3, 14/1; acc.~sg.~\textlink{Havamal}[137]/7; gen.~sg.~Mȧna: sinni \char`~, \textlink{Voluspa}[5]/1; \char`~\ faðir, \textlink{Vafthrudnismal}[23]/1; systur \char`~, \textlink{Reginsmal}[23]/2. ---
  For the \inx[C]{moon} in general see Index to cultural terms above.

\inxitem[P]{Mundlefarer} m.an., ON Mundil·fǿri:
  \emph{father of \inx[P]{Sun} and \inx[P]{Moon};} \textlink{Vafthrudnismal}[23]/1 (n.).  [\Gylfaginning~11:2 has Mundil·fari.  *mundill (= mǫndull ‘mill-handle’) + fǿri ‘driver, mover’.  The simplex form mǫndull has prob.~been influenced by the cognate of OE handle < PWGmc. *handulā (*handu ‘hand’ + *-lā), but the cpd.~element mundil- < PN *mundilaʀ (*mundu ‘hand’ + *-ilaʀ) preserves the original root vowel.]

\inxitem[P]{Narrow} m.wa., ON Nǫrr:
  \emph{father of \inx[P]{Night};} dat.~sg.~Nǫrvi, \textlink{Vafthrudnismal}[25]/2 (n.), \textlink{Alvissmal}[29]/3. [*nǫrr ‘narrow’ (an adj.~only surviving in this name) < PGmc.~*narwaz, cognate with OE nearu ‘narrow, constrained’.  Also found in the weak form Nǫrvi ‘the Narrow One’ in \Gylfaginning\ 10 (cited in \textlink{Vafthrudnismal}[11]/ALL n.).]

\inxitem[P]{Nears} m.~pl., ON Njárar \char`~\ Níarar:
  \emph{Swedish tribe in \textlink{Volundarkvida}, ruled by king \inx[P]{Nithad};} gen.~pl.~Níara: \char`~\ dróttinn, \textlink{Volundarkvida}[7]/1 (n.), 14/1, 30/4.

\inxitem[P]{Nearth} m.u., ON Njǫrðr:
  \emph{a god, one of the \inx[P]{Wanes}, father of \inx[P]{Free} and \inx[P]{Frow};} %
  \textlink{Vafthrudnismal}[38]/3, \textlink{Grimnismal}[16]/1, \textlink{Skirnismal}[P1], \textlink{Lokasenna}[P1], 33/pr, 35/pr, \Gylfaginning\ 23:1, 6, 7, 8, 10 (under \textlink{EddicFragments}[F2][II]); %
  gen.~sg.~Njarðar: \char`~\ bur, \textlink{Grimnismal}[44]/4; %
  sonr \char`~, \textlink{Skirnismal}[P1]; %
  \char`~\ syni, \textlink{Skirnismal}[38]/4, 39/3 \& 41/3 (þar mun Njarðar syni \emph{etc.}); %
  \char`~\ dóttur, \textlink{Thrymskvida}[22]/4;
  voc.~sg.~Njǫrðr, \textlink{Lokasenna}[34]/1, 36/1.
  [= OE *Neorþ]
  --- Cpds.: \textbf{spear-Nearth} \ken{warrior}, ON gęir-Njǫrðr: \textlink{Gudrunarhvot}[8]/3.

\inxitem[P]{Nithad} m.u.,
  \emph{tyrant that imprisons \inx[P]{Wayland}, father of \inx[P]{Beadhild}, lord of the \inx[P]{Nears.}}  ON Níð·uðr, \textlink{Volundarkvida}[P1] (twice), 7/1, 14/1, P2, \emph{31/pr, 32/pr,} 38/2, 41/1;
  gen.~sg.~Níð·aðar, 20/4, 29/2; kvǫ̇n \char`~, \emph{16/1} \& 30/1 (Úti stóð kunnig kvǫ̇n \char`~), \emph{25/2} \& 35/4 (sęndi/sęnda’k kunnigri kvǫ̇n \char`~);
  dat.~sg.~Níð·aði, \textlink{Volundarkvida} T2, 18/1, 20/2, 24/4 \& 35/2 (svęip(’k) útan silfri, sęldi/sęlda’k \char`~);
  voc.~sg.~\textlink{Volundarkvida}[30]/4.
  --- Note: Níð·uðr, Níara dróttinn: \textlink{Volundarkvida}[7]/1, 14/1, 30/4.
  --- OE Níþ·had, \textlink{Deor}[1]/5; gen.~sg.~Níð·hades: \char`~\ mǽg, \textlink{Waldere}[B] 8.
  [PNWGmc. *Níþa-haduz < *níþaz ‘shame, emnity’; *haþuz ‘war, battle’.]

\inxitem[P]{Nithehewer} m., ON Níð·hǫggr:
  \emph{watery dragon and corpse-eater;} \textlink{Voluspa}[38]/4, 62/4, \textlink{Grimnismal}[36]/4; dat.~sg.~Níð·hǫggvi, \textlink{Grimnismal}[32]/4.

\inxitem[P]{Nivelhell} f.ō., ON nifl-hęl:
  \emph{a synonym for \inx[P]{Hell}, lower Underworld;} acc.~sg.~nifl-hęl, \textlink{Vafthrudnismal}[43]/4; gen.~sg.~nifl-hęljar, \textlink{Baldrsdraumar}[2]/3.
  [nifl ‘mist, shade’ + hęl ‘Hell, the Underworld’; ‘Mist-Hell’, ‘Shady Hell’.]

\inxitem[P]{Norns} f.i.pl., ON nornir:
  \emph{(not infrequently malevolent) women who shape the \inx[C]{orlay}s of men;} nom.~sg.~norn, \textlink{Reginsmal}[2]/3 (aumlig \char`~); gen.~sg.~nornar, \textlink{Sigrdrifumal}[15]/12;
  nom.~pl.~nornir, \textlink{HelgakvidaOne}[2]/1, \textlink{Fafnismal}[12]/3, 13/1, \textlink{Sigurdskamma}[7]/3 (ljótar \char`~), \textlink{GudrunTwo}[36]/1;
  acc.~pl.~\char`~, \textlink{HelgakvidaTwo}[20]/2, \textlink{Atlakvida}[18]/4 (n.);
  gen.~pl.~norna, \textlink{Fafnismal}[11]/1 (n.), 44/4, \textlink{Hamdismal}[29]/3, 30/4;
  dat.~pl.~nornum, \textlink{Gudrunarhvot}[13]/1.
  --- See \inx[C]{orlay}, \inx[P]{Dises}, \inx[P]{Mothers}, \inx[P]{Weird}, \textlink{Voluspa}[19].

\inxitem[P]{Oneharriers} m.pl., ON ęin-hęrjar:
  \emph{Weden’s chosen warriors who train in Walhall for the \inx[P]{Rakes of the Reins};}  \textlink{Vafthrudnismal}[41]/1, \textlink{Grimnismal}[18]/4, \textlink{HelgakvidaOne}[38]/3; gen.~pl.~ęin-hęrja, \textlink{Grimnismal}[24]/3; dat.~pl.~ęin-hęrjum, \textlink{Grimnismal}[37]/6, 53/3.

\inxitem[P]{Ossens} f.ōn.pl., ON ǫ̇synjur:
  \emph{the goddesses, wives of the \inx[P]{Eese};} \textlink{Baldrsdraumar}[1]/2, \textlink{Lokasenna}[31]/3, \textlink{Thrymskvida}[14]/2; acc.~pl.~\char`~, \textlink{Hyndluljod}[10]/4; voc.~pl.~\char`~, \textlink{Lokasenna}[11]/1, \textlink{Sigrdrifumal}[4]/1.
  [PGmc. *ansuz ‘Os’ + *-unjō f. ‘-ess, -en’; cf.~vargr, vargynja.]
  --- Note: ok ǫ̇synjur allar ȧ máli: \textlink{Baldrsdraumar}[1]/2, \textlink{Thrymskvida}[14]/2; Hęilir ę̇sir, hęilar ǫ̇synjur: \textlink{Lokasenna}[11]/1, \textlink{Sigrdrifumal}[4]/1.

\inxitem[P]{Osyard} m.a., ON Ǫ̇s·garðr:
  \emph{the stronghold (yard, enclosure) of the \inx[P]{Eese};} acc.~sg.~-garð, \textlink{Thrymskvida}[18]/3; gen.~sg.~-garðs: \char`~\ jǫfurr, \textlink{IcelandicRunePoem}[4]; dat.~sg.~-garði, \textlink{Hymiskvida}[7]/2, \Skaldskaparmal\ 42:2 (in \textlink{EddicFragments}[7][VII]).
  [ǫ̇ss ‘Os, god’ + garðr ‘enclosure’, i.e.~the ‘Enclosure of the \inx[P]{Eese}’.]
  --- See \inx[P]{Middenyard}, \inx[P]{Outyards}.

\inxitem[P]{Oughter} (ON \emph{Óttarr}, OE \emph{Óhthere}, PNWGmc. \emph{*Óhta-harjaʀ})
  Legendary Swedish Ingling king.

\inxitem[P]{Outyards} (ON \emph{Út-garðar})
  Not Eddic.  The ‘Outer Enclosures’, described in \Gylfaginning. See also \inx[P]{Ettinham}, \inx[P]{Middenyard}, \inx[P]{Osyard}.

\inxitem[P]{Rakes of the Age} n.pl., ON aldar rǫk:
  \emph{the \inx[P]{Rakes of the Reins};} acc.~pl.~aldar rǫk, \textlink{Vafthrudnismal}[39]/3.
  [ǫld ‘age, generation’, rǫk ‘unfolding, events’.]

\inxitem[P]{Rakes of the Reins} n.pl., ON ragna rǫk:
  \emph{the destruction of the world in pre-Christian eschatology (cf.~\textlink{Voluspa}[39]--55, \textlink{Vafthrudnismal}[44]--53);} \textlink{Baldrsdraumar}[14]/4 (n.), \textlink{HelgakvidaOne}[40]/2; nom.~pl.~rǫk ragna, \textlink{Atlamal}[21]/3; acc.~pl.~ragna rǫk, \textlink{Voluspa}[43]/4, 47/4, 55/4, \textlink{Vafthrudnismal}[55]/4.
  [ręgin ‘Reins, Gods’, rǫk ‘unfolding, events’.]
  --- See \inx[P]{Rakes of the Age}, \inx[P]{Ripping of the Age}, \inx[P]{Twilight of the Reins}, \inx[P]{Rippers}.

\inxitem[P]{Reading} m.a., ON Hrauðungr:
  \emph{legendary king of the Gots;} \char`~\ konungr, \textlink{Grimnismal}[P1]; gen.~sg.~Hrauðungs: sonum \char`~\ konungs, \textlink{Grimnismal} T1; dat.~sg.~Hrauðungi, \textlink{Hyndluljod}[25]/2.

\inxitem[P]{Reins} (ON \emph{rǫgn}, \emph{ręgin})
  The ‘Powers, Counsels’.  Judging from \textlink{Vafthrudnismal} TODO the term may be more closely associated with the \inx[P]{Wanes} than the \inx[P]{Eese}.

\inxitem[P]{Ripping of the Age} n., ON aldar rof:
  \emph{the \inx[P]{Rakes of the Reins};} \textlink{HelgakvidaTwo}[41]/2.
  [ǫld ‘age, generation’, rof ‘ripping, destruction’.]

\inxitem[P]{Rotholf} (ON \emph{Hrólfr kraki}, OE \emph{Hróþulf}, PNWGmc. \emph{*Hróþi-wulfaʀ})
  A king of the \inx[P]{Shieldings} (see family tree). As foreshadowed in \Beowulf\ 1017--9, 1180--90, he betrays the sons of \inx[P]{Rothgar}, his cousins \inx[P]{Rethric and Rothmund}, in order to take the throne for himself. In the later Icelandic tradition this has been forgotten, and he is consistently portrayed as a heroic king.

\inxitem[P]{Rothgar} (ON \emph{Hróarr}, OE \emph{Hróþgár}, PNWGmc. \emph{*Hróþi-gaiʀaʀ})
  A king of the \inx[P]{Shieldings} (see family tree), one of the main characters in \Beowulf.

\inxitem[P]{Rungner} (ON \emph{Hrungnir})
  Famous ettin fought by Thunder.  The full story is told in \Haustlong\ 14--20 and \Skaldskaparmal\ 24--25, which cites the former.

\inxitem[P]{Sarle} m.an., ON Sǫrli:
  \emph{son of \inx[P]{Enacker} and \inx[P]{Guthrun}, brother of \inx[P]{Hamthew} and \inx[P]{Earp};} \textlink{Gudrunarhvot}[P1], \textlink{Hamdismal}[9]/1, 31/2;
  acc.~sg.~Sǫrla, \textlink{Hamdismal}[21]/2.
  [PGmc. *Sarulô, diminutive of Sarus in Jordanes, \emph{Getica}; see Introduction to \textlink{Gudrunarhvot}.]

\inxitem[P]{Saxons} (ON \emph{saxar}, OE \emph{Seaxan}, \emph{Seaxe})
  TODO.
  Noted members: TODO
  Occurrences: TODO

\inxitem[P]{Shede} f.īn., ON Skaði:
  \emph{goddess, daughter of \inx[P]{Thedse}, wife of \inx[P]{Nearth};} \textlink{Grimnismal}[11]/3, \textlink{Skirnismal}[P1], \textlink{Lokasenna}[P1], \emph{49/pr, 51/pr,} P8, \textlink{Hyndluljod}[29]/5, \Gylfaginning~23:7, 8, 11, 12 (\textlink{EddicFragments}[F2][II]). ---
  [OE *Scędi, PGmc. *Skadį̄, not to be confused with skaði ‘injury’ < PGmc. *skaþô.  Most likely = OE Scędi- in \inx[P]{Shedeny} and the \inx[P]{Shedelands}; Shede was presumably thought to guard the Scandinavian peninsula.]

\inxitem[P]{Shelvings} (ON \emph{skilfingar}, OE \emph{scilfingas}, PNWGmc. \emph{*skilβingóʀ})
  The descendants of \inx[P]{Shelf}; the legendary \inx[P]{Swedes}[Swedish] royal dynasty. The exact difference between the terms Shelvings and \inx[P]{Inglings} is unclear, but the first may have referred to the old royal family in Sweden, while the latter to the Norwegian branch which claimed descent from the former. TODO
  Noted members: TODO
  Occurrences: \textlink{Hyndluljod}[15], 20

\inxitem[P]{Shield} (ON \emph{Skjǫldr}, OE \emph{Scyld}, PNWGmc. \emph{*Skelduz})
  Legendary Danish king, founder of the \inx[P]{Shieldings}.

\inxitem[P]{Shieldings} (ON \emph{skjǫldungar}, OE \emph{Scyldingas}, PNWGmc. \emph{*skeldungóʀ})
  The descendants of \inx[P]{Shield}; the legendary \inx[P]{Danes}[Danish] royal dynasty. With \inx[P]{Harward}’s death after his slaying of \inx[P]{Rotholf} their rule ended. TODO
  Noted members: TODO
  Occurrences: TODO

\inxitem[P]{Shirner} m., ON Skírnir:
  \emph{Free’s servant;} \textlink{Skirnismal}[P1], 1/1, 2/pr, 3/pr, 5/pr, 8/pr, P2, P3, \emph{13/pr, 18/pr, 21/pr, 23/pr, 25/pr, 38/pr,} P4, 40/1, \emph{41/pr,}

\inxitem[P]{Syemund} (ON \emph{Sig-mundr}, OE \emph{Sige-mund}, MHG. \emph{Sieg-mund}, PNWGmc. \emph{*Sigi-munduʀ})
  In the Norse tradition the son of king \inx[P]{Walsing}.  He begets \inx[P]{Siward}, the slayer of the wyrm \inx[P]{Fathomer}.  In \Beowulf\ it is Syemund himself who slays an unnamed wyrm.  Connected with his nephew \inx[P]{Sinfittle}.

\inxitem[P]{Sithguth} (OHG \emph{Sinthgunt}, PNWGmc. \emph{*Sinþa-gunþiʀ}(?))
  Only known from \textlink{MerseburgTwo} as the sister of \inx[P]{Sun}.

\inxitem[P]{Siward} (ON \emph{Sig·urðr})
  A hero of the \inx[P]{Walsings}, slayer of the \inx[C]{wyrm} \inx[P]{Fathomer}.

\inxitem[P]{Sun} (ON \emph{Sól}, OHG \emph{Sunna})
  The personified Sun, who in the Germanic mythology is a woman.  In \textlink{Vafthrudnismal} 22 the daughter of \inx[P]{Mundlefarer} and sister of \inx[P]{Moon}.  In \textlink{MerseburgTwo} the sister of \inx[P]{Sithguth}.

\inxitem[P]{Swanhild} f.i., ON Svan·hildr:
  \emph{daughter of \inx[P]{Siward} and \inx[P]{Guthrun} trampled by King \inx[P]{Ermenric};} \textlink{Sigurdskamma}[55]/3, \textlink{GudrunTwo}[P1], \textlink{Gudrunarhvot}[P1], 15/3, \textlink{Hamdismal}[3]/1; acc.~sg.~-hildi, \textlink{Sigurdskamma}[63]/2, \textlink{Gudrunarhvot}[P1], 8/5, 15/1; gen.~sg.~-hildar, \textlink{Gudrunarhvot}[16]/4, \textlink{Hamdismal}[2]/5.
  (For her historicity see \emph{Getica} 129, cited in Introduction to \textlink{Gudrunarhvot}.)
  [ON svanr ‘swan’ + hildr ‘battle, Hild (a walkirrie)’ but this is prob.~later reanalysis; the form Sunilda (\emph{Getica} 129) suggests Gothic sunja ‘truth’ + *hilds ‘battle, Hild’.]

\inxitem[P]{Swedes} (ON \emph{svíar}, OE \emph{swéon}, PNWGmc. \emph{*swihaníʀ})
  The tribe around the Mälar valley in eastern Sweden.
  Noted members: TODO
  Occurrences: TODO

\inxitem[P]{Tew} m.wa., ON Týr:
  \emph{one-handed god, possibly son of \inx[P]{Hymer} (cf.~\textlink{Hymiskvida}[5]/3 n.);} \textlink{Hymiskvida}[4]/3, \emph{6/2pr,} 33/3, \textlink{Lokasenna}[P1], 37/pr, 39/pr; acc.~sg.~Tý, \textlink{Sigrdrifumal}[6]/4;
  voc.~sg.~Týr: Þęgi þú, \char`~, \textlink{Lokasenna}[38]/1, 40/1.
  [OE Tíw (Tíwes dæg ‘Tuesday’), PGmc.~*tīwaz ‘god’ < PIE *deywós ‘id.’, directly cognate with OI \emph{\textoi{dēvá}} ‘id.’, Latin deus ‘id.’.  Tew is not linguistically or mythologically identical to OI \emph{\textoi{Dyāús}}, Greek \textgreek{Ζεύς}, Latin Iuppiter which derive from a different derivation of the same root.]

\inxitem[P]{Tews} m.wa.pl., ON tívar:
  \emph{the \inx[P]{Gods};} \textlink{Grimnismal}[5]/4; ríkir tívar, \textlink{Baldrsdraumar}[1]/3 \& \textlink{Thrymskvida}[14]/3 (ok umb þat réðu \char`~); mę́rir tívar, \textlink{Hymiskvida}[4]/1; acc.~pl.~tíva, \textlink{Havamal}[160]/2; gen.~sg.~tíva, \textlink{Vafthrudnismal}[38]/1 \& 42/1 (þú \char`~\ rǫk ǫll Vaf·þrúðnir vitir).
  [PGmc. *tīwōz, pl.~of *tīwaz; see Tew above.] ---
  Cpds.: \textbf{slain-Tews}, ON val-tívar: \textlink{Voluspa}[59]/4 (n.), \textlink{Hymiskvida}[1]/1; gen.~pl.~val-tíva, \textlink{Voluspa}[50]/2 (n.).
  \textbf{victory-Tews}, ON sig-tívar: gen.~pl.~sig-tíva: of ragna rǫk, rǫmm \char`~, \textlink{Voluspa}[43]/4, 47/4, 55/4; \char`~\ sonum, \textlink{Grimnismal}[46]/1; \char`~\ synir, \textlink{Lokasenna}[1]/4, 2/2, \textlink{Fafnismal}[24]/2; Guð·ru̇n \char`~, \textlink{Atlakvida}[32]/3.

\inxitem[P]{Thedse} m.an., ON Þjatsi:
  \emph{ettin, father of \inx[P]{Shede};} \textlink{Grimnismal}[11]/1, \textlink{Hyndluljod}[29]/4; acc.~sg.~Þjatsa, \textlink{Harbardsljod}[19]/1; gen.~sg.~Þjatsa: dóttir \char`~\ jǫtuns, \Gylfaginning~23:7 (\textlink{EddicFragments}[F2][II]); dat.~sg.~Þjatsa, \textlink{Lokasenna}[50]/4 \& 51/2.

\inxitem[P]{Thing of the Gods} ON þing goða:
  \emph{the Heavenly Council, the Assembly where the Gods convene and make decisions;} acc.~sg.~\char`~, \textlink{Hymiskvida}[39]/1. --- Indirectly: þingi ‘Thing’ \textlink{Voluspa}[46]/2, \textlink{Baldrsdraumar}[1]/1 (n.), \textlink{Thrymskvida}[14]/1; rǫk-stóla ‘council-seats’, \textlink{Voluspa}[6]/1 (n.), 9/1, 22/1, 24/1; \textlink{Grimnismal}[29]--30. --- See \textlink{Voluspa}[6]/1--2 n., \textlink{Baldrsdraumar}[1]/1--3 n.

\inxitem[P]{Thrim} m., ON Þrymr:
  \emph{ettin who steals \inx[P]{Thunder}’s hammer in \textlink{Thrymskvida};} \textlink{Thrymskvida}[6]/1, \emph{7/pr, 8/pr,} 11/2, 22/1, 25/1, 30/1; acc.~sg.~Þrym \textlink{Thrymskvida}[31]/3; gen.~sg.~Þryms \textlink{Thrymskvida} T1. --- Note: Þrym(r) \dots\ þursa dróttin: \textlink{Thrymskvida}[6]/1, 11/2, 31/3; Þȧ kvað þat Þrymr, þursa dróttin: \textlink{Thrymskvida}[22]/1, 25/1, 30/1.

\inxitem[P]{Thrithham} m., ON Þrúð-hęimr:
  \emph{\inx[P]{Thunder}’s heavenly abode;} dat.~sg.~-hęimi \textlink{Grimnismal}[4]/3.

\inxitem[P]{Thrithyelmer} m.ja., ON Þrúð-gęlmir:
  \emph{ettin, father of \inx[P]{Bareyelmer};} \textlink{Vafthrudnismal}[29]/3.

\inxitem[P]{Thunder} m., (ON Þȯrr, OE *Þunor, OS Þuner, OHG Donar, PNWGmc. *Þonaraʀ)
  ON Þȯrr: \emph{god of Thunder, son of \inx[P]{Weden} and \inx[P]{Earth};} \textlink{Voluspa}[25]/1, \textlink{Grimnismal}[4]/3, 29/2, \textlink{Harbardsljod}[P1], \emph{3/pr, 5/pr, 7/pr, 9/pr,} 11/pr, \emph{13/pr, 15/pr, 17/pr,} 19/pr, 23/pr, 24/4, 25/pr, 26/1, 26/3, 27/pr, 29/pr, 31/pr, 33/pr, 35/pr, 37/pr, 39/pr, 41/pr, 43/pr, 45/pr, 47/pr, 49/pr, 50/2, 51/pr, 53/pr, 55/pr, 57/pr, 59/pr, \textlink{Hymiskvida} T1 (in \Regius), 23/1 (Þȯ\emph{a}rr), \textlink{Lokasenna}[P1], P7, \emph{59/pr,} 60/4, \emph{61/pr, 63/pr,} %
  \textlink{Thrymskvida}[17]/1, \textlink{Alvissmal}[2][\emph{2}]\emph{/pr, 4/pr, 6/pr, 8/pr, 11/pr, 13/pr, 15/pr, 17/pr, 19/pr, 21/pr, 23/pr, 25/pr, 27/pr, 29/pr, 31/pr, 33/pr, 35/pr}, \Skaldskaparmal\ TODO (\textlink{EddicFragments}[F6][VI] P1 (3 times), P2 (2 times), P3 (3 times)), \CanterburyGalder\ 3, N B380/2; %
  acc.~sg.~Þȯr, \textlink{Harbardsljod}[9]/4, 56/4, \textlink{Hymiskvida}[28]/2 (Þȯ\emph{a}r), \textlink{Thrymskvida}[9]/4, 15/3, 19/1, \textlink{Hyndluljod}[4]/1;
  gen.~sg.~Þȯrs, \textlink{Harbardsljod}[P1], \Skaldskaparmal\ TODO  (\textlink{EddicFragments}[F6][VI] P2);
  voc.~sg.~Þȯrr, \textlink{Harbardsljod}[18]/9, 22/3, 28/2, 32/1, 36/1, 38/1, \textlink{Lokasenna}[58]/2; Þęgi þú, \char`~, \textlink{Thrymskvida}[18]/2. ---
  Cpds.: \textbf{Thunder of the Eese}, ON Ȧsa-Þȯrr: gen.~sg.~Ȧsa-Þȯrs, \textlink{Harbardsljod}[52]/1.  \textbf{Wing-Thunder}, ON Ving-Þȯrr: \textlink{Thrymskvida}[1]/1 (n.), \textlink{Alvissmal}[6]/1. --- See \inx[P]{Loride}, \inx[P]{Wighward}, \inx[P]{Wingner}, TODO.

\inxitem[P]{Thurses} (sg. Thurse; ON \emph{þurs}, OE \emph{þyrs}, OS \emph{thuris}, OHG \emph{duris}, PNWGmc. \emph{*þurisaʀ})
  Possibly a poetic synonym for \inx[P]{Ettins}. See also \inx[P]{Rime-Thurses}.
  Noted members: TODO
  Occurrences: TODO

\inxitem[P]{Twilight of the Reins} n., ON ragna røkr:
  \emph{the \inx[P]{Rakes of the Reins};} gen.~sg.~ragna røkrs, \textlink{Lokasenna}[39]/4.
  [ręgin ‘Reins, Gods’, røk(k)r ‘twilight’.]

\inxitem[P]{Ugdrassle’s Ash} m.a., ON askr Ygg·drasils:
  \emph{the world tree and site of the \inx[P]{Thing of the Gods}}; \textlink{Voluspa}[45]/5 (Ygg·drasils askr), \textlink{Grimnismal}[36]/1, 45/1; acc.~sg.~ask Ygg·drasils, \textlink{Voluspa}[18]/1 (Ask vęit’k standa, hęitir Ygg·drasils); dat.~sg.~aski Ygg·drasils, \textlink{Grimnismal}[29]/4 \& 30/5 (dag hvęrn es dǿma fęrr/fara at \char`~), 31/2, 32/2, 34/1.
  [Likely Yggr ‘Ug, Weden’ + *drasill, variant of drǫsull ‘horse, steed’, referring to Weden’s hanging (cf.~\textlink{Havamal}[138]).]
  --- See \inx[P]{Leered}, \inx[P]{Hoardmimer’s Wood}.

\inxitem[P]{Up-heaven} (ON \emph{upp-himinn}, OE \emph{up-heofon}, OS \emph{upp-himil}, OHG \emph{úf-himil})
  Highest Heaven; used in \inx[F]{Earth and Up-heaven}.

\inxitem[P]{Wadyelmer} m.ja., ON Vað·gęlmir:
  acc.~sg.~Vað·gęlmi, \textlink{Reginsmal}[4]/2.
  [\emph{vað} n. ‘shoal, water shallow enough to wade’ + \emph{gęlmir} ‘screamer(?)’]

\inxitem[P]{Waleshelf} f.ō., ON Vala-skjǫlf:
  \emph{mythological place;} \textlink{Grimnismal}[6]/3.

\inxitem[P]{Walfather} (ON \emph{Val-fǫðr})
  ‘Father of the Slain’; name for \inx[P]{Weden}.

  \textlink{Voluspa} 1/3a, 26/4a, 28/4a, \textlink{Grimnismal} 49/2a

\inxitem[P]{Walhall} f., ON Val-hǫll:
  \emph{Weden’s hall;} \textlink{Grimnismal}[8]/2; gen.~sg.~Val-hallar, \textlink{Voluspa}[33]/4, \textlink{HelgakvidaTwo}[P2], Icelandic Rune Poem 4/2, \Skaldskaparmal\ 42 (\textlink{EddicFragments}[F7][VII]); dat.~sg.~Val-hǫllu, \textlink{Grimnismal}[24]/2, \textlink{Hyndluljod}[1]/4, \textlink{Atlakvida}[2]/2(?), 16/2(?).

\inxitem[P]{Walsings} (ON \emph{Vǫlsungar})
  Descendants of king \inx[P]{Walsing}.

\inxitem[P]{Wanes} (ON \emph{vanir}, OE \emph{wan-?})
  A subgroup or tribe of the gods, associated with fertility, harvests and the sea.
  Noted members: \inx[P]{Nearth}, \inx[P]{Free}, \inx[P]{Frow}
  Occurrences: TODO

\inxitem[P]{Wayland} m.u.,
  \emph{legendary smith.}
  ON Vǫlundr, \textlink{Volundarkvida}[P1] (thrice), \emph{4/2}, 5/3 \& 7/2, 9/4, \emph{13/pr, 15/pr,} P2, P3, \emph{22/pr}, 27/pr, \emph{28/pr,} 29/1, 29/3, \emph{33/pr}, 38/1, 41/2 (vit Vǫlundr);
  voc.~sg.~Vǫlundr, \textlink{Volundarkvida}[14]/2 \& 32/1 (\char`~, vísi alfa), 37/2, 40/2 (it Vǫlundr).
  --- Note: Vǫlundr líðandi of langan veg: \textlink{Volundarkvida}[4][\emph{4/2,}] 9/4; ęinn Vǫlundr sat ï Ulf-dǫlum: \textlink{Volundarkvida}[5]/3, 7/2; hlę́jandi Vǫlundr: \textlink{Volundarkvida}[29]/3, 38/1.
  --- OE Wélund, \textlink{Deor}[1]; gen.~sg.~Wélandes, \textlink{Beowulf}[455], \textlink{Waldere}[A] 2; \char`~\ bearn, \textlink{Waldere}[B] 9.

\inxitem[P]{Webthrithner}, m.ja., ON Vaf·þrúðnir:
  \emph{ettin defeated by \inx[P]{Weden} in a wisdom contest;} \textlink{Vafthrudnismal}[6]/1, \emph{7/pr, 9/pr, 11/pr, 13/pr, 15/pr, 17/pr}, 19/pr, 21/pr, 23/pr, 25/pr, 27/pr, 29/pr, 31/pr, 33/pr, 35/pr, 37/pr, 39/pr, 41/pr, \emph{43/pr,} 45/pr, 47/pr, 49/pr, 51/pr, 53/pr, 55/pr; acc.~sg.~Vaf·þrúðni, \textlink{Vafthrudnismal}[2]/4; gen.~sg.~Vaf·þrúðnis, \textlink{Vafthrudnismal} T1, 1/2, 3/3.

\inxitem[P]{Weden} (ON \emph{Óðinn}, OE \emph{Wóden}, \emph{Wéden}, OHG \emph{Wuotan}, PNWGmc. \emph{*Wódanaʀ} ‘Lord of \inx[C]{wode} (poetry, intelligence)’)
  Chief of the \inx[P]{Eese}, God of Wisdom, Galder, Poetry, War.  Husband of \inx[P]{Frie}, and by her father of \inx[P]{Balder}.  Father of \inx[P]{Thunder} by \inx[P]{Earth}.  Brother of \inx[P]{Heener} and \inx[P]{Lother} or \inx[P]{Will} and \inx[P]{Wigh}.

\inxitem[P]{Weird} m.an., ON Urðr:
  \emph{one of three Norns at Ugdrassle’s Ash (for non-personified use see under cultural terms above);} \textlink{GudrunOne}[24]/3; acc.~sg.~Urð, \textlink{Voluspa}[19]/3 (n.).
  [PNGmc. *wurdiz ‘fate, destiny’.]
  --- See \inx[P]{Shild}, \inx[P]{Well of Weird}, \inx[P]{Werthing}.

\inxitem[P]{Wider} m., ON Víð·arr:
  \emph{son and avenger of \inx[P]{Weden} against the \inx[P]{Fenrerswolf}}; \textlink{Voluspa}[52]/2, \textlink{Vafthrudnismal}[51]/1, 53/2, \textlink{Lokasenna}[P1], P4; gen.~sg.~Víð·ars, \textlink{Voluspa}[H48]/4, \textlink{Grimnismal}[17]/2; voc.~sg.~Víð·arr, \textlink{Lokasenna}[10]/1. ---
  [OE *Wíd·hęri, PGmc. *Wída-harjaʀ ‘Broad Warrior’, a fitting name considering he props the jaws of the Wolf apart, which are said to reach up to the very Heaven (\textlink{Voluspa}[H48] n.).]

\inxitem[P]{Wigh} m.an., ON Véi:
  \emph{brother of \inx[P]{Weden} and \inx[P]{Will}, son of \inx[P]{Byre} (cf.~\textlink{Voluspa}[4]/1, 16/2 nn.);} acc.~sg.~Véa, \textlink{Lokasenna}[26]/3 (n.).
  [PNGmc. *wíhá ‘hallower, (heathen) priest’.]

\inxitem[P]{Wighward} m.a., ON Véurr:
  \emph{poetic name for Thunder;} \textlink{Voluspa}[53]/5 (Mið·garðs \char`~), \textlink{Hymiskvida}[11]/5, 17/1, 21/4 (Óðni sifjaðr \char`~).
  [PN \emph{*Wíha-warjaʀ}, \emph{*wíhą} ‘wigh, sanctuary’ + \emph{*warjaz} ‘defender’.]

\inxitem[P]{Will} m.an., ON Vili:
  \emph{brother of \inx[P]{Weden} and \inx[P]{Wigh}, son of \inx[P]{Byre} (cf.~\textlink{Voluspa}[4]/1, 16/2 nn.);} acc.~sg.~Vilja, \textlink{Lokasenna}[26]/3 (n.).
  [PNWGmc. *wiljá ‘willpower’.]

\inxitem[P]{Wingner} m.a., ON Vingnir:
  \emph{poetic name for Thunder;} gen.~sg.~Vingnis, \textlink{Vafthrudnismal}[51]/4.
  [Probably ‘swinger’, viz.~of the Hammer \inx[P]{Millner}.]

\inxitem[P]{Wing-Thunder} m., ON Ving-Þȯrr:
  \emph{name for \inx[P]{Thunder}; see there under Cpds.}

\inxitem[P]{Wode} m.a.(\emph{or} u.?), ON Óðr:
  \emph{obscure husband of \inx[P]{Frow};} gen.~sg.~Óðs: \char`~\ męy, \textlink{Voluspa}[24]/4.
  [OE \emph{Wód}]

\inxitem[P]{Wonnel} m.an., ON Vȧli:
  \emph{god, son of \inx[P]{Weden} and \inx[P]{Wrind} who avenges his brother \inx[P]{Balder} by killing his half-brother \inx[P]{Hath};} \textlink{Voluspa}[H31] (n.), \textlink{Vafthrudnismal}[51], \textlink{Hyndluljod}[28]/3; acc.~sg.~Vȧla, \textlink{Baldrsdraumar}[11][\emph{11/1}].
  [< PGmc. \emph{*Wanilô} ‘the little \inx[P]{Wanes}[Wane]’ (uncertain).]
  --- Indirectly: Baldrs bróðir: \textlink{Voluspa}[32]/3; Óðins sonr: \textlink{Voluspa}[32]/4, \textlink{Baldrsdraumar}[11]/2.

\inxitem[P]{Wolder} m.u., ON Ullr:
  \emph{obscure winter-god;} \textlink{Grimnismal}[5]/1; gen.~sg.~Ullar, \textlink{Grimnismal}[43]/1 (n.), \textlink{Atlakvida}[33]/4.
  [PGmc. *wulþuz ‘glory, splendour’, OE *Wuldor.]

\inxitem[P]{Wrind} f., ON Vrind, Rind:
  \emph{mother of \inx[P]{Wonnel} (cf.~\textlink{Voluspa}[31] n.);}  \textlink{Baldrsdraumar}[11]/1 (n.).
  [‘Ivy’(?); \textlink{Baldrsdraumar}[11]/1 n.]

\inxitem[P]{Yimer} m.ja., ON Ymir:
  \emph{first \inx[P]{Ettins}[ettin] sacrificed by the Gods who created the world from his corpse;} \textlink{Voluspa}[3]/1 (n.); %
  gen.~sg.~Ymis, \textlink{Vafthrudnismal}[21]/1 \& \textlink{Grimnismal}[41]/1 (Ór Ymis holdi \emph{et c.}), \textlink{Vafthrudnismal}[28]/3; %
  dat.~sg.~Ymi, \textlink{Hyndluljod}[32]/4. [OE *Ġyme, PGmc. *Jumjaz.  See \textlink{Voluspa}[3]/1 for etymology.]

\inxitem[P]{Yin-Reins} (ON \emph{ginn-ręgin})
  \inx[C]{yin-} + \inx[P]{Reins}. The sacrosanct, highest Divine Powers.

\inxitem[P]{Yivick} m.an.,
  \emph{king of the \inx[P]{Burgends} (historically from late 4th c.--407 \ce).}
  ON Gjúki: \textlink{GudrunTwo}[P1], 1/3, \textlink{Atlamal}[102]/2; acc.~sg.~Gjúka, \textlink{Sigurdskamma}[1]/1; TODO.
  --- OE Gifica: \textlink{Widsith}[4][19].
  [Cf.~OHG Gibicho, MHG. Gibeche (in legendary ballads).]

\inxitem[P]{Yivickings} m.pl., ON Gjúkungar:
  \emph{descendants of \inx[P]{Yivick};} TODO.
  [Gjúki ‘Yivick’ + -ungar ‘sons, descendants of’]

\end{itemize}

\section{Poetic formulæ (F)}
All formulæ are given in English translation, their attested forms and a Proto-Germanic rendition. For those consisting of two words bound together by a conjunction, \& is written in its place.

\begin{itemize}
\inxitem[F]{Earth and Up-heaven} (ON \emph{jǫrð \& upphiminn}, OE \emph{eorþe \& upheofon}, OS \emph{erþa \& uphimil}, OHG \emph{erdo \& úfhimil}, PGmc. \emph{*erþō \& uphiminaz})
  An old merism; earth and heaven and everything in between, i.e. the whole universe. It has a particular connection to the creation and destruction of the world, and in prayers.
  ON: \textlink{Voluspa} 3/3, \textlink{Vafthrudnismal} 20, \textlink{Thrymskvida} 2, \textlink{Oddrunargratr} 17, SJy 41 (under Galders), Sö 154 (under Runic Poetry);
  OE: \emph{Acreboot};
  OS: \textlink{Heliand} 2886;
  OHG: \textlink{Wessobrunn} 2.

\inxitem[F]{Eese and Elves} (ON \emph{ę́sir \& alfar}, OE \emph{ése \& ielfe}, PNWGmc. \emph{*alβíʀ \& ansiwiʀ})
  A merism; both heavenly and earthly spiritual beings.  Notably the two words always occur in this order (never ‘Elves and Eese’), even in OE.

\inxitem[F]{words and works} (ON \emph{orð \& verk}, OE \emph{word \& weorc}, PGmc. \emph{*wurdó \& werkó})
  \Beowulf\ 289, 1100, 1833


\end{itemize}

%TODO: \section{Primary source citations}
