\title{The Northern Epics:}
\author{Konrad O. L. Rosenberg}

\begin{titlingpage}
  \makeatletter
  \centering
  \HUGE \textsc{\@title} \\
  \Huge The Poetic Edda \\
  \huge and other Old Germanic alliterative poetry \\
  \vspace{1cm}
  \Large\emph{edited and translated by} \\
  \vspace{1cm}
  \huge \@author \\
  \vspace{4cm}
  \Large Compiled \today. \\
  \vspace{1cm}
  THE BOOK IS A WORK IN PROGRESS AND THIS FILE MAY BE OUTDATED. \\
  The reader is kindly asked to periodically download the newest version from https://github.com/martensas/edda.
  \makeatother
\end{titlingpage}

\newpage\thispagestyle{empty}

\vspace*{\fill}

{\leftskip=0pt plus .5fil \rightskip=0pt plus -.5fil \parfillskip=0pt plus 1fil
\noindent {\Large \emph{Lifir hann of allar aldir ok stjórnar ǫllu ríki sínu ok rę́ðr ǫllum hlutum, stórum ok smǫ́m. [...] Hann smíðaði himin ok jǫrð ok lopt’in ok alla eign þeira. [...] Hitt er þó mest, er hann gerði mann’inn ok gaf hónum ǫnd þá, er lifa skal ok aldri týna⸗sk, þó’tt lík-amr fúni at moldu eða brenni at ǫsku; ok skulu allir menn lifa, þeir er rétt eru siðaðir, ok vera með hónum sjǫlfum þar sem heitir Gim·lé eða Vin·gólf.}}
\par}
\begin{flushright}%
\emph{Gylfa ginning} 3:4–7%
\end{flushright}

\vspace{5mm}

\begin{center}{\Large \emph{Vęl kęypts hlutar \hld\ hęf’k vęl notit; \\
fás es fróðum vant; \\
því’t Óð·rǿrir \hld\ es nú upp kominn \\
ȧ alda vés jaðar.}}\end{center}

\begin{flushright}%
\emph{Háva mǫ́l} 106%
\end{flushright}

\vspace{5mm}

\begin{center}{\Large \emph{Dęyr fé, \hld\ dęyja frę̇ndr, \\
dęyr sjalfr hit sama; \\
ek vęit ęinn \hld\ at aldri⸗gi dęyr \\
dómr of dauðan hvęrn.}}\end{center}

\begin{flushright}%
\emph{Háva mǫ́l} 77%
\end{flushright}

\vspace{5mm}

\begin{center}{\Large \emph{Ullar hylli \hld\ hęfr ok allra goða \\
hvęrr’s tękr fyrstr ȧ funa \\
því’t opnir hęimar \hld\ verða umb ȧsa sonum, \\
þȧ’s hęfja af hvera.}}\end{center}

\begin{flushright}%
\emph{Grímnis mǫ́l} 43%
\end{flushright}

\vfill

\newpage\thispagestyle{empty}

\bookStart{Acknowledgments}

{\leftskip=0pt plus .5fil \rightskip=0pt plus -.5fil \parfillskip=0pt plus 1fil
\noindent The following people have been especially helpful in giving suggestions, corrections and providing stimulating discussion: Antonín Ryšavý, C. A., Esma, John Newman, J. Hertling, Joseph S. Hopkins, Nikhil Surya Dwibhashyam, Trevor L. Payne.\par}

\newpage\thispagestyle{empty}

\tableofcontents

\newpage\thispagestyle{empty}

\newpage

\bookStart{Abbreviations}
  \input{sigla.tex}%

\printbibliography% Does it work?

\listoffigures

\bookStart{General Introduction (incomplete!)}

\section{Note}

What is this book?  It is a book of poems, really, of songs—songs dating to the first millenium \ce\ or some time thereafter.  What is their nature?  They were sung by the Germanic peoples about their Gods and heroes through Dark Ages, through the fall of Rome and the volcanic winter of 536, in ships and shores of England, in frozen settlements on Iceland and Greenland, in Danish fields, Swedish forests, and Norwegian firths, in English monasteries and the ear of Charlemagne.

Compared to the epic traditions of Classical Antiquity the Germanic songs of heroes do not tell of a very ancient past.  On the contrary, many of them concern events documented in contemporary Classical sources far more accurate than these illiterate ballad-singers working centuries after-the-fact could ever aspire to be.  But their goal was never historical accuracy.  For them the old times were long past; who could faithfully recall them but the High One, the One-Eyed Himself?  These were bards and storytellers and prophetesses, keepers of an ancient poetic tradition far predating the subjects with which it dealt.  This was a tradition infused with the Indo-European spirit of the unending steppes and the snowcapped mountains, the soaring eagle and the galloping steed.  In their stoic fatalism, in their tales of great love and struggle, deprivation and human tragedy, heroic friendship and betrayal, these stories were far older than the more factually correct mediæval chroniclers which chronologically predated them.  In this way they were, in their oft rough-hewn simplicity, as old as Homer, as old as the Vedas—or maybe older still.

The Northern epics are the songs the Germanic-speaking West for many centuries forgot.  We forgot them when we became just that, West\-erners—Christians, Romans.  It was Alcuin, the English scholar, who in a letter upbraided his less learnèd English brethren who would rather sing pagan Saxon songs about Ingeld than the Psalms of David—“what does Ingeld have to do with Christ?” he asked.  It was Louis the Pious, king of the Franks, who burned his father Charlemagne’s great collection of German epic poetry.  For these new men steeped in Roman learning, whose souls were circumcised, even if their bodies were not, the ethos of the old folk-songs was too warlike, to fatalistic, in a word, too \emph{pagan}.

The echoes of the same stories in time suffered a tragic fate in England and the Continent.  The heavily damaged copies written in idioms and hands far too archaic to be understood by later ages were cut up and used for their thick leather as covers for younger, more modern books.  And yet measuring fate would not have it that they were, like so many other European pagan traditions, forever lost.  It was on Iceland, that northern island outpost on the edge of the known world, where the songs of Gods and heroes lived longest upon the lips of hoary upholders of tradition.  It was that most distant offshoot of Germandom, whose unyielding settlers sailed as far as Greenland and North America, which in time came to preserve what the center had lost.

The Codex Regius of the Poetic Edda, the last surviving storehouse of the songs of the Germanic gods, was rediscovered in Iceland in 1642.  Together with the Prose Edda it laid the ground for the new field of the study of Germanic antiquity, which over time developed in lockstep with the evolution of Western science and culture in general and that of philology and linguistics in particular.  Over the centuries it was joined by the rediscoveries of \emph{Beowulf}, the Exeter Book, \emph{Hildebrandslied}, the Merseburg charms, the Gallehus horns and countless other minor sources.

No work could aim to give a complete overview of all scholarly efforts since the 17th century.  The goal of the present edition must only be to provide a small selection of previous insights, together with some new ones, which (it is hoped) may inspire the curiosity of the English-speaking reader and increase his or her understanding and respect for these very important, frequently very difficult texts.

\section{The historical and cultural context}

  It is presupposed throughout this book that the Germanic-speaking peoples—the English, Scandinavians, Dutch, Frisians, and Germans—were originally united by much more than their languages alone; that these nations, although greatly splintered by the turns of history, ultimately share not only language, but a common ancestral culture, religion, and origin.

  All Germanic languages—like the Romance languages from Latin—ulti\-mately descend from Proto-Germanic.  Unlike its contemporary Classical Latin, Proto-Germanic is a reconstructed language attested only in a few borrowed names and short, late runic inscriptions.  Still, we have enough data to be able to reconstruct it confidently, and such reconstructions will appear throughout the book.

  Proto-Germanic in turn derives from the wholly unattested Proto-Indo-European, from which also spring the Albanian, Armenian, Baltic and Slavic, Celtic, Greek, Italic, and Indo-Iranian tongues along with the now-extinct Anatolian and Tocharian and a likely myriad other now-lost languages.  The Germanic branch most likely broke away from its closest known Indo-European linguistic relatives in the early 3rd millenium \bce\ and has developed on its own since, undergoing important changes in its sounds, word-hoard, word-order, and grammar.

  Around the start of the first millenium \ce\ the Germanic family itself broke apart into three branches: the Northern, Western, and Eastern.  Only the first two are presently extant.  To the first belong the Scandinavian languages descended from Old Norse; to the second English, Frisian, Dutch, Low and High German (including Yiddish); to the third the extinct Gothic.  It is generally held to be the case on the basis of a number of common innovations that Northern and Western branches form a single subclade, Northwest Germanic, to the exclusion of East Germanic \parencite{Kim2024}.  Since the Wielbark culture, which is probably to be identified with the early ancestors of the Gots, emigrated from Scandinavia around 50 \bce, this forms a useful reference point for the time of the splintering of the Germanic language family.

  The cultural affinities of the Germanic-speaking peoples are of course greater the further back in time one looks.  During the Migration Period there was something of a cultural flow between all Germanic peoples, especially those who were not yet Christianized, as seen both archeologically and in later literary sources like \textlink{Widsith} \parencite{Neidorf2013}.  Customs like the husband’s “morning-gift” to his wife on the day of their wedding or the twelve-man jury were spread far and wide by these expanding tribes, as was the two-stemmed naming system also found among other Indo-Europeans (in English e.g. Edwin < Old English \emph{Éad-wine} < Proto-Germanic \emph{*Auda-winiz} ‘Friend of Wealth’).

  Nor were these affinities entirely lost on the tribes themselves.  Both Gots and Longbeards (i.e., Lombards) remembered their old homeland of Scandinavia and many of the related tribes living there (Jordanes, \emph{Getica} 10, 16–25; \emph{Origo Gentis Langbardorum} 1, edited under the appendix to \textlink{Grimnismal}); the Anglo-Saxons, too, never forgot that they had come from Northern Germany and Jutland \parencite{Gretzinger2022}, nor that there were still related people living there \parencite[173--176]{Neidorf2013}.

  As for the religion, the worship of many Gods (the Eese, ON \emph{ę̇sir}, OE \emph{ėse}, Gothic \emph{Ansis} in Jordanes, \emph{Getica} 78, 79; cf. \textlink{Hamdismal}[25]/1 n.) is attested throughout the historical and geographic range of the Germanic peoples, as are the use of sacred rings (the Gothic ‘holy’ ring of Pietroassa \parencites[280]{Antonsen2002}[173--174]{MacLeod2006}, cf. Ambrose of Milan’s letter 10:9 which suggests the wearing of bracelets by Gothic priests \parencite[425]{Capelle2010}; \textlink{Havamal}[110], \textlink{Hildebrandslied} 32), animal sacrifices, and certain forms of ritual warfare (\textlink{Voluspa}[23] n.).
  Also widely attested is the worship of the gods Weden, Ing, and Thunder, the former of whom appears in His name Geat at the head of both Gothic and Anglo-Saxon genealogies (\textlink{Hamdismal}[25]/1 n.).

  Turning to shared ancestry, recent genetic advances have shown that the lines of linguistic expansion line up with signals of genetic kinship, and have in effect confirmed what earlier scholars of the 19th and early 20th centuries presumed.

  The Germanic peoples originate in Northern Europe, chiefly Scandinavia and Northern Germany.  They derive much of their ancestry from the early Indo-Europeans of the Pontic-Caspian steppe, who mixed with Early European Farmers to the west to form the Corded Ware Culture, which over time expanded in all directions \parencite{McColl2024}.
  In the West it formed the Bell Beakers who conquered Ireland and Iberia; in the South it formed the Italic peoples who eventually became the Romans; in the East it formed the Sintashta culture, or Proto-Indo-Iranians, whose descendants became the Iranians and Vedic Indians and who founded the Zoroastrian and Vedic religions; in the Northeast it formed the Balts and Slavs; and in Scandinavia it formed the Battle Axe and Single Grave cultures which eventually gave way to the Nordic Bronze Age, the ancestors of the Germanic peoples.

  Although the genes for light eyes and hair colour were present already on the steppe, it was in the later Corded Ware that the heavy selection for these traits was effected \parencites{Wilde2014}{Ju2021}{Perretti2025}, traits which to this day reach their highest frequency in Northern Europe \parencite{Frost2006}.  In all cases, including Scandinavia, its expansion was that of a warlike and patriarchal culture not unlike its later descendants, whether Roman or Germanic.  Indeed, present-day native Scandinavians, who are largely the direct biological descendants of the Nordic Bronze Age, carry almost no ancestry from the Stone Age Scandinavian Hunter Gatherers that preceded them \parencite{Allentoft2024}.

%  \subsection{Lifestyle and economy}
%    Cattle-based; small farmsteads.
%  \subsection{Morals and Virtues}
%    Honour, personal integrity
%    Notes on the terms \emph{argr} and \emph{ęrgi}
%  \subsection{Religion}
%    Keeping the Powers happy
%    Cosmic cycles
%    Reincarnation
%    Analogies with other Indo-European traditions


\section{Germanic alliterative poetry}

  The historical and literary significance of the poetry in this edition may be thought of in two ways: as the beginning of documented Germanic literature and also as the last stage of an ancient oral tradition with distant Indo-European roots.

  On the one hand, alliterative verse is the medium for the oldest indigenous Germanic-language literary monuments (the Gothic literature, chiefly consisting of Wulfila’s translation of the Bible, being wholly foreign in origin).  It is our earliest native source of knowledge about the Iron Age folk-life of Northern Europe and our most important source for the pre-Christian Germanic religion; it represents the worldview of the Germanics peoples in their earliest historical stages.  As the earliest recorded medium for the transmission of song and story amongst the ancestors of numerous great nations it has to some extent influenced all later Germanic languages and literatures.  The alliterative poetry, albeit together with Latin, Greek, and Hebrew models, laid the ground for the German \emph{Minnessänger}, the \emph{Nibelungenlied}, and Chaucer, which in turn precede such great writers as Goethe, Wagner, and Shakespeare.

  On the other hand, this poetic genre is hardly a new art-form by the time we first encounter it in manuscript form in the 7th century \hyperref[Cadman]{Cadman’s Hymn}, nor is its characteristic feature, alliteration between initial sounds.

  In the 1st century \ce\ Tacitus (\emph{Germania} 2) speaks shortly of the ‘ancient songs’ in which the Germani documented their mythology and history.  He mentions the myth of three brothers, the sons of \emph{Mannus} ‘Man’, who founded three tribes, the \emph{Ingaevones, Herminones} and \emph{Istaevones}.  In their Proto-Germanic forms the names of the three tribes all begin with a vowel, and this is thus the earliest secure example of alliteration.

  Tacitus may also give us the earliest example of the Germanic long-line, the same meter as \textlink{Beowulf} and the Poetic Edda, when he (\emph{Germania} 40) lists the following tribes as worshipping \emph{Nerthus}: \emph{Reudigni, Aviones, Anglii, Varini, Eudoses, Suardones, Nuithones}.  If \emph{Nuithones} is emended to \emph{*Unithones}, which unlike \emph{Nuithones} actually looks like a Germanic word, the list may be arranged into two long-lines, both with vowel-alliteration \parencites{Malone1934}{Malone1936}:

  \begin{verse}
    \emph{Reudigni, Aviones, \hld\ Anglii, Varini,} \\
    \emph{Eudoses, Suardones, \hld\ *Unithones.}
  \end{verse}
  A more secure example of the same phenomenon is found in the 2nd century, when Ptolemy (\emph{Geography} 2, 11:7) lists the following tribes on the Cimbric peninsula: \emph{Saxones, Sigulones, Sabalingii, Cobandi, Chali, Phunusii, Charudes}.  This example has the advantage of not requiring any emendation, and forms two clean long-lines with typical alliterative patterns (a a : a x, a x : a [x]) \parencites{Malone1936}[179]{Neidorf2013}:

  \begin{verse}
    \emph{Saxones, Sigulones, \hld\ Sabalingii, Cobandi,} \\
    \emph{Chali, Phunusii, \hld\ Charudes.}
  \end{verse}

  The most likely explanation for these early alliterations is rather straight-forward: the Germanic informers recited the tribes in the order they had learnt them, namely as alliterative lists or rigamaroles not unlike what we find in later texts like \textlink{Widsith} or the Icelandic poetic lists, the so-called \emph{þulur} ‘thules’ \parencite{Malone1936}.

  The earliest indisputable native attestations of the long-line are in runic inscriptions from the early centuries \ce, the oldest of which is an early 3rd century sword-chape from Thorsberg moor in Schleswig.  These inscription are edited in the pres. ed. under Miscellaneous Runic Poetry.

  In these finds we see not just the fully formed long-line but also examples of stylistic traits common to all later Germanic alliterative poetry.  Thus the 5th century Tjurkö bracteate (DR IK184) calls gold \emph{walha-kurné} (dat. sg.) ‘Roman grain’, a kenning-like reference to the melting of Roman solidi; the 3rd century Thorsberg chape (DR 7) expresses being ‘famous’ with the litotes \emph{ni wajé-máriʀ} ‘not ill-famed’ (cf. \textlink{Atlakvida}[13]/1); and the 6th century Noleby stone (Vg 63) mirrors the formula \emph{rúnó ragina-kundó} (acc. sg.) ‘rune born of the Gods’ later found in the 9th century \hyperref[Havamal:80]{\emph{Háva mǫ́l}, st. 80}.

  Of course, a far greater trove of rich expressions and images, many of great antiquity, is found in the extensive corpus of poetry written in mediæval manuscripts.  In \Beowulf\ and the Edda ‘sea-stallions’ sail across the ocean like in Homer; in \textlink{Grimnismal}[38] the sun is drawn across Heaven in her chariot like in \Rigveda\ and the Trundholm Sun Chariot; in \Beowulf, \textlink{Heliand}, \textlink{Atlakvida} and other poems feasts are held in great chiefly ‘mead-halls’, not unknown from the archæology of the Nordic Bronze and Iron Ages.

  The richness of the poetic tradition is expressed through a language of kennings, stock phrases and exclusive synonyms found only in poetry.  Many of these synonyms are quite ancient, being Indo-European words otherwise extinct in all Germanic languages which find their last refuge in poetry.  Such are the Old English \emph{eoh}, Old Norse \emph{jór}, corresponding to the Sanskrit \emph{áşva}, Latin \emph{equus}, all meaning ‘horse’; Old Norse \emph{týr}, corresponding to Sanskrit \emph{devá}, Latin \emph{deus}, all meaning ‘god’; Old English \emph{folde}, Old Norse \emph{fold}, corresponding to Sanskrit \emph{pṛthivī́}, all meaning ‘land, (mother) earth’.  The fact that many of these relate to the cult perhaps suggests that the Germanic religion was not as innovative as is commonly supposed.

  What a comparative investigation reveals is thus that these poetic motifs and words are seldom uniquely mediæval, nor even inventions of the first centuries \ce; instead they time and time again find parallels in texts and archeological finds as old as the Bronze Age.  The poems themselves are clearly much younger than this, but still not so young that their decidedly unlatinate authors could have been influenced by texts like the Iliad, which were still yet unknown in Northern Europe.  Rather, they seem to retain ancient traditions which must have been passed down through an unbroken chain of poetry going back far into prehistory, to when the Indo-Europeans were still nomadic steppe-pastoralists and much of Europe was still covered by thick forest and marshlands.

  \subsection{Structure}
    To describe the structure of Germanic alliterative poetry it is first necessary to describe the general characteristics of the common Germanic alliterative meter as found in the earliest runic inscriptions, West Germanic poetry, and Norse Eddic poetry in the meter \Fornyrdislag.  The following description is thus not exhaustive, but does describe the bulk of the corpus.

    Alliterative poetry has two primary structural elements: \emph{stress} and \emph{alliteration}.  The exact count of syllables is less important, and end-rhyme is only used as a sporadic flourish.

    Based on alliteration, words are grouped into \emph{lines}.

    \subsubsection{Stress}
    Each syllable is classed as having primary stress (p), secondary stress (s), or no stress (x).  Syllables with p and s are stressed syllables, those with x are unstressed syllables.

    Primary stress is reserved for the root syllable in a word, which is not always the same as the first syllable.  Compare the English word \emph{beginning}, where the stress pattern is xPx; the primary stress falls on the syllable \emph{ginn-}.  Note that not all primary stresses are equal; the general rule is that nouns and adjectives have stronger stress than verbs.

    Secondary stress falls on the second element in a compound word, e.g. \emph{hęimr} in \emph{jǫtun-hęimr} ‘Ettin-ham’.

    No stress falls on inflectional syllables, pronouns and prepositions.  The exception for pronouns and prepositions is that they can take primary stress if it coincides with semantic stress, e.g. in a statement like “It was \emph{you}!”, or when a preposition is placed after the noun it modifies (e.g. \Beowulf\ 19b: \emph{Scęde-landum in} ‘in the Northern lands’).

    \subsubsection{Alliteration}
    Alliteration is the resonance between two stressed syllables beginning with the same phoneme, e.g. \emph{\alst{s}and} with \emph{re\alst{c}eive}, or \emph{\alst{g}reat} with \emph{be\alst{g}in}.

    Any vowel or diphthong can alliterate with any other vowel or dipththong.

    \emph{s} and the clusters \emph{sk, sp} and \emph{st} are counted as four distinct “sounds”.

    Further, in West Germanic poetry, \emph{g} and \emph{j} can alliterate with each other.

    In Norse poetry \emph{j} and \emph{v} can alliterate with vowels, the former always, the latter sporadically.

    In the present edition alliterating sounds are marked with red font.

    \subsubsection{Positions}

    The smallest metrical division is the \emph{position}, a concept related but not identical to the syllable.  The position must be understood as broader than the syllable as several syllables can fill one position, whereas a single syllable can never fill multiple positions.

    A short syllable followed by another syllable can resolve into a single position.  Likewise several unstressed syllables can count as a single position.

    \subsubsection{Verses}

    A \emph{verse} or \emph{half-line} is made up of 4 positions—typically 2 stressed, 2 unstressed—arranged according to set patterns.  These patterns form the core of Sievers’ description of alliterative poetry, and is the description followed here.

    TODO: Sieversian types.

    \subsubsection{Lines}

    2 verses are linked together by shared alliteration to form a couplet or \emph{long-line} (often simply \emph{line}).  They are separated by a short pause or \emph{cæsura}.

    It is not uncommon for editions of Old Norse poetry to print each verse on its own line, e.g. \textlink{Voluspa}[2]/1–2:

    \begin{quote}\begin{small}
      \alst{E}k man \alst{jǫ}tna \\
      \alst{á}r of borna, \\
      þȧ’s \alst{f}orðum mik \\
      \alst{f}ǿdda hǫfðu;
    \end{small}\end{quote}

    Editions of West Germanic poetry typically print each long-line on its own line separated by a space to designate the cæsura:

    \begin{quote}\begin{small}
      \alst{E}k man \alst{jǫ}tna \hspace*{1em} \alst{á}r of borna, \\
      þȧ’s \alst{f}orðum mik \hspace*{1em} \alst{f}ǿdda hǫfðu;
    \end{small}\end{quote}

    The present edition follows the latter convention for all alliterative poetry, Norse or West Germanic, but replaces the space with the interpunct:

      \begin{quote}\begin{small}
        \alst{E}k man \alst{jǫ}tna \hld\ \alst{á}r of borna, \\
        þȧ’s \alst{f}orðum mik \hld\ \alst{f}ǿdda hǫfðu;
      \end{small}\end{quote}

    The first verse (or \emph{a-verse}) may have 1 or 2 alliterations, with preference for alliteration on its first position over the second.

    The second verse (or \emph{b-verse}) must have exactly 1 alliteration, almost always on its first stressed position.

    \subsubsection{Fits and stanzas}
    The long-line is the highest metrical division, but long-lines can be combined into complete poems according to either a \emph{stichic} or \emph{stanzaic} manner.

    West Germanic poetry is typically \emph{stichic}.  This means that long-lines simply follow one other, often with enjambment, so that a new sentence begins in the b-verse.

    In longer epic poems lines may be grouped into \emph{fits} which are sections or cantos.  The fit does not have a fixed length, but is generally around 70-85 lines long.  The 3182-line \Beowulf\ is divided into 42 numbered fits for an average of 76 lines per fit; the surviving 5983 lines of \textlink{Heliand} are divided into 71 for an average of 84 lines per fit.
    It is probably not a coincidence that the length of the fit is similar to the length of shorter legendary poems like \textlink{GudrunThree} or \textlink{Hildebrandslied}.

    In \textlink{Heliand} a new fit can begin on a b-verse.  This does not happen in \Beowulf.

    North Germanic poetry is always more or less \emph{stanzaic} and does not have fits.  Each \emph{stanza} consists of several long-lines, and enjambment is rare.  The regularity of the stanzas varies by poem and meter, but there is a general trend towards stanzas of exactly 4 lines.

    In 4-line stanzas there is generally a semantic division between lines 2 and 3; each group of 2 lines (4 verses) form a \emph{helming} or \emph{half-stanza}.


\section{The corpus}
  The scope of the present corpus is large, and encompasses most of the alliterative poetry extant in Old Germanic languages.  The poetry is divided into the following categories:
  \begin{enumerate}
    \item \textbf{Norse Mythic poetry}, i.e., that which directly treats the Germanic mythology.  This category is exclusively Norse for the simple reason that no West Germanic or Gothic mythic narrative poetry survive.
    \item \textbf{Norse Heroic poetry}, specifically the whole second half of the Codex Regius and then a few other works.  With a few exceptions, subject matter outside of the Walsing cycle is not included.
    \item \textbf{West Germanic Heroic Poetry} in Old English, Old Saxon, and Old High German.
    \item \textbf{Poetry on Christian subjects}.  This category includes explicitly Christian poems where the new religion or its mythology is at the core of the work.  Christian heroic poems depicting native legends, like \Beowulf\ and \textlink{Hildebrandslied}, are not included.
    \item \textbf{Galders}, i.e., alliterative spells and charms, both from runic inscriptions and mediæval manuscripts.
    \item \textbf{Miscellaneous runic poetry}, apart from that already edited under Galders above.
  \end{enumerate}

  \subsection{Exclusions}
    All Norse Scaldic poetry is excluded, as is the Eddic poetry found in the Saws of Icelanders and Saws of Antiquity (\emph{forn-aldar sǫgur}).  Three reasons justify this exclusion.  First, these two categories of poetry have already been edited by the \Skp\ series, in the case of the Scaldic poetry admirably so.  Second, the underlying poetry is often impossible to take out of its prose context.  Third, it is doubtful whether most of the Eddic poetry embedded in prose contexts ever had a life of its own or whether it was composed along with the prose.  For these reasons I would only edit the poetry in its original context, that is, edit the whole Saws, which is an undertaking that falls far outside of the scope of the present edition as an edition of Eddic poetry.

    Most non-legendary Old English poems are excluded.  This includes poems on historical subjects embedded in the Anglo-Saxon Chronicle, the Riddles of the Exeter Book, and the Biblical epics.

  \subsection{Manuscripts}

    See the introduction to each category.

    \subsubsection{Old English poetry}

    The edited Old English poetry primarily derives from a few manuscripts.  Particularly important are the Exeter Book and \Lacnunga.

    \subsubsection{Old Saxon and High German poetry}

    There are no collections of alliterative poetry in these languages; instead the manuscript situation will be disussed in the Introduction to each individual text.

  \subsection{Dating}

  Ways of dating poetry, which naturally differ greatly for each text and tradition, will be brought up in the respective introductions to sections or poems as seen fit.

  The general methodology for dating adopted by the present edition is however one that prioritises linguistic criteria rather than literary analysis or vague historical speulation.  It has been shown for both the Eddic (TODO: cite Males) and Old English poetic traditions (TODO: cite Neidorf Dating of Beowulf) that mediæval scribes did not possess the necessary linguistic knowledge to produce archaizing poetry, or even to reproduce metrically required archaic features (e.g. the consonant cluster \emph{vr-} in ON, for which see \textlink{Havamal}[26]/2 n.), and a text that displays linguistic features belonging to a certain century is thus taken to have been composed during that century.  In this way the present methodology aligns closer with that upheld by scholars like Finnur Jónsson, R. D. Fulk, Leonard Neidorf, and Mikael Males and differs from that of Klaus von See or the “Toronto school” of Beowulf.

\section{The presentation of the text}

  The present edition is divided into two equally large parts, presented side by side.  Each stanza or group of verse lines is presented first in the original language, and immediately below in English translation.  In the right margin is a list of manuscript witnesses for the particular passage.  Below the English translation are two layers of notes.  The first contains critical notes, and the second contains all other notes and commentary.  Broader commentary or particularly important critical notes are linked to ALL, while more specific notes are linked to separate lines and lemmas.

  \subsection{The Old Germanic text}

    The present edition encompasses texts in four Old Germanic languages: Old Norse, Old English, Old Saxon, and Old High German.  All texts have been normalized according to the author’s standardised orthography for the respective language.  The orthographies are designed to follow three core principles:

    \begin{enumerate}
      \item A faithfulness to the spoken language at the time when the texts were written, and the distinctions demonstrably found therein.
      \item A respect for the etymological origin of words, and their distinctions.
      \item A striving for a uniform orthography across the various languages, so that the same sound, where possible, should be written with the same character.
    \end{enumerate}

    These choices often stand in conflict with the orthography of the original manuscripts and with most earlier philological tradition, whence there is some reason to justify them.  My goal is to render the texts themselves in a manner that gives as much philological information to the reader as possible—not to present a facsimile edition for students of paleography.  This follows the philological methods used for printing e.g. the \Rigveda, which is generally printed in an entirely scholarly latinized orthography, not the original \emph{Devanāgarī}.  Regardless, such important traits of the original manuscript tradition as the long \emph{ſ}, arbitrary punctuation, arbitrary spelling, and lack of line breaks, are seldom reproduced in modern editions of Old Germanic poetry.

    \subsubsection{General orthographic conventions}

    The following orthographic conventions are applied for all Old Germanic languages:

    \begin{enumerate}
    \item The voiceless dental fricative is always written with the letter \emph{þ}, never \emph{th}.
    \item Long vowels are marked with the acute accent, never the macron or circumflex;
    \item Excepting those long vowels which have their origin in earlier dipththongs, which are written with the circumflex.
    \item In compounds where the first element has primary stress the elements are separated with a dash (-);
    \item But where the first element is a preposition or unstressed prefix they are separated with an interpunct (·).
    \item Non-inflectional suffixes and derivations are marked with a double oblique hyphen (⸗).
    \end{enumerate}

    Below follow specifications for each specific language.

    \subsubsection{Normalization of Old Norse}

    My Old Norse orthography is inspired by \textcite{FinnurEdda} in that it strives for a more archaic form than that of the surviving mss.; a form that instead represents the poetry as it may (in many cases, must) originally have looked. For this reason, it often has more in common with the proposed orthography of the First Grammatical Treatise than with the standard Old Icelandic orthography seen in most editions. The following list describes the differences from the standard Old Icelandic orthography:

    \begin{enumerate}
    \item I distinguish between short \emph{e} (from etymological short \emph{e}) and short \emph{ę} (from etymological short \emph{a} + \emph{i}-umlaut).
    \item I distinguish between long \emph{á} and \emph{ǫ́}, as done by the First Grammatical Treatise.
    \item I use \emph{ǿ} and \emph{ę́} rather than the usual \emph{œ} and \emph{æ}, to represent the vowels descended from Proto-Norse \emph{ó} and \emph{á} after \emph{i}-umlaut (cf. the short \emph{ø, ę} < \emph{o, a} + \emph{i}-umlaut).
    \item I distinguish long nasal \emph{ȧ, ė, ï, ȯ, u̇} from long oral \emph{á, é, í, ó, ú}, as done in the First Grammatical Treatise.
    \item I restore the old \emph{s} (which in modern Scandinavian and even in most Old Norse manuscripts has become \emph{r}, but which is found in old manuscripts such as AM 237 a fol (c. 1150) and fossilized in forms like \emph{þaz} (i.e. \emph{þat’s}) in \Regius) in the words \emph{es} ‘which, that, where, when’, and in inflections of \emph{vesa} (later \emph{vera}) such as \emph{es} ‘is’ (3rd sg. pres. ind.) and \emph{vas} (3rd sg. pret. ind.).  The following forms of \emph{vesa} retain the \emph{r}, as it is there the result of Verner’s law, and not of this (much younger) sound change: the pl. pres. ind. (\emph{erum} etc.), the pl. pret. ind. (\emph{vǫ́rum} etc.), and the pl. pret. subj. (\emph{vę́rim} etc.)
    \item When metrically benefactory, I contract \emph{ek} ‘I’, \emph{eru} ‘are’, and \emph{es} ‘which; is’ to \emph{’k}, \emph{’ro} and \emph{’s}, respectively.
    \item I resolve \emph{z} to \emph{ts, ds, þs} or \emph{ðs} depending on its underlying elements;
    \item \emph{x} is however still used instead of \emph{ks};
    \item I follow \textcite{FinnurEdda} in distinguishing between the relative particle \emph{es} and the verb \emph{es}: the particle is appended to the previous word with only an apostrophe (e.g. \emph{hann’s} ‘he who’), while the verb is separated by a space (e.g. \emph{hann ’s} ‘he is’).
    \item Where metrically beneficient personal pronouns are removed, often silently (cf. Scaldic poetry, \textlink{Volundarkvida}[21]/1–2 vs. 23/3–4, \textlink{HelgakvidaOne}[32] vs. \textlink{HelgakvidaTwo}[18])
    \end{enumerate}

    \subsubsection{Normalization of Old Swedish and Danish}
    I employ the same conventions as those described for Old Norse above, including the marking of \emph{u}-mutated \emph{a} > \emph{ǫ}.  That this indeed occurred in the Eastern Nordic dialects is most proven by the third-person personal pronoun, which shows \emph{u}-mutation in such forms as Swedish \emph{honom} ‘him’ < \emph{hǫ̇num}, \emph{hon} ‘she’ < \emph{hǫ̇n}).

    According to rule 3 in the general orthographic conventions above, I distinguish between \emph{ǿ} (< \emph{ǿ}) and \emph{ø̂} (< \emph{au, ęy}); \emph{é} (< \emph{é}) and \emph{ê} (< \emph{ęi}).

    Where unstressed vowels have been reduced into an schwa-like sound spelled \emph{e}, this is written with \emph{ę}.

    \subsubsection{Normalization of Old English}
    I spell fronted or brightened etymological \emph{a} and \emph{á} with \emph{æ} and \emph{ǽ}, for instance in \emph{dæg} ‘day’ (< \emph{*dagaʀ}) and \emph{rǽd} ‘advice, counsel’ (< \emph{rádaʀ}).  These are contrasted with \emph{ę} and \emph{ę́}, which represent \emph{i}-mutated \emph{a} and \emph{á}, e.g. in \emph{ęllen} ‘zeal, courage’ (< \emph{*aljaną}).

    An assimilated \emph{n} is marked with an overpoint, like in rule 3 of Old Norse above.

    \subsubsection{Normalization of Old Saxon}

    \subsubsection{Normalization of Old High German}

  \subsection{The English translation}

    There is now a very large number of translations of the most popular alliterative poetic texts, namely \Beowulf\ and the \emph{Poetic Edda}.  These generally fall into two camps:
    \begin{enumerate}
      \item \emph{poetic} translations, which distort the precise meaning of the text for the sake of meter, often quite radically; and
      \item \emph{prose} translations, which nowise preserve the style or feeling of the original.
    \end{enumerate}

    Almost all translations, of both types, also tend toward the following inadequacies: obscuring or glossing over difficult technical and cultural terminology; rendering identically repeated phrases and words (formulae) differently at various places; and simplifying or rewriting kennings and other poetic expressions.  Even worse this is often done with little in the way of notes or commentary, to a point where the reader is sometimes left entirely oblivious to the sense of the original text.

    What sets my translation apart from previous English translations is that it aims to follow the style and register of the original text, without sacrificing the literal sense of the words.  This unfortunately means that literality and consistency at times must sometimes come at the cost of fluid idiomatic English, but it has the advantage of giving the reader an image of not just \emph{what} the original text actually says, but \emph{how} it says it.  The reader should keep in mind that he is in a very foreign land, that he is reading words ancient and long forgotten—not the \emph{New York Times}.

    Maybe this is a pointless effort? One could argue that a translation always is a betrayal, and that those truly interested in the exact meaning of every word in the original text should study just the original (in the original language).  While I do agree that the sufficiently interested reader should study the original texts in the languages in which they were written (something made much easier by the present edition with its notes and parallel edition), it is still a “hard ask” for those readers who are not philologically inclined, but instead students and scholars of history, comparative mythology and religion, anthropology, or literature; those who, for whatever reason, are interested in exploring the oldest poetic heritage of the Germanic peoples of northern Europe.

    \subsection{Anglish proper nouns}
      Perhaps the single most idiosyncratic part of the present translation will be its handling of proper nouns. I have opted to render all cultural and religious terms, names of places, heroes, gods, and other entities by their English cognates (thus \emph{Thunder} for Old Norse \emph{Þórr}) and where such do not exist, their philologically expected English (\emph{Anglish}) forms (e.g. \emph{wallow} for Old Norse \emph{vǫlva}).

      There are two reasons for this.  The first is ideological.  I believe that the Old Germanic myths and poems, their gods and heroes, are a shared heritage of Northern Europe.  When you translate texts from across Germany, England and Scandinavia you quickly come to notice how similar the diction is, how many names reappear. The Scandinavian \emph{Vǫlundr} is the same character as the English \emph{Wélund}; likewise Norse \emph{Óðinn} is the same as English \emph{Wóden}.  These are ultimately mere distinctions in pronunciation.

      The second is aesthetic.  Commonly accepted forms like \emph{Odin} and \emph{Thor} are debased.  They do not even represent the Old Norse pronunciation as accurately as possible within the constraints of English ortography (for instance, \emph{Odin} would be better anglicized as \emph{Othin}).  Many are also difficult for English speakers to pronounce, or lead to absurd confusions.  I shudder at hearing the word \emph{ę́sir} pronounced /aɪˈsɪ:ɹ/; even worse is when \emph{Ǫ́s-garðr} becomes “ass-guard”.
