\book{The Spae of the Wallow. (\emph{Vǫluspǫ́})}\bookStart

% Introduction.

\small{\emph{Vǫluspǫ́}, or the “\inx{Spae} of the \inx{Wallow}”, is the first poem of \Regius. TODO}

% Greeting to the audience, bidding of Weden.

\bvb Of silence I bid all holy kins\footnotemark[1], the greater and lesser sons of Homedale\footnotemark[2]. Wilt thou, Leader of the Slain\footnotemark[3], that I well tell forth the ancient sayings of firs [MEN], those I foremost recall?\footnotemark[4]\evb
\footnotetext[1]{The ‘holy kins’, according to \Finnur\ referring to the gods, but might also be an allusion to the divine origin of men.}
\footnotetext[2]{Cf. with \Rigsthula\, wherein Rig (Homedale) sires the \emph{greater and lesser} human races (\emph{earls}, \emph{churls} and \emph{thralls}). The Wallow (speaking through the poet) addresses not just the human audience, but also the gods.}
\footnotetext[3]{According to \Finnur\ the name (for Weden) is chosen intentionally, as it refers to the final fight at the \inx{Rake of the Powers.}}
\footnotetext[4]{Cf. \Vafthrudnismal\ 34, 35 with very similar phrasing.}

% Wallow reckons what she recalls; the creation and ordering of the world.

\bvb I recall \inx{ettins}, born of yore, those who anciently had nourished me. Nine \inx{homes} I recall, nine \inx{inwithies}, the renowned \inx{Metwood} beneath the earth\footnotemark[1].\evb
\footnotetext[1]{Probably Ugdrassle, still being a seed.}

\bvb It was the beginning of \inx{elds}, there where \inx{Yime} dwelled\footnotemark[1]; was there not sand nor sea, nor cool waves. The earth was found never, nor \inx{up-heaven}; a gap was of ginnings\footnotemark[2], but grass nowhere.\evb
\footnotetext[1]{\Gylfaginning\ “that where nothing was”.}
\footnotetext[2]{See \inx{Gap of Ginnings}. \emph{ginnungr} means ‘hawk’ in the Scoldish poetry, but that meaning is strange here, unless it be an obscure sky-kenning, meaning that there was only an infinite void (but no matter).}

\bvb Before the sons of Bur the flatlands did lift upwards, they who the renowned Middenyard shaped. Sun shone from the south on the stones of the hall, then was the ground grown with green leek.\footnotemark[1]\evb
\footnotetext[1]{The sons of Bur, that is Weden, Will and Wigh, lift the lands out of the primordial chaos (the \inx{Gap of Ginnigs}).}

\bvb Sun cast from the south, — the companion of Moon, — her right hand about heaven’s rim;\footnotemark[1] Sun knew not where she halls owned, stars knew not where they steads owned, Moon knew not what he of power owned.\evb
\footnotetext[1]{The sun heaved herself up over the horizon and rose for the first time?}

\bvb Then the Powers all went onto the rake-seats, the gin-holy gods, and from each other took counsel about that: [...]\evb

\bvb To night and the phases of the moon names they did give; morning they called, and middle day; noon and evening, the years for to reckon.\evb

\bvb Then found the Ease eachother on the \inx{Ithewalds}, they who \inx{harrows} and \inx{hoves} high timbered; forges they laid, wealth they forged, tongs they shaped, and tools they made.\evb

\bvb They played \inx{tables} in the yards, joyous were they, was for them no lack of gold. Until three came, maidens of thurses, much terrifying, out of Ettinham.\footnotemark[1]\evb
\footnotetext[1]{These are immediately forgotten and not again mentioned (unless they are taken to be the norns in v. 21, but they would then be introduced twice). — Clearly there is something missing between this verse and the next, detailing the creation of dwarves.}

% Creation of dwarves; two lists.

\bvb Then the \inx{Powers} all went onto the \inx{rake-seats}, the \inx{gin-holy} gods, and from each other took counsel about that: Who would shape the multitudes of \inx{dwarves}, out of the blood of \inx{Brime}, and out of the legs of \inx{Blown}?\footnotemark[1]\evb
\footnotetext[1]{\Regius\ “Who would shape the multitude of dwarves out of the blood of Brime and out of black legs?”. \Hauksbok\ “Which dwarves would shape multitudes out of bloody Brime, and out of the legs of Blown?”. In any case, they were discussing who would shape the dwarves out of Yime’s corpse.}

\bvb There was Moodsown become the worthiest of all dwarves, but Dorn [was] second. They made men-likenesses many; dwarves out of the earth, as Dorn said.\footnotemark[1]\evb
\footnotetext[1]{A cryptic verse; \emph{manlíkan} ‘man-likeness’ is a hapax. Were the lower dwarves shaped out of soil, by the mightiest dwarves, Moodsown and Dorn, themselves made by the gods from Yime’s flesh and blood? “as Dorn said” would imply that Dorn did not shape the dwarves himself; did he and Moodsown shape the first lower dwarves out of stone, and then command these to finish the creation?}\evb

\bvb New and Nithe, Norther and Suther, Easter and Wester, Allthief, Dwollen, Bewer, Bawer, Bamber, Noor, Own and Owner, great grandfather, Meadwitten.\evb

\bvb Way and Gandelf, Windelf, Thrown, Thetch and Thorn, Throo, Wit and Lit, Naw and Newred, — now have I the dwarves — Rain and Redswinn — rightly counted.\evb

\bvb File, Chile, Founden and Needler, Haftfile, Hanner, Swigher, Fro, Hornborer, Fray and Loon, Earwong, Ere, Oakenshield.\evb

\bvb TODO\evb

\bvb TODO\evb

%

\bvb Until three came out of that host: mighty and loving Ease by the farmstead; they found on land the little availing \inx{Ash} and \inx{Emble}, \inx{orlay}-less.\footnotemark[1]\evb
\footnotetext[1]{For, according to \Snorri, they were pieces of driftwood.}

\bvb Breath they owned not, \inx{wode} they had not, not craft nor sound, nor good complexion. Breath gave \inx{Weden}, wode gave \inx{Heen}, craft gave \inx{Lother}, and good complexion.\evb

\bvb — I know an ash does stand, called \inx{Ugdrassle}: a high tree, poured with white clay\footnotemark[1]. Thence come the dew-drops which in the dales fall; it stands ever green over the \inx{well of Weird}.\evb
\footnotetext[1]{Compare perhaps with the Indian ritual pouring of beverages over the \emph{lingam}. In this case the white clay may stand in for milk.}

\bvb Thence come maidens, much knowing, three out of the lake\footnotemark[1] which stands beneath the pine\footnotemark[2]: \inx{Weird} they call one, the other \inx{Werthing}—carved they on planks—\inx{Shild} the third. They laws laid, they lives chose, for the children of men, the \inx{orlay} of mortals.\evb
\footnotetext[1]{\Hauksbok\ “hall”.}
\footnotetext[2]{But here simply meaning ‘tree’; a sign perhaps that the description of the tree as an ash should not be taken as sure.}

% Battle between Ease and Wanes

\bvb TODO\evb

\bvb TODO\evb

\bvb Then the Powers all went onto the rake-seats, the gin-holy gods, and from each other took counsel about that: whether the Ease should tribute yield, or should the gods all a banquet hold?\evb

\bvb Weden flung [the spear], and into the opposing army did shoot; that was yet the first folk-war in the \inx{home}. Broken was the board-wall of the fortification of the Ease; the Wanes knew by \inx{wigh-spae} tread the fields.\footnotemark[1]\evb
\footnotetext[1]{The Wanes use “wigh-spae” (war magic).}

\bvb Then the Powers all went onto the rake-seats, the gin-holy gods, and from each other took counsel about that:\evb

\bvb TODO\evb

\bvb TODO\evb

\bvb Lone she sat outside, when the old one came, the Terrifier of the Ease, and into [her] eyes looked. “Why inquirest thou me? Why temptest thou me? All I know, Weden, where thine eye thou hidst: in the renowned \inx{well of Mime}. Drinks Mime mead every morning, from the pledge of the \inx{Father of the Slain}; — Know ye yet, or what?”\evb

\bvb TODO\evb

\bvb TODO\evb

\bvb I saw \inx{Bolder}’s, the bloody \inx{tue}’s, the child of Weden’s \inx{orlay}, hidden; grown did stand, higher than the meadows, the slender and much fair mistletoe.\evb

\bvb Of that beam, which meager seemed, became a baneful harm-flier; Hath began to shoot. Bolder’s brother was born early; that son of Weden, one night old, took to striking.\evb

\bvb Hands he never washed, nor head combed, before onto the pyre he did bear the opponent of Bolder. But Frie did lament, in the Fenhalls, the woe of Walhall; — Know ye yet, or what?\evb

\bvb TODO\evb

\bvb TODO\evb

\bvb TODO\evb

\bvb A hall saw she stand, far from the sun, on Neestrand; north face the doors. Fall venom-drops in through the smoke-vent, that hall is wound by the spines of snakes.\evb

\bvb There she sees wade, through heavy streams, oath-breaking men and murderwargs, and the one who confounds another’s understanding\footnotemark[1]. There sucks Nithehew from corpses passed-on; the warg tears men apart; — Know ye yet, or what?\evb
\footnotetext[1]{“who confounds another’s ear-rune,” probably referring to false counsellors.}

\bvb East sat the old woman, in \inx{Ironwood}, and nourished there the kins of \inx{Fenner}; becomes out of them all a single one, the pitch-forker of the moon, in the \inx{hame} of a troll.\footnotemark[1]\evb
\footnotetext[1]{The old hag raises the offspring of the wolf Fenner, of which one will swallow the moon (and according to \Snorri\ TODO the other the sun). See note 2 to the next v.}

\bvb He\footnotemark[1] fills himself with the life-force of \inx{fey} men; he reddens the abode of the Powers with red gore. Black becomes the sunshine about the summers afterwards\footnotemark[2]; the weather all hostile; — Know ye yet, or what?\evb
\footnotetext[1]{The wolf.}
\footnotetext[2]{After the sun is swallowed. But since the wallow does not tell us that this is a different wolf (it seems rather it be one and the same), it may reflect an earlier version of the myth, where one son of Fenner swallowed both the sun and moon. Yet, according to \Vafthrudnismal\ 36-37 it is Fenner himself who will swallow the sun (and thus likely the moon as well,) unless it there be taken as a \inx{hote} for ‘wolf’, which it undoubtedly originally is. See \inx{Moon}. TODO}

\bvb Sat there on the mound, and struck the harp, the troll-woman’s keeper, glad \inx{Edgethew}; above him crowed, in \inx{Gallowwood}, a fair-red cock, he who Fealer is called.\evb

\bvb Above the \inx{Ease} crowed \inx{Goldencombe}, he who wakes men, at the \inx{Father of Hosts}’s [estate]; but another one crows below the earth, a soot-red cock, at the halls of \inx{Hell}.\footnotemark[1]\evb
\footnotetext[1]{Here we have in quick succession a description of three cocks; the fair-red Fealer residing in \inx{Gallowwood} (probably the same as \inx{Ironwood}), Goldencombe in \inx{Osyard}, and a soot-red cock in Hell. TODO Index}

\bvb Barks Garm loudly before the Gnip-caverns; the rope will break, and Freck run. Much she knows of wisdom, forth I see yet further; about the fates of the Powers: the mighty [fates] of the victory-tues.\evb

\bvb Brothers will fight one another, and become one another’s bane; sister’s sons will defile their kinship.\footnotemark[1] ’Tis hard in the Home, great whoredom: halberd-\inx{eld}, short-sword-eld; shields are split. Wind-eld, \inx{warg}-eld; before the world\footnotemark[2] is overthrown, will no man another spare.\evb
\footnotetext[1]{Whether through incest or treachery. TODO: literary evidence of the phrase \emph{spilla sifjum}.}
\footnotetext[2]{\emph{ver-ǫld} ‘world’ might perhaps be better translated as ‘man-eld’, ‘the eld of man’ with the other elds preceding it.}

\bvb The \inx{sons of Mime} play, but the \inx{Metted} is kindled, to [the sound of] the shrill \inx{Horn of Yell}. Loud blows \inx{Homedall}; the horn is in the air; speaks Weden with the head of Mime.\evb

\bvb Trembles the ash of \inx{Ugdrassle}, standing; groans the old tree, and the ettin is loosened. All are frightened on the \inx{Hell-ways}, before \inx{Surt’s kinsman} does devour them.\evb

\bvb — What is with Ease? What is with Elves? Roars all Ettinhome, Ease are at the Thing. Dwarves groan before gates of stone, the princes of the mountain-wall; — Know ye yet, or what?\evb

\bvb Barks Garm loudly before the Gnip-caverns; the rope will break, and Freck run. Much she knows of wisdom, forth I see yet further; about the fates of the Powers: the mighty [fates] of the victory-tues.\evb

\bvb TODO\evb

\bvb A keel travels from the east; come will Muspell’s subjects, but Lock steers it; TODO.\evb

\bvb Surt comes from the south, with the switch-bane\footnotemark[1]; from the sword shines, [like] the sun of the \inx{wal-tues}; TODO men march on the \inx{Hell-ways}, but heaven is sundered.\footnotemark[2]\evb
\footnotetext[1]{Fire and brimstone.}
\footnotetext[2]{The enemies of the gods, led by Lock, march on Osyard. The final battle begins.}

\bvb Then comes \inx{Line}’s second sorrow to pass, as Weden goes to strike against the wolf; but the bane of \inx{Bellow}\footnotemark[1], bright, [goes] against Surt; then will Frie’s love\footnotemark[2] fall.\evb
\footnotetext[1]{\inx{Free}.}
\footnotetext[2]{Weden, her husband.}

\bvb Then comes the great lad of \inx{Sighfather}, Wither, to strike at the murderous beast; he lets his hand plunge the sword into the heart of \inx{Whethring}’s lad\footnotemark[1]; then is the father avenged.\evb
\footnotetext[1]{The son of Lock; the wolf.}

\bvb Then comes the renowned lad of Lathyn: the son of Weden goes the \inx{worm} to meet. Middenyard’s \inx{wigh}-guardian strikes out of wrath; all men will their homesteads empty.\footnotemark[1] The son of Firgyn goes nine paces, pained, away from the loathsome adder.\footnotemark[2]\evb
\footnotetext[1]{This long-line (“all – empty”) looks very out of place here. See the notes to the ON text for \Hauksbok’s order, which may be more original.}
\footnotetext[2]{Thunder, mortally wounded, struggles nine steps away from the serpent before he falls.}

\bvb TODO\evb

\bvb Barks Garm loudly before the Gnip-caverns; the rope will break, and Freck run. Much she knows of wisdom, forth I see yet further; about the fates of the Powers: the mighty [fates] of the victory-tues.\evb

\bvb TODO\evb

\bvb The Ease find each other on the Ithewalds, and about TODO\evb

\bvb TODO\evb

\bvb Unsown fields will grow, evil will all be bettered; Bolder will come. The \inx{wal-tues}, Hath and Bolder, will well inhabit the building-plots of Roft; — Know ye yet or what?\evb

\bvb TODO\evb

\bvb A hall she sees stand, fairer than the sun, thatched with gold, on Gimlee; there the dutiful drights† will dwell, and in their \inx{alder}-days enjoy delight.\evb

\bvb Then comes the shadowy dragon flying; the gleaming serpent down below from the \inx{Nithfells}. He, Nithehew, carries in his feathers—flying over the plain—corpses.” Now she will descend!\footnotemark[1]\evb
\footnotetext[1]{The wallow, referring to herself in third person, sinks back down into her grave, whence Weden woke her.}
