\book{The Speeches of Webthrithner. (Vafþrúðnismǫ́l)}\bookStart

\bvg {\small (Óðinn kvað:)}
\bva Ráð mér nú \alst{F}rigg \hld\ alls mik \alst{f}ara tíðir &
\ind at \alst{v}itja \alst{V}afþrúðnis; &
\alst{f}orvitni mikla \hld\ kveð'k mér á \alst{f}ornum stǫfum &
\ind við þann hinn \alst{a}lsvinna \alst{jǫ}tun.\eva

\bvb \inx{Weden} quoth: “Counsel me now, \inx{Frie}, as I desire to travel to visit \inx{Webthrithner}; greatly curious am I of ancient staves\footnotemark[1] by that all-wise \inx{ettin}."\evb
\footnotetext[1]{Ancient (pieces of) lore; cf. v. 55. — Meaning (from \emph{great} onwards) is clear, but form is very confused.}
\evg


\bvg {\small (Frigg kvað:)}
\bva \alst{H}ęima lętja \hld\ mynda'k \alst{H}ęrjafǫðr &
\ind í \alst{g}ǫrðum \alst{g}oða; &
\alst{ę}ngi \alst{jǫ}tun \hld\ hugða'k \alst{ja}fnramman &
\ind sęm \alst{V}afþrúðni \alst{v}esa.\eva

\bvb Frie quoth: “I would encourage the \inx{Leader of Armies} to [stay at] home in the yards of the gods, for I've judged no ettin be as strong as\footnotemark[3] Webthrithner."\evb
\footnotetext[3]{Lit. ‘equal-strong'.}
\evg


\bvg {\small (Óðinn kvað:)}
\bva Fjǫlð ek fór, \hld\ fjǫlð fręistaða'k, &
\ind fjǫlð ek ręynda ręgin; &
hitt vil'k vita, \hld\ hvé Vafþrúðnis &
\ind salakynni séi.\eva

\bvb Weden quoth: “Much I travelled, much I tried, much I tested the \inx{Reins}\footnotemark[4]. \emph{This} I want to know, how the condition of the halls of Webthrithner might be?"\evb
\footnotetext[4]{The gods.}
\evg


\bvg {\small (Frigg kvað:)}
\bva Hęill þú farir, \hld\ hęill þú aptr komir, &
\ind hęill á sinnum séir; &
ǿði þér dugi \hld\ hvar's skalt, Aldafǫðr, &
\ind orðum mæla jǫtun.\eva

\bvb Frie quoth: “Whole may thou travel, whole may thou return, whole may thou be on thy paths! May thy wisdom suffice, \inx{Leader of Men}, when thou go to exchange words with the ettin."\evb
\evg


\bvg
\bva Fór þá Óðinn \hld\ at fręista orðspęki &
\ind þess hins alsvinna jǫtuns; &
at hǫllu hann kom, \hld\ es\footnotemark[1] átti Íms faðir; &
\ind inn gekk Yggr þegar.\eva
\footnotetext[1]{Ms. \emph{ok} corrected to \emph{es}. Alliteration is lacking in this line, for which reason FJ emends \emph{Íms} to \emph{Hymis}.}

\bvb Then went Weden, to try the word-wisdom of that all-wise ettin. To the hall he came, which the father of \inx{Ime}\footnotemark[5] owned; shortly the \inx{Frightener}\footnotemark[6] walked in.\evb
\footnotetext[5]{Webthrithner.}
\footnotetext[6]{Weden.}
\evg


\bvg {\small (Óðinn kvað:)}
\bva Hęill þú nú, Vafþrúðnir, \hld\ nú em'k í hǫll kominn &
\ind á þik sjalfan séa; &
hitt vilk fyrst vita, \hld\ ef fróðr séir &
\ind eða alsviðr, jǫtunn.\eva

\bvb Weden quoth: “Hail thee now, Webthrithner; now I have come into the hall, to see thee thyself. \emph{This} I want to know first, if knowing thou might be, or all-wise, ettin!"\evb
\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Hvat's þat manna, \hld\ es í mínum sal &
\ind verpumk orði á? &
út þú né kømr \hld\ órum hǫllum frá. &
\ind nema þú inn snotrari séir.\eva

\bvb Webthrithner quoth: “What is that of men\footnotemark[10], that in \emph{my} hall throws words at me? Thou will not come \emph{out}, from \emph{our}\footnotemark[11] halls, unless thou be the wiser [of us two]."\evb
\footnotetext[10]{Ie., ‘what man is that'. The use of the neuter pronoun \emph{hvat} by Web-str. may be seen as an insult or a way of belittling the guest.}
\footnotetext[11]{Prob. again meaning ‘my', unless Web-str. has allies present in the hall, but no such indication is given.}
\evg


\bvg {\small (Óðinn kvað:)}
\bva Gagnráðr\footnotemark[5] hęiti'k, \hld\ nú em'k af gǫngu kominn, &
\ind þyrstr til þinna sala; &
laðar þurfi \hld\ hęf'k lęngi farit &
\ind ok þinna andfanga, jǫtunn.\eva
\footnotetext[5]{R's \emph{Gagnráðr} ‘Gainred', is attested as Gangráðr ‘Journey-adviser' in \emph{Gylf}.}

\bvb Weden quoth: “\inx{Gainred} I am called, I am come from the journey, thirsty to thy halls. I have travelled for a long time in need of hospitality, and of thy reception, ettin!"\evb
\evg


\bvg \bvg {\small (Vafþrúðnir kvað:)}
\bva Hví þú þá, Gagnráðr, \hld\ mælisk af golfi fyrir? &
\ind far þú í sess í sal; &
þá skal fręista, \hld\ hvárr flęira viti, &
\ind gęstr eða hinn gamli þulr.\eva

\bvb Webthrithner quoth: “Why then, Gainred, art thou speaking from the floor before [me]? Take a seat in the hall! Then it shall be proven, which of the two might know more; the guest, or the old \inx{thyle}."\evb
\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Óauðigr maðr, \hld\ es til auðigs kømr, &
\ind mæli þarft eða þęgi; &
ofrmælgi mikil \hld\ hygg at illa geti &
\ind hvęim's við kaldrifjaðan kømr.\eva

\bvb Gainred quoth: “An unwealthy man, who comes to a wealthy [one], ought to speak what is needed, or be silent.\footnotemark[14] Much over-speaking\footnotemark[15], I judge, will be bad for the one who comes to a cold-ribbed\footnotemark[16] [man]."\evb
\footnotetext[14]{Line identical to \emph{High} 18/2. The whole verse strongly reminds of verses from the \emph{Guest-thread} portion of said poem.}
\footnotetext[15]{“Speaking too much".}
\footnotetext[16]{That is, ‘cold-hearted', ‘cunning'.}
\evg


\bvg \bvg {\small (Vafþrúðnir kvað:)}
\bva Sęg mér, Gagnráðr, \hld\ alls á golfi vill &
\ind þíns of fręista frama, &
hvé hęstr hęitir, \hld\ sá's hvęrjan dręgr &
\ind dag of dróttmǫgu.\eva

\bvb Webthrithner quoth: “Say to me, Gainred, since on the floor I will to try thy fame: What is the horse called, which pulls each \emph{day} above the sons of the retinue \ken{Men}?"\evb
\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Skinfaxi hęitir, \hld\ es hinn skíra dręgr &
\ind dag of dróttmǫgu; &
hęsta baztr \hld\ þykkir með Hręiðgotum; &
\ind ęy lýsir mǫn af mari.\eva

\bvb Gainred quoth: “\inx{Shining-fax} [that one] is called, who pulls the bright day above the sons of the retinue. The best of horses he seems among the \inx{Rode-goths}; the mane of that stallion ever shines."\evb
\evg


\bvg (Vafþrúðnir kvað:) &
\bva Sęg þat, Gagnráðr, \hld\ alls á golfi vill &
\ind þíns of fręista frama, &
hvé jór hęitir, \hld\ sá's austan dręgr &
\ind nótt of nýt ręgin.\eva

\bvb Webthrithner quoth: “Say this, Gainred, since on the floor I will to try thy fame: What is the horse called, which from the east pulls night above the useful \inx{Reins}?"\evb
\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Hrímfaxi hęitir, \hld\ es hvęrja dręgr &
\ind nótt of nýt ręgin; &
méldropa fęllir \hld\ morgin hvęrjan; &
\ind þaðan kømr dǫgg of dala.\eva

\bvb Gainred quoth: “\inx{Frost-fax} [that one] is called, who pulls each night above the useful Reins. Every morning he lets foam fall from his bit\footnotemark[26]; thence comes dew in the valleys."\evb
\footnotetext[26]{Lit. “he fells bit-drops".}\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Sęg þat, Gagnráðr, \hld\ alls á golfi vill &
\ind þíns of fręista frama, &
hvé ǫ́ hęitir, \hld\ sú's dęilir með jǫtna sonum &
\ind grund ok með goðum.\eva

\bvb Webthrithner quoth: “Say this, Gainred, since on the floor I will to try thy fame; How the river is called, which divides the ground between the sons of ettins and the gods?"\evb
\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Ífing hęitir ǫ́, \hld\ es dęilir með jǫtna sonum &
\ind grund ok með goðum; &
opin rinna \hld\ hón skal um aldrdaga; &
\ind verðr-at íss á ǫ́.\eva

\bvb Gainred quoth: “\inx{Iving} the river is called, which divides the ground between the sons of ettins and the gods. Throughout [her] life-days she shall flow open; ice forms not on the river."\evb
\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Sęg þat, Gagnráðr, \hld\ alls á golfi vill &
\ind þíns of fręista frama, &
hvé vǫllr hęitir, \hld\ es finnask vigi at &
\ind Surtr ok hin svǫ́su goð.\eva

\bvb Webthrithner quoth: “Say this, Gainred, since on the floor I will to try thy fame: How that valley is called, where \inx{Surt} and the excellent gods find each other at war?"\evb
\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Vígríðr hęitir vǫllr, \hld\ es finnask vígi at &
\ind Surtr ok hin svǫ́su goð; &
hundrað rasta \hld\ hann's á hvęrjan veg; &
\ind sá's þęim vǫllr vitaðr.\eva

\bvb Gainred quoth: “\inx{Battle-rider} is the valley called, where Surt and the cheerful gods find each other at war. A hundred rests\footnotemark[30], he stretches in each direction; that valley is known for them.\footnotemark[31]"\evb
\footnotetext[30]{An old unit of length, from its name prob. the length a horse could travel before resting.}
\footnotetext[31]{That is, known for its great size.}\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Fróðr est nú gęstr, \hld\ far á bękk jǫtuns, &
\ind mælumk í sessi saman; &
hǫfði vęðja \hld\ vit skulum hǫllu í &
\ind gęstr, of gęðspęki.\eva

\bvb Webthrithner quoth: “Knowing art thou now, guest, sit down on the ettin's bench; let us speak while sitting together. In the hall we shall wager a head, guest, over [our] mind-wisdom."\evb
\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Sęg þat hit ęina, \hld\ ef þitt ǿði\footnotemark[10] dugir &
\ind ok þú Vafþrúðnir vitir, &
hvaðan jǫrð of kom \hld\ eða upphiminn &
\ind fyrst, hinn fróði jǫtunn.\eva
\footnotetext[10]{Starting with \emph{ǿði}, the poem is also preserved in 748.}

\bvb Gainred quoth: “Say the first\footnotemark[32], if thy wisdom suffices, and thou, Webthrithner, might know: Whence, O knowing ettin, the earth first came, or \inx{up-heaven}?"\evb
\footnotemark[32]{Lit. ‘one'.}\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Ór Ymis holdi \hld\ vas jǫrð of skǫpuð, &
\ind ęn ór bęinum bjǫrg, &
himinn ór hausi \hld\ hins hrimkalda jǫtuns, &
\ind ęn ór svęita sær.\eva

\bvb Webthrithner quoth: “Out of \inx{Yime's} hull\footnotemark[35], the earth was created, but the mountains out of his bones. Heaven out of the skull of the frost-cold ettin, but the sea out of his sweat.\footnotemark[36]"\evb
\footnotetext[35]{His body.}
\footnotetext[36]{\emph{svęiti} ‘sweat' is a common kenning for blood. — This v. closely resembles \emph{Grím} 40.}\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Sęg þat annat, \hld\ ef þitt ǿði dugir &
\ind ok þú Vafþrúðnir vitir, &
hvaðan máni of kom, \hld\ svát fęrr menn yfir, &
\ind eða sól hit sama.\eva

\bvb Gainred quoth: “Say the second, if thy wisdom suffices, and thou, Webthrithner, might know: Whence the moon came, so that it travels over men, or likewise the sun?"\evb
\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Mundilfari hęitir, \hld\ hann's Mána faðir &
\ind ok svá Solar hit sama; &
himin hverfa \hld\ þau skulu hvęrjan dag &
\ind ǫldum at ártali.\eva

\bvb Webthrithner quoth: “\inx{Moundelfare} [that one] is called, he is the father of the Moon, and likewise of the Sun. They shall circle in the heavens every day, for men to reckon time\footnotemark[40]."\evb
\footnotetext[40]{Lit. “for men to year-reckoning".}\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Sęg þat þriðja, \hld\ alls þik svinnan kveða &
\ind ok þú Vafþrúðnir vitir, &
hvaðan dagr of kom, \hld\ sá's fęrr drótt yfir, &
\ind eða nótt með niðum.\eva

\bvb Gainred quoth: “Say the third, since [they] call thee wise, and thou, Webthrithner, might know: Whence the day came, the one that travels over the rettinue, or night with the moon-phases?"\evb
\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Dęllingr hęitir, \hld\ hann's Dags faðir, &
\ind ęn Nótt vas Nǫrvi borin; &
ný ok nið \hld\ skópu nýt ręgin &
\ind ǫldum at ártali.\eva

\bvb Webthrithner quoth: “\inx{Delling} [that one] is called, he is the father of \inx{Day}, but \inx{Night} was born to \inx{Nare}. The waxing and waning [of the moon], the useful Reins created, for men to reckon time."\evb
\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Sęg þat fjórða, \hld\ alls þik fróðan kveða, &
\ind ok þú Vafþrúðnir vitir, &
hvaðan vetr of kom \hld\ eða varmt sumar &
\ind fyrst með fróð ręgin.\eva

\bvb Gainred quoth: “Say the fourth, since [they] call thee knowing, and thou, Webthrithner, might know: Whence winter first came, or the warm summer, among the knowing Reins?"\evb
\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Vindsvalr hęitir, \hld\ hann's Vetrar faðir, &
\ind ęn Svǫ́suðr Sumars.\footnotemark[15]\eva
\footnotetext[15]{Second half of the v. seems missing.}

\bvb Webthrithner quoth: “\inx{Wind-cool} [that one] is called, he is the father of \inx{Winter}, but \inx{Delightful} of \inx{Summer}."\evb
\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Sęg þat fimta, \hld\ alls þik fróðan kveða, &
\ind ok þú Vafþrúðnir vitir, &
hvęrr ása ęlztr \hld\ eða Ymis niðja &
\ind yrði í árdaga.\eva

\bvb Gainred quoth: “Say the fifth, since [they] call thee knowing, and thou, Webthrithner, might know: Who, in days of yore became the eldest of the \inx{Anses}, or of the descendants of Yime?"\evb
\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Ørófi vetra \hld\ áðr væri jǫrð skǫpuð, &
\ind þá vas Bergęlmir borinn, &
Þrúðgęlmir \hld\ vas þess faðir, &
\ind ęn Aurgęlmir afi.\eva

\bvb Webthrithner quoth: “Uncountable winters before the earth would be created, then \inx{Bear-yeller} was born. \inx{Strength-yeller} was \emph{that one's} father, and \inx{Mud-yeller} the grandfather."\evb
\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Sęg þat sétta, \hld\ alls þik svinnan kveða, &
\ind ok þú Vafþrúðnir vitir, &
hvaðan Aurgęlmir kom \hld\ með jǫtna sonum &
\ind fyrst, hinn fróði jǫtunn.\eva

\bvb Gainred quoth: “Say the sixth, since [they] call thee wise, and thou, Webthrithner, might know: Whence, O knowing ettin, Mud-yeller first came among the sons of ettins?"\evb
\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Ór Élivǫ́gum \hld\ stukku ęitrdropar, &
\ind svá óx unz ór varð jǫtunn; &
órar ættir \hld\ kómu þar allar saman; &
\ind því's þat æ alt til atalt.\footnotemark[20]\eva
\footnotetext[20]{Lines 3–4 missing in R and 748, but quoted in \emph{Gylf}.}

\bvb Webthrithner quoth: “From the \inx{Ell-waves}, poison-drops splashed; thus [it] grew until an ettin emerged. \emph{Our} family lines all together originated there, therefore our race\footnotemark[45] is forever fierce against all.\footnotemark[46]"\evb
\footnotetext[45]{Lit. ‘it' or ‘that'.}
\footnotetext[46]{Somewhat strange phrasing, but the line does not appear damaged. It is clearly an explanation of the fierce and maleficent nature of the ettins, as their first ancestors were created from poison.}\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Sęg þat sjaunda, \hld\ alls þik svinnan kveða, &
\ind ok þú Vafþrúðnir vitir, &
hvé sá bǫrn gat \hld\ hinn baldni\footnotemark[25] jǫtunn, &
\ind es hann hafði-t gýgjar gaman.\eva
\footnotetext[25]{R has \emph{aldni}, ‘aged, old'. This breaks alliteration; \emph{baldni} ‘bold, defiant' has been substituted from 748.}

\bvb Gainred quoth: “Say the seventh, since [they] call thee wise, and thou, Webthrithner, might know: How did that one, the defiant ettin, beget children, when he did not enjoy the [carnal] pleasure of a troll-woman?"\evb
\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Und hęndi vaxa \hld\ kvǫ́ðu hrímþursi &
\ind męy ok mǫg saman; &
fótr við fǿti \hld\ gat hins fróða jǫtuns &
\ind sexhǫfðaðan son.\eva

\bvb Webthrithner quoth: “Neath the hand\footnotemark[50] on the \inx{frost-thurse}, [they] said that a maiden and lad grew together. A foot against a foot begot, for the knowing ettin, a six-headed son."\evb
\footnotetext[50]{\emph{hęndi} (dative of \emph{hǫnd}) means ‘hand', but might here be a poetic circumlocution for ‘arm'.}\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Sęg þat áttunda, \hld\ alls þik fróðan kveða, &
\ind ok þú Vafþrúðnir vitir, &
hvat fyrst of mant \hld\ eða fręmst of vęizt, &
\ind þú est alsviðr jǫtunn.\eva

\bvb Gainred quoth: “Say the eigth, since [they] call thee knowing, and thou, Webthrithner, might know: What dost thou first remember, or earliest know?\footnotemark[55] Thou art all-wise, ettin."\evb
\footnotetext[55]{Cf. Vsp 1.}\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Ørófi vetra \hld\ áðr væri jǫrð of skǫpuð, &
\ind þá vas Bergęlmir borinn; &
þat fyrst um man'k, \hld\ es hinn fróði jǫtunn &
\ind á vas lúðr of lagiðr.\footnotemark[30]\eva
\footnotetext[30]{This verse is quoted in \emph{Gylf}.}

\bvb Webthrithner quoth: “Uncountable winters before the earth would be created, then Bear-yeller was born. \emph{That} I first remember, when the knowing ettin\footnotemark[60] was laid down on the funeral-bed\footnotemark[61]."\evb
\footnotetext[60]{That is, Bear-yeller. Cf. v. 29.}
\footnotetext[61]{\emph{lúðr}, a tricky word.}\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Sęg þat níunda, \hld\ alls þik svinnan kveða, &
\ind ok þú Vafþrúðnir vitir, &
hvaðan vindr of kømr \hld\ svát fęrr vág yfir, &
\ind æ męnn hann sjalfan of séa.\eva

\bvb Gainred quoth: “Say the ninth, since [they] call thee wise, and thou, Webthrithner, might know: Whence the wind comes, so that he travels over the wave; forever men see him himself.\footnotemark[65]"\evb
\footnotetext[65]{Perhaps a negation has been lost here; the wind is never seen by men.}\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Hræsvęlgr hęitir, \hld\ es sitr á himins ęnda, &
\ind jǫtunn í arnar ham; &
af hans vængjum \hld\ kveða vind koma &
\ind alla męnn yfir.\eva

\bvb Webthrithner quoth: “\inx{Corpse-swallower} [that one] is called, which sits at the end of the heavens, an ettin in the shape of an eagle. From his wings, they say [that] the wind comes over all men."\evb
\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Sęg þat tíunda, \hld\ alls þú tíva rǫk &
\ind ǫll Vafþrúðnir vitir, &
hvaðan Njǫrðr of kom \hld\ með niðjum ása. &
Hófum ok hǫrgum \hld\ hann ræðr hundmǫrgum &
\ind ok varð-at hann ǫ́sum alinn.\eva

\bvb Gainred quoth: “Say the \emph{tenth}, since thou, Webthrithner, of all the fates of the \inx{Tues} might know: Whence \inx{Nearth} came into the company of the kinsmen of the \inx{Anses}? He rules an immense number\footnotemark[68] of \inx{hoves} and \inx{heargs}, and he was not begotten among the Anses."\evb
\footnotetext[68]{Lit. “he rules hound-many".}\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Í Vanahęimi \hld\ skópu hann vís ręgin &
\ind ok sęldu at gíslingu goðum, &
í aldar rǫk \hld\ hann mun aptr koma &
\ind hęim með vísum vǫnum.\eva

\bvb Webthrithner quoth: “In \inx{Wane-Home}, the wise \inx{Reins}\footnotemark[69] created him, and sold him as a hostage to the gods. In the fate of the age, he will come back, home among the wise \inx{Wanes}."\evb
\footnotetext[69]{Though \emph{ręgin} usually serves as a direct synonym of \emph{goð} 'god(s)', it here seems to refer specifically to the Wanes, in contrast with the \inx{Eses} or gods.}\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Sęg þat ęllipta, \hld\ hvar ýtar túnum í &
\ind hǫggvask hvęrjan dag; &
val þęir kjósa \hld\ ok ríða vígi frá, &
\ind sitja męir of sáttir saman.\footnotemark[35]\eva
\footnotetext[35]{This and the next v. are damaged in both R and 748; R has only this verse, but splits it in two (the 2nd starting with \emph{val}), while 748 has 40:1 (Ms.: \emph{S. þ. e. XI}) and then jumps to the answer v. 41. They have here been reconstructed, but it is possible some lines are still missing.}

\bvb Gainred quoth: “Say the eleventh, Where men in yards, hew away at each other every day? They choose those destined to die in war, and ride [away] from battle; [then] they sit more content together."\evb
\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Allir ęinhęrjar \hld\ Óðins túnum í &
\ind hǫggvask hvęrjan dag, &
val þeir kjósa \hld\ ok ríða vígi frá, &
\ind sitja męir of sáttir saman.\eva

\bvb Webthrithner quoth: “In Weden's yards, all the \inx{Lone Warriors} hew away at each other every day. They choose those destined to die in war, and ride [away] from battle; [then] they sit more content together."\evb
\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Sęg þat tolpta, \hld\ hví þú tíva rǫk &
\ind ǫll Vafþrúðnir vitir, &
frá jǫtna rúnum \hld\ ok allra goða &
\ind þú hit sannasta sęgir, &
\ind hinn alsvinni jǫtunn.\eva

\bvb Gainred quoth: “Say the twelfth, Why thou, Webthrithner, shouldst know all the fates of the \inx{Tues}\footnotemark[73]? From the \inx{runes} of the ettins and of all the gods, thou, the all-wise ettin, speakest most truly."\evb
\footnotetext[73]{The gods. Formation identical to \emph{ragna rǫk} ‘the fates of the Reins'.}\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Frá jǫtna rúnum \hld\ ok allra goða &
\ind ek kann sęgja satt, &
þvíat hvęrn hęf'k \hld\ heim of komit, &
níu kom'k hęima \hld\ fyr niflhęl neðan; &
\ind hinig dęyja ór hęlju halir.\eva

\bvb Webthrithner quoth: “From the runes of the ettins and of all the gods I can speak truly, for I have been about each \inx{Home}. I was about nine Homes beneath Nivelhell; this way men die out of Hell\footnotemark[1]."\evb
\footnotetext[1]{A difficult verse. Finnur considers \emph{ór hęlju} “out of Hell” a later interpolation.}\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Fjǫlð ek fór, \hld\ fjǫlð fręistaða'k, &
\ind fjǫlð ek ręynda ręgin; &
hvat lifir manna, \hld\ þá's hinn mæra líðr &
\ind fimbulvetr með firum?\eva

\bvb Gainred quoth: “Much I travelled, much I tried, much I tested the \inx{Reins}.\footnotemark[80] What remains\footnotemark[79] of men, when the famous \inx{fimbol-winter} passes among firs\footnotemark[81]?”\evb
\footnotetext[79]{Lit. “lives".}
\footnotetext[80]{Here begins the repetition of the same “mantra" used in v. 3, which continues until the final question (v. 54).}
\footnotetext[81]{Among men.}\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Líf ok Lífþrasir, \hld\ ęn þau lęynask munu &
\ind í holti Hoddmímis; &
morgindǫggvar \hld\ þau sér at mat hafa; &
\ind þaðan af aldir alask.\eva

\bvb Webthrithner quoth: “Life and Lifethrasher, and they will hide themselves in the wood of Hoard-Mime\footnotemark[85]. Morning-dew they [will] have as food; thereof generations [will] be bred.”\evb
\footnotetext[85]{Prob. the same as Uggdrassle.}}\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Fjǫlð ek fór, \hld\ fjǫlð fręistaða'k, &
\ind fjǫlð ek ręynda ręgin; &
hvaðan kømr sól \hld\ á hinn slétta himin, &
\ind es þessa hęfr Fęnrir farit?\eva

\bvb Gainred quoth: “Much I travelled, much I tried, much I tested the Reins. Whence comes sun onto the smooth heaven, when \inx{Fenner} has killed this one\footnotemark[1]?"\evb
\footnotetext[1]{That is, the current incarnation of the sun, as explained in the next v.}\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Ęina dóttur \hld\ berr alfrǫðull, &
\ind áðr hana Fęnrir fari; &
sú skal ríða, \hld\ þá's ręgin dęyja, &
\ind móður brautir mær.\eva

\bvb Webthrithner quoth: “One daughter the elf-wheel <= Sun> bears, before \inx{Fenner} might kill her. When the Reins die, that one, the maiden, shall ride the paths of the mother.”\evb
\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Fjǫlð ek fór, \hld\ fjǫlð fręistaða'k, &
\ind fjǫlð ek ręynda ręgin; &
hvęrjar 'ro męyjar, \hld\ es líða mar yfir, &
\ind fróðgęðjaðar fara.\eva

\bvb Weden quoth: “Much I travelled, much I tried, much I tested the Reins. Which are the maidens that pass over the ocean; wise-minded they go?”\evb
\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Þríar þjóðár \hld\ falla þorp yfir &
\ind męyja Mǫgþrasis; &
hamingjur ęinar \hld\ þær’s í hęimi eru, &
\ind þó þær með jǫtnum alask.\eva

\bvb Webthrithner quoth: “Three great rivers fall over the settlement of the maidens of Maythrasher; the only Hamings that are in the Home,\footnotemark[1] though they are raised among the ettins\footnotemark[2]."\evb
\footnotetext[1]{Either in Ettinhome, or in the entire world.}
\footnotetext[2]{See index entry Maythrasher.}\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Fjǫlð ek fór, \hld\ fjǫlð fręistaða'k, &
\ind fjǫlð ek ręynda ręgin; &
hvęrir ráða æsir \hld\ ęignum goða, &
\ind þá's sloknar Surtalogi?\eva

\bvb Gainred quoth: “Much I travelled, much I tried, much I tested the Reins. Which properties of the gods will the \inx{Anses} [still] rule\footnotemark[105], when the flame of \inx{Surt} burns out?"\evb
\footnotetext[105]{Or ‘control’.}
\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Víðarr ok Váli \hld\ byggva vé goða, &
\ind þá's sloknar Surtalogi; &
Móði ok Magni \hld\ skulu Mjǫlni hafa &
\ind Vingnis at vígþroti.\eva

\bvb Webthrithner quoth: “\inx{Wider} and \inx{Weel} [will] inhabit the sanctuaries of the gods, when the \inx{flame of Surt} burns out. \inx{Mood} and \inx{Main} will own \inx{Meldner}, when \inx{Wingner} can no longer fight\footnotemark[110]."\evb
\footnotetext[110]{Lit. “at Wingner's fight-exhaustion", referring to his death.}
\evg


\bvg {\small (Gagnráðr kvað:)}
\bva Fjǫlð ek fór, \hld\ fjǫlð fręistaða'k, &
\ind fjǫlð ek ręynda ręgin; &
hvat verðr Óðni \hld\ at aldrlagi, &
\ind þá's rjúfask ręgin?\eva

\bvb Gainred quoth: “Much I travelled, much I tried, much I tested the Reins. What brings Weden’s life to an end, when the Reins are broken?"\evb
\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Ulfr glęypa \hld\ mun Aldafǫðr, &
\ind þess mun Víðarr vreka; &
kalda kjapta \hld\ hann klyfja mun &
\ind vitnis vígi at.\eva

\bvb Webthrithner quoth: “The wolf will swallow \inx{Eldfather}; Wider will avenge that. He will cleave the cold jaws of the wolf at the battle."\evb
\evg


\bvg {\small (Gagnráðr kvað:) &
\bva Fjǫlð ek fór, \hld\ fjǫlð fręistaða'k, &
\ind fjǫlð ek ręynda ręgin; &
hvat mælti Óðinn, \hld\ áðr á bál stigi, &
\ind sjalfr í ęyra syni?\eva

\bvb Gainred quoth: “Much I travelled, much I tried, much I tested the Reins. What spoke Weden himself, before [he]\footnotemark[115] would step onto the funeral pyre, into the ear of the son?"\evb
\footnotetext[115]{Prob. Weden's son, that is \inx{Balder}.}\evg


\bvg {\small (Vafþrúðnir kvað:)}
\bva Ęy \edtext{manngi}{\Afootnote{manni \Regius\AM\ \emph{is impossible; we need a nominative here.}}} vęit, \hld\ hvat þú í árdaga &
\ind sagðir í ęyra syni; &
fęigum munni \hld\ mælta’k mína forna stafi &
\ind ok of ragna rǫk. &
Nú við Óðin \hld\ dęilda’k mína orðspęki; &
\ind þú est æ vísastr vera.\eva

\bvb Webthrithner quoth: “Ever no man knows, what thou in days of yore saidst in the ear of the son. With a death-doomed\footnoteB{Webthrithner here realizes that he was bound to die from the moment (v. 19) he proposed the wager, as no being can outwit Weden.} mouth I spoke my ancient utterings, and of the \inx{Rakes of the Reins}. Now with Weden I shared my word-wisdom\footnoteB{The same word-wisdom Weden in v. 5 set out to try.}; thou art ever wisest of beings\footnoteB{Word used is \emph{verr} ‘husband, man’. Perhaps in the broader sense of ‘male being’.}."\evb
\evg
