\book{The Speech of Web-strengthener. (Vafþrúðnismǫ́l)}

(Óðinn kvað:) &
\bva Ráð mér nú \alst{F}rigg \hld alls mik \alst{f}ara tíðir &
at \alst{v}itja \alst{V}afþrúðnis; &
\alst{f}orvitni mikla \hld kveð'k mér á \alst{f}ornum stǫfum &
við þann hinn \alst{a}lsvinna \alst{jǫ}tun.\\ \&

\bvb \textbf{Woden} quoth: \\ “Counsel me now, \textbf{Frie}, as I desire to travel to visit \textbf{Web-strengthener}; I have great curiosity to say ancient staves\footnotemark[1] against that all-wise \textbf{ettin}." \\
\footnotetext[1]{Ancient (pieces of) lore; cf. v. 55.} 

(Frigg kvað:) &
\bva \alst{H}ęima lętja \hld mynda'k \alst{H}ęrjafǫðr &
í \alst{g}ǫrðum \alst{g}oða; &
\alst{ę}ngi \alst{jǫ}tun \hld hugða'k \alst{ja}fnramman &
sęm \alst{V}afþrúðni \alst{v}esa.\\ \&

\bvb Frie quoth: \\ “I would encourage the \textbf{Leader of Armies} to [stay at] home in the yards of the gods, for I've judged no ettin be as strong as\footnotemark[3] Web-strengthener." \\
\footnotetext[3]{Lit. ‘equal-strong'.}

(Óðinn kvað:) &
\bva Fjǫlð ek fór, \hld fjǫlð fręistaða'k, &
fjǫlð ek ręynda ręgin; &
hitt vil'k vita, \hld hvé Vafþrúðnis &
salakynni séi.\\ \&

\bvb Woden quoth: \\ “Much I travelled, much I tried, much I tested the \textbf{Powers}\footnotemark[4]. \emph{This} I want to know, how the condition of the halls of Web-strengthener might be?" \\
\footnotetext[4]{The gods.}

(Frigg kvað:) &
\bva Hęill þú farir, \hld hęill þú aptr komir, &
hęill á sinnum séir; &
ǿði þér dugi \hld hvar's skalt, Aldafǫðr, &
orðum mæla jǫtun.\\ \&

\bvb Frie quoth: \\ “Whole may thou travel, whole may thou return, whole may thou be on thy paths! May, thy wisdom suffice, \textbf{Leader of Men}, when thou go to exchange words with the ettin." \\

\bva Fór þá Óðinn \hld at fręista orðspęki &
þess hins alsvinna jǫtuns; &
at hǫllu hann kom, \hld es\footnotemark[1] átti Íms faðir; &
inn gekk Yggr þegar.\\ \&
\footnotetext[1]{Ms. \emph{ok} corrected to \emph{es}. Alliteration is lacking in this line, for which reason FJ emends \emph{Íms} to \emph{Hymis}.}

\bvb Then went Woden, to try the word-wisdom of that all-wise ettin. To the hall he came, which the father of \textbf{Ime}\footnotemark[5] owned; shortly the \textbf{Frightener}\footnotemark[6] walked in. \\
\footnotetext[5]{Web-strengthener.}
\footnotetext[6]{Woden.}

(Óðinn kvað:) &
\bva Hęill þú nú, Vafþrúðnir, \hld nú em'k í hǫll kominn &
á þik sjalfan séa; &
hitt vilk fyrst vita, \hld ef fróðr séir &
eða alsviðr, jǫtunn.\\ \&

\bvb Woden quoth: \\ “Hail thee now, Web-strengthener; now I have come into the hall, to see thee thyself. \emph{This} I want to know first, if knowing thou might be, or all-wise, ettin!" \\

(Vafþrúðnir kvað:) &
\bva Hvat's þat manna, \hld es í mínum sal &
verpumk orði á? &
út þú né kømr \hld órum hǫllum frá. &
nema þú inn snotrari séir.\\ \&

\bvb Web-strengthener quoth: \\ “What is that of men\footnotemark[10], that in \emph{my} hall throws words at me? Thou will not come \emph{out}, from \emph{our}\footnotemark[11] halls, unless thou be the wiser [of us two]." \\
\footnotetext[10]{Ie., ‘what man is that'. The use of the neuter pronoun \emph{hvat} by Web-str. may be seen as an insult or a way of belittling the guest.}
\footnotetext[11]{Prob. again meaning ‘my', unless Web-str. has allies present in the hall, but no such indication is given.}

(Óðinn kvað:) &
\bva Gagnráðr\footnotemark[5] hęiti'k, \hld nú em'k af gǫngu kominn, &
þyrstr til þinna sala; &
laðar þurfi \hld hęf'k lęngi farit &
ok þinna andfanga, jǫtunn.\\ \&
\footnotetext[5]{R's \emph{Gagnráðr} ‘Gain-adviser', is attested as Gangráðr ‘Journey-adviser' in \emph{Gylf}.}

\bvb Woden quoth: \\ “\textbf{Gain-adviser} I am called, I am come from the journey, thirsty to thy halls. I have travelled for a long time in need of hospitality, and of thy reception, ettin!" \\

(Vafþrúðnir kvað:) &
\bva Hví þú þá, Gagnráðr, \hld mælisk af golfi fyrir? &
far þú í sess í sal; &
þá skal fręista, \hld hvárr flęira viti, &
gęstr eða hinn gamli þulr.\\ \&

\bvb Web-strengthener quoth: \\ “Why then, Gain-adviser, art thou speaking from the floor before [me]? Take a seat in the hall! Then it shall be proven, which of the two might know more; the guest, or the old \textbf{thyle}." \\

(Gagnráðr kvað:) &
\bva Óauðigr maðr, \hld es til auðigs kømr, &
mæli þarft eða þęgi; &
ofrmælgi mikil \hld hygg at illa geti &
hvęim's við kaldrifjaðan kømr.\\ \&

\bvb Gain-adviser quoth: \\ “An unwealthy man, who comes to a wealthy [one], ought to speak what is needed, or be silent.\footnotemark[14] Much over-speaking\footnotemark[15], I judge, will be bad for the one who comes to a cold-ribbed\footnotemark[16] [man]." \\
\footnotetext[14]{Line identical to \emph{High} 18/2. The whole verse strongly reminds of verses from the \emph{Guest-thread} portion of said poem.}
\footnotetext[15]{“Speaking too much".}
\footnotetext[16]{That is, ‘cold-hearted', ‘cunning'.}

(Vafþrúðnir kvað:) &
\bva Sęg mér, Gagnráðr, \hld alls á golfi vill &
þíns of fręista frama, &
hvé hęstr hęitir, \hld sá's hvęrjan dręgr &
dag of dróttmǫgu.\\ \&

\bvb Web-strengthener quoth: \\ “Say to me, Gain-adviser, since on the floor I will to try thy fame: What is the horse called, which drags each \emph{day} above the sons of the retinue\footnotemark[20]?" \\
\footnotetext[20]{Kenning for ‘men', ‘mankind'.}

(Gagnráðr kvað:) &
\bva Skinfaxi hęitir, \hld es hinn skíra dręgr &
dag of dróttmǫgu; &
hęsta baztr \hld þykkir með Hręiðgotum; &
ęy lýsir mǫn af mari.\\ \&

\bvb Gain-adviser quoth: \\ “\textbf{Shining-fax} [that one] is called, who drags the bright day above the sons of the retinue. The best of horses he seems among the \textbf{Rode-goths}; the mane of that stallion ever shines." \\

(Vafþrúðnir kvað:) &
\bva Sęg þat, Gagnráðr, \hld alls á golfi vill &
þíns of fręista frama, &
hvé jór hęitir, \hld sá's austan dręgr &
nótt of nýt ręgin.\\ \&

\bvb Web-strengthener quoth: \\ “Say this, Gain-adviser, since on the floor I will to try thy fame: What is the horse called, which from the east drags night above the useful \textbf{Powers}?" \\

(Gagnráðr kvað:) &
\bva Hrímfaxi hęitir, \hld es hvęrja dręgr &
nótt of nýt ręgin; &
méldropa fęllir \hld morgin hvęrjan; &
þaðan kømr dǫgg of dala.\\ \&

\bvb Gain-adviser quoth: \\ “\textbf{Frost-fax} [that one] is called, who drags each night above the useful Powers. Every morning he lets foam fall from his bit\footnotemark[26]; thence comes dew in the valleys." \\
\footnotetext[26]{Lit. “he fells bit-drops".}

(Vafþrúðnir kvað:) &
\bva Sęg þat, Gagnráðr, \hld alls á golfi vill &
þíns of fręista frama, &
hvé ǫ́ hęitir, \hld sú's dęilir með jǫtna sonum &
grund ok með goðum.\\ \&

\bvb Web-strengthener quoth: \\ “Say this, Gain-adviser, since on the floor I will to try thy fame; How the river is called, which divides the ground between the sons of ettins and the gods?" \\

(Gagnráðr kvað:) &
\bva Ífing hęitir ǫ́, \hld es dęilir með jǫtna sonum &
grund ok með goðum; &
opin rinna \hld hón skal um aldrdaga; &
verðr-at íss á ǫ́.\\ \&

\bvb Gain-adviser quoth: \\ “\textbf{Iving} the river is called, which divides the ground between the sons of ettins and the gods. Throughout [her] life-days she shall flow open; ice forms not on the river." \\

(Vafþrúðnir kvað:) &
\bva Sęg þat, Gagnráðr, \hld alls á golfi vill &
þíns of fręista frama, &
hvé vǫllr hęitir, \hld es finnask vigi at &
Surtr ok hin svǫ́su goð.\\ \&

\bvb Web-strengthener quoth: \\ “Say this, Gain-adviser, since on the floor I will to try thy fame: How that valley is called, where \textbf{Surt} and the excellent gods find each other at war?" \\

(Gagnráðr kvað:) &
\bva Vígríðr hęitir vǫllr, \hld es finnask vígi at &
Surtr ok hin svǫ́su goð; &
hundrað rasta \hld hann's á hvęrjan veg; &
sá's þęim vǫllr vitaðr.\\ \&

\bvb Gain-adviser quoth: \\ “\textbf{Battle-rider} is the valley called, where Surt and the cheerful gods find each other at war. A hundred rests\footnotemark[30], he stretches in each direction; that valley is known for them.\footnotemark[31]" \\
\footnotetext[30]{An old unit of length, from its name prob. the length a horse could travel before resting.}
\footnotetext[31]{That is, known for its great size.}

(Vafþrúðnir kvað:) &
\bva Fróðr est nú gęstr, \hld far á bękk jǫtuns, &
mælumk í sessi saman; &
hǫfði vęðja \hld vit skulum hǫllu í &
gęstr, of gęðspęki.\\ \&

\bvb Web-strengthener quoth: \\ “Knowing art thou now, guest, sit down on the ettin's bench; let us speak while sitting together. In the hall we shall wager a head, guest, over [our] mind-wisdom." \\

(Gagnráðr kvað:) &
\bva Sęg þat hit ęina, \hld ef þitt ǿði\footnotemark[10] dugir &
ok þú Vafþrúðnir vitir, &
hvaðan jǫrð of kom \hld eða upphiminn &
fyrst, hinn fróði jǫtunn.\\ \&
\footnotetext[10]{Starting with \emph{ǿði}, the poem is also preserved in 748.}

\bvb Gain-adviser quoth: \\ “Say the first\footnotemark[32], if thy wisdom suffices, and thou, Web-strengthener, might know: Whence, O knowing ettin, the earth first came, or \textbf{up-heaven}?" \\
\footnotemark[32]{Lit. ‘one'.}

(Vafþrúðnir kvað:) &
\bva Ór Ymis holdi \hld vas jǫrð of skǫpuð, &
ęn ór bęinum bjǫrg, &
himinn ór hausi \hld hins hrimkalda jǫtuns, &
ęn ór svęita sær.\\ \&

\bvb Web-strengthener quoth: \\ “Out of \textbf{Yime's} hull\footnotemark[35], the earth was created, but the mountains out of his bones. Heaven out of the skull of the frost-cold ettin, but the sea out of his sweat.\footnotemark[36]" \\
\footnotetext[35]{His body.}
\footnotetext[36]{\emph{svęiti} ‘sweat' is a common kenning for blood. — This v. closely resembles \emph{Grím} 40.}

(Gagnráðr kvað:) &
\bva Sęg þat annat, \hld ef þitt ǿði dugir &
ok þú Vafþrúðnir vitir, &
hvaðan máni of kom, \hld svát fęrr menn yfir, &
eða sól hit sama.\\ \&

\bvb Gain-adviser quoth: \\ “Say the second, if thy wisdom suffices, and thou, Web-strengthener, might know: Whence the moon came, so that it travels over men, or likewise the sun?" \\

(Vafþrúðnir kvað:) &
\bva Mundilfari hęitir, \hld hann's Mána faðir &
ok svá Solar hit sama; &
himin hverfa \hld þau skulu hvęrjan dag &
ǫldum at ártali.\\ \&

\bvb Web-strengthener quoth: \\ “\textbf{Moundelfare} [that one] is called, he is the father of the Moon, and likewise of the Sun. They shall circle in the heavens every day, for men to reckon time\footnotemark[40]." \\
\footnotetext[40]{Lit. “for men to year-reckoning".}

(Gagnráðr kvað:) &
\bva Sęg þat þriðja, \hld alls þik svinnan kveða &
ok þú Vafþrúðnir vitir, &
hvaðan dagr of kom, \hld sá's fęrr drótt yfir, &
eða nótt með niðum.\\ \&

\bvb Gain-adviser quoth: \\ “Say the third, since [they] call thee wise, and thou, Web-strengthener, might know: Whence the day came, the one that travels over the rettinue, or night with the moon-phases?" \\

(Vafþrúðnir kvað:) &
\bva Dęllingr hęitir, \hld hann's Dags faðir, &
ęn Nótt vas Nǫrvi borin; &
ný ok nið \hld skópu nýt ręgin &
ǫldum at ártali.\\ \&

\bvb Web-strengthener quoth: \\ “\textbf{Delling} [that one] is called, he is the father of \textbf{Day}, but \textbf{Night} was born to \textbf{Nare}. The waxing and waning [of the moon], the useful Powers created, for men to reckon time." \\

(Gagnráðr kvað:) &
\bva Sęg þat fjórða, \hld alls þik fróðan kveða, &
ok þú Vafþrúðnir vitir, &
hvaðan vetr of kom \hld eða varmt sumar &
fyrst með fróð ręgin.\\ \&

\bvb Gain-adviser quoth: \\ “Say the fourth, since [they] call thee knowing, and thou, Web-strengthener, might know: Whence winter first came, or the warm summer, among the knowing Powers?" \\

(Vafþrúðnir kvað:) &
\bva Vindsvalr hęitir, \hld hann's Vetrar faðir, &
ęn Svǫ́suðr Sumars.\footnotemark[15]\\ \&
\footnotetext[15]{Second half of the v. seems missing.}

\bvb Web-strengthener quoth: \\ “\textbf{Wind-cool} [that one] is called, he is the father of \textbf{Winter}, but \textbf{Delightful} of \textbf{Summer}." \\

(Gagnráðr kvað:) &
\bva Sęg þat fimta, \hld alls þik fróðan kveða, &
ok þú Vafþrúðnir vitir, &
hvęrr ása ęlztr \hld eða Ymis niðja &
yrði í árdaga.\\ \&

\bvb Gain-adviser quoth: \\ “Say the fifth, since [they] call thee knowing, and thou, Web-strengthener, might know: Who, in days of yore became the eldest of the \textbf{Anses}, or of the descendants of Yime?" \\

(Vafþrúðnir kvað:) &
\bva Ørófi vetra \hld áðr væri jǫrð skǫpuð, &
þá vas Bergęlmir borinn, &
Þrúðgęlmir \hld vas þess faðir, &
ęn Aurgęlmir afi.\\ \&

\bvb Web-strengthener quoth: \\ “Uncountable winters before the earth would be created, then \textbf{Bear-yeller} was born. \textbf{Strength-yeller} was \emph{that one's} father, and \textbf{Mud-yeller} the grandfather." \\

(Gagnráðr kvað:) &
\bva Sęg þat sétta, \hld alls þik svinnan kveða, &
ok þú Vafþrúðnir vitir, &
hvaðan Aurgęlmir kom \hld með jǫtna sonum &
fyrst, hinn fróði jǫtunn.\\ \&

\bvb Gain-adviser quoth: \\ “Say the sixth, since [they] call thee wise, and thou, Web-strengthener, might know: Whence, O knowing ettin, Mud-yeller first came among the sons of ettins?" \\

(Vafþrúðnir kvað:) &
\bva Ór Élivǫ́gum \hld stukku ęitrdropar, &
svá óx unz ór varð jǫtunn; &
órar ættir \hld kómu þar allar saman; &
því's þat æ alt til atalt.\footnotemark[20]\\ \&
\footnotetext[20]{Lines 3–4 missing in R and 748, but quoted in \emph{Gylf}.}

\bvb Web-strengthener quoth: \\ “From the \textbf{Ell-waves}, poison-drops splashed; thus [it] grew until an ettin emerged. \emph{Our} family lines all together originated there, therefore our race\footnotemark[45] is forever fierce against all.\footnotemark[46]" \\
\footnotetext[45]{Lit. ‘it' or ‘that'.}
\footnotetext[46]{Somewhat strange phrasing, but the line does not appear damaged. It is clearly an explanation of the fierce and maleficent nature of the ettins, as their first ancestors were created from poison.}

(Gagnráðr kvað:) &
\bva Sęg þat sjaunda, \hld alls þik svinnan kveða, &
ok þú Vafþrúðnir vitir, &
hvé sá bǫrn gat \hld hinn baldni\footnotemark[25] jǫtunn, &
es hann hafði-t gýgjar gaman.\\ \&
\footnotetext[25]{R has \emph{aldni}, ‘aged, old'. This breaks alliteration; \emph{baldni} ‘bold, defiant' has been substituted from 748.}

\bvb Gain-adviser quoth: \\ “Say the seventh, since [they] call thee wise, and thou, Web-strengthener, might know: How did that one, the defiant ettin, beget children, when he did not enjoy the [carnal] pleasure of a troll-woman?" \\

(Vafþrúðnir kvað:) &
\bva Und hęndi vaxa \hld kvǫ́ðu hrímþursi &
męy ok mǫg saman; &
fótr við fǿti \hld gat hins fróða jǫtuns &
sexhǫfðaðan son.\\ \&

\bvb Web-strengthener quoth: \\ “Neath the hand\footnotemark[50] on the \textbf{frost-thurse}, [they] said that a girl and boy together grew. A foot against a foot begot, for the knowing ettin, a six-headed son." \\
\footnotetext[50]{\hęndi (dative of \emph{hǫnd}) means ‘hand', but might here be a poetic circumlocution for ‘arm'.}

(Gagnráðr kvað:) &
\bva Sęg þat áttunda, \hld alls þik fróðan kveða, &
ok þú Vafþrúðnir vitir, &
hvat fyrst of mant \hld eða fręmst of vęizt, &
þú est alsviðr jǫtunn.\\ \&

\bvb Gain-adviser quoth: \\ “Say the eigth, since [they] call thee knowing, and thou, Web-strengthener, might know: What dost thou first remember, or earliest know?\footnotemark[55] Thou art all-wise, ettin." \\
\footnotetext[55]{Cf. Vsp 1.}

(Vafþrúðnir kvað:) &
\bva Ørófi vetra \hld áðr væri jǫrð of skǫpuð, &
þá vas Bergęlmir borinn; &
þat fyrst um man'k, \hld es hinn fróði jǫtunn &
á vas lúðr of lagiðr.\footnotemark[30]\\ \&
\footnotetext[30]{This verse is quoted in \emph{Gylf}.}

\bvb Web-strengthener quoth: \\ “Uncountable winters before the earth would be created, then Bear-yeller was born. \emph{That} I first remember, when the knowing ettin\footnotemark[60] was laid down on the funeral-bed\footnotemark[61]." \\
\footnotetext[60]{That is, Bear-yeller. Cf. v. 29.}
\footnotetext[61]{\emph{lúðr}, a tricky word.}

(Gagnráðr kvað:) &
\bva Sęg þat níunda, \hld alls þik svinnan kveða, &
ok þú Vafþrúðnir vitir, &
hvaðan vindr of kømr \hld svát fęrr vág yfir, &
æ męnn hann sjalfan of séa.\\ \&

\bvb Gain-adviser quoth: \\ “Say the ninth, since [they] call thee wise, and thou, Web-strengthener, might know: Whence the wind comes, so that he travels over the wave; forever men see him himself.\footnotemark[65]" \\
\footnotetext[65]{Perhaps a negation has been lost here; the wind is never seen by men.}

(Vafþrúðnir kvað:) &
\bva Hræsvęlgr hęitir, \hld es sitr á himins ęnda, &
jǫtunn í arnar ham; &
af hans vængjum \hld kveða vind koma &
alla męnn yfir.\\ \&

\bvb Web-strengthener quoth: \\ “\textbf{Corpse-swallower} [that one] is called, which sits at the end of the heavens, an ettin in the shape of an eagle. From his wings, they say [that] the wind comes over all men." \\

(Gagnráðr kvað:) &
\bva Sęg þat tíunda, \hld alls þú tíva rǫk &
ǫll Vafþrúðnir vitir, &
hvaðan Njǫrðr of kom \hld með niðjum ása. &
Hófum ok hǫrgum \hld hann ræðr hundmǫrgum &
ok varð-at hann ǫ́sum alinn.\\ \&

\bvb Gain-adviser quoth: \\ “Say the \emph{tenth}, since thou, Web-strengthener, of all the fates of the \textbf{Tues} might know: Whence \textbf{Nearth} came into the company of the kinsmen of the \textbf{Anses}? He rules an immense number\footnotemark[68] of \textbf{hoves} and \textbf{heargs}, and he was not begotten among the Anses." \\
\footnotetext[68]{Lit. “he rules hound-many".}

(Vafþrúðnir kvað:) &
\bva Í Vanahęimi \hld skópu hann vís ręgin &
ok sęldu at gíslingu goðum, &
í aldar rǫk \hld hann mun aptr koma &
hęim með vísum vǫnum.\\ \&

\bvb Web-strengthener quoth: \\ “In \textbf{Wane-Home}, the wise \textbf{Powers}\footnotemark[69] created him, and gave\footnotemark[70] him as a hostage to the gods. In the fate of the age, he will come back, home among the wise \textbf{Wanes}." \\
\footnotetext[69]{The gods, or specifically the Wanes.}
\footnotetext[70]{Lit. “sold".}

(Gagnráðr kvað:) &
\bva Sęg þat ęllipta, \hld hvar ýtar túnum í &
hǫggvask hvęrjan dag; &
val þęir kjósa \hld ok ríða vígi frá, &
sitja męir of sáttir saman.\footnotemark[35]\\ \&
\footnotetext[35]{This and the next v. are damaged in both R and 748; R has only this verse, but splits it in two (the 2nd starting with \emph{val}), while 748 has 40:1 (Ms.: \emph{S. þ. e. XI}) and then jumps to the answer v. 41. They have here been reconstructed, but it is possible some lines are still missing.}

\bvb Gain-adviser quoth: \\ “Say the eleventh, Where men in yards, hew away at each other every day? They choose those destined to die in war, and ride [away] from battle; [then] they sit more content together." \\

(Vafþrúðnir kvað:) &
\bva Allir ęinhęrjar \hld Óðins túnum í &
hǫggvask hvęrjan dag, &
val þeir kjósa \hld ok ríða vígi frá, &
sitja męir of sáttir saman.\\ \&

\bvb Web-strengthener quoth: \\ “In Woden's yards, all the \textbf{Lone Warriors} hew away at each other every day. They choose those destined to die in war, and ride [away] from battle; [then] they sit more content together." \\

(Gagnráðr kvað:) &
\bva Sęg þat tolpta, \hld hví þú tíva rǫk &
ǫll Vafþrúðnir vitir, &
frá jǫtna rúnum \hld ok allra goða &
þú hit sannasta sęgir, &
hinn alsvinni jǫtunn.\\ \&

\bvb Gain-adviser quoth: \\ “Say the twelfth, Why thou, Web-strengthener, should know all the fates of the \textbf{Tues}\footnotemark[73]? Regarding the \textbf{runes} of the ettins and of all the gods, thou, the all-wise ettin, speaks most truly." \\
\footnotetext[73]{The gods. Formation identical to \emph{ragna rǫk} ‘the fates of the Powers'.}

(Vafþrúðnir kvað:) &
\bva Frá jǫtna rúnum \hld ok allra goða &
ek kann sęgja satt, &
þvíat hvęrn hęf'k \hld heim of komit, &
níu kom'k hęima \hld fyr niflhęl neðan; &
hinig dęyja ór hęlju halir.\\ \&

\bvb Web-strengthener quoth: \\ “Regarding the runes of the ettins and of all the gods, I can speak truly, for I have come into each \textbf{Home}. I came [into] \textbf{nine Homes} beneath \textbf{Fog-hell}; thereto men die out of \textbf{Hell}\footnotemark[75]." \\
\footnotetext[75]{A difficult verse. Does it imply that if a man dies when already dead, he goes to Fog-hell?}

(Gagnráðr kvað:) &
\bva Fjǫlð ek fór, \hld fjǫlð fręistaða'k, &
fjǫlð ek ręynda ręgin; &
hvat lifir manna, \hld þá's hinn mæra líðr &
fimbulvetr með firum?\\ \&

\bvb Gain-adviser quoth: \\ “Much I travelled, much I tried, much I tested the \textbf{Powers}.\footnotemark[80] What remains\footnotemark[79] of men, when the famous \textbf{fimbol-winter} passes among firs\footnotemark[81]?" \\
\footnotetext[79]{Lit. “lives".}
\footnotetext[80]{Here begins the repetition of the same “mantra" used in v. 3, which continues until the final question (v. 54).}
\footnotetext[81]{Among men.}

(Vafþrúðnir kvað:) &
\bva Líf ok Lífþrasir, \hld ęn þau lęynask munu &
í holti Hoddmímis; &
morgindǫggvar \hld þau sér at mat hafa; &
þaðan af aldir alask.\\ \&

\bvb Web-strengthener quoth: \\ “\textbf{Life} and \textbf{Life-striver}, and they will hide themselves in the woods of \textbf{Hoard-Mime}\footnotemark[85]. Morning-dew they [will] have as food; thence generations will be bred." \\
\footnotetext[85]{Prob. the same as \emph{Yggdrasill}, the \textbf{Terrifier-Steed.}}

(Gagnráðr kvað:) &
\bva Fjǫlð ek fór, \hld fjǫlð fręistaða'k, &
fjǫlð ek ręynda ręgin; &
hvaðan kømr sól \hld á hinn slétta himin, &
es þessa hęfr Fęnrir farit?\\ \&

\bvb Gain-adviser quoth: \\ “Much I travelled, much I tried, much I tested the Powers. Whence comes sun onto the smooth heaven, when \textbf{Fenner} has killed this one\footnotemark[90]?" \\
\footnotetext[90]{That is, the current incarnation of the sun, as explained in the next v.}

(Vafþrúðnir kvað:) &
\bva Ęina dóttur \hld berr alfrǫðull, &
áðr hana Fęnrir fari; &
sú skal ríða, \hld þá's ręgin dęyja, &
móður brautir mær.\\ \&

\bvb Web-strengthener quoth: \\ “One daughter the elf-wheel\footnotemark[95] bears, before \textbf{Fenner} might kill her. When the Powers die, that one, the girl, shall ride the paths of the mother." \\
\footnotetext[95]{The sun.}

(Gagnráðr kvað:) &
\bva Fjǫlð ek fór, \hld fjǫlð fręistaða'k, &
fjǫlð ek ręynda ręgin; &
hvęrjar 'ro męyjar, \hld es líða mar yfir, &
fróðgęðjaðar fara.\\ \&

\bvb Woden quoth: \\ “Much I travelled, much I tried, much I tested the Powers. Which are the girls that pass over the ocean; wise-minded they go?" \\

(Vafþrúðnir kvað:) &
\bva Þríar þjóðár \hld falla þorp yfir &
męyja Mǫgþrasis; &
hamingjur ęinar \hld þær's í hęimi eru, &
þó þær með jǫtnum alask.\\ \&

\bvb Web-strengthener quoth: \\ “Three great rivers fall over the settlement of the girls of \textbf{Boy-striver}; they are the only \textbf{Hamings} in the Home,\footnotemark[99] though they are raised among the ettins\footnotemark[100]." \\
\footnotetext[99]{In Ettin-Home, or in the entire world?}
\footnotetext[100]{An opaque verse. These three girls may be the same as the three \emph{thurse-maidens} in \emph{Vsp} 9, see note there. Their father, Boy-striver, is not mentioned anywhere else.}

(Gagnráðr kvað:) &
\bva Fjǫlð ek fór, \hld fjǫlð fręistaða'k, &
fjǫlð ek ręynda ręgin; &
hvęrir ráða æsir \hld ęignum goða, &
þá's sloknar Surtalogi?\\ \&

\bvb Gain-adviser quoth: \\ “Much I travelled, much I tried, much I tested the Powers. Which properties of the gods will the \textbf{Anses} [still] rule\footnotemark[105], when the flame of \textbf{Surt} burns out?" \\
\footnotetext[105]{Or ‘control'.}

(Vafþrúðnir kvað:) &
\bva Víðarr ok Váli \hld byggva vé goða, &
þá's sloknar Surtalogi; &
Móði ok Magni \hld skulu Mjǫlni hafa &
Vingnis at vígþroti.\\ \&

\bvb Web-strengthener quoth: \\ “\textbf{Wider} and \textbf{Weel} [will] inhabit the sanctuaries of the gods, when the \textbf{flame of Surt} burns out. \textbf{Mood} and \textbf{Main} will own \textbf{Mealen}, when \textbf{Wingen} can no longer fight\footnotemark[110]." \\
\footnotetext[110]{Lit. “at Wingen's fight-exhaustion", referring to his death.}

(Gagnráðr kvað:) &
\bva Fjǫlð ek fór, \hld fjǫlð fręistaða'k, &
fjǫlð ek ręynda ręgin; &
hvat verðr Óðni \hld at aldrlagi, &
þá's rjúfask ręgin?\\ \&

\bvb Gain-adviser quoth: \\ “Much I travelled, much I tried, much I tested the Powers. What happens to Woden at the end of the age, when the Powers are broken?" \\

(Vafþrúðnir kvað:) &
\bva Ulfr glęypa \hld mun Aldafǫðr, &
þess mun Víðarr vreka; &
kalda kjapta \hld hann klyfja mun &
vitnis vígi at.\\ \&

\bvb Web-strengthener quoth: \\ “The wolf will swallow the \textbf{Leader of Men}; Wider will avenge that. He will cleave the cold jaws of the wolf at the battle." \\

(Gagnráðr kvað:) &
\bva Fjǫlð ek fór, \hld fjǫlð fręistaða'k, &
fjǫlð ek ręynda ręgin; &
hvat mælti Óðinn, \hld áðr á bál stigi, &
sjalfr í ęyra syni?\\ \&

\bvb Gain-adviser quoth: \\ “Much I travelled, much I tried, much I tested the Powers. What spoke Woden himself, before [he]\footnotemark[115] would step onto the funeral pyre, into the ear of the son?" \\
\footnotetext[115]{Prob. Woden's son, that is \textbf{Balder}.}

(Vafþrúðnir kvað:) &
\bva Ęy mann-gi\footnotemark[40] vęit, \hld hvat þú í árdaga &
sagðir í ęyra syni; &
fęigum munni \hld mælta'k mína forna stafi &
ok of ragna rǫk. &
Nú við Óðin \hld dęilda'k mína orðspęki; &
þú est æ vísastr vera.\\ \&
\footnotetext[40]{Emended from R, 748 \emph{manni}; the word must be in the nominative, but \emph{manni} is dative.}

\bvb Web-strengthener quoth: \\ “No man may ever know\footnotemark[119], what thou in days of yore said into the ear of the son. With a death-doomed\footnotemark[120] mouth, I spoke my ancient staves, and about the \textbf{fates of the Powers}\footnotemark[121]. Now against Woden, I have shared my word-wisdom\footnotemark[122]; thou art forever the wisest of men\footnotemark[123]." \\
\footnotetext[119]{Lit. “eternally no man knows".}
\footnotetext[120]{Or “\emph{fey} mouth". Web-strengthener here realizes that he was bound to die from the moment (v. 19) he proposed the wager; no man, nor god, nor ettin can outwit Woden.}
\footnotetext[121]{“fates of the Powers", that is, \emph{ragna rǫk}.}
\footnotetext[122]{The same word-wisdom Woden in v. 5 set out to try.}
\footnotetext[123]{Word used is \emph{ver} ‘husband', ‘man'. Perhaps in the broader sense of ‘male being'.}
