\book{Speeches of the High One. (\emph{Hávamǫ́l})}\bookStart

Introduction.
\small{\emph{Hávamǫ́l}, or “the Speeches of the High One”, is the second poem of \Regius.}

\bvg Advice to wanderers.
\bva \alst{G}áttir allar \hld áðr \alst{g}angi framm &
\ind \edtext{of \alst{sk}oðask \alst{sk}yli}{\lemma{of skoðask skyli}\Bfootnote{\emph{om.} \GylfMS}} &
\ind of \alst{sk}ygnask \alst{sk}yli; &
því’t ó\alst{v}íst's at \alst{v}ita, \hld hvar ó\alst{v}inir &
\ind sitja á \alst{f}lęti \alst{f}yrir.\eva

\bvb All doorways—before one might go forth—should be watched, should be spied at; for uncertain ’tis to know, where enemies sit on the benches inside.\evb
\evg


\bvg
\bva \alst{G}efęndr hęilir, \hld \alst{g}ęstr’s inn kominn, &
\ind hvar skal \alst{s}itja \alst{s}já? &
mjǫk es \alst{b}ráðr \hld sá's á \alst{b}rǫndum skal &
\ind síns of \alst{f}ręista \alst{f}rama.\eva

\bvb Hail the givers\footnoteB{The hosts.}! A guest is come in, where shall that one sit? Greatly hurried is he, who on the fires shall try his fame.\footnoteB{According to \Finnur\ referring a Norwegian folk custom, wherein a guest would sit down on the wood-pile, waiting until being called in. See further TODO.}\evb
\evg


\bvg
\bva \alst{Ę}lds es þǫrf \hld þęim's \alst{i}nn es kominn &
\ind ok á \alst{k}néi \alst{k}alinn, &
\alst{m}atar ok váða \hld es \alst{m}anni þǫrf, &
\ind þęim's hęfr of \alst{f}jall \alst{f}arit.\eva

\bvb Of fire is need, for the one who inside is come, and cold about the knees; of food and clothing is need, for the man who over the fell has fared.\evb
\evg


\bvg
\bva \alst{V}ats es þǫrf \hld þęim's til \alst{v}erðar kømr, &
\ind \alst{þ}ęrru ok \alst{þ}jóðlaðar, &
\alst{g}óðs of ǿðis, \hld —ef sér \alst{g}eta mætti— &
\ind \alst{o}rðs ok \alst{ę}ndrþǫgu.\eva

\bvb Of water is need, for the one who comes for a meal, a towel and a friendly welcome; a good reception—if he might get one—speech, and silence in return.\evb
\evg


\bvg
\bva \alst{V}its es þǫrf \hld þęim's \alst{v}íða ratar; &
\ind dælt es \alst{h}ęima \alst{h}vat; &
at \alst{au}gabragði \hld verðr sá's \alst{ę}kki kann &
\ind ok með \alst{s}notrum \alst{s}itr.\eva

\bvb Of wits is need, for the one who roams widely; all is familiar at home. A laughing stock\Bfootnote{An idiom, \emph{augabragð} lit. ‘twinkling of the eye, moment’.} becomes he who nothing knows, and among the clever sits.\evb
\evg


\bvg
\bva At \alst{h}yggjandi sinni \hld skyli-t maðr \alst{h}rǿsinn vesa, &
\ind hęldr \alst{g}ætinn at \alst{g}ęði, &
þá’s \alst{h}orskr ok þǫgull \hld kømr \alst{h}ęimisgarða til, &
\ind sjaldan verðr \alst{v}íti \alst{v}ǫrum. &
\edtext{því’t óbrigðra vin \hld fær þú aldrigi, &
\ind an \alst{m}anvit \alst{m}ikit.}{\lemma{því ... mikit}\Bfootnote{The shift in person from 3rd to 2nd, along with the abnormal verse length (6 lines instead of 4), strongly suggests that this is an insert.}}\eva

\bvb Of his thinking should man not be boastful; rather guarding of his senses, when sharp and silent, he comes to a homestead; sudden injury seldom strikes the wary. (For thou getst never a more steadfast\footnoteB{Lit. ‘more unfickle’.} friend, than much \inx{manwit}{C}.)\evb
\evg


\bvg
\bva Hinn \alst{v}ari gęstr, \hld es til \alst{v}erðar kømr, &
\ind \alst{þ}unnu hljóði \alst{þ}ęgir; &
\alst{ęy}rum hlýðir, \hld ęn \alst{au}gum skoðar, &
\ind svá nýsisk \alst{f}róðra hvęrr \alst{f}yrir.\eva

\bvb The wary guest, when he comes to a host, is in attentive silence\footnotemark[13]. With ears he listens, but with eyes observes; thus every experienced man looks out in advance.\evb
\footnotetext[13]{Lit. 'with thin heed is silent'}
\evg


\bvg
\bva Hinn es \alst{s}æll, \hld es \alst{s}ér of getr &
\ind \alst{l}of ok \alst{l}íknstafi; &
\alst{ó}dælla es við þat, \hld es \alst{ęi}ga skal &
\ind \alst{a}nnars brjóstum \alst{í}.\eva

\bvb The one is fortunate, who gets praise and high esteem for \emph{himself}. It is not easy regarding that, which [he] must own in another's breast.\evb
\evg


\bvg
\bva \alst{S}á es \alst{s}æll, \hld es \alst{s}jalfr of á &
\ind \alst{l}of ok vit meðan \alst{l}ifir; &
því’t \alst{i}ll rǫ́ð \hld hęfr maðr \alst{o}pt þęgit &
\ind \alst{a}nnars brjóstum \alst{ó}r.\eva

\bvb That one is fortunate, who \emph{himself} owns praise and wits while he lives, for man has oft received ill counsel from another's breast.\evb
\evg


\bvg
\bva \alst{B}yrði \alst{b}ętri \hld berr-at maðr \alst{b}rautu at, &
\ind an sé \alst{m}anvit \alst{m}ikit; &
\alst{au}ði bętra \hld þykkir þat í \alst{ó}kunnum stað; &
\ind slíkt es \alst{v}álaðs \alst{v}era.\eva

\bvb No man carries a better burden on the road than much manwit; in a foreign place, it seems better than wealth; such is the refuge of the wretched.\evb
\evg


\bvg
\bva \alst{B}yrði \alst{b}ętri \hld bęrr-at maðr \alst{b}rautu at, &
\ind an sé \alst{m}anvit \alst{m}ikit; &
\alst{v}egnest \alst{v}erra \hld \alst{v}egr-a \alst{v}ęlli at, &
\ind an sé \alst{o}fdrykkja \alst{ǫ}ls.\eva

\bvb No man carries a better burden on the road than much manwit; on the field he can drag no worse provision along than too much ale drunk.\evb
\evg


\bvg
\bva Es-a svá \alst{g}ótt, \hld sęm \alst{g}ótt kveða, &
\ind \alst{ǫ}l \alst{a}lda sonum; &
því’t \alst{f}æra vęit, \hld es \alst{f}lęira drekkr, &
\ind síns til \alst{g}ęðs \alst{g}umi.\eva

\bvb Is not so good, as good they call it, ale for the sons of men. For less he knows, as more he drinks, man of his own senses.\evb
\evg


\bvg
\bva \alst{Ó}minnishegri hęitir, \hld sá’s of \alst{ǫ}lðrum þrumir, &
\ind hann stelr \alst{g}ęði \alst{g}uma; &
þess \alst{f}ogls \alst{f}jǫðrum \hld ek \alst{f}jǫtraðr vask &
\ind í \alst{g}arði \alst{G}unnlaðar.\eva

\bvb A heron of forgetfulness is called he who loiters about the ale; he robs men of their senses. With that bird’s feathers I fettered was, in the estate of Guthlath.\evb
\evg


\bvg
\bva \alst{Ǫ}lr ek varð, \hld varð \alst{o}frǫlvi, &
\ind at hins \alst{f}róða \alst{F}jalars; &
því es \alst{ǫ}lðr bazt, \hld at \alst{a}ptr of hęimtir &
\ind hvęrr sitt \alst{g}ęð \alst{g}umi.\eva

\bvb Drunk I became, I became the far drunkest, at the knowing Fealer’s. Thus with drinking it is best, that each man come back to his senses.\evb
\evg


\bvg
\bva \alst{Þ}agalt ok hugalt \hld skyli \alst{þ}jóðans barn &
\ind ok \alst{v}ígdjarft \alst{v}esa; &
\alst{g}laðr ok ręifr \hld skyli \alst{g}umna hvęrr, &
\ind unz sinn \alst{b}íðr \alst{b}ana.\eva

\bvb Taciturn and thoughtful should the ruler’s child be, and bold in battle. Glad and happy, should each man be, until he suffer his bane.\evb
\evg


\bvg
\bva \alst{Ó}snjallr maðr \hld hyggsk munu \alst{ę}y lifa, &
\ind ef við \alst{v}íg \alst{v}arask; &
ęn \alst{ę}lli gefr \hld hǫ́num \alst{ę}ngi frið, &
\ind þótt hǫ́num \alst{g}ęirar \alst{g}efi.\eva

\bvb The unvalorous man thinks he will ever live, if he of conflict is wary; but old age gives him no peace, although spears would give him.\evb
\evg


\bvg
\bva \alst{K}ópir afglapi, \hld es til \alst{k}ynnis \alst{k}ømr, &
\ind \alst{þ}ylsk hann umb eða \alst{þ}rumir; &
alt es \alst{s}ęnn, \hld ef \alst{s}ylg of getr, &
\ind uppi es þá \alst{g}ęð \alst{g}uma.\eva

\bvb The oaf gapes, when to a visit he comes; he mumbles about or loiters; all at once—if a sip he gets—are the senses of the man exposed.\evb
\evg


\bvg
\bva Sá ęinn \alst{v}ęit, \hld es \alst{v}íða ratar &
\ind ok hęfr \alst{f}jǫlð of \alst{f}arit, &
hvęrju \alst{g}ęði \hld stýrir \alst{g}umna hvęrr, &
\ind sá es \alst{v}itandi’s \alst{v}its.\eva

\bvb He alone knows, who widely wanders, and has travelled much: his senses does each man control, who is aware of his wits.\evb
\evg


\bvg
\bva \alst{H}aldi-t maðr á kęri, \hld drekki þó at \alst{h}ófi mjǫð, &
\ind mæli \alst{þ}arft eða \alst{þ}ęgi; &
ókynnis \alst{þ}ess \hld váar \alst{þ}ik ęngi maðr, &
\ind at gangir \alst{s}nimma at \alst{s}ofa.\eva

\bvb A man ought not hold onto the cask, rather drink a fitting serving of mead; he ought to speak what is needed or be silent\footnoteB{Identical to a certain verse in \Vafthrudnismal\ TODO: which one}. For that uncouthness will no man blame thee, that thou go early to sleep.\evb
\evg


\bvg
\bva \alst{G}rǫ́ðugr halr, \hld nema \alst{g}ęðs viti, &
\ind \alst{e}tr sér \alst{a}ldrtrega; &
opt fær \alst{h}lǿgis, \hld es með \alst{h}orskum kømr, &
\ind \alst{m}anni hęimskum \alst{m}agi.\eva

\bvb The gluttonous man, unless he know his sense, eats himself a life-sorrow; often the belly, when among sharp ones he comes, brings a foolish man ridicule.\evb
\evg


\bvg
\bva \alst{H}jarðir þat vitu, \hld nær \alst{h}ęim skulu, &
\ind ok \alst{g}anga þá af \alst{g}rasi; &
ęn \alst{ó}sviðr maðr \hld kann \alst{æ}vagi &
\ind síns of \alst{m}ál \alst{m}aga.\eva

\bvb Herds know, when home they must, and then part from the grass; but an unwise man never knows the measure of his belly.\evb
\evg


\bvg
\bva \alst{V}esall maðr \hld ok \alst{i}lla skapi &
\ind \alst{h}lær at \alst{h}vívetna; &
hitki hann \alst{v}ęit, \hld es \alst{v}ita þyrpti, &
\ind at hann es-a \alst{v}amma \alst{v}anr.\eva

\bvb The wretched man, and ill-spirited, laughs at whatever; he knows it not, which he needs to know: that he is not free of blemishes.\evb
\evg


\bvg
\bva \alst{Ó}sviðr maðr \hld vakir of \alst{a}llar nætr &
\ind ok \alst{h}yggr at \alst{h}vívetna; &
þá es \alst{m}óðr, \hld es at \alst{m}orni kømr; &
\ind alt es \alst{v}íl sęm \alst{v}as.\eva

\bvb The unwise man wakes for all nights, and thinks of whatever; then he is weary when the morning comes: his trouble is all as it was.\evb
\evg


\bvg
\bva \alst{Ó}snotr maðr \hld hyggr sér \alst{a}lla vesa &
\ind \alst{v}iðhlæjęndr \alst{v}ini; &
hitki hann \alst{f}iðr, \hld þótt þęir of hann \alst{f}ár lesi, &
\ind ef með \alst{s}notrum \alst{s}itr.\eva

\bvb The unclever man thinks all who laugh at him his friends; it he notices not, though they speak poorly about him, if he sits among the clever.\evb
\evg


\bvg
\bva \alst{Ó}snotr maðr \hld hyggr sér \alst{a}lla vesa &
\ind \alst{v}iðhlæjendr \alst{v}ini; &
\alst{þ}á þat fiðr \hld es at \alst{þ}ingi kømr, &
\ind at á \alst{f}ormælęndr \alst{f}áa.\eva

\bvb The unclever man thinks all who laugh at him his friends; then he finds, when to the \inx{Thing}{C} he comes, that he has spokesmen\Bfootnote{Men ready to take his side.} few.\evb
\evg


\bvg
\bva \alst{Ó}snotr maðr \hld þykkisk \alst{a}lt vita, &
\ind ef á sér i \alst{v}ǫ́ \alst{v}eru; &
hitki hann \alst{v}ęit, \hld hvat hann skal \alst{v}ið kveða, &
\ind ef hans \alst{f}ręista \alst{f}irar.\eva

\bvb The unclever man seems to know everything, if he shelters himself in a nook; it he knows not, what he shall say in return, if men test him.\evb
\evg


\bvg
\bva \alst{Ó}snotr maðr, \hld es með \alst{a}ldir kømr, &
\ind \alst{þ}at’s bazt at hann \alst{þ}ęgi; &
\alst{ę}ngi þat vęit, \hld at hann \alst{ę}kki kann, &
\ind nema hann \alst{m}æli til \alst{m}art. &
\alst{v}ęit-a maðr, \hld hinn’s \alst{v}ætki vęit, &
\ind þótt hann \alst{m}æli til \alst{m}art.\eva

\bvb The unclever man, when among people he comes, it is best that he be silent. None knows that he nothing knows, unless he speak too much. (A man knows not, who nothing knows, although he speak too much.)\evb
\evg


\bvg
\bva \alst{F}róðr sá þykkisk, \hld es \alst{f}regna kann, &
\ind ok \alst{s}ęgja hit \alst{s}ama, &
\alst{ęy}vitu lęyna \hld męgu \alst{ý}ta synir &
\ind því es \alst{g}ęngr of \alst{g}uma.\eva

\bvb Learned seems he who can ask, and answer the same; naught conceal may the sons of men, of that\Bfootnote{Rumours and gossip.} which goes about a man.\evb
\evg


\bvg
\bva \alst{Ǿ}rna mælir, \hld sá’s \alst{æ}va þęgir, &
\ind \alst{st}aðlausu \alst{st}afi; &
\alst{h}raðmælt tunga, \hld nema \alst{h}aldęndr ęigi, &
\ind opt sér ó\alst{g}ótt of \alst{g}ęlr.\eva

\bvb Quite enough does he speak, who is never silent, utterings of absurdity; a quick-spoken tongue—unless it be held in place\Bfootnote{Literally ‘unless holders own it’ or ‘unless it own holders’.}—often sings evil [into existence] for itself.\evb
\evg


\bvg
\bva At \alst{au}gabragði \hld skal-a maðr \alst{a}nnan hafa, &
\ind \edtext{þótt}{\lemma{þótt “although”}\Bfootnote{\emph{Perhaps an error?} es \emph{‘when’ would surely work better in context.}} til \alst{k}ynnis \alst{k}omi; &
margr \alst{f}róðr þykkisk, \hld ef hann \alst{f}reginn es-at &
\ind ok nái hann \alst{þ}urrfjallr \alst{þ}ruma.\eva

\bvb As a laughing stock shall man not have another, although he come to visit; many a man seems learned if he is not asked, and manages to loiter about dry-skinned\footnoteB{This sense of \emph{fjall} is apparently almost non-existent in Old Norse literature, but compare Swedish \emph{fjäll} ‘scale (on fish and reptiles)’. The meaning is in any case figurative; compare the English saying \emph{get your boots wet}.}.\evb
\evg


\bvg
\bva \alst{F}róðr þykkisk \hld sá’s \alst{f}lótta tękr &
\ind \alst{g}ęstr at \alst{g}ęst hæðinn; &
\alst{v}ęit-a gǫrla \hld sá’s of \alst{v}erði glissir, &
\ind þótt með \alst{g}rǫmum \alst{g}lami.\eva

\bvb Learned seems he, who takes to flight, when a guest at a guest is scoffing. He knows not fully, who grins above the food, that he with fiends be prattling.\evb
\evg


\bvg
\bva \alst{G}umnar margir \hld erusk \alst{g}agnhollir, &
\ind ęn at \alst{v}irði \alst{v}rekask; &
\alst{a}ldar róg \hld þat mun \alst{æ} vesa; &
\ind órir \alst{g}ęstr við \alst{g}ęst.\eva

\bvb Many men are loyal to each other, but over a meal drive each other away. The strife of mankind will that ever be; guest raves against guest.\evb
\evg


\bvg
\bva \alst{Á}rliga verðar \hld skyli maðr \alst{o}pt fáa, &
\ind nema til \alst{k}ynnis \alst{k}omi; &
\alst{s}itr ok \alst{s}nópir, \hld lætr sęm \alst{s}olginn sé, &
\ind ok kann \alst{f}regna at \alst{f}ǫ́u.\eva

\bvb An early meal should man often get, unless he is come to visit; he sits and idles haplessly about, makes as if he is starved, and knows to ask about little.\evb
\evg


\bvg
\bva \alst{A}fhvarf mikit \hld es til \alst{i}lls vinar, &
\ind þótt á \alst{b}rautu \alst{b}úi, &
ęn til \alst{g}óðs vinar \hld liggja \alst{g}agnvegir, &
\ind þótt hann sé \alst{f}irr \alst{f}arinn.\eva

\bvb A great detour is it to an ill friend, though on the highway he live; but to a good friend lie the shortest ways, though he be far gone.\evb
\evg


\bvg
\bva \alst{G}anga skal, \hld skal-a \alst{g}ęstr vesa &
\ind \alst{ęy} í \alst{ęi}num stað; &
\alst{l}júfr verðr \alst{l}ęiðr, \hld ef \alst{l}ęngi sitr &
\ind \alst{a}nnars flętjum \alst{á}.\eva

\bvb One shall go, he shall not be a guest ever in one place; lovely becomes the path, if long one sits on another’s bench.\evb
\evg


\bvg
\bva \alst{B}ú es \alst{b}ętra, \hld þótt lítit sé, &
\ind \alst{h}alr es \alst{h}ęima \alst{h}vęrr; &
þótt \alst{t}vær gęitr ęigi \hld ok \alst{t}augręptan sal, &
\ind þat es þó \alst{b}ętra an \alst{b}ǿn.\eva

\bvb A home is better, though little it be: each is a man at home; though two goats he own, and a cord-roofed hall, that is yet better than begging.\evb
\evg


\bvg
\bva \alst{B}ú es \alst{b}ętra, \hld þótt lítit sé, &
\ind \alst{h}alr es \alst{h}ęima \alst{h}vęrr; &
\alst{b}lóðugt es hjarta \hld þęim's \alst{b}iðja skal &
\ind sér í \alst{m}ál hvęrt \alst{m}atar.\eva

\bvb A home is better, though little it be: each is a man at home; bloody is the heart of the one who must beg for each meal of food.\evb
\evg


\bvg
\bva \alst{V}ǫ́pnum sínum \hld skal-a maðr \alst{v}ęlli á &
\ind \alst{f}eti ganga \alst{f}ramar; &
því’t ó\alst{v}íst's at \alst{v}ita, \hld nær verðr á \alst{v}egum úti &
\ind \alst{g}ęirs of þǫrf \alst{g}uma.\eva

\bvb From his weapons shall no man on the field take a step away; for uncertain ’tis to know, when on the ways outside, man comes in need of a spear.\evb
\evg


\bvg
\bva Fann-k-a \alst{m}ildan mann \hld eða svá \alst{m}atar góðan, &
\ind at væri-t \alst{þ}iggja \alst{þ}ęgit; &
eða \alst{s}íns féar \hld \alst{s}vági gløggvan, &
\ind at \alst{l}ęið sé \alst{l}aun, ef þegi.\eva

\bvb I found not a mild man, nor one so good of meat, that a gift was not received; nor of his wealth so frugal, that the rewards be loathed if he took them.\evb
\evg


\bvg
\bva \alst{F}éar síns, \hld es \alst{f}ęngit hęfr, &
\ind skyli-t maðr \alst{þ}ǫrf \alst{þ}ola; &
opt sparir \alst{l}ęiðum \hld þat's hęfr \alst{l}júfum hugat; &
\ind mart gęngr \alst{v}err an \alst{v}arir.\eva

\bvb 40\evb
\evg


\bvg
\bva \alst{V}ǫ́pnum ok \alst{v}ǫ́ðum \hld skulu \alst{v}inir glęðjask; &
\ind þat's á \alst{s}jǫlfum \alst{s}ýnst; &
\alst{v}iðrgefęndr \hld erusk \alst{v}inir lęngst, &
\ind ef þat bíðr at \alst{v}erða \alst{v}ęl.\eva

\bvb With weapons and garments shall friends gladden each other; that is most seen on oneself; mutual givers are friends for the longest, if [the friendship] takes to last long.\evb
\evg


\bvg
\bva \alst{V}in sínum \hld skal maðr \alst{v}inr \alst{v}esa, &
\ind ok \alst{g}jalda \alst{g}jǫf við \alst{g}jǫf; &
\alst{h}látr við \alst{h}látri \hld skyli \alst{h}ǫlðar taka, &
\ind en \alst{l}ausung við \alst{l}ygi.\eva

\bvb With his friend shall a man a friend be, and repay gift against gift; laughter against laughter should men take, but falsehood against lie.\evb
\evg


\bvg
\bva \alst{V}in sínum \hld skal maðr \alst{v}inr vesa, &
\ind \alst{þ}ęim ok \alst{þ}ęss vin; &
en \alst{ó}vinar síns \hld skyli \alst{ę}ngi maðr &
\ind \alst{v}inar \alst{v}inr \alst{v}esa.\eva

\bvb With his friend shall a man a friend be, with him and with his friend; but with his enemy’s, shall no man, friend’s friend be.\evb
\evg


\bvg
\bva \alst{V}ęizt, ef þú \alst{v}in átt, \hld þann's þú \alst{v}ęl trúir &
\ind ok vilt af hǫ́num \alst{g}ótt \alst{g}eta, &
\alst{g}ęði skalt við þann \hld ok \alst{g}jǫfum skipta, &
\ind \alst{f}ara at \alst{f}inna opt.\eva

\bvb Know, if thou hast a friend, which thou trust well, and wilt of him receive good: mind shalt thou with exchange him, and gifts; travel to see him oft.\evb
\evg


\bvg
\bva Ef þú \alst{á}tt \alst{a}nnan, \hld þann's þú \alst{i}lla trúir, &
\ind vildu af hǫ́num þó \alst{g}ótt \alst{g}eta, &
\alst{f}agrt skalt mæla, \hld ęn \alst{f}látt hyggja &
\ind ok gjalda \alst{l}ausung við \alst{l}ygi.\eva

\bvb If thou hast another, which thou trust little, thou wilt of him receive good; fairly shalt thou speak, but deceitfully think, and repay duplicity with lies.\evb
\evg


\bvg
\bva Þat's \alst{ę}nn of þann, \hld es þú \alst{i}lla trúir &
\ind ok þér es \alst{g}runr at \alst{g}ęði, &
\alst{h}læja skalt við þęim \hld ok of \alst{h}ug mæla; &
\ind glík skulu \alst{g}jǫld \alst{g}jǫfum.\eva

\bvb 46\evb
\evg


\bvg
\bva Ungr vas'k \alst{f}orðum, \hld \alst{f}ór'k ęinn saman, &
\ind þá varð'k \alst{v}illr \alst{v}ega; &
\alst{au}ðigr þóttumk, \hld es \alst{a}nnan fann'k, &
\ind \alst{m}aðr es \alst{m}anns gaman.\eva

\bvb Young I was once, I travelled all alone; then I became astray of the paths. Wealthy I thought myself when I another found; man is the joy of man.\evb
\evg


\bvg
\bva \alst{M}ildir frǿknir \hld \alst{m}ęnn bazt lifa, &
\ind \alst{s}jaldan \alst{s}út ala; &
\alst{ó}snjallr maðr \hld \alst{u}ggir hvatvetna, &
\ind sýtir æ \alst{g}løggr við \alst{g}jǫfum.\eva

\bvb 48\evb
\evg


\bvg
\bva \alst{V}áðir mínar \hld gaf'k \alst{v}ęlli at &
\ind \alst{t}vęim \alst{t}rémǫnnum; &
\alst{r}ekkar þat þóttusk, \hld es \alst{r}ipt hǫfðu; &
\ind \alst{n}ęiss es \alst{n}ǫkkviðr halr.\eva

\bvb My garments I gave at the field, to two tree-men. They seemed to be champions, when they cloaks had; ashamed is the naked man.\evb
\evg


\bvg
\bva Hrørnar \alst{þ}ǫll, \hld sú's stęndr \alst{þ}orpi á, &
\ind hlýrat hęnni \alst{b}ǫrkr né \alst{b}arr; &
svá es \alst{m}aðr, \hld sá’s \alst{m}anngi ann; &
\ind hvat skal hann \alst{l}ęngi \alst{l}ifa?\eva

\bvb Wilters the pine, which on the field stands; shield her neither bark nor needle; so is the man, who loves no man; for what shall he live long?\evb
\evg


\bvg
\bva \alst{Ę}ldi hęitari \hld brinnr með \alst{i}llum vinum &
\ind \alst{f}riðr \alst{f}imm daga, &
ęn þá \alst{sl}oknar, \hld es hinn \alst{s}étti kømr, &
\ind ok \alst{v}ersnar allr \alst{v}inskapr.\eva

\bvb Peace burns with ill friends for five days\footnoteB{As \Finnur\ points out, a reference to the five-day week; the number is symbolic.} hotter than fire; but it then goes out, when the sixth comes, and all the friendship worsens.\evb
\evg


\bvg
\bva \alst{M}ikit ęitt \hld skal-a \alst{m}anni gefa; &
\ind opt kaupir sér í \alst{l}ítlu \alst{l}of, &
með \alst{h}ǫlfum \alst{h}lęif \hld ok með \alst{h}ǫllu kęri &
\ind \alst{f}ekk ek mér \alst{f}élaga.\eva

\bvb Much alone, should one not give to a man; often little buys praise. With a half loaf [of bread] and an awry vat, I got me a fellow.\evb
\footnotetext[x]{Lit. “often [one] buys oneself in little praise"}
\evg


\bvg
\bva \alst{L}ítilla sanda, \hld \alst{l}ítilla sæva, &
\ind lítil eru \alst{g}ęð \alst{g}uma; &
því’t \alst{a}llir męnn \hld \alst{u}rðu-t jafnspakir; &
\ind \alst{h}ǫlf es ǫld \alst{h}var.\eva

\bvb Of small sands, of small seas; small are the senses of man. For all have not become evenly wise; half is each man.\footnoteB{Where shores are small, seas are small. Compared to the power of the natural forces, a man is only a grain of sand in the desert, a drop of water in the sea; his wisdom will always be incomplete.}\evb
\evg


\bvg
\bva \alst{M}eðalsnotr \hld skyli \alst{m}anna hvęrr, &
\ind æva til \alst{s}notr \alst{s}é; &
þęim es \alst{f}yrða \hld \alst{f}ęgrst at lifa, &
\ind es \alst{v}ęl mart \alst{v}itu.\eva

\bvb Middle-clever should each man be; never too clever. For those men is it fairest to live, who know well enough.\evb
\evg


\bvg
\bva \alst{M}eðalsnotr \hld skyli \alst{m}anna hvęrr, &
\ind æva til \alst{s}notr \alst{s}é; &
\alst{s}notrs manns hjarta \hld verðr \alst{s}jaldan glatt, &
\ind ef sá’s \alst{a}lsnotr es \alst{á}.\eva

\bvb Middle-clever should each man be; never too clever. The clever man’s heart seldom turns glad, if he is all-clever that owns it.\evb
\evg


\bvg
\bva \alst{M}eðalsnotr \hld skyli \alst{m}anna hvęrr, &
\ind æva til \alst{s}notr \alst{s}é; &
\alst{ø}rlǫg sin \hld viti \alst{ę}ngi fyrir; &
\ind þęim es \alst{s}orgalausastr \alst{s}efi.\eva

\bvb Middle-clever should each man be; never too clever. May no man know his \inx{orlay}{C} ahead; his is the most sorrowless mind\footnoteB{Who knows not his fate. One may contrast Weden who has knowledge of his own inevitable doom.}.\evb
\evg


\bvg
\bva \alst{B}randr af \alst{b}randi \hld \alst{b}rinnr unz \alst{b}runninn es, &
\ind \alst{f}uni kvęykisk af \alst{f}una; &
\alst{m}aðr af \alst{m}anni \hld verðr at \alst{m}áli kuðr; &
\ind ęn til \alst{d}ǿlskr af \alst{d}ul.\eva

\bvb Fire by fire burns until it is burnt; flame is kindled by flame. Man by man becomes known by speech, but the too dull by his conceit.\evb
\evg


\bvg
\bva \alst{Á}r skal rísa, \hld sá’s \alst{a}nnars vill &
\ind \alst{f}é eða \alst{f}jǫr hafa; &
sjaldan \alst{l}iggjandi ulfr \hld \alst{l}ær of getr, &
\ind né \alst{s}ofandi maðr \alst{s}igr.\eva

\bvb Early shall rise, he who another’s cattle or life will have; seldom does the lying wolf get a thigh, or the sleeping man victory.\evb
\evg


\bvg
\bva \alst{Á}r skal rísa, \hld sá's á \alst{y}rkjęndr fáa, &
\ind ok ganga síns \alst{v}erka á \alst{v}it; &
\alst{m}art of dvęlr \hld þann’s of \alst{m}orgin sefr, &
\ind \alst{h}alfr es auðr und \alst{h}vǫtum.\eva

\bvb Early shall rise, he who owns workers few, and go his work to meet; much is kept back from him who in the morning sleeps; half the wealth is due to the bold\Bfootnote{Half of one’s success comes from boldness.}.\evb
\evg


\bvg
\bva \alst{Þ}urra skíða \hld ok \alst{þ}akinna næfra, &
\ind þess kann \alst{m}aðr \alst{m}jǫt, &
ok þess \alst{v}iðar, \hld es \alst{v}innask męgi &
\ind \alst{m}ál ok \alst{m}issęri.\eva

\bvb Dry planks and thatching birch bark, of this man knows the use; and of that firewood, which may be used for a season and half-year\footnoteB{Over the winter.}.\evb
\evg


\bvg
\bva Þvęginn ok \alst{m}ęttr \hld ríði \alst{m}aðr þingi at, &
\ind þótt hann sé-t \alst{v}æddr til \alst{v}ęl; &
\alst{sk}úa ok bróka \hld \alst{sk}ammisk ęngi maðr &
\ind né \alst{h}ęsts in \alst{h}ęldr, \hld \edtext{þótt hann \alst{h}afi't góðan}{\lemma{þótt ... góðan “though ... good one”}\Afootnote{As \Finnur\ points out, surely a later addition.}}.\eva

\bvb Washed and filled may a man ride to the Thing, though he be not dressed too well; of his shoes and breeches ought no man be ashamed, nor indeed of his horse. (Though he have not a good one).\evb
\evg


\bvg
\bva \alst{S}napir ok gnapir, \hld es til \alst{s}ævar kømr, &
\ind \alst{ǫ}rn á \alst{a}ldinn mar; &
svá es \alst{m}aðr, \hld es með \alst{m}ǫrgum kømr &
\ind ok á \alst{f}ormælęndr \alst{f}áa.\eva

\bvb It shuffles and stoops, when to the sea it comes, the eagle on the aged ocean; so is the man who among the many comes, and has spokesmen few.\evb
\evg


\bvg
\bva \alst{F}regna ok sęgja \hld skal \alst{f}róðra hvęrr, &
\ind sá’s vill \alst{h}ęitinn \alst{h}orskr; &
\alst{ęi}nn vita \hld né \alst{a}nnarr skal, &
\ind \alst{þ}jóð vęit ef \alst{þ}rír eru.\eva

\bvb Ask and speak shall every learned man, who will be called sharp; one shall know, but not another: the people knows if there are three.\evb
\evg


\bvg
\bva \alst{R}íki sitt \hld skyli \alst{r}áðsnotra &
\ind \alst{h}vęrr í \alst{h}ófi \alst{h}afa; &
þá hann þat \alst{f}innr, \hld es með \alst{f}rǿknum kømr, &
\ind at \alst{ę}ngi es \alst{ęi}nna hvatastr. &
\alst{O}rða þęira, \hld es maðr \alst{ǫ}ðrum sęgir, &
\ind opt hann \alst{g}jǫld of \alst{g}etr.\eva

\bvb 64\evb
\evg


\bvg
\bva \alst{M}ikilsti snimma \hld kom’k í \alst{m}arga staði, &
\ind ęn til \alst{s}íð í \alst{s}uma; &
\alst{ǫ}l vas drukkit, \hld sumt vas \alst{ó}lagat; &
\ind sjaldan hittir \alst{l}ęiðr í \alst{l}ið.\eva

\bvb 65\evb
\evg


\bvg
\bva \alst{H}ér ok \alst{h}var \hld myndi mér \alst{h}ęim of boðit, &
\ind ef þyrpta'k at \alst{m}ǫ́lungi \alst{m}at, &
eða \alst{t}vau lær hęngi \hld at hins \alst{t}ryggva vinar, &
\ind þar's ek hafða \alst{ęi}tt \alst{e}tit.\eva

\bvb 66\evb
\evg


\bvg
\bva \alst{Ę}ldr es baztr \hld með \alst{ý}ta sonum &
\ind ok \alst{s}ólar \alst{s}ýn, &
\alst{h}ęilyndi sitt, \hld ef \alst{h}afa náir, &
\ind án við \alst{l}ǫst at \alst{l}ifa.\eva

\bvb Fire is best among the sons of men, and the sight of the sun; one’s good health, if one manage to have it, without living with vice.\evb
\evg


\bvg
\bva Es-at maðr \alst{a}lls vesall, \hld þótt sé \alst{i}lla hęill, &
\ind \alst{s}umr es af \alst{s}onum \alst{s}æll, &
\alst{s}umr af frændum, \hld \alst{s}umr af fé ǿrnu, &
\ind sumr af \alst{v}erkum \alst{v}ęl.\eva

\bvb No man is all wretched, though he of poor health be: Some find joy in sons,
some in friends, some in ample \inx{fee}{C}, some in works done well.\evb
\evg


\bvg
\bva Bętra es \alst{l}ifðum, \hld ok sæl\alst{l}ifðum, &
\ind ęy getr \alst{k}vikr \alst{k}ú; &
\alst{ę}ld sá’k \alst{u}pp brinna \hld \alst{au}ðgum manni fyr, &
\ind ęn úti vas \alst{d}auðr fyr \alst{d}urum.\eva

\bvb ’Tis better with the living, and the joyfully living; ever gets the quick\footnoteB{Living.} a cow\footnoteB{A reference to the cattle-based economy, the cow being used as a metonym. The meaning is that new opportunities always present themselves.}. A fire\footnoteB{His funeral-pyre.} I saw burning on high for a wealthy man, but outside he was dead before the door.\evb
\evg


\bvg
\bva \alst{H}altr ríðr \alst{h}rossi, \hld \alst{h}jǫrð rekr \alst{h}andarvanr, &
\ind daufr \alst{v}egr ok \alst{d}ugir; &
\alst{b}lindr es \alst{b}ętri, \hld an \alst{b}ręndr séi; &
\ind \alst{n}ýtr mangi \alst{n}ás.\eva

\bvb The halt man rides a horse, the handless drives a herd, the deaf fights and avails; blind is better than burnèd be: no man has use for a corpse.\evb
\evg


\bvg
\bva \alst{S}onr es bętri, \hld þótt sé \alst{s}íð of alinn &
\ind ęptir \alst{g}inginn \alst{g}uma; &
sjaldan \alst{b}autarstęinar \hld standa \alst{b}rautu nær, &
\ind nema ręisi \alst{n}iðr at \alst{n}ið.\eva

\bvb A son is better, though he late be born, after a passed-on man: seldom beat-stones near the highway stand, unless by kinsman for kinsman raised.\evb
\evg


\bvg
\bva \alst{T}vęir 'ro ęins hęrjar, \hld \alst{t}unga es hǫfuðs bani; &
\ind mér's í \alst{h}eðin \alst{h}vęrn \hld \alst{h}andar væni.\eva

\bvb Two are the harriers of one: the tongue is the head’s bane; in every cloak I expect a hand.\evb
\evg


\bvg
\bva \alst{N}ótt verðr fęginn, \hld sá's \alst{n}esti trúir, &
\ind \alst{sk}ammar 'ro \alst{sk}ips ráar. &
– – – – &
\ind \alst{H}verf es \alst{h}austgríma, &
\alst{f}jǫlð of viðrir \hld á \alst{f}imm dǫgum, &
\ind en \alst{m}ęir á \alst{m}ánaði.\eva

\bvb 73\evb
\evg


\bvg
\bva \alst{V}ęit-a hinn, \hld es \alst{v}ætki \alst{v}ęit, &
\ind margr verðr \edtext{af \alst{au}rum}{\lemma{af aurum}\footnoteA{ ‘aflꜹðrom’ \emph{ms.}}} \alst{a}pi; &
maðr es \alst{au}ðigr, \hld annarr \alst{ó}auðigr, &
\ind skyli-t þann \alst{v}ítka \alst{v}áar.\eva

\bvb He knows not, who nothing knows: treasures make many a man a fool; a man is wealthy, another not wealthy, one oughtn’t curse him for his woe.\evb
\evg


\bvg
\bva \alst{D}ęyr fé, \hld \alst{d}ęyja frændr, &
\ind dęyr \alst{s}jalfr hit \alst{s}ama; &
ęn \alst{o}rðstírr \hld dęyr \alst{a}ldrigi &
\ind hvęim's sér \alst{g}óðan \alst{g}etr.\eva

\bvb \inx{Fee}{C} dies, kinsmen die\Bfootnote{The power of this succinct expression may be less clear to the modern reader. In Germanic Iron Age society wealth was measured by how many heads of cattle a man owned, but he would only be able to keep it if he had a strong clan of male relatives ready to support and protect him.}, oneself dies the same; but a word-glory never dies, for whomever gets himself a good one.\evb
\evg


\bvg
\bva \alst{D}ęyr fé, \hld \alst{d}ęyja frændr, &
\ind dęyr \alst{s}jalfr hit \alst{s}ama; &
\alst{e}k vęit \alst{ęi}nn \hld at \alst{a}ldri dęyr: &
\ind \alst{d}ómr of \alst{d}auðan hvęrn.\eva

\bvb \inx{Fee}{C} dies, kinsmen die, oneself dies the same; I know but one, that never dies: the \inx{Doom}{C} over each man dead.\evb
\evg


\bvg
\bva \alst{F}ullar grindr \hld sá'k fyr \alst{F}itjungs sonum, &
\ind nú bera þęir \alst{v}ánar \alst{v}ǫl; &
svá es \alst{au}ðr \hld sęm \alst{au}gabragð, &
\ind hann es \alst{v}altastr \alst{v}ina. —\eva

\bvb 77\evb
\evg


\bvg
\bva \alst{Ó}snotr maðr, \hld es \alst{ęi}gnask getr &
\ind \alst{f}é eða \alst{f}ljóðs munuð; &
\alst{m}etnaðr hǫ́num þróask, \hld ęn \alst{m}anvit aldrigi; &
\ind framm gęngr hann \alst{d}rjúgt í \alst{d}ul. —\eva

\bvb 78\evb
\evg


\bvg
\bva Þat es þá \alst{r}ęynt, \hld es þú at \alst{r}únum spyrr \hld hinum \alst{r}ęginkunnum, &
\ind þęim's \alst{g}ęrðu \alst{g}innręgin &
\ind ok \alst{f}áði \alst{f}imbulþulr; &
\ind þá hęfr hann bazt, ef þęgir. —\eva

\bvb Then that is proven of which thou inquires the runes, the ones born of the Powers, those which the yin-Powers made, and the fimbol-thyle painted; then he has it best, if he shuts up.\evb
\evg


\bvg
\bva At \alst{k}veldi skal dag lęyfa, \hld \alst{k}onu es bręnd es, &
\alst{m}æki es ręyndr es, \hld \alst{m}ęy es gefin es, &
\alst{í}s es \alst{y}fir kømr, \hld \alst{ǫ}l es drukkit es. —\eva

\bvb At evening [one] shall praise day, a woman when [she] is burned, a sword when [it] is tried, a maiden when [she] is given. Ice when [one] comes over, ale when [it] is drunk.\evb
\evg


\bvg
\bva Í \alst{v}indi skal \alst{v}ið hǫggva, \hld \alst{v}eðri á sæ róa, &
\alst{m}yrkri við \alst{m}an spjalla, \hld \alst{m}ǫrg eru dags augu, &
á \alst{sk}ip skal \alst{sk}riðar orka, \hld ęn á \alst{sk}jǫld til hlífar, &
\alst{m}æki til hǫggs, \hld ęn \alst{m}ęy til kossa. —\eva

\bvb In wind [one] shall cut wood, [in] storm row on the sea, [in] darkness meet with a maid; many are the eyes of day. A ship [one] shall use for its speed, a shield for shelter, a sword for striking, but a maiden for [her] kisses.\evb
\evg


\bvg
\bva Við \alst{e}ld skal \alst{ǫ}l drekka. \hld ęn á \alst{í}si skríða, &
\alst{m}agran \alst{m}ar kaupa, \hld ęn \alst{m}æki saurgan, &
\alst{h}ęima \alst{h}ęst fęita, \hld ęn \alst{h}und á búi. \eva

\bvb 82\evb
\evg


\bvg Regarding the love of women, and Woden's failed love-adventures.
\bva \alst{M}ęyjar orðum \hld skyli \alst{m}angi trúa, &
\ind né því's \alst{k}veðr \alst{k}ona; &
á \alst{h}verfanda \alst{h}véli \hld vǫ́ru þęim \alst{h}jǫrtu skǫpuð, &
\ind \alst{b}rigð í \alst{b}rjóst of lagið\footnotemark[29]. \eva
\footnotetext[29]{The second half of this st. is quoted in \emph{Fbr}, Thott 1768 4°\textsuperscript{x}, ff.: “And he then remembered the short verse, which had been composed about unfaithful women:"}

\bvb The words of a maiden should no man believe, nor that which a woman says. On a spinning wheel were their hearts for them shaped; betrayal laid in their breasts.\evb
\evg


\bvg
\bva \alst{B}restanda \alst{b}oga, \hld \alst{b}rinnanda loga, &
\alst{g}ínanda ulfi, \hld \alst{g}alandi krǫ́ku, &
\alst{r}ýtanda svíni, \hld \alst{r}ótlausum viði, &
\alst{v}axanda \alst{v}ági, \hld \alst{v}ellanda katli,\eva

\bvb Bursting bow, burning flame, gaping wolf, crowing crow, roaring swine, rootless tree, waxing wave, swelling kettle,\evb
\evg


\bvg
\bva \alst{f}ljúganda \alst{f}lęini, \hld \alst{f}allandi bǫ́ru, &
\alst{í}si \alst{ęi}nnættum, \hld \alst{o}rmi hringlęgnum, &
\alst{b}rúðar \alst{b}ęðmǫ́lum \hld eða \alst{b}rotnu sverði, &
\alst{b}jarnar lęiki \hld eða \alst{b}arni konungs,
\alst{s}júkum kalfi, \hld \alst{s}jalfráða þræli, &
\alst{v}ǫlu \alst{v}ilmæli, \hld \alst{v}al nýfęldum. —\eva

\bvb 85\evb
\evg


\bvg
\bva \alst{A}kri \alst{á}rsǫ́num \hld trúi \alst{ę}ngi maðr, &
\ind né til \alst{s}nimma \alst{s}yni; &
\alst{v}eðr ræðr akri, \hld ęn \alst{v}it syni; &
\ind \alst{h}ætt es þęira \alst{h}várt. —\eva

\bvb 86\evb
\evg


\bvg
\bva \alst{B}róður\alst{b}ana sínum \hld þótt á \alst{b}rautu mǿti, &
\alst{h}úsi \alst{h}alfbrunnu, \hld \alst{h}ęsti alskjótum, &
þá's \alst{jó}r \alst{ó}nýtr, \hld ef \alst{ęi}nn fótr brotnar; &
verðit maðr svá \alst{t}ryggr \hld at þessu \alst{t}rúi ǫllu. —\eva

\bvb 87\evb
\evg


\bvg
\bva Svá's \alst{f}riðr kvinna \hld þęira's \alst{f}látt hyggja, &
sęm aki \alst{jó} óbryddum \hld á \alst{í}si hǫ́lum &
(\alst{t}ęitum, \alst{t}vévetrum \hld ok sé \alst{t}amr illa), &
eða í \alst{b}yr óðum \hld \alst{b}ęiti stjórnlausu, &
eða skyli \alst{h}altr \alst{h}ęnda \hld \alst{h}ręin í þáfjalli. —\eva

\bvb 88\evb
\evg


\bvg
\bva \alst{B}ert nú mæli'k, \hld því-at \alst{b}æði vęit'k, &
\ind brigðr es \alst{k}arla hugr \alst{k}onum, &
þá \alst{f}ęgrst mælum, \hld es \alst{f}lást hyggjum; &
\ind þat tælir \alst{h}orska \alst{h}ugi.\eva

\bvb 89\evb
\evg


\bvg
\bva \alst{F}agrt skal mæla \hld ok \alst{f}é bjóða, &
\ind sá's vill \alst{f}ljóðs ǫ́st \alst{f}áa, &
\alst{l}íki \alst{l}ęyfa \hld hins \alst{l}jósa mans, &
\ind sá \alst{f}ær, es \alst{f}ríar.\eva

\bvb 90\evb
\evg


\bvg
\bva \alst{Á}star firna \hld skyli \alst{ę}ngi maðr &
\ind \alst{a}nnan \alst{a}ldrigi; &
opt fáa á \alst{h}orskan, \hld es á \alst{h}ęimskan né fáa, &
\ind \alst{l}ostfagrir \alst{l}itir.\eva

\bvb 91\evb
\evg


\bvg
\bva \alst{Ęy}vitar firna, \hld es maðr \alst{a}nnan skal, &
\ind þess's of margan \alst{g}ęngr \alst{g}uma; &
\alst{h}ęimska ór \alst{h}orskum \hld gęrir \alst{h}ǫlða sonu &
\ind sá hinn \alst{m}átki \alst{m}unr.\eva

\bvb 92\evb
\evg


\bvg
\bva \alst{H}ugr ęinn þat vęit, \hld es býr \alst{h}jarta nær, &
\ind ęinn es hann \alst{s}ér of \alst{s}efa; &
øng es \alst{s}ótt verri \hld hvęim \alst{s}notrum manni &
\ind an sér \alst{ø}ngu at \alst{u}na.\eva

\bvb 93\evb
\evg


\bvg
\bva Þat þá \alst{r}ęyndak, \hld es í \alst{r}ęyri sat'k, &
\ind ok vætta'k \alst{m}íns \alst{m}unar, &
\alst{h}old ok \alst{h}jarta \hld vas mér hin \alst{h}orska mær, &
\ind þęygi hana at \alst{h}ęldr \alst{h}ęf'k.\eva

\bvb 94\evb
\evg


\bvg
\bva \alst{B}illings męy \hld ek fann \alst{b}ęðjum á &
\ind \alst{s}ólhvíta \alst{s}ofa; &
\alst{ja}rls \alst{y}nði \hld þótti mér \alst{ę}kki vesa &
\ind nema við þat \alst{l}ík at \alst{l}ifa.\eva

\bvb 95\evb
\evg


\bvg
\bva “\alst{Au}k nær \alst{a}ptni \hld skaltu \alst{Ó}ðinn koma, &
\ind ef vilt þér \alst{m}æla \alst{m}an, &
\alst{a}lt eru \alst{ó}skǫp, \hld nema \alst{ęi}n vitim &
\ind \alst{s}likan lǫst \alst{s}aman.”\eva

\bvb 96\evb
\evg


\bvg
\bva \alst{A}ptr ek hvarf \hld ok \alst{u}nna þóttumk &
\ind \alst{v}ísum \alst{v}ilja frá; &
\alst{h}itt ek \alst{h}ugða, \hld at \alst{h}afa mynda'k &
\ind \alst{g}ęð hęnnar alt ok \alst{g}aman.\eva

\bvb 97\evb
\evg


\bvg
\bva Svá kom'k \alst{n}æst, \hld at hin \alst{n}ýta vas &
\ind \alst{v}ígdrótt ǫll of \alst{v}akin; &
með \alst{b}rinnǫndum ljósum \hld ok \alst{b}ornum viði, &
\ind svá vas mér \alst{v}ílstígr of \alst{v}itaðr.\eva

\bvb 98\evb
\evg


\bvg
\bva \alst{Au}k nær morni, \hld es vas'k \alst{ę}nn of kominn, &
\ind þá vas \alst{s}aldrótt of \alst{s}ofin; &
\alst{g}ręy ęitt þá fan'k \hld hinnar \alst{g}óðu konu &
\ind \alst{b}undit \alst{b}ęðjum á.\eva

\bvb 99\evb
\evg


\bvg
\bva Mǫrg es \alst{g}óð mær, \hld ef \alst{g}ǫrva kannar, &
\ind \alst{h}ugbrigð við \alst{h}ali; &
þá þat \alst{r}ęynda'k, \hld es hit \alst{r}áðspaka &
\ind tęygða'k á \alst{f}lærðir \alst{f}ljóð. &
\alst{h}ǫ́ðungar \alst{h}vęrrar \hld lęitaði mér hit \alst{h}orska man &
\ind ok hafða'k þess \alst{v}ætki \alst{v}ífs.\eva

\bvb 100\evb
\evg


\bvg Side-composition to the previous poem, starting with a general maxim.
\bva Hęima \alst{g}laðr \hld ok við \alst{g}ęsti ręifr, &
\ind \alst{s}viðr skal of \alst{s}ik vesa; &
\alst{m}innigr ok \alst{m}ǫ́lugr, \hld ef vill \alst{m}argfróðr vesa; &
\ind opt skal \alst{g}óðs \alst{g}eta; &
\alst{f}imbul\alst{f}ambi hęitir, \hld sás \alst{f}átt kann sęgja; &
\ind þat es \alst{ó}snotrs \alst{a}ðal.\eva

\bvb 101\evb
\evg


\bvg
\bva Hinn \alst{a}ldna \alst{jǫ}tun sóttak, \hld nú em'k \alst{a}ptr of kominn; &
\ind fátt gat'k \alst{þ}ęgjandi \alst{þ}ar; &
\alst{m}ǫrgum orðum \hld \alst{m}ælta'k í minn frama &
\ind í \alst{S}uttungs \alst{s}ǫlum.\eva

\bvb The old ettin I sought, now am I come back; I got little silence there. Many words I spoke to my fame, in the halls of Suttung.\evb
\evg


\bvg
\bva \alst{G}unnlǫð mér of \alst{g}af \hld \alst{g}ollnum stóli á &
\ind \alst{d}rykk hins \alst{d}ýra mjaðar; &
\alst{i}ll \alst{i}ðgjǫld \hld lét'k hana \alst{ę}ptir hafa &
\ind síns hins \alst{h}ęila \alst{h}ugar. &
\ind (síns hins \alst{s}vára \alst{s}efa).\eva

\bvb 103\evb
\evg


\bvg
\bva \alst{R}ata munn \hld létumk \alst{r}úms of fáa &
\ind ok of \alst{g}rjót \alst{g}naga; &
\alst{y}fir ok \alst{u}ndir \hld stóðumk \alst{jǫ}tna vegir, &
\ind svá \alst{h}ættak \alst{h}ǫfði til.\eva

\bvb 104\evb
\evg


\bvg
\bva \alst{V}ęl kęypts hlutar \hld hęf'k \alst{v}ęl notit; &
\ind \alst{f}ás es \alst{f}róðum vant; &
\alst{Ó}ðrerir \hld nú \alst{u}pp's kominn &
\ind á \alst{a}lda vé \alst{ja}ðars.\eva

\bvb 105\evb
\evg


\bvg
\bva \alst{I}fi es mér á, \hld at væra'k \alst{ę}nn kominn &
\ind \alst{jǫ}tna gǫrðum \alst{ó}r, &
ef \alst{G}unnlaðar né nyta'k, \hld hinnar \alst{g}óðu konu, &
\ind es lǫgðumk \alst{a}rm \alst{y}fir.\eva

\bvb 106\evb
\evg


\bvg
\bva Hins \alst{h}indra dags \hld gingu \alst{h}rímþursar &
\ind (Háva ráðs at fregna) &
\ind \alst{H}áva \alst{h}ǫllu í, &
at \alst{B}ǫlverki spurðu, \hld ef væri með \alst{b}ǫndum kominn &
\ind eða hęfði hǫ́num \alst{S}uttungr of \alst{s}óit.\eva

\bvb 107\evb
\evg


\bvg
\bva Baugęið \alst{Ó}ðinn \hld hygg at \alst{u}nnit hafi, &
\ind hvat skal hans \alst{t}ryggðum \alst{t}rúa? &
\alst{S}uttung \alst{s}vikvinn \hld hann lét \alst{s}umbli frá &
\ind ok \alst{g}rǿtta \alst{G}unnlǫðu.\eva

\bvb 108\evb
\evg


Advice of the Fimble-Thyle, given to Loddfáfnir.


\bvg
\bva Mál’s at \alst{þ}ylja \hld \alst{þ}ular stóli á; &
\ind \alst{U}rðar brunni \alst{a}t &
\alst{s}á’k ok þagða’k, \hld \alst{s}á’k ok hugða’k, &
\ind hlýdda’k á \alst{m}anna \alst{m}ál; &
of \alst{r}únar hęyrða’k dǿma, \hld né um \alst{r}ǫ́ðumk þǫgðu &
\ind \alst{H}áva \alst{h}ǫllu at, &
\ind \alst{H}áva \alst{h}ǫllu í &
\ind hęyrða'k \alst{s}ęgja \alst{s}vá:\eva

\bvb It is time to \inx{thilly}{C}, upon the chair of the \inx{thyle}{C}. At the well of Weird I saw and I was silent; I saw and I pondered; I heeded the matters of men. Of runes I heard them speak, nor about counsels were they silent; at the hall of the High One; in the hall of the High One, I heard them say thus:\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jǫta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{n}ótt þú rís-at, \hld nema á \alst{n}jósn séir, &
\ind eða lęitir þér \alst{i}nnan \alst{ú}t staðar.\eva

\bvb I counsel thee Loddfathomer, that thou take the counsels; thou wilt benefit if thou take them; they will be good for thee if thou get them: At night thou rise not, unless at scouting thou be, or TODO\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{f}jǫlkunnigri konu \hld skal-at-tu í \alst{f}aðmi sofa, &
\ind svá at hon \alst{l}yki þik \alst{l}iðum. &
Hón svá \alst{g}ęrir \hld at þú \alst{g}áir ęigi &
\ind \alst{þ}ings né \alst{þ}jóðans máls &
\alst{m}at þú vill-at \hld né \alst{m}anskis gaman &
\ind fęrr þú \alst{s}orgafullr at \alst{s}ofa.\eva

\bvb 111\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{a}nnars konu \hld tęyg þér \alst{a}ldrigi &
\ind \alst{ęy}rarúnu \alst{a}t.\eva

\bvb 112\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
á \alst{f}jalli eða \alst{f}irði, \hld ef þik \alst{f}ara tíðir, &
\ind fásk-tu at \alst{v}irði \alst{v}ęl.\eva

\bvb 113\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{i}llan mann \hld lát \alst{a}ldrigi &
\ind óhǫpp at þér vita. &
af \alst{i}llum manni \hld fær þú \alst{a}ldrigi &
\ind \alst{g}jǫld hins \alst{g}óða hugar.\eva

\bvb 114\evb
\evg


\bvg
\bva \alst{O}farla bíta \hld sá'k \alst{ęi}num hal &
\ind \alst{o}rð \alst{i}llrar konu, &
\alst{f}lárǫ́ð tunga \hld varð hǫ́num at \alst{f}jǫrlagi &
\ind ok þęygi of \alst{s}anna \alst{s}ǫk.\eva

\bvb 115\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{v}ęizt ef \alst{v}in átt, \hld þann's \alst{v}el trúir, &
\ind \alst{f}ar þú at \alst{f}inna opt. &
því’t \alst{h}rísi vęx \hld ok \alst{h}óu grasi &
\ind \alst{v}egr, es \alst{v}ætki trøðr,\eva

\bvb 116\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{v}in þínum \hld \alst{v}es þú aldrigi &
\ind \alst{f}yrri at \alst{f}laumslitum. &
\alst{s}org etr hjarta, \hld ef þú \alst{s}ęgja né náir &
\ind \alst{ęi}nhvęrjum \alst{a}llan hug.\eva

\bvb 117\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{g}óðan mann \hld tęyg þér at \alst{g}amanrúnum &
\ind ok nem \alst{l}íknargaldr meðan \alst{l}ifir.\eva

\bvb 118\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
orðum \alst{sk}ipta \hld þú \alst{sk}alt aldrigi &
\ind við \alst{ó}svinna \alst{a}pa.\eva

\bvb 119\evb
\evg


\bvg
\bva Af illum \alst{m}anni \hld \alst{m}undu aldrigi &
\ind \alst{g}óðs laun of \alst{g}eta, &
en \alst{g}óðr maðr \hld mun þik \alst{g}ęrva męga &
\ind \alst{l}íknfastan at \alst{l}ofi.\eva

\bvb 120\evb
\evg


\bvg
\bva \alst{S}ifjum es þá blandit \hld hvęrr es \alst{s}ęgja ræðr &
\ind \alst{ęi}num \alst{a}llan hug; &
alt es \alst{b}ętra \hld an sé \alst{b}rigðum at vesa: &
\ind es-a sá \alst{v}inr es \alst{v}ilt ęitt sęgir.\eva

\bvb 121\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
þrimr orðum sęnna \hld skal-at-tu þér við verra mann, &
\ind opt hinn \alst{b}ętri \alst{b}ilar. &
\ind þás hinn \alst{v}erri \alst{v}egr.\eva

\bvb 122\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{sk}ósmiðr þú verir\hld né \alst{sk}ęptismiðr, &
\ind nema \alst{s}jǫlfum þér \alst{s}éir. &
\alst{Sk}ór's \alst{sk}apaðr illa\hld eða \alst{sk}apt sé rangt, &
\ind þá's þér \alst{b}ǫls \alst{b}eðit.\eva

\bvb 123\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
hvars þú \alst{b}ǫl kant, \hld kveð þér \alst{b}ǫlvi at &
\ind ok gefat þínum \alst{f}jǫ́ndum \alst{f}rið.\eva

\bvb 124\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{i}llu fęginn \hld ves þú \alst{a}ldrigi, &
\ind ęn lát þér at \alst{g}óðu \alst{g}etit.\eva

\bvb 125\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{u}pp líta \hld skalattu í \alst{o}rrostu &
\alst{g}jalti \alst{g}líkir \hld verða \alst{g}umna synir &
\ind síðr þitt of \alst{h}ęilli \alst{h}alir.\eva

\bvb 126\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
Ef vilt þér góða \alst{k}onu \hld \alst{k}vęðja at gamanrúnum &
\ind ok \alst{f}á \alst{f}ǫgnuð af, &
\alst{f}ǫgru skaldu heita \hld ok láta \alst{f}ast vesa; &
\ind lęiðisk mangi \alst{g}ótt ef \alst{g}etr.\eva

\bvb 127\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\ind \alst{v}aran bið'k þik \alst{v}esa &
\ind ok \alst{ęi}gi \alst{o}fvaran, &
ves þú við \alst{ǫ}l varastr. \hld ok við \alst{a}nnars konu &
ok við \alst{þ}at hit \alst{þ}riðja, \hld at \alst{þ}jófar né lęiki.\eva

\bvb 128\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
at \alst{h}áði né \alst{h}látri \hld \alst{h}af þú aldrigi &
\ind \alst{g}ęst né \alst{g}anganda.\eva

\bvb 129\evb
\evg


\bvg
\bva \alst{O}pt vitu \alst{ó}gǫrla, \hld þęir's sitja \alst{i}nni fyrir, &
\ind hvęrs þęir 'ro \alst{k}yns es \alst{k}oma; &
esat maðr svá \alst{g}óðr \hld at \alst{g}alli né fylgi, &
\ind né svá \alst{i}llr at \alst{ęi}nugi dugi.\eva

\bvb 130\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
at \alst{h}ǫ́rum þul \hld \alst{h}læ þú aldrigi, &
\ind opt es \alst{g}ótt þats \alst{g}amlir kveða, &
opt ór \alst{sk}ǫrpum bęlg \hld \alst{sk}ilin orð koma &
\ind þęims \alst{h}angir með \alst{h}ǫ́m &
\ind ok \alst{sk}ollir með \alst{sk}rǫ́m, &
\ind ok \alst{v}áfir með \alst{v}ilmǫgum.\eva

\bvb 131\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{g}ęst né \alst{g}ęyja \hld né á \alst{g}rind hrękir; &
\ind get þú \alst{v}ǫ́luðum \alst{v}el.\eva

\bvb 132\evb
\evg


\bvg
\bva \alst{R}amt es þat tré, \hld es \alst{r}íða skal &
\ind \alst{ǫ}llum at \alst{u}pploki; &
\alst{b}aug þú gef \hld eða þat \alst{b}iðja mun &
\ind þér \alst{l}æs hvęrs á \alst{l}iðu.\eva

\bvb 133\evb
\evg


\bvg
\bva \alst{R}ǫ́ðumk þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
hvars \alst{ǫ}l drekkir \hld kjós þér \alst{ja}rðar męgin, &
því’t \alst{jǫ}rð tękr við \alst{ǫ}ldri, \hld ęn \alst{ę}ldr við sóttum, &
\alst{ęi}k við \alst{a}bbindi, \hld \alst{a}x við fjǫlkyngi, &
\alst{h}ǫll við \alst{h}ýrógi; \hld \alst{h}ęiptum skal mána kvęðja, &
\alst{b}ęiti við \alst{b}itsóttum, \hld ęn við \alst{b}ǫlvi rúnar; &
\ind \alst{f}old skal við \alst{f}lóði taka.\eva

\bvb 134\evb
\evg


Of Woden's taking of the runes.
It is clear that these verses have very little to do with the rest of the poem, but instead are separate. It is for this reason that they are labelled as \emph{Rúnatals þáttr} (The strand of the Runecount) in younger Eddic paper manuscripts. Many give an archaic, pagan impression. It is as if they were drawn from the lips of an Odinic priest.


\bvg
\bva \footnotemark[2] \alst{V}ęit'k at ek hekk \hld \alst{v}indga męiði á &
\ind \alst{n}ætr allar \alst{n}íu, &
\alst{g}ęiri undaðr \hld ok \alst{g}efinn Óðni, &
\ind \alst{s}jalfr \alst{s}jǫlfum mér, &
á þęim \alst{m}ęiði, \hld es \alst{m}angi vęit, &
\ind hvęrs af \alst{r}ótum \alst{r}innr\footnotemark[3].\eva
\footnotetext[3]{This v. clearly belongs together with 138 and 140 and may have been part of a greater now-lost work.}
\footnotetext[3]{Note that \emph{rótum} has lost initial \emph{v-}. Cf. 150/2, which retains it. Unlike the specifically WN. loss of \emph{v} before \emph{r} (cf. 32/2), in this word it occurred in all NGmc. dialects, cf. Swe. \emph{rot}. This may make it less reliable to use for dating the stanza.}

\bvb I know that I hung upon a windy beam [for] nine nights all, gored by spear, and given to Woden, myself to myself, on that tree, which no man knows whereof his roots run.\evb
\evg


\bvg
\bva Við \alst{h}lęifi mik sældu \hld né við \alst{h}ornigi; &
\alst{n}ýsta'k \alst{n}iðr, \hld \alst{n}am'k upp rúnar, &
\alst{ǿ}pandi nam, \hld fell'k \alst{a}ptr þaðan.\eva

\bvb With a loaf [of bread] they gladdened me not, nor with horn's drink. I peered down, I took up the runes, screaming [I] took; then I fell back thence.\evb
\evg


\bvg
\bva \alst{F}imbulljóð níu \hld nam'k af hinum \alst{f}rægja syni &
\ind \alst{B}ǫlþorns, \alst{B}ęstlu fǫður, &
ok ek \alst{d}rykk of gat \hld hins \alst{d}ýra mjaðar &
\ind \alst{au}sinn \alst{Ó}ðreri.\footnotemark[1] —\eva
\footnotetext[1]{It has been noted (FJ) that this verse fits better in the next section of the poem. It is awkwardly placed here, since it mentions \emph{ljóð} '(magical) songs', rather than runes.}

\bvb Nine fimbol-songs, I got from the famous son of \textbf{Balethorn}, the father of \textbf{Bestel}. And a drink I got, of the expensive mead poured to \textbf{Woderearer}.\evb
\evg


\bvg
\bva Þá nam'k \alst{f}rævask \hld ok \alst{f}róðr vesa &
\ind ok \alst{v}axa ok \alst{v}ęl hafask; &
\alst{o}rð mér af \alst{o}rði \hld \alst{o}rðs lęitaði &
\ind \alst{v}erk mér af \alst{v}erki \alst{v}erks.\eva

\bvb Then I began to thrive, and be learned, and grow and have it well. A word for me of a word a word sought out; a work for me of a work a work\footnoteB{Each good word and deed was followed by another.}.\evb
\evg


\bvg
\bva \alst{R}únar munt finna \hld ok \alst{r}áðna stafi, &
\ind mjǫk \alst{st}óra \alst{st}afi, &
\ind mjǫk \alst{st}inna \alst{st}afi, &
\ind es \alst{f}áði \alst{f}imbulþulr &
\ind ok \alst{g}ęrðu \alst{g}innręgin &
\ind ok \alst{r}ęist Hroptr \alst{r}agna\footnotemark[5]. —\eva
\footnotetext[5]{Corrected from \emph{rǫgna}. Cf. Eskál \emph{Vell} 31/2 in SkP I, p. 322.}

\bvb \textbf{Runes} thou wilt find, and interpreted staves; much large staves, much stiff staves, which the \textbf{fimbol-thyle} painted, and the \textbf{ginn}-\textbf{Powers} made, and the \textbf{Roft} of the Powers carved.\evb
\evg


\bvg
\bva \alst{Ó}ðinn með \alst{ǫ́}sum, \hld ęn fyr \alst{ǫ}lfum Dáinn, &
\ind \alst{D}valinn \alst{d}vergum fyrir, &
\ind \alst{Á}sviðr \alst{jǫ}tnum fyrir, &
ek\footnotemark[10] ręist \alst{s}jalfr \alst{s}umar. —\eva
\footnotetext[10]{Since Woden has already been mentioned, it is unclear who the \emph{I} here may be.}

\bvb \textbf{Woden} among the \textbf{Ease}, but before the \textbf{Elves} \textbf{Dowen}, \textbf{Dwollen} before the \textbf{Dwarves}, \textbf{Aswood} before the etins; I myself carved some.\evb
\evg


\bvg
\bva Vęiztu, hvé \alst{r}ísta skal? \hld vęiztu, hvé \alst{r}áða skal? &
vęiztu, hvé \alst{f}áa skal? \hld vęiztu, hvé \alst{f}ręista skal? &
vęiztu, hvé \alst{b}iðja skal? \hld vęiztu, hvé \alst{b}lóta skal? &
vęiztu, hvé \alst{s}ęnda skal? \hld vęiztu, hvé \alst{s}óa skal? —\footnotemark[5]\eva
\footnotetext[5]{This v. bears strong resemblance with Vg 216 (Högstena gealdor). TODO: Elaborate.}

\bvb Knowest thou how to carve? Knowest thou how to interpret? Knowest thou how to paint? Knowest thou how to try? Knowest thou how to bid? Knowest thou how to bloot? Knowest thou how to send? Knowest thou how to \textbf{soo}?\evb
\evg


\bvg
\bva \alst{B}ętra's ó\alst{b}eðit \hld an sé of\alst{b}lótit, &
\ind ęy sér til \alst{g}ildis \alst{g}jǫf; &
bętra's ó\alst{s}ęnt \hld an sé of\alst{s}óit.\footnotemark[6]\eva
\footnotetext[6]{A final line is likely missing here. — Identical word-pairing (\emph{biðja} – \emph{blóta}, \emph{senda} – \emph{sóa}) may reveal this v.'s relation with the previous one.}

\bvb Better is unbid than be excessively blooted; a gift always looks to a tribute. Better is unsent than be excessively sooed.\evb
\evg


\bvg
\bva Svá \alst{Þ}undr of ręist \hld fyr \alst{þ}jóða rǫk &
þar's \alst{u}pp of ręis, \hld es \alst{a}ptr of kom.\eva

\bvb Thus \textbf{Thound} did carve for the fate of the nations; there did rise up, who came back.\footnotemark[8]\evb
\footnotetext[8]{A most cryptic v.}
\evg


\bvg Woden's recounting of his Songs.
\bva Ljóð \alst{þ}au kan'k, \hld es kann-at \alst{þ}jóðans kona &
\ind ok \alst{m}anskis \alst{m}ǫgr. &
\alst{H}jǫlp hęitir ęitt, \hld þat þér \alst{h}jalpa mun &
\ind við \alst{s}orgum ok \alst{s}ǫkum, \hld ok \alst{s}útum gǫrvǫllum.\eva

\bvb I know those \textbf{leeds}, which the prince's woman does not know, and no man's lad. Help one is called, it will help thee, against sorrows and sakes\footnotemark[9], and all kinds of misfortunes.\evb
\footnotetext[9]{Legal proceedings.}
\evg


\bvg
\bva Þat kan'k \alst{a}nnat, \hld es þurfu \alst{ý}ta synir,\footnotemark[10] &
\ind þęir's vilja \alst{l}æknar \alst{l}ifa.\eva
\footnotetext[10]{Identical wording to 163/2.}

\bvb 145\evb
\evg


\bvg
\bva \alst{Þ}at kan'k \alst{þ}riðja, \hld ef mér verðr \alst{þ}ǫrf mikil &
\ind \alst{h}apts við mina \alst{h}ęiptmǫgu, &
\alst{ę}ggjar dęyfi'k \hld minna \alst{a}ndskota, &
\ind bítat þęim \alst{v}ǫ́pn né \alst{v}ęlir.\eva

\bvb 146\evb
\evg


\bvg
\bva Þat kan'k \alst{f}jórða, \hld ef mér \alst{f}yrðar bera &
\ind \alst{b}ǫnd at \alst{b}oglimum, &
svá ek \alst{g}ęl, \hld at \alst{g}anga má'k, &
\ind sprettr mér af \alst{f}ótum \alst{f}jǫturr. &
\ind ęn af \alst{h}ǫndum \alst{h}apt.\eva

\bvb 147\evb
\evg


\bvg
\bva Þat kan'k \alst{f}imta, \hld ef sé'k af \alst{f}ári skotinn &
\ind \alst{f}lęin í \alst{f}olki vaða, &
flýgr-a svá \alst{st}int, \hld at \alst{st}ǫðvigak, &
\ind ef hann \alst{s}jónum of \alst{s}é'k.\eva

\bvb 148\evb
\evg


\bvg
\bva Þat kan'k \alst{s}étta, \hld ef mik \alst{s}ærir þegn &
\ind á \alst{v}rótum hrás \alst{v}iðar. &
þann \alst{h}al, \hld es mik \alst{h}ęipta kvęðr, &
\ind þann eta \alst{m}ęin hęldr an \alst{m}ik.\eva

\bvb 149\evb
\evg


\bvg
\bva Þat kan'k \alst{s}jaunda, \hld ef \alst{s}é'k hóvan loga &
\ind \alst{s}al of \alst{s}essmǫgum, &
\alst{b}rinnrat svá \alst{b}ręitt, \hld at hǫ́num \alst{b}jargigak; &
\ind þann kan'k \alst{g}aldr at \alst{g}ala.\eva

\bvb 150\evb
\evg


\bvg
\bva Þat kan'k \alst{á}tta, \hld es \alst{ǫ}llum es &
\ind \alst{n}ytsamligt at \alst{n}ema, &
\alst{h}var's \alst{h}atr vęx \hld með \alst{h}ildings sonum, &
\ind þat má'k \alst{b}ǿta \alst{b}rátt.\eva

\bvb 151\evb
\evg


\bvg
\bva Þat kan'k \alst{n}íunda, \hld ef mik \alst{n}auðr of stęndr &
\ind at bjarga \alst{f}ari á \alst{f}loti, &
\alst{v}ind ek kyrri \hld \alst{v}ági á &
\ind ok \alst{s}væfi'k allan \alst{s}æ.\eva

\bvb 152\evb
\evg


\bvg
\bva Þat kan'k \alst{t}íunda, \hld ef sé'k \alst{t}únriður &
\ind \alst{l}ęika \alst{l}opti á, &
ek svá \alst{v}in'k, \hld at þær \alst{v}illar fara &
\ind sinna \alst{h}ęim-\alst{h}ama &
\ind sinna \alst{h}ęim-\alst{h}uga.\eva

\bvb 153\evb
\evg


\bvg
\bva Þat kan'k \alst{ę}llipta, \hld ef skal'k til \alst{o}rrostu &
\ind \alst{l}ęiða \alst{l}angvini, &
und \alst{r}andir gęlk, \hld ęn þęir með \alst{r}íki fara, &
\ind \alst{h}ęilir \alst{h}ildar til, &
\ind \alst{h}ęilir \alst{h}ildi frá, &
\ind koma þęir \alst{h}ęilir \alst{h}vaðan.\eva

\bvb 154\evb
\evg


\bvg
\bva Þat kan'k \alst{t}olpta, \hld ef sé'k á \alst{t}ré uppi &
\ind \alst{v}áfa \alst{v}irgilná, &
svá ek \alst{r}íst \hld ok í \alst{r}únum fá'k, &
\ind at sá \alst{g}ęngr \alst{g}umi. &
\ind ok \alst{m}ælir við \alst{m}ik.\eva

\bvb 155\evb
\evg


\bvg
\bva \alst{Þ}at kan'k \alst{þ}rettánda \hld ef skal'k \alst{þ}egn ungan &
\ind \alst{v}erpa \alst{v}atni á,\footnotemark[49] &
munat hann \alst{f}alla, \hld þótt í \alst{f}olk komi, &
\ind \alst{h}nígr-a sá \alst{h}alr fyr \alst{h}jǫrum.\eva
\footnotetext[49]{Describing the pagan ritual of pouring water on a newborn child. Cf. \emph{Rþ} 7/1b, 21/1b, 34/2a.}

\bvb 156\evb
\evg


\bvg
\bva Þat kan'k \alst{f}jogurtánda, \hld ef skal'k \alst{f}yrða liði &
\ind \alst{t}ęlja \alst{t}íva fyrir, &
\alst{á}sa ok \alst{a}lfa \hld ek kann \alst{a}llra skil, &
\ind fár kann ó\alst{s}notr \alst{s}vá.\eva

\bvb 157\evb
\evg


\bvg
\bva \alst{Þ}at kan'k fimtánda, \hld es gól {Þ}jóðrørir &
\ind \alst{d}vergr fyr \alst{D}ęllings \alst{d}urum, &
\alst{a}fl gól \alst{ǫ́}sum, \hld ęn \alst{ǫ}lfum frama, &
\ind \alst{h}yggju \alst{H}roptatý.\eva

\bvb 158\evb
\evg


\bvg
\bva Þat kan'k sextánda, \hld ef vil'k hins svinna mans &
\ind hafa gęð alt ok gaman, &
\alst{h}ugi \alst{h}vęrfi'k \hld \alst{h}vitarmri konu &
\ind ok \alst{s}ný'k hęnnar ǫllum \alst{s}efa.\eva

\bvb 159\evb
\evg


\bvg
\bva Þat kan'k \alst{s}jautjánda \hld at mik \alst{s}ęint mun firrask &
\ind hit \alst{m}anunga \alst{m}an.\eva

\bvb 160\evb
\evg


\bvg
\bva Þat kan'k \alst{á}tjánda, \hld es \alst{æ}va kęnni'k &
\ind \alst{m}ęy né \alst{m}anns konu, &
\alst{a}lt es bętra \hld es \alst{ęi}nn of kann, &
\ind þat fylgir \alst{l}jóða \alst{l}okum, &
nema þęiri \alst{ęi}nni, \hld es mik \alst{a}rmi vęrr, &
\ind eða mín \alst{s}ystir \alst{s}é.\eva

\bvb 161\evb
\evg


\bvg
\bva Nú eru \alst{H}áva mál \hld kveðin \alst{H}áva\alst{h}ǫllu í &
\ind \alst{a}llþǫrf \alst{ý}ta sonum, &
\ind \alst{ó}þǫrf \alst{jǫ}tna sonum; &
hęill sás kvað, \hld hęill sás kann, &
\ind \alst{n}jóti sás \alst{n}am, &
\ind \alst{h}ęilir þęirs \alst{h}lýddu.\footnotemark[50]\eva
\footnotetext[50]{A highly metrically irregular verse; clearly younger than the rest of the poem collection. See the Introduction, p. TODO.}

\bvb 162\evb
\evg
