\book{The Speeches of the High One. (\emph{Hávamǫ́l})}\bookStart

	Introduction.
\small{\emph{Hávamǫ́l}, or "the Speeches of the High One\footnotemark[1]", is the second poem of R.}

	Advice to wanderers.
\bva \alst{G}áttir allar \hld áðr \alst{g}angi framm \\%M
\ind (of \alst{sk}oðask \alst{sk}yli)\footnotemark[2] \\%M
\ind of \alst{sk}yggnask \alst{sk}yli; \\%M
þvíat ó\alst{v}íst's at \alst{v}ita, \hld hvar ó\alst{v}inir sitja\footnotemark[3] \\%M
\ind á \alst{f}lęti \alst{f}yrir.\\%E
\footnotetext[2]{\emph{Gylf}, which quotes this verse as being said by Hár (Woden) welcoming Gylfi, does not have this line. Apart from that, the two are identical.}
\footnotetext[3]{For metrical reasons \emph{sitja} must be part of this l., and not the next. Cf. 6/6 for a similarly short C-line.}

\bvb All doorways—before one might go forth—should be observed, should be investigated; for [it] is uncertain to know, where enemies\footnotemark[1] sit on the floor before [oneself]. \\		
\footnotetext[1]{Lit. 'unfriends'.}

\bva \alst{G}efęndr hęilir, \hld \alst{g}ęstr's inn kominn, \\%M
\ind hvar skal \alst{s}itja \alst{s}já? \\%M
mjǫk es \alst{b}ráðr \hld sá's á \alst{b}rǫndum skal \\%M
\ind síns of \alst{f}ręista \alst{f}rama.\\%E

\bvb Hail the givers\footnotemark[5]! A guest has come in, where shall that one sit? Greatly hurried is he, who on the fires shall try his fame. \\
\footnotetext[5]{The hosts.}

\bva \alst{Ę}lds es þǫrf \hld þęim's \alst{i}nn es kominn \\%M
\ind ok á \alst{k}néi \alst{k}alinn, \\%M
\alst{m}atar ok váða \hld es \alst{m}anni þǫrf, \\%M
\ind þęim's hęfr of \alst{f}jall \alst{f}arit.\\%E

\bvb There is need of fire, for the one who has come in, and is cold around the knees. Food and clothing are of use, to the man who has travelled over the mountains. \\

\bva \alst{V}ats es þǫrf \hld þęim's til \alst{v}erðar kømr, \\%M
\ind \alst{þ}ęrru ok \alst{þ}jóðlaðar, \\%M
\alst{g}óðs of ǿðis, \hld ef sér \alst{g}eta mætti, \\%M
\ind \alst{o}rðs ok \alst{ę}ndrþǫgu.\\%E

\bvb There is need of water, for the one who comes to a host, [likewise] a towel and a friendly welcome, a good reception—if he might get one—speech and silence in return. \\

\bva \alst{V}its es þǫrf \hld þęim's \alst{v}íða ratar; \\%M
\ind dælt es \alst{h}ęima \alst{h}vat; \\%M
at \alst{au}gabragði \hld verðr sá's \alst{ę}kki kann \\%M
\ind ok með \alst{s}notrum \alst{s}itr.\\%E

\bvb There is need of wits, for the one who widely wanders; everything is familiar at home. A laughing-stock\footnotemark[8] becomes he who knows nothing and sits with the wise. \\
\footnotetext[8]{An idiom, literally "becomes into a twinkling of an eye"}

\bva At \alst{h}yggjandi sinni \hld skyli-t maðr \alst{h}rǿsinn vesa, \\%M
\ind hęldr \alst{g}ætinn at \alst{g}ęði, \\%M
þá's \alst{h}orskr ok þǫgull \hld kømr \alst{h}ęimisgarða til, \\%M
\ind sjaldan verðr \alst{v}íti \alst{v}ǫrum. \\%M
þvíat óbrigðra vin \hld fær þú aldrigi, \\%M
\ind an \alst{m}anvit \alst{m}ikit.\\%E

\bvb Of his intelligence no man should be boastful, [but] rather guard his conduct when he, prudent and taciturn, comes to a homestead; sudden injury seldom strikes the wary. For thou getst never a more steadfast\footnotemark[10] friend, than much manwit. \\
\footnotetext[10]{Lit. 'more unfickle'}

\bva Hinn \alst{v}ari gęstr, \hld es til \alst{v}erðar kømr, \\%M
\ind \alst{þ}unnu hljóði \alst{þ}ęgir; \\%M
\alst{ęy}rum hlýðir, \hld ęn \alst{au}gum skoðar, \\%M
\ind svá nýsisk \alst{f}róðra hvęrr \alst{f}yrir.\\%E

\bvb The wary guest who comes to a host, is in attentive silence\footnotemark[13]. With his ears he listens, but with his eyes he observes; thus every experienced man looks out in advance. \\
\footnotetext[13]{Lit. 'with thin heed is silent'}

\bva Hinn es \alst{s}æll, \hld es \alst{s}ér of getr \\%M
\ind \alst{l}of ok \alst{l}íknstafi; \\%M
\alst{ó}dælla es við þat, \hld es \alst{ęi}ga skal \\%M
\ind \alst{a}nnars brjóstum \alst{í}.\\%E

\bvb The one is fortunate, who gets praise and high esteem for \emph{himself}. It is not easy regarding that, which [he] must own in another's breast. \\

\bva \alst{S}á es \alst{s}æll, \hld es \alst{s}jalfr of á \\%M
\ind \alst{l}of ok vit meðan \alst{l}ifir; \\%M
þvíat \alst{i}ll rǫ́ð \hld hęfr maðr \alst{o}pt þęgit \\%M
\ind \alst{a}nnars brjóstum \alst{ó}r.\\%E

\bvb That one is fortunate, who \emph{himself} owns praise and wits while he lives, for man has oft received ill counsel from another's breast. \\

\bva \alst{B}yrði \alst{b}ętri \hld berr-at maðr \alst{b}rautu at, \\%M
\ind an sé \alst{m}anvit \alst{m}ikit; \\%M
\alst{au}ði bętra \hld þykkir þat í \alst{ó}kunnum stað; \\%M
\ind slíkt es \alst{v}álaðs \alst{v}era.\\%E

\bvb No man carries a better burden on the road than much manwit; in a foreign place, it seems better than wealth; such is the refuge of the wretched. \\

\bva \alst{B}yrði \alst{b}ętri \hld bęrr-at maðr \alst{b}rautu at, \\%M
\ind an sé \alst{m}anvit \alst{m}ikit; \\%M
\alst{v}egnest \alst{v}erra \hld \alst{v}egr-a \alst{v}ęlli at, \\%M
\ind an sé \alst{o}fdrykkja \alst{ǫ}ls.\\%E

\bvb No man carries a better burden on the road than much manwit; on the field he can drag no worse provision along than too much ale drunk. \\

\bva Es-a svá \alst{g}ótt, \hld sem \alst{g}ótt kveða, \\%M
\ind \alst{ǫ}l \alst{a}lda sonum; \\%M
þvíat \alst{f}æra vęit, \hld es \alst{f}lęira drekkr, \\%M
\ind síns til \alst{g}ęðs \alst{g}umi.\\%E

\bvb Is not so good, as good they call it, ale for the sons of men. For the more the man drinks, the less he knows of his own conduct. \\

\bva \alst{Ó}minnishegri hęitir, \hld sá's of \alst{ǫ}lðrum þrumir, \\%M
\ind hann stelr \alst{g}ęði \alst{g}uma; \\%M
þess \alst{f}ogls \alst{f}jǫðrum \hld ek \alst{f}jǫtraðr vask \\%M
\ind í \alst{g}arði \alst{G}unnlaðar.\\%E

\bvb A heron of forgetfulness\footnotemark[13] is he called, who thunders above the ale; he robs men of their conduct. With the feathers of that bird I was fettered, in the homestead of Gunnlǫðr. \\
\footnotetext[20]{Lit. 'unmemory'}

\bva \alst{Ǫ}lr ek varð, \hld varð \alst{o}frǫlvi, \\%M
\ind at hins \alst{f}róða \alst{F}jalars; \\%M
því es \alst{ǫ}lðr bazt, \hld at \alst{a}ptr of hęimtir \\%M
\ind hvęrr sitt \alst{g}ęð \alst{g}umi.\\%E

\bvb I became drunk, became too drunk, at the knowing Fealhere's. With that strong drink is best, that each man gets his conduct back. \\

\bva \alst{Þ}agalt ok hugalt \hld skyli \alst{þ}jóðans barn \\%M
\ind ok \alst{v}ígdjarft \alst{v}esa; \\%M
\alst{g}laðr ok ręifr \hld skyli \alst{g}umna hvęrr, \\%M
\ind unz sinn \alst{b}íðr \alst{b}ana.\\%E

\bvb Taciturn and thoughtful, should the prince's child be, and bold in battle. Glad and happy, should each man [be], until he suffers his bane. \\

\bva \alst{Ó}snjallr maðr \hld hyggsk munu \alst{ę}y lifa, \\%M
\ind ef við \alst{v}íg \alst{v}arask; \\%M
ęn \alst{ę}lli gefr \hld hǫ́num \alst{ę}ngi frið, \\%M
\ind þótt hǫ́num \alst{g}ęirar \alst{g}efi.\\%E

\bvb The incapable man, judges [he] will live forever if he avoids battle; but old age gives him no peace, though spears would. \\

\bva \alst{K}ópir afglapi, \hld es til \alst{k}ynnis \alst{k}ømr, \\%M
\ind \alst{þ}ylsk hann umb eða \alst{þ}rumir; \\%M
alt es \alst{s}ęnn, \hld ef \alst{s}ylg of getr, \\%M
\ind uppi es þá \alst{g}ęð \alst{g}uma.\\%E

\bvb 17 \\

\bva Sá ęinn \alst{v}ęit, \hld es \alst{v}íða ratar \\%M
\ind ok hęfr \alst{f}jǫlð of \alst{f}arit, \\%M
hvęrju \alst{g}ęði \hld stýrir \alst{g}umna hvęrr, \\%M
\ind sá es \alst{v}itandi's \alst{v}its.\\%E

\bvb That one alone knows, who wanders widely, and has travelled much: each man controls his [own] conduct\footnotemark[25], who is aware of his wits. \\
\footnotetext[25]{lit. 'each of men controls each conduct'.}

\bva \alst{H}aldi-t maðr á kęri, \hld drekki þó at \alst{h}ófi mjǫð, \\%M
\ind mæli \alst{þ}arft eða \alst{þ}ęgi; \\%M
ókynnis \alst{þ}ess \hld váar \alst{þ}ik ęngi maðr, \\%M
\ind at gangir \alst{s}nimma at \alst{s}ofa.\\%E

\bvb A man should not carouse at the cask\footnotemark[27], but drink a reasonable amount of mead; he ought to speak what is needed, or be silent\footnotemark[26]. TODO thee no man, who might go to sleep early. \\
\footnotetext[26]{Identical to a certain verse in \emph{Web} TODO: elaborate}
\footnotetext[27]{Lit. "should not hold on the cask". The translation as 'carouse' comes from C-V.}

\bva \alst{G}rǫ́ðugr halr, \hld nema \alst{g}ęðs viti, \\%M
\ind \alst{e}tr sér \alst{a}ldrtrega; \\%M
opt fær \alst{h}lǿgis, \hld es með \alst{h}orskum kømr, \\%M
\ind \alst{m}anni hęimskum \alst{m}agi.\\%E

\bvb 20 \\

\bva \alst{H}jarðir þat vitu, \hld nær \alst{h}ęim skulu, \\%M
\ind ok \alst{g}anga þá af \alst{g}rasi; \\%M
ęn \alst{ó}sviðr maðr \hld kann \alst{æ}vagi \\%M
\ind síns of \alst{m}ál \alst{m}aga.\\%E

\bvb 21 \\

\bva \alst{V}esall maðr \hld ok \alst{i}lla skapi \\%M
\ind \alst{h}lær at \alst{h}vívetna; \\%M
hit-ki hann \alst{v}ęit, \hld es \alst{v}ita þyrpti, \\%M
\ind at hann es-a \alst{v}amma \alst{v}anr.\\%E

\bvb 22 \\

\bva \alst{Ó}sviðr maðr \hld vakir of \alst{a}llar nætr \\%M
\ind ok \alst{h}yggr at \alst{h}vívetna; \\%M
þá es \alst{m}óðr, \hld es at \alst{m}orni kømr; \\%M
\ind alt es \alst{v}íl sem \alst{v}as.\\%E

\bvb 23 \\

\bva \alst{Ó}snotr maðr \hld hyggr sér \alst{a}lla vesa \\%M
\ind \alst{v}iðhlæjęndr \alst{v}ini; \\%M
hit-ki hann \alst{f}iðr, \hld þótt þęir of hann \alst{f}ár lesi, \\%M
\ind ef með \alst{s}notrum \alst{s}itr.\\%E

\bvb 24 \\

\bva \alst{Ó}snotr maðr \hld hyggr sér \alst{a}lla vesa \\%M
\ind \alst{v}iðhlæjendr \alst{v}ini; \\%M
\alst{þ}á þat fiðr \hld es at \alst{þ}ingi kømr, \\%M
\ind at á \alst{f}ormælęndr \alst{f}áa.\\%E

\bvb 25 \\

\bva \alst{Ó}snotr maðr \hld þykkisk \alst{a}lt vita, \\%M
\ind ef á sér i \alst{v}ǫ́ \alst{v}eru; \\%M
hit-ki hann \alst{v}ęit, \hld hvat hann skal \alst{v}ið kveða, \\%M
\ind ef hans \alst{f}ręista \alst{f}irar.\\%E

\bvb 26 \\

\bva \alst{Ó}snotr maðr, \hld es með \alst{a}ldir kømr, \\%M
\ind \alst{þ}at's bazt at hann \alst{þ}ęgi; \\%M
\alst{ę}ngi þat vęit, \hld at hann \alst{ę}kki kann, \\%M
\ind nema hann \alst{m}æli til \alst{m}art. \\%M
\alst{v}ęita maðr, \hld hinn's \alst{v}æt-ki vęit, \\%M
\ind þótt hann \alst{m}æli til \alst{m}art.\\%E

\bvb 27 \\

\bva \alst{F}róðr sá þykkisk, \hld es \alst{f}regna kann, \\%M
\ind ok \alst{s}ęgja hit \alst{s}ama, \\%M
\alst{ęy}vitu lęyna \hld męgu \alst{ý}ta synir \\%M
\ind því es \alst{g}ęngr of \alst{g}uma.\\%E

\bvb 28 \\

\bva \alst{Ǿ}rna mælir, \hld sá's \alst{æ}va þęgir, \\%M
\ind \alst{st}aðlausu \alst{st}afi; \\%M
\alst{h}raðmælt tunga, \hld nema \alst{h}aldęndr ęigi, \\%M
\ind opt sér ó\alst{g}ótt of \alst{g}ęlr.\\%E

\bvb 29 \\

\bva At \alst{au}gabragði \hld skal-a maðr \alst{a}nnan hafa, \\%M
\ind þótt til \alst{k}ynnis \alst{k}omi; \\%M
margr \alst{f}róðr þykkisk, \hld ef hann \alst{f}reginn es-at \\%M
\ind ok nái hann \alst{þ}urrfjallr \alst{þ}ruma.\\%E

\bvb 30 \\

\bva \alst{F}róðr þykkisk \hld sá's \alst{f}lótta tękr \\%M
\ind \alst{g}ęstr at \alst{g}ęst hęðinn; \\%M
\alst{v}ęita gǫrla \hld sá's of \alst{v}erði glissir, \\%M
\ind þótt með \alst{g}rǫmum \alst{g}lami.\\%E

\bvb 31 \\

\bva \alst{G}umnar margir \hld erusk \alst{g}agnhollir, \\%M
\ind ęn at \alst{v}irði \alst{v}rekask; \\%M
\alst{a}ldar róg \hld þat mun \alst{æ} vesa; \\%M
\ind órir \alst{g}ęstr við \alst{g}ęst.\\%E

\bvb 32 \\

\bva \alst{Á}rliga verðar \hld skyli-t maðr \alst{o}pt fáa, \\%M
\ind nema til \alst{k}ynnis \alst{k}omi; \\%M
\alst{s}itr ok \alst{s}nópir, \hld lætr sem \alst{s}olginn sé, \\%M
\ind ok kann \alst{f}regna at \alst{f}ǫ́u.\\%E

\bvb 33 \\

\bva \alst{A}fhvarf mikit \hld es til \alst{i}lls vinar, \\%M
\ind þótt á \alst{b}rautu \alst{b}úi, \\%M
ęn til \alst{g}óðs vinar \hld liggja \alst{g}agnvegir, \\%M
\ind þótt hann sé \alst{f}irr \alst{f}arinn.\\%E

\bvb A great detour is to an ill friend, though he might live on the highway; but to a good friend lie short-cuts, though he might be far gone. \\

\bva \alst{G}anga skal, \hld skal-a \alst{g}ęstr vesa \\%M
\ind \alst{ęy} í \alst{ęi}num stað; \\%M
\alst{l}júfr verðr \alst{l}ęiðr, \hld ef \alst{l}ęngi sitr \\%M
\ind \alst{a}nnars flętjum \alst{á}.\\%E

\bvb 35 \\

\bva \alst{B}ú es \alst{b}ętra, \hld an \alst{b}iðja sé, \\%M
\ind \alst{h}alr es \alst{h}ęima \alst{h}vęrr; \\%M
þótt \alst{t}vær gęitr ęigi \hld ok \alst{t}augręptan sal, \\%M
\ind þat es þó \alst{b}ętra an \alst{b}ǿn.\\%E

\bvb 36 \\

\bva \alst{B}ú es \alst{b}ętra, \hld an \alst{b}iðja sé, \\%M
\ind \alst{h}alr es \alst{h}ęima \alst{h}vęrr; \\%M
\alst{b}lóðugt es hjarta \hld þęim's \alst{b}iðja skal \\%M
\ind sér í \alst{m}ál hvęrt \alst{m}atar.\\%E

\bvb 37 \\

\bva \alst{V}ǫ́pnum sínum \hld skal-a maðr \alst{v}ęlli á \\%M
\ind \alst{f}eti ganga \alst{f}ramar; \\%M
ó\alst{v}íst's at \alst{v}ita, \hld nær verðr á \alst{v}egum úti \\%M
\ind \alst{g}ęirs of þǫrf \alst{g}uma.\\%E

\bvb 38 \\

\bva Fann-k-a \alst{m}ildan mann \hld eða svá \alst{m}atar góðan, \\%M
\ind at væri-t \alst{þ}iggja \alst{þ}ęgit; \\%M
eða \alst{s}íns féar \hld \alst{s}vági gløggvan, \\%M
\ind at \alst{l}ęið sé \alst{l}aun, ef þiggr.\\%E

\bvb I found not a mild man, nor [one] so good of meat, TODO \\

\bva \alst{F}éar síns, \hld es \alst{f}ęngit hęfr, \\%M
\ind skyli-t maðr \alst{þ}ǫrf \alst{þ}ola; \\%M
opt sparir \alst{l}ęiðum \hld þat's hęfr \alst{l}júfum hugat; \\%M
\ind mart gęngr \alst{v}err an \alst{v}arir.\\%E

\bvb 40 \\

\bva \alst{V}ǫ́pnum ok \alst{v}ǫ́ðum \hld skulu \alst{v}inir glęðjask; \\%M
\ind þat's á \alst{s}jǫ́lfum \alst{s}ýnst; \\%M
\alst{v}iðrgefęndr \hld erusk \alst{v}inir lęngst, \\%M
\ind ef þat bíðr at \alst{v}erða \alst{v}ęl.\\%E

\bvb 41 \\

\bva \alst{V}in sínum \hld skal maðr \alst{v}inr \alst{v}esa, \\%M
\ind ok \alst{g}jalda \alst{g}jǫf við \alst{g}jǫf; \\%M
\alst{h}látr við \alst{h}látri \hld skyli \alst{h}ǫlðar taka, \\%M
\ind en \alst{l}ausung við \alst{l}ygi.\\%E

\bvb 42 \\

\bva \alst{V}in sínum \hld skal maðr \alst{v}inr vesa, \\%M
\ind \alst{þ}ęim ok \alst{þ}ess vin; \\%M
en \alst{ó}vinar síns \hld skyli \alst{ę}ngi maðr \\%M
\ind \alst{v}inar \alst{v}inr \alst{v}esa.\\%E

\bvb 43 \\

\bva \alst{V}ęizt, ef þú \alst{v}in átt, \hld þann's þú \alst{v}ęl trúir \\%M
\ind ok vilt af hǫ́num \alst{g}ótt \alst{g}eta, \\%M
\alst{g}ęði skalt við þann \hld ok \alst{g}jǫfum skipta, \\%M
\ind \alst{f}ara at \alst{f}inna opt.\\%E

\bvb Know, if thou hast a friend, which thou trust well, and wilt of him receive good: mind shalt thou with exchange him, and gifts; travel to see him oft. \\

\bva Ef þú \alst{á}tt \alst{a}nnan, \hld þann's þú \alst{i}lla trúir, \\%M
\ind vildu af hǫ́num þó \alst{g}ótt \alst{g}eta, \\%M
\alst{f}agrt skalt mæla, \hld ęn \alst{f}látt hyggja \\%M
\ind ok gjalda \alst{l}ausung við \alst{l}ygi.\\%E

\bvb If thou hast another, which thou trust little, thou wilt of him receive good; fairly shalt thou speak, but deceitfully think, and repay duplicity with lies. \\

\bva Þat's \alst{ę}nn of þann, \hld es þú \alst{i}lla trúir \\%M
\ind ok þér es \alst{g}runr at \alst{g}ęði, \\%M
\alst{h}læja skalt við þęim \hld ok of \alst{h}ug mæla; \\%M
\ind glík skulu \alst{g}jǫld \alst{g}jǫfum.\\%E

\bvb 46 \\

\bva Ungr vas'k \alst{f}orðum, \hld \alst{f}ór'k ęinn saman, \\%M
\ind þá varð'k \alst{v}illr \alst{v}ega; \\%M
\alst{au}ðigr þóttumk, \hld es \alst{a}nnan fann'k, \\%M
\ind \alst{m}aðr es \alst{m}anns gaman.\\%E

\bvb Young I was once, I travelled all alone; then I became astray of the paths. Wealthy I thought myself when I another found; man is the joy of man. \\

\bva \alst{M}ildir frǿknir \hld \alst{m}ęnn bazt lifa, \\%M
\ind \alst{s}jaldan \alst{s}út ala; \\%M
\alst{ó}snjallr maðr \hld \alst{u}ggir hvatvetna, \\%M
\ind sýtir æ \alst{g}løggr við \alst{g}jǫfum.\\%E

\bvb 48 \\

\bva \alst{V}áðir mínar \hld gaf'k \alst{v}ęlli at \\%M
\ind \alst{t}vęim \alst{t}rémǫnnum; \\%M
\alst{r}ekkar þat þóttusk, \hld es \alst{r}ipt hǫfðu; \\%M
\ind \alst{n}ęiss es \alst{n}ǫkkviðr halr.\\%E

\bvb My garments I gave at the field, to two tree-men. They seemed to be champions, when they cloaks had; ashamed is the naked man. \\

\bva Hrørnar \alst{þ}ǫll, \hld sú's stęndr \alst{þ}orpi á, \\%M
\ind hlýrat hęnni \alst{b}ǫrkr né \alst{b}arr; \\%M
svá es \alst{m}aðr, \hld sá's \alst{m}anngi ann; \\%M
\ind hvat skal hann \alst{l}ęngi \alst{l}ifa?\\%E

\bvb Wilters the pine, which stands in the village; her shields not bark nor needle; such is the man, who loves nobody; for what shall he live long? \\

\bva \alst{Ę}ldi hęitari \hld brinnr með \alst{i}llum vinum \\%M
\ind \alst{f}riðr \alst{f}imm daga, \\%M
ęn þá \alst{sl}oknar, \hld es hinn \alst{s}étti kømr, \\%M
\ind ok \alst{v}ersnar allr \alst{v}inskapr.\\%E

\bvb Frith burns for five days hotter than fire, with bad friends; but it then goes out, when the sixth comes, and all the friendship worsens. \\

\bva \alst{M}ikit ęitt \hld skal-a \alst{m}anni gefa; \\%M
\ind opt kaupir sér í \alst{l}ítlu \alst{l}of, \\%M
með \alst{h}ǫlfum \alst{h}lęif \hld ok með \alst{h}ǫllu kęri \\%M
\ind \alst{f}ekk ek mér \alst{f}élaga.\\%E

\bvb Much of one [thing], one should not give to a man; often one earns praise for little. With a half loaf [of bread] and an awry vat, I earned me a companion. \\
\footnotetext[x]{Lit. “often [one] buys oneself in little praise"}

\bva \alst{L}ítilla sanda, \hld \alst{l}ítilla sæva, \\%M
\ind lítil eru \alst{g}ęð \alst{g}uma; \\%M
\alst{a}llir męnn \hld \alst{u}rðu-t jafnspakir; \\%M
\ind \alst{h}ǫlf es ǫld \alst{h}var.\\%E

\bvb Of little sands, of little seas; little are the minds of men; all men do not become equal-wise; half is each man. \\

\bva \alst{M}eðalsnotr \hld skyli \alst{m}anna hvęrr, \\%M
\ind æva til \alst{s}notr \alst{s}éi; \\%M
þęim es \alst{f}yrða \hld \alst{f}ęgrst at lifa, \\%M
\ind es \alst{v}ęl mart \alst{v}itu.\\%E

\bvb 55 \\

\bva \alst{M}eðalsnotr \hld skyli \alst{m}anna hvęrr, \\%M
\ind æva til \alst{s}notr \alst{s}éi; \\%M
\alst{s}notrs manns hjarta \hld verðr \alst{s}jaldan glatt, \\%M
\ind ef sá's \alst{a}lsnotr es \alst{á}.\\%E

\bvb 56 \\

\bva \alst{M}eðalsnotr \hld skyli \alst{m}anna hvęrr, \\%M
\ind æva til \alst{s}notr \alst{s}éi; \\%M
\alst{ø}rlǫg sin \hld viti \alst{ę}ngi fyrir; \\%M
\ind þęim es \alst{s}orgalausastr \alst{s}efi.\\%E

\bvb 57 \\

\bva \alst{B}randr af \alst{b}randi \hld \alst{b}rinnr unz \alst{b}runninn es, \\%M
\ind \alst{f}uni kvęykisk af \alst{f}una; \\%M
\alst{m}aðr af \alst{m}anni \hld verðr at \alst{m}áli kuðr; \\%M
\ind ęn til \alst{d}ǿlskr af \alst{d}ul.\\%E

\bvb Flame by flame burns, until it burnt is; fire is kindled by fire. Man by man becomes known by speech, but the fool by his delusions. \\

\bva \alst{Á}r skal rísa, \hld sá's \alst{a}nnars vill \\%M
\ind \alst{f}é eða \alst{f}jǫr hafa; \\%M
\alst{l}iggjandi ulfr \hld \alst{l}ær of getrat, \\%M
\ind né \alst{s}ofandi maðr \alst{s}igr.\\%E

\bvb 58 \\

\bva \alst{Á}r skal rísa, \hld sá's á \alst{y}rkjęndr fáa, \\%M
\ind ok ganga síns \alst{v}erka á \alst{v}it; \\%M
\alst{m}art of dvęlr \hld þann's of \alst{m}orgin sefr, \\%M
\ind \alst{h}alfr es auðr und \alst{h}vǫtum.\\%E

\bvb 59 \\

\bva \alst{Þ}urra skíða \hld ok \alst{þ}akinna næfra, \\%M
\ind þess kann \alst{m}aðr \alst{m}jǫt, \\%M
þess \alst{v}iðar, \hld es \alst{v}innask męgi \\%M
\ind \alst{m}ál ok \alst{m}issęri.\\%E

\bvb 60 \\

\bva Þvęginn ok \alst{m}ęttr \hld ríði \alst{m}aðr þingi at, \\%M
\ind þótt sét \alst{v}æddr til \alst{v}ęl; \\%M
\alst{sk}úa ok bróka \hld \alst{sk}ammisk ęngi maðr \\%M
\ind né \alst{h}ęsts in \alst{h}ęldr. \\%M
\ind (þótt \alst{h}ann \alst{h}afi't góðan).\\%E

\bvb 61 \\

\bva \alst{S}napir ok gnapir, \hld es til \alst{s}ævar kømr, \\%M
\ind \alst{ǫ}rn á \alst{a}ldinn mar; \\%M
svá es \alst{m}aðr, \hld es með \alst{m}ǫrgum kømr \\%M
\ind ok á \alst{f}ormælęndr \alst{f}áa.\\%E

\bvb 62 \\

\bva \alst{F}regna ok sęgja \hld skal \alst{f}róðra hvęrr, \\%M
\ind sá's vill \alst{h}ęitinn \alst{h}orskr; \\%M
\alst{ęi}nn vita \hld né \alst{a}nnarr skal, \\%M
\ind \alst{þ}jóð vęit ef \alst{þ}rír eru.\\%E

\bvb Ask and reply, should each wise man\footnotemark[70] who wants to be called prudent. One may know, but not another; the people knows if there are three. \\
\footnotetext[70]{each of the wise}

\bva \alst{R}íki sitt \hld skyli \alst{r}áðsnotra \\%M
\ind \alst{h}vęrr í \alst{h}ófi \alst{h}afa; \\%M
þá hann þat \alst{f}innr, \hld es með \alst{f}rǿknum kømr, \\%M
\ind at \alst{ę}ngi es \alst{ęi}nna hvatastr. \\%M
\alst{O}rða þęira, \hld es maðr \alst{ǫ}ðrum sęgir, \\%M
\ind opt hann \alst{g}jǫld of \alst{g}etr.\\%E

\bvb 64 \\

\bva \alst{M}ikilsti snimma \hld kom'k í \alst{m}arga staði, \\%M
\ind ęn til \alst{s}íð í \alst{s}uma; \\%M
\alst{ǫ}l vas drukkit, \hld sumt vas \alst{ó}lagat; \\%M
\ind sjaldan hittir \alst{l}ęiðr í \alst{l}ið.\\%E

\bvb 65 \\

\bva \alst{H}ér ok \alst{h}var \hld myndi mér \alst{h}ęim of boðit, \\%M
\ind ef þyrpta'k at \alst{m}ǫ́lungi \alst{m}at, \\%M
eða \alst{t}vau lær hęngi \hld at hins \alst{t}ryggva vinar, \\%M
\ind þar's ek hafða \alst{ęi}tt \alst{e}tit.\\%E

\bvb 66 \\

\bva \alst{Ę}ldr es baztr \hld með \alst{ý}ta sonum \\%M
\ind ok \alst{s}ólar \alst{s}ýn, \\%M
\alst{h}ęilyndi sitt, \hld ef \alst{h}afa náir, \\%M
\ind án við \alst{l}ǫst at \alst{l}ifa.\\%E

\bvb Fire is best among the sons of men, and the sight of the sun; one's good health, if [he] manage to have it, without living by sin. \\

\bva Es-at maðr \alst{a}lls vesall, \hld þótt sé \alst{i}lla hęill, \\%M
\ind \alst{s}umr es af \alst{s}onum \alst{s}æll, \\%M
\alst{s}umr af frændum, \hld \alst{s}umr af fé ǿrnu, \\%M
\ind sumr af \alst{v}erkum \alst{v}ęl.\\%E

\bvb One is not wholly wretched, though health might be poor; some are [made] fortunate by sons, some by kinsmen, some by ample wealth, some by work done well. \\

\bva Bętra es \alst{l}ifðum, \hld an \alst{l}iðnum sé, \\%M
\ind ęy getr \alst{k}vikr \alst{k}ú; \\%M
\alst{ę}ld sá'k brinna \hld \alst{au}ðgum manni fyrir, \\%M
\ind ęn úti vas \alst{d}auðr fyr \alst{d}urum.\\%E

\bvb 69 \\

\bva \alst{H}altr ríðr \alst{h}rossi, \hld \alst{h}jǫrð rekr \alst{h}andarvanr, \\%M
\ind daufr \alst{v}egr ok \alst{d}ugir; \\%M
\alst{b}lindr es \alst{b}ętri, \hld an \alst{b}ręndr séi; \\%M
\ind \alst{n}ýtr mangi \alst{n}ás.\\%E

\bvb The halt rides a horse, the handless can drive a herd, the deaf strikes and is sufficient. Blind is better, than be burned; no man has use for a corpse. \\

\bva \alst{S}onr es bętri, \hld þótt sé \alst{s}íð of alinn \\%M
\ind ępt \alst{g}inginn \alst{g}uma; \\%M
sjaldan \alst{b}autarstęinar \hld standa \alst{b}rautu nær, \\%M
\ind nema ręisi \alst{n}iðr at \alst{n}ið.\\%E

\bvb A son is better, though he might be born late, walking after a man. Seldom large stones stand next to the road, unless raised by kinsman for kinsman. \\

\bva \alst{T}vęir 'ro ęins hęrjar, \hld \alst{t}unga es hǫfuðs bani; \\%M
\ind mér's í \alst{h}eðin \alst{h}vęrn \hld \alst{h}andar væni.\\%E

\bvb 72 \\

\bva \alst{N}ótt verðr fęginn, \hld sá's \alst{n}esti trúir, \\%M
\ind \alst{sk}ammar 'ro \alst{sk}ips ráar. \\%M
– – – – \\%M
\ind \alst{H}verf es \alst{h}austgríma, \\%M
\alst{f}jǫlð of viðrir \hld á \alst{f}imm dǫgum, \\%M
\ind en \alst{m}ęir á \alst{m}ánaði.\\%E

\bvb 73 \\

\bva \alst{V}ęit-a hinn, \hld es \alst{v}æt-ki \alst{v}ęit, \\%M
\ind margr verðr af \alst{au}rum \alst{a}pi; \\%M
maðr es \alst{au}ðigr, \hld annarr \alst{ó}auðigr, \\%M
\ind skyli-t þann \alst{v}ítka \alst{v}áar.\\%E

\bvb 74 \\

\bva \alst{D}ęyr fé, \hld \alst{d}ęyja frændr, \\%M
\ind dęyr \alst{s}jalfr hit \alst{s}ama; \\%M
ęn \alst{o}rðstírr \hld dęyr \alst{a}ldrigi \\%M
\ind hvęim's sér \alst{g}óðan \alst{g}etr.\\%E

\bvb Wealth dies, kinsmen die, the self dies the same; but a word-reputation never dies, for he who gets a good one. \\

\bva \alst{D}ęyr fé, \hld \alst{d}ęyja frændr, \\%M
\ind dęyr \alst{s}jalfr hit \alst{s}ama; \\%M
\alst{e}k vęit \alst{ęi}nn \hld at \alst{a}ldri dęyr: \\%M
\ind \alst{d}ómr of \alst{d}auðan hvęrn.\\%E

\bvb Wealth dies, kinsmen die, the self dies the same; I know \emph{one}, which never dies: the judgement over each dead. \\

\bva \alst{F}ullar grindr \hld sá'k fyr \alst{F}itjungs sonum, \\%M
\ind nú bera þęir \alst{v}ánar \alst{v}ǫl; \\%M
svá es \alst{au}ðr \hld sęm \alst{au}gabragð, \\%M
\ind hann es \alst{v}altastr \alst{v}ina. —\\%E

\bvb 77 \\

\bva \alst{Ó}snotr maðr, \hld es \alst{ęi}gnask getr \\%M
\ind \alst{f}é eða \alst{f}ljóðs munuð; \\%M
\alst{m}etnaðr hǫ́num þróask, \hld ęn \alst{m}anvit aldrigi; \\%M
\ind framm gęngr hann \alst{d}rjúgt í \alst{d}ul. —\\%E

\bvb 78 \\

\bva Þat es þá \alst{r}ęynt, \hld es þú at \alst{r}únum spyrr \hld hinum \alst{r}ęginkunnum, \\%M
\ind þęim's \alst{g}ęrðu \alst{g}innręgin \\%M
\ind ok \alst{f}áði \alst{f}imbulþulr; \\%M
\ind þá hęfr hann bazt, ef þęgir. —\\%E

\bvb Then that is proven of which thou inquires the runes, the ones born of the Powers, those which the yin-Powers made, and the fimbol-thyle painted; then he has it best, if he shuts up. \\

\bva At \alst{k}veldi skal dag lęyfa, \hld \alst{k}onu es bręnd es, \\%M
\alst{m}æki es ręyndr es, \hld \alst{m}ęy es gefin es, \\%M
\alst{í}s es \alst{y}fir kømr, \hld \alst{ǫ}l es drukkit es. —\\%E

\bvb At evening [one] shall praise day, a woman when [she] is burned, a sword when [it] is tried, a maiden when [she] is given. Ice when [one] comes over, ale when [it] is drunk. \\

\bva Í \alst{v}indi skal \alst{v}ið hǫggva, \hld \alst{v}eðri á sæ róa, \\%M
\alst{m}yrkri við \alst{m}an spjalla, \hld \alst{m}ǫrg eru dags augu, \\%M
á \alst{sk}ip skal \alst{sk}riðar orka, \hld ęn á \alst{sk}jǫld til hlífar, \\%M
\alst{m}æki til hǫggs, \hld ęn \alst{m}ęy til kossa. —\\%E

\bvb In wind [one] shall cut wood, [in] storm row on the sea, [in] darkness meet with a maid; many are the eyes of day. A ship [one] shall use for its speed, a shield for shelter, a sword for striking, but a maiden for [her] kisses. \\

\bva Við \alst{e}ld skal \alst{ǫ}l drekka. \hld ęn á \alst{í}si skríða, \\%M
\alst{m}agran \alst{m}ar kaupa, \hld ęn \alst{m}æki saurgan, \\%M
\alst{h}ęima \alst{h}ęst fęita, \hld ęn \alst{h}und á búi. \\%E

\bvb 82 \\

	Regarding the love of women, and Woden's failed love-adventures.

\bva \alst{M}ęyjar orðum \hld skyli \alst{m}angi trúa, \\%M
\ind né því's \alst{k}veðr \alst{k}ona; \\%M
á \alst{h}verfanda \alst{h}véli \hld vǫ́ru þęim \alst{h}jǫrtu skǫpuð, \\%M
\ind \alst{b}rigð í \alst{b}rjóst of lagið\footnotemark[29]. \\%E
\footnotetext[29]{The second half of this st. is quoted in \emph{Fbr}, Thott 1768 4°\textsuperscript{x}, ff.: “And he then remembered the short verse, which had been composed about unfaithful women:"}

\bvb The words of a maiden should no man believe, nor that which a woman says. On a spinning wheel were their hearts for them shaped; betrayal laid in their breasts. \\

\bva \alst{B}restanda \alst{b}oga, \hld \alst{b}rinnanda loga, \\%M
\alst{g}ínanda ulfi, \hld \alst{g}alandi krǫ́ku, \\%M
\alst{r}ýtanda svíni, \hld \alst{r}ótlausum viði, \\%M
\alst{v}axanda \alst{v}ági, \hld \alst{v}ellanda katli,\\%E

\bvb Bursting bow, burning flame, gaping wolf, crowing crow, roaring swine, rootless tree, waxing wave, swelling kettle, \\

\bva \alst{f}ljúganda \alst{f}lęini, \hld \alst{f}allandi bǫ́ru, \\%M
\alst{í}si \alst{ęi}nnættum, \hld \alst{o}rmi hringlęgnum, \\%M
\alst{b}rúðar \alst{b}ęðmǫ́lum \hld eða \alst{b}rotnu sverði, \\%M
\alst{b}jarnar lęiki \hld eða \alst{b}arni konungs,
\alst{s}júkum kalfi, \hld \alst{s}jalfráða þræli, \\%M
\alst{v}ǫlu \alst{v}ilmæli, \hld \alst{v}al nýfęldum. —\\%E

\bvb 85 \\

\bva \alst{A}kri \alst{á}rsǫ́num \hld trúi \alst{ę}ngi maðr, \\%M
\ind né til \alst{s}nimma \alst{s}yni; \\%M
\alst{v}eðr ræðr akri, \hld ęn \alst{v}it syni; \\%M
\ind \alst{h}ætt es þęira \alst{h}várt. —\\%E

\bvb 86 \\

\bva \alst{B}róður\alst{b}ana sínum \hld þótt á \alst{b}rautu mǿti, \\%M
\alst{h}úsi \alst{h}alfbrunnu, \hld \alst{h}ęsti alskjótum, \\%M
þá's \alst{jó}r \alst{ó}nýtr, \hld ef \alst{ęi}nn fótr brotnar; \\%M
verðit maðr svá \alst{t}ryggr \hld at þessu \alst{t}rúi ǫllu. —\\%E

\bvb 87 \\

\bva Svá's \alst{f}riðr kvinna \hld þęira's \alst{f}látt hyggja, \\%M
sęm aki \alst{jó} óbryddum \hld á \alst{í}si hǫ́lum \\%M
(\alst{t}ęitum, \alst{t}vévetrum \hld ok sé \alst{t}amr illa), \\%M
eða í \alst{b}yr óðum \hld \alst{b}ęiti stjórnlausu, \\%M
eða skyli \alst{h}altr \alst{h}ęnda \hld \alst{h}ręin í þáfjalli. —\\%E

\bvb 88 \\

\bva \alst{B}ert nú mæli'k, \hld því-at \alst{b}æði vęit'k, \\%M
\ind brigðr es \alst{k}arla hugr \alst{k}onum, \\%M
þá \alst{f}ęgrst mælum, \hld es \alst{f}lást hyggjum; \\%M
\ind þat tælir \alst{h}orska \alst{h}ugi.\\%E

\bvb 89 \\

\bva \alst{F}agrt skal mæla \hld ok \alst{f}é bjóða, \\%M
\ind sá's vill \alst{f}ljóðs ǫ́st \alst{f}áa, \\%M
\alst{l}íki \alst{l}ęyfa \hld hins \alst{l}jósa mans, \\%M
\ind sá \alst{f}ær, es \alst{f}ríar.\\%E

\bvb 90 \\

\bva \alst{Á}star firna \hld skyli \alst{ę}ngi maðr \\%M
\ind \alst{a}nnan \alst{a}ldrigi; \\%M
opt fáa á \alst{h}orskan, \hld es á \alst{h}ęimskan né fáa, \\%M
\ind \alst{l}ostfagrir \alst{l}itir.\\%E

\bvb 91 \\

\bva \alst{Ęy}vitar firna, \hld es maðr \alst{a}nnan skal, \\%M
\ind þess's of margan \alst{g}ęngr \alst{g}uma; \\%M
\alst{h}ęimska ór \alst{h}orskum \hld gęrir \alst{h}ǫlða sonu \\%M
\ind sá hinn \alst{m}átki \alst{m}unr.\\%E

\bvb 92 \\

\bva \alst{H}ugr ęinn þat vęit, \hld es býr \alst{h}jarta nær, \\%M
\ind ęinn es hann \alst{s}ér of \alst{s}efa; \\%M
øng es \alst{s}ótt verri \hld hvęim \alst{s}notrum manni \\%M
\ind an sér \alst{ø}ngu at \alst{u}na.\\%E

\bvb 93 \\

\bva Þat þá \alst{r}ęyndak, \hld es í \alst{r}ęyri sat'k, \\%M
\ind ok vætta'k \alst{m}íns \alst{m}unar, \\%M
\alst{h}old ok \alst{h}jarta \hld vas mér hin \alst{h}orska mær, \\%M
\ind þęygi hana at \alst{h}ęldr \alst{h}ęf'k.\\%E

\bvb 94 \\

\bva \alst{B}illings męy \hld ek fann \alst{b}ęðjum á \\%M
\ind \alst{s}ólhvíta \alst{s}ofa; \\%M
\alst{ja}rls \alst{y}nði \hld þótti mér \alst{ę}kki vesa \\%M
\ind nema við þat \alst{l}ík at \alst{l}ifa.\\%E

\bvb 95 \\

\bva “\alst{Au}k nær \alst{a}ptni \hld skaltu \alst{Ó}ðinn koma, \\%M
\ind ef vilt þér \alst{m}æla \alst{m}an, \\%M
\alst{a}lt eru \alst{ó}skǫp, \hld nema \alst{ęi}n vitim \\%M
\ind \alst{s}likan lǫst \alst{s}aman.”\\%E

\bvb 96 \\

\bva \alst{A}ptr ek hvarf \hld ok \alst{u}nna þóttumk \\%M
\ind \alst{v}ísum \alst{v}ilja frá; \\%M
\alst{h}itt ek \alst{h}ugða, \hld at \alst{h}afa mynda'k \\%M
\ind \alst{g}ęð hennar alt ok \alst{g}aman.\\%E

\bvb 97 \\

\bva Svá kom'k \alst{n}æst, \hld at hin \alst{n}ýta vas \\%M
\ind \alst{v}ígdrótt ǫll of \alst{v}akin; \\%M
með \alst{b}rinnǫndum ljósum \hld ok \alst{b}ornum viði, \\%M
\ind svá vas mér \alst{v}ílstígr of \alst{v}itaðr.\\%E

\bvb 98 \\

\bva \alst{Au}k nær morni, \hld es vas'k \alst{ę}nn of kominn, \\%M
\ind þá vas \alst{s}aldrótt of \alst{s}ofin; \\%M
\alst{g}ręy ęitt þá fan'k \hld hinnar \alst{g}óðu konu \\%M
\ind \alst{b}undit \alst{b}ęðjum á.\\%E

\bvb 99 \\

\bva Mǫrg es \alst{g}óð mær, \hld ef \alst{g}ǫrva kannar, \\%M
\ind \alst{h}ugbrigð við \alst{h}ali; \\%M
þá þat \alst{r}ęynda'k, \hld es hit \alst{r}áðspaka \\%M
\ind tęygða'k á \alst{f}lærðir \alst{f}ljóð. \\%M
\alst{h}ǫ́ðungar \alst{h}vęrrar \hld lęitaði mér hit \alst{h}orska man \\%M
\ind ok hafða'k þess \alst{v}æt-ki \alst{v}ífs.\\%E

\bvb 100 \\

	Side-composition to the previous poem, starting with a general maxim.

\bva Hęima \alst{g}laðr \hld ok við \alst{g}ęsti ręifr, \\%M
\ind \alst{s}viðr skal of \alst{s}ik vesa; \\%M
\alst{m}innigr ok \alst{m}ǫ́lugr, \hld ef vill \alst{m}argfróðr vesa; \\%M
\ind opt skal \alst{g}óðs \alst{g}eta; \\%M
\alst{f}imbul\alst{f}ambi hęitir, \hld sás \alst{f}átt kann sęgja; \\%M
\ind þat es \alst{ó}snotrs \alst{a}ðal.\\%E

\bvb 101 \\

\bva Hinn \alst{a}ldna \alst{jǫ}tun sóttak, \hld nú em'k \alst{a}ptr of kominn; \\%M
\ind fátt gat'k \alst{þ}ęgjandi \alst{þ}ar; \\%M
\alst{m}ǫrgum orðum \hld \alst{m}ælta'k í minn frama \\%M
\ind í \alst{S}uttungs \alst{s}ǫlum.\\%E

\bvb The old ettin I sought, now am I come back; I got little silence there. Many words I spoke to my fame, in the halls of Suttung. \\

\bva \alst{G}unnlǫð mér of \alst{g}af \hld \alst{g}ollnum stóli á \\%M
\ind \alst{d}rykk hins \alst{d}ýra mjaðar; \\%M
\alst{i}ll \alst{i}ðgjǫld \hld lét'k hana \alst{ę}ptir hafa \\%M
\ind síns hins \alst{h}ęila \alst{h}ugar. \\%M
\ind (síns hins \alst{s}vára \alst{s}efa).\\%E

\bvb 103 \\

\bva \alst{R}ata munn \hld létumk \alst{r}úms of fáa \\%M
\ind ok of \alst{g}rjót \alst{g}naga; \\%M
\alst{y}fir ok \alst{u}ndir \hld stóðumk \alst{jǫ}tna vegir, \\%M
\ind svá \alst{h}ættak \alst{h}ǫfði til.\\%E

\bvb 104 \\

\bva \alst{V}ęl kęypts hlutar \hld hęf'k \alst{v}ęl notit; \\%M
\ind \alst{f}ás es \alst{f}róðum vant; \\%M
\alst{Ó}ðrerir \hld nú \alst{u}pp's kominn \\%M
\ind á \alst{a}lda vé \alst{ja}ðars.\\%E

\bvb 105 \\

\bva \alst{I}fi es mér á, \hld at væra'k \alst{ę}nn kominn \\%M
\ind \alst{jǫ}tna gǫrðum \alst{ó}r, \\%M
ef \alst{G}unnlaðar né nyta'k, \hld hinnar \alst{g}óðu konu, \\%M
\ind es lǫgðumk \alst{a}rm \alst{y}fir.\\%E

\bvb 106 \\

\bva Hins \alst{h}indra dags \hld gingu \alst{h}rímþursar \\%M
\ind (Háva ráðs at fregna) \\%M
\ind \alst{H}áva \alst{h}ǫllu í, \\%M
at \alst{B}ǫlverki spurðu, \hld ef væri með \alst{b}ǫndum kominn \\%M
\ind eða hęfði hǫ́num \alst{S}uttungr of \alst{s}óit.\\%E

\bvb 107 \\

\bva Baugęið \alst{Ó}ðinn \hld hygg at \alst{u}nnit hafi, \\%M
\ind hvat skal hans \alst{t}ryggðum \alst{t}rúa? \\%M
\alst{S}uttung \alst{s}vikvinn \hld hann lét \alst{s}umbli frá \\%M
\ind ok \alst{g}rǿtta \alst{G}unnlǫðu.\\%E

\bvb 108 \\

	Advice of the Fimbol-Thyle, given to Loddfáfnir.

\bva Mál's at \alst{þ}ylja \hld \alst{þ}ular stóli á, \\%M
\ind \alst{U}rðar brunni \alst{a}t \\%M
\alst{s}á'k ok þagða'k, \hld \alst{s}á'k ok hugða'k, \\%M
\ind hlýdda'k á \alst{m}anna \alst{m}ál. \\%M
of \alst{r}únar hęyrða'k dǿma, \hld né um \alst{r}ǫ́ðum þǫgðu \\%M
\ind \alst{H}áva \alst{h}ǫllu at, \\%M
\ind \alst{H}áva \alst{h}ǫllu í \\%M
\ind hęyrða'k \alst{s}ęgja \alst{s}vá:\\%E

\bvb 109 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jǫta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
\alst{n}ótt þú rís-at, \hld nema á \alst{n}jósn séir, \\%M
\ind eða lęitir þér \alst{i}nnan \alst{ú}t staðar.\\%E

\bvb 110 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
\alst{f}jǫlkunnigri konu \hld skal-at-tu í \alst{f}aðmi sofa, \\%M
\ind svá at hon \alst{l}yki þik \alst{l}iðum. \\%M
Hón svá \alst{g}ęrir \hld at þú \alst{g}áir ęigi \\%M
\ind \alst{þ}ings né \alst{þ}jóðans máls \\%M
\alst{m}at þú vill-at \hld né \alst{m}ans-kis gaman \\%M
\ind fęrr þú \alst{s}orgafullr at \alst{s}ofa.\\%E

\bvb 111 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
\alst{a}nnars konu \hld tęyg þér \alst{a}ldrigi \\%M
\ind \alst{ęy}rarúnu \alst{a}t.\\%E

\bvb 112 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
á \alst{f}jalli eða \alst{f}irði, \hld ef þik \alst{f}ara tíðir, \\%M
\ind fásk-tu at \alst{v}irði \alst{v}ęl.\\%E

\bvb 113 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
\alst{i}llan mann \hld lát \alst{a}ldrigi \\%M
\ind óhǫpp at þér vita. \\%M
af \alst{i}llum manni \hld fær þú \alst{a}ldrigi \\%M
\ind \alst{g}jǫld hins \alst{g}óða hugar.\\%E

\bvb 114 \\

\bva \alst{O}farla bíta \hld sá'k \alst{ęi}num hal \\%M
\ind \alst{o}rð \alst{i}llrar konu, \\%M
\alst{f}lárǫ́ð tunga \hld varð hǫ́num at \alst{f}jǫrlagi \\%M
\ind ok þęygi of \alst{s}anna \alst{s}ǫk.\\%E

\bvb 115 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
\alst{v}ęizt ef \alst{v}in átt, \hld þann's \alst{v}el trúir, \\%M
\ind \alst{f}ar þú at \alst{f}inna opt. \\%M
þvíat \alst{h}rísi vęx \hld ok \alst{h}óu grasi \\%M
\ind \alst{v}egr, es \alst{v}æt-ki trøðr,\\%E

\bvb 116 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
\alst{v}in þínum \hld \alst{v}es þú aldrigi \\%M
\ind \alst{f}yrri at \alst{f}laumslitum. \\%M
\alst{s}org etr hjarta, \hld ef þú \alst{s}ęgja né náir \\%M
\ind \alst{ęi}nhvęrjum \alst{a}llan hug.\\%E

\bvb 117 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
\alst{g}óðan mann \hld tęyg þér at \alst{g}amanrúnum \\%M
\ind ok nem \alst{l}íknargaldr meðan \alst{l}ifir.\\%E

\bvb 118 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
orðum \alst{sk}ipta \hld þú \alst{sk}alt aldrigi \\%M
\ind við \alst{ó}svinna \alst{a}pa.\\%E

\bvb 119 \\

\bva Af illum \alst{m}anni \hld \alst{m}undu aldrigi \\%M
\ind \alst{g}óðs laun of \alst{g}eta, \\%M
en \alst{g}óðr maðr \hld mun þik \alst{g}ęrva męga \\%M
\ind \alst{l}íknfastan at \alst{l}ofi.\\%E

\bvb 120 \\

\bva \alst{S}ifjum es þá blandit \hld hvęrr es \alst{s}ęgja ræðr \\%M
\ind \alst{ęi}num \alst{a}llan hug; \\%M
alt es \alst{b}ętra \hld an sé \alst{b}rigðum at vesa: \\%M
\ind es-a sá \alst{v}inr es \alst{v}ilt ęitt sęgir.\\%E

\bvb 121 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
þrimr orðum sęnna \hld skal-at-tu þér við verra mann, \\%M
\ind opt hinn \alst{b}ętri \alst{b}ilar. \\%M
\ind þás hinn \alst{v}erri \alst{v}egr.\\%E

\bvb 122 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
\alst{sk}ósmiðr þú verir\hld né \alst{sk}ęptismiðr, \\%M
\ind nema \alst{s}jǫlfum þér \alst{s}éir. \\%M
\alst{Sk}ór's \alst{sk}apaðr illa\hld eða \alst{sk}apt sé rangt, \\%M
\ind þá's þér \alst{b}ǫls \alst{b}eðit.\\%E

\bvb 123 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
hvars þú \alst{b}ǫl kant, \hld kveð þér \alst{b}ǫlvi at \\%M
\ind ok gefat þínum \alst{f}jǫ́ndum \alst{f}rið.\\%E

\bvb 124 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
\alst{i}llu fęginn \hld ves þú \alst{a}ldrigi, \\%M
\ind ęn lát þér at \alst{g}óðu \alst{g}etit.\\%E

\bvb 125 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
\alst{u}pp líta \hld skalattu í \alst{o}rrostu \\%M
\alst{g}jalti \alst{g}líkir \hld verða \alst{g}umna synir \\%M
\ind síðr þitt of \alst{h}ęilli \alst{h}alir.\\%E

\bvb 126 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
Ef vilt þér góða \alst{k}onu \hld \alst{k}vęðja at gamanrúnum \\%M
\ind ok \alst{f}á \alst{f}ǫgnuð af, \\%M
\alst{f}ǫgru skaldu heita \hld ok láta \alst{f}ast vesa; \\%M
\ind lęiðisk mangi \alst{g}ótt ef \alst{g}etr.\\%E

\bvb 127 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
\ind \alst{v}aran bið'k þik \alst{v}esa \\%M
\ind ok \alst{ęi}gi \alst{o}fvaran, \\%M
ves þú við \alst{ǫ}l varastr. \hld ok við \alst{a}nnars konu \\%M
ok við \alst{þ}at hit \alst{þ}riðja, \hld at \alst{þ}jófar né lęiki.\\%E

\bvb 128 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
at \alst{h}áði né \alst{h}látri \hld \alst{h}af þú aldrigi \\%M
\ind \alst{g}ęst né \alst{g}anganda.\\%E

\bvb 129 \\

\bva \alst{O}pt vitu \alst{ó}gǫrla, \hld þęir's sitja \alst{i}nni fyrir, \\%M
\ind hvęrs þęir 'ro \alst{k}yns es \alst{k}oma; \\%M
esat maðr svá \alst{g}óðr \hld at \alst{g}alli né fylgi, \\%M
\ind né svá \alst{i}llr at \alst{ęi}nugi dugi.\\%E

\bvb 130 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
at \alst{h}ǫ́rum þul \hld \alst{h}læ þú aldrigi, \\%M
\ind opt es \alst{g}ótt þats \alst{g}amlir kveða, \\%M
opt ór \alst{sk}ǫrpum bęlg \hld \alst{sk}ilin orð koma \\%M
\ind þęims \alst{h}angir með \alst{h}ǫ́m \\%M
\ind ok \alst{sk}ollir með \alst{sk}rǫ́m, \\%M
\ind ok \alst{v}áfir með \alst{v}ilmǫgum.\\%E

\bvb 131 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
\alst{g}ęst né \alst{g}ęyja \hld né á \alst{g}rind hrękir; \\%M
\ind get þú \alst{v}ǫ́luðum \alst{v}el.\\%E

\bvb 132 \\

\bva \alst{R}amt es þat tré, \hld es \alst{r}íða skal \\%M
\ind \alst{ǫ}llum at \alst{u}pploki; \\%M
\alst{b}aug þú gef \hld eða þat \alst{b}iðja mun \\%M
\ind þér \alst{l}æs hvęrs á \alst{l}iðu.\\%E

\bvb 133 \\

\bva \alst{R}ǫ́ðum þér Loddfáfnir, \hld at þú \alst{r}ǫ́ð nemir, \\%M
\ind \alst{n}jóta munt ef \alst{n}emr, \\%M
\ind þér munu \alst{g}óð ef \alst{g}etr: \\%M
hvars \alst{ǫ}l drekkir \hld kjós þér \alst{ja}rðar męgin, \\%M
þvíat \alst{jǫ}rð tękr við \alst{ǫ}ldri, \hld ęn \alst{ę}ldr við sóttum, \\%M
\alst{ęi}k við \alst{a}bbindi, \hld \alst{a}x við fjǫlkyngi, \\%M
\alst{h}ǫll við \alst{h}ýrógi; \hld \alst{h}ęiptum skal mána kvęðja, \\%M
\alst{b}ęiti við \alst{b}itsóttum, \hld ęn við \alst{b}ǫlvi rúnar; \\%M
\ind \alst{f}old skal við \alst{f}lóði taka.\\%E

\bvb 134 \\

Of Woden's taking of the runes.
It is clear that these verses have very little to do with the rest of the poem, but instead are separate. It is for this reason that they are labelled as \emph{Rúnatals þáttr} (The strand of the Runecount) in younger Eddic paper manuscripts. Many give an archaic, pagan impression. It is as if they were drawn from the lips of an Odinic priest.

\bva \footnotemark[2] \alst{V}ęit'k at ek hekk \hld \alst{v}indga męiði á \\%M
\ind \alst{n}ætr allar \alst{n}íu, \\%M
\alst{g}ęiri undaðr \hld ok \alst{g}efinn Óðni, \\%M
\ind \alst{s}jalfr \alst{s}jǫlfum mér, \\%M
á þęim \alst{m}ęiði, \hld es \alst{m}angi vęit, \\%M
\ind hvęrs af \alst{r}ótum \alst{r}innr\footnotemark[3].\\%E
\footnotetext[3]{This v. clearly belongs together with 138 and 140 and may have been part of a greater now-lost work.}
\footnotetext[3]{Note that \emph{rótum} has lost initial \emph{v-}. Cf. 150/2, which retains it. Unlike the specifically WN. loss of \emph{v} before \emph{r} (cf. 32/2), in this word it occurred in all NGmc. dialects, cf. Swe. \emph{rot}. This may make it less reliable to use for dating the stanza.}

\bvb I know that I hung upon a windy beam [for] nine nights all, gored by spear, and given to Woden, myself to myself, on that tree, which no man knows whereof his roots run. \\

\bva Við \alst{h}lęifi mik sældu \hld né við \alst{h}ornigi; \\%M
\alst{n}ýsta'k \alst{n}iðr, \hld \alst{n}am'k upp rúnar, \\%M
\alst{ǿ}pandi nam, \hld fell'k \alst{a}ptr þaðan.\\%E

\bvb With a loaf [of bread] they gladdened me not, nor with horn's drink. I peered down, I took up the runes, screaming [I] took; then I fell back thence. \\

\bva \alst{F}imbulljóð níu \hld nam'k af hinum \alst{f}rægja syni \\%M
\ind \alst{B}ǫlþorns, \alst{B}ęstlu fǫður, \\%M
ok ek \alst{d}rykk of gat \hld hins \alst{d}ýra mjaðar \\%M
\ind \alst{au}sinn \alst{Ó}ðreri.\footnotemark[1] —\\%E
\footnotetext[1]{It has been noted (FJ) that this verse fits better in the next section of the poem. It is awkwardly placed here, since it mentions \emph{ljóð} '(magical) songs', rather than runes.}

\bvb Nine fimbol-songs, I got from the famous son of \textbf{Balethorn}, the father of \textbf{Bestel}. And a drink I got, of the expensive mead poured to \textbf{Woderearer}. \\

\bva Þá nam'k \alst{f}rævask \hld ok \alst{f}róðr vesa \\%M
\ind ok \alst{v}axa ok \alst{v}ęl hafask; \\%M
\alst{o}rð mér af \alst{o}rði \hld \alst{o}rðs lęitaði \\%M
\ind \alst{v}erk mér af \alst{v}erki \alst{v}erks.\\%E

\bvb Then I began to thrive, and be knowledgable, and grow and live well. A word for me of a word a word sought out; a work for me of a work a work\footnotemark[10]. \\
\footnotetext[10]{That is, each speech led to another, each productive act to another.}

\bva \alst{R}únar munt finna \hld ok \alst{r}áðna stafi, \\%M
\ind mjǫk \alst{st}óra \alst{st}afi, \\%M
\ind mjǫk \alst{st}inna \alst{st}afi, \\%M
\ind es \alst{f}áði \alst{f}imbulþulr \\%M
\ind ok \alst{g}ęrðu \alst{g}innręgin \\%M
\ind ok \alst{r}ęist Hroptr \alst{r}agna\footnotemark[5]. —\\%E
\footnotetext[5]{Corrected from \emph{rǫgna}. Cf. Eskál \emph{Vell} 31/2 in SkP I, p. 322.}

\bvb \textbf{Runes} thou wilt find, and interpreted staves; much large staves, much stiff staves, which the \textbf{fimbol-thyle} painted, and the \textbf{ginn}-\textbf{Powers} made, and the \textbf{Roft} of the Powers carved. \\

\bva \alst{Ó}ðinn með \alst{ǫ́}sum, \hld ęn fyr \alst{ǫ}lfum Dáinn, \\%M
\ind \alst{D}valinn \alst{d}vergum fyrir, \\%M
\ind \alst{Á}sviðr \alst{jǫ}tnum fyrir, \\%M
ek\footnotemark[10] ręist \alst{s}jalfr \alst{s}umar. —\\%E
\footnotetext[10]{Since Woden has already been mentioned, it is unclear who the \emph{I} here may be.}

\bvb \textbf{Woden} among the \textbf{Eses}, but before the \textbf{Elves} \textbf{Dowen}, \textbf{Dwalen} before the \textbf{Dwarves}, \textbf{Aswood} before the etins; I myself carved some. \\

\bva Vęiztu, hvé \alst{r}ísta skal? \hld vęiztu, hvé \alst{r}áða skal? \\%M
vęiztu, hvé \alst{f}áa skal? \hld vęiztu, hvé \alst{f}ręista skal? \\%M
vęiztu, hvé \alst{b}iðja skal? \hld vęiztu, hvé \alst{b}lóta skal? \\%M
vęiztu, hvé \alst{s}ęnda skal? \hld vęiztu, hvé \alst{s}óa skal? —\footnotemark[5]\\%E
\footnotetext[5]{This v. bears strong resemblance with Vg 216 (Högstena gealdor). TODO: Elaborate.}

\bvb Knowest thou how to carve? Knowest thou how to interpret? Knowest thou how to paint? Knowest thou how to try? Knowest thou how to bid? Knowest thou how to bloot? Knowest thou how to send? Knowest thou how to \textbf{soo}? \\

\bva \alst{B}ętra's ó\alst{b}eðit \hld an sé of\alst{b}lótit, \\%M
\ind ęy sér til \alst{g}ildis \alst{g}jǫf; \\%M
bętra's ó\alst{s}ęnt \hld an sé of\alst{s}óit.\footnotemark[6]\\%E
\footnotetext[6]{A final line is likely missing here. — Identical word-pairing (\emph{biðja} – \emph{blóta}, \emph{senda} – \emph{sóa}) may reveal this v.'s relation with the previous one.}

\bvb Better is unbid than be excessively blooted; a gift always looks to a tribute. Better is unsent than be excessively sooed. \\

\bva Svá \alst{Þ}undr of ręist \hld fyr \alst{þ}jóða rǫk \\%M
þar's \alst{u}pp of ręis, \hld es \alst{a}ptr of kom.\\%E

\bvb Thus \textbf{Thound} did carve for the fate of the nations; there did rise up, who came back.\footnotemark[8] \\
\footnotetext[8]{A most cryptic v.}


	Woden's recounting of his Songs.

\bva Ljóð \alst{þ}au kan'k, \hld es kann-at \alst{þ}jóðans kona \\%M
\ind ok \alst{m}anskis \alst{m}ǫgr. \\%M
\alst{H}jǫlp hęitir ęitt, \hld þat þér \alst{h}jalpa mun \\%M
\ind við \alst{s}orgum ok \alst{s}ǫkum, \hld ok \alst{s}útum gǫrvǫllum.\\%E

\bvb I know those \textbf{leeds}, which the prince's woman does not know, and no man's lad. Help one is called, it will help thee, against sorrows and sakes\footnotemark[9], and all kinds of misfortunes. \\
\footnotetext[9]{Legal proceedings.}

\bva Þat kan'k \alst{a}nnat, \hld es þurfu \alst{ý}ta synir,\footnotemark[10] \\%M
\ind þęir's vilja \alst{l}æknar \alst{l}ifa.\\%E
\footnotetext[10]{Identical wording to 163/2.}

\bvb 145 \\

\bva \alst{Þ}at kan'k \alst{þ}riðja, \hld ef mér verðr \alst{þ}ǫrf mikil \\%M
\ind \alst{h}apts við mina \alst{h}ęiptmǫgu, \\%M
\alst{ę}ggjar dęyfi'k \hld minna \alst{a}ndskota, \\%M
\ind bítat þeim \alst{v}ǫ́pn né \alst{v}ęlir.\\%E

\bvb 146 \\

\bva Þat kan'k \alst{f}jórða, \hld ef mér \alst{f}yrðar bera \\%M
\ind \alst{b}ǫnd at \alst{b}oglimum, \\%M
svá ek \alst{g}ęl, \hld at \alst{g}anga má'k, \\%M
\ind sprettr mér af \alst{f}ótum \alst{f}jǫturr. \\%M
\ind ęn af \alst{h}ǫndum \alst{h}apt.\\%E

\bvb 147 \\

\bva Þat kan'k \alst{f}imta, \hld ef sé'k af \alst{f}ári skotinn \\%M
\ind \alst{f}lęin í \alst{f}olki vaða, \\%M
flýgr-a svá \alst{st}int, \hld at \alst{st}ǫðvigak, \\%M
\ind ef hann \alst{s}jónum of \alst{s}é'k.\\%E

\bvb 148 \\

\bva Þat kan'k \alst{s}étta, \hld ef mik \alst{s}ærir þegn \\%M
\ind á \alst{v}rótum hrás \alst{v}iðar. \\%M
þann \alst{h}al, \hld es mik \alst{h}ęipta kvęðr, \\%M
\ind þann eta \alst{m}ęin hęldr an \alst{m}ik.\\%E

\bvb 149 \\

\bva Þat kan'k \alst{s}jaunda, \hld ef \alst{s}é'k hóvan loga \\%M
\ind \alst{s}al of \alst{s}essmǫgum, \\%M
\alst{b}rinnrat svá \alst{b}ręitt, \hld at hǫ́num \alst{b}jargigak; \\%M
\ind þann kan'k \alst{g}aldr at \alst{g}ala.\\%E

\bvb 150 \\

\bva Þat kan'k \alst{á}tta, \hld es \alst{ǫ}llum es \\%M
\ind \alst{n}ytsamligt at \alst{n}ema, \\%M
\alst{h}var's \alst{h}atr vęx \hld með \alst{h}ildings sonum, \\%M
\ind þat má'k \alst{b}ǿta \alst{b}rátt.\\%E

\bvb 151 \\

\bva Þat kan'k \alst{n}íunda, \hld ef mik \alst{n}auðr of stęndr \\%M
\ind at bjarga \alst{f}ari á \alst{f}loti, \\%M
\alst{v}ind ek kyrri \hld \alst{v}ági á \\%M
\ind ok \alst{s}væfi'k allan \alst{s}æ.\\%E

\bvb 152 \\

\bva Þat kan'k \alst{t}íunda, \hld ef sé'k \alst{t}únriður \\%M
\ind \alst{l}ęika \alst{l}opti á, \\%M
ek svá \alst{v}in'k, \hld at þær \alst{v}illar fara \\%M
\ind sinna \alst{h}ęim-\alst{h}ama \\%M
\ind sinna \alst{h}ęim-\alst{h}uga.\\%E

\bvb 153 \\

\bva Þat kan'k \alst{ę}llipta, \hld ef skal'k til \alst{o}rrostu \\%M
\ind \alst{l}ęiða \alst{l}angvini, \\%M
und \alst{r}andir gęlk, \hld ęn þęir með \alst{r}íki fara, \\%M
\ind \alst{h}ęilir \alst{h}ildar til, \\%M
\ind \alst{h}ęilir \alst{h}ildi frá, \\%M
\ind koma þęir \alst{h}ęilir \alst{h}vaðan.\\%E

\bvb 154 \\

\bva Þat kan'k \alst{t}olpta, \hld ef sé'k á \alst{t}ré uppi \\%M
\ind \alst{v}áfa \alst{v}irgilná, \\%M
svá ek \alst{r}íst \hld ok í \alst{r}únum fá'k, \\%M
\ind at sá \alst{g}ęngr \alst{g}umi. \\%M
\ind ok \alst{m}ælir við \alst{m}ik.\\%E

\bvb 155 \\

\bva \alst{Þ}at kan'k \alst{þ}rettánda \hld ef skal'k \alst{þ}egn ungan \\%M
\ind \alst{v}erpa \alst{v}atni á,\footnotemark[49] \\%M
munat hann \alst{f}alla, \hld þótt í \alst{f}olk komi, \\%M
\ind \alst{h}nígr-a sá \alst{h}alr fyr \alst{h}jǫrum.\\%E
\footnotetext[49]{Describing the pagan ritual of pouring water on a newborn child. Cf. \emph{Rþ} 7/1b, 21/1b, 34/2a.}

\bvb 156 \\

\bva Þat kan'k \alst{f}jogurtánda, \hld ef skal'k \alst{f}yrða liði \\%M
\ind \alst{t}ęlja \alst{t}íva fyrir, \\%M
\alst{á}sa ok \alst{a}lfa \hld ek kann \alst{a}llra skil, \\%M
\ind fár kann ó\alst{s}notr \alst{s}vá.\\%E

\bvb 157 \\

\bva \alst{Þ}at kan'k fimtánda, \hld es gól {Þ}jóðrørir \\%M
\ind \alst{d}vergr fyr \alst{D}ęllings \alst{d}urum, \\%M
\alst{a}fl gól \alst{ǫ́}sum, \hld ęn \alst{ǫ}lfum frama, \\%M
\ind \alst{h}yggju \alst{H}roptatý.\\%E

\bvb 158 \\

\bva Þat kan'k sextánda, \hld ef vil'k hins svinna mans \\%M
\ind hafa gęð alt ok gaman, \\%M
\alst{h}ugi \alst{h}vęrfi'k \hld \alst{h}vitarmri konu \\%M
\ind ok \alst{s}ný'k hennar ǫllum \alst{s}efa.\\%E

\bvb 159 \\

\bva Þat kan'k \alst{s}jautjánda \hld at mik \alst{s}ęint mun firrask \\%M
\ind hit \alst{m}anunga \alst{m}an.\\%E

\bvb 160 \\

\bva Þat kan'k \alst{á}tjánda, \hld es \alst{æ}va kęnni'k \\%M
\ind \alst{m}ęy né \alst{m}anns konu, \\%M
\alst{a}lt es bętra \hld es \alst{ęi}nn of kann, \\%M
\ind þat fylgir \alst{l}jóða \alst{l}okum, \\%M
nema þęiri \alst{ęi}nni, \hld es mik \alst{a}rmi vęrr, \\%M
\ind eða mín \alst{s}ystir \alst{s}é.\\%E

\bvb 161 \\

\bva Nú eru \alst{H}áva mál \hld kveðin \alst{H}áva\alst{h}ǫllu í \\%M
\ind \alst{a}llþǫrf \alst{ý}ta sonum, \\%M
\ind \alst{ó}þǫrf \alst{jǫ}tna sonum; \\%M
hęill sás kvað, \hld hęill sás kann, \\%M
\ind \alst{n}jóti sás \alst{n}am, \\%M
\ind \alst{h}ęilir þęirs \alst{h}lýddu.\footnotemark[50]\\%E
\footnotetext[50]{A highly metrically irregular verse; clearly younger than the rest of the poem collection. See the Introduction, p. TODO.}

\bvb 162 \\
