\book{The Speeches of Grimen. (Grímnismǫ́l)}\bookStart

\begin{verse}
\bva Hęitr est hripuðr \hld ok hęldr til mikill,
\ind gǫngumk firr funi.
Loði sviðnar, \hld þótt á lopt bera'k;
\ind brinnumk feldr fyrir. 
\end{verse}

\bvb Hot art thou, rippeth [\textsc{fire}], and rather too large; go further from me, fire! The shaggy cape is scorched, though I might hold it up in the air; the cloak is burning before me.

\begin{verse}
\bva Átta nætr satk \hld milli ęlda hér,
\ind svát mér mangi mat né bauð
nema ęinn Agnarr, \hld es ęinn skal ráða,
\ind Gęirrøðar sonr, Gotna landi.
\end{verse}

\bvb For eight nights I sat ’tween the fires here, while no man offered me food, except for lone Eynhere, who lone shall rule, that son of Garred, the land of the Gots!

...

Himinbjǫrg eru in áttu \hld ęn þar Hęimdall
\ind kveða valda véum.
þar vǫrðr goða \hld drękkr í væru ranni
\ind glaðr góða mjǫð. 

\bvb Heavenbarrows are the eighth, and there Homedall, they say, wields over wighs. There in the tranquil house the ward of the gods gladly drinks good mead.

Folkvangr es inn níundi \hld en þar Fręyja ræðr
\ind sessa kostum í sal;
halfan val \hld hon kýss hvęrjan dag
\ind ęn halfan Óðinn á. 

\bvb Folkwong is the ninth, and there Frow decides the choice of seats in the hall; half of the slain she chooses each day, but half Weden owns.

Glitnir es inn tíundi; \hld hann es gulli studdr
\ind ok silfri þakðr it sama
ęn þar Forseti \hld byggir flęstan dag
\ind ok svæfir allar sakir. 

\bvb Glitten is the tenth, it is studded by gold, and thatched by silver the same. And there Forset dwells most of the day, and resolves\footnotemark[1] all [legal] matters.
\footnotetext[1]{lit. 'puts to sleep'}

Nóatún eru in ęlliftu \hld ęn þar Njǫrðr hęfir
\ind sér um gǫrva sali,
manna þęngill \hld inn męinsvani
\ind hátimbruðum hǫrgi ræðr. 

\bvb Nowetowns are the tenth, and there Nearth has made himself a hall. The prince of men, the guileless one, rules the high-timbered \textbf{harrow}.

Hrísi vęx \hld ok há grasi
\ind Víðars land, viði,
ęn þar mǫgr of læzk \hld af mars baki
\ind frǿkn at hęfna fǫður.

\bvb TO-DO.

Andhrímnir \hld lætr í Ęldhrímni
\ind Sæhrímni soðinn,
flęska bęzt, \hld ęn þat fáir vitu
\ind við hvat ęinhęrjar alask.

\bvb Andrimen lets in Eldrimen Sowrimen be boiled. Of pork the best, but few know that; by what the Lone Warriors are nourished.\footnotemark[1]
\footnotetext[1]{The cook Andrimen 'face-sooty' has the boar Sowrimen 'sow-sooty' boiled in the cauldron Eldrimen 'fire-sooty'; by this meat are the Lone Warriors nouished. See the index.}

Gera ok Freka \hld sęðr gunntamiðr,
\ind hróðigr Hęrjafǫðr,
ęn við vín ęitt \hld vápngǫfugr
\ind Óðinn æ lifir. 

\bvb Gerr and Freck satiates the battle-accustomed, glorious Father of Hosts; yet by wine alone, does the weapon-worshipful Weden ever life.

Huginn ok Muninn \hld fljúga hvęrjan dag
\ind jǫrmungrund yfir;
óumk of Hugin, \hld at aptr né komit;
\ind þó séumk męir of Munin.

\bvb Highen and Minden fly every day, o'er the ermin-ground. I fear for Highen, that he might not come back, though I worry more for Minden.

Þýtr Þund, \hld unir Þjóðvitnis
\ind fiskr flóði í;
áarstraumr \hld þykkir ofmikill
\ind valglaumi at vaða.

\bvb Thound howls, the fish of Thedewitten is content in the river; the stream of the river seems too great for the cheery slain men to wade.

...

Askr Yggdrasils, \hld hann es ǿztr viða
\ind ęn Skíðblaðnir skipa,
Óðinn ása \hld ęn jóa Slęipnir,
Bilrǫst brúa \hld ęn Bragi skalda,
Hábrók hauka \hld ęn hunda Garmr. 

\bvb The ash of Ugdrassle, that is the noblest of trees, but Shidebladen of ships. Weden of the Ease, but of horses Slopen. Bilrest of bridges, but Bray of scolds. Highbrook of hawks, but of hounds Garm.

