\book{The Speeches of Grimen. (Grímnismǫ́l)}\bookStart

\begin{verse}
\bva Hęitr est hripuðr \hld ok hęldr til mikill,
\ind gǫngumk firr funi.
Loði sviðnar, \hld þótt á lopt bera'k;
\ind brinnumk feldr fyrir. 
\end{verse}

\bvb Hot art thou, rippeth [\textsc{fire}], and rather too large; go further from me, fire! The shaggy cape is scorched, though I might hold it up in the air; the cloak is burning on me.

\begin{verse}
\bva Átta nætr satk \hld milli ęlda hér,
svát mér mangi \hld mat né bauð
nema ęinn Agnarr, \hld es ęinn skal ráða,
Geirrøðar sonr, \hld Gotna landi.
\end{verse}

\bvb For eight nights I sat beneath the fires here, while no man offered me food, except for Eynhere alone, who shall rule alone, that son of Garred, the land of the Gots.

\bvb Heavenbarrows are the eighth, and there Homedall, they say, wields over wighs. There in the tranquil house the ward of the gods gladly drinks good mead.

\bvb Folkwong is the ninth, and there Frow decides the choice of seats in the hall; half of the slain she chooses each day, but half Weden owns.

\bvb Glitten is the tenth, it is studded by gold, and thatched by silver the same. And there Forset dwells most of the day, and resolve all [legal] matters.

\bvb Nowetowns are the tenth, and there Nearth has made himself a hall. The prince of men, the guileless one runes the high-timbered \textbf{harrow}.

\bvb Andrimen lets in Eldrimen Sodrimen be boiled. Of pork the best, but few know that: by what the Lone Warriors are nourished.

\bvb The battle-accustomed, glorious Here-father satiates Gerr and Freck. Yet by wine alone does the weapon-worshipful Weden forever life.

\bvb Highen and Minden fly every day the ermin-ground over. I fear for Highen, that he might not come back, though I worry more for Minden.

\bvb Thound howls, loving the fish of Thedewitten in the river.

\bvb The ash of Ugdrassle, that is the noblest of trees, but Shidebladen of ships. Weden of Eses, but of horses Slopen. Bilrast of bridges, but Bray of scolds. Highbrook of hawks, but of hounds Garm.
