\bookStart{Lay of Attle}[Atlakviða]

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: C10th (0.719)–early C11th (0.212)

\textbf{Meter:} \Malahattr, \Fornyrdislag
\end{flushright}%

A famously archaic poem.

Attle sends his messenger Kneefrith to Guther (1). He arrives at Guther’s hall, where the mood is one of unease, and addresses Guther (2). Kneefrith invites him and his brother Hain to Attle’s court (3), offering them treasures, weapons and land (4–5). Guther asks his brother Hain for advice, since he has not heard of Attle having gold to give away (6).

\sectionline

\section{The Death of Attle (\emph{Dauði Atla})}

\bpg\bpa Guðrún Gjúkadóttir hefndi brǿðra sinna, svá sem frę́gt er orðit. Hon drap fyrst sonu Atla, en eptir drap hon Atla ok brendi hǫllina ok hirðina alla; um þetta er sjá kviða ort.\epa

\bpb Guthrun Yivicksdaughter avenged her brothers, as has become famous. She first killed the sons of Attle, and after that she killed Attle, and burned the hall and the whole hird. Regarding that this lay is wrought.\epb\epg

\sectionline

\bvg\bva \alst{A}tli sęndi \hld\ \alst{á}r til Gunnars &
\alst{k}unnan sęgg at ríða, \hld\ \alst{K}néfrøðr vas sá hęitinn; &
at \alst{g}ǫrðum kom hann \alst{G}júka \hld\ ok at \alst{G}unnars hǫllu, &
\alst{b}ękkjum arin-gręypum \hld\ ok at \alst{b}jóri svǫ́sum.\eva

\bvb Attle sent—of yore–to Guther \\
a well-known messenger to ride; \inx[P]{Kneefrith} he was called. \\
To the yards of Yivick he came, and to the hall of Guther; \\
to the hearth-surrounding benches, and to the lovely beer.\evb\evg


\bvg\bva \alst{D}rukku þar \alst{d}rótt-męgir \hld\ —ęn \edtrans{\alst{d}yljęndr}{concealed ones}{\Bfootnote{\textcite{FinnurEdda} reasonably interprets this as referring to Attle’s spies at Guther’s court.}} þǫgðu— &
\alst{v}ín í \edtrans{\alst{v}al-hǫllu}{the walhall}{\Bfootnote{The interpretation of this compound is difficult in the current context. The first element \emph{val-} could be (1) \emph{valr} ‘falcon’, referring to the aristocratic hunting practice; (2) \emph{valr} ‘\inx[G]{Wales}[Wale]’, cognate with ‘Welsh’ but in ON referring to the French or Romans, stressing the southern location or appearance of the hall; or (3) \emph{valr} ‘(collective) the battle-slain’, foreshadowing the inevitable death (\inx[C]{feyness}) of the \inx[G]{Yivickings}. If (3) is correct the word is linguistically identical to \inx[L]{Walhall}, Weden’s hall, whither the battle-slain go.}}, \hld\ \alst{v}ręiði sǫ́usk þęir Húna; &
\alst{k}allaði þá \alst{K}néfrøðr \hld\ \alst{k}aldri rǫddu, &
\alst{s}ęggr inn \alst{s}uð-rǿni \hld\ \alst{s}at hann á bękk hǫ́m:\eva

\bvb There the dright-lads \ken{warriors} drank—but the concealed ones shut up— \\
wine in the walhall; they feared the wrath of the Huns. \\
Then called Kneefrith with cold voice, \\
the southern messenger, he sat on a high bench:\evb\evg


\bvg\bva „\alst{A}tli mik hingat sęndi \hld\ ríða \alst{ø}ręndi, &
\alst{m}ar inum \alst{m}él-gręypa, \hld\ \alst{M}yrk-við inn ó·kunna &
at \alst{b}iðja yðr, Gunnarr, \hld\ at it á \alst{b}ękk kǿmið &
með \alst{h}jǫlmum arin-gręypum \hld\ at sǿkja \alst{h}ęim Atla.\eva

\bvb “Attle sent me hither to ride with an errand, \\
on the bit-champing steed through Mirkwood uncharted— \\
to ask you, O Guther, that ye two \ken*{= Guther and Hain} on the bench come, \\
with hearth-surrounding helmets, to seek the home of Attle.\evb\evg


\bvg\bva \alst{Sk}jǫldu kneguð þar vęlja \hld\ ok \alst{sk}afna aska, &
\alst{h}jalma gull-roðna \hld\ ok \alst{H}úna męngi, &
\alst{s}ilfr-gyllt \alst{s}ǫðul-klę́ði, \hld\ \alst{s}ęrki val-rauða, &
\alst{d}afar, \alst{d}arraða, \hld\ \alst{d}rǫsla mél-gręypa.\eva

\bvb There ye might choose shields, and shaven ash-spears, \\
helmets gold-reddened, and the multitude of the Huns, \\
silver-gilt saddle-cloths, blood-red serks, \\
daves, spears, bit-champing steeds.\evb\evg


\bvg\bva \alst{V}ǫll létsk ykkr ok myndu gefa \hld\ \alst{v}íðrar Gnita-hęiðar &
af \alst{g}ęiri \alst{g}jallanda \hld\ ok af \alst{g}ylltum stǫfnum, &
\alst{st}órar męiðmar \hld\ ok \alst{st}aði Danpar, &
hrís þat it \alst{m}ę́ra \hld\ es meðr \alst{M}yrk-við kalla.“\eva

\bvb He also declared himself willing to give you two the field of wide Gnit-heath, \\
{[and]} of yelling spears and of gilded prows, \\
great treasures and the place of Danp; \\
the renowned brush which men call Mirkwood.\evb\evg


\bvg\bva \alst{H}ǫfði vatt þá Gunnarr \hld\ ok \alst{H}ǫgna til sagði: &
„Hvat rę́ðr þú okkr, \alst{s}ęggr hinn ǿri, \hld\ alls vit \alst{s}líkt hęyrum? &
\alst{G}ull vissa’k ękki \hld\ á \alst{G}nita-hęiði, &
þat’s vit \alst{ę́}ttim-a \hld\ \alst{a}nnat slíkt.\eva

\bvb His head turned Guther then, and said to Hain: \\
“What dost thou counsel us two, O younger man, as such a thing we hear? \\
I knew of no gold on the Gnit-heath \\
which we two should not own as much of.\evb\evg


\bvg\bva \alst{S}jau ęigu vit \alst{s}al-hús \hld\ \alst{s}verða full, &
\alst{h}vęrju ’ru þęira \hld\ \alst{h}jǫlt ór gulli; &
\alst{m}ínn vęit’k \alst{m}ar bętstan \hld\ en \alst{m}ę́ki hvassastan, &
\alst{b}oga \alst{b}ękk-sǿma \hld\ en \alst{b}rynjur ór gulli;\eva

\bvb We own seven hall-houses filled with swords— \\
on each of them is a golden hilt; \\
I know my horse to be the best and {[my]} sword the sharpest, \\
{[my]} bow bench-fit and {[my]} byrnies golden,\evb\evg


\bvg\bva \alst{h}jalm ok skjǫld \alst{h}vítastan, \hld\ kominn ór \alst{h}ǫll Kíars; &
\alst{ęi}nn ’s mínn bętri \hld\ en sé \alst{a}llra Húna.“\eva

\bvb {[my]} helmet and shield the whitest, come from Choser’s hall; \\
mine alone is better, than [those] of all of the Huns might be!”\evb\evg


\bvg\bva „Hvat hyggr \alst{b}rúði \alst{b}ęndu \hld\ þá’s hón okkr \alst{b}aug sęndi, &
\alst{v}arinn \alst{v}ǫ́ðum hęiðingja? \hld\ Hykk at hón \alst{v}ǫrnuð byði! &
\alst{H}ár fann’k \alst{h}ęiðingja \hld\ riðit í \alst{h}ring rauðum; &
\alst{y}lfskr es vegr \alst{o}kkarr \hld\ at ríða \alst{ø}ręndi.“\eva

\bvb {[Hain quoth:]} \\
“What thinkest thou the bride meant when she sent us a bigh \\
wrapped with a heath-dweller’s cloth \ken{wolf > wolf’s hair}? I think she meant it as a warning! \\
A heath-dweller’s \ken{wolf’s} hair I found wrapped round the red ring: \\
wolven is our road, if we ride that errand!\footnoteB{That it is the more cautious Hain who speaks here is clear from Guther’s response in the following stanzas.  Whereas Hain judges the wolf-hair to be a warning of Hunnish treachery, Guther thinks that it is a warning that wolves will steal his treasure if he does not show up.}”\evb\evg


\bvg\bva \alst{N}iðjar-gi hvǫttu Gunnar \hld\ né \alst{n}áungr annarr, &
\alst{r}ýnęndr né \alst{r}áðęndr, \hld\ né þęir’s \alst{r}íkir vǫ́ru; &
\alst{k}vaddi þá Gunnarr \hld\ sęm \alst{k}onungr skyldi, &
\alst{m}ę́rr í \alst{m}jǫð-ranni \hld\ af \alst{m}óði stórum:\eva

\bvb No kinsmen Guther, nor any other relation, \\
not counselors nor advisors, nor those who were powerful. \\
Then Guther announced—as a king should, \\
renowned in the mead-hall—with great spirit:\evb\evg


\bvg\bva „Rís-tu nú, \edtrans{\alst{F}jǫrnir}{Ferner}{\Bfootnote{An otherwise unknown servant.}}, \hld\ lát-tu á \alst{f}lęt vaða &
\alst{g}reppa \alst{g}ull-skálir \hld\ með \alst{g}umna hǫndum!\eva

\bvb “Rise now, Ferner! Let on the benches wade forth \\
the golden bowls of warriors along the hands of men!\evb\evg


\bvg\bva \alst{U}lfr mun ráða \hld\ \alst{a}rfi Niflunga, &
\alst{g}amlir \alst{g}ran-varðir, \hld\ ef \alst{G}unnars missir; &
\alst{b}irnir \alst{b}lakk-fjallir \hld\ \alst{b}íta þref-tǫnnum, &
\alst{g}amna \alst{g}ręy-stóði, \hld\ ef \alst{G}unnarr né kømr-at.“\eva

\bvb The wolf will rule the inheritance of the Nivlings— \\
the old grey guardians \ken{wolves}—if Guther is absent. \\
Black-furred bears will bite with wrangling teeth— \\
amusing the bitch-pack—if Guther comes not.”\evb\evg


\bvg\bva \alst{L}ęiddu land-rǫgni \hld\ \edtrans{\alst{l}ýðar ó·nęisir}{unshamed men}{\Bfootnote{Compare the long-line on the Thorsberg chape (\char`~\ 160–240 AD): \emph{wlþuþewaʀ \hld\ ni wajē-māriʀ} ‘Wolthew, the not ill-famed \ken{famous}’.}}, &
\alst{g}rátęndr, \alst{g}unn-hvatan, \hld\ ór \alst{g}arði Húna; &
þá kvað þat inn \alst{ø̇}ri \hld\ \alst{ę}rfi-vǫrðr Hǫgna: &
„\alst{H}ęilir farið nú ok \alst{h}orskir \hld\ hvar’s ykkr \alst{h}ugr tęygir!“\eva

\bvb Unshamed men led the lord of the land, \\
weeping, the battle-bold man out of the yards of the Huns. \\
Then quoth this the young inheritance-ward \ken{son} of Hain: \\
“Fare ye two now whole and wise wherever your heart may draw you!”\evb\evg


\bvg\bva \alst{F}etum létu \alst{f}rǿknir \hld\ of \alst{f}jǫll at þyrja &
\alst{m}ar ina \alst{m}él-gręypu, \hld\ \alst{M}yrk-við inn ókunna; &
\alst{h}ristisk ǫll \alst{H}ún-mǫrk \hld\ þar’s \alst{h}arð-móðgir fóru, &
\alst{v}rǫ́ku þęir \alst{v}and-styggva \hld\ \alst{v}ǫllu al-grǿna.\eva

\bvb With strides the braves made the bit-champing steed \\
rush o’er the fells through Mirkwood uncharted. \\
All Hunmark shook where the hard-minded went forth; \\
they drove the whip-shy horse along the allgreen fields.\evb\evg


\bvg\bva \alst{L}and sǫ́u þęir Atla \hld\ ok \alst{l}ið-skjalfar djúpar; &
\alst{B}ikka greppar standa \hld\ á \alst{b}org inni hǫ́u, &
\alst{s}al of \alst{s}uðr-þjóðum, \hld\ \alst{s}lęginn sess-męiðum, &
\alst{b}undnum rǫndum, \hld\ \alst{b}lęikum skjǫldum,\eva

\bvb The land of Attle they saw, and ravines deep, \\
\inx[P]{Bicke}’s soldiers standing on the high stronghold, \\
the hall of the southfolk built with seat-beams, \\
with bound rims, with pale shields,\evb\evg


\bvg\bva \alst{d}afar, \alst{d}arraða; \hld\ en þar \alst{d}rakk Atli &
\alst{v}ín í \alst{v}al-hǫllu; \hld\ \alst{v}ęrðir sǫ́tu úti &
at \alst{v}arða þęim Gunnari \hld\ ef þęir hér \alst{v}itja kǿmi &
með \alst{g}ęiri \alst{g}jallanda \hld\ at vękja \alst{g}ram hildi.\eva

\bvb daves, spears. And there drank Attle \\
wine in the wal-hall—watchmen sat outside \\
to watch for Guther’s men, if they came here to visit, \\
with yelling spears to wake the ruler with war.\evb\evg


\bvg\bva \alst{S}ystir fann þęira \alst{s}nemmst \hld\ at þęir í \alst{s}al kvǫ́mu, &
\alst{b}rǿðr hęnnar \alst{b}áðir, \hld\ \alst{b}jóri vas hón lítt drukkin: &
„\alst{R}áðinn est nú, Gunnarr, \hld\ hvat munt, \alst{r}íkr, vinna &
við \alst{H}úna \alst{h}arm-brǫgðum? \hld\ \alst{H}ǫll gakk þú ór snemma!\eva

\bvb Their sister found soonest they they had come into the hall— \\
her brothers both—on beer was she lightly drunk: \\
“Betrayed art thou now, Guther; what wilt thou, powerful man, work \\
against the Hunnish harm-tricks? Go soon out of the hall!\footnoteB{Before anything evil might happen.}”\evb\evg


\bvg\bva \alst{B}ętr hęfðir þú, \alst{b}róðir, \hld\ at þú í \alst{b}rynju fǿrir, &
sęm \alst{h}jǫlmum arin-gręypum \hld\ at séa \alst{h}ęim Atla; &
\alst{s}ę́tir þú í \alst{s}ǫðlum \hld\ \alst{s}ól-hęiða daga, &
\alst{n}ái \alst{n}auð-fǫlva \hld\ létir \alst{n}ornir gráta,\eva

\bvb Better hadst thou, brother, if thou hadst gone in byrnie \\
with hearth-surrounding helmets, to see the home of Attle; \\
if thou hadst set in the saddle during sun-bright days \\
need-pale corpses; if thou madest the norns cry,\evb\evg


\bvg\bva \alst{H}úna skjald-męyjar \hld\ \alst{h}ęrfi kanna &
en \alst{A}tla sjalfan \hld\ létir í \alst{o}rm-garð koma; &
nú ’s sá \alst{o}rm-garðr \hld\ \alst{y}kkr of folginn.“\eva

\bvb {[and]} the Hunnish shield-maidens to know the harrow;\footnoteB{i.e. if he turned the Hunnish shield-maidens into enslaved farmhands.} \\
and Attle himself hadst thou brought in the snake-pit— \\
now that snake-pit has swallowed you two!”\evb\evg

Guther answers:

\bvg\bva „\alst{S}ęinað ’s nú, systir, \hld\ at \alst{s}amna Niflungum, &
\alst{l}angt ’s at \alst{l}ęita \hld\ \alst{l}ýða sinnis til, &
of \alst{r}osmu-fjǫll \alst{R}ínar, \hld\ \alst{r}ekka ó·nęissa.“\eva

\bvb “’Tis late now, sister, to gather the Nivlings; \\
’tis far to look for the support of men: \\
over the great fells of the Rhine for unshamed warriors.”\evb\evg


\bvg\bva \alst{F}engu þęir Gunnar \hld\ ok í \alst{f}jǫtur sęttu, &
\edtrans{vin \alst{B}orgunda}{the friend of the Burgends}{\Bfootnote{The historic Guther was king of the Burgundians.  The manuscript has a small stroke above the \emph{n} that abbreviates the syllable \emph{ir}, indicating the plural \emph{vinir} ‘friends’, who would then be the people binding Guther.  This is probably due to a scribal misunderstanding of a not uncommon type, since the significance of the kenning had been forgotten.  It is clearly old, for in \Waldere\ 46 Walder addresses Guther, whom he is just about to fight, by the identical phrase \emph{wine Burgenda}.}}, \hld\ ok \alst{b}undu fastla; &
\alst{s}jau hjó Hǫgni \hld\ \alst{s}verði hvǫssu &
en inum \alst{á}tta hratt hann \hld\ í \alst{ę}ld hęitan.\eva

\bvb They caught Guther and in fetters placed him \\
—the friend of the Burgends—and bound him firmly. \\
Hain smote seven with a sharp sword, \\
and the eighth one he threw into hot fire.\evb\evg


\bvg\bva \edtext{Svá skal \alst{f}rǿkn \hld\ \alst{f}jǫ́ndum vęrjask;}{\lemma{Svá \dots\ vęrjask}\Bfootnote{Line moved from the last st. to this one since it seems to connect semantically with the immediately following line, and results in two typical four-line stanzas.}} &
\alst{H}ǫgni varði \hld\ \alst{h}ęndr Gunnars. &
\alst{f}rǫ́gu \alst{f}rǿknan \hld\ ef \alst{f}jǫr vildi &
\alst{G}otna þjóðann \hld\ \alst{g}ulli kaupa.\eva

\bvb So shall a brave guard himself against foes; \\
Hain guarded the hands of Guther. \\
They asked the brave \ken*{Guther} if his \ken*{Hain’s} life he wished— \\
the ruler of the Gots—to buy with gold.\footnoteB{The Huns try to make Guther (the “ruler of the Gots”, cf. sts. 1, 3, 10) pay for Hain’s life.  Guther instead responds with the following.}\evb\evg


\bvg\bva „\alst{H}jarta skal mér \alst{H}ǫgna \hld\ í \alst{h}ęndi liggja &
\alst{b}lóðugt, ór \alst{b}rjósti \hld\ skorit \alst{b}ald-riða, &
\edtrans{\alst{s}axi \alst{s}líðr-bęitu}{slide-biting sax}{\Bfootnote{A short-sword with a blade so sharp that it draws blood when one slides the finger across it.}}, \hld\ \alst{s}yni þjóðans.“\eva

\bvb “The heart of Hain shall lie in my hands: \\
bloody from the breast, cut from the bold rider \ken*{= Hain}, \\
with a slide-biting sax, from the son of the sovereign \ken*{= Hain}.”\evb\evg


\bvg\bva Skǫ́ru þęir \alst{h}jarta \hld\ \alst{H}jalla ór brjósti, &
\alst{b}lóðugt, ok á \alst{b}jóð lǫgðu \hld\ ok \alst{b}ǫ́ru þat fyr Gunnar.\eva

\bvb They cut the heart of Helle from the breast, \\
bloody, and on a platter laid it, and bore it before Guther.\evb\evg


\bvg\bva Þá kvað þat \alst{G}unnarr, \hld\ \alst{g}umna dróttinn: &
„\alst{H}ér hęfi’k \alst{h}jarta \hld\ \alst{H}jalla ins blauða, &
ó·líkt \alst{h}jarta \hld\ \alst{H}ǫgna ins frǿkna, &
es mjǫk \alst{b}ifask \hld\ es á \alst{b}jóði liggr; &
\alst{b}ifðisk hǫlfu męirr \hld\ es í \alst{b}rjósti lá!“\eva

\bvb Then quoth this Guther, the lord of men: \\
“Here have I the heart of Helle the soft—unlike the heart of Hain the bold!— \\
which quivers greatly when on the platter it lies; \\
it quivered twice as much when in the breast it lay.”\evb\evg


\bvg\bva \alst{H}ló þá \alst{H}ǫgni \hld\ es til \alst{h}jarta skǫ́ru &
\alst{k}vikvan \alst{k}umbla-smið \hld\ —\alst{k}løkkva síðst hugði. &
\alst{B}lóðugt þat á \alst{b}jóð lǫgðu \hld\ ok \alst{b}ǫ́ru fyr Gunnar.\eva

\bvb Hain then laughed as to the heart they cut \\
the living wound-smith \ken*{\textsc{warrior} = Hain}; he thought least of sobbing. \\
Bloody on a platter they laid it, and bore it before Guther.\evb\evg


\bvg\bva Mę́rr kvað þat \alst{G}unnarr, \hld\ \alst{G}ęir-Niflungr: &
„\alst{H}ér hęfi’k \alst{h}jarta \hld\ \alst{H}ǫgna ins frǿkna, &
ó·líkt \alst{h}jarta \hld\ \alst{H}jalla ins blauða, &
es lítt \alst{b}ifask \hld\ es á \alst{b}jóði liggr; &
\alst{b}ifðisk svá-gi mjǫk \hld\ þá’s í \alst{b}rjósti lá!\eva

\bvb Renowned Guther quoth this, the Spear-Nivling: \\
“Here have I the heart of Hain the bold \\
—unlike the heart of Helle the soft!— \\
which quivers lightly when on the platter it lies; \\
it quivered not so much when in the breast it lay.\evb\evg


\bvg\bva Svá skalt, \alst{A}tli, \hld\ \alst{au}gum fjarri &
sęm \alst{m}unt \hld\ \alst{m}ęnjum verða; &
es und \alst{ęi}num mér \hld\ \alst{ǫ}ll of folgin &
\alst{h}odd Niflunga: \hld\ lifir-a nú \alst{H}ǫgni!\eva

\bvb Thus shalt thou, Attle, be as far from the eyes \\
as thou wilt from the neck-rings. \\
With me alone is hidden all \\
the hoard of the Nivlings—now Hain lives not!\evb\evg


\bvg\bva Ęy vas mér \alst{t}ýja \hld\ meðan vit \alst{t}vęir lifðum, &
nú ’s mér \alst{ę}ngi \hld\ es \alst{ęi}nn lifi’k; &
\alst{R}ín skal \alst{r}áða \hld\ \alst{r}óg-malmi skatna, &
svinn, \alst{ǫ́}s-kunna \hld\ \alst{a}rfi Niflunga.\eva

\bvb I always had doubt when we two lived; \\
now I have none when I alone live. \\
The Rhine shall rule the strife-ore of princes \ken{gold}: \\
the swift [river] the os-born patrimony of the Nivlings!\evb\evg


\bvg\bva Í \alst{v}eltanda \alst{v}atni \hld\ lýsask \alst{v}al-baugar &
\alst{h}ęldr an á \alst{h}ǫndum gull \hld\ skíni \alst{H}úna bǫrnum.“\eva

\bvb In tumbling water will the Welsh bighs gleam, \\
rather than gold on the hands shine for the children of Huns!”\evb\evg

\sectionline

\bvg\bva “Ýkvið ér hvél-vǫgnum, \hld\ haptr ’s nú í bǫndum!”\eva

\bvb “Turn ye the wheel-wagons, the captive is now in bonds!”\evb\evg


\bvg\bva Atli inn ríki\eva

\bvb TODO\evb\evg


\bvg\bva Svá gangi þér\eva

\bvb TODO\evb\evg


\bvg\bva ok meirr þaðan\eva

\bvb TODO\evb\evg


\bvg\bva \alst{L}ifanda gram \hld\ \alst{l}agði í garð, &
þann’s \alst{sk}riðinn vas, \hld\ \alst{sk}atna męngi, &
\alst{i}nnan \alst{o}rmum. \hld\ En \alst{ęi}nn Gunnarr &
\alst{h}ęipt-móðr hǫrpu \hld\ \alst{h}ęndi kníði; &
\alst{g}lumðu stręngir. \hld\ Svá skal \alst{g}olli &
\alst{f}rǿkn hring-drifi \hld\ við \alst{f}ira halda!\eva

\bvb The living prince was laid in the enclosure \\
(which was crawling) by a multitude of warriors \\
(with snakes inside). And Guther alone \\
spitefully struck the harp with his hand; \\
the strings rang out. \emph{So} shall hold \\
a brave ring-strewer his gold from men.\evb\evg


\bvg\bva Dynr vas í garði,\eva

\bvb TODO\evb\evg


\bvg\bva Út gekk þá Guðrún,\eva

\bvb TODO\evb\evg


\bvg\bva Umðu ǫlskálir\eva

\bvb TODO\evb\evg


\bvg\bva Út gekk þá Guðrún,\eva

\bvb TODO\evb\evg


\bvg\bva Skævaði þá in skírleita\eva

\bvb TODO\evb\evg


\bvg\bva Sona hefir þinna, \eva

\bvb TODO\evb\evg


\bvg\bva Kallar-a þú síðan \eva

\bvb TODO\evb\evg


\bvg\bva Ymr varð á bekkjum, \eva

\bvb TODO\evb\evg


\bvg\bva Gulli seri \eva

\bvb TODO\evb\evg


\bvg\bva Ȯ-varr Atli \hld\ móðan hafði sik drukkit; &
vápn hafði hann ękki, \hld\ varnaði-t við Guðrúnu; &
opt vas sá lęikr bętri \hld\ þá’s þau lint skyldu &
optarr of faðmask \hld\ fyr ǫðlingum.\eva

\bvb Unwary Attle had drunk himself tired; \\
he had no weapons; did not beware Guthrun. \\
Oft their play was better when they gently would \\
more often embrace each other before the athlings.\evb\evg


\bvg\bva Hǫ́n \alst{b}ęð \alst{b}roddi \hld\ gaf \alst{b}lóð at drekka, &
\alst{h}ęndi \alst{h}ęl-fu̇ssi, \hld\ ok \alst{h}velpa lęysti; &
\alst{h}ratt fyr \alst{h}allar dyrr \hld\ ok \alst{h}ús-karla vakði, &
\alst{b}randi, \alst{b}rúðr, hęitum; \hld\ þau lét hǫ́n gjǫld \alst{b}rǿðra.\eva

\bvb With a blade she gave the bed blood to drink, \\
—with a hell-eager hand—and loosened the whelps, \\
blocked the doors of the hall and awoke the housecarls, \\
the bride, with hot flame—so she repaid her brothers!\evb\evg


\bvg\bva \alst{Ę}ldi gaf hón \alst{a}lla \hld\ es \alst{i}nni vǫ́ru &
ok frá \alst{m}orði þęira Gunnars \hld\ komnir vǫ́ru ór \alst{M}yrk-hęimi; &
\alst{f}orn timbr \alst{f}ellu, \hld\ \alst{f}jarg-hús ruku, &
\alst{b}ǿr \alst{B}uðlunga, \hld\ \alst{b}runnu ok skjald-męyjar, &
\alst{i}nni; \alst{a}ldr-stamar \hld\ hnigu í \alst{ę}ld hęitan.\eva

\bvb To the fire she gave all who were within \\
and from the murder of Guther’s men had come from Mirkham. \\
Ancient timbers fell, great houses smoked— \\
the settlement of the Budlungs—also the shield–maidens burned \\
inside; short of life, they sunk into hot fire.\evb\evg


\bvg\bva Full-rǿtt’s umb þetta; \hld\ fęrr ęngi svá síðan &
brúðr í brynju \hld\ brǿðra at hęfna; &
hón hęfir þriggja \hld\ þjóð-konunga &
\edtrans{ban-orð borit}{borne the bane-word}{\Bfootnote{\footnoteB{i.e. “she has caused the deaths of three great kings.” This expression and its Germanic and Indo-European relatives is discussed in detail in \textcite{Watkins1995}[417--422].}}}, \hld\ bjǫrt, áðr sylti.\eva

\bvb ’Tis told fully about this: henceforth none will go so, \\
a bride in byrnie her brothers to avenge. \\
She has of three great kings \\
borne the bane-word—bright woman—before she must die.\evb\evg


\bvg\bva Enn segir gløggra í Atlamálum inum grǿn-lenskum.\eva

\bvb Yet says it more clearly in the Greenlendish Speeches of Attle.\evb\evg

\sectionline
