\bookStart{Dreams of Balder}[Baldrs draumar]

% Introduction.
In ancient manuscripts only preserved in \AM, but the poem also survives in later manuscripts in longer form.

\bvg
\bva Sęnn vǫ́ru ę́sir \hld\ allir á þingi &
ok ǫ́synjur \hld\ allar á máli, &
ok of þat réðu \hld\ ríkir tívar: &
hví vę́ri Baldri \hld\ ballir draumar.\eva

\bvb Soon were the \inx[G]{Ease} all at the \inx[C]{Thing}, and the \inx[G]{Ossens} all at speech, and of this counseled the mighty \inx[G]{Tues}:\footnoteB{Identical to \Thrymskvida\ 13.} why for Balder were baleful dreams.\evb
\evg


\bvg
\bva Upp reis Óðinn, \hld\ aldinn gautr, &
ok hann á Slęipni \hld\ sǫðul of lagði, &
ręið niðr þaðan \hld\ niflhęljar til; &
mǿtti hvelpi, \hld\ þęim’s ór hęlju kom.\eva

\bvb Up rose Weden—the aged Geat—and he on \inx[P]{Slapner} the saddle did lay; rode down thence to \inx[L]{Nivelhell}; met the whelp that out of Hell came.\evb
\evg


\bvg
\bva Sá vas blóðugr \hld\ of brjóst framan, &
ok galdrs fǫður \hld\ gól oflęngi, &
framm ręið Óðinn, \hld\ foldvegr dunði, &
hann kom at hǫ́u \hld\ Hęljar ranni.\eva

\bvb That one was bloody on the front of the chest, and at the father of \inx[C]{galder} \ken*{= Weden} for a long time bayed.—Forth rode Weden, the fold-way \ken{earth} resounded;\footnoteB{A similarity may be noted with the description of \inx[P]{Thunder}’s riding in \Haustlong\ 14: \emph{dunði [...] mána vegr und hǫ́num} ‘the moon’s way \ken{heaven} [...] resounded beneath him’) and \Thrymskvida\ 20; see there for more.} he came to the high house of Hell.\evb
\evg


\bvg
\bva Þá ręið Óðinn \hld\ fyr austan dyrr, &
þar’s hann vissi \hld\ vǫlu lęiði; &
nam hann vittugri \hld\ valgaldr kveða, &
unz nauðug ręis, \hld\ nás orð of kvað:\eva

\bvb Then rode Weden east of the door, there as he knew the wallow’s grave; he took to sing a corpse-\inx[C]{galder}\footnoteB{An incantation to wake her up; cf. \Havamal\ TODO spell section.} for the witchcraft-skilled woman, until forced she rose, a corpse’s words did speak:\evb
\evg


\bvg
\bva „Hvat ’s manna þat \hld\ mér ókunnra, &
es mér hęfr aukit \hld\ ęrfitt sinni; &
vas’k snifin snę́vi, \hld\ ok slęgin regni &
ok drifin dǫggu, \hld\ dauð vas’k lęngi.“\eva

\bvb “What sort of man is that, unknown to me, who has caused for me this toilsome walk?\footnoteB{i.e. out of the grave.} I was snowed by snow and struck by rain, and sprayed with dew;\footnoteB{Cf. \HelgakvidaTwo\ 47–48 (TODO).} dead was I for long.”\evb
\evg


\bvg {\small [Weden quoth:]}
\bva „Vegtamr hęiti’k, \hld\ em’k Valtams sonr, &
sęg mér ór hęlju, \hld\ ek ór hęimi mun, &
hvęim eru bękkir \hld\ baugum sánir? &
flęt fagrliga \hld\ flóuð eru golli.“\eva

\bvb “Waytame I am called, I am Waltame’s son. Tell me about Hell—I will [tell] about the world; for whom are the benches sown with \inx[C]{bigh}[bighs]; the fair rooms are flooded with gold.”\evb
\evg


\bvg {\small [Wallow quoth:]}
\bva „Hér stęndr Baldri \hld\ of brugginn mjǫðr, &
skírar vęigar, \hld\ liggr skjǫldr yfir, &
ęn ásmęgir \hld\ í ofvę́ni; &
nauðug sagða’k, \hld\ nú mun’k þęgja.“\eva

\bvb “Here stands brewed for Balder mead, pure draughts—a shield lies over;\footnoteB{Shields covering casks of mead is a common trope.} but the os-lads \ken*{Ease} [stand] in great suspense; forced I spoke, now I will be silent.”\evb
\evg


\bvg {\small [Weden quoth:]}
\bva „Þęgjat vǫlva, \hld\ þik vil’k fregna, &
unz ’s alkunna, \hld\ vil’k ęnn vita, &
hvęrr mun Baldri \hld\ at bana verða, &
ok Óðins son \hld\ aldri rę́na?“\eva

\bvb “Be not silent, wallow! Thee I wish to ask; until all is known I wish to know further: Who will for Balder become the bane, and Weden’s son \ken*{= Balder} rob of life?”\evb
\evg


\bvg {\small [Wallow quoth:]}
\bva „Hǫðr berr hǫ́van \hld\ hróðrbaðm þinig, &
hann mun Baldri \hld\ at bana verða, &
ok Óðins son \hld\ aldri rę́na; &
nauðug sagða’k, \hld\ nú mun’k þęgja.“\eva

\bvb “\inx[P]{Hath} bears the high, renowned beam \ken{mistletoe} thither; he will for Balder become the bane, and Weden’s son \ken*{= Balder} rob of life; forced I spoke, now I will be silent.”\evb
\evg


\bvg {\small [Weden quoth:]}
\bva „Þęgjat vǫlva, \hld\ þik vil’k fregna, &
unz ’s alkunna, \hld\ vil’k ęnn vita, &
hvęrr mun hęipt Hęði \hld\ hęfnt of vinna, &
eða Baldrs bana \hld\ á bál vega?“\eva

\bvb “Be not silent, wallow! Thee I wish to ask; until all is known I wish to know further: Who will for the evil-doing get revenge on Hath, or bear onto the pyre Balder’s bane \ken*{= Hath}?”\evb
\evg


\bvg {\small [Wallow quoth:]}
\bva „Rindr berr Vála \hld\ í vestrsǫlum, &
sá mun Óðins sonr \hld\ einnę́ttr vega, &
hǫnd of þvę́rat \hld\ né hǫfuð kęmbir, &
áðr á bál of berr \hld\ Baldrs andskota; &
nauðug sagðak, \hld\ nú munk þęgja.“\eva

\bvb “Rind bears \inx[P]{Wonnel} in the western halls; that one will, Weden’s son, one night old, fight. His hand he washes not, nor head combs, before onto the pyre he bears Balder’s opponent \ken*{= Hath}; forced I spoke, now I will be silent.\footnoteB{Note the similarity with \Voluspa\ 34–35 and the irregularity of the verse length, which may suggest that a line (most likely 2) has been inserted.}”\evb
\evg


\bvg {\small [Weden quoth:]}
\bva „Þęgjat vǫlva, \hld\ þik vil’k fregna, &
unz ’s alkunna, \hld\ vil’k ęnn vita, &
hvęrjar ’ró męyjar, \hld\ es at muni gráta &
ok á himin verpa \hld\ halsaskautum?“\eva

\bvb “Be not silent, wallow! Thee I wish to ask; until all is known I wish to know further: Which are the maidens that weep as they wish, and onto heaven throw their throat-corners?\footnoteB{Wat mean...}”\evb
\evg


\bvg {\small [Wallow quoth:]}
\bva „Estat Vegtamr, \hld\ sem ek hugða, &
hęldr est Óðinn, \hld\ aldinn gautr.“ &
„Estat vǫlva \hld\ né vís kona, &
hęldr est þriggja \hld\ þursa móðir.\eva

\bvb “Thou art not Waytame, as I thought; rather art thou Weden, the aged Geat!”—“Thou art not a \inx[C]{wallow}, nor a wise woman; rather art thou of three \inx[G]{Thurses} the mother!”\evb
\evg


\bvg {\small [Weden quoth:]}
\bva „Hęim ríð Óðinn \hld\ ok hróðigr ves, &
svá komit manna \hld\ męirr aptr á vit, &
es lauss Loki \hld\ líðr ór bǫndum &
ok ragna rǫk \hld\ rjúfęndr koma.“\eva

\bvb “Ride home Weden, and be renowned!\footnoteB{A sarcastic statement, the sense being: “Your renown, Weden, will not save you.”} So may no other man come again to visit [me], when loose, Lock passes out of his bonds, and [at] the \inx[P]{Rakes of the Reins}, the renders come.\footnoteB{The \emph{rjúfęndr} ‘renders’ are presumably Surt and Lock with his children, as described in \Voluspa\ 40 ff. The root \emph{rjúf-} ‘\CV: to break, rip up, break a hole in’ is also used in this context in \Vafthrudnismal\ TODO: \emph{þá’s rjúfask ręgin} ‘when the \inx[G]{Reins} are rent’, \Grimnismal\ 4, \Lokasenna\ TODO and \Sigrdrifumal\ TODO: \emph{unz (of) rjúfask ręgin} ‘until the Reins are rent’, all of which attest to this formula. Related is likely also.}”\evb
\evg


Late verses in paper manuscripts? TODO
