\bookStart{Speeches of Fathomer}[Fáfnismǫ́l]

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: C10th (0.442)–early C11th (0.402)

\textbf{Meter:} \Ljodahattr, \Fornyrdislag\ (TODO)%
\end{flushright}

\section{Introduction}

The \textbf{Speeches of Fathomer} (\Fafnismal) are only preserved in \Regius, where it has the title \emph{Frá dauða Fáfnis} ‘From the death of Fathomer’.  It directly continues the narrative of \Reginsmal, and is, like that poem, a prosimetrum.

\sectionline

\section{The Speeches of Fathomer}

\bvg\bva „\alst{S}vęinn ok \alst{s}vęinn! \hld\ Hvęrjum est \alst{s}vęini of borinn? &
\ind Hvęrra est \alst{m}anna \alst{m}ǫgr? &
es þú á \alst{F}áfni rautt \hld\ þínn hinn \alst{f}rána mę́ki; &
\ind stǫndumk til \alst{h}jarta \alst{h}jǫrr!“\eva

\bvb\speakernoteb{[Fathomer quoth:]}%
“O swain and swain! To which swain art thou born; \\
\ind of which men art thou the son? \\
When on Fathomer thou hast reddened this thy gleaming blade; \\
\ind the sword stands unto my heart!”\evb\evg


\bpg\bpa Sigurðr dulði nafns síns fyr því at þat var trúa þeira í forneskju at orð feigs manns mę́tti mikit ef hann bǫlvaði óvin sínum með nafni. Hann kvað:\epa

\bpb Siward hid his name, for it was belief of those in ancient times that a \inx[C]{fey} man’s word could do much if he cursed his foe by his name. He quoth:\epb\epg


\bvg\bva „\alst{G}ǫfugt dýr ek hęiti \hld\ en ek \alst{g}ęngit hef’k &
\ind hinn \alst{m}óður-lausi \alst{m}ǫgr, &
\alst{f}ǫður ek á’kk-a \hld\ sem \alst{f}ira synir, &
\ind gęng \alst{e}k \alst{ęi}nn saman.“\eva

\bvb “Noble Deer am I called, and I have gone \\
\ind as the motherless lad. \\
A father I have not like the sons of men; \\
\ind I go alone.”\evb\evg


\bvg\bva „Vęitst, ef \alst{f}ǫður né átt-at \hld\ sem \alst{f}ira synir, &
\ind af hvęrju vastu \alst{u}ndri \alst{a}linn? &
{[...]}“\eva

\bvb\speakernoteb{[Fathomer quoth:]}%
“Knowest thou, if thou hast no father like the sons of men, \\
\ind by which wonder thou wast begotten?”\evb\evg


\bvg\bva „\alst{Ę́}tterni mitt \hld\ kveð’k þér \alst{ȯ}-kunnigt vesa &
\ind ok mik \alst{s}jalfan hit \alst{s}ama: &
\alst{S}igurðr ek hęiti \hld\ \alst{S}igmundr hét minn faðir &
\ind es hęf’k þik \alst{v}ǫ́pnum \alst{v}egit.“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“My lineage, I say, is unknown to thee, \\
\ind and my self the same.\footnoteB{The sense is that Fathomer would not recognize Siward’s lineage (i.e. his father) or name, since he is an orphan who up until this point has not accomplished much. He is not saying that he is lineage is unknown even to himself, since \emph{sjalfan mik} ‘my self’ is accusative, not dative.} \\
Siward am I called—Syemund was called my father— \\
\ind who with weapons have smitten thee.”\evb\evg


\bvg\bva „\alst{H}vęrr þik \alst{h}vatti, \hld\ hví \alst{h}vętjask lést, &
\ind mínu \alst{f}jǫrvi at \alst{f}ara? &
Hinn \alst{f}rán-ęygi svęinn, \hld\ þú áttir \alst{f}ǫður bitran, &
\ind \edtrans{á-bornu skjór á skęið.}{inborn traits show quickly}{\Bfootnote{The original is cryptic.  \emph{á skęið} means roughly ‘rapidly, quickly’, whence the expression \emph{ríða á skęið} ‘\CV: to ride at full speed’, but the other words are uncertain.  \textcite{LaFargeGlossary} read ‘your innate qualities show quickly’, suggesting two unattested words: an adjective \emph{*áborinn} ‘innate, inborn’ and a verb \emph{*skjóa} ‘to show’. Yet the lack of i-umlaut in the supposed 3rd sg. pres. ind. \emph{skjór} is difficult. We would expect \emph{**skýr}, as in \emph{skjóta} ‘to shoot,’ with 2nd/3rd sg. pres. ind \emph{skýtr}. A solution here would be reading a 2nd sg. pres. subj. \emph{skjóir}, with a vowel TODO}}“\eva

\bvb\speakernoteb{[Fathomer quoth:]}%
“Who goaded thee; why didst thou let thee be goaded \\
\ind my life for to destroy? \\
O gleaming-eyed swain, thou hadst a sharp father; \\
\ind inborn traits show quickly!”\evb\evg


\bvg\bva „\alst{H}ugr mik \alst{h}vatti, \hld\ \alst{h}ęndr mér full-týðu &
\ind ok minn inn \alst{h}vassi \alst{h}jǫrr; &
fár es \alst{h}vatr \hld\ es \alst{h}røðask tękr &
\ind ef í \alst{b}arn-esku es \alst{b}lauðr.“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“My heart goaded me; my hands availed me, \\
\ind and this my sharp sword. \\
Few a man is bold when he takes to grow, \\
\ind if he in youth is soft.”\evb\evg


\bvg\bva „Vęit’k, ef þú \alst{v}axa nę́ðir \hld\ fyr þinna \alst{v}ina brjósti, &
\ind séi-t maðr þik \alst{v}ręiðan \alst{v}ega; &
nú ert \alst{h}aptr \hld\ ok \alst{h}ęr-numinn, &
\ind ę́ kveða \alst{b}andingja \alst{b}ifask.“\eva

\bvb\speakernoteb{[Fathomer quoth:]}%
“I know that if thou hadst managed to grow up at the breasts of thy friends, \\
\ind no man would see thee wrathfully fight. \\
Now art thou a captive and war-taken; \\
\ind the boundling is ever said to tremble.”\evb\evg


\bvg\bva „Því bregðr þú nú mér, \alst{F}áfnir, \hld\ at til \alst{f}jarri sjá’k &
\ind \alst{m}ínum fęðr-\alst{m}unum, &
ęigi em’k \alst{h}aptr \hld\ þótt vę́ra \alst{h}ęr-numi; &
\ind þú fannt, at ek \alst{l}auss \alst{l}ifi!“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“For this thou now upbraidest me, Fathomer, that I be too far \\
\ind from my fathers’ love. \\
I am no captive, though I be war-taken; \\
\ind thou hast found that I live loose!”\evb\evg


\bvg\bva „\alst{H}ęipt-yrði ęin \hld\ tęlr þú þér í \alst{h}ví-vętna &
\ind en ek þér \alst{s}att ęitt \alst{s}ęgi’k: &
It \alst{g}jalla \alst{g}ull \hld\ ok it \alst{g}lóð-rauða fé, &
\ind þér verða þęir \alst{b}augar at \alst{b}ana!“\eva

\bvb\speakernoteb{[Fathomer quoth:]}%
“With hateful words alone dost thou answer anything, \\
\ind but I tell thee truth alone: \\
The clanging gold and the glowing red wealth— \\
\ind those bighs will be thy bane!”\evb\evg


\bvg\bva „\alst{F}éi ráða \hld\ skal \alst{f}yrða hvęrr &
\ind \alst{ę́} til \edtrans{ins \alst{ęi}na dags}{the one day}{\Bfootnote{i.e. his predetermined day of death.  Siward dismisses the curse; he must die regardless of whether he takes the gold or not, and it is better to die wealthy and renowned than wretched and unknown.}} &
því-at \alst{ęi}nu sinni \hld\ skal \alst{a}lda hvęrr &
\ind fara til \alst{h}ęljar \alst{h}eðan.“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“Rule his wealth shall every man, \\
\ind ever, until the one day; \\
for at one time shall every man \\
\ind journey hence to Hell.”\evb\evg


\bvg\bva „\alst{N}orna dóm \hld\ munt \edtrans{fyr \alst{n}ęsjum}{before the headlands}{\Bfootnote{i.e. ‘close at hand, imminent’.  An established metaphor for imminent death, cf. the last st. of \Sonatorrek\ (TODO).}} hafa &
\ind ok \alst{ȯ}-svinns \alst{a}pa; &
í \alst{v}atni þú drukknar \hld\ ef í \alst{v}indi rę́r; &
\ind allt es \alst{f}ęigs \alst{f}orað.“\eva

\bvb\speakernoteb{[Fathomer quoth:]}%
“The doom of the Norns shalt thou have before the headlands, \\
\ind and that of an unwise ape. \\
In water wilt thou drown if thou row in wind; \\
\ind everything is the pit of the \inx[C]{fey}.\footnoteB{The man fated to die will find his death no matter where he turns.}”\evb\evg


\bvg\bva „Sęg mér, \alst{F}áfnir, \hld\ alls þik \alst{f}róðan kveða &
\ind ok \alst{v}ęl mart \alst{v}ita: &
Hvęrjar ’ru þę́r \alst{n}ornir \hld\ \edtrans{es \alst{n}auð-gǫnglar ’ru}{attend in need}{\Bfootnote{lit. ‘are attendant in need’, i.e. help ailing mothers during childbirth.  Cf. \Sigrdrifumal\ 9.}} &
\ind ok kjósa \alst{m}ǿðr frá \alst{m}ǫgum?“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“Tell me, Fathomer, as they call thee wise, \\
\ind and knowing well enough: \\
Who are the Norns that attend in need, \\
\ind and choose mothers from their lads?”\evb\evg


\bvg\bva „\alst{S}undr-bornar mjǫk \hld\ hygg at nornir \alst{s}é, &
\ind \alst{ęi}gu-t þę́r \alst{ę́}tt saman; &
sumar ’ru \alst{ǫ́}s-kunngar, \hld\ sumar \alst{a}lf-kunngar, &
\ind sumar \alst{d}ǿtr \alst{D}valins.“\eva

\bvb\speakernoteb{[Fathomer quoth:]}%
“Of most sundry birth I judge the norns to be, \\
\ind they come not from a common lineage: \\
some are Os-born, some Elf-born, \\
\ind some are the daughters of Dwollen \ken{dwarfesses}.”\evb\evg


\bvg\bva „Sęg mér þat, \alst{F}áfnir, \hld\ alls þik \alst{f}róðan kveða &
\ind ok \alst{v}ęl margt \alst{v}ita, &
hvé sá \alst{h}olmr hęitir \hld\ es blanda \alst{h}jǫr-lęgi &
\ind \alst{S}urtr ok ę́sir \alst{s}aman.“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“Tell me this, Fathomer, as they call thee wise, \\
\ind and knowing well enough: \\
What is the islet called, where Surt and the Eese \\
\ind blend sword-water \ken{blood} together?”\evb\evg


\bvg\bva „\alst{Ó}-skópnir hęitir \hld\ en þar \alst{ǫ}ll skulu &
\ind \alst{g}ęirum lęika \alst{g}oð; &
\alst{B}il-rǫst \alst{b}rotnar \hld\ es á \alst{b}rott fara &
\ind ok svima í \alst{m}óðu \alst{m}arir.“\eva

\bvb\speakernoteb{[Fathomer quoth:]}%
“Unshopner it is called, and there shall all \\
\ind the Gods play with spears \ken{make war}; \\
Bilrest shatters when they go away, \\
\ind and the steeds swim in the sea.”\evb\evg

\sectionline

Fathomer continues speaking, but there is probably something missing here, since the transition is abrupt. Between its paraphrases of st. 15 and of st. 16, \VolsungaMS\ has \emph{Ok enn mę́lti Fáfnir: „Reginn bróðir minn veldr mínum dauða, ok þat hlę́gir mik, er hann veldr ok þínum dauða, ok ferr þá, sem hann vildi.“} ‘And further spoke Fathomer: “My brother Rein causes my death, and it gladdens me that he also causes thy death, and then it will go like he has willed.”’, which may perhaps be a paraphrase of a lost st.

\sectionline

\bvg\bva „\alst{Ǿ}gis hjalm \hld\ bar’k of \alst{a}lda sonum &
\ind \alst{m}eðan of \alst{m}ęnjum lá’k; &
\alst{ęi}nn rammari \hld\ hugðumk \alst{ǫ}llum vesa, &
\ind fann’k-a’k \alst{m}arga \alst{m}ǫgu.“\eva

\bvb “The helmet of terror I carried over the sons of men \\
\ind while on the neckrings I lay; \\
stronger than all I thought me alone to be; \\
\ind I did not find many lads.”\evb\evg


\bvg %NOTE: Heavily formulaic.
\bva „\alst{Ǿ}gis hjalmr \hld\ bergr \alst{ęi}nu-gi &
\ind hvar’s skulu \alst{v}ręiðir \alst{v}ega; &
þá þat \alst{f}innr \hld\ es með \alst{f}lęirum kømr &
\ind at \alst{ę}ngi es \alst{ęi}nna hvatastr.“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“The helmet of terror saves no man, \\
\ind wherever wroth ones should fight; \\
this he then finds, when among the many he comes, \\
\ind that none is the boldest of all.”\evb\evg


\bvg\bva „\alst{Ęi}tri ek fnę́sta \hld\ es á \alst{a}rfi lá’k &
\ind \alst{m}iklum \alst{m}íns fǫður.“\eva

\bvb\speakernoteb{[Fathomer quoth:]}%
“Venom I snorted while I lay on the great \\
\ind inheritance of my father.”\evb\evg


\bvg\bva „Inn rammi ormr, \hld\ þú gørðir frę́s mikla &
\ind ok gatst \alst{h}arðan \alst{h}ug; &
\alst{h}ęipt at męiri \hld\ verðr \alst{h}ǫlða sonum &
\ind at þann \alst{h}jalm \alst{h}afi.“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“O mighty wyrm, thou madest a great snort, \\
\ind and didst win a hard heart; \\
greater hatred arises for the sons of men, \\
\ind who might have that helm.”\evb\evg


\bvg\bva „\alst{R}ę́ð’k þér nú, Sigurðr, \hld\ en þú \alst{r}áð nemir &
\ind ok ríð \alst{h}ęim \alst{h}eðan; &
it \alst{g}jalla \alst{g}ull \hld\ ok it \alst{g}lóð-rauða fé, &
\ind þér verða þęir \alst{b}augar at \alst{b}ana!“\eva

\bvb\speakernoteb{[Fathomer quoth:]}%
“I counsel thee now, Siward—and thou oughtst to take the counsel, \\
\ind and ride home hence: \\
The clanging gold and the glowing red wealth— \\
\ind those bighs will be thy bane!”\evb\evg


\bvg\bva „\alst{R}áð ’s þér \alst{r}áðit \hld\ en ek \alst{r}íða mun &
\ind til þęss gulls es í \alst{l}yngvi \alst{l}iggr, &
en þú, \alst{F}áfnir, ligg \hld\ í \alst{f}jǫr-brotum &
\ind \edtrans{þar’s þik \alst{H}ęl \alst{h}afi}{where Hell may have thee}{\Bfootnote{Formulaic. TODO.}}!“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“Thy counsel has been counseled—but I will ride \\
\ind to the gold which in the heather lies; \\
but thou, Fathomer, do lie in the blood-tracks, \\
\ind where Hell may have thee!”\evb\evg


\bvg% NOTE: Pun.
\bva „Ręginn mik \alst{r}éð, \hld\ hann þik \alst{r}áða mun, &
\ind hann mun okkr verða \alst{b}ǫ́ðum at \alst{b}ana; &
\alst{f}jǫr sitt láta \hld\ hygg at \alst{F}áfnir myni; &
\ind þitt varð nú \alst{m}ęira \alst{m}ęgin.“\eva

\bvb\speakernoteb{[Fathomer quoth:]}%
“Rein fooled \emph{me}; he will fool \emph{thee}; \\
\ind he will become the bane of us both! \\
Let up his life I think that Fathomer will— \\
\ind thy strength was now the greater.”\evb\evg


\bpg\bpa Reginn var á brott horfinn meðan Sigurðr vá Fáfni ok kom þá aptr er Sigurðr strauk blóð af sverðinu. Reginn kvað:\epa

\bpb Rein had disappeared while Siward smote Fathomer, and then came back as Siward wiped the blood off the sword. Rein quoth:\epb\epg


\bvg\bva „Hęill þú nú, \alst{S}igurðr, \hld\ nú hęfir \alst{s}igr vegit &
\ind ok \alst{F}áfni of \alst{f}arit; &
\alst{m}anna þęira \hld\ es \alst{m}old troða &
\ind þik kveð’k \alst{ȯ}-blauðastan \alst{a}linn.“\eva

\bvb “Hail thee now, Siward—now thou hast won victory \\
\ind and Fathomer destroyed! \\
Of those men who tread on the earth \\
\ind I declare thee unsoftest begotten.”\evb\evg


\bvg\bva „Þat ’s \alst{ȯ}-víst at vita \hld\ þá’s komum \alst{a}llir saman, &
\ind \alst{s}ig-tíva \alst{s}ynir, &
\ind hvęrr \alst{ȯ}-blauðastr es \alst{a}linn; &
margr es sá \alst{h}vatr \hld\ es \alst{h}jǫr né rýðr &
\ind \alst{a}nnars brjóstum \alst{ï}.“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“’Tis unsure to know, when we all come together, \\
\ind sons of the victory-Tews \ken{men}, \\
\ind who is unsoftest begotten. \\
Many a man is bold who reddens no sword \\
\ind in another’s chest.”\evb\evg


\bvg\bva „\alst{G}laðr ert nú, Sigurðr, \hld\ ok \alst{g}agni fęginn &
\ind es þú þęrrir \alst{G}ram á \alst{g}rasi; &
\alst{b}róður minn \hld\ hęfir þú \alst{b}ęnjaðan &
\ind ok vęld ek þó \alst{s}jalfr \alst{s}umu.“\eva

\bvb {[Rein quoth:]} \\
“Glad art thou now Siward, and in gain rejoicing \\
\ind when thou driest Gram on the grass. \\
My brother hast thou deathly wounded, \\
\ind and yet I myself played some part.”\evb\evg


\bvg\bva „Þú því \alst{r}étt \hld\ es ek \alst{r}íða skyldak &
\ind \alst{h}eilǫg fjǫll \alst{h}innig; &
\alst{f}éi ok \alst{f}jǫrvi \hld\ réði sá inn \alst{f}ráni ormr &
\ind nema þú frýðir mér \alst{h}vats \alst{h}ugar.“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“\emph{Thou} didst counsel that I should ride \\
\ind the holy fells hither. \\
Wealth and life would the gleaming Wyrm rule, \\
\ind unless thou didst brave my bold heart.”\evb\evg


\bvg\bva Þá gekk Reginn at Fáfni ok skar hjarta ór hánum með sverði er Riðill heitir ok þá drakk hann blóð ór undinni eptir.\eva

\bvb Then Siward walked up to Fathomer and cut the heart out of him with the sword called Riddle, and then he drank blood from the wound afterwards.\evb\evg


\bvg\bva „\alst{S}it-tu nú, \alst{S}igurðr, \hld\ en ek mun \alst{s}ofa ganga &
\ind ok halt \alst{F}áfnis hjarta við \alst{f}una! &
\edtrans{\alst{Ęi}skǫld}{heart-strings}{\Bfootnote{An obscure poetic synonym for heart, it is here in the plural. The translation “heart-strings” is probably inaccurate.}} ek vil \hld\ \alst{e}tin láta &
\ind eptir þęnna \alst{d}ręyra \alst{d}rykk.“\eva

\bvb {[Rein quoth:]} \\
“Sit thou now, Siward—but I will go to sleep— \\
\ind and hold Fathomer’s heart by the fire! \\
The heart-strings I wish to eat, \\
\ind after this drink of blood.”\evb\evg


\bvg\bva „\alst{F}jarri þú gekkt \hld\ meðan ek á \alst{F}áfni rauð’k &
\ind minn inn \alst{h}vassa \alst{h}jǫr; &
\alst{a}fli mínu \hld\ átta’k við \alst{o}rms męgin &
\ind meðan þú í \alst{l}yngvi \alst{l}átt.“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“Far didst thou go while I on Fathomer reddened \\
\ind this my sharp sword. \\
\emph{My} strength I held against the Wyrm’s might, \\
\ind while thou in the heather layst.”\evb\evg


\bvg\bva „\alst{L}ęngi \alst{l}iggja \hld\ létir þú þann \alst{l}yngvi í, &
\ind inn \alst{a}ldna \alst{jǫ}tun, &
ef þú \alst{s}verðs né nytir, \hld\ þess es ek \alst{s}jalfr gørða, &
\ind ok þíns ins \alst{h}vassa \alst{h}jǫrs.“\eva

\bvb {[Rein quoth:]} \\
“Lie long in the heather wouldst thou have let \\
\ind this ancient ettin [me], \\
if the blade thou hadst not used, which I myself made, \\
\ind and this thy sharp sword.”\evb\evg


\bvg\bva „\alst{H}ugr es bętri \hld\ en sé \alst{h}jǫrs męgin &
\ind hvar’s \alst{v}ręiðir skulu \alst{v}ega, &
því at \alst{h}vatan mann \hld\ ek sé \alst{h}arð-liga vega &
\ind með \alst{s}lę́vu \alst{s}verði \alst{s}igr.\eva

\bvb\speakernoteb{[Siward quoth:]}%
“Heart is better than might of sword may be \\
\ind wherever worth men should fight, \\
for a bold man I see fighting a hard \\
\ind victory with sluggish sword.\evb\evg


\bvg\bva \alst{H}vǫtum ’s bętra \hld\ en sé ȯ·\alst{h}vǫtum &
\ind í \alst{h}ildi-lęik \alst{h}afask &
\alst{g}lǫðum es betra \hld\ en sé \alst{g}lúpnanda &
\ind \alst{h}vat sem at \alst{h}ęndi kømr.“\eva

\bvb For the bold it is better than it may be for the unbold, \\
\ind in battle-play to hold themselves; \\
for the glad it is better than for the gloomy, \\
\ind whatever comes to their hands.”\evb\evg


\bpg\bpa Sigurðr tók Fáfnis hjarta ok steikði á teini. Er hann hugði at full-steikt vę́ri ok freyddi sveitinn ór hjartanu þá tók hann á fingri sínum ok skynjaði hvárt full-steikt vę́ri. Hann brann ok brá fingrinum í munn sér. En er hjart-blóð Fáfnis kom á tungu hánum ok skildi hann fugls rǫdd. Hann heyrði at igður klǫkuðu á hrísinum. Igðan kvað:\epa

\bpb Siward took Fathomer’s heart and roasted it on a stick. But when he thought that it was fully roasted and the blood frothed out of the heart then he touched it with his finger to see whether it was fully roasted. He burned himself and threw his finger in his mouth. But when the heart blood of Fathomer came on his tongue and he understood the speech of birds. He heard that some tits were chirping in the brushes. The tit quoth:\epb\epg


\bvg\bva „Þar \alst{s}itr \alst{S}igurðr \hld\ \alst{s}vęita stokkinn, &
\alst{F}áfnis hjarta \hld\ við \alst{f}una stęikir; &
\alst{sp}akr þø̇tti mér \hld\ \alst{sp}illir bauga &
ef hann \alst{f}jǫr-sega \hld\ \alst{f}ránan ę́ti.“\eva

\bvb “There sits Siward spattered by blood, \\
\ind Fathomer’s heart by the fire he roasts. \\
Wise would seem me the spiller of rings \\
\ind if he the gleaming life-muscle ate.”\evb\evg


\bvg\bva „Þar liggr \alst{R}ęginn, \hld\ \alst{r}ę́ðr umb við sik, &
vill \alst{t}ę́la mǫg \hld\ þann’s \alst{t}rúir hǫ́num; &
berr af \alst{v}ręiði \hld\ \alst{v}rǫng orð saman, &
vill \alst{b}ǫlva smiðr \hld\ \alst{b}róður hefna.“\eva

\bvb\speakernoteb{The other one:} \\
“There lies Rein, counsels with himself, \\
\ind wants to betray the lad who trusts in him. \\
From wrath he carries ill words together; \\
\ind the smith of bales wants to avenge his brother.”\evb\evg


\bvg\bva „\alst{H}ǫfði skęmmra \hld\ láti hann inn \alst{h}ára þul &
\ind fara til \alst{h}ęljar \alst{h}eðan! &
\alst{Ǫ}llu gulli \hld\ þá kná hann \alst{ęi}nn ráða, &
\ind \alst{f}jǫlð, því’s und \alst{F}áfni lá.“\eva

\bvb\speakernoteb{The third one:} \\
“Shorter by a head he should make the hoary thyle \\
\ind journey hence to Hell! \\
All the gold he can then wield alone: \\
\ind the trove which under Fathomer lay.”\evb\evg


\bvg\bva „TODO“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“TODO”\evb\evg


\bvg\bva „TODO“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“TODO”\evb\evg


\bvg\bva „TODO“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“TODO”\evb\evg


\bvg\bva „TODO“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“TODO”\evb\evg


\bvg\bva „TODO“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“TODO”\evb\evg


\bvg\bva „Verða svá rík skǫp \hld\ at Ręginn skyli &
\ind mitt ban-orð bera &
því at þęir báðir brǿðr \hld\ skulu brá-liga &
\ind fara til Hęljar heðan.“\eva

\bvb\speakernoteb{[Siward quoth:]}%
“The Shapes will not be so strong that Rein should \\
bear my bane-word, \\
for both those brothers shall hurriedly \\
journey hence to Hell.”\evb\evg


\bpg\bpa Sigurðr hjó hǫfuð af Regin ok þá át hann Fáfnis hjarta ok drakk blóð þeira beggja, Regins ok Fáfnis. Þá heyrði Sigurðr hvað igður mæltu:\epa

\bpb Siward struck the head off Rein and then he ate Fathomer’s heart and drank the blood of them both, Rein and Fathomer. Then Siward heard what the tits spoke:\epb\epg


\bvg\bva „\alst{B}itt þú, Sigurðr, \hld\ \alst{b}auga rauða; &
es-a \alst{k}onung-ligt \hld\ \alst{k}víða mǫrgu. &
\alst{M}ęy vęit’k ęina, \hld\ \alst{m}yklu fęgrsta, &
\alst{g}ulli \alst{g}ǿdda, \hld\ ef þú \alst{g}eta mę́ttir.\eva

\bvb “Bind, O Siward, the red bighs; \\
it is not kingly to fear much. \\
I know one maiden fairest by much, \\
endowed with gold—if thou mighst get her!”\evb\evg


\bvg\bva „Liggja til \alst{G}júka \hld\ \alst{g}rǿnar brautir, &
\alst{f}ramm vísa skǫp \hld\ \alst{f}olk-líðǫndum; &
þar hęfir \alst{d}ýrr konungr \hld\ \alst{d}óttur alna, &
þá \alst{m}unt, Sigurðr, \hld\ \alst{m}undi kaupa.“\eva

\bvb “Toward Yivick lie green highways: \\
the Shapes show [the way] forth for wandering exiles. \\
There the wealthy king has begotten a daughter; \\
her wilt thou, Siward, for a bride-fee buy!”\evb\evg


\bvg\bva „Salr ’s ȧ \alst{h}ǫ́u \hld\ \alst{H}indar-fjalli, &
\alst{a}llr ’s hann \alst{ú}tan \hld\ \alst{ę}ldi svęipinn; &
þann \alst{h}afa \alst{h}orskir \hld\ \alst{h}alir um gǫrvan &
\alst{ó}r \alst{ȯ}-dǫkkum \hld\ \alst{ó}gnar ljóma.“\eva

\bvb “A hall is on the high Hinderfell, \\
it is all outside in a fire enwrapped; \\
that one have wise men made \\
from an un-dark radiance of fear.”\evb\evg


\bvg\bva „Vęit’k á \alst{f}jalli \hld\ \alst{f}olk-vitr sofa &
ok \alst{l}ęikr yfir \hld\ \alst{l}indar váði; &
\alst{Y}ggr stakk þorni— \hld\ \alst{a}ðra fęlldi &
\alst{h}ǫr-Gefn \alst{h}ali \hld\ es \alst{h}afa vildi.“\eva

\bvb “I know on the fell a war-wight sleeping \\
and over her licks the linden’s harm \ken{fire}. \\
Ug has stung her with a thorn; the flax-Yevn \ken{lady} felled \\
the other men who wished to have her.”\evb\evg


\bvg\bva „Knátt, \alst{m}ǫgr, séa \hld\ \alst{m}ęy und hjalmi &
þá’s frȧ \alst{v}ígi \hld\ \alst{V}ingskorni ręið; &
má-at \alst{S}igrdrífar \hld\ \alst{s}vefni bręgða, &
\alst{sk}jǫldunga niðr, \hld\ fyr \alst{sk}ǫpum norna.“\eva

\bvb “Thou canst, lad, see a maiden beneath a helmet, \\
she who from the fray rode on Wingshorner. \\
One may not break Syedrive’s sleep, \\
O kinsman of the Shieldings, before the Shapes of the Norns!”\evb\evg


\bpg\bpa Sigurðr reið eptir slóð Fáfnis til bǿlis hans ok fann þat opit ok hurðir af járni ok gę́tti; af járni vóru ok allir timbr-stokkar í húsinu en grafit í jǫrð niðr. Þar fann Sigurðr stór-mikit gull ok fylldi þar tvę́r kistur. Þar tók hann ǿgis-hjálm ok gull-brynju ok sverðit Hrotta ok marga dýr-gripi ok klyfjaði þar með Grana. En hestrinn vildi eigi fram ganga fyrr en Sigurðr steig á bak hǫ́num.\epa

\bpb Siward rode along Fathomer’s trail to his dwelling and found it open and doors and rabbets of iron. Of iron were also all the timber trunks in the house, and dug down into the earth. There Siward found very much gold and filled there two chests. Then he took the helmet of terror and a golden byrnie and the sword Rotte and many precious things and loaded Grane with them. But the horse did not want to go forth before Siward mounted his back.\epb\epg

\sectionline
