\bookStart{First Lay of Hallow Hundingsbane}[Helgakviða Hundingsbana fyrsta]

\begin{flushright}%
Dating \parencite{Sapp2022}: early C11th (0.103), late C12th (0.805)

Meter: \Fornyrdislag%
\end{flushright}%

\sectionline

Here begins the lay regarding the men of Hallow Hundingsbane and Hathbrod. A lay of the Walsings.

\bvg
\bva\mssnote{\Regius~20r/21}Ár vas alda \hld\ þat’s arar gullu &
hnigu hęilǫg vǫtn \hld\ af Himinfjǫllum; &
þá hafði Hęlga \hld\ inn hugumstóra &
Borghildr borit \hld\ í Brálundi.\eva

\bvb It was the beginning of \inx[C]{eld}[elds], as eagles shrieked; holy waters poured down from the Heavenfells; then Burhild in Browlund gave birth to Hallow the Great-hearted.\evb
\evg


\bvg
\bva\mssnote{\Regius~20r/23}Nǫ́tt varð í bǿ, \hld\ nornir kvǫ́mu, &
þę́r’s ǫðlingi \hld\ aldr of skópu; &
þann bǫ́ðu fylki \hld\ frę́gstan verða &
ok buðlunga \hld\ bęztan þykkja.\eva

\bvb It turned night in the settlement; norns came, those who shaped the age of the nobleman \ken*{= Hallow}. They bade the battle-arrayer be the noblest, and among princes seem the best.\evb
\evg


\bvg
\bva\mssnote{\Regius~20r/25}Sneru þę́r af afli \hld\ ørlǫgþǫ́ttu &
þá’s borgir braut \hld\ í Brálundi; &
þę́r um gręiddu \hld\ gullinsímu &
ok und mána sal \hld\ miðjan fęstu.\eva

\bvb They turned mightily orlay-strands, when castles were broken in Browlund. They arranged a golden band, and beneath the moon-hall \ken{sky} fastened it in the middle.\evb
\evg


\bvg
\bva\mssnote{\Regius~20r/27}Þę́r austr ok vestr \hld\ ęnda fǫ́lu, &
þar átti lofðungr \hld\ land á milli, &
brá nipt Nera \hld\ á norðrvega &
ęinni fęsti, \hld\ ęy bað hon halda.\eva

\bvb They in the east and west hid its ends; there the praised one owned land in between. The kinswoman of Nare \name{unknown person} \ken{norn} tugged onto the northern ways a single cord; she bade it hold forever.\evb
\evg
