\bookStart{Goading of Guthrun}[Guðrúnarhvǫt]
\def\thisBookCode{Gudrunarhvot}

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: early C11th (0.781)–late C11th (0.177)

\textbf{Meter:} \Fornyrdislag
\end{flushright}%

\section{Introduction}

TODO: INTRODUCTION.

\sectionline

\section{From Guthrun (\emph{Frá Guðrúnu})}

\bpg\bpa Guðrún gekk þá til sę́var er hon hafði drepit Atla, gekk út á sę́’inn ok vildi fara sér. Hon mátti eigi søkkva. Rak hana yfir fjǫrð’inn á land Jónakrs konungs. Hann fekk hennar. Þeira synir vóru þeir Sǫrli ok Erpr ok Hamðir. Þar fǿddisk upp Svan·hildr Sig·urðar dóttir. Hon var gift Jǫrmun·rekk inum ríkja. Með hánum var Bikki. Hann réð þat at Rand·vér konungs son skyldi taka hana; þat sagði Bikki konungi. Konungr lét hengja Rand·vé en troða Svan·hildi undir hrossa fótum. En er þat spurði Guð·rún þá kvaddi hon sonu sína.\epa

\bpb Guthrun then went to the sea after she had slain Attle, walked out into the sea and would end herself.  She could not sink.  She was driven across the firth to the land of king Enacker. He got her for wife.  Their sons were Sarle and Earp and Hamthew.  There Swanhild, Siward’s daughter, was raised.  She was married to Ermenric the powerful.  With him was \inx[P]{Bicke}. He counseled that Randwigh, the king’s son, should rape her; this Bicke told the king. The king had Randwigh hanged and Swanhild trampled beneath the hooves of horses. But when Guthrun learned of this she called on her sons.\epb\epg

\sectionline

\section{The Goading of Guthrun}

\bvg\bva%
Þá frá’k \alst{s}ęnnu \hld\ \alst{s}líðr-fęng-ligasta, &
\alst{t}rauð mǫ́l \alst{t}alit \hld\ af \alst{t}rega stórum, &
es \alst{h}arð-\alst{h}uguð \hld\ \alst{h}vatti at vígi &
\alst{g}rimmum orðum \hld\ \alst{G}uðrún sonu:\eva

\bvb That gibing I’ve found most direly caught— \\
hard-pressed speeches told from great grief— \\
when hard-hearted she goaded to war, \\
with fierce words, Guthrun, her sons:\evb\evg


\bvg\bva%
„Hví \alst{s}itið? \hld\ Hví \alst{s}ofið lífi? &
Hví \alst{t}regr-at ykkr \hld\ \alst{t}ęiti at mę́la? &
\edtext{es \alst{Jǫ}rmunrekr \hld\ \alst{y}ðra systur, &
\alst{u}nga at \alst{a}ldri, \hld\ \alst{j}óm of traddi, &
\alst{h}vítum ok svǫrtum \hld\ ȧ \alst{h}ęr-vegi &
\alst{g}rǫ́m, \alst{g}ang-tǫmum \hld\ \alst{G}otna hrossum.}{\lemma{es \dots\ hrossum. ‘when \dots\ horses!’}\Bfootnote{Repeated almost identically in \Hamdismal\ 3.}}\eva

\bvb “Why sit ye two? Why sleep ye your lives away? \\
Why troubles it you not to speak merrily? \\
when Ermenric has had your sister \\
young of age trampled by steeds, \\
by white ones and black on the path of war, \\
by grey, pacing, Gotnish horses!\evb\evg


TODO: Stanzas.


\bvg\bva%
„Berið \alst{h}nossir framm \hld\ \alst{h}ún-konunga, &
\alst{h}ęfir þú okkr \alst{h}vatta \hld\ at \alst{h}jǫr-þingi.“\eva

\bvb “Bring forth the treasures of the Hunnish kings! \\
Thou hast incited us to the thing of swords \ken{battle}.”\evb\evg


\bvg\bva%
\alst{H}lę́jandi Guð·rún \hld\ \alst{h}varf til skęmmu, &
\alst{k}umbl \alst{k}onunga \hld\ ór \alst{k}ęrum valði, &
\alst{s}íðar brynjur \hld\ ok \alst{s}onum fǿrði; &
hlóðusk \alst{m}óðgir \hld\ á \alst{m}ara bógu.\eva

\bvb Laughing, Guthrun turned to her chamber \\
the heirlooms of kings from the chests she picked, \\
the long byrnies, and to her sons brought them; \\
gloomy, they loaded themselves on the backs of steeds.\evb\evg


\bvg\bva%
Þá kvað þat \alst{H}amðir \hld\ inn \alst{h}ugum stóri: &
„Svá kom-a’k \alst{m}ęirr aptr \hld\ \alst{m}óður at vitja &
\alst{g}ęir-Njǫrðr hniginn \hld\ ȧ \alst{G}oð-þjóðu &
at þú \alst{ę}rfi \hld\ at \alst{ǫ}ll oss drykkir, &
at \alst{S}van·hildi \hld\ ok \alst{s}onu þïna.“\eva

\bvb Then quoth this Hamthew the great of heart: \\
“So will I no more come back to visit my mother, \\
{[when I,]} the spear-Nearth \ken{warrior}, fall in the land of the Gots, \\
after thou drinkest a wake for us all, \\
for Swanhild and thy sons.”\evb\evg


\bvg\bva%
\alst{G}uð·rún \alst{g}rátandi, \hld\ \alst{G}júka dóttir, &
gekk \alst{t}reg-liga \hld\ ȧ \alst{t}ȧi sitja &
ok at \alst{t}ęlja, \hld\ \alst{t}ǫ́rug-hlýra, &
\alst{m}óðug spjǫll \hld\ ȧ \alst{m}argan veg:\eva

\bvb Weeping, Guthrun Yivick’s daughter \\
went grieving to sit on the ground, \\
and to tell with teary cheeks \\
her gloomy tale in many ways:\evb\evg


\bvg\bva%
„Þrjȧ vissa’k \alst{ę}lda, \hld\ þrjȧ vissa’k \alst{a}rna, &
\alst{v}as’k þrimr \alst{v}erum \hld\ \alst{v}egin at húsi; &
\alst{ęi}nn vas mér Sig·urðr \hld\ \alst{ǫ}llum bętri &
es \alst{b}rǿðr mínir \hld\ at \alst{b}ana urðu.\eva

\bvb “I knew three fires; I knew three hearths; \\
for three husbands I was brought to the house. \\
Alone was Siward to me better than all— \\
he whose bane my brothers became.\evb\evg


\bvg\bva%
\alst{S}várra \alst{s}ára \hld\ \alst{s}á’k-at ek, né kunn\emph{a} &
\alst{m}ęirr þȯttusk \hld\ \alst{m}ér of stríða &
es mik \alst{ǫ}ðlingar \hld\ \alst{A}tla gǫ́fu.\eva

\bvb Heavy wounds I neither saw nor knew; \\
TODO.\evb\evg


\bvg\bva%
\edtrans{\alst{H}u̇na \alst{h}vassa}{The smart bear-cubs}{\Bfootnote{Her sons with Attle; for the appellation cf. \Volundarkvida\ 24/1.}} \hld\ \alst{h}ét’k mér at ru̇num; &
mátti’g-a’k \alst{b}ǫlva \hld\ \alst{b}ǿtr of vinna &
áðr ek \alst{h}nóf \alst{h}ǫfuð \hld\ af \edtext{\emph{\alst{h}}niflungum}{\Afootnote{metr. emend.; \emph{niflungum} \Regius}}.\eva

\bvb The smart bear-cubs I summoned to me for counsel; \\
I could not find remedies for the bales \\
before I severed the heads from the Nivlings.\evb\evg


\bvg\bva%
\alst{G}ekk ek til strandar, \hld\ \alst{g}rǫm vas’k nornum, &
vilda’k hrinda \hld\ stríð grið þęirra; &
\alst{h}ófu mik, né drękkðu, \hld\ \alst{h}ávar bǫ́rur, &
því \alst{l}and of sté’k \hld\ at \alst{l}ifa skylda’k.\eva

\bvb I walked to the shore; I was wroth against the norns; \\
I wished to break their stubborn peace. \\
The high waves lifted me—drowned me not; \\
thus I stepped aland since I was meant to live.\evb\evg


\bvg\bva%
Gekk ek ȧ \alst{b}ęð \hld\ —hugða’k mér fyr \alst{b}ętra— &
\alst{þ}riðja sinni \hld\ \alst{þ}jóð-konungi; &
\alst{ó}l ek mér \alst{jó}ð, \hld\ \alst{ę}rfi-vǫrðu &
{[...]} \hld\ Jónakrs \edtext{sonu}{\Bfootnote{emend.; \emph{sonum} \Regius}}.\eva

\bvb I mounted the bed (I thought it better for me) \\
of a great king for a third time. \\
I begot babes, heritance-defenders \ken{sons}, \\
{[...]}, Enacker’s sons.\evb\evg


\bvg\bva%
En umb \alst{S}van·hildi \hld\ \alst{s}ǫ́tu þýjar, &
es mïnna \alst{b}arna \hld\ \alst{b}atst full-hugða’k; &
\alst{s}vá vas \alst{S}van·hildr \hld\ ï \alst{s}al mïnum, &
\alst{s}em vę́ri \alst{s}ǿm-lęitr \hld\ \alst{s}ólar gęisli.\eva

\bvb But around Swanhild sat handmaids, \\
she whom of my children I fully loved the most. \\
So was Swanhild in my hall \\
like were she a beautiful sun-ray.\evb\evg


\bvg\bva%
\alst{G}ǿdda’k \alst{g}ulli \hld\ ok \alst{g}uð-vęfjum, &
áðr ek \alst{g}ę́fa’k \hld\ \alst{G}oð-þjóðar til; &
þat ’s mér \alst{h}arðast \hld\ \alst{h}arma mïnna &
of þann inn \alst{h}víta \hld\ \alst{h}add Svan·hildar: &
\alst{au}ri trǫddu \hld\ und \alst{jó}a fótum.\eva

\bvb I endowed her with gold and godweave \\
before I gave her to the folk of the Gots. \\
That is the hardest of my harms, \\
over the white hair of Swanhild— \\
in mud they trampled it beneath the hooves of steeds.\evb\evg


\bvg\bva%
En sá \alst{s}árastr, \hld\ es þęir \alst{S}ig·urð mïnn, &
\alst{s}igri rę́ntan, \hld\ ï \alst{s}ę́ing vǫ́gu; &
en sá \alst{g}rimmastr, \hld\ es þęir \alst{G}unn·ari, &
\alst{f}rȧnir ormar, \hld\ til \alst{f}jǫrs skriðu; &
en sá \alst{h}vassastr, \hld\ es til \alst{h}jarta &
\alst{k}onung ȯ-blauðan \hld\ \alst{k}vikvan skǫ́ru.\eva

\bvb But the sorest when my Siward \\
robbed of victory in his bed they slew, \\
and the cruelest when for Guther \\
gleaming serpents slithered unto his life, \\
and the sharpest when unto the heart \\
the unsoft king \ken*{= Hain}, alive, they cut up.\evb\evg


\bvg\bva%
Fjǫlð man’k bǫlva, \hld\ [...] &
\alst{b}ęit-tu, Sig·urðr, \hld\ inn \alst{b}lakka mar, &
\alst{h}ęst inn \alst{h}rað-fǿra \hld\ lát-tu \alst{h}inig renna! &
\alst{S}itr ęigi hér \hld\ \alst{s}nǫr né dóttir &
sú’s \alst{G}uð·rúnu \hld\ \alst{g}ę́fi hnossir.\eva

\bvb I recall a multitude of bales; [...] \\
Saddle, Siward, thy fallow steed, \\
the quick-pacing horse—let him run hither! \\
Here sits no son’s wife nor daughter \\
who to Guthrun might give treasures.\evb\evg


\bvg\bva%
\alst{M}inns-tu, Sig·urðr, \hld\ hvat vit \alst{m}ę́ltum &
þȧ’s vit ȧ \alst{b}ęð \hld\ \alst{b}ę́ði sǫ́tum? &
at þú \alst{m}yndir \alst{m}ïn \hld\ \alst{m}óðugr vitja, &
\alst{h}alr, ór \alst{h}ęlju, \hld\ en ek þín ór \alst{h}ęimi.\eva

\bvb Recallest thou, Siward, what we said \\
when on the bed we both did sit, \\
that thou wouldst me, gloomy man, \\
visit from Hell—and I thee from the world?\evb\evg


\bvg\bva%
Hlaðið ér, \alst{ja}rlar, \hld\ \alst{ęi}ki-kǫst’inn, &
látið þann und \edtrans{\alst{h}i\emph{mn}i}{heaven}{\Afootnote{emend.; \emph{hilmi} ‘prince’ \Regius}} \hld\ \alst{h}ę́stan verða! &
Męgi \alst{b}ręnna \alst{b}rjóst \hld\ \alst{b}ǫlva-fullt ęldr &
umb hjarta [\dots] \hld\ þiðni sorgir!“\eva

\bvb Load, ye earls, the oaken pile \ken{pyre}! \\
Let it become the highest beneath the heaven! \\
May fire burn my curse-filled chest, \\
unto the heart \dots\ may the sorrows melt away!”\evb\evg


\bvg\bva%
\alst{Jǫ}rlum \alst{ǫ}llum \hld\ \alst{ó}ðal batni, &
\alst{s}nótum ǫllum \hld\ \alst{s}org at minni &
at þetta \alst{t}reg-róf \hld\ of \alst{t}alit vę́ri.\eva

\bvb For all earls may patrimony improve; \\
for all ladies may sorrow decrease \\
after this grief-chain has been recounted!\evb\evg

\sectionline
