\bookStart{Speeches of Hamthew}[Hamðismǫ́l]
\def\thisBookCode{Hamdismal}

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: C10th (0.885)

\textbf{Meter:} \Fornyrdislag, \Malahattr% TODO
\end{flushright}%

\section{Introduction}

Two poems? TODO: discuss the stoning of the sons of Enacker in Brage and Jordanes.

\sectionline

\section{The Speeches of Hamthew}

\bvg\bva Spruttu ȧ \alst{t}ái \hld\ \alst{t}regnar iðir &
\alst{g}rǿti alfa \hld\ in \alst{g}lý-stǫmu &
ár of \alst{m}orgin \hld\ \alst{m}anna bǫlva &
\alst{s}útir hvęrjar \hld\ \alst{s}org of kvęykva.\eva

\bvb TODO.\evb\evg


\bvg\bva Vas-a þat \alst{n}ú \hld\ \alst{n}é ï gę́r &
þat hęfir \alst{l}angt \hld\ \alst{l}iðit síðan &
es \alst{f}átt \alst{f}ornara \hld\ \alst{f}ręmr vas þat hǫlfu &
es hvatti \alst{G}uðrún \hld\ \alst{G}júka borin &
\alst{s}onu \alst{s}ína unga \hld\ at hęfna \alst{S}van-hildar.\eva

\bvb That was not now nor yesterday; \\
a long time has passed since; \\
little is older; TODO; \\
when Guthrun, born to Yivick, goaded \\
on her young sons to avenge Swanhild.\evb\evg


\bvg\bva „\alst{S}ystir vas ykkur \hld\ \alst{S}van-hildr of hęitin &
sú’s \alst{Jǫ}rmunrekr \hld\ \alst{jó}m of traddi &
\alst{h}vítum ok svǫrtum \hld\ ȧ \alst{h}ęr-vegi &
\alst{g}rǫ́m \alst{g}ang-tǫmum \hld\ \alst{g}otna hrossum.\eva

\bvb „The sister of you both was called Swanhild, \\
she whom Erminric trampled with steeds; \\
with whites and blacks on the war-path, \\
with grey, pacing, Gotnish horses!\evb\evg


\bvg\bva Ęptir ’s ykkr \alst{þ}rungit \hld\ \alst{þ}jóð-konunga, &
lifið \alst{ęi}nir \alst{é}r þátta \hld\ \alst{ę́}ttar minnar; &
\alst{ęi}n-stǿð em’k \alst{o}rðin \hld\ sęm \alst{ǫ}sp ï holti, &
\alst{f}allin at \alst{f}rę́ndum \hld\ sem \alst{f}ura at kvisti, &
\alst{v}aðin at \alst{v}ilja \hld\ sem \alst{v}iðr at laufi, &
þȧ’s in \alst{k}vist-skǿða \hld\ \alst{k}ømr umb dag varman.”\eva

\bvb TODO\evb\evg


\bvg\bva Hitt kvað þá \alst{H}amðir \hld\ inn \alst{h}ugum stóri: &
„\alst{L}ítt myndir þȧ, Guðrún, \hld\ \alst{l}ęyfa dǫ́ð Hǫgna &
es þęir \alst{S}igurð \hld\ \alst{s}vefni ór vǫkðu; &
sats-tu ȧ \alst{b}ęð \hld\ en \alst{b}anar hlógu.\eva

\bvb TODO\evb\evg


\bvg\bva \alst{B}ǿkr vóru þïnar \hld\ inar \alst{b}lá-hvítu &
ofnar \alst{v}ǫlundum, \hld\ flutu ï \alst{v}ers dreyra; &
\alst{s}valt þá \alst{S}igurðr, \hld\ \alst{s}atst yfir dauðum, &
\alst{g}lýja né \alst{g}áðir; \hld\ \alst{G}unnarr þér svá vildi.\eva

\bvb TODO\evb\evg


\bvg\bva \alst{A}tla þȯttisk þú stríða \hld\ at \alst{E}rps morði &
ok at \alst{Ęi}tils \alst{a}ldr-lagi, \hld\ þat vas þér \alst{e}nn verra! &
Svá skyldi hvęrr \alst{ǫ}ðrum \hld\ vęrja til \alst{a}ldr-laga &
\alst{s}verði \alst{s}ár-bęitu \hld\ at \alst{s}ér né stríddi-t.“\eva

\bvb TODO\evb\evg


\bvg\bva Hitt kvað þȧ \alst{S}ǫrli, \hld\ \alst{s}vinna hafði hann hyggju: &
Vil’k-at við \alst{m}óður \hld\ \alst{m}ǫ́lum skipta; &
\alst{o}rðs þikkir \alst{e}nn vant \hld\ \alst{y}kkru hvǫ́ru; &
hvęrs biðr nú, \alst{G}uðrún, \hld\ er at \alst{g}ráti né fǿr-at?\eva

\bvb TODO\evb\evg


\bvg\bva \alst{B}rǿðr grát þú þína \hld\ ok \alst{b}uri svása, &
\alst{n}iðja \alst{n}á-borna \hld\ lęidda \alst{n}ę́r rógi; &
okkr skalt ok, \alst{G}uðrún, \hld\ \alst{g}ráta báða &
es hér sitjum \alst{f}ęigir á mǫrum, \hld\ \alst{f}jarri munum dęyja.\eva

\bvb TODO\evb\evg


\bvg\bva \alst{G}engu ór \alst{g}arði \hld\ \alst{g}ǫrvir at eiskra;
liðu þá \alst{y}fir \alst{u}ngir \hld\ \alst{ú}rig fjǫll
\alst{m}ǫrum hún-lęndskum, \hld\ \alst{m}orðs at hefna.\eva

\bvb TODO\evb\evg


\bvg\bva Þȧ kvað þat \alst{E}rpr \hld\ ęinu sinni, &
\alst{m}ę́rr of lék \hld\ á \alst{m}ars baki: &
„Illt ’s \alst{b}lauðum hal \hld\ \alst{b}rautir kenna;” &
kóðu \alst{h}arðan mjǫk \hld\ \alst{h}ornung vesa.\eva

\bvb TODO\evb\evg


\bvg\bva Fundu á \edtrans{strę́ti}{street}{\Bfootnote{A Roman loanword used specifically for a paved road or a street in a town.  It probably emphasises the southern geography in which these events transpire.}} \hld\ stór-brǫgð-óttan: &
„Hvé mun jarp-skammr \hld\ \alst{o}kkr full-tingja?“\eva

\bvb They found on the street the most clever one: \\
“How can the short brownhair assist us?”\evb\evg


\bvg\bva \alst{S}varaði inn \alst{s}undr-mǿðri, \hld\ \alst{s}vá kvaðsk vęita myndu &
\alst{f}ull-ting \alst{f}rę́ndum \hld\ sem \alst{f}ótr ǫðrum. &
„Hvat męgi \alst{f}ótr \hld\ \alst{f}ǿti vęita &
né \alst{h}old-gróin \hld\ \alst{h}ǫnd annarri?“\eva

\bvb He of sundry mother answered; said that he would \\
give his kinsmen assistance like a foot another.— \\
“What can a foot give to a foot, \\
or a flesh-grown hand to another?”\evb\evg


\bvg\bva Drógu ór \alst{sk}íði \hld\ \alst{sk}íði-éarn, &
\alst{m}ę́kis ęggjar \hld\ \edtrans{at \alst{m}un flagði}{at the behest of the ogress}{\Bfootnote{They were acting according to the will of a cruel norn (\emph{flagð}); cf. the next-to-last st./4b (TODO), \Reginsmal\ 2, \Sigurdskamma\ 7.}}; &
\alst{þ}vęrrðu \alst{þ}rótt sinn \hld\ at \alst{þ}riðjungi, &
létu \alst{m}ǫg ungan \hld\ til \alst{m}oldar hníga.\eva

\bvb They drew from their sheaths their sheath-irons \ken{swords}, \\
the blade’s edges, at the behest of the ogress. \\
They used up their strength on the third brother; \\
they let the young lad sink down to the earth.\evb\evg


...TODO: more stanzas...


\bvg\bva \alst{H}itt kvað þȧ \alst{H}amðir \hld\ inn \alst{h}ugum stóri: &
„\alst{Ǿ}stir \alst{Jǫ}rmunrekkr \hld\ \alst{o}kkarrar kvǫ́mu &
\alst{b}rǿðra sam-mǿðra \hld\ innan \alst{b}orgar þïnnar; &
fǿtr \alst{s}ér þú þïna \hld\ hǫndum \alst{s}ér þú þïnum  &
\alst{Jǫ}rmunrekkr \alst{o}rpit \hld\ ï \alst{ę}ld hęitan!“\eva

\bvb This then Hamthew quoth, the great of heart: \\
“Thou didst incite, O Erminric, that we both came, \\
brothers of the same mother, into thy stronghold! \\
Thou seest thy feet, thou seest them with thy hands, \\
O Erminric, hurled into the hot fire!”\evb\evg


\bvg\bva Þȧ \edtext{\alst{r}aut}{\Afootnote{metr. emend.; \emph{hraut} \Regius}} við \hld\ inn \edtrans{\alst{r}ęgin-kunngi}{Reins-begotten}{\Bfootnote{Born of the gods.  As attested by Jordanes, the Gotnish Amals, the house to which Erminric belonged, ultimately claimed descent from the hero \emph{Gapt}.  He is, however, easily identified with ON \emph{Gautr} ‘\inx[P]{Geat}’, a name for the god \inx[P]{Weden}.}} &
\alst{B}aldr ï \alst{b}rynju \hld\ sem \alst{B}jǫrn hryti: &
„\alst{G}rýtið ér ȧ \alst{g}umna \hld\ alls \alst{g}ęirar né bíta &
\alst{ę}ggjar né \alst{jǫ́}rn \hld\ \alst{Jó}nakrs sonu!“\eva

\bvb Then roared back the Reins-begotten \\
Balder in his byrnie \ken*{\textsc{warrior} > = Erminric} like a roaring bear: \\
“Stone ye the men!—for spears will not bite \\
—no edges nor irons—Enacker’s sons!”\evb\evg

... TODO ...


\bvg\bva \alst{V}ęl hǫfum vit \alst{v}egit, \hld\ stǫndum á \alst{v}al Gotna &
\alst{o}fan \alst{ę}gg-móðum \hld\ sem \alst{ę}rnir á kvisti; &
\alst{g}óðs hǫfum tírar fengit \hld\ þótt skylim nú eða í \alst{g}ę́r dęyja, &
\alst{k}vęld lifir maðr ekki \hld\ ęftir \alst{k}við norna.\eva

\bvb “Well have we two fought, we stand on the corpses of the Gots: \\
above the edge-weary \ken{slain} like eagles on a branch. \\
We’ve won great glory, even if we should die now or tomorrow— \\
man lives not one evening after the verdict of the norns!”\evb\evg


\bvg\bva Þar fell \alst{S}ǫrli \hld\ at \alst{s}alar gafli, &
en \alst{H}amðir \alst{h}né \hld\ at \alst{h}ús-baki.\eva

\bvb There fell Sarrel by the hall’s gables, \\
and Hamthew sank down by the back of the house.\evb\evg

\sectionline
