\bookStart{Speeches of the High One}[Hávamǫ́l]

\begin{flushright}%
\textbf{Dating:} See individual sections.

\textbf{Meter:} \Ljodahattr, \Galdralag, \Malahattr
\end{flushright}%

\section{Introduction}

The \textbf{Speeches of the High One} (\Havamal) is the second poem of \Regius, where it follows \Voluspa\ and is followed by \Vafthrudnismal.  \Regius\ is the only medieval witness manuscript for the whole poem, but a few sts. are cited or alluded to in other places, viz. st. 1 in \Gylfaginning, st. 76 in \Hakonarmal\ 21, and st. 84 in \FostrbroedhraSaga\ TODO.

The text before us does not very much seem like a unitary composition, but rather more like a collection of scattered traditional poetry connected by an association with the god Weden in his role as giver of wisdom.  It seems to contain at least two poems of practical life advice, two mythological narratives, scattered gnomic poetry about runes, and a list of galders.

I identify the following “strands”, excepting various lone sts. that are probably later inserts.  In the present edition each of the following is given a separate, short introduction:

\begin{enumerate}
	\item 1–80 The Guest-strand; practical life advice
  \item 81–90 Various scattered sts. of advice
  \item 91–102 Weden’s failed seduction of Billing’s daughter
  \item 103–110 Weden’s obtaining of the Mead of Poetry
  \item 111–137 The Speeches of Loddfathomer; Weden’s advice to Loddfathomer
  \item 138–146 The Rune-tally; various sts. about runes and ritual
  \item 146–165 The Leed-tally; Weden’s listing of 18 galders
\end{enumerate}

It cannot be claimed for certain that these are all separate compositions, but it seems unlikely that e.g. the Guest-Strand and the Speeches of Loddfathomer were originally part of the same large poem; there is so much repetition from the Guest-Strand to the Speeches that the latter seems quite redundant, not to mention the fact that the former has a perfectly fitting conclusion in the form of sts. 76–77.

Now if these indeed are originally separate compositions, there arise some important questions about the circumstances of their redaction into a single poem.

To begin with we may think about the nature of the redaction.  It is surely the case that it need not have happened all at once, but could have taken place successively in the form of layered additions.  Thus the original Guest-Strand probably ended at st. 77, but sts. 78–90 may have been added shortly after, later the two narratives about Weden’s romantic escapades, thereafter the Speeches of Loddfathomer, et c.

A second question is about the context in which the redaction took place; whether in an oral or scribal context, in a Heathen hove or a Catholic monastery.  With regard to this question some clues can be gleaned from a careful reading of the final stanza, 166, which, given that it explicitly mentions the title of the poem, must probably be associated with the final layer of the redaction.  Its blessing of the reciter, hearers, and learners indicate a context of oral transmission, and its invocation of the Heathen dichotomy between the Gods and Ettins (the friends and enemies of Mankind, respectively) and description of the contents of the poem (which include unambiguous Heathen ritual instructions) as “very useful” (\emph{all-þǫrf}) suggest that it was put together by a Heathen.  With this in mind, the poem should have been redacted into something very close to its present form no later than the early 11th century, in an oral, Heathen context.

We may third think about what reason someone would have for redacting numerous disparate sources into a single poem.  The mention of “usefulness” suggests that the reason were utilitarian rather than antiquarian, and a picture then emerges of a redactor arranging a corpus of traditional Odinic poetry deemed “useful”—whether for its practical advice or its mythological subject matter—into a single poem, which would then be learned by heart as a whole and recited for spiritual purposes.  In practice this of course served to create a sort of Odinic “ark”—or “Hoardmimer’s wood”!—in which the bulk of the surviving pre-Christian Norse advice poetry was preserved.  Forever lost were those scattered stanzas which were not included—and many such must have existed.

\sectionline

\section{The Guest-strand (1–79)}

The Guest-Strand (Old Norse: \emph{Gęsta-þáttr}) is a collection of wisdom poetry, taking its outset in the scenario of a lone wanderer’s arriving as a stranger at a farmstead.  It begins by discussing the mutual responsibilites between guest and host, before moving on to describing proper conduct in broader human interactions with a particular focus on drinking, speech, and friendship.

While there are some fine transitions employed in order to move from one theme to another (e.g. between sts. 4–5, or 10–11), there is no clear division into thematic sections, and previous subjects often appear again after having been dropped for a few stanzas.  The spirit of the advice is in any case very consistent throughout, and the poem at all turns advices caution and shrewdness.  Of particular importance is the idea of “manwit” (ON \emph{man-vit}), a word somewhat analogous with the English “common sense” or “street wisdom”.

It seems very likely that the original Guest-Strand ended at st. 77.  This finds strong support in \Hakonarmal\ 21, the final st. of that poem, which likewise begins with the first two lines \emph{dęyr fé \hld\ dęyja frę́ndr}.

\sectionline

\bvg\bva\alst{G}ȧttir allar \hld\ áðr \alst{g}angi framm &
\ind \edtext{of \alst{sk}oðask \alst{sk}yli,}{\Bfootnote{om. \GylfMS}} &
\ind of \alst{sk}yggnask \alst{sk}yli; &
því-at \alst{ȯ}-víst ’s at vita, \hld\ hvar \alst{ȯ}-vinir &
\ind sitja ȧ \alst{f}lęti \alst{f}yrir.\eva

\bvb All doorways—before one might go forth— \\
\ind he should spy round; \\
\ind he should pry round; \\
for it is unsure to know where enemies \\
\ind sit on the benches within.\evb\evg


\bvg\bva \alst{G}efęndr hęilir, \hld\ \alst{g}ęstr ’s inn kominn, &
\ind hvar skal \alst{s}itja \alst{s}já? &
mjǫk es \alst{b}ráðr \hld\ sá’s \edtrans{ȧ \alst{b}rǫndum}{on the fires}{\Bfootnote{Possibly referring a Norwegian folk custom, wherein a guest would sit down on the wood-pile outside of the door, waiting until being let in; see further TODO SOME ARTICLE on this custom.  The speaker is announcing to the hosts (or “givers”) that a guest, frozen, wet and tired, is currently sitting on the wood-pile, and ought to be let in.}} skal &
\ind \edtrans{síns of \alst{f}ręista \alst{f}rama}{test his furtherance}{\Bfootnote{Try his luck, see how far he gets.  The same line is also found in \Vafthrudnismal\ 11, 13, 15, 17.}}.\eva

\bvb O givers, hail! A guest is come in; \\
\ind where shall this one sit? \\
Most hurried is he who on the fires shall \\
\ind test his furtherance.\evb\evg


\bvg\bva\alst{Ę}lds es þǫrf \hld\ þęim’s \alst{i}nn es kominn &
\ind ok ȧ \alst{k}néi \alst{k}alinn, &
\alst{m}atar ok váða \hld\ es \alst{m}anni þǫrf, &
\ind þęim’s hęfr of \alst{f}jall \alst{f}arit.\eva

\bvb Of fire there is need for him who is come in \\
\ind and cold about the knee; \\
of food and of clothing there is need for the man \\
\ind who over the fell has fared.\evb\evg


\bvg\bva\edtext{\alst{V}ats es þǫrf \hld\ þęim’s til \alst{v}erðar kømr, &
\ind \alst{þ}ęrru ok \alst{þ}jóð-laðar, &
\alst{g}óðs of ǿðis, \hld\ —ef sér \alst{g}eta mę́tti— &
\ind \alst{o}rðs ok \edtrans{\alst{ę}ndr-þǫgu}{silence in return}{\Bfootnote{One may note that the verb \emph{þęgja} ‘shut up, be silent’—of which \emph{*þaga}, which only appears in the present cpd., is a derivative formed in the same way as \emph{saga} ‘saw, history’ to \emph{sęgja} ‘say, speak’—and the related noun \emph{þǫgn} ‘silence’ are frequently used at the beginning of Scaldic poems (e.g. Arn \emph{Magndr} 1\textsuperscript{II}: \emph{þegi sęim-brotar} ‘may gold-breakers \ken{generous men} be silent’, Egill \emph{Berdr} 1\textsuperscript{V}: \emph{hyggi \dots\ til þagnar þinn lýðr} ‘may thy retinue focus on silence’, Glúmr \emph{Gráf} 1\textsuperscript{I}: \emph{biðjum vér þagnar} ‘we ask for silence’).}}.}{\lemma{ALL}\Bfootnote{There is a good train of thought throughout the st.: the guest must first wash and dry himself, and then be welcomed to sit and eat at the table.  After the host has furnished him with these amenities the need for proper conduct now shifts onto the guest, who must speak and speak wisely.}}\eva

\bvb Of water there is need for him who comes for a meal; \\
\ind of a towel and a hearty welcome; \\
of a good reception—if he might get it— \\
\ind of a word, of and silence in return.\evb\evg


\bvg\bva\alst{V}its es þǫrf \hld\ þęim’s \alst{v}íða ratar; &
\ind dę́lt es \alst{h}ęima \alst{h}vat; &
\edtrans{at \alst{au}ga-bragði}{Into a laughing-stock}{\Bfootnote{Idomatic.  \emph{auga-bragð} literally means ‘twinkling of an eye, moment’; the sense here is thus something like ‘a quick glance of derision’.}} \hld\ verðr sá’s \alst{ę}kki kann &
\ind ok með \alst{s}notrum \alst{s}itr.\eva

\bvb Of wit there is need for him who widely roams; \\
\ind everything is easy at home. \\
Into a laughing-stock turns he who nothing knows, \\
\ind and among the clever sits.\evb\evg


\bvg\bva At \alst{h}yggjandi sinni \hld\ skyli-t maðr \alst{h}rǿsinn vesa, &
\ind hęldr \alst{g}ę́tinn at \alst{g}ęði, &
þȧ’s \alst{h}orskr ok þǫgull \hld\ kømr \alst{h}ęimis-garða til, &
\ind sjaldan verðr \alst{v}íti \alst{v}ǫrum. &
því-at \alst{ȯ}-brigðra vin \hld\ fę̇r \edtrans{maðr}{man}{\Bfootnote{In \Regius\ abbreviated with the rune ᛘ \textbf{m} “man”, the first of 45 such instances in the present poem.  Whereas Anglo-Saxon Latin-script mss. use several runes ideographically (e.g. ᛟ \textbf{o} for OE \emph{ǿðel} ‘homeland, patrimony’), there do not seem to be any Scandinavian examples with runes other than ᛘ.

The tradition of ideographic runes is in any case ancient, and goes back to the time before Latin writing as shown by the pre-Christian inscriptions from Stentoften (DR 357) and Ingelstad (Ög 43); DR 357 uses the rune ᛃ \textbf{j} for \emph{ár} ‘year, good harvest’ and Ög 43 uses ᛞ \textbf{d} for \emph{dagʀ} ‘day’.  For the names of the runes see the Three Rune Poems, edited below under Miscellaneous Runic Poetry.}} \alst{a}ldri-gi, &
\ind an \alst{m}an-vit \alst{m}ikit.\eva

\bvb Of his thinking should man not be boastful, \\
\ind but rather guarding of his senses \\
when sharp and silent he comes to a homestead; \\
\ind sudden harm seldom strikes the wary, \\
for an unfickler friend man never gets \\
\ind than great \inx[C]{manwit}.\evb\evg


\bvg\bva Hinn \alst{v}ari gęstr \hld\ es til \alst{v}erðar kømr, &
\ind \edtrans{\alst{þ}unnu hljóði}{with sharp hearing}{\Bfootnote{Lit. ‘with thin listening’.}} \alst{þ}ęgir; &
\alst{ęy}rum hlýðir, \hld\ en \alst{au}gum skoðar, &
\ind svá \edtext{nýsisk \alst{f}róðra hvęrr \alst{f}yrir}{\lemma{nýsisk fyrir ‘looks ahead’}\Bfootnote{This verb underlies the noun \emph{for-njósn} as found in \Sigrdrifumal\ 25.}}.\eva

\bvb The wary guest who comes for a meal \\
\ind with sharp hearing shuts up. \\
With ears he listens and with eyes he watches; \\
\ind so looks each learned man ahead.\evb\evg


\bvg\bva Hinn es \alst{s}ę́ll, \hld\ es \alst{s}ér of getr &
\ind \edtrans{\alst{l}of ok \alst{l}íkn-stafi}{praise and staves of liking}{\Bfootnote{\emph{líkn} ‘liking’ is a very interesting word.  It is defined by \ONP\ as: ‘mercy, compassion, relief, comfort, help’.  In the present poem its precise meaning seems to be something like ‘the state of being liked by your surroundings to the point where people are willing to help you out’.  Cf. its two other occurrences in the present poem: sts. 120 and especially 123 (where it is likewise paired with \emph{lof} ‘praise’).}}; &
\alst{ȯ}-dę́lla ’s við þat, \hld\ es \alst{ęi}ga skal &
\ind \alst{a}nnars brjóstum \alst{í}.\eva

\bvb This one is blessed, who for himself does get \\
\ind praise and staves of liking. \\
It is uneasy regarding that which one shall own \\
\ind in another man’s breast.\evb\evg


\bvg\bva%
\edtrans{\alst{S}á}{That one}{\Bfootnote{Contrasting with \emph{hinn} ‘this one’ in the previous stanza.}} es \alst{s}ę́ll, \hld\ es \alst{s}jalfr of á &
\ind \alst{l}of ok vit meðan \alst{l}ifir; &
því-at \alst{i}ll rǫ́ð \hld\ hęfr maðr \alst{o}pt þęgit &
\ind \alst{a}nnars brjóstum \alst{ó}r.\eva

\bvb That one is blessed, who himself does have \\
\ind praise and wits while he lives; \\
for ill counsels has man oft taken \\
\ind out of another man’s breast.\evb\evg


\bvg\bva\alst{B}yrði \alst{b}ętri \hld\ berr-at maðr \alst{b}rautu at, &
\ind an sé \alst{m}an-vit \alst{m}ikit; &
\alst{au}ði bętra \hld\ þykkir þat í \alst{ȯ}-kunnum stað; &
\ind slíkt es \alst{v}á-laðs \alst{v}era.\eva

\bvb A better burden bears man not on the road \\
\ind than be it great manwit. \\
In an unknown place it seems better than wealth; \\
\ind such is the destitute man’s shelter.\evb\evg


\bvg\bva\alst{B}yrði \alst{b}ętri \hld\ berr-at maðr \alst{b}rautu at, &
\ind an sé \alst{m}an-vit \alst{m}ikit; &
\alst{v}eg-nest \alst{v}erra \hld\ vegr-a \edtrans{\alst{v}ęlli at}{on the plain}{\Bfootnote{Formulaic, the word \emph{vǫllr} ‘plain, (uncultivated) field’ is also used in sts. 38 and 49. It is easily understood that the wild heaths and plains of Iron Age Norway were particularly unsafe places where a traveller needed to keep his wits about him, lest he fall victim to robbers or murderers (so st. 38).}}, &
\ind an sé \alst{o}f-drykkja \alst{ǫ}ls.\eva

\bvb A better burden bears man not on the road \\
\ind than be it great manwit. \\
Worse way-provision he drags not along on the plain \\
\ind than a too great drink of ale.\evb\evg


\bvg\bva Es-a svá \alst{g}ótt, \hld\ sęm \alst{g}ótt kveða, &
\ind \alst{ǫ}l \alst{a}lda sonum; &
því-at \alst{f}ę́ra vęit, \hld\ es \alst{f}lęira drekkr, &
\ind síns til \alst{g}ęðs \alst{g}umi.\eva

\bvb It is not so good, as good they say, \\
\ind ale for the sons of men; \\
for the less he knows, as the more he drinks, \\
\ind man of his own senses.\evb\evg


\bvg\bva \edtrans{\alst{Ó}-minnis-hegri}{Forgetfulness-heron}{\Bfootnote{Lit. “unmemory-heron”, the personification of drunkenness as a hovering bird.}} hęitir, \hld\ sá’s yfir \alst{ǫ}lðrum þrumir, &
\ind hann stelr \alst{g}ęði \alst{g}uma; &
\edtext{þess \alst{f}ogls \alst{f}jǫðrum \hld\ ek \alst{f}jǫtraðr vas’k &
\ind í \alst{g}arði \alst{G}unnlaðar.}{\lemma{þess \dots\ Gunnlaðar. ‘By that \dots\ Guthlathe.’}\Bfootnote{Weden stole the mead of poetry from Sutting’s daughter, Guthlathe, who was placed by her father to guard it.  For this myth see introduction to sts. 103–110 below.  In the version told in \Skaldskaparmal\ Weden does indeed drink all of the mead, but he soon spits it out again and shows no adverse effects.  This suggests that the present stanza may be referring to a variant conception, where the mead has drawbacks of normal alcohol.  If this is the case it would lend support to the theory that the Guest-strand and the later parts of \Havamal\ were originally separate compositions.  See note to 14/1–2.}}\eva

\bvb Forgetfulness-heron is he called who hovers over ale-feasts; \\
\ind he robs man of his senses. \\
By that bird’s feathers I was fettered \\
\ind in the yard of \inx[P]{Guthlathe}.\evb\evg


\bvg\bva\edtext{\alst{Ǫ}lr ek varð, \hld\ varð \alst{o}fr-ǫlvi, &
\ind at hins \alst{f}róða \alst{F}jalars}{\lemma{TODO}\Bfootnote{Another reference to the Mead of Poetry, for Fealer was one of the two dwarfs who slew Quasher and made the mead.  Again the sense seems to be that Weden got drunk on it, but in the attested versions of the myth Weden never even meets the two dwarfs.}}; &
því es \alst{ǫ}lðr batst, \hld\ at \alst{a}ptr of hęimtir &
\ind hvęrr sitt \alst{g}ęð \alst{g}umi.\eva

\bvb Drunk I became—became the greatest drunkard— \\
\ind at the learned Fealer’s. \\
That ale-feast is best, where every man \\
\ind gets back to his senses.\evb\evg


\bvg\bva\alst{Þ}agalt ok hugalt \hld\ skyli \alst{þ}jóðans barn &
\ind ok \alst{v}íg-djarft \alst{v}esa; &
\alst{g}laðr ok ręifr \hld\ skyli \alst{g}umna hvęrr, &
\ind unds sinn \alst{b}íðr \alst{b}ana.\eva

\bvb Silent and thoughtful should the king’s child \\
\ind —and battle-bold—be. \\
Glad and cheerful should every man [be], \\
\ind until he suffers his bane.\evb\evg


\bvg\bva\alst{Ȯ}-snjallr maðr \hld\ hyggsk munu \alst{ę}y lifa, &
\ind ef við \alst{v}íg \alst{v}arask; &
en \alst{ę}lli gefr hǫ́num \hld\ \alst{ę}ngi frið, &
\ind þótt hǫ́num \alst{g}ęirar \alst{g}efi.\eva

\bvb The unvalorous man thinks he will forever live \\
\ind if he of war be wary, \\
but old age gives him no peace, \\
\ind although it the spears might give him.\footnoteB{The coward may have been spared by spears, but he cannot avoid old age; since death is unavoidable it is better to live a glorious life in war than a wretched one in peace.  Such an attitude is common in the heroic literature, cf. for instance \Fafnismal\ 10.  Also related is the ancient view of the “straw-death” (TODO).}\evb\evg


\bvg\bva\alst{K}ópir af-glapi, \hld\ es til \alst{k}ynnis kømr, &
\ind \alst{þ}ylsk hann umb eða \alst{þ}rumir; &
allt es \alst{s}ęnn, \hld\ ef \alst{s}ylg of getr, &
\ind uppi ’s þȧ \alst{g}ęð \alst{g}uma.\eva

\bvb The oaf gapes when to visit he comes; \\
\ind he mumbles about or loiters. \\
All at once—if a sip he gets— \\
\ind exposed is then the mind of the man.\evb\evg


\bvg\bva Sá ęinn \alst{v}ęit, \hld\ es \alst{v}íða ratar &
\ind ok \edtrans{hęfr \alst{f}jǫlð of \alst{f}arit}{has journeyed much}{\Bfootnote{Cf. \Vafthrudnismal\ 3, 44, et.c., where Weden repeats: \emph{Fjǫlð ek fór, \hld\ fjǫlð fręistaða’k, // fjǫlð ek ręynda ręgin} ‘Much I journeyed, much I tried, much I tested the \inx[G]{Reins}.’}}, &
hvęrju \alst{g}ęði \hld\ stýrir \alst{g}umna hvęrr, &
\ind sá es \alst{v}itandi ’s \alst{v}its.\eva

\bvb He alone knows who widely roams \\
\ind and has journeyed much, \\
which sort of mind every man wields, \\
\ind who is knowing of his wits.\evb\evg


\bvg\bva\edtrans{\alst{H}aldi-t maðr ȧ kęri}{Man ought not to hold onto the cask}{\Bfootnote{Perhaps referring to a toast wherein the drinking vessel would be passed around in a circle and each recipient would drink in turn.  Such toasts were drunk for a long time in Northern Europe—indeed this is the origin of the Scandinavian toasting-word, \emph{skål} ‘prosit, cheers!’, lit. ‘bowl!’.  “Holding onto” the vessel (and not letting the next person drink) was surely seen as very rude; indeed, in 1519 a man in Jämtland was killed in an argument resulting from his refusal to pass on the bowl \parencite{Sjöberg1907}.  The sense is thus: “Do not refuse a toast when offered, but do not drink too much.”}}, \hld\ drekki þó at \alst{h}ófi mjǫð, &
\ind \edtrans{mę́li \alst{þ}arft eða \alst{þ}ęgi}{ought to speak the needful or shut up}{\Bfootnote{Formulaic, line occurs identically in \Vafthrudnismal\ 10/2.}}; &
\alst{ȯ}-kynnis þess \hld\ váar þik \alst{ę}ngi maðr, &
\ind at gangir \alst{s}nimma at \alst{s}ofa.\eva

\bvb Man ought not to hold onto the cask; ought yet to drink mead in moderation; \\
\ind ought to speak the needful or shut up. \\
For this uncouthness will no man blame thee: \\
\ind that thou go early to sleep.\evb\evg


\bvg\bva\alst{G}rǫ́ðugr halr, \hld\ nema \alst{g}ęðs viti, &
\ind \edtrans{\alst{e}tr sér \alst{a}ldr-trega}{eats himself a life-sorrow}{\Bfootnote{Or, ‘eats himself to death.’}}; &
opt fę̇r \alst{h}lǿgis, \hld\ es með \alst{h}orskum kømr, &
\ind \alst{m}anni hęimskum \alst{m}agi.\eva

\bvb The gluttonous man—unless he know his sense— \\
\ind eats himself a life-sorrow. \\
Oft the belly, when among the sharp he comes, \\
\ind brings the foolish man ridicule.\evb\evg


\bvg\bva\alst{H}jarðir þat vitu, \hld\ nę́r \alst{h}ęim skulu, &
\ind ok \alst{g}anga þȧ af \alst{g}rasi; &
en \alst{ȯ}-sviðr maðr \hld\ kann \alst{ę́}va-gi &
\ind síns of \alst{m}ál \alst{m}aga.\eva

\bvb Herds know when homewards they shall, \\
\ind and then part from the grass; \\
but the unwise man never knows \\
\ind his own belly’s measure.\evb\evg


\bvg\bva\alst{V}e-sall maðr \hld\ ok \alst{i}lla skapi &
\ind \alst{h}lę́r at \alst{h}ví-vetna; &
hitt-ki hann \alst{v}ęit, \hld\ es \alst{v}ita þyrpti, &
\ind at \edtrans{hann es-a \alst{v}amma \alst{v}anr}{he is not free of blemishes}{\Bfootnote{Formulaic, cf. \Lokasenna\ 30: \emph{es-a þér vamma vant} ‘thou art not free of blemishes’.}}.\eva

\bvb The wretched man and ill turned out \\
\ind laughs at anything. \\
This he knows not, which he might need to know: \\
\ind that he is not free of blemishes.\evb\evg


\bvg\bva\alst{Ó}-sviðr maðr \hld\ vakir umb \alst{a}llar nę́tr &
\ind ok \alst{h}yggr at \alst{h}ví-vetna; &
þȧ es \alst{m}óðr, \hld\ es at \alst{m}orni kømr; &
\ind alt es \alst{v}íl sęm \alst{v}as.\eva

\bvb The unwise man is awake for all nights \\
\ind and thinks of anything. \\
Then he is weary when the morning comes; \\
\ind all the trouble is as it was.\evb\evg


\bvg\bva\alst{Ȯ}-snotr maðr \hld\ hyggr sér \alst{a}lla vesa &
\ind \alst{v}ið-hlę́jęndr \alst{v}ini; &
hitt-ki hann \alst{f}iðr, \hld\ þótt of hann \alst{f}ár lesi, &
\ind ef með \alst{s}notrum \alst{s}itr.\eva

\bvb The unclever man thinks all those \\
\ind who laugh with him his friends. \\
This he finds not yet, that they make sport of him, \\
\ind if among the clever he sits.\evb\evg


\bvg\bva\alst{Ȯ}-snotr maðr \hld\ hyggr sér \alst{a}lla vesa &
\ind \alst{v}ið-hlę́jęndr \alst{v}ini; &
\alst{þ}ȧ þat fiðr \hld\ es at \alst{þ}ingi kømr, &
\ind at \edtrans{á \alst{f}or-mę́lęndr \alst{f}áa}{has spokesmen few}{\Bfootnote{Repeated in st. 62.  The Thing was the Germanic legal assembly, where small disputes could easily turn into deadly feuds, so the import is that the fool has few friends ready to take his side and speak up for him “when push comes to shove”.  True friends are proven in conflict, not in drunken revelry.}}.\eva

\bvb The unclever man thinks all those \\
\ind who laugh with him his friends. \\
Then he finds it, when to the \inx[C]{Thing} he comes, \\
\ind that he has spokesmen few.\evb\evg


\bvg\bva\alst{Ȯ}-snotr maðr \hld\ þykkisk \alst{a}llt vita, &
\ind ef á sér í \edtrans{\alst{v}ǫ̇}{nook}{\Bfootnote{From earlier \emph{*vrǫ̇}; cf. Swedish \emph{vrå} ‘corner, nook’, rare English \emph{wroo} ‘id.’  The present stanza is to my knowledge the only Norse attestation of the form \emph{vǫ̇}, which features a rare Western sound change from \emph{vr-} to \emph{v-}.  The more common change \emph{vr-} to \emph{r-} yields \emph{rǫ̇}, which is the normal Norse form. — Tangentially this word is brought up in \FGT\ as an example of a word with nasal \emph{ǫ̇}, and contrasted with oral \emph{ǫ́} in \emph{rǫ́} ‘sailyard’.}} \alst{v}eru; &
hitt-ki hann \alst{v}ęit, \hld\ hvat skal \alst{v}ið kveða, &
\ind ef hans \alst{f}ręista \alst{f}irar.\eva

\bvb The unclever man seems to know everything \\
\ind if he takes shelter in a nook. \\
This he knows not, what he shall answer \\
\ind if men test him.\evb\evg


\bvg\bva\alst{Ȯ}-snotr maðr \hld\ es með \alst{a}ldir kømr, &
\ind \alst{þ}at ’s batst at hann \alst{þ}ęgi; &
\alst{ę}ngi þat vęit, \hld\ at hann \alst{ę}kki kann, &
\ind nema hann \alst{m}ę́li til \alst{m}art. &
\alst{v}ęit-a maðr, \hld\ hinn’s \alst{v}ę́t-ki vęit, &
\ind þótt hann \alst{m}ę́li til \alst{m}art.\eva

\bvb The unclever man who among people comes— \\
\ind it is best that he shut up. \\
No one knows that he nothing knows, \\
\ind unless he speak too much. \\
The man knows not, who nothing knows, \\
\ind that he speak too much.\evb\evg


\bvg\bva\alst{F}róðr sá þykkisk, \hld\ es \edtext{\alst{f}regna kann, &
\ind ok \alst{s}ęgja}{\lemma{fregna \dots\ sęgja ‘ask \dots\ answer’}\Bfootnote{Perhaps specifically in the context of a riddling contest of wisdom.}} hit \alst{s}ama, &
\edtext{\alst{ęy}-vitu lęyna \hld\ męgu \alst{ý}ta synir &
\ind því es \alst{g}ęngr of \alst{g}uma.}{\lemma{ęy-vitu \dots\ guma. ‘In no way \dots\ earthlings.’}\Bfootnote{I.e., when asked a certain question to which one does not know the answer it is not possible to conceal one’s ignorance.}}\eva

\bvb Learned seems he who can ask \\
\ind and answer the same [way]. \\
In no way may the sons of men hide \\
\ind that which eludes earthlings.\evb\evg


\bvg\bva\alst{Ǿ}rna mę́lir, \hld\ sá’s \alst{ę́}va þęgir, &
\ind \alst{st}að-lausu \alst{st}afi; &
\edtext{\alst{h}rað-mę́lt tunga, \hld\ \edtrans{nema \alst{h}aldęndr ęigi}{unless it be held in place}{\Bfootnote{Lit. ‘unless holders own it’ or ‘unless it own holders’; the “holders” perhaps being the teeth which hold the tongue in place.}}, &
\ind opt sér ȯ-\alst{g}ótt of \alst{g}ęlr}{\lemma{hrað-mę́lt \dots\ of gęlr ‘A quick-spoken \dots\ for itself’}\Bfootnote{Formulaic. Cf. \Lokasenna\ 31.}}.\eva

\bvb He who never shuts up speaks plenty many \\
\ind utterings of absurdity. \\
A quick-spoken tongue—unless it be held in place— \\
\ind oft sings evil [into being] for itself.\evb\evg


\bvg\bva At \alst{au}ga-bragði \hld\ skal-a maðr \alst{a}nnan hafa, &
\ind þótt til \alst{k}ynnis \alst{k}omi; &
margr \alst{f}róðr þykkisk, \hld\ ef \alst{f}reginn es-at &
\ind ok nái \edtrans{\alst{þ}urr-fjallr}{dry-skinned}{\Bfootnote{i.e. ‘untested’, equivalent to the English idiom \emph{get one’s feet wet}.  The word \emph{fell} \char`~\ \emph{fjall} ‘skin, pelt’ is rare in Old Norse literature and only occurs in cpds, e.g. \Volundarkvida\ 11: \emph{ber-fjall} ‘bear-pelt’.  It survives in modern Swedish \emph{fjäll} ‘scale (on fish and reptiles)’}} \alst{þ}ruma.\eva

\bvb For a laughing-stock shall man not have another \\
\ind when he comes to visit. \\
Many a one seems learned if he is not asked, \\
\ind and gets to loiter about dry-skinned.\evb\evg


\bvg\bva\alst{F}róðr þykkisk \hld\ sá’s \alst{f}lótta tękr &
\ind \edtrans{\alst{g}ęstr}{guest}{\Bfootnote{The situation hinted at in this and the following stanza is that two guests—unknown to eachother—have come to the same homestead.  The sense is that when mocked by a stranger it is best not to engage, since the dealing may quickly turn violent.  Cf. sts. 122, 123, and 125.}} at \alst{g}ęst hę́ðinn; &
\alst{v}ęit-a gǫrla \hld\ sá’s of \alst{v}erði glissir, &
\ind þótt með \alst{g}rǫmum \alst{g}lami.\eva

\bvb Learned seems he who takes to flight, \\
\ind the guest, from a scoffing guest. \\
He knows not clearly, who grins over the food, \\
\ind that he be flirting with fiends.\evb\evg


\bvg\bva\alst{G}umnar margir \hld\ erusk \alst{g}agn-hollir, &
\ind en \edtrans{at \alst{v}irði \alst{\emph{v}}rekask}{over food drive each other away}{\Bfootnote{The archaic initial \emph{vr-} must be restored for metrical reasons.  Since \emph{vr-} > \emph{r-} is never found in Scaldic poetry younger than 1000 this provides a solid dating criteria.  For a summary of present research especially as it relates to the present poem see \textcite[87--92]{Males2024}.}}; &
\alst{a}ldar róg \hld\ þat mun \alst{ę́} vesa; &
\ind órir \alst{g}ęstr við \alst{g}ęst.\eva

\bvb Many men are well true to each other, \\
\ind but over food drive each other away. \\
The strife of mankind will that ever be; \\
\ind guest raves against guest.\evb\evg


\bvg\bva\alst{Á}r-liga verðar \hld\ skyli maðr \alst{o}pt fȧa, &
\ind nema til \alst{k}ynnis \alst{k}omi; &
\alst{s}itr ok \alst{s}nópir, \hld\ lę́tr sęm \alst{s}olginn sé, &
\ind ok kann \alst{f}regna at \alst{f}ǫ́u.\eva

\bvb An early meal should man oft get, \\
\ind unless he come to visit: \\
he sits and sulks, sounds as if starved, \\
\ind and can ask about little.\evb\evg


\bvg\bva\alst{A}f-hvarf mikit \hld\ es til \alst{i}lls vinar, &
\ind þótt ȧ \alst{b}rautu \alst{b}úi, &
en til \alst{g}óðs vinar \hld\ liggja \alst{g}agn-vegir, &
\ind þótt hann sé \alst{f}irr \alst{f}arinn.\eva

\bvb A great offroad is it to a bad friend, \\
\ind though on the road he live, \\
but to a good friend lie pleasant ways, \\
\ind though he be far gone.\evb\evg


\bvg\bva\alst{G}anga \edtext{skal}{\Afootnote{emend.; om. \Regius}}, \hld\ skal-a \alst{g}ęstr vesa &
\ind \alst{ęy} í \alst{ęi}num stað; &
\alst{l}júfr verðr \alst{l}ęiðr, \hld\ ef \alst{l}ęngi sitr &
\ind \alst{a}nnars flętjum \alst{ȧ}.\eva

\bvb One shall go; he shall not be a guest \\
\ind forever in one place. \\
The loved becomes loathed if for long he sits \\
\ind on another man’s benches.\footnoteB{The customary length of stay in old times was three nights, as noted in \EgilsSaga, ch. 78: \emph{þat var engi siðr, at sitja lengr en þrjár nę́tr at kynni} ‘it was not customary to stay longer than three nights when visiting.’  Compare a much more recent Jutish saying: \emph{en tredje dags gjæst stinker} ‘a third day’s guest stinks’, which closely resembles a maxim popularly attributed to Benjamin Franklin: “Guests, like fish, begin to smell after three days.”  It is probably inspired by such proverbs that Auden and Taylor translate the last two lines of this stanza as “He starts to stink who outstays his welcome, / in a hall that is not his own.”}\evb\evg


\bvg\bva\edtrans{\alst{B}ú es \alst{b}ętra, \hld\ þótt lítit sé}{A dwelling is better though small it be}{\Bfootnote{The b-line is missing the necessary alliteration, but no good emendation suggests itself.}}, &
\ind \alst{h}alr es \alst{h}ęima \alst{h}vęrr; &
þótt \alst{t}vę́r gęitr ęigi \hld\ ok \alst{t}aug-ręptan sal, &
\ind þat ’s þó \alst{b}ętra an \alst{b}ǿn.\eva

\bvb A dwelling is better though small it be; \\
\ind each is a hero at home. \\
Though two goats he own and a cord-roofed hall, \\
\ind it is yet better than begging.\evb\evg


\bvg\bva\alst{B}ú es \alst{b}ętra, \hld\ þótt lítit sé, &
\ind \alst{h}alr es \alst{h}ęima \alst{h}vęrr; &
\alst{b}lóðugt es hjarta \hld\ þęim’s \alst{b}iðja skal &
\ind sér í \alst{m}ál hvęrt \alst{m}atar.\eva

\bvb A dwelling is better though small it be; \\
\ind each is a hero at home. \\
Bloody is the heart in him who shall beg \\
\ind for his every meal of food.\evb\evg


\bvg\bva%
\alst{V}ǫ́pnum sínum \hld\ skal-a maðr \edtrans{\alst{v}ęlli ȧ}{on the plain}{\Bfootnote{Formulaic, see note to st. 11.}} &
\ind \edtrans{\alst{f}eti ganga \alst{f}ramarr}{take one step further}{\Bfootnote{Formulaic. Cf. \Lokasenna\ 1: \emph{svá’t ęinu-gi feti gangir framarr} ‘so that thou not take one step further’.}}, &
því-at ȯ-\alst{v}íst ’s at \alst{v}ita, \hld\ nę́r verðr ȧ \alst{v}egum úti &
\ind \alst{g}ęirs of þǫrf \alst{g}uma.\eva

\bvb From his weapons shall man on the plain \\
\ind not take one step further, \\
for it is unsure to know, when on the ways outside, \\
\ind man comes in need of a spear.\evb\evg


\bvg\bva%
Fann’k-a \alst{m}ildan \alst{m}ann \hld\ eða svá \edtrans{\alst{m}atar góðan}{good of meat}{\Bfootnote{A Viking Age expression; see Index.}}, &
\ind at vę́ri-t \alst{þ}iggja \alst{þ}egit; &
eða \alst{s}íns \edtrans{féar}{fee}{\Bfootnote{‘Wealth, movable property, cattle’; see Index.}} \hld\ \alst{s}vá-gi \edtext{[...]}{\Bfootnote{It is doubtless that a word has been lost here; the meter and sense require it. \textcite{FinnurEdda}\ suggests \emph{gløggvan} ‘miserly, stingy’, giving a litotes ‘so unstingy’, i.e., ‘so generous’.}}, &
\ind at \alst{l}ęið sé \alst{l}aun, ef þegi.\eva

\bvb I found not a generous man or one so \inx[C]{good of meat}, \\
\ind that a gift were not accepted; \\
or one with his \inx[C]{fee} so not [...], \\
\ind that the repayments were loathed, if he accepted [them].\footnoteB{No man is so generous that he would refuse a gift presented to him, nor loathe receiving a favour as thanks for his generosity.}\evb\evg


\bvg\bva%
\alst{F}éar síns, \hld\ es \alst{f}ęngit hęfr, &
\ind skyli-t maðr \alst{þ}ǫrf \alst{þ}ola; &
opt sparir \alst{l}ęiðum \hld\ þat’s hęfr \alst{l}júfum hugat; &
\ind mart gęngr \alst{v}err an \alst{v}arir.\eva

\bvb Of his \inx[C]{fee} which he has earned \\
\ind should man not suffer need. \\
Oft he saves for the loathed what he meant for the loved; \\
\ind much goes worse than he expects.\evb\evg


\bvg\bva%
\edtrans{\alst{V}ǫ́pnum ok \alst{v}ǫ́ðum}{With weapons and garments}{\Bfootnote{i.e. weapons and armour (the “garments” are probably no silks); friends are supposed to help each other and strengthen their “violence capital”.  This alliterative word-pair is formulaic and in other occurences exclusively refers to implements of war; cf. e.g. \Beowulf\ 39, where \inx[P]{Shield}’s pyre-ship is loaded with \emph{hilde-wǽpnum \alst\ ǫnd heaðo-wǽdum} ‘war-weapons and battle-garments’.}} \hld\ skulu \alst{v}inir glęðjask; &
\ind \edtrans{þat ’s ȧ \alst{s}jǫlfum \alst{s}ýnst}{that is best seen on oneself}{\Bfootnote{I.e. in your own lived experience.}}; &
\alst{v}iðr-gefęndr ok ęndr-gefęndr \hld\ erusk \alst{v}inir lęngst, &
\ind ef \edtrans{þat}{it}{\Bfootnote{The friendship.}} bíðr at \alst{v}erða \alst{v}ęl.\eva

\bvb With weapons and garments shall friends gladden each other; \\
\ind that is best seen on oneself. \\
Givers-back and givers-again are friends for the longest \\
\ind if it comes to last long.\evb\evg


\bvg\bva\alst{V}in sínum \hld\ skal maðr \alst{v}inr \alst{v}esa, &
\ind ok \alst{g}jalda \alst{g}jǫf við \alst{g}jǫf; &
\alst{h}látr við \alst{h}látri \hld\ skyli \alst{h}ǫlðar taka, &
\ind en \alst{l}ausung við \alst{l}ygi.\eva

\bvb With his friend shall man be a friend, \\
\ind and pay gift against gift; \\
laughter for laughter should men employ, \\
\ind but duplicity for lie.\evb\evg


\bvg\bva\alst{V}in sínum \hld\ skal maðr \alst{v}inr vesa, &
\ind \alst{þ}ęim ok \alst{þ}ess vin; &
en \alst{ȯ}-vinar síns \hld\ skyli \alst{ę}ngi maðr &
\ind \alst{v}inar \alst{v}inr \alst{v}esa.\eva

\bvb With his friend shall man be a friend, \\
\ind with him and with \emph{his} friend; \\
but his enemy’s, should no man, \\
\ind friend’s friend be.\evb\evg


\bvg\bva\alst{V}ęitst, ef \alst{v}in átt, \hld\ þann’s \alst{v}ęl trúir &
\ind ok vilt af hǫ́num \alst{g}ótt \alst{g}eta, &
\alst{g}ęði skalt við þann \hld\ ok \alst{g}jǫfum skipta, &
\ind \alst{f}ara at \alst{f}inna opt.\eva

\bvb Thou knowest, if thou have a friend whom thou well trust, \\
\ind and wilt get good from him: \\
thoughts and gifts shalt thou exchange with him; \\
\ind journey to find him oft.\footnoteB{Several lines of the present st. are shared with st. 119.}\evb\evg


\bvg\bva Ef þú \alst{á}tt \alst{a}nnan, \hld\ þann’s \alst{i}lla trúir, &
\ind vilt af hǫ́num þó \alst{g}ótt \alst{g}eta, &
\edtext{\alst{f}agrt skalt mę́la við þann, \hld\ en \alst{f}látt hyggja}{\lemma{fagrt \dots\ mę́la, flátt hyggja ‘fairly \dots\ speak, falsely think’}\Bfootnote{Formulaic, cf. sts. 90, 91.}} &
\ind ok gjalda \alst{l}ausung við \alst{l}ygi.\eva

\bvb If thou have another whom thou badly trust, \\
\ind and wilt yet receive good from him: \\
fairly shalt thou speak with him, but falsely think, \\
\ind and pay duplicity for lie.\evb\evg


\bvg\bva Þat ’s \alst{ę}nn umb þann, \hld\ es þú \alst{i}lla trúir &
\ind ok þér es \alst{g}runr at \alst{g}ęði, &
\alst{h}lę́ja skalt við þęim \hld\ ok of \alst{h}ug mę́la; &
\ind \alst{g}lík skulu \alst{g}jǫld \alst{g}jǫfum.\eva

\bvb This is yet about him whom thou trust badly, \\
\ind and about whom thou have doubt: \\
laugh shalt thou with him, and speak with care; \\
\ind repayments shall be equal to gifts.\footnoteB{Equivalent to the last line of the previous st. (“pay duplicity for lie”).}\evb\evg


\bvg\bva Ungr vas’k \alst{f}orðum, \hld\ \alst{f}ór’k ęinn saman, &
\ind þȧ varð’k \alst{v}illr \alst{v}ega; &
\alst{au}ðigr þȯttumk, \hld\ es \alst{a}nnan fann’k, &
\ind \alst{m}aðr es \alst{m}anns gaman.\eva

\bvb Young was I once, I travelled alone; \\
\ind then I became lost of ways. \\
Wealthy I thought me when another I found; \\
\ind man is man’s pleasure.\evb\evg


\bvg\bva\alst{M}ildir frǿknir \hld\ \alst{m}ęnn batst lifa, &
\ind \alst{s}jaldan \alst{s}út ala; &
en \edtext{\alst{ȯ}-snjallr}{\linenum{|3||4}\lemma{ȯ-snjallr, gløggr ‘unvalorous, stingy’}\Bfootnote{Contrasting respectively with \emph{frǿkn, mildr} ‘brave, generous’ in the first half of the stanza; very fine parallelism.}} maðr \hld\ \alst{u}ggir hvat-vetna, &
\ind \edtrans{sýtir ę́ \alst{g}løggr við \alst{g}jǫfum}{the stingy always grieves over gifts}{\Bfootnote{After receiving a gift, one was culturally obliged to give something back.  Cf. sts. 39, 145.}}.\eva

\bvb Generous, brave men live best; \\
\ind seldom they nourish sorrow— \\
but the unvalorous man is frightened by anything, \\
\ind the stingy always grieves over gifts.\evb\evg


\bvg\bva\alst{V}áðir mínar \hld\ gaf’k \alst{v}ęlli at &
\ind \alst{t}vęim \alst{t}ré-mǫnnum; &
\alst{r}ekkar þat þȯttusk, \hld\ es \alst{r}ipt hǫfðu; &
\ind \alst{n}ęiss es \alst{n}ǫkkviðr halr.\eva

\bvb My garments I gave, on the plain, \\
\ind to two tree-men. \\
Champions they seemed when cloaks they had; \\
\ind shameful is the naked hero.\footnoteB{One of the harder sts. in the poem.  The probable sense is that “the clothes make the man” (or warrior): under expensive gear a thin tree-man might be lurking, and likewise even a mighty man (the choice of the word \emph{halr} ‘hero, warrior’ (cf. sts. 36, 37) rather than the more neutral \emph{maðr} ‘man, person’ is surely intentional) can never defend himself against a heavily armoured opponent.  Without his arms, he becomes as vulnerable as the “tree-man” on the plain.}\evb\evg


\bvg\bva Hrørnar \alst{þ}ǫll, \hld\ sú’s stęndr \alst{þ}orpi ȧ, &
\ind \edtext{hlýr-a\emph{t}}{\Afootnote{\emph{hlyrar} \Regius}} hęnni \alst{b}ǫrkr né \alst{b}arr; &
svá es \alst{m}aðr, \hld\ sá’s \alst{m}ann-gi ann; &
\ind hvat skal hann \alst{l}ęngi \alst{l}ifa?\eva

\bvb Wilters the pine that stands on the yard; \\
\ind shields her not bark nor leaf. \\
So is the man who loves no man; \\
\ind why shall he live for long?\evb\evg


\bvg\bva\alst{Ę}ldi hęitari \hld\ brinnr með \alst{i}llum vinum &
\ind \alst{f}riðr \edtrans{\alst{f}imm daga}{for five days}{\Bfootnote{i.e. “for a week”, which was originally five days long.  See also st. 74 and the Index: \inx[C]{five days}.}}, &
en þȧ \alst{sl}oknar, \hld\ es hinn \alst{s}étti kømr, &
\ind ok \alst{v}ersnar allr \alst{v}in-skapr.\eva

\bvb Hotter than fire burns love among bad friends, \\
\ind for \inx[C]{five days}; \\
but then goes out when the sixth one comes, \\
\ind and all the friendship worsens.\evb\evg


\bvg\bva\alst{M}ikit ęitt \hld\ skal-a \alst{m}anni gefa; &
\ind opt kaupir sér í \alst{l}ítlu \alst{l}of, &
með \alst{h}ǫlfum \alst{h}lęif \hld\ ok með \alst{h}ǫllu kęri &
\ind \alst{f}ekk ek mér \edtrans{\alst{f}é-laga}{fellow}{\Bfootnote{A business partner or companion.}}.\eva

\bvb Much at once shall one not give a man; \\
\ind oft one buys him praise for little. \\
With half a loaf and an awry cask \\
\ind I got myself a fellow.\evb\evg


\bvg\bva\edtrans{\alst{L}ítilla sanda, \hld\ \alst{l}ítilla sę́va}{Of small sands, of small seas}{\Bfootnote{Probably a partitive genitive, the sense being that man’s “horizons” are small; the universe will always be far greater than him.}}, &
\ind lítil eru \alst{g}ęð \alst{g}uma; &
\edtext{því-at \alst{a}llir męnn \hld\ \alst{u}rðu-t jafn-spakir; &
\ind \alst{h}ǫlf es ǫld \alst{h}var.}{\lemma{því-at \dots\ ǫld hvar. ‘For \dots\ every man.’}\Bfootnote{On the meaning of the second half of this stanza I find the view of \textcite{Athugasemdir1929} most convincing; namely that every man has both strengths and weaknesses in terms of wisdom.  As nobody can excel at everything, nobody is complete; every person is “half” (and it should be added that ON \emph{halfr} has a more general sense of incompleteness than its English cognate).  This interpretation fits particularly closely with sts. 71 and 132. — This stanza introduces several stanzas dealing with wisdom and foolishness.}}\eva

\bvb Of small sands, of small seas: \\
\ind small are the senses of man. \\
For all have not become evenly knowing; \\
\ind half is every man.\evb\evg


\bvg\bva\alst{M}eðal-snotr \hld\ skyli \alst{m}anna hvęrr, &
\ind ę́va til \alst{s}notr \alst{s}éi; &
þęim es \alst{f}yrða \hld\ \alst{f}ęgrst at lifa, &
\ind es \alst{v}ęl mart \alst{v}itu.\eva

\bvb Middle-clever should each man be; \\
\ind never too clever. \\
For those men it is fairest to live, \\
\ind who know well enough.\evb\evg


\bvg\bva\alst{M}eðal-snotr \hld\ skyli \alst{m}anna hvęrr, &
\ind ę́va til \alst{s}notr \alst{s}éi; &
\alst{s}notrs manns hjarta \hld\ verðr \alst{s}jaldan glatt, &
\ind ef sá ’s \alst{a}l-snotr es \alst{á}.\eva

\bvb Middle-clever should each man be; \\
\ind never too clever. \\
The clever man’s heart is seldom glad, \\
\ind if its owner is all-clever.\evb\evg


\bvg\bva\alst{M}eðal-snotr \hld\ skyli \alst{m}anna hvęrr, &
\ind ę́va til \alst{s}notr \alst{s}éi; &
\alst{ø}r·lǫg sín \hld\ viti \alst{ę}ngi maðr fyrir; &
\ind \edtrans{þęim es \alst{s}orga-lausastr \alst{s}efi.}{his is the most sorrowless mind.}{\Bfootnote{i.e. he who is ignorant of his fate.  It is surely fitting that Weden should say this, having knowledge of the inevitable destruction of the world and himself (see \inx[L]{Rakes of the Reins}).}}\eva

\bvb Middle-clever should each man be; \\
\ind never too clever. \\
His own \inx[C]{orlay} ought no man to know ahead; \\
\ind his is the most sorrowless mind.\evb\evg


\bvg\bva\alst{B}randr af \alst{b}randi \hld\ \alst{b}rinnr unds \alst{b}runninn es, &
\ind \alst{f}uni kvęykisk af \alst{f}una; &
\alst{m}aðr af \alst{m}anni \hld\ verðr at \alst{m}áli kuðr; &
\ind en til \edtrans{\alst{d}ǿlskr}{hickish}{\Bfootnote{Derived from an ablaut variant of \emph{dalr} ‘valley, dale’ + \emph{-iskr} ‘-ish’, the sense being ‘provincial, not having left his (home) valley’.  Cf. the Icelandic tribal names like \emph{vatns-dǿlir} and \emph{lang-dǿlir} ‘inhabitants of \emph{Vatns-dalr} (Waterdale), \emph{Lang-dalr} (Longdale)’.}} af \alst{d}ul.\eva

\bvb Fire by fire burns until it is burned [out]; \\
\ind flame is quickened by flame. \\
Man by man becomes known through speech, \\
\ind but the too hickish from his folly.\evb\evg


\bvg\bva\alst{Á}r skal rísa, \hld\ sá’s \alst{a}nnars vill &
\ind \alst{f}é eða \alst{f}jǫr hafa; &
sjaldan \alst{l}iggjandi ulfr \hld\ \alst{l}ę́r of getr, &
\ind né \alst{s}ofandi maðr \alst{s}igr.\eva

\bvb Early shall he rise who another man’s \\
\ind \inx[C]{fee} or life will have. \\
Seldom gets the lying wolf the thigh, \\
\ind nor the sleeping man victory.\evb\evg


\bvg\bva\alst{Á}r skal rísa, \hld\ sá’s á \alst{y}rkjęndr fáa, &
\ind ok ganga síns \alst{v}erka ȧ \alst{v}it; &
\alst{m}art of dvęlr \hld\ þann’s umb \alst{m}orgin sefr, &
\ind \edtrans{\alst{h}alfr es auðr und \alst{h}vǫtum}{the brisk has half the wealth}{\Bfootnote{i.e. the brisk man has already claimed half of a fortune by simply choosing to wake up early.}}.\eva

\bvb Early shall he rise who has workmen few, \\
\ind and go his work to meet. \\
Much is kept back from him who in the morning sleeps; \\
\ind the brisk has half the wealth.\evb\evg


\bvg\bva\alst{Þ}urra skíða \hld\ ok \alst{þ}akinna nę́fra, &
\ind þess kann \alst{m}aðr \alst{m}jǫt, &
ok þess \alst{v}iðar, \hld\ es \alst{v}innask męgi &
\ind \edtrans{\alst{m}ál ok \alst{m}issęri}{for a season and half-year}{\Bfootnote{Over nine months.}}.\eva

\bvb Of dry billets and thatching birch bark— \\
\ind of \emph{this} man knows the measure— \\
and of that firewood which he may use \\
\ind for a season and half-year.\evb\evg


\bvg\bva\edtrans{\alst{Þ}vęginn ok męttr}{Washed and full}{\Bfootnote{A formulaic collocation.  Cf. \Reginsmal\ 25 (\emph{kęmbðr} ‘combed’ — \emph{þvęginn} ‘washed’ — \emph{męttr} ‘full’) and \Voluspa\ 33: (\emph{þó} ‘washed’ — \emph{kęmbði} ‘combed’).  These examples attest to the importance of personal hygiene in the culture, something further seen by the ubiquity of combs in pre-Christian graves (TODO: archeological reference).
The whole thing reminds of the passage from \emph{Germania} ch. 22: \emph{Statim ē somnō, quem plērumque in diem extrahunt, lavantur, saepius calidā, ut apud quōs plūrimum hiems occupat.  Lautī cibum capiunt: sēparātae singulīs sēdēs et sua cuique mēnsa.  Tum ad negōtia nec minus saepe ad convīvia prōcēdunt armātī.} ‘On waking from sleep, which they generally prolong to a late hour of the day, they take a bath, oftenest of warm water, which suits a country where winter is the longest of the seasons.  After their bath they take their meal, each having a separate seat and table of his own.  Then they go armed to business, or no less often to their festal meetings (\emph{convivia}, i.e., their Things).’}} \hld\ ríði maðr \alst{þ}ingi at, &
\ind þótt sé-t \alst{v}ę́ddr til \alst{v}ęl; &
\alst{sk}úa ok bróka \hld\ \alst{sk}ammisk ęngi maðr &
\ind né \alst{h}ęsts in \alst{h}ęldr, &
\ind \edtrans{þótt hann \alst{h}afi-t góðan}{although he has not a good one}{\Bfootnote{\textcite{FinnurEdda} considers this a late insert, and I agree.  It seems that the inserter was not aware of the rules of the \Ljodahattr\ meter and interpreted the preceding c-verse (\emph{né hęsts in hęldr}) as an a-verse of \Fornyrdislag.}}.\eva

\bvb Washed and full ought a man to ride to the \inx[C]{Thing}, \\
\ind although he be not clothed too well; \\
of his shoes and breeches ought no man to be ashamed, \\
\ind nor the more of his horse, \\
\ind even though he haven’t a good one.\evb\evg

\sectionline

{\small The two following sts. are written in opposite order in \Regius, but a symbol at the start of each indicates that they should switch places.}

\sectionline

\bvg\bva\alst{S}napir ok gnapir, \hld\ es til \alst{s}ę́var kømr, &
\ind \alst{ǫ}rn ȧ \alst{a}ldinn mar; &
svá es \alst{m}aðr, \hld\ es með \alst{m}ǫrgum kømr &
\ind ok \edtrans{á \alst{f}or-mę́lęndr \alst{f}áa}{has spokesmen few}{\Bfootnote{Shared with st. 25.}}.\eva

\bvb He snaps and stoops when to the sea he comes, \\
\ind the eagle on the ancient sea. \\
So is the man who among the many comes, \\
\ind and has spokesmen few.\evb\evg


\bvg\bva\alst{F}regna ok sęgja \hld\ skal \alst{f}róðra hvęrr, &
\ind sá’s vill \alst{h}ęitinn \alst{h}orskr; &
\alst{ęi}nn vita \hld\ né \alst{a}nnarr skal, &
\ind \edtrans{\alst{þ}jóð}{thirty}{\Bfootnote{Or “people, nation”; the sense is in any case “many, everybody”.  For the translation “thirty” cf. \Skaldskaparmal\ 82, a list of poetic expressions for various numerals: \emph{\emph{þjóð} eru þrír tigir} ‘a \emph{nation} is thirty’ etc.}} vęit ef \alst{þ}rír ’ru.\eva

\bvb Ask and answer shall each learned man \\
\ind who wishes to be called sharp. \\
\emph{One} shall know, another shall not; \\
\ind thirty know if there are three.\evb\evg


\bvg\bva\alst{R}íki sitt \hld\ skyli \alst{r}áð-snotra &
\ind hvęrr í \alst{h}ófi \alst{h}afa; &
\edtext{þȧ þat \alst{f}innr, \hld\ es með \alst{f}rǿknum kømr, &
\ind at \alst{ę}ngi es \alst{ęi}nna hvatastr.}{\lemma{þȧ \dots\ ęinna hvatastr ‘then \dots fiercest of all’}\Bfootnote{Almost identical to \Reginsmal\ TODO/3–4, which however has \emph{flęirum} ‘more men’ instead of \emph{frǿknum} ‘the bold’.}}\eva

\bvb His own power should each counsel-clever \\
\ind man use in moderation. \\
This he then finds when among the bold he comes— \\
\ind that noone is fiercest of all.\footnoteB{i.e., every man has his match.}\evb\evg


\bvg\bva\alst{O}rða þęira, \hld\ es maðr \alst{ǫ}ðrum sęgir, &
\ind opt hann \alst{g}jǫld of \alst{g}etr.\eva

\bvb For those words which man says to another \\
\ind he oft gets recompense.\evb\evg


\bvg\bva \edtrans{\alst{M}ikils til}{Much too}{\Afootnote{written as one word \emph{mikilsti} \Regius}} snimma \hld\ kom’k í \alst{m}arga staði, &
\ind en til \alst{s}íð í \alst{s}uma; &
\alst{ǫ}l vas drukkit, \hld\ sumt vas \alst{ȯ}-lagat; &
\ind sjaldan hittir \alst{l}ęiðr í \alst{l}ið.\eva

\bvb Much too early I came to many places, \\
\ind and too late to some: \\
The ale was drunk up, some was unbrewed— \\
\ind seldom finds the loathed his place.\footnoteB{i.e., “there are no wrong times, only wrong people”.}\evb\evg


\bvg\bva\alst{H}ér ok \alst{h}var \hld\ myndi mér \alst{h}ęim of boðit, &
\ind ef þyrpta’k at \alst{m}ǫ́lun-gi \alst{m}at, &
eða \alst{t}vau lę́r hęngi \hld\ at hins \alst{t}ryggva vinar, &
\ind þar’s ek hafða \alst{ęi}tt \alst{e}tit.\eva

\bvb Here and there would I to a home be invited, \\
\ind if at meal-time I needed no food; \\
or if two hams should hang at the trusty friend’s [home], \\
\ind where I had eaten one.\footnoteB{Not everyone is hospitable, especially with regards to food, which was scarce and closely watched among the Norse subsistence farmers.  The poet notes that even a “trusty friend” (possibly sarcastic) would invite him over more often if he brought more food than he ate.}\evb\evg


\bvg\bva\alst{Ę}ldr es batstr \hld\ með \alst{ý}ta sonum &
\ind ok \alst{s}ólar \alst{s}ýn, &
\alst{h}ęilyndi sitt, \hld\ ef maðr \alst{h}afa náir, &
\ind án við \alst{l}ǫst at \alst{l}ifa.\eva

\bvb Fire is best among the sons of men, \\
\ind and the sight of the sun; \\
one’s good health, if he manage to keep it— \\
\ind {[and]} living free from vice.\evb\evg


\bvg\bva\alst{E}s-at maðr \alst{a}lls \edtrans{ve-sall}{unblessed}{\Bfootnote{Or ‘woe-blessed’.  I have elsewhere translated this word as ‘wretched’, but have presently rendered it this way to show the etymological relationship.  The second element in this compound is \emph{sę́ll}, which lacks i-umlaut due to a shortening of the vowel before the umlaut became phonemic.  The ancestral Proto-Norse forms would be \emph{*sāliʀ} and \emph{*wajē-sāliʀ}.  Cf. ᚹᚨᛃᛖ-ᛗᚨᚱᛁᛉ \emph{wajē-mariʀ} ‘infamous’ on the Tjurkö bracteate, where the second element is the ancestor of ON \emph{mę́rr} ‘renowned, famous’; the expected descendant \emph{*ve-marr} is not attested.
I have chosen to translate \emph{sę́ll} as ‘blessed’, but it is not a past participle and could also be rendered as ‘lucky’ or ‘blissful’.  It carries a certain sense of innateness that is foreign to modern Western culture.  Thus a king whose land experiences bountiful harvests (\emph{ár}) is said to be \emph{ár-sę́ll} ‘blessed with harvests’, while one whose kingdom is at peace (\emph{friðr}) is said to be \emph{frið-sę́ll} ‘blessed with peace’.  In this worldview the state of the realm is not due to uncontrollable environmental or political factors, but rather arises from the very person of the king (TODO: Reference PCRN chapter).}}, \hld\ þótt sé \alst{i}lla hęill, &
\ind \alst{s}umr es af \edtext{\alst{s}onum}{\lemma{sonum \dots\ frę́ndum ‘sons \dots\ kinsmen’}\Bfootnote{Cf. st. 72 below, which stresses the importance of sons and kinsmen.}} \alst{s}ę́ll, &
sumr af \alst{f}rę́ndum, \hld\ sumr af \alst{f}é ǿrnu, &
\ind sumr af \alst{v}erkum \alst{v}ęl.\eva

\bvb Man is not all unblessed, though he of poor health be: \\
\ind someone is blessed with sons; \\
someone with kinsmen, someone with ample \inx[C]{fee}, \\
\ind someone with works done well.\evb\evg


\bvg\bva Bętra ’s \alst{l}ifðum, \hld\ \edtrans{an séi ȯ-\alst{l}ifðum}{than with the unliving}{\Afootnote{emend.; \emph{⁊ ſęl lıfðo}m \Regius.}\Bfootnote{The reading of \Regius, which would be normalized as \emph{ok sę́l-lifðum} ‘and for the blessed living’, is metrically defect since \emph{sę́l-} is strongly stressed and should carry alliteration.  For the original form of the line we may instead cf. \Fafnismal\ 30: \emph{Hvǫtum ’s bętra \hld\ an sé ȯ-hvǫtum} ‘It is better for the brisk than it may be for the unbrisk’.  The corruption has probably happened in the following way: \emph{*en} (younger form of \emph{an} ‘than’) in the prototype was misinterpreted as \emph{en} ‘and, but’ and copied as \emph{⁊} (the tironian \emph{et}), while \emph{*séı ólıfðo}m (probably with the words cramped together) became \emph{sęl lıfðo}m.}}, &
\ind \edtrans{ęy getr \alst{k}vikr \alst{k}ú}{always gets the quick a cow}{\Bfootnote{i.e., “new opportunities always present themselves for the living”.  A reference to the cattle-based economy (see also st. 76), the cow being used as a metonym:  (cf. churchly English ‘the \emph{quick} and the dead’, i.e. ‘the \emph{living} and the dead’).}}; &
\alst{ę}ld sá’k \alst{u}pp brinna \hld\ \alst{au}ðgum manni fyr, &
\ind en úti vas \alst{d}auðr fyr \alst{d}urum.\eva

\bvb It is better for the living than it may be for the unliving: \\
\ind ever the quick gets the cow. \\
A fire I saw burning high for a wealthy man, \\
\ind but outside he was dead before the doors.\footnoteB{The fire is presumably the man’s funeral pyre, on which a considerable amount of his wealth has been spent; according to ibn Fadlan (TODO) two thirds of a dead chieftain’s estate was spent on his funeral.  One notes the contrastive \emph{en} ‘but’ and may understand it as follows: “I saw a lavish funeral held for a man, but he was still dead.”  This interpretation is supported by the \Havamal\ 71 below, which expresses the same sentiment.}\evb\evg


\bvg\bva\alst{H}altr ríðr \alst{h}rossi, \hld\ \alst{h}jǫrð rekr \alst{h}andar vanr, &
\ind \alst{d}aufr vegr ok \alst{d}ugir; &
\alst{b}lindr es \alst{b}ętri, \hld\ an \alst{b}ręnndr séi; &
\ind \alst{n}ýtr mann-gi \alst{n}ás.\eva

\bvb A halt man rides a horse; a handless drives a herd; \\
\ind a deaf fights and avails. \\
Blind is better than be burned; \\
\ind no man has use for a corpse.\evb\evg


\bvg\bva \edtrans{\alst{S}onr es bętri}{A son is better}{\Bfootnote{i.e. it is better for a man to have a son and heir than not, even if the father should die some time before he is born. The son can further his father’s lineage and memory (as exemplified by the raising of a “beat-stone”), and as the poet says, it is rare for a non-relative to do so.}}, \hld\ þótt sé \alst{s}íð of alinn &
\ind ęptir \alst{g}inginn \alst{g}uma; &
sjaldan \edtrans{\alst{b}autar-stęinar}{beat-stones}{\Bfootnote{Large standing stones raised in memory of someone.  Numerous such stones with runic inscriptions are known from migration period Norway, often near grave fields.  Some hold only single personal names or short phrases, like the stone from Sunde in Sunnfjord, western Norway (signum \emph{KJ 90}): ᚹᛁᛞᚢᚷᚨᛊᛏᛁᛉ \textbf{widugastiʀ} ‘Woodguest’, or the one from Bø in Rogaland, southwestern Norway (signum \emph{KJ 78}): ᚺᚾᚨᛒᛞᚨᛊ ᚺᛚᚨᛁᚹᚨ \textbf{hnabdas hlaiwa} ‘Naved’s grave’.  Others hold longer inscriptions, like the one from Kjølevik in Rogaland (signum \emph{KJ 75}): ᚺᚨᛞᚢᛚᚨᛁᚲᚨᛉ ᛖᚲᚺᚨᚷᚢᛊᛏᚨᛞᚨᛉ ᚺᛚᚨᚨᛁᚹᛁᛞᛟᛗᚨᚷᚢᛗᛁᚾᛁᚾᛟ \textbf{hadulaikaz ekhagustadaz hlaaiwidomaguminino} ‘Hathlac [lies here].  I, Haystald, buried my lad.’}} \hld\ standa \alst{b}rautu nę́r, &
\ind nema ręisi \alst{n}iðr at \alst{n}ið.\eva

\bvb A son is better, though he late be born \\
\ind after a passed-on man. \\
Seldom beat-stones stand near the road, \\
\ind save by kinsman for kinsman raised.\evb\evg


\bvg\bva \edtext{\edtrans{\alst{T}vęir ’ru ęins hęrjar}{Two are of one host}{\Bfootnote{i.e. “the tongue and head belong to the same body (but the former often leads to the latter’s demise).” — \emph{hęrjar} is an inflected form of \emph{hęrr} ‘host, army’, but its function is ambiguous; it can either be (1) the gen. sg., as adopted here, or (2) the nom. pl. ‘harriers, raiders’ (cf. \emph{ęin-hęrjar} ‘\inx[G]{Oneharriers}’) which would translate as “two are the destroyers of one”, i.e. “the tongue and head often lead to the demise of the body”.}}, \hld\ \edtrans{\alst{t}unga es hǫfuðs bani}{the tongue is the head’s bane}{\Bfootnote{Formulaic or proverbial.  Cf. the Old Swedish “Heathen Law”, which describes how a duel should be conducted following an insult to a man’s honour (my norm. and trans. following \textcite{Läffler1879}): \emph{Fallr þann orð havr givit—glǿpr orða vęrstr,} tunga hovuð-bani—\emph{liggi i ú·gildum akri} ‘If he falls who has given the [insulting] word—an insult is the worst of words, \emph{the tongue the head-bane}—may he lie in an unhallowed field.’}}; &
mér ’s í \alst{h}eðin \alst{h}vęrn \hld\ \edtrans{\alst{h}andar}{a hand}{\Bfootnote{i.e. a hand holding a dagger.}} vę́ni.}{\lemma{ALL}\Bfootnote{The whole st. fits poorly in context, and the metre and style are very out of place; it is probably a later insert.}}\eva

\bvb Two are of one host: the tongue is the head’s bane; \\
in every cloak I expect a hand.\evb\evg


\bvg\bva\alst{N}ǫ́tt verðr fęginn, \hld\ sá’s \alst{n}esti trúir, &
\ind \edtrans{\alst{sk}ammar ’ru \alst{sk}ips ráar}{short are a ship’s sailyards}{\Bfootnote{TODO: Write about the varying interpretations (Finnur, Cleasby, Skp) of this line.}}, &
\ind \alst{h}verf es \alst{h}aust-gríma; &
\alst{f}jǫlð \edtrans{of viðrir}{The winds blow far}{\Bfootnote{I.e., the weather changes; a verb derived from \emph{veðr} ‘weather, wind’.  Consider Weden’s name \emph{Viðrir} ‘Withrer; Stormer, One of the Storm’, which may be analyzed as an agent noun formed to this verb.}} \hld\ ȧ \edtrans{\alst{f}imm dǫgum}{five days}{\Bfootnote{i.e. “in a week” (which was originally five days long), paralleling “month” in the next line.  See note to st. 51 and Index.}}, &
\ind en \alst{m}ęir ȧ \alst{m}ánaði.\eva

\bvb At night he rejoices, who trusts in his provisions; \\
\ind short are a ship’s sailyards; \\
\ind shifty is a stormy fall night. \\
The winds blow far in \inx[C]{five days}; \\
\ind even more in a month.\evb\evg


\bvg\bva\alst{V}ęit-a hinn, \hld\ es \alst{v}ę́tki \alst{v}ęit, &
\ind \edtrans{margr verðr \edtrans{af \alst{au}rum}{from wealth}{\Afootnote{emend. from meaningless \emph{†aflꜹðrom†} \Regius}} \alst{a}pi}{many a man turns an ape from wealth}{\Bfootnote{Cf. \Solarljod\ 34/4: \emph{margan hefr auðr apat} ‘wealth has aped many a man’, which also lends support to the emendation.}}; &
maðr es \alst{au}ðigr, \hld\ annarr \alst{ȯ}-auðigr, &
\ind skyli-t þann \alst{v}ítka \alst{v}áar.\eva

\bvb The one knows not who nothing knows: \\
\ind many a man turns an \inx[C]{ape} from wealth. \\
A man is wealthy, another not wealthy; \\
\ind one oughtn’t to curse him for his woe.\evb\evg


\bvg\bva\edtrans{\alst{D}ęyr \edtext{fé}{\lemma{fé, frę́ndr ‘Fee, kinsmen’}\Bfootnote{In the Germanic Iron Age farming society a man’s wealth was reckoned by how many heads of cattle (and the Norman loan-word \emph{cattle} is itself the same word as \emph{capital}) he owned, and his social power by the number of able male relatives ready to side with him in conflict (cf. st. 72 above and TODO: reference?).  All one’s earthly power will pass away, and so too oneself, but a good reputation, fame and glory can linger on.  For Indo-European poetic analogues, see \textcite[99\psqq]{West2007}.}}, \hld\ \alst{d}ęyja frę́ndr}{Fee dies, kinsmen die}{\Bfootnote{This line is also found in \Hakonarmal\ 21, the final st. of that poem composed ca. 961.}}, &
\ind dęyr \alst{s}jalfr hit \alst{s}ama; &
en \alst{o}rðs-tírr \hld\ dęyr \alst{a}ldri-gi &
\ind hvęim’s sér \alst{g}óðan \alst{g}etr.\eva

\bvb \inx[C]{fee}[Fee] dies, kinsmen die, \\
\ind oneself dies the same [way]; \\
but a word-glory never dies, \\
\ind for whomever gets himself a good one.\evb\evg


\bvg\bva\alst{D}ęyr fé, \hld\ \alst{d}ęyja frę́ndr, &
\ind dęyr \alst{s}jalfr hit \alst{s}ama; &
\alst{e}k vęit \alst{ęi}nn \hld\ at \alst{a}ldri-gi dęyr: &
\ind \edtrans{\alst{d}ómr}{Doom}{\Bfootnote{Here meaning ‘judgment, glory’.  See Index.}} of \alst{d}auðan hvęrn.\eva

\bvb Fee dies, kinsmen die, \\
\ind oneself dies the same [way]. \\
I know one that never dies: \\
\ind the \inx[C]{Doom} o’er each man dead.\evb\evg

\sectionline

{\small It is likely that the original Guest-Strand ended here.  The three following stanzas, especially the third, are poorly placed and seem like later inserts.}

\sectionline

\bvg\bva\alst{F}ullar grindr \hld\ sá’k fyr \alst{F}itjungs sonum, &
\ind nú bera þęir \edtrans{\alst{v}ánar \alst{v}ǫl}{the staff of hope}{\Bfootnote{A beggar’s staff.}}; &
svá es \alst{au}ðr \hld\ sęm \alst{au}ga-bragð, &
\ind hann es \alst{v}altastr \alst{v}ina.\eva

\bvb Full pens I saw for the sons of Fitting; \\
\ind now they carry the staff of hope. \\
So is wealth like the twinkling of an eye: \\
\ind it is the ficklest of friends.\evb\evg


\bvg\bva\alst{Ȯ}-snotr maðr \hld\ es \alst{ęi}gnask getr &
\ind \alst{f}é eða \alst{f}ljóðs mun-úð; &
\alst{m}etnaðr hǫ́num þróask, \hld\ en \alst{m}an-vit aldri-gi; &
\ind framm gęngr hann \alst{d}rjúgt í \alst{d}ul.\eva

\bvb The unclever man who comes to own \\
\ind fee or a girl’s loving grace: \\
his pride flourishes, but never his manwit; \\
\ind he goes forth far in folly.\evb\evg


\bvg\bva Þat ’s þȧ \alst{r}ęynt, es þú at \edtext{\alst{r}únum spyrr, \hld\ hinum \alst{r}ęgin-kunnum}{\lemma{rúnum \dots\ ręgin-kunnum ‘runes \dots\ born of the Reins’}\Bfootnote{This expression also appears on the C4th–6th Noleby stone (in the acc. sg. \emph{rúnó ragina-kundó} ‘a rune born of the Reins’), which proves that the Eddic rune-magic is (at least in part) founded in oral tradition going back to the Heathen age. See also Index \inx[C]{rune}.}}, &
\ind \edtext{þęim’s \alst{g}ørðu \alst{g}inn-ręgin &
\ind ok \alst{f}áði \alst{F}imbul-þulr;}{\lemma{þęim’s \dots\ Fimbul-þulr ‘those which \dots\ Fimble-Thyle’}\Bfootnote{Formulaic. Cf. st. 142 where these two lines occur almost identically, but in reverse order.}} &
\ind \alst{þ}ȧ hęfr hann batst, ef hann \alst{þ}ęgir.\eva

\bvb That is then proven, which thou learnest from the runes, those born of the Reins, \\
\ind those which the \inx[G]{yin-Reins} made, \\
\ind and the Fimble-Thyle \name{= Weden} painted.— \\
\ind Then he has it best, if he shuts up.\footnoteB{This stanza, which deals with runic magic and shares expressions with sts. in the Rune-Tally section (beginning with st. 138 below), hardly fits in its current place.  The last line with its shift in person is likely to be a later insert.}\evb\evg

\sectionline

\section{Scattered stanzas of practical advice (81–90)}

{\small The following stanzas are distinguished by a prevalence of \Malahattr\ and the common subject matter.}

\sectionline

\bvg\bva%
At \alst{k}veldi skal dag lęyfa, \hld\ \alst{k}onu es bręnnd es, &
\alst{m}ę́ki es ręyndr es, \hld\ \alst{m}ęy es \edtrans{gefin}{given}{\Bfootnote{In marriage.}} es, &
\alst{í}s es \alst{y}fir kømr, \hld\ \alst{ǫ}l es drukkit es.\eva

\bvb At evening shall one praise day, a woman when she is burned, \\
a sword when it is tried, a maiden when she is given, \\
ice when one comes over it, ale when it is drunk.\evb\evg


\bvg\bva%
Í \alst{v}indi skal \alst{v}ið hǫggva, \hld\ \edtrans{\alst{v}eðri}{weather}{\Bfootnote{i.e. ‘in good weather’; elsewhere the word \emph{veðr} typically means ‘storm’, but that can hardly be the sense here.}} ȧ sę́ róa, &
\alst{m}yrkri við \alst{m}an spjalla— \hld\ \alst{m}ǫrg eru dags augu— &
ȧ \alst{sk}ip skal \alst{sk}riðar orka, \hld\ en ȧ \alst{sk}jǫld til hlífar, &
\alst{m}ę́ki til hǫggs, \hld\ en \alst{m}ęy til kossa.\eva

\bvb In wind shall one cut wood, in weather row at sea, \\
in darkness speak with a maiden—many are the eyes of day. \\
A ship shall one have for speed, and a shield for protection; \\
a sword for striking, and a maiden for kisses.\evb\evg


\bvg\bva%
Við \alst{ę}ld skal \alst{ǫ}l drekka, \hld\ en ȧ \alst{í}si skríða, &
\alst{m}agran \edtext{\alst{m}ar kaupa, \hld\ en \alst{m}ę́ki}{\lemma{mar \dots\ mę́ki ‘steed \dots\ sword’}\Bfootnote{Formulaic pair, also occurring in \Lokasenna\ 12/1, \Volundarkvida\ 33/3, \Atlakvida\ 7/3.}} saurgan, &
\alst{h}ęima \alst{h}ęst fęita, \hld\ en \alst{h}und ȧ búi.\eva

\bvb One shall drink ale by fire and skate on ice; \\
buy a starved steed and a rusty sword; \\
fatten the horse at home and the hound in its dwelling.\evb\evg


\bvg\bva%
\alst{M}ęyjar orðum \hld\ skyli \alst{m}ann-gi trúa, &
\ind né því’s \alst{k}veðr \alst{k}ona; &
\edtext{\edtext{því-at}{\Afootnote{om. \FostrbroedhraSaga}} ȧ \alst{h}verfanda \alst{h}véli \hld\ \edtext{vǫ́ru}{\Afootnote{\emph{er} \FostrbroedhraSaga}} þęim \edtrans{\alst{h}jǫrtu skǫpuð}{hearts shaped}{\Afootnote{\emph{hjarta skapat} ‘heart shaped’ \FostrbroedhraSaga}}, &
\ind \edtext{\alst{b}rigð}{\Afootnote{ok brigð \FostrbroedhraSaga}} í \alst{b}rjóst of \edtext{lagit}{\Afootnote{\emph{laginn} \FostrbroedhraSaga}}.}{\lemma{þvít \dots\ lagið}\Bfootnote{Quoted in slightly divergent form in \FostrbroedhraSaga\ (Thott 1768 4°\textsuperscript{x}, fol. 210r) introduced with the words: \emph{Kom honum þá í hug kviðlingr sá, er kveðinn hafði verit um lausungar-konur:} ‘And then he remembered the ditty which had been composed about loose women:’}}\eva

\bvb A maiden’s words should no man trust, \\
\ind nor that which a woman speaks. \\
For on a whirling wheel their hearts were shaped; \\
\ind fickleness laid in their breasts.\evb\evg


\bvg\bva%
\alst{B}restanda \alst{b}oga, \hld\ \alst{b}rinnanda loga, &
\alst{g}ínanda ulfi, \hld\ \alst{g}alandi krǫ́ku, &
\alst{r}ýtanda svíni, \hld\ \alst{r}ót-lausum viði, &
\alst{v}axanda \alst{v}ági, \hld\ \alst{v}ellanda katli,\eva

\bvb In bursting bow, in burning flame, \\
in yawning wolf, in crowing crow, \\
in roaring swine, in rootless tree, \\
in waxing wave, in boiling kettle,\evb\evg


\bvg\bva%
\alst{f}ljúganda \alst{f}lęini, \hld\ \alst{f}allandi bǫ́ru, &
\alst{í}si \alst{ęi}n-nę́ttum, \hld\ \alst{o}rmi hring-lęgnum, &
\alst{b}rúðar \alst{b}ęð-mǫ́lum \hld\ eða \alst{b}rotnu sverði, &
\alst{b}jarnar lęiki \hld\ eða \alst{b}arni konungs,\eva

\bvb in flying spear, in falling billow, \\
in one-night old ice, in coiled-up serpent, \\
in bride’s bed-speech, or in broken sword, \\
in bear’s play, or in king’s child,\evb\evg


\bvg\bva%
\alst{s}júkum kalfi, \hld\ \alst{s}jalf-ráða þrę́li, &
\edtrans{\alst{v}ǫlu \alst{v}il-mę́li}{in wallow’s pleasing speech}{\Bfootnote{i.e. in a favourable prophecy (\inx[C]{spae}).}}, \hld\ \alst{v}al ný-fęldum.\eva

\bvb in sick calf, in self-willing thrall, \\
in wallow’s pleasing speech, in newly felled corpses,\evb\evg

\sectionline

{\small In \Regius\ the following two sts. come in the opposite order, but it seems probable from its \Malahattr\ meter and the dative case of the words that 89 should follow 87.  On the other hand st. 88, with its \Ljodahattr\ meter and self-enclosed form seems a separate composition, and was probably inserted after 87 due to its first line (\emph{akri ár-sǫ́num}), which is also in the dative.}

\sectionline

\bvg\bva[89]%
\alst{b}róður-\alst{b}ana sínum \hld\ þótt ȧ \alst{b}rautu mǿti, &
\alst{h}úsi \alst{h}alf-brunnu, \hld\ \alst{h}ęsti al-skjótum, &
þȧ ’s \alst{jó}r \alst{ȯ}-nýtr, \hld\ ef \alst{ęi}nn fótr brotnar; &
verðr-it maðr svá \alst{t}ryggr \hld\ at þessu \alst{t}rúi ǫllu!\eva

\bvb in one’s brother’s bane—though on the road ye meet— \\
in half-burned house, in all-fleet horse— \\
the steed is useless if one foot breaks. \\
No man be so trusting that he trust in all this!\evb\evg


\bvg\bva[88]%
\alst{A}kri \alst{á}r-sǫ́num \hld\ trúi \alst{ę}ngi maðr, &
\ind né til \alst{s}nimma \alst{s}yni; &
\alst{v}eðr rę́ðr akri, \hld\ en \alst{v}it syni; &
\ind \alst{h}ę́tt es þęira \alst{h}várt.\eva

\bvb In an early sown field ought no man to trust, \\
\ind nor too soon in a son. \\
The weather rules the field and the wits the son: \\
\ind there is risk to them both.\evb\evg


\bvg\bva Svá ’s \alst{f}riðr kvinna \hld\ þęira’s \alst{f}látt hyggja, &
sęm \alst{a}ki \alst{jó} ȯ-bryddum \hld\ ȧ \alst{í}si hǫ́lum &
\alst{t}ęitum, \alst{t}vé-vetrum \hld\ ok sé \alst{t}amr illa, &
eða í \alst{b}yr óðum \hld\ \alst{b}ęiti stjórn-lausu, &
eða skyli \alst{h}altr \alst{h}ęnda \hld\ \alst{h}ręin \edtrans{í þá-fjalli}{on a thawing fell}{\Bfootnote{i.e. in springtime, when the melting ice on the ground is most slippery.}}.\eva

\bvb So is the love of those women who falsely think \\
like one rode an unshod horse on slippery ice— \\
a merry one, two winters old, and ill-tamed— \\
or in mad wind tacked a rudderless [ship], \\
or a halt man should catch a reindeer on a thawing fell.\evb\evg

\sectionline

\section{Weden’s failed seduction of Billing’s daughter (91–102)}

The following sts. are united by their meter, \Ljodahattr\ (unlike most of the preceding sts., see introduction to them above), style and content.  The strand begins with general maxims about love and relations between the sexes, before moving on to the narrative about Billing’s daughter.

\sectionline

\bvg\bva\alst{B}ęrt nú mę́li’k, \hld\ því-at \edtrans{\alst{b}ę́ði}{both}{\Bfootnote{i.e. both sides, both sexes.  The (male) poet declares that he will not attack the fair sex unfairly; he is also aware of men’s faults.}} vęit’k, &
\ind brigðr es \alst{k}arla hugr \alst{k}onum, &
\edtext{þȧ \alst{f}ęgrst mę́lum, \hld\ es \alst{f}lást hyggjum}{\lemma{fęgrst mę́lum \dots\ flást hyggjum ‘speak fairest \dots\ think falsest’}\Bfootnote{Formulaic.  Cf. st. 45.}}; &
\ind \edtrans{þat tę́lir \alst{h}orska \alst{h}ugi}{that entraps sharp minds}{\Bfootnote{Love (or sexual infatuation—the poet does not distinguish between them) turns even wise men into liars or otherwise dishonest persons.  Cf. \Malshattakvadi\ 20/1–2, which is probably partly based on this stanza:
\emph{Ást-blindir ’ru seggir svá \hld\ sumir, at þykkja mjǫk fás gá;} \\
\emph{þannig verðr um man-sǫng mę́lt: \hld\ marga hefr þat hyggna tę́lt.}
‘Some men are so love-blind, that they seem to heed very little; // for that sake it is said about love-song: many thinking men has it entrapped.’}}.\eva

\bvb Plainly I now speak, for I know them both: \\
\ind fickle is men’s thought towards women. \\
Fairest we speak when falsest we think; \\
\ind that entraps sharp minds.\evb\evg


\bvg\bva \edtrans{\alst{F}agrt skal mę́la}{Fairly shall speak}{\Bfootnote{Formulaic. Cf. st. 45.}} \hld\ ok \alst{f}é bjóða, &
\ind sá’s vill \alst{f}ljóðs ǫ́st \alst{f}ȧa, &
\alst{l}íki \alst{l}ęyfa \hld\ hins \alst{l}jósa mans, &
\ind \edtrans{sá \alst{f}ę̇r, es \alst{f}ríar}{he wins, who woos}{\Bfootnote{I.e., “he who courts her will win her hand.”}}.\eva

\bvb Fairly shall speak, and offer \inx[C]{fee}, \\
\ind he who will win a woman’s love; \\
praise the body of the bright girl— \\
\ind he wins, who woos.\evb\evg


\bvg\bva\alst{Á}star firna \hld\ skyli \alst{ę}ngi maðr &
\ind \alst{a}nnan \alst{a}ldri-gi; &
opt fȧa ȧ \alst{h}orskan, \hld\ es ȧ \alst{h}ęimskan né fȧa, &
\ind \edtrans{\alst{l}ost-fagrir \alst{l}itir}{lust-fair hues}{\Bfootnote{i.e. a (woman with a) countenance so beautiful that men cannot help but lust after her.}}.\eva

\bvb For [matters of] love should no man \\
\ind ever blame another; \\
oft they seize the sharp when they seize not the foolish, \\
\ind the lust-fair hues.\evb\evg


\bvg\bva\alst{Ęy}-vitar firna, \hld\ es maðr \alst{a}nnan skal, &
\ind þess es of margan \alst{g}ęngr \alst{g}uma; &
\alst{h}ęimska ór \alst{h}orskum \hld\ gęrir \alst{h}ǫlða sonu &
\ind sá hinn \alst{m}átki \alst{m}unr.\eva

\bvb In no way shall man blame another \\
\ind for that which happens to many a man; \\
from sharp to fools are the sons of men made \\
\ind by this mighty thing, love.\evb\evg


\bvg\bva\alst{H}ugr ęinn þat vęit, \hld\ es býr \alst{h}jarta nę́r, &
\ind ęinn es hann \alst{s}ér of \alst{s}efa; &
øng es \alst{s}ótt verri \hld\ hvęim \alst{s}notrum manni &
\ind an sér \alst{ø}ngu at \alst{u}na.\eva

\bvb The mind alone knows what dwells close to the heart; \\
\ind it is alone with its thoughts. \\
No sickness is worse for each clever man \\
\ind than with nothing to be content.\evb\evg


\bvg\bva Þat þȧ \alst{r}ęynda’k, \hld\ es í \alst{r}ęyri sat’k, &
\ind ok vę̇tta’k \alst{m}íns \alst{m}unar, &
\alst{h}old ok \alst{h}jarta \hld\ vas mér hin \alst{h}orska mę́r, &
\ind þęygi hana at \alst{h}ęldr \alst{h}ęf’k.\eva

\bvb It I found out when I sat in the reed \\
\ind and awaited my love. \\
My flesh and heart was that sharp maiden— \\
\ind I have her none the more.\evb\evg


\bvg\bva\alst{B}illings \edtrans{męy}{maiden}{\Bfootnote{i.e. unmarried (virgin) daughter.}} \hld\ ek fann \alst{b}ęðjum ȧ &
\ind \alst{s}ól-hvíta \alst{s}ofa; &
\alst{ja}rls \alst{y}nði \hld\ þȯtti mér \alst{ę}kki vesa &
\ind nema við þat \alst{l}ík at \alst{l}ifa.\eva

\bvb Billing’s maiden I found on the beds, \\
\ind sun-white, asleep. \\
An earl’s pleasure seemed me naught to be, \\
\ind save living alongside that body.\evb\evg


\bvg\bva\speakernote{[Billings mę́r:]}%
„\alst{Au}k nę́r \alst{a}ptni \hld\ skalt \alst{Ó}ðinn koma, &
\ind ef vilt þér \alst{m}ę́la \alst{m}an, &
\alst{a}llt eru \alst{ȯ}-skǫp, \hld\ nema \alst{ęi}n vitim &
\ind \alst{s}likan lǫst \alst{s}aman.“\eva

\bvb\speakernoteb{[Billing’s maiden:]}
“And by evening shalt thou, Weden, come, \\
\ind if thou wilt get for thee the girl [me]; \\
everything is misshapen unless we alone should know \\
\ind such a vice together.”\evb\evg


\bvg\bva\alst{A}ptr ek hvarf \hld\ ok \alst{u}nna þȯttumk &
\ind \edtrans{\alst{v}ísum \alst{v}ilja frȧ}{away from my wise will}{\Bfootnote{i.e., “against my better judgment”; the wise choice would have been to walk away.}}; &
\alst{h}itt ek \alst{h}ugða, \hld\ at \alst{h}afa mynda’k &
\ind \alst{g}ęð hęnnar allt ok \alst{g}aman.\eva

\bvb Back I turned—and thought myself in love— \\
\ind away from my wise will; \\
\emph{this} I thought, that I would have \\
\ind her senses all, and pleasure.\evb\evg


\bvg\bva Svá kom’k \alst{n}ę́st, \hld\ at hin \edtrans{\alst{n}ýta}{useful}{\Bfootnote{Sarcastic. Billing’s daughter had apparently summoned a lynch mob.}} vas &
\ind \alst{v}íg-drótt ǫll of \alst{v}akin, &
með \alst{b}rinnǫndum ljósum \hld\ ok \edtrans{\alst{b}ornum viði}{carried sticks}{\Bfootnote{The mob was armed with clubs.}}, &
\ind svá vas mér \edtrans{\alst{v}íl-stígr}{sad path}{\Bfootnote{Ambiguous, referring either to the beating he would have received at the hands of the mob, or to his walk of shame away from the hall.  The latter is perhaps more likely.}} of \alst{v}itaðr.\eva

\bvb So I came next as the useful \\
\ind war-troop was all awake; \\
with burning lights and with carried sticks; \\
\ind so a sad path was marked out for me.\evb\evg


\bvg\bva \edtrans{\alst{Au}k nę́r morni}{And by morning}{\Bfootnote{Mirroring the beginning of st. 97 above.}}, \hld\ es vas’k \alst{ę}nn of kominn, &
\ind þȧ vas \alst{s}al-drótt of \alst{s}ofin; &
\edtrans{\alst{g}ręy ęitt}{A lone bitch}{\Bfootnote{The insult is clearly understood; Weden is compared to a horny dog, and mockingly asked to make love to one—“this is all you get, you dog!”}} þȧ fann’k \hld\ hinnar \edtrans{\alst{g}óðu}{good}{\Bfootnote{Possibly not sarcastic, but rather referring to her chastity.}} konu &
\ind \alst{b}undit \alst{b}ęðjum ȧ.\eva

\bvb And by morning when I had come again, \\
\ind then was the hall-troop asleep. \\
A lone bitch I then found, by the good woman \\
\ind bound upon the beds.\evb\evg


\bvg\bva Mǫrg es \edtrans{\alst{g}óð mę́r}{good maiden}{\Bfootnote{A formulaic expression; the “goodness” here refers to faithfulness and chastity.  Cf. \Skirnismal\ 12, TODO.}}, \hld\ ef \alst{g}ǫrva kannar, &
\ind \alst{h}ug-brigð við \alst{h}ali; &
þȧ þat \alst{r}ęynda’k, \hld\ es hit \alst{r}áð-spaka &
\ind tęygða’k ȧ \alst{f}lę́rðir \alst{f}ljóð; &
\alst{h}ǫ́ðungar \alst{h}vęrrar \hld\ lęitaði mér hit \alst{h}orska man &
\ind ok hafða’k þess \alst{v}ę́t-ki \alst{v}ífs.\eva

\bvb Many a good maiden—if one comes to know her well— \\
\ind is heart-fickle towards men. \\
I found that out when the counsel-clever \\
\ind lady into sins I lured; \\
every disgrace that sharp girl sought out for me, \\
\ind and I had naught of the woman.\evb\evg

\sectionline

\section{Weden’s theft of the Mead of Poetry (103–110)}

The intricate myth of how Weden came to own the Mead of Poetry is told more fully in \Skaldskaparmal\ 5–6. That narrative goes as follows, with minor details left out:

\begin{quote}
	{\small After the war between the Eese and Wanes, the two tribes of gods reconcile through spitting into a vat. Not wanting to discard this token of their truce, they instead create a man out of the spit, calling him \inx[P]{Quasher}; he is so wise that he can answer any question posed to him, and so travels around the world in order to share his wisdom with humans.

	Quasher eventually comes to the dwelling of two dwarfs, Fealer and Galer. They kill him and drain his blood into three vessels: two vats named Soon and Bothem, and a kettle named \inx[P]{Woderearer}. Through mixing the blood with honey they make a mead, with the power to turn anyone who drinks from it “a scold or man of learning (\emph{skald eða frǿða-maðr})”. The dwarfs then lie to the Eese about the murder, telling them that Quasher drowned in his own wisdom.

	Some time later, the dwarfs murder an ettin named \inx[P]{Gilling} and his wife. Gilling’s son, \inx[P]{Sutting}, learns of this and prepares to drown the dwarfs. In exchange for their lives and as recompense for his father’s slaying, the dwarfs offer Sutting the “dear mead” (\emph{mjǫðinn dýra}; cf. here sts. 105 and 140). Sutting accepts the ransom and takes the mead home with him. He makes his daughter \inx[P]{Guthlathe} guard it.

	Some time later, Weden is out journeying, and finds nine thralls mowing hay. He sharpens their scythes with a special whetstone, and the mowing improves greatly. He then throws it in the air and the thralls shortly kill each other over it. By evening Weden comes to the owner of the thralls, Bigh, Sutting’s brother. Bigh laments the death of his workmen, and so Weden, who calls himself \inx[P]{Baleworker}, offers to do the work of the thralls over the summer, in exchange for one drink of Sutting’s mead. Bigh tells him that Sutting alone owns the mead, but that he will accompany Baleworker to Sutting to ask for the drink.

	The two arrive at Sutting, who as expected refuses to give any part of the mead away. Baleworker then tells Bigh that he will get to it anyway; he takes out the drill \inx[P]{Rate}, and tells Bigh to drill through the mountain, into the room where the mead is stored. Bigh first attempts to trick him by only drilling halfway, but eventually creates a narrow passage. Baleworker turns himself into a snake and crawls through it; as he does, Bigh tries to strike him the drill, but misses.

	After passing through, Baleworker sees Guthlathe watching over the mead. He goes on to sleep with her for three nights, after which she promises him three sips of the mead. With each sip he swallows the contents of one of the three vessels, so that all of the mead ends up in his belly.

	Having taken the mead, he dons his eagle-hame and flies away from the mountain. Sutting sees him, takes his own eagle-hame, and gives chase. The Eese see Weden in flight, and set out several large vat on the ground, into which Weden, still flying, spits out the mead. At this point Sutting has almost caught up with him, and so Weden “sends back” (\emph{sęnda aptr}, usually interpreted as being “sent out” from behind) some of the mead, presumably into his face.  This portion becomes the lot of foolish poets (\emph{skald-fífla hlutr}), while the rest of the mead is given to the Eese and to skilled poets (\emph{þęim mǫnnum, er yrkja kunnu} ‘those men who can compose [poetry]’).}
\end{quote}

The core of this many-twisted myth is old. A close parallel is found in \Rigveda\ hymns 4.26–27. In these two hymns the \emph{soma} plant (who in the Vedic mythology is not just the plant and its resulting drink, but also a god, perhaps somewhat like Quasher) is first held within “a hundred iron forts” (4.27.1c: \emph{śatám púraḥ ā́yasīḥ}) by the archer \emph{Kr̥şānu}, before being stolen by a sweeping falcon. The falcon brings \emph{Soma} to \emph{Manu}, the ancestor of the Aryans and first sacrificer.

The resemblance to the last part of the \Skaldskaparmal\ account should be obvious, but, notably, the detail of the falcon is not found in any of the sts. below. This shows that the narrative of \Skaldskaparmal\ cannot be exclusively based on the sts. here below, but instead also relies on other, now-lost sources. This is also supported by the present sts. leaving out the narratives about Quasher, the two dwarfs, and Baye, along with some subtler narrative differences.

The order of the present sts. follows that of \Regius, their main witness manuscript. The strand begins with some social advice (103), after which the narrative follows (104–110). It is narrated in the first person by Weden himself. The sts. do not tell the myth in chronological order and leave much up to the listener; they are surely composed for an audience that already knows the story. The following narrative details are given:

\begin{enumerate}
	\setcounter{enumi}{103}
	\item Weden visits Sutting’s home, but does not receive a good reception.
	\item Guthlate falls in love with Weden, and gives him a drink of the Mead.
	\item Weden has to bore through the mountains with the drill Rate.
	\item Weden has “bought [the Mead] well”; possibly a euphemistic reference to sleeping with Guthlathe for it.
	\item Guthlathe indeed does sleep with Weden, though not expressely in exchange for the Mead.
	\item The following day (\emph{hins hindra dags}, see note to this word in the edited text below), a group of Rime-Thurses come to Weden’s hall, to ask him whether a Baleworker is among the Gods, or if he has been slain by Sutting.
	\item Switching to the third person (which may indicate that this is his answer to the Rime-Thurses), Weden says that he “thinks” that Weden has sworn an oath, but that his words cannot be trusted. After the “simble” (i.e. drinking feast, banquet; probably referring to the drink of the Mead), Weden betrayed Sutting and made Guthlathe weep.
\end{enumerate}

The underlying narrative seems to generally agree with that of \Skaldskaparmal, but unlike its more transactional affair, we here find a stronger emphasis on Weden’s cruel betrayal of Guthlathe. A notable detail not found in \Skaldskaparmal\ is Weden’s oath in st. 109. The content of the oath was most likely that Weden would marry Guthlathe, something supported by the language used (see note to st. 108: \emph{hins hindra dags}). The recipient of the oath, which Weden clearly broke, was either Sutting or Guthlathe. That Weden swore it to Sutting, and thus asked him for Guthlathe’s hand in marriage, may be suggested by the description of Sutting as \emph{svikvinn} ‘betrayed’ in st. 109. This view, however, has an internal narrative problem: in st. 103 Weden describes his interaction with Sutting as poor, and in st. 105 Weden is said to have had to bore through the mountains, but this may just have been to reach Sutting, rather than Guthlathe as in \Skaldskaparmal.
The recipient of the oath being Guthlathe would agree better with the \Skaldskaparmal\ narrative, and Sutting’s betrayer would instead be her.

\sectionline

\bvg\bva Hęima \alst{g}laðr \alst{g}umi \hld\ ok við \alst{g}ęsti ręifr, &
\ind \alst{s}viðr skal of \alst{s}ik vesa; &
\alst{m}innigr ok \alst{m}ǫ́lugr, \hld\ ef vill \alst{m}arg-fróðr vesa; &
\ind opt skal \alst{g}óðs \alst{g}eta; &
\alst{f}imbul-\alst{f}ambi hęitir, \hld\ sá’s \alst{f}átt kann sęgja; &
\ind þat es \alst{ȯ}-snotrs \alst{a}ðal.\eva

\bvb At home shall man be glad and giving with the guest, \\
\ind wise about himself. \\
Of good memory and speech, if he wishes to be many-learned; \\
\ind oft shall he speak of good. \\
A fimble-fool is he called who little can say; \\
\ind that is the unclever man’s nature.\evb\evg


\bvg\bva Hinn \alst{a}ldna \alst{jǫ}tun sótta’k, \hld\ nú em’k \alst{a}ptr of kominn; &
\ind fátt gat’k \alst{þ}ęgjandi \alst{þ}ar; &
\alst{m}ǫrgum orðum \hld\ \alst{m}ę́lta’k í minn frama &
\ind í \alst{S}uttungs \alst{s}ǫlum.\eva

\bvb The old ettin \ken*{= Sutting} I sought, now am I come back; \\
\ind I got little hearing there. \\
Many words I spoke to my furtherance, \\
\ind in the halls of Sutting.\evb\evg


\bvg\bva\alst{G}unn-lǫð mér of \alst{g}af \hld\ \alst{g}ullnum stóli ȧ &
\ind \alst{d}rykk hins \alst{d}ýra mjaðar; &
\alst{i}ll \alst{i}ð-gjǫld \hld\ lét’k hana \alst{ę}ptir hafa &
\ind síns hins \alst{h}ęila \alst{h}ugar, &
\ind síns hins \alst{s}vára \alst{s}efa.\eva

\bvb \inx[P]{Guthlathe} gave me on the golden throne \\
\ind a drink of the dear mead; \\
evil recompense I let her have afterwards, \\
\ind for her whole heart, \\
\ind for her severe affection.\evb\evg


\bvg\bva\alst{R}ata munn \hld\ létumk \alst{r}úms of fȧa &
\ind ok of \alst{g}rjót \alst{g}naga; &
\alst{y}fir ok \alst{u}ndir \hld\ stóðumk \alst{jǫ}tna vegir, &
\ind svá \alst{h}ę́tta’k \alst{h}ǫfði til.\eva

\bvb Rate’s mouth I made to bring me room, \\
\ind and gnaw away at the rocks. \\
Over and under me stood the roads of the ettins \ken{mountains}; \\
\ind so I risked my head.\evb\evg


\bvg\bva \edtext{\alst{V}ęl kęypts hlutar \hld\ hęf’k \alst{v}ęl notit; &
\ind \alst{f}ás es \alst{f}róðum vant; &
því-at \edtrans{\alst{Ó}ð-rǿrir}{Woderearer}{\Bfootnote{One of the vessels in with the Mead of Poetry was held (see introduction to the present section above), here standing in for all the Mead.}} \hld\ es nú \alst{u}pp kominn &
\ind ȧ \alst{a}lda vés \edtrans{\alst{ja}ðar}{rim}{\Bfootnote{metr. emend.; \emph{jarðar} \Regius\ has a long root-syllable, and does not fit grammatically.}}.}{\lemma{Vęl \dots\ jaðar}\Bfootnote{Taken on its own this st. would be somewhat difficult, but in context the import is clear: Weden says that He has made good use of the Mead of Poetry by bringing it to earth, making poetry (and surely likewise other intellectual disciplines) available to men.}}\eva

\bvb The well bought thing \ken*{Mead of Poetry} have I used well— \\
\ind little do the learned lack, \\
for Woderearer is now come up \\
\ind over the rim of the \inx[C]{wigh} of men \ken*{= Middenyard}.\evb\evg


\bvg\bva\alst{I}fi ’s mér \alst{ȧ}, \hld\ at vę́ra’k \alst{ę}nn kominn &
\ind \alst{jǫ}tna gǫrðum \alst{ó}r, &
ef \alst{G}unn-laðar né nyta’k, \hld\ hinnar \alst{g}óðu konu, &
\ind es lǫgðumk \alst{a}rm \alst{y}fir.\eva

\bvb There is doubt in me, if I would yet be come \\
\ind out of the yards of the Ettins, \\
if Guthlathe I had not used, that good woman \\
\ind whom I laid my arm over.\evb\evg


\bvg\bva \edtrans{\alst{H}ins \alst{h}indra dags}{The following day}{\Bfootnote{This is the only occurrence of the comparative \emph{hindra} ‘following, next’ in the Norse (i.e. ‘belonging to Norway and its colonies’) literature. The superlative \emph{hindstr} ‘last, final’ does occur more often (e.g. \emph{indsta sinni} ‘the last time’, with loss of the \emph{h-}; see \CV: \emph{hindri}), and the possible derivative \emph{hindar-dags} ‘day after tomorrow, two days after’ is found twice, both times in the \Gulatingslog, chh. 37 and 266.  If we, however, search in the broader Scandinavian sphere, we find in the Swedish provicial laws an exact equivalent of the present phrase, namely OSwe. \emph{hindra-dagher}, a law-word referring specifically to the ‘day after the wedding’, used both on its own and in the expression \emph{hindra-dags gięf} ‘morning gift’.  If this is indeed the sense in the present stanza, two interpretations are possible: it either (i) refers sarcastically to Weden’s sleeping with Guthlathe (as would be done on the wedding night), or (ii) means that Weden actually married, or promised to marry, Guthlathe.  The latter interpretation may find support in st. 109, see notes there.}} \hld\ gingu \alst{h}rím-þursar &
\alst{H}áva ráðs at fregna, \hld\ \alst{H}áva \alst{h}ǫllu í, &
at \alst{B}ǫl-verki spurðu, \hld\ ef vę́ri með \alst{b}ǫndum kominn &
\ind eða hęfði hǫ́num \alst{S}uttungr of \alst{s}óit.\eva

\bvb The following day went the Rime-Thurses \\
\ind to ask for the High One’s counsel, in the High One’s hall. \\
About Baleworker \name{= Weden} they asked, if he were come among the bonds \name{gods}, \\
\ind or if Sutting had slain him.\evb\evg


\bvg\bva \edtext{Baug-ęið \alst{Ó}ðinn \hld\ hygg at \alst{u}nnit hafi, &
\ind hvat skal hans \alst{t}ryggðum \alst{t}rúa? &
\alst{S}uttung \alst{s}vikvinn \hld\ hann lét \alst{s}umbli frȧ &
\ind ok \alst{g}rǿtta \alst{G}unn-lǫðu}{\lemma{Baug-ęið \dots\ Gunn-lǫðu ‘A bigh-oath \dots\ brought to tears™}\Bfootnote{The exact narrative referred to in the stanza is hard to pin down, but I find the following most likely: Weden swore an oath on a bigh, its contents being that he would marry Guthlathe. Sutting then hosted a simble (banquet, drinking feast) for the new couple (cf. \emph{hins hindra dags} in st. 108), and Weden slept with her, but after. \emph{svikvinn} ‘betrayed’ and \emph{grǿtta} ‘brought to tears’ are (respectively masc. and fem.) acc. sg. past participles of the transitive verbs \emph{svíkva} ‘to betray’ and \emph{grǿta} ‘to make weep, bring to tears’. I read \emph{lét} as meaning ‘left, abandoned, forsook’.}}.\eva

\bvb A \inx[C]{bigh-oath} I ween that Weden has sworn— \\
\ind how shall one trust his truces? \\
Away from the \inx[C]{simble} he left Sutting betrayed, \\
\ind and Guthlathe, made to weep.\evb\evg

\sectionline

\section{The Speeches of Loddfathomer (\emph{Loddfáfnis mǫ́l}, 111–137)}

A series of advice stanzas addressed to \inx[P]{Loddfathomer}, an otherwise unknown figure who is clearly mythological.  The name is a compound: the first element, \emph{lodd-}, is related to ON \emph{loddari} ‘juggler, tramp’, OE \emph{loddere} ‘pauper, beggar’; the second, \emph{Fáfnir} (\inx[P]{Fathomer}), is the name of a famous Wyrm and literally means ‘embracer’.  This name gives a picture of an archetypal greedy fool; he is taught by Weden, his opposite.

The section division is found in \Regius.  Stanza 111 has a large initial \emph{M}, albeit smaller than those which introduce new chapters and poems, and the beginning of the following section, the \emph{Rune-Tally}, is also clearly marked by an initial.

\sectionline

\bvg\bva Mál ’s at \alst{þ}ylja \hld\ \alst{þ}ular stóli ȧ; &
\ind \alst{U}rðar brunni \alst{a}t &
\alst{s}á’k ok þagða’k, \hld\ \alst{s}á’k ok hugða’k, &
\ind hlýdda’k ȧ \alst{m}anna \alst{m}ál; &
of \alst{r}únar hęyrða’k dǿma, \hld\ né of \alst{r}ǫ́ðum þǫgðu &
\ind \alst{H}áva \alst{h}ǫllu at, &
\ind \alst{H}áva \alst{h}ǫllu í &
\ind hęyrða’k \alst{s}ęgja \alst{s}vá:\eva

\bvb It is time to \inx[C]{thill}, upon the \inx[C]{thyle}’s chair. \\
\ind At the \inx[L]{Well of Weird} \\
I saw and shut up; I saw and I thought; \\
\ind I heeded the matters of men. \\
Of runes I heard them speak, nor were they silent of counsels \\
\ind at the High One’s hall, \\
\ind in the High One’s hall; \\
\ind I heard them say so:\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ at \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{n}ǫ́tt þú rís-at, \hld\ nema ȧ \alst{n}jósn séir, &
\ind eða \edtrans{lęitir þér \alst{i}nnan \alst{ú}t staðar}{or thou look for thy place outside}{\Bfootnote{Lit. word-for-word “or thou look for thee from within out a place”, which becomes nonsensical.  \emph{lęita sér staðar} ‘look for one’s place’ is a euphemism, i.e. “to relieve oneself”, which was done outside.}}.\eva

\bvb I counsel thee, O Loddfathomer, that thou learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
At night do not rise, unless thou be scouting, \\
\ind or thou look for thy place outside.\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ at \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{f}jǫl-kunnigri konu \hld\ skal-at-tu í \alst{f}aðmi sofa, &
\ind svá’t hon \alst{l}yki þik \alst{l}iðum.\eva

\bvb I counsel thee, O Loddfathomer, that thou learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
By a \inx[C]{many-cunning} woman’s bosom shalt thou never sleep, \\
\ind lest she lock thee in [her?] limbs.\evb\evg


\bvg\bva Hǫ́n svá \alst{g}ørir \hld\ at \edtrans{\alst{g}ȧir}{heed}{\Bfootnote{The nasal vowel here is based on Elfdalian \emph{gą̊}.}} ęigi &
\ind \alst{þ}ings né \alst{þ}jóðans máls; &
\alst{m}at þú vill-at \hld\ né \alst{m}anns-kis gaman &
\ind fęrr þú \alst{s}orga-fullr at \alst{s}ofa.\eva

\bvb She makes it so that thou heed not \\
\ind \inx[C]{Thing}’s or ruler’s speech; \\
thou hast no wish for food nor any man’s pleasure; \\
\ind thou goest sorrowful to sleep.\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ at \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{a}nnars konu \hld\ tęyg þér \alst{a}ldri-gi &
\ind \edtrans{\alst{ęy}ra-rúnu}{ear-whisperer \ken{lover}}{\Bfootnote{This word is also used in \Voluspa\ 38, in which male seducers of married women are among those being forced to wade through “heavy streams” in the afterlife.}} \alst{a}t.\eva

\bvb I counsel thee, O Loddfathomer, that thou learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
Another man’s woman do never tug \\
\ind into becoming thy ear-whisperer \ken{lover}.\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\edtrans{\alst{f}jalli eða \alst{f}irði}{on fell or firth}{\Bfootnote{i.e. ‘hiking through mountains or travelling at sea’; a very Norwegian expression.  This word pair is a formulaic merism; this is its only poetic attestation, but it is found a few times in the Old Norwegian laws.}}, \hld\ ef þik \alst{f}ara tíðir, &
\ind fȧsk-tu at \alst{v}irði \alst{v}ęl.\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
on fell or firth—if thou desire to journey— \\
\ind furnish thyself well with food.\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{i}llan mann \hld\ lát \alst{a}ldri-gi &
\ind \edtext{\alst{ȯ}-hǫpp at þér \alst{v}ita}{\Bfootnote{An unambiguous instance of \emph{v} alliterating with a vowel.}}, &
því-at af \alst{i}llum manni \hld\ fę̇r \alst{a}ldri-gi &
\ind \alst{g}jǫld hins \alst{g}óða hugar.\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
An evil man do never let \\
\ind know of thy misfortunes; \\
for from an evil man gettest thou never \\
\ind rewards for thy good will.\evb\evg


\bvg\bva\edtrans{\alst{O}far-la}{Sorely}{\Bfootnote{Contraction of \emph{ofar-liga} ‘\CV: high up, in the upper part’, presumably meaning that the words were particularly grievous or insulting, i.e., they “got to him”.  Whether he was murdered or committed suicide is not clear.}} bíta \hld\ sá’k \alst{ęi}num hal &
\ind \alst{o}rð \alst{i}llrar konu, &
\edtrans{\alst{f}lá-rǫ́ð tunga}{a false-counseling tongue}{\Bfootnote{Cf. \Lokasenna\ 31/1: \emph{flǫ́ ’s þér tunga} ‘false is thy tongue’.}} \hld\ varð hǫ́num at \alst{f}jǫr-lagi &
\ind ok þęygi of \alst{s}anna \alst{s}ǫk.\eva

\bvb Sorely biting I saw at a lonely man \\
\ind the words of an evil woman; \\
a false-counseling tongue brought his life to its end, \\
\ind and in no way over a truthful charge.\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{v}ęitst, ef \alst{v}in átt, \hld\ þann’s \alst{v}ęl trúir, &
\ind \alst{f}ar þú at \alst{f}inna opt; &
því-at \edtrans{\alst{h}rísi vęx \hld\ ok \alst{h}ǫ́u grasi}{with brushwood and with tall grass grows}{\Bfootnote{Identical to \Grimnismal\ 17/1.}} &
\ind \alst{v}egr, es \alst{v}ę́t-ki trøðr.\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
Thou knowest, if thou have a friend whom thou well trust: \\
\ind journey to find him oft; \\
for with brushwood and tall grass grows \\
\ind the way which no one treads.\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{g}óðan mann \hld\ tęyg þér at \edtrans{\alst{g}aman-rúnum}{pleasure-runes}{\Bfootnote{Here “rune” appears to carry its root meaning of ‘whisper, counsel, speech’, thus ‘pleasing speech’.  Cf. st. 129 where this word reoccurs.}} &
\ind ok nem \edtrans{\alst{l}íknar-galdr}{liking-galders}{\Bfootnote{i.e. ways of speaking which will make one liked or popular.  For \emph{líkn} ‘liking’ see sts. 8 (with note) and 123.}} meðan \alst{l}ifir.\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
A good man do tug toward thee with pleasure-runes, \\
\ind and learn liking-galders while thou livest.\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{v}in þínum \hld\ \alst{v}es aldri-gi &
\ind \alst{f}yrri at \alst{f}laum-slitum. &
\alst{s}org etr hjarta, \hld\ ef þú \edtext{\alst{s}ęgja né náir &
\ind \alst{ęi}n-hvęrjum \alst{a}llan hug}{\lemma{sęgja \dots\ ęin-hvęrjum allan hug ‘tell anyone thy whole mind’}\Bfootnote{Cf. st. 123 which uses almost the same expression.}}.\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
With thy friend be thou never the first \\
\ind to tear the relation apart. \\
Sorrow will eat thy heart if thou canst not tell \\
\ind anyone thy whole mind.\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\edtext{\alst{o}rðum skipta \hld\ skalt \alst{a}ldri-gi &
\ind við \edtrans{\alst{ȯ}-svinna \alst{a}pa}{unwise apes}{\Bfootnote{Formulaic; cf. \Grimnismal\ 33, \Fafnismal\ 11.}}}{\lemma{orðum \dots\ apa ‘Words \dots\ apes’}\Bfootnote{Cf. st. 125 which gives similar advice.}},\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
Words shalt thou never exchange \\
\ind with unwise apes,\evb\evg


\bvg\bva \edtext{því-at af \alst{i}llum manni \hld\ munt \alst{a}ldri-gi &
\ind \alst{g}óðs laun of \alst{g}eta}{\lemma{því-at \dots\ geta ‘For \dots\ praise’}\Bfootnote{Cf. st. 117/6–7.}}, &
en \alst{g}óðr maðr \hld\ mun þik \alst{g}ørva męga &
\ind \edtrans{\alst{l}íkn-fastan}{steadfast in liking}{\Bfootnote{The first element \emph{líkn} ‘liking’ is somewhat difficult; see sts. 8 (with note) and 120.  For the present cpd \textcite{LaFargeGlossary} give a tentative ‘assured of favour’, while \CV\ gives ‘fast in goodwill, beloved’.}} at \alst{l}ofi.\eva

\bvb for from an evil man wilt thou never \\
\ind get a reward for thy goodness, \\
but a good man will know to make thee \\
\ind steadfast in liking by [his] praise.\evb\evg


\bvg\bva\alst{S}ifjum ’s þȧ blandit \hld\ hvęrr es \edtext{\alst{s}ęgja rę́ðr &
\ind \alst{ęi}num \alst{a}llan hug}{\lemma{sęgja \dots\ \alst{ęi}num \alst{a}llan hug ‘tell one man his whole mind’}\Bfootnote{Cf. st. 121 which uses almost the same expression.}}; &
alt es \alst{b}ętra \hld\ an sé \alst{b}rigðum at vesa: &
es-a sá \alst{v}inr ǫðrum \hld\ es \alst{v}ilt ęitt sęgir.\eva

\bvb Kinship is blended for whomever resolves to tell \\
\ind one man his whole mind. \\
Everything is better than to be with the fickle; \\
he is no friend to another who tells only what is pleasant.\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\edtrans{þrimr \alst{o}rðum}{With three words}{\Bfootnote{i.e. ‘not even with three words’. If one understands \emph{orð} to mean ‘speech’, it may be interpreted as that if one says something (the first speech) to which another man responds insultingly (the second speech), one should not respond a third time and turn it into a fight.}} sęnna \hld\ skal-at-tu þér við \alst{v}erra mann; &
\ind opt hinn \alst{b}ętri \alst{b}ilar, &
\ind þȧ’s hinn \alst{v}erri \alst{v}egr.\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
With three words shalt thou not flyte with a worse man; \\
\ind oft the better man breaks \\
\ind when the worse man strikes.\footnoteB{Cf. st. 121.}\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{sk}ó-smiðr þú vesir \hld\ né \alst{sk}ępti-smiðr, &
\ind nema \alst{s}jǫlfum þér \alst{s}éir. &
\alst{Sk}ór ’s \alst{sk}apaðr illa \hld\ eða \alst{sk}apt sé rangt, &
\ind þȧ ’s þér \alst{b}ǫls \alst{b}eðit.\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
Be not a shoe-maker nor shaft-maker, \\
\ind unless thou be one for thyself. \\
The shoe is shaped badly or the shaft be crooked— \\
\ind then for thee a \inx[C]{bale} is bid.\footnoteB{i.e. the customer will place a curse on you if he dislikes the wares.}\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
hvar’s \alst{b}ǫl kant, \hld\ kveð þér \alst{b}ǫlvi at &
\ind ok gef-at þínum \alst{f}jǫ́ndum \alst{f}rið.\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
Wherever thou knowest a bale, call it a bale against thee, \\
\ind and give not thy foes peace.\footnoteB{i.e. “if somebody puts a curse on you, do not ignore it, but respond decisively”.  This st. has often been interpreted as a command to call out evil, even when committed towards somebody else, and while there is nothing in it that speaks clearly against that interpretation, it does not agree with the general spirit of the \Havamal, which is one of caution and shrewdness.}\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{i}llu fęginn \hld\ ves \alst{a}ldri-gi, &
\ind \edtrans{en lát þér at \alst{g}óðu \alst{g}etit}{but [rather] let thyself be pleased by good}{\Bfootnote{This construction is equivalent to \CV: \emph{geta}, A. IV. with acc.}}.\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
Rejoicing in evil be thou never, \\
\ind but let thyself be pleased by good.\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{u}pp líta \hld\ skal-at-tu í \alst{o}rrostu; &
—\alst{g}jalti \alst{g}líkir \hld\ verða \alst{g}umna synir— &
\ind síðr þitt of \alst{h}ęilli \edtrans{\alst{h}alir}{warriors}{\Bfootnote{Some sort of “supernatural sky warriors”, to quote \textcite{PettitEdda}—perhaps even the \inx[G]{Oneharriers}.}}.\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
Look upward shalt thou not in battle \\
—alike to a madman become the sons of men— \\
\ind lest warriors bewitch thee.\footnoteB{An obscure superstition; the interpretation hinges on the word \emph{gjalti} ‘madman’ dat. sg., which must be compared with closely related phrase \emph{verða at gjalti} ‘to be turned into a “gelt”’.

(1) \CV\ explains it as an old dative of \emph{gǫltr} ‘boar, hog’.  This neccessitates an irregular breaking of \emph{ja} < \emph{ę}, since \emph{gǫltr} (< Proto-Norse \emph{*galtuʀ}) is an u-stem and should have dat. sg. \emph{gęlti} (< \emph{*galtiu}, cf. \textbf{kunimudiu}, dat. sg. of \emph{*Kunimunduʀ}, on the Tjurkö 1 bracteate).

(2) The generally accepted explanation in modern scholarship seems to be a borrowing from Old Irish \emph{geilt} ‘insane, mad’ (so \textcite{LaFargeGlossary} and others).  The closest parallel to the present stanza is the C12th or C13th tale of Suibhne mac Colmáin, who was cursed by saint Rónán Finn to become mad; the curse took effect when he looked into the sky during a battle, after which he was known as Suibhne geilt.  Earlier attestations of the same story are found as early as the 840s (see \textcite[100]{Males2024}).}\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
Ef vilt þér \alst{g}óða konu \hld\ kvęðja at \edtrans{\alst{g}aman-rúnum}{pleasure-runes}{\Bfootnote{While easily interpreted as ‘sexual intercourse’, the word is used in st. 120 in a decidedly non-sexual sense.  Its base meaning is probably ‘good conversation’.}} &
\ind ok \alst{f}ȧa \alst{f}ǫgnuð af, &
\alst{f}ǫgru skalt hęita \hld\ ok láta \alst{f}ast vesa; &
\ind lęiðisk mann-gi \alst{g}ótt ef \alst{g}etr.\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
If thou wilt for thyself greet a good woman to pleasure-runes, \\
\ind and get good cheer from her; \\
fair things shalt thou promise, and let it be fast; \\
\ind no man loathes a good thing if he gets it.\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{v}aran bið’k þik \alst{v}esa \hld\ ok ęigi of·\alst{v}aran, &
ves við \alst{ǫ}l varastr, \hld\ ok við \alst{a}nnars konu &
ok við \alst{þ}at hit \alst{þ}riðja, \hld\ at \alst{þ}jófar né lęiki.\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
Wary I ask thee to be, and not over-wary; \\
be thou wariest with ale, and with another man’s woman, \\
and with the third, that thieves do not outplay [thee].\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
at \alst{h}áði né \alst{h}látri \hld\ \alst{h}af aldri-gi &
\ind \alst{g}ęst né \alst{g}anganda.\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
In scorn or laughter do never have \\
\ind a guest or wanderer.\evb\evg


\bvg\bva\alst{O}pt vitu \alst{ȯ}-gǫrla, \hld\ þęir’s sitja \alst{i}nni fyrir, &
\ind hvęrs þęir ’ru \alst{k}yns es \alst{k}oma; &
es-at maðr svá \alst{g}óðr \hld\ at \alst{g}alli né fylgi, &
\ind né svá \alst{i}llr at \alst{ęi}nu-gi dugi.\eva

\bvb Oft they know unclearly, who sit further within, \\
\ind of what kind are those who come; \\
there is no man so good that no flaw follows, \\
\ind nor so bad that he for nothing avails.\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
at \alst{h}ǫ́rum þul \hld\ \alst{h}lę́ aldri-gi, &
\ind opt ’s \alst{g}ótt þat’s \alst{g}amlir kveða, &
opt ór \alst{sk}ǫrpum bęlg \hld\ \alst{sk}ilin orð koma &
\ind þęim’s \alst{h}angir með \alst{h}ǫ́um &
\ind ok \alst{sk}ollir með \alst{sk}rǫ́um, &
\ind ok \alst{v}áfir með \alst{v}íl-mǫgum.\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
At a hoary thyle do never laugh; \\
\ind oft is good that which old men sing. \\
Oft from scorched leather come discerning words; \\
\ind from him who hangs with hides, \\
\ind and dangles with dry skins, \\
\ind and sways among lads of toil \ken{thralls}.\footnoteB{TODO: Some note. \emph{vil-mǫgum} meaning ‘veal-stomachs’? Cf. Crawford’s video and Finnur on this.}\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
\alst{g}ęst þú né \alst{g}ęyj-a \hld\ \edtrans{né ȧ \alst{g}rind hrę́kir}{nor spit at the gate}{\Bfootnote{The guest is presumably standing behind gate waiting for the farmer to open it and let him in.}}; &
\ind get þú \alst{v}ǫ́-luðum \alst{v}ęl.\eva

\bvb I counsel thee, O Loddfathomer—and thou oughtst to learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
At a guest bark not, nor spit at the gate; \\
\ind furnish the destitute well.\evb\evg


\bvg\bva\alst{R}ammt es þat tré, \hld\ es \alst{r}íða skal &
\ind \alst{ǫ}llum at \alst{u}pp-loki; &
\alst{b}aug þú gef \hld\ eða þat \alst{b}iðja mun &
\ind þér \alst{l}ę́s hvęrs ȧ \alst{l}iðu.\eva

\bvb Strong is that wood which shall swing \\
\ind to open up for all.\footnoteB{i.e. the beam of the gate in front of the farm.} \\
Do give a bigh, or it will bid \\
\ind every kind of guile onto thy limbs.\evb\evg


\bvg\bva\alst{R}ǫ́ðumk þér Loddfáfnir, \hld\ en \alst{r}ǫ́ð nemir, &
\ind \alst{n}jóta munt ef \alst{n}emr, &
\ind þér munu \alst{g}óð ef \alst{g}etr: &
hvar’s \alst{ǫ}l drekkir \hld\ kjós þér \alst{ja}rðar męgin, &
því-at \alst{jǫ}rð tękr við \alst{ǫ}lðri, \hld\ en \alst{ę}ldr við sóttum, &
\alst{ęi}k við \alst{a}bbindi, \hld\ \alst{a}x við fjǫl-kyngi, &
\alst{h}ǫll við \alst{h}ýrógi; \hld\ \edtrans{\alst{h}ęiptum skal Mána kvęðja}{in feuds shall one hail Moon}{\Bfootnote{Cf. \Voluspa\ 5 which mentions the “Moon’s might”; for which He is presumably here invoked.  For \emph{kvęðja} ‘hail, invoke’ cf. \Lokasenna\ P3.}}, &
\alst{b}ęiti við \alst{b}it-sóttum, \hld\ en við \alst{b}ǫlvi rúnar; &
\ind \alst{f}old skal við \alst{f}lóði taka.\eva

\bvb I counsel thee, O Loddfathomer, that thou learn the counsels; \\
\ind thou wilt have use if thou learn, \\
\ind they will be good for thee if thou get: \\
Wherever thou drinkest ale choose thee Earth’s might, \\
for earth takes against drunkenness, and fire against sicknesses; \\
oak against dysentery; the ear [of corn] against sorcery; \\
bearded rye against hernia—in feuds shall one hail Moon— \\
heather against bite-sicknesses, and \inx[C]{rune}[runes] against a \inx[C]{bale};\footnoteB{cf. sts. 126, 152.} \\
\ind earth shall one have against flood.\evb\evg

\sectionline

\section{The Rune-Tally (138–146)}

This group of stanzas is introduced by a large initial in \Regius, marking the beginning of a new section.  In younger paper manuscripts they have the header \emph{Rúna-tals þáttr} ‘Strand of the Rune-Tally’, and generally give an archaic, mystic impression; at times one gets a feeling that they were drawn from the lips of an Odinic priest.

Apart from these stanzas there are a few other manuscript attestations of similar Runic magic.  Closest at hand is st. 80 above, which would fit seamlessly into the present section.  Outside of \Havamal\ there is \Sigrdrifumal\ 5–17, also preserved in \Regius.

\sectionline

\bvg\bva\alst{V}ęit’k at ek hekk \hld\ \edtrans{\alst{v}indga-męiði}{the windy beam}{\Bfootnote{Generally understood to be a variant of \emph{vinga-męiðr} ‘gallows tree’ found in three Scaldic stanzas (\Skp\ signa: Egill Lv 14, Eyv \emph{Hál} 5, Anon (FoGT) 17).}} ȧ &
\ind \alst{n}ę́tr allar \alst{n}íu, &
\alst{g}ęiri undaðr \hld\ ok \alst{g}efinn Óðni, &
\ind \alst{s}jalfr \alst{s}jǫlfum mér, &
ȧ þęim \alst{m}ęiði, \hld\ es \alst{m}ann-gi vęit, &
\ind hvęrs af \alst{r}ótum \alst{r}innr.\eva

\bvb I know that I hung on the windy beam \\
\ind for nine nights all, \\
wounded by spear and given to Weden, \\
\ind myself to myself— \\
on that beam where no man knows \\
\ind of whose roots it runs.\evb\evg


\bvg\bva Við \edtext{\alst{h}lęifi mik sǿldu-t \hld\ né við \alst{h}orni-gi}{\lemma{hlęifi \dots\ horni-gi ‘loaf \dots\ horn’}\Bfootnote{i.e. “I got neither bread to eat nor ale to drink.”}}; &
\alst{n}ýsta ek \alst{n}iðr, \hld\ \alst{n}am’k upp rúnar, &
\alst{ǿ}pandi nam, \hld\ fell’k \alst{a}ptr þaðan.\eva

\bvb With loaf they relieved me not, nor with any horn. \\
I peered down; I took up the runes; \\
screaming I took—I fell back thence.\evb\evg


\bvg\bva\edtrans{\alst{F}imbul-ljóð níu}{Nine fimble-leeds}{\Bfootnote{Nine very great chants or spells (\inx[C]{galders}), compare the eighteen leeds below (st. 147 onward).  It is unclear what this has to do with Weden’s Hanging; this stanza may be an insert.}} \hld\ nam’k af \edtext{hinum \alst{f}rę́gja syni &
\ind \alst{B}ǫlþorns, \alst{B}ęstlu fǫður,}{\lemma{hinum frę́gja syni Bǫlþorns, Bęstlu fǫður ‘the famed son of Balethorn, Bestle’s father’}\Bfootnote{According to \Gylfaginning\ 6: \emph{[Borr] fekk þeirar konu, er Bestla hét, dóttir Bǫlþorns jötuns, ok fengu þau þrjá sonu; hét einn Óðinn, annarr Vili, þriði Vé [\dots]} ‘[Byre] got the wife called Bestle, the daughter of the ettin Balethorn, and they had three sons: one was called Weden, the other Will, the third Wigh.’  Balethorn’s son is then Weden’s uncle, an instance of the old Indo-European custom of sending sons away to be fostered by the mother’s male relations.  Cf. TODO: some reference.}} &
ok ek \alst{d}rykk of gat \hld\ hins \alst{d}ýra mjaðar &
\ind \alst{au}sinn \alst{Ó}ð-rǿri.\eva

\bvb Nine \inx[C]{fimble}-leeds I learned from the famed son \\
\ind of \inx[P]{Balethorn}, \inx[P]{Bestle}’s father— \\
and a drink I got of the dear mead \\
\ind poured from \inx[P]{Woderearer}.\evb\evg


\bvg\bva Þȧ \edtrans{nam’k \alst{f}rę́vask}{I began to flourish}{\Bfootnote{A notorious mistranslation popularized by \textcite{Greenberg1988} has rendered these words as “I took semen”.  They would supposedly reference Weden stealing the ejaculate from hanged men in order to replenish his own powers—something not otherwise attested.  This preposterous notion makes no sense in the context of the text and has no philological grounding.  While Old Norse \emph{frę́} does mean “seed”, it only refers to the seeds of plants, not the seed animals or men.  Regardless, \emph{frę́vask} is without doubt a reflexive verb literally meaning something like ‘cultivate oneself’.}} \hld\ ok \alst{f}róðr vesa &
\ind ok \alst{v}axa ok \alst{v}ęl hafask; &
\edtext{\alst{o}rð mér af \alst{o}rði \hld\ \alst{o}rðs lęitaði &
\alst{v}erk mér af \alst{v}erki \hld\ \alst{v}erks lęitaði.}{\lemma{orð \dots\ lęitaði. ‘My word \dots sought out.’}\Bfootnote{Every good speech led to another; every good deed likewise.}}\eva

\bvb Then I took to flourish and be wise, \\
\ind and grow and have it well. \\
My word from a word a word sought out; \\
my work from a work a work sought out.\evb\evg


\bvg\bva\edtrans{\alst{R}únar munt finna \hld\ ok \alst{r}áðna stafi}{Runes wilt thou find, and interpreted staves’}{\Bfootnote{A strong resemblance is found in the long-line on the medieval runestone N 13: \emph{rúnar ek ríst \hld\ ok ráðna stafi} ‘runes I carve, and interpreted staves.’}}, &
\ind mjǫk \alst{st}óra \alst{st}afi, &
\ind mjǫk \alst{st}inna \alst{st}afi, &
\ind es \alst{f}áði \alst{F}imbul-þulr &
\ind ok \alst{g}ørðu \alst{g}inn-ręgin &
\ind ok \alst{r}ęist Hroptr \edtrans{\alst{r}agna}{of the Reins}{\Afootnote{\emph{‘rǫgna’} \Regius}}.\eva

\bvb \inx[C]{rune}[Runes] wilt thou find, and interpreted staves: \\
\ind very large staves, \\
\ind very stiff staves, \\
\ind which \inx[P]{Fimble-Thyle} \name{= Weden} painted, \\
\ind and the \inx[G]{yin-Reins} made, \\
\ind and Roft \name{= Weden} of the Reins carved.\evb\evg


\bvg\bva\alst{Ó}ðinn með \alst{ǫ̇}sum, \hld\ en fyr \alst{ǫ}lfum Dáinn, &
\ind \alst{D}valinn \alst{d}vergum fyrir, &
\ind \alst{Á}sviðr \alst{jǫ}tnum fyrir, &
\ind \edtrans{ek}{I}{\Bfootnote{The identity of the speaker is unclear; one would expect it to be Weden, but He is already named in line 1.}} ręist \alst{s}jalfr \alst{s}umar.\eva

\bvb \inx[P]{Weden} among the \inx[G]{Eese} and \inx[P]{Dowen} for the \inx[G]{Elves}; \\
\ind \inx[P]{Dwollen} for the \inx[G]{Dwarfs}; \\
\ind \inx[P]{Oswith} for the Ettins; \\
\ind I myself carved some.\evb\evg


\bvg\bva Vęitst, hvé \alst{r}ísta skal? \hld\ Vęitst, hvé \alst{r}áða skal? &
Vęitst, hvé \alst{f}áa skal? \hld\ Vęitst, hvé \alst{f}ręista skal? &
Vęitst, hvé \alst{b}iðja skal? \hld\ Vęitst, hvé \alst{b}lóta skal? &
Vęitst, hvé \alst{s}ęnda skal? \hld\ Vęitst, hvé \alst{s}óa skal?\eva

\bvb Knowest thou how one shall carve? Knowest thou how one shall read? \\
Knowest thou how one shall paint? Knowest thou how one shall try? \\
Knowest thou how one shall bid? Knowest thou how one shall \inx[C]{bloot}? \\
Knowest thou one shall send? Knowest thou how one shall \inx[C]{soo}?\footnoteB{The first four verbs refer to runes: carving, interpreting, colouring (with blood?), and divining.  The latter four refer to sacrifice: praying, worshipping, sending (the sacrifice or the prayer to the gods), and wasting the victim.  See further relevant Index entries: bloot, soo.}\footnoteB{The meter of this st. is unusual, but bears some resemblance to Vg 216 (the Högstena galder). TODO: Elaborate.}\evb\evg


\bvg\bva%
\edtext{\alst{B}ętra ’s ȯ-\alst{b}eðit \hld\ an sé of·\alst{b}lótit, &
\ind ęy sér til \alst{g}ildis \alst{g}jǫf; &
bętra ’s ȯ-\alst{s}ęnt \hld\ an sé of·\alst{s}óit; &
\edtext{[...]}{\Bfootnote{For metrical reasons it is very likely that a line has been lost here.}}}{\lemma{ALL}\Bfootnote{An identical progression of four verbs suggests a close relation with the previous st. — I believe \textcite{Males2024} has the correct interpretation: since a gift always requires recompense, an excessive sacrifice could be seen as offensive and upset the relationship with the god.  Males draws the analogy with an episode in \EgilsSaga, where a rival poet leaves an expensive shield for Eyel and rides off; the latter understands this as a demand to compose a poem about the shield, and is greatly angered.  The cycle of gifts and rewards between men and the gods is very important in Indo-European pagan religions; compare the Sanskrit phrase \emph{Dehí me, dádāmi te} ‘Give to me, I give to thee’ and Latin \emph{dō ut dēs} ‘I give that thou might give’.}}\eva

\bvb It is better unbid than over-\inx[C]{bloot}[blooted]; \\
\ind a gift always looks for recompense. \\
It is better unsent than over-\inx[C]{soo}[sooed]; \\
{[...]}\evb\evg


\bvg\bva Svá \alst{Þ}undr of ręist \hld\ fyr \alst{þ}jóða rǫk, &
þar’s \alst{u}pp of ręis, \hld\ es \alst{a}ptr of kom.\eva

\bvb Thus did \inx[P]{Thound} \name{= Weden} carve for the rakes of nations, \\
where up he rose as back he came.\footnoteB{TODO: A very cryptic st.}\evb\evg

\sectionline

\section{The Leed-Tally (147–165)}

This section of \Havamal, the so-called the Leed-Tally (\emph{Ljóðatal}), is not separated from the preceding section (which is marked out with a large initial), but is usually taken as separate since it is a self-contained list not much concerned with runes.  The speaker, Weden, addressing Loddfathomer, lists eighteen galders or spells he knows.  The spells themselves are not given; only their purpose.  They are aristocratic and Odinic in character and deal with such things as battle (3, 4, 5, 8, 11, 13), healing (galder 2, 12), countering sorcery (6, 10), controlling the elements (7, 9), and seduction (16, 17).  The eighteenth and last spell is a mystery; not even its purpose is told, and it is known only by Weden and his closest women.

There is a clear relation to other known Germanic galders.  The fourth bears a strong likeness to \Grougaldr\ 10, and its effect (removing fetters) is shared with the High German \MerseburgOne, an actual galder of that type.  The mysterious eighteenth spell finds an interesting parallel in the unknowable eighteenth question posed by Weden in \Vafthrudnismal\ 54.

\sectionline

\bvg\bva Ljóð \alst{þ}au kann’k, \hld\ es kann-at \alst{þ}jóðans kona &
\ind ok \alst{m}anns-kis \alst{m}ǫgr. &
\alst{H}jǫlp hęitir ęitt, \hld\ þat þér \alst{h}jalpa mun &
við \alst{s}orgum ok \edtrans{\alst{s}ǫkum}{sakes}{\Bfootnote{Legal charges, the first element of English \emph{sakeless}.}}, \hld\ ok \alst{s}útum gǫrv-ǫllum.\eva

\bvb Those \inx[C]{leed}[leeds] I know, which no king’s wife knows, \\
\ind and no man’s lad. \\
Help is one called, it will help thee \\
against sorrows and sakes, and all kinds of griefs.\evb\evg


\bvg\bva Þat kann’k \alst{a}nnat, \hld\ es \edtrans{þurfu \alst{ý}ta synir}{the sons of men need}{\Bfootnote{Cf. the similar wording in 166/2.}}, &
\ind þęir’s vilja \alst{l}ę́knar \alst{l}ifa.\eva

\bvb This I know second, which the sons of men need, \\
\ind who wish to live as leechers.\evb\evg


\bvg\bva Þat kann’k \alst{þ}riðja, \hld\ ef mér verðr \alst{þ}ǫrf mikil &
\ind \alst{h}apts við mína \alst{h}ęipt-mǫgu, &
\alst{ę}ggjar dęyfi’k \hld\ minna \alst{a}nd-skota, &
\ind bíta-t þęim \alst{v}ǫ́pn né \edtrans{\alst{v}ęlir}{staffs}{\Bfootnote{Plural of \emph{vǫlr}, here referring to the magic staff or sceptre used by witches and warlocks; the word \emph{vǫlva} ‘\inx[C]{wallow}’ (seeress, prophetess) is probably derived from this word.  The reading \emph{vélir} ‘wiles, tricks, deceits’ must be excluded for metrical reasons, since a c-verse in \Ljodahattr\ cannot end in a trochée.}}.\eva

\bvb This I know third, if I come in great need \\
\ind of hindrance against my feud-lads \ken{foes}; \\
I dull the edges of my opponents; \\
\ind for them bite no weapons nor staffs.\evb\evg


\bvg\bva \edtext{Þat kann’k \alst{f}jórða, \hld\ ef mér \alst{f}yrðar bera &
\ind \alst{b}ǫnd at \alst{b}óg-limum, &
svá ek \alst{g}ęl, \hld\ at \alst{g}anga má’k, &
\ind sprettr mér af \alst{f}ótum \alst{f}jǫturr, &
\ind en af \alst{h}ǫndum \alst{h}apt.}{\lemma{ALL}\Bfootnote{Cf. \Grougaldr\ 10, which is very similar to the present stanza, and \MerseburgOne\ (edited below under Galders), a galder that seems actually to have been used for loosening fetters.}}\eva

\bvb This I know fourth, if men bear \\
\ind bonds onto my shoulder-limbs: \\
\emph{so} I gale that I may walk; \\
\ind springs from my feet the fetter, \\
\ind and from my hands the bond.\evb\evg


\bvg\bva Þat kann’k \alst{f}imta, \hld\ ef sé’k af \alst{f}ári skotinn &
\ind \alst{f}lęin í \alst{f}olki vaða, &
flýgr-a svá \alst{st}int, \hld\ at \alst{st}ǫðvi’g-a’k, &
\ind ef hann \alst{s}jónum of \alst{s}é’k.\eva

\bvb This I know fifth, if I see a dangerously shot \\
\ind arrow in the troop wading: \\
it flies not so stiff that I may not stop it, \\
\ind if I see it with my sights.\evb\evg


\bvg\bva Þat kann’k \alst{s}étta, \hld\ \edtext{ef mik \alst{s}ę́rir þegn &
\ind ȧ \alst{r}ótum \edtrans{\alst{r}ás}{raw/sappy}{\Bfootnote{The normal form of this word is \emph{hrár} (cf. \Skirnismal\ 32), but the required alliteration with \emph{rótum} makes it impossible here.}} viðar}{\lemma{ef mik sę́rir þegn ȧ rótum rás viðar ‘if a thane wounds me on the roots of a raw/sappy tree’}\Bfootnote{I.e., “if a man carves a runic curse against me”.  The sappy wood was apparently of importance for the curse; cf. the curious account of \Grettissaga\ 79, where a hag curses Gretter in the following way: after finding a small tree and planing a small smooth surface onto a scorched side of it, she carves runes in its roots and reddens them with her own blood.  She then chants \inx[C]{galder}[galders] while walking counter-clockwise around it.  Lastly she pushes it out to sea, praying for it to drift to Gretter’s homestead and curse him.  Cf. also \Skirnismal\ 32 where Shirner goes to a \emph{hrár viðr} ‘raw/sappy tree’ to get a certain curse-object.}}, &
þann \alst{h}al, \hld\ es mik \alst{h}ęipta kveðr, &
\ind þann eta \alst{m}ęin hęldr an \alst{m}ik.\eva

\bvb This I know sixth, if a thane wounds me \\
\ind on the roots of a raw/sappy tree: \\
that man who sings hatred against me, \\
\ind \emph{him} the harms eat instead of me.\evb\evg


\bvg\bva Þat kann’k \alst{s}jaunda, \hld\ ef \alst{s}é’k hǫ́van loga &
\ind \alst{s}al of \alst{s}ess-mǫgum, &
\alst{b}rinnr-at svá \alst{b}ręitt, \hld\ at hǫ́num \alst{b}jargi’g-a’k; &
\ind þann kann’k \alst{g}aldr at \alst{g}ala.\eva

\bvb This I know seventh, if I see a high hall \\
\ind blazing over seat-lads \ken{warriors}: \\
it burns not so broadly that I cannot save it— \\
\ind that galder I can gale.\evb\evg


\bvg\bva Þat kann’k \alst{á}tta, \hld\ es \alst{ǫ}llum es &
\ind \alst{n}yt-sam-ligt at \alst{n}ema, &
\alst{h}var’s \edtrans{\alst{h}atr}{hatred}{\Bfootnote{i.e. with regard to the father’s inheritance.}} vęx \hld\ með \alst{h}ildings sonum, &
\ind þat má’k \alst{b}ǿta \alst{b}rátt.\eva

\bvb This I know eighth, which for all men is \\
\ind useful to learn: \\
wherever hatred grows among a prince’s sons, \\
\ind it I may shortly mend.\evb\evg


\bvg\bva Þat kann’k \alst{n}íunda, \hld\ ef mik \alst{n}auðr of stęndr &
\ind at bjarga \alst{f}ari mínu ȧ \alst{f}loti, &
\alst{v}ind ek kyrri \hld\ \alst{v}ági ȧ &
\ind ok \alst{s}vę́fi’k allan \alst{s}ę́.\eva

\bvb This I know ninth, if I come in need \\
\ind of saving my ride on a floater \ken{ship}: \\
the wind I calm upon the wave, \\
\ind and put all the sea asleep.\evb\evg


\bvg\bva Þat kann’k \alst{t}íunda, \hld\ ef sé’k \edtrans{\alst{t}ún-riður}{town-rideresses}{\Bfootnote{The \emph{riður} ‘rideresses’ were witches who would leave their original human shapes or skins (\emph{hamir}) in order to fly around in the air tormenting and poisoning villagers.  Their original bodies would then be lying in a coma-like state, something like “astral projection”.  It was not the case that their whole mental faculties would disconnect from their bodies, but rather they would leave behind something of their humanity, which was thought to be inextricably linked to their human bodies.  Through his second sight, Weden was could see these riders, and would then use his superior magical wisdom to confuse them so that they would not be able to return to their human “home-shapes” or minds, but would instead be forced to stray as tormented bodyless ghosts; a cruel fate.  Weden also brags about tricking riders in \Harbardsljod\ 20.}} &
\ind \alst{l}ęika \alst{l}opti ȧ, &
ek svá \alst{v}inn’k, \hld\ at \edtrans{þę́r \alst{v}illar fara}{they (\emph{fem.}) go astray}{\Afootnote{emend.; \emph{þęir villir fara} ‘they (\emph{masc.}) go astray’ \Regius}} &
\ind sinna \alst{h}ęim-\alst{h}ama &
\ind sinna \alst{h}ęim-\alst{h}uga.\eva

\bvb This I know tenth, if I see \inx[G]{town-rideresses} \\
\ind playing aloft: \\
I work it so that they go astray \\
\ind of their home-\inx[C]{hame}[hames], \\
\ind of their home-minds.\evb\evg


\bvg\bva Þat kann’k \alst{ę}llipta, \hld\ ef skal’k til \alst{o}rrostu &
\ind \alst{l}ęiða \edtrans{\alst{l}ang-vini}{old friends}{\Bfootnote{In Germanic paganism the followers and protégés of a god are his friends (\emph{vinir}).  Already in \Beowulf\ we see that the Shieldings are called the \emph{Ing-wine} ‘friends of \inx[P]{Ing}’, and in \Hymiskvida\ 11 Thunder is called the \emph{vinr ver-liða} ‘friend of manly retinues’.  Two other places where it is used of Weden’s followers in particular are \Grimnismal\ 54 and \Sonatorrek\ 22, where Eyel speaks about his friendship (\emph{vin-átt}) with Weden.}}, &
und \alst{r}andir gęl’k, \hld\ en þęir með \alst{r}íki fara, &
\ind \alst{h}ęilir \alst{h}ildar til, &
\ind \alst{h}ęilir \alst{h}ildi frȧ, &
\ind koma þęir \alst{h}ęilir \alst{h}vaðan.\eva

\bvb This I know eleventh, if I shall into the fray \\
\ind lead old friends: \\
beneath the shields I gale, and they go with power \\
\ind healthy to the battle, \\
\ind healthy from the battle; \\
\ind they return healthy anywhence.\evb\evg


\bvg\bva Þat kann’k \alst{t}olpta, \hld\ ef sé’k ȧ \alst{t}ré uppi &
\ind \alst{v}áfa \alst{v}irgil-ná, &
svá ek \alst{r}íst \hld\ ok í \alst{r}únum fá’k, &
\ind at sá \alst{g}ęngr \alst{g}umi. &
\ind ok \alst{m}ę́lir við \alst{m}ik.\eva

\bvb This I know twelfth, if I see in a tree up high \\
\ind a gallow-corpse dangling: \\
so I carve and paint in the runes, \\
\ind that that man walks \\
\ind and speaks with me.\evb\evg


\bvg\bva Þat kann’k \alst{þ}rettánda \hld\ \edtext{ef skal’k \alst{þ}egn ungan &
\ind \alst{v}erpa \alst{v}atni ȧ,}{\lemma{ef skal’k þegn ungan verpa vatni ȧ ‘if on a young thane I shall sprinkle water’}\Bfootnote{A reference to the Heathen name-giving ceremony in which the infant would be sprinkled with water; cf. the attestations in \Rigsthula\ 7, 21, 34.}} &
mun-at hann \alst{f}alla \hld\ þótt í \alst{f}olk komi, &
\ind \alst{h}nígr-a sá \alst{h}alr fyr \alst{h}jǫrum.\eva

\bvb This I know thirteenth, if on a young thane \\
\ind I shall sprinkle water: \\
he will not fall though he come into battle; \\
\ind that warrior sinks not down before swords.\evb\evg


\bvg\bva Þat kann’k \alst{f}jórtánda, \hld\ ef skal’k \alst{f}yrða liði &
\ind \alst{t}ęlja \alst{t}íva fyr, &
\alst{ȧ}sa ok \alst{a}lfa \hld\ ek kann \alst{a}llra \edtrans{skil}{discernments}{\Bfootnote{Their unique traits.  Cf. \Hymiskvida\ 38, where the corresponding verb \emph{skilja} ‘to discern, understand’ is used in the context of god-lore.}}, &
\ind fár kann ȯ-\alst{s}notr \alst{s}vá.\eva

\bvb This I know fourteenth, if before a retinue of men \\
\ind I shall count forth the Tews: \\
of the Eese and Elves all I know the discernments; \\
\ind few unwise men can do so.\evb\evg


\bvg\bva\alst{Þ}at kann’k fimtánda, \hld\ es gól \alst{Þ}jóð-rǿrir &
\ind \alst{d}vergr fyr \alst{D}ęllings \alst{d}urum, &
\alst{a}fl gól \alst{ǫ́}sum, \hld\ en \alst{ǫ}lfum frama, &
\ind \alst{h}yggju \alst{H}ropta-tý.\eva

\bvb This I know fifteenth, which Thedrearer galed, \\
\ind the dwarf, before Delling’s doors. \\
Strength he galed for the Eese, and fame for the Elves, \\
\ind thought for Roft-Tew \name{= Weden}.\evb\evg


\bvg\bva Þat kann’k \alst{s}extánda, \hld\ ef vil’k hins \alst{s}vinna mans &
\ind hafa \alst{g}ęð allt ok \alst{g}aman, &
\alst{h}ugi \alst{h}vęrfi’k \hld\ \alst{h}vit-armri konu &
\ind ok \alst{s}ný’k hęnnar ǫllum \alst{s}efa.\eva

\bvb This I know sixteenth, if I will from the smart girl \\
\ind have her senses all, and pleasure: \\
the heart I change of the white-armed woman, \\
\ind and I twist her whole mind.\evb\evg


\bvg\bva Þat kann’k \alst{s}jautjánda \hld\ at mik \alst{s}ęint mun firrask &
\ind hit \alst{m}an-unga \alst{m}an.\eva

\bvb This I know seventeenth, that the girl-young girl \\
\ind will lately shun me.\evb\evg


\bvg\bva\alst{L}jóða þessa \hld\ munt \alst{L}oddfáfnir &
\ind lęngi \alst{v}anr \alst{v}esa; &
\ind þó sé þér \alst{g}óð ef \alst{g}etr, &
\ind \alst{n}ýt ef \alst{n}emr, &
\ind \alst{þ}ǫrf ef \alst{þ}iggr.\eva

\bvb These leeds wilt thou, Loddfathomer, \\
\ind for long be lacking! \\
\ind Though they would be good for thee if thou get, \\
\ind useful if thou learn, \\
\ind needful if thou receive.\evb\evg


\bvg\bva Þat kann’k \alst{á}tjánda, \hld\ es \alst{ę́}va kęnni’k &
\ind \alst{m}ęy né \alst{m}anns konu, &
—\alst{a}llt es bętra \hld\ es \alst{ęi}nn of kann, &
\ind þat fylgir \alst{l}jóða \alst{l}okum— &
nema þęiri \alst{ęi}nni, \hld\ es \edtrans{mik \alst{a}rmi vęrr}{holds me in her arm}{\Bfootnote{A similar expression is also used \Volundarkvida\ 2.  The one who wraps Weden in her arm may be his wife, \inx[P]{Frie}.}}, &
\ind eða mín \alst{s}ystir \alst{s}éi.\eva

\bvb This I know eighteenth, which I will never teach \\
\ind a maiden nor man’s woman, \\
(everything is better which one alone knows; \\
\ind that follows the last of the leeds,) \\
save for her alone who holds me in her arm, \\
\ind or is my sister.\evb\evg

\sectionline

\bvg\bva Nú eru \alst{H}áva mǫ́l kveðin \hld\ \alst{H}áva \alst{h}ǫllu í; &
\ind \alst{a}ll-þǫrf \alst{ý}ta sonum, &
\ind \alst{ȯ}-þǫrf \edtrans{\alst{jǫ}tna}{ettins}{\Afootnote{corrected in margin from \emph{ýta} ‘men’ \Regius}} sonum; &
hęill sá’s \edtext{\alst{k}vað, \hld\ hęill sá’s \alst{k}ann, &
\ind \alst{n}jóti sá’s \alst{n}am, &
\ind \alst{h}ęilir þęir’s \alst{h}lýddu}{\lemma{kvað, kann, nam, hlýddu ‘sang, knows, learned, heeded’}\Bfootnote{The implied subject is the speeches, i.e. ‘hail he who sang them, hail he who knows them,’ et.c.}}.\eva

\bvb Now are the High One’s speeches sung in the High One’s hall, \\
\ind of great use for the sons of men, \\
\ind of harm for the sons of ettins. \\
Hail he who sang; hail he who knows; \\
\ind may he use who learned; \\
\ind hail they who heeded!\evb\evg

\sectionline
