\bookStart{Leeds of Hindle}[Hyndluljóð]
\setBookCode{Hyndluljod}

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: late C11th (0.996)

\textbf{Meter:} \Fornyrdislag%
\end{flushright}%

\section{Introduction}

The \textbf{Leeds of Hindle} (\Hyndluljod) is a poorly preserved poem found only in \FlatMS.

\section{The Leeds of Hindle}

\bvg\bva%
„Vaki \alst{m}ę́r \alst{m}ęyja, \hld\ vaki \alst{m}ín vina, &
\alst{H}yndla systir, \hld\ es í \alst{h}ęlli býr; &
nú ’s \alst{r}økr \alst{r}økra, \hld\ \alst{r}íða vit skulum &
til \alst{V}al-hallar \hld\ ok til \alst{v}és hęilags.\eva

\bvb\speakernoteb{[Frow quoth:]}%
“Wake, maiden of maidens! Wake, my friend! \\
O Hindle, sister, who livest in the cave! \\
Now’s the twilight of twilights; we two shall ride \\
to Walhall and to the holy \inx[C]{wigh}!\evb\evg


\bvg\bva%
Biðjum \alst{H}ęrja-fǫðr \hld\ í \alst{h}ugum sitja, &
\edtrans{hann \alst{g}eldr ok \alst{g}efr \hld\ \alst{g}ull verðugum}{he pays and gives gold to the worthy}{\Bfootnote{Closely related to \textlink{HelgakvidaOne} 9/3, which is why \textcite{FinnurEdda}, \textcite{GudniEdda} emend \emph{verðugum} ‘the worthy’ to \emph{verðungu} ‘the retinue’.}}, &
gaf hann \alst{H}ęrmóði \hld\ \alst{h}jalm ok brynju, &
en \alst{S}igmundi \hld\ \alst{s}verð at þiggja.\eva

\bvb Let us bid the Father of Hosts \name{= Weden} to remain in good spirits;  \\
he pays and gives gold to the worthy.  \\
He gave \inx[P]{Harmod} helmet and byrnie, \\
and \inx[P]{Syemund} a sword to receive.\evb\evg


\bvg\bva%
Gefr hann \alst{s}igr \alst{s}onum, \hld\ en \alst{s}vinnum \edtrans{aura}{ounces}{\Bfootnote{Of silver.}}, &
\alst{m}ę́lsku \alst{m}ǫrgum \hld\ ok \alst{m}an-vit firum, &
\alst{b}yri gefr \alst{b}rǫgnum, \hld\ en \alst{b}rag skǫldum, &
gefr hann \alst{m}ann-sęmi \hld\ \alst{m}ǫrgum rekki.\eva

\bvb He gives victory to sons and ounces to the wise, \\
speech to many and \inx[C]{manwit} to men. \\
Fair wind he gives to nobles and praise-song to \inx[C]{scald}[scalds]; \\
he gives manly valour to many a champion.\evb\evg


\bvg\bva%
\alst{Þ}ór mun’k blóta, \hld\ \alst{þ}ess mun’k biðja, &
at hann \alst{ę́} við þik \hld\ \alst{ęi}n-art láti; &
þó ’s hǫ́num \alst{ȯ}-títt \hld\ við \alst{jǫ}tuns brúðir.\eva

\bvb To Thunder I will \inx[C]{bloot}; of this I will bid, \\
that he always be upright with thee \\
even though he hates the ettin’s brides.\evb\evg


\bvg\bva%
Nú tak-tu \alst{u}lf þinn \hld\ \alst{ęi}nn af stalli, &
lát hann \alst{r}inna \hld\ með \alst{r}una mínum.“— &
„Sęinn es \alst{g}ǫltr þinn \hld\ \alst{g}oð-veg troða, &
vil’k-at \alst{m}ar \alst{m}ínn \hld\ \alst{m}ę́tan hlǿða.\eva

\bvb Now take thy one wolf from the stable; \\
let him run alongside my boar.”— \\
\speakernoteb{[Hindle quoth:]}%
“Slow is thy boar to tread the Godways; \\
I wish not to load my noble steed.\evb\evg


\bvg\bva%
\alst{F}lǫ́ ert \alst{F}ręyja, \hld\ es \alst{f}ręistar mín, &
\edtext{vísar þú \alst{au}gum \hld\ á \alst{o}ss þannig, &
es hafir ver þinn \hld\ í val-sinni}{\lemma{vísar \dots\ val-sinni ‘thou showest \dots\ slain-ways’}\Bfootnote{i.e., “You only show favour to me because you want me to help your lover”.  For the expression cf. \textlink{Sigrdrifumal} 3/3 and note.}} &
Óttar unga \hld\ Innstęins bur.“\eva

\bvb False art thou, Frow, who temptest me; \\
thou showest thy eyes on us this way \\
since thou hast thy lover on the slain-path: \\
the young Oughter, Instone's offspring.”\evb\evg


\bvg\bva%
„\alst{D}ulið est Hyndla, \hld\ \alst{d}raums ę́tla’k þér, &
es kveðr \alst{v}er minn \hld\ í \alst{v}al-sinni.\eva

\bvb\speakernoteb{[Frow quoth:]}%
Deluded art thou, Hindle; I think thee dreamy \\
as thou sayest that my man is on the slain-path.\evb\evg


\bvg\bva%
Þar’s \alst{g}ǫltr \alst{g}lóar \hld\ \alst{G}ullinbursti, &
\edtrans{\alst{H}ildisvíni}{Hildswine}{\Bfootnote{The ‘battle-swine’, presumably an alternative name of Goldenbristle.}}, \hld\ es mér \alst{h}agir gęrðu, &
\alst{d}vergar tvęir \hld\ \alst{D}áinn ok Nabbi.\eva

\bvb There where the boar Goldenbristle glows, \\
the Hildswine, which for me made \\
the two skilful dwarfs Dowen and Nab.\evb\evg


\bvg\bva%
\alst{S}ęnn í \alst{s}ǫðlum \hld\ \alst{s}itja vit skulum &
ok of \alst{jǫ}fra \hld\ \alst{ę́}ttir dǿma, &
\alst{g}umna þęira, \hld\ es frá \alst{g}oðum kómu.\eva

\bvb Soon in the saddles we two shall sit, \\
and of rulers’ lineages speak, \\
of those men who came from the gods.\evb\evg


\bvg\bva%
Þęir hafa \alst{v}ęðjat \hld\ \alst{v}ala malmi &
\alst{Ó}ttarr \alst{u}ngi \hld\ ok \alst{A}ngantýr; &
\alst{sk}ylt ’s at vęita, \hld\ svá’t \alst{sk}ati hinn ungi &
\alst{f}ǫður-lęifð hafi \hld\ ępt \alst{f}rę́ndr sína.\eva

\bvb They have wagered the Welsh ore \ken{gold}, \\
young Oughter and Ongenthew— \\
it must be granted so that the young prince \\
may have the patrimony of his kinsmen.\evb\evg


\bvg\bva%
\alst{H}ǫrg hann mér gęrði \hld\ \alst{h}laðinn stęinum; &
nú ’s \alst{g}rjót þat \hld\ at \alst{g}lęri orðit; &
rauð hann í \alst{n}ýju \hld\ \alst{n}auta blóði; &
\alst{ę́} trúði \alst{Ó}ttarr \hld\ á \alst{ǫ́}synjur.\eva

\bvb A \inx[C]{harrow} he made me, loaded with stones; \\
now that stone-pile has turned into glass. \\
He reddened it in the fresh blood of oxen; \\
always did Oughter trust on the \inx[G]{Ossens}.\evb\evg


\bvg\bva%
\alst{N}ú lát forna \hld\ \alst{n}iðja talða &
ok \alst{u}pp-bornar \hld\ \alst{ę́}ttir manna &
hvat ’s Skjǫldunga, \hld\ hvat ’s Skilfinga, &
hvat ’s \alst{Ǫ}ðlinga \hld\ hvat ’s \alst{Y}lfinga & &
hvat ’s \alst{h}ǫld-borit, \hld\ hvat ’s \alst{h}ęrs-borit &
\alst{m}ęst \alst{m}anna val \hld\ \alst{u}nd Mið-garði?“\eva

\bvb Now let ancient kinsmen be counted, \\
and the high born lineages of men: \\
What’s of Shieldings? What’s of Shilvings? \\
What’s of Athlings? What’s of Wolvings? \\
What’s born of hero? What’s born of chief, \\
the greatest choice of men within Middenyard?”\evb\evg


\bvg\bva%
„Þú ert \alst{Ó}ttarr \hld\ borinn \alst{I}nnstęini, &
en \alst{I}nnstęinn vas \hld\ \alst{A}lfi inum gamla, &
\alst{A}lfr vas \alst{U}lfi, \hld\ \alst{U}lfr Sę́fara, &
en \alst{S}ę́fari \hld\ \alst{S}van inum rauða.\eva

\bvb\speakernoteb{[Hindle quoth:]}%
“Thou\footnoteB{Hindle, maybe in a trance-like state, speaks straight to Oughter.} art, Oughter, born to Instone, \\
and Instone was born to Elf the old, \\
Elf was to Wolf, Wolf to Seafarer, \\
and Seafarer to Swan the red.\evb\evg


\bvg\bva%
\alst{M}óður átti faðir þinn \hld\ \alst{m}ęnjum gǫfga, &
\alst{h}ygg at \alst{h}éti \hld\ \alst{H}lédís gyðja, &
\alst{F}róði vas \alst{f}aðir þęirar, \hld\ en \edtext{\alst{F}rí\emph{und}}{\Afootnote{emend. from meaningless \emph{†friaut†} \FlatMS}} móðir; &
\alst{ǫ}ll þótti \alst{ę́}tt sú \hld\ með \alst{y}fir-mǫnnum.\eva

\bvb Thy father won thy esteemed mother with torcs, \\
I think that she was called Leedise the \inx[C]{gidden}. \\
Frood was her father and Friend her mother; \\
all that lineage seemed to be among \inx[C]{overmen}.\evb\evg


\bvg\bva%
\alst{Au}ði vas \alst{á}ðr \hld\ \alst{ǫ}flgastr manna, &
\alst{H}alfdanr fyrri \hld\ \alst{h}ę́str Skjǫldunga, &
\alst{f}rę́g vǫ́ru \alst{f}olk-víg, \hld\ þau’s \alst{f}ramir gęrðu, &
\alst{h}varfla þóttu verk \hld\ með \alst{h}imins skautum.\eva

\bvb Ead was once the strongest of men, \\
Halfdane earlier the highest of Shieldings. \\
Famous were the troop-wars which the brave ones made; \\
his \name{= Halfdane’s} works seemed to whirl along the corners of heaven.\evb\evg


\bvg\bva%
\edtrans{\alst{Ę}flðisk}{became the in-law}{\Bfootnote{Lit. “was strengthened by”.  Elmwey was Eanmund’s daughter or sister.}} við \alst{Ęy}mund \hld\ \alst{ǿ}ðstan manna &
en vá \alst{S}igtrygg \hld\ með \alst{s}vǫlum ęggjum, &
\alst{ęi}ga gekk \alst{A}lmvęig, \hld\ \alst{ǿ}ðsta kvinna, &
\alst{ó}lu þau ok \alst{ǫ́}ttu \hld\ \alst{á}tján sonu.\eva

\bvb He \name{= Halfdane} became the in-law of Eanmund, the noblest of men, \\
but he slew Syetrue with cool edges. \\
He went to have Elmwey, the noblest of women; \\
they begot and had eighteen sons.\evb\evg


\bvg\bva%
Þaðan eru \alst{S}kjǫldungar, \hld\ þaðan eru \alst{S}kilfingar, &
þaðan eru \alst{Ǫ}ðlingar, \hld\ þaðan eru \alst{Y}nglingar, &
þaðan es \alst{h}ǫld-borit, \hld\ þaðan es \alst{h}ęrs-borit, &
\alst{m}est \alst{m}anna val \hld\ und \alst{M}ið-garði; &
allt ’s þat ę́tt þín, \hld\ Óttarr hęimski.\eva

\bvb Thence come Shieldings! Thence come Shilvings! \\
Thence come Athlings! Thence come Inglings!\footnote{Note the contradiction with v. 12. Since the Inglings have already been mentioned (under the name Shilvings, for the difference between the two see Index), it seems likely that Wolvings is the original reading.} \\
Thence is born of hero! Thence is born of chief \\
the greatest choice of men within Middenyard! \\
This is all thy lineage, O foolish Oughter!”\evb\evg


\bvg\bva%
Vas Hildigunnr \hld\ hęnnar móðir, &
Svǫ́fu barn \hld\ ok Sę́-konungs; &
alt ’s þat ę́tt þín, \hld\ Óttarr hęimski. &
varði at viti svá, \hld\ viltu ęnn lęngra?\eva

\bvb Hildguth was her mother, \\
the child of Sweve and Sea-king. \\
This is all thy lineage, O foolish Oughter!— \\
It is meaningful that one might know thus; wilt thou yet further?\evb\evg


\bvg\bva%
Dagr átti Þóru \hld\ dręngja móður, &
ólusk í ę́tt þar \hld\ ǿðstir kappar, &
Fraðmarr ok Gyrðr \hld\ ok Frekar báðir, &
Ámr ok Jǫsurmarr, \hld\ Alfr hinn gamli. &
varðar at viti svá, \hld\ viltu ęnn lęngra?\eva

\bvb Day had Thure, the mother of valiant men; \\
in that lineage were begotten the noblest champions: \\
Fradmer and Yird, and both Frekes; \\
Ame and Essirmer; Elf the old.— \\
It is meaningful that one might know thus; wilt thou yet further?\evb\evg


\bvg\bva%
Kętill hét vinr þęira \hld\ Klypps arf-þęgi, &
vas hann móður-faðir \hld\ móður þinnar; &
þar vas Fróði \hld\ fyrr ęnn Kári, &
en Hildi vas \hld\ Hóalfr of getinn.\eva

\bvb Kettle was their friend, the heir of Clip; \\
he was the father of thy mother's mother. \\
There was Frood, yet earlier Keer, \\
but by Hild was Highelf begotten.\evb\evg

... %TODO More dialogue

\sectionline
