\bookStart{Fragments from the Saw of the Walsings}

\section{Introduction}

In \Regius, \Sigrdrifumal\ ends abruptly at stanza 27, after which a number of pages have gone missing; the so-called “great lacuna”.  The poetry contained in them undoubtedly belonged to the Walsing cycle, specifically concerning the life of Siward.

The author of \VolsungaSaga\ drew heavily from a collection of Walsing-cycle poetry closely related to \Regius.  He quotes many stanzas known from \Regius, but also some which do not survive anywhere else—these are the stanzas edited here.  They correspond to the story which would have been found in the great lacuna, and it is probable that they derive from the now-lost poems found there.

\sectionline

\bvg\bva%
Ristu af \alst{m}agni \hld\ \alst{m}ikla hellu, &
\alst{S}igmundr hjǫrvi \hld\ ok \alst{S}infjǫtli.\eva

\bvb They carved with strength the great stone, \\
\inx[P]{Syemund} with sword, and \inx[P]{Sinfittle}.\evb\evg

\sectionline

\bvg\bva%
\alst{Ę}ldr nam at \alst{ǿ}sask \hld\ en \alst{jǫ}rð at skjalfa &
ok \alst{h}ár logi \hld\ við \alst{h}imni gnę́fa; &
\alst{f}ár tręystisk þar \hld\ \alst{f}ylkis rekka &
\alst{ę}ld at ríða \hld\ né \alst{y}fir stíga.\eva

\bvb Fire took to rage and earth to shake \\
and high flame to rise against heaven. \\
Few there dared of the marshall’s champions \\
the fire to ride or to step over.\evb\evg


\bvg\bva%
\alst{S}igurðr Grana \hld\ \alst{s}verði kęyrði; &
\alst{ę}ldr sloknaði \hld\ fyr \alst{ǫ}ðlingi; &
\alst{l}ogi allr \alst{l}ę́gðisk \hld\ fyr \alst{l}of-gjǫrnum; &
bliku \alst{r}ęiði, \hld\ es \alst{R}eginn átti.\eva

\bvb Siward drove Grane on by sword; \\
the fire went out before the athling; \\
the flame all lowered before the praise-eager man; \\
the harness flashed which Rein had owned.\evb\evg

\sectionline

\bvg\bva%
\alst{S}igurðr vá at ormi, \hld\ en þat \alst{s}íðan mun &
\alst{ø}ngum fyrnask, \hld\ meðan \alst{ǫ}ld lifir. &
En \alst{h}lýri þinn \hld\ \alst{h}várki þorði &
\alst{ę}ld at ríða \hld\ né \alst{y}fir stíga.\eva

\bvb Siward smote the Wyrm, and that will afterwards \\
by none be forgotten while mankind lives, \\
but thy brother dared not either \\
the fire to ride or to step over.\evb\evg

\sectionline

\bvg\bva%
\alst{Ú}t gekk Sigurðr \hld\ \alst{a}nn-spjalli frá, &
\alst{h}oll-vinr lofða, \hld\ ok \alst{h}nípaði, &
svá at \alst{g}anga nam \hld\ \alst{g}unnar-fúsum &
\alst{s}undr of \alst{s}íður \hld\ \alst{s}erkr járn-ofinn.\eva

\bvb TODO: translation.\evb\evg


TODO: More stanzas?


\sectionline
