\bookStart{The Lay of Thrim}[Þrymskviða]


\bvg
\bva \edtext{\alst{V}ręiðr}{\lemma{Vręiðr}\Afootnote{TODO: Note about ambiguity of alliteration.}} vas þá Ving-Þórr \hld\ es hann vaknaði &
ok síns hamars \hld\ of saknaði, &
skegg nam at hrista, \hld\ skǫr nam at dýja, &
réð Jarðar burr \hld\ umb at þreifask.\eva

\bvb Wroth was then Wing-Thunder when he woke, and of his hammer was bereaved. His beard he took to shake, his locks he took to pull; resolved the son of Earth to look about.\evb
\evg


\bvg
\bva Ok hann þat orða \hld\ allz fyrst of kvað: &
“Hęyrðu nú, Loki, \hld\ hvat ek nú mę́li &
es ęigi vęit \hld\ jarðar hvęrgi &
né upphimins: \hld\ áss es stolinn hamri!”\eva

\bvb And he that word, first of all did speak: “Hear thou now, Lock, what I now speak, which nowhere is known, not on earth nor \inx{up-heaven}{G}: the \inx{os}[G] \ken*{= Thunder = I} has been robbed of his hammer!”\evb
\evg


\bvg
\bva Gengu þęir fagra \hld\ Fręyju túna &
ok hann þat orða \hld\ allz fyrst of kvað: &
“Muntu mér, Fręyja, \hld\ fjaðrhams ljá &
ef ek mínn hamar \hld\ mę́tta’k hitta?”\eva

\bvb Went they to the fair yards of Frow, and he that word, first of all did speak: “Wilt thou me, Frow, the \inx{feather-hame}{P} lend, if I my hammer might find?”\evb
\evg


\bvg {\small [Fręyja kvað:]}
\bva “Þó mynda’k gefa þér \hld\ þótt ór gulli vę́ri &
ok þó sęlja \hld\ at vę́ri ór silfri.” &
Fló þá Loki, \hld\ fjaðrhamr dunði, &
unz fyr útan kom \hld\ ása garða &
ok fyr innan kom \hld\ jǫtna hęima.\eva

\bvb Frow quoth: “I would yet give it to thee, though it were out of gold, and yet offer\footnoteB{\emph{sęlja} ‘sell’ here has its earlier meaning, cf. Gothic \emph{saljan} ‘\emph{opfern}; θύειν’ (Streitberg 1910:116).} it to thee, if it were out of silver.”\footnoteB{Regaining the hammer is of such importance to the gods (cf. v. 17; without it the Ease stand powerless against the \inx{Ettins}{T}), that Frow would lend the feather-hame to the greedy and unreliable Lock, even if it were made out of solid gold or silver.} Flew then Lock\footnoteB{Though Thunder is the one asking for the hame (“\emph{my} hammer”), Lock is the one that takes off flying.}—the feather-hame rustled—until outside he came of the yards of the Ease, and inside he came of the homes of the Ettins.
\evg

\bvg
\bva Þrymr sat á haugi, \hld\ þursa dróttinn, &
gręyjum sínum \hld\ gullbǫnd snøri &
ok mǫrum sínum \hld\ mǫn jafnaði.\eva

\bvb Thrim sat on the mound, the lord of thurses: on his greyhounds the golden leashes he twisted, and on his mares the manes he cut even.\evb
\evg


\bvg {\small [Þrymr kvað ok Loki:]}
\bva »Hvat es með ǫ́sum? \hld\ Hvat es með ǫlfum? &
Hví estu ęinn kominn \hld\ í jǫtunhęima?» &
»Illt es með ǫ́sum, \hld\ [illt es með ǫlfum!]\footnoteB{Second half-line inserted in analogy with the first pair, regardless it is needed for metrical reasons.} &
Hęfir þú Hlórriða \hld\ hamar of folginn?»\eva

\bvb [Thrim quoth:] “What is with the Ease? What is with the elves? Why art thou alone come into the Ettin-homes?” [Lock quoth:] “’Tis ill with the Ease, ’tis ill with the elves! Hast thou the hammer of Loride hidden?”\evb
\evg


\bvg {\small [Þrymr kvað ok Loki:]}
\bva Ek hęfi Hlórriða \hld\ hamar of folginn &
átta rǫstum \hld\ fyr jǫrð neðan; &
hann ęngi maðr \hld\ aptr of hęimtir &
nęma fǿri mér \hld\ Fręyju at kvę́n.\eva

\bvb [Thrim quoth:] “I have the hammer of Loride hidden, eight \inx{rests} beneath the earth; it no man will fetch again, lest he bring me Frow as wife.”\evb
\evg


\bvg
\bva Fló þá Loki, \hld\ fjaðrhamr dunði, &
unz fyr útan kom \hld\ jǫtna hęima &
ok fyr innan kom \hld\ ása garða; &
mǿtti hann Þór \hld\ miðra garða &
ok þat hann orða \hld\ allz fyrst of kvað: &
Hęfir þú ørendi \hld\ sem ęrfiði?\eva

\bvb Flew then Lock—the feather-hame rustled—until outside he came of the homes of the Ettins, and inside he came of the yards of the Ease. He met Thunder in the middle of the yards, and he that word, first of all did say: “Hast thou an errand, as hardship\footnoteB{A difficult phrase to translate; Thunder asks Lock if he has bad news.}?\evb
\evg


\bvg
\bva Segðu á lopti \hld\ lǫng tíðendi! &
Opt sitjanda \hld\ sǫgur of fallask &
ok liggjandi \hld\ lygi of bęllir.\eva

\bvb Say thou aloft, the long tidings! Often [when] sitting, tales fail each other, and [when] lying down, lies are dealt.”\footnoteB{Proverbial, meaning being that if one sits down and thinks too much over bad news, things will be left out, excuses thought up. Thus it is best that Lock immediately tell Thunder what he has learned.}\evb
\evg


\bvg
\bva Hefi ek ørindi \hld\ erfiði ok: &
Þrymr hęfir þinn hamar, \hld\ þursa dróttinn; &
hann ęngi maðr \hld\ aptr of hęimtir &
nęma hǫ́num fǿri \hld\ Fręyju at kvę́n.\eva

\bvb [Lock quoth:] “I have an errand, hardship also: Thrim has thy hammer, the lord of thurses; it no man will fetch again, lest he bring him Frow as wife.”\evb
\evg


\bvg
\bva Ganga þęir fagra \hld\ Fręyju at hitta &
ok hann þat orða \hld\ allz fyrst of kvað: &
Bittu þik, Fręyja, \hld\ brúðar líni! &
Vit skulum aka tvau \hld\ í jǫtunhęima.\eva

\bvb Go they the fair Frow to find, and he that word, first of all did say: “Bind thee, Frow, with a bride’s linen\footnoteB{A linen band tied around the bride’s head. TODO: Reference this note.}! We two shall drive into the Ettin-homes.”\evb
\evg


\bvg
\bva Vręið varð þá Fręyja \hld\ ok fnasaði, &
allr ása salr \hld\ undir bifðisk, &
stǫkk þat it mikla \hld\ męn Brísinga: &
Mik vęiztu verða \hld\ vergjarnasta &
ef ek ęk með þér \hld\ í jǫtunhęima.\eva

Wroth became then Frow, and snorted—the whole hall of the Ease trembled below—threw she off the great necklace of the Brisings: “Thou knowest that I will become the most man-eager, if I drive with thee into the Ettin-homes.”
\evg


\bvg
\bva Senn vǫ́ru ę́sir \hld\ allir á þingi &
ok ásynjur \hld\ allar á máli, &
ok of þat réðu \hld\ ríkir tívar &
hvé þęir Hlórriða \hld\ hamar of sǿtti.\eva

Soon the Ease were all at the Thing, and the Osennies all at speech, and of that decided the rich Tues: how they the hammer of Loride would regain.\footnoteB{\emph{sǿkja} ‘attack, seek’; they decide how to handle the hammer issue.}
\evg


\bvg
\bva Þá kvað þat Heimdallr, \hld\ hvítastr ása, &
vissi hann vel fram \hld\ sęm vanir aðrir: &
“Bindu vér Þór þá \hld\ brúðar líni; &
hafi hann it mikla \hld\ męn Brísinga!\eva

Then quoth that Homedall, the whitest of the Ease — he knew well the future, like the other Wanes —: “Let us bind Thunder with the bride’s linen; he may have the great necklace of the Brisings.
\evg


\bvg
\bva Lǫ́tum und hǫ́num \hld\ hrynja lukla &
ok kvenváðir \hld\ umb kné falla &
en á brjósti \hld\ bręiða stęina &
ok hagliga \hld\ umb hǫfuð typpum!”\eva

Let us have by him jingling keys\footnoteB{As the everyday manager of the household, keys were the mark of the married housewife.}, and women’s garments falling down about his knees, but on the breast broad stones\footnoteB{Large jewels.}, and skillfully let us tip his head!\footnoteB{With a (pointed) hood (TODO: literature).}”
\evg


\bvg
\bva Þá kvað þat Þórr, \hld\ þrúðugr áss: &
“Mik munu ę́sir \hld\ argan kalla &
ef ek bindask lę́t \hld\ brúðar líni!”\eva

Then quoth that Thunder, the mighty Os: “\emph{Me}, the Ease would call an arrow† man, if I let myself be bound with bride’s linen!”
\evg


\bvg
\bva Þá kvað þat Loki \hld\ Laufęyjar sonr: &
“Þęgi þú, Þórr, \hld\ þęira orða! &
Þegar munu jǫtnar \hld\ Ásgarð búa &
nęma þú þinn hamar \hld\ þér of hęimtir.”\eva

Then quoth that Lock, the son of Leafie: “Be silent thou, Thunder, of those words! Shortly the ettins will Osyard inhabit, unless thou thy hammer for thyself dost fetch!”
\evg


\bvg
\bva Bundu þęir Þór þá \hld\ brúðar líni &
ok inu mikla \hld\ męni Brísinga, &
létu und hǫ́num \hld\ hrynja lukla &
ok kvenváðir \hld\ umb kné falla &
ęn á brjósti \hld\ bręiða stęina &
ok hagliga \hld\ of hǫfuð typpðu.\eva

Bound they Thunder then, with bride’s linen, and with the great necklace of the Brisings. They had by him keys jingling, and women’s garments falling down to his knees, but on the breast broad stones, and skillfully they tipped his head.
\evg


\bvg
\bva Þá kvað þat Loki \hld\ Laufęyjar sonr: &
“Mun ek ok með þér \hld\ ambǫ́tt vesa, &
vit skulum aka tvau \hld\ í jǫtunhęima!”\eva

Then quoth that Lock, the son of Leafie: “I will also with thee be a handmaid; we two\footnoteB{The form used, \emph{tvau}, is the neuter plural, ie. one of the pair is female and the other male. This is either an error due to mindless copying of v. 11, or a backhanded insult against Thunder.} shall drive into the Ettin-homes.”
\evg


\bvg
\bva Senn vǫ́ru hafrar \hld\ hęim of vreknir, &
skyndir at skǫklum, \hld\ skyldu vel renna; &
bjǫrg brotnuðu, \hld\ brann jǫrð loga, &
ók Óðins sonr \hld\ í jǫtunhęima.\eva

Soon he-goats were driven home, hasted onto the cart-poles; they were to run well. Crags burst, the earth burned with flame; the son of Weden drove into the Ettin-homes.\footnoteB{For a similar account of Thunder driving, see \emph{Haustlǫng}, v. 14-16 in \emph{Skp} 3 (pp. 453-456).}
\evg


\bvg
\bva Þá kvað þat Þrymr, \hld\ þursa dróttinn: &
“Standið upp, jǫtnar, \hld\ ok stráið bękki! &
Nú fǿrið mér \hld\ Fręyju at kván, &
Njarðar dóttur \hld\ ór Nóatúnum.”\eva

Then quoth that Thrim, the lord of thurses: “Stand ye up, ettins, and strew ye the benches! Now bring ye me Frow as wife, the daughter of Nearth of the Nowetowns.
\evg


\bvg
\bva Ganga hér at garði \hld\ gullhyrnðar kýr, &
øxn alsvartir, \hld\ jǫtni at gamni, &
fjǫlð á’k męiðma, \hld\ fjǫlð á’k męnja; &
ęinnar mér Fręyju \hld\ ávant þykkir.\eva

Here march to the estate golden-horned cows, all-black oxen, to the enjoyment of the ettin; a multitude I own of treasures, a multitude I own of necklaces; of Frow alone methinks is missing.”
\evg


\bvg
\bva Vas þar at kveldi \hld\ of komit snimma &
ok fyr jǫtna \hld\ ǫl framm borit. &
Ęinn át oxa, \hld\ átta laxa, &
krásir allar, \hld\ þę́r’s konur skyldu, &
drakk Sifjar verr \hld\ sáld þrjú mjaðar.\eva

There was the evening come quickly, and for the ettins ale brought forth. [Thunder] ate one ox, eight salmons, and all the delicacies, which the women were supposed to; drank the husband of Sib \ken{Thunder}[1] three sieves of mead.\footnoteB{Compare \Hymiskvida\ 15 for a strikingly similar description of Thunder’s great eating.}
\evg


\bvg
\bva Þá kvað þat Þrymr, \hld\ þursa dróttinn: &
“Hvar sáttu brúðir \hld\ bíta hvassara? &
Sá’k-a brúðir \hld\ bíta ęnn bręiðara &
né ęnn męira mjǫð \hld\ męy of drekka!”\eva

Then quoth that Thrim, the lord of thurses: “Where sawest thou brides bite sharper? Saw I never brides bite yet broader, nor yet more mead a maiden drink.”
\evg


\bvg
\bva Sat in alsnotra \hld\ ambǫ́tt fyr &
es orð of fann \hld\ við jǫtuns máli: &
“Át vę́tr Fręyja \hld\ átta nóttum, &
svá vas hón óðfús \hld\ í jǫtunhęima.”\eva

Sat the allclever maid-servant\footnoteB{Lock in disguise.} in front, when she a word did find against the speech of the ettin: “\emph{Ate} Frow naught, for eight nights; so madly was she longing for the Ettin-homes.”
\evg


\bvg
\bva Laut und línu, \hld\ lysti at kyssa, &
ęn hann útan stǫkk \hld\ ęndlangan sal: &
“Hví eru ǫndótt \hld\ augu Fręyju? &
Þykki mér ór \hld\ augum brenna!”\eva

He looked ’neath the linen, he lusted for a kiss, but he from the outside leapt back, across the length of the hall: “Why are the eyes of Frow fiery? Methinks there is fire coming out of the eyes!\footnoteB{Lit. “Methinks out of the eyes burn.”}”
\evg


\bvg
\bva Sat in alsnotra \hld\ ambǫ́tt \edtext{fyrir}{\Afootnote{‘ſ.’ \emph{add.} \Regius \emph{possibly representing a lost word}}} &
es orð of fann \hld\ við jǫtuns máli: &
“Svaf vę́tr Fręyja \hld\ átta nóttum, &
svá vas hón óðfús \hld\ í jǫtunhęima.”\eva

Sat the allclever maid-servant in front, when she a word did find against the speech of the ettin: “\emph{Slept} Frow naught, for eight nights; so madly was she longing for the Ettin-homes.”
\evg


\bvg
\bva Inn kom in arma \hld\ jǫtna systir, &
hin es brúðfjár \hld\ biðja þorði: &
“Láttu þér af hǫndum \hld\ hringa rauða &
ef þú ǫðlask vill \hld\ ástir mínar, &
ástir mínar, \hld\ alla hylli!”\eva

In came the wretched sister of the ettins, the one who for the bride-price had dared ask: “Take off from thy hands the red rings, if thou wilt win my loves; my loves, [and] all favour.”\footnoteB{The sister, who already asked for the hammer, now has the audacity to ask “Frow” (Thunder) to take the rings off her hands and give them to her.}
\evg


\bvg
\bva Þá kvað þat Þrymr, \hld\ þursa dróttinn: &
“Berið inn hamar \hld\ brúði at vígja, &
leggið Mjǫllni \hld\ í męyjar kné, &
vígið okkr saman \hld\ Várar hęndi!”\eva

Then quoth that Thrim, the lord of thurses: “Bear ye in the hammer, the bride to bless; lay ye Millner in the maiden’s knee, bless ye us two together by the hand of Ware!\footnoteB{A minor goddess presumably presiding over marriage.}”
\evg


\bvg
\bva Hló Hlórriða \hld\ hugr í brjósti &
es harðhugaðr \hld\ hamar of þękkði; &
Þrym drap hann fyrstan, \hld\ þursa dróttin, &
ok ę́tt jǫtuns \hld\ alla lamði.\eva

The heart of Loride laughed in his breast, when, hard-hearted, he recognized the hammer. Thrim he slew first, the lord of thurses, and all the aught† of the ettin he thrashed.
\evg


\bvg
\bva Drap hann ina ǫldnu \hld\ jǫtna systur, &
hin es brúðfjár \hld\ of beðit hafði; &
hón skell of hlaut \hld\ fyr skillinga &
en hǫgg hamars \hld\ fyr hringa fjǫlð.\eva

He slew the old sister of the ettins, the one who for the bride-price had asked; she received a smiting before shillings, and a strike of the hammer before a multitude of rings.
\evg


\bvg
\bva Svá kom Óðins sonr \hld\ ęndr at hamri.\eva

Thus Weden’s son regained his hammer.
\evg
