\book{The Lay of Ettle. (Atlakviða)}\bookStart

\begin{verse}
\bva Hjarta skal mér Hǫgna \hld í hęndi liggja  \\%M
blóðugt, ór brjósti \hld skorit baldriða, \\%M
saxi slíðrbęitu, \hld syni þjóðans.\\%E
\end{verse}

\bvb (Guthhere quoth:) \\ “The heart of Hayne shall in my hands lie, bloody out of the breast, cut from the bold rider; with a razor-sharp seax, from the son of the sovereign.”

\begin{verse}
\bva Skǫ́ru þęir hjarta \hld Hjalla ór brjósti \\%M
blóðugt ok á bjóð lǫgðu \hld ok bǫ́ru þat fyr Gunnar.\\%E
\end{verse}

\bvb They cut the heart of Helle out of the breast; bloody on a platter they laid it, and carried it before Guthhere.

\begin{verse}
\bva Þá kvað þat Gunnarr, \hld gumna dróttinn: \\%M
Hér hęfi ek hjarta \hld Hjalla ins blauða, \\%M
ólíkt hjarta \hld Hǫgna ins frǿkna, \\%M
es mjǫk bifask \hld es á bjóði liggr; \\%M
bifðisk hǫlfu męirr \hld es í brjósti lá!\\%E
\end{verse}

\bvb Then Guthhere quoth this, the lord of men: “Here have I the heart of Helle the soft—unlike the heart of Hayne the bold!—which much trembles when on the platter it lies; it trembled twice as much when in the breast it lay.”

\begin{verse}
\bva Hló þá Hǫgni \hld es til hjarta skǫ́ru \\%M
kvikvan kumblasmið \hld kløkkva hann sízt hugði \\%M
blóðugt þat á bjóð lǫgðu \hld ok bǫ́ru fyr Gunnar.\\%E
\end{verse}

\bvb Hayne laughed then, when to the heart they cut; the living landmark-smith thought least of sobbing. Bloody on a platter they laid it, and carried it before Guthhere.

\begin{verse}
\bva Mærr kvað þat Gunnarr, \hld Gęir-Niflungr: \\%M
Hér hęfi ek hjarta \hld Hǫgna ins frǿkna, \\%M
ólíkt hjarta \hld Hjalla ins blauða, \\%M
es lítt bifask \hld es á bjóði liggr; \\%M
bifðisk svági mjǫk \hld þá’s í brjósti lá!\\%E
\end{verse}

\bvb Renowned Guthhere quoth this, the Gore-Nifling: “Here have I the heart of Hayne the bold—unlike the heart of Helle the soft!—which little trembles when on the platter it lies; it trembled even less when in the breast it lay.

\begin{verse}
\bva Svá skaltu, Atli, \hld augum fjarri \\%M
sęm munt \hld męnjum verða; \\%M
es und ęinum mér \hld ǫll of folgin \\%M
hodd Niflunga: \hld Lifir-a nú Hǫgni!\\%E
\end{verse}

\bvb Thus shalt thou, Ettle, be as far from the eyes, as thou wilt from the neck-rings. With me alone it is all hidden, the hoard of the Niflings; now Hayne does not live.

\begin{verse}
\bva Ęy vas mér týja \hld  meðan vit tvęir lifðum, \\%M
nú es mér ęngi \hld es ęinn lifi’k; \\%M
Rín skal ráða \hld rógmalmi skatna, \\%M
svinn, ǫ́skunna \hld arfi Niflunga.\\%E
\end{verse}

\bvb I was always doubtful, when we two lived; now I am not, when I alone live. The Rhine shall rule the strife-ore of princes; wisely the os-born inheritance of the Niflings.

\begin{verse}
\bva Í veltanda vatni \hld lýsask valbaugar \\%M
hęldr an á hǫndum gull \hld skíni Húna bǫrnum.\\%E
\end{verse}

\bvb In tumbling water the Welsh rings [will] gleam, rather than [as] gold on hands shine on the children of Huns.”
