\bookStart{The Leed of Hindle}[Hyndluljóð]

\bvg
\bva „Vaki mǽr męyja, \hld\ vaki mín vina, &
Hyndla systir, \hld\ es í hęlli býr; &
nú ’s røkr røkra, \hld\ ríða vit skulum &
til Valhallar \hld\ ok til vés hęilags.\eva

\bvb Frow quoth:
“Wake maiden of maidens, wake my friend, sister Hindle, who lives in the rock-face. Now is the twilight of twilights, we two shall ride to Walhall, and to the holy wigh†!\evb
\evg


\bvg
\bva Biðjum Hęrjafǫðr \hld\ í hugum sitja, &
hann geldr ok gefr \hld\ gull \edtext{verðugum}{\Afootnote{verðungu ‘to the retinue’ \emph{\Finnur\ \GudniEdda} \emph{emend.}}}, &
gaf hann Hęrmóði \hld\ hjalm ok brynju, &
ęn Sigmundi \hld\ sverð at þiggja.\eva

\bvb Let us bid the Father of Hosts \ken{Weden}[1] to be in his favour; he rewards and gives gold to the worthy. Gave he to Heremood helmet and byrnie, but Sighmund a sword to receive.\evb
\evg


\bvg
\bva Gefr hann sigr sumum\footnotetext[1], \hld\ ęn sumum\footnotetext[2] aura, &
mǽlsku mǫrgum \hld\ ok manvit firum, &
byri gefr brǫgnum, \hld\ ęn brag skǫldum, &
gefr hann mannsęmi \hld\ mǫrgum rekki. &
\footnotetext[1] ms. \emph{sonum}
\footnotetext[2] ms. \emph{suinnum}\eva

\bvb He gives victory to some, but to some silver\footnotemark[1]; speech to many, and manwit to men. Fair wind he gives to noble ones, and poetry to scolds†; he gives valour to many a champion.
\footnotemark[1] Lit. "ounces".\evb
\evg


\bvg
\bva Þór munk blóta, \hld\ þess munk biðja, &
at hann ǽ við þik \hld\ einart láti; &
þó ’s hǫ́num ótítt \hld\ við jǫtuns brúðir.\eva

\bvb To Thunder I will bloot†, of this I will bid, that he always show friendliness to thee, though he is prejudiced against the brides of the ettins\footnotemark[1].
\footnotetext[1] Lit. “though [it] is to him infrequent with ettin's brides”.\evb
\evg


\bvg
\bva Nú taktu ulf þinn \hld\ ęinn af stalli, &
lát hann rinna \hld\ með runa mínum.“ &
[Hyndla kvað:] „Sęinn es gǫltr þinn \hld\ goðveg troða, &
vilkat mar minn \hld\ mǽtan hlǿða.\eva

\bvb Now take thy single wolf from the stable; let him run with my boar.” [Hindle quoth:] “Slow is thy boar, to tread the Godways; I wish not lade my dear steed.”\evb
\evg


\bvg
\bva Flǫ́ est Fręyja, \hld\ es fręistar mín, &
visar þú augum \hld\ á oss þannig, &
es hafir ver þinn \hld\ í valsinni &
Óttar unga \hld\ Innsteins bur.“\eva

\bvb Deicitful art thou, Frow, as thou temptest me; thou showest thy eyes on us this way, as thou hast thy man on the Walways: the young Oughthere, Instone's offspring.”\evb
\evg


\bvg
\bva Fręyja kvað:
„Dulið est Hyndla, \hld\ draums ǽtlak þér, &
es kveðr ver minn \hld\ í valsinni.\eva

\bvb Frow quoth:
Thou art foolish, Hindle, I think thee dreamy, who sayest that my man is on the Walways.\evb
\evg


\bvg
\bva Þar’s gǫltr glóar \hld\ Gullinbursti, &
Hildisvíni, \hld\ es mér hagir gęrðu, &
dvergar tvęir \hld\ Dáinn ok Nabbi.\eva

\bvb Where the boar glows, Goldenbristle; the hildswine\footnotemark[1], which the skillful for me made: the two dwarves Dowen and Nab.
\footnotemark[1] \emph{Hildisvíni} 'battle-swine', in this case probably an alternative name for Goldenbristle.\evb
\evg


\bvg
\bva Sęnn í sǫðlum \hld\ sitja vit skulum &
ok of jǫfra \hld\ ǽttir dǿma, &
gumna þęira, \hld\ es frá goðum kómu.\eva

\bvb Soon in the saddles we two shall sit, and judge about the aughts† of princes, of those men who came from the gods.\evb
\evg


\bvg
\bva Þęir hafa vęðjat \hld\ Vala malmi &
Óttarr ungi \hld\ ok Angantýr; &
skylt ’s at vęita, \hld\ svá’t skati hinn ungi & &
fǫðurlęifð hafi \hld\ ępt frǽndr sína.\eva

\bvb They have wagered the Welsh ore [GOLD], young Oughter and Ongenthew; it is required to grant, so that the young prince might have the fatherly inheritance left behind by his kinsmen.\footnotemark[1]
\footnotemark[1] Lit. 'the father-remains after his kinsmen'. — Happening seems to be that Oughthere and Ongenthew each lay claim the inheritance. In order to settle the matter (in Oughthere's favour) Hindle must (\emph{skylt es} “it is required, obligated”) divulge (\emph{vęita} ‘to grant, to give away’) what she knows about his lineage.\evb
\evg


\bvg
\bva Hǫrg hann mér gęrði \hld\ hlaðinn stęinum; &
nú ’s grjót þat \hld\ at glęri orðit; &
rauð hann í nýju \hld\ nauta blóði; &
ǽ trúði Óttarr \hld\ á ǫ́synjur.\footnotemark[1]
\footnotemark[1] Frow argues yet further in favour of Oughthere, bringing up his piety shown towards the godesses.\eva

\bvb A harrow† he made for me, loaded with stones; now that stone-pile is become into glass. He reddened [it] in fresh blood of oxen; Oughthere ever trusted on the osennies†.\evb
\evg


\bvg
\bva Nú lát-tu forna \hld\ niðja talða &
ok uppbornar \hld\ ǽttir manna &
hvat ’s Skjǫldunga, \hld\ hvat ’s Skilfinga, &
hvat ’s Ǫðlinga \hld\ hvat ’s Ylfinga & &
hvat ’s hǫldborit, \hld\ hvat ’s hęrsborit &
męst manna val \hld\ und Miðgarði?“\eva

\bvb Now let be recounted the ancient lines of kinsmen, and the upborn\footnote[1] aughts† of men: What is of the Shieldings? What is of the Shilvings? What is of the Athlings? What is of the Wolvings? What is born of hero? What is born of chief, the mightiest choice of men in Midyard?”
\footnote[1] Noble.\evb
\evg


\bvg
\bva „Þú est Óttarr \hld\ borinn Innstęini, &
ęn Innstęinn vas \hld\ Alfi inum gamla, &
Alfr vas Ulfi, \hld\ Ulfr Sǽfara, &
ęn Sǽfari \hld\ Svan inum rauða.\eva

\bvb Hindle quoth:
“Thou\footnote[1] art, Oughthere, born to Instone, but Instone was born to Elf the old, Elf to Wolf, Wolf to Seafare, but Seafare to Swan the red.
\footnote[1] Hindle, apparently in a trance-like state, speaks straight to Oughthere.\evb
\evg


\bvg
\bva Móður átti faðir þinn \hld\ męnjum gǫfga, &
hygg at héti \hld\ Hlédís gyðja, &
Fróði vas faðir þęirar, \hld\ ęn Fríund\footnotemark[1] móðir; &
ǫll þótti ǽtt sú \hld\ með yfirmǫnnum.\eva
\footnotemark[1] Emended from the meaningless ms. reading \emph{friaut}.

\bvb Thy father had thy mother, beautiful with neck-rings, I think that she was called Leedise yidde†. Frood was her father, but Friend her mother; all her aught seemed to be among overmen.\evb
\evg


\bvg
\bva Auði vas áðr \hld\ ǫflgastr manna, &
Halfdanr fyrri \hld\ hǽstr Skjǫldunga, &
frǽg vǫ́ru folkvíg, \hld\ þaus framir gęrðu, &
hvarfla þóttu verk \hld\ með himins skautum.\eva

\bvb Ed was before [that] the most powerful of men, Halfdane earlier the highest of Shieldings. Renowned were the troop-battles which the famous ones performed; his <= Halfdane's> works seemed to travel around the corners of heaven.\evb
\evg


\bvg
\bva Ęflðisk við Ęymund \hld\ ǿztan manna &
ęn vá Sigtrygg \hld\ með svǫlum ęggjum, &
ęiga gekk Almvęig, \hld\ ǿzta kvinna, &
ólu þau ok ǫ́ttu \hld\ átján sonu.\eva

\bvb He <= Halfdane> became the in-law of Iemund\footnotemark[1], the noblest of men, but he slew Sightrue with cool edges. He went on to have Elmwey, the noblest of women; they begot and had eighteen sons.
\footnotemark[1] Lit. "[he] was strengthened by". Parallelism of "noblest of men/women" makes the meaning yet clearer. Elmwey was Iemund's daughter or sister.\evb
\evg


\bvg
\bva Þaðan eru Skjǫldungar, \hld\ þaðan eru Skilfingar, &
þaðan eru Ǫðlingar, \hld\ þaðan eru Ynglingar, &
þaðan es hǫldborit, \hld\ þaðan es hęrsborit, &
mest mannaval \hld\ und Miðgarði; &
alt ’s þat ǽtt þín, \hld\ Óttarr heimski.\eva

\bvb Thereof are the Shieldings! Thereof are the Shilvings! Thereof are the Inglings!\footnotemark[1] Thereof is born of hero! Thereof is born of chief, the mightiest choice of men in Midyard! That is all thy aught†, foolish Oughthere!”
\footnotemark[1] Note the contradiction with v. 12. Since the Inglings have already been mentioned (under the name Shilvings, of the difference between the two see the index), it seems likely that Wolvings is the original reading.\evb
\evg


\bvg
\bva Vas Hildigunnr \hld\ hęnnar móðir, &
Svǫ́fu barn \hld\ ok sǽkonungs; &
alt ’s þat ǽtt þín, \hld\ Óttarr hęimski. &
varðar\footnote[1] at viti svá, \hld\ viltu ęnn lęngra?\eva
\footnote[1] Emended from ms. \emph{varði}.

\bvb Hildguth was her mother, the child of Swabe and Seaking; that is all thy aught†, foolish Oughthere!—It is meaningful that one might know thus; wilt thou [go] yet further?\evb
\evg


\bvg
\bva Dagr átti Þóru \hld\ dręngjamóður, &
ólusk í ǽtt þar \hld\ ǿztir kappar, &
Fraðmarr ok Gyrðr \hld\ ok Frekar báðir, &
Ámr ok Jǫsurmarr, \hld\ Alfr hinn gamli. &
varðar at viti svá, \hld\ viltu ęnn lęngra?\eva

\bvb Day had Thure, the mother of valiant men; in that aught were begotten the noblest champions: Fradmer and Yird, and both Frecks; Ame and Essirmer; Elf the old.—It is meaningful that one might know thus; wilt thou [go] yet further?\evb
\evg


\bvg
\bva Kętill hét vinr þęira \hld\ Klypps arfþęgi, &
vas hann móðurfaðir \hld\ móður þinnar; &
þar vas Fróði \hld\ fyrr ęnn Kári, &
ęn Hildi vas \hld\ Hóalfr of getinn.\eva

\bvb Kettle, the inheritor of Clip, was their friend; he was the father of thy mother's mother. There was Frood, yet earlier Keer, but Highelf was by Hild begotten.\evb
\evg

... %TODO More dialogue
