\bookStart{The Leed of Hoarbeard}[Hárbarðsljóð]

BPG
BPAÞórr fór ór austrvegi ok kom at sundi einu. Ǫðrum megum sundsins var ferjukarlinn með skipit. Þórr kallaði:EPA

BPBThunder travelled out of the eastern ways and came to a sound. At the other side of the sound was the ferryman with the ship. Thunder called out:EPB
EPG

\bvg
\bva „Hvęrr ’s sá svęinn svęina \hld es stęndr fyr sundit handan?“\eva

\bvb “Who is that swain of swains, that stands across the sound?”\evb
\evg


\bvg
\bva Hann svaraði:
„Hvęrr ’s sá karl karla \hld es kallar of váginn?“\eva

\bvb He answered:
“Who is that churl of churls, that calls out over the wave?”\evb
\evg


\bvg
\bva „Fęr þú mik of sundit, \hld fǿði’k þik á morgun; &
męis hęfi’k á baki, \hld verðr-a matrinn bętri.“\eva

\bvb [Thunder:]
“Ferry me over the sound, I feed thee in the morning! A basket\footnoteB{TODO: Note about \emph{męis} and its connection with the later verse previously interpreted as referring to the sctroum.} I have on my back, the food does not get better.”\evb
\evg


\bvg
\bva „Át ek í hvíld \hld áðr ek hęiman fór, &
síldr ok hafra; \hld saðr em’k ęnn þęss.“\eva

\bvb [Ferryman:]
“I ate for a while before I travelled from home, herring and hegoats; I am still full from that.\evb
\evg


\bvg
\bva Árligum verkum \hld hrósar þú verðinum; &
veizt at u fyr görla, \hld döpr eru þín heimkynni, &
dauð hygg ek að þín móðir sé.„\eva

\bvb Gaming\evb
\evg

%TODO: rest of verses

\bvg Þórr kvað:
\bva „Skammt mun nú mál okkat vesa, \hld allz þú mér skǿtingu ęinni svarar; &
launa mun ek þér farsynjun \hld ef vit finnumk í sinn annat! &
\ind Farþú nú þar’s þik hafi allan gramir!“\eva

\bvb Thunder quoth:
“Now our speech will be short, as thou answers me with taunts alone; I will reward thee for this ferry-refusal if we meet another time! Now go whither the fiends may have all of thee!”\evb
\evg
