\bookStart{Thule of Righ}[Rígsþula]

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: early C11th (0.240), late C11th (0.204), late C12th (0.195), C13th (0.280)

\textbf{Meter:} \Fornyrdislag%
\end{flushright}

Dumezil hypothesis. Irish influence? Many interesting things to write here!

The language of \Rigsthula\ is highly formulaic, but also often unique to it. Of particular note are the alliteration between the adverb \emph{męirr} ‘further’ and \emph{miðra}, e.g. in st. 2/1: \emph{gekk męirr at þat}

\sectionline

\bpg
\bpa\mssnote{\Wormianus~78r/1}Svá sęgja męnn í fornum sǫgum, at ęinn-hvęrr af ǫ́sum, sá er Hęimdallr hét, fór fęrðar sinnar ok framm með sjóvar-strǫndu nǫkkurri, kom at ęinum húsa-bǿ ok nęfndisk Rigr; ęptir þęiri sǫgu er kvę́ði þetta.\epa

\bpb So say men in ancient \inx[C]{saw}[saws] that one of the \inx[G]{Eese}, he who was called \inx[P]{Homedal}, went on his journey and came forth along a certain lake shore, came upon a lone homestead and called himself Righ—according to that saw is this poem.\epb
\epg


\bvg\bva\mssnote{\Wormianus~78r/TODO}\edtrans{Ár}{Of yore}{\Afootnote{sens. emend. (see note); \emph{at} \Wormianus}\Bfootnote{Formulaic. It is very common for poems to begin with \emph{ár} ‘of yore, in the beginning, in the dawn’. Cf. \Voluspa\ 3/1, \Hymiskvida\ 1/1, \HelgakvidaOne\ 1/1, \GudrunOne\ 1/1, \Sigurdskamma\ 1/1}} kvǫ́ðu \alst{g}anga \hld\ \alst{g}rǿnar brautir &
\alst{ǫ}flgan ok \alst{a}ldinn \hld\ \alst{ǫ́}s kunnigan, &
\alst{r}amman ok \alst{r}ǫskvan \hld\ \alst{R}íg stíganda.\eva

\bvb Of yore, they said, did walk on green highways \\
a mighty and ancient \inx[G]{Eese}[os], cunning: \\
the strong and brisk Righ, striding.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}Gekk \alst{m}ęirr at þat \hld\ \alst{m}iðrar brautar, &
kom hann at \alst{h}úsi, \hld\ \alst{h}urð vas á gę́tti; &
\alst{i}nn nam at ganga, \hld\ \alst{ę}ldr vas á golfi, &
\alst{h}jón sǫ́tu þar \hld\ \alst{h}ǫ́r \edtext{at}{\Afootnote{sens. emend.; \emph{af} \Wormianus}} arni, &
\alst{Á}i ok \alst{Ę}dda \hld\ \alst{a}ldin-falda.\eva

\bvb Went he further after that in the middle of the road; \\
came to a house—the door was wide open. \\
He took to go inside; fire was on the floor. \\
A couple sat there, hoary by the hearth: \\
Great-Grandfather and Great-Grandmother, old-fashioned.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}\alst{R}igr kunni þęim \hld\ \alst{r}ǫ́ð at sęgja; &
\alst{m}ęirr sęttisk hann \hld\ \alst{m}iðra flętja &
en á \alst{h}lið \alst{h}vára \hld\ \alst{h}jón sal-kynna.\eva

\bvb Righ knew to tell them counsels, \\
further he set himself down on the middle of the bench, \\
and on either side the couple of the hall.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}%
Þá tók \alst{Ę}dda \hld\ \alst{ø}kkvinn hlęif, &
\alst{þ}ungan ok \alst{þ}ykkvan, \hld\ \alst{þ}runginn sǫ́ðum, &
bar hǫ́n \alst{m}ęirr at þat \hld\ \alst{m}iðra skutla, &
\alst{s}oð vas í bolla \hld\ \alst{s}ętti á bjóð; &
vas \alst{k}alfr soðinn \hld\ \alst{k}rása bętstr; &
\alst{r}ęis hann upp þaðan, \hld\ \alst{r}éðsk at sofna;\eva

\bvb Then took Great-Grandmother a lumpy loaf, \\
heavy and thick, stuffed with chaff, \\
she carried it further after that on the middle of a trencher, \\
broth was in a bowl, she set it on a plate— \\
a cooked calf was the best dainty; \\
he \ken*{= Righ} rose up thence, resolved to sleep.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}\alst{R}igr kunni þęim \hld\ \alst{r}ǫ́ð at sęgja; &
\alst{m}ęirr lagðisk hann \hld\ \alst{m}iðrar rękkju, &
en á \alst{h}lið \alst{h}vára \hld\ \alst{h}jón sal-kynna.\eva

\bvb Righ knew to tell them counsels; \\
further he laid himself down in the middle of the bed, \\
and on either side the couple of the hall.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}\alst{Þ}ar vas hann at þat \hld\ \alst{þ}rjár nę́tr saman; &
gekk hann \alst{m}ęirr at þat \hld\ \alst{m}iðrar brautar; &
liðu \alst{m}ęirr at þat \hld\ \alst{m}ǫ́nuðr níu.\eva

\bvb There he was after that for three nights in all; \\
went he further after that in the middle of the road; \\
passed further after that nine months.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}\alst{Jó}ð \alst{ó}l \alst{Ę}dda, \hld\ \edtrans{\alst{jó}su vatni}{they sprinkled it with water}{\Bfootnote{A reference to the Heathen naming ceremony wherein water would be poured on a newborn, somewhat resembling the Christian baptism.  See \Havamal\ 156.}} &
\edtrans{\alst{h}ǫrund-svartan}{swarthy of skin}{\Afootnote{emend.; \emph{hǫrfi svartan} ‘swarthy with flax(?)’ \Wormianus}}, \hld\ \alst{h}étu Þrę́l.\eva

\bvb Great-Grandmother begot a child—they sprinkled it with water: \\
swarthy of skin, they called it Thrall.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}Hann nam at \alst{v}axa \hld\ ok \alst{v}ęl dafna; &
vas þar á \alst{h}ǫndum \hld\ \alst{h}rokkit skinn, &
\alst{k}ropnir \alst{k}núar, \hld\ [...] &
\alst{f}ingr digrir, \hld\ \alst{f}úlligt and-lit, &
\alst{l}otr hryggr, \hld\ \alst{l}angir hę́lar.\eva

\bvb He took to grow and have it well; \\
there on his hands was wrinkled skin, \\
crooked knuckles, [...], \\
stubby fingers, loathsome face, \\
stooping back, long heels.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}Nam \alst{m}ęirr at þat \hld\ \alst{m}agns of kosta, &
\alst{b}ast at \alst{b}inda, \hld\ \alst{b}yrðar gørva; &
bar \alst{h}ęim at þat \hld\ \alst{h}rís gęrstan dag.\eva

\bvb He took further after that to try his strength: \\
bast to bind, burdens to make; \\
he carried home after that brushwood on a gloomy day.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}Þar kom at \alst{g}arði \hld\ \edtrans{\alst{g}ęngil-bęina}{gangle-boned woman}{\Bfootnote{Derogatory, somebody who (due to poverty) only travels by foot.}}, &
\alst{au}rr vas á \alst{i}ljum, \hld\ \alst{a}rmr sól-brunninn, &
\alst{n}iðr-bjúgt es \alst{n}ęf, \hld\ \alst{n}ęfndisk \edtrans{Þír}{Thew}{\Bfootnote{The name probably means ‘maid-servant’ or ‘female slave’. Unlike Thrall, it is not attested in any prose texts, but probably corresponds to OS \emph{thiwi} ‘maid(-servant)’, being further root-related to \emph{þéa \char`~\ þjá} ‘to enthral’, Proto-Norse \textbf{þewaʀ} ‘servant’, OE \emph{þéow} ‘slave, servant’,.}}.\eva

\bvb There came to the farm a gangle-boned woman: \\
mud was on her footsoles, her arm sunburnt, \\
downturned her face—she called herself Thew.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}\edtext{Męirr sęttisk hǫ́n \hld\ miðra flętja,}{\lemma{Męirr \dots\ flętja}\Afootnote{emend. based on other sts.; \emph{miðra flętja \hld\ męirr sęttisk hǫ́n} \Wormianus}} &
\alst{s}at hjá hęnni \hld\ \alst{s}onr húss, &
\alst{r}ǿddu ok \alst{r}ýndu, \hld\ \alst{r}ękkju gørðu &
\alst{Þ}rę́ll ok \alst{Þ}ír \hld\ \alst{þ}rungin dǿgr.\eva

\bvb Further she set herself down on the middle of the bench; \\
by her sat the son of the house \ken*{= Thrall}. \\
They spoke and whispered, made a bed— \\
Thrall and Thew—in hard-pressed nights.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}\alst{B}ǫrn ólu þau, \hld\ \alst{b}juggu ok unðu; &
\alst{h}ygg’k at \alst{h}éti \hld\ \alst{H}ręimr ok Fjósnir, &
\alst{K}lúrr ok \alst{K}lęggi, \hld\ \alst{K}ęfsir, Fúlnir, &
\alst{D}rumbr, \alst{D}igraldi, \hld\ \alst{D}rǫttr ok Hǫsvir, &
\alst{L}útr ok \alst{L}ęggjaldi; \hld\ \alst{l}ǫgðu garða, &
\alst{a}kra tǫddu, \hld\ \alst{u}nnu at svínum, &
\alst{g}ęita \alst{g}ę́ttu, \hld\ \alst{g}rófu torf.\eva

\bvb Children they begot—they settled and were content— \\
I think that they were called Rame and Feesner, \\
Clour and Cledge, Chafser, Foulner, \\
Drumber, Digrald, Drant and Hazer, \\
Lout and Ledgald.—They laid yard-fences, \\
dunged fields, fed swine, \\
herded goats, dug turf.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}\alst{D}ǿtr vǫ́ru þę́r \hld\ \alst{D}rumba ok Kumba, &
\alst{Ø}kkvin-kalfa \hld\ ok \alst{A}rin-nęfja, &
\alst{Y}sja ok \alst{A}mbǫ́tt, \hld\ \alst{Ęi}kin-tjasna, &
\alst{T}ǫtrug-hypja \hld\ ok \alst{T}rǫnu-bęina; &
\alst{þ}aðan eru komnar \hld\ \alst{þ}rę́la ę́ttir.\eva

\bvb The daughters were Drumb and Cumb; \\
Inkencalf and Arn-neb, \\
Yeaze and Ambight, Oakentezen, \\
Tattryhip and Tranebone— \\
from thence are come the lines of thralls.\evb\evg


\sectionline


\bvg\bva\mssnote{\Wormianus~78r/TODO}Gekk \alst{R}ígr at þat \hld\ \alst{r}éttar brautir &
kom hann at \edtrans{\alst{h}ǫllu}{hall}{\Afootnote{sens. and metr. emend., cf. st. TODO; om. \Wormianus}} \hld\ \alst{h}urð vas á skiði &
\alst{i}nn nam at ganga, \hld\ \alst{ę}ldr vas á golfi &
\alst{h}jón sǫ́tu þar \hld\ \alst{h}eldu á syslu.\eva

\bvb Went Righ after that on straight highways; \\
he came to a hall—the TODO. \\
He took to go inside; fire was on the floor. \\
A couple sat there, busy with their chores:\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}\alst{M}aðr tęlgði þar \hld\ \alst{m}ęið til rifjar, &
vas \alst{sk}ęgg \alst{sk}apat, \hld\ \alst{sk}ǫr vas fyr ęnni &
\alst{sk}yrtu þrǫngva \hld\ \alst{sk}okkr vas á golfi.\eva

\bvb A man there carved a stick into a loom-beam. \\
His beard was shapely, locks hung down his forehead, \\
his shirt tight; a toolbox was on the floor.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}\alst{S}at þar kona, \hld\ \alst{s}vęigði rokk, &
\alst{b}ręiddi faðm, \hld\ \alst{b}jó til váðar; &
\alst{s}vęigr vas á hǫfði, \hld\ \alst{s}mokkr vas á bríngu, &
\alst{d}úkr vas á halsi, \hld\ \alst{d}vergar á ǫxlum; &
\alst{A}fi ok \alst{A}mma \hld\ \alst{ǫ́}ttu hús.\eva

\bvb There sat a woman, twirled a distaff, \\
stretched out her arms, readied a cloth. \\
A scarf was on her head, a smock on her breast, \\
a kerchief on her throat, brooches on her shoulders— \\
Grandfather and Grandmother owned a house.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}\alst{R}ígr kunni þęim \hld\ \alst{r}ǫ́ð at sęgja, &
\alst{r}ęis frá borði \hld\ \alst{r}éð at sofna. &
\alst{M}ęirr lagðisk hann \hld\ \alst{m}iðrar rękkju &
en á \alst{h}lið hvára \hld\ \alst{h}jón sal-kynna. &
\alst{Þ}ar vas hann at þat \hld\ \alst{þ}rjár nę́tr saman &
liðu \alst{m}ęirr at þat \hld\ \alst{m}ǫ́nuðr níu.\eva

\bvb Righ knew to tell them counsels; \\
rose from the table, resolved to sleep. \\
Further he laid himself down in the middle of the bed, \\
and on either side the couple of the hall. \\
There he was after that for three nights in all; \\
passed further after that nine months.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}Jóð ól Amma, \hld\ jósu vatni, &
\alst{k}ǫlluðu \alst{K}arl \hld\ \alst{k}ona svęip ripti &
\alst{r}auðan ok \alst{r}jóðan \hld\ \alst{r}iðuðu augu.\eva

\bvb Grandmother begot a child, they sprinkled it with water, \\
called it Churl; the woman wrapped him in cloth, \\
red and ruddy; his eyes trembled.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}Hann nam at \alst{v}axa \hld\ ok \alst{v}ęl dafna, &
ǫxn nam at tęmja \hld\ arðr at gørva &
hús at timbra \hld\ ok hlǫður smíða &
karta at gørva \hld\ ok kęyra plóg.\eva

\bvb He took to grow and have it well; \\
oxen he took to tame, the ard to make, \\
houses to timber and barns to craft, \\
carts to make and drive the plough.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}Hęim óku þá \hld\ Hangin-luklu &
gęita kyrtlu \hld\ giptu Karli. &
Snǫr hęitir sú, \hld\ sęttisk und ripti. &
Bjuggu hjón, \hld\ bauga dęildu, &
bręiddu blę́jur, \hld\ ok bú gørðu.\eva

\bvb Homewards then drove Hangenkey, \\
TODO, married her to Churl. \\
Daughter-in-law she is called; she set herself under a cloth. \\
The couple settled, shared their money, \\
spread fine cloth and made a home.\evb\evg


\bvg\bva\mssnote{\Wormianus~78r/TODO}\alst{B}ǫrn ólu þau, \hld\ \alst{b}juggu ok unðu; &
hét Halr ok Drengr, \hld\ Hǫldr, Þegn ok Smiðr, &
Bręiðr, Bóndi, \hld\ Bundin-skęggi, &
Búi ok Boddi \hld\ Bratt-skęggr ok Sęggr.\eva

\bvb Children they begot—they settled and were content— \\
they were called Hale and Drang, Haled, Thane and Smith, \\
Broad, Bond, Boundenshag, \\
Bower and Bod, Brantshag and Sedge.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/1}Enn hétu svá \hld\ ǫðrum nǫfnum &
Snot, Brúðr, Svanni, \hld\ Svarri, Sprakki, &
Fljóð, Sprund, ok Víf, \hld\ Fęima, Ristill— &
þaðan eru \alst{k}omnar \hld\ \alst{k}arla ę́ttir.\eva

\bvb Yet some were called so with other names: \\
Snot, Bride, Swannie, Swarrie, Sprackie, \\
Fleed, Sprund and Wife, Fome, Ristle— \\
from thence are come the lines of churls.\evb\evg


\sectionline


\bvg\bva\mssnote{\Wormianus~78v/TODO}Gekk Rigr þaðan \hld\ réttar brautir &
kom hann at sal, \hld\ suðr horfðu dyrr, &
vas hurð hnigin, \hld\ hringr vas í gę́tti.\eva

\bvb TODO: Translation.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Gekk hann inn at þat \hld\ golf vas stráat &
sǫ́tu hjón \hld\ sǫ́usk í augu &
faðir ok móðir \hld\ fingrum at lęika.\eva

\bvb TODO: Translation.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Sat hús-gumi \hld\ ok snøri stręng &
alm of bęndi \hld\ ǫrvar skępti; &
en hús-kona \hld\ hugði at ǫrmum, &
strauk of ripti \hld\ sterti ęrmar.\eva%TODO: sterti or stęrti?

\bvb Sat the man of the houise and twisted the bow-string, \\
bent the elmwood, shafted arrows— \\
but the housewife minded her arms, \\
smoothened the fabric, tightened the sleeves.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Kęisti fald, \hld\ kinga vas á bringu, &
síðar slǿður, \hld\ sęrk blá-fáan; &
brún bjartari, \hld\ brjóst ljósara, &
hals hvítari \hld\ hręinni mjǫllu.\eva

\bvb The linen hood jutted out, a brooch was on her chest, \\
a long-hanging gown, her serk dyed blue; \\
her brow was brighter, her chest lighter, \\
her throat whiter than purest snow.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Rigr kunni þęim \hld\ rǫ́ð at sęgja; &
męirr sęttisk hann \hld\ miðra flętja &
en á hlið hvára \hld\ hjón sal-kynna.\eva

\bvb Righ knew to tell them counsels, \\
further he set himself down on the middle of the floor-bench, \\
and on either side: the couple of the hall.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Þá tók móðir \hld\ męrktan dúk, &
hvítan af hǫrvi, \hld\ hulði bjóð; &
hón tók at þat \hld\ hlęifa þunna, &
hvíta af hvęiti, \hld\ ok hulði dúk.\eva

\bvb Then took Mother a patterned cloth, \\
white of flax—she covered a platter. \\
She took after that thin loaves, \\
white of wheat—and covered the cloth.\footnoteB{Note the strong parallelism. The household can afford an excess of expensive fabric and bread; Mother can cover the platter with a patterned (\emph{męrktr}) flaxen cloth, and then cover the cloth with wheat-bread.}\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Framm sętti hón \hld\ skutla fulla &
silfri varða á bjóð &
fán ok flęski \hld\ ok fugla stęikta &
vín vas i kǫnnu \hld\ varðir kalkar; &
drukku ok dǿmðu; \hld\ dagr vas á sinnum.\eva

\bvb TODO: Translation.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Rigr kunni þęim \hld\ rǫ́ð at sęgja, &
ręis Rigr at þat, \hld\ rękkju gørði.\eva

\bvb Righ knew to tell them counsels, \\
rose Righ after that, made the bed.\evb\evg

\bvg\bva\mssnote{\Wormianus~78v/TODO}Þar vas hann at þat \hld\ þrjár nę́tr saman; &
gekk hann męirr at þat \hld\ miðrar brautar; &
liðu męirr at þat \hld\ mǫ́nuðr níu.\eva

\bvb There he was after that for three nights in all; \\
went he further after that on the middle of the road; \\
passed further after that nine months.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Svęin ól móðir, \hld\ silki vafði, &
jósu vatni— \hld\ Jarl létu hęita; &
blęikt vas hár, \hld\ bjartir vangar, &
\edtext{ǫtul vǫ́ro augu \hld\ sem yrmlingi}{\lemma{ǫtul \dots\ yrmlingi ‘fierce \dots\ the young serpent’}\Bfootnote{A person of noble stock being recognised as such through their appearance is a motif in Norse literature. Cf. esp. the incident at the beginning of \HelgakvidaTwo, where Hallow, disguised as a thrall-woman, is almost caught due to his unslavelike eyes, which are, as in the present stanza, likewise said to be \emph{ǫtul} ‘fierce, terrible’.}}.\eva

\bvb Mother begot a swain, swaddled him in silk; \\
they sprinkled him with water—let him be called Earl. \\
Pale was his hair, bright his cheeks, \\
fierce were his eyes, like the young serpent.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Upp óx þar \hld\ Jarl á flętjum; &
lind nam at skęlfa, \hld\ lęggja stręngi, &
alm at bęygja, \hld\ ǫrvar skępta, &
flęin at flęyja, \hld\ frǫkkur dýja, &
hęstum ríða, \hld\ hundum verpa, &
sverðum bregða, \hld\ sund at fręmja.\eva

\bvb Up grew Earl there on the floor-benches; \\
he took to shake shields, fasten bow-strings, \\
bend elmwood, shaft arrows, \\
throw javelins, hoist frankish spears, \\
ride horses, throw hounds (TODO) \\,
brandish swords, practice swimming.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}\edtext{Kom þar ór runni \hld\ Rigr gangandi, &
Rigr gangandi, \hld\ rúnar kęnndi; &
sitt gaf hęiti, \hld\ son kveðsk ęiga; &
þann bað hann ęignask \hld\ óðal-vǫllu, &
óðal-vǫllu, \hld\ aldnar bygðir.}{\lemma{Kom \dots\ bygðir.}\Bfootnote{Righ approaches his son, Earl. He reveals himself as his father and initiates him into the warrior aristocracy through teaching him the runes and giving him the noble title Righ (henceforth he will be known as Righ Earl). Finally he instructs him to set out and win land for himself, which Righ Earl soon does.}}\eva

\bvb There came out of a brush Righ, walking: \\
Righ, walking, taught runes; \\
he gave his own name; said that he had a son; \\
he bade \emph{him} take the ethel-plains: \\
the ethel-plains, the ancient villages.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Ręið hann męirr þaðan \hld\ myrkan við &
hélug fjǫll \hld\ unds at hǫllu kom; &
skapt nam at dýja, \hld\ skęlfði lind, &
hęsti hlęypti, \hld\ ok hjǫrvi brá; &
víg nam at vękja, \hld\ vǫll nam at rjóða, &
val nam at fęlla, \hld\ vá til landa.\eva

\bvb He \ken{= Righ-Earl} rode further thence through the mirky wood, \\
through the frosty fells, until to a hall he came— \\
the shaft he took to hoist, shook the linden shield, \\
leapt with the horse, and brandished the sword; \\
war he took to rouse, the plain he took to redden, \\
men he took to fell—he won the land.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Réð hann ęinn at þat \hld\ átján búum; &
auð nam skipta \hld\ ǫllum vęita &
męiðmar ok mǫsma, \hld\ mara svang-rifja; &
\edtrans{hringum hręytti}{rings he scattered}{\Bfootnote{Cf. StarkSt Frag 1/2a \emph{hring-hręytanda} ‘ring-scattererer \ken{generous man}’ which contains the same words.}}, \hld\ hjó sundr baug.\eva

\bvb He alone ruled, after that, eighteen homesteads. \\
Wealth he took to hand out; to give all men \\
gifts and treasures, [and] slender-ribbed steeds; \\
rings he scattered; he cut apart a bigh.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}\edtext{Óku}{\Afootnote{\emph{okū} \Wormianus}} ę́rir \hld\ úrgar brautir &
kvǫ́mu at hǫllu \hld\ þar’s hęrsir bjó: &
mǿtti [...] \hld\ \edtext{mjó-fingraðri}{\Afootnote{the grammar requires \emph{-ri}; mjó-fingraði \Wormianus}} &
hvítri ok horskri, \hld\ hétu Ęrna.\eva

\bvb Messengers drove through drizzling roads, \\
came to the hall where a ruler lived; \\
met a slender-fingered, \\
white and wise—they called her Erne.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Bǫ́ðu hęnnar \hld\ ok hęim óku, &
giptu Jarli, \hld\ \edtrans{gekk hón und líni}{she went ’neath the linen}{\Bfootnote{i.e. she donned the bridal veil; cf. \Thrymskvida\ 27.}}; &
saman bjuggu þau \hld\ ok sér unðu, &
ę́ttir jóku \hld\ ok aldrs nutu.\eva

\bvb They asked for her hand and drove home, \\
married her off to Earl—she went under the linen. \\
They settled together and were content with themselves, \\
grew their lineage and enjoyed life.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Burr vas hinn ęlsti, \hld\ en Barn annat; &
Jóð ok Aðal, \hld\ Arfi, Mǫgr, &
Niðr ok Niðjungr, \hld\ (nǫ́mu lęika) &
Sonr ok Svęinn, \hld\ (sund ok tafl) &
Kundr hét ęinn; \hld\ Konr vas hinn yngsti.\eva

\bvb Byre was the oldest, and Bairn another; \\
TODO: Translation. \\
TODO: Translation (they learned to play)
Son and Swain (swimming and Tavel)
Kund was one called; Kin was the youngest.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Upp óxu þar \hld\ Jarli bornir: &
hęsta tǫmðu, \hld\ hlífar bęndu, &
skęyti skófu, \hld\ skęlfðu aska. &
En \edtrans{Konr ungr}{Kin the Young}{\Bfootnote{The name is clearly a folk etymological pun on ON \emph{konungr} ‘king’, who held the highest social rank, above even the earls.}} \hld\ kunni rúnar: &
ę́vin-rúnar \hld\ ok aldr-rúnar.\eva

\bvb There grew up the sons of Earl: \\
horses they tamed, shield-rims they bent, \\
smoothened shafts, shook ashen spears.— \\
But Kin the Young knew runes: \\
ever-runes and life-runes.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}%
Męirr kunni hann \hld\ mǫnnum bjarga, &
ęggjar dęyfa, \hld\ ę́gi lę́gja; &
klǫk nam fugla, \hld\ kyrra ęlda, &
sǿfa ok svęfja, \hld\ sorgir lę́gja, &
afl ok ęljun \hld\ átta manna.\eva

\bvb Further he knew men to save, \\
blades to dull, the sea to lower; \\
he learned the chirping of birds, to calm fires, \\
to soothe and lull to sleep, to lower sorrows; \\
the strength and zeal of eight men.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Hann við Rig Jarl \hld\ rúnar dęildi; &
brǫgðum bęitti \hld\ ok bętr kunni; &
þá ǫðladisk \hld\ ok þá ęiga gat, &
Rigr at hęita, \hld\ rúnar kunna.\eva

\bvb With Righ-Earl he shared runes; \\
TODO. \\
then he earned for himself, and got to own, \\
Righ to be called, runes to know.\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Ręið Konr ungr \hld\ kjǫrr ok skóga; &
kolfi flęygði \hld\ kyrði fugla; &
þá kvað þat kráka \hld\ —sat kvisti ęin— &
„Hvat skalt, Konr ungr, \hld\ kyrra fugla? &
Hęldr mę́tti þér \hld\ hęstum ríða &
{[...]} \hld\ ok hęr fęlla.\eva

\bvb Kin the Young rode through brushes and woods, \\
flung bolts, hunted birds. \\
Then quoth a crow—sat on a branch alone— \\
“Why shalt thou, Kin the Young, hunt birds? \\
Better it fit thee horses to ride, \\
{[...]}, and armies to fell.”\evb\evg


\bvg\bva\mssnote{\Wormianus~78v/TODO}Á Danr ok Danpr \hld\ dýrar hallir; &
ǿðra \edtrans{óðal}{ethel}{\Bfootnote{Ancestral farmland, in this case the eighteen homesteads owned by Earl.}} \hld\ an \edtrans{ér}{ye}{\Afootnote{metr. emend.; \emph{þér} ‘id.’ \Wormianus, which is simply a younger form of \emph{ér}, and shows that the poem has been linguistically modernised.}} hafið; &
þęir kunnu vel \hld\ \edtrans{kjól at riða}{ship to ride}{\Bfootnote{i.e. to sail.}}, &
\edtrans{ęgg at kęnna}{the blade to teach}{\Bfootnote{i.e. to fight, wage war.  A euphemism; to “teach someone the blade” is to fight him.}}, \hld\ undir rjúfa.\eva

\bvb Dan and Danp own costly halls: \\
nobler ethel than ye do— \\
they know well the ship to ride, \\
the blade to teach, wounds to tear.\evb\evg

\sectionline

At this point leaf 78 ends. The rest of the poem is lost.

\sectionline
