\bookStart{The Icelandic Rune Poem}

\begin{flushright}%
Dating: Medieval.%TODO

Meter: Unclear.
\end{flushright}%

The poem is highly formulaic.  All lines begin with the respective rune’s name, followed by three synonyms.  It is only attested in late manuscripts which often have major disagreements with each other.

\sectionline

\bvg\bva \alst{F}é es \alst{f}rę́nda róg \hld\ ok \alst{f}lǿðar viti &
\ind ok \alst{g}raf-sęiðs \alst{g}ata.\eva

\bvb Fee is strife of kinsmen and beacon of the sea \\
and grave-saithe’s \ken{serpent’s} street.\evb\evg


\bvg\bva Úr es \alst{sk}ýja grátr \hld\ ok \alst{sk}ára þvęrrir &
\ind ok \alst{h}irðis \alst{h}atr.\eva

\bvb Drizzle is weeping of clouds and ... \\
and shepherd’s hatred.\evb\evg


\bvg\bva Þurs es \alst{k}venna \alst{k}vǫl \hld\ ok \alst{k}letta í·búi &
\ind ok \alst{v}arð-rúnar \alst{v}err.\eva

\bvb Thurse is women’s torment and indweller of hills \\
and husband of the weird-whisperess \ken{giantess}.\evb\evg


\bvg\bva \alst{Ǫ́}ss es \alst{a}ldinn gautr \hld\ ok \alst{Ǫ́}s-garðs jǫfurr, &
\ind ok \alst{V}al-hallar \alst{v}ísi.\eva

\bvb Os is ancient Geat, and Osyard’s chief, \\
and Walhall’s overseer.\evb\evg


\bvg\bva Ręið es \alst{s}itjandi \alst{s}ę́la \hld\ ok \alst{s}núðig fęrð &
\ind ok \alst{jó}s \alst{ę}rfiði.\eva

\bvb Chariot is sitting bliss and twirling journey \\
and horse’s heavy work.\evb\evg


\bvg\bva Kaun es \alst{b}arna \alst{b}ǫl \hld\ ok \alst{b}ar-dagi &
\ind ok \alst{h}old-fúa \alst{h}ús.\eva

\bvb Boil is children’s curse and TODO \\
and house of flesh-rot.\evb\evg


\bvg\bva Hagall es \alst{k}alda \alst{k}orn \hld\ ok \alst{k}nappa drífa &
\ind ok \alst{s}náka \alst{s}ótt.\eva

\bvb Hail is cold kernel and storm of beads \\
and sickness of snakes.\evb\evg


\bvg\bva Nauð es \alst{þ}ýjar \alst{þ}rǫ́ \hld\ ok \alst{þ}ungr kostr &
\ind ok \alst{v}ás-samlig \alst{v}erk.\eva

\bvb Need is maidservant’s yearning and scant choice \\
and working in wet-cold weather.\evb\evg


\bvg\bva \alst{Í}ss es \alst{á}ar bǫrkr \hld\ ok \alst{u}nnar þękja &
\ind ok \alst{f}ęigra manna \alst{f}ár.\eva

\bvb Ice is river’s bark and wave’s roof \\
and fey men’s danger.\evb\evg


\bvg\bva Ár es \alst{g}umna \alst{g}óði \hld\ ok \alst{g}ótt sumar &
\ind \emph{ok} \alst{a}l-gróinn \alst{a}kr.\eva

\bvb Year is the boon of men and good summer \\
(and) all-grown acre.\evb\evg


\bvg\bva Sól es \alst{sk}ýja \alst{sk}jǫldr \hld\ ok \alst{sk}ínandi rǫðull &
\ind ok \alst{í}sa \alst{a}ldr-tregi.\eva

\bvb Sun is the shield of clouds and shining wheel \\
and life-grief of ice.\evb\evg


\bvg\bva Týr es \alst{ęi}n-hęndr \alst{ǫ́}ss \hld\ ok \alst{u}lfs lęifar &
\ind ok \alst{h}ofa \alst{h}ilmir.\eva

\bvb Tew is the one-handed Os and the wolf’s leftovers \\
and lord of hoves.\evb\evg


\bvg\bva Bjarkan es \alst{l}aufgat \alst{l}im \hld\ ok \alst{l}ítit tré &
\ind ok \alst{u}ng-samligr \alst{v}iðr.\eva

\bvb Birch is leafy branch and little tree \\
and youthful wood.\evb\evg


\bvg\bva \alst{M}aðr es \alst{m}anns gaman \hld\ ok \alst{m}oldar auki &
\ind ok \alst{sk}ipa \alst{sk}ręytir.\eva

\bvb Man is man’s joy and the product of dust \\
and adorner of ships.\evb\evg


\bvg\bva Lǫgr es \alst{v}ellanda \alst{v}atn \hld\ ok \alst{v}íðr kętill &
\ind ok \alst{g}lǫmmungr \alst{g}rund.\eva

\bvb Liquid is boiling watter and wide kettle \\
and TODO.\evb\evg


\bvg\bva Ýr es \alst{b}ęndr bogi \hld\ ok \alst{b}rot-gjarnt járn &
\ind ok \alst{f}ęnju \alst{f}lęygir.\eva

\bvb Yew is a bent bow and easily broken iron \\
and arrow’s hurler.\evb\evg

\sectionline
