\bookStart{Second Lay of Hallow Hundingsbane}[Helgakviða Hundingsbana aðra]

BPG
BPA Helgi fekk Sigrúnar ok áttu þau sonu; var Helgi eigi gamall. Dagr Hǫgna sonr blótaði Óðin til fǫðurhefnda. Óðinn léði Dag geirs síns. Dagr fann Helga, mág sinn, þar sem heitir at Fjǫturlundi. Hann lagði í gǫgnum Helga með geirnum. Þar fell Helgi en Dagr reið til fjalla ok sagði Sigrúnu tíðindi:

BPB Hallow got Sighrun, and they owned sons; Hallow was not old. Day, son of Hain, \inx[C]{bloot}[blooted] to Weden to take revenge for his father. Weden lent Day his spear. Day found Hallow, his brother-in-law, at a place called Fetterlund; he laid the spear through Hallow. There fell Hallow, but Day rode to the fells and told Sighrun the news:
EPB


\bvg
\bva „Trauðr em ek, systir, \hld\ trega þér at sęgja &
þvíat ek hęfi nauðigr \hld\ nipti grę́tta: &
Fell í morgun \hld\ und Fjǫturlundi &
buðlungr sá’s vas \hld\ bęztr í hęimi &
ok hildingum \hld\ á halsi stóð.“\eva

\bvb “Regretful am I, sister, to grieve thee by saying—for, forced must I cause my kinswoman to cry: This morning fell, ’neath Fetterlund, that prince who was in the world the best, and on the throats of rulers stood.”\evb
\evg

...

\bvg
\bva „Fyrr vil’k kyssa \hld\ konung ólifðan &
an þú blóðugri \hld\ brynju kastir; &
hár es þitt, Helgi, \hld\ hélu þrungit, &
allr es vísi \hld\ valdǫgg slęginn, &
hęndr úrsvalar \hld\ Hǫgna mági; &
hvé skal’k þér, buðlungr, \hld\ þess bót of vinna?“\eva

\bvb “Sooner would I kiss the unliving king, than thou the bloody byrnie mightst cast away. Thy hair is, Hallow, with hoarfrost thick: the prince is all with corpse-dew whipped:\footnoteB{For the formulation cf. \Baldrsdraumar\ 5.} the hands wet-cold on the kinsman of Hain. How shall I for thee, lord, remedy that?”\evb
\evg


\bvg
\bva „Ęin vęldr þú, Sigrún \hld\ frá Sefafjǫllum, &
es Hęlgi es \hld\ harmdǫgg slęginn: &
Grę́tr þú, gullvarit, \hld\ grimmum tǫ́rum, &
sólbjǫrt suðrǿn, \hld\ áðr þú sofa gangir, &
hvęrt fęllr blóðugt \hld\ á brjóst grami, &
úrsvalt, innfjalgt \hld\ ękka þrungit.“\eva

\bvb “Thou alone causest, Sighrun from the Sevefells, that Hallow be by harm-dew whipped; thou criest, gold-covered, bitter tears, sun-bright southern lady, before thou to sleep mightst go. Each one falls bloody on the breast of the ruler, wet-cold and stifled, pressed forth by grief.”\evb
\evg
