\bookStart{The Speeches of Shirner}[Skírnismǫ́l]

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: C10th (0.897)

\textbf{Meter:} \Ljodahattr, \Galdralag\ (TODO)%
\end{flushright}

% Introduction

The whole poem is attested in both \Regius\ and \AM. The name \emph{Skírnismǫ́l} ‘\textbf{Speeches of Shirner}’ comes from \AM; \Regius\ has in the typical titular red ink \emph{Fǫr Skírnis} ‘Shirner’s journey’.

The same myth is told in prose in \Gylfaginning\ 37.  A single stanza of the present poem is quoted there, namely the last one, with some minor differences in wording that would seem to stem from oral tradition (see Note to st. 42 below).  It is unlikely that the author of \Gylfaginning\ knew of the narrative through an oral tradition which included only the last verse, chiefly since his paraphrase does not add a single detail not found in the present poem, but on the other hand condenses and abbreviates.  So, Shirner’s journey and curse (roughly sts. 10–38 here) is simply summarized in the following manner: “Then Shirner journeyed and requested the woman [i.e. Gird] for him [i.e. Free], and received her promise, that nine nights later she would come to the place which is called Barrey, and have a wedding with Free.”  The summarising of a narrative mythic poem with a single verse quotation in the form of a dialogue-stanza is something done several times in \Gylfaginning; see Eddic fragments from Snorre’s Edda below.

On the other hand, the paragraph in \Gylfaginning\ 37 corresponding to what is here P1 is much more detailed and reads: “Gymer was a man called, and his woman Earbode; she was of the lineage of mountain-risers. Their daughter is Gird, who is fairest of all women.  It was one day when Free had gone to Lithshelf and looked about all the Homes, but when he looked to the north he saw on a farm a great and fine house, and to that house walked a woman, and when she lifted her hands and closed the doors before her, then it did shine from her hands both into the air and onto the waters, and all the homes were brightened by her.  And that beauty, which he had seen in that holy seat, harmed him so that he walked away filled with pain, and when he came home he spoke nothing; he neither slept nor drank; nobody dared to get words out of him.  Then Nearth had Shirner, Free’s shoe-swain, called unto himself, and asked him to go to Free and ask him to speak, [...]”

% It seems to me that this circumstance, where the part corresponding to the poem is a short paraphrase, but the part corresponding to the prose passage is much more detailed, can only have arisen if the former already had a fixed form, whereas the latter was freer and could vary with each retelling. For this, see further TODO.

\sectionline

\bpg
\bpa\mssnote{\Regius~11r/10, \AM~2r/11}%
Freyr, sonr Njarðar, hafði einn dag setsk í Hlið-skjálf ok sá um heima alla; hann sá í Jǫtun-heima ok sá þar mey fagra, þá er hon gekk frá skála fǫður síns til skemmu; þar af fekk hann hug-sóttir miklar. Skírnir hét skó-sveinn Freys. Njǫrðr bað hann kveðja Frey máls. Þá mę́lti Skaði:\epa

\bpb \inx[P]{Free}, son of \inx[P]{Nearth}, had one day set himself in \inx[L]{Lithshelf} and looked about all the \inx[C]{Homes}.  He looked into the \inx[L]{Ettinhomes} and saw there a fair maiden as she walked from her father’s hall to her bower; thereof he got great heart-aches.  \inx[P]{Shirner} was called the shoe-swain of Free.  Nearth asked him to speak with Free.  Then \inx[P]{Shede} spoke:\epb\epg


\bvg\bva\mssnote{\Regius~11r/14, \AM~2r/15}%
„\edtext{Rís-tu nú Skírnir \hld\ ok gakk at bęiða}{\lemma{rís \dots\ bęiða ‘Rise \dots\ ask’}\Bfootnote{Alliteration is missing here. A simple solution would be to replace \emph{gakk} ‘go’ with a synonym like \emph{rinn} ‘run’ or \emph{ráð} ‘resolve’, but this lessens the semantic mirroring with l. 2/2 below (though, the insertion of the verb \emph{ganga} in the present stanza may in fact be due to influence from 2/2).}} &
\ind okkarn \alst{m}ála \alst{m}ǫg, &
ok þess at \alst{f}regna \hld\ hvęim hinn \alst{f}róði séi &
\ind \alst{o}f-ręiði \edtrans{\alst{a}fi}{man}{\Bfootnote{While this word usually means ‘father’ or ‘grandfather’, it must here certainly mean ‘man’ without a connotation of old age. See further \CV.}}.“\eva

\bvb “Rise thou now, Shirner, and go to ask \\
our lad \ken*{= Free} for speech; \\
and to learn at whom the wise \\
man \ken*{= Free} might be cross.”\evb\evg

\bvg\bva\speakernote{Skírnir kvað:}\mssnote{\Regius~11r/15, \AM~2r/17}%
„\alst{I}llra \alst{o}rða \hld\ es mér \alst{ó}n at ykkrum syni, &
\ind ef ek gęng at \alst{m}ę́la við \alst{m}ǫg, &
ok þess at \alst{f}regna, \hld\ hvęim hinn \alst{f}róði séi &
\ind \alst{o}f-ręiði \alst{a}fi.“\eva

\bvb Shirner quoth: “Bad words I expect from your son \ken*{= Free},  \\
if I go with the lad to speak; \\
and to learn at whom the wise \\
man might be cross.”\evb\evg

\sectionline

\bvg\bva\speakernote{Skírnir:}\mssnote{\Regius~11r/17, \AM~2r/18}%
„Sęg þat \alst{F}ręyr, \hld\ \alst{f}olk-valdi goða, &
\ind ok ek \alst{v}ilja \alst{v}ita, &
hví þú \alst{ęi}nn sitr \hld\ \alst{ę}nd-langa sali, &
\ind minn \alst{d}róttinn, of \alst{d}aga?“\eva

\bvb Shirner [quoth]: “Tell it, O Free, troop-wielder of the gods; \\
I too would wish to know: \\
why thou sittest alone in the endlong halls, \\
my lord, during the days?”\evb\evg


\bvg\bva\speakernote{Fręyr:}\mssnote{\Regius~11r/19, \AM~2r/20}%
„Hví of \alst{s}ęgja’k þér, \hld\ \alst{s}ęggr hinn ungi, &
\ind \alst{m}ikinn \alst{m}óð-trega? &
því-at \alst{a}lf-rǫðull \hld\ lýsir of \alst{a}lla daga &
\ind ok þęygi at \alst{m}ínum \alst{m}unum.“\eva

\bvb Free [quoth]: “Why should I tell thee, O young youth, \\
{[of my]} great mood-grief? \\
For the elf-wheel \ken{sun} shines during all days, \\
and naught to my liking.”\evb\evg


\bvg\bva\speakernote{Skírnir:}\mssnote{\Regius~11r/20, \AM~2r/21}%
„\alst{M}uni þína \hld\ hykk-a svá \alst{m}ikla vesa, &
\ind at þú mér \edtrans{\alst{s}ęggr}{youth}{\Bfootnote{This word usually means simply ‘man’, but it seems to have a specific connotation with youth. Its original meaning is ‘messenger’, and the semantic shift is thus: ‘messenger’ > ‘young man’ > ‘warrior/man’. The sense of ‘young man’ is also seen in \Volundarkvida\ 23, where it is used in reference to king Nithad’s two young sons. In the present stanza it answers Free’s addressing Shirner as \emph{sęggr hinn ungi} ‘the young youth’; Shirner points out that the two are of equal age, and so Free is as much of a young man as he.}} né \alst{s}ęgir; &
\alst{u}ngir saman \hld\ vǫ́rum í \alst{á}r-daga, &
\ind vęl mę́ttim \alst{t}vęir \alst{t}rúask.“\eva

\bvb Shirner [quoth]: “Thy liking I do not think so great, \\
that thou, O youth, should not tell me [of it]. \\
Young together were we in days of yore; \\
we two might well trust each other.”\evb\evg


\bvg\bva\speakernote{Fręyr:}\mssnote{\Regius~11r/22, \AM~2r/23}%
„Í \alst{G}ymis gǫrðum \hld\ ek \alst{g}anga sá &
\ind \alst{m}ér tíða \alst{m}ęy; &
\alst{a}rmar lýstu, \hld\ en \alst{a}f þaðan &
\ind allt \edtrans{\alst{l}opt ok \alst{l}ǫgr}{air and sea}{\Bfootnote{Formulaic and very old, also paralleled in the Anglo-Saxon. TODO.}}.\eva

\bvb Free [quoth]: “In Gymer’s yards I saw walking \\
a maiden, dear to me. \\
The arms shone, but thereof \\
all the air and sea.\evb\evg


\bvg\bva\mssnote{\Regius~11r/24, \AM~2r/24}%
\alst{M}ę́r ’s mér tíðari \hld\ an \alst{m}anna hvęim &
\ind \alst{u}ngum í \alst{á}r-daga; &
\alst{á}sa ok \alst{a}lfa \hld\ þat vill \alst{ę}ngi maðr, &
\ind at vit \alst{s}átt \alst{s}éim.“\eva

\bvb The maiden is dearer to me than to any man \\
young in days of yore. \\
Of the \inx[F]{Eese and Elves} does no man\footnoteB{i.e. ‘person’. For other examples of gods being called men see note to final st. of \Vafthrudnismal\ (TODO).} wish \\
that we two should be brought together.”\evb\evg


\bvg\bva\speakernote{Skírnir:}\mssnote{\Regius~11r/25, \AM~2r/25}%
„\alst{M}ar gef mér þá, \hld\ es mik of \alst{m}yrkvan beri &
\ind \alst{v}ísan \alst{v}afr-loga, &
ok þat \alst{s}verð, \hld\ es \alst{s}jalft vegisk &
\ind við \alst{jǫ}tna \alst{ę́}tt.“\eva

\bvb Shirner [quoth]: “The steed then give me, which might bear me over the dark, \\
wise wavering-flame; \\
and that sword, which by itself might strike \\
against the line of the \inx[G]{Ettins}.”\evb\evg


\bvg\bva\speakernote{Fręyr:}\mssnote{\Regius~11r/27, \AM~2r/27}%
„\alst{M}ar þér þann gef’k, \hld\ es þik of \alst{m}yrkvan \edtext{berr &
\ind \alst{v}ísan \alst{v}afr-loga, &
auk þat \alst{s}verð, \hld\ es \alst{s}jalft mun vegask, &
\ind ef sá ’s \alst{h}orskr es \alst{h}ęfr.“}{\lemma{berr ‘bears’; mun vegask, ef sá ’s horskr es hęfr ‘will strike, if he is wise who owns it’}\Bfootnote{In his response Free replaces the subjunctive verb forms (\emph{beri} ‘might bear’, \emph{vegisk} ‘might strike’) with indicative and future forms, giving a sense of certainity and authority. The steed and sword are faultless, and if Shirner fails on the mission, it would be only due to his own fault (“if he is sharp who owns it.”).}}\eva
%TODO? Change the line numbering from 1–4 to 1, 3–4.

\bvb Free [quoth]: “That steed I give thee, which bears thee over the dark, \\
wise wavering-flame; \\
and that sword which by itself will strike, \\
if he is wise who owns it.”\evb\evg


\bpg\bpa Skírnir mę́lti við hest’inn:\epa
\bpb Shirner spoke with the horse:\epb\epg


\bvg\bva\mssnote{\Regius~11r/29, \AM~2r/28}%
„\alst{M}yrkt es úti, \hld\ \alst{m}ál kveð’k okkr fara &
\ind \alst{ú}rig fjǫll \alst{y}fir &
\ind \edtrans{\alst{þ}ursa}{of the Thurses}{\Afootnote{so \AM; \emph{þyria} \Regius}} \alst{þ}jóð yfir; &
\alst{b}áðir vit komumk \hld\ eða okkr \alst{b}áða tękr &
\ind sá hinn \edtrans{\alst{á}m-átki \alst{jǫ}tunn}{uncanny ettin}{\Bfootnote{Formulaic. See note to \Voluspa\ 8.}}.“\eva

\bvb “’Tis dark outside; I declare it time for us to journey \\
over the drizzling mountains, \\
over the tribe of the \inx[G]{Thurses}. \\
Both two [shall] we come [over], or us both does take \\
that uncanny ettin.\footnoteB{Shirner declares his intention not to abandon the horse given to him by his lord; they will either both make it, or both perish.}”\evb\evg


\bpg
\bpa\mssnote{\Regius~11r/31, \AM~2v/1}%
Skírnir reið i Jǫtun-heima til Gymis garða; þar váru hundar ólmir ok bundnir fyrir skíð-garðs hliði þess, er um sal Gerðar var. Hann reið at þar, er fé-hirðir sat á haugi, ok kvaddi hann: \epa

\bpb Shirner rode into the Ettinhomes, to Gymer’s yards. There were fierce hounds bound in front of the slope of the wooden fence which surrounded Gird’s\footnoteB{It is first now that we are informed of the maiden’s name.} hall. He rode to where a shepherd sat on a mound, and greeted him:\epb\epg


\bvg\bva\mssnote{\Regius~11v/2, \AM~2v/4}%
„Sęg þat \alst{h}irðir, \hld\ es á \alst{h}augi sitr &
\ind ok \alst{v}arðar alla \alst{v}ega: &
hvé ek at \alst{a}nd-spilli \hld\ komumk hins \alst{u}nga mans &
\ind fyr \alst{g}ręyjum \alst{G}ymis.“\eva

\bvb “Say this, O herdsman, who on the mound dost sit, \\
and watchest all the ways: \\
How I to discourse might come with the young girl \ken*{= Gird}, \\
past the greyhounds of Gymer?”\evb\evg


\bvg\bva\speakernote{[Hirðir] kvað:}\mssnote{\Regius~11v/4, \AM~2v/5}%
„Hvárt est \alst{f}ęigr, \hld\ eða est \alst{f}ramm ginginn &
\ind [...]; &
\alst{a}nd-spillis vanr \hld\ þú skalt \alst{ę́} vesa &
\ind \edtrans{\alst{g}óðrar męyjar}{good maiden}{\Bfootnote{Formulaic, carrying with it a sense of chastity.  See note to \Havamal\ 102/1 for further occurrences.}} \alst{G}ymis.“\eva

\bvb {[The herdsman]} quoth: \\
“Either art thou fey, or gone forth \ken{dead}; \\
{[...]}. \\
Discourse-less shalt thou always be, \\
with the good maiden of Gymer \ken*{= Gird}.”\evb\evg


\bvg\bva\speakernote{[Skírnir] kvað:}\mssnote{\Regius~11v/6, \AM~2v/7}
„\edtrans{\alst{K}ostir}{Choices}{\Bfootnote{i.e. ‘alternatives, other ways’.}} ’ru bętri \hld\ \edtrans{an}{than}{\Afootnote{so \AM; \emph{hęldr an at} ‘rather than to [be]’ \Regius}} \alst{k}løkkva séi &
\ind hvęim es \alst{f}úss es \alst{f}ara, &
\alst{ęi}nu dǿgri \hld\ mér vas \alst{a}ldr of skapaðr &
\ind ok alt \alst{l}íf of \alst{l}agit.“\eva

\bvb {[Shirner]} quoth: \\
“Choices are better than sobbing might be \\
for whomever is eager to journey. \\
In one half-day my age was shaped, \\
and all my life laid down.\footnoteB{An excellent example of the fatalistic Germanic worldview, in which one’s course of life was determined (“laid down”) at birth (“in one half-day”).  Presumably after uttering these words Shirner rides through the fire surrounding the fortress. — The causative \emph{lęgja} ‘to lay (down, in place)’ is closely connected to fate; the expression is formulaic.  Cf. \Lokasenna\ 48: \emph{í ár-daga vas þér hit ljóta líf of lagit} ‘in days of yore was thy ugly life laid down’ and \Voluspa\ 19: \emph{þę́r lǫg lǫgðu} ‘they [= the Norns] laid down laws’.}”\evb\evg


\bvg\bva\speakernote{[Gęrðr] kvað:}\mssnote{\Regius~11v/7, \AM~2v/8}%
„Hvat ’s þat \alst{h}lym \alst{h}lymja \hld\ es \alst{h}lymja hęyri’k nú til &
\ind \alst{o}ssum rǫnnum \alst{í}? &
\alst{jǫ}rð bifask, \hld\ en \alst{a}llir fyr &
\ind skjalfa \alst{g}arðar \alst{G}ymis.“\eva

\bvb {[Gird]} quoth: \\
“What is that din of dins, which I of dins now hear \\
in our halls? \\
The earth quakes, but before [me] tremble \\
all Gymer’s yards.”\evb\evg


\bvg\bva\speakernote{Ambǫ́tt kvað:}\mssnote{\Regius~11v/9, \AM~2v/10}%
„\alst{M}aðr ’s hér úti, \hld\ stiginn af \alst{m}ars baki, &
\ind \alst{jó} lę́tr til \alst{ja}rðar taka.“\eva

\bvb A servant-woman quoth: \\
“A man is here outside, stepped down off horseback; \\
he lets take his steed to the ground.\footnoteB{According to \textcite{FinnurEdda} a still known (in his time) Icelandic expression; Shirner lets his horse graze.}”\evb\evg


\bvg\bva\speakernote{[Gęrðr] kvað:}\mssnote{\Regius~11v/10, \AM~2v/11}%
„\alst{I}nn bið þú hann ganga \hld\ í \alst{o}kkarn sal &
\ind ok drekka hinn \alst{m}ę́ra \alst{m}jǫð, &
þó ek hitt \alst{ó}umk, \hld\ at hér \alst{ú}ti séi &
\ind minn \alst{b}róður-\alst{b}ani.“\eva

\bvb {[Gird]} quoth: \\
“Bid thou him to go in into our hall, \\
and to drink the renowned mead; \\
though I fear that here outside should be  \\
my brother’s bane.”\evb\evg

\sectionline

\bvg\bva\speakernote{[Gęrðr] kvað:}\mssnote{\Regius~11v/12, \AM~2v/13}%
„Hvat ’s þat \alst{a}lfa \hld\ né \alst{á}sa sona, &
\ind né \alst{v}íssa \alst{v}ana; &
hví \alst{ęi}nn of komt \hld\ \alst{ęi}kinn fúr yfir &
\ind ór \alst{s}al-kynni at \alst{s}éa?“\eva

\bvb {[Gird quoth:]} \\
“What sort is that, not of Elves, nor of sons of the Eese, \\
nor of wise Wanes? \\
Why camest thou alone over the raging fire, \\
to see the state of our hall?”\evb\evg


\bvg\bva\speakernote{[Skírnir kvað:]}\mssnote{\Regius~11v/14}%
„\alst{E}m’k-at \alst{a}lfa \hld\ né \alst{á}sa sona &
\ind né \alst{v}íssa \alst{v}ana, &
þó \alst{ęi}nn of kom’k \hld\ \alst{ęi}kinn fúr yfir &
\ind yður \alst{s}al-kynni at \alst{s}éa.\eva

\bvb {[Shirner quoth:]} \\
“I am not of Elves, nor of sons of the Eese, \\
nor of wise Wanes— \\
yet I came alone over the raging fire, \\
to see the state of your hall.\evb\evg


\bvg\bva\mssnote{\Regius~11v/15, \AM~2v/14}%
\alst{Ę}pli \alst{ę}llifu \hld\ hér hef’k \alst{a}l-gullin, &
\ind þau mun’k þér \alst{G}ęrðr \alst{g}efa, &
\alst{f}rið at kaupa, \hld\ at þú þér \alst{F}ręy kveðir &
\ind ó·\alst{l}ęiðastan at \alst{l}ifa.“\eva

\bvb Eleven apples have I here, all-golden; \\
those I will to thee, O Gird, give \\
to buy [thy] love, that thou callest Free for thee \\
most unloathsome [lovely] in life.\footnoteB{\emph{at lifa} here means seems to mean ‘in life/living’ rather than the typical infinitive sense ‘to live’; cf. st. 22 \emph{at dęila} ‘in sharing’ below. This is possibly an archaism.}”\evb\evg


\bvg\bva\speakernote{[Gęrðr] kvað:}\mssnote{\Regius~11v/17, \AM~2v/15}%
„\alst{Ę}pli \alst{ę}llifu \hld\ ek þigg \alst{a}ldri-gi &
\ind at \alst{m}anns-kis \alst{m}unum, &
né vit \alst{F}ręyr, \hld\ meðan okkart \alst{f}jǫr lifir, &
\ind \alst{b}yggum \alst{b}ę́ði saman.“\eva

\bvb {[Gird quoth:]} \\
“Eleven apples will I never take, \\
to any man’s liking; \\
nor will I and Free—while our lifeblood lives— \\
dwell both together.”\evb\evg


\bvg\bva\speakernote{[Skírnir kvað:]}\mssnote{\Regius~11v/19, \AM~2v/17 (ll. 1–2)}%
„\alst{B}aug þér þá gef’k, \hld\ þann’s \alst{b}ręndr of vas &
\ind með \alst{u}ngum \alst{Ó}ðins syni; &
\edtext{\alst{á}tta ’ru \alst{ja}fn-hǫfgir, \hld\ es \alst{a}f drjúpa &
\ind hina \alst{n}íundu hvęrja \alst{n}ǫ́tt.“}{\lemma{átta ... nǫ́tt ‘Eight ... night.’}\Bfootnote{In \AM\ these lines and 22:1–2 are missing.  Instead 1–2 here and 22:3–4 are combined into one.}}\eva

\bvb {[Shirner quoth:]} \\
“The \inx[C]{bigh} I then give thee, it which was burned \\
with Weden’s young son \ken*{= Balder}. \\
Eight are even-heavy, which from it drip, \\
every ninth night.\footnoteB{The bigh, while not named, is clearly Dreepner as known from \Gylfaginning\ 49, describing Balder’s funeral: “Weden laid on the pyre that gold ring which is called Dreepner. Its nature was such that every ninth night, eight even-heavy golden rings dripped from it.” When \inx[P]{Harmod} later comes to \inx[L]{Hell} to try to bring Balder back, Balder tells him to bring the ring back to Weden, as a token of memory.}”\evb\evg


\bvg\bva\speakernote{[Gęrðr] kvað:}\mssnote{\Regius~11v/21, \AM~2v/18 (ll. 3–4)}%
„\alst{B}aug þikk-a’k, \hld\ þótt \alst{b}ręndr séi, &
\ind með \alst{u}ngum \alst{Ó}ðins syni; &
es-a mér \alst{g}ulls vant \hld\ í \alst{g}ǫrðum \alst{G}ymis &
\ind at dęila \alst{f}é \alst{f}ǫður.“\eva

\bvb {[Gird quoth:]} \\
“The bigh I take not, though it may have been burned \\
with Weden’s young son \ken*{= Balder}; \\
I have no want of gold in Gymer’s yards, \\
in sharing the \inx[C]{fee} of my father.”\evb\evg


\bvg\bva\speakernote{[Skírnir kvað:]}\mssnote{\Regius~11v/23, \AM~2v/19}%
„Sér þú \alst{m}ę́ki, \alst{m}ę́r, \hld\ \alst{m}jóvan, \edtrans{\alst{m}ál-fáan}{picture-painted}{\Bfootnote{The sword is inlaid with metal (perhaps gold or silver) forming a pattern.  The expression is formulaic; cf. TODO.}}, &
\ind es \alst{h}ęf’k í \alst{h}ęndi \alst{h}ér? &
\alst{h}ǫfuð \alst{h}ǫggva \hld\ mun’k þér \alst{h}alsi af, &
\ind nema mér \alst{s}ę́tt \alst{s}ęgir.“\eva

\bvb {[Shirner quoth:]} \\
“Seest thou this sword, O maiden—slender, pictured-painted—, \\
which I have in my hand here? \\
Hew the head will I, off thy neck, \\
unless thou come to terms with me.”\evb\evg


\bvg\bva\speakernote{[Gęrðr kvað:]}\mssnote{\Regius~11v/25, \AM~2v/20}%
„\alst{Á}-nauð þola \hld\ vil’k \alst{a}ldri-gi &
\ind at \edtrans{\alst{m}anns-kis}{any man’s (lit. ‘no man’s)}{\Afootnote{\emph{manns ęnskis} \AM}} \alst{m}unum, &
þó hins \alst{g}et’k, \hld\ ef it \alst{G}ymir finniðsk &
\alst{v}ígs ó·trauðir \hld\ at ykkr \alst{v}ega tíði.“\eva

\bvb {[Gird quoth:]} \\
“Stand coercion will I never, \\
to any man’s liking; \\
though I get this, if thou and Gymer meet— \\
men unreluctant of conflict—that ye two will wish to fight.\footnoteB{Gird says that she will never let herself be forced to marry Free, even if that means that her father and Shirner should fight over her.}”\evb\evg


\bvg\bva\speakernote{[Skírnir kvað:]}\mssnote{\Regius~11v/27, \AM~2v/22}%
„Sér þú \alst{m}ę́ki, \alst{m}ę́r, \hld\ \alst{m}jóvan, \alst{m}ál-fáan, &
\ind es \alst{h}ęf’k í \alst{h}ęndi \alst{h}ér? &
fyr þessum \alst{ę}ggjum \hld\ hnígr sá hinn \alst{a}ldni jǫtunn, &
\ind verðr þinn \alst{f}ęigr \alst{f}aðir.\eva

\bvb {[Shirner quoth:]} \\
“Seest thou this sword, O maiden—slender, pictured-painted—, \\
which I have in my hand here? \\
By these edges sinks the aged ettin \ken*{= Gymer} down; \\
\inx[C]{fey} becomes thy father.\evb\evg


\bvg\bva\mssnote{\Regius~11v/28, \AM~2v/24}%
\edtrans{\alst{T}ams-vęndi}{taming-wand}{\Bfootnote{Has been interpreted as a sword, TODO.}} þik drep’k, \hld\ ęn þik \alst{t}ęmja mun’k, &
\ind \alst{m}ę́r, at mínum \alst{m}unum, &
þar skalt \alst{g}anga \hld\ es þik \alst{g}umna synir &
\ind \alst{s}íðan ę́va \alst{s}éi.\eva

\bvb With the taming-wand I strike thee—and tame thee I will, \\
O maiden, to my liking! \\
There shalt thou go, where thee the sons of men \\
never since may see.\evb\evg


\bvg\bva\mssnote{\Regius~11v/30, \AM~2v/26}%
\edtrans{\alst{A}ra þúfu \alst{á} \hld\ skalt \alst{á}r sitja}{On an eagle’s perch shalt thou sit at dawn}{\Afootnote{\emph{ár skalt sitja \hld\ ara þúfu á} ‘at dawn shalt thou sit on an eagle’s perch’ \AM}}, &
\ind \edtext{\alst{h}orfa \alst{h}ęimi ór; &
\ind snugga \alst{h}ęljar til}{\lemma{horfa hęimi ór; snugga hęljar til ‘turn out of the world; hanker after Hell’}\Afootnote{\emph{horfa ok snugga hęljar til} ‘turn and hanker after Hell’ \AM}\Bfootnote{i.e. “you will look toward and yearn for the underworld”.}}; &
\alst{m}atr sé þér męir lęiðr \hld\ an \alst{m}anna hvęim &
\ind hinn \alst{f}ráni ormr með \edtext{\alst{f}irum}{\Bfootnote{This is the last word of fol. 2v of \AM, after which the text cuts off.}}.\eva

\bvb On an eagle’s perch shalt thou sit at dawn; \\
turn out of the world; \\
hanker after \inx[L]{Hell}. \\
Food be for thee more loathsome, than to any one \\
the gleaming serpent \ken*{= the Middenyardswyrm} among men.\footnoteB{Her food will be more disgusting than the \inx[C]{Middenyardswyrm}, for which cf. \Hymiskvida\ 22.}\evb\evg


\bvg\bva\mssnote{\Regius~11v/32}%
At \alst{u}ndr-sjónum verðir \hld\ es \alst{ú}t of kømr, &
\ind á þik \alst{H}rímnir \alst{h}ari &
\ind á þik \alst{h}ot-vetna stari, &
\alst{v}íð-kunnari \alst{v}erðir \hld\ an \alst{v}ǫrðr með goðum, &
\ind \alst{g}api þú \alst{g}rindum frá.\eva

\bvb A wondrous sight mayst thou become when thou comest out; \\
at thee may Rimner ogle; \\
at thee may anyone stare. \\
More widely known mayst thou become than the Watchman among the Gods \ken*{= Homedal}; \\
mayst thou gape from the gates.\evb\evg


\bvg\bva\mssnote{\Regius~12r/2}%
\edtrans{\alst{T}ópi ok ópi, \hld\ \alst{t}jǫsull ok ó·þoli}{Toop and woop, tessle and restlessness}{\Bfootnote{The first three of these four words are magic curse words without clear meaning; I have left them untranslated.}}, &
\ind vaxi þér \alst{t}ǫ́r með \alst{t}rega; &
\alst{s}ętsk þú niðr \hld\ en mun’k \alst{s}ęgja þér &
\ind \alst{s}váran \alst{s}ús-breka, &
\ind ok \alst{t}vinnan \alst{t}rega.\eva

\bvb Toop and woop, tarsle and restlessness— \\
may thy tears grow with grief! \\
Sit thyself down, and I will tell thee \\
a heavy roaring-breaker, \\
and a twined grief.\evb\evg


\bvg\bva\mssnote{\Regius~12r/3}%
Tramar \alst{g}nęypa \hld\ þik skulu \alst{g}ęrstan dag &
\ind \alst{jǫ}tna gǫrðum \alst{í}, &
til \alst{h}rím-þursa \alst{h}allar \hld\ þú skalt \alst{h}vęrjan dag &
\ind \alst{k}ranga \alst{k}osta-laus; &
\ind \alst{k}ranga \alst{k}osta-vǫn; &
\alst{g}rát at \alst{g}amni \hld\ skalt í \alst{g}ǫgn hafa &
\ind ok lęiða með \alst{t}ǫ́rum \alst{t}rega.\eva

\bvb Fiends shall pine thee during gloomy day, \\
in the yards of the Ettins. \\
To the hall of Rime-thurses shalt thou every day \\
crawl choice-less; \\
crawl choices-lacking. \\
Weeping for joy shalt thou have in exchange, \\
and nurse with tears [thy] grief.\evb\evg


\bvg\bva\mssnote{\Regius~12r/7}%
Með \alst{þ}ursi \alst{þ}rí-hǫfðuðum \hld\ \alst{þ}ú skalt ę́ nara &
\ind eða \alst{v}er-laus \alst{v}esa, &
\ind þitt \alst{g}ęð \alst{g}rípi; &
\ind þik \alst{m}orn \alst{m}orni &
ves þú sem \alst{þ}istill, \hld\ sá’s \alst{þ}runginn vas &
\ind í \alst{o}fan-verða \alst{ǫ́}nn.\eva

\bvb With a thurse three-headed shalt thou always subsist, \\
or be husband-less. \\
May thy senses seize; \\
may murrain mourn thee; \\
be thou like the thistle that was pressed \\
during highest harvest!\evb\evg


\bvg\bva\mssnote{\Regius~12r/9}%
Til \alst{h}olts ek gekk \hld\ ok til \alst{h}rás viðar &
\ind \edtrans{\alst{g}amban-tęin}{gombentoe}{\Bfootnote{Perhaps “curse-twig”.  A compound consisting of the very rare word \emph{gamban} ‘magic/curse?’ and \emph{tęinn} ‘twig, branch’ (cf. \emph{mistil-tęinn} ‘mistle-toe’).  This may be the stick on which the runic curse in st. 36 below should be carved, or it is to be identified with the \emph{tams-vǫndr} ‘taming-wand’ of st. 26 above.  Cf. \Havamal\ 152, which speaks about a runic curse carved on \emph{rótum rás viðar} ‘the roots of a raw/sappy tree’.}} at \alst{g}eta &
\ind \alst{g}amban-tęin ek \alst{g}at.\eva

\bvb To the wood I went, and to the raw/sappy tree, \\
the \inx[C]{gombentoe} for to get; \\
the gombentoe I got.\evb\evg


\bvg\bva\mssnote{\Regius~12r/10}%
\alst{R}ęiðr ’s þér Óðinn, \hld\ \alst{r}ęiðr ’s þér Ása-bragr, &
\ind þik skal \alst{F}ręyr \alst{f}íask, &
hin \alst{f}irin-illa mę́r, \hld\ en \alst{f}ingit hęfr &
\ind \alst{g}amban-ręiði \alst{g}oða.\eva

\bvb Wroth with thee is Weden; wroth with thee is Bray of the Eese \name{= Thunder}; \\
thee shall Free come to hate, \\
O most wicked maiden, if thou hast earned \\
the gomben-wrath of the gods.\evb\evg


\bvg\bva\mssnote{\Regius~12r/12}%
\alst{H}ęyri jǫtnar, \hld\ \alst{h}ęyri \alst{h}rím-þursar, &
\alst{s}ynir \alst{S}uttunga, \hld\ \alst{s}jalfir ás-liðar, &
hvé \alst{f}yrir býð’k, \hld\ hvé \alst{f}yrir banna’k &
\ind \alst{m}anna glaum \alst{m}ani, &
\ind \alst{m}anna nyt \alst{m}ani.\eva

\bvb Hear may Ettins, hear may Rime-thurses, \\
sons of Sutting, the very Os-Troops \ken*{= Eese} themselves,— \\
how I forbid, how I forban \\
men’s fellowship from the maid, \\
men’s joy from the maid!\evb\evg


\bvg\bva\mssnote{\Regius~12r/14}%
\alst{H}rím-grímnir hęitir þurs, \hld\ es þik \alst{h}afa skal &
\ind fyr \alst{n}á-grindr \alst{n}eðan, &
þar þér \alst{v}íl-męgir \hld\ á \alst{v}iðar rótum &
\ind \alst{g}ęita-hland \alst{g}efi; &
\alst{ǿ}ðri drykkju \hld\ fá þú \alst{a}ldri-gi, &
\ind \alst{m}ę́r, af þínum \alst{m}unum, &
\ind \alst{m}ę́r, at \alst{m}ínum \alst{m}unum.\eva

\bvb Rimegrimner is called the thurse who thee shall have \\
down beneath Nawgrind, \\
where the lads of toil \ken{thralls} on the roots of a tree, \\
goat-piss will give thee. \\
A finer drink do thou never get, \\
O maiden, against thy liking, \\
O maiden, to my liking!\evb\evg


\bvg\bva\mssnote{\Regius~12r/16}%
\edtrans{\alst{Þ}urs}{thurse}{\Bfootnote{Thurse is the name of the \textbf{þ}-rune (ᚦ); it is carved as part of the curse.}} ríst’k \alst{þ}ér \hld\ ok \edtrans{\alst{þ}ría stafi}{three staves}{\Bfootnote{Three runic letters (or phrases) representing the three following words (\emph{ęrgi} ‘queerness, degeneracy’ etc.).  The ritual practice of carving “three staves” is first found on the C7th Gummarp stone: \textbf{h\textsc{a}þuwol\textsc{a}fʀ s\textsc{a}te st\textsc{a}b\textsc{a} þri\textsc{a} fff} ‘Hathwolf placed three staves: fff’, where the \textbf{f}-rune (ᚠ) stands for its name \inx[C]{fee} (i.e. ‘wealth, cattle’) and is thus meant to bring wealth.}}, &
\ind \edtrans{\alst{ę}rgi ok \alst{ǿ}ði ok \alst{ó}·þola}{queerness and madness and restlessness}{\Bfootnote{Both \emph{ęrgi} ‘queerness, degeneracy’ and \emph{ó·þoli} ‘restlessness’ (here probably from strong lust) are found in the love magic charm on the rune stick B257 from Bryggen (edited below under Galders).  \emph{ęrgi} is also found in the curse-formula on the C7th Proto-Norse runestones from Stentoften and Björketorp.  See further introduction to B257.}}, &
svá ek þat \alst{a}f ríst \hld\ sem ek þat \alst{á} ręist, &
\ind ef gørask \alst{þ}arfar \alst{þ}ęss.“\eva

\bvb \inx[G]{Thurses}[Thurse] I carve for thee, and three staves: \\
\inx[C]{queerness} and madness and restlessness.— \\
So I carve it \emph{off}, like I carved it \emph{on}, \\
if there be need for that.\footnoteB{Shirner has carved the curse (which will make true the curse), but tells Gird that he will scrape it off if she accepts his demands. She promptly does.}”\evb\evg


\bvg\bva\speakernote{[Gęrðr kvað:]}\mssnote{\Regius~12r/19}%
„\edtext{\alst{H}ęill ves þú \alst{h}ęldr, svęinn, \hld\ ok tak við \alst{h}rím-kalki &
\ind \alst{f}ullum \alst{f}orns mjaðar,}{\lemma{Hęill \dots\ mjaðar ‘Hale \dots\ mead’}\Bfootnote{Formulaic; the same lines occur in \Lokasenna\ 53.}} &
þó hafða’k \alst{ę́}tlat, \hld\ at mynda’k \alst{a}ldri-gi &
\ind unna \edtrans{\alst{v}aningja}{the Waning \ken*{= Free}}{\Bfootnote{lit. ‘descendant of the \inx[G]{Wanes}’.  A rare word.  Its only other occurence in the Norse corpus is in a \inx[C]{thule} of boar-names.  Boars were sacred to Free, TODO.}} \alst{v}ęl.“\eva

\bvb {[Gird quoth:]} \\
“Hale be thou rather, O swain, and receive the rime-chalice, \\
full of ancient mead— \\
though I had intended that I never would \\
love the Waning \ken*{= Free} well.”\evb\evg


\bvg\bva\speakernote{[Skírnir kvað:]}\mssnote{\Regius~12r/21}%
„\alst{Ø}rendi mín \hld\ vil’k \alst{ǫ}ll vita, &
\ind áðr ríða’k \alst{h}ęim \alst{h}eðan, &
nę́r á \alst{þ}ingi \hld\ munt hinum \alst{þ}roska &
\ind \alst{n}ęnna \alst{N}jarðar syni.“\eva

\bvb {[Shirner quoth:]} \\
“My errands all I wish to know, \\
before I ride home hence: \\
when on the \inx[C]{Thing} wilt thou with the vigorous \\
son of Nearth \ken*{= Free} be joined?”\evb\evg


\bvg\bva\speakernote{[Gęrðr kvað:]}\mssnote{\Regius~12r/23}%
„\alst{B}arri hęitir, \hld\ es vit \alst{b}ę́ði vitum, &
\ind \alst{l}undr \alst{l}ogn-fara, &
en ępt \alst{n}ę́tr \alst{n}íu, \hld\ þar mun \alst{N}jarðar syni &
\ind \alst{G}ęrðr unna \alst{g}amans.“\eva

\bvb {[Gird quoth:]} \\
“Barrey is called—as we both know— \\
a grove of calm rushes, \\
and after nine nights there will to the son of Nearth \\
Gird her pleasure grant.”\evb\evg


\bpg\bpa\mssnote{\Regius~12r/24}%
Þá reið Skírnir heim. Freyr stóð úti ok kvaddi hann ok spurði tíðenda:\epa

\bpb Then Shirner rode home. Free stood outside and greeted him and asked for the tidings:\epb\epg


\bvg\bva\mssnote{\Regius~12r/25}%
„\alst{S}ęg mér, Skírnir, \hld\ áðr verpir \alst{s}ǫðli af mar &
\ind ok stígir \alst{f}eti \alst{f}ramarr, &
hvat \alst{á}rnaðir \hld\ í \alst{Jǫ}tun-hęima &
\ind þíns eða \alst{m}íns \alst{m}unar?“\eva

\bvb “Tell me, O Shirner, before thou throw the saddle off the steed, \\
and take a step further: \\
what hast thou accomplished in the \inx[L]{Ettinhomes}, \\
to thy or my liking?”\evb\evg


\bvg\bva\speakernote{[Skírnir kvað:]}\mssnote{\Regius~12r/27}%
„\alst{B}arri hęitir, \hld\ es vit \alst{b}áðir vitum, &
\ind \alst{l}undr \alst{l}ogn-fara, &
en ępt \alst{n}ę́tr \alst{n}íu, \hld\ þar mun \alst{N}jarðar syni &
\ind \alst{G}ęrðr unna \alst{g}amans.“\eva

\bvb {[Shirner quoth:]} \\
“Barrey is called—as we both know— \\
a grove of calm rushes, \\
and after nine nights there will to the son of Nearth \\
Gird her pleasure grant.”\evb\evg


\bvg\bva\speakernote{[Fręyr kvað:]}\mssnote{\Regius~12r/28, \GylfMS}%
\alst{L}ǫng es nǫ́tt, \hld\ \edtrans{\alst{l}angar ’u tvę́r}{long are two}{\Afootnote{\emph{lǫng es ǫnnur} ‘long is another’ \GylfMS}}, &
\ind \edtext{hvé of \alst{þ}ręyja’k \alst{þ}ríar?}{\Afootnote{\emph{hvé męga’k þręyja þríar} \GylfMS}} &
opt \alst{m}ér \alst{m}ánaðr \hld\ \alst{m}inni þótti &
\ind an sjá \alst{h}ǫlf \alst{h}ý-nǫ́tt.\eva

\bvb {[Free quoth:]} \\
Long is a night; long are two; \\
how should I yearn for three? \\
Oft a month to me seemed less, \\
than this half wedding-night.\footnoteB{The wedding-night (TODO: it's a hapax so explain the etymology?) is presumably half as it is not consumated.}\evb\evg

\sectionline
