\bookStart{Fragment of a Lay of Siward}[Brot af Sigurðarkviða]

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: C10th (0.974)

\textbf{Meter:} \Fornyrdislag%
\end{flushright}

\section{Introduction}

A fragment of a longer lay about Siward and Byrnhild, following the lacuna.  According to the following prose (see \GudrunOne) the poem began with Siward’s death.  TODO: Translation is in progress.

\sectionline

\section{Fragment of a Lay of Siward}

\bvg\bva%
„\emph{hvat hęfir Sigurð} \hld\ saka unnit &
es þú frǿknan vill \hld\ fjǫrvi nę́ma?“\eva

\bvb “[What has Siward] done for a crime, \\
that thou wilt deprive the brave of life?”\evb\evg


\bvg\bva%
„Mér hęfir Sigurðr \hld\ sęlda ęiða &
ęiða sęlda \hld\ alla logna &
þȧ vélti hann mik \hld\ es hann vesa skyldi &
allra ęiða \hld\ ęinn full-trúi.“\eva

\bvb “To me has Siward given oaths, \\
oaths given, all lies. \\
He betrayed me when he should have been \\
of all oaths the one true keeper.”\evb\evg


\bvg\bva%
Þik hęfir Brynhildr \hld\ bǫl at gęrva &
hęiptar hvattan \hld\ harm at vinna. &
fyrr man hǫ́n Guðrúnu \hld\ góðra ráða &
enn síðan þér \hld\ sín at njóta.\eva

\bvb TODO: Translation.\evb\evg


\bvg\bva%
Sumir ulf sviðu, \hld\ sumir orm sniðu, &
sumir Gothormi \hld\ af gęra dęildu, &
áðr þęir mę́tti \hld\ męins of lystir &
á horskum hal \hld\ hęndr of lęggja.\eva

\bvb Some roasted a wolf; some cut up a snake; \\
some shared wolf-flesh with Godthorm, \\
TODO..\evb\evg


\bvg\bva%
Úti stóð Guðrún \hld\ Gjúka dóttir &
ok hǫ́n þat \alst{o}rða \hld\ \alst{a}lls fyrst of kvað: &
„Hvar es nú Sigurðr \hld\ sęggja dróttinn &
es frę́ndr mínir \hld\ fyrri ríða?“\eva

\bvb Outside stood Guthrun, Yivick’s daughter, \\
and she this word first of all did say: \\
“Where is now Siward, the lord of men, \\
when my kinsmen ride in front?”\evb\evg


\bvg\bva%
Ęinn því Hǫgni \hld\ and-svǫr vęitti: &
„Sundr hǫfum Sigurð \hld\ sverði hǫgginn; &
\alst{g}napir ę́ \alst{g}rár jór \hld\ yfir \alst{g}ram dauðum.“\eva

\bvb Alone did Hain this answer grant: \\
“We have cut Siward asunder by sword; \\
the grey steed always neighs over the dead prince.”\evb\evg


\bvg\bva%
Þȧ kvað þat \alst{B}rynhildr \hld\ \alst{B}uðla dóttir: &
„\alst{v}ęl skuluð njóta \hld\ \alst{v}ápna ok níu landa; &
\alst{ęi}nn \edtext{myndiv}{\Afootnote{\emph{myndiv} \Regius}} Sigurðr \hld\ \alst{ǫ}llu ráða &
ef hann \alst{l}ęngr \alst{l}ítlu \hld\ \alst{l}ifi hęldi.“\eva

\bvb Then quoth this Byrnhild, Budle’s daughter: \\
“Well shall ye enjoy weapons and nine lands! \\
Alone would Siward rule them all \\
if a little longer he had held his life.”\evb\evg


\bvg\bva%
„Vę́ri-a þat \alst{s}ǿmt \hld\ at hann \alst{s}vá réði &
\alst{G}júka arfi \hld\ ok \alst{g}ota męngi &
es hann \alst{f}imm sonu \hld\ at \alst{f}olk-rǿði &
\alst{g}unnar fu̇sa \hld\ \alst{g}etna hafði.“\eva

\bvb TODO: Translation.\evb\evg


\bvg\bva%
Hló þá \alst{B}rynhildr \hld\ —\alst{b}ǿr allr dunði— &
\alst{ęi}nu sinni \hld\ af \alst{ǫ}llum hug: &
„vęl skuluð njóta \hld\ landa ok þegna &
es þér \alst{f}rǿknan gram \hld\ \alst{f}alla létuð.“\eva

\bvb Then Byrnhild laughed—the farm all resounded— \\
a single time out of her whole heart: \\
“Well shall ye enjoy the lands and thanes, \\
since ye made the brave prince to fall.”\evb\evg


\bvg\bva%
Þȧ kvað þat Guðrún \hld\ Gjúka dóttir: &
„Mjǫk mę́lir þú \hld\ miklar firnar &
gramir hafi Gunnar \hld\ gǫtvað Sigurðar &
hęipt-gjarns hugar \hld\ hęfnt skal verða.“\eva

\bvb Then quoth this Guthrun, Yivick’s daughter: \\
“TODO.”\evb\evg


\bvg\bva%
Soltinn varð Sigurðr \hld\ sunnan Rínar &
hrafn at męiði \hld\ hátt kallaði: &
„Ykkr mun Atli \hld\ ęggjar rjóða &
munu víg-skáa \hld\ of viða ęiðar.“\eva

\bvb Dead was Siward to the south of the Rhine; \\
a raven on a branch loudly called out: \\
“On you two will Attle redden his blades; \\
the warriors will be destroyed by the oaths!”\evb\evg


\bvg\bva%
Framm vas kvelda \hld\ fjǫlð vas drukkit &
þȧ vas hví-vetna \hld\ vil-mál talit. &
sofnuðu allir \hld\ es ï sę́ing kvǫ́mu.\eva

\bvb TODO: Translation.\evb\evg


\bvg\bva%
\alst{Ęi}nn vakði Gunnarr \hld\ \alst{ǫ}llum lęngr &
\alst{f}ót nam at hrǿra \hld\ \alst{f}jǫlð nam at spjalla &
\alst{h}itt \alst{h}ęr-glǫtuðr \hld\ \alst{h}yggja téði, &
hvat þęir ï \alst{b}ǫðvi \hld\ \alst{b}áðir sǫgðu &
\alst{h}rafn ęy ok ǫrn \hld\ es þęir \alst{h}ęim riðu.\eva

\bvb Alone did Guther wake longer than all; \\
his foot he took to move, much he took to speak. \\
Of that the army-destroyer thought: \\
what in the fray they both had said, \\
the raven always and the eagle, when home they rode.\evb\evg


\bvg\bva%
Vaknaði \alst{B}rynhildr \hld\ \alst{B}uðla dóttir &
\alst{d}ís skjǫldunga \hld\ fyr \alst{d}ag lítlu: &
„\alst{h}vętið mik eða lętið mik \hld\ \alst{h}armr es unninn &
\alst{s}org at \alst{s}ęgja \hld\ eða \alst{s}vá láta.“\eva

\bvb Byrnhild awoke, Budle’s daughter, \\
the dise of shieldings a little before day: \\
“TODO.”\evb\evg


\bvg\bva%
Þǫgðu allir \hld\ við því orði &
fár kunni þęim \hld\ fljóða lǫ́tum &
es hǫ́n grátandi \hld\ gǫrðisk at sęgja &
þat’s hlę́jandi \hld\ hǫlða bęiddi.\eva

\bvb All men shut up at that word; \\
TODO.\evb\evg


\bvg\bva%
„Hugða’k mér, \alst{G}unnarr, \hld\ \alst{g}rimmt ï svefni, &
\alst{s}valt allt ï \alst{s}al \hld\ ę́tta’k \alst{s}ę́ing kalda, &
en þú \alst{g}ramr riðir \hld\ \alst{g}laums and-vani &
\alst{f}jǫtri \alst{f}atlaðr \hld\ ï \alst{f}jánda lið.\eva

\bvb “I had a cruel thought, Guther, in my sleep: \\
everything died in the hall, I had a cold bed, \\
and thou, prince, didst ride without cheerful fellows, \\
bound by fetters, into a troop of foes.\evb\evg


\bvg\bva%
Svá mun ǫll yður \hld\ ę́tt niflunga &
afli gęngin— \hld\ eruð ęið-rofa.\eva

\bvb So will all your line of Nivlings \\
part from strength—ye are oath-breakers.\evb\evg


\bvg\bva%
Mant-at-tu \alst{G}unnarr \hld\ til \alst{g}ǫrva þat &
es \edtrans{þit \alst{b}lóði ï spor \hld\ \alst{b}áðir ręnnduð}{your blood into your tracks ye both did drive}{\Bfootnote{Referring to a ritual of blood-brotherhood, wherein the “brothers” would mix their blood into their footprints on the ground.  This ritual is mentioned in \textcite{Saxo} 1.6.7: \emph{Siquidem icturi foedus veteres vestigia sua mutui sanguinis aspersione perfundere consueverant, amicitiarum pignus alterni cruoris commercio firmaturi} ‘Now our ancestors, when they meant to strike a pact, would sprinkle their combined blood in their footprints and mingle it, so as to strengthen the pledge of their fellowship.’.  For discussion see PCRN History and Structures III:845 ff.}}, &
nú hęfir þú hǫ́num þat \alst{a}llt \hld\ \alst{i}llu launat &
es hann \alst{f}ręmstan sik \hld\ \alst{f}inna vildi.\eva

\bvb Thou didst not recall, O Guther, clearly enough \\
that your blood into your tracks ye both did drive. \\
Now hast thou for all that cruelly repaid him, \\
TODO.\evb\evg


\bvg\bva%
Þȧ \alst{r}ęyndi þat \hld\ es \alst{r}iðit hafði &
\alst{m}óðigr ȧ vit \hld\ \alst{m}ín at biðja. &
\alst{h}vé \alst{h}ęr-glǫtuðr \hld\ \alst{h}afði fyrri &
\alst{ęi}ðum haldit \hld\ við inn \alst{u}nga gram.\eva

\bvb TODO: Translation.\evb\evg


\bvg\bva%
\alst{B}ęn-vǫnd of lét \hld\ \alst{b}rugðinn \edtext{gulli}{\Afootnote{\emph{gylli} \Regius}} &
\alst{m}arg dýrr konungr \hld\ ȧ \alst{m}eðal okkar; &
\alst{ę}ldi vǫ́ru \alst{ę}ggjar \hld\ \alst{ú}tan gǫrvar &
ęnn \alst{ęi}tr-dropum \hld\ \alst{i}nnan fáðar.“\eva

\bvb TODO: Translation.\evb\evg

\sectionline
