\bookStart{The Speeches of Sighdrive}[Sigrdrífumǫ́l]

% Introduction

Many of the verses are quoted in \VolsungaSaga, but notably the two prayer-verses are missing; possibly an instance of Christian censorship. TODO

\bvg {\small [Sighdrive quoth:]}
\bva „Lęngi ek svaf, \hld\ lęngi ek sofnuð vas, &
\ind lǫng eru lýða lę́; &
Óðinn því vęldr \hld\ es ęigi mátta’k &
\ind bregða blundstǫfum.“\eva

\bvb “Long I slept, long was I asleep, long are the deceits”\evb
\evg

BPG
BPA Sigurðr sęttisk niðr ok spyrr hana nafns. Hón tók þá horn fullt mjaðar ok gaf hǫ́num minnisvęig.EPA

BPB Siward set himself down, asking for her name. Then she took a horn full of mead, and gave him a mind-draught:EPB
EPG


\bvg
\bva Hęill Dagr, \hld\ hęilir Dags synir, &
\ind heil Nǫ́tt ok nipt! &
Óręiðum augum \hld\ lítið okkr þinig &
\ind ok gefið sitjǫndum sigr!\eva

\bvb “Hail \inx[P]{Day}! Hail the sons of Day!\footnoteB{TODO. Who?} Hail Night and [her] kinswoman \ken*{= Earth}!\footnoteB{According to \Gylfaginning\ TODO, Earth is the daughter of Night.} With unwrathful eyes look ye upon us two, and give the sitting ones \ken*{= us} victory.\evb
\evg


\bvg
\bva Hęilir ę́sir, \hld\ hęilar ǫ́synjur, &
\ind hęil sjá in fjǫlnýta fold! &
Mál ok manvit \hld\ gefið okkr mę́rum tvęim &
\ind ok lę́knishęndr meðan lifum!\eva

\bvb Hail the \inx[G]{Ease}! Hail the \inx[G]{Ossens}! Hail this bountiful fold \name{= Earth}! Speech and \inx[C]{manwit} give ye us renowned two, and \inx[C]{healing-hands}\footnoteB{Hands with the power to heal (perhaps supernaturally). This word also occurs in the semi-Christianized prayer on a c. 1300 stick from Ribe, Denmark (signum DR EM85;493).} while we live.”\evb
\evg


BPG
BPA Hon nefndisk Sigrdrífa ok var valkyrja. Hon sagði, at tveir konvngar bǫrðusk. Hét annarr Hjalmgunnarr; hann var þá gamall ok inn mesti hermaðr, ok hafði Óðinn hánum sigri heitit. En annarr hét Agnarr, · Auðu bróðir · er vę́tr engi · vildi þiggja. Sigrdrífa felldi Hjalmgunnar í orrostunni. En Óðinn stakk hana svefnþorni í hefnd þess ok kvað hana aldri skyldu síðan sigr vega í orrostu, ok kvað hana giftask skyldu, „en sagða’k hánum at strengða’k heit þar í mót, at giptask øngom þeim manni er hrę́ðask kynni.“ Hann segir ok biðr hana kenna sér speki ef hon\footnoteA{hánom ms.} vissi tíðendi ór ǫllum heimum. Sigrdrífa kvað:EPA

BPB She called herself Sighdrive and was a walkirrie. She said that two kings fought. One of them was called Helmguther; he was then old and the greatest harrier, and Weden had promised him victory. But another one was called Eyner, Eade’s brother, who in no way wished to accept.\footnoteB{i.e. ‘wished to lose’ TODO} Sighdrive felled Helmguther in the battle, but Weden pierced her with the sleeping-thorn as revenge for that, and said that she would never thenceforth win victory in battle, and said that she must marry, “but I told him that I made a vow against that, to marry no man who could be frightened.” He [i.e. Siward] speaks and asks her to teach him wisdom, if she knew any tidings out of all the \inx[C]{Home}[Homes]. Sighdrive quoth: EPB
EPG


\bvg
\bva „Bjór fǿri’k þér, \hld\ brynþings apaldr, &
magni blandinn \hld\ ok męgintíri, &
fullr ’s hann ljóða \hld\ ok líknstafa, &
góðra galdra \hld\ ok gamanrúna.\eva

\bvb Beer I bring thee—apple-tree of the byrnie-\inx[C]{Thing} \ken{battle > warrior}—mixed with might, and might-glory; it is filled with \inx[C]{leed}[leeds], and grace-staves; good \inx[C]{galder}[galders], and pleasure-\inx[C]{rune}[runes].\evb
\evg


\bvg
\bva Sigrúnar skalt kunna, \hld\ ef vilt sigr hafa, &
\ind ok rísta á hjalti hjǫrs, &
sumar á véttrimum, \hld\ sumar á valbǫstum, &
\ind ok nęfna tysvar Tý.\eva

\bvb Victory-runes shalt thou know, if thou wilt have victory, and carve on the hilt of the sword; some on weight-rims;\footnoteB{Unclear.} some on walbasts\footnoteB{Possibly the sword-pommel, the word also occurs in \HelgakvidaHjorvardssonar\ 9.}, and name \inx[P]{Tue} twice.\evb
\evg


\bvg
\bva Ǫlrúnar skalt kunna \hld\ ef þv vill aɴarſ qvęn &
\ind vęli t þic i trꝩgd ef þv trvir. &
a horni ſcal þęr riſta \hld\ oc a handar baki &
\ind oc merkia a nagli nꜹþ.\eva

\bvb Ale-runes shalt thou know, if TODO\evb
\evg


\bvg
\bva Full skal signa \hld\ ok við fári sjá &
\ind ok verpa lauki í lǫg; &
\edtext{þá þat vęitk, \hld\ at þér verðr aldri &
męini blandinn mjǫðr.}{\lemma{þá ... mjǫðr}\Bfootnote{\emph{thus} \VolsungaSaga, \emph{om.} \Regius}}\eva

\bvb TODO\evb
\evg

...


\bvg
\bva Þá mę́lti \hld\ Míms hǫfuð &
\ind fróðligt it fyrsta orð, &
\ind ok sagði sanna stafi.\eva

\bvb Then spoke the head of Mime learnedly the first word, and said true staves:\evb
\evg


\bvg
\bva Á skildi kvað ristnar \hld\ þęim’s stęndr fyr skínanda goði, &
á ęyra Árvakrs, \hld\ ok á Alsvinnz hófi, &
á því hvéli es snýz \hld\ undir ręið Hrungnis, &
á Slęipnis tǫnnum \hld\ ok á slęða fjǫtrum, &
á bjarnar hrammi \hld\ ok á Braga tungu, &
á ulfs klóm \hld\ ok á arnar nęfi, &
á blóðgum vę́ngjum \hld\ ok á brúar sporði, &
á lausnar lófa \hld\ ok á líknar spori, &
á glęri ok á gulli \hld\ ok á gumna hęillum, &
í víni ok virtri \hld\ ok vilisessi.\eva

\bvb On a shield it said were carved [runes]—[the shield] that stands before the shining god\footnoteB{According to \Grimnismal\ 39 the sun is covered by a shield, protecting the earth from its heat. Without it, the whole world would burn up.} \ken{sun}—[also] on the ear of Yorewaker, on the hoof of Allswith,\footnoteB{The two horses that pull the sun across the heavens; cf. \Grimnismal\ 38.} on that wheel which turns beneath the chariot of Rungner, on the teeth of Slopner, and on the fetters of sleds, on the paw of the bear, and on the tongue of Bray, on the claws of the wolf, and on the beak of the eagle, on bloody wings, and on the supports of the bridge, on the palm of release, and the track of grace, on glass and on gold, and on the good healths of men, in wine and beerwort, and on the comfortable seat.\evb
\evg


\bvg
\bva Á Gungnis oddi \hld\ ok á Grana brjósti, &
á nornar nagli \hld\ ok á nęfi uglu; &
allar vǫ́ru af skafnar, \hld\ þę́r es vǫ́ru á ristnar, &
\ind ok hvęrfðar við inn hęlga mjǫð &
\ind ok sęndar á víða vega.\eva

\bvb On the point of Gungner, and on the breast of Grane, on the nail of a norn, and on the beak of an owl;—all were shaven off—those that were carved on—and thrown into the holy mead, and sent on wide ways:\evb
\evg


\bvg
\bva Þę́r ’ró með ǫ́sum, \hld\ þę́r ’ró með ǫlfum, &
sumar með vísum vǫnum, \hld\ sumar hafa męnskir męnn.\eva

\bvb They are among Ease, they are among Elves; some among wise Wanes, some are had by manly men.\evb
\evg
