\bookStart{The Third Lay of Guthrun}[Guðrúnarkviða þriðja]

BPG
BPA Herkja hét ambǫ́tt Atla; hón hafði verit frilla hans. Hón sagði Atla at hón hefði sét Þjóðrek ok Guðrúnu bę́ði saman. Atli var þá allókátr. Þá kvað Guðrún: EPA

BPB Hark was named the female thrall of Attle; she had been his concubine. She told Attle that she had seen Thederick and Guthrun both together. Attle was then wholly displeased. Then Guthrun quoth: EPB
EPG


\bvg
\bva “Hvat es þér, Atli? \hld ę́, Buðla sonr, &
es þér hryggt í hug; \hld hví hlę́r þú ę́va? &
Hitt myndi ǿðra \hld jǫrlum þykkja &
at við menn mę́ltir \hld ok mik sę́ir.”\eva

\bvb What is with thee, Attle? Always, son of Bodle, art thou sad at heart; why laughest thou never? TO-DO\evb
\evg


\bvg
\bva “Tregr mik þat, Guðrún, \hld Gjúka dóttir, &
mér í hǫllu \hld Hęrkja sagði &
at þit Þjóðrekr \hld undir þaki svę́fið &
ok léttliga \hld líni vęrðið.”\eva

\bvb It troubles me, Guthrun, Yivick’s daughter, which in the hall Hark has said me: that thou and Thederick beneath thatched roof slept, and ye lightly warded the linen.\footnoteB{i.e., they threw off their clothes and slept together.}\evb
\evg


\bvg
\bva “Þér mun’k alls þęss \hld ęiða vinna &
at inum hvíta \hld helga stęini. &
at ek við Þjóðmar \hld þat-ki átta’k &
es vǫrðr né verr \hld vinna knátti.\eva

\bvb GAGAGGAGAG\evb
\evg


\bvg
\bva Nema ek halsaða \hld hęrja stilli, &
jǫfur óneisinn, \hld ęinu sinni; &
aðrar vǫ́ru \hld okkrar spękjur &
es við hǫrmug tvau \hld hnigum at rúnum.\eva

\bvb TESTETET STET T\evb
\evg


\bvg
\bva Hér kom Þjóðrekr \hld með þrjá tǫgu, &
lifa þęir né ęinir, \hld þriggja tega manna; &
hrinktu mik at brǿðrum \hld ok at brynjuðum, &
hrinktu mik at ǫllum \hld á hǫfuðniðjum.\eva

\bvb TESTE TEST EST TES\evb
\evg


\bvg
\bva Sęntu at Saxa, \hld sunnmanna gram; &
hann kann hęlga \hld hver vellanda;” &
sjau hundruð manna \hld í sal gengu &
áðr kvę́n konungs \hld í kętil tǿki.\eva

\bvb Send for Saxe, the prince of southmen; he knows how to hallow a swelling cauldron!” — Seven hundred men went into the hall, before the wife of the king might touch the kettle.\evb
\evg


\bvg
\bva “Kęmr-a nú Gunnarr, \hld kalli’k-a Hǫgna,
sé’k-a síðan \hld svása brǿðr;
sverði myndi Hǫgni \hld slíks harms reka,
nú verð’k sjǫlf fyr mik \hld synja lýta.”\eva

\bvb “Now Guthhere comes not, I call not on Hain; I see not hence [my] sweet brothers. With sword would Hain drive away such an affront; now I will for myself disprove the slanders.”\evb
\evg


\bvg
\bva Brá hón til botns \hld bjǫrtum lófa &
ok hón upp of tók \hld jarknastęina: &
Sé nú sęggir \hld sykn em ek orðin &
hęilagliga— \hld hvé sjá hverr velli.\eva

\bvb Brought she the bright palms to the bottom, and she up did take the earkenstones: “See now, men—I am proven innocent, through holy means—how this cauldron boils!”\evb
\evg


\bvg
\bva Hló þá Atla \hld hugr í brjósti &
es hann hęilar sá \hld hęndr Guðrúnar: &
Nú skal Hęrkja \hld til hvers ganga, &
sú er Guðrúnu \hld grandi vę́nti. \eva

\bvb Then the heart of Attle laughed in his breast, when he saw the hands of Guthrun unscathed: “Now shall Hark go to the cauldron, she who to Guthrun hoped to cause harm.”\evb
\evg


\bvg
\bva Sá-at maðr armligt, \hld hvęrr es þat sá at, &
hvé þar á Hęrkju \hld hęndr sviðnuðu; &
lęiddu þá męy \hld í mýri fúla, &
svá þá Guðrún \hld sinna harma.\eva

\bvb Each man saw not something so pitiful, who saw that: how there on Hark the hands were scorched. Led they the maiden into the foul bog; thus was Guðrún reconstituted for her affronts.\evb
\evg
