\bookStart{Spae of the Wallow}[Vǫluspǫ́]
\def\thisBookCode{Voluspa}

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: C10th (0.865)–early C11th (0.121)

\textbf{Meter:} \Fornyrdislag%
\end{flushright}

\section{Introduction}

The \textbf{Spae of the Wallow} (abbrev. \Voluspa) is probably the most important Norse mythological text surviving from Heathen times.  It is a \inx[C]{spae} (ON \emph{spǫ́} ‘prophecy’) told by a \inx[C]{wallow} (\emph{vǫlva} ‘seeress, sibyl, prophetess’) who has been summoned by \inx[P]{Weden}, the chief of the Gods, in order to relate the mythic history of the World.

The figure of the wallow represents the Germanic belief that women, at least in elder age, were invested with a certain psychic or prophetic power; this power became even greater when they died and came into possession of such knowledge unknown even to the chief god Weden himself.  It was for this reason that Weden in his incessant lust for wisdom undertook frequent journeys in order to question various beings, especially ones dead, about mythological lore.  \Voluspa\ is not the only Eddic poem reflecting this motif.  Most similar is \Baldrsdraumar, wherein Weden summons a wallow from her grave in order to find out why the god \inx[P]{Balder} is having ominous nightmares, but Weden’s journeys to commune with the dead are also alluded to in \Harbardsljod\ TODO.  For his general quest for knowledge there is also \textlink{Vafthrudnismal}, wherein Weden defeats the wise ettin \inx[P]{Webthrithner} to a wisdom contest, his self-hanging (\Havamal\ 138, 139), and his giving of an eye to Mimer (\Voluspa\ 28).

In its role as a mythic catalogue \Voluspa\ resembles \Grimnismal, \Sigrdrifumal, and \Alvissmal, but it differs from all aforementioned works in two key ways: \Voluspa\ is a prophecy rather than a motley collection of scattered mythological lore and a monologue rather than a dialogue or riddle contest.  Its unique importance lies in the fact that it offers a chronological overview of the Norse mythic timeline of the World from its creation to its coming destruction and rebirth.  This timeline is, unfortunately for the study of Norse mythology, very much troubled by the fact that the wallow continually speaks in the most obscure terms.  Events are related in an allusive fashion that presupposes that the audience be already familiar with them, and while we can connect many of the allusions to more complete narratives in Snorre’s Edda (which, to complicate things further, often relies on \Voluspa), others are entirely unknown to us.  Things are made even harder by the fact that the witness manuscripts often disagree over the content and order of stanzas.

\subsection{Preservation}

\Voluspa\ is attested in two full independent recensions.  The first and most important is \Regius\ (foll. 1r–3r), where it is the first poem.  The other is \Hauksbok\ (foll. 20r–21r), where it is found in a large collection of saws and Catholics works.

Many stanzas from \Voluspa\ are also cited or paraphrased in \Gylfaginning, the first part of Snorre’s Edda, which in many parts closely follows the poem.  Snorre clearly had access to a third, now-lost, recension of \Voluspa, and therefore his cited stanzas are very important.  The paraphrases are harder to use, but can still provide important critical readings and serve as a tiebreaker for readings that differ between \Regius\ and \Hauksbok\ (e.g. in st. 19, where \emph{sal} ‘hall’ in the paraphrase agrees with \Hauksbok\ against \Regius\ \emph{sę́} ‘lake’).  For the four mss. of \Gylfaginning—\RegiusProse, \Trajectinus, \Wormianus, and \Upsaliensis—see the Introduction to Mythic Poetry.

When it comes to the order of stanzas, which varies greatly across mss., I have prepared the following table.  Stanzas in \Gylfaginning\ which are quoted independently are marked with plus signs; sequences of uninterrupted quotations of several sts. are abbreviated with an incrementing alphabetic symbol, and each stanza is given a number based on its position, so that \emph{B1} is the first stanza in the second sequence, and so on.  When a stanza found in a manuscript is strongly divergent (e.g. st. 10, where \Gylfaginning\ omits the first two half-lines), its number is followed by a star.  The stanzas beginning with the formula \emph{Þȧ gingu ręgin ǫll} ‘Then went all the Reins’ et c. are represented by the immediately following half-line.

\begin{small}\begin{longtabu} to \textwidth {|c c c c c c|}
	\hline
	\multicolumn{2}{|c}{\emph{pres. ed.}} & \Regius & \Hauksbok & \RegiusProse\Trajectinus\Wormianus & \Upsaliensis \\ [0.5ex]
	\hline\hline\endhead
	\hline\endfoot
	1 & Hljóðs bið’k allar & 1 & 1 & − & − \\
	2 & Ek man jǫtna & 2 & 2 & − & − \\
	3 & Ár vas alda & 3 & 3 & + & + \\
	4 & áðr Burs synir & 4 & 4 & − & − \\
	5 & Sól varp sunnan & 5 & 5 & +* & +* \\
	6 & \dots\ nǫ́tt ok niðjum & 6 & 6 & − & − \\
	7 & Hittusk ę̇sir & 7 & 7 & − & − \\
	8 & Tęflðu ï tu̇ni & 8 & 8 & − & − \\
	9 & \dots\ hvęrr skyldi dverga & 9 & 9 & B1 & B1 \\
	10 & Þar vas Móðsognir & 10 & 10 & B2* & B2* \\
	11–15 & \emph{Dwarf-tallies} & 11–15 & 11–16 & + & + \\
	16 & Und’s þrír kvǫ̇mu & 16 & 17 & − & − \\
	17 & Ǫnd þau né ǫ́ttu & 17 & 18 & − & − \\
	18 & Ask vęit’k standa & 18 & 19 & + & + \\
	19 & Þaðan koma męyjar & 19–20 & 20–21 & − & − \\
	20 & Þat man hǫ̇n folk-víg & 21–22 & 27 & − & − \\
	21 & Hęiði hétu & 23 & 28 & − & − \\
	22 & \dots\ hvárt skyldu ę̇sir & 24 & 29 & − & − \\
	23 & Flęygði Óðinn & 25 & 30 & − & − \\
	24 & \dots\ hvęrr hęfði lopt alt & 26 & 22 & C1 & C1 \\
	25 & Þȯrr ęinn þar vá & 27 & 23 & C2* & C2* \\
	26 & Vęit hǫ̇n Hęimdalar & 28 & 24 & − & − \\
	27 & Ęin sat hǫ̇n úti & 29 & − & − & − \\
	28 & Alt vęit’k, Óðinn & 29 & − & + & + \\
	29 & Valði hęnni Hęr-fǫðr & 30 & − & − & − \\
	30 & Sá hǫ̇n val-kyrjur & 31 & − & − & − \\
	31 & Ek sá Baldri & 32 & − & − & − \\
	32 & Varð af męiði & 33 & − & − & − \\
	33 & Þó hann ę́va hęndr & 34 & − & − & − \\
	H1 & Þȧ kná Váli & − & 31 & − & − \\
	34a & Hapt sá hǫ̇n liggja & 35a & − & − & − \\
	34b & þar sitr Sigyn & 35b & 32 & − & − \\
	35 & Ǫ́ fęllr austan & 36 & − & − & − \\
	36 & Stóð fyr norðan & 36 & − & − & − \\
	37 & Sal sá hǫ̇n standa & 37 & 36 & E1 & E1 \\
	38 & Sér hǫ̇n þar vaða & 38 & 37 & E2* & E2* \\
	39 & Austr býr hin aldna & 39 & 25 & A1 & A1 \\
	40 & Fyllisk fjǫrvi & 40 & 26 & A2 & A2 \\
	41 & Sat þar ȧ haugi & 41 & 34 & − & − \\
	42 & Gól of ǫ̇sum & 42 & 35 & − & − \\
	43, 48, 56 & Gęyr (nú) Garmr mjǫk & 43, 46, 55 & 33, 38, 43, 48, 51 & − & − \\
	44 & Brǿðr munu bęrjask & 44 & 39 & − & − \\
	45 & Lęika Mïms synir & 45 & 40 & D1* & D1* \\
	H2 & Hrę́ðask allir & − & 41 & − & − \\
	46 & Hvat ’s með ǫ̇sum? & 49 & 42 & D2 & D2* \\
	48 & Hrymr ękr austan & 47 & 44 & D3 & − \\
	49 & Kjóll fęrr austan & 48 & 45 & D4 & − \\
	50 & Surtr fęrr sunnan & 50 & 46 & +, D5 (cited twice) & + \\
	51 & Þȧ kømr Hlïnar & 51 & 47 & D6 & − \\
	52 & Þȧ kømr hinn mikli & 52 & − & D7 & − \\
	H3 & Gïnn lopt yfir & − & 48 & — & − \\
	53 & Þȧ kømr hinn mę́ri & 53* & 49* & D8 & − \\
	54 & Sól tér sortna & 54 & 50 & D9 & − \\
	56 & Sér hǫ̇n upp koma & 56 & 52 & − & − \\
	57 & Finnask ę̇sir & 57* & 53 & − & − \\
	58 & Þar munu ęptir & 58 & 54 & − & − \\
	59 & Munu ȯ·sánir & 59 & 55 & − & − \\
	60 & Þȧ kná Hø̇nir & 60 & 56 & − & − \\
	61 & Sal sér hǫ̇n standa & 61 & 57 & + & + \\
	H4 & Þȧ kømr hinn ríki & − & 58 & − & − \\
	62 & Þar kømr hinn dimmi & 62 & 59 & − & − \\ [1ex]
	\hline
\end{longtabu}\end{small}

\subsection{Summary}

It seems that immediately prior to the poem’s commencing, Weden has summoned the wallow from her grave (probably in a manner similar to that in \Baldrsdraumar) and commanded her to speak.

The wallow’s \emph{spae} begins with a bid for silence (1), followed by her earliest memories of the world before its Creation by the Gods out of the body of Yimer (2).  She recounts the creation (3–6) and the following golden age (7–8), which was brought to an end by the intrusion of three unidentified ettin-maidens (8).  After this she describes the making of the dwarfs (9–10), whose names are listed in several separate \emph{dwarf-tallies} which are without doubt later inserts (11–15).  She then describes how the Gods gave life to the first human beings (16–17), and then describes the Ugdrassle’s Ash (18), and the three \inx[G]{norns} living under it (19).

Following st. 19 the order of stanzas in the two full recension—\Regius\ and \Hauksbok—diverges significantly.  In \Regius—whose ordering is the one adopted for the present edition—the wallow recalls how a woman named Goldwey was sacrificed and reborn three times (20), and how she, under the name Heath, practiced sorcery and witchcraft (21).  She then recalls the first war in the world, between the Eese and Wanes (22–23), and alludes to the killing of the smith who according to \Gylfaginning\ 42 was promised \inx[P]{Frow} and the sun and moon in exchange for building the wall of Osyard (24-25).  This is followed by a cryptic verse describing Homedal’s hidden silence or hearing (26).

In \Hauksbok\ the structure is quite different.  The Eese immediately go to decide what action to take regarding the promising of Frow to the ettin (24-25), and Homedal’s hearing is described (26). Then follow the two sts about the wolves that will swallow the sun and moon (40-41), and after this come sts 20–23 in the same order as \Regius.

TODO.

\newpage

\section{The Spae of the Wallow}

\bvg\bva\mssnote{\Regius~1r/2, \Hauksbok~20r/1}%
\edtext{„\edtrans{\alst{H}ljóðs bið’k}{For hearing I ask}{\Bfootnote{The same introductory expression is found in st. 2 of Eyel’s Head-ransom (Egill \emph{Hfl} in \Skp\ 5): \emph{hljóðs biðjum hann} ‘for hearing we [I] ask him’.}} allar \hld\ \edtrans{\alst{h}ęlgar}{holy}{\Afootnote{so \Hauksbok; om. \Regius}\Bfootnote{That the omission of this word in \Regius\ is nothing more than a scribal error is clearly shown by the meter; the a-verse in \emph{Hljóðs bið \alst{e}k \hld\ \alst{a}llar kindir} is only three syllables long, and has highly unnatural alliteration on the unstressed \emph{ek} rather than the expected first nominal \emph{hljóðs}.}} kindir, &
\edtrans{\alst{m}ęiri ok \alst{m}inni}{greater and lesser}{\Bfootnote{It is ambiguous to which noun phrase these adjectives belong.  It can either be (a) “all greater and lesser holy kindreds”, which might be equivalent to the phrase \inx[F]{Eese and Elves}—earthly and heavenly supernatural beings; see Index for occurrences—, or (b) “greater and lesser lads of Homedal”.  (b) is to be preferred as the syntactically simpler reading, in which case “greater or lesser” most likely refers literally to physical size (the younger and older members of the audience) rather than more figuratively to varying social classes.}} \hld\ \edtrans{\alst{m}ǫgu Hęimdalar}{lads of Homedal \ken{men}}{\Bfootnote{Homedal sired the three castes of men, as told in \Rigsthula.}}; &
\alst{v}ilt at, \edtrans{\alst{V}al-fǫðr}{Walfather}{\Bfootnote{That is, “Father of the Slain”, a name for Weden.  The wallow probably uses this name since Weden has awoken her from her grave (perhaps with a \emph{val-galdr} ‘slain-galder’ as in \Baldrsdraumar\ 4/3b).  Cf. st. 62/4b.}}, \hld\ \alst{v}ęl fram tęlja’k &
\alst{f}orn spjǫll \alst{f}ira, \hld\ \edtrans{þau’s \alst{f}ręmst of man}{which I earliest recall}{\Bfootnote{Cf. \textlink{Vafthrudnismal}[34]–35 with similar phrasing.}}?}{\lemma{ALL}\Bfootnote[nonum]{The wallow begins by asking for the silence of both gods and men, a meristic expression \parencite[99--100]{West2007}.  The whole introductory formula has Indo-European parallels; see \textcite[63,92--93,312]{West2007}.}}\eva

\bvb “{\huge F}\textsc{or hearing} I ask all holy races \ken{gods}, \\
greater and lesser lads of \inx[P]{Homedal} \ken{men}! \\
Wilt thou, \inx[P]{Walfather}, that I well tell forth \\
the ancient sayings of men which I earliest recall?\evb\evg


\bvg\bva\mssnote{\Regius~1r/4, \Hauksbok~20r/2}%
\alst{E}k man \alst{jǫ}tna \hld\ \alst{á}r of borna, &
þȧ’s \alst{f}orðum mik \hld\ \alst{f}ǿdda hǫfðu; &
\alst{n}íu man’k hęima, \hld\ \alst{n}íu \edtext{ïviðjur}{\Afootnote{so all.  \Regius\ has previously been as read \emph{‘iviði’}, but this was made obsolete by an x-ray scan undertaken by \textcite{StefanKarlsson1979} revealing a tiny abbreviation mark for \emph{-ur}.}\Bfootnote{Evil-working women or ogresses; this word also appears in a list of names for trollwomen (Þul \emph{Trollkvenna} 3 in \Skp\ 3).  The word is a fem. \emph{jōn}-stem.  A commonly suggested etymology is \emph{í} ‘in’ + \emph{viðr} ‘wood’ (i.e. forest-dwellers), but this would be an unusual formation, and leaves the \emph{-j-} unexplained.  A more plausible etymology is an agent-noun based on \emph{*ïvið} ‘guile, malice’, attested in the cpd. \emph{ïvið-gjarn} (\Volundarkvida\ 28).  This etymology can also explain the \emph{-j-}, since its WGmc. cognates OE \emph{inwid}, OS \emph{inwid}, and OHG \emph{inwit} show it to be a neutr. \emph{ja}-stem.}}, &
\edtrans{\alst{m}jǫt-við \alst{m}ę́ran \hld\ fyr \alst{m}old neðan.}{the famed Measure-Tree beneath the soil.}{\Bfootnote{Probably \inx[L]{Ugdrassle’s Ash}, still but a seed.  The sense of \emph{mjǫt-viðr} ‘Measure-Tree’ is not clear, but it seems to be the same word as \emph{mjǫtuðr} in 45/1b below.}}\eva

\bvb I recall \inx[G]{Ettins} born of yore, \\
those who formerly had nourished me. \\
Nine \inx[C]{Home}[Homes] I recall; nine \inx[G]{Inwithies}; \\
the famed Measure-Tree beneath the soil.\evb\evg


\bvg\bva\mssnote{\Regius~1r/6, \Hauksbok~20r/4, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{Á}r vas \alst{a}lda \hld\ \edtrans{þar’s \alst{Y}mir byggði}{where Yimer dwelled}{\Afootnote{\emph{þat’s ękki vas} ‘when nothing was’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}}, &
vas-a \alst{s}andr né \alst{s}ę́r, \hld\ né \alst{s}valar unnir; &
\edtext{\alst{jǫ}rð fannsk \alst{ę́}va \hld\ né \alst{u}pp-himinn}{\lemma{jǫrð \dots\ né upp-himinn ‘Earth \dots\ nor Up-heaven’}\Bfootnote{A well-attested formulaic cosmological word-pair found in all four Old Germanic languages with alliterative poetic traditions (ON, OE, OS, and OHG), especially in the context of the creation and destruction of the world.  See Index: \inx[F]{Earth and Upheaven}.}}; &
\edtrans{\alst{g}ap vas \alst{g}innunga}{there was the gap of hawks \ken{air/midspace}}{\Bfootnote{\emph{gap ginnunga} ‘gap of hawks’ has usually been interpreted on the basis of the cosmogony in \Gylfaginning, where Snorre presents \emph{ginnunga-gap} as a primordial physical location where cold and hot elements combined to form the acidic rivers known as the Ilewaves in the place where the Earth would later be situated.  For this reason it is typically translated into English simply as “Ginnungagap”, and explained as a sort of supernatural void.  However, the present stanza is the only conjunction of the words \emph{gap} and \emph{ginnunga} outside of Snorre’s Edda, and his cosmogony is very peculiar.  There is therefore some reason to reexamine these two words in the present context to see what the “gap” may actually be.

To begin, I reject the traditional translation “yawning chaos” as etymologically unfounded: the oft connected ON \emph{gína} ‘to yawn, gape’ has a different root structure than \emph{ginnunga}, which appears to be a gen. sg. or pl. and not an adjective; the explanation ‘width, breadth’ (from the lost cognate of OE \emph{ginne} ‘broad, vast’) is somewhat more plausible, but “gap of widths” is still a strange and decidedly non-Norse construction.  Instead, I agree with Meissner in reading \emph{gap ginnunga} as a kenning “gap of hawks \ken{air/midspace}”, where \emph{ginnunga} is gen. pl. of the attested poetic word \emph{ginnungr} ‘hawk’.  The kenning-type “land, path of the bird \ken{air}” is conventional \parencite[108]{Meissner1921}, and the determinant \emph{ginnungr} is also found in a kenning in \Haustlong\ 15: \emph{ǫll endi-lǫ́g ginnunga vé} ‘all the low mansions of hawks \ken{lower skies} from end to end’.  This interpretation is confirmed by an underappreciated passage in \Skaldskaparmal\ 74, which lists \emph{ginnunga-gap} among poetic synonyms for the air: \emph{Lopt heitir ginnunga-gap ok meðal-heimr, fogl-heimr, veðr-heimr.} ‘Air is called gap of ginnings and middle-home, bird-home, weather-home.’

Having determined the sense of the word, we should consider its sense in the present context.  The Old Germanic cosmology, like many other ancient religions (e.g. the Vedic with its two World-Halves (Sanskrit \emph{ródasī}), \Rigveda\ 1.10.8, 5.85.3 etc.) sees the World as consisting of two halves, Earth and Upheaven.  The former is the land of Men, the latter the land of the Gods—both are believed to be physically concrete, formed from different parts of Yimer’s body (\Grimnismal\ 41, \textlink{Vafthrudnismal}[21]).  Sandwiched between them are the lower atmospheric skies, the air (ON \emph{lopt}) or the Midspace (ON \emph{meðal-hęimr} ‘middle realm’, equivalent to the \emph{antárikṣam} of the Vedic hymns, e.g. in \Rigveda\ 5.85.3).  This is seen in \Haustlong, which in st. 15 speaks of the “low \textsc{skies}”, contrasting them with Upheaven in st. 16.  The primordial state expressed in the present stanza is thus one where only the airy lower skies exist, while Heaven and Earth are still conjoined in the monstrous body of Yimer.}}, \hld\ en \alst{g}ras \edtrans{hvęrgi}{nowhere}{\Afootnote{\emph{ękki} ‘not’ \Hauksbok}},%
\Ballnote{Told in sts. 3–4 is the creation of the Earth and Heavens.  Other poetic sources for it are \textlink{Grimnismal}[41]–42, \textlink{Vafthrudnismal}[21].  A more extensive creation narrative is found in \Gylfaginning\ 4–5, according to which the world first consisted of two extremities, viz. the frozen Nivelham in the north and the burning Muspellsham in the south.  From Nivelham rushed the icey venom rivers called the \inx[L]{Ilewaves} until they froze to ice, while burning lava flowed from Muspellsham.  The ice and lava met in the Gap of Ginnings (\emph{Ginnunga-gap}), “which was as calm as windless air”, and there combined to form the first being, \inx[P]{Yimer}, who was the ancestor of the ettins.}\eva%TODO: bibliography, Meissner is 20. Luft, b

\bvb It was early of ages where \inx[P]{Yimer} dwelled; \\
there was not sand nor sea nor cool waves. \\
\inx[L]{Earth} was never found nor \inx[L]{Up-heaven}; \\
there was the \inx[L]{gap of hawks} \ken{air/midspace}, but grass nowhere,\evb\evg


\bvg\bva\mssnote{\Regius~1r/8, \Hauksbok~20r/5}%
áðr \edtrans{\alst{B}urs synir}{the Sons of Byre}{\Bfootnote{In \Gylfaginning\ 6 identified as the three brothers Weden, Will, and Wigh, the first Gods.  They sacrificed Yimer and created the World out of his body, for which cf. \Grimnismal\ 41–42, \textlink{Vafthrudnismal}[21].}} \hld\ \alst{b}jǫðum of ypðu, &
þęir es \edtrans{\alst{M}ið-garð}{Middenyard}{\Bfootnote{The Middle Enclosure, which was created as the home of Men (\Grimnismal\ 42).}} \hld\ \alst{m}ę́ran skópu; &
\alst{s}ól skęin \alst{s}unnan \hld\ ȧ \alst{s}alar stęina; &
þȧ vas \alst{g}rund \alst{g}róin \hld\ \edtrans{\alst{g}rø̇num lauki}{green leek}{\Bfootnote{A sign of the golden age, for the leek was in ancient times held to be the noblest plant. See Index.}}.\eva

\bvb before the \inx[P]{Sons of Byre} revealed the flatlands, \\
they who created famed \inx[L]{Middenyard}. \\
The sun shone from the south on the stones of the hall; \\
then was the ground grown with green leek.\evb\evg


\bvg\bva\mssnote{\Regius~1r/11, \Hauksbok~20r/7, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\edtext{\alst{S}ól varp \alst{s}unnan, \hld\ \edtrans{\alst{s}inni Mȧna}{Moon’s companion}{\Bfootnote{At times mistranslated as ‘her moon’, understanding \emph{sinni} as dat. sg. f. of \emph{sïnn} ‘its (reflexive)’.  This cannot be correct since ON possessives are inflected based on the gender of the noun possessed, not the gender of the possessor.  \emph{mȧni} ‘moon’ is masculine, and so ‘her moon’ would be \emph{*sïnum Mȧna}.}}, &
\alst{h}ęndi hinni \alst{h}ǿgri \hld\ of \edtrans{\alst{h}imin-jǫður}{heaven’s rim}{\Afootnote{composite; \emph{himin †iodyr†} \Regius; \emph{ioður} \Hauksbok.}\Bfootnote{Some recent editors have taken it upon themselves to normalize the reading of \Regius\ as \emph{himin-jó-dýr} ‘heaven-horse-beast’, which is not just nonsensical but also unmetrical due the stress pattern.  On the other hand the reading of \Hauksbok\ (norm. \emph{jǫður} ‘rim, edge’) is clearly deficient since it lacks the necessary alliteration on \emph{h}.
If we instead see \emph{iodyr} \Regius\ as corrupted from an archetypal \emph{*iodur} shared by \Hauksbok, we can restore \emph{*himin-jǫður}, as done here.}}}{\lemma{Sól \dots\ himin-jǫður ‘Sun \dots\ heaven’s rim’}\Afootnote{om. \RegiusProse\Trajectinus\Upsaliensis\Wormianus.}\Bfootnote{Probably a poetic description of the dawn; the Sun lifted herself up over the horizon and rose for the first time.}}; &
\alst{S}ól þat né vissi, \hld\ hvar hǫ̇n \alst{s}ali átti; &
\edtext{\edtrans{\alst{M}ȧni þat né vissi, \hld\ hvat hann \alst{m}ęgins átti}{the Moon knew not what sort of might he had}{\Bfootnote{For a concrete example of this “might” cf. \Havamal\ 137/7, where the moon is to be invoked in a blood feud.  A belief in the power of the moon is found in most ancient religions.  Its influence on the human mind was remarked already by \textcite{PlinyNatHist}, 2.102, 18.75; witness in folklore the influence of the full moon on the werewolf, or in English the word \emph{lunacy} (from Latin \emph{lūna} ‘moon’).  Women are particularly affected by the moon through the synchronisation of the monthly and menstrual cycles (the latter word deriving from Latin \emph{mēnsis} ‘month’, originally ‘moon’) as caused by the moon’s gravitational pull on the body \parencite{Forster2025}.}}; &
\alst{st}jǫrnur þat né vissu, \hld\ hvar þę́r \alst{st}aði ǫ́ttu}{\lemma{Mȧni \dots\ ǫ́ttu}\Afootnote{In \Regius\Hauksbok\ these two lines switch places, so that the order is Sun, Stars, Moon.}}; &\eva

\bvb The Sun cast from the south—the Moon’s companion— \\
her right hand over heaven’s rim. \\
The Sun knew not where halls she had; \\
the Moon knew not what sort of might he had; \\
the Stars knew not where seats they had.\evb\evg


\bvg\bva\mssnote{\Regius~1r/13, \Hauksbok~20r/9}%
\edtext{Þȧ gingu \edtrans{\alst{r}ęgin ǫll}{all the Reins}{\Bfootnote{All the Gods (\emph{ręgin} ‘Reins’, i.e. ‘Powers, Counsels’; a pantheistic word), here acting as one body.}} \hld\ ȧ \edtrans{\alst{r}ǫk-stóla}{rake-seats}{\Bfootnote{Their thrones of judgment.}}, &
\alst{g}inn-hęilǫg \alst{g}oð, \hld\ ok umb þat \alst{g}ę̇ttusk}{\lemma{Þȧ \dots\ gę̇ttusk ‘Then \dots\ of this.’}\Bfootnote{A formulaic expression for the convening of the \inx[L]{Thing of the Gods}, identically repeated below in sts. 9/1–2, 22/1–2, and 24/1–2.  Cf. also the formula shared between \Baldrsdraumar\ 1/1–3 and \Thrymskvida\ 14/1–3, which follows the structure of the present formula very closely: \emph{Sęnn vǫ́ru ę̇sir \hld\ allir ȧ þingi // ok ǫ̇synjur \hld\ allar ȧ máli, // ok umb þat réðu \hld\ ríkir tívar.} ‘Soon were the \inx[G]{Eese} all at the \inx[C]{Thing}, // and the \inx[G]{Ossens} all at speech, // and of this counseled the mighty \inx[G]{Tews}.’ — In the five occurrences of these two formulae outside of the present stanza, the demonstrative pronoun \emph{þat} ‘this’ clearly refers to an immediately following question introduced by a \emph{hv}-word (e.g. \Thrymskvida\ 14/4: \emph{hvé þęir Hlórriða \hld\ hamar of sǿtti?} ‘how they Loride’s \name{= Thunder’s} hammer would find?’)  Following this pattern we would expect to find such a question following \emph{umb þat gę̇ttusk} ‘took counsel of that’ in the present stanza, and it seems reasonable plausible (but not certain) that one has been lost in transmission.}}. &
\edtrans{\alst{N}ǫ́tt ok \alst{n}iðjum}{night and the moon-phases}{\Bfootnote{Also found in \textlink{Vafthrudnismal}[24]/4.}} \hld\ \edtrans{\alst{n}ǫfn of gǫ́fu}{names they gave}{\Bfootnote{The giving of names to the natural phenomena is seen as a part of the Godly ordering of the universe; in fact it is the very act of naming and thereby categorizing these phenomena which sets them apart from the disordered Natural chaos.  These categories, rather than being arbitrarily invented by Men, are thus seen as stemming directly from the Gods.  In a modern scientific context this passage can be read as a celebration of the work of such men as Linnæus, who by their actions in classifying and rationalizing nature are thus carrying out the will of the Gods.  This is not the same as the positive Christian attitude towards science, which is based on the idea of understanding the perfect design of the intelligent Creator.  In the Germanic worldview, however, the raw matter already exists (in the form of Yimer and other Ettins) before the Gods, and the Creation consists in shaping and classifying it according to Godly Will (which appears synonymous with rationality)—in the present stanza in order to reckon time and keep dates.}}, &
\alst{m}orgin hétu \hld\ ok \alst{m}iðjan dag, &
\alst{u}ndurn ok \alst{a}ptan, \hld\ \edtrans{\alst{ǫ́}rum at tęlja}{the years for to tally}{\Bfootnote{Cf. \textlink{Vafthrudnismal}[22]–25, where it is said that the Gods made the sun and moon turn round in heaven and created the moon-phases \emph{ǫldum at ár-tali} ‘for mankind’s tally of years’.}}\eva

\bvb Then went all the Reins onto the rake-seats: \\
the Yin-holy Gods, and from each other took counsel of that. \\
To night and the moon-phases names they gave; \\
morning they named, and middle day, \\
afternoon and evening, the years for to tally.\evb\evg


\bvg\bva\mssnote{\Regius~1r/16, \Hauksbok~20r/10}%
Hittusk \alst{ę̇}sir \hld\ ȧ \edtrans{\alst{I}ða-vęlli}{Idewolds}{\Bfootnote{“The plains of industry.”}}, &
\edtext{þęir’s \alst{h}ǫrg ok \alst{h}of \hld\ \alst{h}ǫ́-timbruðu}{\lemma{þęir’s \dots\ hǫ́-timbruðu ‘they who \dots\ timbered on high’}\Afootnote{\emph{afls kostuðu \hld\ alls freistuðu} ‘[their] strength they tried; everything they tempted’ \Hauksbok}\Bfootnote{Two formulæ. — \emph{hǫrgr ok hof} ‘harrow and hove’ is a merism, i.e. ritual structures made of stone and wood; cf. \textlink{Vafthrudnismal}[38] and \HelgakvidaHjorvardssonar\ TODO, as well as the Norwegian Christian laws that impose ‘the burning of hoves and the breaking of harrows’ (\emph{brenna hof ok brjóta hǫrga}). — \emph{hǫ́-timbra} ‘timber on high’ is a rare compound. Its only other occurrence in the ON corpus is in \Grimnismal\ 16, where it describes a harrow ruled by Nearth. — This passage has often been wondered at; why would the Gods themselves make cultic buildings—whom do \emph{they} worship?  Let it be added that they partake in the ritual slaughter of beasts, and the following divination and feasting on their flesh (e.g. \Voluspa\ 61, \Hymiskvida\ 1, 39, \Lokasenna, \Haustlong\ 2).  This question is really not as difficult as it is made out to be, however, since the answer is surely that the behaviour of the Gods serves as the model for virtuous human behaviour and justifies the ideology of settlement.  Colonising new lands, tilling the earth, building houses and enclosures and ritual structures, sacrificing beasts—if these are virtuous for humanity, they must have a precedent among the Gods, which does not mean that the Gods worship themselves (or anything else).  The Gods operate on the level of the macrocosm, and it is from them that the Godly Will flows.  Humans, on the level of the microcosm, connect with this Godly Will through worship, among other behaviour, which mirrors the Gods on a smaller scale.  Thus the great enclosure of Middenyard and the sacrifice of Yimer (who was the Earth and Heaven conjoined) become smaller ritual and settlement enclosures and sacrifices of horses and oxen; for more discussion cf. \textlink{Grimnismal}[41]–43 and notes.}}; &
\alst{a}fla lǫgðu, \hld\ \alst{au}ð smíðuðu, &
\alst{t}angir skópu \hld\ ok \alst{t}ól gęrðu.\eva

\bvb The Eese met each other on the \inx[L]{Idewolds}, \\
they who \inx[C]{harrow} and \inx[C]{hove} timbered on high. \\
Hearths they laid, wealth they smithed, \\
tongs they shaped and tools they made.\evb\evg


\bvg\bva\mssnote{\Regius~1r/18, \Hauksbok~20r/12}%
\edtext{\edtrans{\alst{T}ęflðu}{played Tables}{\Bfootnote{A verb derived from \emph{tafl} ‘board game’, an old borrowing from Latin \emph{tabula}.  “Tables” is used as a cognate translation; the exact type of board game referred to is unimportant.}} ï \alst{t}u̇ni, \hld\ \alst{t}ęitir vǫ́ru, &
\edtrans{\alst{v}as þęim \edtrans{\alst{v}éttir-gis}{nothing}{\Bfootnote{An archaic gen. of \emph{vę́tt-ki} ‘nothing’; the \emph{-ir} representing a fossilized i-stem genitive, for \emph{véttr} ‘thing’ comes from PGmc. \emph{*wihtiʀ}.  The only other occurrence of this form is in the highly linguistically archaic Icelandic Homily Book (ms. Holm perg 15 4°, fol. 36v/30).}} \hld\ \alst{v}ant ór gulli}{for them was nothing golden wanting}{\Bfootnote{Indeed even the bricks they played with were of gold. See st. 58.}}, &
und’s \edtext{\alst{þ}ríar kvǫ̇mu \hld\ \alst{þ}ursa męyjar}{\lemma{þríar \dots\ þursa męyjar ‘three maidens of Thurses’}\Bfootnote{These three maidens are never mentioned again (unless they are taken to be the three norns in st. 19, but they would then be introduced twice). It is possible that an additional stanza giving further information about them has been lost. If it originally existed, it was already absent in the version used for \Gylfaginning, since no additional information is found there.}}, &
\edtrans{\alst{ȧ}m-átkar}{uncanny}{\Bfootnote{The word \emph{ám-áttigr} has a clear association with supernatural beings; trolls and ettins. It occurs in four other places in \Regius. In \Grimnismal\ 11, \Skirnismal\ 10 and \HelgakvidaHjorvardssonar\ 17 it modifies \emph{jǫtunn} ‘ettin’ in a \Ljodahattr\ c-line. In \HelgakvidaHjorvardssonar\ 14 it is used by the daughter of an ettin to refer to a human hero.}} mjǫk, \hld\ ór \alst{Jǫ}tun-hęimum.}{\lemma{ALL}\Bfootnote[nonum]{The whole stanza is paraphrased in \Gylfaginning\ ch. 14: \emph{Ok því nę́st smíðuðu þeir málm ok stein ok tré ok svá gnóg-liga þann málm, er gull heitir, at ǫll bús-gǫgn ok ǫll reiði-gǫgn hǫfðu þeir af gulli, ok er sú ǫld kǫlluð gull-aldr, áðr en spilltist af til-kvámu kvinnanna; þę́r kómu ór Jǫtun-heimum.} ‘And after this they smithed ore and stone and wood, and so abundantly [did they smith] that ore which is called gold, that all their house tools and riding tools were golden. And that age is called the golden age, before it was spoiled by the arrival of the women; they came from Ettinham.’}}\eva

\bvb They played \inx[C]{Tables} in the yard; merry were they; \\
for them was nothing golden wanting— \\
until three maidens of \inx[G]{Thurses} came, \\
most uncanny, out of \inx[L]{Ettinham}.\evb\evg


\bvg\bva\mssnote{\Regius~1r/20, \Hauksbok~20r/14, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\edtext{Þȧ gingu \alst{r}ęgin ǫll \hld\ ȧ \alst{r}ǫk-stóla, &
\alst{g}inn-hęilǫg \alst{g}oð, \hld\ ok umb þat \alst{g}ę̇ttusk: &
\edtrans{Hvęrr skyldi \alst{d}verga}{Who would \dots\ of dwarfs’}{\Afootnote{so \Regius\Wormianus\Upsaliensis; \emph{at skyldi dverga} ‘That they would \dots\ of dwarfs’ \RegiusProse\Trajectinus; \emph{hverir skyldu dvergar} ‘Which dwarfs would [shape the retinues]’ \Hauksbok}} \hld\ \edtrans{\alst{d}rótt}{the retinue}{\Afootnote{so \RegiusProse\Trajectinus\Upsaliensis\Wormianus; \emph{drotin} ‘the lord’ \Regius; \emph{dróttir} ‘the retinues’ \Hauksbok}} \edtrans{of skępja}{shape}{\Afootnote{\emph{spekia} ‘soothe’ \Upsaliensis}} &
\edtrans{ór \edtrans{\alst{b}rimi \alst{b}lóðgu}{bloody surf}{\Afootnote{so \Hauksbok\RegiusProse\Wormianus\Upsaliensis; \emph{Brimis blóði} ‘the blood of Brimmer’ \Regius\Trajectinus}} \hld\ ok ór \edtrans{\alst{b}lǫ́um}{blue-black}{\Afootnote{metr. emend. from \emph{blám} \Regius; \emph{Bláins} ‘Blown’s’ \Hauksbok\Wormianus; \emph{Bláms} \RegiusProse\Trajectinus\Upsaliensis\ is prob. a corrupt form of \emph{Bláins}}} lęggjum}{out of the bloody surf and the blue-black legs}{\Bfootnote{I think that the poem simply telling of “the bloody surf” and “the blue-black legs” fits better with its general allusive style, but this requires a composite reading.  If we read \emph{Bláinn} ‘Blown’ (named in the \inx[C]{thule}[thules] as a dwarf) instead of \emph{blǫ́um} ‘blue-black’, then following Gurevich (\emph{Skp} 2017, p. 693) we may see a kenning “the legs of Blown \name{dwarf} \ken{stones}”.  Blown has otherwise usually been read as a poetic name for Yimer, but it is not attested anywhere else. — The “blood” and “legs” are in any case those of Yimer; from his bones were made the rocks, and from his blood the sea (see \Grimnismal\ 41, \textlink{Vafthrudnismal}[21]).  Dwarfs of course dwell in rocks and earth; cf. for instance \Ynglingatal\ 2, where the Swedish king Swayther (\emph{Svęigðir}) runs into a rock in pursuit of a dwarf.  More difficult to explain is the creation of dwarfs from the sea.  Einheri suggests that it may be referring to the formation of salt-stones by means of evaporating salty seawater.}}?}%
{\lemma{ALL}\Bfootnote[nonum]{After the Golden Age is spoiled, the Gods must get their metal in some other way.  For this they need the dwarfs, who are connected with finding minerals, perhaps through techniques similar to dousing.  Ancient ideas about the spontaneous generation of maggots in flesh (likened to minerals in the earth) are also clearly at play. — \Gylfaginning\ 14 continues with its paraphrase: \emph{Þar nę́st settust goðin upp í sę́ti sín ok réttu dóma sína ok minntust, hvaðan dvergar hǫfðu kviknat í moldinni ok niðri í jǫrðunni, svá sem maðkar í holdi. Dvergarnir hǫfðu skipazt fyrst ok tekit kviknun í holdi Ymis ok váru þá maðkar, en af atkvę́ðum goðanna urðu þeir vitandi mann-vits ok hǫfðu manns líki ok búa þó í jǫrðu ok í steinum. Móðsognir var ǿðstr ok annarr Durinn. Svá segir í Vǫluspá:} ‘Thereafter the gods set themselves up in their seats and made their judgments and remembered whence the dwarfs had come to life in the ground and down in the earth like maggots in flesh. The dwarfs had first taken shape and come to life in Yimer’s flesh and were then maggots, but by the decrees of the gods they became knowing of manwit and had a man’s likeness, and even so they live in the earth and in stones. Moodsowner was the highest in rank, and second Dorn. So it says in the Spae of the Wallow:’ after which the text quotes the present st. and 10/3–4.}}\eva

\bvb Then went all the Reins onto the rake-seats: \\
the Yin-holy Gods, and from each other took counsel of this: \\
Who would shape the retinue of \inx[G]{Dwarfs}, \\
out of the bloody surf and the blue-black legs?\evb\evg


\bvg\bva\mssnote{\Regius~1r/21, \Hauksbok~20r/15, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\edtext{\edtext{Þar vas \alst{M}óðsognir}{\Afootnote{so \Hauksbok; \emph{Þar †mótſognir vitnir†} ‘there Mootsowner wolf(?)’ \Regius. The prose of \Gylfaginning\ 14 agrees with \Hauksbok\ that the correct form of the name is \emph{Móðsognir}, not \emph{Mótsognir}.}} \hld\ \alst{m}ę́tstr of orðinn &
\alst{d}verga allra, \hld\ en \alst{D}urinn annarr;}{\lemma{Þar \dots\ annarr ‘There \dots\ second’}\Bfootnote{om. \RegiusProse\Trajectinus\Upsaliensis\Wormianus, but the author must have had the full stanza, since he paraphrases these lines (see note to ALL for st. 9 above).}} &
\edtext{\edtrans{þęir \edtrans{\alst{m}an-líkun}{man-likenesses}{\Bfootnote{The Dwarfs were in the shape of Men, but not quite human; thus the creation is seen as progressive, with the man-shaped Dwarfs preceding the human beings.  The Dwarfs, arising out of the earth, also lack the Godly spark found in Men.}} \hld\ \alst{m}ǫrg of gørðu}{Many man-likenesses they made’}{\Afootnote{so \Regius\Hauksbok\Upsaliensis; \emph{þar man-líkun \hld\ mǫrg of gęrðusk} ‘There many man-likenesses were made’ \RegiusProse\Trajectinus\Wormianus}}, &
\alst{d}vergar \edtrans{ï}{in}{\Afootnote{so \RegiusProse\Trajectinus\Upsaliensis\Wormianus\Hauksbok; \emph{ór} ‘out of’ \Regius}} jǫrðu, \hld\ \edtrans{sęm \alst{D}urinn sagði}{as Dorn said}{\Afootnote{so \Regius\Hauksbok\RegiusProse\Wormianus; \emph{sem †dur menn† sagði} ‘as door-men(?) said’ \Trajectinus; \emph{sem †þeim dyrinn kendi†} ‘as the beasts(?) taught them’ \Upsaliensis}}.}{\lemma{þęir \dots\ sagði ‘They \dots\ said.’}\Bfootnote{The mss. readings offer two conflicting versions of the creation of the dwarfs.  Either they arose on their own; this is supported by the prose of \Gylfaginning\ 14 (see note to previous st.) and by the form of the stanza quoted there in the non-\Upsaliensis\ mss., but it may have been changed to correspond to the author’s vision.  On the other hand, \Regius, \Hauksbok\ and \Upsaliensis\ have the dwarfs Moodsowner and Dorn shaping “man-likenesses” out of soil.  The present ed. follows \Regius\Hauksbok\Upsaliensis.}}\eva

\bvb There was Moodsowner made the worthiest \\
of all Dwarfs, and Dorn the second. \\
Many man-likenesses they made: \\
Dwarfs in the earth, as Dorn said.\evb\evg

\sectionline\setcounter{stanza}{16}

For the \emph{Dwarf-tallies} see the end of the poem.

\sectionline

\bvg\bva\mssnote{\Regius~1v/1, \Hauksbok~20r/26}%
\edtrans{Und’s \edtext{\alst{þ}r\emph{í}r}{\Afootnote{emend.; \emph{þrjár} \Regius\Hauksbok}} kvǫ̇mu \hld\ \edtext{ór \alst{þ}ví liði}{\Afootnote{so \Regius; \emph{þussa brúðir} \Hauksbok.}}%
}{Until three came out of that host}{\Bfootnote{This line, beginning with the conjunction “until”, is clearly supposed to connect to a previous stanza (cf. st. 4 \emph{áðr}, TODO: other occurrences).  If the dwarf-tallies (11–15), as seems likely, are later inserts, that would be st. 10, in which case the creation of men is perhaps to be understood as superseding the creation of the dwarfs; cf. note to st. 10/3: \emph{man-líkun}.

Both mss. show influence from st. 8 in using the fem. \emph{þrjár} instead of masc. \emph{*þrír}.  \Hauksbok\ goes further in replacing \emph{ór því liði} ‘out of that host’ with \emph{þussa brúðir} ‘brides of thurses’.  That these are errors is clearly shown by the masculine \emph{ǫflgir ok ȧstkir ę̇sir} in l. 2.}} &
\edtrans{\alst{ǫ}flgir ok \edtrans{\alst{ȧ}stkir}{loving}{\Bfootnote{The creation of men was an act of love; for humanity the Gods later made Middenyard (\Grimnismal\ 42); the moon-phases had already been created for our time-reckoning (st. 6 above).}}}{strong and loving}{\Afootnote{\emph{ȧstkir ok ǫflgir} (norm.) ‘loving and strong’ \Hauksbok}} \hld\ \alst{ę̇}sir \edtrans{at húsi}{along the houses}{\Bfootnote{An adverbial; the Gods were walking on the outskirts of their settlement.}}; &
fundu ȧ \alst{l}andi \hld\ \alst{l}ítt męgandi &
\edtrans{\alst{A}sk ok \alst{Ę}mblu}{Ash and Emble}{\Bfootnote{Ash (nom. \emph{Askr}) is easily identified with the same-named wood species (\emph{Fraxinus excelsior}), but the etymology of Emble (nom. \emph{Ęmbla}) is much more difficult to explain.  Her name is often translated as “Elm” (so Neil Price), but the ON word for that tree is the masc. \emph{almr} ‘elm’.  Metathesis from earlier \emph{*Ęlma}, a derivative of the same type as \emph{þęlla} ‘young fir tree’ < \emph{þǫll} ‘fir tree’, is possible but uncertain.}} \hld\ \alst{ø}r·lǫg-lausa.%
\Ballnote{This stanza and the next are paraphrased in \Gylfaginning\ 9: \emph{Þá er þeir gengu með sę́var-strǫndu Bors synir, fundu þeir tré tvau ok tóku upp trén ok skǫpuðu af menn. Gaf inn fyrsti ǫnd ok líf, annarr vit ok hrę́ring, þriði á·sjónu, mál ok heyrn ok sjón. Gáfu þeim klę́ði ok nǫfn; hét karl-maðr’inn Askr, en kona’n Embla, ok ólst þaðan af mann-kind’in, sú er byggð’in var gefinn undir Mið-garði.} ‘When the sons of Byre (cf. st. 4) walked along the sea-shore they found two trees (\emph{tré}, alt. ‘pieces of wood’) and they took up the trees and shaped men out of them.  The first one gave breath and life; the second wit and movement; the third outward appearance, speech and hearing and sight.  They gave them clothes and names: the male was called Ash and the woman Emble.  And from them was begotten mankind, to which the dwelling within Middenyard was given.’

Based on \Gylfaginning\ the creation of men is traditionally seen as referring to pieces of driftwood, but that may be a later Icelandic or Snorronean interpretation.  As pointed out by \textcite{Hultgård2006}, the evidence from comparative mythology suggests that the first humans were originally seen as living, growing trees, and there is really nothing in the \Voluspa\ that speaks against such an interpretation.  This myth is probably the reason why words for trees are used extensively by Norse poets in kennings for men and women (see \textciteshorttitle{SkP} I, p. lxxv ff., \cite[245,266--272, 410]{Meissner1921}), more commonly in Scaldic poetry, but at times also in Eddic poetry (e.g. \Sigrdrifumal\ 5: \emph{bryn-þings apaldr} ‘apple-tree of the byrnie-\inx[C]{Thing} \ken{battle > warrior}’).}\eva

\bvb Until three came out of that host, \\
strong and loving Eese along the houses. \\
They found on land the little availing \\
Ash and Emble, \inx[C]{orlay}-less.\evb\evg


\bvg\bva\mssnote{\Regius~1v/3, \Hauksbok~20r/27}%
\edtrans{\alst{Ǫ}nd}{Breath}{\Bfootnote{The breath (animating spirit) of life, which sets living things apart from the unliving.  Cf. \Gylfaginning\ 3: \emph{Hitt er þó mest, er hann gerði manninn ok gaf honum ǫnd þá, er lifa skal ok aldri týnast, þótt líkaminn fúni at moldu eða brenni at ǫsku} ‘Yet the greatest thing is when he [= Weden the Allfather] made man and gave him that “breath” which shall live and never perish even though the body molders to dust or burns to ashes.’  On Christian Scandinavian memorial runestones from the C11th onwards this word is used interchangably with the Anglo-Saxon borrowing \emph{sál} ‘soul’; compare e.g. Sö 10 \emph{Guð hjalpi ǫnd hans} ‘God help his “breath”’, Sö 8 \emph{Guð hjalpi sǫ́lu hans} ‘God help his soul’, and the frequent (at least 14 separate inscriptions) pairing of the two, like e.g. U 358 \emph{Guð hjalpi hans ǫnd ok sálu} ‘God help his “breath” and soul.’  It seems likely that this idea of an immortal “breath”, instead of being pagan, stems from the Latin \emph{spiritus} which means both ‘breath’ and ‘spirit’.  In old poems a person gives up his “breath” when he dies and stops breathing, cf. \HelgakvidaHjorvardssonar, \Sigrdrifumal, \Sigurdskamma\ TODO.}} þau né \alst{ǫ́}ttu, \hld\ \alst{ó}ð þau né hǫfðu, &
\alst{l}ǫ́ né \alst{l}ę́ti \hld\ né \alst{l}itu góða; &
\alst{ǫ}nd gaf \alst{Ó}ðinn, \hld\ \alst{ó}ð gaf Hø̇nir, &
\alst{l}ǫ́ gaf \alst{L}óðurr \hld\ ok \alst{l}itu góða.\eva

\bvb Breath they owned not, \inx[C]{wode} they had not, \\
not craft nor sound nor good colour. \\
Breath gave Weden, wode gave Heener, \\
craft gave Lother, and good colour.\evb\evg


\bvg\bva\mssnote{\Regius~1v/5, \Hauksbok~20r/29, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{A}sk vęit’k \edtext{standa}{\lemma{standa ‘standing’}\Afootnote{so \Regius\Hauksbok\Upsaliensis; \emph{ausinn} ‘sprinkled’ \RegiusProse\Trajectinus\Wormianus}}, \hld\ hęitir \edtext{\alst{Y}gg-drasill}{\Afootnote{\emph{Ygg-drasils} \RegiusProse}}, &
\alst{h}ǫ́r \edtrans{baðmr}{beam}{\Afootnote{\emph{borinn} ‘born’ \Upsaliensis\ wo. doubt corrupt.}}, \edtrans{\edtrans{ausinn}{sprinkled}{\Afootnote{\emph{hęilagr} ‘holy’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}} \hld\ \alst{h}víta auri}{sprinkled with white mud}{\Bfootnote{Possibly relevant is the Indian ritual pouring of beverages like milk onto the phallic \emph{líṅga}, although Nikhil Surya Dwibhashyam considers this an indigenous Indian practice foreign to the old Vedic religion.  Cf. st. 26 below.}}; &
þaðan koma \alst{d}ǫggvar \hld\ \edtext{þę́r’s}{\Afootnote{\emph{es} \RegiusProse\Trajectinus}} ï \alst{d}ala falla; &
stęndr \edtext{\alst{ę́}}{\Afootnote{\emph{om.} \Upsaliensis}} \alst{y}fir \edtext{grø̇nn}{\Afootnote{\emph{†grvnn†} \RegiusProse; \emph{†grein†} \Upsaliensis}} \hld\ \alst{U}rðar brunni.\eva

\bvb An ash I know standing, ’tis called \inx[L]{Ugdrassle}: \\
a high beam \ken{tree} sprinkled with white mud. \\
Thence come the dew-drops which fall in the dales; \\
it stands ever green over \inx[L]{Weird’s Well}.\evb\evg


\bvg\bva\mssnote{\Regius~1v/8, \Hauksbok~20r/31}%
Þaðan koma \alst{m}ęyjar \hld\ \alst{m}args vitandi &
\alst{þ}ríar ór þęim \edtrans{sal}{hall}{\Afootnote{so \Hauksbok, \RegiusProse\Trajectinus\Upsaliensis\Wormianus\ (in the paraphrase); \emph{sę́} ‘lake’ \Regius}} \hld\ es \edtrans{und}{under}{\Afootnote{\emph{ȧ} ‘on’ \Hauksbok}} \edtrans{\alst{þ}olli}{tree}{\Bfootnote{Literally ‘fir’, but the word is only used for the alliteration.  The same may perhaps apply to \emph{askr} ‘ash’ above, the species being indeterminate.}} stęndr; &
\alst{U}rð hétu \alst{ęi}na, \hld\ \alst{a}ðra Verðandi, &
—\edtrans{\alst{sk}ǫ́ru ȧ \alst{sk}íði}{they scored billets}{\Bfootnote{Unclear; perhaps they carve tallies for the number of years allotted to each human being.}}— \hld\ \alst{Sk}uld hina þriðju &
þę́r \alst{l}ǫg \alst{l}ǫgðu, \hld\ þę́r \alst{l}íf køru, &
\alst{a}lda bǫrnum, \hld\ \alst{ø}r·lǫg \edtrans{sęggja}{of youths}{\Afootnote{\emph{at sęgja} ‘to say’ \Hauksbok}}.%
\Ballnote{The st. is paraphrased in \Gylfaginning\ 15: \emph{Þar stendr salr einn fagr undir aski’num við brunni’n, ok ór þeim sal koma þrjár meyjar, þę́r er svá heita: Urðr, Verðandi, Skuld.  Þessar meyjar skapa mǫnnum aldr; þę́r kǫllum vér nornir.} ‘There stands a single fair hall beneath the ash-tree by the well, and out of that hall come three maidens, who are called so: Weird, Werthing, Shild. These maidens shape the ages of men; we call them norns.’}\eva

\bvb Thence come maidens much knowing: \\
three out of the hall which stands beneath the tree. \\
Weird they called one, the other Werthing \\
—they scored billets—Shild the third. \\
Laws they laid down; lives they chose \\
for the children of mankind, the \inx[C]{orlay} of youths.\evb\evg


\bvg\bva\mssnote{\Regius~1v/11, \Hauksbok~20v/5}%
Þat man hǫ̇n \edtrans{\alst{f}olk-víg}{troop-conflict}{\Bfootnote{\emph{folk} here carries its older meaning ‘troop, band’, as seen in the Slavic borrowing exemplified by Russian \textrussian{полк} ‘regiment, host, army’.}} \hld\ \alst{f}yrst ï hęimi, &
es \alst{G}ull-vęigu \hld\ \alst{g}ęirum studdu &
ok ï \alst{h}ǫll \alst{H}á\emph{a}rs \hld\ \alst{h}ȧna bręnndu, &
\edtext{\alst{þ}rysvar bręnndu}{\Afootnote{\emph{†þrysvar brendv þrysvar brendv†} \Hauksbok}} \hld\ \alst{þ}rysvar borna, &
\alst{o}pt, \alst{ȯ}-sjaldan, \hld\ þó hǫ̇n \alst{ę}nn lifir.\edtext{}{\lemma{ALL}\Bfootnote[nonum]{The wallow begins to describe herself in the third person; a very cryptic stanza.  The name Goldwey is not mentioned in any other source, nor is the underlying story, but she was apparently slain, burned and reborn three times (in short succession?) by the Eese.}}\eva

\bvb That troop-conflict she recalls first in the \inx[C]{Home}, \\
when Goldwey with spears they pierced \\
and in the hall of \inx[P]{Hougher} \name{= Weden} \ken*{= Walhall} burned her; \\
thrice they burned the thrice born, \\
often, unseldom, though she still lives.\evb\evg


\bvg\bva\mssnote{\Regius~1v/13, \Hauksbok~20v/7}%
\alst{H}ęiði \alst{h}étu, \hld\ hvar’s til \alst{h}úsa kom, &
\edtext{\alst{v}ǫlu}{\Afootnote{\emph{ok vǫlu} \Hauksbok}} \alst{v}ęl-spáa, \hld\ \alst{v}itti ganda; &
\alst{s}ęið hǫ́n \edtrans{hvar’s hǫ́n kunni}{where she could}{\Afootnote{so \Hauksbok; \emph{hǫ́n kunni} ‘she knew’ \Regius}}, \hld\ \alst{s}ęið hǫ́n \edtrans{hug lęikinn}{deluded minds}{\Afootnote{so \Hauksbok; \emph{leikinn} \Regius}}; &
\alst{ę́} vas hǫ̇n \alst{a}ngan \hld\ \alst{i}llrar brúðar.\eva

\bvb Heath they called—where to houses she came— \\
the well-spaeing \inx[C]{wallow}; she bewitched \inx[C]{gand}[gands]. \\
She sorcered where she could; she sorcered deluded minds; \\
she was always the love of any evil bride.\evb\evg


\bvg\bva\mssnote{\Regius~1v/16, \Hauksbok~20v/9}%
Þȧ gingu \alst{r}ęgin ǫll \hld\ ȧ \alst{r}ǫk-stóla, &
\alst{g}inn-hęilǫg goð, \hld\ ok umb þat \alst{g}ę̇ttusk: &
Hvárt skyldu \alst{ę̇}sir \hld\ \alst{a}f-ráð gjalda, &
eða skyldu \edtrans{\alst{g}oð’in ǫll}{all the Gods}{\Bfootnote{The clitic definite \emph{-in} is very rare in older Norse poetry; this is its only occurence in \Voluspa. — Here “all the Gods” (viz., the Eese \emph{and} the Wanes) seem to be contrasted with the \inx[G]{Eese}, a subset.}} \hld\ \alst{g}ildi ęiga?\eva

\bvb Then went all the Reins onto the rake-seats: \\
the Yin-holy Gods, and from each other took counsel of this: \\
Whether the Eese should yield tribute, \\
or should all the Gods hold a banquet?\evb\evg


\bvg\bva\mssnote{\Regius~1v/17, \Hauksbok~20v/11}%
\edtrans{\alst{F}lęygði Óðinn \hld\ ok ï \alst{f}olk of skaut}{Weden hurled and shot into the troop}{\Bfootnote{The object, a spear, is understood.  This first spear-throw was reenacted in a ritual well attested in Icelandic literature, wherein the king leading his troops would hurl the first spear into the opposing host, typically with the phrase \emph{Óðinn á yðr alla} ‘Weden owns you all!’  The battle-slain were thusly devoted to Weden, and they would join him as \inx[G]{Oneharriers} in \inx[L]{Walhall}.  The sacrifice of an entire army or nation was not uncommon in ancient warfare, and examples are also found among the Hebrews (the \textgreek{חֵרֶם} \emph{ḥērem}) and the Romans (the \emph{devotio}, Livy 8:9).  Weden is also described as “owning” dead warriors in \Harbardsljod\ TODO, and in runic inscription \emph{N B380} (edited below under Galders), a sort of greeting wherein the receiver is wished to be owned by Weden (and “received” by Thunder).  For further literature see \textciteshorttitle{PCRN-HS} II:24, p. 560, II:25, p. 617, and especially III:42, p. 1166ff.}}; &
þat vas ęnn \alst{f}olk-víg \hld\ \edtrans{\alst{f}yrr}{earlier}{\Afootnote{so \Hauksbok; \emph{fyrst} ‘first’ \Regius. The \Regius\ reading cannot be correct as this st. is describing a different war, and thus not the first. It has probably arisen due to the similarity with st. 20/1.}} ï hęimi; &
\alst{b}rotinn vas \alst{b}orð-vęggr \hld\ \alst{b}orgar ȧsa, &
knǫ́ttu \alst{v}anir \edtrans{\alst{v}íg-spǫ́}{war-spae}{\Bfootnote{The Wanes used a magic prophecy (\emph{spǫ́} ‘spae’) to win the battle and sack \inx[L]{Osyard}, the stronghold of the Eese.}} \hld\ \alst{v}ǫllu sporna.\eva

\bvb Weden hurled and shot into the troop; \\
that was yet a troop-conflict earlier in the \inx[L]{Home}. \\
Broken was the plank-wall of the stronghold of the Eese; \\
the Wanes by a war-\inx[C]{spae} did tread the fields.\evb\evg


\bvg\bva\mssnote{\Regius~1v/19, \Hauksbok~20r/34, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\edtext{Þȧ gingu \alst{r}ęgin ǫll \hld\ ȧ \alst{r}ǫk-stóla, &
\alst{g}inn-hęilǫg \alst{g}oð, \hld\ ok umb þat \alst{g}ę̇ttusk: &
Hvęrr hęfði \alst{l}opt alt \hld\ \alst{l}ę́vi blandit &
eða \alst{ę́}tt \alst{jǫ}tuns \hld\ \alst{Ó}ðs męy gefna?}{\lemma{ALL}\Bfootnote[nonum]{After their stronghold, protected only by a plank-wall (\emph{borð-vęggr}), is sacked by the Wanes, the Eese decide to build a stronger wall.  The story of the wall-builder is told in \Gylfaginning\ 42, which ends by quoting sts. 24–25.  An \inx[G]{Ettins}[ettin] craftsman approached the Eese and asked to build them a great wall.  His price was \inx[P]{Frow}’s hand, and the \inx[P]{Sun} and \inx[P]{Moon}, but only if he could complete the entire wall alone in a single winter.  He also asked for permission to use his workhorse, \inx[P]{Swaddlefare}, which Lock granted him.  The agreement was sealed with strong oaths.  The horse was, however, unexpectedly strong, and when three days were left before summer the wall was almost finished.  The panicked Eese then turned to Lock and forced him to deal with the horse.  His solution was to turn into a mare to distract the ettin’s workhorse, which worked; the two were out all night, and Lock was made pregnant, later giving birth to \inx[P]{Slapner}.  When the ettin realised that he would not finish the wall on time he came into his greatest ettin-wrath, at which point the Eese called on Thunder; he showed up and quickly slew the builder.}}\eva

\bvb Then went all the Reins onto the rake-seats: \\
the Yin-holy Gods, and from each other took counsel of this: \\
Who might have blended all the air with deceit, \\
or to the ettin’s lineage given \inx[P]{Wode}’s maiden \ken*{= Frow}?\evb\evg


\bvg\bva\mssnote{\Regius~1v/20, \Hauksbok~20r/36, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\edtext{\alst{Þ}ȯrr ęinn \edtrans{\alst{þ}ar vá}{fought there}{\Afootnote{so \Hauksbok\Trajectinus\Upsaliensis; \emph{þar var} ‘was there’ \Regius; \emph{þat vann} ‘accomplished it’ \RegiusProse; \emph{þat vá} ‘fought it’ \Wormianus}} \hld\ \alst{þ}runginn móði, &
\edtrans{hann \alst{s}jaldan \alst{s}itr \hld\ es \alst{s}líkt of fregn;}{he seldom sits when of such he learns}{\Bfootnote{When he learns of an ettin encroaching on the gods (see note to 24/ALL).  Thunder is the defender of the gods (\Thrymskvida\ 18, Þdís \emph{Þórr} in \Skp\ III) and is willing to break even oaths sworn to an ettin for this purpose (cf. \Lokasenna\ 57–64).}} &
\edtext{\alst{ȧ} gingusk \alst{ęi}ðar, \hld\ \alst{o}rð ok sǿri, &
\alst{m}ǫ́l ǫll \alst{m}ęgin-lig, \hld\ es ȧ \alst{m}eðal \edtext{fóru}{\lemma{fóru ‘had gone’}\Afootnote{\emph{vǫ́ru} ‘had been’ \Hauksbok\Trajectinus}}.}{\lemma{ȧ \dots\ fóru.}\Afootnote{om. \Wormianus}}}{\lemma{ALL}\Afootnote[nonum]{The order of the lines is that of \Regius\Hauksbok; in \RegiusProse\Trajectinus\Upsaliensis\Wormianus\ the two helmings (\emph{Þȯrr \dots\ fregn;} and \emph{ȧ \dots\ fóru.}) are reversed.}}\eva

\bvb Thunder alone fought there, pressed by wrath; \\
he seldom sits when of such he learns. \\
Trampled were oaths, speeches and vows, \\
all the mighty treaties which had gone between them.\evb\evg

\bvg\bva\mssnote{\Regius~1v/23, \Hauksbok~20v/1}%
Vęit hǫ̇n \alst{H}ęimdalar \hld\ \alst{h}ljóð of folgit &
und \edtrans{\alst{h}ęið-vǫnum}{shady}{\Bfootnote{Literally ‘light-less’, \emph{hęiðr} referring especially to the light of a clear sky.}} \hld\ \alst{h}ęlgum baðmi; &
\alst{ǫ́} sér hǫ̇n \alst{au}sask \hld\ \edtrans{\alst{au}rgum}{muddy}{\Bfootnote{Which should be the same mud (\emph{aurr}) as in st. 19, there said of Weird’s Well.}} forsi &
af \edtrans{\alst{v}ęði \alst{V}al-fǫðrs}{Walfather’s pledge}{\Bfootnote{Weden placed his eye in Mimer’s well, which gives wisdom to any man who drinks from it.  So \Gylfaginning\ 15: \emph{Þar kom Alfǫðr ok beiddisk eins drykkjar af brunninum, en hann fekk eigi, fyrr en hann lagði auga sitt at veði.} ‘There came Allfather and asked for a single drink from the well, but he did not get it before he laid down his eye as a pledge.’}}. \hld\ \edtrans{\alst{V}ituð ér ęnn eða hvat?}{Know ye yet, or what?}{\Bfootnote{“Do you, Weden, know enough now, or what?”, repeated in 28, 33, 34, 38, 40, 47, 60, and 61.  Similar refrains are found in \Baldrsdraumar\ and \Hyndluljod.}}\eva

\bvb She knows Homedal’s sound \ken*{= Horn of Yell?} hidden \\
beneath the shady, hallowed beam \ken*{= Ugdrassle’s Ash?}. \\
A river she sees being fed by a muddy torrent \\
from Walfather’s pledge \ken*{= Mimer’s well}.—Know ye yet, or what?”\evb\evg

\sectionline

\bvg\bva\mssnote{\Regius~1v/25}%
\edtrans{\alst{Ęi}n sat hǫ̇n \alst{ú}ti}{Alone sat she outside}{\Bfootnote{To \emph{sitja úti} ‘sit outside’ has a cultural connotation of meditation in order to connect or communicate with the otherworld; cf. the noun \emph{úti-seta}.  This line is directly repeated in \Sigurdskamma\ 6/1a.}}, \hld\ þȧ’s hinn \alst{a}ldni kom &
\alst{y}ggjungr \alst{ȧ}sa \hld\ ok ï \alst{au}gu lęit: &
‚hvęrs \alst{f}regnið mik? \hld\ hví \edtrans{\alst{f}ręistið}{tempt}{\Bfootnote{\emph{fręista} ‘tempt’ has a sense of testing someone, especially intellectually.  Cf. \Havamal\ 2, 26, \textlink{Vafthrudnismal}[3], 5.}} mïn?\eva

\bvb Alone sat she outside when the old one came, \\
the Terrifier of the Eese \ken*{= Weden}, and looked into her eyes. \\
\speakernoteb{[The Wallow:]}%
‘Of what ask ye me? Why tempt ye me?\evb\evg


\bvg\bva\mssnote{\Regius~1v/26, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{A}lt vęit’k, \alst{Ó}ðinn, \hld\ hvar \alst{au}ga falt &
\edtrans{ï hinum \alst{m}ę́ra}{in the famed}{\Afootnote{so \Wormianus; \emph{þitt} (corr.) \emph{i enom męra} ‘id.’ \Regius; \emph{j þeim enom meira} ‘in the greater’ \Trajectinus; \emph{i þeim envm mæra} ‘in the famed’ \Upsaliensis; \emph{vr þeim envm mę́ra} ‘out of the famed’ \RegiusProse}} \hld\ \alst{M}ímis brunni; &
drekkr \alst{m}jǫð \alst{M}ímir \hld\ \alst{m}orgin hvęrjan &
af \edtrans{\alst{v}ęði}{pledge}{\Afootnote{\emph{†veiði†} \RegiusProse}} \alst{V}al-fǫðrs.‘ \hld\ \alst{V}ituð ér ęnn eða hvat?\eva

\bvb I know it all, Weden, where thine eye thou hidst: \\
in the famed \inx[L]{Mimer’s Well} \\
drinks Mimer mead every morning \\
from Walfather’s pledge.’—Know ye yet, or what?\evb\evg


\bvg\bva\mssnote{\Regius~1v/29}%
Valði hęnni \alst{H}ęr-fǫðr \hld\ \alst{h}ringa ok męn, &
\edtrans{fekk \alst{sp}jǫll \alst{sp}ak-lig}{got foresighted tidings}{\Afootnote{emend.; \emph{fe spioll spaclig} \Regius}\Bfootnote{The reading of \Regius\ may be interpreted either as (1): \emph{fé-spjǫll spak-lig} ‘foresighted wealth-spells’ or (2) \emph{fé, spjǫll spak-lig} ‘wealth, foresighted tidings’; both are metrically deficient.  In (1) a second element in a cpd. like \emph{fé-spjǫll} cannot carry alliteration, and (2) has three strongly stressed nominals; in both cases \emph{fé} which stands first would be expected to carry the alliteration.  The word \emph{fé} ‘wealth, cattle’ also makes little sense in context, since Weden is the one giving her expensive jewellery.

The emendation places the verb \emph{fekk} ‘got, received’ for \emph{fé}.  Verbs carry less stress than verbs, and the line is thus metrically equivalent to 28/3b \emph{drekkr mjǫð Mímir}.  The line parallels st. 1, where the wallow likewise says that she will relate \emph{spjǫll} ‘tidings, sayings’ (cf. English \emph{gospel}, lit. ‘good news’ which originally translates the Greek \textgreek{εὐαγγέλιον}).  For discussion on this reading see \textcite[51--53]{Haukur2020}, \textcite[16]{Males2023}.}} \hld\ ok \edtrans{\alst{sp}á-ganda}{spae-gands}{\Bfootnote{Spirits sent out in order to gather hidden wisdom and spaes.  See relevant Index entries.}}; &
sá \alst{v}ítt ok umb \alst{v}ítt \hld\ of \alst{v}er-ǫld hvęrja.\eva

\bvb Host-father \name{= Weden} chose for her rings and a necklace, \\
he got foresighted tidings and \inx[C]{spae}-\inx[C]{gands}— \\
she saw widely and more widely, o’er every world.\evb\evg


\bvg\bva\mssnote{\Regius~1v/30}%
Sá hǫ̇n \alst{v}al-kyrjur \hld\ \alst{v}ítt of komnar, &
\alst{g}ǫrvar at ríða \hld\ til \edtrans{\alst{g}oð-þjóðar}{folk of the Gots}{\Bfootnote{Ambiguous; ON \emph{goð-þjóð} may mean either (1) ‘folk of the Gots’ or (2) ‘folk of the Gods’, for the difficult cluster \emph{tþ} in \emph{Got-þjóð} ‘folk of the Gots’ was at some point changed to \emph{ðþ}.  Alternative 1 is preferred since it is attested in three other places in \Regius, viz. \Helreid\ TODO and \Gudrunarhvot\ TODO and TODO, whereas 2 is entirely unattested. — It is interesting that ON \emph{Got-þjóð} reflects the attested Gotnish self-name, \emph{Gut-þiuda} (found in the October 29 entry of the Gotnish calender, TODO: reference).  The Walkirries have a particular association with the Gots, who fought the greatest battles of the Migration Period; cf. note to \Volundarkvida\ 1/1b.}}: &
\edtext{\alst{Sk}uld hélt \alst{sk}ildi, \hld\ en \alst{Sk}ǫgul ǫnnur, &
\alst{G}unnr, Hildr, \alst{G}ǫndul \hld\ ok \alst{G}ęir-skǫgul; &
\alst{n}ú eru talðar \hld\ \edtrans{\alst{N}ǫnnur Hęrjans}{Nans of Harn \name{= Weden}}{\Bfootnote{\emph{Nanna} ‘\inx[P]{Nan}’ (the name itself is a nursing word) was the wife of \inx[P]{Balder}, but the word is here certainly being used to refer generically to ‘maidens, women’.  Cf. Þul \emph{Ásynja} (\Skp\ 3), where the Walkirries are kenned \emph{Óðins męyjar} ‘Weden’s maidens’.}}, &
\alst{g}ǫrvar at ríða \hld\ \alst{g}rund, val-kyrjur.}{\lemma{Skuld \dots\ val-kyrjur. ‘Shild \dots\ lands.’}\Bfootnote{Judging especially by the out-of-place phrase \emph{nú eru talðar} ‘now are tallied’, these four lines seem to be a later insert from a \inx[C]{thule} counting the Walkirries.}}\eva

\bvb She saw \inx[G]{Walkirries} come from afar, \\
ready to ride to the folk of the \inx[C]{Gots}. \\
Shild held a shield and Shagle another, \\
Guth, Hild, Gandle and Goreshagle— \\
now are tallied the Nans of Harn \name{= Weden}, \\
Walkirries ready to ride o’er the earth.\evb\evg


\bvg\bva\mssnote{\Regius~2r/2}%
Ek sá \alst{B}aldri, \hld\ \alst{b}lóðgum \edtrans{tífur}{victim’s}{\Bfootnote{This word is rather difficult and possibly corrupt.  It may be connected with \emph{týr} ‘tew, god’, but the dat. sg. of \emph{týr} is \emph{tívi} and the intrusive \emph{r} is unexplained.  A better explanation is given by \CV, who connect it with OE \emph{tiber, tifer} ‘victim, hostage’, but this also has some problems.  \emph{blóðgum} ‘bloody’ is masc. dat. sg., but OE \emph{tiber} is neuter.  If we are dealing with a masc. noun \emph{*tífurr} with the same declension as \emph{jǫfurr}, we would expect dat. sg. \emph{*tífri}, not \emph{tífur} (which would however be the expected acc. sg.).}}, &
\alst{Ó}ðins barni, \hld\ \alst{ø}r·lǫg \edtrans{folgin}{sealed}{\Bfootnote{Or “hidden”.  The verb \emph{fela} ‘hide, conceal’ is used in poetry to describe burial in mounds, as in \Ynglingatal\ 24 (“[...] And afterwards the victory-havers hid (\emph{fǫ́lu}) the ruler on Borrey.”) or the C10th Karlevi stone (“Hidden (\textbf{fulkin} \emph{folginn}) in this mound lies he whom the greatest deeds followed; [...]”)}}; &
stóð of \alst{v}axinn \hld\ \alst{v}ǫllum hę́ri &
\alst{m}jór ok \alst{m}jǫk fagr \hld\ \alst{m}istil-tęinn.%
\Ballnote{Told allusively in \Voluspa\ 31–33 is the myth about Balder’s death.  Balder, the son of Weden and Frie, was slain with an arrow shot by his blind half-brother Hath, whose hand was guided by Lock.  Weden could not slay Hath, who was his son, and so he seduced the woman Rind, apparently through love-magic (Cormac Awmundson’s TODO: \emph{sęið Yggr til rindar} ‘Ug won Rind through sorcery’).  Rind gave birth to Wonnel, who grew very fast; after just one day he was big enough to kill Hath, which he also did, avenging Balder’s death.  The other important sources for this myth are \Baldrsdraumar\ 8–11, \Gylfaginning\ 49, and \textcite{Saxo} 3.4.1–8.

The language of \Baldrsdraumar\ is so similar to the present sts. that they must be of common origin; \Baldrsdraumar\ 11/2–4 is near-identical to \Voluspa\ 32/4–33/2.  The biggest narrative difference is that \Baldrsdraumar\ mentions Rind, who is not found in \Voluspa.

The most elaborate narrative is found in \Gylfaginning\ 49, which may be shortly summarised as follows: Balder has terrible nightmares about his own death, and so his mother Frie makes all sorts of things (fire, water, venom, metals, stones, trees, diseases, beasts, et c.) swear oaths not to harm him.  After this the Eese make sport of shooting and striking at him, since he cannot be harmed.  Lock is annoyed by this and approaches Frie while disguised as a woman.  He finds out from her that there is one thing that did not swear the oath—the mistletoe, which was thought too young.  Lock takes a mistletoe and a bow and gives it to the blind god Hath, showing him where to shoot.  Hath does so, and kills Balder.  After this \Gylfaginning\ describes Balder’s funeral (treated poetically in Wolf Ugson’s fragmentary \emph{House-drape}, ÚlfrU \emph{Húsdr} in \Skp\ III) and how the gods attempted to “weep Balder out of hell”, which failed (see Eddic Fragments in the present ed.)  \Gylfaginning\ 50 goes on to describe how the Eese punished Lock (see st. 34 below.)

It is notable that \Gylfaginning\ 49–50 fails to mention Wonnel.  This part of the myth may have been left out for moral reasons, but was certainly known to the author of the Prose Edda; cf. \Gylfaginning\ 30: \emph{Áli eða Váli heitir einn, sonr Óðins ok Rindar. Hann er djarfr í orrostum ok mjǫk happ-skęytr} ‘Onnel or Wonnel one is called, the son of Weden and Rind. He is brave in battles and a very lucky shot’ and \Skaldskaparmal\ 19: \emph{Hvernig skal kenna Vála? Svá, at kalla hann son Óðins ok Rindar, [\dots] hefni-ás Baldrs, dólg Haðar ok bana hans [\dots]} ‘How shall one ken Wonnel? Namely, by calling him the son of Weden and Rind, [\dots] avenging \inx[C]{os} of Balder, the foe of Hath and his bane [\dots].’

The last source is \textcite{Saxo} 3.4.1–8, who relates the revenge narrative in his typical euhemerized form, turning the gods Hath and Balder into flawed human rulers.  The outline of the narrative is as follows: After Balder (\emph{Balderus}) has succumbed to a mortal wound dealt him by Hath (Latin \emph{Høtherus}) with a treacherous sword, his father Weden (\emph{Othinus}) takes counsel from a group of wizards.  One of them, Horsethief the Finn (\emph{Rostiophus Phinnicus}), foresees that Rind (\emph{Rinda}), daughter of the Russian king, must bear him a son to avenge Balder.  Weden soon enlists in the Russian king’s army and leads it to great victories, but is spurned by his daughter.  He tries various disguises but is continually refused, until at last he disguises himself as an old woman and volunteers himself as her private physician.  When the daughter becomes ill, he binds her, ostensibly in order to administer a certain foul potion—instead he rapes her, apparently with her father’s consent.  Their son, Bo, grows up to become a fierce raider.  One day Weden summons him and reminds him of his duty to avenge his half-brother Balder.  Bo slays Hath in a fierce duel but soon himself perishes from his wounds.}\eva

\bvb I saw Balder’s—the bloody victim’s, \\
Weden’s child’s—\inx[C]{orlay} sealed: \\
there stood grown—higher than the plains, \\
slender and most fair—the mistletoe.\evb\evg


\bvg\bva\mssnote{\Regius~2r/4}%
Varð af \alst{m}ęiði, \hld\ þęim’s \alst{m}ę́r sýndisk, &
\alst{h}arm-flaug \alst{h}ę̇ttlig, \hld\ \alst{H}ǫðr nam skjóta. &
\alst{B}aldrs \alst{b}róðir vas \hld\ of \alst{b}orinn snimma, &
sá nam, \alst{Ó}ðins sonr, \hld\ \alst{ęi}n-nę́ttr vega.\eva

\bvb Of the tree which slender seemed \\
became a baneful harm-flier—Hath took to shoot. \\
Balder’s brother \ken*{= Wonnel} was born early; \\
he took, Weden’s son, one night old, to fight.\evb\evg


\bvg\bva\mssnote{\Regius~2r/6}%
\edtext{Þó}{\lemma{Þó \dots\ kęmbði ‘washed \dots\ combed’}\Bfootnote{A collocation, see note to \Havamal\ 61 for discussion and other examples. Wonnel, being oathbound and on the mission to avenge his brother, could not engage in such acts of personal vanity.}} ę́va \alst{h}ęndr \hld\ né \alst{h}ǫfuð kęmbði, &
áðr ȧ \alst{b}ál of \alst{b}ar \hld\ \alst{B}aldrs and-skota; &
en \alst{F}rigg of grét \hld\ ï \alst{F}ęn-sǫlum &
\edtrans{\alst{v}ǫ́ \alst{V}al-hallar}{the woe of Walhall}{\Bfootnote{The deaths of two sons; Balder and Hath.}}. \hld\ \alst{V}ituð ér ęnn eða hvat?\eva

\bvb He washed ne’er his hands nor combed his head, \\
before onto the pyre he bore Balder’s opponent \ken*{= Hath}, \\
and Frie lamented in the Fenhalls \\
the woe of \inx[L]{Walhall}.—Know ye yet, or what?\evb\evg


\bvg\bva\mssnote{\Regius~2r/8, \Hauksbok~20v/13}%
\edtext{\alst{H}apt sá hǫ̇n liggja \hld\ und \alst{H}vera-lundi &
\edtrans{\alst{l}ę́-gjarns}{guile-eager}{\Bfootnote{A formulaic epithet of Lock. See note to TODO for other examples and discussion.}} líki \hld\ \alst{L}oka ȧ-þękkjan;}{\lemma{Hapt \dots\ ȧ-þękkjan ‘A captive \dots\ to Lock,’}\Afootnote{Replaced with H1 \Hauksbok.}} &
\alst{þ}ar sitr Sigyn \hld\ \alst{þ}ęygi of sïnum &
\alst{v}eri \alst{v}ęl-glýjuð. \hld\ \alst{V}ituð ér ęnn eða hvat?%
\Ballnote{After Balder was avenged, the Eese went to catch Lock.  They bound him with his son’s intestines.  A snake was then placed above his face to spit venom onto it.  His wife, Syein, sat over him and collected the venom in a small basin, but when she had to empty it he writhed so greatly from the pain of the dripping venom that the earth shook.  This event is told in detail in \Lokasenna\ P8 and \Gylfaginning\ 50.}\eva

\bvb A captive \ken*{= Lock} she saw lying beneath Wharlund: \\
a guile-eager man’s form, alike to Lock, \\
There sits Syein not at all cheerful, \\
o’er her husband.—Know ye yet, or what?\evb\evg


\bvg\bva\mssnote{\Regius~2r/10}%
\alst{Ǫ́} fęllr \alst{au}stan \hld\ of \alst{ęi}tr-dala &
\alst{s}ǫxum ok \alst{s}verðum, \hld\ \edtrans{\alst{S}líðr}{Slide}{\Bfootnote{An adjective describing a blade so sharp that the finger is cut when slid across it.  Cf. \Atlakvida\ 23: \emph{sax slíðr-bęitt} ‘slide-biting sax’.}} hęitir sú.%
\Ballnote{The function of this half-stanza is very unclear; it is not alluded to in any way in \Gylfaginning.  The river Slide is also mentioned in \Grimnismal\ 28/4 in a long list of rivers.}\eva

\bvb A river falls from the east above the venom-dales: \\
{[a river]} of saxes and swords—Slide is it called.\evb\evg


\bvg\bva\mssnote{\Regius~2r/11}%
Stóð fyr \alst{n}orðan \hld\ ȧ \edtrans{\alst{N}iða-vǫllum}{Nithwolds}{\Afootnote{\emph{Niða-fjǫllum} ‘Nithfells’ \Regius\Wormianus\ (paraphrase); \emph{fjǫllom nǫkkurum} ‘some certain fells’ \Trajectinus}} &
\alst{s}alr ór gulli \hld\ \alst{S}indra ę́ttar; &
en \alst{a}nnarr stóð \hld\ ȧ \alst{Ȯ}·kólni, &
\alst{b}jór-salr jǫtuns, \hld\ \edtrans{en sá \alst{B}rimir hęitir}{and it is called Brimmer}{\Bfootnote{It is not clear whether this is the name of the ettin or the hall itself.  \Gylfaginning\ clearly considers it the name of the hall.}}.%
\Ballnote{Sts. 36–38 are paraphrased in \Gylfaginning\ 52:

\begin{quote}
	\emph{Þá mę́lti Gangleri: „Hvat verðr þá eptir, er brenndr er himinn ok jǫrð ok heimr allr, ok dauð goð’in ǫll ok allir Ein-herjar ok alt mann-folk, ok hafið ér áðr sagt, at hverr maðr skal lifa í nǫkkvǫrum heimi um allar aldir?“}

	\emph{Þá svarar Þriði: „Margar eru þá vistir góðar ok margar illar; batst er þá at vera á Gim-léi á himni, ok all-gótt er til góðs drykkjar þeim, er þat þykkir gaman, í þeim sal, er Brimir heitir; hann stendr ok á himni. Sá er ok góðr salr, er stendr á Niða-fjǫllum, gǫrr af rauðu gulli; sá heitir Sindri. Í þessum sǫlum skulu byggja góðir menn ok sið-látir.}

	\emph{Á Ná-strǫndum er mikill salr ok illr ok horfa norðr dyrr; hann er ok ofinn allr orma-hryggjum sem vanda-hús, en orma hǫfuð ǫll vitu inn í hús’it ok blása eitri, svá at eptir sal’num renna eitr-ár, ok vaða þę́r ár eið-rofar ok morð-vargar, svá sem hér segir:“}
\end{quote}

\begin{quote}
	‘Then spoke Gangler: “What will then remain, when heaven and earth and the whole world are burned, and the gods all are dead and all the Oneharriers and all man-kind—and still ye have said earlier, that every man will live in a certain home throughout all ages?”

	Then answers Third: “Many good dwellings are there then, and many evil; then it is best to be in Gimlee in the heaven.  And in the hall which is called Brimmer it is all-furnished with good drink for those who find joy in that; that one also stands in the heaven.  Another good hall is the one which stands on the Nithfells, made out of red gold; it is called Sinder.  In these halls shall dwell good and virtuous men.

	On Neestrand is a great hall and evil, and its doors face north.  It is all interlaced with the spines of serpents like a wicker-house, but the heads of the serpents all face into the house and blow venom, so that there run through the hall rivers of venom, and in those rivers wade oath-breakers and murder-wargs, as is said here:”’
\end{quote}

after which the text quotes sts. 37 and 38/1–2, followed by the prose: \emph{En í Hver-gelmi er verst} ‘But in Wharyelmer is is worst’ and finally 38/4.}\eva

\bvb Stood to the north on the Nithwolds, \\
a hall of gold, of Sinder’s lineage \ken{dwarfs}. \\
But another one stood on Uncolner, \\
an ettin’s beer-hall, and it is called Brimmer.\evb\evg


\bvg\bva\mssnote{\Regius~2r/13, \Hauksbok~20v/19, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{S}al \edtrans{sá hǫ̇n}{she saw}{\Afootnote{\emph{vęit’k} ‘I know’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus; cf. st. 61.}} standa \hld\ \alst{s}ólu fjarri &
\alst{N}á-strǫndu ȧ, \hld\ \alst{n}orðr horfa dyrr; &
falla \alst{ęi}tr-dropar \hld\ \alst{i}nn umb ljóra, &
sá ’s \alst{u}ndinn salr \hld\ \alst{o}rma hryggjum.\eva

\bvb A hall she saw standing, far from the sun, \\
on Neestrand; north face its doors. \\
Venom-drops fall in through the smoke-vent; \\
that hall is wound with the spines of snakes.\evb\evg


\bvg\bva\mssnote{\Regius~2r/15, \Hauksbok~20v/21, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\edtrans{Sá hǫ̇n}{she saw}{\Afootnote{so \Regius; \emph{ser hon} ‘she sees’ \Hauksbok; \emph{skulu} ‘shall [be]’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}} \alst{þ}ar vaða \hld\ \alst{þ}unga strauma &
\alst{m}ęnn \alst{m}ęin-svara \hld\ ok \edtrans{\alst{m}orð-varga}{murder-wargs}{\Bfootnote{Murderous outlaws.}} &
ok þann’s \alst{a}nnars glępr \hld\ \alst{ęy}ra-ru̇nu. &
Þar \edtrans{saug}{sucked}{\Afootnote{so \Hauksbok; \emph{†súg†} \Regius; \emph{kvęlr} ‘torments’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}} \alst{N}íð-hǫggr \hld\ \alst{n}ái fram-gingna; &
slęit \alst{v}argr \alst{v}era. \hld\ \alst{V}ituð ér ęnn eða hvat?%
\Ballnote{Watery punishment in the Heathen afterlife, as described in this stanza, is also found in \Reginsmal\ 3–4 and possibly in \Grimnismal\ 21.  The crimes are what one might expect from a source in Germanic antiquity: perjury, shameful murder, and adultery with a married woman.  In Anglo-Saxon and Nordic laws the committer of such crimes got the title of \inx[C]{nithing}, that is, one afflicted with \inx[C]{nithe} (severe shame).  It is fitting, then, that such “nithings” should be tortured by a creature named Nithehewer (‘Nithe-striker’).  The practice of burying criminals, not just in the wilderness, but specifically in bogs and flood-marks, is well attested in sources relating to Germanic cultures from Tacitī \emph{Germania} onwards, and the “heavy streams” of the present stanza are clearly related to such customs.  This is further elaborated on in \textcite{GermanicGems2}.}\eva

\bvb There she saw wading through heavy streams \\
false-swearing men and murder-wargs, \\
and him who beguiles another man’s ear-whisperer \ken{wife}. \\
There sucked \inx[P]{Nithehewer} on corpses passed-on; \\
the warg tore at men.—Know ye yet, or what?\evb\evg


\bvg\bva\mssnote{\Regius~2r/17, \Hauksbok~20v/2, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\edtrans{\alst{Au}str}{In the east}{\Bfootnote{The cardinal direction associated with ettins and other monstrosities.}} \edtrans{býr}{dwells}{\Afootnote{so \Hauksbok\RegiusProse\Trajectinus\Upsaliensis\Wormianus; \emph{sat} ’sat/stayed’ \Regius}} hin \edtrans{\alst{a}ldna}{old}{\Afootnote{\emph{arma} ‘wretched’ \Upsaliensis}} \hld\ ï \edtrans{\alst{Éa}rn-viði}{Ironwood}{\Afootnote{metr. emend.; \emph{Járnviði} \Regius\Hauksbok\RegiusProse\Wormianus\Upsaliensis; \emph{Járn-viðjum} ‘Ironwoods’ \Trajectinus}} &
ok \edtrans{\alst{f}ǿðir}{nourishes}{\Afootnote{so \Hauksbok\RegiusProse\Trajectinus\Upsaliensis\Wormianus; \emph{fǿddi} ‘nourished’ \Regius}} þar \hld\ \alst{F}ęnris kindir; &
verðr \edtext{af}{\Afootnote{\emph{ór} \Trajectinus\RegiusProse}} þęim \alst{ǫ}llum \hld\ \alst{ęi}nna nøkkurr &
\alst{t}ungls \edtrans{\alst{t}júgari}{seizer}{\Afootnote{\emph{†tuigan†} \Trajectinus; \emph{tregari} ‘griever’ \Upsaliensis. As the young agentive suffix \emph{-ari} is found nowhere else in the poem it is possible that this word is corrupt. If it is, it must have occurred early in the transmission, as reflexes of \emph{tjúgari} are found in all surviving mss.}} \hld\ ï \alst{t}rolls hami.%
\Ballnote{The old hag (who is not identified) raises the cubs of the wolf Fenrer; one of these will eventually swallow the moon.  For the myth of the wolves that chase the sun and moon cf. \textlink{Grimnismal}[40], according to which the sun is chased by the wolf Scoll, while another wolf, Hater, runs in front of it, and \textlink{Vafthrudnismal}[46]–47, where the Sun is said to be swallowed by Fenrer.

\Gylfaginning\ 12 says that Hater will swallow the moon, and continues:

\emph{Gýgr ein býr fyrir austan Miðgarð í þeim skógi, er Járnviðr heitir.  [...]  In gamla gýgr fǿðir at sonum marga jǫtna ok alla í vargs líkjum, ok þaðan af eru komnir þessir ulfar. Ok svá er sagt, at af ę́ttinni verðr sá einna máttkastr, er kallaðr er Mána-garmr. Hann fyllist með fjǫrvi allra þeira manna, er deyja, ok hann gleypir tungl, en stǫkkvir blóði himin ok loft ǫll.  Þaðan týnir sól skini sínu, ok vindar eru þá ó·kyrrir ok gnýja heðan ok handan.  Svá segir í Vǫlu-spá:}

‘A lonely ogress lives to the east of Middenyard in the forest called Ironwood.  The old ogress raises many ettins as her sons, all in the likenesses of wolves, and thereof these wolves [Scoll and Hater] come.  And so it is said that from that lineage a single one becomes the mightiest, and he is called \inx[P]{Moongarm}.  He fills himself with the life of all those men who die; and he swallows the moon and stains the heaven and all the air with blood. From that the sun loses her shine, and then the winds are violent and howl hither and thither, and thus it says in the Spae of the Wallow:’

after which it quotes \Voluspa\ 40–41.  \Gylfaginning\ is clearly systematizing from several sources, and it seems likely that the Heathen poets themselves were not entirely in agreement about the exact details of these events.}\eva

\bvb In the east dwells the old woman in \inx[L]{Ironwood} \\
and there nourishes the brood of \inx[P]{Fenrer} \ken{wolves}. \\
Out of them all one most certain arises: \\
a seizer of the Moon in a troll’s \inx[C]{hame}.\evb\evg


\bvg\bva\mssnote{\Regius~2r/19, \Hauksbok~20v/4, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{F}yllisk \alst{f}jǫrvi \hld\ \alst{f}ęigra manna, &
\alst{r}ýðr \alst{r}agna sjǫt \hld\ \alst{r}auðum dręyra, &
\alst{s}vǫrt verða \alst{s}ól-skin \hld\ of \alst{s}umur ęptir, &
\alst{v}eðr ǫll \alst{v}á-lynd. \hld\ \alst{V}ituð ér ęnn eða hvat?\eva

\bvb He fills himself with the lifeblood of \inx[C]{fey} men; \\
he reddens the seat of the \inx[G]{Reins} \ken{heaven} with red gore. \\
Black turn the sun’s rays in the summers thereafter, \\
the winds all woeful.—Know ye yet, or what?\evb\evg


\bvg\bva\mssnote{\Regius~2r/21, \Hauksbok~20v/16}%
\edtrans{\alst{S}at þar ȧ haugi}{There sat on the mound}{\Bfootnote{The motif of ettins sitting on burial mounds is also found in \Thrymskvida\ 6 and \Skirnismal\ P2.  The significance of this is uncertain,.}} \hld\ ok \alst{s}ló hǫrpu &
\alst{g}ýgjar hirðir, \hld\ \alst{g}laðr Ęggþér; &
\alst{g}ól of hǫ̇num \hld\ ï \edtrans{\alst{G}agl-viði}{Galewood}{\Bfootnote{An otherwise unknown location; the first element is \emph{gagl} ‘wild goose’.  Galewood is perhaps the same as Ironwood.}} &
\alst{f}agr-rauðr hani, \hld\ sá’s \alst{F}jalarr hęitir.%
\Ballnote{Edgethew is apparently the keeper of the monstrous wolves; this detail is not found in \Gylfaginning.}\eva

\bvb There sat on the mound and struck the harp \\
the gow’s herdsman, glad \inx[P]{Edgethew}. \\
Over him crowed in \inx[L]{Galewood} \\
a fair-red cock, he who is called Feller.\evb\evg


\bvg\bva\mssnote{\Regius~2r/23, \Hauksbok~20v/18}%
\alst{G}ól of ǫ̇sum \hld\ \alst{G}ullin-kambi, &
sá vękr \alst{h}ǫlða \hld\ at \alst{H}ęrja-fǫðrs, &
en \alst{a}nnarr gęlr \hld\ fyr \alst{jǫ}rð neðan &
\alst{s}ót-rauðr hani \hld\ at \alst{s}ǫlum Hęljar.\eva

\bvb Over the Eese crowed Goldencomb; \\
he wakes men at the Father of Hosts’s \name{= Weden’s}— \\
but another one crows beneath the earth: \\
a soot-red cock at the halls of Hell.\evb\evg


\bvg\bva\mssnote{\Regius~2r/25}%
\alst{G}ęyr \alst{G}armr mjǫk \hld\ fyr \alst{G}nipa-hęlli, &
\alst{f}ęstr mun slitna, \hld\ en \alst{F}reki rinna; &
\alst{f}jǫlð vęit hǫ̇n \alst{f}rǿða, \hld\ \alst{f}ramm sé’k lęngra &
of \alst{r}agna \alst{r}ǫk, \hld\ \alst{r}ǫmm sig-tíva.%
\Ballnote{With the crowing of these three cocks—the first in Ettinham, the second in Walhall, the third in Hell—the destruction of the world begins, as indicated by the first occurrence of the refrain stanza (ON \emph{stęf}).}\eva

\bvb Garm barks loud before the Gnip-caves; \\
the rope will tear and the Wolf will run. \\
She knows much lore; I foresee further \\
about the mighty \inx[L]{Rakes of the Reins}, of the Victory-Tews \ken{gods}.\evb\evg


\bvg\bva\mssnote{\Regius~2r/28, \Hauksbok~20v/24, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{B}rǿðr munu \alst{b}ęrjask \hld\ ok at \alst{b}ǫnum verðask, &
munu \edtrans{\alst{s}ystrungar}{the children of sisters}{\Afootnote{\emph{†stystrungar†} \Trajectinus}} \hld\ \edtrans{\alst{s}ifjum spilla}{defile the kinship}{\Bfootnote{I.e. “commit incest”, probably referring to marriages between first cousins.  Compare related words found in the laws, like \emph{frę́nd-semis-spell} ‘incest’ and especially \emph{sifja-spell} ‘id.’ — The idea of incest as a sign of later ages is also found in \Rigveda\ 10.10.10a–b (norm. and tr., Nikhil S. Dwibhashyam. (2025, Aug. 31). \emph{Véda quote 6}. \url{https://nikhilsd.com/dvq/6/}): \emph{Ā́ gʰā tā́ gaccʰān \hld\ úttarā yugā́ni, // yátra jāmáyaḥ \hld\ kr̥ṇávann ájāmi} ‘There shall come indeed those later ages when relatives shall do (acts) not (fit for) relatives.’}}; &
\alst{h}art ’s \edtrans{ï \alst{h}ęimi}{in the Home}{\Afootnote{so \Regius\Hauksbok\Upsaliensis; \emph{með hǫlðum} ‘among men’ \RegiusProse\Trajectinus\Wormianus}}, \hld\ \alst{h}ór-dȯmr mikill, &
\alst{sk}ęggj-ǫld, \alst{sk}alm-ǫld, \hld\ \edtrans{\alst{sk}ildir}{shields}{\Afootnote{\emph{’ru} ‘are’ add. \Regius}} \edtrans{klofnir}{split}{\Afootnote{\emph{klofna} ‘become split’ \Upsaliensis}}, &
\edtrans{\alst{v}ind-ǫld}{wind-age}{\Bfootnote{In \Hauksbok\ the \emph{v} is capitalized, marking the beginning of a new stanza.}}, \alst{v}arg-ǫld, \hld\ \edtrans{áðr}{before}{\Afootnote{\emph{unz} (norm.) ‘until’ \Upsaliensis}} \edtrans{\alst{v}er-ǫld}{man-age}{\Bfootnote{Translated as such since it stands next to various other compounds ending in \emph{ǫld} ‘age’.  ON \emph{ver-ǫld} is cognate with English “world”, but in ON that sense is usually expressed with \emph{hęimr} (e.g. l. 3 of the present stanza).}} \edtrans{stęypisk}{tumbles down}{\Bfootnote{\emph{grundir gjalla \hld\ gífr fljúgandi} (norm.) ‘foundations shrill, fiends flying’ add. after this l. \Hauksbok}} &
\edtext{mun \edtext{\alst{ę}ngi}{\Afootnote{\emph{†enn†} \Upsaliensis}} maðr \hld\ \alst{ǫ}ðrum þyrma.}{\lemma{mun \dots\ þyrma ‘before \dots\ spare’}\Bfootnote{om. \RegiusProse\Trajectinus\Wormianus}}\eva

\bvb Brothers will fight and become each other’s slayers; \\
the children of sisters will defile the kinship. \\
’Tis hard in the Home; whoredom is great: \\
axe-age, sword-age—shields are split— \\
wind-age, warg-age! Before the man-age tumbles \\
will no man another spare.\evb\evg


\bvg\bva\mssnote{\Regius~2r/32, \Hauksbok~20v/27, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\edtext{Lęika \alst{M}íms synir, \hld\ en \alst{m}jǫtuðr kyndisk &
at hinu \alst{g}alla \hld\ \alst{G}jallar-horni;}{\lemma{Lęika \dots\ Gjallar-horni; ‘Mime’s \dots\ Yell.’}\Bfootnote{om. \RegiusProse\Trajectinus\Upsaliensis\Wormianus}} &
\alst{h}ǫ́tt blę́ss \alst{H}ęimdallr, \hld\ \alst{h}orn ’s ȧ lopti; &
\edtrans{\alst{m}ę́lir}{speaks}{\Afootnote{\emph{†mey†} \RegiusProse; \emph{†nie†} \Trajectinus}} Óðinn \hld\ við \alst{M}íms hǫfuð; &
\edtext{skęlfr \alst{Y}ggdrasils \hld\ \alst{a}skr standandi, &
\alst{y}mr it \alst{a}ldna tré, \hld\ en \alst{jǫ}tunn \edtext{losnar}{\Afootnote{add. H2 \Hauksbok.}}.}{\lemma{Skęlfr \dots\ losnar ‘Ugdrassle’s \dots\ loosens’}\Afootnote{so \Hauksbok\RegiusProse\Trajectinus\Upsaliensis\Wormianus; these lines are reversed in \Regius.}}%
\Ballnote{With the omission of the refrain-stanza 47, sts. 45–54 are cited in sequence in \Gylfaginning\ 51.}\eva

\bvb Mime’s sons play and the Measure-Tree is kindled \\
after [the sounding of] the shrill Horn of Yell. \\
High blows Homedal; the horn is aloft; \\
Weden speaks with the head of Mime. \\
Ugdrassle’s Ash trembles, standing: \\
the old tree whines and the ettin loosens.\evb\evg


\bvg\bva\mssnote{\Regius~2v/8, \Hauksbok~20v/30, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\edtrans{Hvat ’s með \alst{ǫ̇}sum? \hld\ hvat ’s með \edtrans{\alst{ǫ}lfum}{Elves}{\Afootnote{\emph{ǫ̇synjum} ‘Ossens’ \Upsaliensis}}}{What is with the Eese? What is with the Elves?}{\Bfootnote{Also occuring in \Thrymskvida\ 7/1.}}? &
\edtext{gnýr \alst{a}llr \alst{Jǫ}tun-hęimr, \hld\ \alst{ę̇}sir ’ru ȧ \edtrans{þingi}{the Thing}{\Bfootnote{The \inx[L]{Thing of the Gods} or the Divine Council; see note to st 6/1–2 and Index.}},}{\lemma{gnýr \dots\ þingi}\Afootnote{om. \Upsaliensis}} &
\alst{st}ynja dvergar \hld\ fyr \edtext{\alst{st}ęin-durum}{\Afootnote{\emph{stęins} \Upsaliensis; \emph{stęin-dyrum} \Hauksbok\Wormianus\Upsaliensis}} &
\edtext{\edtext{\alst{v}ęgg-bergs}{\Afootnote{\emph{veg-bergs} \Hauksbok\Trajectinus\Wormianus}} \alst{v}ísir}{\Afootnote{om. \Upsaliensis}}. \hld\ \alst{V}ituð ér ęnn eða hvat?\eva

\bvb What is with the Eese? What is with the Elves? \\
All Ettinham roars—the Eese are at the Thing! \\
Dwarfs groan before gates of stone \\
the cliff-side princes.—Know ye yet, or what?\evb\evg


\bvg\bva\mssnote{\Regius~2v/4, \Hauksbok~20v/32}%
\alst{G}ęyr nú \alst{G}armr mjǫk \hld\ fyr \alst{G}nipa-hęlli, &
\alst{f}ęstr mun slitna, \hld\ en \alst{f}reki rinna; &
\alst{f}jǫlð vęit hǫ̇n \alst{f}rǿða, \hld\ \alst{f}ramm sé’k lęngra &
of \alst{r}agna \alst{r}ǫk \hld\ \alst{r}ǫmm sig-tíva.\eva

\bvb Now Garm barks loud before the Gnip-caves; \\
the rope will tear and the Wolf will run. \\
She knows much lore; I foresee further \\
about the mighty Rakes of the Reins, of the Victory-Tews \ken{gods}.\evb\evg


\bvg\bva\mssnote{\Regius~2v/4, \Hauksbok~20v/32, \RegiusProse\Trajectinus\Wormianus}%
\alst{H}rymr ękr austan, \hld\ \alst{h}ęfsk lind fyrir, &
snýsk \alst{Jǫ}rmun-gandr \hld\ ï \alst{jǫ}tun-móði, &
\alst{o}rmr knýr \alst{u}nnir, \hld\ \edtrans{en \alst{a}ri hlakkar}{and the eagle screams}{\Afootnote{\emph{ǫrn mun hlakka} ‘the eagle will scream’ \RegiusProse\Trajectinus}}, &
slítr \alst{n}ái \alst{n}ęf-fǫlr; \hld\ \alst{N}agl-far losnar.\eva

\bvb Rim drives from the east; he holds his shield before him; \\
Ermingand writhes in ettin-wrath. \\
The Wyrm pushes waves and the eagle screams: \\
the pale-beak tears at corpses—Nailfare loosens.\evb\evg


\bvg\bva\mssnote{\Regius~2v/6, \Hauksbok~20v/34, \RegiusProse\Trajectinus\Wormianus}%
\alst{K}jóll fęrr austan \hld\ \alst{k}oma munu Múspells &
of \alst{l}ǫg \alst{l}ýðir, \hld\ en \alst{L}oki stýrir; &
\alst{f}ara \alst{f}ífl-męgir \hld\ með \alst{f}reka allir, &
þęim es \alst{b}róðir \hld\ \alst{B}ýlęists ï fǫr.\eva

\bvb A ship fares from the east—come will Muspell’s \\
subjects o’er the sea—and Lock steers it. \\
The devil-lads all journey with the Wolf; \\
with them comes the brother of Bylest \ken*{= Lock} along.\evb\evg


\bvg\bva\mssnote{\Regius~2v/10, \Hauksbok~20v/36, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\edtext{\alst{S}urtr}{\Afootnote{\emph{Svartr} \Upsaliensis}} fęrr \alst{s}unnan \hld\ með \alst{s}viga lę́vi, &
\edtrans{skïnn af \alst{s}verði \hld\ \alst{s}ól val-tíva}{from the sword of the slaughter-Tews it shines [like] the sun}{\Bfootnote{A difficult line in several regards.  First, \emph{val-tíva} may be read either (1) as gen. pl. of \emph{val-tívar} ‘slain-Tews \ken{gods}’ (so \textcite{LaFargeGlossary}; cf. below st. 59 and \Hymiskvida\ 1) or (2) as gen. sg. of an unattested masc. \emph{n}-stem \emph{*val-tívi} ‘slaughter-tew \ken*{= Surt}’ (so \CV, Fritzner and others).
Here, (1) is adopted as the simpler explanation, since it does not presuppose an otherwise unattested derivative of \emph{týr} ‘tew, god’, and since the pl. \emph{val-tívar} demonstrably occurs below in \Voluspa\ 59, but the sense of the line is still very difficult.
A second issue arises over whether the gen. \emph{val-tíva} modifies (a) \emph{sverði} ‘sword’ or (b) \emph{sól} ‘sun’.  In (1a) and (2a) the sense would simply be that the flashing sword shines like the sun, reflecting the intensity of the battle.  Although one may expect a pl. \emph{sverðum} ‘swords’ for case (1a), it is not a major problem, since sg. forms are often used with pl. possessors in Old Germanic languages.  Case (1b) must probably be rejected, since the possessor of the sword would be Surt, and the sentence “the sun of the Gods shines from [Surt’s] sword” makes little sense.  If (2b) is adopted we may see a kenning “sun of Surt \ken{fire}”, but “from his sword shines the \ken{fire}” seems excessively repetitive when compared to l. 1.}}; &
\alst{g}rjót-bjǫrg \alst{g}nata, \hld\ en \edtrans{\alst{g}ífr rata}{the fiends are on the march}{\Afootnote{\emph{guðar hrata} ‘the gods stagger’ \Upsaliensis}\Bfootnote{The reading of \Upsaliensis\ is wo. doubt corrupt; the anachronistic masc. pl. ending \emph{-ar} is proof enough, for the word \emph{goð} \char`~\ \emph{guð} ‘gods’ was always neuter in heathen times.}}, &
troða \alst{h}alir \edtrans{\alst{h}ęl-veg}{Hellway}{\Bfootnote{The road on which one has to travel after death to reach his final resting place.  Cf. \Helreid.}}, \hld\ en \alst{h}iminn klofnar.\eva

\bvb Surt fares from the south with the twig’s betrayer \ken{fire}; \\
from the sword of the slaughter-Tews it shines [like] the sun. \\
Rocky cliffs clash and the fiends are on the march; \\
men tread the \inx[L]{Hellway} and heaven is rent.\evb\evg


\bvg\bva\mssnote{\Regius~2v/13, \Hauksbok~20v/37, \RegiusProse\Trajectinus\Wormianus}%
Þȧ kømr \edtrans{\alst{H}línar \hld\ \alst{h}armr annarr}{Line’s second sorrow}{\Bfootnote{The first sorrow being the death of Balder.  Line is described in \Gylfaginning\ 35 as a minor goddess \emph{sett til gę́zlu yfir þeim mǫnnum, er Frigg vill forða við háska nǫkkurum} ‘placed to watch over those men which Frie wishes to save from any particular danger’.  In spite of this almost all translators and editors have understood Line as synonymous with Frie and questioned whether her existence as a distinct goddess is not something invented by the author of \Gylfaginning.  \textcite{Hopkins2017} reasonably argues that this need not be the case; as Frie’s maidservant, Line’s two sorrows would consist in her failure to protect both the son and husband of her mistress.}} framm, &
es \alst{Ó}ðinn fęrr \hld\ við \alst{u}lf vega, &
—en \edtrans{\alst{b}ani \alst{B}ęlja}{the bane of Bellower \ken*{= Free}}{\Bfootnote{Bellower (ON \emph{Bęli}) was slain by Free in an obscure duel; see Index.}} \hld\ \alst{b}jartr at Surti— &
þȧ mun \alst{F}riggjar \hld\ \alst{f}alla \edtext{angan}{\Afootnote{so \Hauksbok\RegiusProse\Trajectinus\Upsaliensis\Wormianus; \emph{angantyr} \Regius}}.%
\Ballnote{Sts. 51–52 describe how Weden fights the Wolf and dies, and how he is then avenged by Wider.  This event is also mentioned in \textlink{Vafthrudnismal}[53].}\eva

\bvb Then comes \inx[P]{Line}’s second sorrow to pass, \\
when Weden goes to fight the Wolf \\
—but the bane of Bellower \ken*{= Free}, bright, against Surt— \\
then will Frie’s beloved \ken*{= Weden} fall.\evb\evg


\bvg\bva\mssnote{\Regius~2v/15, \RegiusProse\Trajectinus\Wormianus}%
\edtrans{Þȧ kømr hinn \alst{m}ikli \hld\ \alst{m}ǫgr Sig-fǫður}{Then comes the great lad of Syefather}{\Afootnote{\emph{Gęngr Óðins sonr \hld\ við ulf vega} ‘Weden’s son goes the Wolf to fight’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus.}}, &
\alst{V}íðarr \edtext{\alst{v}ega}{\Afootnote{\emph{of veg} \RegiusProse\Trajectinus\Upsaliensis\Wormianus}} \hld\ at \alst{v}al-dýri; &
lę́tr \alst{m}ęgi \edtrans{Hveðrungs}{Whethring}{\Bfootnote{An obscure name for \inx[P]{Lock}, whose son is the Wolf.}} \hld\ \alst{m}und of standa &
\alst{h}jǫr til \alst{h}jarta; \hld\ þȧ ’s \alst{h}efnt fǫður.\eva

\bvb Then comes the great lad of \inx[P]{Syefather}, \\
Wider, to fight that slaughter-beast. \\
He lets his hand through \inx[P]{Whethring}’s lad \ken*{= the Wolf} \\
drive the sword to the heart—then the father is avenged!\evb\evg


\bvg\bva\mssnote{\Regius~2v/17, \Hauksbok~20v/41, \RegiusProse\Trajectinus\Wormianus}%
\edtrans{\edtrans{Þȧ kømr}{Then comes}{\Afootnote{\emph{Gęngr} ‘Goes’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}} hinn \alst{m}ę́ri \hld\ \alst{m}ǫgr \edtrans{Hlǫðynjar}{Lathyn}{\Afootnote{add. \emph{gęngr Óðins sonr \hld\ við orm vega.} ‘Weden’s son goes the Wyrm to fight.’ \Regius.}},}{Then comes the famed lad of Lathyn}{\Afootnote{om. \Hauksbok.}} &
\edtrans{gęngr \alst{f}et níu \hld\ \alst{F}jǫrgynjar burr}{nine paces takes Firgyn’s son}{\Afootnote{om. \RegiusProse\Trajectinus\Upsaliensis\Wormianus.}} &
\alst{n}ęppr frȧ \alst{n}aðri \hld\ \alst{n}íðs ȯ-kvíðnum; &
\edtrans{munu \alst{h}alir allir \hld\ \alst{h}ęim-stǫð ryðja}{All men will clear their homesteads}{\Bfootnote{After Thunder is slain the Earth is no longer habitable.  Cf. \Harbardsljod\ TODO, \Thrymskvida\ 18.}} &
\edtext{es af \alst{m}óði drepr}{\Afootnote{\emph{drepr hann af móði} \Regius}} \hld\ \edtrans{\alst{M}ið-garðs véurr}{Middenyard’s Wighward}{\Bfootnote{“The Guardian of the Sanctuaries of Middenyard”; a fitting kenning.}}.%
\Ballnote{Thunder and the Middenyardswyrm slay each other, fulfilling the unfinished fight seen in the Fishing Expedition; see \Hymiskvida.

The present version of the stanza is an amalgamation of all three ms. traditions (\Regius, \Hauksbok\ and \RegiusProse\Trajectinus\Upsaliensis\Wormianus) based most closely on the latter two, which have the last 3 lines in the same order.  \Regius\ has the lines in the following order: 1, 5, 4, 2, 3, and also inserts its own line between 1 and 5.}\eva

\bvb Then comes the famed lad of Lathyn; \\
nine paces takes Firgyn’s son \\
pained, away from the shameful Adder. \\
All men will clear their homesteads \\
when Middenyard’s Wighward strikes out of wrath.\evb\evg


\bvg\bva\mssnote{\Regius~2v/20, \Hauksbok~21r/1, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{S}ól tér \alst{s}ortna, \hld\ \edtrans{\edtrans{\alst{s}økkr}{sinks}{\Afootnote{so \RegiusProse\Trajectinus\Wormianus; \emph{sígr} ‘descends’ \Regius\Hauksbok\Upsaliensis}} fold ï mar}{the fold sinks into the sea}{\Bfootnote{The reading \emph{søkkr} ‘sinks’ is supported by Arn \emph{Þorfdr} 24 (\Skp\ II), which is probably based on the present line: \emph{Bjǫrt verðr sól at svartri; \hld\ søkkr fold ï mar døkkvan;} ‘Bright, the sun turns to black; the fold sinks into the dark sea’.}}, &
\alst{h}verfa af \alst{h}imni \hld\ \alst{h}ęiðar stjǫrnur; &
gęisar \alst{ęi}mi \hld\ við \alst{a}ldr-nara; &
lęikr \alst{h}ǫ́r \alst{h}iti \hld\ við \alst{h}imin sjalfan.\eva

\bvb The Sun doth blacken; land sinks in sea; \\
from heaven fade the shining stars. \\
Smoke billows from the nourisher of life \ken{fire}; \\
the high heat licks the very heaven.\evb\evg


\bvg\bva\mssnote{\Regius~2v/22, \Hauksbok~21r/2}%
\alst{G}ęyr nú \alst{G}armr mjǫk \hld\ fyr \alst{G}nipa-hęlli, &
\alst{f}ęstr mun slitna, \hld\ en \alst{f}reki rinna; &
\alst{f}jǫlð vęit hǫ̇n \alst{f}rǿða, \hld\ \alst{f}ramm sé’k lęngra &
of \alst{r}agna \alst{r}ǫk, \hld\ \alst{r}ǫmm sig-tíva.%
\Ballnote{With the sinking of the Earth into the sea and the blackening of the sky the destruction has reached its climax, as signalled by the final repetition of the refrain stanza.  To indicate the dramatic pause, a page break has been inserted.}\eva

\bvb Now Garm barks loud before the Gnip-caves; \\
the rope will tear and the Wolf will run. \\
She knows much lore; I foresee further \\
about the mighty Rakes of the Reins, of the Victory-Tews \ken{gods}.\evb\evg

%TODO: Insert image?

\sectionline\newpage

\bvg\bva\mssnote{\Regius~2v/23, \Hauksbok~21r/4}%
Sér hǫ̇n \alst{u}pp koma \hld\ \edtrans{\alst{ǫ}ðru sinni}{a second time}{\Bfootnote{The first time probably being the lifting of the Earth in st. 4.}} &
\alst{jǫ}rð ór \alst{ę́}gi \hld\ \alst{i}ðja-grø̇na; &
\alst{f}alla \alst{f}orsar, \hld\ \alst{f}lýgr ǫrn yfir, &
sá’s ȧ \alst{f}jalli \hld\ \alst{f}iska vęiðir.%
\Ballnote{Sts. 56–59 are paraphrased in \Gylfaginning\ ch. 53:

\begin{quote}
	\emph{Þá mę́lti Gangleri: „Hvárt lifa nǫkkur goð’in þá, eða er þá nǫkkur jǫrð eða himinn?“ Hárr segir: „Upp skýtr jǫrðu’nni þá ór sę́’num, ok er þá grǿn ok fǫgr. Vaxa þá akrar ó·sánir. Víðarr ok Váli lifa, svá at eigi hefir sę́r’inn ok Surta-logi grandat þeim, ok byggja þeir á Iða-velli, þar sem fyrr var Ás-garðr, ok þar koma þá synir Þórs, Móði ok Magni, ok hafa þar Mjǫllni. Því nę́st koma þar Baldr ok Hǫðr frá Heljar, setjast þá allir samt, ok talast við, ok minnast á rúnar sínar, ok rǿða of tíðendi þau, er fyrrum hǫfðu verit, of Mið·garðs-orm ok um Fenris-ulf.  Þá finna þeir í grasi’nu gull-tǫflur þę́r, er ę́sir’nir hǫfðu átt. Svá er sagt:“}
\end{quote}

\begin{quote}
	‘Then spoke Gangler: “Do any of the gods live then, or is there then any earth or heaven?”  High says: “Then the earth shoots up from the seas, and it is then green and fair.  Then acres grow unsown.  Wider and Wonnel live, for the sea and Surt’s flame have not scathed them, and they settle on the Idewolds where Osyard once stood.  And then the sons of Thunder—Mood and Main—come there, and there they have Millner.  Next come Balder and Hath from Hell; then they all sit down together and make speech and think back on their runes and discuss the events of antiquity, about the Middenyardswyrm and about the Fenrerswolf.  Then they find in the grass those golden game-bricks which the Eese had owned.  So it is said:”’
\end{quote}

after which it quotes \textlink{Vafthrudnismal}[51].}\eva

\bvb She sees coming up a second time \\
Earth from the ocean, ever green anew. \\
Torrents fall, flies the eagle above, \\
which on the fells catches fish.\evb\evg


\bvg\bva\mssnote{\Regius~2v/24, \Hauksbok~21r/5}%
\edtrans{Finnask}{find each other}{\Afootnote{\emph{hittask} \Hauksbok\ provides closer parallelism with st. 7, but for the same reason it may also have replaced earlier \emph{finnask}.}} \alst{ę̇}sir \hld\ ȧ \alst{I}ða-vęlli &
ok umb \edtrans{\alst{m}old-þinur}{Earth-cord}{\Bfootnote{Cf. the kenning for the Middenyardswyrm in ÚlfrU \emph{Húsdr} 4: \emph{stirð-þinull storðar} ‘the stiff cord of the land \ken*{= Middenyardswyrm}’}} \hld\ \alst{m}ǫ́tkan dø̇ma, &
\edtrans{ok \alst{m}innask þar \hld\ ȧ \alst{m}ęgin-dȯma}{and there think back on mighty verdicts}{\Afootnote{om. \Regius}} &
ok ȧ \alst{F}imbul-týs \hld\ \alst{f}ornar ru̇nar.\eva

\bvb The Eese find each other on the Idewolds \\
and of the mighty Earth-cord \ken*{= Middenyardswyrm} speak, \\
and there think back on mighty verdicts \\
and on Fimble-Tew’s \name{= Weden’s} ancient runes.\evb\evg


\bvg\bva\mssnote{\Regius~2v/26, \Hauksbok~21r/7}%
Þar munu \alst{ę}ptir \hld\ \edtext{\alst{u}ndr-samligar &
\alst{g}ullnar tǫflur}{\lemma{undr-samligar gullnar tǫflur ‘wondersome golden game-bricks’}\Bfootnote{A fine literary device.  In st. 8 the golden age of the Eese, exemplified by their playing board games, was spoiled by the three ettin-women.  The rediscovering of the golden board game then betokens a new golden age.}} \hld\ ï \alst{g}rasi finnask, &
þę́r’s \edtext{\emph{\alst{ę̇}sir}}{\Afootnote{emend.; om. \Regius\Hauksbok}\Bfootnote{The verb \emph{hǫfðu} ‘had’ requires a plural subject, and this is found in the paraphrase of \Gylfaginning\ 53: \emph{Þá finna þeir í grasi’nu gull-tǫflur þę́r, er ę́sir’nir hǫfðu átt.} (see above note to st. 56).}} ï \alst{á}r-daga \hld\ \alst{á}ttar hǫfðu.\eva

\bvb There will again the wondersome \\
golden game-bricks in the grass be found, \\
which in days of yore the Eese had owned.\evb\evg


\bvg\bva\mssnote{\Regius~2v/28, \Hauksbok~21r/9}%
Munu \alst{ȯ}-sánir \hld\ \alst{a}krar vaxa, &
\edtrans{\alst{b}ǫls}{the bale}{\Bfootnote{The evil of Hath’s slaying Balder will be forgotten as the two live together in peace.}} mun alls \alst{b}atna, \hld\ mun \alst{B}aldr koma; &
búa \alst{H}ǫðr ok Baldr \hld\ \alst{H}ropts sig-toptir, &
\alst{v}ęl \alst{v}al-tívar. \hld\ \alst{V}ituð ér ęnn eða hvat?\eva

\bvb Unsown will acres grow; \\
the bale will all be bettered; Balder will come. \\
Hath and Balder bedwell Roft’s \name{= Weden’s} victory-plots \\
well, the slain-Tews.—Know ye yet, or what?\evb\evg


\bvg\bva\mssnote{\Regius~2v/30, \Hauksbok~21r/11}%
Þȧ kná \alst{H}ø̇nir \hld\ \edtrans{\alst{h}laut-við kjósa}{choose the leat-wood}{\Bfootnote{Foresee the future by means of lots, specifically twigs drenched in holy blood.  See \Hymiskvida\ 1 and Index: \inx[C]{leat}.}} &
ok \alst{b}urir \alst{b}yggva \hld\ \edtrans{\alst{b}rǿðra tvęggja}{the two brothers}{\Bfootnote{The present translation understands \emph{tvęggja} as the gen. pl. of \emph{tvęir} ‘two’; the two brothers are presumably Hath and Balder, mentioned in the previous stanza.
Since the original ms. does not capitalize proper nouns one could also read \emph{brǿðra Tvęggja} ‘the brothers of Tway \name{= Weden}’.  Weden’s brothers are attested in \Gylfaginning\ 6 as \inx[P]{Will} and \inx[P]{Wigh}; they are never said to have children.}} &
\alst{v}ind-hęim \alst{v}íðan. \hld\ \alst{V}ituð ér ęnn eða hvat?\eva

\bvb Then does Heener choose the \inx[C]{leat}-wood, \\
and the sons of the two brothers settle \\
the wide wind-home \ken{sky/heaven}.—Know ye yet, or what?\evb\evg


\bvg\bva\mssnote{\Regius~2v/31, \Hauksbok~21r/12, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{S}al \edtrans{sér hǫ̇n}{she sees}{\Afootnote{\emph{vęit’k} ‘I know’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}} standa \hld\ \alst{s}ólu fęgra, &
\edtrans{\alst{g}ulli þakðan}{thatched with gold}{\Afootnote{\emph{gulli bętra} ‘better than gold’ \RegiusProse\Trajectinus}}, \hld\ ȧ \edtext{\alst{G}imléi}{\Afootnote{metr. emend.; \emph{Gimlé} \Regius\Hauksbok\RegiusProse\Trajectinus\Upsaliensis\Wormianus}}; &
\edtrans{þar}{there}{\Afootnote{\emph{þann} ‘[in] that [hall]’ \Trajectinus\Wormianus}} skulu \alst{d}yggvar \hld\ \alst{d}róttir byggva &
ok umb \alst{a}ldr-daga \hld\ \alst{y}nðis njóta.\eva

\bvb A hall she sees standing, fairer than the sun, \\
thatched with gold, on Gemlee. \\
There shall faithful folk settle, \\
and in their days of life enjoy delight.\evb\evg


\bvg\bva\mssnote{\Regius~3r/2, \Hauksbok~21r/15}%
Þar kømr hinn \alst{d}immi \hld\ \alst{d}ręki fljúgandi, &
\alst{n}aðr frȧnn \alst{n}eðan \hld\ frȧ \alst{N}iða-fjǫllum; &
berr sér ï \alst{f}jǫðrum \hld\ —\alst{f}lýgr vǫll yfir— &
\alst{N}íð-hǫggr \alst{n}ái; \hld\ \edtrans{\alst{n}ú mun hǫ̇n søkkvask}{Now she will sink!}{\Bfootnote{The spae is concluded and the wallow, referring to herself in third person, descends back down into the grave whence Weden woke her.  Cf. the very last half-line of \Helreid: \emph{søkkst-u, gýgjar-kyn} ‘sink, thou gow’s kin!’}}.\eva

\bvb Then comes the gloomy dragon flying, \\
the gleaming adder up from the \inx[L]{Nithfells}. \\
He carries in his feathers—he flies over the field— \\
Nithehewer, corpses.—Now she will sink!”\evb\evg

\sectionline

\section{Appendix}

\subsection{Dwarf-tallies}

The following sts. (11–15) contain two originally distinct lists of dwarf names.  At least some of them are certainly later inserts; it is proof enough that there is a repetition of names (Oakenshield, Great-grandfather) and more than one formulaic conclusion.  They interrupt the flow of the whole poem so much that they have been moved hither.

Sts. 11–13, having no repeated names, seem to belong together. If they do, st. 12, which contains the formulaic conclusion to the list, should probably switch places with 13.

Sts. 14–15 form the second group, having an introduction and a conclusion which both mention the dwarf Loffer.

\bvg\bva[11]\mssnote{\Regius~1r/23, \Hauksbok~20r/17, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{N}ýi ok \alst{N}iði, \hld\ \alst{N}orðri, Suðri, &
\alst{Au}stri, Vestri, \hld\ \alst{A}l-þjófr, Dvalinn, &
\alst{B}ívurr, \alst{B}ávurr, \hld\ \alst{B}ǫmburr, Nóri, &
\alst{Ȧ}nn ok \alst{Ȧ}narr, \hld\ \alst{Á}i, Mjǫð-vitnir.\eva

\bvb New and Nithe, Norther and Souther, \\
Easter and Wester, Allthief, Dwollen, \\
Bewer, Bower, Bamber, Noor, \\
Own and Owner, Great-grandfather, Meadwitner.\evb\evg


\bvg\bva[12]\mssnote{\Regius~1r/25, \Hauksbok~20r/18, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{V}ęigr ok Gand-alfr, \hld\ \alst{V}ind-alfr, Þráinn, &
\alst{Þ}ękkr ok \alst{Þ}orinn, \hld\ \alst{Þ}rór, Vitr ok Litr, &
\alst{N}ár ok \alst{N}ý-ráðr— \hld\ \alst{n}ú hęf’k dverga &
—\alst{R}ęginn ok \alst{R}áð-sviðr— \hld\ \alst{r}étt of talða.\eva

\bvb Wey and Gandelf, Windelf, Thrown, \\
Thetch and Thorn, Threw, Wit and Lit, \\
Nee and Newred—now have I the dwarfs— \\
Rain and Redswith—rightly tallied.\evb\evg


\bvg\bva[13]\mssnote{\Regius~1r/28, \Hauksbok~20r/20, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{F}íli, Kíli, \hld\ \alst{F}undinn, Náli, &
\alst{H}ępti, Víli, \hld\ \alst{H}annarr, Svíurr, &
\alst{F}rár, Horn-bori, \hld\ \alst{F}rę́gr ok Lȯni, &
\alst{Au}r-vangr, \alst{Ja}ri, \hld\ \alst{Ęi}kin-skjaldi.\eva

\bvb Filer, Chiler, Found and Needler, \\
Hefter, Wiler, Hanner, Swigher, \\
Fraw, Hornborer, Fray and Looner, \\
Earwong, Earer, Oakenshield.\evb\evg


\bvg\bva[14]\mssnote{\Regius~1r/30, \Hauksbok~20r/22, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
Mál es \alst{d}verga \hld\ ï \alst{D}valins liði &
\alst{l}jȯna kindum \hld\ til \alst{L}ofars tęlja, &
\edtext{þęir}{\Afootnote{\emph{þeim} \Hauksbok}} es \alst{s}óttu \hld\ frȧ \alst{s}alar stęini &
\alst{Au}r-vanga sjǫt \hld\ til \alst{Jǫ}ru-valla.%
\Ballnote{A standard genealogical introduction (cf. \Haleygjatal\ 1: \emph{meðan hans ę́tt \dots\ til goða tęljum} ‘while we tally his line \dots\ back to the gods’); the patrilineal line of dwarfs is to be counted back to their progenitor, Loffer.  This possibly disagrees with st. 10, where Moodsowner is said to be the foremost (and presumably the oldest) of the dwarfs, and Loffer is not mentioned, but such details were probably not very important.

Sts. 14–15 is paraphrased in \Gylfaginning\ 14: “But these came from Swornshigh (\emph{Svarinshaugr}) to the Earwongs on the Erwolds, and from them Loffer is come—these are their names: Sherper (\emph{Skirpir}), Werper (\emph{Virpir}), Showfind, Great-grandfather, Elf and Ing (\emph{Ingi}), Oakenshield, Fale (\emph{Falr}), Frost, Finn, Ginner.”}\eva

\bvb It is time to tally the dwarfs in Dwollen’s troop \\
back to Loffer for the races of men; \\
they who sought, from the stone of the hall, \\
the seat of the \inx[L]{Earwongs} unto the \inx[L]{Erwolds}.\evb\evg


\bvg\bva[15]\mssnote{\Regius~1r/32, \Hauksbok~20r/24, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
Þar vas \alst{D}raupnir \hld\ ok \alst{D}olg-þrasir, &
\alst{H}ár, \alst{H}aug-spori, \hld\ \alst{H}lé-vangr, Glói, &
\alst{Sk}irfir, Virfir, \hld\ \alst{Sk}áfiðr, Ái, &
\alst{A}lfr ok \alst{Y}ngvi, \hld\ \alst{Ęi}kin-skjaldi, &
\alst{F}jalarr ok \alst{F}rosti, \hld\ \alst{F}innr ok Ginnarr; &
Þat mun \edtext{\alst{ę́}}{\Afootnote{om. \Regius}} \alst{u}ppi, \hld\ meðan \alst{ǫ}ld lifir, &
\alst{l}ang-niðja-tal \hld\ \edtext{til}{\Afootnote{om. \Hauksbok}} \alst{L}ofars hafat.\eva

\bvb There was Dreepner and Dollowthrasher, \\
High, Highspurer, Leewong, Glower, \\
Sherver, Werver, Showfind, Great-grandfather, \\
Elf and Ing, Oakenshield, \\
Feller and Frost, Finn and Ginner.— \\
It will ever be remembered while the age lives,\footnoteB{Two archaic formulæ. The first literally ‘that will ever [be] up above’, cf. \HervararSaga\ TODO: “We two are cursed, brother, thy bane am I become! That will ever be remembered (\emph{þat mun ę́ uppi}, but both mss. \emph{þat mun enn uppi}), evil is the doom of the norns!” The second is found in a runic inscription, U 323 (980–1015): “Ever will lie—while the age lives (\textbf{meþ + altr + lifiʀ} \emph{með aldr lifir})—the hard-hammered bridge, broad, after a good man.” An especially close parallel is found in Þstf \emph{Stuttdr} (st. 5, Kari Ellen Gade ed. in \Skp\ II): \emph{Ęy mun uppi \hld\ Ęndils, meðan stęndr // sól-borgar salr, \hld\ svǫr-gǿðis fǫr.} ‘Always will be remembered—while the hall of the sun’s stronghold \ken{sky/heaven > earth} stands—the journey of the fattener of Andle’s bird \ken{raven/eagle > warrior}.’} \\
the tally of kinsmen lifted to Lofer.\evb\evg

\subsection{Stanzas from \emph{Hauksbók}}

\Hauksbok\ has a few substantial inserts and differences from \Regius.  Their style strongly suggests that they are later compositions.

\bvg\bva[H1]\mssnote{\Hauksbok~20v/12}%
Þȧ kná \edtrans{\alst{V}áli}{Wonnel}{\Afootnote{emend.; \emph{Vála} \Hauksbok}} \hld\ \alst{v}íg-bǫnd snúa &
\alst{h}ęldr vǫ́ru \alst{h}arð-gǫr \hld\ \alst{h}ǫpt ór þǫrmum.%
\Ballnote{34/1–2 are replaced by these lines in \Hauksbok.}\eva

\bvb Then did \inx[C]{Wonnel} the war-bonds twist: \\
the most sturdy fetters were made from intestines.\evb\evg


\bvg\bva[H2]\mssnote{\Hauksbok~20v/28}%
\alst{H}rę́ðask allir \hld\ ȧ \alst{h}ęl-vegum &
áðr \alst{S}urtar þann \hld\ \alst{s}efi of glęypir.%
\Ballnote{45/5–6 is followed by these lines in \Hauksbok, in effect forming another four-line stanza.  They seem to be referring to an ettin swallowing the World Tree.}\eva

\bvb All are frightened on the Hell-ways \\
before Surt’s kinsman does devour it.\evb\evg


\bvg\bva[H3]\mssnote{\Hauksbok~20v/39}%
Gïnn \alst{l}opt yfir \hld\ \alst{l}indi jarðar, &
gapa \alst{ý}gs kjaptar \hld\ \alst{o}rms ï hę́ðum; &
mun \alst{Ó}ðins son \hld\ \edtrans{\alst{ęi}tri}{venom}{\Afootnote{emend.; \emph{ormi} ‘Wyrm’ \Hauksbok.}\Bfootnote{Cf. \Gylfaginning\ 51: “Thunder bears the bane-word from the Middenyardswyrm and strides nine paces away from it. Then he falls dead to the earth for the venom (\emph{ęitri}) which the Wyrm blows on him.”}} mǿta &
\alst{v}args at \edtext{da\emph{uða}}{\Afootnote{\emph{‘da...’} \Hauksbok}} \hld\ \alst{V}íðars niðja.%
\Aallnote{The last part of the stanza is almost completely illegible in the ms.  I have relied on the transliteration of \textcite[13,44\psqq]{JonHelgason1971}.}
\Ballnote{This stanza appears between 52 and 53 in \Hauksbok.}\eva

\bvb Over the air yawns the Girdle of the Earth \ken*{= Middenyardswyrm}; \\
the jaws of the fierce Wyrm gape in the heights. \\
Weden’s son \ken*{= Thunder} will meet the venom \\
of the Warg, after the deaths of Wider’s kinsmen \ken*{= the Eese}.\evb\evg


\bvg\bva[H4]\mssnote{\Hauksbok~21r/14}%
Þȧ kømr hinn \alst{r}íki \hld\ at \alst{r}ęgin-dȯmi &
\alst{ǫ}flugr \alst{o}fan \hld\ sá’s \alst{ǫ}llu rę́ðr.%
\Ballnote{This half-stanza appears between 61 and 62; it is generally held to be a late Christian insert.}\eva

\bvb Then comes the mighty one to the great power, \\
strong from above, he who rules everything.\evb\evg

\sectionline
