\bookStart{The Lay of Wayland}[Vǫlundarkviða]

\begin{flushright}%
Dating \parencite{Sapp2022}: C10th (0.428)–early C11th (0.475)

Meter: \Fornyrdislag%
\end{flushright}%

% Introduction

The \textbf{Lay of Wayland} (\Volundarkvida) is a story of immense psychological complexity, one of the masterpieces of Norse poetry.

The poem begins with a prose introduction, which survives in both \Regius\ and \AM.

Wayland gets his revenge on the whole royal household. He murders Nithad’s two young sons (affectionately, his “bear-cubs”) and thus ends his male lineage. Likewise he defangs Nithad’s “cunning wife” (she is never called anything else) by reducing her once powerful counsels to cold words; and finally he rapes Beadhild, depriving her of her maidenhood and value in marriage. They are thus reduced to the same state of complete powerlessness as he himself experienced, something clearly seen in the repetition of the adjective \emph{viljalauss} ‘powerless’; in v. 12 it describes Wayland after he wakes in shackles, but in v. 31 Nithad uses it to refer to his own mental state after the deaths of his sons. This sense of hopelessness is also seen in Beadhild’s haunting concluding speech. “I knew by naught struggle against him; I could by naught struggle against him.”

From the other versions of the story it is known that Beadhild gave birth to a son, Woody (OE \emph{Wudga}, \ThidreksSaga\ \emph{Viðga}, in Danish ballads \emph{Vidrik Verlandsøn}). He went on to become a great hero, and in the later heroic ballads by far eclipses his father. His birth seems heavily foreshadowed by Wayland forcing Nithad to swear an oath in v. 33, but he is nowhere directly mentioned in the poem, probably for artistic reasons.

Apart from this lay there is one other telling of the full story, namely the Strand of Wayland the Smith in \ThidreksSaga. While written in Old Norse, it is clear from the proper names and content that it is based on German sources (probably heroic ballads). Thus the native form \emph{Vǫlundr} is replaced with \emph{Velent} [\emph{sic}], \emph{Níðuðr} with \emph{Niðungr}. Interestingly there is a note within it showing that the native form was still known, namely about “Velent, the excellent smith, whom Warrings (\emph{Væringjar}) call Wayland (\emph{Vǫlundr})”. Apparently Wayland was so famous that “all men seem to praise his workmanship so, that the maker of any smith’s work which is made better than other works, is called a Wayland (\emph{Vǫlundr}) with regards to workmanship.”

Far more stark than minor differences of language is that of tone. The psychological complexity and tension of the older redaction is almost entirely gone: Wayland is no longer a mysterious wild man, but a chivalrous knight who can escape from any peril through his ingenuity and craftmanship. He is not kidnapped out of Nithad’s greed, nor hamstrung out of the suspicion of his cruel wife, but rather a loyal servant of Nithad’s, banished from the kingdom after defending himself against the king’s corrupt steward, and hamstrung after being caught attempting to poison the king’s food in revenge.

Most frustratingly the personality of Beadhild is entirely expulged. She is the anonymous “king’s daughter”, an unnamed maiden (\emph{jungfrú}, a borrowing from Low German) who is peacefully seduced by Wayland and quickly falls in love with him. Likewise the person of Nithad’s cunning wife is completely gone, and the murder of his sons no longer ends his lineage, since he has another, older son who survives him and takes over the kingdom. Wayland still flies away laughing after telling Nithad what he has done, but only four years (his son with Beadhild is three years old) later reconciliates with Nithad’s son, retrieves Beadhild and their son and lives a long life as a famous craftsman.

With this it is clearly seen that the story by the time of the \ThidreksSaga\ had been heavily distorted, a tragic victim of medieval romantic sensibilities. It does not have any high literary value, but is of interest since it shows the wide reception and variation of the narrative.

Finally there are also traces of the story in the Anglo-Saxon tradition, where it is alluded to in both \Waldere\ and \Deor, the latter of which particularly emphasising the powerlessness felt by Wayland and Beadhild (thus being much closer in spirit to the present poem than to \ThidreksSaga). Parts of the narrative are depicted on the early C8th Frank’s casket, where it is as prominent as the depiction of the Adoration of the Magi—a true testament to the weight with which it was regarded within that culture.

To illustrate the narrative correspondences and differences of the various redactions, I present the following table:
\begin{longtabu} to \textwidth {|c c c c c c|}
	\hline
	Person & \Volundarkvida & \ThidreksSaga & \Deor & \Waldere & Frank’s casket \\ [0.5ex]
	\hline\hline
  Wayland & \emph{Vǫlundr} & \emph{Velent} & \emph{Welund} & \emph{Weland} & + \\
  Wayland’s brothers & Agle and Slayfinn & Agle & − & − & Agle? \\
  Father of the brothers & A Finnish king & The riser Wade (\emph{Vaði}) of Zealand & − & − & − \\
  Nithad & \emph{Níðuðr}, lord of the Nears in Sweden & \emph{Niðungr}, king of \inx[G]{Thede} in Jutland & \emph{Niðhad} & \emph{Niðhad} & − \\
  Nithad’s daughter (Beadhild) & \emph{Bǫðvildr} & Unnamed & \emph{Beadohild} & − & + \\
  Nithad’s sons & Two & Three & More than one & − & At least one \\
  Wayland and Beadhild’s son (Woody) & − & Viðga & − & Widia & − \\
  Wives of the brothers & Wayland and Allwit, Agle and Alerune, Slayfinn and Swanwhite & Agle and Alerune & − & − & − \\ [0.5ex]
  \hline
  — & Wayland and his brothers ski and hunt animals. They settle in the Wolfdales, and one day find their wives. When they suddenly leave, Slayfinn and Agle go out to search for them, while Wayland remains alone, smithing rings and longing for his wife. & Wade sends Wayland away to learn smithing from dwarfs, and he becomes exceptionally skilled. He enters into the service of Nithad and becomes his trusted friend. One day Nithad asks Wayland to retrieve his victory-stone (the owner of which will always have victory in battle), in exchange for which he will be given half of his kingdom and his daughter in marriage. After retrieving it, Wayland is ambushed by Nithad’s steward, who asks him for the stone. When he refuses, the steward attacks him, but Wayland easily slays the steward, whose men flee. Wayland is banished for this, but returns and attempts to poison the king’s food. & − & − & − \\
  — & Nithad learns that Wayland is alone, and arrives with a large number of warriors. He abducts him, and on the counsel of his wife has him hamstrung. & After the trick is discovered, Wayland is caught and hamstrung by the king. & − & − & − \\
  — & Wayland is placed on the island Seastead, and forced to make jewellery for him. & Wayland pretends to reconcile with the king, acknowledging his error, and saying that he will never flee, even if he is able to. In return for this he is given a smithy, and as much gold and silver as he asks for. & − & − & − \\ [1ex]
	\hline
\end{longtabu}

\sectionline

\section{Regarding Wayland (\emph{Frá Vǫlundi})}

\bpg\bpa Níðuðr hét konungr í Svíþjóð.
Hann átti tvá sonu ok eina dóttur. Hon hét Bǫðvildr.
Brę́ðr váru þrír, synir Finnakonungs.
Hét einn Slagfiðr, annarr Egill, þriði Vǫlundr.
Þeir skriðu ok veiddu dýr. Þeir kómu í Úlfdali ok gerðu sér þar hús.
Þar er vatn, er heitir Úlfsjár.
Snemma of morgin fundu þeir á vatnsstrǫndu konur þrjár, ok spunnu lín.
Þar váru hjá þeim álftarhamir þeira. Þat váru valkyrjur.
Þar váru tvę́r dę́tr Hlǫðvés konungs, Hlaðguðr svanhvít ok Hervǫr alvitr, in þriðja var Ǫlrún Kjársdóttir af Vallandi.
Þeir hǫfðu þę́r heim til skála með sér. Fekk Egill Ǫlrúnar, en Slagfiðr Svanhvítrar, en Vǫlundr Alvitrar.
Þau bjuggu sjau vetr. Þá flugu þę́r at vitja víga ok kómu eigi aftr.
Þá skreið Egill at leita Ǫlrúnar, en Slagfiðr leitaði Svanhvítrar, en Vǫlundr sat í Úlfdǫlum.
Hann var hagastr maðr, svá’t menn viti í fornum sǫgum.
Níðuðr konungr lét hann hǫndum taka, svá sem hér er um kveðit:\epa

\bpb Nithad was named a king in Sweden.
He owned two sons and one daughter; she was called Beadhild.
There were three brothers, the sons of a king of the Finns.
One was called Slayfinn, another Agle, the third Wayland.
They travelled on skis and hunted wild animals. They came into the Wolfdales and made for themselves houses there.
There is a water there, called Wolfsea.
Early in the morning they found on the lake-shore three women, and they were spinning linen.
By them were their swan-\inx[C]{hame}[hames]; they were Walkirries.
Two of them were the daughters of king Ladwigh: Ladguth Swanwhite and Harware Allwit, the third was Alerune, daughter of \inx[P]{Kear} of \inx[G]{Walland}\footnoteB{The Roman emperor; see Encyclopedia.}.
The brothers brought the maidens with them to their halls. Agle got Alerune, but Slayfinn Swanwhite, but Wayland Allwit.
They lived there for seven winters, then they left to attend battles, and did not return.
Then Agle left on skis to look for Alerune, but Slayfinn sought out Swanwhite; but Wayland stayed in the Wolfdales.
He was the most skilled craftsman, as men know, in the ancient saws.
King Nithad had him captured, about which this has been sung:\epb\epg

\sectionline

\bvg
\bva \alst{M}ęyjar flugu sunnan \hld\ \alst{M}yrk-við í gǫgnum &
\alst{a}l-vitr \alst{u}ngar, \hld\ \edtrans{\alst{ø}r-lǫg drýgja;}{fulfill [their] orlay}{\Bfootnote{That is, to fulfill their already laid-down destinies, as described in P1 and st. 3. I disagree with \textcite{MCR2005}[103], who translates this phrase as ‘engage in war’, seeing the latter word as a borrowing from OE (cf. Dutch \emph{oorlog} ‘war’). In fact, the expression \emph{drýgja ørlǫg} is also attested in OE, namely in l. 29 of a poem on the Christian Doomsday (TODO?), about a man going to Hell for his sins: \emph{ond þonne á tó ealdre \hld\ orleg dreógeð} ‘And then (the sinner) suffers his orlay for ever and ever’}} &
þę́r á \alst{s}ę́var-strǫnd \hld\ \alst{s}ęttusk at hvílask &
\alst{d}rósir suð-rǿnar, \hld\ \alst{d}ýrt lín spunnu.\eva

\bvb Maidens flew from the south through Mirkwood\footnoteB{Mirkwood is surely referenced for its association with the war-ravaged lands of the Gots and Huns; a natural environment for Walkirries.}—young allwits\footnoteB{Maybe look at what this means. TODO.}—to fulfill [their] \inx[C]{orlay}. They on the lake-shore set down to rest; the southern ladies span expensive linen.\evb
\evg


\bvg
\bva \alst{Ęi}n nam þęira \hld\ \alst{Ę}gil at vęrja &
\alst{f}ǫgr mę́r \alst{f}ira \hld\ \alst{f}aðmi ljósum; &
ǫnnur vas \alst{S}vanhvít, \hld\ \alst{s}van-fjaðrar dró, &
\edtext{[...]}{\Bfootnote{A line mentioning the name of Slayfinn has certainly gone missing here.}} &
en hin \alst{þ}riðja \hld\ \alst{þ}ęira systir &
varði \alst{h}vítan \hld\ \alst{h}als Vǫlundar.\eva

\bvb One of them began—the fair maiden of men—to embrace Agle in her light bosom. Another was Swanwhite—her swan-feathers she rustled; but the third of the sisters warded the white throat of Wayland.\evb
\evg


\bvg
\bva \alst{S}ǫ́tu \alst{s}íðan \hld\ \alst{s}jau vetr at þat, &
en hinn \alst{á}tta \hld\ \alst{a}llan þrǫ́ðu, &
en hinn \alst{n}íunda \hld\ \alst{n}auðr of skilði, &
\alst{m}ęyjar fýstusk \hld\ á \alst{m}yrkvan við, &
\alst{a}l-vitr \alst{u}ngar \hld\ \alst{ø}r-lǫg drýgja.\eva

\bvb Then they stayed for seven winters at that, but all the eighth they yearned, but the ninth did need divorce them: the maidens longed for the mirky wood: the young allwits, to fulfill orlay.\footnoteB{As Walkirries the \inx[C]{orlay} (already laid-down destiny) of the sisters is to preside over battles for Weden. Remembering this duty they become increasingly restless, until they one day decide to leave when their husbands are out hunting. For the significance of Mirkwood, see note to st. 1.}\evb
\evg


\bvg
\bva Kom þar af \alst{v}ęiði \hld\ \alst{v}eðr-ęygr skyti &
Vǫlundr \alst{l}íðandi \hld\ of \alst{l}angan veg, &
\alst{S}lagfiðr ok Ęgill, \hld\ \alst{s}ali fundu auða, &
gingu \alst{ú}t ok \alst{i}nn \hld\ ok \alst{u}mb sǫ́usk.\eva

\bvb Came there from the hunt the weather-eyed shooter: Wayland passing over a long way. Slayfinn and Agle found the halls deserted; they walked out and in, and looked about.\evb
\evg


\bvg
\bva \alst{Au}str skręið \alst{Ę}gill \hld\ at \alst{Ǫ}lrúnu, &
en \alst{s}uðr \alst{S}lagfiðr \hld\ at \alst{S}vanhvítu, &
en \alst{ęi}nn Vǫlundr \hld\ sat í \alst{U}lf-dǫlum.\eva

\bvb East skied Agle for Alerune, but south Slayfinn for Swanwhite; but alone Wayland stayed in the Wolfdales.\evb
\evg


\bvg
\bva Hann sló \alst{g}oll rautt \hld\ við \alst{g}im fastan, &
\alst{l}ukði hann alla \hld\ \alst{l}inn-baugum vęl; &
\alst{s}vá bęið hann \hld\ \alst{s}innar ljóssar &
\alst{k}vánar, ef hǫ́num \hld\ of \alst{k}oma gęrði.\eva

\bvb He struck red gold by gemstone fastened, enclosed he all the serpent-\inx[C]{bigh}[bighs]\footnoteB{Armlets, torcs resembling serpents, perhaps even literally shaped like them; cf. the Viking age armlet found in a hoard in Undrom, Ångermanland, northern Sweden. Museum ID 108822 HST. TODO: Maybe include photo?} well; thus awaited he his bright wife, if to him she might come.\evb
\evg


\bvg
\bva Þat spyrr \alst{N}íðuðr, \hld\ \alst{N}íara dróttinn, &
at \alst{ęi}nn Vǫlundr \hld\ sat í \alst{U}lf-dǫlum; &
\alst{n}ǫ́ttum fóru sęggir, \hld\ \alst{n}ęglðar vǫ́ru brynjur, &
\alst{sk}ildir bliku þęira \hld\ við hinn \alst{sk}arða mána.\eva

\bvb This learns Nithad, lord of the \inx[G]{Nears}, that alone Wayland stayed in the Wolfdales. By night travelled warriors—nailed were their byrnies\footnoteB{The soldiers had plated armour.}—their shields gleamed by the waning moon.\evb
\evg


\bvg
\bva Stigu ór \alst{s}ǫðlum \hld\ at \alst{s}alar gafli, &
\edtext{gingu \alst{i}nn þaðan \hld\ \alst{ę}nd-langan sal}{\lemma{gingu \dots\ sal ‘went \dots\ hall’}\Bfootnote{Formulaic. The fixed variant line \emph{hón/hann inn of gekk \hld\ ęnd-langan sal} ‘he/she inside did go the endlong hall’ occurs in three other places: sts. 16 and 30 of the present poem, and st. 3 of \Oddrunargratr. \emph{ęnd-langr salr} ‘endlong hall’ occurs in two additional places: st. 27 of \Thrymskvida\ and st. 3 of \Skirnismal. — \emph{ęnd-langr} ‘endlong’ may be rendered as ‘throughout, the entire (length of)’.}}, &
sǫ́u þęir á \alst{b}ast \hld\ \alst{b}auga dręgna, &
\alst{s}jau hundruð allra, \hld\ es sá \alst{s}ęggr átti.\eva

\bvb They stepped down from the saddles by the hall’s gables; went thence inside the endlong hall; saw they on a bast-rope bighs drawn up: seven hundred in all which that man \ken*{= Wayland} owned.\evb
\evg


\bvg
\bva Ok þęir \alst{a}f tóku \hld\ ok þęir \alst{á} létu &
fyr \alst{ęi}nn \alst{ú}tan, \hld\ es \alst{a}f létu; &
kom þar af \alst{v}ęiði \hld\ \alst{v}eðr-ęygr skyti &
Vǫlundr \alst{l}íðandi \hld\ of \alst{l}angan veg.\eva

\bvb And they slid [them] off, and they slid [them] on; but for one, which off they slid.\footnoteB{Nithad’s men take off all the seven hundred rings (presumably to count them) and then put them back on, but they keep just one. This bigh is probably the one mentioned in sts. 17 and 26, since Beadhild has it already when Wayland is brought back after being captured. \textcite{FinnurEdda}\ writes (\emph{My translation from the Danish.}): “The ring which Nithad kept must have had special properties, and distinguished itself before others. There is no doubt that the ring is a flight ring; whether this was clear to the poet is however questionable. This much is certain, that Wayland seems to be able to fly away only after he has got back the ring; that is, the one which Beadhild brings him.” —The reader may for himself judge the plausibility of this, but it seems that Wayland, being an exceptionally handy craftsman, may just as well have crafted wings for himself without need for magical rings. This agrees with the Low German verison and the Daedalus myth, for both of which see the introduction to the present poem.}—Came there from the hunt the weather-eyed shooter: Wayland passing over a long way.\evb
\evg


\bvg
\bva Gekk \alst{b}rúnni \hld\ \alst{b}eru hold stęikja, &
\alst{á}r brann hrísi \hld\ \alst{a}ll-þurru fura, &
\alst{v}iðr hinn \alst{v}ind-þurri, \hld\ fyr \alst{V}ǫlundi.\eva

\bvb Went he the brown she-bear’s hull to roast; in early morning burned the twigs of all-dry pine—the wind-dry wood—before Wayland.\evb
\evg


\bvg
\bva Sat á \alst{b}er-fjalli, \hld\ \alst{b}auga talði, &
\alst{a}lfa ljóði \hld\ \alst{ęi}ns saknaði; &
\alst{h}ugði at \alst{h}ęfði \hld\ \alst{H}lǫðvés dóttir, &
\alst{A}l-vitr \alst{u}nga, \hld\ vę́ri \alst{a}ptr komin.\eva

\bvb Sat he on the bear-pelt, bighs he counted—the prince of elves was missing one! Thought he that Ladwigh’s daughter \ken*{= Harware} might have it; that the young allwit might be come back.\evb
\evg


\bvg
\bva \alst{S}at \alst{s}vá lęngi, \hld\ at \alst{s}ofnaði, &
ok \alst{v}aknaði \hld\ \alst{v}ilja-lauss; &
vissi sér á \alst{h}ǫndum \hld\ \alst{h}ǫfgar nauðir, &
en á \alst{f}ótum \hld\ \alst{f}jǫtur of spęntan.\eva

\bvb Sat he so long that asleep he fell, and he awoke, powerless. He knew on his hands tortuous restraints, and on his feet were fetters tightened.\evb
\evg


\bvg
\bva „Hvęrir ’ru \alst{jǫ}frar \hld\ þęir’s \alst{á} lǫgðu &
\alst{b}ęsti-síma \hld\ ok \alst{b}undu mik?“\eva

\bvb {[Wayland quoth:]} “Which are the princes, those that laid on thick bast-ropes, and bound me?”\evb
\evg


\bvg
\bva Kallaði \alst{n}ú \alst{N}íðuðr, \hld\ \alst{N}íara dróttinn: &
„Hvar gazt \alst{V}ǫlundr, \hld\ \alst{v}ísi alfa, &
\alst{ó}ra \alst{au}ra, \hld\ í \alst{U}lf-dǫlum? &
\alst{G}oll vas þar ęigi \hld\ á \alst{G}rana lęiðu, &
\alst{f}jarri hugða’k várt land \hld\ \alst{f}jǫllum Rínar.“\eva

\bvb Now called Nithad, lord of the Nears: “Where gottest thou, Wayland, leader of elves, \emph{our} ounces in the Wolfdales? Gold was there not on \inx[P]{Grane}’s path; far I thought our land from the fells of the Rhine.\footnoteB{Grane was the horse of the legendary hero \inx[P]{Siward}, slayer of the dragon \inx[P]{Fathomer}. These events were thought to have taken place in Germany. The sense of the is thus sarcastic: “Where did you get that gold? A dragon’s hoard?”.}”\evb
\evg


\bvg
\bva „\alst{M}an’k at \alst{m}ęiri \hld\ \alst{m}ę́ti ǫ́ttum, &
es vér \alst{h}ęil \alst{h}jú \hld\ \alst{h}ęima vǫ́rum: &
\alst{H}laðguðr ok \alst{H}ęrvǫr \hld\ borin vas \alst{H}lǫðvé, &
\alst{k}unn vas Ǫlrún \hld\ \alst{K}íars dóttir.“\eva

\bvb {[Wayland quoth:]} “I remember that we owned greater wealth, when we a whole household were at home: Ladguth and Harware were born to Ladwigh; known was Alerune, Kear’s daughter.”\footnoteB{Wayland responds rather cryptically. It seems that by asserting the noble lineages of the three swan-wives he gives a legitimate reason for his wealth, although he seems to be aware, judging by the tone, that the greedy Nithad neither cares nor believes him.}\evb
\evg

\sectionline

\bvg
\bva Úti stóð \alst{k}unnig \hld\ \alst{k}vǫ́n Níðaðar, &
\edtext{hón \alst{i}nn of gekk \hld\ \alst{ę}nd-langan sal}{\lemma{hon inn \dots\ sal ‘she inside \dots\ hall’}\Bfootnote{Formulaic, also occuring in st. 30 of the present poem and in \Oddrunargratr\ 3.}}, &
\alst{st}óð á golfi, \hld\ \alst{st}ilti rǫddu: &
„es-a sá nú \alst{h}ýrr, \hld\ es ór \alst{h}olti fęrr.\eva

\bvb Outside stood the cunning wife of Nithad; she inside did go the endlong hall; stood on the floor, steered her voice: “That one \ken*{= Wayland} is not mild now, who comes out of the wood.\evb
\evg


\bvg
\bva \alst{T}ęnn hǫ́num \alst{t}ęygjask \hld\ es hǫ́num’s \alst{t}ét sverð &
ok hann \alst{B}ǫðvildar \hld\ \alst{b}aug of þękkir, &
\alst{ǫ́}mun eru \alst{au}gu \hld\ \alst{o}rmi hinum frána; &
\alst{s}níðið ér hann \hld\ \alst{s}ina magni, &
ok \alst{s}ętið hann \alst{s}íðan \hld\ í \alst{S}ę́varstǫð.“\eva

\bvb His teeth are bared when he is shown the sword, and he recognizes Beadhild’s bigh; reminiscent are the eyes to the gleaming serpent’s.—Snithe ye from him the might of his sinews, and set him thereafter on Seastead!”\evb
\evg


\bpg
\bpa Svá var gǫrt, at skornar váru sinar í knés-fótum ok settr í holm einn, er þar var fyrir landi, er hét Sę́varstaðr. Þar smíðaði hann konungi alls-kyns gǫr-simar; engi maðr þorði at fara til hans, nema konungr einn. Vǫlundr kvað:\epa

\bpb Thus was done, that the sinews in his houghs were cut, and he was placed on a lonely islet lying there before the land, which was called Seastead. There he smithed for the king all manner of jewels. No man dared journey to him, save for the king alone. Wayland quoth:\epb
\epg


\bvg
\bva „\alst{S}é’k Níðaði \hld\ \alst{s}verð á linda, &
þat’s ek \alst{h}vęsta \hld\ sęm \alst{h}agast kunna’k &
ok ek \alst{h}ęrða’k \hld\ sęm \alst{h}ǿgst þótti; &
sá ’s mér \alst{f}ránn mę́kir \hld\ ę́ \alst{f}jarri borinn; &
\alst{s}é’kk-a þann Vǫlundi \hld\ til \alst{s}miðju borinn.\eva

\bvb “I see a sword on Nithad’s belt, that one I sharpened as most handily I knew, and hardened as most pleasingly seemed. Now that gleaming blade is ever far from me carried; I see it not for Wayland to the smithy carried.\evb
\evg


\bvg
\bva Nú \alst{b}err \alst{B}ǫðvildr \hld\ \alst{b}rúðar minnar &
—\alst{b}íð’k-a þęss \alst{b}ót— \hld\ \alst{b}auga rauða.“\eva

\bvb Now Beadhild bears my bride’s—I await no bettering for that—red bighs.”\evb
\evg


\bvg
\bva \edtrans{\alst{S}at—né \alst{s}vaf á-valt—}{He sat—he slept not—}{\Bfootnote{Compare \Gudrunarhvot\ TODO: \emph{hófu mik—né drękkðu—} ‘lifted me—drowned [me] not—’.}} \hld\ ok \alst{s}ló hamri; &
vél gęrði \alst{h}ęldr \hld\ \alst{h}vatt Níðaðí; &
\alst{d}rifu ungir tvęir \hld\ á \alst{d}ýr séa &
\alst{s}ynir Níðaðar \hld\ í \alst{S}ę́varstǫð.\eva

\bvb He sat—he slept not—and struck the hammer; he very boldly planned wiles for Nithad.—Two young ones drifted to look at precious things: Nithad’s sons, onto Seastead.\evb
\evg


\bvg
\bva \alst{K}vǫ́mu til \alst{k}istu, \hld\ \alst{k}rǫfðu lukla, &
\alst{o}pin vas \alst{i}llúð, \hld\ es \alst{í} sǫ́u, &
fjǫlð vas þar \alst{m}ęina, \hld\ es \alst{m}ǫgum sýndisk &
at vę́ri \alst{g}oll rautt \hld\ ok \alst{g}ǫr-simar.\eva

\bvb Came they to the chest, demanded the keys; open was the evil when inside they looked. A great deal was there of harms, which to the lads seemed like were it red gold and jewels.\evb
\evg


\bvg
\bva „Komið \alst{ęi}nir tvęir, \hld\ komið \alst{a}nnars dags; &
ykkr lę́t’k þat \alst{g}oll \hld\ of \alst{g}efit verða; &
\alst{s}ęgið-a męyjum \hld\ né \alst{s}al-þjóðum, &
\alst{m}anni ęngum, \hld\ at \alst{m}ik fyndið.“\eva

\bvb {[Wayland quoth:]} “Come alone ye two, come another day; to you I will let that gold be given. Say not to maidens nor to the people of the hall—to no man—that ye met me!”\evb
\evg


\bvg
\bva \alst{S}nimma kallaði \hld\ \alst{s}ęggr á annan, &
\alst{b}róðir á \alst{b}róður: \hld\ „gǫngum \alst{b}aug séa!“ &
\alst{K}vǫ́mu til \alst{k}istu, \hld\ \alst{k}rǫfðu lukla, &
\alst{o}pin vas \alst{i}llúð \hld\ es \alst{í} litu.\eva

\bvb Early called one youth to another, brother to brother: “Let us go see the bighs!” Came they to the chest, demanded the keys; open was the evil when inside they looked.\evb
\evg


\bvg
\bva Snęið af \alst{h}ǫfuð \hld\ \alst{h}úna þęira &
ok und \alst{f}ęn \alst{f}jǫturs \hld\ \alst{f}ǿtr of lagði, &
ęn þę́r \alst{sk}álar, \hld\ es und \alst{sk}ǫrum vǫ́ru, &
\alst{s}vęip útan \alst{s}ilfri, \hld\ \alst{s}ęldi Níðaði.\eva

\bvb He sliced off the heads of those bear-cubs\footnoteB{An affectionate term for the young boys. TODO: Relate to Bearserks?} \ken{boys}, and under the fetter’s fen\footnoteB{Very unclear. TODO.} their feet did lay; but the bowls which were under their curls \ken{skulls}, he coated with silver and gave to Nithad.\evb
\evg


\bvg
\bva \alst{E}n ór \alst{au}gum \hld\ \alst{ja}rkna-stęina &
sęndi \alst{k}unnigri \hld\ \alst{k}vǫ́n Níðaðar; &
en ór \alst{t}ǫnnum \hld\ \alst{t}vęggja þęira &
\alst{s}ló brjóst-kringlur, \hld\ \alst{s}ęndi Bǫðvildi.\eva

\bvb But out of the eyes earkenstones he sent to the cunning wife of Nithad; but out of the teeth of the two he struck breast-brooches, sent to Beadhild.\evb
\evg

\sectionline

\bvg
\bva Þá nam \alst{B}ǫðvildr \hld\ \alst{b}augi at hrósa &
\edtext{[...]}{\Bfootnote{The meter requires a half-line here, likely containing a more specific description of the bigh.}}\ \hld\ es brotit hafði, &
„\alst{þ}ori’k-a’k sęgja, \hld\ nema \alst{þ}ér ęinum.“\eva

\bvb Then Beadhild began to praise the ring,\footnoteB{The verse is without doubt incomplete, but the story can be gleaned: Beadhild breaks the bigh she has been given by her parents (previously mentioned in vv. 10 (see note there) and 17), and is afraid that her parents may become upset. She thus goes to Wayland in secret, asking him to repair it.} [...] which she had broken, “I dare not tell it, save to thee alone.”\evb
\evg


\bvg
\bva „Ek \alst{b}ǿti svá \hld\ \alst{b}rest á golli, &
at \alst{f}ęðr þínum \hld\ \alst{f}ęgri þykkir, &
ok \alst{m}ǿðr þinni \hld\ \alst{m}iklu bętri, &
ok \alst{s}jalfri þér \hld\ at \alst{s}ama hófi.“\eva

\bvb {[Wayland quoth:]} “I mend such the crack on the gold, that to thy father it fairer seems, and to thy mother far better, and to thyself of the same rank.”\evb
\evg


\bvg
\bva \alst{B}ar hann hána \alst{b}jóri, \hld\ því’t hann \alst{b}ętr kunni, &
\alst{s}vá’t hón í \alst{s}essi \hld\ of \alst{s}ofnaði. &
„Nú \alst{h}ęf’k \alst{h}ęfnt \hld\ \alst{h}arma minna &
\alst{a}llra nema \alst{ę}inna \hld\ \alst{í}-við-gjǫrnum.“\eva

\bvb He overcame her with beer—for he knew better\footnoteB{i.e. was more cunning, experienced than her.}—so that she in the seat asleep did fall. “Now have I avenged my harms—all but one\footnoteB{Presumably the deprivation of his mobility due to the hamstringing, which he resolves in the following stanza.}—on the insidious ones.\footnoteB{King Nithad and his family.}”\evb
\evg


\bvg
\bva „\alst{V}ęl ek,“ kvað \alst{V}ǫlundr, \hld\ „\alst{v}erða’k á fitjum, &
þęim’s mik \alst{N}íðaðar \hld\ \alst{n}ǫ́mu rekkar.“ &
\alst{H}lę́jandi Vǫlundr \hld\ \alst{h}ófsk at lopti, &
\alst{g}rátandi Bǫðvildr \hld\ \alst{g}ekk ór ęyju. &
tregði \alst{f}ǫr \alst{f}riðils \hld\ ok \alst{f}ǫður vręiði.\eva

\bvb “Well I”, quoth Wayland, “fall on my paddles; those which Nithad’s men bereaved me of!\footnoteB{\emph{C-V}: \emph{fit} ‘the webbed foot of water-birds’, the reader may picture for himself. Wayland has crafted a mechanism to take flight, regaining his mobility which he lost when he was hamstrung.}” Laughing Wayland threw himself in the air; weeping Beadhild went from the island: she grieved the lover’s flight, and the father’s fury.\evb
\evg

\sectionline

\bvg
\bva Úti stóð \alst{k}unnig \hld\ \alst{k}vǫ́n Níðaðar, &
ok hón \alst{i}nn of gekk \hld\ \alst{ę}nd-langan sal, &
en hann á \alst{s}al-garð \hld\ \alst{s}ęttisk at hvílask, &
„Vakir þú \alst{N}íðuðr, \hld\ \alst{N}íara dróttinn?“\eva

\bvb Outside stood the cunning wife of Nithad, and she inside did go the endlong hall—but he, on the courtyard, set down to rest. “Art thou awake, Nithad, lord of the Nears?”\evb
\evg


\bvg
\bva „\alst{V}aki’k á-\alst{v}alt \hld\ \alst{v}ilja-lauss, &
\alst{s}ofna’k minst, \hld\ síz \alst{s}onu dauða, &
\alst{k}ęll mik í hǫfuð, \hld\ \alst{k}ǫld erumk rǫ́ð þín, &
\alst{v}ilnumk þęss nú, \hld\ at við \alst{V}ǫlund dǿma’k.“\eva

\bvb {[Nithad quoth:]} “I am always awake, powerless; I fall asleep the least, since the death of my sons. My head freezes; cold are thy counsels—I wish now but that: to speak with Wayland.”\evb
\evg

\sectionline

\bvg
\bva „Sęg mér þat \alst{V}ǫlundr, \hld\ \alst{v}ísi alfa, &
af \alst{h}ęilum \alst{h}vat varð \hld\ \alst{h}únum mínum?“\eva

\bvb {[Nithad quoth:]} “Say it to me, Wayland, leader of elves: what became of my healthy bear-cubs \ken{boys}?”\evb
\evg


\bvg
\bva „\alst{Ęi}ða skalt mér \alst{á}ðr \hld\ \alst{a}lla vinna, &
at \alst{sk}ips borði \hld\ ok at \alst{sk}jaldar rǫnd, &
at \alst{m}ars bǿgi \hld\ ok at \alst{m}ę́kis ęgg &
at þú \alst{k}vęlj-at \hld\ \alst{k}vǫ́n Vǫlundar, &
né \alst{b}rúði minni \hld\ at \alst{b}ana verðir, &
þótt kvǫ́n \alst{ęi}gim, \hld\ þá’s \alst{é}r kunnið, &
eða \alst{jó}ð \alst{ęi}gim \hld\ \alst{i}nnan hallar.\eva

\bvb {[Wayland quoth:]} “Before that shalt thou swear to me all oaths:—by the deck of the ship and the rim of the shield, by the bough of the steed and the edge of the sword—that thou wilt not torment the wife of Wayland, nor of my bride become the bane, though a wife we might own, which ye might know; or a babe might own within the hall.\footnoteB{Wayland has Nithad swear an oath that he will not harm Beadhild, nor their (yet unborn) child. For the form of the oaths cf. TODO.}\evb
\evg


\bvg
\bva \alst{G}akk til smiðju, \hld\ es \alst{g}ęrðir þú, &
þar fiðr þú \alst{b}ęlgi \hld\ \alst{b}lóði stokna, &
snęið’k af \alst{h}ǫfuð \hld\ \alst{h}úna þinna &
ok und \alst{f}ęn \alst{f}jǫturs \hld\ \alst{f}ǿtr of lagða’k.\eva

\bvb Go to the smithy, which thou madest; there wilt thou find bellows, sprinkled with blood. I sliced off the heads of thy bear-cubs \ken{boys}, and under the fetter’s fen their feet did I lay.\evb
\evg


\bvg
\bva En þę́r \alst{sk}álar, \hld\ es und \alst{sk}ǫrum vǫ́ru, &
\alst{s}vęip’k útan \alst{s}ilfri, \hld\ \alst{s}ęlda’k Níðaði, &
\alst{e}n ór \alst{au}gum \hld\ \alst{ja}rkna-stęina, &
sęnda’k \alst{k}unnigri \hld\ \alst{k}vǫ́n Níðaðar.\eva

\bvb But the bowls, which were under their curls, I coated with silver and gave to Nithad. But out of the eyes earkenstones I sent to the cunning wife of Nithad.\evb
\evg


\bvg
\bva En ór \alst{t}ǫnnum \hld\ \alst{t}vęggja þęira &
\alst{s}ló’k brjóst-kringlur, \hld\ \alst{s}ęnda’k Bǫðvildi; &
nú gęngr \alst{B}ǫðvildr \hld\ \alst{b}arni aukin, &
\edtrans{\alst{ęi}nga dóttir \hld\ \alst{y}kkur bęggja.}{the only daughter of you both}{\Bfootnote{Formulaic, near-identical to \HervararSaga\ st. 25/1–2: (\emph{Vaki, Angantýr, \hld\ vękr þik Hęrvǫr, // ęinga dóttir \hld\ ykkr Svǫ́fu.} ‘Wake, Ongentew: Harware awakes thee, the only daughter of thee and Sweve.’ Cf. also \Beowulf\ 375a, 2997b: \emph{ángan dohtor} ‘only daughter (accusative)’.)}}“\eva

\bvb But out of the teeth of the two, I struck breast-brooches, sent to Beadhild. Now walks Beadhild, swollen with child; the only daughter of you both.”\evb
\evg


\bvg
\bva „\alst{M}ę́ltir-a þú þat \alst{m}ál, \hld\ es mik \alst{m}ęir tregi, &
né þik \alst{v}ilja’k \alst{V}ǫlundr \hld\ \alst{v}err of níta; &
es-at svá maðr \alst{h}ǫ́r, \hld\ at þik af \alst{h}ęsti taki, &
\alst{n}é svá ǫflugr, \hld\ at þik \alst{n}eðan skjóti, &
þar’s þú \alst{sk}ollir \hld\ við \alst{sk}ý uppi.“\eva

\bvb {[Nithad quoth:]} “Thou couldst not have spoken that speech which might grieve me more; nor could I worse wish, Wayland, to deny thee. There is no man so high that he from horse might take thee, nor so mighty that he might shoot thee from below, there as thou jeerest against the cloud-cover on high!”\evb
\evg


\bvg
\bva \alst{H}lę́jandi Vǫlundr \hld\ \alst{h}ófsk at lopti, &
en \alst{ó}-kátr Níðuðr \hld\ þá \alst{ę}ptir sat.\eva

\bvb Laughing Wayland threw himself in the air, but gloomy Nithad thereafter stayed.\evb
\evg

\sectionline

\bvg
\bva „Upp rís \alst{Þ}akkráðr, \hld\ \alst{þ}rę́ll minn bazti, &
\alst{b}ið \alst{B}ǫðvildi, \hld\ \edtext{męy hina \alst{b}rá-hvítu, &
gangi \alst{f}agr-varið}{\lemma{męy hina brá-hvítu \dots\ fagr-varið ‘the brow-white maiden \dots\ fair-clothed’}\Bfootnote{With these expressions Nithad strongly stresses the purity of his daughter (\emph{mę́r} ‘maiden’ here simply meaning ‘virgin’). Perhaps he thinks that her innocence can be restored if she dresses in fair clothes, but it will not be so.}} \hld\ við \alst{f}ǫður rǿða.“\eva

\bvb {[Nithad quoth:]} “Rise up, Thankred, my best thrall! Ask Beadhild—the brow-white maiden—to go fair-clothed with her father to counsel.”\evb
\evg

\sectionline

\bvg
\bva „Es þat \alst{s}att Bǫðvildr, \hld\ es \alst{s}ǫgðu mér, &
\alst{s}ǫ́tuð it Vǫlundr \hld\ \alst{s}aman í holmi?“\eva

\bvb {[Nithad quoth:]} “Is it true, Beadhild, as they said to me: stayed thou and Wayland together on the islet?”\evb
\evg


\bvg
\bva „\alst{S}att ’s þat Níðuðr \hld\ es \alst{s}agði þér: &
\alst{s}ǫ́tum vit Vǫlundr \hld\ \alst{s}aman í holmi &
\alst{ęi}na \alst{ǫ}gur-stund, \hld\ \alst{ę́}va skyldi; &
ek \alst{v}ę́tr hǫ́num \hld\ \alst{v}inna kunna’k, &
ek \alst{v}ę́tr hǫ́num \hld\ \alst{v}inna mátta’k.“\eva

\bvb {[Beadhild quoth:]} “’Tis true, Nithad, as \emph{he} said\footnoteB{Beadhild, knowing that the only one who is aware of what happened is Wayland, makes the subtle change in the conjugation, from her father’s general plural (“what \emph{they} said”), to the specific singular (“what \emph{he} said”).} to thee: stayed I and Wayland together on the islet, for one heavy hour—it should never [have been]! I knew by naught struggle against him; I could by naught struggle against him.\footnoteB{She was both mentally (\emph{kunna} ‘to know, understand’) and physically (\emph{mega} ‘to have strength to do, avail’) incapable of struggling against him. — As \textcite{FinnurEdda} comments, an unsurpassed final verse.}”\evb
\evg
