\bookStart{Speeches of Webthrithner}[Vafþrúðnismǫ́l]

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: C10th (0.894)

\textbf{Meter:} \Ljodahattr%
\end{flushright}%

\section{Introduction}

A wisdom contest poem, known by the author of \Gylfaginning.

Far from being a loose collection of pieces of mythic information, the poem is tightly structured.  There is a logical plan to the questions asked, and they are clearly divided into groups.  It is probably not a coincidence that Weden asks exactly 18 questions, this being the same number as the spells in the Leed-tally section (sts. 147–165) of \Havamal.  Eighteen is of course a multiple of the sacred number nine.  It is of special note that the eighteenth item in both lists is an unknowable mystery.

Weden first asks his wife, Frie, for counsel, as he is curious about the ancient wisdom which the ettin Webthrithner might possess (1). Frie expresses worry, as she considers Webthrithner wiser than all other ettins (2), but Weden says that he has travelled far and wide, and wishes to know what Webthrithner’s hall is like (3). Frie wishes Weden good luck against the ettin (4) and he departs, to challenge Webthrithner’s \emph{orð-spęki} ‘word-wisdom’ (5). He arrives at hall of Webthrithner (6), who promptly declares that Weden will not come out of the hall unless he be wiser than him (7). Weden introduces himself as Gainred, saying that he has travelled far in need of Webthrithner’s hospitality (8). Webthrithner invites this Gainred to sit down (9), but he instead utters a gnomic stanza (10) not unlike those of the first section of \Havamal.

Webthrithner begins by asking four mythological questions, each answered by Gainred in turn. The questions concern which horses pull the day (11–12) and night (13–14), the river which divides the gods and ettins (15–16), and the plain where \inx[P]{Surt} and the gods will fight (17–18).

Webthrithner calls the god learned and invites him to sit.  He declares that the loser of the contest must give his head (19).  The roles are now reversed, and Gainred poses twelve numbered questions to the ettin.  He asks about the origins of earth and heaven (20–21), of sun and moon (22–23), of day, night, and the phases of the moon (24–25), and of winter and summer (26–27); then about the earliest being, namely the ettin \inx[P]{Earyelmer} (28–29), his origins (30–31) and how he reproduced asexually (32–33). Gainred continues by asking what Webthrithner himself first remembers (34–35), about the origin of the wind (36–37), the god \inx[P]{Nearth} (38–39), Walhall and the Oneharriers (40–41), and where Webthrithner has learned all this wisdom (42–43).

After this the structure and tone of the questions change, and each of the six final question-stanzas begins with the same first half as st. 3; they all concern the end times. Gainred asks which humans will survive after the \inx[L]{Fimblewinter} (44–45), how the sun can rise after Fenrer has destroyed it (46–47), about some obscure maidens (48–49; see there), which Eese will survive after the flame of Surt goes out (50–51) and how Weden will die (52–53). Finally, he asks what Weden spoke in the ear of Balder before he was burned on the pyre (54). Webthrithner at last understands the identity of his challenger, since only Weden himself could know the answer to that question. He laconically accepts his imminent death and the futility of his wisdom; the poem ends with his admission that Weden will always be the wisest (55).

\sectionline

\section{The Speeches of Webthrithner}

\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~7v/9}%
„Ráð mér nú \alst{F}rigg \hld\ alls mik \alst{f}ara tíðir &
\ind at \alst{v}itja \alst{V}af-þrúðnis; &
\alst{f}or-vitni mikla \hld\ kveð’k mér á \alst{f}ornum stǫfum &
\ind við þann hinn \alst{a}l-svinna \alst{jǫ}tun.“\eva

\bvb\speakernoteb{[\inx[P]{Weden} quoth:]}
“Counsel me now, \inx[P]{Frie}, as I desire to journey \\
\ind to visit \inx[P]{Webthrithner}; \\
Very curious am I of ancient staves \\
\ind from that all-wise \inx[G]{Ettins}[ettin].\footnoteB{i.e. ‘I am very curious to learn his ancient wisdom.’ Cf. st. 55.}”\evb\evg


\bvg\bva\speakernote{[Frigg kvað:]}\mssnote{\Regius~7v/12}„\alst{H}ęima lętja \hld\ mynda’k \alst{H}ęrja-fǫðr &
\ind í \alst{g}ǫrðum \alst{g}oða; &
því-at \alst{ę}ngi \alst{jǫ}tun \hld\ hugða’k \alst{ja}fn-ramman &
\ind sęm \alst{V}af-þrúðni \alst{v}esa.“\eva

\bvb\speakernoteb{[Frie quoth:]}
“At home would I keep the Father of Hosts \ken*{= Weden}, \\
\ind in the yards of the Gods— \\
for no ettin have I judged to be \\
\ind as strong as Webthrithner.”\evb\evg


\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~7v/13}„\alst{F}jǫlð ek \alst{f}ór, \hld\ \alst{f}jǫlð \alst{f}ręistaða’k, &
\ind fjǫlð ek \alst{r}ęynda \alst{r}ęgin; &
hitt \alst{v}il’k \alst{v}ita, \hld\ hvé \alst{V}af-þrúðnis &
\ind \alst{s}ala-kynni \alst{s}éi.“\eva

\bvb\speakernoteb{[Weden quoth:]}
“Much I journeyed, much I tried, \\
\ind much I tested the \inx[G]{Reins}. \\
This I wish to know: how Webthrithner’s \\
\ind halls may be.”\evb\evg


\bvg\bva\speakernote{[Frigg kvað:]}\mssnote{\Regius~7v/15}„\alst{H}ęill þú farir, \hld\ \alst{h}ęill þú aptr komir, &
\ind hęill á \alst{s}innum \alst{s}éir; &
\alst{ǿ}ði þér dugi \hld\ hvar’s skalt, \alst{A}lda-fǫðr, &
\ind \alst{o}rðum mę́la \alst{jǫ}tun.“\eva

\bvb\speakernoteb{[Frie quoth:]}
“Whole journey thou, whole come thou back, \\
\ind whole be thou on thy paths! \\
Thy wisdom avail thee where thou, Father of Men, \\
\ind with words shalt address the ettin!”\evb\evg


\bvg\bva\mssnote{\Regius~7v/17}%
\alst{F}ór þá Óðinn \hld\ at \alst{f}ręista orð-spęki &
\ind þess hins \alst{a}l-svinna \alst{jǫ}tuns; &
at \alst{h}ǫllu kom, \hld\ \edtext{es}{\Afootnote{ok \Regius}} átti \edtext{\alst{H}ymis}{\lemma{Hymis}\Afootnote{metr. emend. following \textcite{FinnurEdda}; \emph{Íms} \Regius}} faðir; &
\ind \alst{i}nn gekk \alst{Y}ggr þegar.\eva

\bvb Then journeyed Weden to test the word-wisdom \\
\ind of that all-wise ettin. \\
He came to the hall which \inx[P]{Hymer}’s father \ken*{= Webthrithner} owned; \\
\ind \inx[P]{Ug} \name{= Weden} went soon inside.\evb\evg


\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~7v/18}%
„\alst{H}ęill þú nú, Vaf-þrúðnir, \hld\ nú em’k í \alst{h}ǫll kominn &
\ind á þik \alst{s}jalfan \alst{s}éa; &
hitt vil’k \alst{f}yrst vita, \hld\ ef \alst{f}róðr séir &
\ind eða \alst{a}l-sviðr, \alst{jǫ}tunn.“\eva

\bvb\speakernoteb{[Weden quoth:]}%
“Hail thee now, Webthrithner! now I’ve come into the hall, \\
\ind to see thy very self! \\
This I wish first to know, if learned thou be, \\
\ind or all-wise, O ettin.”\evb\evg


\bvg\bva\speakernote{[Vafþrúðnir kvað:]}\mssnote{\Regius~7v/20}%
„Hvat ’s þat \alst{m}anna, \hld\ es í \alst{m}ínum sal &
\ind verpumk \alst{o}rði \alst{á}? &
\alst{ú}t þú né kømr \hld\ \alst{ó}rum hǫllum frá, &
\ind nema þú inn \alst{s}notrari \alst{s}éir.“\eva

\bvb\speakernoteb{[Webthrithner quoth:]}%
“What sort of man is that, who in \emph{my} hall \\
\ind throws a word at me? \\
Out thou comest not from \emph{our} halls, \\
\ind unless thou be the smarter man.”\evb\evg


\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~7v/22}%
„\edtext{\alst{G}agnráðr}{\Afootnote{Gang-ráðr ‘Gangred; Journey-adviser’ \GylfMS\ (paraphrased).}} hęiti’k, \hld\ nú em’k af \alst{g}ǫngu kominn, &
\ind \alst{þ}yrstr til \alst{þ}inna sala; &
\alst{l}aðar þurfi \hld\ hęf’k \alst{l}ęngi farit &
\ind ok þinna \alst{a}nd-fanga, \alst{jǫ}tunn.“\eva

\bvb\speakernoteb{[Weden quoth:]}%
“\inx[P]{Gainred} I am called; now am I come from walking, \\
\ind thirsty, to thy halls. \\
In need of a welcome have I journeyed for long, \\
\ind and of thy reception, ettin!”\evb\evg


\bvg\bva\speakernote{[Vafþrúðnir kvað:]}\mssnote{\Regius~7v/24}%
„Hví þú þá, \alst{G}agnráðr, \hld\ mę́lisk af \alst{g}olfi fyrir? &
\ind far þú í \alst{s}ess í \alst{s}al; &
þá skal \alst{f}ręista, \hld\ hvárr \alst{f}lęira viti, &
\ind \alst{g}ęstr eða hinn \alst{g}amli þulr.“\eva

\bvb\speakernoteb{[Webthrithner quoth:]}%
“Why then, Gainred, speakest thou from the floor before me? \\
\ind Take a seat in the hall! \\
Then it shall be tried, which of the two might know more: \\
\ind the guest, or the old \inx[C]{thyle}.”\evb\evg


\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~7v/26}%
„\alst{Ó}-auðigr maðr, \hld\ es til \alst{au}ðigs kømr, &
\ind \edtrans{mę́li \alst{þ}arft eða \alst{þ}ęgi}{ought to speak the needful or shut up}{\Bfootnote{Formulaic, this line occurs identically in \Havamal\ 19.}}; &
\alst{o}fr-mę́lgi mikil \hld\ hygg’k at \alst{i}lla geti &
\ind hvęim’s við \edtrans{\alst{k}ald-rifjaðan}{cold-ribbed}{\Bfootnote{i.e. ‘cold-hearted, cunning’.}} \alst{k}ømr.“\eva

\bvb\speakernoteb{[Weden quoth:]}%
“An unwealthy man who to a wealthy comes \\
\ind ought to speak the needful or shut up. \\
Great over-speaking I think brings ill \\
\ind for him who comes by a cold-ribbed man.”\evb\evg


\bvg\bva\speakernote{[Vafþrúðnir kvað:]}\mssnote{\Regius~7v/28}%
„Sęg mér, \alst{G}agnráðr, \hld\ alls á \alst{g}olfi vill &
\ind \edtrans{þíns of \alst{f}ręista \alst{f}rama}{test thy furtherance}{\Bfootnote{i.e. “try your luck, see how far you get”.  Formulaic; cf. \Havamal\ 2.}}, &
hvé \alst{h}ęstr \alst{h}ęitir, \hld\ sá’s \alst{h}vęrjan dręgr &
\ind \alst{d}ag of \alst{d}rótt-mǫgu?“\eva

\bvb\speakernoteb{[Webthrithner quoth:]}%
“Tell me, Gainred, since on the floor thou wilt \\
\ind test thy furtherance, \\
what the horse is called which pulls every \\
\ind day over the lads of the retinue \ken{men}?”\evb\evg


\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~7v/30}%
„\alst{Sk}in-faxi hęitir, \hld\ es hinn \alst{sk}íra dręgr &
\ind \alst{d}ag of \alst{d}rótt-mǫgu; &
\alst{h}ęsta batstr \hld\ þykkir með \alst{H}ręið-gotum; &
\ind ęy lýsir \alst{m}ǫn af \alst{m}ari.“\eva

\bvb\speakernoteb{[Weden quoth:]}%
“\inx[P]{Shinefax} is he called who pulls the bright \\
\ind day over the lads of the retinue. \\
The best of horses he seems among the \inx[G]{Reth-Gots}; \\
\ind ever shines that stallion’s mane.”\evb\evg


\bvg\bva\speakernote{[Vafþrúðnir kvað:]}\mssnote{\Regius~7v/32}%
„Sęg þat, \alst{G}agn-ráðr, \hld\ alls á \alst{g}olfi vill &
\ind þíns of \alst{f}ręista \alst{f}rama, &
hvé \alst{jó}r hęitir, \hld\ sá’s \alst{au}stan dręgr &
\ind \alst{n}ǫ́tt of \alst{n}ýt ręgin?“\eva

\bvb\speakernoteb{[Webthrithner quoth:]}%
“Tell this, Gainred, since on the floor thou wilt \\
\ind test thy furtherance, \\
what the steed is called which from east does pull \\
\ind night over the useful \inx[G]{Reins}?”\evb\evg


\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~7v/33}%
„\alst{H}rím-faxi \alst{h}ęitir, \hld\ es \alst{h}vęrja dręgr &
\ind \alst{n}ǫ́tt of \alst{n}ýt ręgin; &
\alst{m}él-dropa fęllir \hld\ \alst{m}orgin hvęrjan; &
\ind \edtrans{þaðan kømr \alst{d}ǫgg of \alst{d}ala}{from thence comes the dew of the dales}{\Bfootnote{For another explanation of the origin of dew, see \Voluspa\ 18.}}.“\eva

\bvb\speakernoteb{[Weden quoth:]}%
“\inx[P]{Rimefax}\ is he called who pulls every \\
\ind night over the useful Reins. \\
Drool from his bit he lets fall each morning; \\
\ind from thence comes the dew of the dales.”\evb\evg


\bvg\bva\speakernote{[Vafþrúðnir kvað:]}\mssnote{\Regius~8r/1}%
„Sęg þat, \alst{G}agnráðr, \hld\ alls á \alst{g}olfi vill &
\ind þíns of \alst{f}ręista \alst{f}rama, &
hvé \alst{ǫ́} hęitir, \hld\ sú’s dęilir með \alst{jǫ}tna sonum &
\ind \alst{g}rund, ok með \alst{g}oðum.“\eva

\bvb\speakernoteb{[Webthrithner quoth:]}%
“Tell this, Gainred, since on the floor thou wilt \\
\ind test thy furtherance, \\
what the river is called which divides the land \\
\ind between the sons of ettins and the gods.”\evb\evg


\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~8r/2}%
„\alst{Í}fing hęitir \alst{ǫ́}, \hld\ es dęilir með \alst{jǫ}tna sonum &
\ind \alst{g}rund, ok með \alst{g}oðum; &
\alst{o}pin rinna \hld\ hón skal umb \alst{a}ldr-daga; &
\ind verðr-at \alst{í}ss á \alst{ǫ́}u.“\eva

\bvb\speakernoteb{[Weden quoth:]}%
“\inx[L]{Iving} is the river called which divides the land \\
\ind between the sons of ettins and the gods. \\
Open shall it flow through its days of life; \\
\ind there forms no ice on that river.”\evb\evg


\bvg\bva\speakernote{[Vafþrúðnir kvað:]}\mssnote{\Regius~8r/3}%
„Sęg þat, \alst{G}agnráðr, \hld\ alls á \alst{g}olfi vill &
\ind þíns of \alst{f}ręista \alst{f}rama, &
hvé \alst{v}ǫllr hęitir, \hld\ es finnask \alst{v}igi at &
\ind \alst{S}urtr ok hin \alst{s}vǫ́su goð?“\eva

\bvb\speakernoteb{[Webthrithner quoth:]}%
“Tell this, Gainred, since on the floor thou wilt \\
\ind test thy furtherance, \\
what the plain is called where they find each other at war, \\
\ind \inx[P]{Surt} and the excellent Gods.”\evb\evg


\bvg\bva\speakernote{Óðinn:}\mssnote{\Regius~8r/4}%
„\alst{V}ígríðr hęitir \alst{v}ǫllr, \hld\ es finnask \alst{v}ígi at &
\ind \alst{S}urtr ok hin \alst{s}vǫ́su goð; &
\alst{h}undrað rasta \hld\ hann ’s á \alst{h}vęrjan veg; &
\ind sá ’s þęim \alst{v}ǫllr \alst{v}itaðr.“\eva

\bvb\speakernoteb{Weden:}%
“\inx[L]{Wighride} is the plain called where they find each other at war, \\
\ind Surt and the excellent gods. \\
A hundred \inx[C]{rest}[rests] it stretches in every way; \\
\ind for them that plain is marked out.”\evb\evg


\bvg\bva\speakernote{Vafþrúðnir:}\mssnote{\Regius~8r/6}%
„\alst{F}róðr est nú gęstr, \hld\ \alst{f}ar á bękk jǫtuns, &
\ind ok mę́lumk í \alst{s}essi \alst{s}aman; &
\alst{h}ǫfði vęðja \hld\ vit skulum \alst{h}ǫllu í &
\ind \alst{g}ęstr, of \alst{g}oð-spęki.“\eva

\bvb\speakernoteb{Webthrithner:}%
“Learned art thou now, guest; take the ettin’s bench \\
\ind and let us speak on the seat together! \\
Wager a head shall we two in the hall, \\
\ind O guest, over god-wisdom!”\evb\evg

\sectionline

{\small \Regius\ begins here has the header \emph{capitulum} ‘(new) chapter’ and begins st. 20 with a large initial.}

\sectionline

\bvg\bva\speakernote{Óðinn:}\mssnote{\Regius~8r/9, \AM~3r/1}%
„Sęg þat hit \alst{ęi}na, \hld\ ef þitt \edtext{\alst{ǿ}ði}{\lemma{ǿði}\Bfootnote{The first word on fol. 3r. of \AM; from this point we have the poem in both manuscripts.}} dugir &
\ind ok þú \alst{V}af-þrúðnir \alst{v}itir, &
hvaðan \alst{jǫ}rð of kom, \hld\ eða \alst{u}pp-himinn &
\ind \alst{f}yrst, hinn \alst{f}róði jǫtunn.“\eva

\bvb\speakernoteb{Weden:}%
“Tell this one, if thy wisdom avails, \\
\ind and thou, Webthrithner, mightst know, \\
from whence Earth did come, or \inx[L]{Up-heaven}, \\
\ind first, O learned ettin.”\evb\evg


\bvg\bva\speakernote{Vafþrúðnir:}\mssnote{\Regius~8r/10, \AM~3r/2}%
„Ór \alst{Y}mis holdi \hld\ vas \alst{jǫ}rð of skǫpuð, &
\ind en ór \alst{b}ęinum \alst{b}jǫrg, &
\alst{h}iminn ór \alst{h}ausi \hld\ hins \alst{h}rím-kalda jǫtuns, &
\ind en ór \edtrans{\alst{s}vęita}{blood}{\Bfootnote{Cf. \Sonatorrek\ 3/3 \emph{jǫtuns hals \hld\ undir þjóta} ‘the neck-wounds of the ettin \ken{seas} roar’.  In poetry \emph{svęiti} ‘sweat’ almost always means ‘blood’. This is shared with OE \emph{swât}, as seen e.g. in \Beowulf\ 1286a: \emph{sweord \emph{swâte} fâh} ‘sword stained with “sweat”’, 2689b–2690: \emph{hé ge-blódegod wearð // sâwul-dríore; \hld\ \emph{swât} ýðum wéoll.} ‘he was bloodied in soul-gore; the “sweat” gushed in waves’.}} \alst{s}ę́r.“\eva

\bvb\speakernoteb{Webthrithner:}%
“From \inx[P]{Yimer}’s flesh was the earth shaped, \\
\ind and from his bones the mountains; \\
heaven from the skull of that rime-cold ettin, \\
\ind and from his blood the sea.\footnoteB{The present st. very closely resembles \Grimnismal\ 41; lines 1 and 4 here are identical to lines 1–2 there, and lines 2 and 3a here are very similar to line 3a and 4 there. Although the stanzas are clearly related, they are still distinct enough that the one cannot be a direct scribal copy of the other. I think that the relation is most likely to be oral, and that the two stanzas were both composed in the same, most likely West Norwegian, community of poets, or perhaps even by the same individual.}”\evb\evg


\bvg\bva\speakernote{Óðinn:}\mssnote{\Regius~8r/12, \AM~3r/3}%
„Sęg þat \alst{a}nnat, \hld\ ef þitt \alst{ǿ}ði dugir &
\ind ok þú \alst{V}af-þrúðnir \alst{v}itir, &
hvaðan \alst{M}áni of kom, \hld\ svá’t fęrr \alst{m}ęnn yfir, &
\ind eða \alst{S}ól hit sama?“\eva

\bvb\speakernoteb{Weden:}%
“Tell this other, if thy wisdom avails, \\
\ind and thou, Webthrithner, mightst know, \\
from whence Moon did come, so that he journeys over men, \\
\ind or Sun likewise.”\evb\evg


\bvg\bva\speakernote{Vafþrúðnir:}\mssnote{\Regius~8r/13, \AM~3r/4}%
„\alst{M}undil-fari hęitir, \hld\ hann ’s \alst{M}ána faðir &
\ind ok svá \alst{S}ólar hit \alst{s}ama; &
\alst{h}imin \alst{h}verfa \hld\ þau skulu \alst{h}vęrjan dag &
\ind \edtrans{\alst{ǫ}ldum at \alst{á}r-tali}{for mankind’s tally of years}{\Bfootnote{Cf. \Voluspa\ 6, where the Reins gave names to night, the moon-phases, morning, midday, afternoon, and evening \emph{ǫ́rum at tęlja} ‘the years for to tally’.}}.“\eva

\bvb\speakernoteb{Webthrithner:}%
“\inx[P]{Mundlefare} he is called—he is the father of Moon, \\
\ind and so of Sun likewise. \\
Turn round heaven shall they every day, \\
\ind for mankind’s tally of years.”\evb\evg


\bvg\bva\speakernote{Óðinn:}\mssnote{\Regius~8r/15, \AM~3r/6}„\alst{S}ęg þat þriðja, \hld\ alls þik \alst{s}vinnan kveða &
\ind ok þú \alst{V}af-þrúðnir \alst{v}itir, &
hvaðan \alst{D}agr of kom, \hld\ sá’s fęrr \alst{d}rótt yfir, &
\ind eða \alst{N}ǫ́tt með \alst{n}iðum?“\eva

\bvb\speakernoteb{Weden:}%
“Tell this third, since they call thee wise, \\
\ind and thou, Webthrithner, mightst know, \\
from whence Day did come, who journeys over the folk, \\
\ind or Night with the moon-phases.”\evb\evg


\bvg\bva\speakernote{Vafþrúðnir:}\mssnote{\Regius~8r/17, \AM~3r/8}%
„\alst{D}ęllingr hęitir, \hld\ hann ’s \alst{D}ags faðir, &
\ind en \alst{N}ǫ́tt vas \alst{N}ǫrvi borin; &
\edtrans{\alst{n}ý ok \alst{n}ið}{The waxing and waning}{\Bfootnote{i.e. “the moon-phases”.  Cf. \Voluspa\ 6.}} \hld\ skópu \alst{n}ýt ręgin &
\ind \alst{ǫ}ldum at \alst{á}r-tali.“\eva

\bvb\speakernoteb{Webthrithner:}%
“\inx[P]{Delling} he is called—he is the father of \inx[P]{Day}, \\
\ind but \inx[P]{Night} was born to \inx[P]{Narrow}. \\
The waxing and waning did the useful Reins create \\
\ind for mankind’s tally of years.”\evb\evg


\bvg\bva\speakernote{Óðinn kvað:}\mssnote{\Regius~8r/18, \AM~3r/9}%
„Sęg þat \alst{f}jórða, \hld\ alls þik \alst{f}róðan kveða, &
\ind ok þú \alst{V}af-þrúðnir \alst{v}itir, &
hvaðan \alst{v}etr of kom \hld\ eða \alst{v}armt sumar &
\ind \alst{f}yrst með \alst{f}róð ręgin?“\eva

\bvb\speakernoteb{Weden quoth:}%
“Tell this fourth, since they call thee learned, \\
\ind and thou, Webthrithner, mightst know, \\
from whence winter did come, or warm summer, \\
\ind first, amidst the learned Reins.”\evb\evg


\bvg\bva\speakernote{Vafþrúðnir:}\mssnote{\Regius~8r/20, \AM~3r/10}%
„\alst{V}ind-svalr hęitir, \hld\ hann’s \alst{V}etrar faðir, &
\ind en \alst{S}vǫ́suðr \alst{S}umars.“ &
\edtext{[...]}{\Bfootnote{The second half of the st. seems to be missing; its contents are completely unknown.  No gap is indicated in the mss.}}\eva

\bvb\speakernoteb{Webthrithner:}%
“\inx[P]{Windswoll}\ is he called; he is \inx[P]{Winter}’s father; \\
\ind but \inx[P]{Sosuth}\ [is] \inx[P]{Summer}’s.”\evb\evg


\bvg\bva\speakernote{Óðinn kvað:}\mssnote{\Regius~8r/21, \AM~3r/11}%
„Sęg þat \alst{f}imta, \hld\ alls þik \alst{f}róðan kveða, &
\ind ok þú \alst{V}af-þrúðnir \alst{v}itir, &
hvęrr \alst{á}sa \alst{ę}ldstr \hld\ eða \alst{Y}mis niðja &
\ind \alst{y}rði í \alst{á}r-daga?“\eva

\bvb\speakernoteb{Weden quoth:}%
“Tell this fifth, since they call thee learned, \\
\ind and thou, Webthrithner, mightst know: \\
Who oldest of the \inx[G]{Eese}, or of Yimer’s kinsmen \ken{ettins}, \\
\ind arose in days of yore?\footnoteB{i.e. “which was the oldest, first being?” Cf. the question on the C9th Malt Stone (DR NOR1988;5): \textbf{huaʀisi : alistiąsa}, perhaps \emph{Hvaʀ es inn ęlisti ȧsa?} ‘Who is the eldest of the Eese?’}”\evb\evg


\bvg\bva\speakernote{Vafþrúðnir:}\mssnote{\Regius~8r/22, \AM~3r/12}%
„\alst{Ø}r-ófi vetra \hld\ áðr vę́ri \alst{jǫ}rð of skǫpuð, &
\ind þá vas \alst{B}er-gęlmir \alst{b}orinn, &
\alst{Þ}rúð-gęlmir \hld\ vas \alst{þ}ess faðir, &
\ind en \alst{Au}r-gęlmir \alst{a}fi.“\eva

\bvb\speakernoteb{Webthrithner:}%
“Uncountable winters before the Earth was created, \\
\ind then was \inx[P]{Bareyelmer} born. \\
\inx[P]{Thrithyelmer} was that one’s father, \\
\ind and \inx[P]{Earyelmer} the grandfather.”\evb\evg


\bvg\bva\speakernote{Óðinn kvað:}\mssnote{\Regius~8r/23, \AM~3r/14}%
„\alst{S}ęg þat \alst{s}étta, \hld\ alls þik \alst{s}vinnan kveða, &
\ind ok þú \alst{V}af-þrúðnir \alst{v}itir, &
hvaðan \alst{Au}r-gęlmir kom \hld\ með \alst{jǫ}tna sonum &
\ind \alst{f}yrst, hinn \alst{f}róði jǫtunn.“\eva

\bvb\speakernoteb{Weden quoth:}%
“Tell this sixth, since they call thee wise, \\
\ind and thou, Webthrithner, mightst know: \\
From whence Earyelmer came amidst the sons of ettins, \\
\ind first, O learned ettin?”\evb\evg


\bvg\bva\speakernote{Vafþrúðnir:}\mssnote{\Regius~8r/25, \AM~3r/15, \GylfMS}%
„\edtext{\alst{Ó}r \alst{É}li-vǫ́gum \hld\ stukku \alst{ęi}tr-dropar, &
\ind svá \alst{ó}x unds ór varð \alst{jǫ}tunn; &
\edtext{þar \alst{ó}rar \alst{ę́}ttir \hld\ kómu \alst{a}llar saman; &
\ind því’s \edtrans{þat}{it}{\Bfootnote{i.e. the ettin race.}} \alst{ę́} \alst{a}lt til \alst{a}talt.“}{\lemma{órar \dots\ atalt ‘Our \dots\ fierce’}\Bfootnote{so \GylfMS; om. \Regius\AM.}}}{\lemma{ALL}\Bfootnote{Over æons the splashing venom-drops combined until they formed a sentient being: this was Earyelmer, whom \Gylfaginning\ identifies with \inx[P]{Yimer}.  This stanza is cited in support of the lengthy and embellished creation narrative found in \Gylfaginning, but it is not certain that this is what our poet had in mind.

The Ilewaves are probably a reflex of the chaotic primeval Waters found in many West Eurasian mythologies, including Genesis 1:1–3 and \Rigveda\ 10.129.  Of these two foundational religious sources the latter is closer to the present stanza, and probably holds the more archaic conception.  Where we find in the Jewish narrative a proper \emph{creation}; at the very beginning of time God’s spirit is on the Waters and He makes the light shine over them, we find in these two Indo-European texts a \emph{spontaneous emergence} of a single primeval entity long before the Gods are born—here from the violent splashing of venom, in \Rigveda\ 10.129.3 from “the power of heat” (\emph{tápasaḥ mahinā́}).  This entity in turn asexually begets sexual beings—here through rubbing his limbs together, in \Rigveda\ 10.129.4 simply giving rise to “desire” (\emph{kā́ma}) which serves as the “primal seed of thought” (\emph{mánasaḥ rétaḥ prathamám})—and it is from these that the world is populated.}}\eva

\bvb\speakernoteb{Webthrithner:}%
“From the \inx[L]{Ilewaves} splashed venom-drops; \\
\ind so it grew until it formed an ettin. \\
Our lineages came there all together, \\
\ind thus it is ever all too fierce.”\evb\evg


\bvg\bva\speakernote{Óðinn kvað:}\mssnote{\Regius~8r/26, \AM~3r/16}%
„\alst{S}ęg þat \alst{s}jaunda, \hld\ alls þik \alst{s}vinnan kveða, &
\ind ok þú \alst{V}af-þrúðnir \alst{v}itir, &
hvé sá \alst{b}ǫrn gat \hld\ hinn \edtrans{\alst{b}aldni}{stubborn}{\Afootnote{so \AM; \emph{aldni} ‘the aged, old’ \Regius\ breaks alliteration}} jǫtunn, &
\ind es hann hafði-t \alst{g}ýgjar \alst{g}aman.“\eva

\bvb\speakernoteb{Weden quoth:}%
“Tell this seventh, since they call thee wise, \\
\ind and thou, Webthrithner, mightst know: \\
How that one begot bairns, the stubborn ettin, \\
\ind when he knew not a troll-woman’s pleasure?”\evb\evg


\bvg\bva\speakernote{Vafþrúðnir kvað:}\mssnote{\Regius~8r/27, \AM~3r/17}%
„\edtext{Und \alst{h}ęndi vaxa \hld\ kvǫ́ðu \alst{h}rím-þursi &
\ind \alst{m}ęy ok \alst{m}ǫg saman; &
\alst{f}ótr við \alst{f}ǿti}{\lemma{Und hęndi \dots\ fótr við fǿti ‘Within the hand \dots\ Foot against foot’}\Bfootnote{The image is masturbatory.  The stanza is paraphrased in \Gylfaginning\ 5: \emph{En svá er sagt, at þá er hann svaf, fekk hann sveita. Þá óx undir vinstri hendi honum maðr ok kona, ok annarr fótr hans gat son við ǫðrum, en þaðan af kómu ę́ttir.} ‘But so is said, that when he slept he began to sweat.  Then grew within his left hand a man and a woman, and one foot of his begat a son by the other, and thereof come the lineages [of Ettins].’}} \hld\ gat hins \alst{f}róða jǫtuns &
\ind \alst{s}ex-hǫfðaðan \alst{s}on.“\eva

\bvb\speakernoteb{Webthrithner quoth:}%
“In the hand of the \inx[G]{Rime-Thurses}[rime-thurse], they said, did grow \\
\ind a maiden and lad together. \\
Foot by a foot begat for the learned ettin \\
\ind a six-headed son.”\evb\evg


\bvg\bva\speakernote{Óðinn kvað:}\mssnote{\Regius~8r/29, \AM~3r/18}%
„Sęg þat ǫ́ttunda, \hld\ alls þik fróðan kveða, &
\ind ok þú \alst{V}af-þrúðnir \alst{v}itir, &
hvat \alst{f}yrst of mant \hld\ eða \alst{f}ręmst of vęitst, &
\ind þú est \alst{a}l-sviðr \alst{jǫ}tunn.“\eva

\bvb\speakernoteb{Weden quoth:}%
“Tell this eighth, since they call thee learned, \\
\ind and thou, Webthrithner, mightst know: \\
What recallest thou first, or knowest foremost? \\
\ind Thou art all-wise, ettin!”\evb\evg


\bvg\bva\speakernote{Vafþrúðnir kvað:}\mssnote{\Regius~8r/30, \AM~3r/19, \GylfMS}%
„\alst{Ø}r-ófi vetra \hld\ áðr vę́ri \alst{jǫ}rð of skǫpuð, &
\ind þá vas \alst{B}er-gęlmir \alst{b}orinn; &
þat \alst{f}yrst of man’k, \hld\ es hinn \alst{f}róði jǫtunn &
\ind á vas \alst{l}úðr of \alst{l}agiðr.“\eva

\bvb\speakernoteb{Webthrithner quoth:}%
“Uncountable winters before the Earth was created, \\
\ind then was Bareyelmer born. \\
It I first remember, when the learned ettin \\
\ind on the tree-trunk was laid.\footnoteB{An obscure mythological reference.  According to the prose of \Gylfaginning, after the sons of \inx[P]{Byre} (that is, \inx[P]{Weden}, \inx[P]{Will} and \inx[P]{Wigh}) slew Yimer, so much blood flew from his wounds that all the race of Ettins were drowned, save for Bareyelmer and his family, who survived by getting up on his \emph{lúðr}.  This is clearly a variant of the flood myth, but it may be of Biblical origin.

In regular prose, \emph{lúðr} usually means ‘trumpet, blowing horn’, less commonly ‘flour-bin’; the underlying sense seems to be ‘hollowed-out wood’.  Considering the transitive nature of Bareyelmer being laid (\emph{of lagiðr}) upon it, the stanza might instead be referring a ship burial, so that the first thing Webthrithner remembers is Bareyelmer’s funeral.}”\evb\evg


\bvg\bva\speakernote{Óðinn kvað:}\mssnote{\Regius~8r/32, \AM~3r/21}%
„\alst{S}ęg þat níunda, \hld\ alls þik \alst{s}vinnan kveða, &
\ind ok þú \alst{V}af-þrúðnir \alst{v}itir, &
hvaðan \alst{v}indr of kømr \hld\ svá’t fęrr \alst{v}ág yfir, &
\ind \edtrans{ę́ męnn hann \alst{s}jalfan of \alst{s}éa}{men always see his very self}{\Bfootnote{Most likely a negative clitic \emph{-t} has been lost from the verb \emph{séa} ‘see’, which would have given the proper reading: “men \emph{never} see his very self”.}}.“\eva

\bvb\speakernoteb{Weden quoth:}%
“Tell this ninth, since they call thee wise, \\
\ind and thou, Webthrithner, mightst know: \\
From whence the wind comes which fares over the wave; \\
\ind men always see his very self?”\evb\evg


\bvg\bva\speakernote{Vafþrúðnir:}\mssnote{\Regius~8r/34, \AM~3r/22}%
„\alst{H}rę́-svęlgr \alst{h}ęitir, \hld\ es sitr á \alst{h}imins ęnda, &
\ind \alst{jǫ}tunn í \alst{a}rnar ham; &
af hans \alst{v}ę́ngjum \hld\ kveða \alst{v}ind koma &
\ind \alst{a}lla męnn \alst{y}fir.“\eva

\bvb\speakernoteb{Webthrithner:}%
“\inx[P]{Rawswallower} is he called who sits at heaven’s end; \\
\ind an ettin in an eagle’s \inx[C]{hame}. \\
From his wings they say that the wind comes \\
\ind over all men.”\evb\evg


\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~8v/1, \AM~3r/24}%
„Sęg þat \alst{t}íunda, \hld\ alls þú \alst{t}íva rǫk &
\ind ǫll \alst{V}afþrúðnir \alst{v}itir, &
hvaðan Njǫrðr of kom \hld\ með ása sonum; &
\edtrans{\alst{h}ofum ok \alst{h}ǫrgum}{hoves and harrows}{\Bfootnote{A formulaic merism, see note to \Voluspa\ 7 for other occurrences. This stanza seems to be referring to the large count of cultic places named after Nearth in Norway (TODO: source this); cf. here \Grimnismal\ 16, where it is said that Nearth \emph{rę́ðr hǫ́-timbruðum hǫrgi} ‘rules a high-timbered harrow’. Also of interest is \Lokasenna\ 51, where a goddess speaks about her \emph{véum ok vǫngum} ‘wighs and wongs’, other cultic names. All of these examples suggest something about the Heathen view of shrines.}} \hld\ rę́ðr \alst{h}und-mǫrgum &
\ind ok varð-at \alst{ǫ́}sum \alst{a}linn.“\eva

\bvb\speakernoteb{[Weden quoth:]}%
“Tell this tenth, since thou of the \inx[L]{Rakes of the Tews} \\
\ind all, O Webthrithner, mightst know: \\
From whence \inx[P]{Nearth} did come amidst the sons of the \inx[G]{Eese}? \\
\inx[C]{hove}[Hoves] and \inx[C]{harrow}[harrows] he rules hound-many, \\
\ind and he was not by the Eese begotten.”\evb\evg


\bvg\bva\speakernote{[Vafþrúðnir kvað:]}\mssnote{\Regius~8v/3, \AM~3r/26}%
„Í \alst{V}ana-hęimi \hld\ skópu hann \alst{v}ís \edtrans{ręgin}{Reins}{\Bfootnote{\emph{ręgin} ‘the Reins, Powers’ is generally used simply to refer to the gods as a collective, but here seems to refer specifically to the \inx[G]{Wanes}, in contrast with the \inx[G]{Eese}.}} &
\ind ok sęldu at \alst{g}íslingu \alst{g}oðum, &
í \edtrans{\alst{a}ldar rǫk}{the Rakes of the Age}{\Bfootnote{The \inx[L]{Rakes of the Reins}; the time of the destruction of the world.}} \hld\ hann mun \alst{a}ptr koma &
\ind hęim með \alst{v}ísum \alst{v}ǫnum.“\eva

\bvb\speakernoteb{[Webthrithner quoth:]}%
“In \inx[L]{Waneham} the wise \inx[G]{Reins}created him, \\
\ind and sold him as a hostage to/for the gods. \\
In the \inx[L]{Rakes of the Age} he will come back \\
\ind home amidst the wise \inx[G]{Wanes}.”\evb\evg

\sectionline

{\small Stanzas 40–41 are malformed in both \Regius\ and \AM, and thus have to be reconstructed.  \Regius\ has only what is here st. 40; \AM\ has only the very beginning of st. 40 (“Tell this eleventh”), followed by the full st. 41.  In diplomatic transcription:

\textbf{R} \emph{Segðv þat et .xi. hvar ytar tvnom i hǫggvaz hverian dag. Val þeir kíosa oc riþa vígi fra sitia meirr vm sáttir saman.} \\
\textbf{A} \emph{Sæg þat et .xi. allir eins hæriar oðins tvnvm i hǫgguaz hværian dag. Val þæir kiosa ok riða vigi fra sitja mæirr vm sattir saman.}}

\sectionline

\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~8v/5, \AM~3r/28}%
„Sęg þat \alst{ę}llipta, \hld\ hvar \alst{ý}tar túnum í &
\ind \alst{h}ǫggvask \alst{h}vęrjan dag; &
\edtrans{\alst{v}al þęir kjósa}{The slain they choose}{\Bfootnote{It is from this verbal phrase that the female agent noun \emph{val-kyrja} ‘\inx[G]{walkirries}[walkirrie]’ is derived.}} \hld\ ok ríða \alst{v}ígi frá, &
\ind \alst{s}itja męirr of \alst{s}áttir \alst{s}aman.“\eva

\bvb\speakernoteb{[Weden quoth:]}%
“Tell this eleventh, where men in yards \\
\ind strike at each other every day? \\
The slain they choose and they ride from the fray; \\
\ind then they sit at peace together.”\evb\evg


\bvg\bva\speakernote{[Vafþrúðnir kvað:]}\mssnote{\AM~3r/28}%
„\alst{A}llir \alst{ęi}n-hęrjar \hld\ \alst{Ó}ðins túnum í &
\ind \alst{h}ǫggvask \alst{h}vęrjan dag, &
\alst{v}al þęir kjósa \hld\ ok ríða \alst{v}ígi frá, &
\ind \alst{s}itja męirr of \alst{s}áttir \alst{s}aman.“\eva

\bvb\speakernoteb{[Webthrithner quoth:]}%
“All the \inx[G]{Oneharriers} in Weden’s yards \\
\ind strike at each other every day. \\
The slain they choose and they ride from the fray; \\
\ind then they sit at peace together.”\evb\evg


\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~8v/6, \AM~3v/1}%
„Sęg þat \alst{t}olpta, \hld\ hví þú \alst{t}íva rǫk &
\ind ǫll \alst{V}af-þrúðnir \alst{v}itir? &
Frá \alst{jǫ}tna rúnum \hld\ ok \alst{a}llra goða &
\ind þú hit \alst{s}annasta \alst{s}ęgir, &
\ind hinn \alst{a}l-svinni \alst{jǫ}tunn.“\eva

\bvb\speakernoteb{[Weden quoth:]}%
“Tell this twelfth, why thou the Rakes of the Tews \\
\ind all, Webthrithner, shouldst know? \\
From the \inx[C]{rune}[runes] of the ettins and of all the gods \\
\ind dost thou speak the most truly, \\
\ind O all-wise ettin!”\evb\evg


\bvg\bva\speakernote{[Vafþrúðnir kvað:]}\mssnote{\Regius~8v/8, \AM~3v/2}„Frá \alst{jǫ}tna rúnum \hld\ ok \alst{a}llra goða &
\ind ek kann \alst{s}ęgja \alst{s}att, &
\ind því-at \alst{h}vęrn hęf’k \alst{h}ęim of komit, &
\edtext{\alst{n}íu kom’k hęima \hld\ fyr \alst{n}ifl-hęl neðan; &
\ind \alst{h}inig dęyja ór \alst{h}ęlju \alst{h}alir.}{\lemma{níu \dots\ halir. ‘Into nine \dots\ of Hell.’}\Bfootnote{Perhaps lower infernal underworlds.  \textcite{FinnurEdda}\ considers \emph{ór hęlju} ‘out of Hell’ a later interpolation, probably for metrical reasons.}}“\eva

\bvb\speakernoteb{[Webthrithner quoth:]}%
“From the runes of the ettins and of all the gods \\
\ind I can speak truly, \\
\ind for I have come into each \inx[C]{Home}. \\
Into nine Homes I came beneath \inx[L]{Nivelhell}; \\
\ind that way men die out of \inx[L]{Hell}.”\evb\evg

\sectionline

\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~8v/11, \AM~3v/4}%
„\alst{F}jǫlð ek \alst{f}ór, \hld\ \alst{f}jǫlð \alst{f}ręistaða’k, &
\ind fjǫlð ek \alst{r}ęynda \alst{r}ęgin; &
hvat lifir \alst{m}anna, \hld\ þá’s hinn \alst{m}ę́ra líðr &
\ind \alst{f}imbul-vetr með \alst{f}irum?“\eva

\bvb\speakernoteb{[Weden quoth:]}%
“Much I journeyed, much I tried, \\
\ind much I tested the Reins. \\
What remains of men when the renowned \inx[L]{Fimble-winter} \\
\ind passes amidst the folk?”\evb\evg


\bvg\bva\speakernote{[Vafþrúðnir kvað:]}\mssnote{\Regius~8v/13, \AM~3v/6}%
„\alst{L}íf ok \alst{L}ífþrasir, \hld\ en þau \alst{l}ęynask munu &
\ind í \edtrans{\alst{h}olti \alst{H}odd-mímis}{in Hoardmimer’s wood}{\Bfootnote{Perhaps the hollowed-out \inx[L]{Uggdrassle’s Ash.}}}; &
\alst{m}orgin-dǫggvar \hld\ þau sér at \alst{m}at hafa; &
\ind þaðan af \alst{a}ldir \alst{a}lask.“\eva

\bvb\speakernoteb{[Webthrithner quoth:]}%
“\inx[P]{Life} and \inx[P]{Lifethrasher}—but they will hide themselves \\
\ind in \inx[P]{Hoardmimer}’s wood. \\
Morning dew will they have for food; \\
\ind from thence is mankind begotten.”\evb\evg


\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~8v/15, \AM~3v/8}%
„\alst{F}jǫlð ek \alst{f}ór, \hld\ \alst{f}jǫlð \alst{f}ręistaða’k, &
\ind fjǫlð ek \alst{r}ęynda \alst{r}ęgin; &
hvaðan kømr \alst{s}ól \hld\ á hinn \alst{s}létta himin, &
\ind es \edtrans{þessa}{this one}{\Bfootnote{The present sun, as explained in the following st.}} hęfr \edtrans{\alst{F}ęnrir}{Fenrer}{\Bfootnote{Perhaps not the same “Fenrerswolf” that fights against Weden (cf. st. 53 below); the word, which originally prob. means “Fen-creature”, may here simply mean “Wolf”.  For the wolves who chase the sun and moon see \Voluspa\ 40 and \Grimnismal\ 40.}} \alst{f}arit?\eva

\bvb\speakernoteb{[Weden quoth:]}%
“Much I journeyed, much I tried, \\
\ind much I tested the Reins! \\
From whence comes Sun onto the smooth heaven, \\
\ind when \inx[P]{Fenrer} has destroyed this one?”\evb\evg


\bvg\bva\speakernote{[Vafþrúðnir kvað:]}\mssnote{\Regius~8v/16, \AM~3v/9}%
„\alst{Ęi}na dóttur \hld\ berr \alst{a}lf-rǫðull, &
\ind áðr hana \alst{F}ęnrir \alst{f}ari; &
sú skal \alst{r}íða, \hld\ þá’s \alst{r}ęgin dęyja, &
\ind \alst{m}óður brautir \alst{m}ę́r.“\eva

\bvb\speakernoteb{[Webthrithner quoth:]}%
“One daughter the elf-wheel \ken*{= Sun} bears \\
\ind before Fenrer might slay her. \\
She shall ride—when the Reins die— \\
\ind the maiden, her mother’s paths.”\evb\evg


\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~8v/18, \AM~3v/10}%
\alst{F}jǫlð ek \alst{f}ór, \hld\ \alst{f}jǫlð \alst{f}ręistaða’k, &
\ind fjǫlð ek \alst{r}ęynda \alst{r}ęgin; &
\edtext{hvęrjar ’ru \alst{m}ęyjar, \hld\ es líða \alst{m}ar yfir, &
\ind \alst{f}róð-gęðjaðar \alst{f}ara?}{\lemma{hvęrjar \dots\ fara? ‘Who \dots\ go?’}\Bfootnote{The identity of these maidens is very mysterious, and Webthrithner’s answer in the next st. does not give much more information.  Considering all other questions introduced with the words \emph{fjǫlð ek fór} et.c. have something to do with the end times, this one should as well.  With this in mind they are probably to be identified with the maidens Weden asks about in \Baldrsdraumar\ 12.}}\eva

\bvb\speakernoteb{[Weden quoth:]}%
“Much I journeyed, much I tried, \\
\ind much I tested the Reins! \\
Who are the maidens that pass over the ocean; \\
\ind wise-minded they go?”\evb\evg


\bvg\bva\speakernote{[Vafþrúðnir kvað:]}\mssnote{\Regius~8v/19, \AM~3v/11}%
\alst{Þ}ríar \alst{þ}jóð-áar \hld\ falla \alst{þ}orp yfir &
\ind \alst{m}ęyja \alst{M}ǫg-þrasis; &
\alst{h}amingjur ęinar \hld\ þę́r’s í \alst{h}ęimi eru, &
\ind þó þę́r með \alst{jǫ}tnum \alst{a}lask.\eva

\bvb\speakernoteb{[Webthrithner quoth:]}%
“Three great rivers fall over the house \\
\ind of the maidens of Maythrasher; \\
they are the only Hamings in the Home, \\
\ind although they are raised amidst ettins.”\evb\evg


\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~8v/21, \AM~3v/13}%
„\alst{F}jǫlð ek \alst{f}ór, \hld\ \alst{f}jǫlð \alst{f}ręistaða’k, &
\ind fjǫlð ek \alst{r}ęynda \alst{r}ęgin; &
hvęrir ráða \alst{ę́}sir \hld\ \alst{ęi}gnum goða, &
\ind þá’s \alst{s}loknar \alst{S}urta-logi?“\eva

\bvb\speakernoteb{[Weden quoth:]}%
“Much I journeyed, much I tried, \\
\ind much I tested the Reins! \\
Which Eese rule the ownings of the Gods \\
\ind when the flame of \inx[P]{Surt} goes out?”\evb\evg


\bvg\bva\speakernote{[Vafþrúðnir kvað:]}\mssnote{\Regius~8v/22, \AM~3v/14}%
„\alst{V}íðarr ok \alst{V}ȧli \hld\ byggva \alst{v}é goða, &
\ind þá’s \alst{s}loknar \alst{S}urta-logi; &
\alst{M}óði ok \alst{M}agni \hld\ skulu \alst{M}jǫllni hafa &
\ind \edtrans{\alst{V}ingnis at \alst{v}íg-þroti}{after Wingner’s end in battle}{\Bfootnote{When Thunder dies.}}.“\eva

\bvb\speakernoteb{[Webthrithner quoth:]}%
“\inx[P]{Wider} and \inx[P]{Wonnel} bedwell the \inx[C]{wigh}[wighs] of the gods \\
\ind when the flame of Surt goes out. \\
\inx[P]{Mood} and \inx[P]{Main} shall have \inx[P]{Millner} \\
\ind after \inx[P]{Wingner}’s end in battle.”\evb\evg


\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~8v/24, \AM~3v/16}%
„\alst{F}jǫlð ek \alst{f}ór, \hld\ \alst{f}jǫlð \alst{f}ręistaða’k, &
\ind fjǫlð ek \alst{r}ęynda \alst{r}ęgin; &
hvat verðr \alst{Ó}ðni \hld\ at \alst{a}ldr-lagi, &
\ind \edtrans{þá’s \alst{r}júfask \alst{r}ęgin?}{when the Reins are ripped?}{\Bfootnote{Formulaic; see note to \Baldrsdraumar\ 14/1.}}“\eva

\bvb\speakernoteb{[Weden quoth:]}%
“Much I journeyed, much I tried, \\
\ind much I tested the Reins! \\
What brings Weden’s life to an end, \\
\ind when the Reins are ripped?”\evb\evg


\bvg\bva\speakernote{[Vafþrúðnir kvað:]}\mssnote{\Regius~8v/25, \AM~3v/17}%
„\alst{U}lfr glęypa \hld\ mun \alst{A}lda-fǫðr, &
\ind þęss mun \alst{V}íðarr \alst{v}reka; &
\alst{k}alda \alst{k}japta \hld\ hann \alst{k}lyfja mun &
\ind \alst{v}itnis \alst{v}ígi at.“\eva

\bvb\speakernoteb{[Webthrithner quoth:]}%
“The Wolf will devour the Father of Men: \\
\ind that will Wider avenge. \\
The cold jaws he will split apart \\
\ind of the beast at the battle.”\evb\evg


\bvg\bva\speakernote{[Óðinn kvað:]}\mssnote{\Regius~8v/27, \AM~3v/19}%
„\alst{F}jǫlð ek \alst{f}ór, \hld\ \alst{f}jǫlð \alst{f}ręistaða’k, &
\ind fjǫlð ek \alst{r}ęynda \alst{r}ęgin; &
hvat mę́lti Óðinn, \hld\ áðr \edtrans{á bál stigi}{step onto the pyre}{\Bfootnote{The phrase \emph{stíga á} ‘step onto, mount’ is also used to refer to one stepping aboard a ship or mounting a horse (see \CV: \emph{stíga} for citations).  Its use for a person being borne onto the funeral pyre has been compared with \Beowulf\ 1118b: \emph{gu̇ð-rinc á·stâh} ‘the war-champion mounted [his pyre]’, but the interpretation of that line is controversial; \textcite{KlaeberBeowulf}[186] follow Grundtvig in emending \emph{gu̇ð-rinc} to \emph{gu̇ð-réc} ‘war-smoke’ and compare it with \Beowulf\ 3144b (\emph{wudu-réc á·stâh} ‘wood-smoke rose up’, which also describes a cremation; (according to them) the present stanza “almost certainly refers not to Baldr but to Óðinn, probably imagined to mount the pyre in order to set fire to it.”}}, &
\ind \alst{s}jalfr í ęyra \alst{s}yni?“\eva

\bvb\speakernoteb{[Weden quoth:]}%
“Much I journeyed, much I tempted, \\
\ind much I tested the Reins! \\
What spoke Weden, before he would step onto the pyre, \\
\ind himself in his son’s \ken*{= Balder’s} ear?”\evb\evg


\bvg\bva\speakernote{[Vafþrúðnir kvað:]}\mssnote{\Regius~8v/28, \AM~3v/19}%
„\alst{Ęy} \edtext{mann-gi}{\Afootnote{\emph{manni} dat. sg. \Regius\AM\ is impossible; a subject is needed.}} vęit, \hld\ hvat þú í \alst{á}r-daga &
\ind \alst{s}agðir í ęyra \alst{s}yni; &
\edtrans{\alst{f}ęigum}{fey}{\Bfootnote{A word with strong fatalistic connections. Webthrithner realises that he was bound to die from the moment he proposed the wager (st. 19), as no being can outwit Weden.}} munni \hld\ mę́lta’k \edtrans{mína \alst{f}orna stafi}{my ancient staves}{\Bfootnote{Referencing st. 1.}} &
\ind ok of \alst{r}agna \alst{r}ǫk; &
nú við \alst{Ó}ðin \hld\ dęilda’k mína \edtrans{\alst{o}rð-spęki}{word-wisdom}{\Bfootnote{Referencing st. 5.}}; &
\ind þú est ę́ \alst{v}ísastr \edtrans{\alst{v}era}{of men}{\Bfootnote{\emph{verr} means ‘husband, man’ and is here used for reasons of alliteration; it does not imply that Weden is not a God.}}.“\eva

\bvb\speakernoteb{[Webthrithner quoth:]}%
“Never will man know what thou in days of yore \\
\ind saidst in the ear of thy son. \\
With a \inx[C]{fey} mouth I spoke my ancient \inx[C]{stave}[staves], \\
\ind and about the Rakes of the Reins. \\
Now with Weden have I shared my word-wisdom— \\
\ind thou art ever wisest of men!”\evb\evg

\sectionline
