\bookStart{Widsith}[Wídsïþ]

\begin{flushright}%
\textbf{Dating:} 600–700s (Neidorf 2013)

\textbf{Meter:} \Fornyrdislag%para
\end{flushright}%

\section{Introduction}

An archaic heroic poem.

\sectionline

\section{Widsith}

\bvg\bva \alst{W}íd-sïð maðolade, \hld\ \alst{w}ord-hord ǫn·leac, &
sé þe \alst{m}æ̂st \hld\ \alst{m}ærþa ofer eorþan, &
\alst{f}olca geond·\alst{f}ǿrde; \hld\ oft hé \alst{f}lętte ge·þah &
\alst{m}yne-lícne \alst{m}âþþum. \hld\ Hine frǫm \alst{M}yrgingum &
\alst{æ}þele \alst{ǫ}n·wócon. \hld\ He mid \alst{E}alh-hilde, &
\alst{f}æ̂lre \edtrans{\alst{f}reoþu-wębban}{peace-weaveress}{\Bfootnote{A woman used in a political marriage to bring peace between two tribes or families, in this case between King Edwin of the Mirgings (see ll. 97–98) and Erminric of the Gots.}}, \hld\ \alst{f}orman sïþe &
\edtrans{\alst{H}reð-cyninges}{Reth-King}{\Bfootnote{The king of the \inx[G]{Reth-Gots}, which is apparently just a poetic name for the (Eastern) Gots; cf. ll. 18, 57, 88–89.}} \hld\ \alst{h}âm ge·sóhte &
\alst{éa}stan of \alst{Ǫ}ngle, \hld\ \alst{Eo}rman-ríces, &
\alst{w}râþes \alst{w}ær-logan. \hld\ Ǫn·gǫnn þá \alst{w}orn sprecan:\eva

\bvb Widsith spoke, unlocked his word-hoard, \\
he who mots through tribes on earth \\
and nations had journeyed. Oft on the bench had he received \\
delightful treasures. From the Mirgings \\
his ancestry stemmed. Along with Elhild, \\
the good peace-weaveress, for the first time \\
had he sought out the Reth-King’s realm, \\
east of the Angles, [the realm of] \inx[P]{Erminric}, \\
the fierce oath-breaker.  He then began a long speech:\evb\evg


\bvg\bva „Fela ic \alst{m}ǫnna ge·frægn \hld\ \alst{m}ægþum wealdan. &
Sceal \alst{þ}eóda ge·hwylc \hld\ \alst{þ}éawum lifgan, &
\alst{eo}rl æfter \alst{ȯ}þrum \hld\ \alst{ǿ}ðle rǽdan, &
sé þe his \alst{þ}eóden-stól \hld\ ge·\alst{þ}éon wile.\eva

\bvb “A great deal of men I’ve learned ruling tribes. \\
Every person shall live in virtue; \\
each earl after the other lead his homeland, \\
he who on his ruling-seat will prosper.\evb\evg


\bvg\bva Þâra wæs Wala \hld\ hwíle sélast, &
ǫnd \alst{A}lexandreas \hld\ \alst{ea}lra rícost &
\alst{m}ǫnna cynnes, \hld\ ǫnd he \alst{m}æ̂st ge·þȧh &
þâra þe ic ofer \alst{f}oldan \hld\ ge·\alst{f}rægen hæbbe.\eva

\bvb Of them was Wale for a while the most blessed, \\
and Alexander of all the strongest \\
of mankind, and he prospered most \\
of those men over the earth of whom I’ve learned.\evb\evg


\bvg\bva Ætla weold Húnum, \hld\ Eorman-ríc Gotum, &
Becca Baningum, \hld\ Burgendum Gifica. &
Câsere weold Créacum \hld\ ǫnd Cælic Finnum, &
Hagena Holm-rycum \hld\ ǫnd Henden Glommum.\eva

\bvb Attle ruled the Huns, Erminric the Gots, \\
Bicke the Banings, the Burgends Yivick. \\
Choser ruled the Greeks and Calic the Finns, \\
Hain the Holmrighs and Henden the Glams.”\evb\evg


\bvg\bva Witta weold Swǽfum, \hld\ Wada Hælsingum, &
Meaca Myrgingum, \hld\ Mearc-healf Hundingum. &
Þeód-ríc weold Frǫncum, \hld\ Þyle Rǫndingum, &
Breoca Brǫndingum, \hld\ Billing Wernum.\eva

\bvb TODO.\evb\evg


\bvg\bva Ȯswine weold Eowum \hld\ ǫnd Ytum Gef-wulf, &
Finn Folc-walding \hld\ Fresna cynne. &
Sige-hęre lęngest \hld\ Sæ̂-dęnum weold, &
Hnæf Hocingum, \hld\ Helm Wulfingum, &
Wald Wóingum, \hld\ Wód Þyringum, &
Sæ̂-ferð Sycgum, \hld\ Swéom Ongend-þeow, &
Sceaft-hęre Ymbrum, \hld\ Sceafa Lǫng-beardum, &
Hún Hæt-werum \hld\ ǫnd Holen Wrosnum; &
Hring-wald wæs hâten \hld\ Hęre-farena cyning.\eva

\bvb TODO.\evb\evg


\bvg\bva Offa weold Ǫngle, \hld\ Ale-wíh Dęnum; &
sé wæs þâra manna \hld\ módgast ealra, &
no hwæþre he ofer Offan \hld\ eorl-scype fręmede, &
ac Offa ge·slóg \hld\ æ̂rest mǫnna, &
cniht-wesende, \hld\ cyne-ríca mæ̂st.\eva

\bvb Offe ruled the Angles, Alewigh the Danes; \\
of those men he was the bravest of all, \\
but he never furthered greater earlship than Offe, \\
for Offe won—youngest of men, \\
still a boy—the greatest of kingdoms.\evb\evg


\bvg\bva Nænig efen-eald him \hld\ eorl-scipe mâran &
ǫn orette: \hld\ âne sweorde &%TODO: orette.
męrce ge·mæ̂rde \hld\ wið Myrgingum &
bi Fifel-dore; \hld\ heoldon forð siþþan &
Ęngle ǫnd Swǽfe, \hld\ swá hit Offa ge·slóg.\eva

\bvb No man of his age accomplished \\
greater earlship: with but one sword \\
he marked the border against the Mirgings, \\
by Fiveldoor. It was thenceforth held \\
by the Angles and Sweves as Offe had won it.\evb\evg


\bvg\bva Hróþ-wulf ǫnd Hróð-gâr \hld\ heoldon lęngest &
sibbe æt·somne \hld\ suhtor-fædran, &
siþþan hý for·wrǽcon \hld\ Wícinga cynn &
ǫnd Ingeldes \hld\ ord for·bigdan, &
for·heowan æt Heorote \hld\ Heaðo-beardna þrym.\eva

\bvb Rotholf and Rothgar held for the longest \\
the peace together, uncle and nephew, \\
since they drove away the race of Wikings, \\
and bent down Ingeld’s spear-point; \\
at Hart they cut down the host of the Hathbeards.\evb\evg

\sectionline

\bvg\bva Swá ic geond·fǿrde fela \hld\ fręmdra lǫnda &
geond ginne grund. \hld\ Gódes ǫnd yfles &
þær ic cunnade; \hld\ cnósle bi·dæ̂led, &
fréo-mǽgum feor \hld\ folgade wíde.\eva

\bvb So I journeyed through a great deal of strange lands \\
through the wide world. Of good and evil \\
I there became acquainted; of kin deprived, \\
far from dear kinsmen, I strayed widely.\evb\evg


\bvg\bva For·þǫn ic mæg singan \hld\ ǫnd sęcgan spell, &
mæ̂nan fore męngo \hld\ in meodu-healle &
hú mé cyne-góde \hld\ cystum dohten.\eva

\bvb Therefore I can sing and tell tales, \\
recall before the many in the mead-hall, \\
how men of good kin treated me with grace.\evb\evg


\bvg\bva Ic wæs mid Húnum \hld\ ǫnd mid Hreð-gotum, &
mid Swéom ǫnd mid Géatum \hld\ ǫnd mid Su̇þ-dęnum. &
Mid Wenlum ic wæs ǫnd mid Wærnum \hld\ ǫnd mid wícingum; &
mid Gefþum ic wæs ǫnd mid Winedum \hld\ ǫnd mid Gefflegum; &
mid Englum ic wæs ǫnd mid Swǽfum \hld\ ǫnd mid Ænenum; &
mid Seaxum ic wæs ǫnd Sycgum \hld\ ǫnd mid Sweord-werum; &
mid Hronum ic wæs ǫnd mid Deanum \hld\ ǫnd mid Heaþo-réamum.\eva

\bvb I was among Huns and among Reth-Gots, \\
among Swedes and among Geats, and among South-Danes. \\
Among Wendles I was and among Warns, and among Wikings; \\
among Yefths I was and among Wends, and among Yefflegs; \\
among Angles I was and among Sweves, and among Anens; \\
among Saxes I was and among Sidges, and among Sword-weres; \\
among Ranes I was and among Deans, and among Hath-Reams.\evb\evg


\bvg\bva Mid \alst{Þ}yringum ic wæs \hld\ ǫnd mid \alst{Þ}rowendum, &
ǫnd mid \alst{B}urgendum, \hld\ þær ic \alst{b}éag ge·þâh; &
mé þær \alst{G}u̇ð-hęre for·\alst{g}eaf \hld\ \alst{g}læd-lícne maþþum &
\alst{s}ǫnges to léane. \hld\ Næs þæt \alst{s}æne cyning!\eva%TODO: Check sæne.

\bvb Among Thirings I was and among Throwends, \\
and among the Burgends, where I received a bigh. \\
There Guther gladdened me with treasures, \\
as reward for my song. That was not a bad king!\evb\evg


\bvg\bva Mid \alst{F}rǫncum ic wæs ǫnd mid \alst{F}rysum \hld\ ǫnd mid \alst{F}rumtingum; &
mid \alst{R}ugum ic wæs ǫnd mid Glommum \hld\ ǫnd mid \alst{R}úm-walum.\eva

\bvb Among Franks I was and among Frises, and among Frumtings; \\
among Ruges I was and among Glams, and among Rome-Wales.\evb\evg

\sectionline

\bvg\bva Swylce ic wæs ǫn \alst{Ea}tule \hld\ mid \alst{Æ}lf-wine, &
sé hæfde \alst{m}ǫn-cynnes, \hld\ \alst{m}íne ge·fræge, &
\alst{l}eohteste hǫnd \hld\ \alst{l}ofes tó wyrcenne, &
\alst{h}eortan un·\alst{h}neaweste \hld\ \alst{h}ringa ge·dâles, &
\alst{b}eorhtra \alst{b}éaga, \hld\ \alst{b}earn Éad-wines.\eva

\bvb Likewise was I in Italy with Elfwin; \\
of mankind he had—as far as I have learned— \\
the lightest hand in the winning of praise, \\
the unstingiest heart in the dealing of rings \\
and bright bighs, that child of Edwin.\evb\evg


\bvg\bva Mid \alst{S}ercingum ic wæs \hld\ ǫnd mid \alst{S}eringum; &
mid \alst{C}reacum ic wæs ǫnd mid Finnum \hld\ ǫnd mid \alst{C}âsere, &
sé þe \alst{w}in-burga \hld\ ge·\alst{w}eald áhte, &
\alst{w}iolena ǫnd \alst{w}ilna, \hld\ ǫnd \alst{W}ala rices.\eva

\bvb TODO.\evb\evg


\bvg\bva Mid Scottum ic wæs ǫnd mid Peohtum \hld\ ǫnd mid Scríde-finnum; &
mid Líd-wícingum ic wæs ǫnd mid Léonum \hld\ ǫnd mid Lǫng-beardum, &
mid hæ̂ðnum ǫnd mid hæleþum \hld\ ǫnd mid Hundingum.\eva

\bvb Among Scots I was and among Picts, and among Shride-Finns; \\
among Lid-Wikings I was among Leans, and among Longbeards; \\
among heathens and among heroes and among Hundings.\evb\evg


\bvg\bva Mid Israhelum ic wæs \hld\ ǫnd mid Exsyringum, &
mid Ebreum ǫnd mid Indeum \hld\ ǫnd mid Egyptum. &
Mid Moidum ic wæs ǫnd mid Persum \hld\ ǫnd mid Myrgingum, &
ǫnd Mofdingum \hld\ ǫnd on·gend Myrgingum, &
ǫnd mid Amothingum. \hld\ Mid Éast-þyringum ic wæs &
ǫnd mid Eolum ǫnd mid Istum \hld\ ǫnd Idumingum.\eva

\bvb Among Israelites I was and among Assyrians, \\
among Hebrews and among Indians and among Egyptians. \\
Among the Medes I was and among Persians, and among Mirgings \\
and Mofdings and again the Mirgings \\
and among Amothings. Among East-Thirings I was \\
and among Eals and among Ists, and Idumings.\evb\evg


\bvg\bva Ǫnd ic wæs mid Eorman-ríce \hld\ ealle þráge, &
þær mé Gotena cyning \hld\ góde dohte; &
sé mé béag for·geaf, \hld\ burg-warena fruma, &
ǫn þam siex hund wæs \hld\ smǽtes goldes, &
ge·scyred sceatta \hld\ scilling-ríme; &
þǫne ic Ead-gilse \hld\ ǫn æht sealde, &
mínum hléo-dryhtne, \hld\ þa ic to hâm bi·cwǫm, &
leófum to léane, \hld\ þæs þe hé mé lǫnd for·geaf, &
mínes fæder ǿþel, \hld\ fréa Myrginga.\eva

\bvb And I was with Ermenric for the longest time, \\
where the king of the Gots treated me well. \\
He gave me a bigh—that chief of city-dwellers— \\
in which were reckoned six hundred shats \\
of purest gold in shilling-count. \\
I gave it in the possession of Edgils \\
my dear shelter and lord, when I came home, \\
as repayment for his giving me land, \\
—that lord of Mirgins—my father’s ethel.\evb\evg


\bvg\bva Ǫnd mé þá Ealh-hild \hld\ ȯþerne for·geaf, &
dryht-cwén duguþe, \hld\ dohtor Éad-wines. &
Hyre lof lęngde \hld\ geond lǫnda fela, &
þǫnne ic be sǫnge \hld\ sęcgan sceolde &
hwær ic under swegl \hld\ sélast wisse &
gold-hrodene cwén \hld\ giefe bryttian.\eva

\bvb And then Elhild gave me another, \\
the noble queen of the veterans, daughter of Edwin. \\
Her praise stretched further through a multitude of lands; \\
then I in song should say, \\
where beneath the heaven I know the most blessed \\
gold-adorned queen dispensing gifts.\evb\evg


\bvg\bva Þǫnne wit Scilling \hld\ scíran reorde &
for uncrum sige-dryhtne \hld\ sǫng a·hófan, &
hlúde bí hearpan, \hld\ hleoþor swinsade, &
þǫnne mǫnige męnn, \hld\ módum wlǫnce, &
wordum sprécan, \hld\ þá þe wel cu̇þan, &
þæt hí næ̂fre sǫng \hld\ séllan ne hýrdon.\eva

\bvb Then I and Shilling with clear voices, \\
before our victorious lord raised up a song,
loudly by the harp—the tune rang out. \\
Then many men proud of heart \\
told with words—those who knew well— \\
that they never had heard a better song.\evb\evg

\sectionline

\bvg\bva Ðǫnan ic ealne geond·hwearf \hld\ ǿþel Gotena, &
sóhte ic â síþa \hld\ þá sélestan; &
þæt wæs inn-weorud \hld\ Earman-rices.\eva

\bvb Then I passed through all the ethel of the Gots; \\
TODO.\evb\evg


\bvg\bva Heðcan sóhte ic ǫnd Beadecan \hld\ ǫnd Hęre-lingas, &
Emercan sóhte ic ǫnd Fridlan \hld\ ǫnd Éast-gotan, &
fródne ǫnd gódne \hld\ fæder Un-wenes.\eva

\bvb TODO\evb\evg


\bvg\bva Seccan sóhte ic ǫnd Beccan, \hld\ Seafolan ǫnd Þeód-ríc, &
Heaþo-ríc ǫnd Sifecan, \hld\ Hliþe ǫnd Incgen-þeow. &
Éad-wine sóhte ic ǫnd Elsan, \hld\ Ægel-mund ǫnd Hún-gâr, &
ǫnd þá wlǫncan ge·dryht \hld\ Wiþ-myrginga.\eva

\bvb TODO\evb\evg


\bvg\bva Wulf-hęre sóhte ic ǫnd Wyrm-hęre; \hld\ ful oft þær wíg ne a·læg, &
þǫnne Hræda hęre \hld\ heardum sweordum &
ymb Wistla-wudu \hld\ węrgan sceoldon &
ealdne ǿþel-stól \hld\ Ætlan leódum.\eva

\bvb I sought out Wolfer and Wyrmer—very seldom did the warring there stop, \\
when the Reth-army, with hard swords, \\
in the Wistlewood had to defend \\
the old homeland-seat against Attle’s people.\evb\evg


\bvg\bva Rǽd-hęre sóhte ic ǫnd Rǫnd-hęre, \hld\ Rúm-stân ǫnd Gisl-hęre, &
Wiþer-gield ǫnd Freoþe-ric, \hld\ Wudgan ǫnd Hâman; &
ne wǽran þæt ge·síþa \hld\ þá sǽmestan, &
þéah þe ic hý a·níhst \hld\ nęmnan sceolde.\eva

\bvb TODO.\evb\evg


\bvg\bva Ful oft of þâm héape \hld\ hwínende fléag &
\edtrans{giellende gâr}{a yelling spear}{\Bfootnote{Formulaic.}} \hld\ ǫn grǫme þeóde; &
wræccan þær weoldan \hld\ wundnan golde &
werum ǫnd wífum, \hld\ Wudga ǫnd Hâma.\eva

\bvb Most often from that troop whistling did fly \\
a yelling spear into the fiendish host; \\
there ruled the exiles Woody and Homer \\
twisted gold, men and women.\evb\evg


\bvg\bva Swá ic þæt symle ǫn·fǫnd \hld\ ǫn þæ̂re fęringe, &
þæt sé biþ leófast \hld\ lǫnd-búendum &
sé þe him God syleð \hld\ gumena ríce &
to ge·healdenne, \hld\ þenden hé hér leofað.“\eva

\bvb So I always did find while on that journey, \\
that he is dearest to land-dwellers \ken{men}, \\
whom God grants the realm of men \\
for to hold while here he lives.”\evb\evg

\sectionline

\bvg\bva Swá \alst{sc}ríþende \hld\ ge·\alst{sc}eapum hweorfað &
\alst{g}leó-męnn \alst{g}umena \hld\ geond \alst{g}runda fela, &
\alst{þ}earfe sęcgað, \hld\ \alst{þ}ǫnc-word sprecaþ, &
\alst{s}imle \alst{s}u̇ð oþþe norð \hld\ \alst{s}umne ge·mǿtað &
\alst{g}ydda \alst{g}leawne, \hld\ \alst{g}eofum un·hneawne, &
sé þe fore \alst{d}uguþe wile \hld\ \alst{d}óm a·ræ̂ran, &
\alst{eo}rl-scipe \alst{æ}fnan, \hld\ oþþæt \alst{ea}l scæceð, &
\alst{l}eoht ǫnd \alst{l}if sǫmod; \hld\ \alst{l}of sé ge·wyrceð, &
\alst{h}afað under \alst{h}eofonum \hld\ \alst{h}éah-fæstne dóm.\eva

\bvb So passing through fates they wander, \\
the song-men of mankind, through many lands; \\
they say their needs, speak thoughtful words; \\
whether in the south or north they meet some one, \\
gay in songs, unstingy with gifts, \\
who for the old troop will rear up \inx[C]{doom}, \\
accomplish earlship until all goes away, \\
light and life together.  He who works praise \\
has under the heavens a high, firm doom.\evb\evg

\sectionline
