Poems

WSp. — Spae of the Wale (\emph{Vǫluspǫ́})
High. — Speeches of the High One (\emph{Brot af Sigurðarkviða})
Web. — Speeches of Webthrithen (\emph{Brot af Sigurðarkviða})
Reed. — From the Sons of king Reeding (\emph{Frá sonum Hrauðungs konungs})
Grim. — Speeches of Grimen (\emph{Grímnismǫ́l})
Shea. — Speeches of Shearen (\emph{Skírnismǫ́l})
Hoar. — Leed of Hoarbeard (\emph{Hárbarðsljóð})
Hyme. — Lay of Hyme (\emph{Hymiskviða})

Way. — Lay of Wayland
Alw. — Speeches of Allwise (\emph{Alvíssmǫ́l})
I HHb. — First Lay of Hallow Hundingsbane
HHw. — Lay of Hallow Herwardsson
II HHb. — Second Lay of Hallow Hundingsbane

SiL. — Fragment of the Lay of Siward (\emph{Brot af Sigurðarkviða})
SDe. — From the Death of Siward. (\emph{Frá dauða Sigurðar})
I Gr. — First Lay of Guthrun. (\emph{Guðrúnarkviða I})
ShS. — Short Lay of Siward (\emph{Sigurðarkviða in skamma})
BHel. — Hellride of Byrnhild (\emph{Helreið Brynhildar})
Nifl. — The Killing of the Niflings (\emph{Dráp Niflunga})
II Gr. — Second Lay of Guthrun. (\emph{Guðrúnarkviða II})
III Gr. — Third Lay of Guthrun. (\emph{Guðrúnarkviða III})
BnOr. — From Burgny and Ordrun (\emph{Frá Borgnýju ok Oddrúnu})
OdW. — Weeping of Ordrun (\emph{Oddrúnargrátr})
EtD. — Death of Ettle (\emph{Dauði Atla})
EtL. — Lay of Ettle (\emph{Atlakviða})
EtS. — Speeches of Ettle (\emph{Atlamǫ́l})
FGr. — From Guthrun. (\emph{Frá Guðrúnu})
GrB. — Boldness of Guthrun (\emph{Guðrúnarhvǫt})
Ham. — Speeches of Hamthew (\emph{Hamðismǫ́l}).

Gods and men

\textbf{Ettle} (ON. \emph{Atli}, OE. \emph{Ætla}, MHG. \emph{Etzel}) — \emph{Attila}, ruler of the \emph{Huns} from 434—453.
 I HHb. 54, SiL. 11, I Gr. 23, ShS. 28, 29, 33, 37, 54, 56, 57, II Gr. 26, 38, 45, III Gr. 1, 9, BnOr. 0, OdW. A, 2, 22, 23, 25, 26, 30, 31, EtD. 0, EtL. 1, 3, 15, 17, 18, 27, 31, 32, 34, 36, 37, 38, 41, 43, B, EtS. 2, 4, 21, 22, 44, 52, 60, 64, 71, 73, 77, 80, 86, 87, 97, 98, 108, 113, 117, FGr. 0, GrB. 12, Ham. 6

Tribes, lands and clans

\textbf{Huns} (ON. \emph{Húnir}, OE. \emph{Hūne}, OHG. \emph{Hūni}, \emph{Hunni}) — TO-DO.
