\bookStart{Dreams of Balder}[Baldrs draumar]
\setBookCode{Baldrsdraumar}

% Introduction.
\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: C10th (0.890)

\textbf{Meter:} \Fornyrdislag%
\end{flushright}

\section{Introduction}

The \textbf{Dreams of Balder} (abbrev.~\Baldrsdraumar) is a short Eddic poem that sees Weden ride down to Hell, the Underworld, to ask a dead wallow about Balder’s future.

\subsection{Textual preservation and transmission}

\Baldrsdraumar\ is preserved only in \AM\ on foll.~1v/17--2r/11.
A younger redaction characterised by a number of post-mediæval inserts is transmitted in several copies in later paper mss.  Metrical characteristics make it clear that these inserts are very late \parencite{Adalsteinsson2009}, and they are therefore excluded from this edition without further comment.

\subsection{Meter and stanza}

\Baldrsdraumar\ is a short poem consisting of 57 long-lines across 14 stanzas.  The meter is \Fornyrdislag.  The typical length of the stanza is 4 (13/14: 1--10, 12--14), and this is upheld consistently.  A single stanza, 11, has 5 lines.

\subsection{Origin and motifs}

The main source for the death and subsequent avenging of the god Balder is \Gylfaginning~49--50 (for a summary of which see \textlink{Voluspa}[31] n.).  Compared to that narrative, \Baldrsdraumar\ corresponds only to the very beginning, namely Balder’s ominous dreams, and does not go into great depth about its subject; at just 14 sts.~the poem is the shortest surviving mythological Eddic poem.  In isolation, the purpose of \Baldrsdraumar\ seems somewhat unclear, and it must originally have been part of a longer cycle of lost poetry dealing with Balder’s death.  That such a cycle likely existed is supported by the fragmentary stanza \textlink{EddicFragments}[F5][V].

\subsection{Dating}

\Baldrsdraumar\ has a particularly high rate of the archaic particle \emph{of}, with a frequency of 0.463 occurrences per 10 lines \parencite[77]{Sapp2022}, although it must be noted that \Baldrsdraumar\ is so short that this number may not be as reliable as for longer poems.  St.~3/4a may require a hiatus form, suggesting a date of composition before ca.~1150 \ce, but is an uncertain attestation (see note).  The alliteration of \emph{vr-} with \emph{v-} in 11/1 suggests a date before ca.~1000 \ce\ \parencite{Haukur2016}, but unfortunately that line is partly emended, and the presence of \emph{vr-} in \emph{*Vrindr} (> \emph{Rindr}) cannot be considered fully certain due to a lack of parallels.  To conclude, \Baldrsdraumar\ is likely an old poem, but it is too short to contain clear diagnostics.

\subsection{Summary}

The poem begins \emph{in medias res}: \inx[P]{Balder} has been having nightmares, which the Gods meet at their Thing to discuss (1).  \inx[P]{Weden} rides down to \inx[P]{Hell} where he has an encounter with a bloody hound; he passes it and continues to “the high house of \inx[P]{Hell}” (2--3) from which he rides west to the grave of a certain \inx[C]{wallow} whom he revives using magic (4).

The wallow asks who it is that has forced her out of the grave (5) and Weden introduces himself as Waytame, before asking for whom the benches of Hell are covered with gold (6).  The wallow responds that barrels of mead stand brewed for Balder and that the Gods must be very anxious (7).  Weden asks her who will slay Balder (8), and she responds that it is Hath, carrying a “tall, famous wood” (9).  The god next asks who will avenge Balder’s death (10), and the wallow responds that \inx[P]{Wrind} will give birth to Weden’s son \inx[P]{Wonnel}, who will slay Hath when only one night old (11).  Weden then asks about some mysterious maidens (12), which apparently betrays his identity.  The wallow announces that she now knows that her accoster is Weden, not Waytame; the god in turn retorts that she is not a wallow but rather the “mother of three thurses” (13).  The wallow tells him to ride home and “be renowned”, taunting him over his unavoidable death at the \inx[P]{Rakes of the Reins} (14).

\newpage

\section{Text}

\subsection{The Dreams of Balder}

\bvg\bva\mssnote{\AM~1v/18}%
\edtext{Sęnn vǫ́ru \alst{ę̇}sir \hld\ \alst{a}llir ȧ þingi &
ok \alst{ǫ̇}synjur \hld\ \alst{a}llar ȧ máli, &
ok umb þat \alst{r}éðu \hld\ \alst{r}íkir tívar:}{\lemma{Sęnn \dots\ tívar ‘Soon \dots\ Tews’}\Bfootnote{Identically shared with \textlink{Thrymskvida}[14]/1--3. ---

The ‘Thing’ is the Thing of the Gods (cf.~\textlink{Voluspa}[6]/1--2 n., \textlink{Hymiskvida}[39]/1), but the detail that the male Gods (\emph{ę̇sir} ‘Eese’) are \emph{ȧ þingi} ‘at, upon the Thing’, while the Goddesses (\emph{ǫ̇synjur} ‘Ossens’) are on the sidelines chatting requires elaboration.
Being \emph{ȧ þingi} is probably to be understood as being seated within the ritual enclosure at the legal assembly.  This reflects historical Norse Thing-proceedings where the judgments were decided by a select group of twelve men (sometimes several groups of twelve, a number still surviving in the twelve-man jury of English common law) seated inside a circular holy enclosure (the \emph{vé-bǫnd} ‘wigh-bonds’, i.e., ‘holy cords’) into which all others were forbidden to enter.  So \EgilsSaga\ 56 in a description of the Gole-Thing in Norway: \emph{En þar er dómr’inn var settr, var vǫllr sléttr ok settar niðr hesli-stengr í vǫll’inn í hring, en lǫgð um útan snǿri um·hverfis.  Vǫ́ru þat kǫlluð vé-bǫnd.  En fyrir innan í hringi’num sǫ́tu dómendr [...]; þę́r þrennar tylftir manna skyldu þar dǿma um mál manna.} ‘But where the court was established there was a smooth plain and hazelwood-poles were set down into the plain in a ring, and cords tied around them on the outside; those were called the ‘wigh-bonds’.  But inside within the ring sat the judges [...]; those three groups of twelve men each would there judge the cases of men.’  The Norwegian New Law of the Land, I. 3 \parenciteshorthand[14]{NGL2} specifies the proceedings of the Gole-Thing and notes that: \emph{Svá er ok mę́lt at engi maðr þeirra er eigi er nefndr skal firir innan vé-bǫnd setjask útan or·lofs.} ‘So it is also said that no man among them who was not named shall be seated inside the wigh-bonds without permission.’}} &
hví vę́ri \edtext{\alst{B}aldri}{\Afootnote{corr.~from \emph{baldrs} above line in same hand \AM}} \hld\ \alst{b}allir draumar?\eva

\bvb%
{\huge S}\textsc{oon were the \inx[P]{Eese}} all at the \inx[C]{Thing}, \\
and the \inx[P]{Ossens} all at speech, \\
and of this counseled the mighty \inx[P]{Tews}: \\
Why did Balder have troubling dreams?\evb\evg


\bvg\bva\mssnote{\AM~1v/19}%
\alst{U}pp ręis \alst{Ó}ðinn, \hld\ \edtext{\alst{a}ld\emph{inn}}{\Afootnote{emend.~(cf.~13/2 below); \emph{alda} \AM}} Gautr, &
\edtrans{ok hann ȧ \alst{S}lęipni \hld\ \alst{s}ǫðul of lagði}{and he on Slopner the saddle did lay}{\Bfootnote{Possibly formulaic; cf.~\textlink{Oddrunargratr}[2]/4.}}, &
ręið \alst{n}iðr þaðan \hld\ \alst{n}ifl-hęljar til; &
mǿtti \edtrans{\alst{h}velpi, \hld\ þęim’s ór \alst{h}ęlju kom}{the whelp that came out of Hell}{\Bfootnote{An otherwise unknown dog, sometimes identified with \inx[P]{Garm}.  The motif of the Hellhound guarding the underworld is well known throughout world mythology, most famously in the form of the Greek \textgreek{Κέρβερος} (\Theogony\ 300–312).}}.\eva

\bvb Up rose Weden, the olden Geat, \\
and he on \inx[P]{Slopner} the saddle did lay, \\
rode down thence to \inx[P]{Nivelhell}, \\
met the whelp that came out of Hell.\evb\evg


\bvg\bva\mssnote{\AM~1v/21}%
Sá vas \alst{b}lóðugr \hld\ umb \alst{b}rjóst framan, &
ok \edtrans{\alst{g}aldrs fǫður}{the father of \inx[C]{galder} \ken*{= Weden}}{\Bfootnote{An unparalleled expression for Weden, the master of magical spells and chants (galders).  Cf.~\textlink{Havamal}[140], 147--165.}} \hld\ \edtext{\alst{g}ól \emph{of}}{\Afootnote{\emph{‘golv’} (corruption of \emph{*gol ū}) \AM}} lęngi, &
\edtrans{\alst{f}ramm ręið Óðinn, \hld\ \alst{f}old-vegr dunði}{Forth rode Weden---the fold-way \ken{earth} resounded}{\Bfootnote{The movement of gods often causes cosmic disturbance; cf.~the description of \inx[P]{Thunder}’s riding in \Haustlong\ 14: \emph{Ók at ísarn-lęiki \hld\ Jarðar sunr ok dunði \dots\ mȧna vegr und hǫ̇num} ‘The Son of Earth drove to the iron-play \ken{battle} \dots\ and the moon’s way \ken{sky/heaven} resounded beneath him’.  See further \textlink{Thrymskvida}[21].}}, &
\edtext{\alst{h}ann kom at \alst{h}ǫ́u}{\Bfootnote{The hiatus form in \emph{hǫ́u} ‘high’ (\emph{‘háfu’} \AM) is not diagnostic for the dating since it may be due to analogy with other forms in \emph{hǫ́v-} in the adjective, that is, it may an old uncontracted \emph{hǫ́u} (which would be contracted to \emph{hǫ́} post-1150), but it may also be a young analogical \emph{hǫ́vu}.  Regardless of the presence of the analogical glide in \emph{hǫ́(v)u}, the line may be read either as a type A1 (with stress on \emph{hann}) or a type A3.}} \hld\ \alst{H}ęljar ranni.\eva

\bvb It was bloody at the front of its chest \\
and at the father of \inx[C]{galder} \ken*{= Weden} for a long time bayed. \\
Forth rode Weden---the fold-way \ken{earth} resounded--- \\
he came to the high house of Hell.\evb\evg


\bvg\bva\mssnote{\AM~1v/22}%
Þȧ ręið \alst{Ó}ðinn \hld\ fyr \alst{au}stan dyrr, &
þar’s hann \alst{v}issi \hld\ \alst{v}ǫlu lęiði; &
nam hann \alst{v}ittugri \hld\ \edtrans{\alst{v}al-galdr}{slain-galder}{\Bfootnote{A galder to quicken the dead, in this case the wallow.  Cf.~\textlink{Havamal}[158] where Weden tells how He can bring hanged men back to life with a galder.}} kveða, &
und’s \alst{n}auðug ręis, \hld\ \alst{n}ás orð of kvað:\eva

\bvb Then Weden rode east from its doors, \\
whither he knew the wallow’s grave. \\
He took to sing a slain-\inx[C]{galder} for the cunning woman \\
until forced she rose; a corpse’s words quoth she:\evb\evg


\bvg\bva\mssnote{\AM~1v/24}%
\llap{„}Hvat ’s \alst{m}anna þat \hld\ \alst{m}ér ȯ·kunnra, &
es mér hęfr \alst{au}kit \hld\ \edtrans{\alst{ę}rfitt sinni}{this toilsome journey}{\Bfootnote{The “journey” out of the grave.}}? &
\edtext{Vas’k \alst{s}nifin \alst{s}njóvi, \hld\ ok \alst{s}lęgin regni, &
ok \alst{d}rifin \alst{d}ǫggu, \hld\ \alst{d}auð vas’k lęngi.}{\lemma{Vas’k snifin \dots\ lęngi. ‘I was snowed \dots\ long.’}\Bfootnote{Cf.~the similar description of a buried person in \textlink{HelgakvidaTwo}[44]--45.}}“\eva

\bvb “What sort of man is this, unknown to me, \\
who has caused for me this toilsome journey? \\
I was snowed by snow and struck by rain \\
and bespattered by dew---dead was I for long.”\evb\evg


\bvg\bva\mssnote{\AM~1v/25}%
\speakernote{[Óðinn kvað:]}%
\llap{„}\edtrans{\alst{V}eg·tamr ek hęiti, \hld\ sonr em’k \alst{V}al·tams}{Waytame I am called, I am Waltame’s son}{\Bfootnote{Neither name is attested elsewhere, but both are Odinic in character, the first alluding to Weden’s role as protector of wayfarers (\textlink{Havamal}, \textlink{Grimnismal}), the second to his role as gatherer of dead warriors (\textlink{Voluspa}[23]/1, \textlink{Vafthrudnismal}[40]--41, \textlink{Grimnismal}[8]--10, \textlink{Harbardsljod}[24]).}}, &
sęg mér ór \alst{h}ęlju, \hld\ ek man ór \alst{h}ęimi; &
hvęim eru \alst{b}ękkir \hld\ \alst{b}augum sánir, &
\alst{f}lęt \alst{f}agr⸗liga \hld\ \alst{f}lóuð golli?“\eva

\bvb\speakernoteb{[Quoth Weden:]}%
“Waytame I am called; I am Waltame’s son; \\
tell me [the tidings] from Hell---I will [tell those] from the Home. \\
For whom are the benches sown with \inx[C]{bigh}[bighs], \\
the floors fairly flooded with gold?”\evb\evg


\bvg\bva\mssnote{\AM~1v/27}%
\speakernote{[Vǫlva kvað:]}%
\llap{„}Hér stęndr \alst{B}aldri \hld\ of \alst{b}rugginn mjǫðr, &
\alst{sk}írar vęigar, \hld\ liggr \alst{sk}jǫldr yfir, &
ęn \alst{ǫ̇}s-męgir \hld\ ï \alst{o}f·vę̇ni; &
\alst{n}auðug sagða’k, \hld\ \alst{n}ú mun’k þęgja.“\eva

\bvb\speakernoteb{[Quoth the wallow:]}%
“Here for Balder mead stands brewed, \\
pure draughts---a shield lies over them--- \\
but the os-lads \ken*{= Eese} stand in great suspense--- \\
forced I spoke, now I will shut up!”\evb\evg


\bvg\bva\mssnote{\AM~1v/29}%
\speakernote{[Óðinn kvað:]}%
\llap{„}\alst{Þ}ęgj⸗at-tu vǫlva, \hld\ \alst{þ}ik vil’k fregna, &
\alst{u}nd’s \alst{a}l-kunna, \hld\ vil’k \alst{ę}nn vita: &
hvęrr man \alst{B}aldri \hld\ at \alst{b}ana verða, &
ok \alst{Ó}ðins son \hld\ \alst{a}ldri rę̇na?“\eva

\bvb\speakernoteb{[Quoth Weden:]}%
“Shut not up, wallow---thee I wish to ask! \\
Until all is known I wish yet to know: \\
Who will become Balder’s bane \\
and rob Weden’s son of life?”\evb\evg


\bvg\bva\mssnote{\AM~2r/1}%
\speakernote{[Vǫlva kvað:]}%
\llap{„}\alst{H}ǫðr berr \alst{h}ǫ́van \hld\ \edtext{\alst{h}róðr-ba\emph{ð}m}{\Afootnote{emend.; \emph{hróðr-barm} \AM}} þinig, &
hann man \alst{B}aldri \hld\ at \alst{b}ana verða, &
ok \alst{Ó}ðins son \hld\ \alst{a}ldri rę̇na; &
\alst{n}auðug sagða’k, \hld\ \alst{n}ú mun’k þęgja.“\eva

\bvb\speakernoteb{[Quoth the wallow:]}%
“\inx[P]{Hath} bears the tall, famous wood \ken{mistletoe} thither; \\
he will become Balder’s bane \\
and rob Weden’s son of life--- \\
forced I spoke, now I will shut up!”\evb\evg


\bvg\bva\mssnote{\AM~2r/3}%
\speakernote{[Óðinn kvað:]}%
\llap{„}\alst{Þ}ęgj⸗at-tu vǫlva, \hld\ \alst{þ}ik vil’k fregna, &
\alst{u}nd’s \alst{a}l-kunna, \hld\ vil’k \alst{ę}nn vita, &
hvęrr man \alst{h}ęipt \alst{H}ęði \hld\ \alst{h}ęfnt of vinna, &
eða \alst{B}aldrs \alst{b}ana \hld\ ȧ \alst{b}ál vega?“\eva

\bvb\speakernoteb{[Quoth Weden:]}%
“Shut not up, wallow---thee I wish to ask! \\
Until all is known I wish yet to know: \\
Who will avenge that evil on Hath, \\
or drag onto the pyre Balder’s bane?”\evb\evg


\bvg\bva\mssnote{\AM~2r/4}%
\speakernote{[Vǫlva kvað:]}%
\llap{„}\edtrans{\emph{\alst{V}}rindr}{Wrind}{\Afootnote{metr. emend.; \emph{Rindr} \AM}\Bfootnote{As the first nominal in the verse, ms. \emph{Rindr} is expected to carry alliteration with \emph{vestr-sǫlum}.  This is obtained by restoring an archaic initial \emph{vr-} cluster, for which cf.~\textlink{Havamal}[26]/2~n.  Unlike other instances of this cluster, \emph{*Vrindr} is etymologically obscure.  A likely theory connects the name with Gutnish \emph{rind} m.~‘ivy, common clubmoss’, in which case Wrind may be a minor plant goddess \parencite{Lundberg1913}.  If so, there is prob.~an obscure ritual significance to the notion that it is the son of the Ivy who is born to avenge the harm done by the mistletoe.
\citeauthor{Lundberg1913} argues that worship of Wrind is indicated by the OSwe.~place name \emph{Vrinda-vi} (modern-day Vrinnevid, supposedly gen.~sg.~of \emph{*Vrindr} + \emph{ví} ‘wigh, shrine’), although \textcite[66, 78--80]{Sahlgren1924} considers this etymology false and instead explains the name as containing the second element OSwe.~\emph{viði} ‘wood, forest’.}} berr \edtrans{\emph{\alst{V}ȧla}}{Wonnel}{\Afootnote{required by meter and sense; om. \AM}\Bfootnote{The name must be restored.  Without it the verse is only two syllables long, which is absolutely forbidden in Norse \Fornyrdislag\ meter, and the verb \emph{berr} ‘bears, gives birth to’ is left without an object.}} \hld\ ï \alst{v}estr-sǫlum, &
\edtext{sá man \alst{Ó}ðins sonr \hld\ \alst{ęi}n-nę́ttr vega; &
\alst{h}ǫnd of þvę́r-\edtext{\emph{at}}{\Afootnote{required by meter and sense; om. \AM}} \hld\ né \alst{h}ǫfuð kęmbir, &
áðr ȧ \alst{b}ál of \alst{b}err \hld\ \alst{B}aldrs and·skota;}{\lemma{sá \dots\ and-skota ‘he will \dots\ shooter’}\Bfootnote{These lines are, apart from the tense of the verbs, almost identical to \textlink{Voluspa}[32]/4--33/2.  The direction of influence is unclear, and it is not impossible that both are borrowing from a now-lost third poem.}} &
\alst{n}auðug sagða’k, \hld\ \alst{n}ú mun’k þęgja.“\eva

\bvb\speakernoteb{[Quoth the wallow:]}%
“Wrind bears \inx[P]{Wonnel} in the western halls: \\
he will, Weden’s son, one night old, fight. \\
He washes not his hand nor combs his head \\
before onto the pyre he bears Balder’s shooter--- \\
forced I spoke, now I will shut up.”\evb\evg


\bvg\bva\mssnote{\AM~2r/6}%
\speakernote{[Óðinn kvað:]}%
\Ballnote{The question in this stanza is quite obscure.  If it is connected to the myth of Balder’s death in any way, it ought to be the episode where the all things in the world wept for Balder.
According to \Gylfaginning~49:70--82 (see \textlink{EddicFragments}[F5][V] below), Hell promised to give Balder back to the Eese if all things in the world would weep for him.  The Eese relayed this message, and all things wept for Balder, save Lock in disguise.
It may thus be that these maidens were among the weepers (perhaps they were the Walkirries, and this is what reveals Weden’s identity?), but in any case their identity is otherwise unknown.  They may tentatively be identified with the equally mysterious daughters of Maythrasher in \textlink{Vafthrudnismal}[49], who are also young women (\emph{męyjar}) who appear to have something to do with the End Times.}%
\llap{„}\alst{Þ}ęgj⸗at-tu vǫlva, \hld\ \alst{þ}ik vil’k fregna, &
\alst{u}nd’s \alst{a}l-kunna, \hld\ vil’k \alst{ę}nn vita, &
hvęrjar ’ro \alst{m}ęyjar, \hld\ es at \alst{m}uni gráta &
ok \edtrans{ȧ \alst{h}imin verpa \hld\ \alst{h}alsa-skautum}{onto heaven throw the front-sheets}{\Bfootnote{Unexplained.  Perhaps an astrological allusion.}}?“\eva

\bvb\speakernoteb{[Quoth Weden:]}%
“Shut not up, wallow---thee I wish to ask! \\
Until all is known I wish yet to know: \\
Which are the maidens that heartily weep, \\
and onto heaven throw the front-sheets?”\evb\evg


\bvg\bva\mssnote{\AM~2r/8}%
\speakernote{[Vǫlva kvað:]}%
\llap{„}\alst{E}st⸗at Veg·tamr, \hld\ sęm \alst{e}k hugða, &
hęldr est \alst{Ó}ðinn, \hld\ \edtext{\alst{a}ldinn}{\Afootnote{corr.~from \emph{alda} above line in same hand \AM}} Gautr!“ &
\speakernote{[Óðinn kvað:]}%
\llap{„}Est⸗at \alst{v}ǫlva \hld\ né \alst{v}ís kona, &
hęldr est \alst{þ}riggja \hld\ \alst{þ}ursa móðir!“\eva

\bvb\speakernoteb{[Quoth the wallow:]}%
“Thou art not Waytame as I thought, \\
rather art thou Weden, the olden Geat!”--- \\
\speakernoteb{[Quoth Weden:]}%
“Thou art no \inx[C]{wallow} nor wise woman, \\
rather art thou three \inx[P]{Thurses}’ mother!”\evb\evg


\bvg\bva\mssnote{\AM~2r/9}%
\speakernote{[Vǫlva kvað:]}%
\llap{„}\alst{H}ęim ríð Óðinn \hld\ \edtrans{ok ves \alst{h}róðigr}{and be renowned}{\Bfootnote{A sarcastic taunt, the sense being: “Your fame, Weden, will not save you!”}}, &
svá komi⸗t \alst{m}anna \hld\ \alst{m}ęirr aptr ȧ vit, &
es \alst{l}auss \alst{L}oki \hld\ \alst{l}íðr ór bǫndum &
ok \alst{r}agna \alst{r}ǫk \hld\ \edtrans{\alst{r}júfęndr}{rippers}{\Bfootnote{Presumably Surt and Lock with his children, as described in \textlink{Voluspa}[39]--40, 48--50.
The verb \emph{rjúfa} ‘to break, rip up, break a hole in’ (\CV) is used in the same context in the formulaic \emph{(of) rjúfa⸗sk ręgin} ‘the \inx[P]{Reins} are ripped’ (\textlink{Vafthrudnismal}[52]/4, \textlink{Grimnismal}[4]/4, \textlink{Lokasenna}[41]/2, \textlink{Sigrdrifumal}[17]/7).}} koma.“\eva
%NOTE: If printing mythological Eddic poems separately.  One may also compare the similar sounding (but not or only very distantly related) verb \emph{rifna} ‘be riven, rent apart’ used in reference to the destruction of the world in Runic inscription Sö 154 (Skarpåker), and Arn \emph{Hryn} (in \Skp\ 2 pp. 185--6, ll. 3/7--8, see also note there): \emph{meiri verði þinn an þeira \hld\ þrifnuðr allr, und’s himinn rifnar.} ‘greater than theirs may thy whole wealth be, until heaven is riven.’}} koma.“

\bvb\speakernoteb{[Quoth the wallow:]}%
“Ride home, Weden, and be renowned! \\
So may no man come again to visit \\
when, loose, Lock slips out of his bonds, \\
and at the \inx[P]{Rakes of the Reins} the rippers come.”\evb\evg

\sectionline
