\bookStart{Leeds of Hindle}[Hyndlu ljóð]
\setBookCode{Hyndluljod}

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: late C11th (0.996)

\textbf{Meter:} \Fornyrdislag%
\end{flushright}%

\section{Introduction}

The \textbf{Leeds of Hindle} (\textlink{Hyndluljod}) is a poorly preserved poem found only in \FlatMS.  St.~32 is cited in \Gylfaginning\ and attributed to the \emph{Vǫluspǫ́ in skamma} ‘Short Spae of the Wallow’.  The relation of that poem to the \Hyndluljod\ in its present form is unclear.

NOTE: Both translation and edition are currently incomplete.

\newpage

\section{Text}

\bvg\bva%
\speakernote{[Fręyja kvað:]}%
\llap{„}Vaki \alst{m}ę́r \alst{m}ęyja, \hld\ vaki \alst{m}ïn vina, &
\alst{H}yndla systir, \hld\ es ï \alst{h}ęlli býr; &
nú ’s \alst{r}økr \alst{r}økra, \hld\ \alst{r}íða vit skulum &
til \alst{V}al-hallar \hld\ ok til \alst{v}és hęilags.\eva

\bvb\speakernoteb{[Frow quoth:]}%
“{\huge W}\textsc{ake, maiden of maidens!} Wake, my friend! \\
O Hindle, sister, who livest in the cave! \\
Now is the twilight of twilights; we two shall ride \\
to Walhall and to the holy \inx[C]{wigh}!\evb\evg


\bvg\bva%
Biðjum \edtext{\alst{H}ęrja*·fǫðr}{\Afootnote{emend.; \emph{‘herians faudur’} \FlatMS}} \hld\ ï \alst{h}ugum sitja, &
\edtrans{hann \alst{g}eldr ok \alst{g}efr \hld\ \alst{g}oll verðugum}{he pays and gives gold to the worthy}{\Bfootnote{Closely related to \textlink{HelgakvidaOne}[9]/3, which is why \textcite{FinnurEdda} and \textcite{GudniEdda} emend \emph{verðugum} ‘the worthy’ to \emph{verðungu} ‘the retinue’.}}, &
gaf hann \alst{H}ęr·móði \hld\ \alst{h}jalm ok brynju, &
ęn \alst{S}ig·mundi \hld\ \alst{s}verð at þiggja.\eva

\bvb Let us bid the Father of Hosts \name{= Weden} to remain in good thoughts;  \\
he pays and gives gold to the worthy.  \\
He gave to \inx[P]{Harmod} helmet and byrnie, \\
but to \inx[P]{Syemund} a sword to take.\evb\evg


\bvg\bva%
Gefr hann \alst{s}igr \alst{s}onum, \hld\ ęn \alst{s}vinnum aura, &
\alst{m}ę́lsku \alst{m}ǫrgum \hld\ ok \alst{m}an-vit firum, &
\alst{b}yri gefr \alst{b}rǫgnum, \hld\ ęn \alst{b}rag skǫldum, &
gefr hann \alst{m}ann-sęmi \hld\ \alst{m}ǫrgum rekki.\eva

\bvb He gives victory to sons and ounces [of silver] to the wise, \\
speech to many and \inx[C]{manwit} to men. \\
Fair wind he gives to nobles and praise-song to \inx[C]{scald}[scalds]; \\
he gives manly valour to many a champion.\evb\evg


\bvg\bva%
\alst{Þ}ór mun hǫ́n blóta, \hld\ \alst{þ}ess mun hǫ́n biðja, &
at hann \alst{ę́} við þik \hld\ \alst{ęi}n-art láti; &
þó ’s hǫ̇num \alst{ȯ}·títt \hld\ við \alst{jǫ}tuns brúðir.\eva

\bvb To Thunder she [I] will \inx[C]{bloot}, of this she [I] will pray: \\
that he always be upright with thee, \\
although he bears hatred for the ettin’s brides.\evb\evg


\bvg\bva%
Nú tak-tu \alst{u}lf þïnn \hld\ \alst{ęi}nn af stalli, &
lát hann \alst{r}inna \hld\ með \alst{r}una mínum.“--- &
\speakernote{[Hyndla kvað:]}
\llap{„}Sęinn es \alst{g}ǫltr þinn \hld\ \alst{g}oð-veg troða, &
vil’k⸗\emph{at} \alst{m}ar \alst{m}ïnn \hld\ \alst{m}ę́tan hlǿða.\eva

\bvb Now take thy one wolf from the stable; \\
let him run alongside my boar.”--- \\
\speakernoteb{[Hindle quoth:]}%
“[Too] slow is thy boar to tread the Godways; \\
I wish not to load my noble steed.\evb\evg


\bvg\bva%
\alst{F}lǫ́ est \alst{F}ręyja, \hld\ es \alst{f}ręistar mín, &
\edtext{vísar þú \alst{au}gum \hld\ ȧ \alst{o}ss þannig, &
es hafir \alst{v}er þinn \hld\ ï \alst{v}al-sinni}{\lemma{vísar \dots\ val-sinni ‘thou showest \dots\ slain-ways’}\Bfootnote{I.e., “You only show favour to me because you want me to help your lover”.  For the expression cf.~\textlink{Sigrdrifumal}[3]/3 n.}} &
\alst{Ó}tt·ar \alst{u}nga \hld\ \alst{I}nn·stęins bur.“\eva

\bvb False art thou, Frow, who temptest me; \\
thou showest thy eyes on us this way \\
since thou hast thy lover on the slain-path: \\
the young Oughter, Instone's offspring.”\evb\evg


\bvg\bva%
\speakernote{[Fręyja kvað:]}%
\llap{„}\alst{D}ulið est Hyndla, \hld\ \alst{d}raums ę́tla’k þér, &
es kveðr \alst{v}er minn \hld\ ï \alst{v}al-sinni, &
þar’s \alst{g}ǫltr \alst{g}lóar \hld\ \alst{G}ollin·bursti, &
\edtrans{\alst{H}ildi·svíni}{Hildswine}{\Bfootnote{The ‘battle-swine’, another name for Goldenbristle.}}, \hld\ es mér \alst{h}agir gęrðu, &
\alst{d}vergar tvęir \hld\ \alst{D}áinn ok Nabbi.\eva

\bvb\speakernoteb{[Frow quoth:]}%
Deluded art thou, Hindle; I think thee dreamy, \\
who sayest that my man is on the slain-path \\
where the boar Goldenbristle glows, \\
Hildswine, which for me made \\
the two skilful dwarfs Dowen and Nab.\evb\evg


\bvg\bva%
\edtext{\alst{S}ęnn*}{\Afootnote{emend.; \emph{‘sennum’} \FlatMS}} \edtext{\emph{ï}}{\Afootnote{emend.; \emph{vit ór} \FlatMS}} \alst{s}ǫðlum \hld\ \alst{s}itja vit skulum &
ok of \alst{jǫ}fra \hld\ \alst{ę́}ttir dǿma, &
\alst{g}umna þęira, \hld\ es frȧ \alst{g}oðum kómu.\eva

\bvb Soon in the saddles we two shall sit, \\
and of the lines of chieftains speak, \\
of those men who came from the gods.\evb\evg


\bvg\bva%
Þęir hafa \alst{v}ęðjat \hld\ \alst{v}ala malmi &
\alst{Ó}tt·arr \alst{u}ngi \hld\ ok \alst{A}ngantýr; &
\alst{sk}ylt ’s at vęita, \hld\ svá’t \alst{sk}ati inn ungi &
\alst{f}ǫður-lęifð hafi \hld\ ępt \alst{f}rę́ndr sína.\eva

\bvb They have wagered the Welsh ore \ken{gold}, \\
young Oughter and Ongenthew. \\
It must be granted so that the young prince \\
may have the patrimony after his kinsmen.\evb\evg


\bvg\bva%
\alst{H}ǫrg hann mér gęrði \hld\ \alst{h}laðinn stęinum; &
nú ’s \alst{g}rjót þat \hld\ at \alst{g}lęri orðit; &
rauð hann ï \alst{n}ýju \hld\ \alst{n}auta blóði; &
\alst{ę́} trúði \alst{Ó}tt·arr \hld\ ȧ \alst{ǫ̇}synjur.\eva

\bvb A \inx[C]{harrow} he made for me, loaded with stones; \\
now that stone-pile has turned into glass. \\
He reddened it in the fresh blood of oxen; \\
always did Oughter trust in the \inx[P]{Ossens}.\evb\evg


\bvg\bva%
\alst{N}ú lát forna \hld\ \alst{n}iðja talða &
ok \alst{u}pp-bornar \hld\ \alst{ę́}ttir manna &
hvat ’s \alst{sk}jǫldunga, \hld\ hvat ’s \alst{sk}ilfinga, &
hvat ’s \alst{ǫ}ðlinga \hld\ hvat ’s \alst{y}lfinga &
hvat ’s \alst{h}ǫlð-borit, \hld\ hvat ’s \alst{h}ęrs-borit &
\alst{m}ęst \alst{m}anna val \hld\ \alst{u}nd Mið·garði?“\eva

\bvb Now let ancient ancestors be counted, \\
and the high-born lines of men. \\
What is of Shieldings? What is of Shilvings? \\
What is of Athlings? What is of Wolvings? \\
What is born of hero? What is born of chief, \\
the greatest choice of men in Middenyard?”\evb\evg


\bvg\bva%
\Ballnote{Hindle begins the genealogy; she speaks directly to Oughter.}%
\llap{„}Þú est \alst{Ó}tt·arr \hld\ borinn \alst{I}nn·stęini, &
ęn \alst{I}nn·stęinn vas \hld\ \alst{Ǫ́}lfi inum gamla, &
\alst{Ǫ́}lfr vas \alst{U}lfi, \hld\ \alst{U}lfr Sę́·fara, &
ęn \alst{S}ę́·fari \hld\ \alst{S}van inum rauða.\eva

\bvb\speakernoteb{[Hindle quoth:]}%
“Thou art, Oughter, born to Instone, \\
and Instone was born to Elf the old, \\
Elf was to Wolf, Wolf to Seafarer, \\
and Seafarer to Swan the red.\evb\evg


\bvg\bva%
\alst{M}óður átti faðir þïnn \hld\ \alst{m}ęnjum gǫfga, &
\alst{h}ygg at \alst{h}éti \hld\ \alst{H}lédís gyðja, &
\alst{F}róði vas \alst{f}aðir þęirar, \hld\ ęn \edtext{\alst{F}rí\emph{und}}{\Afootnote{emend. from meaningless \emph{†friaut†} \FlatMS}} móðir; &
\alst{ǫ}ll þȯtti \alst{ę́}tt sú \hld\ með \alst{y}fir-mǫnnum.\eva

\bvb Thy father got thy mother, beautiful with torcs; \\
I think that she was called Leedise the \inx[C]{gidden}. \\
Frood was her father, but Friend her mother; \\
all that lineage seemed to be among \inx[C]{overmen}.\evb\evg


\bvg\bva%
\alst{Au}ði vas \alst{á}ðr \hld\ \alst{ǫ}flgastr manna, &
\alst{H}alf·danr fyrri \hld\ \alst{h}ę́str skjǫldunga, &
\alst{f}rę́g vǫ́ru \alst{f}olk-víg, \hld\ þau’s \alst{f}ramir gęrðu, &
\alst{h}varfla þȯttu hans verk \hld\ með \alst{h}imins skautum.\eva

\bvb Ed was once the strongest of men, \\
Halfdane yet earlier the highest of Shieldings. \\
Famous were the troop-wars which the brave ones made; \\
his \name{= Halfdane’s} works seemed to whirl along the corners of heaven.\evb\evg


\bvg\bva%
\edtrans{\alst{Ę}flði⸗sk}{became the in-law}{\Bfootnote{Lit. ‘was strengthened by’.  Elmwey was Eanmund’s daughter or sister.}} við \alst{Ęy}·mund \hld\ \alst{ǿ}ðstan manna &
ęn vá \alst{S}ig·trygg \hld\ með \alst{s}vǫlum ęggjum, &
\alst{ęi}ga gekk \alst{A}lm·vęig, \hld\ \alst{ǿ}ðsta kvinna, &
\alst{ó}lu þau ok \alst{ǫ́}ttu \hld\ \alst{á}tján sonu.\eva

\bvb He became the in-law of Eanmund, the noblest of men, \\
but he slew Syetrue with cool edges. \\
He went to have Elmwey, the noblest of women; \\
they begot and had eighteen sons.\evb\evg


\bvg\bva%
Þaðan eru \alst{sk}jǫldungar, \hld\ þaðan eru \alst{sk}ilfingar, &
þaðan eru \alst{ǫ}ðlingar, \hld\ þaðan eru \edtrans{\alst{y}nglingar}{Inglings}{\Bfootnote{Note the variation from st.~12, which has \emph{ylfingar} ‘Wolvings’.  Since the Inglings have already been mentioned (under the name of Shilvings), it seems likely that st.~12’s Wolvings is the original reading.}}, &
þaðan es \alst{h}ǫld-borit, \hld\ þaðan es \alst{h}ęrs-borit, &
\alst{m}est \alst{m}anna val \hld\ und \alst{M}ið·garði; &
\alst{a}llt ’s þat \alst{ę́}tt þïn, \hld\ \alst{Ó}tt·arr hęimski.\eva

\bvb Thereof come the Shieldings! Thereof come the Shilvings! \\
Thereof come the Athlings! Thereof come the Inglings! \\
Thereof is born of hero! Thereof is born of chief \\
the greatest choice of men within Middenyard! \\
That is all thy lineage, foolish Oughter!”\evb\evg


\bvg\bva%
Vas \alst{H}ildi·gunnr \hld\ \alst{h}ęnnar móðir, &
\alst{S}vǫ́fu barn \hld\ ok \alst{S}ę́·konungs; &
\alst{a}llt ’s þat \alst{ę́}tt þïn, \hld\ \alst{Ó}tt·arr hęimski. &
\alst{v}arði at \alst{v}iti svá, \hld\ \alst{v}ilt ęnn lęngra?\eva

\bvb Hildguth was her \name{= Elmwey’s} mother, \\
the child of Sweve and Sea-king. \\
That is all thy lineage, foolish Oughter! \\
It is meaningful that one might know thus; wilt thou yet further?\evb\evg


\bvg\bva%
\alst{D}agr átti Þóru \hld\ \alst{d}ręngja móður, &
\alst{ó}lu⸗sk ï \alst{ę́}tt þar \hld\ \alst{ǿ}ðstir kappar, &
\alst{F}rað·marr ok Gyrðr \hld\ ok \alst{F}rekar báðir, &
\alst{Á}mr ok Jǫsur·marr, \hld\ \alst{Ǫ́}lfr inn gamli. &
\alst{v}arðar at \alst{v}iti svá, \hld\ \alst{v}ilt ęnn lęngra?\eva

\bvb Day got Thure, the mother of valiant men; \\
in that lineage were begotten the noblest champions: \\
Fradmer and Yird and both Frecks, \\
Ame and Essermer, Athelf the old.--- \\
It is meaningful that one might know thus; wilt thou yet further?\evb\evg


\bvg\bva%
\alst{K}ętill hét vinr þęira \hld\ \alst{K}lypps arf-þęgi, &
vas hann \alst{m}óður-faðir \hld\ \alst{m}óður þinnar; &
þar vas \alst{F}róði \hld\ \alst{f}yrr ęnn Kári, &
hinn \alst{ę}ldri vas \hld\ \alst{Ǫ́}lfr of getinn.\eva

\bvb Kettle was the name of their friend, the heir of Clip; \\
he was the father of thy mother's mother. \\
There was Frood, yet earlier Keer, \\
the elder Atholf was mentioned.\evb\evg


\bvg\bva%
\edtext{\emph{\alst{N}}anna}{\Afootnote{emend.; \emph{Manna} \FlatMS}} vas \alst{n}ę́st þar \hld\ \alst{N}ǫkkva dóttir
vas \alst{m}ǫgr hennar \hld\ \alst{m}ágr þïns fǫður. &
\alst{f}yrnd ’s sú mę́gð \hld\ \alst{f}ramm tęl’k lęngra &
\edtrans{kunna’k \alst{b}áða \hld\ \alst{B}rodd ok Hǫrvi}{I know both Brord and Harrow}{\Bfootnote{This l.~is probably in error here; it reoccurs identically as 24/1 below.}} &
\alst{a}llt ’s þat \alst{ę́}tt þïn, \hld\ \alst{Ó}tt·arr hęimski.\eva

\bvb Nan was next there, the daughter of Nake; \\
her son was the in-law of thy father. \\
Ancient is that family; I foresee further. \\
I know both Brord and Harrow. \\
That is all thy lineage, foolish Oughter!\evb\evg


\bvg\bva%
\alst{Í}s·ólfr ok \alst{Ǫ̇}s·ólfr \hld\ \alst{Ǫ}l·móðs synir &
ok \alst{Sk}ur·hildar \hld\ \alst{Sk}ękkils dóttur, &
\alst{sk}alt til tęlja \hld\ \alst{sk}atna margra &
\alst{a}llt ’s þat \alst{ę́}tt þïn, \hld\ \alst{Ó}tt·arr hęimski.\eva

\bvb Icewolf and Osulf, the sons of Alemod \\
and Showerhild Shackle’s daughter--- \\
them shalt thou recount before the many princes. \\
That is all thy lineage, foolish Oughter!\evb\evg


\bvg\bva%
\alst{G}unn·arr balkr, \hld\ \alst{G}rímr arð-skafi, &
\alst{Já}rn-skjǫldr, Þȯrir, \hld\ \alst{U}lfr gïnandi, &
\alst{B}úi ok \alst{B}rami, \hld\ \alst{B}arri ok Ręifnir, &
\alst{T}indr ok \alst{T}yrfingr \hld\ ok \alst{t}vęir Haddingjar. &
\alst{a}llt ’s þat \alst{ę́}tt þïn, \hld\ \alst{Ó}tt·arr hęimski.\eva

\bvb Guther the balk, Grim the ard-shaver, \\
Ironshield, Thurwigh, Wolf the grinning, \\
Bowe and Brame, Barr and Rafner, \\
Tind and Tirving and the two Hardings. \\
That is all thy lineage, foolish Oughter!\evb\evg


\bvg\bva%
\alst{Ȧ}ni, \alst{Ȯ}mi \hld\ \alst{v}ǫ́ru bornir &
\alst{A}rn·gríms synir \hld\ ok \alst{Ęy}·furu; &
\alst{b}rǫkun \alst{b}er-sęrkja \hld\ \alst{b}ǫls margs-konar &
of \alst{l}ǫnd ok of \alst{l}ǫg \hld\ sem \alst{l}ogi fǿri. &
\alst{a}llt ’s þat \alst{ę́}tt þïn, \hld\ \alst{Ó}tt·arr hęimski.\eva

\bvb Ane, Ome were born \\
the sons of Arngrim and Eanfore. \\
The battles of the bear-serks of the manifold evils \\
over land and over sea went forth like a flame. \\
That is all thy lineage, foolish Oughter!\evb\evg


\bvg\bva%
Kunna’k \alst{b}áða \hld\ \alst{B}rodd ok Hǫrvi &
vǫ́ru þęir ï \alst{h}irð \hld\ \alst{H}rólfs ins gamla, &
\alst{a}llir bornir \hld\ frȧ \alst{Jǫ}rmun·reki &
\alst{S}ig·urðar mági \hld\ (hlýð \alst{s}ǫgu mïnni!) &
\alst{f}olkum gríms \hld\ þęss ’s \alst{F}áfni vá.\eva

\bvb I know both Brord and Harrow; \\
they were in the troop of Rotholf the old, \\
all born from Ermenric, \\
the inlaw of Siward---heed my tale!--- \\
the terror of hosts who slew Fathomer.\evb\evg


\bvg\bva%
Sá vas \alst{v}ísir \hld\ frȧ \alst{V}ǫlsungi &
ok \alst{H}jǫr·dís \hld\ frȧ \alst{H}rauðungi &
ęn \alst{Ęy}·limi \hld\ frȧ \alst{Ǫ}ðlingum. &
\alst{a}llt ’s þat \alst{ę́}tt þïn, \hld\ \alst{Ó}tt·arr hęimski.\eva

\bvb That ruler came from Walsing, \\
and Hardise from Reading, \\
but Eanlime from the Athlings. \\
That is all thy lineage, foolish Oughter!\evb\evg


\bvg\bva%
\alst{G}unnarr ok Hǫgni \hld\ \alst{G}júka arfar &
ok it \alst{s}ama Guð·ru̇n \hld\ \alst{s}ystir þęira. &
Ęigi vas \alst{G}uð·þormr \hld\ \alst{G}júka ę́ttar &
þó vas hann \alst{b}róðir \hld\ \alst{b}ęggja þęira &
\alst{a}llt ’s þat \alst{ę́}tt þïn, \hld\ \alst{Ó}tt·arr hęimski.\eva

\bvb Guther and Hawn, Yivick’s heirs, \\
and likewise Guthrun, their sister. \\
Godthorm was not of Yivick’s lineage, \\
though he was the brother of them both. \\
That is all thy lineage, foolish Oughter!\evb\evg


\bvg\bva%
\alst{H}araldr \alst{h}ildi-tǫnn \hld\ borinn \alst{H}rǿ·reki &
\alst{s}løngvan-bauga \hld\ \alst{s}on*r vas hann Auðar. &
\alst{Au}ðr djúp-audga \hld\ \alst{Í}·vars dóttir &
ęn \alst{R}áð·barðr \hld\ vas \alst{R}and·vé*s faðir &
þęir vǫ́ru \alst{g}umnar \hld\ \alst{g}oðum signaðir. &
\alst{a}llt ’s þat \alst{ę́}tt þïn, \hld\ \alst{Ó}tt·arr hęimski.\eva

\bvb Harold Hild-tooth was born to Rethric \\
the slinger of bighs; he was the son of Ead. \\
Ead the deep-minded was Inwar’s daughter, \\
but Redebeard was Randwigh’s father. \\
They were men dedicated to the gods. \\
That is all thy lineage, foolish Oughter!\evb\evg


\bvg\bva%
Vǫ́ru \alst{ę}llifu \hld\ \alst{ę̇}sir taldir &
\alst{B}aldr es hné \hld\ við \alst{b}ana-þúfu &
þęss létsk \alst{V}ȧli \hld\ \alst{v}erðr at hęfna &
\alst{s}ïns bróður \hld\ \alst{s}ló hand-bana; &
\alst{a}llt ’s þat \alst{ę́}tt þïn, \hld\ \alst{Ó}tt·arr hęimski.\eva

\bvb Eleven were the Eese tallied \\
when Balder sank down against the tuft of his bane. \\
That Wonnel declared himself worthy to avenge; \\
he slew the hand-bane of his brother. \\
That is all thy lineage, foolish Oughter!\evb\evg


\bvg\bva%
Vas \alst{B}aldrs faðir \hld\ \alst{B}urs arf-þegi, &
Fręyr átti \alst{G}ęrdi, \hld\ hȯn vas \alst{G}*ymis dóttir &
\alst{jǫ}tna \alst{ę́}ttar \hld\ ok \alst{Au}r·boðu; &
\alst{þ}ó vas \alst{Þ}jatsi \hld\ \alst{þ}ęira frę̇ndi &
\alst{sk}aut-gjarn jǫtunn; \hld\ hans vas \alst{Sk}aði dóttir.\eva

\bvb Balder’s father was the heir of Byre. \\
Free had Gird; she was Gymer’s daughter \\
of the lineage of the ettins, and Earbode’s. \\
Yet Thedse was their kinsman, \\
the shot-eager ettin; Shede was his daughter.\evb\evg


\bvg\bva%
\alst{M}art sęgjum þér \hld\ ok \alst{m}unum flęira &
\alst{v}ǫrumsk at \alst{v}iti svá, \hld\ \alst{v}ilt ęnn flęira.\eva

\bvb Much we tell thee and remember more; \\
I warn against one knowing such; wilt thou yet more?\evb\evg


\bvg\bva%
\alst{H}aki vas \alst{H}vęðnu \hld\ \alst{H}óti bętstr sona &
ęn \alst{H}vęðnu vas \hld\ \alst{H}jǫr·varðr faðir &
\alst{H}eiðr ok \alst{H}ross·þjófr \hld\ \alst{H}rímnis kindar.\eva

\bvb TODO. \\
TODO. \\
TODO.\evb\evg


\bvg\bva%
\Ballnote{This st.~is quoted following \Gylfaginning~5:6--9, which describes Yimer (cf.~\textlink{Voluspa}[3]/1 n.) and \Gylfaginning~5:10--11 which introduces it with the words \emph{ok eru þaðan komnar ę́ttir hrím-þursa, svá sem segir í Vǫlu-spá hinni skǫmmu} ‘and thereof are come the lineages of Rime-Thurses, as it says in the short Spae of the Wallow’.}%
Eru \alst{v}ǫlur allar \hld\ frȧ \alst{V}íð·ólfi, &
\alst{v}ítkar allir \hld\ frȧ \alst{V}il·męiði, &
\edtext{\alst{s}ęið-berendr}{\Afootnote{\emph{skil-berendr} \FlatMS}} \hld\ frȧ \alst{S}vart-hǫfða, &
\alst{jǫ}tnar \alst{a}llir \hld\ frȧ \alst{Y}mi komnir.\eva

\bvb All wallows are from Widolf, \\
all warlocks from Wilmad, \\
soth-bearers from Swarthead, \\
all ettins are come from Yimer.\evb\evg


\bvg\bva%
\alst{M}art sęgjum þér \hld\ ok \alst{m}unum flęira &
\alst{v}ǫrumsk at \alst{v}iti svá, \hld\ \alst{v}ilt ęnn lęngra?\eva

\bvb Much we tell thee and remember more; \\
I warn against one knowing such; wilt thou yet further?\evb\evg


\bvg\bva%
Varð \alst{ęi}nn borinn \hld\ ï \alst{á}r-daga &
\alst{r}amm-aukinn mjǫk \hld\ \alst{r}ǫgna kindar; &
\alst{n}íu bǫ́ru þann \hld\ \alst{n}ad\emph{d-g}ǫfgan mann &
\alst{jǫ}tna męyjar \hld\ við \alst{ja}rðar þrǫm.\eva

\bvb One was born in days of yore, \\
much increased in strength by the race of the Reins. \\
Nine gave birth to that arrow-famed man: \\
the maidens of ettins at the rim of the earth.\evb\evg


\bvg\bva%
\alst{M}art sęgjum þér \hld\ ok \alst{m}unum flęira &
\alst{v}ǫrumsk at \alst{v}iti svá, \hld\ \alst{v}ilt ęnn lęngra?\eva

\bvb Much we tell thee and remember more; \\
I warn against one knowing such; wilt thou yet further?\evb\evg


\bvg\bva%
Hann \alst{G}jǫlp of bar \hld\ hann \alst{G}ręip of bar, &
bar hann \alst{Ęi}stla \hld\ ok \alst{Ęy}r·gjafa; &
hann bar \alst{U}lf·ru̇n \hld\ ok \alst{A}ng·ęyja, &
\alst{Í}mdr ok \alst{A}tla \hld\ ok \alst{Ja}rn·saxa.\eva

\bvb Him Yelp did bear; him Grope did bear; \\
Astle bore him and Earyive; \\
Wolfrun bore him and Angie, \\
Imd and Attle and Ironsaxe.\evb\evg


\bvg\bva%
Sá vas \alst{au}kinn \hld\ \alst{ja}rðar magni &
\alst{s}val-kǫldum \alst{s}ę́ \hld\ ok \alst{s}ónar dręyra.\eva

\bvb He was increased with the might of the earth, \\
the freezing cold sea and the guts of a boar.\evb\evg


\bvg\bva%
\alst{M}art sęgjum þér \hld\ ok \alst{m}unum flęira &
\alst{v}ǫrumsk at \alst{v}iti svá, \hld\ \alst{v}ilt ęnn lęngra?\eva

\bvb Much we tell thee and remember more; \\
I warn against one knowing such; wilt thou yet further?\evb\evg


\bvg\bva%
\alst{Ó}l \alst{U}lf Loki \hld\ við \alst{A}ngr·boðu &
ęn \alst{S}lęipni gat \hld\ við \alst{S}vaðil·fara; &
\edtext{\alst{ęi}tt þȯtti skars}{\lemma{ęitt \dots\ skars ‘One troll-woman’}\Bfootnote{Viz.~Hell, Lock’s daughter.}} \hld\ \alst{a}llra fęiknast &
þat vas \alst{b}róður frȧ \hld\ \alst{B}ý·lęists komit.\eva

\bvb Lock begot the Wolf with Angerbode, \\
but Slapner he begat with Swaddlefare. \\
One troll-woman seemed wickedest of all; \\
she was come from the brother of Bilast.\evb\evg


\bvg\bva%
\alst{L}oki á\emph{t} hjarta \hld\ \alst{l}indi bręnndu &
fann hann \alst{h}alf-sviðinn \hld\ \alst{h}ug-stęin ko\emph{n}u; &
varð Loptr \alst{k}viðugr \hld\ af \alst{k}onu illri &
þaðan es ȧ \alst{f}oldu \hld\ \alst{f}lagð hvęrt komit.\eva

\bvb Lock ate a heart burned with linden-wood; \\
he found the half-scorched thought-stone \ken{heart} of a woman. \\
Loft became pregnant from that evil woman; \\
thereof is come every troll upon the earth.\evb\evg


\bvg\bva%
\Ballnote{A prediction of the Rakes of the Reins (as told at length in \textlink{Voluspa}[43]--54), likely the reason why this passage is called the Short Spae of the Wallow.}%
\alst{H}af gęngr \alst{h}ríðum \hld\ við \alst{h}imin sjalfan &
\alst{l}íðr \alst{l}ǫnd yfir \hld\ ęn \alst{l}opt bilar; &
þaðan koma \alst{s}njóvar \hld\ ok \alst{s}narir vindar &
þȧ ’s ï \alst{r}áði \hld\ at \alst{r}ęg\emph{i}n of þrjóti.\eva

\bvb The sea rises in storms against the very heaven, \\
goes over the lands, but the air gives way. \\
Thence come snows and swift winds; \\
then it is in the motions that the Reins expire.\evb\evg


\bvg\bva%
Varð \alst{ęi}nn borinn \hld\ \alst{ǫ}llum męiri &
sá vas \alst{au}kinn \hld\ \alst{ja}rðar m\emph{a}gni. &
þann kveða \alst{st}illi \hld\ \alst{st}ór-auðgastan &
\alst{S}if \alst{s}ifjaðan \hld\ \alst{s}jǫtum gǫrv-ǫllum.\eva

\bvb One was born greater than all; \\
he was increased with the might of the earth. \\
That ruler they say is the very wealthiest, \\
related to Sib, to each and every host.\evb\evg


\bvg\bva%
Þȧ kømr \alst{a}nnarr \hld\ \alst{ę}nn mátkari &
þó \alst{þ}ori’k ęigi \hld\ \alst{þ}ann at nęfna; &
\alst{f}áir sjá nú \hld\ \alst{f}ramm of lęngra &
an \alst{Ó}ðinn man \hld\ \alst{U}lfi mǿta.“\eva

\bvb Then comes another one yet mightier, \\
though I dare not say his name. \\
Few do now see further ahead, \\
than when Weden will meet the Wolf.”\evb\evg


\bvg\bva%
„Ber \alst{m}innis-ǫl \hld\ \alst{m}ïnum gęsti &
svá hann \alst{ǫ}ll muni \hld\ \alst{o}rð at tïna &
\alst{þ}essa rǿðu \hld\ ȧ \alst{þ}ridja morni &
þȧ’s þęir \alst{A}ngan·týr \hld\ \alst{ę́}ttir rękna.“\eva

\bvb “Bear an ale of remembrance unto my guest, \\
that he may remember to recite all the words \\
of this speech on the third morning, \\
when he and Ongentew reckon lineages.”\evb\evg


\bvg\bva%
„\alst{S}nú þú burt heðan \hld\ \alst{s}ofa lystir mik &
\alst{f}ę̇r þú \alst{f}átt af mér \hld\ \alst{f}ríðra kosta. &
hlęypr þú edlvina \hld\ úti ȧ nǫ́ttum &
sem með hǫfrum \hld\ Hęið·ru̇n fari.\eva

\bvb Turn away hence!  I desire to sleep; \\
from me wilt thou get few pretty things. \\
TODO. \\
TODO.\evb\evg


\bvg\bva%
Rant at \alst{Ǿ}ði \hld\ \alst{ęy} þręyjandi &
skutusk þér flęiri \hld\ und fyrir skyrtu. &
hleypr þu edlvina \hld\ úti ȧ nǫ́ttum &
sem með hǫfrum \hld\ Hęið·ru̇n fari.“\eva

\bvb TODO. \\
TODO. \\
TODO. \\
TODO.\evb\evg


\bvg\bva%
„\alst{E}k slę́ \alst{ę}ldi \hld\ af \alst{ï}·viðju &
svá’t \alst{ęi}\emph{gi} kømsk \hld\ \alst{ȧ} burt heðan.“ &
„hleypr þu edlvina \hld\ úti ȧ nǫ́ttum &
sem með \alst{h}ǫfrum \hld\ \alst{H}ęið·ru̇n fari.\eva

\bvb “I strike a fire around the Inwithy, \\
that thou wilt not come away hence!” \\
TODO. \\
TODO.\evb\evg


\bvg\bva%
\alst{H}yr sé’k brenna \hld\ ęn \alst{h}auðr loga &
verða \alst{f}lęstir \hld\ \alst{f}jǫr-lausn þola; &
\alst{b}er Ótt·ari \hld\ \alst{b}jór at hęndi &
\alst{ęi}tri blandinn mjǫk \hld\ \alst{i}llu hęilli.“\eva

\bvb A fire I see burning, but the earth does blaze; \\
most must suffer the saving of life. \\
Bring beer to Oughter’s hands, \\
mixed with much poison, of ill omen!”\evb\evg


\bvg\bva%
„\alst{O}rð-hęill þïn \hld\ skal \alst{ę}ngu ráða &
þótt \alst{b}rúðr jǫtuns \hld\ \alst{b}ǫlvi hęitir; &
hann skal \alst{d}rekka \hld\ \alst{d}ýrar vęigar &
bið’k \alst{Ó}tt·ari \hld\ \alst{ǫ}ll goð duga.“\eva

\bvb “Thy word-omens shall have no power, \\
though the ettin’s bride does summon a bale. \\
He shall drink costly draughts; \\
Oughter I bid all gods avail!”\evb\evg

\sectionline
