\bookStart{Leeds of Hoarbeard}[Hár·barðs ljóð]
\setBookCode{Harbardsljod}

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: early C11th (0.578)--late C11th (0.377)

\textbf{Meter:} Varied%
\end{flushright}

\section{Introduction}

\subsection{Preservation}

The \textbf{Leeds of Hoarbeard} (\textlink{Harbardsljod}) is preserved in full (sts.~1--60) in \Regius\ on foll.~12r/30--13v/26.  Sts.~14 (n.)--60 are found in \AM\ on foll.~1r/1--1v/17.

\subsection{Contents}

\Harbardsljod\ is a dialogue between the two figures Thunder (\emph{Þȯrr}) and Hoarbeard (\emph{Hár·barðr}, i.e. ‘grey-beard’), who meet at a sound.  Thunder asks Hoarbeard, the ferryman, for passage; an argument ensues, and Thunder is forced to depart without passage.

Thunder is of course the well-known god, but Hoarbeard merits discussion.  Considering his name, which appears in \textlink{Grimnismal}[51] as a name for Weden, and his generally Odinic character (cf.~esp.~st.~24), he is near-universally considered to be the god Weden in disguise.

The difference in personality between the two gods is striking, so that \Harbardsljod\ is far from a flyting between, say, two warriors of equal stature.  Thunder is portrayed as moral and altruistic, but also as rather naive and the foil for Hoarbeard’s insults.  Hoarbeard on the other hand is egoistic and sly.  He is the first to insult Thunder, and gets the last laugh by denying him passage.

The poem might be seen as an allegory on class relations, namely between the self-owning yeomen farmers and the warlike earls, represented by their patron gods.  TODO.

\subsection{Meter}

Of all Eddic poems \textlink{Harbardsljod} is probably the strangest in terms of form.  It consists of 60 “stanzas”, if all of them can accurately be called that.  No narrative prose is embedded apart from P1 at the very beginning, which merely sets the stage for the dialogue.

Stanza length varies greatly.  4 lines is by no means the standard (in fact only a single fully regular \Ljodahattr-stanza of type \emph{abcabc} occurs, viz.~st.~47), and many sts.~consist of single lines.

The meter is a mix of a peripheral form of \Malahattr\ and \Ljodahattr.  Many lines are of an obscene length somewhat reminiscent of, though certainly not related to \textlink{Heliand} (e.g.~sts.~4/3b, 56/5a, 59/1--2); others have no metrical qualities at all (17/1, 31/1, 41/1, 44--46, 57/1, 60/1).

\subsection{Language}

The young clitic definite \emph{-inn} is employed with unique frequency throughout \Harbardsljod\ in both the prose and poetic lines.  This is probably the clearest identifiable young feature of the poem.  The clitic, although occasionally attested in Wiking Age runic inscriptions, only sporadically appears in poetry older than the 14th c., suggesting that it was not considered proper to use in poetic composition.  (TODO: cite this.)

The normal forms \emph{hingat} and \emph{þangat} are used in the respective sense ‘hither’ and ‘thither’ for the older \emph{hinig} and \emph{þinig}.

On the other hand there is the presence of the rare and prob.~old words \emph{ǫgurr} ‘burden(?)’ (st.~13) and \emph{gamban-tęinn} (st.~20).  The old particle \emph{of} is also used before finite verbs, although not very frequently.

\subsection{Possibilities of performance}

The loose meter and non-formal language demand discussion.  To begin with, we should disregard the notion that \Harbardsljod\ is heavily corrupt and requires restoration.  It must be stated clearly that no amount of emendation can “restore” \Harbardsljod\ to pure \Ljodahattr\ like we find in \textlink{Havamal} or \textlink{Lokasenna}.  Most lines, including those of unusual length or without metrical qualities, are readily understandable and fit well both within their immediate context and the poem as a whole.  If there is such great corruption, it has produced a remarkably consistent and intelligible text.  Even so, the poem is unconventional.

As all Norse poetry, at least of sufficient age (see below), was originally oral, \Harbardsljod\ must have been performed in some way.  The problems described above are solved quite nicely if we assume that it was acted out in a sort of carnivalesque theatre with two actors standing on a stage, perhaps separated by a staged body of water, and shouting at each other; alternatively with one actor playing both roles by the use of props and acting.

The whole poem seems remarkably fit for such a performance, and with this assumption many pieces fall into place.  Take Thunder’s talk of his breakfast and the pack on his back (st.~3), which invites the actor to point to his back and then rub his belly; Hoarbeard’s insult of Thunder’s dress (st.~6), which would work much better if there was in fact an actor with his pants pulled up to his knees; or Hoarbeard’s declaration that he will not cross the water but instead remain \emph{hér} ‘here’ (st. 14), which invites stomping on the ground.  A theatric performance would also explain the metrical variations and the use of informal non-poetic features like the clitic definite as vulgar elements.
The prose lines may even have been shouted out, which would explain their lack of meter as a discordant stylistic feature meant to catch the audience by surprise; the regular metrical sts., like 47, become all the more powerful when standing immediately surrounded by prose.

(TODO: mention the concept of the "double scene" by Lars Lönnroth?)

Circumstantial support is found in the structure of the dialogue, which has a remarkable affinity to a fixture in modern comedic duos: the straight man (Thunder) and the funny man (Hoarbeard).  The by Wiking Age standards ever-moral Thunder serves as the perfect foil to the egoistic Hoarbeard and falls effortlessly for the ferryman’s “trolling”.  It is hard not to imagine the audience laughing more and more as Thunder gets angrier and angrier.

The god really is the butt of the joke for much of the poem.  As nooted above, his threat in st.~47 is quite impressive, but it is soon neutralized by Hoarbeard’s accusation that his wife Sib is currently engaged in adultery behind his back.  Even so, Thunder’s great feats are never made fun of directly (beyond st.~37--39, where Thunder has the last word); in the pagan religion such stories served not just as entertainment, but also as ritual etiology \parencite{Lindow1988}.

\subsection{Dating}

Formal and linguistic criteria can be taken to point towards a later date for \Harbardsljod\ than for most other mythological poems, but as discussed above, the loose meter and use of the clitic definite may be stylistic features belonging to the relaxed and vulgar register of the poem, which is clearly meant to entertain the audience.

There are also features that speak against a late date.  Among content-based criteria, there is a thorough understanding of the personalities of the two gods, which would seem unlikely to have survived long after the conversion of Iceland and which does not come off as antiquarian.  The use of a variant ettin-name for the attested story in st.~26 is notable since it points away from antiquarian composition, which would more likely have inserted the familiar name \emph{Skrymir} ‘Shrimer’.

It is shown by \textlink{Lokasenna} (certainly an old poem, as proven by the metrically required \emph{vr} in \textlink{Lokasenna}[15]/3 (n.), 18/4, 27/4) and \textlink{Thrymskvida} (generally held to be a particularly old poem) that pre-Christian poetry could be light-hearted in its depiction of the gods, and \Harbardsljod\ certainly falls neatly into that pattern.

The model devised by \textcite{Sapp2022} gives the poem a 57.8\% likelihood of being from the early 11th c., and a 37.7\% likelihood of being from the late 11th.  These percentages are most similar to those obtained by \textlink{Gripisspa} (a poem recognized as very young---not from the 11th c.) but \Harbardsljod\ shares so little in common with it that the similarity is must be considered unreliable.

\newpage

\section{Text}

\subsection{The Leeds of Hoarbeard}

\bpg\bpa\mssnote{\Regius~12r/30}%
Þórr fór ór austr-vegi ok kom at sundi einu.  Ǫðrum megum sunds’ins var ferju-karl’inn með skip’it. Þórr kallaði:\epa

\bpb {\huge T}\textsc{hunder journeyed} out of the Eastern Way and came to a sound.  At the other side of the sound was the ferryman with the ship.  Thunder called out:\epb\epg


\bvg\bva\mssnote{\Regius~12r/32}%
„Hvęrr ’s sá \edtrans{\alst{s}vęinn \alst{s}vęina}{swain of swains}{\Bfootnote{The word \emph{svęinn} ‘swain, lad’ generally refers to a prepubescent boy, later on a page (the servant of a knight, who was aged approx.~7--14), but can also mean ‘servant’ (cf.~English boy as a term of address), as is most likely the case here.  It is presumably offended by this address that the decidedly non-servile Hoarbeard decides to insult Thunder. ---
The construction “X of Xs” may ultimately stem from Biblical language, although it appears fairly early on in Norse poetry; cf.~\textlink{GudrunTwo}[1]/1.}} \hld\ es stęndr fyr \alst{s}und’it handan?“\eva

\bvb%
“Who is this swain of swains that stands yonder across the sound?”\evb\evg


\bvg\bva\mssnote{\Regius~12v/1}%
\speakernote{Hann svaraði:}%
„Hvęrr ’s sá \edtrans{\alst{k}arl \alst{k}arla}{churl of churls}{\Bfootnote{Hoarbeard directs an insult to Thunder by denigrating his social status as a lowly peasant.  The construction should not be read as a compliment like modern Swedish \emph{karlakarl} ‘a real man, manly man’.
Although \emph{karl} ‘churl’ is an unmarked word for ‘adult human being’ in later NGmc., including modern Icelandic, in the oldest Norse poetry it specifically refers to a ‘churl, peasant, old man’ (so \textlink{Hymiskvida}[10]/4, 31/3, 32/3, \textlink{HelgakvidaTwo}[2]/2, \textlink{Reginsmal}[18]/3, \textlink{Rigsthula}[22]/4; in \Skp: Vitg Lv 1, KormǪ Lv 16, EVald \emph{Þórr} 1, Gunnl Lv 5) unless occurring in the collocation \emph{konur ok karlar} ‘women and menfolk’ (\textlink{Havamal}[91]/2, \textlink{Helreid}[14]/2, \textlink{Atlamal}[70]/1).
The only exception to this in the Poetic Edda is \textlink{Atlamal}[63]/3, but \textlink{Atlamal} is a fairly late poem.}} \hld\ es \alst{k}allar of vág’inn?“\eva

\bvb\speakernoteb{He answered:}%
“Who is this churl of churls that calls out o’er the wave?”\evb\evg


\bvg\bva\mssnote{\Regius~12v/2}%
\speakernote{[Þórr kvað:]}%
„\alst{F}ęr þú mik of sund’it, \hld\ \alst{f}ǿði’k þik á morgun; &
\alst{m}ęis hęfi’k ȧ baki, \hld\ verðr⸗a \alst{m}atr inn bętri. &
Át’k ï \alst{h}víld \hld\ áðr ek \alst{h}ęiman fór, &
\alst{s}íldr ok \edtrans{hafra}{oatmeal/he-goats}{\Bfootnote{The easiest reading (preferred by \textcite{FinnurEdda}, \textcite{PettitEdda} and others) is an acc.~pl.~of \emph{hafr} ‘he-goat’.  Thunder also eats his goats in \Gylfaginning\ 44, where he butchers and cooks them in the evening and brings them back to life at dawn by blessing them with his hammer (cf.~\textlink{Hymiskvida}[37]--38), and this seems to be a common motif of the god.  For his goats cf.~\textlink{Hymiskvida}[20], 31, \textlink{Thrymskvida}[21].

Other scholars instead read an acc. pl. of \emph{hafri} ‘oat’, i.e. ‘porridge, oatmeal’.  \textcite{Stiles2023} connects this with the porridge-eating of the Vedic god \textoi{Pūṣán} (\Rigveda\ 6.56.1, 57.2), who is “partner and yokemate” (\Rigveda\ 6.56.2) of \textoi{Índra}, Thunder’s Vedic equivalent.  Another similarity Stiles notes between Thunder and \textoi{Pūṣán} is that both have chariots driven by goats (\Rigveda\ 6.57.3: “Goats are the draft-animals for the one”, 6.58.2: “Having goats as his horses”).  Whether the Vedic tradition has split the Thunder-god in two or whether the Germanic Thunder has absorbed elements of an earlier yokemate is hard to say.  Like \textoi{Índra}, Thunder is frequently joined by companions, as is in fact one of his most identifying characteristics \parencite[123]{Lindow1988}.}}; \hld\ \alst{s}aðr em’k ęnn þęss.“\eva

\bvb\speakernoteb{[Thunder quoth:]}%
“Ferry me o’er the sound, I’ll feed thee in the morning! \\
A basket I have on my back; better food will not be found. \\
I ate, while at rest before I journeyed from home, \\
herring and oatmeal/he-goats---I am still full thereof.”\evb\evg


\bvg\bva\mssnote{\Regius~12v/5}%
\speakernote{[Hár·barðr kvað:]}%
„\edtrans{Ár⸗ligum \alst{v}erkum \hld\ hrósar þú, \alst{v}ęrði’num}{Of early works thou boastest, of thy eating}{\Bfootnote{A difficult line, not unlikely corrupt.  The present interpretation sees it as an insult directed against Thunder for bragging about the quality of his food; that is not the type of \emph{ár⸗ligt verk} ‘early morning work’ a real man should boast about (cf.~\textlink{Havamal}[33], 58--59).}}; &
\ind \alst{v}ęitst⸗at-tu fyr gǫr⸗la, &
\alst{d}ǫpr ’ro þín hęim-kynni, \hld\ \alst{d}auð hygg’k at þïn móðir sé.“\eva

\bvb%
“Of early morning works thou boastest---of thy eating! \\
\ind Thou seest not clearly ahead; \\
dire is the state of thy home; I think that thy mother is dead!”\evb\evg


\bvg\bva\mssnote{\Regius~12v/6}%
\speakernote{[Þórr kvað:]}%
„\alst{Þ}at sęgir þú nú \hld\ es hvęrjum \alst{þ}ikkir &
\alst{m}ęst at vita: \hld\ at mín \alst{m}óðir dauð sé.“\eva

\bvb%
“Now thou sayest that which to every man seems \\
of most weight to know---that my mother is dead!”\evb\evg


\bvg\bva\mssnote{\Regius~12v/8}%
\speakernote{[Hár·barðr kvað:]}%
„\alst{Þ}ęy⸗gi ’s sęm \alst{þ}ú \hld\ \alst{þ}rjú bú góð ęigir; &
\alst{b}ęr-\alst{b}ęinn þú stęndr \hld\ ok hęfir \alst{b}rautinga gørvi, &
\ind þat-ki at þú hafir \alst{b}rǿkr þïnar.“\eva

\bvb%
“Yet it’s not as if \emph{thou} mightst own three good farms--- \\
bare-legged thou standest and hast the gear of a tramp; \\
\ind it’s hardly as if thou own thy breeches!”\evb\evg


\bvg\bva\mssnote{\Regius~12v/9}%
\speakernote{[Þórr kvað:]}%
„\alst{St}ýr þú hingat ęikju’nni, \hld\ ek mun þér \alst{st}ǫð’na kęnna &
eða \alst{h}vęrr á skip’it \hld\ es þú \alst{h}ęldr við land’it?“\eva

\bvb%
“Steer hither the boat! I will show thee to the harbour--- \\
or who owns the ship which thou holdest by the shore?”\evb\evg


\bvg\bva\mssnote{\Regius~12v/11}%
\speakernote{[Hár·barðr kvað:]}%
„\alst{H}ild·ólfr sá \alst{h}ęitir \hld\ es mik \alst{h}alda bað, &
\alst{r}ekkr inn \alst{r}áð-svinni \hld\ es býr ï \alst{R}áðs-ęyjar-sundi; &
bað⸗at hann \alst{h}lęnni-męnn flytja \hld\ eða \alst{h}rossa-þjófa, &
\alst{g}óða ęina \hld\ ok þá’s ek \alst{g}ǫrva kunna; &
\alst{s}ęg-ðu til nafns þíns \hld\ ef þú vill of \alst{s}und’it fara.“\eva

\bvb%
“Hildolf is he called who bade me hold it, \\
the counsel-wise man who lives in Redeseysound. \\
He bade me not ferry highwaymen nor horse-thieves--- \\
good ones alone, and those I know well--- \\
speak to thy name if thou wilt come o’er the sound!”\evb\evg


\bvg\bva\mssnote{\Regius~12v/15}%
\speakernote{[Þórr kvað:]}%
„\alst{S}ęgja mun’k til nafns míns \hld\ þó’tt ek \edtrans{\alst{s}ękr}{charged}{\Bfootnote{I.e.~involved in an unresolved blood feud.}} sjá’k &
ok til \alst{a}lls \alst{ø}ðlis: \hld\ Ek em \alst{Ó}ðins sonr, &
\alst{M}ęila bróðir \hld\ ęn \alst{M}agna faðir, &
\alst{þ}rúð-valdr goða \hld\ við \alst{Þ}ȯr knátt-u hér dǿma! &
\alst{H}ins vil’k nú spyrja, \hld\ \alst{h}vat þú hęitir.“\eva

\bvb%
“I will speak to my name---though I might be charged--- \\
and to all my origin: I am Weden’s son, \\
Male’s brother and Main’s father, \\
the strength-wielder of the Gods---with Thunder dost thou here speak! \\
One thing will I now ask: what thou art called.”\evb\evg


\bvg\bva\mssnote{\Regius~12v/18}%
\speakernote{[Hár·barðr kvað:]}%
„\alst{H}ár·barðr ek \alst{h}ęiti, \hld\ \alst{h}yl’k umb nafn sjaldan.“\eva

\bvb%
“Hoarbeard I am called; I seldom conceal my name.”\evb\evg


\bvg\bva\mssnote{\Regius~12v/18}%
\speakernote{[Þórr kvað:]}%
„Hvat skalt-u umb \alst{n}afn hylja \hld\ \alst{n}ema þú sakar ęigir?“\eva

\bvb%
“Why shalt thou conceal thy name, unless thou have charges?”\evb\evg


\bvg\bva\mssnote{\Regius~12v/19}%
\speakernote{[Hár·barðr kvað:]}%
„Ęn þó’tt ek \alst{s}akar ęiga, \hld\ fyr \alst{s}líkum sem þú est &
þȧ mun’k \alst{f}orða \alst{f}jǫrvi mïnu \hld\ nema ek \edtrans{\alst{f}ęigr}{fey}{\Bfootnote{Doomed (by fate) to die shortly.}} sé.“\eva

\bvb%
“Yet though I had charges, for such a one as thou art \\
I would then protect my life, unless I should be \inx[C]{fey}.”\evb\evg


\bvg\bva\mssnote{\Regius~12v/21}%
\speakernote{[Þórr kvað:]}%
„\edtrans{Harm ljótan \hld\ mér \alst{þ}ikkir ï \alst{þ}ví}{An ugly harm it seems to me}{\Bfootnote{The meter is clearly faulty.  The whole line comes off rhythmically as an a- and b-verse of \Ljodahattr, but the verses are not linked by alliteration, which on the other hand occurs on \emph{þ} in the b-verse.}} &
\ind at \alst{v}aða of \alst{v}ág’inn til þïn &
\ind ok \alst{v}ę́ta \edtrans{ǫgur}{burden}{\Bfootnote{The sense of this word is not clear, though it is probably the same as the first element of the compound \emph{ǫgur-stund} ‘burdensome hour’ found in \textlink{Volundarkvida}[42].
I find the most convincing interpretation to be that it refers to the food Thunder carries on his back, as previously mentioned in st.~3.  Thunder then uses his desire not to get his food wet as an excuse not to wade.  Some authors have read it as a euphemism for ‘bollocks’, but this is discordant with the angered but quite serious tone of the rest of the stanza and with Thunder’s general role in the poem as the “straight man” next to the scurrilous Hoarbeard.
It is also a very poor excuse for Thunder’s unwillingness to wade, since wading is a frequent attribute of this god (e.g. \textlink{Grimnismal}[29], \textlink{EddicFragments}[F6][6]), who on those occasions shows no fear of “getting his bollocks wet”.}} mïnn; &
\edtext{skylda’k launa \alst{k}ǫgur-svęini \hld\ þïnum \alst{k}angin-yrði}{\Bfootnote{Cf.~\Gylfaginning\ 45: \emph{Ekki munu hirð-menn Útgarða-Loka vel þola því·líkum kǫgur-sveinum kǫpur-yrði.} ‘TODO.’}} &
\ind ef ek \alst{k}om⸗umk yfir sund’it.“\eva

\bvb%
“An ugly harm it seems to me \\
\ind to wade o’er the wave to thee, \\
\ind and wet this burden of mine. \\
I would repay thee, swaddle-swain, for thy mocking words, \\
\ind if I might get myself over the sound.”\evb\evg


\bvg\bva\mssnote{\Regius~12v/23}%
\speakernote{[Hár·barðr kvað:]}%
„\alst{H}ér mun’k standa \hld\ ok þïn \alst{h}eðan bíða; &
fannt⸗a-tu mann inn \alst{h}arðara \hld\ at \edtrans{\alst{H}rungni}{Rungner}{\Bfootnote{Rungner was a famous ettin slain by Thunder in a fierce battle (\Skaldskaparmal\ 24--25, \Haustlong\ 14--20).  Hoarbeard’s mention of him sets off a long argument over their respective accomplishments which takes up the bulk of the poem (sts.~14--40).}} dauðan.“\eva

\bvb%
“\emph{Here} will I stand, and from \emph{here} await thee. \\
Thou hast not found a harder man since \inx[P]{Rungner} died!”\evb\evg


\bvg\bva\mssnote{\Regius~12v/25}%
\speakernote{[Þórr kvað:]}%
„\alst{H}ins vilt-u nú geta \hld\ es vit \alst{H}rungnir dęildum, &
sá inn \alst{st}ór-úðgi jǫtunn, \hld\ \edtrans{es ór \alst{st}ęini vas hǫfuð’it ȧ}{on whom the head was of stone}{\Bfootnote{Cf.~\textlink{Hymiskvida}[29]--30, where the ettin Hymer’s head is harder than stone.  This characteristic part of ettin-physiology can probably be explained by reference to Germanic cosmology.
In numerous Indo-European cosmologies the Firmament is believed to be made of stone, as is seen in the PIE root \emph{*h₂éḱmō} whose descendants can mean both ‘heaven, sky’ and ‘stone’, sometimes varying even within languages \parencites[342]{West2007}[3]{Calin1996}; cf.~e.g. Sanskrit \emph{áşman} ‘stone, rock’ with Old Persian \emph{asman} ‘sky, heaven’ and ON \emph{himinn} ‘sky, heaven’ with the derivative \emph{hamarr} ‘cliff, rock’ \parencite[220,206--207]{Kroonen2013}.
In the Germanic cosmology this “Stoney Heaven” was originally the skull of \inx[P]{Yimer}, the primordial ettin sacrificed by the Gods (\textlink{Voluspa}[3]--4, \textlink{Vafthrudnismal}[21], \textlink{Grimnismal}[41]), and as the ancestor of the Ettins he thus passed his stone-skull on to his descendants.}}, &
þó lét’k hann \alst{f}alla \hld\ ok \alst{f}yrir hníga; &
\ind \alst{h}vat vannt-u þȧ meðan, \alst{H}ár·barðr?“\eva

\bvb%
“This wilt thou now bring up, when I and Rungner dealt with each other, \\
that great-minded ettin on whom the head was of stone. \\
Yet I made him fall and kneel down before me--- \\
\ind what didst thou accomplish in the meantime, Hoarbeard?”\evb\evg


\bvg\bva\mssnote{\Regius~12v/27}%
\speakernote{[Hár·barðr kvað:]}%
„Vas’k með \alst{F}jǫl·vari \hld\ \alst{f}imm vetr alla &
ï \alst{ęy} þęiri \hld\ es \alst{A}l-grø̇n hęitir; &
\alst{v}ega vér þar knǫ́ttum \hld\ ok \alst{v}al fęlla, &
\edtrans{\alst{m}args at fręista, \hld\ \alst{m}ans at kosta}{many a girl did we tempt and win}{\Bfootnote{I read \emph{margs} ‘many a’ as modifying \emph{mans} ‘girl’.}}.“\eva

\bvb%
“I was with Felwar for five winters all \\
in that island which is called Allgreen. \\
There we did fight and fell the slain, \\
many a girl did we tempt and win.”\evb\evg


\bvg\bva\mssnote{\Regius~12v/30}%
\speakernote{[Þórr kvað:]}%
„Hversu snúnuðu yðr konur yðrar?“\eva

\bvb%
“How did your women turn out for you?”\evb\evg


\bvg\bva\mssnote{\Regius~12v/30}%
\speakernote{[Hár·barðr kvað:]}%
„\alst{Sp}arkar ǫ́ttum vér konur \hld\ ef oss at \alst{sp}ǫkum yrði; &
\alst{h}orskar ǫ́ttum vér konur \hld\ ef oss \alst{h}ollar vę́ri, &
\edtext{þę́r ór \alst{s}andi \hld\ \alst{s}íma undu &
\ind ok ór \alst{d}ali \alst{d}júpum &
\ind \alst{g}rund of \alst{g}rófu}{\lemma{þę́r ór sandi \dots\ of grófu ‘Out of the sand \dots\ the ground.’}\Bfootnote{TODO: Very unclear.}}; &
varð’k þęim ęinn \alst{ǫ}llum \hld\ \alst{ø}fri at rǫ́ðum; &
\ind hvílda’k hjá \alst{s}ystrum \alst{s}jau &
\ind ok hafða’k \alst{g}ęð þęira allt ok \alst{g}aman; &
\ind hvat vannt-u \alst{þ}ȧ meðan, \alst{Þ}ȯrr?“\eva

\bvb%
“We had smart women if they were to our enjoyment; \\
we had wise women if they were \inx[C]{hold} toward us. \\
Out of the sand they unwound a rope, \\
\ind and out of a deep dale \\
\ind dug up the ground. \\
I alone became superior to them all in counsel; \\
\ind I rested beside those sisters seven \\
\ind and had their senses all, and pleasure--- \\
\ind what didst thou accomplish in the meantime, Thunder?”\evb\evg


\bvg\bva\mssnote{\Regius~13r/2, \AM~1r/1}%
\Ballnote{For the slaying of Thedse cf.~\textlink{Lokasenna}[50] n.  In this stanza we seem to have a rare example of native Germanic star-lore.  TODO: is the exact constellation identifiable?  Speculation time!}%
\speakernote{Þórr kvað:}%
„Ek drap \alst{Þ}jatsa, \hld\ hinn \alst{þ}rúð-móðga jǫtun, &
\alst{u}pp ek varp \alst{au}gum \hld\ \alst{A}ll·valda sonar &
\ind ȧ þann inn \alst{h}ęiða \alst{h}imin; &
þau ’ro \alst{m}ęrki \alst{m}ęst \hld\ \edtext{\alst{m}inna}{\Afootnote{From this word forwards the poem is also attested in \AM.}} verka, &
\ind þau’s allir męnn \edtext{\alst{s}íðan}{\Afootnote{om. \AM}} of \alst{s}é\emph{a}; &
\ind \alst{h}vat vannt-u þȧ meðan, \alst{H}ár·barðr?“\eva

\bvb%
“I slew \inx[C]{Thedse}, the strength-minded ettin; \\
I threw up the eyes of Allwald’s son \ken*{= Thedse} \\
\ind onto the clear heaven. \\
Those are the greatest marks of my works, \\
\ind those which all men since do see--- \\
\ind what didst thou accomplish in the meantime, Hoarbeard?”\evb\evg


\bvg\bva\mssnote{\Regius~13r/5, \AM~1r/1}%
\speakernote{Hár·barðr kvað:}%
„\alst{M}iklar \alst{m}an-vélar \hld\ hafða’k við \edtrans{\alst{m}yrk-riður}{mirk-rideresses}{\Bfootnote{The \emph{riður} ‘rideresses, fevers’ were witches thought to torment people and cause disease and delirium.  See \textlink{Havamal}[156] for discussion.}} &
\ind þȧ’s ek \alst{v}élta þę́r \edtrans{frȧ \alst{v}erum}{from men}{\Bfootnote{Alternatively ‘from their husbands’, but that makes less sense in the context.  For \emph{verar} ‘men, husbands’ in the general sense ‘mankind’ cf.~st. 53/3 below, \textlink{Lokasenna}[24]/3.}}. &
\alst{H}arðan jǫtun \hld\ hugða’k \alst{H}lé·barð vesa; &
\ind \alst{g}af hann mér \edtrans{\alst{g}amban-tęin}{gombentoe}{\Bfootnote{An unknown magical object also occuring in \textlink{Skirnismal}[32]/2.  Etymologically it is clearly some type of stick (\emph{tęinn} ‘stick, branch’), perhaps for carving runes into.}} &
\ind ęn ek \alst{v}élta hann ór \alst{v}iti.“\eva

\bvb%
“Great girl-tricks I had against \inx[C]{mirk-rideresses}, \\
\ind when I lured them away from men. \\
A hard ettin I judged Leebeard to be; \\
\ind he gave me a \inx[C]{gombentoe}, \\
\ind but I tricked him out of his wits.”\evb\evg


\bvg\bva\mssnote{\Regius~13r/7, \AM~1r/3}%
\speakernote{Þórr kvað:}%
„Illum huga launaðir þú \edtext{þá}{\Afootnote{om. \AM}} \alst{g}óðar \alst{g}jafar.“\eva

\bvb%
“With an evil heart didst thou then repay the good gifts.”\evb\evg


\bvg\bva\mssnote{\Regius~13r/8, \AM~1r/4}%
\speakernote{Hár·barðr kvað:}%
„\edtrans{Þat hęfir \alst{ęi}k \hld\ es af \alst{a}nnarri skęfr}{The oak has that which it chafes from the other}{\Bfootnote{Proverbial, also appearing in \Malshattakvadi\ 26/3a.}}; &
\ind umb \alst{s}ik es hvęrr ï \alst{s}líku--- &
\ind hvat vannt-u \alst{þ}ȧ meðan, \alst{Þ}ȯrr?“\eva

\bvb%
“The oak has that which it chafes from the other; \\
\ind it is every man for himself in such matters--- \\
\ind what didst thou accomplish in the meantime, Thunder?”\evb\evg


\bvg\bva\mssnote{\Regius~13r/9, \AM~1r/4}%
\Ballnote{Thunder here makes a moral response to Hoarbeard’s egoism as expressed in the previous stanza.  In the religion, Thunder is the defender of the Gods and Middenyard (the home of Men) against the Ettins, for which cf.~\textlink{Voluspa}[25], 53, \textlink{Thrymskvida}[18].  For Thunder’s killing of ettin-women in particular cf.~sts. 37--39 below, \textlink{EddicFragments}[F6][6], and \textcite{Lindow1988}.}%
\speakernote{Þórr kvað:}%
„\alst{E}k vas \alst{au}str \hld\ ok \alst{jǫ}tna barða’k &
\alst{b}rúðir \alst{b}ǫl-vísar \hld\ es til \alst{b}jargs gingu; &
mikil myndi \alst{ę́}tt \alst{jǫ}tna \hld\ ef \alst{a}llir lifði, &
vę́tr myndi \alst{m}anna \hld\ undir \alst{M}ið-garði--- &
\ind \alst{h}vat vannt-u þȧ meðan, \alst{H}ár·barðr?\eva

\bvb%
“I was in the east and bashed Ettins, \\
bale-wise brides who walked to the mountain. \\
Great would the line of Ettins be if they all lived, \\
naught would remain of men within Middenyard--- \\
\ind what didst thou accomplish in the meantime, Hoarbeard?”\evb\evg


\bvg\bva\mssnote{\Regius~13r/11, \AM~1r/6}%
\Ballnote{Weden brushes off Thunder’s moralism and instead expresses a characteristic aristocratic disregard for lower life and life as mere life.  Where the altruistic Thunder boasts of protecting mankind, the ambivalent Weden sarcastically responds that he incited war and conflict in order to take the best of them for himself.}%
\speakernote{Hár·barðr kvað:}%
„\alst{V}as’k ȧ \alst{V}allandi \hld\ ok \alst{v}ígum fylgða’k, &
\alst{a}tta ek \alst{jǫ}frum \hld\ ęn \alst{a}ldri sę́tta’k; &
\edtrans{\alst{Ó}ðinn á}{Weden owns}{\Bfootnote{For Weden’s “owning” of men see \textlink{Voluspa}[23] n.}} \alst{ja}rla \hld\ þȧ’s ï \alst{v}al falla &
\ind ęn \edtrans{\alst{Þ}ȯrr á \alst{þ}rę́la kyn}{Thunder owns the race of thralls}{\Bfootnote{This has been connected (TODO: cite) with the sacrifice of prisoners and criminals to Thunder, which if true would support a pre-Christian origin for the poem since such an allusion would soon fade out of relevance following conversion.}}.“\eva

\bvb%
“I was in \inx[P]{Walland} and followed wars; \\
I incited princes and never reconciled them. \\
Weden owns the earls which fall among the slain, \\
\ind but Thunder owns the race of thralls.”\evb\evg


\bvg\bva\mssnote{\Regius~13r/13, \AM~1r/8}%
\speakernote{Þórr kvað:}%
„\alst{Ȯ}·jafnt skipta \hld\ es þú myndir með \edtext{\alst{ǫ̇}sum}{\Afootnote{\emph{ása} \AM}} liði &
\ind ef þú ę́ttir \alst{v}il-gi mikils \alst{v}ald.“\eva

\bvb%
“Unfairly wouldst thou deal out troops amidst the Eese \\
\ind if thou hadst unrestrained power.”\evg


\bvg\bva\mssnote{\Regius~13r/14, \AM~1r/9}%
\Ballnote{The story alluded to in this st.~is also mentioned in \textlink{Lokasenna}[60] and told in full in \Gylfaginning\ 45: Lock, Thunder, and his servants Thelve and Wrash had journeyed east for a long time when they came upon a large hall, with an opening on one end as wide as the building.  They rested inside, but in the middle of the night they were awakened by a great earthquake.  Thunder rose and led the party into a side-room to the right in the middle of the hall.  He stood closest to the opening with his hammer ready, while the terrified others were further inside.  At daybreak they left the hall and found the huge ettin \emph{Skrymir} (\inx[P]{Shrimer}) asleep outside.  His snoring had caused the earthquakes, and the hall was his mitten; the side-room was its thumb.}%
\speakernote{Hár·barðr kvað:}%
„Þȯrr á \alst{a}fl \alst{ǿ}rit \hld\ ęn \alst{ę}kki hjarta; &
af \alst{h}rę́ðslu ok \alst{h}ug-blęyði \hld\ \edtext{vas þér}{\Afootnote{\emph{þér vas} \Regius}} ï \alst{h}andska troðit &
\ind ok \alst{þ}ȯtti⸗sk⸗a þú \alst{þ}ȧ \alst{Þ}ȯrr vesa; &
\alst{h}vár⸗ki þú þá þorðir \hld\ fyr \alst{h}rę́ðslu þïnni &
\edtrans{hnjósa né \alst{f}ísa}{sneeze or fart}{\Afootnote{\emph{físa né hnjósa} ‘fart or sneeze’ \AM}} \hld\ svá’t \alst{F}jalarr hęyrði.“\eva

\bvb%
“Thunder has ample strength, but little heart. \\
Out of fear and heart-softness wast thou driven into a glove, \\
\ind and then seemedst thou not to be Thunder. \\
Thou daredst neither for thy fear \\
sneeze nor fart lest Feller should hear.”\evb\evg


\bvg\bva\mssnote{\Regius~13r/17, \AM~1r/11}%
\speakernote{Þórr kvað:}%
„\alst{H}ár·barðr inn ragi, \hld\ ek munda þik ï \alst{h}ęl drepa &
\ind ef ek mę́tta \alst{s}ęila⸗sk of \edtext{\alst{s}und}{\Afootnote{\emph{sundit} \AM}}.“\eva

\bvb%
“Hoarbeard the \inx[C]{queer}! I would strike thee into \inx[P]{Hell}, \\
\ind if I could sail o’er the sound!”\evb\evg


\bvg\bva\mssnote{\Regius~13r/18, \AM~1r/12}%
\speakernote{Hár·barðr kvað:}%
„Hvat \edtext{skyldir}{\Afootnote{\emph{skalt-u} \AM}} of \alst{s}und \alst{s}ęila⸗sk \hld\ es \edtext{\alst{s}akir}{\Afootnote{\emph{sakar} \AM}} ’ro alls øngar? &
\ind hvat vannt-u \alst{þ}ȧ meðan, \alst{Þ}ȯrr?“\eva

\bvb%
“Why shouldst thou sail o’er the sound when there are no charges at all? \\
\ind What didst thou accomplish in the meantime, Thunder?”\evb\evg


\bvg\bva\mssnote{\Regius~13r/19, \AM~1r/13}%
\speakernote{Þórr kvað:}%
„\alst{E}k vas \alst{au}str \hld\ ok \alst{ǫ́}’na varða’k &
\ind þȧ’s \edtext{mik \alst{s}óttu þęir}{\Afootnote{\emph{þęir sóttu mik} \AM}} \alst{S}várangs \alst{s}ynir; &
\alst{g}rjóti mik bǫrðu, \hld\ \alst{g}agni urðu \edtext{þó}{\Afootnote{om. \AM}} lítt fęgnir, &
þó urðu mik \alst{f}yrri \hld\ \alst{f}riðar at biðja--- &
\ind \alst{h}vat vannt-u þȧ meðan, \alst{H}ár·barðr?“\eva

\bvb%
“I was in the east and guarded the river \\
\ind when I was set upon by Sweering’s sons. \\
With rocks they bashed me, yet in victory rejoiced little; \\
yet they soon had to beg me for peace--- \\
\ind what didst thou accomplish in the meantime, Hoarbeard?”\evb\evg


\bvg\bva\mssnote{\Regius~13r/22, \AM~1r/15}%
\speakernote{Hár·barðr kvað:}%
„\alst{E}k vas \alst{au}str \hld\ ok við \edtext{\alst{ęi}n-hvęrja}{\Afootnote{\emph{‘æinhæriu’} \AM}} dǿmða’k, &
\alst{l}ék’k við ina \alst{l}ind-hvítu \hld\ ok \edtrans{\alst{l}aun-þing}{secret trysts}{\Afootnote{so \AM; \emph{laung þing} ‘long trysts’ \Regius}} háða’k, &
\alst{g}ladda’k ina \edtrans{\alst{g}oll-bjǫrtu}{gold-bright}{\Afootnote{\emph{goll-hvítu} ‘gold-white’ \AM}}, \hld\ \alst{g}amni mę́r unði.“\eva

\bvb%
“I was in the east and flirted with a certain someone; \\
I played with the linen-white one and held secret trysts: \\
I gladdened the gold-bright one---the maiden enjoyed pleasure.”\evb\evg


\bvg\bva\mssnote{\Regius~13r/24, \AM~1r/17}%
\speakernote{Þórr kvað:}%
„\edtext{Góð ǫ́ttu þęir man-kynni þar þȧ.}{\Bfootnote{Clearly prose.}}“\eva

\bvb%
“Then they had good girls there.”\evb\evg


\bvg\bva\mssnote{\Regius~13r/24, \AM~1r/17}%
\speakernote{Hár·barðr kvað:}%
„\alst{L}iðs þïns \edtext{vę́ra’k}{\Afootnote{\emph{vas’k} \AM}} þȧ þurfi, Þȯrr, \hld\ at ek hęlda þęiri inni \alst{l}ín-hvítu męy.“\eva

\bvb%
“I might have needed thy aid then, Thunder, that I might hold that linen-white maiden.”\evb\evg


\bvg\bva\mssnote{\Regius~13r/25, \AM~1r/18}%
\speakernote{Þórr kvað:}%
„Ek mynda þér \edtext{þȧ þat}{\Afootnote{\emph{þat þá} \AM}} \alst{v}ęita \hld\ ef ek \alst{v}iðr of \edtext{kǿm⸗umk}{\Afootnote{so \AM\ (\emph{‘kæmvmz’}); \emph{kǿmist} \Regius}}.“\eva

\bvb%
“I would then have given it to thee, if I were able.”\evb\evg


\bvg\bva\mssnote{\Regius~13r/26, \AM~1r/18}%
\speakernote{Hár·barðr kvað:}%
„Ek mynda þér þá \alst{t}rúa, \hld\ nema mik ï \alst{t}ryggð véltir.“\eva

\bvb%
“I would then have trusted thee, unless thou wouldst betray my trust.”\evb\evg


\bvg\bva\mssnote{\Regius~13r/27, \AM~1r/19}%
\speakernote{Þórr kvað:}%
„\edtrans{Em’k⸗at ek sá \alst{h}ę̇l-bítr \hld\ sem \alst{h}úð-skór forn ȧ vár}{I am not such a heel-biter as an old hide-shoe in spring}{\Bfootnote{Likely proverbial and a pun; a heel-biter is someone who betrays his companions, and the old shoe-leather becomes stiff over the winter.}}.“\eva

\bvb%
“I am not such a heel-biter as an old hide-shoe in spring.”\evb\evg


\bvg\bva\mssnote{\Regius~13r/28, \AM~1r/20}%
\speakernote{Hár·barðr kvað:}%
„Hvat vannt-u \alst{þ}ȧ meðan, \alst{Þ}ȯrr?“\eva

\bvb%
“What didst thou accomplish in the meantime, Thunder?”\evb\evg


\bvg\bva\mssnote{\Regius~13r/28, \AM~1r/20}%
\speakernote{Þórr kvað:}%
„\alst{B}rúðir \alst{b}er-sęrkja \hld\ \alst{b}arða’k ï \edtext{Hlés-ęyju}{\Afootnote{\emph{Hlés-ęy} \AM}}; &
þę́r hǫfðu \alst{v}ęrst unnit, \hld \alst{v}élta þjóð alla.“\eva

\bvb%
“The brides of bearserks I bashed in Leesey; \\
they had done the worst thing: tricked the whole people.”\evb\evg


\bvg\bva\mssnote{\Regius~13r/29, \AM~1r/21}%
\speakernote{Hár·barðr kvað:}%
„\alst{K}lę́ki vannt-u þá, Þórr, \hld\ es þú \edtext{ȧ}{\Afootnote{\emph{‘ǽ’} with small corr. above \AM}} \alst{k}onum barðir.“\eva

\bvb%
“A disgrace didst thou accomplish, Thunder, when thou bashedest women!”\evb\evg


\bvg\bva\mssnote{\Regius~13r/30, \AM~1r/22}%
\speakernote{Þórr kvað:}%
„\alst{V}argynjur \edtext{vǫ́ru þę́r}{\Afootnote{\emph{þat vǫ́ru} \AM}} \hld\ ęn \alst{v}ar⸗la konur, &
\alst{sk}ęlldu \alst{sk}ip mitt \hld\ es \alst{sk}orðat hafða’k, &
\alst{ǿ}gðu \edtext{mér}{\Afootnote{add. \emph{þęim} \AM}} \alst{já}rn-lurki \hld\ ęn \alst{ę}ltu Þjálfa--- &
\ind \alst{h}vat vannt-u þȧ meðan, \alst{H}ár·barðr?“\eva

\bvb%
“She-wolves were they, but hardly women; \\
they overturned my ship which I had propped, \\
terrorised me with iron cudgels and chased \inx[P]{Thelve} around--- \\
\ind what didst thou accomplish in the meantime, Hoarbeard?”\evb\evg


\bvg\bva\mssnote{\Regius~13r/32, \AM~1r/23}%
\speakernote{Hár·barðr kvað:}%
„Ek vas’k ï \alst{h}ęr’num \hld\ es \alst{h}ingat gørði⸗sk &
\alst{g}nę́fa \alst{g}unn-fana, \hld\ \alst{g}ęir at rjóða.“\eva

\bvb%
“I was in the warband when it readied itself hither \\
to raise the war-standard, to redden the spear.”\evb\evg


\bvg\bva\mssnote{\Regius~13v/1, \AM~1r/24}%
\Ballnote{The exchange shifts away from the comparison of deeds.}%
\speakernote{Þórr kvað:}%
„Þęss vilt-u nú geta, es þú fórt oss \edtext{ȯ·ljúfan}{\Afootnote{\emph{‘óliyfan’} \AM; \emph{†olubann†} \Regius}\lemma{ȯ·ljúfan ‘enmity’}\Bfootnote{Lit.~‘un-lovely’, a rare m.~dat.~sg.~adverbial.}} at bjóða!“\eva

\bvb%
“This wilt thou now bring up, that thou didst set out to bring enmity against us!”\evb\evg


\bvg\bva\mssnote{\Regius~13v/2, \AM~1r/25}%
\speakernote{Hár·barðr kvað:}%
„\alst{B}ǿta skal þér \edtext{þat þá}{\Afootnote{om. \AM}} \hld\ munda \alst{b}augi &
sem \alst{ja}fnęndr \alst{u}nnu \hld\ \edtext{þęir’s \alst{o}kkr vilja sę́tta}{\Afootnote{\emph{þęir’s okkr vilja sę́tt hafa} \AM}}.“\eva

\bvb%
“Then I shall repay thee for that with a hand-bigh, \\
bestowed by the mediators who wish to reconcile us two.”\evb\evg


\bvg\bva\mssnote{\Regius~13v/3, \AM~1r/26}%
\speakernote{Þórr kvað:}%
„\alst{H}var namt þęssi \hld\ in \alst{h}nǿfi⸗ligu orð &
es \alst{h}ęyrða’k aldri⸗gi \hld\ \edtext{in}{\Afootnote{so \AM; om. \Regius}} \alst{h}nǿfi⸗ligri?“\eva

\bvb%
“Where didst thou learn these sarcastic words, \\
when I never heard more sarcastic ones?”\evb\evg


\bvg\bva\mssnote{\Regius~13v/5, \AM~1r/27}%
\speakernote{Hár·barðr kvað:}%
„Nam’k at \edtext{mǫnnum}{\Afootnote{om. \AM}} þęim inum aldr-ǿnum es búa í hęimis-skógum.“\eva

\bvb%
“I learned them from the elderly men who dwell in the homely forests.”\evb\evg


\bvg\bva\mssnote{\Regius~13v/5, \AM~1v/1}%
\speakernote{Þórr kvað:}%
„Þó gefr þú gótt nafn \edtrans{dysjum}{poor cairns}{\Bfootnote{A reference to Weden’s consultation of the dead for knowledge as attested e.g. in \textlink{Voluspa} and \textlink{Baldrsdraumar}.}}, es þú kallar þat hęimis-skóga.“\eva

\bvb%
“Though thou givest a good name to poor cairns when thou callest them ‘homely forests’.”\evb\evg


\bvg\bva\mssnote{\Regius~13v/6, \AM~1v/2}%
\speakernote{Hár·barðr kvað:}%
„Svá dǿmi’k \edtext{umb}{\Afootnote{om. \AM}} slíkt far.“\eva

\bvb%
“So do I speak about such matters.”\evb\evg


\bvg\bva\mssnote{\Regius~13v/7, \AM~1v/2}%
\Ballnote{This perfectly regular \Ljodahattr-stanza marks a sudden shift in the exchange from the previous laconic one-line prose lines.  Thunder’s threat of hammer-based violence is parallelled by \textlink{Lokasenna}[57]--63.}%
\speakernote{Þórr kvað:}%
„\alst{O}rð-kringi þïn \hld\ mun þér \alst{i}lla koma &
\ind ef ek rę́ð ȧ \alst{v}ág at \alst{v}aða; &
\alst{u}lfi hę́rra \hld\ hygg’k \edtext{at \alst{ǿ}pa mynir}{\Afootnote{\emph{þik ǿpa munu} \AM}} &
\ind ef \alst{h}lýtr af \alst{h}amri \alst{h}ǫgg.“\eva

\bvb%
“Thy glibness of word will bring thee ill \\
\ind if I decide to wade on the wave. \\
Louder than a wolf I think thou wilt scream \\
\ind if thou gettest a strike from the hammer!”\evb\evg


\bvg\bva\mssnote{\Regius~13v/9, \AM~1v/4}%
\speakernote{Hár·barðr kvað:}%
„Sif á \edtrans{\alst{h}ó}{lover}{\Bfootnote{Most editors take this acc.~sg.~word as an alternative form of \emph{hórr} m. ‘adulterer’ (gen. \emph{hórs}), containing the same root as \emph{hóra} f. ‘whore, prostitute’, \emph{hór} n. ‘adultery, fornication’, ModEngl. whore.  The \emph{-r} has presumably been reinterpreted as the m.~nom.~sg.~ending, yielding nom.~sg.~\emph{*hór}, gen.~sg.~\emph{*hós}, although neither form is attested.  The same accusation is found in \textlink{Lokasenna}[54], where Lock says that he has been Sib’s lover (\emph{hórr}).

On formal grounds \textciteshorthand[s.v. \textit{hór}]{CleasbyVigfusson} interpret this word as the unrelated \emph{hór} m. ‘pot-hook’, “insinuating that Thor busied himself with cooking and dairy-work.”  If that reading is correct \emph{hans, þann} ‘him’ should be translated ‘it’ instead, but the whole thing seems rather unlikely when we consider Thunder’s strong response: ‘I think thou liest’, which fits well against an accusation of his wife’s adultery, not so much about her owning a pot-hook.}} \alst{h}ęima, \hld\ \alst{h}ans munt fund vilja, &
\alst{þ}ann munt \alst{þ}ręk drýgja, \hld\ \alst{þ}at ’s þér \edtext{skyldara}{\Afootnote{\emph{skyldra} \AM}}.“\eva

\bvb%
“Sib has a lover at home; \emph{him} wilt thou wish to seek out; \\
on \emph{him} wilt thou use thy strength---that is more urgent for thee!”\evb\evg


\bvg\bva\mssnote{\Regius~13v/10, \AM~1v/5}%
\speakernote{Þórr kvað:}%
„\alst{M}ę́lir þú at \alst{m}unns ráði \hld\ svá’t \alst{m}ér skyldi vęrst þikkja, &
\alst{h}alr inn \alst{h}ug-blauði, \hld\ \alst{h}ygg’k at þú ljúgir.“\eva

\bvb%
“Thou speakest after thy mouth’s counsel what should seem worst to me--- \\
O heart-soft man, I think thou liest!”\evb\evg


\bvg\bva\mssnote{\Regius~13v/12, \AM~1v/6}%
\speakernote{Hár·barðr kvað:}%
„\alst{S}att hygg’k \edtext{mik}{\Afootnote{\emph{þik} \AM}} \alst{s}ęgja, \hld\ \alst{s}ęinn est at fǫr þïnni, &
\alst{l}angt myndir nú kominn, Þórr, \hld\ ef þú \edtrans{\alst{l}itum fǿrir}{changed colour}{\Bfootnote{An unclear expression, probably metaphorically ‘if thou hadst not been so stubborn’.}}.“\eva

\bvb%
“I think I speak truly.  Thou art late on thy journey; \\
far wouldst thou have come by now, Thunder, if thou hadst changed colour.”\evb\evg


\bvg\bva\mssnote{\Regius~13v/14, \AM~1v/8}%
\speakernote{Þórr kvað:}%
„\alst{H}ár·barðr inn ragi, \hld\ \alst{h}ęldr hęfir nú mik \edtext{dvalðan}{\Afootnote{\emph{dvalit} \AM}}!“\eva

\bvb%
“Hoarbeard the queer! Thou hast now greatly delayed me!”\evb\evg


\bvg\bva\mssnote{\Regius~13v/14, \AM~1v/8}%
\speakernote{Hár·barðr kvað:}%
„\edtext{\alst{Ȧ}sa-Þȯrs}{\Afootnote{\emph{Ȧsa-Þȯr} \AM}\lemma{Ȧsa-Þȯrs ‘Thunder of the Eese’}\Bfootnote{A reverent, majestic epithet, here used sarcastically, i.e.~‘the high and mighty Thunder’.}} \hld\ hugða’k \alst{a}ldri⸗gi myndu &
\ind glępja \alst{f}é-hirði \alst{f}arar.“\eva

\bvb%
“The journey of Thunder of the Eese I never thought \\
\ind a shepherd would divert.”\evb\evg


\bvg\bva\mssnote{\Regius~13v/15, \AM~1v/9}%
\speakernote{Þórr kvað:}%
„\alst{R}áð mun’k þér nú \alst{r}áða: \hld\ \alst{r}ó hingat báti’num, &
\alst{h}ę̇ttum \alst{h}ǿtingi, \hld\ \alst{h}itt fǫður Magna!“\eva

\bvb%
“I will now counsel thee counsel!  Row the boat hither; \\
let us cease the taunting; meet the father of Main \ken*{= Thunder = me}!”\evb\evg


\bvg\bva\mssnote{\Regius~13v/17, \AM~1v/10}%
\speakernote{Hár·barðr kvað:}%
„\alst{F}ar þú \edtext{\alst{f}irr}{\Afootnote{\emph{frá} \AM}} sundi, \hld\ þér skal \alst{f}ars synja!“\eva

\bvb%
“Go far away from the sound; passage shall be denied thee!”\evb\evg


\bvg\bva\mssnote{\Regius~13v/17, \AM~1v/11}%
\speakernote{Þórr kvað:}%
„\alst{V}ísa þú mér \edtext{nú}{\Afootnote{om. \AM}} lęið’ina, \hld\ alls þú vill mik ęigi of \alst{v}ág’inn fęrja.“\eva

\bvb%
“Show me now the way, as thou wilt not ferry me o’er the wave.”\evb\evg


\bvg\bva\mssnote{\Regius~13v/18, \AM~1v/11}%
\speakernote{Hár·barðr kvað:}%
„\alst{L}ítit ’s \edtext{at}{\Afootnote{om. \Regius}} synja, \hld\ \alst{l}angt ’s at fara; &
\alst{st}und ’s til \edtext{\alst{st}okks’ins}{\Afootnote{\emph{stokks} \AM}}, \hld\ ǫnnur til \edtext{\alst{st}ęins’ins}{\Afootnote{\emph{stęins} \AM}}, &
halt svá til \alst{v}instra \edtext{\alst{v}egs’ins}{\Afootnote{\emph{vegs} \AM}} \hld\ und’s þú hittir \edtrans{\alst{V}er-land}{Wereland}{\Afootnote{\emph{Valland} \AM}\Bfootnote{The ‘Land of Men’.}}; &
\alst{þ}ar mun \edtrans{Fjǫrgyn}{Firgyn}{\Bfootnote{Another name for Earth, Thunder’s mother.}} \hld\ hitta \alst{Þ}ȯr, son sïnn, &
ok mun hǫ̇n kęnna hǫ̇num \alst{ǫ́}ttunga brautir \hld\ til \alst{Ó}ðins landa.“\eva

\bvb%
“It is a little thing to deny. It is long to journey: \\
an hour to the log, another to the stone; \\
hold so to the left road until thou findest Wereland. \\
There will Firgyn find Thunder, her son, \\
and she will show him the ancestral roads to Weden’s lands \ken*{= Osyard}.”\evb\evg


\bvg\bva\mssnote{\Regius~13v/22, \AM~1v/14}%
\speakernote{Þórr kvað:}%
„Mun’k taka þangat \edtext{ï dag}{\Afootnote{\emph{ȧ dęgi} \AM}}?“\eva

\bvb%
“Will I get thither today?”\evb\evg


\bvg\bva\mssnote{\Regius~13v/22, \AM~1v/14}%
\speakernote{Hár·barðr kvað:}%
„Taka við víl \edtext{ok}{\Afootnote{\emph{við} \AM}} \alst{ę}rfiði \hld\ at \edtext{\alst{u}pp-vesandi}{\Afootnote{\emph{upp-rennandi} \AM}} sólu &
\ind es ek get þána.“\eva

\bvb%
“[Thou wilt] get thither with toil and hardship by the rising of the sun, \\
\ind as I believe it is thawing.”\evb\evg


\bvg\bva\mssnote{\Regius~13v/23, \AM~1v/15}%
\speakernote{Þórr kvað:}
„\alst{Sk}ammt mun nú mál okkat vesa, \hld\ alls þú mér \alst{sk}ǿtingu ęinni svarar; &
launa mun’k þér \alst{f}ar-synjun \hld\ ef vit \alst{f}innum⸗sk ï sinn annat.“\eva

\bvb%
“Short will now our speech be since thou answerest me with scoffing alone. \\
I will reward thee for this ferry-denial if we meet another time!”\evb\evg


\bvg\bva\mssnote{\Regius~13v/25, \AM~1v/17}%
\speakernote{Hár·barðr kvað:}%
\Ballnote{The speaker is uncertain.  This line is introduced with the speech marker \emph{‘H. q.’} (= \emph{Hár·barðr kvað} ‘Hoarbeard quoth’) in \AM; in \Regius\ it belongs to 59 and is thus spoken by Thunder.  Since it seems more likely for a speech marker to be dropped than erroneously added the \AM\ reading is preferred.}%
„Far þú nú þar’s \edtrans{þik hafi allan gramir}{the fiends may have thee whole}{\Bfootnote{The \emph{gramir} ‘fiends’, lit. ‘wroth, cross ones’ are (male) dæmons.  The adjective \emph{gramr} is commonly used specifically with a connotation of divine wrath and they may thus be supernatural avengers for severe crimes.  The same curse is found in \textlink{Grimnismal}[P1] with \emph{smyl} for \emph{gramir} and in related form in \textlink{Brot}[10].}}!“\eva

\bvb%
“Go now whither the fiends may have thee whole!”\evb\evg

\sectionline
