\bookStart{Speeches of Shirner}[Skírnis mǫ́l]
\setBookCode{Skirnismal}

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: C10th (0.897)

\textbf{Meter:} \Ljodahattr, \Galdralag\ (TODO)%
\end{flushright}

\section{Introduction}

The \textbf{Speeches of Shirner} (\textlink{Skirnismal}) is a mythological dialogue-poem dealing with Free’s servant Shirner who is sent to get the hand of the beautiful ettin’s daughter Gird for his master by any means necessary.

\subsection{Preservation}

\Skirnismal\ is attested in full in both \Regius\ and \AM.  The name \emph{Skírnis-mǫ́l} ‘Speeches of Shirner’ comes from \AM; \Regius\ instead has \emph{Fǫr Skírnis} ‘Shirner’s journey’.

\subsection{Parallels}

The whole of \Skirnismal\ is paraphrased in \Gylfaginning~37, for comparative purposes edited in full in the Appendix to this poem.  \Gylfaginning~37:15 cites a variant form of \textlink{Skirnismal}[42], the final stanza.
It is also of note that the prose of \Skirnismal\ P1 and \Gylfaginning~37:1--6 share striking similiarities in certain expressions like Shirner being a \emph{skó-svęinn} ‘shoe-swain’ or Free one day having sat down in Lithshelf and looked about all the homes.

It seems near-certain that Snorre, the author of \Gylfaginning, had access to \textlink{Skirnismal} directly rather than a prose retelling of the story accompanied by a solitary stanza.
The most important reason for this assumption is that there is no detail in Snorre’s paraphrase that is not found in the version of the poem in \Regius\AM, although there are many things found in \Regius\AM\ not found in \Gylfaginning, most notably Shirner’s curse.  Snorre’s prose however has many more details.

These circumstances are perhaps best explained if the version of \textlink{Skirnismal} underlying \Gylfaginning~37 was available to Snorre based on oral tradition rather than a manuscript.  This would explain the small variation in the final stanza and the greater variation in the introductory prose; whereas poetry is a bound form and less liable to drastic change, a prose narrative can vary greatly with each retelling.
On the other hand it is also possible that Snorre chose to expand a preexisting written prose passage by adding his own details, something likely done to the prose of \textlink{Lokasenna}[P8] (see there) and then supplied the stanza with its minor variants based on his own memory.  To sum up a narrative mythic poem in prose form and then quote one or two stanzas from it is something likely done elsewhere in \Gylfaginning; see \textlink{EddicFragments} below.

\subsection{Interpretations}

TODO.

Fertility? Free gives away his sword for sex (st.~8).

\newpage

\section{Text}

\subsection{The Speeches of Shirner}

\bpg\bpa\mssnote{\Regius~11r/10, \AM~2r/11}%
Freyr, sonr Njarðar, hafði einn dag set⸗sk í \edtrans{Hlið-skjálf}{Lithshelf}{\Bfootnote{The high lookout-post of the Gods.}} ok sá um heima alla; hann sá í Jǫtun-heima ok sá þar mey fagra, þá er hón gekk frá skála fǫður síns til skemmu; þar af fekk hann hug-sóttir miklar. Skírnir hét \edtrans{skó-sveinn}{shoe-swain}{\Bfootnote{A young boy who ties the shoes for his master.}} Freys.  Njǫrðr bað hann kveðja Frey máls. Þá mę́lti Skaði:\epa

\bpb \inx[P]{Free}[{\huge F}\textsc{ree}], son of \inx[P]{Nearth}, had one day sat down in the \inx[P]{Lithshelf}, and saw throughout all the \inx[C]{Homes}.  He looked into the \inx[P]{Ettinhomes} and saw there a fair maiden as she walked from her father’s hall to her bower; thereof he got great heart-aches.  \inx[P]{Shirner} was the name of Free’s shoe-swain; Nearth asked him to get Free to speak.  Then \inx[P]{Shede} spoke:\epb\epg


\bvg\bva\mssnote{\Regius~11r/14, \AM~2r/15}%
\llap{„}\edtext{Rís-tu nú Skírnir \hld\ ok gakk at bęiða}{\lemma{Rís-tu \dots\ bęiða ‘Rise \dots\ beg’}\Bfootnote{Alliteration is missing here. A simple solution would be to replace \emph{gakk} ‘go’ with a synonym like \emph{rinn} ‘run’ or \emph{ráð} ‘resolve’, but this lessens the semantic mirroring with l. 2/2 below (though the verb \emph{ganga} in the present stanza may in fact be due to influence from 2/2).}} &
\ind okkarn \alst{m}ála \alst{m}ǫg, &
ok þęss at \alst{f}regna \hld\ hvęim hinn \alst{f}róði séi &
\ind \alst{o}f-ręiði \edtrans{\alst{a}fi}{man}{\Bfootnote{While this word usually means ‘father’ or ‘grandfather’, it should here mean ‘man’ without a connotation of old age. See further \CV.}}.“\eva

\bvb%
“{\huge R}\textsc{ise now, Shirner}, and go to beg \\
\ind our lad for speech, \\
and to ask at whom the wise \\
\ind man might be cross.”\evb\evg


\bvg\bva\speakernote{Skírnir kvað:}\mssnote{\Regius~11r/15, \AM~2r/17}%
\llap{„}\alst{I}llra \alst{o}rða \hld\ es mér \alst{ȯ}n at ykkrum syni, &
\ind ef ek gęng at \alst{m}ę́la við \alst{m}ǫg, &
ok þęss at \alst{f}regna, \hld\ hvęim hinn \alst{f}róði séi &
\ind \alst{o}f-ręiði \alst{a}fi.“\eva

\bvb\speakernoteb{Shirner quoth:}%
“Ill words I expect from your son \\
\ind if I go to speak with the lad, \\
and to ask at whom the wise \\
\ind man might be cross.”\evb\evg


\bvg\bva\speakernote{Skírnir:}\mssnote{\Regius~11r/17, \AM~2r/18}%
\llap{„}Sęg þat \alst{F}ręyr, \hld\ \alst{f}olk-valdi goða, &
\ind ok ek \alst{v}ilja \alst{v}ita, &
hví þú \alst{ęi}nn sitr \hld\ \alst{ę}nd-langa sali, &
\ind mïnn \alst{d}róttinn, of \alst{d}aga?“\eva

\bvb\speakernoteb{Shirner [quoth]:}%
“Tell this, O Free, troop-wielder of the gods--- \\
\ind I too would wish to know \\
why thou sittest alone in the endlong halls, \\
\ind my lord, during the days.”\evb\evg


\bvg\bva\speakernote{Fręyr:}\mssnote{\Regius~11r/19, \AM~2r/20}%
\llap{„}Hví of \alst{s}ęgja’k þér, \hld\ \alst{s}ęggr hinn ungi, &
\ind \alst{m}ikinn \alst{m}óð-trega? &
því’t \edtrans{\alst{a}lf-rǫðull}{Elf-wheel}{\Bfootnote{A rare poetic synonym (\emph{hęiti}) for the sun; see note to \textlink{Vafthrudnismal}[47]/1.}} \hld\ lýsir of \alst{a}lla daga &
\ind ok þęy⸗gi at \alst{m}ïnum \alst{m}unum.“\eva

\bvb\speakernoteb{Free [quoth]:}%
“Why should I tell thee, O young youth, \\
\ind of my great heartache? \\
For the Elf-wheel \ken{sun} shines during all days \\
\ind and nowise to my liking.”\evb\evg


\bvg\bva\speakernote{Skírnir:}\mssnote{\Regius~11r/20, \AM~2r/21}%
\llap{„}\alst{M}uni þïna \hld\ hykk⸗a svá \alst{m}ikla vesa, &
\ind at þú mér \edtrans{\alst{s}ęggr}{youth}{\Bfootnote{TODO: etymological discussion.  This word, which usually means simply ‘man’, here seems to have a specific connotation with daring youth.  The sense of ‘young man’ is also seen in \textlink{Volundarkvida}[23], where it is used in reference to king Nithad’s two young sons.  In the present context it answers Free’s addressing Shirner as \emph{sęggr hinn ungi} ‘the young youth’; Shirner points out that the two are of equal age, and so Free is as much of a young man as he.}} né \alst{s}ęgir; &
\alst{u}ngir saman \hld\ vǫ́rum ï \alst{á}r-daga, &
\ind vęl mę́ttim \alst{t}vęir \alst{t}rúa⸗sk.“\eva

\bvb\speakernoteb{Shirner [quoth]:}%
“Thy liking I do not think so great \\
\ind that thou, O youth, should not tell me. \\
Young together were we in days of yore; \\
\ind we two might well trust each other.”\evb\evg


\bvg\bva\speakernote{Fręyr:}\mssnote{\Regius~11r/22, \AM~2r/23}%
\llap{„}İ \alst{G}ymis gǫrðum \hld\ ek \alst{g}anga sá &
\ind \alst{m}ér tíða \alst{m}ęy; &
\alst{a}rmar lýstu, \hld\ en \alst{a}f þaðan &
\ind allt \edtrans{\alst{l}opt ok \alst{l}ǫgr}{air and sea}{\Bfootnote{Formulaic and very old, also parallelled in OE. TODO.}}.\eva

\bvb\speakernoteb{Free [quoth]:}%
“In Gymer’s yards I saw walking \\
\ind a maiden desirable to me. \\
Her arms shone, and thereof [shone] \\
\ind all the air and sea.\evb\evg


\bvg\bva\mssnote{\Regius~11r/24, \AM~2r/24}%
\alst{M}ę́r ’s mér tíðari \hld\ an \alst{m}anna hvęim &
\ind \alst{u}ngum ï \alst{á}r-daga; &
\alst{ȧ}sa ok \alst{a}lfa \hld\ þat vill \edtrans{\alst{ę}ngi maðr}{no one}{\Bfootnote{Lit. ‘no man’, where “man” just means person.  Cf.~note to \textlink{Vafthrudnismal}[55].}}, &
\ind at vit \alst{s}ǫ́tt \alst{s}éim.“\eva

\bvb The maiden is more desirable to me than to any young \\
\ind man in days of yore. \\
Of the \inx[F]{Eese and Elves} does no one wish \\
\ind that we two should be agreed.”\evb\evg


\bvg\bva\speakernote{Skírnir:}\mssnote{\Regius~11r/25, \AM~2r/25}%
\llap{„}\alst{M}ar gef mér þȧ, \hld\ es mik of \alst{m}yrkvan beri &
\ind \alst{v}ísan \alst{v}afr-loga, &
ok \edtrans{þat \alst{s}verð}{that sword}{\Bfootnote{This sword appears to have filled greater function in an earlier version of the poem, since according to \Gylfaginning~37:16 Shirner gave it away, perhaps as a dowry to Gird, and Free did not get it back (cf.~\textlink{Voluspa}[51]/3 and the insult in \textlink{Lokasenna}[42]).  It is notable that Free, a god especially associated with the erect phallus, trades away his actual sword (a symbol of martial power) for sex.}}, \hld\ es \alst{s}jalft vegi⸗sk &
\ind við \alst{jǫ}tna \alst{ę́}tt.“\eva

\bvb\speakernoteb{Shirner [quoth]:}%
“The steed then give me which might bear me over the dark, \\
\ind wise wavering-flame, \\
and that sword which by itself might strike \\
\ind against the race of \inx[P]{Ettins}.”\evb\evg


\bvg\bva\speakernote{Fręyr:}\mssnote{\Regius~11r/27, \AM~2r/27}%
\llap{„}\alst{M}ar þér þann gef’k, \hld\ es þik of \alst{m}yrkvan \edtext{berr &
\ind \alst{v}ísan \alst{v}afr-loga, &
auk þat \alst{s}verð, \hld\ es \alst{s}jalft mun vega⸗sk, &
\ind ef sá ’s \alst{h}orskr es \alst{h}ęfr.“}{\lemma{berr ‘bears’; mun vega⸗sk, ef sá ’s horskr es hęfr ‘will strike if he be wise who holds it’}\Bfootnote{In his response Free replaces the subjunctive verb forms (\emph{beri} ‘might bear’, \emph{vegi⸗sk} ‘might strike’) with indicative and future forms, giving a sense of certainity and authority.  The steed and sword are faultless, and if Shirner fails on the mission, it would be only due to his own fault (“if he be wise who holds it.”).}}\eva
%TODO? Change the line numbering from 1--4 to 1, 3--4.

\bvb\speakernoteb{Free [quoth]:}%
“That steed I give thee which bears thee over the dark, \\
\ind wise wavering-flame, \\
and that sword which by itself will strike \\
\ind if he be wise who holds it.”\evb\evg


\bpg\bpa Skírnir mę́lti við hest’inn:\epa

\bpb Shirner spoke with the horse:\epb\epg


\bvg\bva\mssnote{\Regius~11r/29, \AM~2r/28}%
\Ballnote{Shirner expresses his resolute loyalty to Free.  He will not abandon the horse given to him by his lord; either they both make it or both perish.}%
\llap{„}\alst{M}yrkt es úti, \hld\ \alst{m}ál kveð’k okkr fara &
\ind \alst{ú}rig fjǫll \alst{y}fir &
\ind \edtrans{\alst{þ}ursa}{of the Thurses}{\Afootnote{so \AM; \emph{‘þyria’} \Regius}} \alst{þ}jóð yfir; &
\alst{b}áðir vit komumk \hld\ eða okkr \alst{b}áða tękr &
\ind sá hinn \edtrans{\alst{á}m-át⸗ki \alst{jǫ}tunn}{uncanny ettin}{\Bfootnote{Formulaic; the adjective \emph{ám-átt⸗igr} ‘uncanny’ is used exclusively for evil supernatural beings.  See note to \textlink{Voluspa}[8].}}.“\eva

\bvb “It is dark outside; I call it time for us to journey \\
\ind over the drizzling mountains, \\
\ind over the tribe of \inx[P]{Thurses}. \\
{[Either]} we both make it through or he takes us both, \\
\ind that uncanny ettin.”\evb\evg


\bpg\bpa\mssnote{\Regius~11r/31, \AM~2v/1}%
Skírnir reið i Jǫtun-heima til Gymis garða; þar vǫ́ru hundar ólmir ok bundnir fyrir skíð-garðs hliði þess, er um sal \edtrans{Gerðar}{Gird}{\Bfootnote{It is only now that we find out the maiden’s name.}} var. Hann reið at þar, er fé-hirðir sat á haugi, ok kvaddi hann: \epa

\bpb Shirner rode into the Ettinhomes to Gymer’s yards.  Thereat there were hounds fierce and bound before the slope of the paled fence which surrounded Gird’s hall.  He rode up to where a shepherd sat on a mound and greeted him:\epb\epg


\bvg\bva\mssnote{\Regius~11v/2, \AM~2v/4}%
\llap{„}Sęg þat \alst{h}irðir, \hld\ es \edtrans{ȧ \alst{h}augi sitr}{who sittest on the mound}{\Bfootnote{For this motif cf.~\textlink{Voluspa}[41]/1 n.}} &
\ind ok \alst{v}arðar alla \alst{v}ega: &
hvé ek at \alst{a}nd-spilli \hld\ komumk hins \alst{u}nga mans &
\ind fyr \alst{g}ręyjum \alst{G}ymis.“\eva

\bvb “Tell this, O herdsman who sittest on the mound \\
\ind and watchest all the ways, \\
how I to discourse might come with the young girl \\
\ind past Gymer’s greyhounds.”\evb\evg


\bvg\bva\speakernote{[Hirðir] kvað:}\mssnote{\Regius~11v/4, \AM~2v/5}%
\llap{„}Hvárt est \alst{f}ęigr, \hld\ eða est \alst{f}ramm ginginn &
\ind \edtext{[...]}{\Bfootnote{No gap is indicated in the mss. but it is certain that a c-line is missing here.}}; &
\alst{a}nd-spillis vanr \hld\ þú skalt \alst{ę́} vesa &
\ind \edtrans{\alst{g}óðrar męyjar}{good maiden}{\Bfootnote{Formulaic, carrying with it a sense of chastity.  See note to \textlink{Havamal}[102]/1 for further occurrences.}} \alst{G}ymis.“\eva

\bvb\speakernoteb{[The herdsman] quoth:}%
“Regardless of whether thou art fey or passed-on, \\
\ind {[...]} \\
discourseless shalt thou forever be \\
\ind with Gymer’s good maiden!”\evb\evg


\bvg\bva\speakernote{[Skírnir] kvað:}\mssnote{\Regius~11v/6, \AM~2v/7}
\Ballnote{An excellent formulation of the Old Germanic fatalistic world-view.  Presumably after uttering these words Shirner rides through the fire surrounding the fortress.}%
\llap{„}\edtrans{\alst{K}ostir}{Choices}{\Bfootnote{i.e. ‘alternatives, other ways’.}} ’ro bętri \hld\ \edtrans{an}{than}{\Afootnote{so \AM; \emph{hęldr an at} ‘rather than to [be]’ \Regius}} \alst{k}løkkva séi &
\ind hvęim es \alst{f}úss es \alst{f}ara, &
\edtrans{\alst{ęi}nu dǿgri \hld\ mér vas \alst{a}ldr of skapaðr}{In one half-day my age was shaped}{\Bfootnote{Formulaic.  The single half-day is both the time of one’s birth when the norns or dises arrive (cf.~\textlink{HelgakvidaOne}[1]--4 and \Hakonarmal\ 19/1--2: \emph{Góðu dǿgri \hld\ verðr sá gramr of borinn, / es sér getr slíkan sefa} ‘On a good half-day will that lord be born who gets for himself such heart’) and the fixed time of one’s inevitable death (\textlink{Gripisspa}[25]/4).
For the norns shaping one’s age (ON \emph{(of) skapa aldr}) cf.~\textlink{HelgakvidaOne}[2]/2 and \Gylfaginning~15:30: \emph{Enn eru fleiri nornir, þę́r er koma til hvers barns, er borit er, at skapa aldr} ‘There are yet more norns, those which come to every child which is born to shape its age’; for the use of the verb shape in this context cf.~also \textlink{Reginsmal}[2]/3, 6/3, \textlink{Sigurdskamma}[7]/3.}} &
\ind ok \edtrans{alt \alst{l}íf of \alst{l}agit}{all my life laid down}{\Bfootnote{The causative \emph{lęggja} ‘to lay (down, in place)’ is closely connected to fate; the expression is formulaic.  Cf.~\textlink{Lokasenna}[48]: \emph{ï ár-daga vas þér hit ljóta líf of lagit} ‘in days of yore was thy ugly life laid down’ and \textlink{Voluspa}[19]: \emph{þę́r lǫg lǫgðu} ‘they [= the Norns] laid down laws’.}}.“\eva

\bvb\speakernoteb{[Shirner] quoth:}%
“Choices are better than sobbing might \\
\ind for whomever is eager to journey. \\
In one half-day my age was shaped, \\
\ind and all my life laid down.”\evb\evg


\bvg\bva\speakernote{[Gęrðr] kvað:}\mssnote{\Regius~11v/7, \AM~2v/8}%
\llap{„}Hvat ’s þat \alst{h}lym \alst{h}lymja \hld\ es \alst{h}lymja hęyri’k nú til &
\ind \alst{o}ssum rǫnnum \alst{ï}? &
\alst{jǫ}rð bifa⸗sk, \hld\ en \alst{a}llir fyr &
\ind skjalfa \alst{g}arðar \alst{G}ymis.“\eva

\bvb\speakernoteb{[Gird] quoth:}%
“What is this din of dins which I now hear dinning \\
\ind in our houses? \\
The earth trembles and before us quake \\
\ind all Gymer’s yards.”\evb\evg


\bvg\bva\speakernote{Ambǫ́tt kvað:}\mssnote{\Regius~11v/9, \AM~2v/10}%
\llap{„}\alst{M}aðr ’s hér úti, \hld\ stiginn af \alst{m}ars baki, &
\ind \edtrans{\alst{jó} lę́tr til \alst{ja}rðar taka}{he lets his steed graze the earth}{\Bfootnote{Lit. ‘he lets his steed take to the earth’.  According to \textcite{FinnurEdda} this expression is still used in modern Icelandic (or was, at his time).}}.“\eva

\bvb\speakernoteb{A servant-woman quoth:}%
“A man is here outside stepped down from horse-back; \\
\ind he lets his steed graze the earth.”\evb\evg


\bvg\bva\speakernote{[Gęrðr] kvað:}\mssnote{\Regius~11v/10, \AM~2v/11}%
\llap{„}\alst{I}nn bið hann ganga \hld\ ï \alst{o}kkarn sal &
\ind ok drekka hinn \alst{m}ę́ra \alst{m}jǫð, &
þó ek hitt \alst{ó}umk, \hld\ at hér \alst{ú}ti séi &
\ind mïnn \alst{b}róður-\alst{b}ani.“\eva

\bvb\speakernoteb{[Gird] quoth:}%
“Bid him go in, into our hall \\
\ind and drink the renowned mead; \\
yet I fear that here outside might be \\
\ind my brother’s bane.”\evb\evg


\bvg\bva\speakernote{[Gęrðr] kvað:}\mssnote{\Regius~11v/12, \AM~2v/13}%
\llap{„}Hvat ’s þat \alst{a}lfa \hld\ né \alst{ȧ}sa sona, &
\ind né \alst{v}íssa \alst{v}ana; &
hví \alst{ęi}nn of komt \hld\ \alst{ęi}kinn fúr yfir &
\ind ór \alst{s}al-kynni at \alst{s}éa?“\eva

\bvb\speakernoteb{[Gird quoth:]}%
“What kind [of creature] is this not of Elves nor of sons of the Eese \\
\ind nor of wise Wanes? \\
Why camest thou alone o’er the raging fire, \\
\ind to see the state of our hall?”\evb\evg


\bvg\bva\speakernote{[Skírnir kvað:]}\mssnote{\Regius~11v/14}%
\llap{„}\alst{E}m’k⸗at \alst{a}lfa \hld\ né \alst{ȧ}sa sona &
\ind né \alst{v}íssa \alst{v}ana, &
þó \alst{ęi}nn of kom’k \hld\ \alst{ęi}kinn fúr yfir &
\ind yður \alst{s}al-kynni at \alst{s}éa.\eva

\bvb\speakernoteb{[Shirner quoth:]}%
“I am not of Elves, nor of sons of the Eese, \\
\ind nor of wise Wanes--- \\
yet I came alone o’er the raging fire, \\
\ind to see the state of your hall.\evb\evg


\bvg\bva\mssnote{\Regius~11v/15, \AM~2v/14}%
\Ballnote{Shirner begins his coercion of Gird, which takes up sts.~19--36.
It has been noted by \textcite{Brate1913} that the structure of this coercion bears some resemblance to Weden’s seduction of Wrind in \textcite{Saxo} 3.4.1--8.  Shirner first offers Gird gold (sts.~19--22), then threatens her with violence (23--25) and finally curses her with a long spell (26--36), which breaks her.  In Weden’s seduction of Wrind he disguises as a warrior, then as a gold-smith, then again as a warrior but is spurned each time.  Finally he makes out to be a witch, and having tied Wrind down, proceeds to rape her (cf.~\textlink{Voluspa}[31] n.)  If the first disguise is omitted, we see in both texts a three-fold progression consisting of giving gold, using force of arms, and at last employing magic.}%
\edtrans{\alst{Ę}pli \alst{ę}llifu}{Eleven apples}{\Bfootnote{Probably the apples of Idun, which grant perpetual youth to the gods.  It is not clear whether Shirner has the apples in his possession or whether he is merely promising big.}} \hld\ hér hef’k \alst{a}l-gollin, &
\ind þau mun’k þér \alst{G}ęrðr \alst{g}efa, &
\alst{f}rið at kaupa, \hld\ at þú þér \alst{F}ręy kveðir &
\ind ȯ·\alst{l}ęiðastan \edtrans{at \alst{l}ifa}{in life}{\Bfootnote{\emph{at lifa} here seems to mean ‘in life/living’ rather than the typical infinitive sense ‘to live’; cf.~st.~22 \emph{at dęila} ‘in sharing’ below.  This may be an archaism.}}.“\eva

\bvb Eleven apples have I here, all-golden; \\
\ind these will I to thee, Gird, give \\
to buy thy love, that thou callest Free for thee \\
\ind most unloathsome [lovely] in life.”\evb\evg


\bvg\bva\speakernote{[Gęrðr] kvað:}\mssnote{\Regius~11v/17, \AM~2v/15}%
\llap{„}\alst{Ę}pli \alst{ę}llifu \hld\ ek þigg \alst{a}ldri⸗gi &
\ind at \alst{m}anns-kis \alst{m}unum, &
né vit \alst{F}ręyr, \hld\ meðan okkart \alst{f}jǫr lifir, &
\ind \alst{b}yggjum \alst{b}ę́ði saman.“\eva

\bvb\speakernoteb{[Gird quoth:]}%
“Eleven apples will I never take \\
\ind to any man’s liking, \\
nor will I and Free---while our life remains--- \\
\ind dwell both together.”\evb\evg


\bvg\bva\speakernote{[Skírnir kvað:]}\mssnote{\Regius~11v/19, \AM~2v/17}%
\llap{„}\edtrans{\alst{B}aug}{The bigh}{\Bfootnote{Bigh, meaning ‘armlet’ or ‘ring’.  While not named, the bigh is clearly Dreepner.  It is known from \Gylfaginning~49:54, which describes Balder’s funeral and probably draws on the present stanza: \emph{Óðinn lagði á bál’it gull-hring þann, er Draupnir heitir.  Hónum fylgði sú náttúra, at ina níundu hverja nótt drupu af hónum átta gull-hringar jafn-hǫfgir.} ‘Weden laid on the pyre that gold ring which is called Dreepner.  Its nature was such that every ninth night eight even-heavy golden rings dripped from it.’
The Eese later regained the bigh, which is presumably why Shirner can promise to give it.  \Gylfaginning~49:73--74 tells us that when the god \inx[P]{Harmod} rode to \inx[P]{Hell} to try to bring Balder back, Balder instead gave him the bigh, and told him to bring it back to Weden as a token by which he would be remembered.}} þér þȧ gef’k, \hld\ þann’s \alst{b}ręnndr vas &
\ind með \alst{u}ngum \alst{Ó}ðins syni; &
\edtext{\alst{á}tta ’ro \alst{ja}fn-hǫfgir, \hld\ es \alst{a}f drjúpa &
\ind hina \alst{n}íundu hvęrja \alst{n}ǫ́tt.“}{\lemma{átta \dots\ nǫ́tt ‘Eight \dots\ night.’}\Afootnote{om. by haplography \AM}}\eva

\bvb\speakernoteb{[Shirner quoth:]}%
“The \inx[C]{bigh} I then give thee, which was burned \\
\ind with Weden’s young son \ken*{= Balder}. \\
Eight even-heavy ones drip thereof \\
\ind every ninth night.”\evb\evg


\bvg\bva\speakernote{[Gęrðr] kvað:}\mssnote{\Regius~11v/21, \AM~2v/18}%
\llap{„}\edtext{\alst{B}aug þikk⸗a’k, \hld\ þó’tt \alst{b}ręnndr séi, &
\ind með \alst{u}ngum \alst{Ó}ðins syni;}{\lemma{Baug \dots\ syni ‘The bigh \dots\ son’}\Afootnote{om. by haplography \AM}} &
es⸗a mér \alst{g}olls vant \hld\ ï \alst{g}ǫrðum \alst{G}ymis &
\ind at dęila \alst{f}é \alst{f}ǫður.“\eva

\bvb\speakernoteb{[Gird quoth:]}%
“The bigh will I not take even though it may have been burned \\
\ind with Weden’s young son \ken*{= Balder}. \\
I have no want of gold in Gymer’s yards, \\
\ind partaking of the money of my father.”\evb\evg


\bvg\bva\speakernote{[Skírnir kvað:]}\mssnote{\Regius~11v/23, \AM~2v/19}%
\llap{„}Sér þú \edtext{\alst{m}ę́ki, \alst{m}ę́r, \hld\ \alst{m}jóvan, \alst{m}ál-fáan}{\lemma{mę́ki \dots\ mál-fáan ‘sword \dots\ picture-painted’}\Bfootnote{The sword is inlaid with metal (perhaps gold or silver) forming a pattern.  The expression \emph{mę́ki mál-fáan} (acc. sg.) ‘picture-painted sword’ also occurs in \textlink{Sigurdskamma}[4]/2.}}, &
\ind es \alst{h}ęf’k ï \alst{h}ęndi \alst{h}ér? &
\alst{h}ǫfuð \alst{h}ǫggva \hld\ mun’k þér \alst{h}alsi af, &
\ind nema mér \alst{s}ę́tt \alst{s}ęgir.“\eva

\bvb\speakernoteb{[Shirner quoth:]}%
“Seest thou the sword, maiden---slender, picture-painted--- \\
\ind which I have in my hand here? \\
I will cut the head off thy neck \\
\ind unless thou agree with me!”\evb\evg


\bvg\bva\speakernote{[Gęrðr kvað:]}\mssnote{\Regius~11v/25, \AM~2v/20}%
\llap{„}\alst{Ȧ}·nauð þola \hld\ vil’k \alst{a}ldri⸗gi &
\ind at \edtrans{\alst{m}anns-kis}{any man’s (lit. ‘no man’s)}{\Afootnote{\emph{manns ęnskis} \AM}} \alst{m}unum, &
\edtrans{þó}{although}{\Bfootnote{The interpretation of this word changes the tone of the stanza.  If it means ‘although, even though’ then Gird is expressing her fear of violence between Shirner and her father, i.e., “I will not give in, even if it should lead to violence, as I fear”.  If, on the other hand, it means ‘still, yet’, she may be expressing pride, i.e., “I will not give in, and what is more, my father will be eager to fight thee”.  I find the former understanding more natural.}} hins \alst{g}et’k, \hld\ ef it \alst{G}ymir finnið⸗sk &
\alst{v}ígs ȯ·trauðir \hld\ at ykkr \alst{v}ega tíði.“\eva

\bvb\speakernoteb{[Gird quoth:]}%
“I will never stand coercion \\
\ind to any man’s liking, \\
although I guess that if thou and Gymer meet--- \\
men unreluctant of conflict---ye two will wish to fight.”\evb\evg


\bvg\bva\speakernote{[Skírnir kvað:]}\mssnote{\Regius~11v/27, \AM~2v/22}%
\llap{„}Sér þú \alst{m}ę́ki, \alst{m}ę́r, \hld\ \alst{m}jóvan, \alst{m}ál-fáan, &
\ind es \alst{h}ęf’k ï \alst{h}ęndi \alst{h}ér? &
fyr þęssum \alst{ę}ggjum \hld\ hnígr sá hinn \alst{a}ldni jǫtunn, &
\ind verðr þinn \alst{f}ęigr \alst{f}aðir.\eva

\bvb\speakernoteb{[Shirner quoth:]}%
“Seest thou the sword, maiden---slender, picture-painted--- \\
\ind which I have in my hand here? \\
Before these edges that olden ettin sinks down; \\
\ind \inx[C]{fey} becomes thy father!\evb\evg


\bvg\bva\mssnote{\Regius~11v/28, \AM~2v/24}%
\Ballnote{With this stanza Shirner’s turns away from direct threats of violence and begins his long curse (sts.~26--36).  He curses Gird to be loveless, joyless, captive, and the subject of sexual abuse at the hands of loathsome ettins if she does not give in to his demands.}%
\edtrans{\alst{T}ams-vęndi}{taming-wand}{\Bfootnote{Probably a cane or whip, perhaps of a magical nature.  The imagery is uncomfortably sexual, and Shirner threatens to ‘tame’ Gird like an animal.}} þik drep’k, \hld\ ęn ek þik \alst{t}ęmja mun, &
\ind \alst{m}ę́r, at mïnum \alst{m}unum, &
þar skalt \alst{g}anga \hld\ es þik \alst{g}umna synir &
\ind \alst{s}íðan ę́va \alst{s}éi.\eva

\bvb With the taming-wand I strike thee, and I will tame thee, \\
\ind maiden, to my liking! \\
Thou shalt go whither the sons of men \\
\ind never since will see thee!\evb\evg


\bvg\bva\mssnote{\Regius~11v/30, \AM~2v/26}%
\Ballnote{Shirner threatens Gird with captivity on a high mountain.  The structure of sts.~27 and 28 is identical; both feature a direct threat in ll. 1--3 followed by a mythological hyperbole in ll. 5--6.}%
\edtrans{\alst{A}ra þúfu \alst{ȧ} \hld\ skalt \alst{á}r sitja}{On an eagle’s perch shalt thou sit for long}{\Afootnote{\emph{ár skalt sitja \hld\ ara þúfu ȧ} ‘for long shalt thou sit on an eagle’s perch’ \AM}}, &
\ind \edtext{\alst{h}orfa \alst{h}ęimi \edtext{ór}{\Afootnote{add. in marg. \Regius}}; &
\ind snugga \alst{h}ęljar til}{\lemma{horfa hęimi ór; / snugga hęljar til ‘turn away from the world; / hanker after Hell’}\Afootnote{\emph{horfa ok snugga hęljar til} ‘turn and hanker after Hell’ \AM}\Bfootnote{On the mountain-perch Gird will be able to see both Earth and the Underworld, but her situation will be so miserable that she will turn her face away from the former and long for her own death.}}; &
\edtext{\alst{m}atr séi þér męir lęiðr \hld\ an \alst{m}anna hvęim &
\ind hinn \alst{f}rȧni ormr með \edtext{\alst{f}irum}{\Afootnote{The last word on \AM~2v, after which the text cuts off.  Apart from the very last stanza, the rest of the poem is preserved only in \Regius.}}.}{\lemma{matr \dots\ firum ‘Be \dots\ folk.’}\Bfootnote{Gird’s disgust over her food will be greater than the hatred of the \inx[C]{Middenyardswyrm} among mankind, which is understood to be extreme.  For this hatred cf.~\textlink{Hymiskvida}[22].}}\eva

\bvb On an eagle’s perch shalt thou sit for long, \\
\ind turn away from the world, \\
\ind hanker after \inx[P]{Hell}! \\
Be the food more loathsome to thee than to any man \\
\ind the gleaming Wyrm among folk.\evb\evg


\bvg\bva\mssnote{\Regius~11v/32}%
At \alst{u}ndr-sjȯnum verðir \hld\ es þú \alst{ú}t kømr, &
\ind ȧ þik \edtrans{\alst{H}rímnir}{Rimner}{\Bfootnote{An unknown ettin whose name derives from \emph{hrím} ‘rime ice’; he is one the many who will jeer at Gird.}} \alst{h}ari &
\ind ȧ þik \alst{h}ot-vetna stari, &
\alst{v}íð-kunnari \alst{v}erðir \hld\ an \edtrans{\alst{v}ǫrðr með goðum}{the watchman with the Gods}{\Bfootnote{Homedal, who guards Bivrest from the ettins; cf.~\textlink{EddicFragments}[F3][III] P1.  For the epithet cf.~\textlink{Grimnismal}[13]}}, &
\ind \alst{g}api þú \alst{g}rindum frȧ.\eva

\bvb A wondrous sight mayst thou be when thou comest out; \\
\ind at thee may Rimner ogle; \\
\ind at thee may anyone stare! \\
Mayst thou be more widely known than the watchman with the Gods; \\
\ind mayst thou gape from the gates.\evb\evg


\bvg\bva\mssnote{\Regius~12r/2}%
\edtrans{\alst{T}ópi ok ópi, \hld\ \alst{t}jǫsull}{Toop and woop, tarsle}{\Bfootnote{Otherwise unknown magic curse words without clear meaning; I have left them untranslated.  \emph{ópi} resembles \emph{óp} ‘scream’, and \emph{tjǫsull} ‘tarsle’ may perhaps be related to OE \emph{teors} ‘penis’ and mean ‘little phallus’.}} ok ȯ·þoli, &
\ind vaxi þér \alst{t}ǫ́r með \alst{t}rega; &
\alst{s}ęt⸗sk þú niðr \hld\ en mun’k \alst{s}ęgja þér &
\ind \edtrans{\alst{s}váran \alst{s}ús-breka}{hard roaring-breaker}{\Bfootnote{Unclear.  A curse that will hit Gird like a breaking wave (\emph{breki})?}}, &
\ind ok \alst{t}vinnan \alst{t}rega.\eva

\bvb Toop and woop, tarsle and restlessness--- \\
\ind may thy tears grow with grief! \\
Sit thyself down and I will tell thee \\
\ind a hard roaring-breaker \\
\ind and a twined grief.\evb\evg


\bvg\bva\mssnote{\Regius~12r/3}%
\edtrans{Tramar}{Fiends}{\Bfootnote{TODO: reference ordsifjabók.}} \alst{g}nęypa \hld\ þik skulu \alst{g}ęrstan dag &
\ind \alst{jǫ}tna gǫrðum \alst{ï}, &
til \alst{h}rím-þursa \alst{h}allar \hld\ þú skalt \alst{h}vęrjan dag &
\ind \alst{k}ranga \alst{k}osta-laus; &
\ind \alst{k}ranga \alst{k}osta-vǫn; &
\alst{g}rát at \alst{g}amni \hld\ skalt ï \alst{g}ǫgn hafa &
\ind ok lęiða með \alst{t}ǫ́rum \alst{t}rega.\eva

\bvb Fiends shall pine thee on a dismal day \\
\ind in the yards of the Ettins. \\
To the hall of Rime-Thurses shalt thou every day \\
\ind crawl choice-less; \\
\ind crawl choice-lacking. \\
Weeping shalt thou have in exchange for joy, \\
\ind and nurse thy grief with tears.\evb\evg


\bvg\bva\mssnote{\Regius~12r/7}%
Með \edtrans{\alst{þ}ursi \alst{þ}rí-hǫfðuðum}{three-headed thurse}{\Bfootnote{Ettins often have an abnormal number of body parts.  For their many-headed-ness see note to \textlink{Hymiskvida}[8]/2.}} \hld\ \alst{þ}ú skalt ę́ nara &
\ind eða \alst{v}er-laus \alst{v}esa; &
\ind þitt \alst{g}ęð \alst{g}rípi, &
\ind þik \alst{m}orn \alst{m}orni; &
\edtext{ves þú sem \alst{þ}istill, \hld\ sá’s \alst{þ}runginn vas &
\ind ï \alst{o}fan-verða \alst{ǫ}nn.}{\lemma{ves \dots\ ǫnn ‘Be \dots\ harvest’}\Bfootnote{The thistle was considered a weed.  Cf.~the English galder against a cattle-thief (Charm IX in margins of CCCC 41. TODO: edit this!) cursing him to be \emph{swá bréðel swa séo þystel} ‘as wretched as the thistle’.  The ‘pressing’ of the thistle has been connected with an Estonian folk-ritual wherein a thistle would be crushed by stones at the beginning of harvest to prevent further weeds (TODO: reference).}}\eva

\bvb With a three-headed thurse shalt thou always dwell \\
\ind or be husband-less. \\
\ind May thy senses seize; \\
\ind may murrain mourn thee! \\
Be thou like the thistle which was pressed \\
\ind during highest harvest!\evb\evg


\bvg\bva\mssnote{\Regius~12r/9}%
Til \alst{h}olts ek gekk \hld\ ok \edtrans{til \alst{h}rás viðar}{to the raw/sappy tree}{\Bfootnote{The wood of a sapling was apparently thought to be the most effective for magic; cf.~\textlink{Havamal}[152], which speaks about a runic curse carved on \emph{rótum rás viðar} ‘the roots of a raw/sappy tree’.}} &
\ind \edtrans{\alst{g}amban-tęin}{gomben-toe}{\Bfootnote{Perhaps ‘mighty twig’, also found in \textlink{Harbardsljod}[20]/4.  The word is a cpd.~of the very rare prefix \emph{gamban-} ‘magic/curse?’ (cf.~\textlink{Lokasenna}[8]/4 n.) and \emph{tęinn} ‘twig, branch’ (cf.~\emph{mistil-tęinn} ‘mistle-toe’).  This may be the stick on which the runic curse in st.~36 below should be carved, or it is to be identified with the \emph{tams-vǫndr} ‘taming-wand’ of st.~26 above.  \textlink{Harbardsljod}[21] suggests that it is an object of some value.}} at \alst{g}eta &
\ind \alst{g}amban-tęin ek \alst{g}at.\eva

\bvb To the grove I went and to the raw/sappy tree, \\
\ind the \inx[C]{gombentoe} for to get; \\
\ind the gombentoe I got.\evb\evg


\bvg\bva\mssnote{\Regius~12r/10}%
\alst{R}ęiðr ’s þér Óðinn, \hld\ \alst{r}ęiðr ’s þér \edtrans{Ȧsa-bragr}{Eesebray}{\Bfootnote{‘The pinnacle of the Eese’, Thunder.}}, &
\ind þik skal \alst{F}ręyr \alst{f}ía⸗sk, &
hin \alst{f}irin-illa mę́r, \hld\ en \alst{f}ingit hęfr &
\ind \alst{g}amban-ręiði \alst{g}oða.\eva

\bvb Wroth with thee is Weden; wroth with thee is Eesebray \name{= Thunder}; \\
\ind thee shall Free come to hate, \\
O most wicked maiden, if thou hast earned \\
\ind the gomben-wrath of the gods.\evb\evg


\bvg\bva\mssnote{\Regius~12r/12}%
\alst{H}ęyri jǫtnar, \hld\ \alst{h}ęyri \alst{h}rím-þursar, &
\alst{s}ynir \alst{S}uttunga, \hld\ \alst{s}jalfir ǫ̇s-liðar, &
hvé ek \alst{f}yrir·býð, \hld\ hvé ek \alst{f}yrir·banna &
\ind \alst{m}anna glaum \alst{m}ani, &
\ind \alst{m}anna nyt \alst{m}ani.\eva

\bvb Hear Ettins, hear Rime-thurses, \\
sons of Sutting \ken{ettins}, the very Os-troops \ken*{= Eese}, \\
how I forbid, how I forban \\
\ind men’s fellowship from the maiden, \\
\ind men’s joy from the maiden!\evb\evg


\bvg\bva\mssnote{\Regius~12r/14}%
\alst{H}rím-grímnir hęitir þurs, \hld\ es þik \alst{h}afa skal &
\ind fyr \alst{N}á-grindr \alst{n}eðan, &
þar þér \alst{v}íl-męgir \hld\ ȧ \alst{v}iðar rótum &
\ind \alst{g}ęita-hland \alst{g}efi; &
\alst{ǿ}ðri drykkju \hld\ fȧ þú \alst{a}ldri⸗gi, &
\ind \alst{m}ę́r, af þïnum \alst{m}unum, &
\ind \alst{m}ę́r, at \alst{m}ïnum \alst{m}unum.\eva

\bvb Rimegrimner is the name of the thurse who shall have thee \\
\ind down beneath Neegrind, \\
where lads of toil \ken{thralls} on the roots of a tree \\
\ind goat-piss will give thee. \\
Never fetch a finer drink, \\
\ind O maiden, against thy liking, \\
\ind O maiden, to \emph{my} liking!\evb\evg


\bvg\bva\mssnote{\Regius~12r/16}%
\Ballnote{With the carving of the rune-stick Shirner threatens to fulfill the curse predicted in sts.~26--35, but tells Gird that he will scrape the runes off---thus nullifying the curse---if she will accept his demands.}%
\edtrans{\alst{Þ}urs}{thurse}{\Bfootnote{Thurse is the name of the \textbf{þ}-rune (ᚦ); it is carved as part of the curse.}} ríst’k \alst{þ}ér \hld\ ok \edtrans{\alst{þ}ría stafi}{three staves}{\Bfootnote{Three runic letters (or phrases) representing the three following words (\emph{ęrgi} ‘queerness, degeneracy’ etc.).  The ritual practice of carving “three staves” is first found on the C7th Gummarp stone: \textbf{h\textsc{a}þuwol\textsc{a}fʀ s\textsc{a}te st\textsc{a}b\textsc{a} þri\textsc{a} fff} ‘Hathwolf placed three staves: fff’, where the \textbf{f}-rune (ᚠ) stands for its name \inx[C]{fee} (i.e. ‘wealth, cattle’) and is thus meant to bring wealth.}}, &
\ind \edtrans{\alst{ę}rgi ok \alst{ǿ}ði ok \alst{ȯ}·þola}{queerness and madness and restlessness}{\Bfootnote{Both \emph{ęrgi} ‘queerness, degeneracy’ and \emph{ȯ·þoli} ‘restlessness’ (here probably from strong lust) are found in the love magic charm on the rune stick B257 from Bryggen (edited below under Galders).  \emph{ęrgi} is also found in the curse-formula on the C7th Proto-Norse runestones from Stentoften and Björketorp.  See further introduction to B257.}}, &
svá ek þat \alst{a}f ríst \hld\ sem ek þat \alst{ȧ} ręist, &
\ind ef gøra⸗sk \alst{þ}arfar \alst{þ}ęss.“\eva

\bvb \inx[P]{Thurses}[Thurse] I carve for thee, and three staves: \\
\ind \inx[C]{queerness} and madness and restlessness. \\
So will I carve it \emph{off} like I carved it \emph{on}, \\
\ind if there arise need for that.”\evb\evg


\bvg\bva\speakernote{[Gęrðr kvað:]}\mssnote{\Regius~12r/19}%
\llap{„}\edtext{\alst{H}ęill ves þú \alst{h}ęldr, svęinn, \hld\ ok tak við \edtrans{\alst{h}rím-kalki}{rime-chalice}{\Bfootnote{Some kind of expensive glazed drinking vessel; the second element \emph{kalkr} ‘chalice’ is a borrowing from Latin \emph{calix} and suggests a Roman origin.  Cf.~the \emph{kalkr} in \textlink{Hymiskvida}[28]/4b.}} &
\ind \alst{f}ullum \alst{f}orns mjaðar,}{\lemma{Hęill \dots\ mjaðar ‘Hale \dots\ mead’}\Bfootnote{Formulaic; repeated identically in \textlink{Lokasenna}[53]/1--2.}} &
þó hafða’k \alst{ę́}tlat, \hld\ at mynda’k \alst{a}ldri⸗gi &
\ind unna \edtrans{\alst{v}aningja}{the Waning \ken*{= Free}}{\Bfootnote{The ‘descendant of the \inx[P]{Wanes}’.  A rare word.  Its only other occurence in the Norse corpus is in a \inx[C]{thule} of boar-names.  Boars were sacred to Free, TODO.}} \alst{v}ęl.“\eva

\bvb\speakernoteb{[Gird quoth:]}%
“Hale be thou rather, swain, and receive the rime-chalice, \\
\ind full of ancient mead, \\
although I had intended that I never would \\
\ind love the Waning \ken*{= Free} well.”\evb\evg


\bvg\bva\speakernote{[Skírnir kvað:]}\mssnote{\Regius~12r/21}%
\llap{„}\alst{Ø}rendi mïn \hld\ vil’k \alst{ǫ}ll vita, &
\ind áðr ríða’k \alst{h}ęim \alst{h}eðan, &
nę́r ȧ \alst{þ}ingi \hld\ munt hinum \alst{þ}roska &
\ind \alst{n}ęnna \alst{N}jarðar syni?“\eva

\bvb\speakernoteb{[Shirner quoth:]}%
“My errands all I wish to know \\
\ind before I ride home hence: \\
when on the \inx[C]{Thing} wilt thou to the virile \\
\ind son of Nearth \ken*{= Free} submit?”\evb\evg


\bvg\bva\speakernote{[Gęrðr kvað:]}\mssnote{\Regius~12r/23}%
\llap{„}\edtrans{\alst{B}arri}{Barrey}{\Bfootnote{Derived from \emph{barr} ‘leaf, foliage’; perhaps a reference to a cornfield (of wheat or barley, not maize!) which would fit in an interpretation of the poem as an analogy for agricultural fertility.}} hęitir, \hld\ es vit \alst{b}ę́ði vitum, &
\ind \alst{l}undr \edtrans{\alst{l}ogn-fara}{calm breezes}{\Bfootnote{A \emph{hapax} compound, here read as a gen. pl. of \emph{logn} n. ‘calm (wind)’ + \emph{far} ‘trail, passage’, i.e. ‘where one travels calmly’.  Another interpretation is a gen. sg. of an unattested proper noun \emph{Logn·fari} ‘Calm-farer’, ‘the one who causes calm winds’ (cf.~\emph{Mundil·fari} in \textlink{Vafthrudnismal}[23]), probably a name for Free in his role as a fertility god.}}, &
en ępt \alst{n}ę́tr \alst{n}íu, \hld\ þar mun \alst{N}jarðar syni &
\ind \alst{G}ęrðr unna \alst{g}amans.“\eva

\bvb\speakernoteb{[Gird quoth:]}%
“Barrey is the name---as we both do know--- \\
\ind of a grove of calm breezes, \\
and after nine nights there will to the son of Nearth \\
\ind Gird her pleasure grant.”\evb\evg


\bpg\bpa\mssnote{\Regius~12r/24}%
Þá reið Skírnir heim.  Freyr stóð úti ok kvaddi hann ok spurði tíðenda:\epa

\bpb Then Shirner rode home.  Free stood outside and greeted him and asked for the tidings:\epb\epg


\bvg\bva\mssnote{\Regius~12r/25}%
\llap{„}\alst{S}ęg mér, Skírnir, \hld\ áðr verpir \alst{s}ǫðli af mar &
\ind \edtrans{ok stígir \alst{f}eti \alst{f}ramarr}{and take a step further}{\Bfootnote{Formulaic; a variant of \emph{feti ganga/gangir framarr} (\textlink{Havamal}[38]/2, \textlink{Lokasenna}[1]/2).}}, &
hvat \alst{á}rnaðir \hld\ ï \alst{Jǫ}tun-hęima &
\ind þïns eða \alst{m}ïns \alst{m}unar?“\eva

\bvb “Tell me, Shirner, before thou mightst throw the saddle off the steed \\
\ind and take a step further, \\
what thou accomplished in the \inx[P]{Ettinhomes} \\
\ind to thy or my liking?”\evb\evg


\bvg\bva\speakernote{[Skírnir kvað:]}\mssnote{\Regius~12r/27}%
\llap{„}\alst{B}arri hęitir, \hld\ es vit \alst{b}áðir vitum, &
\ind \alst{l}undr \alst{l}ogn-fara, &
en ępt \alst{n}ę́tr \alst{n}íu, \hld\ þar mun \alst{N}jarðar syni &
\ind \alst{G}ęrðr unna \alst{g}amans.“\eva

\bvb\speakernoteb{[Shirner quoth:]}%
“Barrey is the name---as we both do know--- \\
\ind of a grove of calm breezes, \\
and after nine nights there will to the son of Nearth \\
\ind Gird her pleasure grant.”\evb\evg


\bvg\bva\speakernote{[Fręyr kvað:]}\mssnote{\Regius~12r/28, \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\llap{„}\alst{L}ǫng es nǫ́tt, \hld\ \edtrans{\alst{l}angar ’ro tvę́r}{long are two}{\Afootnote{\emph{lǫng es ǫnnur} ‘long is another’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}}, &
\ind \edtext{hvé of \alst{þ}ręyja’k \alst{þ}ríar?}{\Afootnote{\emph{hvé męga’k þręyja þríar} \RegiusProse\Trajectinus\Upsaliensis\Wormianus}} &
opt \alst{m}ér \alst{m}ȧnaðr \hld\ \alst{m}inni þȯtti &
\ind an sjá \edtrans{\alst{h}ǫlf \alst{h}ý-nǫ́tt}{half wedding-night}{\Bfootnote{\emph{hý-nǫ́tt} is a hapax cpd.~of \emph{hý-} + \emph{nǫ́tt} ‘night’.  \emph{hý-} has been variously explained, e.g.~as a form of the demonstrative pronoun stem PGmc.~\emph{*hiz}, but is most likely the same prefix found in \emph{hý-býli} ‘farmstead’, equivalent to Gothic \emph{heiwa-} ‘marriage, household’, OHG \emph{hí-} ‘marriage’; for the root cf.~\textcite[227]{Kroonen2013}.

On the basis of the first element the most likely sense of the cpd.~is the wedding-night and more specifically the consummation of the marriage.  The word \emph{hǫlf} ‘half’ should then probably not be taken as a measurement of time (like ‘half a month’) but rather as meaning ‘incomplete’ or ‘divided’ (cf.~e.g.~\textlink{Havamal}[53]/4).  The wedding-night may be called divided since although the two are now betrothed, they are still separated; each is ‘half’.}}.“\eva

\bvb\speakernoteb{[Free quoth:]}%
“Long is a night, long are two--- \\
\ind how can I yearn for three? \\
Oft a month to me seemed less \\
\ind than this half wedding-night!”\evb\evg

\sectionline

\section{Appendix}

\subsection{\Gylfaginning~37:1--15}

NOTE: The following excerpt has not yet been critically edited.%TODO

\Gylfaginning~37:1--15 deals with the relevant myth.  The remainder of the chapter (37:16--19) deals with Free’s lack of his sword and is cited in \textlink{Voluspa}[51]/3 (n.).  Cf.~also st.~8 above and \textlink{Lokasenna}[42].

\Gylfaginning~37:1--7 corresponds to \textlink{Skirnismal}[P1]--2 and 37:8--13 paraphrases sts.~3--9.  The rest of the poem is summed up very succinctly by 37:14.

\bpg\bpa[0]%
\pva Gymir hét maðr, en kona hans Aur·boða. Hón var berg-risa ę́ttar.
\pva Dóttir þeira er Gerðr, er allra kvinna var fegrst.
\pva Þat var einn dag, at Freyr hafði gengit í Hlið-skjálf ok sá of heima alla.
\pva En er hann leit í norðr-ę́tt, þá sá hann á einum bǿ mikit hús ok fagrt, ok til þess húss gekk kona,
\pva ok er hón tók upp hǫndum ok lauk hurð fyrir sér, þá lýsti af hǫndum hennar bę́ði í lopt ok á lǫg, ok allir heimar birtusk af henni.
\pva Ok svá hefnði hónum þat mikla mikil-lę́ti, er hann hafði setsk í þat it helga sę́ti, at hann gekk í braut fullr af harmi.
\pva Ok er hann kom heim, mę́lti hann ekki. Ekki svaf hann, ekki drakk hann. Engi þorði ok at krefja hann orða.\epa

\bpb\pvb Gymer was the name of a man, and his woman was Earbode; she was of the lineage of barrow-risers.
\pvb Their daughter is Gird who was fairest of all women.
\pvb It was one day when Free had gone to the Lithshelf and looked out over all the Homes.
\pvb But when he looked north he saw on a farm a large and fair house, and to that house walked a woman,
\pvb and when she raised her hands and closed the doors behind her then it shone from her hands into both the air and the waters, and all the homes were brightened by her.
\pvb And so greatly did the greatness which he had beheld in the holy seat harm him that he walked away filled with grief,
\pvb and when he came home he said nothing.  He slept nothing; he drank nothing.  No one dared to get words out of him.\epb\epg


\bpg\bpa[0]\pva[8]%
Þá lét Njǫrðr kalla til sín Skírni, skó-svein Freys, ok bað hann ganga til Freys ok beiða hann orða ok spyrja, hverjum hann vę́ri svá reiðr, at hann mélti ekki við menn.
\pva En Skírnir létsk ganga mundu ok eigi fúss ok kvað illra svara vera ván af honum.
\pva En er hann kom til Freys, þá spurði hann, hví Freyr var svá hnipinn ok mélti ekki við menn.
\pva Þá svarar Freyr ok sagði, at hann hefði sét konu fagra ok fyrir hennar sakir var hann svá harm-fullr, at eigi myndi hann lengi lifa, ef hann skyldi eigi ná henni,
\pva \llap{„}ok nú skaltu fara ok biðja hennar mér til handa ok hafa hana heim hingat, hvárt er faðir hennar vill eða eigi, ok skal ek þat vel launa þér.“
\pva Þá svarar Skírnir, sagði svá, at hann skal fara sendi-ferð, en Freyr skal fá honum sverð sitt. Þat var svá gott sverð, at sjálft vást. En Freyr lét eigi þat til skorta ok gaf honum sverðit.\epa

\bpb\pvb[8]%
Then Nearth had Shirner, Free’s shoe-swain, summoned before him and asked him to go to Free and bid him to speak and ask at whom he was so wroth that he would not speak with men.
\pvb And Shirner said that he would go, although not eagerly, and said that he expected ill answers from him.
\pvb And when he came to Free he asked why Free was so downcast and spoke nothing with men.
\pvb Then Free answers, and said that he had seen a fair woman and for her sakes was he so full of grief that he would not live long if he should not reach her,
\pvb “and now shalt thou journey to ask for her hand on my behalf, and bring her home hither whether her father wants to or not, and I shall reward thee well for that.”
\pvb Then Shirner answers; said so, that he will go on the errand-journey, but Free shall give him his sword; it was such a good sword that it struck by itself.  And Free did not deny him that and gave him the sword.\epb\epg


\bpg\bpa[0]\pva[14]%
Þá fór Skírnir ok bað honum konu’nnar ok fekk heit hennar, ok níu nóttum síðar skyldi hon þar koma, er Barrey heitir, ok ganga þá at brul-laupinu með Frey.
\pva En er Skírnir sagði Frey sitt erindi, þá kvað hann þetta: \edtext{[...]}{\Bfootnote{Here the text cites its own variant of st.~42; see there for critical notes.}}\epa

\bpb\pvb[14]%
Then Shirner journeyed and asked for the woman’s hand for him [Free] and got her promise that nine nights later she would come to that place which is called Barrey and have a wedding with Free.
\pvb And when Shirner told Free his errand then he quoth this: [...]\epb\epg
