\bookStart{Spae of the Wallow}[Vǫlu spǫ́]
\setBookCode{Voluspa}

\begin{flushright}%
\textbf{Dating} \parencite{Sapp2022}: C10th (0.865)–early C11th (0.121)

\textbf{Meter:} \Fornyrdislag%
\end{flushright}

\section{Introduction}

The \textbf{Spae of the Wallow} (ON \emph{Vǫlu spǫ́}, abbrev. \textlink{Voluspa}) is probably the most important Norse mythological text surviving from Heathen times.  It is a \inx[C]{spae} (\emph{spǫ́} ‘prophecy’) told by a \inx[C]{wallow} (\emph{vǫlva} ‘seeress, sibyl, prophetess’) who has been summoned by \inx[P]{Weden}, the chief of the Gods, in order to relate the mythic history of the World.

The figure of the wallow represents the Germanic belief that women, at least in elder age, were invested with a certain psychic or prophetic power; this power became even greater when they died and came into possession of such knowledge unknown even to the chief god Weden himself.  It was for this reason that Weden in his incessant lust for wisdom undertook frequent journeys in order to question various beings, especially ones dead, about mythological lore.  \textlink{Voluspa} is not the only Eddic poem reflecting this motif.  Most similar is \textlink{Baldrsdraumar}, wherein Weden summons a wallow from her grave in order to find out why the god \inx[P]{Balder} is having ominous nightmares, but Weden’s journeys to commune with the dead are also alluded to in \textlink{Harbardsljod} TODO.  For his general quest for knowledge there is also \textlink{Vafthrudnismal}, wherein Weden defeats the wise ettin \inx[P]{Webthrithner} to a wisdom contest, his self-hanging (\textlink{Havamal} 138, 139), and his giving of an eye to Mimer (\textlink{Voluspa} 28).

As a repository of mythic lore \textlink{Voluspa} resembles \textlink{Vafthrudnismal}, \textlink{Grimnismal}, \textlink{Sigrdrifumal}, and \textlink{Alvissmal}, but it differs from all aforementioned works in two key ways: \textlink{Voluspa} is a prophecy rather than a motley collection of scattered mythological lore and a monologue rather than a dialogue or riddle contest.  Its unique importance lies in the fact that it offers a chronological overview of the Norse mythic timeline of the World from its creation to its coming destruction and rebirth.  This timeline is, unfortunately for the study of Norse mythology, very much troubled by the fact that the wallow continually speaks in the most obscure terms.  Events are related in an allusive fashion that presupposes that the audience be already familiar with them, and while we can connect many of the allusions to more complete narratives in Snorre’s Edda (which, to complicate things further, often relies on \textlink{Voluspa}), others are entirely unknown to us.  Things are made even harder by the fact that the witness manuscripts often disagree over the content and order of stanzas.

\subsection{Preservation}

\textlink{Voluspa} is attested in two full independent recensions.  The first and most important is \Regius\ (foll. 1r–3r), where it is the first poem.  The other is \Hauksbok\ (foll. 20r–21r), where it is found in a large collection of saws and Catholics works.

Many stanzas from \textlink{Voluspa} are also cited or paraphrased in \Gylfaginning, the first part of Snorre’s Edda, which in many parts closely follows the poem.  Snorre clearly had access to a third, now-lost, recension of \textlink{Voluspa}, and therefore his cited stanzas are very important.  The paraphrases are harder to use, but can still provide important critical readings and serve as a tiebreaker for readings that differ between \Regius\ and \Hauksbok\ (e.g. in st. 19, where \emph{sal} ‘hall’ in the paraphrase agrees with \Hauksbok\ against \Regius\ \emph{sę́} ‘lake’).  For the four mss. of \Gylfaginning—\RegiusProse, \Trajectinus, \Wormianus, and \Upsaliensis—see the Introduction to Mythic Poetry.

The order of stanzas, which varies greatly across mss., is illustrated by the following table.  Stanzas in \Gylfaginning\ which are quoted independently are marked with plus signs; sequences of uninterrupted quotations of several sts. are abbreviated with an incrementing alphabetic symbol, and each stanza is given a number based on its position, so that \emph{B1} is the first stanza in the second sequence, and so on.  When a stanza found in a manuscript is strongly divergent (e.g. st. 10, where \Gylfaginning\ omits the first two half-lines), its number is followed by a star.  The stanzas beginning with the formula \emph{Þȧ gingu ręgin ǫll} ‘Then went all the Reins’ et c. are represented by the immediately following half-line.

\begin{small}\begin{longtabu} to \textwidth {|c c c c c c|}
	\hline
	\multicolumn{2}{|c}{\emph{pres. ed.}} & \Regius & \Hauksbok & \RegiusProse\Trajectinus\Wormianus & \Upsaliensis \\ [0.5ex]
	\hline\hline\endhead
	\hline\endfoot
	1 & Hljóðs bið’k allar & 1 & 1 & − & − \\
	2 & Ek man jǫtna & 2 & 2 & − & − \\
	3 & Ár vas alda & 3 & 3 & + & + \\
	4 & áðr Burs synir & 4 & 4 & − & − \\
	5 & Sól varp sunnan & 5 & 5 & +* & +* \\
	6 & \dots\ nǫ́tt ok niðjum & 6 & 6 & − & − \\
	7 & Hittu⸗sk ę̇sir & 7 & 7 & − & − \\
	8 & Tęflðu ï tu̇ni & 8 & 8 & − & − \\
	9 & \dots\ hvęrr skyldi dverga & 9 & 9 & B1 & B1 \\
	10 & Þar vas Móð·sognir & 10 & 10 & B2* & B2* \\
	11–15 & \emph{Dwarf-tallies} & 11–15 & 11–16 & + & + \\
	16 & Und’s þrír kvǫ̇mu & 16 & 17 & − & − \\
	17 & Ǫnd þau né ǫ́ttu & 17 & 18 & − & − \\
	18 & Ask vęit’k standa & 18 & 19 & + & + \\
	19 & Þaðan koma męyjar & 19–20 & 20–21 & − & − \\
	20 & Þat man hǫ̇n folk-víg & 21–22 & 27 & − & − \\
	21 & Hęiði hétu & 23 & 28 & − & − \\
	22 & \dots\ hvárt skyldu ę̇sir & 24 & 29 & − & − \\
	23 & Flęygði Óðinn & 25 & 30 & − & − \\
	24 & \dots\ hvęrr hęfði lopt alt & 26 & 22 & C1 & C1 \\
	25 & Þȯrr ęinn þar vá & 27 & 23 & C2* & C2* \\
	26 & Vęit hǫ̇n Hęim·dalar & 28 & 24 & − & − \\
	27 & Ęin sat hǫ̇n úti & 29 & − & − & − \\
	28 & Alt vęit’k, Óðinn & 29 & − & + & + \\
	29 & Valði hęnni Hęr-fǫðr & 30 & − & − & − \\
	30 & Sá hǫ̇n val-kyrjur & 31 & − & − & − \\
	31 & Ek sá Baldri & 32 & − & − & − \\
	32 & Varð af męiði & 33 & − & − & − \\
	33 & Þó hann ę́va hęndr & 34 & − & − & − \\
	34 & þar sitr Sigyn & 35 & 32* & − & − \\
	35 & Ǫ́ fęllr austan & 36 & − & − & − \\
	36 & Stóð fyr norðan & 36 & − & − & − \\
	37 & Sal sá hǫ̇n standa & 37 & 36 & E1 & E1 \\
	38 & Sér hǫ̇n þar vaða & 38 & 37 & E2* & E2* \\
	39 & Austr býr hin aldna & 39 & 25 & A1 & A1 \\
	40 & Fylli⸗sk fjǫrvi & 40 & 26 & A2 & A2 \\
	41 & Sat þar ȧ haugi & 41 & 34 & − & − \\
	42 & Gól of ǫ̇sum & 42 & 35 & − & − \\
	43, 47, 55 & Gęyr (nú) Garmr mjǫk & 43, 46, 55 & 33, 38, 43, 48, 51 & − & − \\
	44 & Brǿðr munu bęrja⸗sk & 44 & 39 & − & − \\
	45 & Lęika Mïms synir & 45 & 40 & D1* & D1* \\
	46 & Hvat ’s með ǫ̇sum? & 49 & 42 & D2 & D2* \\
	48 & Hrymr ękr austan & 47 & 44 & D3 & − \\
	49 & Kjóll fęrr austan & 48 & 45 & D4 & − \\
	50 & Surtr fęrr sunnan & 50 & 46 & +, D5 (cited twice) & + \\
	51 & Þȧ kømr Hlïnar & 51 & 47 & D6 & − \\
	52 & Þȧ kømr hinn mikli & 52 & − & D7 & − \\
	53 & Þȧ kømr hinn mę́ri & 53* & 49* & D8 & − \\
	54 & Sól tér sortna & 54 & 50 & D9 & − \\
	56 & Sér hǫ̇n upp koma & 56 & 52 & − & − \\
	57 & Finna⸗sk ę̇sir & 57* & 53 & − & − \\
	58 & Þar munu ęptir & 58 & 54 & − & − \\
	59 & Munu ȯ·sánir & 59 & 55 & − & − \\
	60 & Þȧ kná Hø̇nir & 60 & 56 & − & − \\
	61 & Sal sér hǫ̇n standa & 61 & 57 & + & + \\
	62 & Þar kømr hinn dimmi & 62 & 59 & − & − \\
	H1 & Þȧ kná Váli & − & 31 & − & − \\
	H2 & Hrę́ða⸗sk allir & − & 41 & − & − \\
	H3 & Gïnn lopt yfir & − & 48 & − & − \\
	H4 & Þȧ kømr hinn ríki & − & 58 & − & − \\ [1ex]

	\hline
\end{longtabu}\end{small}

\subsection{Summary}

On the basis of its refrains the poem may be divided into the following sections.  1–2: introduction.  3–17: creation.  18–42: cosmological and mythological exposition.  43–55: destruction.  56–62: the new world.

It seems that immediately prior to the poem’s commencing, Weden has awakened the wallow out of her grave (probably in a manner similar to that in \textlink{Baldrsdraumar}).  He commands her to speak.

The wallow’s \emph{spae} begins with a bid for silence (1), followed by her earliest memories of being raised amongst ettins (2).  She begins by recounting the Creation (3–6) and the following Golden Age (7–8), which was brought to an end by the intrusion of three unidentified ettin-maidens (8).  After this she describes the making of the dwarfs (9–10), whose names are listed in several separate \emph{dwarf-tallies} which are without doubt later inserts (11–15).  She mentions how the Gods gave life to the first human beings (16–17), and then describes the Ugdrassle’s Ash (18) and the three \inx[P]{norns} living under it (19).

Following st. 19 the order of stanzas in the two full recension—\Regius\ and \Hauksbok—diverges significantly and they do not come to agree until st. 43.

In \Regius—whose order of stanzas is (with a single exception) adopted for the present edition—the wallow recalls how a woman named Goldwey was sacrificed and reborn three times (20), and how she, under the name Heath, practiced sorcery and witchcraft (21).  She then recalls the first war in the world, between the Eese and Wanes (22–23), and the killing of the smith who was promised \inx[P]{Frow} and the sun and moon in exchange for building the wall of Osyard (24-25).  This is followed by a cryptic verse describing Homedal’s hidden silence or hearing (26).  She then seems to tell of her own origin; how she was approached by Weden, who gave her rings and jewels in exchange for her wisdom (27–29).  She then tells of the walkirries (30) and relates the death of Balder (31–33) and the subsequent binding of Lock (34).  After this she describes a sword-river (35) and two beer-halls (36), then the torture of criminals in the afterlife by Nithehewer (37–38).  She sees Ironwood in the East, where the wolves who will swallow the Sun and Moon are being reared (39–40).  She foresees the crowing of three cocks: one in Ironwood (41), the other among the Gods, and the third in \inx[P]{Hell} or the Underworld (42).  After this we get the first occurrence of the refrain-stanza; it describes how the wolf breaks free, and is repeated several times throughout the part of the spae that deals with the destruction of the world (43).

In \Hauksbok\ the structure is quite different.  The Eese immediately go to decide what action to take regarding the wall-builder (24-25), and Homedal’s hearing is described (26). Then follow the two sts. about the wolves that will swallow the sun and moon (40-41), and after this Goldwey and the war between the Eese and Wanes (20–23).  The stanzas about her origin (27–29), the walkirries (30) and the death of Balder (31–34) are left out of \Hauksbok\ along with the sword-river and beer-halls (35–36), but she does describe the binding of Lock (34), after which we get the first occurrence of the refrain-stanza; then the crowing of the three cocks (41–42), and lastly the punishment of criminals (38–39).  \Hauksbok\ then repeats the refrain-stanza (43) a second time, and \Regius\ and \Hauksbok\ are now almost completely aligned for the rest of the poem.

The destruction of the world begins with numerous catastrophes and a decay in human morality: kinslaying, incest, harlotry, and great wars become widespread (44).  Homedal blows the great horn of Yell, the Ash of Ugdrassle is set ablaze, and a trapped ettin comes loose (45).  The Eese, Elves and Dwarfs are all distressed (46).  Here the refrain-stanza is repeated again (47).  Now the enemies of the Gods launch their attack: the ettin Rim drives from the east; the Wyrm Ermingand writhes in the sea and propels huge waves; and the ship Nailfare comes loose (48).  The ship drives from the east and is led by Lock and his son the Wolf who are now free (49).  From the south Surt brings a flaming sword to set fire to the world; there is a crack in Heaven and Hell is joined by many men (50).  Weden goes out to fight the Wolf and falls, as does Free against Surt (51).  Wider avenges Weden by killing the Wolf (52), while Thunder and the Wyrm kill each other; the great battle leaves the world uninhabited (53).  Instead of 52 \Hauksbok\ has a stanza describing how the Wyrm gapes over the entire world, and how Thunder will ‘meet its venom’ (H3).  The destruction culminates in the sun turning black, the stars disappearing, and earth sinking into the sea.  The whole world burns (54).  The refrain-stanza repeats a final time (55).

Yet the world is reborn; the earth rises from the sea and soon turns green again; nature lives on (56).  The surviving Eese find eachother on the Idewolds, where they had first settled, and reminisce (57).  A new Golden Age begins, and the fields grow high without toil.  Hath and Balder reconcile.  Heener becomes the head priest of the Gods, and the sons of Hath and Balder settle the sky (58–60).  The wallow sees a fair, golden hall in which virtuous generations reside (61), but her vision concludes with the sight of a flying, dusky dragon—it is Nithehewer, carrying corpses in its mouth, a sign that there is still some evil left in the world.  With this portent she sinks back into her grave (62).

\newpage

\section{Text}

\bvg\bva\mssnote{\Regius~1r/2, \Hauksbok~20r/1}%
\Ballnote{The wallow has just been awoken from her grave (cf. st. 62/4b below) and commences her speech by asking for the silence of gods and men, a meristic expression \parencite[99--100]{West2007}.  The whole introductory formula has Indo-European parallels; see \textcite[63,92--93,312]{West2007}.}%
„\edtrans{\alst{H}ljóðs bið’k}{For hearing I ask}{\Bfootnote{The same introductory expression is found in st. 2 of Eyel’s Head-ransom (Egill \emph{Hfl} in \Skp\ 5): \emph{hljóðs biðjum hann} ‘for hearing we [I] ask him’.}} allar \hld\ \edtrans{\alst{h}ęlgar}{holy}{\Afootnote{so \Hauksbok; om. \Regius}\Bfootnote{That the omission of this word in \Regius\ is nothing more than a scribal error is clearly shown by the meter; the a-verse in \emph{Hljóðs bið \alst{e}k \hld\ \alst{a}llar kindir} is only three syllables long, and has highly unnatural alliteration on the unstressed \emph{ek} rather than the expected first nominal \emph{hljóðs}.}} kindir, &
\edtrans{\alst{m}ęiri ok \alst{m}inni}{greater and lesser}{\Bfootnote{It is ambiguous to which noun phrase these adjectives belong.  It can either be (a) “all greater and lesser holy races”, which might be equivalent to the phrase \inx[F]{Eese and Elves} (earthly and heavenly supernatural beings; see Index for occurrences), or (b) “the greater and lesser sons of Homedal”.  (b) is to be preferred as the simpler reading since it avoids enjambment, in which case the sense of “greater or lesser” is more likely to be literal and physical (i.e., the younger and older members of the audience) rather than figurative (i.e., the varying social classes).}} \hld\ \edtrans{\alst{m}ǫgu Hęim·dalar}{sons of Homedal \ken{men}}{\Bfootnote{As told in \textlink{Rigsthula}, Homedal sired the three castes of men.}}; &
\alst{v}ilt at, \edtrans{\alst{V}al-fǫðr}{Walfather}{\Bfootnote{That is, “Father of the Slain”, a name for Weden.  The wallow probably uses this name since Weden, acting in his role as necromancer and psychopomp, has just awoken her from her grave (perhaps with a \emph{val-galdr} ‘slain-galder’ as in \textlink{Baldrsdraumar}[4]/3b).}}, \hld\ \alst{v}ęl fram tęlja’k &
\alst{f}orn spjǫll \alst{f}ira, \hld\ \edtrans{þau’s \alst{f}ręmst of man}{which I earliest recall}{\Bfootnote{Cf. \textlink{Vafthrudnismal}[34]–35 with similar phrasing.}}?\eva

\bvb “{\huge F}\textsc{or hearing I ask} all holy races \ken{gods}, \\
the greater and lesser sons of \inx[P]{Homedal} \ken{men}! \\
Wilt thou, \inx[P]{Walfather}, that I well tell forth \\
the ancient sayings of men which I earliest recall?\evb\evg


\bvg\bva\mssnote{\Regius~1r/4, \Hauksbok~20r/2}%
\alst{E}k man \alst{jǫ}tna \hld\ \alst{á}r of borna, &
þȧ’s \alst{f}orðum mik \hld\ \alst{f}ǿdda hǫfðu; &
\alst{n}íu man’k hęima, \hld\ \alst{n}íu \edtext{ï·viðjur}{\Afootnote{so all; previously erroneously \emph{‘iviði’} \Regius.}\Bfootnote{Evil-working women or ogresses; their identity is by no means clear.  The fem. \emph{jōn}-stem \emph{ï·viðja} also appears in a list of poetic synonyms for trollwomen (Þul \emph{Trollkvenna} 3 in \Skp\ 3).  A commonly suggested etymology is \emph{í} ‘in’ + \emph{viðr} ‘tree, forest’ (i.e. ‘tree, forest-dwellers’), but this would be an unusual formation and leaves the \emph{-j-} unexplained.  A more plausible etymology is a derivative of \emph{*ï·við} ‘malice’, attested in the cpd. \emph{ï·við-gjarn} ‘malicious’ (\textlink{Volundarkvida}[28]).
This derivation can also explain the \emph{-j-}, since the WGmc. cognates OE \emph{in·wid}, OS \emph{in·wid}, and OHG \emph{in·wit} show it to be a neutr. \emph{ja}-stem. — The old reading \emph{íviði} ‘in the tree’ \Regius\ was made obsolete by an x-ray scan undertaken by \textcite{StefanKarlsson1979} which revealed a previously overlooked abbreviation mark standing for \emph{-ur}.}}, &
\edtrans{\alst{m}jǫt-við \alst{m}ę́ran \hld\ fyr \alst{m}old neðan.}{the famed Measure-Tree beneath the soil.}{\Bfootnote{Probably \inx[P]{Ugdrassle’s Ash}, still but a seed.  The sense of \emph{mjǫt-viðr} ‘Measure-Tree’ is not clear, but it seems to be the same word as \emph{mjǫtuðr} in 45/1b below.}}\eva

\bvb I recall \inx[P]{Ettins} born of yore, \\
those who formerly had reared me. \\
Nine \inx[C]{Home}[Homes] I recall, nine \inx[P]{Inwithies}, \\
the famed Measure-Tree beneath the soil.\evb\evg


\bvg\bva\mssnote{\Regius~1r/6, \Hauksbok~20r/4, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\Ballnote{Told in sts. 3–4 is the creation of the Earth and the Heavens.  Other Norse poetic sources for the creation are \textlink{Grimnismal}[41]–42, \textlink{Vafthrudnismal}[21].  A Christian creation hymn apparently drawing on older pagan sources is the OHG \textlink{Wessobrunn} edited below under Poetry on Christian Subjects; cf. also \textlink{Cadman} in the same section.  The Norse creation narrative is treated at length in \Gylfaginning\ 4–8, according to which the world first consisted of two extremities, viz. the frozen Nivelham in the north and the burning Muspellsham in the south.  From Nivelham in the north the icy venomous rivers called the \inx[P]{Ilewaves} rushed forth until they froze to ice, while from Muspellsham in the south there flowed burning lava.  The ice and the lava eventually met in the Gap of Ginnings (\emph{Ginnunga-gap}, which \emph{var svá hlę́tt sem lopt vind-laust} ‘was as calm as windless air’) and there combined to form the first being, \inx[P]{Yimer}, who was the ancestor of the ettins and the victim of the First Sacrifice.}%
\alst{Á}r vas \alst{a}lda \hld\ \edtrans{þar’s \alst{Y}mir byggði}{where Yimer dwelled}{\Afootnote{\emph{þat’s ękki vas} ‘when nothing was’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}}, &
vas⸗a \alst{s}andr né \alst{s}ę́r, \hld\ né \alst{s}valar unnir; &
\edtext{\alst{jǫ}rð fann⸗sk \alst{ę́}va \hld\ né \alst{u}pp-himinn}{\lemma{jǫrð \dots\ né upp-himinn ‘Earth \dots\ nor Up-heaven’}\Bfootnote{A well-attested formulaic cosmological word-pair found in all four Old Germanic languages with alliterative poetic traditions (ON, OE, OS, and OHG), especially in the context of the creation and destruction of the world.  See Index: \inx[F]{Earth and Upheaven}.}}; &
\edtrans{\alst{g}ap vas \alst{g}innunga}{there was the gap of hawks \ken{air/midspace}}{\Bfootnote{\emph{gap ginnunga} ‘gap of hawks’ has usually been interpreted on the basis of the cosmogony in \Gylfaginning\ 4–5, where Snorre presents \emph{Ginnunga-gap} ‘the Gap of Ginnings’ as a primordial physical location where cold and hot elements combined to form the first sentient being, Yimer, in the place where the Earth would later be situated.  For this reason the phrase is typically rendered into English simply as “Ginnungagap” and explained as a sort of supernatural void.  The present stanza, however, is the only conjunction of the words \emph{gap} and \emph{ginnunga} outside of Snorre’s Edda, and his cosmogony is very peculiar.  There is therefore good reason to reexamine these two words in the present context in order to determine the exact nature of the ‘gap’.

To begin, we must reject the traditional translation “yawning chaos” as linguistically unfounded.  The oft-connected ON verb \emph{gína} ‘yawn, gape’ has a different root (\emph{gín-}) than \emph{ginnunga} (\emph{ginn-}), which further appears to be a gen. sg. or pl. rather than an adjective.  The explanation ‘width, breadth’ (from a lost equivalent to OE \emph{ginne} ‘broad, vast’ + the nominal suffix \emph{-ung}) is somewhat more plausible, but ‘gap of widths’ is still a strange and decidedly non-Norse construction.

I instead agree with \citeauthor{Meissner1921} in reading \emph{gap ginnunga} as a kenning ‘gap of hawks \ken{air/midspace}’, where \emph{ginnunga} is gen. pl. of the attested poetic word \emph{ginnungr} ‘hawk’.
The proposed kenning is conventional and belongs to the type “land, path of the bird \ken{air}” \parencite[108]{Meissner1921}; indeed, the determinant \emph{ginnungr} is also found in a \textsc{sky}-kenning in \Haustlong\ 15: \emph{ǫll endi-lǫ́g ginnunga vé} ‘all the low mansions of hawks \ken{skies} from end to end’.
This interpretation is further confirmed by an underappreciated passage in \Skaldskaparmal\ 74 which lists \emph{ginnunga-gap} among poetic synonyms for the air: \emph{Lopt heitir ginnunga-gap ok meðal-heimr, fogl-heimr, veðr-heimr.} ‘Air is called \emph{ginnunga-gap} (‘gap of hawks’) and middle-home, bird-home, weather-home.’  The sense of \Voluspa ’s \emph{gap ginnunga} is thus the same as \Haustlong’s \emph{vé ginnunga}, viz. ‘air, atmospheric sky’.

Having determined the sense of the phrase, we should consider its significance in the present context.  This requires a cosmological excursion.
The Old Germanic pagan cosmology, like that of many other ancient religions (e.g. the Vedic with its two World-Halves—OI \emph{ródasī}—\Rigveda\ 1.10.8, 5.85.3 et c.) sees the cosmos as consisting of two world-halves: Earth and Upheaven, the land of Men and the land of Gods.  The Earth is held to be a flat disc, while the Heaven is a round dome above it, upheld by four dwarfs lest it fall down (\Gylfaginning\ 8, cf. Arn \emph{Þorfdr} 24 in \Skp\ 2).
Both Earth and Heaven are believed to be physically solid realms and made out of different parts of Yimer’s body: the Earth from his torso, the Heaven from his skull (\textlink{Grimnismal}[41], \textlink{Vafthrudnismal}[21]).  Like the Biblical Firmament, the Heaven is made out of stone (\textlink{Harbardsljod}[15]), which also explains the falling of meteors.

Between the two solid World-halves are the atmospheric skies, the Air (ON \emph{lopt}) or the Midspace (ON \emph{meðal-hęimr} ‘middle realm’, equivalent to the \emph{antárikṣam} of the Vedic hymns, e.g. \Rigveda\ 4.40.5, 5.85.3) wherein the clouds—Yimer’s brains—float (\textlink{Grimnismal}[42]).  The distinction between skies and Heaven is further seen in the aforementioned \Haustlong, which distinguishes between the ‘low \textsc{skies}’ (\Haustlong\ 15) and \emph{upp-himinn} ‘Upheaven’ (\Haustlong\ 16).  Thus, the primordial state expressed in the present stanza is one where there is only air, while Earth and Stoney Heaven are still conjoined in the body of Yimer.}}, \hld\ ęn \alst{g}ras \edtrans{hvęr⸗gi}{nowhere}{\Afootnote{\emph{ękki} ‘not’ \Hauksbok}},\eva%TODO: bibliography, Meissner is 20. Luft, b

\bvb It was early of ages where \inx[P]{Yimer} dwelled; \\
there was not sand nor sea nor cool waves. \\
\inx[P]{Earth} was never found nor \inx[P]{Up-heaven}; \\
there was the \inx[P]{gap of hawks} \ken{air/midspace} but grass nowhere,\evb\evg


\bvg\bva\mssnote{\Regius~1r/8, \Hauksbok~20r/5}%
\Ballnote{The Sons of Byre—the first Gods—sacrificed the primordial ettin Yimer through decapitation and created the World out of his body, as told in \textlink{Grimnismal}[41]–42, \textlink{Vafthrudnismal}[21], and \Gylfaginning\ 7–8.  This may be seen as contradictory to the notion of their lifting the lands, namely out of from the waters which covered them, something clearly implied in st. 56 below.  This contradiction can be explained in two ways.  (1) According to \Gylfaginning\ 7 the killing of Yimer flooded the world with so much blood that it drowned all the ettins except for Bearyelmer and his family, who survived by taking refuge on his canoe (ON \emph{lúðr}, see \textlink{Vafthrudnismal}[35] for discussion on this word).  Thus, the present stanza likewise alludes to this primeval post-sacrificial flood.  (2) If the \Gylfaginning\ (\RegiusProse\Trajectinus\Upsaliensis\Wormianus) reading of 3/1b (\emph{þat’s ękki vas} ‘when nothing was’) is authentic, the wallow may not have intended to allude to the myth of Yimer at all but instead have been thinking of something like Genesis 1:2, 9–10, where Elohim makes dry earth rise up from the deep waters.  The problem with (2) is that st. 3/2 clearly states that there was not yet any sea, and as such (1) should be preferred.}%
áðr \edtrans{\alst{B}urs synir}{the Sons of Byre}{\Bfootnote{In \Gylfaginning\ 6–8 identified as the three brothers Weden, Will, and Wigh, the first Gods.}} \hld\ \edtrans{\alst{b}jǫðum of ypðu}{revealed the flatlands}{\Bfootnote{The verb \emph{yppa} (here ‘reveal’) is a causative derived from ON \emph{upp} ‘up’, \emph{uppi} ‘above, up, on high, remembered’ (cf. below st. 15/6a).  The translation as ‘reveal’ is based on its poetic sense ‘extol, divulge’ and its OE cognate \emph{yppan} ‘bring up, bring forth, reveal’; if the intended sense were simply a ‘lifting’, the verb \emph{hęfja} (in the 3pl. past. ind. \emph{hófu}) would be expected.}}, &
þęir es \edtrans{\alst{M}ið-garð}{Middenyard}{\Bfootnote{The Middle Enclosure, which was created as the home of Men (\textlink{Grimnismal}[42]).}} \hld\ \alst{m}ę́ran skópu; &
\alst{s}ól skęin \alst{s}unnan \hld\ ȧ \alst{s}alar stęina; &
þȧ vas \alst{g}rund \alst{g}róin \hld\ \edtrans{\alst{g}rø̇num lauki}{green leek}{\Bfootnote{A sign of the golden age, for the leek was in ancient times held to be the noblest plant. See Index.}}.\eva

\bvb before the \inx[P]{Sons of Byre} revealed the flatlands, \\
they who created famed \inx[P]{Middenyard}. \\
The sun shone from the south on the stones of the hall; \\
then was the ground grown with green leek.\evb\evg


\bvg\bva\mssnote{\Regius~1r/11, \Hauksbok~20r/7, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\edtext{\alst{S}ól varp \alst{s}unnan, \hld\ \edtrans{\alst{s}inni Mȧna}{Moon’s companion}{\Bfootnote{At times mistranslated as ‘her moon’, understanding \emph{sinni} as dat. sg. f. of \emph{sïnn} ‘its (reflexive)’.  This cannot be correct since ON possessives are inflected based on the gender of the noun possessed, not the gender of the possessor.  \emph{mȧni} ‘moon’ is masculine, and so ‘her moon’ would be \emph{*sïnum Mȧna}.}}, &
\alst{h}ęndi hinni \alst{h}ǿgri \hld\ of \edtrans{\alst{h}imin-jǫður}{heaven’s rim}{\Afootnote{composite; \emph{‘himin †iodyr†’} \Regius; \emph{jǫður (‘ioður’)} \Hauksbok.}\Bfootnote{Some recent editors have taken it upon themselves to normalize the reading of \Regius\ as \emph{himin-jó-dýr} ‘heaven-horse-beast’, which is not just nonsensical but also unmetrical due the stress pattern.  On the other hand the reading of \Hauksbok\ (norm. \emph{jǫður} ‘rim, edge’) is clearly deficient since it lacks the necessary alliteration on \emph{h}.
If we instead see \emph{iodyr} \Regius\ as corrupted from an archetypal \emph{*iodur} shared by \Hauksbok, we can restore \emph{*himin-jǫður}, as done here.}}}{\lemma{Sól \dots\ himin-jǫður ‘Sun \dots\ heaven’s rim’}\Afootnote{om. \RegiusProse\Trajectinus\Upsaliensis\Wormianus.}\Bfootnote{Probably a poetic description of the dawn; the Sun lifted herself up over the horizon and rose for the first time.}}; &
\alst{S}ól þat né vissi, \hld\ hvar hǫ̇n \alst{s}ali átti; &
\edtext{\edtrans{\alst{M}ȧni þat né vissi, \hld\ hvat hann \alst{m}ęgins átti}{the Moon knew not what sort of might he had}{\Bfootnote{For a concrete example of this “might” cf. \textlink{Havamal} 137/7, where the moon is to be invoked in a blood feud.  A belief in the power of the moon is found in most ancient religions.  Its influence on the human mind was remarked already by \textcite{PlinyNatHist}, 2.102, 18.75; witness in folklore the influence of the full moon on the werewolf, or in English the word \emph{lunacy} (from Latin \emph{lūna} ‘moon’).  Women are particularly affected by the moon through the synchronisation of the monthly and menstrual cycles (the latter word deriving from Latin \emph{mēnsis} ‘month’, originally ‘moon’) as caused by the moon’s gravitational pull on the body \parencite{Forster2025}.}}; &
\alst{st}jǫrnur þat né vissu, \hld\ hvar þę́r \alst{st}aði ǫ́ttu}{\lemma{Mȧni \dots\ ǫ́ttu ‘the Moon \dots\ they had.’}\Afootnote{These lines are reversed (i.e., Sun, Stars, Moon) in \Regius\Hauksbok.}}.\eva

\bvb The Sun cast from the south—the Moon’s companion— \\
her right hand over heaven’s rim. \\
The Sun knew not where halls she had; \\
the Moon knew not what sort of might he had; \\
the Stars knew not where seats they had.\evb\evg


\bvg\bva\mssnote{\Regius~1r/13, \Hauksbok~20r/9}%
\edtext{Þȧ gingu \edtrans{\alst{r}ęgin ǫll}{all the Reins}{\Bfootnote{All the Gods (\emph{ręgin} ‘Reins’, i.e. ‘Powers, Counsels’; a pantheistic word), here acting as one body.}} \hld\ ȧ \edtrans{\alst{r}ǫk-stóla}{rake-seats}{\Bfootnote{Their thrones of judgment.}}, &
\alst{g}inn-hęilǫg \alst{g}oð, \hld\ ok umb þat \alst{g}ę̇ttu⸗sk}{\lemma{Þȧ \dots\ gę̇ttu⸗sk ‘Then \dots\ of that.’}\Bfootnote{A refrain (\emph{stęf}) describing the convening of the \inx[P]{Thing of the Gods} repeated below in sts. 9/1–2, 22/1–2, and 24/1–2.  Cf. also the three lines shared between \textlink{Baldrsdraumar}[1]/1–3 and \textlink{Thrymskvida}[14]/1–3, which follow the same structure: \emph{Sęnn vǫ́ru ę̇sir \hld\ allir ȧ þingi // ok ǫ̇synjur \hld\ allar ȧ máli, // ok umb þat réðu \hld\ ríkir tívar.} ‘Soon were the \inx[P]{Eese} all at the \inx[C]{Thing}, // and the \inx[P]{Ossens} all at speech, // and of this counseled the mighty \inx[P]{Tews}.’

In every occurrence of either formulae outside of the present stanza, the demonstrative pronoun \emph{þat} ‘this, that’ clearly refers to an immediately following question introduced by a \emph{hv}-word which describes a problem the Gods have to deal with (e.g. \textlink{Thrymskvida}[14]/4: \emph{hvé þęir Hlór·riða \hld\ hamar of sǿtti?} ‘how they Loride’s \name{= Thunder’s} hammer would find?’)  Following this pattern we would expect to find such a question following \emph{umb þat gę̇ttu⸗sk} ‘took counsel of that’ in the present stanza, and it seems reasonably plausible (but not certain) that such has been lost in transmission.}}. &
\edtrans{\alst{N}ǫ́tt ok \alst{n}iðjum}{night and the moon-phases}{\Bfootnote{Also found in \textlink{Vafthrudnismal}[24]/4.}} \hld\ \edtrans{\alst{n}ǫfn of gǫ́fu}{names they gave}{\Bfootnote{The giving of names to the natural phenomena is seen as a part of the Godly ordering of the universe; in fact it is the very act of naming and thereby categorizing these phenomena which sets them apart from the disordered Natural chaos.  These categories, rather than being arbitrarily invented by Men, are thus seen as stemming directly from the Gods.  In a modern scientific context this passage can be read as a celebration of the work of such men as Linnæus, who by their actions in classifying and rationalizing nature are thus carrying out the will of the Gods.  This is not the same as the positive Christian attitude towards science, which is based on the idea of understanding the perfect design of the intelligent Creator.  In the Germanic worldview, however, the raw matter already exists (in the form of Yimer and other Ettins) before the Gods, and the Creation consists in shaping and classifying it according to Godly Will (which appears synonymous with rationality)—in the present stanza in order to reckon time and keep dates.}}, &
\alst{m}orgin hétu \hld\ ok \alst{m}iðjan dag, &
\alst{u}ndurn ok \alst{a}ptan, \hld\ \edtrans{\alst{ǫ́}rum at tęlja}{the years for to tally}{\Bfootnote{Cf. \textlink{Vafthrudnismal}[22]–25, where it is said that the Gods made the sun and moon turn round in heaven and created the moon-phases \emph{ǫldum at ár-tali} ‘for mankind’s tally of years’.}}.\eva

\bvb Then went all the Reins onto the rake-seats: \\
the Yin-holy Gods, and from each other took counsel of that. \\
To night and the moon-phases names they gave; \\
morning they named, and middle day, \\
afternoon and evening, the years for to tally.\evb\evg


\bvg\bva\mssnote{\Regius~1r/16, \Hauksbok~20r/10}%
Hittu⸗sk \alst{ę̇}sir \hld\ ȧ \edtrans{\alst{I}ða-vęlli}{Idewolds}{\Bfootnote{“The plains of industry.”}}, &
\edtext{þęir’s \alst{h}ǫrg ok \alst{h}of \hld\ \alst{h}ǫ́-timbruðu}{\lemma{þęir’s \dots\ hǫ́-timbruðu ‘they who \dots\ timbered on high’}\Afootnote{\emph{afls kostuðu \hld\ alls freistuðu} ‘[their] strength they tried; everything they tempted’ \Hauksbok}\Bfootnote{Two formulæ. — \emph{hǫrgr ok hof} ‘harrow and hove’ is a merism, i.e. ritual structures made of stone and wood; cf. \textlink{Vafthrudnismal}[38] and \textlink{HelgakvidaHjorvardssonar} TODO, as well as the Norwegian Christian laws that impose ‘the burning of hoves and the breaking of harrows’ (\emph{brenna hof ok brjóta hǫrga}). — \emph{hǫ́-timbra} ‘timber on high’ is a rare compound. Its only other occurrence in the ON corpus is in \textlink{Grimnismal}[16], where it describes a harrow ruled by Nearth. — This passage has often been wondered at.  Why would the Gods themselves make cultic buildings?  Whom do \emph{they} worship?
Let it be added to the present instance that they partake in the ritual slaughter of beasts, and the following divination and feasting on their flesh (e.g. \textlink{Voluspa}[61], \textlink{Hymiskvida}[1], 39, \textlink{Lokasenna}, \Haustlong\ 2).  This question is really not as difficult as it is made out to be, however, since the answer is surely that the behaviour of the Gods serves as the model for virtuous human behaviour and justifies the ideology of settlement.  Colonising new lands, tilling the earth, building houses and enclosures and ritual structures, sacrificing beasts—if these are virtuous for humanity, they must find their model among the Gods.  That does not mean that the Gods worship themselves (or anything else); the Gods operate on the level of the macrocosm, and it is from them that the Godly Will flows.  Humans, on the level of the microcosm, connect with this Godly Will through imitating the Gods on a smaller scale, including by worship.  Thus the sacrifice of Yimer (who was the Earth and Heaven conjoined) and the enclosing of the great Middenyard are reenacted through the sacrifices of horses and oxen within ritual and settlement enclosures.  See further \textlink{Grimnismal}[41]–43 and notes.}}; &
\alst{a}fla lǫgðu, \hld\ \alst{au}ð smíðuðu, &
\alst{t}angir skópu \hld\ ok \alst{t}ól gęrðu.\eva

\bvb The Eese met each other on the \inx[P]{Idewolds}, \\
they who \inx[C]{harrow} and \inx[C]{hove} timbered on high. \\
Hearths they laid, wealth they smithed, \\
tongs they shaped and tools they made.\evb\evg


\bvg\bva\mssnote{\Regius~1r/18, \Hauksbok~20r/12}%
\Ballnote{The whole stanza is paraphrased in \Gylfaginning\ ch. 14: \emph{Ok því nę́st smíðuðu þeir málm ok stein ok tré ok svá gnóg⸗liga þann málm, er gull heitir, at ǫll bús-gǫgn ok ǫll reiði-gǫgn hǫfðu þeir af gulli, ok er sú ǫld kǫlluð gull-aldr, áðr en spillti⸗st af til-kvámu kvinna’nna; þę́r kómu ór Jǫtun-heimum.} ‘And after this they forge ore and stone and wood, and so abundantly [did they forge] that ore which is called gold, that all their household tools and riding tools were golden.  And that age is called the golden age, before it was spoiled by the arrival of the women; they came out of the Ettinhomes.’}%
\edtrans{\alst{T}ęflðu}{played Tables}{\Bfootnote{\emph{tęfla} is a causative verb derived from \emph{tafl} ‘board game’, an old borrowing from Latin \emph{tabula}.  “Tables” is used as a cognate translation; the exact type of board game referred to is unimportant, although it is probably a game of strategy (e.g. \emph{hnefa-tafl} or nine men’s morris, for chess had not yet been introduced in the late C10th) rather than one of chance.}} ï \alst{t}u̇ni, \hld\ \alst{t}ęitir vǫ́ru, &
\edtrans{\alst{v}as þęim \edtrans{\alst{v}éttir-gis}{nothing}{\Bfootnote{An archaic gen. of \emph{vę́tt-ki} ‘nothing’; the \emph{-ir} representing a fossilized i-stem genitive, for \emph{véttr} ‘thing’ comes from PGmc. \emph{*wihtiʀ}.  The only other occurrence of this form is in the highly linguistically archaic Icelandic Homily Book (ms. Holm perg 15 4°, fol. 36v/30).}} \hld\ \alst{v}ant ór golli}{for them was nothing golden wanting}{\Bfootnote{Indeed even the bricks they played with were of gold. See st. 58.}}, &
und’s \edtext{\alst{þ}ríar kvǫ̇mu \hld\ \alst{þ}ursa męyjar}{\lemma{þríar \dots\ þursa męyjar ‘three maidens of Thurses’}\Bfootnote{These three maidens are never mentioned again (unless they are taken to be the three norns in st. 19, but they would then be introduced twice). It is possible that an additional stanza giving further information about them has been lost. If it originally existed, it was already absent in the version used for \Gylfaginning, since no additional information is found there.}}, &
\edtrans{\alst{ȧ}m-átkar}{uncanny}{\Bfootnote{The word \emph{ám-áttigr} has a clear association with supernatural beings; trolls and ettins. It occurs in four other places in \Regius. In \textlink{Grimnismal}[11], \textlink{Skirnismal}[10] and \textlink{HelgakvidaHjorvardssonar}[17] it modifies \emph{jǫtunn} ‘ettin’ in a \Ljodahattr\ c-line. In \textlink{HelgakvidaHjorvardssonar}[14] it is used by the daughter of an ettin to refer to a human hero.}} mjǫk, \hld\ ór \alst{Jǫ}tun-hęimum.\eva

\bvb They played \inx[C]{Tables} in the yard, merry were they, \\
for them was nothing golden wanting \\
until three maidens of \inx[P]{Thurses} came, \\
most uncanny, out of the \inx[P]{Ettinham}[Ettinhomes].\evb\evg


\bvg\bva\mssnote{\Regius~1r/20, \Hauksbok~20r/14, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\Ballnote{After the Golden Age is spoiled, the Gods must get their gold in some other way.  For this they need the Dwarfs, who are associated with precious metals and ore-veins, perhaps through techniques ancestral to dousing (i.e., the dwarf being the entity that moves the dousing rod).  The ancient conception of spontaneous generation is clearly at play.  According to that view, maggots and worms were thought to spontaneously generate in rotting flesh and earth, and the dwarfs were apparently likewise thought to generate in rocks.  Thus is stated in \Gylfaginning\ 14, which continues with its paraphrase:

\emph{Þar nę́st settu⸗st goð’in upp í sę́ti sín ok réttu dóma sína ok minntu⸗st, hvaðan dvergar hǫfðu kviknat í mold’inni ok niðri í jǫrðu’nni, svá sem maðkar í holdi.  Dvergar’nir hǫfðu skipat⸗st fyrst ok tekit kviknun í holdi Ymis ok váru þá maðkar, en af at·kvę́ðum goða’nna urðu þeir vitandi mann-vits ok hǫfðu manns líki ok búa þó í jǫrðu ok í steinum.  Móð·sognir var ǿðstr ok annarr Durinn.  Svá segir í Vǫluspá:}

‘Thereafter the gods set themselves high in their seats and made their judgments and remembered whence the Dwarfs had come to life in the ground and down in the earth like maggots in flesh.  The dwarfs had first taken shape and come to life in Yimer’s flesh and were then maggots, but by the decrees of the gods they became knowing of manwit and had a man’s form, and yet they live in the earth and in stones.  Moodsowner was the highest in rank and second Dorn.  So it says in the Spae of the Wallow:’

after which the text quotes the present st. and 10/3–4.}%
Þȧ gingu \alst{r}ęgin ǫll \hld\ ȧ \alst{r}ǫk-stóla, &
\alst{g}inn-hęilǫg \alst{g}oð, \hld\ ok umb þat \alst{g}ę̇ttu⸗sk: &
\edtrans{Hvęrr skyldi \alst{d}verga}{Who would \dots\ of dwarfs’}{\Afootnote{so \Regius\Wormianus\Upsaliensis; \emph{at skyldi dverga} ‘That they would \dots\ of dwarfs’ \RegiusProse\Trajectinus; \emph{hverir skyldu dvergar} ‘Which dwarfs would [shape the retinues]’ \Hauksbok}} \hld\ \edtrans{\alst{d}rótt}{the retinue}{\Afootnote{so \RegiusProse\Trajectinus\Upsaliensis\Wormianus; \emph{drotin} ‘the lord’ \Regius; \emph{dróttir} ‘the retinues’ \Hauksbok}} \edtrans{of skępja}{shape}{\Afootnote{\emph{spekia} ‘soothe’ \Upsaliensis}} &
\edtrans{ór \edtrans{\alst{b}rimi \alst{b}lóðgu}{bloody surf}{\Afootnote{so \Hauksbok\RegiusProse\Wormianus\Upsaliensis; \emph{Brimis blóði} ‘the blood of Brimmer’ \Regius\Trajectinus}} %
\hld\ ok ór \edtrans{\alst{b}lǫ́um}{blue-black}{\Afootnote{metr. emend. from \emph{blám} \Regius; \emph{Bláins} ‘Blown’s’ \Hauksbok\Wormianus; \emph{Bláms} \RegiusProse\Trajectinus\Upsaliensis\ is prob. a corrupt form of \emph{Bláins}}} lęggjum}{out of the bloody surf and the blue-black legs}{\Bfootnote{The poem simply telling of “the bloody surf” and “the blue-black legs” fits better with its general allusive style, but this requires a composite reading.
If we read \emph{Bláinn} ‘Blown’ (named in the \inx[C]{thule}[thules] as a dwarf) instead of \emph{blǫ́um} ‘blue-black’, then following Gurevich (\emph{Skp} 2017, p. 693) we may see a kenning “the legs of Blown \name{dwarf} \ken{stones}”.  Blown has otherwise usually been read as a poetic name for Yimer, but it is not attested anywhere else. —
The “blood” and “legs” are in any case those of Yimer; from his bones were made the rocks and from his blood the sea (\textlink{Grimnismal}[41], \textlink{Vafthrudnismal}[21]).  Dwarfs of course dwell in rocks and earth; cf. for instance \Ynglingatal\ 2, where the Swedish king Swayther (ON \emph{Svęigðir}) runs into a rock in pursuit of a dwarf.  More difficult to explain is how the dwarfs would be made from the sea.  Einheri suggests that it may be referring to the formation of salt-stones by means of evaporating salty seawater.  A simpler reading is that it refers merely to underwater stones as in \textsc{rock}-kennings of the type \emph{sę́var bęin} ‘bone of the sea’ (\Haleygjatal\ 2), \emph{fjarð-ępli} ‘firth-apple’ (\Thorsdrapa\ 16) \parencite[90]{Meissner1921}.}}?\eva

\bvb Then went all the Reins onto the rake-seats: \\
the Yin-holy Gods, and from each other took counsel of this: \\
Who would shape the retinue of \inx[P]{Dwarfs}, \\
out of the bloody surf and the blue-black legs?\evb\evg


\bvg\bva\mssnote{\Regius~1r/21, \Hauksbok~20r/15, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\edtext{\edtext{Þar vas \alst{M}óð·sognir}{\Afootnote{so \Hauksbok; \emph{Þar Mót·sognir †vitnir†} ‘there Mootsowner wolf(?)’ \Regius. The prose of \Gylfaginning\ 14 agrees with \Hauksbok\ that the correct form of the name is \emph{Móð·sognir}, not \emph{Mót·sognir}.}} \hld\ \alst{m}ę́tstr of orðinn &
\alst{d}verga allra, \hld\ en \alst{D}urinn annarr;}{\lemma{Þar \dots\ annarr ‘There \dots\ second’}\Afootnote{om. \RegiusProse\Trajectinus\Upsaliensis\Wormianus}{\Bfootnote{These lines are not found in \Gylfaginning, but the author must have had the full stanza since he paraphrases them (see st. 9 n. above).}}} &
\edtext{\edtrans{þęir \edtrans{\alst{m}an-líkun}{man-likenesses}{\Bfootnote{The Dwarfs were in the shape of Men, but not quite human; thus the creation is seen as progressive, with the man-shaped Dwarfs preceding the human beings.  The Dwarfs, arising out of the earth, also lack the Godly spark found in Men.}} \hld\ \alst{m}ǫrg of gørðu}{Many man-likenesses they made’}{\Afootnote{so \Regius\Hauksbok\Upsaliensis; \emph{þar man-líkun \hld\ mǫrg of gørðu⸗sk} ‘There many man-likenesses were made’ \RegiusProse\Trajectinus\Wormianus}}, &
\alst{d}vergar \edtrans{ï}{in}{\Afootnote{so \RegiusProse\Trajectinus\Upsaliensis\Wormianus\Hauksbok; \emph{ór} ‘out of’ \Regius}} jǫrðu, \hld\ \edtrans{sęm \alst{D}urinn sagði}{as Dorn said}{\Afootnote{so \Regius\Hauksbok\RegiusProse\Wormianus; \emph{sem †dur menn† sagði} ‘as door-men(?) said’ \Trajectinus; \emph{sem †þeim dyrinn kendi†} ‘as the beasts(?) taught them’ \Upsaliensis}}.}{\lemma{þęir \dots\ sagði ‘They \dots\ said.’}\Bfootnote{The mss. readings offer two conflicting versions of the creation of the dwarfs.  Either they arose on their own; this is supported by the prose of \Gylfaginning\ 14 (see note to previous st.) and by the form of the stanza quoted there in the non-\Upsaliensis\ mss., but it may have been changed to correspond to the author’s vision.  On the other hand, \Regius, \Hauksbok\ and \Upsaliensis\ have the dwarfs Moodsowner and Dorn shaping “man-likenesses” out of soil.  The present ed. follows \Regius\Hauksbok\Upsaliensis.}}\eva

\bvb There was Moodsowner made the noblest \\
of all Dwarfs, but Dorn the second. \\
Many man-likenesses they made: \\
Dwarfs in the earth, as Dorn said.\evb\evg

\sectionline\setcounter{stanza}{15}

\begin{center}(For sts. \textbf{11–15}—the \emph{Dwarf-tallies}—see Appendix to the poem!)\end{center}

\sectionline

\bvg\bva\mssnote{\Regius~1v/1, \Hauksbok~20r/26}%
\Ballnote{Sts. 16–17 are paraphrased in \Gylfaginning\ 9: \emph{Þá er þeir gengu með sę́var-strǫndu Bors synir, fundu þeir tré tvau ok tóku upp trén ok skǫpuðu af menn.  Gaf inn fyrsti ǫnd ok líf, annarr vit ok hrę́ring, þriði á·sjónu, mál ok heyrn ok sjón.  Gǫ́fu þeim klę́ði ok nǫfn; hét karl-maðr’inn Askr, en kona’n Embla, ok ólst þaðan af mann-kind’in, sú er byggð’in var gefinn undir Mið-garði.} ‘When the sons of Byre (cf. st. 4) were walking along the sea-shore they found two trees (ON \emph{tré}, alt. ‘pieces of wood’) and they took up the trees and from them created Men.  The first one gave breath and life, the second intelligence and movement, the third outward appearance, speech, and hearing and sight.  They gave them clothes and names; the male was called Ash and the woman Emble, and from them was begotten mankind, to which the dwelling within Middenyard was given.’

Based on \Gylfaginning\ the creation of men is traditionally seen as referring to pieces of driftwood, but that may be a later Icelandic or Snorronean interpretation.  As pointed out by \textcite{Hultgård2006}, the evidence from comparative mythology suggests that the first humans were originally seen as living, growing trees, and there is really nothing in \textlink{Voluspa} that speaks against such an interpretation.  This myth is probably the reason why words for trees are used extensively by Norse poets in kennings for men and women (see \Skp\ I, p. lxxv ff., \cite[245,266--272, 410]{Meissner1921}), more commonly in Scaldic poetry, but at times also in Eddic poetry (e.g. \textlink{Sigrdrifumal}[5]: \emph{bryn-þings apaldr} ‘apple-tree of the byrnie-\inx[C]{Thing} \ken{battle > warrior}’).}%
\edtrans{Und’s \edtext{\alst{þ}r\emph{í}r}{\Afootnote{emend.; \emph{þrjár} \Regius\Hauksbok}} kvǫ̇mu \hld\ \edtext{ór \alst{þ}ví liði}{\Afootnote{so \Regius; \emph{þussa brúðir} \Hauksbok.}}%
}{Until three came out of that host}{\Bfootnote{This line, beginning with the conjunction “until”, is clearly supposed to connect to a previous stanza (cf. st. 4 \emph{áðr}, TODO: other occurrences).  If the dwarf-tallies (11–15), as seems likely, are later inserts, that would be st. 10, in which case the creation of men is perhaps to be understood as superseding the creation of the dwarfs; cf. note to st. 10/3: \emph{man-líkun}.

Both mss. show influence from st. 8 in using the fem. \emph{þrjár} instead of masc. \emph{*þrír}.  \Hauksbok\ goes further in replacing \emph{ór því liði} ‘out of that host’ with \emph{þussa brúðir} ‘brides of thurses’.  That these are errors is clearly shown by the masculine \emph{ǫflgir ok ȧstkir ę̇sir} in l. 2.}} &
\edtrans{\alst{ǫ}flgir ok \edtrans{\alst{ȧ}stkir}{loving}{\Bfootnote{The creation of men was an act of love; for humanity the Gods later made Middenyard (\textlink{Grimnismal}[42]); the moon-phases had already been created for our time-reckoning (st. 6 above).}}}{strong and loving}{\Afootnote{\emph{ȧstkir ok ǫflgir} (norm.) ‘loving and strong’ \Hauksbok}} \hld\ \alst{ę̇}sir \edtrans{at húsi}{along the houses}{\Bfootnote{An adverbial; the Gods were walking on the outskirts of their settlement.}}; &
fundu ȧ \alst{l}andi \hld\ \alst{l}ítt megandi &
\edtrans{\alst{A}sk ok \alst{Ę}mblu}{Ash and Emble}{\Bfootnote{Ash (nom. \emph{Askr}) is easily identified with the same-named wood species (\emph{Fraxinus excelsior}), but the etymology of Emble (nom. \emph{Ęmbla}) is much more difficult to explain.  Her name is often translated as “Elm” (so Neil Price), but the ON word for that tree is the masc. \emph{almr} ‘elm’.  Metathesis from earlier \emph{*Ęlma}, a derivative of the same type as \emph{þęlla} ‘young fir tree’ < \emph{þǫll} ‘fir tree’, is possible but uncertain.}} \hld\ \alst{ø}r·lǫg-lausa.\eva

\bvb Until three came out of that host, \\
strong and loving Eese along the houses. \\
They found on land the little availing \\
Ash and Emble, \inx[C]{orlay}-less.\evb\evg


\bvg\bva\mssnote{\Regius~1v/3, \Hauksbok~20r/27}%
\edtrans{\alst{Ǫ}nd}{Breath}{\Bfootnote{The breath of life or life force which sets living beings apart from dead matter.  The word \emph{ǫnd} originally means simply ‘breath’ (cf. Gothic \emph{uz·anan} ‘to breathe out’) but the semantic development \textsc{breath} > \textsc{vitality} is by no means unparalleled as seen by the Latin root-cognate \emph{animus} ‘life force, vitality; (later) soul’ vs. the Greek \textgreek{ᾰ̓́νεμος} ‘wind, breeze’.  All these words are ultimately from PIE \emph{*h₂enh₁-} ‘breathe’ \parencite[26--27]{Kroonen2013}. —
After Christianity (from the C11th onwards) ON \emph{ǫnd} takes on the additional sense ‘spirit, soul’, as seen on Scandinavian memorial runestones where it is used interchangably with the Anglo-Saxon borrowing \emph{sǫ́l} ‘soul’ (e.g. Sö 10 \emph{Guð hjalpi ǫnd hans} ‘God help his breath’ vs. Sö 8 \emph{Guð hjalpi sǫ́lu hans} ‘God help his soul’) and paired with it in at least 14 separate inscriptions (e.g. U 358 \emph{Guð hjalpi hans ǫnd ok sǫ́lu} ‘God help his breath and soul.’)
This development, probably based on a calque the Latin \emph{spiritus} ‘breath; spirit’, is also charted in the Scaldic corpus, where \emph{ǫnd} (regardless of sense) makes it first securely dateable appearance in Hallfrith’s memorial poem to Anlaf Trueson (Hfr \emph{ErfÓl} 27 in \Skp\ 1) dated to ca. 1001 CE.
This circumstance may lead one to question the authenticity of the sense ‘life (force)’ in the pagan period, but I judge that it must have existed to begin with, in order for the new sense of ‘spirit’ to have been acceptable to converts; if \emph{ǫnd} had meant merely ‘respiration’ it would probably have been displaced entirely by the thoroughly Christian \emph{sǫ́l} ‘soul’ rather than have seen acceptance even amongst former pagan poets like Hallfrith.
In any case the former sense is attested in the archaic Eddic poem \textlink{Sigurdskamma}[33], 53 (\emph{ǫndu látit} ‘lost his life’), 60 (\emph{ǫndu tẏna} ‘id.’)—whereas one could hardly say that a person who dies “loses his soul”.
Also in later poetry a person who dies is often said to \emph{láta} or \emph{tẏna ǫndu} (in the pres. ed. \textlink{HelgakvidaHjorvardssonar}[37]; in \Skp\ Ótt \emph{Knútdr} 9, Þorm Lv 13, ÞSjár \emph{Þórdr} 3, Ív \emph{Sig} 30, ESk \emph{Geisl} 60, Bjbp \emph{Jóms} 41, Þórðh Lv 9, ǪrvOdd \emph{Ævdr} 12, 39, Anon \emph{Brúðv} 27), where this is, again, not said once of the \emph{sǫ́l}.

The use of the word in \Gylfaginning\ 3—printed at the beginning of this book for its simple beauty—is distinctly with the Christian sense of ‘spirit’: \emph{Hitt er þó mest, er hann gerði mann’inn ok gaf honum ǫnd þá, er lifa skal ok aldri týna⸗st, þó’tt líkam’inn fúni at moldu eða brenni at ǫsku} ‘Yet the greatest thing is when he [= Weden the Allfather] made man and gave him that “breath” which shall live and never perish even though the body should moulder to dust or burn to ashes.’}} þau né \alst{ǫ́}ttu, \hld\ \alst{ó}ð þau né hǫfðu, &
\alst{l}ǫ́ né \alst{l}ę́ti \hld\ né \alst{l}itu góða; &
\alst{ǫ}nd gaf \alst{Ó}ðinn, \hld\ \alst{ó}ð gaf Hø̇nir, &
\alst{l}ǫ́ gaf \alst{L}óðurr \hld\ ok \alst{l}itu góða.\eva

\bvb Breath they had not, \inx[C]{wode} they had not, \\
not craft nor sound nor good hues. \\
Breath gave Weden, wode gave Heener, \\
craft gave Lother and good hues.\evb\evg


\bvg\bva\mssnote{\Regius~1v/5, \Hauksbok~20r/29, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{A}sk vęit’k \edtext{standa}{\lemma{standa ‘standing’}\Afootnote{so \Regius\Hauksbok\Upsaliensis; \emph{ausinn} ‘sprinkled’ \RegiusProse\Trajectinus\Wormianus}}, \hld\ hęitir \edtext{\alst{Y}gg-drasill}{\Afootnote{\emph{Ygg-drasils} \RegiusProse}}, &
\alst{h}ǫ́r \edtrans{baðmr}{beam}{\Afootnote{\emph{borinn} ‘born’ \Upsaliensis\ wo. doubt corrupt.}}, \edtrans{\edtrans{ausinn}{sprinkled}{\Afootnote{\emph{hęilagr} ‘holy’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}} \hld\ \alst{h}víta auri}{sprinkled with white mud}{\Bfootnote{Possibly relevant is the Indian ritual pouring of beverages like milk onto the phallic \emph{líṅga}, although Nikhil Surya Dwibhashyam considers this an indigenous Indian practice foreign to the old Vedic religion.  Cf. st. 26 below.}}; &
þaðan koma \alst{d}ǫggvar \hld\ \edtext{þę́r’s}{\Afootnote{\emph{es} \RegiusProse\Trajectinus}} ï \alst{d}ala falla; &
stęndr \edtext{\alst{ę́}}{\Afootnote{\emph{om.} \Upsaliensis}} \alst{y}fir \edtext{grø̇nn}{\Afootnote{\emph{†grvnn†} \RegiusProse; \emph{†grein†} \Upsaliensis}} \hld\ \alst{U}rðar brunni.\eva

\bvb An ash I know standing, ’tis called \inx[P]{Ugdrassle}: \\
a high tree sprinkled with white mud. \\
Thence come the dew-drops which fall in the dales; \\
it stands ever green over \inx[P]{Weird’s Well}.\evb\evg


\bvg\bva\mssnote{\Regius~1v/8, \Hauksbok~20r/31}%
\Ballnote{The st. is paraphrased in \Gylfaginning\ 15: \emph{Þar stendr salr einn fagr undir aski’num við brunni’n, ok ór þeim sal koma þrjár meyjar, þę́r er svá heita: Urðr, Verðandi, Skuld.  Þessar meyjar skapa mǫnnum aldr; þę́r kǫllum vér nornir.} ‘There stands a single fair hall beneath the ash-tree by the well, and out of that hall come three maidens who are called thus: Weird, Werthing, Shild.  These maidens shape the ages of men; we call them norns.’}%
Þaðan koma \alst{m}ęyjar \hld\ \alst{m}args vitandi &
\alst{þ}ríar ór þęim \edtrans{sal}{hall}{\Afootnote{so \Hauksbok, \RegiusProse\Trajectinus\Upsaliensis\Wormianus\ (in the paraphrase); \emph{sę́} ‘lake’ \Regius}} \hld\ es \edtrans{und}{under}{\Afootnote{\emph{ȧ} ‘on’ \Hauksbok}} \edtrans{\alst{þ}olli}{tree}{\Bfootnote{Literally ‘fir’, but the word is only used for the alliteration.  The same may perhaps apply to \emph{askr} ‘ash’ above, the species being indeterminate.}} stęndr; &
\alst{U}rð hétu \alst{ęi}na, \hld\ \alst{a}ðra Verðandi, &
—\edtrans{\alst{sk}ǫ́ru ȧ \alst{sk}íði}{they scored billets}{\Bfootnote{Unclear; perhaps they carve tallies for the number of years allotted to each human being.}}— \hld\ \alst{Sk}uld hina þriðju &
þę́r \alst{l}ǫg \alst{l}ǫgðu, \hld\ þę́r \alst{l}íf køru, &
\alst{a}lda bǫrnum, \hld\ \alst{ø}r·lǫg \edtrans{sęggja}{of youths}{\Afootnote{\emph{at sęgja} ‘to say’ \Hauksbok}}.\eva

\bvb Thence come maidens much knowing: \\
three out of the hall which stands beneath the tree. \\
Weird they called one, the other Werthing \\
—they scored billets—Shild the third. \\
Laws they laid down; lives they chose \\
for the children of mankind, the \inx[C]{orlay} of youths.\evb\evg


\bvg\bva\mssnote{\Regius~1v/11, \Hauksbok~20v/5}%
\Ballnote{A very cryptic stanza.  The name Goldwey is not mentioned in any other source, nor is the underlying story, but she was apparently slain, burned and reborn three times (in short succession?) by the Eese; she is perhaps to be identified with the wallow herself, which would explain the cryptic “yet she still lives”.}%
Þat man hǫ̇n \edtrans{\alst{f}olk-víg}{troop-conflict}{\Bfootnote{\emph{folk} here carries its older meaning ‘troop, band’, as seen in the Slavic borrowing exemplified by Russian \textrussian{полк} ‘regiment, host, army’.}} \hld\ \alst{f}yrst ï hęimi, &
es \alst{G}oll·vęigu \hld\ \alst{g}ęirum studdu &
ok ï \alst{h}ǫll \alst{H}á\emph{a}rs \hld\ \alst{h}ȧna bręnndu, &
\edtext{\alst{þ}rysvar bręnndu}{\Afootnote{\emph{†þrysvar brendv þrysvar brendv†} \Hauksbok}} \hld\ \alst{þ}rysvar borna, &
\alst{o}pt, \alst{ȯ}·sjaldan, \hld\ þó hǫ̇n \alst{ę}nn lifir.\eva

\bvb That troop-conflict she recalls first in the \inx[C]{Home}, \\
when Goldwey with spears they pierced \\
and in the hall of \inx[P]{Hougher} \name{= Weden} \ken*{= Walhall} burned her. \\
Thrice they burned the one thrice born, \\
often, unseldom—yet she still lives.\evb\evg


\bvg\bva\mssnote{\Regius~1v/13, \Hauksbok~20v/7}%
\alst{H}ęiði \alst{h}étu, \hld\ hvar’s til \alst{h}úsa kom, &
\edtext{\alst{v}ǫlu}{\Afootnote{\emph{ok vǫlu} \Hauksbok}} \alst{v}ęl-spáa, \hld\ \alst{v}itti ganda; &
\alst{s}ęið hǫ̇n \edtrans{hvar’s hǫ̇n kunni}{where she could}{\Afootnote{so \Hauksbok; \emph{hǫ̇n kunni} ‘she knew’ \Regius}}, \hld\ \alst{s}ęið hǫ̇n \edtrans{hug lęikinn}{deluded minds}{\Afootnote{so \Hauksbok; \emph{leikinn} \Regius}}; &
\alst{ę́} vas hǫ̇n \alst{a}ngan \hld\ \alst{i}llrar brúðar.\eva

\bvb Heath they called—where to houses she came— \\
the well-spaeing \inx[C]{wallow}; she bewitched \inx[C]{gand}[gands]. \\
She sorcered where she could; she sorcered deluded minds; \\
she was ever the delight of an evil bride.\evb\evg


\bvg\bva\mssnote{\Regius~1v/16, \Hauksbok~20v/9}%
\Ballnote{Sts. 22–23 deal with the obscure war between the Eese and the Wanes, the two tribes of the Gods.  It is most fully described in \YnglingaSaga\ 4, which also features a long description of the hostages: \emph{Óðinn fór með her á hendr Vǫnum, en þeir urðu vel við ok vǫrðu land sitt, ok hǫfðu ymsir sigr; herjuðu hvárir á land annarra ok gerðu skaða.  En er þat leiddi⸗st hvárum-tveggjum, lǫgðu þeir milli sín séttar-stefnu, ok gerðu frið ok seldu⸗st gíslar.} ‘Weden went with an warband against the Wanes, but they responded well and defended their land, and both sides had victory in succession; either troop raided the land of the other and did harm.  But when that seemed loathsome to both sides they held a conciliation-settlement between themselves and made peace and traded hostages.’
It is also alluded to in \Gylfaginning\ 23 (edited in the pres. ed. as \textlink{EddicFragments}[F2][2]), which mentions how Nearth was one of the hostages sent by the Wanes, and \Skaldskaparmal\ 5, which explains the origin of the mead of poetry (for which cf. \textlink{Havamal}[103]–110): \emph{Þat váru upp-hǫf til þess, at goðin hǫfðu ó·sę́tt við þat folk, er Vanir heita. En þeir lǫgðu með sér frið-stefnu ok settu grið á þá lund, at þeir gengu hvárir-tveggju til eins kers ok spýttu í hráka sínum} ‘This was the origin of it, that the Gods had hostile relations with the folk called the Wanes.  But they held a peace-settlement amongst each other and made a truce by the following means, that each of them went to a vat and spit their spittle into it.’

As the relevant sources are all closely related, the war, indeed the very distinction between the Eese and the Wanes, has been taken by some scholars (TODO: cite) to be a Snorronean extrapolation based on weak evidence.  Still, a straight-forward reading of st.~33/3–4 sees the ‘Wanes’ breaking down the fortification of the ‘Eese’, and when compared with \textlink{Vafthrudnismal}[38]–39 which explicitly contrasts the ‘Wanes’ and the ‘Eese’ they really do seem to be different groups also in the Poetic Edda.}%
Þȧ gingu \alst{r}ęgin ǫll \hld\ ȧ \alst{r}ǫk-stóla, &
\alst{g}inn-hęilǫg goð, \hld\ ok umb þat \alst{g}ę̇ttu⸗sk: &
Hvárt skyldu \alst{ę̇}sir \hld\ \alst{a}f·ráð gjalda, &
eða skyldu \edtrans{\alst{g}oð’in ǫll}{all the Gods}{\Bfootnote{The clitic definite \emph{-in} is very rare in older ON poetry; this is its only occurence in \textlink{Voluspa}. — Here “all the Gods” (viz., the Eese \emph{and} the Wanes) seem to be contrasted with the \inx[P]{Eese}, a subset.}} \hld\ \alst{g}ildi ęiga?\eva

\bvb Then went all the Reins onto the rake-seats: \\
the Yin-holy Gods, and from each other took counsel of this: \\
Whether the Eese should yield tribute— \\
or all the Gods should hold a banquet?\evb\evg


\bvg\bva\mssnote{\Regius~1v/17, \Hauksbok~20v/11}%
\edtrans{\alst{F}lęygði Óðinn \hld\ ok ï \alst{f}olk of skaut}{Weden hurled and shot into the troop}{\Bfootnote{The object, a spear, is understood. — This first spear-throw was ritually reenacted during battle; so \EyrbyggjaSaga\ 44: \emph{þá skaut Stein·þórr spjóti at fornum sið til heilla sér yfir flokk Snorra} ‘then Stanthur shot a spear according to ancient custom for his fortune over Snorre’s troop’ and \StyrbjarnarThattr\ 2, where the Swedish king Eric devotes himself as a sacrifice within a period of ten years in exchange for victory.  In return Weden gives him a reed and tells him to hurl it into the opposing host with the phrase \emph{Óðinn á yðr alla} ‘Weden owns you all!’  Eric does so, and the reed turns into a spear in the air (for this motif cf. \GautreksSaga\ 7 as summarized in \textlink{Havamal}[138] n.).  As it flies over his foes they are stricken by terror and blindness before being buried in a landslide.

It was customary among the Germanic pagans that the battle-slain (\emph{valr}) were devoted to Weden; cf. Þhorn \emph{Harkv} 12 (\Skp\ 1): \emph{Valr lá þar ȧ sandi \hld\ vitinn inum ęin-ęygja // Friggjar faðm-byggvi} ‘The slain lay there on the sand set apart for the one-eyed bosom-dweller \ken{husband} of Frie \ken*{= Weden}’.
The consecration could also be done as a thank-offering after a victory, as in \HervararSaga\ 7: \emph{Heið·rekr kveð⸗st nú gjalda fyrir son sinn þetta lið allt, er drepit var, ok gaf hann nú þennan val Óðni} ‘Heathric now declares that he, in exchange for his son[’s life], pays up the whole part of the army which was killed, and he now gave these battle-slain to Weden.’
In the afterlife these men would join the god as \inx[P]{Oneharriers} in \inx[P]{Walhall} (cf. \textlink{Grimnismal}[14], \textlink{Harbardsljod}[24], also runic inscription \emph{N B380} edited below under Galders).  The sacrifice also included captives, who were at times executed by hanging (\textlink{Havamal}[138] n.) and which may be the reason why Weden is also called \emph{Hapta-guð} ‘God of Captives’ (\Gylfaginning\ 20).  As seen below, even the weapons of the vanquished were devoted.

This type of devotion of an entire army was not uncommon in Germanic antiquity and is securely attested in both history and archeology.  Among historical sources I particularly wish to mention Orosius, \emph{History} 5.16.5–6, describing the battle of Arausio fought in 105 BCE between the Romans and the Germanic tribes of the Cimbri and Teutons: “After capturing the two Roman camps and a vast amount of booty, the enemy destroyed everything that they had laid their hands on in some new, unexpected form of curse.  6 Clothing was ripped up and discarded, gold and silver thrown into the river, the men’s armour was torn apart, the horses’ harness scattered and the horses themselves drowned in the river, while the men had nooses tied round their necks and were hanged from trees. In this way the victor knew no booty nor the vanquished any mercy.” (\cite[235]{FearOrosius}).

The most relevant archeological finds are the various Northern European bogs and wetlands where horses and arms, presumably belonging to defeated foes, have been mutilated and defaced (e.g. by whipping the horses, bending the swords) before their submersion.  This defacing is to be understood by the fact that the objects, having been devoted and made holy unto the god, had to be made materially worthless and unusable to men.  Such is confirmed by Tacitus, \emph{Annals} 13.57 who describes that the Chatti, \emph{victores diversam aciem Marti ac Mercurio sacravere, quo voto equi, viri, cuncta occidioni dantur} ‘had devoted, in the event of victory, the enemy’s army to Mars and Mercury, a vow which consigns horses, men, everything indeed (on the vanquished side) to destruction.’
For Germanic devotional sacrifice cf. further \textciteshorttitle{PCRN-HS} II:25, pp. 617; III:42, p. 1166–1168; for bog burials cf. \textlink{Voluspa}[38] n. below; for human sacrifice to Weden generally cf. \textlink{Havamal}[138] n.

Outside of the Germanic sphere similar customs were practiced among both the Romans (Latin \emph{devotio})—cf. Livy 8.9–10, where the Roman consul of 340 BCE Publius Decius Mus, like king Eric in \StyrbjarnarThattr, devotes not just the opposing legions but also himself, although Livy comments that this is not strictly necessary—and the Jews (Hebrew \texthebrew{חֵרֶם} \emph{ḥērem}, \cite{Lohfink1986}), who would devote entire cities or nations to YHWH (Deuteronomy 20:16–17, Joshua 6:17-21, Joshua 8:1–29; note in Joshua 8:18 the detail of the god-given spear that leads to victory), after which, like in the Germanic bog deposits, also the arms and property of the vanquished were to be wasted or otherwise set aside for the god (Leviticus 27:28–29).}}; &
þat vas ęnn \alst{f}olk-víg \hld\ \edtrans{\alst{f}yrr}{earlier}{\Afootnote{so \Hauksbok; \emph{fyrst} ‘first’ \Regius}\Bfootnote{The \Regius\ reading cannot be correct as this st. is describing a different war than the first; it has probably arisen due to the similarity with st. 20/1.}} ï hęimi; &
\alst{b}rotinn vas \edtrans{\alst{b}orð-vęggr}{plank-wall}{\Bfootnote{A dipax legomenon in the Old Norse corpus.  Its only other occurrence is in Þloft \emph{Glækv} 6 (\Skp\ 1) from ca. 1032, where it describes some kind of wooden structure, probably part of a stave-church, in which bells ring.}} \hld\ \alst{b}orgar ȧsa, &
knǫ́ttu \alst{v}anir \edtrans{\alst{v}íg-spǫ́}{war-spae}{\Bfootnote{The Wanes used a magic prophecy (\emph{spǫ́} ‘spae’) to win the battle and sack \inx[P]{Osyard}, the stronghold of the Eese.}} \hld\ \alst{v}ǫllu sporna.\eva

\bvb Weden hurled and shot into the troop; \\
that was yet a troop-conflict earlier in the \inx[P]{Home}. \\
Broken was the plank-wall of the stronghold of the Eese; \\
the Wanes by a war-\inx[C]{spae} did tread the fields.\evb\evg


\bvg\bva\mssnote{\Regius~1v/19, \Hauksbok~20r/34, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\Ballnote{After the stronghold of the Eese, protected only by a plank-wall (the \emph{borð-vęggr} of st. 23), is sacked by the Wanes they decide to build stronger fortifications.  This leads into the story of the wall-builder as told at length in the following way in \Gylfaginning\ 42, which concludes by quoting \Voluspa\ 24–25:  An \inx[P]{Ettins}[ettin] craftsman arrives at Osyard and offers to build the Eese a great wall in exchange for \inx[P]{Frow}’s hand and the \inx[P]{Sun} and \inx[P]{Moon}.  He further promises that he will finish the entire wall alone over a single winter, or else the contract will be void.  This seems reasonable to the Eese who assume that he cannot possibly finish it in time, but things go awry when he asks for permission to have the help of his workhorse \inx[P]{Swaddlefare}.  Lock, without consulting the other gods, grants him this, after which the agreement is sealed with strong oaths.  The horse turns out to be unexpectedly strong, and when three days are left before summer the wall is almost finished.  The panicked Eese blame Lock and threaten him to get rid of the horse.  His solution is to turn into a mare and seduce it.  The trick works and the two stay out all night; Lock is made pregnant, later giving birth to \inx[P]{Slapner}.  When the ettin realises that he has been tricked he comes into his greatest ettin wrath, at which point the Eese call on Thunder who shortly appears and kills him.}%
Þȧ gingu \alst{r}ęgin ǫll \hld\ ȧ \alst{r}ǫk-stóla, &
\alst{g}inn-hęilǫg \alst{g}oð, \hld\ ok umb þat \alst{g}ę̇ttu⸗sk: &
Hvęrr hęfði \alst{l}opt alt \hld\ \alst{l}ę́vi blandit &
eða \alst{ę́}tt \alst{jǫ}tuns \hld\ \edtrans{\alst{Ó}ðs męy}{Wode’s maiden \ken*{= Frow}}{\Bfootnote{According to \Gylfaginning\ 35, \Skaldskaparmal\ 28, 44 \emph{Óðr} ‘Wode’ (not to be confused with \emph{Óðinn} ‘Weden’!) was Frow’s husband.

The ettins seem to have a particular obsession with abducting Frow, as seen also in \textlink{Thrymskvida}[8].  This may be explained if Frow is the Indo-European dawn-goddess, her nature as which being suggested not only by her association with the Sun and Moon in the wall-builder myth but also by her weeping of golden tears (\Gylfaginning\ 35, \Skaldskaparmal\ 40, 44; possibly a poetic description of the dawn), her epithet Meredall (\emph{Mar·dǫll} ‘Sea-brilliant’, another likely poetic description; for the second element see \textlink{Vafthrudnismal}[25]) and her possession of the fire-symbolic Blaze-Necklace (\emph{brísinga-męn}, cf. \textlink{Thrymskvida}[13]/3 n.)
The attempted abduction of the Dawn has a parallel in the Vedic \emph{Valá}-myth (part of which is told in \Rigveda\ 10.108; \cite{JamBrer2020}, pp. 70, 71).  In that myth the \emph{Paṇi}-s, enemies of the Gods, hide the Dawns in the form of kine in the \emph{Valá}-cave and thereby prevent the natural cycle between Night and Day.  With his great song the Storm-god \emph{Índra} breaks open the cave, killing \emph{Valá} and releasing the dawns.}} gefna?\eva

\bvb Then went all the Reins onto the rake-seats: \\
the Yin-holy Gods, and from each other took counsel of this: \\
Who might have blended all the air with deceit, \\
or to the ettin’s lineage given \inx[P]{Wode}’s maiden \ken*{= Frow}?\evb\evg


\bvg\bva\mssnote{\Regius~1v/20, \Hauksbok~20r/36, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\Aallnote{The order of the lines follows \Regius\Hauksbok; they are 3, 4, 1, 2 in \RegiusProse\Trajectinus\Upsaliensis\Wormianus.}%
\alst{Þ}ȯrr ęinn \edtrans{\alst{þ}ar vá}{fought there}{\Afootnote{so \Hauksbok\Trajectinus\Upsaliensis; \emph{þar var} ‘was there’ \Regius; \emph{þat vann} ‘accomplished it’ \RegiusProse; \emph{þat vá} ‘fought it out’ \Wormianus}} \hld\ \alst{þ}runginn móði, &
\edtrans{hann \alst{s}jaldan \alst{s}itr \hld\ es \alst{s}líkt of fregn;}{he seldom sits when of such he learns}{\Bfootnote{When he learns of an ettin encroaching on the gods (see note to 24/ALL).  Thunder, as the defender of the gods (\textlink{Thrymskvida}[18], Þdís \emph{Þórr} in \Skp\ 3), is willing to break even oaths sworn to an ettin for this purpose (cf. \textlink{Lokasenna}[57]–64).}} &
\edtext{\alst{ȧ} gingu⸗sk \alst{ęi}ðar, \hld\ \alst{o}rð ok sǿri, &
\alst{m}ǫ́l ǫll \alst{m}ęgin⸗lig, \hld\ es ȧ \alst{m}eðal \edtext{fóru}{\lemma{fóru ‘had gone’}\Afootnote{\emph{vǫ́ru} ‘had been’ \Hauksbok\Trajectinus}}.}{\lemma{ȧ \dots\ fóru.}\Afootnote{om. \Wormianus}}\eva

\bvb Thunder alone fought there, pressed by wrath; \\
he seldom sits when of such he learns. \\
Trampled were oaths, speeches and vows, \\
all the mighty treaties which between them had gone.\evb\evg


\bvg\bva\mssnote{\Regius~1v/23, \Hauksbok~20v/1}%
Vęit hǫ̇n \alst{H}ęim·dalar \hld\ \alst{h}ljóð of folgit &
und \edtrans{\alst{h}ęið-vǫnum}{shady}{\Bfootnote{Literally ‘light-less’, \emph{hęiðr} referring especially to the light of a clear sky.}} \hld\ \alst{h}ęlgum baðmi; &
\alst{ǫ́} sér hǫ̇n \alst{au}sa⸗sk \hld\ \edtrans{\alst{au}rgum}{muddy}{\Bfootnote{Which should be the same mud (\emph{aurr}) as in st. 19, there said of Weird’s Well.}} forsi &
af \edtrans{\alst{v}ęði \alst{V}al-fǫðrs}{Walfather’s pledge}{\Bfootnote{Weden placed his eye in Mimer’s well, which gives wisdom to any man who drinks from it.  So \Gylfaginning\ 15: \emph{Þar kom Al·fǫðr ok beiddi⸗st eins drykkjar af brunni’num, en hann fekk eigi, fyrr en hann lagði auga sitt at veði.} ‘There came Allfather and asked for a single drink from the well, but he did not get it before he laid down his eye as a pledge.’}}. \hld\ \edtrans{\alst{V}ituð ér ęnn eða hvat?}{Know ye yet, or what?}{\Bfootnote{“Do you, Weden, know enough now, or what?”, repeated in 28, 33, 34, 38, 40, 47, 60, and 61.  Similar refrains are found in \textlink{Baldrsdraumar} and \textlink{Hyndluljod}.}}\eva

\bvb She knows Homedal’s sound \ken*{= Horn of Yell?} hidden \\
beneath the shady, hallowed beam \ken*{= Ugdrassle’s Ash?}. \\
A river she sees being fed by a muddy torrent \\
from Walfather’s pledge \ken*{= Mimer’s well}. \emph{Know ye yet, or what?}\evb\evg


\bvg\bva\mssnote{\Regius~1v/25}%
\edtrans{\alst{Ęi}n sat hǫ̇n \alst{ú}ti}{Alone sat she outside}{\Bfootnote{ON \emph{sitja úti} ‘sit outside’ has a cultural connotation of meditating in order to communicate with the otherworld; cf. the noun \emph{úti-seta}.  This line is directly repeated in \textlink{Sigurdskamma}[6]/1a.}}, \hld\ þȧ’s hinn \alst{a}ldni kom &
\alst{y}ggjungr \alst{ȧ}sa \hld\ ok ï \alst{au}gu lęit: &
‚hvęrs \alst{f}regnið mik? \hld\ hví \edtrans{\alst{f}ręistið}{tempt}{\Bfootnote{\emph{fręista} ‘tempt’ has a sense of testing someone, especially intellectually.  Cf. \textlink{Havamal}[26], \textlink{Vafthrudnismal}[3], 5.}} mïn?\eva

\bvb Alone sat she outside when the old one came, \\
the Terrifier of the Eese \ken*{= Weden}, and looked into her eyes. \\
\speakernoteb{[The Wallow:]}%
‘Of what ask ye me? Why tempt ye me?\evb\evg


\bvg\bva\mssnote{\Regius~1v/26, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{A}lt vęit’k, \alst{Ó}ðinn, \hld\ hvar \alst{au}ga falt &
\edtrans{ï hinum \alst{m}ę́ra}{in the famed}{\Afootnote{so \Wormianus; \emph{þitt} (corr.) \emph{i enom męra} ‘id.’ \Regius; \emph{j þeim enom meira} ‘in the greater’ \Trajectinus; \emph{i þeim envm mæra} ‘in the famed’ \Upsaliensis; \emph{vr þeim envm mę́ra} ‘out of the famed’ \RegiusProse}} \hld\ \alst{M}ímis brunni; &
drekkr \alst{m}jǫð \alst{M}ímir \hld\ \alst{m}orgin hvęrjan &
af \edtrans{\alst{v}ęði}{pledge}{\Afootnote{\emph{†veiði†} \RegiusProse}} \alst{V}al-fǫðrs.‘ \hld\ \alst{V}ituð ér ęnn eða hvat?\eva

\bvb I know it all, Weden, where thine eye thou hidst: \\
in the famed \inx[P]{Mimer’s Well}! \\
Mimer drinks mead every morning \\
from Walfather’s pledge.’ \emph{Know ye yet, or what?}\evb\evg


\bvg\bva\mssnote{\Regius~1v/29}%
Valði hęnni \alst{H}ęr-fǫðr \hld\ \alst{h}ringa ok męn, &
\edtrans{fe\emph{kk} \alst{sp}jǫll \alst{sp}ak⸗lig}{got foreseen tidings}{\Afootnote{emend.; \emph{‘fe spioll spaclig’} \Regius}\Bfootnote{The reading of \Regius\ may be interpreted either as (1) \emph{fé-spjǫll spak⸗lig} ‘foreseen wealth-spells’ or (2) \emph{fé, spjǫll spak⸗lig} ‘wealth, foreseen tidings’.  Both are metrically deficient; in (1) a second element in a cpd. like \emph{fé-spjǫll} cannot carry alliteration, and (2) has three strongly stressed nominals.  In both cases the first nominal \emph{fé} would be expected to carry the alliteration.  The word \emph{fé} ‘wealth, cattle’ also makes little sense in context, since Weden is the one giving her expensive jewellery.
The emendation places the verb \emph{fekk} ‘got, received’ for \emph{fé}.  Verbs carry less stress than verbs and the line thus becomes metrically equivalent to 28/3b \emph{drekkr mjǫð Mímir}.  Semantically we get a parallel to st. 1, where the wallow likewise says that she will relate \emph{spjǫll} ‘tidings, sayings’ (cf. English \emph{gospel}, lit. ‘good news’ which originally translates the Greek \textgreek{εὐαγγέλιον}).  For further discussion on the reading see \textcite[51--53]{Haukur2020}, \textcite[16]{Males2023}.}} \hld\ ok \edtrans{\alst{sp}á-ganda}{spae-gands}{\Bfootnote{Spirits sent out in order to gather hidden wisdom and spaes.  See relevant Index entries.}}; &
sá \alst{v}ítt ok umb \alst{v}ítt \hld\ of \alst{v}er-ǫld hvęrja.\eva

\bvb Host-father \name{= Weden} chose for her rings and necklaces; \\
he got foresighted tidings and \inx[C]{spae}-\inx[C]{gands}— \\
she saw far and yet further o’er every world.\evb\evg


\bvg\bva\mssnote{\Regius~1v/30}%
Sá hǫ̇n \alst{v}al-kyrjur \hld\ \alst{v}ítt of komnar, &
\alst{g}ǫrvar at ríða \hld\ til \edtrans{\alst{G}oð-þjóðar}{land of the Gots}{\Bfootnote{Uncertain, for in \Regius\ the difficult cluster \emph{tþ} in \emph{Got-þjóð} ‘land of the Gots’ (attested in \HervararSaga\ st. 94/8b, \Skp\ 8) has been changed \emph{ðþ} which means \emph{Goð-þjóð} may contain either (1) \emph{goti} (thus ‘land of the Gots’) or (2) \emph{goð} (‘land of the Gods’).  ‘Land of the Gots’ is preferable since that sense is attested in three other places in \Regius\ (\textlink{Helreid}[8], \textlink{Gudrunarhvot}[8], 16), whereas (2) is entirely unattested.
The Walkirries do in fact have a particular association with the Gots, who were remembered as having fought the greatest battles of the Migration Period; cf. note to \textlink{Volundarkvida}[1]/1b. —
It is notable that ON \emph{Got-þjóð} reflects the attested Gotnish self-name, \emph{Gut-þiuda} (found in the October 29 entry of the Gotnish calender, TODO: reference).}}: &
\edtext{\alst{Sk}uld hélt \alst{sk}ildi, \hld\ en \alst{Sk}ǫgul ǫnnur, &
\alst{G}unnr, Hildr, \alst{G}ǫndul \hld\ ok \alst{G}ęir-skǫgul; &
\alst{n}ú eru talðar \hld\ \edtrans{\alst{N}ǫnnur Hęrjans}{Nans of Harn \name{= Weden}}{\Bfootnote{\emph{Nanna} ‘\inx[P]{Nan}’ (the name itself is a nursing word) was the wife of \inx[P]{Balder}, but the word is here certainly being used to refer generically to ‘maidens, women’.  Cf. Þul \emph{Ásynja} (\Skp\ 3), where the Walkirries are kenned \emph{Óðins męyjar} ‘Weden’s maidens’.}}, &
\alst{g}ǫrvar at ríða \hld\ \alst{g}rund, val-kyrjur.}{\lemma{Skuld \dots\ val-kyrjur. ‘Shild \dots\ lands.’}\Bfootnote{Judging especially by the out-of-place phrase \emph{nú eru talðar} ‘now are tallied’, these four lines seem to be a later insert from a \inx[C]{thule} counting the Walkirries.}}\eva

\bvb She saw \inx[P]{Walkirries} come from afar, \\
ready to ride to the land of the \inx[C]{Gots}. \\
Shild held a shield and Shagle another, \\
Guth, Hild, Gandle and Goreshagle— \\
now are tallied the Nans of Harn \name{= Weden}, \\
Walkirries ready to ride o’er the earth.\evb\evg


\bvg\bva\mssnote{\Regius~2r/2}%
\Ballnote{Told allusively in sts. 31–33 is the myth about Balder’s death.  The outline of this myth is that Balder, the son of Weden and Frie, was killed by a mistletoe-arrow shot by his blind half-brother Hath, whose hand was guided by Lock.  As the father of Balder it was Weden’s duty to avenge him, but he could not kill Hath since he was also his son.
He thus devised a plan, and through magic seduced the woman Wrind (so KormǪ \emph{Sigdr} 3, \Skp\ 3: \emph{sęið Yggr til Rindar} ‘Ug won Wrind through sorcery’).
Wrind gave birth to Wonnel, who grew up fast, and after just one day was big enough to kill Hath and avenge Balder.
Other sources for the same myth include the poems \textlink{Baldrsdraumar}[8]–11, \textlink{Lokasenna}[28], and the prose texts \Gylfaginning\ 49 and \textcite{Saxo} 3.4.1–8.
The Gods also punished Lock for his role in the killing; see st. 34 below.

The most closely related text to \Voluspa\ 31–33 is \textlink{Baldrsdraumar}[8]–11, with which it shares three near-identical lines (\textlink{Voluspa}[32]/4–33/2 = \textlink{Baldrsdraumar}[11]/2–4).  The biggest narrative difference between the two poems is that \textlink{Baldrsdraumar} mentions Wrind, who is not found in \Voluspa.

The narrative is treated at greater length in prose by \Gylfaginning\ 49, which may be shortly summarised as follows: Balder has terrible nightmares about his own death, and so his mother Frie makes all sorts of things (fire, water, venom, metals, stones, trees, diseases, animals) swear oaths not to harm him.  After this the Eese make sport of shooting and cutting him since he cannot be harmed.  Lock is annoyed by the spectacle and tricks Frie into telling him that there was one thing that did not swear the oath—the mistletoe, which was thought too young.
Lock makes a mistletoe-arrow and gives it to the blind god Hath, guiding his hand towards Balder.  Hath shoots and Balder dies.  After this \Gylfaginning\ 49 describes at some length Balder’s funeral (treated poetically in ÚlfrU \emph{Húsdr} (in \Skp\ 3)) and how the gods failed to “weep Balder out of hell” (see \textlink{EddicFragments}[F5][5]), before moving on to the binding of Lock in \Gylfaginning\ 50 (cf. st. 34 below.)

Like \Voluspa\ 31–33, \Gylfaginning\ 49 fails to mention Wonnel and Wrind.  That part of the myth may have been left out for moral reasons (rape, kinslaying) but was certainly known to the author, as proven by \Gylfaginning\ 30: \emph{Áli eða Váli heitir einn, sonr Óðins ok Rindar. Hann er djarfr í orrostum ok mjǫk happ-skęytr} ‘Onnel or Wonnel one is called, the son of Weden and Wrind. He is brave in battles and a very lucky shot’ and \Skaldskaparmal\ 19: \emph{Hvernig skal kenna Vála? Svá, at kalla hann son Óðins ok Rindar, [\dots] hefni-ás Baldrs, dólg Haðar ok bana hans [\dots]} ‘How shall one ken Wonnel? Namely, by calling him the son of Weden and Wrind, [\dots] avenging \inx[C]{os} of Balder, the foe of Hath and his bane [\dots]’

The other lengthy prose retelling is \textcite{Saxo} 3.4, which relates the revenge narrative in typical euhemerized form, turning the gods Hath and Balder into flawed human rulers.  The outline of that version is as follows: After Balder (Latin \emph{Balderus}) has succumbed to a mortal wound dealt him by Hath (\emph{Høtherus}) with a treacherous sword, his father Weden (\emph{Othinus}) takes counsel from a group of wizards.  One of them, Horsethief the Finn (\emph{Rostiophus Phinnicus}), foresees that Wrind (\emph{Rinda}), daughter of the Russian king, will bear him a son to avenge Balder.  Weden soon enlists in the Russian king’s army and leads it to great victories, but is spurned by his daughter.  He tries various disguises but is continually refused, until at last he disguises himself as an old witch and volunteers himself as her private physician.  When Wrind becomes ill, he binds her, ostensibly in order to administer a certain foul potion, but instead rapes her, implicitly with her father’s consent.  Their son Bo grows up to become a fierce raider.  One day Weden summons him and reminds him of his duty to avenge his half-brother Balder; Bo slays Hath in a fierce duel but soon himself perishes from his wounds.
I may here note that the seduction of Wrind as told by \citeauthor{Saxo} is unlikely to be a fantasy of that author, but on the contrary seems to derive from a now-lost West Norse source and has notable parallels to Shirner’s coercion of Gird in \textlink{Skirnismal}; see further \textcite{Brate1913}}%
Ek sá \alst{B}aldri, \hld\ \alst{b}lóðgum \edtrans{tífur}{victim’s}{\Bfootnote{This word is rather difficult and possibly corrupt.  It may be connected with \emph{týr} ‘tew, god’, but the dat. sg. of \emph{týr} is \emph{tívi} and the intrusive \emph{r} is unexplained.  A better explanation is given by \CV, who connect it with OE \emph{tiber, tifer} ‘victim, hostage’, but this also has some problems.  \emph{blóðgum} ‘bloody’ is masc. dat. sg., but OE \emph{tiber} is neuter.  If we are dealing with a masc. noun \emph{*tífurr} with the same declension as \emph{jǫfurr}, we would expect dat. sg. \emph{*tífri}, not \emph{tífur} (which would however be the expected acc. sg.).}}, &
\alst{Ó}ðins barni, \hld\ \alst{ø}r·lǫg \edtrans{folgin}{sealed}{\Bfootnote{The verb \emph{fela} ‘hide, conceal’, of which \emph{folgin} is the n. acc. pl. past participle, is used in poetry to describe burial in mounds, as in \Ynglingatal\ 24 (“[...] And afterwards the victory-havers hid (\emph{fǫ́lu}) the ruler on Borrey.”) or the C10th Karlevi stone (“Hidden (\textbf{fulkin} \emph{folginn}) in this mound lies he whom the greatest deeds followed; [...]”)}}; &
stóð of \alst{v}axinn \hld\ \alst{v}ǫllum hę́ri &
\alst{m}jór ok \alst{m}jǫk fagr \hld\ \alst{m}istil-tęinn.\eva

\bvb I saw Balder’s, the bloody victim’s, \\
Weden’s child’s \inx[C]{orlay} sealed: \\
there stood grown higher than the plains \\
a slender and most fair mistletoe.\evb\evg


\bvg\bva\mssnote{\Regius~2r/4}%
Varð af \alst{m}ęiði, \hld\ þęim’s \alst{m}jór sẏndi⸗sk, &
\alst{h}arm-flaug \alst{h}ę̇ttlig, \hld\ \alst{H}ǫðr nam skjóta. &
\alst{B}aldrs \alst{b}róðir vas \hld\ of \alst{b}orinn snimma, &
sá nam, \alst{Ó}ðins sonr, \hld\ \alst{ęi}n-nę́ttr vega.\eva

\bvb Of that tree which slender seemed \\
came a baneful harm-flier—Hath took to shoot. \\
Balder’s brother \ken*{= Wonnel} was born early; \\
he took, Weden’s son, one night old, to fight.\evb\evg


\bvg\bva\mssnote{\Regius~2r/6}%
\edtext{Þó}{\lemma{Þó \dots\ kęmbði ‘washed \dots\ combed’}\Bfootnote{A collocation, see note to \textlink{Havamal} 61 for discussion and other examples. Wonnel, being oathbound and on the mission to avenge his brother, could not engage in such acts of personal vanity.}} ę́va \alst{h}ęndr \hld\ né \alst{h}ǫfuð kęmbði, &
áðr ȧ \alst{b}ál of \alst{b}ar \hld\ \alst{B}aldrs and·skota; &
en \alst{F}rigg of grét \hld\ ï \alst{F}ęn-sǫlum &
\edtrans{\alst{v}ǫ́ \alst{V}al-hallar}{the woe of Walhall}{\Bfootnote{The deaths of two sons; Balder and Hath.}}. \hld\ \alst{V}ituð ér ęnn eða hvat?\eva

\bvb He washed ne’er his hands nor combed his head, \\
before onto the pyre he bore Balder’s opponent \ken*{= Hath}— \\
but Frie lamented in the Fenhalls \\
the woe of \inx[P]{Walhall}. \emph{Know ye yet, or what?}\evb\evg


\bvg\bva\mssnote{\Regius~2r/8, \Hauksbok~20v/13}%
\Ballnote{After Balder was avenged, the Eese went to catch Lock.  They bound him with his son’s intestines.  A snake was then placed above his face to spit venom on it.  His wife, Syein, sat over him and collected the venom in a small basin, but when she had to empty it he writhed so greatly from the pain of the dripping venom that the earth shook.  (\textlink{Lokasenna}[P8], \Gylfaginning\ 50).}%
\edtext{\alst{H}apt sá hǫ̇n liggja \hld\ und \alst{H}vera-lundi &
\edtrans{\alst{l}ę́-gjarns}{guile-eager}{\Bfootnote{A formulaic epithet of Lock. See note to TODO for other examples and discussion.}} líki \hld\ \alst{L}oka ȧ·þękkjan;}{\lemma{Hapt \dots\ ȧ·þękkjan ‘A captive \dots\ to Lock,’}\Afootnote{Replaced with H1 \Hauksbok.}} &
\alst{þ}ar sitr Sigyn \hld\ \alst{þ}ęy⸗gi of sïnum &
\alst{v}eri \alst{v}ęl-glýjuð. \hld\ \alst{V}ituð ér ęnn eða hvat?\eva

\bvb A captive \ken*{= Lock} she saw lying beneath Wharlund: \\
a guile-eager man’s form, alike to Lock. \\
There sits Syein not at all cheerful, \\
o’er her husband. \emph{Know ye yet, or what?}\evb\evg


\bvg\bva\mssnote{\Regius~2r/10}%
\Ballnote{Having described the death of Balder and how the gods took revenge on Lock, the wallow moves on to describing the places of evil (sts. 35–41).  The function of the present half-stanza is very unclear, probably because half of it is missing.  It is not in any way alluded to in \Gylfaginning, nor is it found in \Hauksbok.  The river Slide is also mentioned in \textlink{Grimnismal}[28]/4 in a long list of rivers.}%
\alst{Ǫ́} fęllr \alst{au}stan \hld\ of \alst{Ęi}tr-dala &
\alst{s}ǫxum ok \alst{s}verðum, \hld\ \edtrans{\alst{S}líðr}{Slide}{\Bfootnote{An adjective describing a blade so sharp that the finger is cut when slid across it.  Cf. \textlink{Atlakvida}[23]: \emph{saxi slíðr-bęitu} (dat. sg.) ‘slide-biting sax’.}} hęitir sú.\eva

\bvb A river falls from the east above the Venom-dales: \\
{[a river]} of saxes and swords—Slide is it called.\evb\evg


\bvg\bva\mssnote{\Regius~2r/11}%
\Ballnote{Sts. 36–38 are paraphrased in \Gylfaginning\ 52:

\emph{Þá mę́lti Gangleri: „Hvat verðr þá eptir, er brenndr er himinn ok jǫrð ok heimr allr, ok dauð goð’in ǫll ok allir Ein-herjar ok alt mann-folk, ok hafið ér áðr sagt, at hverr maðr skal lifa í nǫkkvǫrum heimi um allar aldir?“  Þá svarar Þriði: „Margar eru þá vistir góðar ok margar illar; batst er þá at vera á Gim-léi á himni, ok all-gótt er til góðs drykkjar þeim, er þat þykkir gaman, í þeim sal, er Brimir heitir; hann stendr ok á himni. Sá er ok góðr salr, er stendr á Niða-fjǫllum, gǫrr af rauðu gulli; sá heitir Sindri. Í þessum sǫlum skulu byggja góðir menn ok sið-látir.  Á Ná-strǫndum er mikill salr ok illr ok horfa norðr dyrr; hann er ok ofinn allr orma-hryggjum sem vanda-hús, en orma hǫfuð ǫll vitu inn í hús’it ok blása eitri, svá at eptir sal’num renna eitr-ár, ok vaða þę́r ár eið-rofar ok morð-vargar, svá sem hér segir:“}

‘Then spoke Gangler: “What will then remain, when heaven and earth and the whole world are burned, and the gods all are dead and all the Oneharriers and all man-kind—and still ye have said earlier, that every man will live in a certain home throughout all ages?”  Then answers Third: “Many good dwellings are there then, and many evil; then it is best to be in Gemlee in the heaven.  And in the hall which is called Brimmer it is all-furnished with good drink for those who find joy in that; that one also stands in the heaven.  Another good hall is the one which stands on the Nithfells, made out of red gold; it is called Sinder.  In these halls shall dwell good and virtuous men.  On Neestrand is a great hall and evil, and its doors face north.  It is all interlaced with the spines of serpents like a wicker-house, but the heads of the serpents all face into the house and blow venom so that there run through the hall rivers of venom; and in those rivers wade oath-breakers and murder-wargs, as it says here:”’

after which it quotes sts. 37 and 38/1–2, followed by the prose: \emph{En í Hver-gelmi er verst} ‘But in Wharyelmer it is the worst’ and finally 38/4.}%
Stóð fyr \alst{n}orðan \hld\ ȧ \edtrans{\alst{N}iða-vǫllum}{Nithwolds}{\Afootnote{\emph{Niða-fjǫllum} ‘Nithfells’ (paraphrase) \RegiusProse\Wormianus; \emph{fjǫllum nǫkkurum} ‘some certain fells’ \Trajectinus.}} &
\alst{s}alr ór golli \hld\ \alst{S}indra ę́ttar; &
en \alst{a}nnarr stóð \hld\ ȧ \alst{Ȯ}·kólni, &
\alst{b}jór-salr jǫtuns, \hld\ \edtrans{en sá \alst{B}rimir hęitir}{and he/it is called Brimmer}{\Bfootnote{It is syntactically ambiguous whether Brimmer is the name of the ettin or the hall itself.  \Gylfaginning\ clearly considers it the name of the hall.}}.\eva

\bvb To the north on the Nithwolds stood \\
a hall made out of gold of Sinder’s lineage \ken{dwarfs}, \\
but another one stood on Uncolner, \\
an ettin’s beer-hall, and he/it is called Brimmer.\evb\evg


\bvg\bva\mssnote{\Regius~2r/13, \Hauksbok~20v/19, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{S}al \edtrans{sá hǫ̇n}{she saw}{\Afootnote{\emph{vęit’k} ‘I know’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus; cf. st. 61.}} standa \hld\ \alst{s}ólu fjarri &
\alst{N}á-strǫndu ȧ, \hld\ \alst{n}orðr horfa dyrr; &
falla \alst{ęi}tr-dropar \hld\ \alst{i}nn umb ljóra, &
sá ’s \alst{u}ndinn salr \hld\ \alst{o}rma hryggjum.\eva

\bvb A hall she saw standing far from the sun, \\
on Neestrand—north face its doors. \\
Venom-drops fall in through the smoke-vent; \\
that hall is wound with the spines of snakes.\evb\evg


\bvg\bva\mssnote{\Regius~2r/15, \Hauksbok~20v/21, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\Ballnote{Watery punishment in the Heathen afterlife, as described in this stanza, is also found in \textlink{Reginsmal}[3]–4 and possibly in \textlink{Grimnismal}[21].  The crimes are what one might expect from a source in Germanic antiquity: perjury, shameful murder, and adultery with a married woman.  In Anglo-Saxon and Nordic laws the committer of such crimes got the title of \inx[C]{nithing}, that is, one afflicted with \inx[C]{nithe} (severe shame).  It is fitting, then, that such “nithings” should be tortured by a creature named Nithehewer (‘Nithe-striker’).  The practice of burying criminals not just in wilderness but specifically in bogs and flood-marks is well attested in sources relating to Germanic cultures from Tacitus, \emph{Germania} onwards and the “heavy streams” of the present stanza are clearly related to such customs.  This is discussed at length in \textcite{GermanicGems2}.}%
\edtrans{Sá hǫ̇n}{she saw}{\Afootnote{so \Regius; \emph{sér hǫ̇n} ‘she sees’ \Hauksbok; \emph{skulu} ‘shall [be]’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}} \alst{þ}ar vaða \hld\ \alst{þ}unga strauma &
\alst{m}ęnn \alst{m}ęin-svara \hld\ ok \edtrans{\alst{m}orð-varga}{murder-wargs}{\Bfootnote{Murderous outlaws.}} &
ok þann’s \alst{a}nnars glępr \hld\ \alst{ęy}ra-ru̇nu. &
Þar \edtrans{saug}{sucked}{\Afootnote{so \Hauksbok; \emph{†súg†} \Regius; \emph{kvęlr} ‘torments’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}} \alst{N}íð·hǫggr \hld\ \alst{n}ái fram-gingna; &
slęit \alst{v}argr \alst{v}era. \hld\ \alst{V}ituð ér ęnn eða hvat?\eva

\bvb There she saw wading through heavy streams \\
false-swearing men and murder-wargs, \\
and him who seduces another man’s ear-whisperer \ken{wife}. \\
There \inx[P]{Nithehewer} sucked corpses passed-on; \\
the warg tore at men. \emph{Know ye yet, or what?}\evb\evg


\bvg\bva\mssnote{\Regius~2r/17, \Hauksbok~20v/2, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\Ballnote{The old hag (who is not identified) raises the cubs of the wolf Fenrer; one of these will eventually swallow the moon.  For the myth of the wolves that chase the sun and moon cf. \textlink{Grimnismal}[40], according to which the sun is chased by the wolf Sholl, while another wolf, Hater, runs in front of it, and \textlink{Vafthrudnismal}[46]–47, where the Sun is said to be swallowed by Fenrer.

\Gylfaginning\ 12 says that Hater will swallow the moon, and continues:

\emph{Gýgr ein býr fyrir austan Mið-garð í þeim skógi, er Járn-viðr heitir.  [...]  In gamla gýgr fǿðir at sonum marga jǫtna ok alla í vargs líkjum, ok þaðan af eru komnir þessir ulfar. Ok svá er sagt, at af ę́tt’inni verðr sá einna máttkastr, er kallaðr er Mána-garmr. Hann fylli⸗st með fjǫrvi allra þeira manna, er deyja, ok hann gleypir tungl, en støkkvir blóði himin ok loft ǫll.  Þaðan týnir sól skini sínu, ok vindar eru þá ó·kyrrir ok gnýja heðan ok handan.  Svá segir í Vǫlu-spá:}

‘A lonely ogress lives to the east of Middenyard in the forest called Ironwood.  The old ogress raises many ettins as her sons, all in the likenesses of wolves, and thereof these wolves [Sholl and Hater] come.  And so it is said that from that lineage a single one becomes the mightiest, and he is called \inx[P]{Moongarm}.  He fills himself with the life of all those men who die; and he swallows the moon and stains the heaven and all the air with blood. From that the sun loses her shine, and then the winds are violent and howl hither and thither, and thus it says in the Spae of the Wallow:’

after which it quotes sts. 40–41.  \Gylfaginning\ is clearly synthesizing from several sources, and it seems likely that the Heathen poets themselves were not entirely in agreement about the exact details of these events.}%
\edtrans{\alst{Au}str}{In the east}{\Bfootnote{The cardinal direction associated with ettins and other monstrosities.}} \edtrans{býr}{dwells}{\Afootnote{so \Hauksbok\RegiusProse\Trajectinus\Upsaliensis\Wormianus; \emph{sat} ’sat/stayed’ \Regius}} hin \edtrans{\alst{a}ldna}{old}{\Afootnote{\emph{arma} ‘wretched’ \Upsaliensis}} \hld\ ï \edtrans{\alst{Éa}rn-viði}{Ironwood}{\Afootnote{metr. emend.; \emph{Járnviði} \Regius\Hauksbok\RegiusProse\Wormianus\Upsaliensis; \emph{Járn-viðjum} ‘Ironwoods’ \Trajectinus}} &
ok \edtrans{\alst{f}ǿðir}{rears}{\Afootnote{so \Hauksbok\RegiusProse\Trajectinus\Upsaliensis\Wormianus; \emph{fǿddi} ‘reared’ \Regius}} þar \hld\ \alst{F}ęnris kindir; &
verðr \edtext{af}{\Afootnote{\emph{ór} \Trajectinus\RegiusProse}} þęim \alst{ǫ}llum \hld\ \alst{ęi}nna nøkkurr &
\alst{t}ungls \edtrans{\alst{t}júgari}{seizer}{\Afootnote{\emph{†tuigan†} \Trajectinus; \emph{tregari} ‘griever’ \Upsaliensis. As the young agentive suffix \emph{-ari} is found nowhere else in the poem it is possible that this word is corrupt. If it is, it must have occurred early in the transmission, as reflexes of \emph{tjúgari} are found in all surviving mss.}} \hld\ ï \alst{t}rolls hami.\eva

\bvb In the east dwells the old woman in the \inx[P]{Ironwood} \\
and there rears the brood of \inx[P]{Fenrer} \ken{wolves}. \\
Out of them all arises one most certain: \\
a seizer of the Moon in a troll’s \inx[C]{hame}.\evb\evg


\bvg\bva\mssnote{\Regius~2r/19, \Hauksbok~20v/4, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{F}ylli⸗sk \alst{f}jǫrvi \hld\ \alst{f}ęigra manna, &
\alst{r}ýðr \alst{r}agna sjǫt \hld\ \alst{r}auðum dręyra, &
\alst{s}vǫrt verða \alst{s}ól-skin \hld\ of \alst{s}umur ęptir, &
\alst{v}eðr ǫll \alst{v}á-lynd. \hld\ \alst{V}ituð ér ęnn eða hvat?\eva

\bvb He fills himself with the lifeblood of \inx[C]{fey} men; \\
he reddens the seat of the \inx[P]{Reins} \ken{heaven} with red gore. \\
Black turn the sun’s rays in the summers thereafter, \\
the winds all woeful. \emph{Know ye yet, or what?}\evb\evg


\bvg\bva\mssnote{\Regius~2r/21, \Hauksbok~20v/16}%
\edtrans{\alst{S}at þar ȧ haugi}{There he sat on the mound}{\Bfootnote{The motif of ettins sitting on burial mounds is also found in \textlink{Thrymskvida}[6] and \textlink{Skirnismal}[11].  The significance of this is uncertain.}} \hld\ ok \alst{s}ló hǫrpu &
\alst{g}ýgjar hirðir, \hld\ \alst{g}laðr Ęgg·þér; &
\edtrans{\alst{g}ól of hǫ̇num}{Over him crowed}{\Bfootnote{The crowing of three cocks—the first in Ettinham, the second in Walhall, the third in Hell—heralds the destruction of the world.}} \hld\ ï \edtrans{\alst{G}agl-viði}{Galewood}{\Bfootnote{An otherwise unknown location; the first element is \emph{gagl} ‘wild goose’.  Galewood is perhaps the same as Ironwood.}} &
\alst{f}agr-rauðr hani, \hld\ sá’s \alst{F}jalarr hęitir.\eva

\bvb There he sat on the mound and struck the harp \\
the gow’s herdsman, glad \inx[P]{Edgethew}. \\
Over him crowed in the \inx[P]{Galewood} \\
the fair-red cock, he who is called Feller.\evb\evg


\bvg\bva\mssnote{\Regius~2r/23, \Hauksbok~20v/18}%
\alst{G}ól of ǫ̇sum \hld\ \alst{G}ollin·kambi, &
sá vękr \alst{h}ǫlða \hld\ at \alst{H}ęrja-fǫðrs, &
en \alst{a}nnarr gęlr \hld\ fyr \alst{jǫ}rð neðan &
\alst{s}ót-rauðr hani \hld\ at \alst{s}ǫlum Hęljar.\eva

\bvb Over the Eese crowed Goldencomb; \\
he wakes the heroes at the Father of Hosts’s \name{= Weden’s}— \\
but another one crows beneath the earth: \\
a soot-red cock in the halls of Hell.\evb\evg


\bvg\bva\mssnote{\Regius~2r/25}%
\Ballnote{The first occurrence of the refrain stanza (\emph{stęf}), repeated several times in the section of the poem dealing with the Rakes of the Reins (ON \emph{ragna rǫk}), culminating with st. 55.  The destruction of the world, involving the deaths of all major Gods save Balder, is treated at length in \Gylfaginning\ 51.  Further it is the subject of a wisdom contest in \textlink{Vafthrudnismal}[44]–53 and alluded to in \textlink{Lokasenna}[39], 41.}%
\alst{G}ęyr \alst{G}armr mjǫk \hld\ fyr \alst{G}nipa-hęlli, &
\alst{f}ęstr mun slitna, \hld\ en \alst{F}reki rinna; &
\alst{f}jǫlð vęit hǫ̇n \alst{f}rǿða, \hld\ \alst{f}ramm sé’k lęngra &
of \alst{r}agna \alst{r}ǫk, \hld\ \alst{r}ǫmm sig-tíva.\eva

\bvb Garm barks loud before the Gnip-caves; \\
the rope will tear and the Wolf will run. \\
She knows much lore; I foresee further \\
about the mighty \inx[P]{Rakes of the Reins}, of the Victory-Tews \ken{gods}.\evb\evg


\bvg\bva\mssnote{\Regius~2r/28, \Hauksbok~20v/24, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\Ballnote{This st. is cited in \Gylfaginning\ 51, where it is introduced by the text: \emph{Þá drepa⸗st brǿðr fyrir á·girni sakar, ok engi þyrmir fǫður eða syni í mann-drǫ́pum eða sifja-sliti.  Svá segir í Vǫlu-spǫ́:} ‘Then brothers slay each other for reasons of greed, and no man spares his father or son in murder of men or slaying of kin.  So it says in the Spae of the Wallow:’}%
\alst{B}rǿðr munu \alst{b}ęrja⸗sk \hld\ ok at \alst{b}ǫnum verða⸗sk, &
munu \edtrans{\alst{s}ystrungar}{the children of sisters}{\Afootnote{\emph{†stystrungar†} \Trajectinus}} \hld\ \edtrans{\alst{s}ifjum spilla}{defile the kinship}{\Bfootnote{I.e. “commit incest”, apparently referring to marriages between first cousins, which are prohibited in all mediæval Scandinavian law codes.  Compare related words found in the laws, like \emph{frę́nd-semis-spell} ‘incest’ and especially \emph{sifja-spell} ‘id.’ — The idea of incest as a sign of later ages is also found in \Rigveda\ 10.10.10a–b (norm. and tr., Nikhil S. Dwibhashyam. (2025, Aug. 31). \emph{Véda quote 6}. \url{https://nikhilsd.com/dvq/6/}): \emph{Ā́ gʰā tā́ gaccʰān \hld\ úttarā yugā́ni, // yátra jāmáyaḥ \hld\ kr̥ṇávann ájāmi} ‘There shall come indeed those later ages when relatives shall do (acts) not (fit for) relatives.’}}; &
\alst{h}art ’s \edtrans{ï \alst{h}ęimi}{in the Home}{\Afootnote{so \Regius\Hauksbok\Upsaliensis; \emph{með hǫlðum} ‘among men’ \RegiusProse\Trajectinus\Wormianus}}, \hld\ \alst{h}ór-dȯmr mikill, &
\alst{sk}ęggj-ǫld, \alst{sk}alm-ǫld, \hld\ \edtrans{\alst{sk}ildir}{shields}{\Afootnote{add. \emph{’ro} ‘are’ \Regius}} \edtrans{klofnir}{split}{\Afootnote{\emph{klofna} ‘become split’ \Upsaliensis}}, &
\edtrans{\alst{v}ind-ǫld}{wind-age}{\Bfootnote{In \Hauksbok\ the \emph{v} is capitalized, marking the beginning of a new stanza.}}, \alst{v}arg-ǫld, \hld\ \edtrans{áðr}{before}{\Afootnote{\emph{unz} (norm.) ‘until’ \Upsaliensis}} \edtrans{\alst{v}er-ǫld}{man-age}{\Bfootnote{Translated as such since it stands next to various other compounds with the second element \emph{ǫld} ‘age’.  ON \emph{ver-ǫld} is cognate with English \emph{world}, but in ON the sense “cosmos, universe” is usually expressed by \emph{hęimr} (e.g. l. 3 of the present stanza).}} \edtrans{stęypi⸗sk}{tumbles down}{\Bfootnote{\emph{grundir gjalla \hld\ gífr fljúgandi} (norm.) ‘foundations shrill, fiends flying’ add. after this l. \Hauksbok}} &
\edtext{mun \edtext{\alst{ę}ngi}{\Afootnote{\emph{†enn†} \Upsaliensis}} maðr \hld\ \alst{ǫ}ðrum þyrma.}{\lemma{mun \dots\ þyrma ‘Before \dots\ spare’}\Bfootnote{om. \RegiusProse\Trajectinus\Wormianus}}\eva

\bvb Brothers will fight and become each other’s bane; \\
the children of sisters will defile the kinship. \\
It is hard in the Home! Great whoredom, \\
axe-age, sword-age, shields split asunder, \\
wind-age, warg-age. Before the man-age tumbles, \\
no man will another spare.\evb\evg


\bvg\bva\mssnote{\Regius~2r/32, \Hauksbok~20v/27, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\Ballnote{Sts. 45–54 are cited (with the omission of the refrain-stanza 47) in sequence in \Gylfaginning\ 51.}%
\edtext{Lęika \alst{M}íms synir, \hld\ en \alst{m}jǫtuðr kyndi⸗sk &
at hinu \alst{g}alla \hld\ \alst{G}jallar-horni;}{\lemma{Lęika \dots\ Gjallar-horni; ‘Mime’s \dots\ Yell.’}\Bfootnote{om. \RegiusProse\Trajectinus\Upsaliensis\Wormianus}} &
\alst{h}ǫ́tt blę́ss \alst{H}ęim·dallr, \hld\ \alst{h}orn ’s ȧ lopti; &
\edtrans{\alst{m}ę́lir}{speaks}{\Afootnote{\emph{†mey†} \RegiusProse; \emph{†nie†} \Trajectinus}} Óðinn \hld\ við \alst{M}íms hǫfuð; &
\edtext{skęlfr \alst{Y}gg·drasils \hld\ \alst{a}skr standandi, &
\alst{y}mr it \alst{a}ldna tré, \hld\ en \alst{jǫ}tunn \edtext{losnar}{\Afootnote{add. H2 \Hauksbok.}}.}{\lemma{Skęlfr \dots\ losnar ‘Ugdrassle’s \dots\ comes loose’}\Afootnote{so \Hauksbok\RegiusProse\Trajectinus\Upsaliensis\Wormianus; these lines are reversed in \Regius.}}\eva

\bvb Mime’s sons play and the Measurer ... \\
after [the sounding of] the shrill Horn of Yell. \\
High blows Homedal—the horn is aloft; \\
Weden speaks with the head of Mime. \\
Ugdrassle’s Ash quakes, standing: \\
the old tree whines and the ettin comes loose.\evb\evg


\bvg\bva\mssnote{\Regius~2v/8, \Hauksbok~20v/30, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\edtrans{Hvat ’s með \alst{ǫ̇}sum? \hld\ hvat ’s með \edtrans{\alst{ǫ}lfum}{Elves}{\Afootnote{\emph{ǫ̇synjum} ‘Ossens’ \Upsaliensis}}}{What is with the Eese? What is with the Elves?}{\Bfootnote{Also occuring in \textlink{Thrymskvida}[7]/1.}}? &
\edtext{gnýr \alst{a}llr \alst{Jǫ}tun-hęimr, \hld\ \alst{ę̇}sir ’ro ȧ \edtrans{þingi}{the Thing}{\Bfootnote{The \inx[P]{Thing of the Gods} or the Divine Council; see note to st 6/1–2 and Index: Thing of the Gods.}},}{\lemma{gnýr \dots\ þingi}\Afootnote{om. \Upsaliensis}} &
\alst{st}ynja dvergar \hld\ fyr \edtext{\alst{st}ęin-durum}{\Afootnote{\emph{stęins} \Upsaliensis; \emph{stęin-dyrum} \Hauksbok\Wormianus\Upsaliensis}} &
\edtext{\edtext{\alst{v}ęgg-bergs}{\Afootnote{\emph{veg-bergs} \Hauksbok\Trajectinus\Wormianus}} \alst{v}ísir}{\Afootnote{om. \Upsaliensis}}. \hld\ \alst{V}ituð ér ęnn eða hvat?\eva

\bvb What is with the Eese? What is with the Elves? \\
All Ettinham roars—the Eese are at the Thing! \\
Dwarfs groan before gates of stone, \\
the lords of the cliff-side. \emph{Know ye yet, or what?}\evb\evg


\bvg\bva\mssnote{\Regius~2v/4, \Hauksbok~20v/32}%
\alst{G}ęyr nú \alst{G}armr mjǫk \hld\ fyr \alst{G}nipa-hęlli, &
\alst{f}ęstr mun slitna, \hld\ en \alst{f}reki rinna; &
\alst{f}jǫlð vęit hǫ̇n \alst{f}rǿða, \hld\ \alst{f}ramm sé’k lęngra &
of \alst{r}agna \alst{r}ǫk \hld\ \alst{r}ǫmm sig-tíva.\eva

\bvb Now Garm barks loud before the Gnip-caves; \\
the rope will tear and the Wolf will run. \\
She knows much lore; I foresee further \\
about the mighty Rakes of the Reins, of the Victory-Tews \ken{gods}.\evb\evg


\bvg\bva\mssnote{\Regius~2v/4, \Hauksbok~20v/32, \RegiusProse\Trajectinus\Wormianus}%
\alst{H}rymr ękr austan, \hld\ \alst{h}ęf⸗sk lind fyrir, &
sný⸗sk \alst{Jǫ}rmun·gandr \hld\ ï \alst{jǫ}tun-móði, &
\alst{o}rmr knýr \alst{u}nnir, \hld\ \edtrans{en \alst{a}ri hlakkar}{and the eagle screams}{\Afootnote{\emph{ǫrn mun hlakka} ‘the eagle will scream’ \RegiusProse\Trajectinus}}, &
slítr \alst{n}ái \alst{n}ęf-fǫlr; \hld\ \edtrans{\alst{N}agl·far}{Nailfare}{\Bfootnote{The ship on which the fiends sail from the east.  It is described in \Gylfaginning\ 51: \emph{Þá geysist haf’it á lǫnd’in, fyrir því at þá sný⸗st Mið·garðs-ormr í jǫtun-móð ok sǿkir upp á land’it.  Þá verðr ok þat, at Nagl·far losnar, skip þat, er svá heitir.  Þat er gert af nǫglum dauðra manna, ok er þat fyrir því varnanar vert, ef maðr deyr með ó·skornum nǫglum, at sá maðr eykr mikit efni til skips’ins Nagl·fars, er goð’in ok menn vildi seint, at gert yrði.  En í þessum sę́var-gang flýtr Nagl·far.} ‘Then the sea surges onto the lands because the Middenyardswyrm writhes in ettin-wrath and attacks the land.  Then it also happens that Nailfare comes loose—the ship which is called thus; it is made from the nails of dead men, and therefore it is worth a warning if a man dies with uncut nails, that that man greatly increases the material for the ship Nailfare, which the Gods and men would rather were not done.  But in this motion of the sea Nailfare floats.’

The idea of the Devil rowing a boat made out of improperly discarded nail clippings is widespread in Scandinavian and Finnish folklore and thus the “nail-ship” cannot be a mere Snorronean invention.  Specific rituals for disposing cut nails (and hair) are found in many Indo-European cultures, and clearly described ramifications for not observing them are also found in Zoroastrianism (\emph{Vidēvdāt} 17.2–3: “[W]hen one arranges and cuts his hair and clips his nails and then lets them fall into holes in the earth or into furrows, [...] demons come forth, and from these improprieties monsters come forth from the earth which mortals call lice and which devour the grain in the fields and the clothes in the closets”) \parencite{Lincoln1977}.  Beyond the Indo-Europeans, Judaism also has a taboo against discarding nail-clippings wherever a pregnant woman might walk, lest they (magically) cause her to miscarry.

An interesting psychological commonality presents itself in all three traditions, namely that the improper clipping or disposal of nails is thought to threaten the fundamental stability of the cosmos by bringing about the principial fear of each culture.  In Judaism, singularly concerned with the fertility of its adherents (witness its famous ban on masturbation, \emph{Shulchan Arukh, Even HaEzer} 23:1: \texthebrew{ועון זה חמור מכל עבירות שבתורה} ‘more severe than all Torah transgressions’) it is the miscarriage; in the dualistic Zoroastrianism the proliferation of demons; and in the Germanic religion, with its Wiking Age eschatological obsession (of which \textlink{Voluspa} is an example, as is the Eddic \textlink{Vafthrudnismal}, the Scaldic poems \Hakonarmal\ and \emph{Eirm}, and the Runic Sö 154 Skarpåker) it is the end times.}} losnar.\eva

\bvb Rim drives from the east, holds his shield before him; \\
Ermengand writhes in ettin-wrath. \\
The Wyrm drives forth the waves and the eagle screams: \\
the pale-beak tears at corpses; Nailfare comes loose.\evb\evg


\bvg\bva\mssnote{\Regius~2v/6, \Hauksbok~20v/34, \RegiusProse\Trajectinus\Wormianus}%
\alst{K}jóll fęrr austan \hld\ \alst{k}oma munu Múspells &
of \alst{l}ǫg \alst{l}ýðir, \hld\ en \alst{L}oki stýrir; &
\alst{f}ara \alst{f}ífl-męgir \hld\ með \alst{f}reka allir, &
þęim es \alst{b}róðir \hld\ \alst{B}ýlęists ï fǫr.\eva

\bvb The ship fares from the east—come will Muspell’s \\
subjects o’er the sea—and Lock steers it. \\
The devil-lads all fare with the Wolf; \\
with them comes the brother of Bylest \ken*{= Lock} along.\evb\evg


\bvg\bva\mssnote{\Regius~2v/10, \Hauksbok~20v/36, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\edtext{\alst{S}urtr}{\Afootnote{\emph{Svartr} \Upsaliensis}} fęrr \alst{s}unnan \hld\ með \alst{s}viga lę́vi, &
\edtrans{skïnn af \alst{s}verði \hld\ \alst{s}ól val-tíva}{from the sword of the slaughter-Tews [it] shines [like] the sun}{\Bfootnote{A difficult line in several regards.  First, \emph{val-tíva} may be read either (1) as gen. pl. of \emph{val-tívar} ‘the slain-Tews \ken{gods}’ (so \textcite{LaFargeGlossary}; cf. below st. 59 and \textlink{Hymiskvida} 1) or (2) as gen. sg. of an unattested masc. \emph{n}-stem \emph{*val-tívi} ‘the slaughter-tew \ken*{= Surt}’ (so \CV, Fritzner and others).
(1) is adopted here as the simpler explanation since it does not presuppose an otherwise unattested derivative of \emph{týr} ‘tew, god’ and since the pl. \emph{val-tívar} demonstrably occurs below in the poem in st. 59, but the sense of the line is still very difficult.

A second issue arises over whether the gen. \emph{val-tíva} modifies (a) \emph{sverði} ‘sword’ or (b) \emph{sól} ‘sun’.  In cases (1a) and (2a) the sense would simply be that the flashing sword shines like the sun, reflecting the intensity of the battle.  Although one may expect a pl. \emph{sverðum} ‘swords’ for case (1a) that is not a major problem, since sg. forms are often used with pl. possessors in Old Germanic languages.  Case (1b) must probably be rejected, since the possessor of the sword has to be Surt, and “the sun of the \textsc{gods} shines from Surt’s sword” makes little sense.  If (2b) is adopted we may see a kenning “sun of Surt \ken{fire}”, but “from his sword shines the \ken{fire}” seems excessively repetitive when compared to l. 1.  In conclusion, the preferred sense is case (1a), although it is still not perfect.}}; &
\alst{g}rjót-bjǫrg \alst{g}nata, \hld\ en \edtrans{\alst{g}ífr rata}{the fiends are on the march}{\Afootnote{\emph{guðar hrata} ‘the gods stagger’ \Upsaliensis}\Bfootnote{The reading of \Upsaliensis\ is wo. doubt corrupt; the anachronistic masc. pl. ending \emph{-ar} is proof enough, for the word \emph{goð} \char`~\ \emph{guð} ‘gods’ was always neuter in heathen times.}}, &
troða \alst{h}alir \edtrans{\alst{h}ęl-veg}{the Hellway}{\Bfootnote{The road between the earth and the underworld on which one has to travel after death to reach one’s final resting place (cf. the story told in \textlink{Helreid}).  According to \Gylfaginning\ 51, Lock is followed by \emph{allir Hęljar sinnar} ‘all companions of Hell’, who are presumably thought to be some sort of revenant zombies.}}, \hld\ en \alst{h}iminn klofnar.\eva

\bvb Surt fares from the south with the twig’s betrayer \ken{fire}; \\
from the sword of the slaughter-Tews [it] shines [like] the sun. \\
Rocky cliffs clash and the fiends are on the march; \\
men tread the \inx[P]{Hellway} and the heaven is split.\evb\evg


\bvg\bva\mssnote{\Regius~2v/13, \Hauksbok~20v/37, \RegiusProse\Trajectinus\Wormianus}%
\Ballnote{Sts. 51–53 describe the deaths of the major gods Weden, Free, and Thunder.  The battle takes place on the great plain Wighride as told in \textlink{Vafthrudnismal}[18]; the fight between the Wolf and Weden, and his subsequent avenging by Widar is mentioned in \textlink{Vafthrudnismal}[53].  According to \Gylfaginning\ 51: \emph{Ę́sir her-vę́ða sik ok allir Ein-herjar ok sǿkja fram á vǫllu’na.  Ríðr fyrstr Óðinn með gull-hjalm’inn ok fagra brynju ok geir sinn, er Gungnir heitir; stefnir hann móti Fenris-ulf, en Þórr fram á aðra hlið honum, ok má hann ekki duga honum, því at hann hefir fullt fang at berja⸗st við Miðgarðs-orm.  Freyr ber⸗st móti Surti, ok verðr harðr sam-gangr, áðr Freyr fellr.  Þat verðr hans bani, er hann missir þess ins góða sverðs, er hann gaf Skírni.} ‘The Eese and all the Oneharriers clothe themselves for war and rush forth on the plains.  Weden rides first with the golden helmet and fair byrnie and his spear which is called Gungner; he faces against the Fenrerswolf, but Thunder goes forth on one of his sides; and he cannot avail him, for he has his hands full in fighting against the Middenyardswyrm.  Free fights against Surt and it becomes a hard-fought struggle before Free falls.  It will be his bane that he is missing the good sword which he gave Shirner.’}%
Þȧ kømr \edtrans{\alst{H}línar \hld\ \alst{h}armr annarr}{Line’s second sorrow}{\Bfootnote{The first sorrow being the death of Balder.  Line is described in \Gylfaginning\ 35 as a minor goddess \emph{sett til gę́zlu yfir þeim mǫnnum, er Frigg vill forða við háska nǫkkurum} ‘placed to watch over those men which Frie wishes to save from any particular danger’.  In spite of this almost all translators and editors have understood Line as synonymous with Frie and questioned whether her existence as a distinct goddess is not something invented by the author of \Gylfaginning.  \textcite{Hopkins2017} reasonably argues that this need not be the case; as Frie’s maidservant, Line’s two sorrows would consist in her failure to protect both the son and husband of her mistress.}} framm, &
es \alst{Ó}ðinn fęrr \hld\ við \alst{u}lf vega, &
—\edtrans{en \alst{b}ani \alst{B}ęlja \hld\ \alst{b}jartr at Surti}{but the bane of Bellower \ken*{= Free}, bright, towards Surt}{\Bfootnote{In a single line the wallow tells about the death of Free; cf. note to ALL.  Free’s obscure fight against Bellower (ON \emph{Bęli}) is mentioned in \Gylfaginning\ 37, which also explains the curious detail of the sword (for which cf. \textlink{Skirnismal} Introduction and st. 8): \emph{Þessi sǫk er til þess, er Freyr var svá vápn-lauss, er hann barði⸗st við Belja ok drap hann með hjartar-horni.} ‘This event (viz. Free giving Shirner his sword) is the reason for why Free was so unarmed when he fought against Bellower and slew him with a hart’s antler.’}}— &
þȧ mun \alst{F}riggjar \hld\ \alst{f}alla \edtext{angan}{\Afootnote{so \Hauksbok\RegiusProse\Trajectinus\Upsaliensis\Wormianus; \emph{angantyr} \Regius}}.\eva

\bvb Then comes \inx[P]{Line}’s second sorrow to pass, \\
when Weden goes to fight against the Wolf \\
—but the bane of Bellower \ken*{= Free}, bright, towards Surt— \\
then will Frie’s beloved \ken*{= Weden} fall.\evb\evg


\bvg\bva\mssnote{\Regius~2v/15, \RegiusProse\Trajectinus\Wormianus}%
\edtrans{Þȧ kømr hinn \alst{m}ikli \hld\ \alst{m}ǫgr Sig-fǫður}{Then comes the great lad of Syefather}{\Afootnote{\emph{Gęngr Óðins sonr \hld\ við ulf vega} ‘Weden’s son goes the Wolf to fight’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus.}}, &
\alst{V}íðarr \edtext{\alst{v}ega}{\Afootnote{\emph{of veg} \RegiusProse\Trajectinus\Upsaliensis\Wormianus}} \hld\ at \alst{v}al-dýri; &
lę́tr \alst{m}ęgi \edtrans{Hveðrungs}{Whethring}{\Bfootnote{An obscure name for \inx[P]{Lock}, whose son is the Wolf.}} \hld\ \alst{m}und of standa &
\alst{h}jǫr til \alst{h}jarta; \hld\ þȧ ’s \alst{h}efnt fǫður.\eva

\bvb Then comes the great lad of \inx[P]{Syefather}, \\
Wider, to fight that slaughter-beast. \\
He lets his hand through \inx[P]{Whethring}’s lad \ken*{= the Wolf} \\
drive the sword to the heart—then the father is avenged!\evb\evg


\bvg\bva\mssnote{\Regius~2v/17, \Hauksbok~20v/41, \RegiusProse\Trajectinus\Wormianus}%
\Ballnote{Thunder and the Middenyardswyrm kill each other, fulfilling the unfinished fight seen in the Fishing Expedition; see \textlink{Hymiskvida}.  From a comparative perspective it is notable that the Norse religion makes the ancient \emph{Chaoskampf} motif of a Storm God facing off against a watery Serpent a key part of its eschatology.  It is for this reason that the Wyrm cannot die in the Fishing Expedition (as it must have done originally, and as it actually does in some of the Scaldic retellings), since it must remain alive to fight at the end of days.}%
\Aallnote{The present version of the stanza is an amalgamation of all three ms. traditions (\Regius, \Hauksbok, and \RegiusProse\Trajectinus\Upsaliensis\Wormianus) based most closely on the latter two, which have the last 3 lines in the same order.  \Regius\ has the lines in the order 1, 5, 4, 2, 3, and inserts an additional line between 1 and 5.}%
\edtrans{\edtrans{Þȧ kømr}{Then comes}{\Afootnote{\emph{Gęngr} ‘Goes’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}} hinn \alst{m}ę́ri \hld\ \alst{m}ǫgr \edtrans{Hlǫðynjar}{Lathyn}{\Afootnote{add. \emph{gęngr Óðins sonr \hld\ við orm vega.} ‘Weden’s son goes the Wyrm to fight.’ \Regius.}},}{Then comes the famed lad of Lathyn}{\Afootnote{om. \Hauksbok.}} &
\edtrans{gęngr \alst{f}et níu \hld\ \alst{F}jǫrgynjar burr}{nine paces takes Firgyn’s son}{\Afootnote{om. \RegiusProse\Trajectinus\Upsaliensis\Wormianus.}} &
\alst{n}ęppr frȧ \alst{n}aðri \hld\ \alst{n}íðs ȯ·kvíðnum; &
\edtrans{munu \alst{h}alir allir \hld\ \alst{h}ęim-stǫð ryðja}{All men will clear their homesteads}{\Bfootnote{After Thunder is slain the realm of men is no longer habitable.  Cf. \textlink{Harbardsljod}[23], \textlink{Thrymskvida}[18].}} &
\edtext{es af \alst{m}óði drepr}{\Afootnote{\emph{drepr hann af móði} \Regius}} \hld\ \edtrans{\alst{M}ið-garðs véurr}{Middenyard’s Wighward}{\Bfootnote{“The Guardian of the Sanctuaries of Middenyard”; a fitting kenning.}}.\eva

\bvb Then comes the famed lad of Lathyn; \\
nine paces takes Firgyn’s son \\
pained, away from the shameful Adder. \\
All men will clear their homesteads \\
when Middenyard’s Wighward strikes out of wrath.\evb\evg


\bvg\bva\mssnote{\Regius~2v/20, \Hauksbok~21r/1, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\Ballnote{This st. is echoed in Arn \emph{Þorfdr} 24/1–2 (\Skp\ 2) in an instance of \emph{adynaton}:
\emph{Bjǫrt verðr sól at svartri; \hld\ søkkr fold ï mar døkkvan; / brestr ęrfiði Austra; \hld\ allr glymr sę́r ȧ fjǫllum} ‘Bright, the sun will turn to black; the land will sink into the dark sea; // Eastre’s toil \ken{heaven} will shatter; all the sea will roar over the fells’.}%
\alst{S}ól tér \alst{s}ortna, \hld\ \edtrans{\edtrans{\alst{s}ígr}{sinks}{\Afootnote{ \emph{søkkr} \RegiusProse\Trajectinus\Wormianus}} fold ï mar}{the land sinks into sea}{\Bfootnote{The reading \emph{søkkr} ‘sinks’ is probably influenced by Arn \emph{Þorfdr} 24/1b (\Skp\ 2) but is stemmatically excluded.}}, &
\alst{h}verfa af \alst{h}imni \hld\ \alst{h}ęiðar stjǫrnur; &
gęisar \alst{ęi}mi \hld\ við \alst{a}ldr-nara; &
lęikr \alst{h}ǫ́r \alst{h}iti \hld\ við \alst{h}imin sjalfan.\eva

\bvb The Sun doth blacken; the land sinks into sea; \\
from the heaven fade the shining stars. \\
Smoke billows from the nourisher of life \ken{fire}; \\
high heat licks the very heaven.\evb\evg


\bvg\bva\mssnote{\Regius~2v/22, \Hauksbok~21r/2}%
\Ballnote{With the sinking of the earth into the sea and the blackening of the heaven the destruction has reached its climax, as signalled by the final repetition of the refrain stanza.  To indicate the dramatic pause, a black page has been inserted.}%
\alst{G}ęyr nú \alst{G}armr mjǫk \hld\ fyr \alst{G}nipa-hęlli, &
\alst{f}ęstr mun slitna, \hld\ en \alst{f}reki rinna; &
\alst{f}jǫlð vęit hǫ̇n \alst{f}rǿða, \hld\ \alst{f}ramm sé’k lęngra &
of \alst{r}agna \alst{r}ǫk, \hld\ \alst{r}ǫmm sig-tíva.\eva

\bvb Now Garm barks loud before the Gnip-caves; \\
the rope will tear and the Wolf will run. \\
She knows much lore; I foresee further \\
about the mighty Rakes of the Reins, of the Victory-Tews \ken{gods}.\evb\evg

\newpage
\pagecolor{black}
\thispagestyle{empty}

\null
\vspace*{\fill}
%\color{white}
%\begin{center}
%\Huge Darkness.
%\end{center}

\newpage
\nopagecolor
\color{black}

\bvg\bva\mssnote{\Regius~2v/23, \Hauksbok~21r/4}%
\Ballnote{Sts. 56–59 are paraphrased in \Gylfaginning\ ch. 53:

	\emph{Þá mę́lti Gangleri: „Hvárt lifa nǫkkur goð’in þá, eða er þá nǫkkur jǫrð eða himinn?“ Hárr segir: „Upp skýtr jǫrðu’nni þá ór sę́’num, ok er þá grǿn ok fǫgr. Vaxa þá akrar ó·sánir. Víðarr ok Váli lifa, svá at eigi hefir sę́r’inn ok Surta-logi grandat þeim, ok byggja þeir á Iða-velli, þar sem fyrr var Ás-garðr, ok þar koma þá synir Þórs, Móði ok Magni, ok hafa þar Mjǫllni. Því nę́st koma þar Baldr ok Hǫðr frá Heljar, setja⸗st þá allir samt, ok tala⸗st við, ok minna⸗st á rúnar sínar, ok rǿða of tíðendi þau, er fyrrum hǫfðu verit, of Mið·garðs-orm ok um Fenris-ulf.  Þá finna þeir í grasi’nu gull-tǫflur þę́r, er ę́sir’nir hǫfðu átt. Svá er sagt:“}

	‘Then spoke Gangler: “Do any of the gods live then, or is there then any earth or heaven?”  High says: “Then the earth shoots up from the seas, and it is then green and fair.  Then acres grow unsown.  Wider and Wonnel live, for the sea and Surt’s flame have not scathed them, and they settle on the Idewolds where Osyard once stood.  And then the sons of Thunder—Mood and Main—come there, and there they have Millner.  Next come Balder and Hath from Hell; then they all sit down together and make speech and think back on their runes and discuss the events of antiquity, about the Middenyardswyrm and about the Fenrerswolf.  Then they find in the grass those golden game-bricks which the Eese had owned.  So it is said:”’

after which the text quotes \textlink{Vafthrudnismal}[51].}%
Sér hǫ̇n \alst{u}pp koma \hld\ \edtrans{\alst{ǫ}ðru sinni}{a second time}{\Bfootnote{The first time probably being the lifting of the Earth in st. 4.}} &
\alst{jǫ}rð ór \alst{ę́}gi \hld\ \alst{i}ðja-grø̇na; &
\alst{f}alla \alst{f}orsar, \hld\ \alst{f}lýgr ǫrn yfir, &
sá’s ȧ \alst{f}jalli \hld\ \alst{f}iska vęiðir.\eva

\bvb She sees coming up for a second time \\
the Earth from the ocean ever green anew. \\
Torrents fall; an eagle flies o’er them, \\
he who in the fells catches fish.\evb\evg


\bvg\bva\mssnote{\Regius~2v/24, \Hauksbok~21r/5}%
\edtrans{Finna⸗sk}{find each other}{\Afootnote{\emph{hitta⸗sk} \Hauksbok\ provides closer parallelism with st. 7, but for the same reason it may also have replaced earlier \emph{finna⸗sk}.}} \alst{ę̇}sir \hld\ ȧ \alst{I}ða-vęlli &
ok umb \edtrans{\alst{m}old-þinur}{Earth-cord}{\Bfootnote{Cf. the kenning for the Middenyardswyrm in ÚlfrU \emph{Húsdr} 4: \emph{stirð-þinull storðar} ‘the stiff cord of the land \ken*{= Middenyardswyrm}’}} \hld\ \alst{m}ǫ́tkan dø̇ma, &
\edtrans{ok \alst{m}inna⸗sk þar \hld\ ȧ \alst{m}ęgin-dȯma}{and there think back on mighty verdicts}{\Afootnote{so \Hauksbok; om. \Regius}} &
ok ȧ \alst{F}imbul-týs \hld\ \alst{f}ornar ru̇nar.\eva

\bvb The Eese find each other on the Idewolds \\
and of the mighty Earth-Cord \ken*{= Middenyardswyrm} speak, \\
and there think back on mighty verdicts \\
and on Fimble-Tew’s \name{= Weden’s} ancient runes.\evb\evg


\bvg\bva\mssnote{\Regius~2v/26, \Hauksbok~21r/7}%
Þar munu \edtrans{\alst{ę}ptir}{again}{\Afootnote{\emph{ę̇sir} ‘the Eese’ \Hauksbok}} \hld\ \edtext{\alst{u}ndr-sam⸗ligar &
\alst{g}ollnar tǫflur}{\lemma{undr-sam⸗ligar gollnar tǫflur ‘wondersome golden game-bricks’}\Bfootnote{A fine literary device.  In st. 8 the golden age of the Eese, exemplified by their playing board games, was spoiled by the three ettin-women.  The rediscovering of the golden board game then betokens a new golden age.}} \hld\ ï \alst{g}rasi \edtrans{finna⸗sk}{be found}{\Afootnote{\emph{finna} ‘find’ \Hauksbok}}, &
þę́r’s \edtext{\emph{\alst{ę̇}sir}}{\Afootnote{emend.; om. \Regius\Hauksbok}\Bfootnote{The verb \emph{hǫfðu} ‘had’ requires a plural subject, and this is found in the paraphrase of \Gylfaginning\ 53: \emph{Þá finna þeir í grasi’nu gull-tǫflur þę́r, er ę́sir’nir hǫfðu átt} (see st. 56 n. above).}} ï \alst{á}r-daga \hld\ \alst{á}ttar hǫfðu.\eva

\bvb There will again the wondersome \\
golden game-bricks in the grass be found, \\
which in days of yore the Eese had owned.\evb\evg


\bvg\bva\mssnote{\Regius~2v/28, \Hauksbok~21r/9}%
Munu \alst{ȯ}·sánir \hld\ \alst{a}krar vaxa, &
\edtrans{\alst{b}ǫls}{the bale}{\Bfootnote{The evil of Hath’s slaying Balder will be forgotten as the two live together in peace.}} mun alls \alst{b}atna, \hld\ mun \alst{B}aldr koma; &
búa \alst{H}ǫðr ok Baldr \hld\ \alst{H}ropts sig-toptir, &
\alst{v}ęl \alst{v}al-tívar. \hld\ \alst{V}ituð ér ęnn eða hvat?\eva

\bvb Unsown will acres grow; \\
the bale will all be bettered; Balder will come. \\
Hath and Balder live on Roft’s \name{= Weden’s} victory-plots \\
well, the slain-Tews. \emph{Know ye yet, or what?}\evb\evg


\bvg\bva\mssnote{\Regius~2v/30, \Hauksbok~21r/11}%
Þȧ kná \alst{H}ø̇nir \hld\ \edtrans{\alst{h}laut-við kjósa}{choose the leat-wood}{\Bfootnote{Foresee the future by means of lots, specifically twigs drenched in blood of holy beasts.  See \textlink{Hymiskvida}[1] and Index: \inx[C]{leat}.}} &
ok \alst{b}urir \alst{b}yggva \hld\ \edtrans{\alst{b}rǿðra tvęggja}{of the two brothers}{\Bfootnote{The present translation understands \emph{tvęggja} as the gen. pl. of \emph{tvęir} ‘two’; the two brothers are presumably Hath and Balder, mentioned in the previous stanza.  Since the original ms. does not capitalize proper nouns one could also read \emph{brǿðra Tvęggja} ‘the brothers of Tway \name{= Weden}’.  Weden’s brothers are attested in \Gylfaginning\ 6 as \inx[P]{Will} and \inx[P]{Wigh}, but they are never said to have children.}} &
\alst{v}ind-hęim \alst{v}íðan. \hld\ \alst{V}ituð ér ęnn eða hvat?\eva

\bvb Then does Heener choose the \inx[C]{leat}-wood, \\
and the sons of the two brothers settle \\
the wide wind-home \ken{sky/heaven}. \emph{Know ye yet, or what?}\evb\evg


\bvg\bva\mssnote{\Regius~2v/31, \Hauksbok~21r/12, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{S}al \edtrans{sér hǫ̇n}{she sees}{\Afootnote{\emph{vęit’k} ‘I know’ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}} standa \hld\ \alst{s}ólu fęgra, &
\edtrans{\alst{g}olli þakðan}{thatched with gold}{\Afootnote{\emph{golli bętra} ‘better than gold’ \RegiusProse\Trajectinus}}, \hld\ ȧ \edtext{\alst{G}im·léi}{\Afootnote{metr. emend.; \emph{Gim·lé} \Regius\Hauksbok\RegiusProse\Trajectinus\Upsaliensis\Wormianus}}; &
\edtrans{þar}{there}{\Afootnote{\emph{þann} ‘[in] that [hall]’ \Trajectinus\Wormianus}} skulu \alst{d}yggvar \hld\ \alst{d}róttir byggva &
ok umb \alst{a}ldr-daga \hld\ \alst{y}nðis njóta.\eva

\bvb A hall she sees standing fairer than the sun, \\
thatched with gold on Gemlee. \\
There shall faithful folk settle \\
and in their days of life enjoy delight.\evb\evg


\bvg\bva\mssnote{\Regius~3r/2, \Hauksbok~21r/15}%
Þar kømr hinn \alst{d}immi \hld\ \alst{d}ręki fljúgandi, &
\alst{n}aðr frȧnn \alst{n}eðan \hld\ frȧ \alst{N}iða-fjǫllum; &
berr sér ï \alst{f}jǫðrum \hld\ —\alst{f}lýgr vǫll yfir— &
\alst{N}íð·hǫggr \alst{n}ái; \hld\ \edtrans{\alst{n}ú mun hǫ̇n søkkva⸗sk}{Now will she sink!}{\Bfootnote{The spae is concluded and the wallow, referring to herself in third person, descends back down into the grave whence Weden woke her.  Cf. the very last half-line of \textlink{Helreid}: \emph{søkkst-u, gýgjar-kyn} ‘sink, thou gow’s kin!’}}.\eva

\bvb Then comes the dusky dragon flying, \\
the gleaming adder up from the \inx[P]{Nithfells}. \\
He carries in his feathers—he flies o’er the field— \\
Nithehewer, corpses.—Now will she sink!”\evb\evg

\sectionline

\newpage

\section{Appendix}

\subsection{Dwarf-tallies}

The following sts. (11–15) contain two or three originally distinct lists of dwarf-names.  They interrupt the flow of the whole poem so much that they have been moved to this appendix.

That the lists are originally distinct is seen by the repetition of names (13/4, 15/4: \emph{Ęikin·skjaldi} ‘Oakenshield’; 11/4, 15/3: \emph{Ái} ‘Great-grandfather’) and existence of two formulaic conclusions (12/3–4, 15/6–7).  Sts. 11–13, having no repeated names, seem to belong together. If they do, st. 12, which contains the formulaic conclusion to the list, should probably switch places with 13.  Sts. 14–15 form the second group, having an introduction and a conclusion which both mention the dwarf Loffer.

\bvg\bva[11]\mssnote{\Regius~1r/23, \Hauksbok~20r/17, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{N}ýi ok \alst{N}iði, \hld\ \alst{N}orðri, Suðri, &
\alst{Au}stri, Vestri, \hld\ \alst{A}l·þjófr, Dvalinn, &
\alst{B}ívurr, \alst{B}ávurr, \hld\ \alst{B}ǫmburr, Nóri, &
\alst{Ȧ}nn ok \alst{Ȧ}narr, \hld\ \alst{Á}i, Mjǫð·vitnir.\eva

\bvb New and Nithe, Norther and Souther, \\
Easter and Wester, Allthief, Dwollen, \\
Bewer, Bower, Bamber, Noor, \\
Own and Owner, Great-grandfather, Meadwitner.\evb\evg


\bvg\bva[12]\mssnote{\Regius~1r/25, \Hauksbok~20r/18, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{V}ęigr ok Gand·alfr, \hld\ \alst{V}ind·alfr, Þráinn, &
\alst{Þ}ękkr ok \alst{Þ}orinn, \hld\ \alst{Þ}rór, Vitr ok Litr, &
\alst{N}ár ok \alst{N}ý·ráðr— \hld\ \alst{n}ú hęf’k dverga &
—\alst{R}ęginn ok \alst{R}áð·sviðr— \hld\ \alst{r}étt of talða.\eva

\bvb Wey and Gandelf, Windelf, Thrown, \\
Thetch and Thorn, Threw, Wit and Lit, \\
Nee and Newred—now have I the dwarfs— \\
Rain and Redswith—rightly tallied.\evb\evg


\bvg\bva[13]\mssnote{\Regius~1r/28, \Hauksbok~20r/20, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\alst{F}íli, Kíli, \hld\ \alst{F}undinn, Náli, &
\alst{H}ępti, Víli, \hld\ \alst{H}annarr, \edtext{Svíurr}{\Afootnote{\emph{†Sviðrr†, / Nár ok Náinn, \hld\ Nípingr, Dáinn / Billingr, Brúni \hld\ Bildr ok Búri} ‘\dots\ Nee and Nowen, Nipng, Dowen, Billing, Brown, Bild and Boor’ \Hauksbok}}, &
\alst{F}rár, Horn·bori, \hld\ \alst{F}rę́gr ok Lȯni, &
\alst{Au}r·vangr, \alst{Ja}ri, \hld\ \alst{Ęi}kin·skjaldi.\eva

\bvb Filer, Chiler, Found and Needler, \\
Hefter, Wiler, Hanner, Swigher, \\
Fraw, Hornborer, Fray and Looner, \\
Earwong, Earer, Oakenshield.\evb\evg


\bvg\bva[14]\mssnote{\Regius~1r/30, \Hauksbok~20r/22, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
\Ballnote{A standard genealogical introduction (cf. \Haleygjatal\ 1: \emph{meðan hans ę́tt \dots\ til goða tęljum} ‘while we tally his line \dots\ back to the gods’); the patrilineal line of dwarfs is to be counted back to their progenitor, Loffer.  This possibly disagrees with st. 10, where Moodsowner is said to be the foremost (and presumably the oldest) of the dwarfs, and Loffer is not mentioned, but such details were probably not very important.

Sts. 14–15 is paraphrased in \Gylfaginning\ 14: “But these came from Swornshigh (\emph{Svarinshaugr}) to the Earwongs on the Erwolds, and from them Loffer is come—these are their names: Sherper (\emph{Skirpir}), Werper (\emph{Virpir}), Showfind, Great-grandfather, Elf and Ing (\emph{Ingi}), Oakenshield, Fale (\emph{Falr}), Frost, Finn, Ginner.”}%
Mál es \alst{d}verga \hld\ ï \alst{D}valins liði &
\alst{l}jȯna kindum \hld\ til \alst{L}ofars tęlja, &
\edtext{þęir}{\Afootnote{\emph{þeim} \Hauksbok}} es \alst{s}óttu \hld\ frȧ \alst{s}alar stęini &
\alst{Au}r-vanga sjǫt \hld\ til \alst{Jǫ}ru-valla.\eva

\bvb It is time to tally the dwarfs in Dwollen’s troop \\
back to Loffer for the races of men; \\
they who sought, from the stone of the hall, \\
the seat of the \inx[P]{Earwongs} unto the \inx[P]{Erwolds}.\evb\evg


\bvg\bva[15]\mssnote{\Regius~1r/32, \Hauksbok~20r/24, \\ \RegiusProse\Trajectinus\Upsaliensis\Wormianus}%
Þar vas \alst{D}raupnir \hld\ ok \alst{D}olg·þrasir, &
\alst{H}ár, \alst{H}aug·spori, \hld\ \alst{H}lé·vangr, Glói, &
\alst{Sk}irfir, Virfir, \hld\ \alst{Sk}á·fiðr, Ái, &
\alst{A}lfr ok \alst{Y}ngvi, \hld\ \alst{Ęi}kin·skjaldi, &
\alst{F}jalarr ok \alst{F}rosti, \hld\ \alst{F}innr ok Ginnarr; &
\edtrans{Þat mun \edtext{\alst{ę́}}{\Afootnote{so \Hauksbok; om. \Regius}} \alst{u}ppi, \hld\ meðan \alst{ǫ}ld lifir}{This will ever be remembered while the age lives}{\Bfootnote{Both verses are formulaic.  For the first cf. \HervararSaga\ 119 (\Skp\ 7): \emph{þat mun ę́ uppi; \hld\
illr es dómr norna!} ‘That will always be remembered; evil is the judgement of the norns!’
For the second cf. runic inscription U 323 (ca. 980–1015), edited below.  What is essentially a Scaldic rewrite of the whole line is found in Þstf \emph{Stuttdr} 5/1–2 (\Skp\ 2): \emph{Ęy mun uppi \hld\ Ęndils, meðan stęndr // sól-borgar salr, \hld\ svǫr-gǿðis fǫr} ‘Ever will be remembered—while the hall of the sun’s stronghold \ken{sky/heaven > earth} stands—the journey of the fattener of Andle’s bird \ken{raven/eagle > warrior}.’}}, &
\alst{l}ang-niðja-tal \hld\ \edtext{til}{\Afootnote{om. \Hauksbok}} \alst{L}ofars hafat.\eva

\bvb There was Dreepner and Dollowthrasher, \\
High, Highspurer, Leewong, Glower, \\
Sherver, Werver, Showfind, Great-grandfather, \\
Elf and Ing, Oakenshield, \\
Feller and Frost, Finn and Ginner. \\
This will ever be remembered while the age lives, \\
the tally of ancestors lifted to Loffer.\evb\evg

\subsection{Stanzas from \emph{Hauksbók}}

\Hauksbok\ has a few substantial inserts and differences from \Regius.  Their style strongly suggests that they are later compositions.

\bvg\bva[H1]\mssnote{\Hauksbok~20v/12}%
\Ballnote{34/1–2 are replaced by these lines in \Hauksbok.}%
Þȧ kná \edtrans{\alst{V}áli}{Wonnel}{\Afootnote{emend.; \emph{Vála} \Hauksbok}} \hld\ \alst{v}íg-bǫnd snúa &
\alst{h}ęldr vǫ́ru \alst{h}arð-gǫr \hld\ \alst{h}ǫpt ór þǫrmum.\eva

\bvb Then did \inx[P]{Wonnel} the war-bonds twist: \\
the most sturdy fetters were made from entrails.\evb\evg


\bvg\bva[H2]\mssnote{\Hauksbok~20v/28}%
\Ballnote{45/5–6 is followed by these lines in \Hauksbok, in effect forming another four-line stanza.  They seem to be referring to an ettin swallowing the World Tree.}%
\alst{H}rę́ða⸗sk allir \hld\ ȧ \alst{h}ęl-vegum &
áðr \alst{S}urtar þann \hld\ \alst{s}efi of glęypir.\eva

\bvb All are frightened on the Hell-ways \\
before Surt’s kinsman does devour it.\evb\evg


\bvg\bva[H3]\mssnote{\Hauksbok~20v/39}%
\Aallnote{The last part of the stanza is almost completely illegible in the ms.  I have relied on the transliteration of \textcite[13,44\psqq]{JonHelgason1971}.}%
\Ballnote{This stanza appears between 52 and 53 in \Hauksbok.  It is not found in the sequence of stanzas cited in \Gylfaginning\ 51, but some details in the accompanying prose are not entirely dissimilar: \emph{en Fenris-ulfr ferr með gapandi munn, ok er inn neðri kjǫptr við jǫrðu, en in øfri við himin.  Gapa myndi hann meira, ef rúm vę́ri til.  Eldar brenna ór augum hans ok nǫsum.  Miðgarðs-ormr blę́ss svá eitri’nu, at hann dreifir lopt ǫll ok lǫg, ok er hann all-ógur⸗ligr, ok er hann á aðra hlið ulfi’num.} ‘But the Fenrerswolf fares with gaping mouth; and the lower jaw is on the earth, but the upper is in the heaven; he would gape yet broader if there were room for it.  Fires burn from his eyes and nostrils.  The Middenyardswyrm blows the venom so far that it covers the whole air and sea, and he is most terrible, and he is by the side of the Wolf.’}%
Gïnn \alst{l}opt yfir \hld\ \alst{l}indi jarðar, &
gapa \alst{ý}gs kjaptar \hld\ \alst{o}rms ï hę́ðum; &
mun \alst{Ó}ðins son \hld\ \edtrans{\emph{\alst{ęi}tr}i}{venom}{\Afootnote{emend.; \emph{ormi} ‘Wyrm’ \Hauksbok.}\Bfootnote{Cf. \Gylfaginning\ 51: “Thunder bears the bane-word from the Middenyardswyrm and strides nine paces away from it. Then he falls dead to the earth for the venom (\emph{ęitri}) which the Wyrm blows on him.”}} mǿta &
\alst{v}args at \edtext{da\emph{uða}}{\Afootnote{\emph{‘da...’} \Hauksbok}} \hld\ \alst{V}íðars niðja.\eva

\bvb Over the air yawns the Girdle of the Earth \ken*{= Middenyardswyrm}; \\
the jaws of the fierce Wyrm gape in the heights. \\
Weden’s son \ken*{= Thunder} will meet the venom \\
of the Warg, after the deaths of Wider’s kinsmen \ken*{= the Eese}.\evb\evg


\bvg\bva[H4]\mssnote{\Hauksbok~21r/14}%
\Ballnote{This half-stanza appears between 61 and 62; it is generally held to be a late Christian insert.}%
Þȧ kømr hinn \alst{r}íki \hld\ at \alst{r}ęgin-dȯmi &
\alst{ǫ}flugr \alst{o}fan \hld\ sá’s \alst{ǫ}llu rę́ðr.\eva

\bvb Then comes the mighty one to the great power, \\
strong from above, he who rules everything.\evb\evg

\sectionline
