\bookStart{Fight at Finnsbury}
\setBookCode{Finnsburg}

\begin{flushright}%
\textbf{Dating:} TODO

\textbf{Meter:} \Fornyrdislag%para
\end{flushright}%

\section{Introduction}

Along with \textlink{Waldere} and \Beowulf, the \textbf{Fight at Finnsbury} (\Finnsburg) is one of three Anglo-Saxon narrative epics dealing with heroic legend.

\subsection{Preservation}

\Finnsburg\ only survives in the printed edition of \textcite[\nopp vol. 1, pp. 192--3]{Hickes1703}, based on the original manuscript which was most likely destroyed in the Cottonian fire of 1731.

Hickes introduces the poem with the following words (original emphasis kept):

\begin{quote}
  \emph{Eodem metro conditum forte reperi fragmenti poetici singulare folium, in codice \textsc{ms.} homiliarum} Semi-Saxonicarum \emph{qui extat in Bibliotheca} Lambethana. \emph{Fragmentum autem subsequitur.}

  ‘By chance, I found a single leaf of a poetic fragment composed in the same meter, in a manuscript codex of Semi-Saxon homilies that is housed in the Lambeth Library. The fragment itself follows.’
\end{quote}

Hickes’ edition seems generally reliable, and the singular folio appears to have been in good shape.  The majority of his errors are transparent spacing errors, e.g. writing two words as one (3 \emph{‘herdraca’} for \emph{*her draca}) or splitting one word in two (9 \emph{‘þis ne’} for \emph{*þisne}, 11 \emph{‘Hie geaþ’} for \emph{*Hicgeaþ}).  Others are plain copying errors, e.g. reading \emph{*a} as \emph{u} in l.~3 \emph{‘eastun’} (for \emph{*eastan}) or \emph{*þ} as the very similar \emph{ꞃ} in l.~2 \emph{‘hearo’} (for \emph{*heaþo}).

\subsection{Contents}

TODO.  A similar episode appears in \Beowulf.

Prior to the fragment’s commencing the poem would have seen Naf and his warriors encamped and asleep in the Finnsbury hall late at night.  One of the men, perhaps a lookout, sees the blinking of swords and chainmail from the east and loudly asks what it may be; is it the dawn, or a dragon, or (and here the fragment picks up) are the gables of the hall burning (l. 1)?  King Naf responds in parallel: it is not the dawn, nor a dragon, nor do the gables burn---rather it is a hostile army (ll. 2--9).  Naf makes a speech to rouse his warriors (ll. 10--12).

TODO

\newpage

\section{Text}

\bvg\bva[]%
„{[...]} \edtext{\emph{hor}nas}{\Afootnote{reconstructed based on l.~4; \emph{‘*nas’} Hickes}} byrnað nǽfre.“\eva

\bvb “{[...]} gables never burn.”\evb\evg


\bvg\bva[][2]%
\alst{H}leoðrode þá \hld\ \edtext{\alst{h}eaðo-geong}{\Afootnote{\emph{‘hearo geong’} (for \emph{*heaþo geong}) Hickes}} cyning: &
„Né þis ne \alst{d}agað \edtext{éast\emph{a}n}{\Afootnote{\emph{‘Eastun’} Hickes}}, \hld\ né \edtext{hér \alst{d}raca}{\Afootnote{\emph{‘herdraca’} Hickes}} ne fléogeð, &
né \alst{h}ér þisse \alst{h}ealle \hld\ \alst{h}ornas ne byrnað, &
ac hér \edtext{\alst{f}orð berað}{\Afootnote{\emph{‘forþberað’} Hickes}} \hld\ \alst{f}ugelas singað, &
\alst{g}ylleð \alst{g}rǽg-hama, \hld\ \alst{g}u̇ð-wudu hlynneð, &
\alst{sc}yld \alst{sc}ęfte on·cwyð. \hld\ Nú \edtext{\alst{sc}ïneð}{\Afootnote{\emph{scýneð} Hickes}} þes mȯna &
\alst{w}aðol under \alst{w}olcnum; \hld\ nú á·rísað \alst{w}éa-dǽda, &
þe \edtext{þisne}{\Afootnote{\emph{‘ðis ne’} Hickes}} \alst{f}olces níð \hld\ \alst{f}ręmman willað. &
\edtext{Ac on·\alst{w}acnigeað nú, \hld\ \alst{w}ígend mïne, &
\alst{h}ębbað éowre \edtext{\emph{\alst{h}}anda}{\Afootnote{\emph{landa} Hickes}}, \hld\ \edtext{\alst{h}i\emph{c}geað}{\Afootnote{\emph{‘Hie geaþ’} Hickes}} on ęllen, &
\edtext{\alst{w}in\emph{n}að}{\Afootnote{\emph{‘Windað’} Hickes}} on orde, \hld\ \alst{w}esað on móde!“}{\lemma{Ac \dots\ móde! ‘Yet \dots\ spirits!’}\Bfootnote{It has often been noted that Naf’s rousing of the sleeping warriors bears a strong resemblance to Beader’s speech in \emph{Bjarkamál in fornu} 1, ll.~3--4 (Anon \emph{Bjark} in \Skp\ 3): \emph{Vaki ę́ ok vaki, \hld\ vina haufuð, // allir inir ǿðstu \hld\ Aðils of sinnar.} ‘Wake now and wake, ye chief of friends; all the noblest companions of Edyils!’}}\eva

\bvb Then bellowed out the battle-young king: \\
“This is no dawning from the east, nor does a dragon fly hither, \\
nor here on this hall do the gables burn--- \\
but here they bear forth arms, the birds sing, \\
the grey-pelt \ken{wolf} howls, the war-wood \ken{shield} dins, \\
the shield clashes against the spear.  Now shines the moon, \\
wandering behind the clouds; now arise the woeful deeds \\
which wish to fulfill the evil of this host. \\
Yet awaken now, ye my warriors! \\
Raise your hands, put your minds to zeal, \\
win by the spear-point, be in brave spirits!”\evb\evg


\bvg\bva[][13]%
Þá á·râs mǫnig \alst{g}old-hladen þegn, \hld\ \edtrans{\alst{g}yrde hine his swurde}{girded on his sword}{\Bfootnote{Cf.~\textlink{Hildebrandslied} 5b.}}; &
þá tó \alst{d}ura éodon \hld\ \alst{d}riht⸗líce cęmpan, &
\alst{S}ige·ferð and Éaha, \hld\ hyra \alst{s}weord ge·tugon, &
and æt \alst{ȯ}ðrum durum \hld\ \alst{O}rd·lâf and Gu̇ð·lâf, &
and \alst{H}ęngest sylf; \hld\ \alst{h}wearf him on lâste.\eva

\bvb Then arose many a gold-laden thane, girded on his sword. \\
Then to the door went the lordly champions \\
Syeferth and Eah---they drew their swords--- \\
and at the other door Ordlaf and Guthlaf \\
and Hengest himself; he followed in their tracks.\evb\evg


\bvg\bva[][18]%
Þá gít \alst{G}âr·ulf \hld\ \alst{G}u̇ð·ere styrode, &
þæt híe swá \alst{f}réo⸗líc \alst{f}eorh \hld\ \edtext{\alst{f}orman}{\Afootnote{\emph{‘for man’} Hickes}} sïðe &
tó þǽre \alst{h}ealle durum \hld\ \alst{h}yrsta ne bǽron, &
nú hyt \alst{n}íða heard \hld\ á·\alst{n}yman wolde: &
ac hé frægn ofer \alst{ea}l \hld\ \alst{u}n·dearninga, &
\alst{d}éor-mód hæleð, \hld\ hwá þá \alst{d}uru héolde.\eva

\bvb TODO: Translation\evb\evg


\bvg\bva[][24]%
„\alst{S}ige·ferð is mïn nama“, \hld[cwæð hé] „ic eom \alst{S}ecgena léod, &
\alst{w}ręcca \alst{w}íde cu̇ð. \hld\ Fela ic \alst{w}éana ge·bâd, &
\alst{h}eardra \alst{h}ilda; \hld\ þé is gýt \alst{h}ér witod, &
\alst{s}wæðer þú \alst{s}ylf tó mé \hld\ \alst{s}ǿcean wylle.“\eva

\bvb TODO: Translation\evb\evg


\bvg\bva[][28]%
Þá wæs on \alst{w}ealle \hld\ \alst{w}æl-slihta ge·hlyn, &
sceolde \alst{c}elæs bord \hld\ \alst{c}ǿnum on handa &
\alst{b}ân-helm \alst{b}erstan. \hld\ \alst{B}urụh-þelu dynede, &
oð þæt æt þǽre \alst{g}u̇ðe \hld\ \alst{G}âr·ulf ge·crang, &
\alst{ea}lra \alst{ǽ}rest \hld\ \alst{eo}rð-búendra, &
\alst{G}u̇ð·lâfes sunu; \hld\ ymbe hine \alst{g}ódra fela &
\alst{h}wearf-latra \alst{h}rǽw. \hld\ \alst{H}ræfẹn wandrode &
\alst{s}weart and \alst{s}ealo-bru̇n; \hld\ \alst{s}wurd-léoma stód &
swylce eal \alst{F}innes-burụh \hld\ \alst{f}ýrenu wǽre.\eva

\bvb TODO: Translation \\
Until in that battle Garulf perished \\
earliest of all earth-dwellers \ken{men}, \\
the son of Guthlaf.  Around him a great many of good men, \\
the corpses of those late to flight.  A raven wandered, \\
black and dusky brown.  The sword-flash stood \\
as if all Finnsbury were a burning beacon.\evb\evg


\bvg\bva[][37]%
Ne ge·frægn ic nǽfre \alst{w}urð⸗lícor \hld\ æt \alst{w}era hilde &
\alst{s}ixtig \alst{s}ige-beorna \hld\ \alst{s}ǿl ge·bǽran, &
ne nǽfre \alst{s}wânas \alst{s}wǿtne \hld\ medo \alst{s}ǿl for·gyldan, &
þonne \alst{H}næfe guldon \hld\ his \alst{h}æg-stealdas. &
Hig \alst{f}uhton \alst{f}ïf dagas, \hld\ swá hyra nân ne \alst{f}éol &
\alst{d}riht-ge·sïða, \hld\ ac hig þá \alst{d}uru héoldon.\eva

\bvb I have not learned that in a worthier manner in the fight of men \\
ever sixty triumph-warriors better carried themselves \\
nor ever lads better repaid sweet mead \\
than Naf was repaid by his young warriors. \\
They fought for five days while not one fell \\
among that troop-fellowship, yet they held the doors.\evb\evg


\bvg\bva[][43]%
Þá ge·\alst{w}ât him \alst{w}und hæleð \hld\ on \alst{w}æg gangan, &
sǽde þæt his \alst{b}yrne \hld\ á·\alst{b}rocen wǽre, &
\alst{h}ęre-sceorp un·\alst{h}rór, \hld\ and éac wæs his \alst{h}elm þyrl. &
Þá hine sȯna \alst{f}rægn \hld\ \alst{f}olces hyrde, &
hú þá \alst{w}ígend \hld\ hyra \alst{w}unda ge·nǽson &
oððe hwæðer þǽra hyssa\eva

\bvb Then went a wounded hero, walking away; \\
he said that his byrnie was broken up, \\
his war-gear powerless, and his helmet, too, was pierced. \\
Then he was soon asked by the herdsman of the troop \ken{leader} \\
how the warriors were treating their wounds, \\
or ... of the boys ...\evb\evg

\sectionline
