\bookStart{Lay of Hildbrand}[Hildebrandslied]
\setBookCode{Hildebrandslied}

\begin{flushright}%
\textbf{Dating:} C8th

\textbf{Meter:} \Fornyrdislag%para
\end{flushright}%

\section{Introduction}

The \textbf{Lay of Hildbrand} (abbrev. \Hildebrandslied) is the only surviving High German alliterative poem dealing with native legendary material and is such of immense value.

\subsection{Preservation}

\Hildebrandslied\ is preserved in the 2° Ms. theol. 54 of the Kassel university library (ca.~820 \ce, henceforth labelled \HildMS) where it is written in a different hand (ca.~830 \ce) on the outer covers, foll.~1r and 76v.  76v ends in the middle of a line and the poem is thus fragmentary.

\subsection{Language}

The language of \Hildebrandslied\ is one of the great mysteries of German philology.  It is a strange mix of High German and Old Saxon.

Let us begin with an overview of features.  For transparency, the quoted forms will be in their manuscript orthography and not normalized.

TODO: list is incomplete!

The following features are clearly Old Saxon:

\begin{enumerate}

\item The Ingvaeonic nasal spirant law \textipa{/\*VnT/} > \textipa{/\~{\*V}:T/} (\emph{guð-} 5, \emph{ódre} 12, \emph{chud} 13, 28, \emph{usere} 15, \emph{gudea} 59)
\item Masc.~nom.~pl.~\emph{-os} (6 \emph{helidos})
\item Loss of \emph{*-ʀ} after a vowel in monosyllabic words (\emph{mi} 12a, 12b, 15, 41; but cf.~TODO below)
\item Unshifted \emph{-k} in \emph{ik} ‘I’ (cf.~below)

\end{enumerate}

The following features are clearly High German:

\begin{enumerate}

\item \emph{*wr-} > \emph{r-}, as required by the alliteration between \emph{reccheo} and \emph{riche} (l.~47 n.)
\item Preserved \emph{*-ʀ} (> \emph{-r}) after a vowel in monosyllabic words.
\item Neuter adj.~nom./acc.~sg. \emph{-at} (\emph{suasat} 52).
\item \emph{*þ} > \emph{d} (see below).
\item The High German consonant shift (see below).

\end{enumerate}

\subsection{\emph{*þ} > \emph{d} and the High German consonant shift}

These two features affect such a great number of words that they ought to be treated separately.

\subsubsection{\emph{*þ} > \emph{d}}

\textipa{/T/} (\emph{ð}: \emph{ðat} 1, 2, \emph{haðubrant} 3, \emph{guð-} 5; \emph{th}: \emph{theotrihhe} 19) > \textipa{/d/} (\emph{hęlidos} 6, \emph{do} 6, 32, 62, 64, \emph{dero} 6, 60, \emph{eddo} 11, 53, \emph{du} 11, 12, 31, 38, 40, 47, 47, 54, 56, \emph{de} 12, 59, \emph{ódre} 12, \emph{chud} 13, 28, \emph{irmindeot} 13, %
\emph{hadubrant} 14, 17, 35, \emph{dat} 15, 17, 24, 31, 34, 41, 42, 46, 47, 63, \emph{dea} 16, \emph{nid} 18, \emph{degano} 19, 26, \emph{des} 23, \emph{sid} 23, \emph{detrihhe} 23, \emph{darba} 23, 26*, \emph{dechisto} 26, \emph{deotrichhe} 26, \emph{quad} 30, 48, 57, \emph{dana} 31, \emph{dinc} 31, \emph{der} 33, 57, 58, \emph{dir} 34, 38, 54, 58, %
\emph{huldi} 34, \emph{ƿidar} 37, \emph{dinem} 39, 45, \emph{dinu} 39, \emph{seolidante} 41, \emph{desemo} 47, \emph{dar} 50, 56, \emph{ƿerdan} 53, \emph{doh} 54, 57, \emph{aodlihho} 54, \emph{din} 54, \emph{dih} 58, \emph{gudea} 59, \emph{ƿerdar} 60, \emph{erdo} 61, \emph{desero} 61, %
\emph{bedero} 61, \emph{dem} 63). %

\subsubsection{The High German consonant shift}

The High German consonant shift consists of two related changes, whose distribution however varies across German dialects.

\begin{enumerate}
  \item The spirantization of voiceless stops \emph{p, t, k} > \emph{f, z, x} or \emph{pf, tz, kx}, depending on their position in the word;
  \item The devoicing of voiced stops \emph{b, d, g} > \emph{p, t, k}.
\end{enumerate}

The following changes are attested in \Hildebrandslied:

\begin{enumerate}
  \item Spirantization: \begin{enumerate}%
  \item \emph{k} (\emph{ik} 1, 12, \emph{cnuosles} 11) > \emph{x} (\emph{ch}: \emph{folche} 10, \emph{chind} 13, 52, \emph{\textbf{ch}unincriche} 13, \emph{chud} 13, 28, \emph{otachres} 18, \emph{otachre} 25, \emph{dechisto} 26, \emph{folches} 27, \emph{chonnem} 28, \emph{cheisuringu} 33, %
  \emph{chuning} 33, \emph{chludun} 64; %
  word-finally \emph{h}: \emph{sih} 2, 5, 60, \emph{ih} 17, 29, 34, 45, 49, 53, \emph{mih} 39, 50, 52, \emph{dih} 58); %
  \emph{kk} > \emph{kkx} (\emph{reccheo} 47);
  \emph{k} after long vowel > \emph{xx} (\emph{hh}: \emph{ƿelihhes} 11, \emph{theotrihhe} 19, \emph{detrihhe} 23, \emph{aodlihho} 54; \emph{chh}: \emph{deotrichhe} 26; \emph{cc}: \emph{harmlicco} 65; \emph{ch}: \emph{chunincri\textbf{ch}e} 13, \emph{riche} 47). %
  \end{enumerate}%

  \item Devoicing: \begin{enumerate}%
  \item \emph{b} (\emph{hiltibrant} 3, 7, 17, 30, 43, 44, 48, 57, \emph{hadubrant} 3, 14, 17, 35, \emph{ubar} 6, 42, \emph{heribrantes} 7, 43, 44, \emph{ibu} 12, 54, 56, \emph{hiltibrantes} 14, 35, \emph{bure} 21, \emph{barn} 21, \emph{arbeo-} 22, \emph{darba} 23, 26, \emph{lib} 29, \emph{habbe} 29, \emph{obana} 30, \emph{ab} 30, %
  \emph{bauga} 32, \emph{bi} 34, 47, \emph{gibu} 34, \emph{geba} 36, \emph{bist} 38, \emph{habes} 46, 56, \emph{burc} 51, \emph{banun} 51, \emph{breton} 53, \emph{billiu} 53, \emph{banin} 53, \emph{rauba} 56, \emph{bihrahanen} 56, \emph{brunnono} 61, \emph{bedero} 61, %
  \emph{-bort} 64, \emph{ƿabnum} 67) > \emph{p} (\emph{prut} 21, \emph{leop} 27, \emph{gap} 33, \emph{pist} 40); %
  \emph{bb} > \emph{pp} (\emph{pp}: \emph{sippan} 31),

  \item \emph{d} (only in \emph{chind} 13, 52) > \emph{t} (\emph{t}: \emph{gihorta} 1, \emph{hiltibrant} 3, 7, 17, 30, 43, 44, 48, 57, \emph{enti} 3, 19, 49, \emph{hadubrant} 3, 14, 17, 35, \emph{untar} 3, \emph{sunufatarungo} 4, \emph{garutun} 5, \emph{gurtun} 5, \emph{suert} 5, \emph{hiltiu} 6, \emph{ritun} 6, \emph{gimahalta} 7, 14, 35, 44, %
  \emph{heribrantes} 7, 43, 44, \emph{frotoro} 8, \emph{gistuon\textbf{t}} 8, \emph{uuortum} 9, \emph{fater} 9, 17, \emph{irmindeot} 13, \emph{hiltibrantes} 14, 35, \emph{sagetun} 15, 41, \emph{liuti} 15, \emph{alte} 16, \emph{anti} 16, \emph{frote} 16, \emph{otachre} 18, 25, \emph{miti} 19, 53, 67, %
  \emph{theo\textbf{t}rihhe} 19, \emph{lante} 20, 49, \emph{prut} 21, \emph{raet} 22, \emph{detrihhe} 23, \emph{gistuon\textbf{t}um} 23, \emph{fatereres} 24, \emph{friuntlaos} 24, \emph{unti} 26, \emph{deotrichhe} 26, \emph{giston\textbf{t}un} 26, \emph{ente} 27, \emph{irmingot} 30, %
  \emph{halt} 31, \emph{mit} 31, 36, 39, \emph{gileitos} 31, \emph{ƿant} 32, \emph{ƿuntane} 32, \emph{gitan} 33, \emph{\textbf{t}ruhtin} 34, \emph{ort} 37, \emph{orte} 37, \emph{alter} 38, \emph{ƿuortun} 39, \emph{gialtet} 40, \emph{inƿit} 40, \emph{fortos} 40, \emph{seolidante} 41, %
  \emph{ƿentilseo} 42, \emph{tot} 43, \emph{goten} 46, \emph{ƿurti} 47, \emph{ƿaltant} 48, \emph{got} 48, \emph{ƿeƿurt} 48, \emph{skihit} 48, \emph{ƿallota} 49, \emph{scerita} 50, \emph{sceotan\textbf{t}ero} 50, \emph{suertu} 52, \emph{breton} 53 (n.), \emph{taoc} 54, \emph{liuto} 57, %
  \emph{lusti\textbf{t}} 58, \emph{hiutu} 60, \emph{uualtan} 61, \emph{scritan} 62, \emph{sciltim} 63, \emph{ston\textbf{t}} 63, \emph{bort} 64, \emph{scilti} 65, \emph{lintun} 66, \emph{ƿurtun} 66), %

  \item \emph{g} (\emph{gihorta} 1, \emph{seggen} 1, \emph{sunufatarungo} 4, \emph{garutun} 5, \emph{gudhamun} 5, \emph{gurtun} 5, \emph{ringa} 6, \emph{gimahalta} 7, 14, 35, 44, \emph{fragen} 8, \emph{gistuont} 8, \emph{sages} 12, \emph{sagetun} 15, 41, \emph{gihueit} 18, \emph{degano} 19, 26, %
  \emph{gistuontum} 23, \emph{gistontun} 26, \emph{irmingot} 30, \emph{gileitos} 31, \emph{bauga} 32, \emph{cheisuringu} 33, \emph{gitan} 33, \emph{chuning} 33, \emph{gap} 33, \emph{gibu} 34, \emph{geru} 36, \emph{geba} 36, \emph{gialtet} 40, \emph{gisihu} 45, \emph{goten} 46, \emph{ƿelaga} 48, %
  \emph{got} 48, \emph{enigeru} 51, \emph{gifasta} 51, \emph{giƿinnan} 55, \emph{argosto} 57, \emph{ƿiges} 58, \emph{gudea} 59, \emph{gimeinun} 59, \emph{hregilo} 60, \emph{giƿigan} 67) %
  > \emph{k} (only at the end of words: \emph{chuni\textbf{c}riche} 13, \emph{dinc} 31, \emph{ƿic} 42, \emph{sehstic} 49, \emph{burc} 51, \emph{taoc} 54, \emph{enic} 56).%
  \end{enumerate}

\end{enumerate}

\subsubsection{Orthography}

For the orthography of the edition I have found it impossible to produce a normalized text without too heavily distorting the received text, being as it is, a blend of several dialects.  One need only observe the treatment of the name Thedric, which appears thrice, and each time in a markedly different form.  Apart from the typical practices as described in the General Introduction, and using acute accents to signify long vowels, circumflex accents to signify now-monophthongised original diphthongs, and overdots to mark nasal vowels, I have carried out the following changes in order to clarify etymological relationships and make the text somewhat less unwieldy. Of these changes, 7--9 have also been noted in the apparatus where they occur:
\begin{enumerate}
  \item Replaced both \emph{ƿ} (wynn) and \emph{uu} with \emph{w}.
  \item Replaced \emph{c} with \emph{k}.
  \item Replaced \emph{qu} with \emph{kw}.
  \item Replaced \emph{t} with \emph{t̨} where corresponding to OHG \emph{z}.
  \item Replaced \emph{th} with \emph{þ}.
  \item Replaced \emph{e} with \emph{ę} when reflecting an original a-vowel affected by \emph{i}-mutation.
  \item Replaced unetymological double \emph{nn} with \emph{n}.
  \item Restored initial \emph{h-} where etymological and/or metrically required.
  \item Removed initial \emph{h-} where unetymological and/or metrically deficient.
\end{enumerate}

The punctuation of the original, entirely consisting of interpuncts, at times representing metrical breaks, at others sporadically placed, has not been retained.

Where it appears in the cæsura, the extrametrical interjection \emph{kwad Hilti·brant} ‘quoth Hildbrand’ (found in ll. 30, 49, and 58) replaces the usual interpunct, to indicate that the pause of the cæsura has been filled with an indication of the speaker.  Outside of \textlink{Hildebrandslied}, similar interjections are found throughout early Germanic poetry: in Old Norse (e.g. \textlink{Reginsmal}[3]/1, Anon \emph{Eirm} 1/1 (in \Skp\ 1)), Old Saxon (e.g. \textlink{Heliand}[3][226], \textlink{SaxonGenesis} 1), and Old English (e.g. \textlink{Finnsburg} 24).  The chronological distribution of these interjections is such that they cannot be mere scribal additions (Old Norse poetry was first written down in the C12th, several centuries after the alliterative meter had gone extinct in Germany); instead, they appear to be genuine remnants of oral performance going back to the common Germanic tradition.

\subsection{Summary}

The poet begins with a short formulaic introduction; he is relating older stories (1--2).  The two duellists, Hildbrand and Hathbrand, father and son, arm themselves and ride into battle at the head of two opposing armies (3--6). They speak, and Hildbrand asks Hathbrand for his name and lineage (7--13). Hathbrand gives his name and ancestry; his father was the warrior Hildbrand, who abandoned him as a newborn. This was long ago, and Hathbrand does not think him still alive (14--29). Hearing this, Hildbrand calls on God as witness, and offers his son a golden torc as a token of loyalty (30--34). Hathbrand takes this as an insulting tricks. He proclaims that wealth should be won by struggle alone and accuses Hildbrand of having grown old through treachery (35--40); he has heard from sailors on the Mediterranean that his father is dead (41--43).

After this straight-forward narrative sequence three short speeches follow, in the ms. all spoken by Hildbrand. The second is certainly spoken by Hildbrand, but the other two may be misplaced or misattributed.

1. Hildbrand reflects on his son’s prosperity: from his clothes he can tell that he has a good lord, and that he, unlike himself, has not suffered the fate of exile (44--47).

2. Hildbrand calls on God, and laments that, after thirty years at war, he is now forced to fight against his own son. Still, Hathbrand should easily be able to kill such an old man as Hildbrand, if he has strength and fate on his side (48--56).

3. Hildbrand (or Hathbrand, and there is a case for emending here) says that only the most cowardly easterner could refuse the fight so greatly desired. Let both men fight their hardest, and when the duel is over the winner will strip the armour of the other (57--61).

The two men then throw their javelins into each other’s shields and rush at each other, hacking away at their shields until they become worthless (62--67).  Here the poem ends abruptly in the middle of the battle.

\sectionline

\newpage

\section{Text}

\bvg\bva[]\mssnote{\HildMS~1r/1--5}%%
Ik gi·hôrta dat̨ sęggen &
dat̨ sih \alst{u}r·hêt̨t̨un \hld\ \edtrans{\alst{ae}non}{in single combat}{\Bfootnote{A dative plural adverbial, lit. ‘alone, singly’.}} muot̨ín: &
\alst{H}ilti·brant ęnti \alst{H}adu·brant \hld\ \edtrans{untar \alst{h}ęrjun t̨wêm}{under two hosts}{\Bfootnote{Either man was a champion of his army.}}; &
\alst{s}unu-fatar·ungo \hld\ iro \alst{s}aro rihtun &
\alst{g}arutun sé iro \edtrans{\alst{g}u̇d-hamun}{warcloth}{\Bfootnote{Their chainmail.  Cf.~l. 61.}} \hld\ \edtrans{\alst{g}urtun sih iro swert ana}{girded on their swords}{\Bfootnote{Cf.~\textlink{Finnsburg} 13b.}} &
\alst{h}ęlidos ubar \edtext{\emph{\alst{h}}ringa}{\Afootnote{\emph{ringa} \HildMS}} \hld\ dó sie t̨ó dero \alst{h}iltu ritun.\eva

\bvb {\huge I} \textsc{have heard it said} \\
that two champions met in single combat: \\
Hildbrand and Hathbrand under two hosts. \\
The son and father set their armour, \\
readied their warcloth, girded on their swords, \\
the heroes, over the chain-links, when to that fray they rode.\evb\evg


\bvg\bva[][7]\mssnote{\HildMS~1r/5--11}%
\alst{H}ilti·brant \edtext{gi·mahalta}{\Afootnote{\emph{heribrantes sunu} ‘Harbrand’s son’ add. \HildMS}} \hld\ ---her was \alst{h}êróro man &
\edtrans{\alst{f}erạhes \alst{f}rótóro}{more learned of life}{\Bfootnote{Possibly formulaic; cf.~\emph{Maldon} 317a: \emph{Ic eom fród feores.} ‘I am learned of life’.}}--- \hld\ her \alst{f}rágén gi·stuont &
\alst{f}ôhém wortum \hld\ \edtext{\emph{h}wer}{\Afootnote{\emph{wer} \HildMS}} sín \alst{f}ater wári &
\edtext{\alst{f}ir\emph{ih}o}{\Afootnote{\emph{fireo} \HildMS}\lemma{fir\emph{ih}o ‘men’}\Bfootnote{Gen.~of \emph{fir(a)ha}, a \emph{plurale tantum} poetic synonym for ‘men, people’ found throughout the Old Germanic languages.  In OHG it is only otherwise found in \textlink{Wessobrunn} 1 and \textlink{Muspilli} 55, never in prose \parenciteshorthand[s.v. \textit{fir(a)ha}]{EWA}, although it is also found in the cpd.~\emph{smala-firihi} m.~pl.~‘the common people’ \parenciteshorthand[s.v. \textit{ferah}]{EWA}.

The word stems from PGmc.~\emph{firhwijōz}, a \emph{ja}-stem derivative of PGmc.~\emph{*ferhwą} ‘life, lifeblood’.  It is cognate to OS \emph{firihos}, OE \emph{fíras} ‘men’, and ON \emph{firar, fyrvar, fjǫrvar, fyrðar}.  The fact that the aforementioned cognates are also exclusively poetic suggests that this was likely the case already in PGmc., or at the very least in PNWGmc.}} in \alst{f}olkhe \hld\ {[...]} &
{[...]} \hld\ „eddo \edtext{hwe-líhhes}{\Afootnote{\emph{welihhes} \HildMS}} \alst{k}nuosles dú sís &
ibu dú mí \alst{ê}nan sagés \hld\ ik mí de \alst{ȯ}dre wêt &
\edtrans{\alst{kh}ind}{my child}{\Bfootnote{Here Hildbrand is addressing Hathbrand as his junior (cf.~modern English son), but the word also subtly foreshadows his paternity even though at this point in the narrative he is not yet aware of it.}} in \edtext{\alst{kh}un*ink-ríkhe}{\Afootnote{\emph{chunnincriche} \HildMS}} \hld\ \alst{kh}u̇d ist \edtext{mí*}{\Afootnote{\emph{mín} \HildMS}} al irmin-deot.“\eva

\bvb Hildbrand spoke---he was the hoarier man, \\
more learned of life.  He began to ask \\
in few words who his father might be \\
of men in the troop, [...] \\
{[...]} “or of which lineage thou be. \\
If thou tell me one I will recall the others, \\
my child, in the kingdom I know all great men.”\evb\evg


\bvg\bva[][14]\mssnote{\HildMS~1r/11--16}%
\alst{H}adu·brant gi·mahalta \hld\ \alst{H}ilti·brantes sunu: &
„\edtrans{Dat̨ sagetun mí \hld\ u̇sere liuti}{This our liegemen said to me}{\Bfootnote{The scansion of this line is inscrutable (cf.~l. 42), but the required alliteration is missing.}} &
\alst{a}lte anti fróte \hld\ dea \alst{ê}r hina wárun &
dat̨ \alst{H}ilti·brant hǽt̨t̨i mín fater \hld\ ih hęit̨t̨u \alst{H}adu·brant. &
Forn her \alst{ô}star \edtext{gi·*węit̨}{\Afootnote{\emph{gihueit} \HildMS}}; \hld\ flôh her \alst{Ô}t·akhres níd &
hina miti \edtext{\emph{\alst{D}}eot·ríhhe}{\Afootnote{\emph{theotrihhe} with pre-shift consonant \HildMS}} \hld\ ęnti sínero \alst{d}egano filu.\eva

\bvb Hathbrand spoke, Hildbrand’s son: \\
“This our liegemen said to me--- \\
the old and learned who lived long ago--- \\
that Hildbrand was the name of my father; my name is Hathbrand. \\
Long ago he turned east; he fled Edwaker’s hate \\
thither with Thedric and his multitude of thanes.\evb\evg


\bvg\bva[][20]\mssnote{\HildMS~1r/16--20}%
Her fur·\alst{l}ǽt̨ in \alst{l}ante \hld\ \edtrans{\alst{l}út̨t̨ila}{a little wife}{\Bfootnote{The word is the fem.~acc.~sg.~of \emph{luzzil} ‘little, small’, but ‘wife’ is supplied in the translation to resolve the ambiguity between the bride and the child.}} sit̨t̨en &
\edtext{\emph{\alst{b}}rút}{\Afootnote{\emph{prut} \HildMS}} in \alst{b}úre, \hld\ \alst{b}arn un·wahsan &
\alst{a}rbjo-laosa; \hld\ \edtext{he\emph{r} raet}{\Afootnote{\emph{heraet} \HildMS}} \alst{ô}star hina, &
des síd \alst{D}e\emph{o}t·ríhhe \hld\ \alst{d}arba \edtext{gi·stuontu\emph{n}}{\Afootnote{\emph{gistuontum} \HildMS}} &
\edtext{\alst{f}ater*es}{\Afootnote{\emph{fatereres} \HildMS}} mínes: \hld\ dat̨ was só \alst{f}riunt-laos man.\eva

\bvb He left behind in this land a little wife to stay: \\
a bride in the bower, a child ungrown, \\
inheritance-less.  He rode thither to the east, \\
since at that time Thedric was in great need \\
of my father---that was so friendless a man!\evb\evg


\bvg\bva[][25]\mssnote{\HildMS~1r/20--24}%
Her was \alst{Ô}t·akhre \hld\ \alst{u}m·met̨ t̨irri, &
\alst{d}egano \alst{d}ękhisto \hld\ unti \edtext{\alst{D}eot·ríkhhe*}{\Afootnote{\emph{darba gistontun} add. \HildMS}}; &
her was eo \alst{f}olkhes at̨ ęnte, \hld\ imo was eo \edtext{\emph{\alst{f}}ehẹta}{\Afootnote{\emph{peheta} \HildMS}} t̨i leop; &
\alst{kh}u̇d was her [...] \hld\ \edtext{\alst{kh}ón*ém}{\Afootnote{\emph{chonnem} \HildMS}} mannum: &
ni wániu ih iu líb habbe.“\eva

\bvb He was towards Edwaker utterly hostile; \\
the dearest of thanes under Thedric. \\
He was ever at the head of the troop; he always loved the fight too much; \\
known was he among keen men. \\
I do not expect he could still be alive.”\evb\evg


\bvg\bva[][30]\mssnote{\HildMS~1r/24--76v/4}%
„Wêt̨t̨u \edtrans{\alst{I}rmin got}{Ermen-God}{\Bfootnote{These words may be interpreted in two ways, either as a cpd. \emph{irmin-got} ‘great God’ referring to the Christian God, or as an invocation of the pre-Christian god Ermen (OHG \emph{Irmin}, ON \emph{Jǫrmunr} ‘Weden’), the ancestor of the \emph{Herminones} of Tacitus (\emph{Germania} 2).
The problem with the former interpretation is that such a cpd is unparallelled in all of Germanic, including in related Christian epic poetry like \textlink{Heliand} and \textlink{SaxonGenesis}, whereas something like \emph{waltant got} (l. 48) is very widely parallelled.  For this reason we are most likely looking at a pagan expression, although this does not necessarily imply that the poet himself was a pagan; he may simply have borrowed it from earlier poetry without understanding its original significance.

For the invocation of Ermen in oaths we have Widukind of Corvey’s \emph{History of the Saxons} 1:12 (ca. 968 \ce):
\emph{quia Hirmin vel Hermis Graece Mars dicitur; quo vocabulo ad laudem vel ad vituperationem usque hodie etiam ignorantes utimur} ‘For \emph{Hirmin} or \emph{Hermes} is what they call Mars in Greek, a term which we use even to this day, without knowing its meaning, for praise or for blame.’  This passage seems to imply that the name of this god persisted into the 10th c.~in vulgar expletives or mild oaths, gradually losing its power as knowledge of the pagan religion declined.  Yet in the present circumstance it is certainly not used in such a way, but rather with great power by the warrior Hildbrand.}}“ \hld[kwad Hilti·brant] „\alst{o}bana ab \edtrans{hevane}{heaven}{\Afootnote{\emph{heuane} \HildMS}\Bfootnote{A likely Old Saxon form which merits some discussion on the relation between the synonymous \emph{himil} and \emph{hevan} in West Germanic.  \emph{himil} is found in both OS and OHG, but a cognate of \emph{hevan} is never found in OHG.  Nor is the use of OS \emph{hevan} without note; it is never used in prose, and in poetry (\textlink{Heliand} and \textlink{SaxonGenesis}) its use is heavily stereotyped, being restricted to 5 cpds and 3 formulaic expressions where it is never used only in the gen. sg.  Still, it must have existed in the language seeing as it has left descendants in modern Low German dialects.

In any case this word adds further difficulty to the question of providence; if \textlink{Hildebrandslied} were an originally OHG text (cf.~l. 47 n.) haphazardly “translated” into OS in a scribal context it seems strange that the translator should have replaced the neutral \emph{himil} with the rare, stereotyped \emph{hevan}, but on the other hand the presence of \emph{hevan} in the OHG archetype would be a major anomaly since that word is not attested in any known variety of High German ancient or modern.}} &
dat̨ dú neo \alst{d}ana halt mit sus sippan man \hld\ \alst{d}ink ni gi·lęitós.“ &
\alst{W}ant her dó ar arme \hld\ \edtrans{\alst{w}untane bauga}{twisted bighs}{\Bfootnote{A spiral cord of precious metal, in this case probably gold.  The association between \inx[C]{bighs} (metal armlets) and a warrior’s honour is well attested; see Index.  This encounter is particularly reminiscent of \textlink{Harbardsljod}[42].}} &
\edtrans{\alst{kh}ęisur·ingu gi·tán}{made of Coser’s coin}{\Bfootnote{Coser (< OE \emph{Cáser}), i.e. Caesar.  A cultural memory of the melting down of Roman \emph{solidī} by Germanic mercenaries.}} \hld\ só imo sie der \alst{kh}uning gap, &
\edtrans{\alst{H}unjo truhtin}{lord of the Huns}{\Bfootnote{Almost certainly \inx[P]{Attle}, although he is not mentioned by name in the poem.}}, \hld\ „\edtrans{dat̨ ih dír it̨ nú}{that I now \dots}{\Bfootnote{Grammatically this clause connects to the previous one and is thus part of Hildbrand’s oath to Ermen-God; he swears that he is giving the bigh out of goodwill (\emph{bí huldí}), i.e., without any secret motive.}} bí \alst{h}uldí gibu.“\eva

\bvb “I call as my witness Ermen-God,” quoth Hildbrand, “in heaven above, \\
that thou never hence with such close kin shouldst lead dispute,” \\
---then he unwound from his arm twisted \inx[C]{bigh}[bighs] \\
made of Coser’s coin, which the king had given him, \\
the lord of the Huns---“that I now give thee this out of goodwill.”\evb\evg


\bvg\bva[][35]\mssnote{\HildMS~76v/4--10}%
\alst{H}adu·brant gi·mahalta \hld\ \alst{H}ilti·brantes sunu: &
„\edtrans{Mit \alst{g}êru skal man \hld\ \alst{g}eba in·fȧhan}{By the spear shall man win gifts}{\Bfootnote{The line comes off as proverbial, although it probably isn’t.  It is however notable that the same warrior-ethic was codified by the Indians as part of the \textoi{kṣatra-dʰarmá}, the honour-code of the Warrior-caste (\textoi{kṣatríya}), which explicitly forbade them from taking gifts.
So in \Mahabharata\ 12.192.73, a \textoi{kṣatríya} king refuses a gift from a priest (\textoi{brāhmaṇá}) since “it is the duty prescribed for a \textoi{kṣatríya} that he must fight and protect (people).  Kṣatriya are said to be the givers, then, how can I take (this) from you?” (transl. \cite{Hara1974}, see further there).}} &
\alst{o}rt widar \alst{o}rte! &
Dú bist dir \alst{a}ltér Hun \hld\ \alst{u}m·met̨ spáhér; &
\alst{sp}ęnis mih mit díném wortun: \hld\ wili mih dínu \alst{sp}eru werpan! &
\edtrans{Pist al-só gi·\alst{a}ltét man \hld\ só dú êwín \edtrans{\alst{i}n·wit}{treachery}{\Bfootnote{A hapax in OHG, although the related form \emph{in·witte} is attested in a gloss \parenciteshorthand[s.v. \textit{inwitti}]{EWA}.  The word is cognate with OS \emph{in·wid}, OE \emph{in·wid}, and ON \emph{ï·við-} (in \textlink{Volundarkvida}[28]/4).
It has been argued that \emph{in·witti} is in fact the only High German form and that \Hildebrandslied’s \emph{in·wit} is a product of Saxon scribal influence \parenciteshorthand{EWA}, but it seems unlikely for metrical reasons that the present line should ever have had a trisyllabic form of this word.}} fórtós.}{Indeed, thou hast become an old man by always employing treachery}{\Bfootnote{Lit.~‘thou art indeed an aged man since thou always didst employ treachery.’  Hathbrand accuses Hildbrand of being an imposter and views his old age as suspect; no brave warrior (it is implied) should be so old, so he must have relied on cunning and cowardly blows instead.}} &
Dat̨ \alst{s}agetun mí \hld\ \alst{s}êo-lídante &
\alst{w}estạr ubar \edtrans{\alst{W}ęntil-sêo}{Wendle-sea}{\Bfootnote{The Mediterranean Sea, the name referring to the \emph{Vandali}, who for a time ruled North Africa.}} \hld\ dat̨ \edtext{\emph{in}an}{\Afootnote{emend.; \emph{man} \HildMS}} \alst{w}ík fur·nam: &
tôt ist \alst{H}ilti·brant \hld\ \alst{H}ęri·brantes suno!“\eva

\bvb Hathbrand spoke, Hildbrand’s son: \\
“By the spear shall man win gifts, \\
point against point! \\
Thou art, old Hun, utterly clever; \\
thou temptest me with thy words---at me wilt thou hurl thy spear! \\
Indeed, thou hast become an old man by always employing treachery. \\
This seafarers said to me \\
west o’er the Wendle-sea, that war took him away. \\
Dead is Hildbrand, Harbrand’s son!”\evb\evg


\bvg\bva[][44]\mssnote{\HildMS~76v/11--13}%
\alst{H}ilti·brant gi·mahalta \hld\ \alst{H}ęri·brantes suno: &
„Wela gi·sihu ih in díném hrustim &
dat̨ dú \alst{h}abés \alst{h}ême \hld\ \alst{h}êrron góten, &
dat̨ dú noh bí desemo \alst{r}íkhe \hld\ \edtrans{\alst{r}ękkhjo}{exile}{\Bfootnote{An important word for identifying the poem’s original dialect.  \emph{rękkhjo} (< PWGmc.~\emph{*wrakkjō}) originally began with a \emph{w-} (cf.~OE \emph{wręcca}, OS \emph{wrękkjo}), but its alliteration with \emph{ríkhe} (< PWGmc.~\emph{*ríkija}, cf.~OE \emph{ríce}, OS \emph{ríki}) proves that this sound had been lost by the time of the poem’s composition.  Since the \emph{w-} is retained in OS (cf.~\textlink{Heliand} 631, 671) and to this day in Low German, \Hildebrandslied\ must have been composed in the High German dialect area.}} ni wurti.\eva

\bvb Hildbrand spoke, Harbrand’s son: \\
“Well do I see upon thy armour, \\
that thou hast at home a good lord; \\
that thou yet in this realm hast not become an exile.\evb\evg


\bvg\bva[][48]\mssnote{\HildMS~76v/13--22}%
\alst{W}elaga nú \edtrans{\alst{w}altant got}{Wielding God!}{\Bfootnote{Cf.~OE \emph{wealdend god}, OS \emph{waldand god}.  Apparently a common West Germanic poetic expression.}},“ \hld[kwad Hilti·brant] „\edtrans{\alst{w}ê-wurt}{woeful Weird}{\Bfootnote{The sense of \emph{wurt} ‘weird’ bears explaining.  Rather than the Norse norn Weird (\textlink{Voluspa}[19]), its meaning here ought to be understood as ‘inexorable course of events decreed by fate’, although this sense is still tied to the pagan norn and ultimately of pre-Christian origin.  The same sense underlies OS \emph{wurd(i)} at various points in \textlink{Heliand} (e.g.~l.~4619), OE \emph{wyrd} (e.g. \Beowulf\ 572b, quoted in the note to l. 54 below), and ON \emph{grimmar urðir} ‘grim weirds’ in \textlink{Sigurdskamma}[5]/4.}} skihit. &
Ih wallóta \edtrans{\alst{s}umaro ęnti wintro \hld\ \alst{s}ehs-tik}{sixty summers and winters}{\Bfootnote{Thirty years.  Cf.~\Beowulf\ 1498, 1769: \emph{hund misséra} ‘a hundred half-years’, i.e. ‘fifty years’.  Hathbrand is then thirty years old while Hildbrand is probably in his late fourties or fifties.}} ur lante &
dár man mih eo \alst{sk}ęrita \hld\ in folk \edtrans{\alst{sk}eot̨antero}{shooters}{\Bfootnote{Cf.~\Beowulf\ 702, where the OE cognate \emph{sceótend} stands for “warriors” in general.}}, &
só man mir at̨ \alst{b}urk ênigeru \hld\ \alst{b}anun ni gi·fasta. &
Nú skal mih \alst{s}wásat̨ khind \hld\ \alst{s}wertu hauwan &
\edtrans{\alst{b}retón}{beat}{\Bfootnote{A hapax.  TODO: cite \parenciteshorthand[s.v.\textit{bretôn}]{EWA}}} mit sínu \edtrans{\alst{b}illju}{blade}{\Bfootnote{Instr.~sg.~of \emph{billi}, a hapax in OHG whose exact sense is unclear \parenciteshorthand[s.v. \textit{billi}]{EWA}.  It is cognate with OE \emph{bil} ‘a type of weapon’ which is exclusively used in poetry.  The rarity of this word in the old languages suggests that it is probably simply being used as a poetic synonym for ‘sword’ or ‘sharp weapon’, whence the present translation ‘blade’.}} \hld\ eddo ih imo t̨i \alst{b}anin werdan. &
Doh maht dú nú \alst{ao}d-líhho \hld\ \edtrans{ibu dir dín \alst{ę}llen taok}{if thy zeal avails thee}{\Bfootnote{Formulaic.  Cf.~\Beowulf\ 572b--573: \emph{Wyrd oft nęrið //
un·fǽgnǽ eorl \hld\ þǫnnǽ his ęllin déah.} ‘Weird often saves the un-\inx[C]{fey} \inx[C]{earl} when his zeal avails.’}} &
in sús \alst{h}êremo man \hld\ \alst{h}rusti gi·winnan &
\alst{r}auba \edtext{bi·*\alst{r}ahanen}{\Afootnote{\emph{bihrahanen} \HildMS}} \hld\ \edtrans{ibu dú dár êníg \alst{r}eht habés!}{if thou hast any right thereto!}{\Bfootnote{The last three lines of Hildbrand’s speech bear a remarkable similarity to \textlink{Waldere}[B]/14 ff.  From comparison to such passages it is clear these lines must be understood in the cultural context of the flyting and taunting speeches of Germanic warriors as a direct provocation, as is most clearly seen in this final half-line.  This in turn is what motivates Hathbrand’s hostile answer in ll. 57--61.}}“\eva

\bvb Well now, Wielding God!” quoth Hildbrand, “The woeful Weird comes to pass. \\
I roamed for sixty summers and winters away from this land, \\
whenas I always was placed in the troop of shooters, \\
while at no stronghold my bane was fastened.--- \\
Now shall my beloved child strike me with his sword, \\
beat me with his blade---or I become his bane. \\
Yet mayst thou now easily---if thy zeal avails thee--- \\
from such a hoary man win the armour, \\
bear away the booty---if thou hast any right thereto!”\evb\evg


\bvg\bva[][57]\mssnote{\HildMS~76v/22--26}%
\Ballnote{That the speaker of ll. 57--61 is in fact Hathbrand, the son, rather than Hildbrand, the father, is suggested by the speaker’s hostile tone (unlike the more resigned, albeit taunting tone of ll. 48--56) and his prejudice against \emph{ôstar-liuto} ‘easterners’, i.e. the Huns, against whom Hathbrand is clearly negatively disposed and to whom he thinks the imposter Hildbrand belongs (cf.~l. 38).  A switch of speaker is further implied by the finality of Hildbrand’s speech in l. 56b.

On its own, the speech-marker in the cæsura of l. 57 (\emph{kwad X} ‘quoth X’) is not proof that we are dealing with a new speaker (cf.~l. 48), but it does yield additional support to it considering the high likelihood of corruption due to the formulaic nature of the marker and the numerous proofs of clear inattention on the part of the scribe(s) of the poem as a whole (e.g. the dittography in l. 26, the highly inconsistent handling of the name Thedric ll. 19, 23, 26, and the various unmetrical lines).}%
„Der sí doh nú \alst{a}rgósto“ \edtext{\hld[kwad \emph{Hadu}·brant]}{\lemma{kwad \emph{Hadu}·brant}\Afootnote{emend.; \emph{‘quad hiltibrant’} \HildMS}} „\alst{ô}star-liuto &
der dir nú \alst{w}íges \alst{w}arne \hld\ nú dih es só \alst{w}el lustit &
gu̇dja gi·\alst{m}ęinun; \hld\ niuse de \alst{m}ót̨t̨i &
\edtext{\emph{h}wę*dar}{\Afootnote{\emph{werdar} \HildMS}} sih \edtext{\emph{\alst{h}iutu dêro}}{\Afootnote{metr. emend.; rev. \HildMS}} \edtext{\alst{h}ręgilo \hld\ \edtext{\alst{h}ru\emph{o}men}{\Afootnote{\emph{hrumen} \HildMS}} muot̨t̨i &
\edtext{e\emph{d}do}{\Afootnote{\emph{erdo} \HildMS}} desero \alst{b}runnóno \hld\ \alst{b}êdero waltan}{\lemma{hręgilo hruomen muot̨t̨i \dots\ desero brunnóno bêdero waltan ‘of these garments will boast \dots\ both these byrnies wield’}\Bfootnote{Like in the Iliad, the winner is expected to strip the slain of his armour.}}!“\eva

\bvb “Yet he would now,” quoth Hathbrand, “be the cowardliest of easterners \\
whoever should deny thee the fight now that thou hast such desire \\
for a battle between us.  Let him try who might, \\
whichever one today of these garments will boast, \\
or both these byrnies wield!”\evb\evg


\bvg\bva[][62]\mssnote{\HildMS~76v/26--29}%
Dó lét̨t̨un sé \alst{æ̂}rist \hld\ \edtext{\alst{a}skim}{\Afootnote{\emph{asckim} \HildMS}} skrítan &
\edtrans{\alst{sk}arpén \alst{sk}úrim}{in sharp showers}{\Bfootnote{Formulaic half-line describing a hail of projectiles, also occurring in \textlink{Heliand} 5136a.}} \hld\ dat̨ in dem \alst{sk}iltim stónt &
dó \alst{st}óptun t̨ó·samane \hld\ \edtrans{\alst{st}aim-bort}{painted boards}{\Bfootnote{A cpd., transparently segmented as \emph{staim-} + \emph{bort} ‘board, plank’.  In OE poetry \emph{bord} can also mean ‘shield’ (e.g. \Beowulf\ 2259, 2524, 2673, in cpds.~1243, 2203, 2559, 2895, 2980), as can OS \emph{bord} (only once: \textlink{Heliand} 5767) and ON \emph{borð} (e.g. Þmáhl \emph{Máv} 8/2 (in \Skp\ 5)), so that this is probably an inherited poetic sense and certainly the sense here.

The first element is more difficult, but is generally understood to be an assimilated form of \emph{stęin} ‘stone’ (though note that in the ms.~the digraph \emph{ai} is only found here, and that the etymological reflex of PGmc.~\emph{*ai} is otherwise spelled \emph{‘ei’} or \emph{‘ae’}).  This has often been interpreted as ‘jewel’ (an attested sense), so that the sense would be ‘bejewelled shield’, yet the existence of such shields in the late 8th c.~must be considered highly doubtful.  On the basis of the ON cognate \emph{stęinn}, which in addition to ‘stone’ can also mean ‘mineral colour, paint’ (cf.~ModEngl.~stain, e.g. in “stained glass”, an ON loan), I propose that the word mean ‘paint’ in the sense ‘painted shield’.
Such painted shields are in fact attested in near-contemporary Norse poetry (\Ragnarsdrapa\ 1, ca.~850 \ce), which, although not identical to High German poetry, still belongs to the same Germanic epic tradition.}} \edtext{chlu\emph{b}un}{\Afootnote{\emph{chludun} \HildMS}} &
\alst{h}ewun harm-líkko \hld\ \alst{h}wít̨t̨e skilti &
unti imo iro \alst{l}intún \hld\ \alst{l}út̨t̨ilo wurtun &
gi·\alst{w}igan miti \alst{w}ábnum \hld\ \edtext{[...]}{\Bfootnote{At this point \HildMS\ 76v ends in the middle of the line.  The rest of the poem would have been found on the now-lost following page(s); see Introduction above.}}\eva

\bvb Then they first let their ashen spears glide \\
in sharp showers so that in the shields they stuck. \\
Then they charged at each other, split the painted boards \ken{shields}; \\
they hewed harmfully at the white shields, \\
till before them their linden-planks \ken{shields} became small, \\
worn down by the weapons, [...]\evb\evg

\sectionline
