\bookStart{Widsith}[Wíd·sïþ]
\setBookCode{Widsith}

\begin{flushright}%
\textbf{Dating:} C7th–8th \parencite{Neidorf2013}

\textbf{Meter:} \Fornyrdislag%para
\end{flushright}%

\section{Introduction}

An archaic heroic poem found in the Exeter Book.

\sectionline

\newpage

\section{Text}

\bvg\bva%
\alst{W}íd-sïð maðolade, \hld\ \alst{w}ord-hord ǫn·leac, &
sé þe \alst{m}æ̂st \hld\ \alst{m}ærþa ofer eorþan, &
\alst{f}olca geond·\alst{f}ǿrde; \hld\ oft hé \alst{f}lętte ge·þâh &
\alst{m}yne-lícne \alst{m}âþþum. \hld\ Hine frǫm \alst{M}yrgingum &
\alst{æ}þele \alst{ǫ}n·wócon. \hld\ He mid \alst{E}alh·hilde, &
\alst{f}æ̂lre \edtrans{\alst{f}reoþu-wębban}{peace-weaveress}{\Bfootnote{A woman used in a political marriage to bring peace between two tribes or families, in this case between a king Edwin (st. 20) and Ermenric of the Gots.}}, \hld\ \alst{f}orman sïþe &
\edtrans{\alst{H}reð-cyninges}{Reth-King}{\Bfootnote{The king of the \inx[P]{Reth-Gots}, which is apparently just a poetic name for the (Eastern) Gots; cf. sts. 4, 12, 19.}} \hld\ \alst{h}âm ge·sóhte &
\alst{éa}stan of \alst{Ǫ}ngle, \hld\ \alst{Eo}rman-ríces, &
\alst{w}râþes \alst{w}ær-logan. \hld\ Ǫn·gǫnn þȧ \alst{w}orn sprecan:\eva

\bvb%
{\huge W}\textsc{idsith spoke}, unlocked his word-hoard, \\
he who furthest through tribes on the earth \\
and nations had journeyed.  Oft on the bench had he received \\
delightful treasures.  From the Mirgings \\
his ancestors came.  Along with Elhild \\
the good peace-weaveress for the very first time \\
he sought the Reth-King’s realm, \\
east of the Angles: [the realm] of \inx[P]{Ermenric}, \\
the fierce oath-breaker.  He then began a long speech:\evb\evg


\bvg\bva%
„Fela ic \alst{m}ǫnna ge·frægn \hld\ \alst{m}ægþum wealdan. &
Sceal \alst{þ}éoda ge·hwylc \hld\ \alst{þ}éawum lifgan, &
\alst{eo}rl æfter \alst{ȯ}þrum \hld\ \alst{ǿ}ðle rǽdan, &
sé þe his \alst{þ}éoden-stól \hld\ ge·\alst{þ}éon wile.\eva

\bvb “A great deal of men I’ve learned ruling tribes. \\
Every person shall live in virtue, \\
each earl after the other lead his homeland \\
who on his sovereign seat will prosper.\evb\evg


\bvg\bva%
Þâra wæs Wala \hld\ hwíle sǿlast, &
ǫnd \alst{A}lexandreas \hld\ \alst{ea}lra rícost &
\alst{m}ǫnna cynnes, \hld\ ǫnd he \alst{m}æ̂st ge·þȧh &
þâra þe ic ofer \alst{f}oldan \hld\ ge·\alst{f}rægen hæbbe.\eva

\bvb Of them was Wale for a while the best, \\
and Alexander of all the strongest \\
of mankind, and he prospered most \\
of those men over the earth of whom I’ve learned.\evb\evg


\bvg\bva%
\alst{Æ}tla weold Húnum, \hld\ \alst{Eo}rman·ríc Gotum, &
\alst{B}ecca \alst{B}aningum, \hld\ \alst{B}urgendum Gifica. &
\alst{C}âsere weold \alst{C}réacum \hld\ ǫnd \alst{C}ælic Finnum, &
\alst{H}agena \alst{H}olm-rycum \hld\ ǫnd \alst{H}enden Glommum.\eva

\bvb Attle ruled the Huns, Ermenric the Gots, \\
Bicke the Banings, Yivick the Burgends. \\
Coser ruled the Greeks and Calic the Finns, \\
Hain the Holmrighs and Henden the Glams.”\evb\evg


\bvg\bva%
\alst{W}itta weold Swǽfum, \hld\ \alst{W}ada Hælsingum, &
\alst{M}eaca \alst{M}yrgingum, \hld\ \alst{M}earch·ealf Hundingum. &
\alst{Þ}éod·ríc weold Frǫncum, \hld\ \alst{Þ}yle Rǫndingum, &
\alst{B}reoca \alst{B}rǫndingum, \hld\ \alst{B}illing Wernum. &
\alst{Ȯ}s·wine weold Eowum \hld\ ǫnd \alst{Y}tum Gef·wulf, &
\alst{F}inn \alst{F}olc·walding \hld\ \alst{F}resna cynne.\eva

\bvb Wit ruled the Sweves, Wade the Helsings, \\
Mack the Mirgings, Marchalf the Hundings. \\
Thedric ruled the Franks, Thyle the Randings, \\
Brack the Brandings, Billing the Werns. \\
Oswin ruled the Eawes and Gefwolf the Ytes, \\
Finn Folkwalding the race of Frisians.\evb\evg


\bvg\bva%
\alst{S}ige·hęre lęngest \hld\ \alst{s}æ̂-Dęnum weold, &
\alst{H}næf \alst{H}ocingum, \hld\ \alst{H}elm Wulfingum, &
\alst{W}ald \alst{W}óingum, \hld\ \alst{W}ód Þyringum, &
\alst{S}æ̂·ferð \alst{S}ycgum, \hld\ \alst{S}wéom Ongend·þeow, &
\alst{Sc}eaft·hęre Ymbrum, \hld\ \alst{Sc}eafa Lǫng-beardum, &
\alst{H}ún \alst{H}æt-werum \hld\ ǫnd \alst{H}olen Wrosnum; &
\alst{H}ring·wald wæs \alst{h}âten \hld\ \alst{H}ęre-farena cyning.\eva

\bvb Sigar for the longest ruled the sea-Danes, \\
Naf the Hockings, Helm the Wolvings, \\
Wald the Wooings, Wood the Thirings, \\
Seaferth the Sidges, Angenthew the Swedes, \\
Shafter the Imbres, Sheaf the Longbeards, \\
Hown the Hatweres, and Holen the Wrasens; \\
Ringwald was the name of the king of Herefarers.\evb\evg


\bvg\bva%
\Ballnote{The great praise here accorded to Offe, the legendary ruler of the continental Angles from whom the Anglo-Saxons claimed desent, is paralleled by \Beowulf\ 1954–1960a.}%
\alst{O}ffa weold Ǫngle, \hld\ \alst{A}le·wíh Dęnum; &
sé wæs þâra \alst{m}anna \hld\ \alst{m}ódgast ealra, &
no hwæþre he ofer \alst{O}ffan \hld\ \alst{eo}rl-scype fręmede, &
ac \alst{O}ffa ge·slóg \hld\ \alst{æ̂}rest mǫnna, &
\alst{c}niht-wesende, \hld\ \alst{c}yne-ríca mæ̂st.\eva

\bvb Offe ruled the Angles, Alewigh the Danes; \\
of these men he was bravest of all, \\
but he did not further greater earlship than Offe, \\
for Offe conquered—youngest of men, \\
still a boy—the greatest of kingdoms.\evb\evg


\bvg\bva%
Nænig \alst{e}fen-\alst{ea}ld him \hld\ \alst{eo}rl-scipe mâran &
\alst{ǫ}n or·\alst{ę}tte: \hld\ \alst{â}ne sweorde &
\alst{m}ęrce ge·\alst{m}æ̂rde \hld\ wið \alst{M}yrgingum &
bi \alst{F}ïfel-dore; \hld\ heoldon \alst{f}orð siþþan &
\alst{Ę}ngle ǫnd Swǽfe, \hld\ swá hit \alst{O}ffa ge·slóg.\eva

\bvb No man of his age accomplished \\
greater earlship: with but one sword \\
he marked the border against the Mirgings \\
by Fiveldoor.  It was thenceforth held \\
by the Angles and Sweves as Offe had won it.\evb\evg


\bvg\bva%
\alst{H}róþ·wulf ǫnd \alst{H}róð·gâr \hld\ \alst{h}eoldon lęngest &
\alst{s}ibbe æt·\alst{s}ǫmne \hld\ \alst{s}uhtor-fædran, &
siþþan hý for·\alst{w}rǽcon \hld\ \alst{W}ícinga cynn &
ǫnd \alst{I}ngeldes \hld\ \alst{o}rd for·bigdon, &
for·\alst{h}eowon æt \edtrans{\alst{H}eorote}{Hart}{\Bfootnote{The legendary hall of the Danish kings, the Shieldings.  It is described at length in \Beowulf.}} \hld\ \alst{H}eaðo-beardna þrym.\eva

\bvb Rotholf and Rothgar held for the longest \\
the peace together, uncle and nephew, \\
since they drove away the race of Wikings, \\
and bent down Ingeld’s spear-point; \\
at Hart they cut down the host of the Hathbeards.\evb\evg


\bvg\bva%
Swá ic geond·\alst{f}ǿrde \alst{f}ela \hld\ \alst{f}ręmdra lǫnda &
geond \alst{g}inne \alst{g}rund. \hld\ \alst{G}ódes ǫnd yfles &
þǽr ic \alst{c}unnade; \hld\ \alst{c}nósle bi·dæ̂led, &
\alst{f}réo-mǽgum \alst{f}eor \hld\ \alst{f}olgade wíde.\eva

\bvb So I journeyed through a great deal of strange lands \\
through the wide world.  Of good and evil \\
I there became acquainted.  Of kin deprived, \\
far from dear kinsmen, I strayed widely.\evb\evg


\bvg\bva%
For·þǫn ic mæg \alst{s}ingan \hld\ ǫnd \alst{s}ęcgan spell, &
\alst{m}æ̂nan fore \alst{m}ęngo \hld\ in \alst{m}eodu-healle &
hú mé \alst{c}yne-góde \hld\ \alst{c}ystum dohten.\eva

\bvb Therefore I can sing and tell tales, \\
recall before the many in the mead-hall, \\
how men of good kin treated me with grace.\evb\evg


\bvg\bva%
Ic wæs mid \alst{H}únum \hld\ ǫnd mid \alst{H}reð-Gotum, &
mid \alst{S}wéom ǫnd mid Géatum \hld\ ǫnd mid \alst{S}u̇þ-Dęnum. &
Mid \alst{W}ęnlum ic wæs ǫnd mid \alst{W}ærnum \hld\ ǫnd mid \alst{W}ícingum; &
mid \alst{G}efþum ic wæs ǫnd mid Winedum \hld\ ǫnd mid \alst{G}efflegum; &
mid \alst{Ę}nglum ic wæs ǫnd mid Swǽfum \hld\ ǫnd mid \alst{Æ}nenum; &
mid \alst{S}eaxum ic wæs ǫnd \alst{S}ycgum \hld\ ǫnd mid \alst{S}weord-werum; &
mid \alst{H}ronum ic wæs ǫnd mid Déanum \hld\ ǫnd mid \alst{H}eaþo-Réamum.\eva

\bvb I was among Huns and among Reth-Gots, \\
among Swedes and among Geats, and among South-Danes. \\
Among Wendles I was and among Warns, and among Wikings; \\
among Yefths I was and among Wends, and among Yefflegs; \\
among Angles I was and among Sweves, and among Anens; \\
among Saxes I was and among Sidges, and among Sword-weres; \\
among Ranes I was and among Deans, and among Hath-Reams.\evb\evg


\bvg\bva%
Mid \alst{Þ}yringum ic wæs \hld\ ǫnd mid \alst{Þ}rowendum, &
ǫnd mid \alst{B}urgendum, \hld\ þǽr ic \alst{b}éag ge·þâh; &
mé þǽr \alst{G}u̇ð·hęre for·\alst{g}eaf \hld\ \alst{g}læd-lícne maþþum &
\alst{s}ǫnges to léane. \hld\ Næs þæt \alst{s}æne cyning!\eva%TODO: Check sæne.

\bvb Among Thirings I was and among Throwends \\
and among the Burgends where I received a bigh; \\
there Guther gladdened me with treasures \\
as reward for my song. That was not a bad king!\evb\evg


\bvg\bva%
Mid \alst{F}rǫncum ic wæs ǫnd mid \alst{F}rysum \hld\ ǫnd mid \alst{F}rumtingum; &
mid \alst{R}ugum ic wæs ǫnd mid Glommum \hld\ ǫnd mid \alst{R}úm-walum.\eva

\bvb Among Franks I was and among Frises, and among Frumtings; \\
among Ruges I was and among Glams, and among Rome-Wales.\evb\evg


\bvg\bva%
Swylce ic wæs ǫn \alst{Ea}tule \hld\ mid \edtrans{\alst{Æ}lf·wine}{Elfwin}{\Bfootnote{Latin \emph{Alboinus} (530s – 28 June 572), king of the Longbeards.  As king he succeeded his father Edwin (Latin \emph{Audoinus}) and led the Longbeards to settle in Northern Italy, where they established their kingdom and gave the region its present name, Lombardy.}}, &
sé hæfde \alst{m}ǫn-cynnes, \hld\ \alst{m}ïne ge·fræge, &
\alst{l}eohteste hǫnd \hld\ \alst{l}ofes tó wyrcenne, &
\alst{h}eortan un·\alst{h}neaweste \hld\ \alst{h}ringa ge·dâles, &
\alst{b}eorhtra \alst{b}éaga, \hld\ \alst{b}earn Éad·wines.\eva

\bvb Likewise was I in Italy with Elfwin. \\
Of mankind he had—as I have learned— \\
the lightest hand in working generosity, \\
the unstingiest heart in the dealing of rings \\
and bright bighs, the child of Edwin.\evb\evg


\bvg\bva%
Mid \edtrans{\alst{S}ęrcingum}{Serkings}{\Bfootnote{Probably related to ‘Saracen’.  Cf. ON \emph{Sęrk-land} ‘Serkland’, a name for Arabia or North Africa, which appears to be from the same Germanic source.}} ic wæs \hld\ ǫnd mid \alst{S}eringum; &
mid \alst{C}reacum ic wæs ǫnd mid Finnum \hld\ ǫnd mid \edtrans{\alst{C}âsere}{Coser}{\Bfootnote{Caesar, the Roman Emperor, but no particular historical emperor is intended.  Cf. \textlink{Volundarkvida}[P1], 15, \textlink{Hildebrandslied} 33.}}, &
sé þe \edtrans{\alst{w}ïn-burga}{wine-towns}{\Bfootnote{Roman wineyards.}} \hld\ ge·\alst{w}eald áhte, &
\alst{w}iolena ǫnd \alst{w}ilna, \hld\ ǫnd \edtrans{\alst{W}ala ríces}{the realm of the Wales}{\Bfootnote{The Roman Empire, or more specifically Gaul (which in ON was known as \emph{Val-land} from the same root).  For the use of \emph{Walas} ‘Wales’ in the sense ‘Romans’ see st. 14/2 above}}.\eva

\bvb Among Serkings I was and among Searings; \\
among Greeks I was and among Finns, and with Coser, \\
he who held dominion over wine-towns, \\
over wealth and concubines and the realm of the Wales.\evb\evg


\bvg\bva%
Mid \alst{Sc}ottum ic wæs ǫnd mid Peohtum \hld\ ǫnd mid \alst{Sc}ríde-Finnum; &
mid \alst{L}íd-wícingum ic wæs ǫnd mid \alst{L}éonum \hld\ ǫnd mid \alst{L}ǫng-beardum, &
mid \alst{h}æ̂ðnum ǫnd mid \alst{h}æleþum \hld\ ǫnd mid \alst{H}undingum.\eva

\bvb Among Scots I was and among Picts, and among Shrithe-Finns; \\
among Lid-Wikings I was among Leans, and among Longbeards; \\
among heathens and among heroes and among Hundings.\evb\evg


\bvg\bva%
Mid \alst{I}srahelum ic wæs \hld\ ǫnd mid \alst{E}xsyringum, &
mid \alst{E}breum ǫnd mid \alst{I}ndeum \hld\ ǫnd mid \alst{E}gyptum. &
Mid \alst{M}oidum ic wæs ǫnd mid Persum \hld\ ǫnd mid \alst{M}yrgingum, &
ǫnd \alst{M}ofdingum \hld\ ǫnd on·gęnd \alst{M}yrgingum, &
ǫnd mid \edtext{\alst{A}moþingum}{\Afootnote{\emph{‘Amothingum’} ms.}}. \hld\ Mid \alst{Éa}st-Þyringum ic wæs &
ǫnd mid \alst{Eo}lum ǫnd mid \alst{I}stum \hld\ ǫnd \alst{I}dumingum.\eva

\bvb Among Israelites I was and among Assyrians, \\
among Hebrews and among Indians and among Egyptians. \\
Among the Medes I was and among Persians, and among Mirgings \\
and Mofdings and again the Mirgings \\
and among Amothings. Among East-Thirings I was \\
and among Eals and among Ists, and Idumings.\evb\evg


\bvg\bva%
\alst{Ǫ}nd ic wæs mid \alst{Eo}rman·ríce \hld\ \alst{ea}lle þráge, &
þǽr mé \alst{G}otena cyning \hld\ \alst{g}óde dohte; &
sé mé \alst{b}éag for·geaf, \hld\ \alst{b}urg-warena fruma, &
ǫn þam \alst{s}iex hund wæs \hld\ \alst{s}mǽtes goldes, &
ge·\alst{sc}yred \alst{sc}eatta \hld\ \alst{sc}illing-ríme; &
þǫne ic \alst{Éa}d·gilse \hld\ ǫn \alst{ǽ}ht sealde, &
mínum \alst{h}léo-dryhtne, \hld\ þá ic tó \alst{h}âm bi·cwǫm, &
\alst{l}éofum to \alst{l}éane, \hld\ þæs þe hé mé \alst{l}ǫnd for·geaf, &
mínes \alst{f}æder ǿþel, \hld\ \alst{f}réa Myrginga.\eva

\bvb And I was with Ermenric for the longest time, \\
where the king of the Gots treated me well. \\
He gave me a bigh—that chief of city-dwellers— \\
in which were reckoned six hundred shats \\
of purest gold in shilling-count. \\
I gave it in the possession of Edgils, \\
my dear shelter and lord, when I came home, \\
as repayment for his giving me land, \\
—that lord of Mirgins—my father’s ethel.\evb\evg


\bvg\bva%
\alst{Ǫ}nd mé þá \alst{Ea}lh·hild \hld\ \alst{ȯ}þerne for·geaf, &
\alst{d}ryht-cwén \alst{d}uguþe, \hld\ \alst{d}ohtor Éad·wines; &
hyre \alst{l}of \alst{l}ęngde \hld\ geond \alst{l}ǫnda fela. &
Þǫnne ic be \alst{s}ǫnge \hld\ \alst{s}ęcgan sceolde &
hwær ic under \alst{s}wegl \hld\ \alst{s}ǿlast wisse &
\alst{g}old-hrodene cwén \hld\ \alst{g}iefe bryttian.\eva

\bvb And then Elhild gave me another one, \\
the noble queen of the old troop, daughter of Edwin; \\
her praise stretched far through many lands. \\
Then in my song I had to say, \\
where beneath the sun I knew the best \\
gold-adorned queen dispensing gifts.\evb\evg


\bvg\bva%
Þǫnne wit \alst{Sc}illing \hld\ \alst{sc}íran reorde &
for uncrum \alst{s}ige-dryhtne \hld\ \alst{s}ǫng a·hófon, &
\alst{h}lúde bí \alst{h}earpan, \hld\ \alst{h}leoþor swinsade, &
þǫnne \alst{m}ǫnige \alst{m}ęnn, \hld\ \alst{m}ódum wlǫnce, &
\alst{w}ordum sprécan, \hld\ þá þe \alst{w}el cu̇þan, &
þæt hí næ̂fre \alst{s}ǫng \hld\ \alst{s}ǿllan ne hýrdon.\eva

\bvb Then I and Shilling with clear voices, \\
for our victorious lord raised up a song, \\
loud by the harp—the tune rang out. \\
Then many men proud of heart \\
told with words—those who knew well— \\
that they never had heard a better song.\evb\evg


\bvg\bva%
Ðǫnan ic \alst{ea}lne geond·hwearf \hld\ \alst{ǿ}þel Gotena, &
\alst{s}óhte ic â \alst{s}ïþa \hld\ þá \alst{s}ǿlestan; &
þæt wæs \alst{i}nn-weorud \hld\ \alst{Ea}rman·rices.\eva

\bvb Thence I passed through all the ethel of the Gots; \\
I always sought out the best companions; \\
that was the inner retinue of Ermenric.\evb\evg


\bvg\bva%
\alst{H}eðcan sóhte ic ǫnd Beadecan \hld\ ǫnd \alst{H}ęre-lingas, &
\alst{E}mercan sóhte ic ǫnd Fridlan \hld\ ǫnd \alst{Éa}st-Gotan, &
\alst{f}ródne ǫnd gódne \hld\ \alst{f}æder Un·wø̇nes.\eva

\bvb TODO\evb\evg


\bvg\bva%
\alst{S}eccan sóhte ic ǫnd Beccan, \hld\ \alst{S}eafolan ǫnd Þéod·ríc, &
\alst{H}eaþo·ríc ǫnd Sifecan, \hld\ \alst{H}liþe ǫnd Incgen·þeow. &
\alst{Éa}d·wine sóhte ic ǫnd \alst{E}lsan, \hld\ \alst{Æ}gel·mund ǫnd Hún·gâr, &
ǫnd þá \alst{w}lǫncan ge·dryht \hld\ \alst{W}iþ-myrginga.\eva

\bvb TODO\evb\evg


\bvg\bva%
\alst{W}ulf·hęre sóhte ic ǫnd \alst{W}yrm·hęre; \hld\ ful oft þǽr \alst{w}íg ne á·læg, &
þǫnne \alst{H}ræda hęre \hld\ \alst{h}eardum sweordum &
ymb \alst{W}istla-\alst{w}udu \hld\ \alst{w}ęrgan sceoldon &
\alst{ea}ldne \alst{ǿ}þel-stól \hld\ \alst{Æ}tlan léodum.\eva

\bvb I sought out Wolfer and Wyrmer—most seldom there the warring ceased \\
when the Reth-host with hard swords \\
in the Wistlewood had to defend \\
the old ethel-seat against Attle’s folk.\evb\evg


\bvg\bva%
\alst{R}ǽd·hęre sóhte ic ǫnd \alst{R}ǫnd·hęre, \hld\ \alst{R}úm·stân ǫnd Gisl·hęre, &
\alst{W}iþer·gield ǫnd Freoþe·ric, \hld\ \alst{W}udgan ǫnd Hâman; &
ne wǽron þæt ge·\alst{s}ïþa \hld\ þá \alst{s}ǽmestan, &
þéah þe ic hý a·\alst{n}íhst \hld\ \alst{n}ęmnan sceolde.\eva

\bvb I sought out Reeder and Rander, Roomstan and Yisler, \\
Witheryeld and Frithric, Woody and Homer; \\
nor were those the worst of companions, \\
although I last should mention them.\evb\evg


\bvg\bva%
Ful oft of þâm \alst{h}éape \hld\ \alst{h}wínende fléag &
\edtrans{\alst{g}iellende \alst{g}âr}{yelling spear}{\Bfootnote{Formulaic.  Cf. in ON \textlink{Atlakvida}[5]/2a: \emph{af gęiri gjallanda}, Egill Lv 9/2a (\Skp\ 5): \emph{ok gjallanda gęiri}.}} \hld\ ǫn \alst{g}rǫme þéode; &
\alst{w}ræccan þǽr \alst{w}eoldan \hld\ \alst{w}undnan golde &
\alst{w}erum ǫnd \alst{w}ífum, \hld\ \alst{W}udga ǫnd Hâma.\eva

\bvb Most often from that troop did fly a whizzing, \\
yelling spear into the fiendish folk. \\
There ruled the exiles Woody and Homer \\
twisted gold, men and women.\evb\evg


\bvg\bva%
Swá ic þæt symle ǫn·\alst{f}ǫnd \hld\ ǫn þæ̂re \alst{f}ęringe, &
þæt sé biþ \alst{l}éofast \hld\ \alst{l}ǫnd-búendum &
sé þe him \alst{G}od syleð \hld\ \alst{g}umena ríce &
to ge·\alst{h}ealdenne, \hld\ þenden hé \alst{h}ér leofað.“\eva

\bvb So I always found while on that journey, \\
that he is dearest to land-dwellers \ken{men}, \\
whom God grants sovereignity of men \\
for to hold while here he lives.”\evb\evg


\bvg\bva%
\Ballnote{Widsith’s speech ends, and the poet reflects on the importance of generosity.}%
Swá \alst{sc}ríþende \hld\ ge·\alst{sc}eapum hweorfað &
\alst{g}léo-męnn \alst{g}umena \hld\ geond \alst{g}runda fela, &
\alst{þ}earfe sęcgað, \hld\ \alst{þ}ǫnc-word sprecaþ, &
\alst{s}imle \alst{s}u̇ð oþþe norð \hld\ \alst{s}umne ge·mǿtað &
\alst{g}ydda \alst{g}leawne, \hld\ \alst{g}eofum un·hneawne, &
sé þe fore \alst{d}uguþe wile \hld\ \alst{d}óm a·ræ̂ran, &
\alst{eo}rl-scipe \alst{æ}fnan, \hld\ oþ·þæt \alst{ea}l scæceð, &
\alst{l}eoht ǫnd \alst{l}if sǫmod; \hld\ \alst{l}of sé ge·wyrceð, &
\alst{h}afað under \alst{h}eofonum \hld\ \alst{h}éah-fæstne dóm.\eva

\bvb So passing through fates they wander, \\
the song-men of mankind, through many lands. \\
They tell their needs, speak thoughtful words; \\
whether in the south or north they meet some one, \\
gay in songs, unstingy with gifts, \\
who for the old troop will rear up fame, \\
accomplish earlship until all goes away, \\
light and life together.  He who works generosity \\
has under the heavens high, lasting fame.\evb\evg

\sectionline
