\bookStart{Wolf and Edwaker}[Wulf and Éad·wacer]
\setBookCode{Wulf}

\begin{flushright}%
\textbf{Dating:} TODO

\textbf{Meter:} \Fornyrdislag, \Ljodahattr%para
\end{flushright}%

\section{Introduction}

\textbf{Wolf and Edwaker} (abbrev.~\Wulf) is a mysterious stanzaic poem.

\subsection{Preservation}

\Wulf\ is found in \ExeterBook\ on foll.~100v/14--101r/4.  It comes after the heroic elegy \textlink{Deor} and is followed by Riddles 1--59 (\emph{Hwylc is hæleþa þæs horsc} \dots).
rom the Exeter Book.

\newpage

\section{Text}

\bvg\bva%
\alst{L}éodum is mïnum \hld\ swylce him mǫn \alst{l}âc gife; &
willað hý hine á·\alst{þ}ęcgan, \hld\ gif hé ǫn \alst{þ}réat cymeð. &
\ind \alst{U}n-ge·líc is \alst{u̇}s.\eva

\bvb For my people it is as if one gave them an offering; \\
they wish to devour him if he comes upon that host. \\
\ind It is not so for us.\evb\evg


\bvg\bva%
Wulf is ǫn \alst{ie}ge, \hld\ ic ǫn \alst{ȯ}þerre; &
\alst{f}æst is þæt eg-lǫnd, \hld\ \alst{f}ęnne bi·worpen. &
Sindon \alst{w}æl-reowe \hld\ \alst{w}eras þǽr ǫn ige; &
willað hý hine á·\alst{þ}ęcgan, \hld\ gif hé ǫn \alst{þ}réat cymeð. &
\ind \alst{U}n-ge·líce is \alst{u̇}s.\eva

\bvb Wolf is on an island, I on another; \\
fast is that island, surrounded by water. \\
There are slaughter-fierce warriors there on the island; \\
they wish to devour him if he comes upon that host. \\
\ind It is not so for us.\evb\evg


\bvg\bva%
\alst{W}ulfes ic mïnes \alst{w}íd-lâstum \hld\ \alst{w}ėnum dogode, &
þǫnne hit wæs \alst{r}énig weder \hld\ ǫnd ic \alst{r}éotugu sæt; &
þǫnne mec sé \alst{b}eadu-câfa \hld\ \alst{b}ógum bi·lęgde, &
wæs mé \alst{w}yn tó þǫn, \hld\ \alst{w}æs mé hwæþre éac lâð.\eva

\bvb The far journeys of my Wolf I’ve thought of in my yearnings, \\
when it was rainy weather and I sat grieving, \\
when the battle-bold man laid his arms around me, \\
I found joy in that, yet also found it loathsome.\evb\evg


\bvg\bva%
\alst{W}ulf, mïn \alst{W}ulf, \hld\ \alst{w}ėna mé þïne &
\alst{s}éoce ge·dydon, \hld\ þïne \alst{s}eld-cymas, &
\alst{m}urnende \alst{m}ód, \hld\ nales \alst{m}ęte-liste.\eva

\bvb Wolf, my Wolf! My yearnings for thee \\
have made me sick, and thy rare comings, \\
a mourning mind---nowise want of food!\evb\evg


\bvg\bva%
Ge·hýrest þú, \alst{Éa}d·wacer? \hld\ Uncerne \alst{ea}rne hwelp &
\ind bireð \alst{W}ulf to \alst{w}uda; &
þæt mǫn éaþe tó·\alst{s}liteð \hld\ þætte nǽfre ge·\alst{s}ǫmnad wæs, &
\ind uncer \alst{g}iedd \alst{g}eador.\eva

\bvb Hearest thou Edwaker?  Our wretched whelp \\
\ind is carried by Wolf to the woods. \\
One easily tears apart that which was never put together: \\
\ind our song together.\evb\evg


\sectionline
