\bookStart{Muspell}[Muspilli]
\setBookCode{Muspilli}

\begin{flushright}%
\textbf{Dating:} C9th

\textbf{Meter:} \Fornyrdislag%para
\end{flushright}%

\section{Introduction}

The \textbf{Muspell} (\Muspilli) is an Old High German Christian poem dealing with the Day of Judgment.

\subsection{Preservation}

\Muspilli\ survives in a single copy, found scribbled in a Latin-language theological manuscript from the 820s \ce\ with the signum BSB Clm 14098 (in the apparatus of the pres. poem denoted as 14098).  Lines 1--18a are found on fol.~61r while the rest of the poem is found on foll.~119v--121v.  Foll. 61r and 120v--121v are blank and there the poem occupies the whole page; on 119v--120r the main text is in Latin and the poem occupies the lower third of the page in the bottom margin below the text.

From 61v--120r the main text of 14098 is a pseudo-Augustinian sermon, which ends on 120r with a dedication by the archbishop of Salzburg Ethelram (Latin \textlatin{Adalrammus}) to the young duke Ludwigh (Latin \textlatin{Hludowicus}, later Louis the German).  It is important to note that the poem and main text are written in different hands.  The hand of the poem is shaky and ugly which may suggest that its author was not very used to writing; he may perhaps even have been Ludwigh himself.  From the start on fol.~61r we can presume that the writer of the poem originally meant to write down just a part of it as marginalia, beginning on the back leaf (61r) of the already written sermon and continuing on its last two (119v--120r).  When he ran out of space he acquired another blank leaf, 121r--v.  The abrupt ending of the poem may suggest that further blank leaves were also present.

Due to the low quality of the digitally available scans of the already very poorly preserved manuscript, the pres. ed. has had to rely partly on the text found in the 17th edition of Braune’s \emph{Althochdeutsche Lesebuch} published in 1994.%TODO: add when true:  All critical choices have, however, been controllied by the editor against scans of the manuscript.

\subsection{Content}

\Muspilli\ is a Catholic sermon that describes the events of the Judgment Day and advises its readers to do penance and give alms to assure their safety.  The poem is not a masterpiece and in the original at times comes off as stilted.  Still it has a certain charm, and its meter is notably regular when compared to the other comparable High German alliterative poems \textlink{Hildebrandslied} and \textlink{Wessobrunn}.

\subsubsection{Relation to Otfrid}

The use of end-rhymed Otfridian meter in ll.~60--61 suggests a relation between \Muspilli\ and Otfrid’s \emph{Evangelienbuch} (written 863--871 \ce), something which is strongly supported by the exact correspondence between \Muspilli\ 14 and \emph{Evangelienbuch} 1.18.9.  Since there appears to be a relation between these two texts, we ought to try to determine the direction of influence.  I believe that we may reasonably conclude on the basis of literary history and the internal metrical variation of the \emph{Evangelienbuch} that the direction of influence is more likely to be from \Muspilli\ to the \emph{Evangelienbuch} rather than the reverse, and will lay out my argument as follows.

With the exception of a very small number of irregular, often alliterating lines found exclusively in the first book (e.g. 1.4.9, 1.5.5, 1.7.19, 1.7.27--28, 1.18.9; cf.~1.5.6, 1.5.10--12 which combine alliteration with end-rhyme), the \emph{Evangelienbuch} is completely end-rhymed and non-alliterative.  With the exception of 1.18.9 these lines are all found in the first fourth of the first book, which runs to 28 chapters.  This distribution suggests that Otfrid was still finding his footing, as it were, and either borrowed older lines or experimented with compensating for a lack of rhyme with alliteration before deciding against it.

Of these lines the only one that fully accords with Sieversian metrics is 1.18.9.  This unusual regularity along with its late relative position points to the borrowing of 1.18.9 from another source.  Already by the 18th chapter of book 1, Otfrid was comfortable enough with the end-rhymed meter not to need to insert an alliterative line; what is more, even in his irregular lines he does not follow the Sieversian rule of four positions, nor always the linking of alliteration between couplets.

From the viewpoint of literary history it is certain that it would be highly unusual for a poet active in a clerical context after the success of the \emph{Evangelienbuch} to still employ alliteration, especially when he was clearly familiar with the end-rhymed form as proven by ll.~60--61.  The natural choice for such a writer would be to use end-rhyme throughout.  Indeed, the \emph{Evangelienbuch}, along with related works like the \emph{Georgslied} and \emph{Ludwigslied}, seems to have definitively put an end to the alliterative tradition as a whole in Germany, since no German alliterative poetry younger than the 9th century survives.

Morever, while \Muspilli\ 14 is a perfectly regular alliterative long-line, its direct equivalent \emph{Evangelienbuch} 1.18.9 is, from the perspective of Otfridian meter, deficient.  This again gives support to a \emph{Muspilli}-an primacy, for it makes sense for an alliterative poet working after Otfrid to borrow just that particular deficient line for his own work.

It may be said shortly that \Muspilli\ is more archaic than Otfrid not only in meter, but also in its use of originally pre-Christian expressions like \emph{poum, perga} (l.~50) and \emph{Mú·spell} (l.~56).  These may, however, simply be due to its more rustic form and less theologically informed content and therefore do not necessarily support an earlier relative dating.

If \Muspilli\ indeed is the older poem, there arises the question of how Otfrid would have come into contact with it.  Yet of all problems associated with the relation between the two texts this is the most easily answered.  Several decades before he dedicated his \emph{Evangelienbuch} to Ludwig the German in ca.~870 \ce, Otfrid had spent a number of years as a scribe at Ludwigh’s court \parencite[406]{Haubrichs1978}---just where the ms. containing \Muspilli\ must have been located after its dedication (see Preservation above).  In poring over the pseudo-Augustinian sermon the young Otfrid would have discovered \Muspilli\ written in the margins and put it to memory.  When as an old man he set out to write his \emph{Evangelienbuch} and came to the description of Paradise, he remembered, if not the whole \emph{Muspilli}, at least this one evocative line---“There life is without death, light without darkness”---and included it.

\subsubsection{Language}

The dialect of the poem is that of the southern High German area, most likely Bavaria, where  Regensburg was the seat of Emperor Ludwigh’s government.  This is seen by the consistent application of the most extensive form of the second sound shift, where \emph{g, b, k} change to \emph{k, p, kh}.  That this sound change had occurred at the time of composition and is not merely a scribal feature is proven by the fact that Germanic roots originally beginning with \emph{b} consistently alliterate with Latin borrowings beginning with \emph{p}, namely in:

\begin{itemize}
  \item l.~5: \emph{pehhe} (< Latin \emph{pix}) : \emph{págant} (= ON \emph{bága}),
  \item l.~16: \emph{pú} (= OS \emph{bú}) : \emph{pardísu} (< Latin \emph{paradīsum}),
  \item l.~21: \emph{piutit} (= OS \emph{biudid}) : \emph{pehhes}, \emph{pína} (< Latin \emph{poena}),
  \item l.~25: \emph{prinnan} (= OS \emph{brinnan}), \emph{palw-} (= OS \emph{baluw-}) : \emph{pehhe}.
\end{itemize}

Interestingly, the meter also shows that the poet retained old \emph{h-} before \emph{l} (l.~72), by extension almost certainly also before \emph{r} and \emph{n}, and probably also before \emph{w} (l.~7).  This sound is, however, consistently omitted by the scribe.

\subsection{Dating}

Since the poem is marginalia, the dating of the manuscript can unfortunately only serve as a \emph{terminus post quem} for the poem itself, viz. the 820s \ce.  The likely direction of influence to the \emph{Evangelienbuch} (863--871 \ce) may serve as a \emph{terminus ante quem}, and the poem was therefore most likely composed in Bavaria some time between 820 and 870 \ce.

\subsection{Summary}

The poem begins with a description of the fate of the soul immediately after death (ll.~1--29).  This is followed by a short sermon about the importance of doing good works before the Judgment (ll.~30--35).  After this follows a description of the destruction of the world and the Judgment Day.  First the fight between Elijah and the Anti-Christ (ll.~36--49), then the burning of the world and the Muspell (ll.~50--61).  This is followed by a second sermon warning against corruption by describing how the Devil keeps track of evil deeds to recount them at the Judgment (ll.~62--71).  After this the Judgment itself begins.  An angel blows a horn and God and His Angels are on the move (72--80); then the dead come back to life to be judged (81--84).  They gather in a court before God and the Angels (85--89); the poet stresses that all must become known at the Judgment, and even a man’s own limbs will bear witness against him (90--96); the only way to avoid punishment is through good deeds on earth (97-100).  After this the Holy Cross is brought forth, and the poem suddenly ends (101--104).

\sectionline
\newpage

\section{Text}

\bvg\bva[]%
\edtrans{Sín \alst{t}ak pi·kweme, \hld\ daz er \alst{t}ouwan skal}{His day will come when he must die.}{\Bfootnote{It is traditionally assumed that the beginning of the poem is incomplete, there originally being other lines before this one, but there is nothing in 14098 to suggest this.  I find the opening line with its generic \emph{sín, er} ‘his, he’ rather powerful.}}. &
\emph{H}wanta \alst{s}ár só sih diu \alst{s}êla \hld\ in den \alst{s}ind ar·hęvit, &
ęnti sí den \alst{l}íh-hamun \hld\ \edtext{\alst{l}ikkan lázzit}{\Bfootnote{The technical double alliteration in the second half-line is unusual, but probably not due to any scribal corruption.}}, &
só kwimit ęin \alst{h}ęri \hld\ fona \alst{h}imil-zungalon; &
daz andar fona \edtrans{\alst{p}ehhe}{the Pitch}{\Bfootnote{The burning hellfire.}}: \hld\ dár \alst{p}ágant siu umpi.\eva

\bvb {\huge H}\textsc{is day will come} when he must die. \\
For as soon as the soul lifts itself on its way \\
and leaves the body behind to lie, \\
then comes an army from the heavenly stars, \\
the other from the Pitch; there they fight over it.\evb\evg


\bvg\bva[][6]%
\alst{S}orgén mak diu \alst{s}êla, \hld\ unzi diu \alst{s}uona ar·gét, &
za \edtext{\emph{\alst{h}}wederemo}{\Afootnote{\emph{wederemo} 14098}\Bfootnote{Restoration of the initial \emph{h-} is not strictly required for the line to alliterate properly, but is done on the basis of l.~72.}} \alst{h}ęrje \hld\ si gi·\alst{h}alót werde. &
\emph{H}wanta ipu sia daz \alst{S}atanazses \hld\ ki·\alst{s}indi ki·winnit, &
daz \alst{l}ęitit sia sár \hld\ dár iru \alst{l}ęid wirdit, &
in \alst{f}uir ęnti in \alst{f}instrí: \hld\ daz ist rehto \alst{v}irin-líh ding.\eva

\bvb The soul must worry until the [time of] atonement begins \\
about which of the armies will receive it, \\
for if Satan’s following wins it \\
then it is soon led whither it must face torment \\
into fire and darkness.  That is surely a wicked thing.\evb\evg


\bvg\bva[][11]%
Upi sia \edtext{*avar}{\Afootnote{\emph{hauar} 14098}} ki·\alst{h}alónt die \hld\ die dár fona \alst{h}imile kwemant, &
ęnti sí dero \alst{ę}ngilo \hld\ \alst{ęi}gan wirdit, &
\edtext{die pringent sia sár úf \hld\ in himilo ríhi:}{\Bfootnote{The line lacks alliteration, but no obvious emendation presents itself.  On the basis of the occasional spelling of vowel-words with \emph{h-}, suggesting confusion over that sound, it is possible that the alliteraiton should fall on \emph{úf} and \emph{’imilo}, but that is most speculative.}} &
\edtrans{dár ist \alst{l}íp áno tôd, \hld\ \edtext{\alst{l}i\emph{oh}t}{\Afootnote{\emph{lihot} 14098}} áno \edtext{finst\emph{r}í}{\Afootnote{\emph{finsti} 14098}}}{There life is without death, light without darkness}{\Bfootnote{This line also appears in Otfrid’s \emph{Evangelienbuch} 1.18.9, where it begins a description of Paradise, in the form: \emph{Thár ist líb ána tôd, \hld\ lioht ána finstri.}
It is one of Otfrid’s rhymeless lines where alliteration compensates for the expected end-rhyme.  For the relevance of this shared line to the relation between \Muspilli\ and \emph{Evangelienbuch} see Introduction above.}}, &
\alst{s}ęlida áno \edtext{\alst{s}orgu\emph{n}}{\Afootnote{partly illegible in 14098}}: \hld\ \edtext{\emph{dár n·ist}}{\Afootnote{emend.; illegible in 14098}} neo-man \alst{s}iuh. &
Denne der man in \edtext{\alst{p}ar\emph{dí}su}{\Afootnote{partly illegible in 14098}} \hld\ \alst{p}ú ki·winnit, &
\alst{h}ús in \alst{h}imile, \hld\ dár kwimit imo \alst{h}ilfa ki·nuok.\eva

\bvb If, however, it is received by those who come from heaven \\
and the angels are to have it for themselves, \\
they soon bring it up into the realm of the heavens. \\
There life is without death, light without darkness, \\
happiness without sorrow; there no man is sick. \\
When the man wins his dwelling in Paradise, \\
his house in heaven, he will get help aplenty.\evb\evg


\bvg\bva[][18]%
Pi·diu ist durft \alst{m}ihhil allero \alst{m}anno \edtext{\emph{h}we-líhemo}{\Afootnote{end 61r, the text continues on 119v.}}, \hld\ daz in es sín \alst{m}uot ki·spane, &
daz er \alst{k}otes willun \hld\ \alst{k}erno tu\emph{e} &
ęnti \alst{h}ęlla fuir \hld\ \alst{h}arto wíse, &
\alst{p}ehhes \alst{p}ína: \hld\ dár \alst{p}iutit \edtrans{der Satanasz altist}{the oldest Satan}{\Bfootnote{Not a reference to any specific ‘Satan’ but rather denoting Satan as the primeval, ancient enemy of God.}} &
\alst{h}ęizzan lauk. \hld\ Só mak \alst{h}ukkan za diu, &%NOTE: za] 120r
\alst{s}orgén dráto, \hld\ der sih \alst{s}untigen węiz.\eva

\bvb Therefore it is of great need for every man that he persuade his heart \\
that it shall gladly do the will of God \\
and stalwartly avoid the fire of Hell, \\
the torture of the Pitch; there the oldest Satan offers \\
hot flame.  So he must think about that, \\
worry greatly, whoever knows that he sins.\evb\evg


\bvg\bva[][24]%
Wê demo in \alst{v}instrí skal \hld\ síno \alst{v}iriná stúén, &
\alst{p}rinnan in \edtext{\alst{p}\emph{e}hhe}{\Afootnote{\emph{phhe} 14098}}: \hld\ daz ist rehto \alst{p}alwík dink, &
daz der man \alst{h}arét ze gote \hld\ ęnti imo \alst{h}ilfa ni kwimit. &
\alst{W}ánit sih ki·náda \hld\ diu \alst{w}ênaga sêla: &%NOTE: wánit] 120v
ni ist in ki·\alst{h}uktin \hld\ \alst{h}imiliskin gote, &
\emph{h}wanta hiar in \alst{w}er-olti \hld\ after ni \alst{w}erkóta.\eva

\bvb Woe to him who in the darkness shall atone for his wickedness, \\
burn in the Pitch; that is surely a baleful thing, \\
that the man calls out to God and no help for him comes. \\
It expects mercy, that wretched soul; \\
it is not in the thoughts of the heavenly God, \\
for here in this world it left no works behind.\evb\evg


\bvg\bva[][30]%
Só denne der \alst{m}ahtigo khunink \hld\ daz \alst{m}ahal ki·pannit, &
dara skal \alst{k}weman \hld\ \alst{kh}unno ki·líhaz: &
denne ni ki·tar \edtrans{\alst{p}arno}{the children}{\Bfootnote{The children of men, i.e., human beings.}} nohhęin \hld\ den \alst{p}an furi·sizzan, &
ni allero \alst{m}anno \emph{h}we-líh; \hld\ ze demo \alst{m}ahale skuli. &
Dár skal er vora demo \alst{r}íhhe \hld\ az \alst{r}ahhu stantan, &
pí daz er in \alst{w}er-olti eo \hld\ ki·\alst{w}erkót hapéta.\eva

\bvb So, when the mighty King summons the court \\
whither every race shall come, \\
then none of the children dare sit out the summons, \\
nor whatsoever man; he must go to the court. \\
There shall he before that power stand on trial \\
for the works he has done earlier in the world.\evb\evg


\bvg\bva[][36]%
Daz hôrt’ ih \alst{r}ahhón \hld\ dia wer-olt-\alst{r}eht-wíson, &
daz skuli der \alst{A}nti-khristo \hld\ mit \alst{E}líase págan. &
Der \alst{w}arkh ist ki·\alst{w}áfanit, \hld\ denne wirdit untar in \alst{w}ík ar·hapan. &
\alst{Kh}ęnfun sint só \alst{k}ręftík; \hld\ diu \alst{k}ósa ist só mihhil. &
\alst{E}lías strítit \hld\ pí den \alst{ê}wigon líp, &
wili dén \alst{r}eht-kernón \hld\ daz \alst{r}íhhi ki·starkan: &
pi·diu skal imo \alst{h}elfan \hld\ der \alst{h}imiles ki·waltit.\eva

\bvb This I heard tell the righteous of the world, \\
that the Anti-Christ shall fight with Elijah; \\
the outlaw is armed; then he will commence the battle. \\
The champions are so strong, the strife is so great; \\
Elijah fights for the eternal life, \\
wishes to strengthen the power of the righteous; \\
therefore he will have help from Him Who rules heaven.\evb\evg


\bvg\bva[][43]%
Der \alst{A}nti-khristo \hld\ stét pí demo \edtrans{\alst{a}lt-fíante}{Fiend of Old}{\Bfootnote{Or ‘the Ancient Enemy.’  Satan.}}, &
stét pí demo \alst{S}atanase, \hld\ der inan var·\alst{s}ęnkan skal: &
pi·diu skal er in deru \alst{w}ík-stęti \hld\ \alst{w}unt pi·vallan &
ęnti in demo \alst{s}inde \hld\ \alst{s}iga-lôs werdan.\eva

\bvb The Anti-Christ stands beside the Fiend of Old, \\
stands beside Satan who shall cause him to fall. \\
Therefore shall he on that battlefield fall wounded \\
and in the incursion be victoryless.\evb\evg


\bvg\bva[][47]%
Doh wánit des vila got-manno, &
daz Elías in demo \alst{w}íge \hld\ ar·\alst{w}artit werde, &%NOTE: werde] 121r
só daz \alst{E}líases pluot \hld\ in \alst{e}rda ki·triufit.\eva

\bvb Yet many Godly men expect \\
that Elijah in that battle will be maimed, \\
so that Elijah’s blood pours into the earth.\evb\evg


\bvg\bva[][50]%
Só in·\alst{p}rinnant die \edtext{\alst{p}erga, \hld\ \alst{p}oum}{\lemma{perga \dots\ poum ‘mountains \dots\ tree’}\Bfootnote{Formulaic word-pair; see note to \textlink{Wessobrunn} 3.}} ni ki·stęntit &
\alst{ê}nihk in \alst{e}rdu, \hld\ \alst{a}há ar·truknént, &
muor var·\alst{s}wilhit sih, \hld\ \alst{s}wilizót lougiu der himil, &
\alst{m}áno vallit, \hld\ prinnit \alst{m}ittila-gart, &
\edtext{\alst{st}ên ni ki·\alst{st}ęntit*}{\Afootnote{add. \emph{ênikh in erdu} 14098 is unmetrical dittography from l.~50b--51a.}}, \hld\ vęrit denne \alst{st}úa-tago in lant, &
\alst{v}ęrit mit diu \alst{v}uiru \hld\ \edtrans{\alst{v}iriho}{folk}{\Bfootnote{Gen.~of \emph{fir(a)ha}, a \emph{plurale tantum} poetic synonym for ‘men, people’ found throughout the Old Germanic languages.  See \textlink{Hildebrandslied} 10 n.~for etymology and distribution.}} wísón.\eva

\bvb So the mountains burn down, no tree stands \\
anywhere upon the earth, the rivers turn dry, \\
the swamp is swallowed up, the heaven glows with flame, \\
the moon falls, Middenyard burns, \\
no stone stands.  Then the Judgment Day comes upon the land, \\
comes with the fire to visit the folk.\evb\evg


\bvg\bva[][56]%
Dár ni mak denae \alst{m}ák andremo \hld\ helfan vora demo \edtrans{\alst{M}ú·spille}{the Muspell}{\Bfootnote{The word which has given the poem its modern name, probably ‘the Destruction of the Earth’.  \emph{Mú·spille} is a hapax in OHG, but corresponds to OS \emph{Múd·spelli} \char`~\ \emph{Mút·spelli} (\textlink{Heliand}[31][2591], 4358) and ON \emph{Mú·spell} (see \textlink{Voluspa}[49]/1 n.).
Its exact sense and origin are hard to pin down.

In all cases the word seems to refer to at least an element of the eschatological destruction of the world, whether Heathen (as in the Norse \textlink{Voluspa} and \Gylfaginning) or Christian (as in the High German \Muspilli\ and the Saxon \textlink{Heliand}).  In \Muspilli\ the word refers to the culmination of the destruction of the world by fire.  In \textlink{Heliand} 2591--92 \emph{Múd·spelles męgin} ‘Muspell’s might’ is said to pass over mankind together with the \emph{ęndi þesaro wer-oldes} ‘end of this world’ and in \textlink{Heliand} 4358--4361 Muspell is said to come like a thief in the night along with the last day.  In \textlink{Voluspa}[49] \emph{Mú·spells lýðir} ‘Muspell’s people’ are said to come from the south on a great ship, while in \Gylfaginning\ 14 Muspell or Muspellsham is said to be a burning realm in the south ruled over by the fire-ettin Surt.

Etymologically the word is clearly a cpd., segmented between \emph{mú(d)} and \emph{spilli}; it is a \emph{ja}-stem in OS and OHG and an \emph{a}-stem in ON; its gender is masc.~in ON and masc.~or neuter in OS and OHG \parenciteshorthand[s.v. \textit{mū̆spili}]{EWA}.

The first element, the root length of which is uncertain, has been explained as reflecting either PGmc. \emph{*mūhô} ‘heap (of earth)’ or, on the basis of the OS dental, PGmc. \emph{*munþaz} ‘mouth’ \parenciteshorthand{EWA}.  On the basis of sound-laws the latter is unlikely since \textipa{/\*VnT/} > \textipa{/\~{\*V}:T/} (part of the Ingvaeonic nasal spirant law) does not affect OHG or ON, which for ‘mouth’ have \emph{mund} and \emph{munnr}, respectively.  Still, the OS dental poses a formal difficulty and opens up the possibility of borrowing, although the direction is unclear.  It is notable that both ON and OHG, at the geographical extremes, agree about the lack of a dental.

The second element is prob. an agent noun derived either from the verb \emph{*spilþijaną} ‘to spoil, ruin; kill, murder’ (OE \emph{spillan, spildan}, OS \emph{spildjan}, ON \emph{spilla}) or from \emph{*spellōną} ‘to tell, talk’ \parenciteshorthand{EWA}.  Due to the lack of breaking in ON the former is probably more likely.

To conclude, due to the possibility of borrowing between languages, the lack of any glosses, and the formal difficulties, any derivation must be considered uncertain.  Still, its undoubtedly Germanic linguistic origin does make a pagan origin more likely.  If it really is a purely Christian word one has to ask why it does not appear in West Germanic prose or even non-alliterative poetry, and why it should signify a fiery realm in the Norse mythology.}}, &
denne daz \alst{p}ręita wasal \hld\ allaz var·\alst{p}rinnit, &
ęnti \alst{v}uir ęnti luft \hld\ iz allaz ar·\alst{f}urpit. &
\edtrans{\emph{H}wár ist denne diu \alst{m}arha, \hld\ dár man dár eo mit sínén \alst{m}ágon piehk?}{Where is then the ground where man once with his kinsmen fought?}{\Bfootnote{An example of the \emph{ubi sunt} motif common in mediæval literature from both Europe and the Arab world.  This literary motif, originally drawing on the Book of Baruch (3:16--19) and in Old Germanic poetry most famously found in \emph{The Wanderer} (ll.~92--96), illustrates the vanity of life by listing asking ‘where are they?’  The German poet here extends the motif to the Day of Judgment and takes the opportunity to comment on his own time by condemning the internecine feuds that plagued all mediæval monarchies.}} &
\edtext{Diu marha ist far·prunnan, \hld\ diu sêla stét pi·d\emph{w}ungan, &
ni węiz mit \emph{h}wiu puaze: \hld\ só vęrit sí za wíze.}{\lemma{Diu \dots\ wíze}\Bfootnote{In these two lines the poet replaces the usual alliteration with end-rhyme between each half-line (60: \emph{prunnan} : \emph{dwungan}, 61: \emph{puaze} : \emph{wíze}).  The very same meter, including the looseness of the rhymes, is used by Otfrid throughout the whole of his \emph{Evangelienbuch}, written some time between 863 and 871 \ce.  The direction of influence between \Muspilli\ and that work is uncertain owing to the difficulties of dating the present poem, for which see l.~14 and Introduction above.}}\eva

\bvb There may no kinsman help another one before the Muspell \\
when the broad damp earth all burns up \\
and fire and air cleanse it all. \\
Where is then the ground where man once with his kinsmen fought? \\
The ground is burnt up; the soul is defeated, \\
knows no remedy; so it goes to punishment.\evb\evg


\bvg\bva[][62]%
Pi·diu ist demo \alst{m}anne só guot, \hld\ denne ’r ze demo \alst{m}ahale kwimit, &
daz er \alst{r}ahóno \emph{h}we-líha \hld\ \alst{r}ehto ar·tęile; &
denne ni darf er \alst{s}orgén, \hld\ denne er ze deru \alst{s}uonu kwimit. &
Ni \alst{w}ęiz der \alst{w}ênago man, \hld\ \emph{h}wie-líhan \alst{w}artil er habét, &
denne ’r mit den \alst{m}iatón \hld\ \alst{m}arrit daz rehta, &
daz der \alst{t}iuval dár pí \hld\ ki·\alst{t}arnit stęntit.\eva

\bvb Therefore it is so good for the man when he comes to the court \\
that he rightly consider every charge, \\
then he need not worry when he comes to the atonement. \\
The wretched man knows not what sort of overseer he has, \\
when with bribes he obstructs what is right, \\
that the Devil stands hidden by his side.\evb\evg


\bvg\bva[][68]%
\edtrans{Der}{\emph{He}}{\Bfootnote{The Devil, as just mentioned.}} hapét in \alst{r}uovu \hld\ \alst{r}ahóno \emph{h}we-líha, &
daz der man \alst{ê}r ęnti síd \hld\ \alst{u}piles ki·frumita, &
daz er iz allaz ki·\alst{s}agét, \hld\ denne er ze deru \alst{s}uonu kwimit; &
ni skolta síd \alst{m}anno nohhęin \hld\ \alst{m}iatun int·fáhan.\eva%NOTE: ni: 121v

\bvb \emph{He} will in silence keep every charge, \\
which the man, early or late, has committed, \\
so that he tells it all when he comes to the atonement; \\
for that sake should no man take bribes.\evb\evg


\bvg\bva[][72]%
Só daz \alst{h}imiliska \alst{h}orn \hld\ \edtrans{ki·\emph{\alst{h}}lútit}{sounded}{\Afootnote{\emph{kilutit} 14098}\Bfootnote{Restoration of the cluster \emph{hl-} is required by the alliteration; cf.~l.~7.}} wirdit, &
ęnti sih der \alst{s}uanari \hld\ ana den \alst{s}ind ar·hęvit &
der dár suannan skal \hld\ tôten ęnti lepentén; &
denne \alst{h}ęvit sih mit imo \hld\ \alst{h}ęrjo męista, &
daz ist allaz só \alst{p}ald, \hld\ daz imo nio-man ki·\alst{p}ágan ni mak. &
Denne vęrit er ze deru \alst{m}ahal-stęti, \hld\ deru dár ki·\alst{m}arkhót ist: &
dár wirdit diu \alst{s}uona, \hld\ dia man dár io \alst{s}agéta.\eva

\bvb So the heavenly horn is sounded, \\
and the Judge lifts Himself on His way, \\
He who shall judge the dead and the living. \\
Then rises with Him the greatest of armies; \\
it is all so bold that no man may fight it. \\
Then He journeys to the place of the court which is there marked out; \\
there begins the atonement as was once foretold.\evb\evg


\bvg\bva[][79]%
Denne varant \alst{ę}ngila \hld\ \alst{u}per dio marha, &
\alst{w}ękhant deota, \hld\ \alst{w}íssant ze dinge. &
Denne skal \alst{m}anno gi·líh \hld\ fona deru \alst{m}oltu ar·stén, &
\alst{l}ôssan sih ar dero \alst{l}éwo vazzón: \hld\ \edtext{skal imo avar sín \alst{l}íp pi·kweman}{\lemma{skal imo \dots\ sín líp pi·kweman ‘he shall come to life’}\Bfootnote{Lit. ‘his life shall come to him.’}}, &
daz er sín \alst{r}eht allaz \hld\ ki·\alst{r}ahhón muozzi, &
ęnti imo after sínén \alst{t}átin \hld\ ar·\alst{t}ęilit werde.\eva

\bvb Then the angels journey over the ground, \\
they wake the nations, show them to the Thing. \\
Then shall every man arise out of the earth, \\
free himself from the burdens of the grave; further he shall come to life \\
that he may fully plead his case \\
and according to his deeds be considered.\evb\evg


\bvg\bva[][85]%
Denne der gi·\alst{s}izzit, \hld\ der dár \alst{s}uonnan skal &
ęnti ar·\alst{t}ęillan skal \hld\ \alst{t}ôtén ęnti kwekkhén, &
denne stét dár \alst{u}mpi \hld\ \alst{ę}ngilo męnigí, &
\alst{g}uotero \alst{g}omóno: \hld\ \alst{g}art ist só mihhil: &
dara kwimit ze deru \alst{r}ihtungu só vilo \hld\ dia dár ar \alst{r}ęstí ar·stént.\eva

\bvb Then sits he who there shall judge \\
and shall consider the dead and the living. \\
Then there around him stands a throng of angels, \\
{[and]} of good men.  The enclosure is so great; \\
there to that tribunal come so many who have arisen from their rest.\evb\evg


\bvg\bva[][90]%
Só dár \alst{m}anno nohhęin \hld\ wiht pi·\alst{m}ídan ni mak; &
\edtext{dár skal denne \alst{h}ant sprehhan, \hld\ \alst{h}oupit sagén, &
allero \alst{l}ido \emph{h}we-líhk \hld\ unzi in den \alst{l}uzígun vinger}{\lemma{dár skal denne hant sprehhan, \hld\ houpit sagén, // allero lido \emph{h}we-líhk \hld\ unzi in den luzígun vinger ‘there the hand shall speak, the head shall tell, [and] every single limb down to the little finger’}\Bfootnote{The idea that a man’s own limbs shall bear witness against him is well known from the Islamic Day of Judgment (\emph{Quran} 36:65, 41:20--21, 24:24), but is not found in the New Testament.}}, &
\emph{h}waz er untar desen \alst{m}annun \hld\ \alst{m}ordes ki·frumita.\eva

\bvb So there may no man elude the smallest thing; \\
there the hand shall speak, the head shall tell, \\
{[and]} every single limb down to the little finger, \\
what sort of murder it committed in the midst these men.\evb\evg


\bvg\bva[][94]%
Dár ni ist eo só \alst{l}istík man \hld\ der dár io·wiht ar·\alst{l}iugan męgi, &
daz er ki·\alst{t}arnan męgi \hld\ \alst{t}áto dehhęina, &
niz al fora demo \alst{kh}uninge \hld\ ki·\alst{kh}undit werde, &
\alst{ú}zzan er \alst{i}z \hld\ mit \alst{a}lamusanu furi·męgi &
ęnti mit \alst{f}astún \hld\ dio \alst{v}iriná ki·puazti. &
Denne der \alst{p}aldét \hld\ der gi·\alst{p}uazzit hapét, &
denne ’r ze deru suonu kwimit.\eva

\bvb There no man is ever so clever that he may lie about the smallest thing, \\
that he may hide away any deed, \\
so that it would not fully be made known to the King, \\
unless he overcame it through the giving of alms \\
and through fasting did penance for the wickedness. \\
Then is he emboldened who has done penance, \\
when he comes to the atonement.\evb\evg


\bvg\bva[][101]%
Wirdit denne \alst{f}uri ki·tragan \hld\ daz \alst{f}rôno khrúki, &
dár der \alst{h}êligo Khrist \hld\ ana ar·\alst{h}angan ward. &
Denne augit er dio \alst{m}ásún, \hld\ dio er in deru \alst{m}ęnniskí an·fénk, &
dio er duruh desse \alst{m}an-kunnes \hld\ \alst{m}inna far·doléta.\eva

\bvb Then is brought forth the Lordly Cross, \\
whereon the Holy Christ was made to hang. \\
Then He makes seen the wounds which He as a man received \\
which for the love of Mankind He suffered.\evb\evg

\sectionline
