\bookStart{Wessobrunn Hymn}
\setBookCode{Wessobrunn}

\begin{flushright}%
\textbf{Dating:} late C8th

\textbf{Meter:} \Fornyrdislag%para
\end{flushright}%

\section{Introduction}

\subsection{Preservation}

The so-called \textbf{Wessobrunn Hymn} (abbrev. \Wessobrunn) is found on foll. 65v--66r of the Bavarian ms.~BSB Clm 22053 (henceforth known as 22053), previously located at the Wessobrunn abbey in Upper Bavaria.  22053 consists of 100 foll. containing a number of Latin texts dealing with varied themes, including theology, chronology, geography, and Latin-German glosses.  On the basis of a note on 99v mentioning the death of Charlemagne (748--814), 22053 is dated to ca. 814 \ce\ or some years before.  \Wessobrunn\ is written in the same uncial hand as the rest of the ms., and is therefore not marginalia. \parencite[83--84]{Bierbrauer1990}.

\subsection{Contents}

\Wessobrunn\ is introduced by the Latin heading \emph{De poeta} ‘By the poet’.  The text itself is divided by the scribe into three sections, each introduced by a capital letter adorned with dots of red ink.  The first two sections are poetic (“the poem”), and the third is in prose (“the prayer”)

The poem consists of 9 long-lines in alliterative meter, detailing the earliest beginning of the world.  The first 5 lines describe “the greatest of wonders”, namely that the universe was once void, without earth or heaven, wood or mountain, sun or moon or sea.  These lines are very similar to pre-Christian Norse stanzas about the creation of the world, and in fact contain formulaic word-pairs also found in those stanzas (see notes to ll. 2, 3), suggesting a repurposing of older Heathen motifs and expressions in the new, Christian context.

With this in mind, we see that the latter 4 lines constitute a subversion of the earlier Heathen tradition, for they place into this early emptiness the Almighty God, Jesus Christ, and His many ghosts---presumably the Heavenly Host or the Angels.  This is the Christian creation \emph{ex nihilo}, rather than the Indo-European creation \emph{ex materia} through sacrifice of a primordial being (cf. \textlink{Vafthrudnismal}[21], \textlink{Grimnismal}[41]--42).

The prayer is in prose.  The author first thanks God for creating the earth and heaven---this is presumably why the poem was included---and for giving boons to mankind.  He then asks for faith, strength, and wisdom to help him in his mission.

\subsection{Dating}

Since \Wessobrunn\ is not marginalia the \emph{terminus ante quem} of its composition is 814 \ce.  However, on the basis of the evidently poor preservation of the poem (ll. 3--4) it is unlikely that the scribe of 22053 is the direct author of \Wessobrunn; the title \emph{De poeta} ‘By the poet’ also suggests that the scribe was not, in fact, the poet.  With this in mind we are probably dealing with an older composition, allowing us to push the date back to the late 8th century.

\newpage

\section{Text}

\subsection{De poeta (‘By the poet’)}

\bvg\bva%
\edtrans{Dat \edtext{ga·\alst{f}ręgin}{\Afootnote{\emph{ga-} is abbrev. by the rune-like symbol \emph{ᛡ}.  This symbol is used for all other occurrences of \emph{ga-} in the present text except for \emph{ga·náda} and \emph{ga·laupa} in P1 below.}} ih mit \edtrans{\alst{f}irạhim}{men}{\Bfootnote{Dat.~of \emph{fir(a)ha}, a \emph{plurale tantum} poetic synonym for ‘men, people’ found throughout the Old Germanic languages.  See \textlink{Hildebrandslied} 10 n.~for etymology and distribution.}} \hld\ \alst{f}iri-wizzó męista}{I have learned among men this greatest of wonders}{\Bfootnote{A formulaic opening to a Germanic poem, expressing that the narrating poet’s knowledge of the events described is indirect.  For other parallels see note to \textlink{Gudrunarhvot}[1]/1, which contains a word-for-word cognate expression.}}, &
dat \edtext{\edtext{\alst{e}r\emph{d}o}{\Afootnote{\emph{ero} 22053}\Bfootnote{The ms. form \emph{ero} has been interpreted in a variety of ways, usually as a masc. \emph{wan}-stem derived from the same root as the more common \emph{*erþō} ‘earth’.  This is however problematic, since the expected form in that case would be \emph{*er(a)wo} \parenciteshorthand[s.v. \textit{ero}]{EWA}.
The present emendation is based on the formulaic nature of the merism, but also has problems, since the final vowel of this word in OHG is \emph{-a}, not \emph{-o} \parenciteshorthand[s.v. \textit{ero}]{EWA}.}} ni was \hld\ noh \alst{ú}f-himil}{\lemma{er\emph{d}o \dots\ úf-himil ‘Earth \dots\ Up-heaven’}\Bfootnote{A formulaic merism attested in OHG, OS, OE, and ON; it expresses the totality of the universe and is usually used in the context of the creation or destruction of the world.
For its use in creation cf.~esp. \textlink{Voluspa}[3]/3, where it is said, about the time before the World existed, that “Earth and Up-heaven were never found” and \textlink{Vafthrudnismal}[21], where the god Weden asks the ettin Webthrithner about the origin of “Earth and Up-heaven”.  For all other occurrences see Index: \inx[F]{Earth and Upheaven}.}} &
\edtext{noh \alst{p}aum \hld\ noh \alst{p}erek ni was}{\lemma{paum \dots\ perek ‘wood \dots\ mountain’}\Bfootnote{The same word-pair is found in OHG (\textlink{Muspilli} 50, describing the Christian destruction of the world prior to the Judgment) and in ON (\textlink{Grimnismal}[40], describing the creation of the world from Yimer’s body by the Gods). --- The line is clearly defective; \emph{noh paum} is not a metrically acceptable a-verse.}} &
ni [...] nohh-ęinig \hld\ noh sunna ni \edtext{s\emph{k}ęin}{\Afootnote{emend. after \textcite[237]{Grienberger1921}, who notes that the \emph{ſt}-ligature here is slightly different than at other places in the text, reflecting original \emph{*ſc}; \emph{‘ſtein’} 22053}} &
noh \alst{m}áno ni liuhta \hld\ noh der \alst{m}árjo sêo.\eva

\bvb%
{\huge I} \textsc{have learnt among men} that greatest of wonders, \\
that Earth was not, nor Up-heaven, \\
nor wood, nor was there mountain, \\
nor did any [...], nor did the sun shine, \\
nor the moon give off light, nor the famous sea.\evb\evg


\bvg\bva[][6]%
Dó dar ni·\alst{w}iht ni \alst{w}as \hld\ ęntjó ni \alst{w}ęntjó, &
ęnti dó was der \alst{ęi}no \hld\ \alst{a}l-mahtiko kot, &
\edtrans{\alst{m}anno \alst{m}iltisto}{the most giving of men}{\Bfootnote{The expression refers to God in the form of Jesus, who is held to be coeternal with God the Father.  Cf. \textlink{Heliand} 338, et.c.: \emph{allaro barno bętst} ‘best of all children’, Otfrid’s \emph{Evangelienbuch} 1.22.43: \emph{manno liobosta} ‘loveliest of all men’.}}, \hld\ ęnti dar wárun auh \alst{m}anaké mit inan &
\alst{k}ót-líhhé \alst{g}ęistá, \hld\ ęnti \alst{k}ot hęilak.\eva

\bvb%
Then there was no kind of end nor border, \\
and then was the one Almighty God, \\
the most giving of men, and there were also many \\
glorious ghosts with Him, and Holy God.\evb\evg


\bpg\bpa Kot al-mahtiko, dú himil ęnti erda ga·worahtós, ęnti dú mannun só manak kót for·gápi,
for·gip mir in dína ga·náda rehta ga·laupa, ęnti kótan willjon; wís-tóm ęnti spáhida ęnti kraft tiuflun za widar·stantanne, ęnti ark za pi·wísanne, ęnti dínan willjon za ga·wurkhanne.\epa

\bpb O God almighty! Thou wroughtest heaven and earth and Thou gavest men so much good.
Give me in Thy mercy right belief and good will, wisdom and foresight and power, to withstand devils and to reproach queerness and to work thy will.\epb\epg

\sectionline
